### 2014-10-11 manuscript/modelSelection.Rnw
added model selection description etc/
along with model selection description, change the growth model to a
beta likelihood to deal with constrained proportion data. Uploaded DIC
values for each climate model.

\documentclass[12pt]{article}
\usepackage{graphics}
\usepackage{graphicx, verbatim}
\usepackage{amssymb,amsmath}
% \usepackage[T1]{fontenc}
% \usepackage[utf8]{inputenc}
\usepackage{authblk}
\textwidth=6.2in
\textheight=8.5in
%\parskip=.3cm
\oddsidemargin=.1in
\evensidemargin=.1in
\headheight=-.3in

\begin{document}
\SweaveOpts{concordance=TRUE}
\SweaveOpts{highlight=TRUE, tidy=TRUE, keep.space=TRUE, keep.blank.space=FALSE, keep.comment=TRUE}
\SweaveOpts{prefix.string=Fig}
%To put figures in subfolder
%\SweaveOpts{prefix.string=figures/fig}
\DefineVerbatimEnvironment{Sinput}{Verbatim} {xleftmargin=2em}
\DefineVerbatimEnvironment{Soutput}{Verbatim}{xleftmargin=2em}
\DefineVerbatimEnvironment{Scode}{Verbatim}{xleftmargin=2em}

% \title{\LARGE How coexistence mechanisms mediate temporal stability}
% \author[1]{\large Andrew T. Tredennick}
% \author[1]{\large Peter B. Adler}
% \author[2]{\large Frederick Adler}
% \affil[1]{\footnotesize Department of Wildland Resources and the Ecology Center, Utah State University}
% \affil[2]{\footnotesize Departments of Biology and Mathematics, University of Utah}
% \maketitle
\section{Stastical models of vital rates}
The first step in building our single-species population models was to fit statistical models of vital rates at both levels of inference: individual-level and quadrat-level. Here we describe the general statistical approach at the individual-level, but the same approach applies at the quadrat-level. We modeled survival and growth as functions of genet size (quadrat cover in the quadrat-level models) and climate covariates (described in more detail below). We maintained a consistent random effects structure for both models that included three terms: (1) a random year effect on the intercept, (2) a random year effect on the coefficient for plant cover (either individual or quadrate level), and (3) a random effect of group (see Data set description) on the intercept. Vital rates for each species are modeled separately.

We used logistic regression to model survival probability:

\[ 
logit(s) = \beta_{0,t} + \beta_{s,t}x + \beta_{Q} + \beta_{c,1}\theta_{1,t}\\
+ \cdots +  \beta_{c,i}\theta_{i,t} + \varepsilon_{t}
\]

\noindent where $x$ is the log of genet size (or log of quadrat areal cover), $\beta_{0,t}$ is a year-specific intercept, $\beta_{s,t}$ is the year-specific slope parameter for size, $\beta_{Q}$ is the random effect of quadrat group location, $\theta$ is a matrix of $i$ climate effects over $t$ years, $\beta_{c,i}$ is the fixed parameter for the effect of the $i$th climate covariate, and $\varepsilon_{t}$ is the error term.

We modeled growth as gaussian process describing genet size (or quadrat cover) at time $t+1$ as a function of size at $t$ and climate covariates:

\[ 
x_{t+1} = \beta_{0,t} + \beta_{s,t}x_{t} + \beta_{Q} + \beta_{c,1}\theta_{1,t}\\
+ \cdots +  \beta_{c,i}\theta_{i,t} + \varepsilon_{t}
\]

\noindent where $x$ is genet size and all other paramters are as described for the survival regression. For the quadrat-level approach we modeled growth as a process describing proportional cover within a quadrat at time $t+1$ as a function of proportional cover at time $t$ and climate covariates. Thus, instead of a gaussian process, with a normal likelihood, we modeled growth at the quadrat level as above but with a beta likelihood.

Add section on recruitment...

\section{Selecting climate covariates}
For both the individual-level IPM (IPM) and the quadrat-based IBM (QBM) we followed the same model selection approach for including climate covariates in vital rate regressions.  Working within the random effects structure described previously, we fit a set of models with all possible combinations of four climate covariates: fall through spring precipitation at \emph{t}-1 and \emph{t}-2 (ppt1 and ppt2, respectively) and mean spring temperature at \emph{t}-1 and \emph{t}-2 (TmeanSpr1 and TmeanSpr2, respectively), where \emph{t} is the observation year. The combinations of climate covariates we fit included same-year interactions between precipitation and temperature (e.g., a ppt1$\times$TmeanSpr1 interaction). Finally, we also fit a model with no climate covariates included as fixed effects. In total, this resulted in a set of 25 candidate models.

We used the analytical Bayesian software INLA to fit the candidate models and selected the model with the lowest Deviance Information Criteria (DIC). However, for many species there were several models with $\Delta$DIC < 4, indicating that each model received generally the same amount of support. So, among the models with $\Delta$DIC < 4, we chose the model with the fewest number of parameters. In some cases this resulted in selecting a model with no climate covariates.

\end{document}

### 2014-10-11 montana/quadBM/vitalRateRegressions/growthModelSelection.R

@@ -13,30 +13,31 @@ rm(list=ls(all=TRUE))
 ####
 #### POSSIBLE FIXED-EFFECTS -----------------------
 ####
-fixed.forms <- list(form1 = "log(totCover) ~ log(lag.cover)+ppt1*TmeanSpr1+ppt2*TmeanSpr2",
-                    form2 = "log(totCover) ~ log(lag.cover)+ppt1*TmeanSpr1+ppt2+TmeanSpr2",
-                    form3 = "log(totCover) ~ log(lag.cover)+ppt1+TmeanSpr1+ppt2*TmeanSpr2",
-                    form4 = "log(totCover) ~ log(lag.cover)+ppt1*TmeanSpr1+ppt2",
-                    form5 = "log(totCover) ~ log(lag.cover)+ppt1*TmeanSpr1+TmeanSpr2",
-                    form6 = "log(totCover) ~ log(lag.cover)+ppt1+ppt2*TmeanSpr2",
-                    form7 = "log(totCover) ~ log(lag.cover)+TmeanSpr1+ppt2*TmeanSpr2",
-                    form8 = "log(totCover) ~ log(lag.cover)+ppt1+TmeanSpr1+ppt2+TmeanSpr2",
-                    form9 = "log(totCover) ~ log(lag.cover)+ppt1+TmeanSpr1+ppt2",
-                    form10 = "log(totCover) ~ log(lag.cover)+ppt1+TmeanSpr1+TmeanSpr2",
-                    form11 = "log(totCover) ~ log(lag.cover)+ppt1+ppt2+TmeanSpr2",
-                    form12 = "log(totCover) ~ log(lag.cover)+TmeanSpr1+ppt2+TmeanSpr2",
-                    form13 = "log(totCover) ~ log(lag.cover)+ppt1+ppt2",
-                    form14 = "log(totCover) ~ log(lag.cover)+ppt1+TmeanSpr2",
-                    form15 = "log(totCover) ~ log(lag.cover)+TmeanSpr1+ppt2",
-                    form16 = "log(totCover) ~ log(lag.cover)+TmeanSpr1+TmeanSpr2",
-                    form17 = "log(totCover) ~ log(lag.cover)+ppt1*TmeanSpr1",
-                    form18 = "log(totCover) ~ log(lag.cover)+ppt2*TmeanSpr2",
-                    form19 = "log(totCover) ~ log(lag.cover)+ppt1+TmeanSpr1",
[…]

### 2014-10-24 manuscript/ms_MicroMeso.Rnw
renamed manuscript file

--- 2014-10-11 manuscript/modelSelection.Rnw
@@
 
-% \title{\LARGE How coexistence mechanisms mediate temporal stability}
-% \author[1]{\large Andrew T. Tredennick}
-% \author[1]{\large Peter B. Adler}
-% \author[2]{\large Frederick Adler}
-% \affil[1]{\footnotesize Department of Wildland Resources and the Ecology Center, Utah State University}
-% \affil[2]{\footnotesize Departments of Biology and Mathematics, University of Utah}
-% \maketitle
-\section{Stastical models of vital rates}
+\title{\LARGE Micro- versus meso-scale population models for ecological forecasting under climate change}
+\author[1]{\large Andrew T. Tredennick} 
+\author[1]{\large Peter B. Adler}
+\affil[1]{\footnotesize Department of Wildland Resources and the Ecology Center, Utah State University}
+\maketitle
+
+\begin{abstract}
+The basic idea here is to do a head-to-head comparison of complex, individual-based integral projection models and less complex, ``patch-based" models.
+The problem is this: as ecologists we want to forecast the impacts of global change, like climate change, at scales relevant to land management decisions and policy intervention.
+However, almost all studies of plant population dynamics rely on data at the sub 1 meter scale.
+This data is really hard to collect.
+Can we use average plot cover to forecast plant population dynamics just as well as when we use complex IPMs?
+Can ``averaged-out" population models at the plot level adequately capture climate drivers?
+These are the questions we seek to answer here.
+\end{abstract}
+
+\section{Introduction}
+Producing skillfull forecasts at spatial scales relevant to global environmental changes that play out at the landscape and regional scale has proven challenging for population ecologists.
+The major hurdle is that most population models, at least for plant species, are built using data from small, localized plots because parameterizing vital rate regressions requires tracking the fates of individuals.
+These models are difficult to scale up from the micro to meso-scales because the fitted parameters do not fully represent the spatial variation present at spatial scales beyond that at which the data are collected.
+
+Recently, Freckelton et al., building on work by Hastings and NAME, have proposed density-structured population models that focus on the transition of populations among discrete states, rather than the traditional approach of modeling the transitions of individuals.
+Such an approach could be extremely valuable because the data needed to parameterize density-structured population models is much easier, and less costly, to collect (Queenborough et al.
+year).
+For example, using a density-structured approach, one could build a population model using a time series of annual plot-based censuses of species percent cover.
+However, a major assumption of the density-structured approach is that the aggregate dynamics of the population observed at coarse spatial resolution faithfully represent, and correspond to, the fates of individual plants.
+In other words, using a density-structured approach requires a leap of faith that important covariates (e.g., climate variables) at the level of the individual are captured adequately at the population level.
+If we seek to forecast the impacts of climate change on plant populations, then clearly this assumption requires testing.
+
+\section{Materials and Methods}
+\subsection{Study site and data}
+Our demographic data comes from the Fort Keogh Livestock and Range Research Laboratory in eastern Montana's northern mixed prairie near Miles City, Montana, USA (46 deg.
+19' N, 105 deg 48' W).
+The dataset is freely available on Ecological Archives (CITE), and interested readers should refer to the metadata therin for a complete description.
+The site is about 800 m above sea level and mean annual precipitation (1878-2009) is 334 mm, with most annual precipitaion falling from April through September (76).
+The site is grass dominated and, for the purposes of our study, we focus on the four most abundant graminoid species: \emph{Bouteloua gracilis}, \emph{Hesperostipa comata}, \emph{Pascopyrum smithii}, and \emph{Poa secunda}.
+
+From 1932 to 1945 individual plants were identified and mapped annualy in 44 1-m2 quadrats using a pantograph.
+The quadrats were distributed in six pastures, each assigned a grazing treatment of light (1.24 ha/animal unit month), moderate (0.92 ha/aum), and heavy (0.76 ha/aum) stocking rates (two pastures per treatment).
+In this analysis we account for potential differences among the grazing treatments, but do not focus on grazing$\times$climate interactions.
+The annual maps of the quadrats were digitized and the fates of individual plants tracked and extracted using a computer program.
+Daily climate data, which we aggregated into climate variables of interest, are available for the duration of the data collection period (1932 - 1945) from the Miles City airport, Wiley Field, 9 km from the study site.
+
+\subsection{Stastical models of vital rates}
 The first step in building our single-species population models was to fit statistical models of vital rates at both levels of inference: individual-level and quadrat-level.

### 2014-10-24 manuscript/ms_MicroMeso.tex
added climate cont table to ms

--- 2014-10-24 manuscript/ms_MicroMeso.Rnw
@@
 \usepackage{amssymb,amsmath}
+\usepackage{tabularx}
 % \usepackage[T1]{fontenc}
@@
 
+\usepackage{Sweave}
 \begin{document}
-\SweaveOpts{concordance=TRUE}
-\SweaveOpts{highlight=TRUE, tidy=TRUE, keep.space=TRUE, keep.blank.space=FALSE, keep.comment=TRUE}
-\SweaveOpts{prefix.string=Fig}
+\input{ms_MicroMeso-concordance}
+
 %To put figures in subfolder
@@
 \end{abstract}
+
+\noindent{}\footnotesize{\textbf{Keywords:} population model, forecasting, integral projection model, etc.}
 
@@
 
+\section{Results}
+We assessed the statistical importance of the climate covariates included the final vital rate regressions by comparing the residual deviance of models with climate covariates and temporal random effects, climate covariates only, and temporal random effects only.
+When a model includes climate covariates, this comparison shows the relative contribution of the climate covariates in explaining the total interannual variability (Adler et al.
+2012).
+
+% latex table generated in R 3.1.0 by xtable 1.7-3 package
+% Fri Oct 24 17:26:36 2014
+\begin{table}[ht]
+\centering
+\caption{The proportion of interannual variability in  vital rates explained by the climate covariates.} 
+\begin{tabularx}{\textwidth}{lllllX}
+  \hline
+ & N & Constant model & Climate model & Full model & Contribution of climate covariates \\ 
+  \hline
+Growth &  &  &  &  &  \\ 
+  BOGR & 213 & -881.25 & -881.25 & -938.35 & 0 \\ 
+  HECO & 171 & -1365.35 & -1372.06 & -1381.61 & 0.41 \\ 
+  PASM & 217 & -2107.02 & -2116.98 & -2164.8 & 0.17 \\ 
+  POSE & 197 & -1503.69 & -1502.32 & -1551.44 & -0.03 \\ 
+  Survival &  &  &  &  &  \\ 
+  BOGR & 218 & 44.62 & 44.62 & 44.53 & 0 \\ 
+  HECO & 179 & 51.21 & 51.21 & 51.01 & 0 \\ 
+  PASM & 221 & 40.14 & 40.14 & 40.51 & 0 \\ 
+  POSE & 204 & 46.41 & 46.41 & 46.25 & 0 \\ 
+  Colonization &  &  &  &  &  \\ 
+  BOGR & 29 & 38.32 & 39.24 & 44.53 & 0.15 \\ 
+  HECO & 68 & 75.78 & 75.78 & 51.01 & 0 \\ 
+  PASM & 26 & 32.35 & 32.35 & 40.51 & 0 \\ 
+  POSE & 43 & 54.05 & 54.05 & 46.25 & 0 \\ 
+   \hline
+\end{tabularx}
+\end{table}
+
 \end{document}

### 2014-10-24 manuscript/ms_MicroMeso.Rnw
updated table caption

@@
 
-\[ 
+\begin{equation}
 logit(s) = \beta_{0,t} + \beta_{s,t}x + \beta_{Q} + \beta_{c,1}\theta_{1,t}\\
 + \cdots +  \beta_{c,i}\theta_{i,t} + \varepsilon_{t}
-\]
+\end{equation}
 
@@
 
-\[ 
+\begin{equation}
 x_{t+1} = \beta_{0,t} + \beta_{s,t}x_{t} + \beta_{Q} + \beta_{c,1}\theta_{1,t}\\
 + \cdots +  \beta_{c,i}\theta_{i,t} + \varepsilon_{t}
-\]
+\end{equation}
 
@@
 For the quadrat-level approach we modeled growth as a process describing proportional cover within a quadrat at time $t+1$ as a function of proportional cover at time $t$ and climate covariates.
-Thus, instead of a gaussian process, with a normal likelihood, we modeled growth at the quadrat level as above but with a beta likelihood.
+Thus, instead of a gaussian process, with a normal likelihood, we modeled growth at the quadrat level as above but with a beta likelihood and a logit link to the linear predictors.
 
@@
 
+\section{Results}
+We assessed the statistical importance of the climate covariates included the final vital rate regressions by comparing the residual deviance of models with climate covariates and temporal random effects, climate covariates only, and temporal random effects only.
+When a model includes climate covariates, this comparison shows the relative contribution of the climate covariates in explaining the total interannual variability (Adler et al.
+2012).
+
+<<echo=FALSE,results=tex>>=
+library(xtable)
+growthSSRTable <- read.csv("../montana/quadBM/vitalRateRegressions/finalModels/growthClimateContributionTable.csv")
+survivalSSRTable <- read.csv("../montana/quadBM/vitalRateRegressions/finalModels/survivalClimateContributionTable.csv")
+colonizationSSRTable <- read.csv("../montana/quadBM/vitalRateRegressions/finalModels/colonizationClimateContributionTable.csv")
+
+growthSSRTable[,3:6] <- apply(X = growthSSRTable[,3:6], MARGIN = 2, FUN = function(x){round(x, digits = 2)})
+survivalSSRTable[,3:6] <- apply(X = survivalSSRTable[,3:6], MARGIN = 2, FUN = function(x){round(x, digits = 2)})
+colonizationSSRTable[,3:6] <- apply(X = colonizationSSRTable[,3:6], MARGIN = 2, FUN = function(x){round(x, digits = 2)})
+
+growthSSRTable$Species <- as.character(growthSSRTable$Species)
+survivalSSRTable$Species <- as.character(survivalSSRTable$Species)
+colonizationSSRTable$Species <- as.character(colonizationSSRTable$Species)
+
+allTableQBM <- rbind(c("Growth", rep("", times=5)), growthSSRTable, 
+                     c("Survival", rep("", times=5)), survivalSSRTable, 
+                     c("Colonization", rep("", times=5)), colonizationSSRTable) 
+colnames(allTableQBM) <- c("", "N", "Constant model", "Climate model", "Full model", "Contribution of climate covariates")
+
+print(xtable(allTableQBM, align=c("l", "l", "l", "l", "l", "l", "X"), 
+             caption="The proportion of interannual variability in  vital rates explained by the climate covariates.
+The contribution is defined as: (Climate model - Constant Model)/(Full model - Constant model).
+Note that in some cases this will be slightly negative where DIC selected a model with climate covariates but a model without climate covariates had a very similar DIC value."),
+      tabular.environment="tabularx", caption.placement="top",
+      width="\\textwidth", include.rownames=FALSE)
+@
+
 \end{document}

### 2014-10-24 manuscript/ms_MicroMeso.tex
updated table caption

@@
 % latex table generated in R 3.1.0 by xtable 1.7-3 package
-% Fri Oct 24 17:26:36 2014
+% Fri Oct 24 17:36:45 2014
 \begin{table}[ht]
 \centering
-\caption{The proportion of interannual variability in  vital rates explained by the climate covariates.} 
+\caption{The proportion of interannual variability in  vital rates explained by the climate covariates.
+The contribution is defined as: (Climate model - Constant Model)/(Full model - Constant model).
+Note that in some cases this will be slightly negative where DIC selected a model with climate covariates but a model without climate covariates had a very similar DIC value.}
 \begin{tabularx}{\textwidth}{lllllX}

### 2014-10-25 manuscript/ms_MicroMeso.Rnw
again updated caption.
updated affiliations

@@
 \title{\LARGE Micro- versus meso-scale population models for ecological forecasting under climate change}
-\author[1]{\large Andrew T. Tredennick} 
-\author[1]{\large Peter B. Adler}
-\affil[1]{\footnotesize Department of Wildland Resources and the Ecology Center, Utah State University}
+\author[]{\large Andrew T. Tredennick} 
+\author[]{\large Peter B. Adler}
+\affil[]{\footnotesize \emph{Department of Wildland Resources and the Ecology Center, Utah State University}}
 \maketitle
@@
 
-\section{Selecting climate covariates}
+\subsection{Selecting climate covariates}
 For both the individual-level IPM (IPM) and the quadrat-based IBM (QBM) we followed the same model selection approach for including climate covariates in vital rate regressions.
@@
 The contribution is defined as: (Climate model - Constant Model)/(Full model - Constant model).
+The values are residual deviance.
 Note that in some cases this will be slightly negative where DIC selected a model with climate covariates but a model without climate covariates had a very similar DIC value."),

### 2014-10-25 manuscript/ms_MicroMeso.tex
again updated caption.
updated affiliations

@@
 % latex table generated in R 3.1.0 by xtable 1.7-3 package
-% Fri Oct 24 17:36:45 2014
+% Sat Oct 25 11:51:52 2014
 \begin{table}[ht]
@@
 The contribution is defined as: (Climate model - Constant Model)/(Full model - Constant model).
+The values are residual deviance.
 Note that in some cases this will be slightly negative where DIC selected a model with climate covariates but a model without climate covariates had a very similar DIC value.}

### 2014-10-26 manuscript/ms_MicroMeso.Rnw
fixed font size in ms; updated plot in script

@@
 \author[]{\large Peter B. Adler}
-\affil[]{\footnotesize \emph{Department of Wildland Resources and the Ecology Center, Utah State University}}
+\affil[]{\footnotesize{\emph{Department of Wildland Resources and the Ecology Center, Utah State University}}}
 \maketitle
@@
 
-\noindent{}\footnotesize{\textbf{Keywords:} population model, forecasting, integral projection model, etc.}
+\noindent{}\textbf{Keywords:} population model, forecasting, integral projection model, etc.
 

### 2014-10-26 manuscript/ms_MicroMeso.tex
fixed font size in ms; updated plot in script

@@
 
-% latex table generated in R 3.1.0 by xtable 1.7-3 package
-% Sat Oct 25 11:51:52 2014
+% latex table generated in R 3.1.1 by xtable 1.7-3 package
+% Sat Oct 25 19:49:12 2014
 \begin{table}[ht]

### 2014-10-26 montana/quadBM/simulations/baselineQBM.R

@@ -9,6 +9,8 @@ rm(list=ls(all=TRUE))
 
 library(reshape2)
 library(plyr)
+library(ggplot2)
+library(ggthemes)
 
 #bring in data
 allD <- read.csv("../../speciesData/quadAllCover.csv")
@@ -89,7 +91,7 @@ Nsave[,1] <- mean(subset(allD, year==yearsID[1])$propCover)
 
 for(yr in 2:yearsN){
   N <- Nsave[,yr-1]
-  climate <- subset(climD, year==years[yr])
+  climate <- subset(climD, year==years[yr-1])
   climate <- melt(climate)
   
   NforG <- N[N>0]
@@ -111,11 +113,22 @@ colnames(dN) <- years
 nM <- melt(dN)
 nM$sim <- rep(1:nSim, length(years))
 
-ggplot(nM, aes(x=variable, y=value*100, group=sim))+
-  geom_line(alpha=0.02, color="purple")+
-  theme_bw()+
-  scale_y_continuous(limits=c(0,25))
+quadD <- ddply(allD, .variables = c("year"), .fun = summarise,
+               year = mean(year),
+               cover = mean(propCover))
+quadD$year <- unique(nM$variable)
+
+ggplot()+
+  geom_line(data=nM, aes(x=variable, y=value*100, group=sim),alpha=0.03, color="purple")+
+  geom_line(data=quadD, aes(x=year, y=cover*100, group=NA), color="grey25")+
+#   geom_point(data=quadD, aes(x=year, y=cover*100), size=6, color="white")+
+  geom_point(data=quadD, aes(x=year, y=cover*100), size=4, color="grey25")+
+  theme_few()+
+  scale_y_continuous(limits=c(0,10))+
+  ylab("Mean Cover (%)")+
+  xlab("Year")
 
 
+  
 
 

### 2014-10-26 manuscript/ms_MicroMeso.Rnw
ms formatting updates

@@
 \headheight=-.3in
+\usepackage{lineno}
+\linenumbers
+\modulolinenumbers[2]
+\usepackage{setspace}
+\doublespacing
 
@@
 \title{\LARGE Micro- versus meso-scale population models for ecological forecasting under climate change}
-\author[]{\large Andrew T. Tredennick} 
+\author[]{\large Andrew T. Tredennick\footnote{\emph{email:} atredenn@gmail.com}}
 \author[]{\large Peter B. Adler}
@@
 
-Add section on recruitment...
+While our approach for modeling survival and growth is similar at the individual and quadrat levels, the addition of new genets at the inidividual level or occupied sites at the quadrat level requires two separate approaches.
 
@@
 
-We used the analytical Bayesian software INLA to fit the candidate models and selected the model with the lowest Deviance Information Criteria (DIC).
+We based model selection on the Deviance Information Criteria (DIC).
 However, for many species there were several models with $\Delta$DIC < 4, indicating that each model received generally the same amount of support.
-So, among the models with $\Delta$DIC < 4, we chose the model with the fewest number of parameters.
+Given this ambiguity, we opted to simply select the candidate model with lowest DIC, regardless of the statistical nearness of other models, as the final model to use for simulating the population dynamics.
 In some cases this resulted in selecting a model with no climate covariates.
+We had to choose a single statistical model for each vital rate for our population models, but when comparing climate covariate selection between the individual and quadrat levels we acknowledge climate covariates selected in all models with $\Delta$DIC < 4.
+
+\subsection{Computing}
+\subsubsection{Model fitting}
+All of our analyses (model selection, fitting, and simulating) were conducted in R.
+We used the analytical Bayesian software INLA to fit candidate vital rate models at the individual and quadrat levels for computational efficiency.
+This is the case except for the recruitment model at the individual level which required a more complex hierarchical approach.
+In that case we used the MCMC sampler in JAGS to estimate the posterior distributions of model parameters.
+After model selection, we used MCMC samplers, again in JAGS, to estimate the posterior distributions of model parameters for all vital rate models at the individual and quadrat levels.
+We obtained posterior distributions for all model parameters from three MCMC chains run for 10,000 iterations, after discarding the initial 5,000 iterations.
+We assessed convergence visually and using the Gelman diagnostic in the R package `coda.'
+
+\subsubsection{Population models}
+With the posterior distribution of the vital rate statistical models in hand, it is straightforward to simulate the population models.
 
@@
 
+% # <<echo=FALSE, fig=TRUE,height=3,width=8.5 >>=
+% # library(ggplot2)
+% # plotD <- data.frame(Species = rep(growthSSRTable$Species, 3),
+% #                     Model = rep(c("Growth", "Survival", "Colonization"), each=4),
+% #                     Contribution = c(growthSSRTable[,6],
+% #                                      survivalSSRTable[,6],
+% #                                      colonizationSSRTable[,6]))
+% # ggplot(plotD, aes(x=Species, y=Contribution))+
+% #   geom_bar(stat="identity")+
+% #   facet_grid(.~Model)
+% # @
+
 \end{document}

### 2014-10-26 manuscript/ms_MicroMeso.tex
ms formatting updates

@@
 % latex table generated in R 3.1.1 by xtable 1.7-3 package
-% Sat Oct 25 19:49:12 2014
+% Sun Oct 26 08:34:04 2014
 \begin{table}[ht]
@@
 \end{table}
+% # <<echo=FALSE, fig=TRUE,height=3,width=8.5 >>=
+% # library(ggplot2)
+% # plotD <- data.frame(Species = rep(growthSSRTable$Species, 3),
+% #                     Model = rep(c("Growth", "Survival", "Colonization"), each=4),
+% #                     Contribution = c(growthSSRTable[,6],
+% #                                      survivalSSRTable[,6],
+% #                                      colonizationSSRTable[,6]))
+% # ggplot(plotD, aes(x=Species, y=Contribution))+
+% #   geom_bar(stat="identity")+
+% #   facet_grid(.~Model)
+% # @
 

### 2014-10-27 manuscript/ms_MicroMeso.Rnw
updated abstract

@@
 \begin{abstract}
-The basic idea here is to do a head-to-head comparison of complex, individual-based integral projection models and less complex, ``patch-based" models.
-The problem is this: as ecologists we want to forecast the impacts of global change, like climate change, at scales relevant to land management decisions and policy intervention.
-However, almost all studies of plant population dynamics rely on data at the sub 1 meter scale.
-This data is really hard to collect.
-Can we use average plot cover to forecast plant population dynamics just as well as when we use complex IPMs?
-Can ``averaged-out" population models at the plot level adequately capture climate drivers?
-These are the questions we seek to answer here.
+\noindent Forecasts of populations under future climate change are constrained by the spatial and temporal scales at which we can gather data.
+For plants, ecologists often require individual level data to build population models; data that is difficult and expensive to collect at spatial scales greater than one square meter.
+Models built using such data are difficult to scale-up to spatial scales relevant to land management decisions and policy intervention.
+A possible route forward are density-structured models based on coarse-scale census data, which are much easier to collect over large spatial extents.
+However, to use coarse-scale data in a climate change context requires the assumption that the climate signals in population dynamics identified using individual-level data can also be identified when individual-level data are averaged over.
+Here we use a longterm dataset to model population dynamics using individual-level and plot-level data.
 \end{abstract}

### 2014-10-27 manuscript/ms_MicroMeso.tex
just a test
test again
updated abstract

@@
 % latex table generated in R 3.1.1 by xtable 1.7-3 package
-% Sun Oct 26 08:34:04 2014
+% Sun Oct 26 18:09:01 2014
 \begin{table}[ht]

### 2014-10-27 montana/quadBM/TableFigureScripts/DIC_Table.R
scripts for DIC tables in latex

@@ -10,7 +10,7 @@ source("../vitalRateRegressions/fixedEffectsModels_Growth.R")
 tabG <- read.csv("../vitalRateRegressions/topModels_Growth.csv")
 index <- tabG$modelNum
 modVec <- fixed.forms[index]
-tabG$model <- as.vector(modVec)
+tabG$Model <- as.vector(modVec)
 
 
 ####
@@ -20,7 +20,7 @@ source("../vitalRateRegressions/fixedEffectsModels_Survival.R")
 tabS <- read.csv("../vitalRateRegressions/topModels_Survival.csv")
 index <- tabS$modelNum
 modVec <- fixed.forms[index]
-tabS$model <- as.vector(modVec)
+tabS$Model <- as.vector(modVec)
 
 
 ####
@@ -30,5 +30,5 @@ source("../vitalRateRegressions/fixedEffectsModels_Colonization.R")
 tabC <- read.csv("../vitalRateRegressions/topModels_Colonization.csv")
 index <- tabC$modelNum
 modVec <- fixed.forms[index]
-tabC$model <- as.vector(modVec)
+tabC$Model <- as.vector(modVec)
 

### 2014-10-27 manuscript/ms_MicroMeso.Rnw
little update for saving 1000 iterations for model

@@
 After model selection, we used MCMC samplers, again in JAGS, to estimate the posterior distributions of model parameters for all vital rate models at the individual and quadrat levels.
-We obtained posterior distributions for all model parameters from three MCMC chains run for 10,000 iterations, after discarding the initial 5,000 iterations.
-We assessed convergence visually and using the Gelman diagnostic in the R package `coda.'
+We obtained posterior distributions for all model parameters from three parallel MCMC chains run for 20,000 iterations, after discarding the initial 10,000 iterations.
+We assessed convergence visually and using the Gelman diagnostic in the R package `coda.' Scale reduction factors for all parameters were less than 1.02, indicating convergence.
+For the purposes of introducing stochasticity in our population models, we saved the final 1,000 iterations from each chain for all parameters to be used as randomly drawn values during population simulation.
 

### 2014-11-18 manuscript/ms_MicroMeso.Rnw
new models commit and results

@@
 library(xtable)
-growthSSRTable <- read.csv("../montana/quadBM/vitalRateRegressions/finalModels/growthClimateContributionTable.csv")
-survivalSSRTable <- read.csv("../montana/quadBM/vitalRateRegressions/finalModels/survivalClimateContributionTable.csv")
-colonizationSSRTable <- read.csv("../montana/quadBM/vitalRateRegressions/finalModels/colonizationClimateContributionTable.csv")
+deviance <- data.frame(VitalRate = c("Growth", "Survival", "Colonization"),
+                       ConstantModel = rep(NA,3),
+                       ClimateModel = rep(NA,3),
+                       FullModel = rep(NA,3),
+                       ClimateContribution = rep(NA,3))
+deviance$FullModel[1] <- round(read.csv("../montana/quadBM/vitalRateRegressions/growth/growthDeviance.csv")[,2],2)
+deviance$ClimateModel[1] <- round(read.csv("../montana/quadBM/vitalRateRegressions/growth/growthDevianceCLIMATE.csv")[,2],2)
+deviance$ConstantModel[1] <- round(read.csv("../montana/quadBM/vitalRateRegressions/growth/growthDevianceCONSTANT.csv")[,2],2)
+deviance$ClimateContribution[1] <- with(deviance[1,], (ClimateModel-ConstantModel)/(FullModel-ConstantModel))
+deviance$ClimateModel[2] <- round(read.csv("../montana/quadBM/vitalRateRegressions/survival/survivalDeviance.csv")[,2],2)
+deviance$ConstantModel[2] <- round(read.csv("../montana/quadBM/vitalRateRegressions/survival/survivalDevianceCONSTANT.csv")[,2],2)
+deviance$ClimateModel[3] <- round(read.csv("../montana/quadBM/vitalRateRegressions/colonization/colonizationDeviance.csv")[,2],2)
+deviance$ConstantModel[3] <- round(read.csv("../montana/quadBM/vitalRateRegressions/colonization/colonizationDevianceCONSTANT.csv")[,2],2)
+deviance$ClimateContribution[2:3] <- with(deviance[2:3,], (ConstantModel-ClimateModel)/(ConstantModel))
 
-growthSSRTable[,3:6] <- apply(X = growthSSRTable[,3:6], MARGIN = 2, FUN = function(x){round(x, digits = 2)})
-survivalSSRTable[,3:6] <- apply(X = survivalSSRTable[,3:6], MARGIN = 2, FUN = function(x){round(x, digits = 2)})
-colonizationSSRTable[,3:6] <- apply(X = colonizationSSRTable[,3:6], MARGIN = 2, FUN = function(x){round(x, digits = 2)})
+colnames(deviance) <- c("Vital Rate", "Constant model", "Climate model", "Full model", "Contribution of climate covariates")
 
-growthSSRTable$Species <- as.character(growthSSRTable$Species)
-survivalSSRTable$Species <- as.character(survivalSSRTable$Species)
-colonizationSSRTable$Species <- as.character(colonizationSSRTable$Species)
-
-allTableQBM <- rbind(c("Growth", rep("", times=5)), growthSSRTable, 
-                     c("Survival", rep("", times=5)), survivalSSRTable, 
-                     c("Colonization", rep("", times=5)), colonizationSSRTable) 
-colnames(allTableQBM) <- c("", "N", "Constant model", "Climate model", "Full model", "Contribution of climate covariates")
-
-print(xtable(allTableQBM, align=c("l", "l", "l", "l", "l", "l", "X"), 
+print(xtable(deviance, align=c("l", "l", "l", "l", "l", "X"), 
              caption="The proportion of interannual variability in  vital rates explained by the climate covariates.
-The contribution is defined as: (Climate model - Constant Model)/(Full model - Constant model).
-The values are residual deviance.
-Note that in some cases this will be slightly negative where DIC selected a model with climate covariates but a model without climate covariates had a very similar DIC value."),
+The contribution for growth is defined as: (Climate model - Constant Model)/(Full model - Constant model).
+The contribution for survival and colonization, where we could not estimate a full model with year random effects, is defined as: (Constant Model - Climate Model)/Constant Model.
+The values are model deviance estimated from the `coda' package in R."),
       tabular.environment="tabularx", caption.placement="top",

### 2014-11-18 manuscript/ms_MicroMeso.tex
new models commit and results

@@
 % latex table generated in R 3.1.1 by xtable 1.7-3 package
-% Sun Oct 26 18:09:01 2014
+% Tue Nov 18 09:15:14 2014
 \begin{table}[ht]
@@
 \caption{The proportion of interannual variability in  vital rates explained by the climate covariates.
-The contribution is defined as: (Climate model - Constant Model)/(Full model - Constant model).
-The values are residual deviance.
-Note that in some cases this will be slightly negative where DIC selected a model with climate covariates but a model without climate covariates had a very similar DIC value.}
-\begin{tabularx}{\textwidth}{lllllX}
+The contribution for growth is defined as: (Climate model - Constant Model)/(Full model - Constant model).
+The contribution for survival and colonization, where we could not estimate a full model with year random effects, is defined as: (Constant Model - Climate Model)/Constant Model.
+The values are model deviance estimated from the `coda' package in R.}
+\begin{tabularx}{\textwidth}{llllX}
   \hline
- & N & Constant model & Climate model & Full model & Contribution of climate covariates \\ 
+Vital Rate & Constant model & Climate model & Full model & Contribution of climate covariates \\ 
   \hline
-Growth &  &  &  &  &  \\ 
-  BOGR & 213 & -881.25 & -881.25 & -938.35 & 0 \\ 
-  HECO & 171 & -1365.35 & -1372.06 & -1381.61 & 0.41 \\ 
-  PASM & 217 & -2107.02 & -2116.98 & -2164.8 & 0.17 \\ 
-  POSE & 197 & -1503.69 & -1502.32 & -1551.44 & -0.03 \\ 
-  Survival &  &  &  &  &  \\ 
-  BOGR & 218 & 44.62 & 44.62 & 44.53 & 0 \\ 
-  HECO & 179 & 51.21 & 51.21 & 51.01 & 0 \\ 
-  PASM & 221 & 40.14 & 40.14 & 40.51 & 0 \\ 
-  POSE & 204 & 46.41 & 46.41 & 46.25 & 0 \\ 
-  Colonization &  &  &  &  &  \\ 
-  BOGR & 29 & 38.32 & 39.24 & 44.53 & 0.15 \\ 
-  HECO & 68 & 75.78 & 75.78 & 51.01 & 0 \\ 
-  PASM & 26 & 32.35 & 32.35 & 40.51 & 0 \\ 
-  POSE & 43 & 54.05 & 54.05 & 46.25 & 0 \\ 
+Growth & -5566.24 & -5735.32 & -6167.98 & 0.28 \\ 
+  Survival & 171.55 & 170.37 &  & 0.01 \\ 
+  Colonization & 203.98 & 193.65 &  & 0.05 \\ 
    \hline

### 2014-11-18 montana/ipm/vitalRateRegs/growth/growthAllSppCLIMATE_MCMC.R

@@ -70,44 +70,16 @@ dataJ <- list(Y = log(growD$area.t1),
 ####
 iterations <- 50000
 adapt <- 10000
-mod <- jags.model("growthAllSpp_JAGS.R", data=dataJ, n.chains=3, n.adapt=adapt)
+mod <- jags.model("growthAllSppCLIMATE_JAGS.R", data=dataJ, n.chains=3, n.adapt=adapt)
 update(mod, n.iter = (iterations))
-out <- coda.samples(mod, c("intYr", "beta", "intG", "temp1", "temp2", "rain1", "rain2"),
-                    n.iter=iterations, n.thin=10)
 dic <- jags.samples(mod, c("deviance"),
                     n.iter=iterations, n.thin=10)
 
 ####
-#### Check for convergence
-####
-gelmDiag <- gelman.diag(out)
-# heidel.diag(out)
-# gelman.plot(out)
-
-pdf("growthOutPlots.pdf")
-plot(out, auto.layout=FALSE)
-dev.off()
-
-####
 #### Convert to dataframe for export and get other summaries
 ####
-outC <- rbind(out[[1]][(iterations-999):iterations,], 
-              out[[2]][(iterations-999):iterations,], 
-              out[[3]][(iterations-999):iterations,])
-
-outStat <- as.data.frame(summary(out)$stat)
-outQuant <- as.data.frame(summary(out)$quantile)
 outDeviance <- as.data.frame(summary(dic$deviance, mean)$stat)
-
-sppNames <- c(rep(sppList, 13+6+13+4))
-outStat$species <- sppNames
-outQuant$species <- sppNames
-
-saveRDS(outC, file = "growthParamsMCMC.rds")
-write.csv(gelmDiag[[1]], file="growthGelman.csv")
-write.csv(outStat, file="growthStats.csv")
-write.csv(outQuant, file="growthQuants.csv")
-write.csv(outDeviance, file="growthDeviance.csv")
+write.csv(outDeviance, file="growthDevianceCLIMATE.csv")
 
 
 

### 2014-11-18 montana/ipm/vitalRateRegs/growth/growthAllSppCONSTANT_JAGS.R

@@ -1,0 +1,20 @@
+model{
+  #process model and likelihood
+  for(i in 1:nObs){
+    mu[i] <- intercept[spp[i]] + intG[spp[i],grp[i]] + betaSpp[spp[i]]*X[i]
+
+    Y[i] ~ dnorm(mu[i], error[spp[i]])
+  }
+  
+  #priors
+  for(s in 1:nSpp){
+    betaSpp[s] ~ dnorm(0, 1e-6)
+    intercept[s] ~ dnorm(0, 1e-6)
+    error[s] ~ dgamma(0.001, 0.001)
+    intVarG[s] ~ dgamma(2, 0.5) 
+    for(g in 1:nGrp){
+      intG[s,g] ~ dnorm(0, intVarG[s])
+    }
+  }
+  
+}

### 2014-11-18 montana/ipm/vitalRateRegs/growth/growthAllSppCONSTANT_MCMC.R

@@ -70,44 +70,16 @@ dataJ <- list(Y = log(growD$area.t1),
 ####
 iterations <- 50000
 adapt <- 10000
-mod <- jags.model("growthAllSpp_JAGS.R", data=dataJ, n.chains=3, n.adapt=adapt)
+mod <- jags.model("growthAllSppCONSTANT_JAGS.R", data=dataJ, n.chains=3, n.adapt=adapt)
 update(mod, n.iter = (iterations))
-out <- coda.samples(mod, c("intYr", "beta", "intG", "temp1", "temp2", "rain1", "rain2"),
-                    n.iter=iterations, n.thin=10)
 dic <- jags.samples(mod, c("deviance"),
                     n.iter=iterations, n.thin=10)
 
 ####
-#### Check for convergence
-####
-gelmDiag <- gelman.diag(out)
-# heidel.diag(out)
-# gelman.plot(out)
-
-pdf("growthOutPlots.pdf")
-plot(out, auto.layout=FALSE)
-dev.off()
-
-####
 #### Convert to dataframe for export and get other summaries
 ####
-outC <- rbind(out[[1]][(iterations-999):iterations,], 
-              out[[2]][(iterations-999):iterations,], 
-              out[[3]][(iterations-999):iterations,])
-
-outStat <- as.data.frame(summary(out)$stat)
-outQuant <- as.data.frame(summary(out)$quantile)
 outDeviance <- as.data.frame(summary(dic$deviance, mean)$stat)
-
-sppNames <- c(rep(sppList, 13+6+13+4))
-outStat$species <- sppNames
-outQuant$species <- sppNames
-
-saveRDS(outC, file = "growthParamsMCMC.rds")
-write.csv(gelmDiag[[1]], file="growthGelman.csv")
-write.csv(outStat, file="growthStats.csv")
-write.csv(outQuant, file="growthQuants.csv")
-write.csv(outDeviance, file="growthDeviance.csv")
+write.csv(outDeviance, file="growthDevianceCONSTANT.csv")
 
 
 

### 2014-11-18 montana/ipm/vitalRateRegs/growth/growthWrapper.R

@@ -1,0 +1,7 @@
+####
+#### Wrapper script to call each growth model
+####
+
+source("growthAllSpp_MCMC.R")
+source("growthAllSppCLIMATE_MCMC.R")
+source("growthAllSppCONSTANT_MCMC.R")

### 2014-11-18 montana/ipm/vitalRateRegs/recruitment/recruitJAGS.R

@@ -1,0 +1,38 @@
+model{
+  for(i in 1:N){
+    for(j in 1:Nspp){
+      y[i,j]~dnegbin(q[i,j],theta[j])
+      q[i,j]<-theta[j]/(theta[j]+lambda[i,j])
+      lambda[i,j]<-trueP1[i,j]*fecundity[i,j]
+      log(fecundity[i,j])<-intcpt.yr[year[i],j]+intcpt.gr[Group[i],j]+inprod(dd[1:Nspp,j],trueP2[i,1:Nspp])
+      trueP1[i,j]<-parents1[i,j]*u[j]+parents2[i,j]*(1-u[j])
+      trueP2[i,j]<-sqrt(trueP1[i,j])
+    }
+  }
+  
+  #priors
+  
+  
+  # intercepts
+  for(s in 1:Nspp){
+    u[s]~dunif(0,1)
+    theta[s]~dgamma(0.001,0.001)
+    g.tau[s]~dgamma(2,0.5)
+    for(g in 1:Ngroups){
+      intcpt.gr[g,s]~dnorm(0,g.tau[s])
+    }
+    for(k in 1:Nspp){
+      
+      dd[k,s]~dunif(-10,-0.001)
+      
+      #dd[k,s]~dnorm(0,0.001)
+      #dd[k,s]~djl.dnorm.trunc(0,0.001,-0.0001,-10) # truncated upper limit at -0.001
+    }
+    intcpt.mu[s]~dnorm(0,0.001)
+    intcpt.tau[s]~dgamma(0.001,0.001)
+    for(m in 1:Nyrs){
+      intcpt.yr[m,s]~dnorm(intcpt.mu[s],intcpt.tau[s])  
+    }
+  }
+  
+}

### 2014-11-18 montana/ipm/vitalRateRegs/recruitment/recruitmentAll.R

@@ -1,0 +1,160 @@
+# try discrete distributions for number recruits produced
+#clear everything, just to be safe 
+rm(list=ls(all=TRUE))
+
+sppList=c("ARTR","HECO","POSE","PSSP")
+outfile="recruit_params_GnoG.csv"
+#--------------------------------------------------------
+
+# get recruitment data
+Nspp=length(sppList)
+for(i in 1:Nspp){
+  infile1=paste("../../speciesData/",sppList[i],"/recArea.csv",sep="")
+  tmpD=read.csv(infile1)
+  
+  tmpD$Group=as.factor(substr(tmpD$quad,1,1)) #add by Chengjin
+
+  
+  tmpD=tmpD[,c("quad","year","NRquad","totParea","Group")]
+  names(tmpD)[3]=paste("R.",sppList[i],sep="")
+  names(tmpD)[4]=paste("cov.",sppList[i],sep="")
+  if(i==1){
+    D=tmpD
+  }else{
+    D=merge(D,tmpD,all=T)
+  }
+}
+D[is.na(D)]=0  # replace missing values 
+
+
+##then we moved some specific points:
+tmp2<-which(D$quad=="A12" & D$year==44)
+tmp3<-which(D$quad=="B1"  & D$year==44)
+
+
+tmpONE<-c(tmp2,tmp3)
+if(length(tmpONE)>0) D<-D[-tmpONE,]
+
+
+# calculate mean cover by group and year
+tmpD=D[,c("quad","year","Group",paste("cov.",sppList,sep=""))]
+tmpD=aggregate(tmpD[,4:NCOL(tmpD)],by=list("year"=tmpD$year,"Group"=tmpD$Group),FUN=mean)
+names(tmpD)[3:NCOL(tmpD)]=paste("Gcov.",sppList,sep="")
+
+D=merge(D,tmpD,all.x=T)
+
+
+
+
+###if using square transform, there is no need of the following code
+####################################################################
+###here you need to check: both parents1 and parents2 are equal to 0 at the same time
+parents1=as.matrix(D[,c(paste("cov.",sppList,sep=""))])/100 ##convert from absolute cover to [1,100] range
[…]

### 2014-11-18 montana/ipm/vitalRateRegs/survival/survivalAllSppCLIMATE_JAGS.R

@@ -1,0 +1,30 @@
+model{
+  #process model and likelihood
+  for(i in 1:nObs){
+    logit(p[i]) <- intercept[spp[i]] + intG[spp[i],grp[i]] + betaSpp[spp[i]]*X[i] + temp1[spp[i]]*TmeanSpr1[i] + temp2[spp[i]]*TmeanSpr2[i] + rain1[spp[i]]*ppt1[i] + rain2[spp[i]]*ppt2[i]
+    Y[i] ~ dbern(p[i])
+  }
+  
+  #priors
+  for(s in 1:nSpp){
+    betaSpp[s] ~ dnorm(0, 1e-6)
+    temp1[s] ~ dnorm(temp1Mu, temp1Var)
+    temp2[s] ~ dnorm(temp2Mu, temp2Var)
+    rain1[s] ~ dnorm(rain1Mu, rain1Var)
+    rain2[s] ~ dnorm(rain2Mu, rain2Var)
+    intercept[s] ~ dnorm(0, 1e-6)
+    intVarG[s] ~ dgamma(2, 0.5) 
+    for(g in 1:nGrp){
+      intG[s,g] ~ dnorm(0, intVarG[s])
+    }
+  }
+  
+  temp1Mu ~ dnorm(0,1e-6)
+  temp2Mu ~ dnorm(0,1e-6)
+  rain1Mu ~ dnorm(0,1e-6)
+  rain2Mu ~ dnorm(0,1e-6)
+  temp1Var ~ dgamma(0.001, 0.001)
+  temp2Var ~ dgamma(0.001, 0.001)
+  rain1Var ~ dgamma(0.001, 0.001)
+  rain2Var ~ dgamma(0.001, 0.001)
+}

### 2014-11-18 montana/ipm/vitalRateRegs/survival/survivalAllSppCLIMATE_MCMC.R

@@ -1,0 +1,73 @@
+#!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!#
+# Set working directory to location of this source file #
+#!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!#
+
+#clear everything, just to be safe 
+rm(list=ls(all=TRUE))
+
+#load libraries
+library(rjags)
+library(coda)
+load.module("dic")
+
+sppList=sort(c("BOGR","HECO","PASM","POSE"))
+
+####
+#### Read in data by species and make one long data frame
+####
+outD <- data.frame(X=NA,
+                   quad=NA,
+                   year=NA,
+                   trackID=NA,
+                   area=NA,
+                   survives=NA,
+                   age=NA,
+                   distEdgeMin=NA,
+                   allEdge=NA,
+                   species=NA)
+
+for(spp in 1:length(sppList)){
+  doSpp <- sppList[spp]
+  sppD <- read.csv(paste("../../../speciesData/", doSpp, "/survD.csv", sep=""))
+  sppD$species <- doSpp
+  outD <- rbind(outD, sppD)
+}
+
+survD <- outD[2:nrow(outD),]
+climD <- read.csv("../../../weather/Climate.csv")
+climD[3:6] <- scale(climD[3:6], center = TRUE, scale = TRUE)
+climD$year <- climD$year-1900
+survD <- merge(survD,climD)
+survD$Group=as.factor(substr(survD$quad,1,1))
+
+####
+#### Set up data for JAGS model
+####
+dataJ <- list(Y = survD$survives,
+              X = log(survD$area),
+              nObs = nrow(survD),
+              grp = as.numeric(survD$Group),
+              nGrp = length(unique(survD$Group)),
+              spp = as.numeric(as.factor(survD$species)),
+              nSpp = length(unique(survD$species)),
+              yrs = (survD$year - 31),
+              nYrs = length(unique(survD$year)),
+              TmeanSpr1 = survD$TmeanSpr1,
+              TmeanSpr2 = survD$TmeanSpr2,
+              ppt1 = survD$ppt1,
+              ppt2 = survD$ppt2)
+
+####
+#### Run MCMC from JAGS
+####
+iterations <- 50000
+adapt <- 10000
[…]

### 2014-11-18 montana/ipm/vitalRateRegs/survival/survivalAllSppCONSTANT_JAGS.R

@@ -1,0 +1,18 @@
+model{
+  #process model and likelihood
+  for(i in 1:nObs){
+    logit(p[i]) <- intercept[spp[i]] + intG[spp[i],grp[i]] + betaSpp[spp[i]]*X[i]
+    Y[i] ~ dbern(p[i])
+  }
+  
+  #priors
+  for(s in 1:nSpp){
+    betaSpp[s] ~ dnorm(0, 1e-6)
+    intercept[s] ~ dnorm(0, 1e-6)
+    intVarG[s] ~ dgamma(2, 0.5) 
+    for(g in 1:nGrp){
+      intG[s,g] ~ dnorm(0, intVarG[s])
+    }
+  }
+
+}

### 2014-11-18 montana/ipm/vitalRateRegs/survival/survivalAllSppCONSTANT_MCMC.R

@@ -1,0 +1,73 @@
+#!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!#
+# Set working directory to location of this source file #
+#!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!#
+
+#clear everything, just to be safe 
+rm(list=ls(all=TRUE))
+
+#load libraries
+library(rjags)
+library(coda)
+load.module("dic")
+
+sppList=sort(c("BOGR","HECO","PASM","POSE"))
+
+####
+#### Read in data by species and make one long data frame
+####
+outD <- data.frame(X=NA,
+                   quad=NA,
+                   year=NA,
+                   trackID=NA,
+                   area=NA,
+                   survives=NA,
+                   age=NA,
+                   distEdgeMin=NA,
+                   allEdge=NA,
+                   species=NA)
+
+for(spp in 1:length(sppList)){
+  doSpp <- sppList[spp]
+  sppD <- read.csv(paste("../../../speciesData/", doSpp, "/survD.csv", sep=""))
+  sppD$species <- doSpp
+  outD <- rbind(outD, sppD)
+}
+
+survD <- outD[2:nrow(outD),]
+climD <- read.csv("../../../weather/Climate.csv")
+climD[3:6] <- scale(climD[3:6], center = TRUE, scale = TRUE)
+climD$year <- climD$year-1900
+survD <- merge(survD,climD)
+survD$Group=as.factor(substr(survD$quad,1,1))
+
+####
+#### Set up data for JAGS model
+####
+dataJ <- list(Y = survD$survives,
+              X = log(survD$area),
+              nObs = nrow(survD),
+              grp = as.numeric(survD$Group),
+              nGrp = length(unique(survD$Group)),
+              spp = as.numeric(as.factor(survD$species)),
+              nSpp = length(unique(survD$species)),
+              yrs = (survD$year - 31),
+              nYrs = length(unique(survD$year)),
+              TmeanSpr1 = survD$TmeanSpr1,
+              TmeanSpr2 = survD$TmeanSpr2,
+              ppt1 = survD$ppt1,
+              ppt2 = survD$ppt2)
+
+####
+#### Run MCMC from JAGS
+####
+iterations <- 50000
+adapt <- 10000
[…]

### 2014-11-18 montana/ipm/vitalRateRegs/survival/survivalAllSpp_JAGS.R

@@ -1,0 +1,37 @@
+model{
+  #process model and likelihood
+  for(i in 1:nObs){
+    logit(p[i]) <- intYr[spp[i],yrs[i]] + intG[spp[i],grp[i]] + beta[spp[i],yrs[i]]*X[i] + temp1[spp[i]]*TmeanSpr1[i] + temp2[spp[i]]*TmeanSpr2[i] + rain1[spp[i]]*ppt1[i] + rain2[spp[i]]*ppt2[i]
+    Y[i] ~ dbern(p[i])
+  }
+  
+  #priors
+  for(s in 1:nSpp){
+    betaSpp[s] ~ dnorm(0, 1e-6)
+    temp1[s] ~ dnorm(temp1Mu, temp1Var)
+    temp2[s] ~ dnorm(temp2Mu, temp2Var)
+    rain1[s] ~ dnorm(rain1Mu, rain1Var)
+    rain2[s] ~ dnorm(rain2Mu, rain2Var)
+    intercept[s] ~ dnorm(0, 1e-6)
+    intVaryY[s] ~ dgamma(0.001, 0.001)
+    betaVar[s] ~ dgamma(0.001, 0.001)
+    error[s] ~ dgamma(0.001, 0.001)
+    intVarG[s] ~ dgamma(2, 0.5) 
+    for(y in 1:nYrs){
+      intYr[s,y] ~ dnorm(intercept[s], intVaryY[s])
+      beta[s,y] ~ dnorm(betaSpp[s], betaVar[s])
+    }
+    for(g in 1:nGrp){
+      intG[s,g] ~ dnorm(0, intVarG[s])
+    }
+  }
+  
+  temp1Mu ~ dnorm(0,1e-6)
+  temp2Mu ~ dnorm(0,1e-6)
+  rain1Mu ~ dnorm(0,1e-6)
+  rain2Mu ~ dnorm(0,1e-6)
+  temp1Var ~ dgamma(0.001, 0.001)
+  temp2Var ~ dgamma(0.001, 0.001)
+  rain1Var ~ dgamma(0.001, 0.001)
+  rain2Var ~ dgamma(0.001, 0.001)
+}

### 2014-11-18 montana/quadBM/simulations/baselineQBM.R

@@ -23,11 +23,9 @@ climD <- read.csv("../../weather/Climate.csv")
 doSpp <- sppList[1]
 
 #load vital rate parameters
-files <- list.files("../vitalRateRegressions/finalModels/vitalRateParameters/")
-paramFiles <- files[grep(doSpp, files)]
-pCol <- readRDS(paste("../vitalRateRegressions/finalModels/vitalRateParameters/", paramFiles[1], sep=""))
-pGrow <- readRDS(paste("../vitalRateRegressions/finalModels/vitalRateParameters/", paramFiles[2], sep=""))
-pSurv <- readRDS(paste("../vitalRateRegressions/finalModels/vitalRateParameters/", paramFiles[3], sep=""))
+pCol <- readRDS("../vitalRateRegressions/colonization/colonizationParamsMCMC.rds")
+pGrow <- readRDS("../vitalRateRegressions/growth/growthParamsMCMC.rds")
+pSurv <- readRDS("../vitalRateRegressions/survival/survivalParamsMCMC.rds")
 
 #antilogit function
 antilogit <- function(x) { exp(x) / (1 + exp(x) ) }

### 2014-11-18 montana/quadBM/vitalRateRegressions/colonization/colonizationWrapper.R

@@ -2,5 +2,5 @@
 #### Colonization MCMC wrapper
 ####
 
-source("colonizationAllSppHierarchy_MCMC")
-source("colonizationAllSppHierarchyCONSTANT_MCMC")
+source("colonizationAllSppHierarchy_MCMC.R")
+source("colonizationAllSppHierarchyCONSTANT_MCMC.R")

### 2014-11-18 montana/quadBM/vitalRateRegressions/growth/growthAllSppHierarchy_MCMC.R

@@ -120,11 +120,11 @@ sppNames <- c(rep(sppList, 13+6+13+4))
 outStat$species <- sppNames
 outQuant$species <- sppNames
 
-saveRDS(outC, file = "survivalParamsMCMC.rds")
-write.csv(gelmDiag[[1]], file="survivalGelman.csv")
-write.csv(outStat, file="survivalStats.csv")
-write.csv(outQuant, file="survivalQuants.csv")
-write.csv(outDeviance, file="survivalDeviance.csv")
+saveRDS(outC, file = "growthParamsMCMC.rds")
+write.csv(gelmDiag[[1]], file="growthGelman.csv")
+write.csv(outStat, file="growthStats.csv")
+write.csv(outQuant, file="growthQuants.csv")
+write.csv(outDeviance, file="growthDeviance.csv")
 
 
 

### 2014-11-18 montana/quadBM/vitalRateRegressions/survival/survivalWrapper.R

@@ -2,5 +2,5 @@
 #### Survival MCMC wrapper
 ####
 
-source("survivalAllSppHierarchy_MCMC")
-source("survivalAllSppHierarchyCONSTANT_MCMC")
+source("survivalAllSppHierarchy_MCMC.R")
+source("survivalAllSppHierarchyCONSTANT_MCMC.R")

### 2014-11-25 manuscript/ms_MicroMeso.Rnw
updated model selection section

@@
 \noindent where $x$ is the log of genet size (or log of quadrat areal cover), $\beta_{0,t}$ is a year-specific intercept, $\beta_{s,t}$ is the year-specific slope parameter for size, $\beta_{Q}$ is the random effect of quadrat group location, $\theta$ is a matrix of $i$ climate effects over $t$ years, $\beta_{c,i}$ is the fixed parameter for the effect of the $i$th climate covariate, and $\varepsilon_{t}$ is the error term.
+At the quadrat level we did not fit random year effects for survival because those parameters did not reach convergence (see "Computing").
 
@@
 
-\subsection{Selecting climate covariates}
-For both the individual-level IPM (IPM) and the quadrat-based IBM (QBM) we followed the same model selection approach for including climate covariates in vital rate regressions.
-Working within the random effects structure described previously, we fit a set of models with all possible combinations of four climate covariates: fall through spring precipitation at \emph{t}-1 and \emph{t}-2 (ppt1 and ppt2, respectively) and mean spring temperature at \emph{t}-1 and \emph{t}-2 (TmeanSpr1 and TmeanSpr2, respectively), where \emph{t} is the observation year.
-The combinations of climate covariates we fit included same-year interactions between precipitation and temperature (e.g., a ppt1$\times$TmeanSpr1 interaction).
-Finally, we also fit a model with no climate covariates included as fixed effects.
-In total, this resulted in a set of 25 candidate models.
-
-We based model selection on the Deviance Information Criteria (DIC).
-However, for many species there were several models with $\Delta$DIC < 4, indicating that each model received generally the same amount of support.
-Given this ambiguity, we opted to simply select the candidate model with lowest DIC, regardless of the statistical nearness of other models, as the final model to use for simulating the population dynamics.
-In some cases this resulted in selecting a model with no climate covariates.
-We had to choose a single statistical model for each vital rate for our population models, but when comparing climate covariate selection between the individual and quadrat levels we acknowledge climate covariates selected in all models with $\Delta$DIC < 4.
+\subsection{Including climate covariates (and avoiding model selection)}
+For both the individual-level IPM (IPM) and the quadrat-based IBM (QBM) we included four climate covariates in each vital rate regression: fall through spring precipitation at \emph{t}-1 and \emph{t}-2 (ppt1 and ppt2, respectively) and mean spring temperature at \emph{t}-1 and \emph{t}-2 (TmeanSpr1 and TmeanSpr2, respectively), where \emph{t} is the observation year.
+We included climate covariates as additive effects.
+We did not consider interactions among climate covariates or among climate covariates and plant size or proportional cover.
+Climate effects ($\beta_{c}$s) were modeled hierarchically in each vital rate regression so that each species specific climate effect is drawn from an interspecific climate effect distribution.
+This was necessary to achieve convergence at the quadrat level and we used the same approach at the individual level for uniformity.
 
@@
 \subsubsection{Model fitting}
-All of our analyses (model selection, fitting, and simulating) were conducted in R.
-We used the analytical Bayesian software INLA to fit candidate vital rate models at the individual and quadrat levels for computational efficiency.
-This is the case except for the recruitment model at the individual level which required a more complex hierarchical approach.
-In that case we used the MCMC sampler in JAGS to estimate the posterior distributions of model parameters.
-After model selection, we used MCMC samplers, again in JAGS, to estimate the posterior distributions of model parameters for all vital rate models at the individual and quadrat levels.
-We obtained posterior distributions for all model parameters from three parallel MCMC chains run for 20,000 iterations, after discarding the initial 10,000 iterations.
+All of our analyses (model fitting and simulating) were conducted in Program R.
+We used the MCMC sampler in JAGS to estimate the posterior distributions of model parameters.
+We obtained posterior distributions for all model parameters from three parallel MCMC chains run for 50,000 iterations, after discarding the initial 20,000 iterations.
 We assessed convergence visually and using the Gelman diagnostic in the R package `coda.' Scale reduction factors for all parameters were less than 1.02, indicating convergence.

### 2014-11-25 manuscript/ms_MicroMeso.tex
updated model selection section

@@
 
-\subsection{Selecting climate covariates}
-For both the individual-level IPM (IPM) and the quadrat-based IBM (QBM) we followed the same model selection approach for including climate covariates in vital rate regressions.
-Working within the random effects structure described previously, we fit a set of models with all possible combinations of four climate covariates: fall through spring precipitation at \emph{t}-1 and \emph{t}-2 (ppt1 and ppt2, respectively) and mean spring temperature at \emph{t}-1 and \emph{t}-2 (TmeanSpr1 and TmeanSpr2, respectively), where \emph{t} is the observation year.
-The combinations of climate covariates we fit included same-year interactions between precipitation and temperature (e.g., a ppt1$\times$TmeanSpr1 interaction).
-Finally, we also fit a model with no climate covariates included as fixed effects.
-In total, this resulted in a set of 25 candidate models.
-
-We based model selection on the Deviance Information Criteria (DIC).
-However, for many species there were several models with $\Delta$DIC < 4, indicating that each model received generally the same amount of support.
-Given this ambiguity, we opted to simply select the candidate model with lowest DIC, regardless of the statistical nearness of other models, as the final model to use for simulating the population dynamics.
-In some cases this resulted in selecting a model with no climate covariates.
-We had to choose a single statistical model for each vital rate for our population models, but when comparing climate covariate selection between the individual and quadrat levels we acknowledge climate covariates selected in all models with $\Delta$DIC < 4.
+\subsection{Including climate covariates (and avoiding model selection)}
+For both the individual-level IPM (IPM) and the quadrat-based IBM (QBM) we included four climate covariates in each vital rate regression: fall through spring precipitation at \emph{t}-1 and \emph{t}-2 (ppt1 and ppt2, respectively) and mean spring temperature at \emph{t}-1 and \emph{t}-2 (TmeanSpr1 and TmeanSpr2, respectively), where \emph{t} is the observation year.
+We included climate covariates as additive effects.
+We did not consider interactions among climate covariates or among climate covariates and plant size or proportional cover.
 
@@
 
-% latex table generated in R 3.1.1 by xtable 1.7-3 package
-% Tue Nov 18 09:15:14 2014
+% latex table generated in R 3.1.0 by xtable 1.7-3 package
+% Tue Nov 25 11:51:40 2014
 \begin{table}[ht]

### 2014-12-04 manuscript/ms_MicroMeso.Rnw
updated ms figures

@@
 We modeled survival and growth as functions of genet size (quadrat cover in the quadrat-level models) and climate covariates (described in more detail below).
-We maintained a consistent random effects structure for both models that included three terms: (1) a random year effect on the intercept, (2) a random year effect on the coefficient for plant cover (either individual or quadrate level), and (3) a random effect of group (see Data set description) on the intercept.
+We maintained a consistent random effects structure for both models that included three terms: (1) a random year effect on the intercept, (2) a random year effect on the coefficient for plant cover (either individual or quadrat level), and (3) a random effect of group (see Data set description) on the intercept.
 Vital rates for each species are modeled separately.
@@
 We used the MCMC sampler in JAGS to estimate the posterior distributions of model parameters.
-We obtained posterior distributions for all model parameters from three parallel MCMC chains run for 50,000 iterations, after discarding the initial 20,000 iterations.
+We obtained posterior distributions for all model parameters from three parallel MCMC chains run for 50,000 iterations, after a 50,000 iteration burn in.
 We assessed convergence visually and using the Gelman diagnostic in the R package `coda.' Scale reduction factors for all parameters were less than 1.02, indicating convergence.
@@
 
-print(xtable(deviance, align=c("l", "l", "l", "l", "l", "X"), 
-             caption="The proportion of interannual variability in  vital rates explained by the climate covariates.
+# print(xtable(deviance, align=c("l", "l", "l", "l", "l", "X"), 
+#              caption="The proportion of interannual variability in  vital rates explained by the climate covariates.
 The contribution for growth is defined as: (Climate model - Constant Model)/(Full model - Constant model).
@@
 The values are model deviance estimated from the `coda' package in R."),
-      tabular.environment="tabularx", caption.placement="top",
-      width="\\textwidth", include.rownames=FALSE)
+#       tabular.environment="tabularx", caption.placement="top",
+#       width="\\textwidth", include.rownames=FALSE)
 @
 
-% # <<echo=FALSE, fig=TRUE,height=3,width=8.5 >>=
-% # library(ggplot2)
-% # plotD <- data.frame(Species = rep(growthSSRTable$Species, 3),
-% #                     Model = rep(c("Growth", "Survival", "Colonization"), each=4),
-% #                     Contribution = c(growthSSRTable[,6],
-% #                                      survivalSSRTable[,6],
-% #                                      colonizationSSRTable[,6]))
-% # ggplot(plotD, aes(x=Species, y=Contribution))+
-% #   geom_bar(stat="identity")+
-% #   facet_grid(.~Model)
-% # @
+<<echo=FALSE, include=FALSE, fig=TRUE, label=ClimFig, width=5, height=3>>=
+library(ggplot2)
+devGQ <- data.frame(VitalRate = c("Growth", "Survival", "Colonization"),
+                       ConstantModel = rep(NA,3),
+                       ClimateModel = rep(NA,3),
+                       FullModel = rep(NA,3),
+                       ClimateContribution = rep(NA,3))
+devGQ$FullModel[1] <- round(read.csv("../montana/ipm/vitalRateRegs/growth/growthDeviance.csv")[,2],2)
+devGQ$ClimateModel[1] <- round(read.csv("../montana/ipm/vitalRateRegs/growth/growthDevianceCLIMATE.csv")[,2],2)
+devGQ$ConstantModel[1] <- round(read.csv("../montana/ipm/vitalRateRegs/growth/growthDevianceCONSTANT.csv")[,2],2)
+devGQ$ClimateContribution[1] <- with(devGQ[1,], (ClimateModel-ConstantModel)/(FullModel-ConstantModel))
+devM <- devGQ[,c(1,5)]
+devM$Level <- rep("Individual", 3)
+tmpD <- deviance[,c(1,5)]
+tmpD$Level <- rep("Quadrat", 3)
+colnames(tmpD)[2] <- "value"
+colnames(tmpD)[1] <- "VitalRate"
+colnames(devM)[2] <- "value"
+plotD <- rbind(tmpD, devM)
+ggplot(data=plotD, aes(x=VitalRate, y=value, fill=Level))+
+  geom_bar(stat="identity", position="dodge")+
+  scale_fill_manual(values=c("steelblue", "darkorange"))+
+  ylab("Contribution of Climate Effects")+
+  xlab("Vital Rate")+
+  theme_bw()
+@
+
+\begin{figure}[th!]
+\begin{center}
+\includegraphics{Fig-ClimFig}
+\end{center}
+\caption{The proportion of interannual variability in  vital rates explained by the climate covariates.
+The contribution for growth is defined as: (Climate model - Constant Model)/(Full model - Constant model).
+The contribution for survival and colonization at the quadrat level, where we could not estimate a full model with year random effects, is defined as: (Constant Model - Climate Model)/Constant Model.
+The values are model deviance estimated from the `coda' package in R.}
+\label{fig:ClimFig}
+\end{figure}
 

### 2014-12-04 manuscript/ms_MicroMeso.tex
updated ms figures

@@
 We did not consider interactions among climate covariates or among climate covariates and plant size or proportional cover.
+Climate effects ($\beta_{c}$s) were modeled hierarchically in each vital rate regression so that each species specific climate effect is drawn from an interspecific climate effect distribution.
+This was necessary to achieve convergence at the quadrat level and we used the same approach at the individual level for uniformity.
 
@@
 \subsubsection{Model fitting}
-All of our analyses (model selection, fitting, and simulating) were conducted in R.
-We used the analytical Bayesian software INLA to fit candidate vital rate models at the individual and quadrat levels for computational efficiency.
-This is the case except for the recruitment model at the individual level which required a more complex hierarchical approach.
-In that case we used the MCMC sampler in JAGS to estimate the posterior distributions of model parameters.
-After model selection, we used MCMC samplers, again in JAGS, to estimate the posterior distributions of model parameters for all vital rate models at the individual and quadrat levels.
-We obtained posterior distributions for all model parameters from three parallel MCMC chains run for 20,000 iterations, after discarding the initial 10,000 iterations.
+All of our analyses (model fitting and simulating) were conducted in Program R.
+We used the MCMC sampler in JAGS to estimate the posterior distributions of model parameters.
+We obtained posterior distributions for all model parameters from three parallel MCMC chains run for 50,000 iterations, after a 50,000 iteration burn in.
 We assessed convergence visually and using the Gelman diagnostic in the R package `coda.' Scale reduction factors for all parameters were less than 1.02, indicating convergence.
@@
 
-% latex table generated in R 3.1.0 by xtable 1.7-3 package
-% Tue Nov 25 11:51:40 2014
-\begin{table}[ht]
-\centering
+\begin{figure}[th!]
+\begin{center}
+\includegraphics{Fig-ClimFig}
+\end{center}
 \caption{The proportion of interannual variability in  vital rates explained by the climate covariates.
@@
 The values are model deviance estimated from the `coda' package in R.}
-\begin{tabularx}{\textwidth}{llllX}
-  \hline
-Vital Rate & Constant model & Climate model & Full model & Contribution of climate covariates \\ 
-  \hline
-Growth & -5566.24 & -5735.32 & -6167.98 & 0.28 \\ 
-  Survival & 171.55 & 170.37 &  & 0.01 \\ 
-  Colonization & 203.98 & 193.65 &  & 0.05 \\ 
-   \hline
-\end{tabularx}
-\end{table}
-% # <<echo=FALSE, fig=TRUE,height=3,width=8.5 >>=
-% # library(ggplot2)
-% # plotD <- data.frame(Species = rep(growthSSRTable$Species, 3),
-% #                     Model = rep(c("Growth", "Survival", "Colonization"), each=4),
-% #                     Contribution = c(growthSSRTable[,6],
-% #                                      survivalSSRTable[,6],
-% #                                      colonizationSSRTable[,6]))
-% # ggplot(plotD, aes(x=Species, y=Contribution))+
-% #   geom_bar(stat="identity")+
-% #   facet_grid(.~Model)
-% # @
+\label{fig:ClimFig}
+\end{figure}
 

### 2014-12-04 manuscript/ms_MicroMeso.Rnw
updated ms

@@
 
-<<echo=FALSE,results=tex>>=
+<<echo=FALSE,results=tex, label=GetQuadDeviance>>=
 library(xtable)
@@
 
-<<echo=FALSE, include=FALSE, fig=TRUE, label=ClimFig, width=5, height=3>>=
+<<echo=FALSE, include=FALSE, fig=TRUE, label=DevFig, width=4, height=3>>=
 library(ggplot2)
@@
 devGQ$ConstantModel[1] <- round(read.csv("../montana/ipm/vitalRateRegs/growth/growthDevianceCONSTANT.csv")[,2],2)
+devGQ$FullModel[2] <- round(read.csv("../montana/ipm/vitalRateRegs/survival/survivalDeviance.csv")[,2],2)
+devGQ$ClimateModel[2] <- round(read.csv("../montana/ipm/vitalRateRegs/survival/survivalDevianceCLIMATE.csv")[,2],2)
+devGQ$ConstantModel[2] <- round(read.csv("../montana/ipm/vitalRateRegs/survival/survivalDevianceCONSTANT.csv")[,2],2)
+devGQ$FullModel[3] <- round(read.csv("../montana/ipm/vitalRateRegs/recruitment/recruitmentDeviance.csv")[,2],2)
+devGQ$ClimateModel[3] <- round(read.csv("../montana/ipm/vitalRateRegs/recruitment/recruitmentDevianceCLIMATE.csv")[,2],2)
+devGQ$ConstantModel[3] <- round(read.csv("../montana/ipm/vitalRateRegs/recruitment/recruitmentDevianceCONSTANT.csv")[,2],2)
 devGQ$ClimateContribution[1] <- with(devGQ[1,], (ClimateModel-ConstantModel)/(FullModel-ConstantModel))
+devGQ$ClimateContribution[2] <- with(devGQ[2,], (ConstantModel-ClimateModel)/(ConstantModel))
+devGQ$ClimateContribution[3] <- with(devGQ[3,], (ConstantModel-ClimateModel)/(ConstantModel))
+
 devM <- devGQ[,c(1,5)]
@@
 plotD <- rbind(tmpD, devM)
-ggplot(data=plotD, aes(x=VitalRate, y=value, fill=Level))+
+gClimCont <- ggplot(data=plotD, aes(x=VitalRate, y=value, fill=Level))+
   geom_bar(stat="identity", position="dodge")+
@@
   theme_bw()
+gClimCont
+@
+
+<<echo=FALSE, include=FALSE, fig=TRUE, label=CoefFig, width=8.5, height=8>>=
+gQC <- read.csv("../montana/quadBM/vitalRateRegressions/growth/growthQuants.csv")
+gQC <- gQC[129:144,]
+coefNames <- c("ppt1", "ppt2", "temp1", "temp2")
+gQC$X <- rep(coefNames, each=4)
+gQC$rate <- "growth"
+sQC <- read.csv("../montana/quadBM/vitalRateRegressions/survival/survivalQuants.csv")
+sQC <- sQC[33:48,]
+coefNames <- c("ppt1", "ppt2", "temp1", "temp2")
+sQC$X <- rep(coefNames, each=4)
+sQC$rate <- "survival"
+rQC <- read.csv("../montana/quadBM/vitalRateRegressions/colonization/colonizationQuants.csv")
+rQC <- rQC[29:44,]
+coefNames <- c("ppt1", "ppt2", "temp1", "temp2")
+rQC$X <- rep(coefNames, each=4)
+rQC$rate <- "colonization"
+plotQC <- rbind(gQC,sQC,rQC)
+gQ <- ggplot(plotQC, aes(x=species, y=X50., color=rate))+
+  geom_hline(aes(yintercept=0), linetype="dashed")+
+  geom_errorbar(aes(ymin=X2.5., ymax=X97.5.), width=0, position=position_dodge(width=0.5))+
+  geom_errorbar(aes(ymin=X25., ymax=X75.), width=0, size=1.5, position=position_dodge(width=0.5))+
+  geom_point(size=4, position=position_dodge(width=0.5))+
+  facet_grid(X~.)+
+  coord_flip()+
+  theme_bw()+
+  guides(color=FALSE)
+
+gIC <- read.csv("../montana/ipm/vitalRateRegs/growth/growthQuants.csv")
+gIC <- gIC[133:148,]
+coefNames <- c("ppt1", "ppt2", "temp1", "temp2")
+gIC$X <- rep(coefNames, each=4)
+gIC$rate <- "growth"
+sIC <- read.csv("../montana/ipm/vitalRateRegs/survival/survivalQuants.csv")
+sIC <- sIC[133:148,]
+coefNames <- c("ppt1", "ppt2", "temp1", "temp2")
+sIC$X <- rep(coefNames, each=4)
+sIC$rate <- "survival"
+rIC <- read.csv("../montana/ipm/vitalRateRegs/recruitment/recruitmentQuants.csv")
+rIC <- rIC[81:96,]
+coefNames <- c("ppt1", "ppt2", "temp1", "temp2")
+rIC$X <- rep(coefNames, each=4)
+rIC$rate <- "colonization"
+plotIC <- rbind(gIC,sIC,rIC)
+gI <- ggplot(plotIC, aes(x=species, y=X50., color=rate))+
+  geom_hline(aes(yintercept=0), linetype="dashed")+
+  geom_errorbar(aes(ymin=X2.5., ymax=X97.5.), width=0, position=position_dodge(width=0.5))+
+  geom_errorbar(aes(ymin=X25., ymax=X75.), width=0, size=1.5, position=position_dodge(width=0.5))+
+  geom_point(size=4, position=position_dodge(width=0.5))+
+  facet_grid(X~.)+
+  coord_flip()+
+  theme_bw()+
+  guides(color=FALSE)
+
+library(gridExtra)
+g <- grid.arrange(gQ, gI, ncol=2, nrow=1)
+g
 @
@@
 \begin{center}
-\includegraphics{Fig-ClimFig}
+\includegraphics{Fig-DevFig}
 \end{center}
@@
 The contribution for growth is defined as: (Climate model - Constant Model)/(Full model - Constant model).
-The contribution for survival and colonization at the quadrat level, where we could not estimate a full model with year random effects, is defined as: (Constant Model - Climate Model)/Constant Model.
-The values are model deviance estimated from the `coda' package in R.}
-\label{fig:ClimFig}
+The contribution for survival and colonization, where we could not estimate a full model with year random effects at the quadrat level, is defined as: (Constant Model - Climate Model)/Constant Model.}
+\label{fig:DevFig}
+\end{figure}
+
+\begin{figure}[th!]
+\begin{center}
+\includegraphics{Fig-CoefFig}
+\end{center}
+\caption{Posterior means (points), 75\% credible intervals (heavy lines), and 95\% credible intervals (light lines) of climate effects in each vital rate statistical model.
+The quadrat level estimates are shown in left panels and individual level estimates are in right panels.
+Color codes: blue = colonization/recruitment, green = growth, red = survival.}
+\label{fig:CoefFig}
 \end{figure}

### 2014-12-04 manuscript/ms_MicroMeso.tex
updated ms

@@
 
+\begin{Schunk}
+\begin{Soutput}
+NULL
+\end{Soutput}
+\end{Schunk}
+
 \begin{figure}[th!]
 \begin{center}
-\includegraphics{Fig-ClimFig}
+\includegraphics{Fig-DevFig}
 \end{center}
@@
 The contribution for growth is defined as: (Climate model - Constant Model)/(Full model - Constant model).
-The contribution for survival and colonization, where we could not estimate a full model with year random effects, is defined as: (Constant Model - Climate Model)/Constant Model.
-The values are model deviance estimated from the `coda' package in R.}
-\label{fig:ClimFig}
+The contribution for survival and colonization, where we could not estimate a full model with year random effects at the quadrat level, is defined as: (Constant Model - Climate Model)/Constant Model.}
+\label{fig:DevFig}
+\end{figure}
+
+\begin{figure}[th!]
+\begin{center}
+\includegraphics{Fig-CoefFig}
+\end{center}
+\caption{Posterior means (points), 75\% credible intervals (heavy lines), and 95\% credible intervals (light lines) of climate effects in each vital rate statistical model.
+The quadrat level estimates are shown in left panels and individual level estimates are in right panels.
+Color codes: blue = colonization/recruitment, green = growth, red = survival.}
+\label{fig:CoefFig}
 \end{figure}

### 2014-12-04 manuscript/ms_MicroMeso.Rnw
ms updates
added deviance and coefficient figures for climate effects

@@
 
-<<echo=FALSE, include=FALSE, fig=TRUE, label=DevFig, width=4, height=3>>=
+<<echo=FALSE, include=FALSE, fig=TRUE, label=DevFig, width=5, height=3, results=hide>>=
 library(ggplot2)
@@
 
-<<echo=FALSE, include=FALSE, fig=TRUE, label=CoefFig, width=8.5, height=8>>=
+<<echo=FALSE, include=FALSE, fig=TRUE, label=CoefFig, width=8.5, height=10, results=hide>>=
 gQC <- read.csv("../montana/quadBM/vitalRateRegressions/growth/growthQuants.csv")
@@
   geom_point(size=4, position=position_dodge(width=0.5))+
+  ylab("Standardized Coefficient Value")+
+  xlab("Species")+
   facet_grid(X~.)+
@@
   theme_bw()+
-  guides(color=FALSE)
+  scale_color_manual(values=c("grey25", "steelblue", "slateblue"))+
+  guides(color=guide_legend(title=""))+
+  theme(legend.position="top")
 
@@
   theme_bw()+
-  guides(color=FALSE)
+  scale_color_manual(values=c("grey25", "steelblue", "slateblue"))+
+#   guides(color=FALSE)+
+  guides(color=guide_legend(title=""))+
+  theme(legend.position="top")
 
@@
 \begin{center}
-\includegraphics{Fig-CoefFig}
+\includegraphics[ width=\textwidth] {Fig-CoefFig}
 \end{center}
 \caption{Posterior means (points), 75\% credible intervals (heavy lines), and 95\% credible intervals (light lines) of climate effects in each vital rate statistical model.
-The quadrat level estimates are shown in left panels and individual level estimates are in right panels.
-Color codes: blue = colonization/recruitment, green = growth, red = survival.}
+The quadrat level estimates are shown in left panels and individual level estimates are in right panels.}
 \label{fig:CoefFig}

### 2014-12-04 manuscript/ms_MicroMeso.Rnw
cosmetic updates to figures

@@
   guides(color=guide_legend(title=""))+
-  theme(legend.position="top")
+  theme(legend.position="top")+
+  ggtitle("Quadrat-level")
 
@@
   theme_bw()+
-  scale_color_manual(values=c("grey25", "steelblue", "slateblue"))+
+  scale_color_manual(values=c("grey25", "steelblue", "slateblue"),
+                     labels=c("recruitment", "growth", "survival"))+
 #   guides(color=FALSE)+
   guides(color=guide_legend(title=""))+
-  theme(legend.position="top")
+  theme(legend.position="top")+
+  ggtitle("Individual-level")
 

### 2014-12-04 manuscript/ms_MicroMeso.Rnw
figure update

--- 2014-12-04 manuscript/ms_MicroMeso.Rnw
@@
   geom_errorbar(aes(ymin=X2.5., ymax=X97.5.), width=0, position=position_dodge(width=0.5))+
-  geom_errorbar(aes(ymin=X25., ymax=X75.), width=0, size=1.5, position=position_dodge(width=0.5))+
-  geom_point(size=4, position=position_dodge(width=0.5))+
+  geom_errorbar(aes(ymin=X25., ymax=X75.), width=0, size=1.25, position=position_dodge(width=0.5))+
+  geom_point(size=4, position=position_dodge(width=0.5), shape="|")+
   ylab("Standardized Coefficient Value")+
@@
 \end{center}
-\caption{Posterior means (points), 75\% credible intervals (heavy lines), and 95\% credible intervals (light lines) of climate effects in each vital rate statistical model.
+\caption{Posterior means (vertical ticks), 75\% credible intervals (heavy lines), and 95\% credible intervals (light lines) of climate effects in each vital rate statistical model.
 The quadrat level estimates are shown in left panels and individual level estimates are in right panels.}

### 2014-12-04 manuscript/ms_MicroMeso.tex
ms updates
added deviance and coefficient figures for climate effects
figure update

@@
 
-\begin{Schunk}
-\begin{Soutput}
-NULL
-\end{Soutput}
-\end{Schunk}
-
 \begin{figure}[th!]
@@
 \begin{center}
-\includegraphics{Fig-CoefFig}
+\includegraphics[ width=\textwidth] {Fig-CoefFig}
 \end{center}
-\caption{Posterior means (points), 75\% credible intervals (heavy lines), and 95\% credible intervals (light lines) of climate effects in each vital rate statistical model.
-The quadrat level estimates are shown in left panels and individual level estimates are in right panels.
-Color codes: blue = colonization/recruitment, green = growth, red = survival.}
+\caption{Posterior means (vertical ticks), 75\% credible intervals (heavy lines), and 95\% credible intervals (light lines) of climate effects in each vital rate statistical model.
+The quadrat level estimates are shown in left panels and individual level estimates are in right panels.}
 \label{fig:CoefFig}

### 2014-12-05 manuscript/ms_MicroMeso.Rnw
updated ms

@@
 
-\subsection{Computing}
-\subsubsection{Model fitting}
+\subsection{Model fitting}
 All of our analyses (model fitting and simulating) were conducted in Program R.
@@
 
-\subsubsection{Population models}
+\subsection{Population models}
 With the posterior distribution of the vital rate statistical models in hand, it is straightforward to simulate the population models.
@@
   theme(legend.position="top")+
-  ggtitle("Quadrat-level")
+  ggtitle("Quadrat level")
 
@@
   theme(legend.position="top")+
-  ggtitle("Individual-level")
+  ggtitle("Individual level")
 
@@
 \begin{center}
-\includegraphics{Fig-DevFig}
+\includegraphics[ width=0.5\textwidth ]{Fig-DevFig}
 \end{center}
@@
 \caption{Posterior means (vertical ticks), 75\% credible intervals (heavy lines), and 95\% credible intervals (light lines) of climate effects in each vital rate statistical model.
-The quadrat level estimates are shown in left panels and individual level estimates are in right panels.}
+The dashed vertical line is at 0, indicating no effect.}
 \label{fig:CoefFig}

### 2014-12-05 montana/quadBM/simulations/oneStepForecastQBM.R
added one-step-head forecast script
this will be used to assess model performance based on historical data
via residuals

@@ -1,0 +1,185 @@
+#Quad-Based Model simulations
+
+#!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!#
+# Set working directory to location of this source file #
+#!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!#
+
+#clear everything, just to be safe 
+rm(list=ls(all=TRUE))
+
+library(reshape2)
+library(plyr)
+library(ggplot2)
+library(ggthemes)
+library(gridExtra)
+
+#bring in data
+allD <- read.csv("../../speciesData/quadAllCover.csv")
+allD <- allD[,2:ncol(allD)] #get rid of X ID column
+allD$percCover <- allD$totCover/10000
+head(scale(allD$percCover, center=TRUE, scale=TRUE))
+sppList <- as.character(unique(allD$Species))
+
+#bring in climate data
+climD <- read.csv("../../weather/Climate.csv")
+climD[3:6] <- scale(climD[3:6], center = TRUE, scale = TRUE)
+
+doSpp <- sppList[1]
+
+#load vital rate parameters
+pCol <- readRDS("../vitalRateRegressions/colonization/colonizationParamsMCMC.rds")
+pGrow <- readRDS("../vitalRateRegressions/growth/growthParamsMCMC.rds")
+pSurv <- readRDS("../vitalRateRegressions/survival/survivalParamsMCMC.rds")
+
+####
+#### Organize parameter values
+####
+#colonization
+pCol2 <- melt(pCol)
+pCol2$Spp <- c(rep(rep(sppList, each=3000), times=6),
+               rep(rep(sppList, each=3000), times=5))
+pCol2$Coef <- c(rep("gInt", times=6*4*3000),
+                rep("int", times=4*3000),
+                rep("rain1", times=4*3000),
+                rep("rain2", times=4*3000),
+                rep("temp1", times=4*3000),
+                rep("temp2", times=4*3000))
+colnames(pCol2)[1] <- "Iter"
+pCol <- pCol2[,c(1,3:5)]; rm(pCol2)
+
+#survival
+pSurv2 <- melt(pSurv)
+pSurv2$Spp <- c(rep(rep(sppList, each=3000), times=1),
+                rep(rep(sppList, each=3000), times=6),
+               rep(rep(sppList, each=3000), times=5))
[…]

### 2014-12-05 manuscript/ms_MicroMeso.Rnw
complete one step simulation; ms updates

@@
 
-\title{\LARGE Micro- versus meso-scale population models for ecological forecasting under climate change}
+\title{\LARGE Do we need detailed demographic data to forecast plant population responses to climate change?}
 \author[]{\large Andrew T. Tredennick\footnote{\emph{email:} atredenn@gmail.com}}
@@
 
+<<echo=FALSE, include=FALSE, fig=TRUE, label=ResidFig, width=5, height=6, results=hide>>=
+library(plyr)
+library(reshape2)
+resD <- readRDS("../montana/quadBM/simulations/quadBM_oneStep_Residuals.rds")
+nSims <- length(unique(resD$sim))
+diffD <- ddply(resD, c("year", "species"), summarise,
+               avgDiff = mean(yearDiff))
+
+ggplot(resD)+
+  geom_hline(aes(yintercept=0))+
+#   geom_line(data=diffD, aes(x=year, y=avgDiff, group=species), color="purple", alpha=0.5)+
+  stat_smooth(data=diffD, aes(x=year, y=avgDiff, group=species), color="lightblue", se = FALSE, size=1)+
+  geom_boxplot(aes(x=year, y=Residuals), outlier.shape = 1, width=0.3, outlier.size = 0.5)+
+  facet_grid(species~., scales = "free_y")+
+  xlab("Year")+
+  ylab("Residuals (% cover difference)")+
+  theme_bw()+
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+@
+
 \begin{figure}[th!]
@@
 \caption{Posterior means (vertical ticks), 75\% credible intervals (heavy lines), and 95\% credible intervals (light lines) of climate effects in each vital rate statistical model.
-The dashed vertical line is at 0, indicating no effect.}
+The dashed vertical line is at 0, indicating no effect.
+Horizontal line at 0 indicates perfect agreement between mean observed cover in that year and the model predictions.}
 \label{fig:CoefFig}
@@
 
+\begin{figure}[th!]
+\begin{center}
+\includegraphics{Fig-ResidFig}
+\end{center}
+\caption{Boxplots of model residuals for one-step-ahead forecasts at each observation year.
+Each one-step forecast was simulated \Sexpr{nSims} times.
+Note that the y-axes vary across panels.
+The light blue line shows the difference between the observed-year percent cover and the average cover observed across all years.
+The models tend to underpredict and perform poorly when observed cover in a given year is a large deviant from the mean.}
+\label{fig:ResidFig}
+\end{figure}
+
 \end{document}

### 2014-12-05 manuscript/ms_MicroMeso.tex
complete one step simulation; ms updates

@@
 
+\begin{figure}[th!]
+\begin{center}
+\includegraphics{Fig-ResidFig}
+\end{center}
+\caption{Boxplots of model residuals for one-step-ahead forecasts at each observation year.
+Each one-step forecast was simulated 100 times.
+Note that the y-axes vary across panels.
+The light blue line shows the difference between the observed-year percent cover and the average cover observed across all years.
+The models tend to underpredict and perform poorly when observed cover in a given year is a large deviant from the mean.}
+\label{fig:ResidFig}
+\end{figure}
+
 \end{document}

### 2014-12-10 manuscript/ms_MicroMeso.Rnw
updated figs and residual analysis
residuals set up to compare cover change as opposed to just absolute
cover in each year. Saved output can still be used to calculate
“regular” residuals.

@@
 
-\noindent where $x$ is the log of genet size (or log of quadrat areal cover), $\beta_{0,t}$ is a year-specific intercept, $\beta_{s,t}$ is the year-specific slope parameter for size, $\beta_{Q}$ is the random effect of quadrat group location, $\theta$ is a matrix of $i$ climate effects over $t$ years, $\beta_{c,i}$ is the fixed parameter for the effect of the $i$th climate covariate, and $\varepsilon_{t}$ is the error term.
+\noindent where $x$ is the log of genet size (quadrat areal cover), $\beta_{0,t}$ is a year-specific intercept, $\beta_{s,t}$ is the year-specific slope parameter for size, $\beta_{Q}$ is the random effect of quadrat group location, $\theta$ is a matrix of $i$ climate effects over $t$ years, $\beta_{c,i}$ is the fixed parameter for the effect of the $i$th climate covariate, and $\varepsilon_{t}$ is the error term.
 At the quadrat level we did not fit random year effects for survival because those parameters did not reach convergence (see "Computing").
@@
 
-<<echo=FALSE, include=FALSE, fig=TRUE, label=DevFig, width=5, height=3, results=hide>>=
+<<echo=FALSE, include=FALSE, fig=TRUE, label=DevFig, width=7, height=3, results=hide>>=
 library(ggplot2)
@@
 plotD <- rbind(tmpD, devM)
-gClimCont <- ggplot(data=plotD, aes(x=VitalRate, y=value, fill=Level))+
-  geom_bar(stat="identity", position="dodge")+
-  scale_fill_manual(values=c("steelblue", "darkorange"))+
+gClimCont1 <- ggplot(data=subset(plotD, Level=="Individual"), aes(x=VitalRate, y=value))+
+  geom_bar(stat="identity", position="dodge", fill="grey30")+
   ylab("Contribution of Climate Effects")+
   xlab("Vital Rate")+
+  ggtitle("Individual level")+
+  scale_y_continuous(limits=c(0, 0.5))+
+  scale_x_discrete(labels=c("Recruitment", "Growth", "Survival"))+
   theme_bw()
-gClimCont
+gClimCont2 <- ggplot(data=subset(plotD, Level=="Quadrat"), aes(x=VitalRate, y=value))+
+  geom_bar(stat="identity", position="dodge", fill="grey30")+
+  ylab("Contribution of Climate Effects")+
+  xlab("Vital Rate")+
+  ggtitle("Quadrat level")+
+  scale_y_continuous(limits=c(0, 0.5))+
+  scale_x_discrete(labels=c("Colonization", "Cover change", "Extirpation"))+
+  theme_bw()
+library(gridExtra)
+gClim <- grid.arrange(gClimCont1, gClimCont2, ncol=2, nrow=1)
+gClim
 @
 
-<<echo=FALSE, include=FALSE, fig=TRUE, label=CoefFig, width=8.5, height=10, results=hide>>=
+<<echo=FALSE, include=FALSE, fig=TRUE, label=CoefFig, width=4, height=10, results=hide>>=
 gQC <- read.csv("../montana/quadBM/vitalRateRegressions/growth/growthQuants.csv")
@@
 
-library(gridExtra)
-g <- grid.arrange(gQ, gI, ncol=2, nrow=1)
-g
+####
+#### Plot just for growth
+####
+growthIC <- subset(plotIC, rate=="growth")
+growthQC <- subset(plotQC, rate=="growth")
+growthIC$Level <- "Individual"
+growthQC$Level <- "Quadrat"
+growthAll <- rbind(growthIC, growthQC)
+
+gGrowth <- ggplot(growthAll, aes(x=species, y=X50., color=Level))+
+  geom_hline(aes(yintercept=0), linetype="dashed")+
+  geom_errorbar(aes(ymin=X2.5., ymax=X97.5.), width=0, position=position_dodge(width=0.5))+
+  geom_errorbar(aes(ymin=X25., ymax=X75.), width=0, size=1.25, position=position_dodge(width=0.5))+
+  geom_point(size=4, position=position_dodge(width=0.5), shape="|")+
+  ylab("Standardized Coefficient Value")+
+  xlab("Species")+
+  facet_grid(X~.)+
+  coord_flip()+
+  theme_bw()+
+  scale_color_manual(values=c("grey25", "slateblue"))+
+#   guides(color=FALSE)+
+  guides(color=guide_legend(title=""))+
+  theme(legend.position="top")
+print(gGrowth)
 @
@@
   stat_smooth(data=diffD, aes(x=year, y=avgDiff, group=species), color="lightblue", se = FALSE, size=1)+
-  geom_boxplot(aes(x=year, y=Residuals), outlier.shape = 1, width=0.3, outlier.size = 0.5)+
+  geom_boxplot(aes(x=year, y=covChangeResiduals), outlier.shape = 1, width=0.3, outlier.size = 0.5)+
   facet_grid(species~., scales = "free_y")+
@@
 \begin{center}
-\includegraphics[ width=0.5\textwidth ]{Fig-DevFig}
+\includegraphics[ width=1\textwidth ]{Fig-DevFig}
 \end{center}
@@
 \begin{center}
-\includegraphics[ width=\textwidth] {Fig-CoefFig}
+\includegraphics[ width=0.5\textwidth] {Fig-CoefFig}
 \end{center}
-\caption{Posterior means (vertical ticks), 75\% credible intervals (heavy lines), and 95\% credible intervals (light lines) of climate effects in each vital rate statistical model.
+\caption{Posterior means (vertical ticks), 75\% credible intervals (heavy lines), and 95\% credible intervals (light lines) of climate effects on growth at both levels of inferences.
 The dashed vertical line is at 0, indicating no effect.

### 2014-12-10 montana/quadBM/simulations/oneStepForecastQBM.R

@@ -7,6 +7,8 @@
 #clear everything, just to be safe 
 rm(list=ls(all=TRUE))
 
+NumberSimsPerYear <- 100
+
 library(reshape2)
 library(plyr)
 library(ggplot2)
@@ -150,7 +152,7 @@ outD <- data.frame(year=NA, cover=NA, sim=NA, species=NA)
 
 for(i in 1:length(sppList)){
   sppSim <- sppList[i]
-  nSim <- 100
+  nSim <- NumberSimsPerYear
   yearsN <- length(unique(allD$year))
   years <- unique(allD$year)+1900
   yearsID <- unique(allD$year)
@@ -191,9 +193,29 @@ sppD <- ddply(allD, .variables = c("Species"), .fun = summarise,
 quadD <- merge(quadD1, sppD)
 quadD$year <- quadD$year+1900
 quadD$yearDiff <- with(quadD, cover*100-avgcover*100)
+tmp <- which(outD$year==1932)
+tmp2 <- which(quadD$year==1932)
+outD[tmp, "cover"] <- quadD[tmp2, "cover"]
 
-resD <- merge(outD, quadD, by.x = c("species", "year"), by.y = c("Species", "year"))
-resD$Residuals <- with(resD, cover.x*100 - cover.y*100)
+#create lag cover variable
+lagD <- outD
+lagD$year2 <- as.numeric(lagD$year)+1
+lagD <- lagD[, c(2,4,5)]
+colnames(lagD) <- c("lagCover", "species", "year")
+outD2 <- merge(outD, lagD, by=c("species", "year"))
+
+lagQ <- quadD
+lagQ$year2 <- as.numeric(lagQ$year)+1
+lagQ <- lagQ[, c(1,3,6)]
+colnames(lagQ) <- c( "Species", "lagCover", "year")
+quadD2 <- merge(quadD, lagQ, by=c("Species", "year"))
+
+#calculate expected and observed cover change
+quadD2$coverChange <- with(quadD2, cover*100-lagCover*100)
+outD2$coverChange <- with(outD2, cover*100-lagCover*100)
+
+resD <- merge(outD2, quadD2, by.x = c("species", "year"), by.y = c("Species", "year"))
+resD$covChangeResiduals <- with(resD, coverChange.x - lagCover.y)
 resD <- subset(resD, year!=1932)
 saveRDS(resD, file = "quadBM_oneStep_Residuals.rds")
 

### 2014-12-11 manuscript/manuscript.Rmd
adding markdown ms file

--- 2014-12-05 manuscript/ms_MicroMeso.tex
@@
-\documentclass[12pt]{article}
-\usepackage{graphics}
-\usepackage{graphicx, verbatim}
-\usepackage{amssymb,amsmath}
-\usepackage{tabularx}
-% \usepackage[T1]{fontenc}
-% \usepackage[utf8]{inputenc}
-\usepackage{authblk}
-\textwidth=6.2in
-\textheight=8.5in
-%\parskip=.3cm
-\oddsidemargin=.1in
-\evensidemargin=.1in
-\headheight=-.3in
-\usepackage{lineno}
-\linenumbers
-\modulolinenumbers[2]
-\usepackage{setspace}
-\doublespacing
-
-\usepackage{Sweave}
-\begin{document}
-\input{ms_MicroMeso-concordance}
-
-%To put figures in subfolder
-%\SweaveOpts{prefix.string=figures/fig}
-\DefineVerbatimEnvironment{Sinput}{Verbatim} {xleftmargin=2em}
-\DefineVerbatimEnvironment{Soutput}{Verbatim}{xleftmargin=2em}
-\DefineVerbatimEnvironment{Scode}{Verbatim}{xleftmargin=2em}
-
-\title{\LARGE Micro- versus meso-scale population models for ecological forecasting under climate change}
-\author[]{\large Andrew T. Tredennick\footnote{\emph{email:} atredenn@gmail.com}}
-\author[]{\large Peter B. Adler}
-\affil[]{\footnotesize{\emph{Department of Wildland Resources and the Ecology Center, Utah State University}}}
-\maketitle
-
-\begin{abstract}
-\noindent Forecasts of populations under future climate change are constrained by the spatial and temporal scales at which we can gather data.
+---
+layout: 12pt,review
+linenumbers: true
+title: "Do we need detailed demographic data to forecast plant population responses to climate change?"
+author: 
+  - name: Andrew T. Tredennick
+    affiliation: usu
+    email: atredenn(at)gmail.com
+    footnote: Corresponding author
+  - name: Peter B. Adler
+    affiliation: usu
+address: 
+  - code: usu
+    address: | 
+      Department of Wildland Resources and the Ecology Center, 
+      Utah State University,
+      Logan, UT 84322, USA
+abstract: |
+  Forecasts of populations under future climate change are constrained by the spatial and temporal scales at which we can gather data.
 For plants, ecologists often require individual level data to build population models; data that is difficult and expensive to collect at spatial scales greater than one square meter.
@@
 Here we use a longterm dataset to model population dynamics using individual-level and plot-level data.
-\end{abstract}
 
-\noindent{}\textbf{Keywords:} population model, forecasting, integral projection model, etc.
+keywords:
+  - ecological forecasts
+  - population models
+  - integral projection model
+  - climate change
+  - grassland
 
-\section{Introduction}
+bibliography: components/references.bib
+csl: components/ecology.csl
+documentclass: components/elsarticle
+
+## rmarkdown render options
+output:
+  pdf_document:
+    fig_caption: true
+    template: components/elsarticle.latex
+    keep_tex: true
+
+---
+
+```{r caching, include=FALSE}
+library("methods")
+library("knitr")
+basename <- "manuscript"
+opts_chunk$set(fig.path = paste("components/figure/", basename, "-", sep=""),
+               cache.path = paste("components/cache/", basename, "/", sep=""))
+opts_chunk$set(cache = 2)
+opts_chunk$set(tidy=FALSE, warning=FALSE, message=FALSE, 
+               comment = NA, verbose = TRUE, echo=FALSE)
+
+# PDF-based figures
+opts_chunk$set(dev='pdf')
+```
+
+```{r plot-options, message=FALSE, warning=FALSE, include=FALSE, echo=FALSE}
+cbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", 
+               "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
+
+opts_chunk$set(dev='pdf', dev.args=c(version="1.3"))
+
+```
+
+Introduction
+============
 Producing skillfull forecasts at spatial scales relevant to global environmental changes that play out at the landscape and regional scale has proven challenging for population ecologists.
@@
 
-\section{Materials and Methods}
-\subsection{Study site and data}
+Materials and Methods
+=====================
+### Study site and data
 Our demographic data comes from the Fort Keogh Livestock and Range Research Laboratory in eastern Montana's northern mixed prairie near Miles City, Montana, USA (46 deg.
@@
 
-\subsection{Stastical models of vital rates}
+### Stastical models of vital rates
 The first step in building our single-species population models was to fit statistical models of vital rates at both levels of inference: individual-level and quadrat-level.
@@
 
-\noindent where $x$ is the log of genet size (or log of quadrat areal cover), $\beta_{0,t}$ is a year-specific intercept, $\beta_{s,t}$ is the year-specific slope parameter for size, $\beta_{Q}$ is the random effect of quadrat group location, $\theta$ is a matrix of $i$ climate effects over $t$ years, $\beta_{c,i}$ is the fixed parameter for the effect of the $i$th climate covariate, and $\varepsilon_{t}$ is the error term.
+where $x$ is the log of genet size (quadrat areal cover), $\beta_{0,t}$ is a year-specific intercept, $\beta_{s,t}$ is the year-specific slope parameter for size, $\beta_{Q}$ is the random effect of quadrat group location, $\theta$ is a matrix of $i$ climate effects over $t$ years, $\beta_{c,i}$ is the fixed parameter for the effect of the $i$th climate covariate, and $\varepsilon_{t}$ is the error term.
 At the quadrat level we did not fit random year effects for survival because those parameters did not reach convergence (see "Computing").
@@
 
-\noindent where $x$ is genet size and all other paramters are as described for the survival regression.
+where $x$ is genet size and all other paramters are as described for the survival regression.
 For the quadrat-level approach we modeled growth as a process describing proportional cover within a quadrat at time $t+1$ as a function of proportional cover at time $t$ and climate covariates.
@@
 
-\subsection{Including climate covariates (and avoiding model selection)}
+### Including climate covariates (and avoiding model selection)
 For both the individual-level IPM (IPM) and the quadrat-based IBM (QBM) we included four climate covariates in each vital rate regression: fall through spring precipitation at \emph{t}-1 and \emph{t}-2 (ppt1 and ppt2, respectively) and mean spring temperature at \emph{t}-1 and \emph{t}-2 (TmeanSpr1 and TmeanSpr2, respectively), where \emph{t} is the observation year.
@@
 
-\subsection{Model fitting}
+### Model fitting
 All of our analyses (model fitting and simulating) were conducted in Program R.
@@
 
-\subsection{Population models}
+### Population models
 With the posterior distribution of the vital rate statistical models in hand, it is straightforward to simulate the population models.
 
-\section{Results}
+Results
+=======
 We assessed the statistical importance of the climate covariates included the final vital rate regressions by comparing the residual deviance of models with climate covariates and temporal random effects, climate covariates only, and temporal random effects only.
@@
 
-\begin{figure}[th!]
-\begin{center}
-\includegraphics[ width=0.5\textwidth ]{Fig-DevFig}
-\end{center}
-\caption{The proportion of interannual variability in  vital rates explained by the climate covariates.
+```{r figure_1, dependson="plot-options", fig.cap="The proportion of interannual variability in  vital rates explained by the climate covariates.
 The contribution for growth is defined as: (Climate model - Constant Model)/(Full model - Constant model).
-The contribution for survival and colonization, where we could not estimate a full model with year random effects at the quadrat level, is defined as: (Constant Model - Climate Model)/Constant Model.}
-\label{fig:DevFig}
-\end{figure}
+The contribution for survival and colonization, where we could not estimate a full model with year random effects at the quadrat level, is defined as: (Constant Model - Climate Model)/Constant Model.", fig.width=7, fig.height=3, cache=FALSE, results='hide'}
+deviance <- data.frame(VitalRate = c("Growth", "Survival", "Colonization"),
+                       ConstantModel = rep(NA,3),
+                       ClimateModel = rep(NA,3),
+                       FullModel = rep(NA,3),
+                       ClimateContribution = rep(NA,3))
+deviance$FullModel[1] <- round(read.csv("../montana/quadBM/vitalRateRegressions/growth/growthDeviance.csv")[,2],2)
+deviance$ClimateModel[1] <- round(read.csv("../montana/quadBM/vitalRateRegressions/growth/growthDevianceCLIMATE.csv")[,2],2)
+deviance$ConstantModel[1] <- round(read.csv("../montana/quadBM/vitalRateRegressions/growth/growthDevianceCONSTANT.csv")[,2],2)
+deviance$ClimateContribution[1] <- with(deviance[1,], (ClimateModel-ConstantModel)/(FullModel-ConstantModel))
+deviance$ClimateModel[2] <- round(read.csv("../montana/quadBM/vitalRateRegressions/survival/survivalDeviance.csv")[,2],2)
+deviance$ConstantModel[2] <- round(read.csv("../montana/quadBM/vitalRateRegressions/survival/survivalDevianceCONSTANT.csv")[,2],2)
+deviance$ClimateModel[3] <- round(read.csv("../montana/quadBM/vitalRateRegressions/colonization/colonizationDeviance.csv")[,2],2)
+deviance$ConstantModel[3] <- round(read.csv("../montana/quadBM/vitalRateRegressions/colonization/colonizationDevianceCONSTANT.csv")[,2],2)
+deviance$ClimateContribution[2:3] <- with(deviance[2:3,], (ConstantModel-ClimateModel)/(ConstantModel))
 
-\begin{figure}[th!]
-\begin{center}
-\includegraphics[ width=\textwidth] {Fig-CoefFig}
-\end{center}
-\caption{Posterior means (vertical ticks), 75\% credible intervals (heavy lines), and 95\% credible intervals (light lines) of climate effects in each vital rate statistical model.
-The dashed vertical line is at 0, indicating no effect.}
-\label{fig:CoefFig}
-\end{figure}
+colnames(deviance) <- c("Vital Rate", "Constant model", "Climate model", "Full model", "Contribution of climate covariates")
+library(ggplot2)
+devGQ <- data.frame(VitalRate = c("Growth", "Survival", "Colonization"),
+                       ConstantModel = rep(NA,3),
+                       ClimateModel = rep(NA,3),
+                       FullModel = rep(NA,3),
+                       ClimateContribution = rep(NA,3))
+devGQ$FullModel[1] <- round(read.csv("../montana/ipm/vitalRateRegs/growth/growthDeviance.csv")[,2],2)
+devGQ$ClimateModel[1] <- round(read.csv("../montana/ipm/vitalRateRegs/growth/growthDevianceCLIMATE.csv")[,2],2)
+devGQ$ConstantModel[1] <- round(read.csv("../montana/ipm/vitalRateRegs/growth/growthDevianceCONSTANT.csv")[,2],2)
+devGQ$FullModel[2] <- round(read.csv("../montana/ipm/vitalRateRegs/survival/survivalDeviance.csv")[,2],2)
+devGQ$ClimateModel[2] <- round(read.csv("../montana/ipm/vitalRateRegs/survival/survivalDevianceCLIMATE.csv")[,2],2)
+devGQ$ConstantModel[2] <- round(read.csv("../montana/ipm/vitalRateRegs/survival/survivalDevianceCONSTANT.csv")[,2],2)
+devGQ$FullModel[3] <- round(read.csv("../montana/ipm/vitalRateRegs/recruitment/recruitmentDeviance.csv")[,2],2)
+devGQ$ClimateModel[3] <- round(read.csv("../montana/ipm/vitalRateRegs/recruitment/recruitmentDevianceCLIMATE.csv")[,2],2)
+devGQ$ConstantModel[3] <- round(read.csv("../montana/ipm/vitalRateRegs/recruitment/recruitmentDevianceCONSTANT.csv")[,2],2)
+devGQ$ClimateContribution[1] <- with(devGQ[1,], (ClimateModel-ConstantModel)/(FullModel-ConstantModel))
+devGQ$ClimateContribution[2] <- with(devGQ[2,], (ConstantModel-ClimateModel)/(ConstantModel))
+devGQ$ClimateContribution[3] <- with(devGQ[3,], (ConstantModel-ClimateModel)/(ConstantModel))
 
-\end{document}
+devM <- devGQ[,c(1,5)]
+devM$Level <- rep("Individual", 3)
+tmpD <- deviance[,c(1,5)]
+tmpD$Level <- rep("Quadrat", 3)
+colnames(tmpD)[2] <- "value"
+colnames(tmpD)[1] <- "VitalRate"
+colnames(devM)[2] <- "value"
+plotD <- rbind(tmpD, devM)
+gClimCont1 <- ggplot(data=subset(plotD, Level=="Individual"), aes(x=VitalRate, y=value))+
+  geom_bar(stat="identity", position="dodge", fill="grey30")+
+  ylab("Contribution of Climate Effects")+
+  xlab("Vital Rate")+
+  ggtitle("Individual level")+
+  scale_y_continuous(limits=c(0, 0.5))+
+  scale_x_discrete(labels=c("Recruitment", "Growth", "Survival"))+
+  theme_bw()+
+  theme(axis.text.x = element_text(angle = 25, hjust = 1))
+gClimCont2 <- ggplot(data=subset(plotD, Level=="Quadrat"), aes(x=VitalRate, y=value))+
+  geom_bar(stat="identity", position="dodge", fill="grey30")+
+  ylab("Contribution of Climate Effects")+
+  xlab("Vital Rate")+
+  ggtitle("Quadrat level")+
+  scale_y_continuous(limits=c(0, 0.5))+
+  scale_x_discrete(labels=c("Colonization", "Cover change", "Extirpation"))+
+  theme_bw()+
+  theme(axis.text.x = element_text(angle = 25, hjust = 1))
+library(gridExtra)
+gClim <- grid.arrange(gClimCont1, gClimCont2, ncol=2, nrow=1)
+gClim
+
+```

### 2014-12-11 manuscript/manuscript.tex
adding markdown ms file

--- 2014-12-10 manuscript/ms_MicroMeso.tex
@@
-\documentclass[12pt]{article}
-\usepackage{graphics}
-\usepackage{graphicx, verbatim}
+\documentclass[author-year, 12pt,review]{components/elsarticle} %review=doublespace preprint=single 5p=2 column
+%%% Begin My package additions %%%%%%%%%%%%%%%%%%%
+\usepackage[hyphens]{url}
+\usepackage{lineno} % add 
+  \linenumbers % turns line numbering on 
+\bibliographystyle{elsarticle-harv}
+\biboptions{sort&compress} % For natbib
+\usepackage{graphicx}
+\usepackage{booktabs} % book-quality tables
+%% Redefines the elsarticle footer
+\makeatletter
+\def\ps@pprintTitle{%
+ \let\@oddhead\@empty
+ \let\@evenhead\@empty
+ \def\@oddfoot{\it \hfill\today}%
+ \let\@evenfoot\@oddfoot}
+\makeatother
+
+% A modified page layout
+\textwidth 6.75in
+\oddsidemargin -0.15in
+\evensidemargin -0.15in
+\textheight 9in
+\topmargin -0.5in
+%%%%%%%%%%%%%%%% end my additions to header
+
+\usepackage[T1]{fontenc}
+\usepackage{lmodern}
 \usepackage{amssymb,amsmath}
-\usepackage{tabularx}
-% \usepackage[T1]{fontenc}
-% \usepackage[utf8]{inputenc}
-\usepackage{authblk}
-\textwidth=6.2in
-\textheight=8.5in
-%\parskip=.3cm
-\oddsidemargin=.1in
-\evensidemargin=.1in
-\headheight=-.3in
-\usepackage{lineno}
-\linenumbers
-\modulolinenumbers[2]
-\usepackage{setspace}
-\doublespacing
+\usepackage{ifxetex,ifluatex}
+\usepackage{fixltx2e} % provides \textsubscript
+% use upquote if available, for straight quotes in verbatim environments
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
+\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
+  \usepackage[utf8]{inputenc}
+\else % if luatex or xelatex
+  \usepackage{fontspec}
+  \ifxetex
+    \usepackage{xltxtra,xunicode}
+  \fi
+  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
+  \newcommand{\euro}{€}
+\fi
+% use microtype if available
+\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
+\usepackage{graphicx}
+% We will generate all images so they have a width \maxwidth. This means
+% that they will get their normal width if they fit onto the page, but
+% are scaled down if they would overflow the margins.
+\makeatletter
+\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth
+\else\Gin@nat@width\fi}
+\makeatother
+\let\Oldincludegraphics\includegraphics
+\renewcommand{\includegraphics}[1]{\Oldincludegraphics[width=\maxwidth]{#1}}
+\ifxetex
+  \usepackage[setpagesize=false, % page size defined by xetex
+              unicode=false, % unicode breaks when used with xetex
+              xetex]{hyperref}
+\else
+  \usepackage[unicode=true]{hyperref}
+\fi
+\hypersetup{breaklinks=true,
+            bookmarks=true,
+            pdfauthor={},
+            pdftitle={Do we need detailed demographic data to forecast plant population responses to climate change?},
+            colorlinks=true,
+            urlcolor=blue,
+            linkcolor=magenta,
+            pdfborder={0 0 0}}
+\urlstyle{same}  % don't use monospace font for urls
+\setlength{\parindent}{0pt}
+\setlength{\parskip}{6pt plus 2pt minus 1pt}
+\setlength{\emergencystretch}{3em}  % prevent overfull lines
+\setcounter{secnumdepth}{0}
+% Pandoc toggle for numbering sections (defaults to be off)
+\setcounter{secnumdepth}{0}
+% Pandoc header
 
-\usepackage{Sweave}
 \begin{document}
-\input{ms_MicroMeso-concordance}
+\begin{frontmatter}
 
-%To put figures in subfolder
-%\SweaveOpts{prefix.string=figures/fig}
-\DefineVerbatimEnvironment{Sinput}{Verbatim} {xleftmargin=2em}
-\DefineVerbatimEnvironment{Soutput}{Verbatim}{xleftmargin=2em}
-\DefineVerbatimEnvironment{Scode}{Verbatim}{xleftmargin=2em}
+  \title{Do we need detailed demographic data to forecast plant population
+responses to climate change?}
+    \author[usu]{Andrew T. Tredennick\corref{c1}}
+   \ead{atredenn(at)gmail.com} 
+   \cortext[c1]{Corresponding author}
+    \author[usu]{Peter B. Adler}
 
-\title{\LARGE Do we need detailed demographic data to forecast plant population responses to climate change?}
-\author[]{\large Andrew T. Tredennick\footnote{\emph{email:} atredenn@gmail.com}}
-\author[]{\large Peter B. Adler}
-\affil[]{\footnotesize{\emph{Department of Wildland Resources and the Ecology Center, Utah State University}}}
-\maketitle
+      \address[usu]{Department of Wildland Resources and the Ecology Center, Utah State
+University, Logan, UT 84322, USA}
 
   \begin{abstract}
-\noindent Forecasts of populations under future climate change are constrained by the spatial and temporal scales at which we can gather data.
-For plants, ecologists often require individual level data to build population models; data that is difficult and expensive to collect at spatial scales greater than one square meter.
-Models built using such data are difficult to scale-up to spatial scales relevant to land management decisions and policy intervention.
-A possible route forward are density-structured models based on coarse-scale census data, which are much easier to collect over large spatial extents.
-However, to use coarse-scale data in a climate change context requires the assumption that the climate signals in population dynamics identified using individual-level data can also be identified when individual-level data are averaged over.
-Here we use a longterm dataset to model population dynamics using individual-level and plot-level data.
+  Forecasts of populations under future climate change are constrained by
+  the spatial and temporal scales at which we can gather data. For plants,
+  ecologists often require individual level data to build population
+  models; data that is difficult and expensive to collect at spatial
+  scales greater than one square meter. Models built using such data are
+  difficult to scale-up to spatial scales relevant to land management
+  decisions and policy intervention. A possible route forward are
+  density-structured models based on coarse-scale census data, which are
+  much easier to collect over large spatial extents. However, to use
+  coarse-scale data in a climate change context requires the assumption
+  that the climate signals in population dynamics identified using
+  individual-level data can also be identified when individual-level data
+  are averaged over. Here we use a longterm dataset to model population
+  dynamics using individual-level and plot-level data.
   \end{abstract}
+   \begin{keyword} ecological forecasts \sep population models \sep integral projection model \sep climate change \sep grassland \sep \end{keyword}
+ \end{frontmatter}
 
-\noindent{}\textbf{Keywords:} population model, forecasting, integral projection model, etc.
+\section{Introduction}\label{introduction}
 
-\section{Introduction}
-Producing skillfull forecasts at spatial scales relevant to global environmental changes that play out at the landscape and regional scale has proven challenging for population ecologists.
-The major hurdle is that most population models, at least for plant species, are built using data from small, localized plots because parameterizing vital rate regressions requires tracking the fates of individuals.
-These models are difficult to scale up from the micro to meso-scales because the fitted parameters do not fully represent the spatial variation present at spatial scales beyond that at which the data are collected.
+Producing skillfull forecasts at spatial scales relevant to global
+environmental changes that play out at the landscape and regional scale
+has proven challenging for population ecologists. The major hurdle is
+that most population models, at least for plant species, are built using
+data from small, localized plots because parameterizing vital rate
+regressions requires tracking the fates of individuals. These models are
+difficult to scale up from the micro to meso-scales because the fitted
+parameters do not fully represent the spatial variation present at
+spatial scales beyond that at which the data are collected.
 
-Recently, Freckelton et al., building on work by Hastings and NAME, have proposed density-structured population models that focus on the transition of populations among discrete states, rather than the traditional approach of modeling the transitions of individuals.
-Such an approach could be extremely valuable because the data needed to parameterize density-structured population models is much easier, and less costly, to collect (Queenborough et al.
-year).
-For example, using a density-structured approach, one could build a population model using a time series of annual plot-based censuses of species percent cover.
-However, a major assumption of the density-structured approach is that the aggregate dynamics of the population observed at coarse spatial resolution faithfully represent, and correspond to, the fates of individual plants.
-In other words, using a density-structured approach requires a leap of faith that important covariates (e.g., climate variables) at the level of the individual are captured adequately at the population level.
-If we seek to forecast the impacts of climate change on plant populations, then clearly this assumption requires testing.
+Recently, Freckelton et al., building on work by Hastings and NAME, have
+proposed density-structured population models that focus on the
+transition of populations among discrete states, rather than the
+traditional approach of modeling the transitions of individuals. Such an
+approach could be extremely valuable because the data needed to
+parameterize density-structured population models is much easier, and
+less costly, to collect (Queenborough et al. year). For example, using a
+density-structured approach, one could build a population model using a
+time series of annual plot-based censuses of species percent cover.
+However, a major assumption of the density-structured approach is that
+the aggregate dynamics of the population observed at coarse spatial
+resolution faithfully represent, and correspond to, the fates of
+individual plants. In other words, using a density-structured approach
+requires a leap of faith that important covariates (e.g., climate
+variables) at the level of the individual are captured adequately at the
+population level. If we seek to forecast the impacts of climate change
+on plant populations, then clearly this assumption requires testing.
 
-\section{Materials and Methods}
-\subsection{Study site and data}
-Our demographic data comes from the Fort Keogh Livestock and Range Research Laboratory in eastern Montana's northern mixed prairie near Miles City, Montana, USA (46 deg.
-19' N, 105 deg 48' W).
-The dataset is freely available on Ecological Archives (CITE), and interested readers should refer to the metadata therin for a complete description.
-The site is about 800 m above sea level and mean annual precipitation (1878-2009) is 334 mm, with most annual precipitaion falling from April through September (76).
-The site is grass dominated and, for the purposes of our study, we focus on the four most abundant graminoid species: \emph{Bouteloua gracilis}, \emph{Hesperostipa comata}, \emph{Pascopyrum smithii}, and \emph{Poa secunda}.
+\section{Materials and Methods}\label{materials-and-methods}
 
-From 1932 to 1945 individual plants were identified and mapped annualy in 44 1-m2 quadrats using a pantograph.
-The quadrats were distributed in six pastures, each assigned a grazing treatment of light (1.24 ha/animal unit month), moderate (0.92 ha/aum), and heavy (0.76 ha/aum) stocking rates (two pastures per treatment).
-In this analysis we account for potential differences among the grazing treatments, but do not focus on grazing$\times$climate interactions.
-The annual maps of the quadrats were digitized and the fates of individual plants tracked and extracted using a computer program.
-Daily climate data, which we aggregated into climate variables of interest, are available for the duration of the data collection period (1932 - 1945) from the Miles City airport, Wiley Field, 9 km from the study site.
+\subsubsection{Study site and data}\label{study-site-and-data}
 
-\subsection{Stastical models of vital rates}
-The first step in building our single-species population models was to fit statistical models of vital rates at both levels of inference: individual-level and quadrat-level.
-Here we describe the general statistical approach at the individual-level, but the same approach applies at the quadrat-level.
-We modeled survival and growth as functions of genet size (quadrat cover in the quadrat-level models) and climate covariates (described in more detail below).
-We maintained a consistent random effects structure for both models that included three terms: (1) a random year effect on the intercept, (2) a random year effect on the coefficient for plant cover (either individual or quadrat level), and (3) a random effect of group (see Data set description) on the intercept.
-Vital rates for each species are modeled separately.
+Our demographic data comes from the Fort Keogh Livestock and Range
+Research Laboratory in eastern Montana's northern mixed prairie near
+Miles City, Montana, USA (46 deg. 19' N, 105 deg 48' W). The dataset is
+freely available on Ecological Archives (CITE), and interested readers
+should refer to the metadata therin for a complete description. The site
+is about 800 m above sea level and mean annual precipitation (1878-2009)
+is 334 mm, with most annual precipitaion falling from April through
+September (76). The site is grass dominated and, for the purposes of our
+study, we focus on the four most abundant graminoid species:
+\emph{Bouteloua gracilis}, \emph{Hesperostipa comata},
+\emph{Pascopyrum smithii}, and \emph{Poa secunda}.
+
+From 1932 to 1945 individual plants were identified and mapped annualy
+in 44 1-m2 quadrats using a pantograph. The quadrats were distributed in
+six pastures, each assigned a grazing treatment of light (1.24 ha/animal
+unit month), moderate (0.92 ha/aum), and heavy (0.76 ha/aum) stocking
+rates (two pastures per treatment). In this analysis we account for
+potential differences among the grazing treatments, but do not focus on
+grazing$\times$climate interactions. The annual maps of the quadrats
+were digitized and the fates of individual plants tracked and extracted
+using a computer program. Daily climate data, which we aggregated into
+climate variables of interest, are available for the duration of the
+data collection period (1932 - 1945) from the Miles City airport, Wiley
+Field, 9 km from the study site.
+
+\subsubsection{Stastical models of vital
+rates}\label{stastical-models-of-vital-rates}
+
+The first step in building our single-species population models was to
+fit statistical models of vital rates at both levels of inference:
+individual-level and quadrat-level. Here we describe the general
+statistical approach at the individual-level, but the same approach
+applies at the quadrat-level. We modeled survival and growth as
+functions of genet size (quadrat cover in the quadrat-level models) and
+climate covariates (described in more detail below). We maintained a
+consistent random effects structure for both models that included three
+terms: (1) a random year effect on the intercept, (2) a random year
+effect on the coefficient for plant cover (either individual or quadrat
+level), and (3) a random effect of group (see Data set description) on
+the intercept. Vital rates for each species are modeled separately.
 
@@
 
-\noindent where $x$ is the log of genet size (quadrat areal cover), $\beta_{0,t}$ is a year-specific intercept, $\beta_{s,t}$ is the year-specific slope parameter for size, $\beta_{Q}$ is the random effect of quadrat group location, $\theta$ is a matrix of $i$ climate effects over $t$ years, $\beta_{c,i}$ is the fixed parameter for the effect of the $i$th climate covariate, and $\varepsilon_{t}$ is the error term.
-At the quadrat level we did not fit random year effects for survival because those parameters did not reach convergence (see "Computing").
+where $x$ is the log of genet size (quadrat areal cover), $\beta_{0,t}$
+is a year-specific intercept, $\beta_{s,t}$ is the year-specific slope
+parameter for size, $\beta_{Q}$ is the random effect of quadrat group
+location, $\theta$ is a matrix of $i$ climate effects over $t$ years,
+$\beta_{c,i}$ is the fixed parameter for the effect of the $i$th climate
+covariate, and $\varepsilon_{t}$ is the error term. At the quadrat level
+we did not fit random year effects for survival because those parameters
+did not reach convergence (see ``Computing'').
 
-We modeled growth as gaussian process describing genet size (or quadrat cover) at time $t+1$ as a function of size at $t$ and climate covariates:
+We modeled growth as gaussian process describing genet size (or quadrat
+cover) at time $t+1$ as a function of size at $t$ and climate
+covariates:
 
@@
 
-\noindent where $x$ is genet size and all other paramters are as described for the survival regression.
-For the quadrat-level approach we modeled growth as a process describing proportional cover within a quadrat at time $t+1$ as a function of proportional cover at time $t$ and climate covariates.
-Thus, instead of a gaussian process, with a normal likelihood, we modeled growth at the quadrat level as above but with a beta likelihood and a logit link to the linear predictors.
+where $x$ is genet size and all other paramters are as described for the
+survival regression. For the quadrat-level approach we modeled growth as
+a process describing proportional cover within a quadrat at time $t+1$
+as a function of proportional cover at time $t$ and climate covariates.
+Thus, instead of a gaussian process, with a normal likelihood, we
+modeled growth at the quadrat level as above but with a beta likelihood
+and a logit link to the linear predictors.
 
-While our approach for modeling survival and growth is similar at the individual and quadrat levels, the addition of new genets at the inidividual level or occupied sites at the quadrat level requires two separate approaches.
+While our approach for modeling survival and growth is similar at the
+individual and quadrat levels, the addition of new genets at the
+inidividual level or occupied sites at the quadrat level requires two
+separate approaches.
 
-\subsection{Including climate covariates (and avoiding model selection)}
-For both the individual-level IPM (IPM) and the quadrat-based IBM (QBM) we included four climate covariates in each vital rate regression: fall through spring precipitation at \emph{t}-1 and \emph{t}-2 (ppt1 and ppt2, respectively) and mean spring temperature at \emph{t}-1 and \emph{t}-2 (TmeanSpr1 and TmeanSpr2, respectively), where \emph{t} is the observation year.
-We included climate covariates as additive effects.
-We did not consider interactions among climate covariates or among climate covariates and plant size or proportional cover.
-Climate effects ($\beta_{c}$s) were modeled hierarchically in each vital rate regression so that each species specific climate effect is drawn from an interspecific climate effect distribution.
-This was necessary to achieve convergence at the quadrat level and we used the same approach at the individual level for uniformity.
+\subsubsection{Including climate covariates (and avoiding model
+selection)}\label{including-climate-covariates-and-avoiding-model-selection}
 
-\subsection{Model fitting}
-All of our analyses (model fitting and simulating) were conducted in Program R.
-We used the MCMC sampler in JAGS to estimate the posterior distributions of model parameters.
-We obtained posterior distributions for all model parameters from three parallel MCMC chains run for 50,000 iterations, after a 50,000 iteration burn in.
-We assessed convergence visually and using the Gelman diagnostic in the R package `coda.' Scale reduction factors for all parameters were less than 1.02, indicating convergence.
-For the purposes of introducing stochasticity in our population models, we saved the final 1,000 iterations from each chain for all parameters to be used as randomly drawn values during population simulation.
+For both the individual-level IPM (IPM) and the quadrat-based IBM (QBM)
+we included four climate covariates in each vital rate regression: fall
+through spring precipitation at \emph{t}-1 and \emph{t}-2 (ppt1 and
+ppt2, respectively) and mean spring temperature at \emph{t}-1 and
+\emph{t}-2 (TmeanSpr1 and TmeanSpr2, respectively), where \emph{t} is
+the observation year. We included climate covariates as additive
+effects. We did not consider interactions among climate covariates or
+among climate covariates and plant size or proportional cover. Climate
+effects ($\beta_{c}$s) were modeled hierarchically in each vital rate
+regression so that each species specific climate effect is drawn from an
+interspecific climate effect distribution. This was necessary to achieve
+convergence at the quadrat level and we used the same approach at the
+individual level for uniformity.
 
-\subsection{Population models}
-With the posterior distribution of the vital rate statistical models in hand, it is straightforward to simulate the population models.
+\subsubsection{Model fitting}\label{model-fitting}
 
-\section{Results}
-We assessed the statistical importance of the climate covariates included the final vital rate regressions by comparing the residual deviance of models with climate covariates and temporal random effects, climate covariates only, and temporal random effects only.
-When a model includes climate covariates, this comparison shows the relative contribution of the climate covariates in explaining the total interannual variability (Adler et al.
-2012).
+All of our analyses (model fitting and simulating) were conducted in
+Program R. We used the MCMC sampler in JAGS to estimate the posterior
+distributions of model parameters. We obtained posterior distributions
+for all model parameters from three parallel MCMC chains run for 50,000
+iterations, after a 50,000 iteration burn in. We assessed convergence
+visually and using the Gelman diagnostic in the R package `coda.' Scale
+reduction factors for all parameters were less than 1.02, indicating
+convergence. For the purposes of introducing stochasticity in our
+population models, we saved the final 1,000 iterations from each chain
+for all parameters to be used as randomly drawn values during population
+simulation.
 
-\begin{figure}[th!]
-\begin{center}
-\includegraphics[ width=1\textwidth ]{Fig-DevFig}
-\end{center}
-\caption{The proportion of interannual variability in  vital rates explained by the climate covariates.
-The contribution for growth is defined as: (Climate model - Constant Model)/(Full model - Constant model).
-The contribution for survival and colonization, where we could not estimate a full model with year random effects at the quadrat level, is defined as: (Constant Model - Climate Model)/Constant Model.}
-\label{fig:DevFig}
-\end{figure}
+\subsubsection{Population models}\label{population-models}
 
-\begin{figure}[th!]
-\begin{center}
-\includegraphics[ width=0.5\textwidth] {Fig-CoefFig}
-\end{center}
-\caption{Posterior means (vertical ticks), 75\% credible intervals (heavy lines), and 95\% credible intervals (light lines) of climate effects on growth at both levels of inferences.
-The dashed vertical line is at 0, indicating no effect.
-Horizontal line at 0 indicates perfect agreement between mean observed cover in that year and the model predictions.}
-\label{fig:CoefFig}
-\end{figure}
+With the posterior distribution of the vital rate statistical models in
+hand, it is straightforward to simulate the population models.
 
-\begin{figure}[th!]
-\begin{center}
-\includegraphics{Fig-ResidFig}
-\end{center}
-\caption{Boxplots of model residuals for one-step-ahead forecasts at each observation year.
-Each one-step forecast was simulated 100 times.
-Note that the y-axes vary across panels.
-The light blue line shows the difference between the observed-year percent cover and the average cover observed across all years.
-The models tend to underpredict and perform poorly when observed cover in a given year is a large deviant from the mean.}
-\label{fig:ResidFig}
+\section{Results}\label{results}
+
+We assessed the statistical importance of the climate covariates
+included the final vital rate regressions by comparing the residual
+deviance of models with climate covariates and temporal random effects,
+climate covariates only, and temporal random effects only. When a model
+includes climate covariates, this comparison shows the relative
+contribution of the climate covariates in explaining the total
+interannual variability (Adler et al. 2012).
+
+\begin{figure}[htbp]
+\centering
+\includegraphics{components/figure/manuscript-figure_1.pdf}
+\caption{The proportion of interannual variability in vital rates
+explained by the climate covariates. The contribution for growth is
+defined as: (Climate model - Constant Model)/(Full model - Constant
+model). The contribution for survival and colonization, where we could
+not estimate a full model with year random effects at the quadrat level,
+is defined as: (Constant Model - Climate Model)/Constant Model.}
 \end{figure}

### 2014-12-11 manuscript/manuscript.Rmd
replaced ms file with markdown format

@@
 ```
+
+```{r figure_2, dependson="plot-options", fig.cap="Posterior means (vertical ticks), 75% credible intervals (heavy lines), and 95% credible intervals (light lines) of climate effects on growth at both levels of inferences.
+The dashed vertical line is at 0, indicating no effect.
+Horizontal line at 0 indicates perfect agreement between mean observed cover in that year and the model predictions.", fig.width=3, fig.height=7, cache=FALSE, results='hide'}
+gQC <- read.csv("../montana/quadBM/vitalRateRegressions/growth/growthQuants.csv")
+gQC <- gQC[129:144,]
+coefNames <- c("ppt1", "ppt2", "temp1", "temp2")
+gQC$X <- rep(coefNames, each=4)
+gQC$rate <- "growth"
+sQC <- read.csv("../montana/quadBM/vitalRateRegressions/survival/survivalQuants.csv")
+sQC <- sQC[33:48,]
+coefNames <- c("ppt1", "ppt2", "temp1", "temp2")
+sQC$X <- rep(coefNames, each=4)
+sQC$rate <- "survival"
+rQC <- read.csv("../montana/quadBM/vitalRateRegressions/colonization/colonizationQuants.csv")
+rQC <- rQC[29:44,]
+coefNames <- c("ppt1", "ppt2", "temp1", "temp2")
+rQC$X <- rep(coefNames, each=4)
+rQC$rate <- "colonization"
+plotQC <- rbind(gQC,sQC,rQC)
+gQ <- ggplot(plotQC, aes(x=species, y=X50., color=rate))+
+  geom_hline(aes(yintercept=0), linetype="dashed")+
+  geom_errorbar(aes(ymin=X2.5., ymax=X97.5.), width=0, position=position_dodge(width=0.5))+
+  geom_errorbar(aes(ymin=X25., ymax=X75.), width=0, size=1.25, position=position_dodge(width=0.5))+
+  geom_point(size=4, position=position_dodge(width=0.5), shape="|")+
+  ylab("Standardized Coefficient Value")+
+  xlab("Species")+
+  facet_grid(X~.)+
+  coord_flip()+
+  theme_bw()+
+  scale_color_manual(values=c("grey25", "steelblue", "slateblue"))+
+  guides(color=guide_legend(title=""))+
+  theme(legend.position="top")+
+  ggtitle("Quadrat level")
+
+gIC <- read.csv("../montana/ipm/vitalRateRegs/growth/growthQuants.csv")
+gIC <- gIC[133:148,]
+coefNames <- c("ppt1", "ppt2", "temp1", "temp2")
+gIC$X <- rep(coefNames, each=4)
+gIC$rate <- "growth"
+sIC <- read.csv("../montana/ipm/vitalRateRegs/survival/survivalQuants.csv")
+sIC <- sIC[133:148,]
+coefNames <- c("ppt1", "ppt2", "temp1", "temp2")
+sIC$X <- rep(coefNames, each=4)
+sIC$rate <- "survival"
+rIC <- read.csv("../montana/ipm/vitalRateRegs/recruitment/recruitmentQuants.csv")
+rIC <- rIC[81:96,]
+coefNames <- c("ppt1", "ppt2", "temp1", "temp2")
+rIC$X <- rep(coefNames, each=4)
+rIC$rate <- "colonization"
+plotIC <- rbind(gIC,sIC,rIC)
+gI <- ggplot(plotIC, aes(x=species, y=X50., color=rate))+
+  geom_hline(aes(yintercept=0), linetype="dashed")+
+  geom_errorbar(aes(ymin=X2.5., ymax=X97.5.), width=0, position=position_dodge(width=0.5))+
+  geom_errorbar(aes(ymin=X25., ymax=X75.), width=0, size=1.25, position=position_dodge(width=0.5))+
+  geom_point(size=4, position=position_dodge(width=0.5), shape="|")+
+  ylab("Standardized Coefficient Value")+
+  xlab("Species")+
+  facet_grid(X~.)+
+  coord_flip()+
+  theme_bw()+
+  scale_color_manual(values=c("grey25", "steelblue", "slateblue"),
+                     labels=c("recruitment", "growth", "survival"))+
+#   guides(color=FALSE)+
+  guides(color=guide_legend(title=""))+
+  theme(legend.position="top")+
+  ggtitle("Individual level")
+
+####
+#### Plot just for growth
+####
+growthIC <- subset(plotIC, rate=="growth")
+growthQC <- subset(plotQC, rate=="growth")
+growthIC$Level <- "Individual"
+growthQC$Level <- "Quadrat"
+growthAll <- rbind(growthIC, growthQC)
+
+gGrowth <- ggplot(growthAll, aes(x=species, y=X50., color=Level))+
+  geom_hline(aes(yintercept=0), linetype="dashed")+
+  geom_errorbar(aes(ymin=X2.5., ymax=X97.5.), width=0, position=position_dodge(width=0.5))+
+  geom_errorbar(aes(ymin=X25., ymax=X75.), width=0, size=1.25, position=position_dodge(width=0.5))+
+  geom_point(size=4, position=position_dodge(width=0.5), shape="|")+
+  ylab("Standardized Coefficient Value")+
+  xlab("Species")+
+  facet_grid(X~.)+
+  coord_flip()+
+  theme_bw()+
+  scale_color_manual(values=c("grey25", "slateblue"))+
+#   guides(color=FALSE)+
+  guides(color=guide_legend(title=""))+
+  theme(legend.position="top")
+print(gGrowth)
+```
+
+```{r figure_3, dependson="plot-options", fig.cap="Boxplots of model residuals for one-step-ahead forecasts at each observation year.
+Each one-step forecast was simulated ``` r nSims``` times.
+Note that the y-axes vary across panels.
+The light blue line shows the difference between the observed-year percent cover and the average cover observed across all years.
+The models tend to underpredict and perform poorly when observed cover in a given year is a large deviant from the mean.", fig.width=5, fig.height=6, cache=FALSE, results='hide'}
+library(plyr)
+library(reshape2)
+resD <- readRDS("../montana/quadBM/simulations/quadBM_oneStep_Residuals.rds")
+nSims <- length(unique(resD$sim))
+diffD <- ddply(resD, c("year", "species"), summarise,
+               avgDiff = mean(yearDiff))
+
+ggplot(resD)+
+  geom_hline(aes(yintercept=0))+
+#   geom_line(data=diffD, aes(x=year, y=avgDiff, group=species), color="purple", alpha=0.5)+
+  stat_smooth(data=diffD, aes(x=year, y=avgDiff, group=species), color="lightblue", se = FALSE, size=1)+
+  geom_boxplot(aes(x=year, y=covChangeResiduals), outlier.shape = 1, width=0.3, outlier.size = 0.5)+
+  facet_grid(species~., scales = "free_y")+
+  xlab("Year")+
+  ylab("Residuals (% cover difference)")+
+  theme_bw()+
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+```

### 2014-12-11 manuscript/manuscript.tex
replaced ms file with markdown format

@@
 
+\begin{figure}[htbp]
+\centering
+\includegraphics{components/figure/manuscript-figure_2.pdf}
+\caption{Posterior means (vertical ticks), 75\% credible intervals
+(heavy lines), and 95\% credible intervals (light lines) of climate
+effects on growth at both levels of inferences. The dashed vertical line
+is at 0, indicating no effect. Horizontal line at 0 indicates perfect
+agreement between mean observed cover in that year and the model
+predictions.}
+\end{figure}
+
+\begin{figure}[htbp]
+\centering
+\includegraphics{components/figure/manuscript-figure_3.pdf}
+\caption{Boxplots of model residuals for one-step-ahead forecasts at
+each observation year. Each one-step forecast was simulated
+\texttt{r nSims} times. Note that the y-axes vary across panels. The
+light blue line shows the difference between the observed-year percent
+cover and the average cover observed across all years. The models tend
+to underpredict and perform poorly when observed cover in a given year
+is a large deviant from the mean.}
+\end{figure}
+
 \end{document}

### 2014-12-12 montana/readme.Rmd
new readme for subdirectories

---
title: "READ ME"
author: "Andrew Tredennick"
date: "December 11, 2014"
output: html_document
---

These are the base folders for the Montana dataset analysis.

### Folder descriptions
* `ipm`: contains subdirectories for vital rate statistical models based on individual level demographic data and code to run single-species integral projection models.

* `quadBM`: contains subdirectories for vital rate statistical models based on plot level data and code to run the quadrat-based population model

* `speciesData`: contains subdirectories for each analysed species (the dominant species) and the average quadrat cover for each species. Species directories hold raw data on individual cover through time in each quadrat

* `weather`: contains the climate file with pre-aggregated variables

### 2014-12-12 manuscript/manuscript.Rmd
new readme for subdirectories
just trying to get refs to work

@@
 
-Recently, Freckelton et al., building on work by Hastings and NAME, have proposed density-structured population models that focus on the transition of populations among discrete states, rather than the traditional approach of modeling the transitions of individuals.
+Recently, @Freckleton2011, building on work by Hastings and NAME, have proposed density-structured population models that focus on the transition of populations among discrete states, rather than the traditional approach of modeling the transitions of individuals.
 Such an approach could be extremely valuable because the data needed to parameterize density-structured population models is much easier, and less costly, to collect (Queenborough et al.
@@
 ```{r figure_3, dependson="plot-options", fig.cap="Boxplots of model residuals for one-step-ahead forecasts at each observation year.
-Each one-step forecast was simulated ``` r nSims``` times.
+Each one-step forecast was simulated `r nSims` times.
 Note that the y-axes vary across panels.
@@
 ```
+
+References
+==========

### 2014-12-17 manuscript/manuscript.Rmd
ms ref updates

@@
 
-bibliography: components/references.bib
+bibliography: ~/Dropbox/Bibliography/MicroMeso.bib
 csl: components/ecology.csl
@@
 ============
-Producing skillfull forecasts at spatial scales relevant to global environmental changes that play out at the landscape and regional scale has proven challenging for population ecologists.
-The major hurdle is that most population models, at least for plant species, are built using data from small, localized plots because parameterizing vital rate regressions requires tracking the fates of individuals.
-These models are difficult to scale up from the micro to meso-scales because the fitted parameters do not fully represent the spatial variation present at spatial scales beyond that at which the data are collected.
+Population models are important tools for predicting the impacts of environmental change on species.
+But reconciling the scales at which population models are parameterized and the scales at which environmental changes play out remains a challenge [@Freckleton2011; @Queenborough2011].
+The major hurdle is that most population models, at least for plant species, are built using data from small, localized plots because parameterizing traditional population models requires tracking the fates of individuals.
+These models are difficult to scale up from the micro to meso-scales because the fitted parameters do not fully represent the spatial variation present at scales beyond that at which the data are collected [@Sæther2007].
+Thus, our ability to use population models to predict the consequences of climate change is limited when we rely on individual-level data.
 
-Recently, @Freckleton2011, building on work by Hastings and NAME, have proposed density-structured population models that focus on the transition of populations among discrete states, rather than the traditional approach of modeling the transitions of individuals.
-Such an approach could be extremely valuable because the data needed to parameterize density-structured population models is much easier, and less costly, to collect (Queenborough et al.
-year).
+In contrast to individual-level data, population-level data arising from large-scale census efforts
+
+Recently, @Freckleton2011, building on work by @Taylor2004, have proposed density-structured population models that focus on the transition of populations among discrete states, rather than the traditional approach of modeling the transitions of individuals.
+Such an approach could be extremely valuable because the data needed to parameterize density-structured population models is much easier, and less costly, to collect.
+
 For example, using a density-structured approach, one could build a population model using a time series of annual plot-based censuses of species percent cover.
@@
 We assessed the statistical importance of the climate covariates included the final vital rate regressions by comparing the residual deviance of models with climate covariates and temporal random effects, climate covariates only, and temporal random effects only.
-When a model includes climate covariates, this comparison shows the relative contribution of the climate covariates in explaining the total interannual variability (Adler et al.
-2012).
+When a model includes climate covariates, this comparison shows the relative contribution of the climate covariates in explaining the total interannual variability (@Adler2012).
 
@@
 
+----------
+
 References

### 2014-12-17 manuscript/manuscript.tex
new readme for subdirectories
ms ref updates

@@
 
-Producing skillfull forecasts at spatial scales relevant to global
-environmental changes that play out at the landscape and regional scale
-has proven challenging for population ecologists. The major hurdle is
-that most population models, at least for plant species, are built using
-data from small, localized plots because parameterizing vital rate
-regressions requires tracking the fates of individuals. These models are
+Population models are important tools for predicting the impacts of
+environmental change on species. But reconciling the scales at which
+population models are parameterized and the scales at which
+environmental changes play out remains a challenge (Freckleton et al.
+2011, Queenborough et al. 2011). The major hurdle is that most
+population models, at least for plant species, are built using data from
+small, localized plots because parameterizing traditional population
+models requires tracking the fates of individuals. These models are
 difficult to scale up from the micro to meso-scales because the fitted
-parameters do not fully represent the spatial variation present at
-spatial scales beyond that at which the data are collected.
+parameters do not fully represent the spatial variation present scales
+beyond that at which the data are collected (Sæther et al. (2007)).
+Thus, our ability to use population models to predict the consequences
+of climate change is limited.
 
-Recently, Freckelton et al., building on work by Hastings and NAME, have
-proposed density-structured population models that focus on the
-transition of populations among discrete states, rather than the
+Recently, Freckleton et al. (2011), building on work by Hastings and
+NAME, have proposed density-structured population models that focus on
+the transition of populations among discrete states, rather than the
 traditional approach of modeling the transitions of individuals. Such an
@@
 contribution of the climate covariates in explaining the total
-interannual variability (Adler et al. 2012).
+interannual variability (Adler et al. (2012)).
 
@@
 
+\begin{center}\rule{3in}{0.4pt}\end{center}
+
+\section{References}\label{references}
+
+Adler, P. B., H. J. Dalgleish, and S. P. Ellner. 2012. Forecasting plant
+community impacts of climate variability and change: when do competitive
+interactions matter? Journal of Ecology 100:478--487.
+
+Freckleton, R. P., W. J. Sutherland, A. R. Watkinson, and S. A.
+Queenborough. 2011. Density-structured models for plant population
+dynamics. American Naturalist 177:1--17.
+
+Queenborough, S. A., K. M. Burnet, W. J. Sutherland, A. R. Watkinson,
+and R. P. Freckleton. 2011. From meso- to macroscale population
+dynamics: A new density-structured approach. Methods in Ecology and
+Evolution 2:289--302.
+
+Sæther, B. E., S. Engen, V. Grøtan, W. Fiedler, E. Matthysen, M. E.
+Visser, J. Wright, A. P. Møller, F. Adriaensen, H. {Van Balen}, D.
+Balmer, M. C. Mainwaring, R. H. McCleery, M. Pampus, and W. Winkel.
+2007. The extended Moran effect and large-scale synchronous fluctuations
+in the size of great tit and blue tit populations. Journal of Animal
+Ecology 76:315--325.
+
 \end{document}

### 2014-12-19 manuscript/manuscript.Rmd
writing methods

@@
 
+In this paper, we model populations based on two levels of data: individual and quadrat.
+The individual data is the "raw" data.
+For the quadrat level we data we simply sum individual areal cover for each quadrat by species.
+This is equivalent to a perfect census of quadrat percent cover, so we do not need to consider measurement error.
+Based on these two datasets we can compare population models built using individual level data and aggregated quadrat level data.
+
 ### Stastical models of vital rates
-The first step in building our single-species population models was to fit statistical models of vital rates at both levels of inference: individual-level and quadrat-level.
-Here we describe the general statistical approach at the individual-level, but the same approach applies at the quadrat-level.
-We modeled survival and growth as functions of genet size (quadrat cover in the quadrat-level models) and climate covariates (described in more detail below).
-We maintained a consistent random effects structure for both models that included three terms: (1) a random year effect on the intercept, (2) a random year effect on the coefficient for plant cover (either individual or quadrat level), and (3) a random effect of group (see Data set description) on the intercept.
-Vital rates for each species are modeled separately.
+At both levels of inference (individual and quadrat), the building blocks of our population models are vital rate regressions.
+For individual level data we fit models for survival, growth, and recruitment of new individuals for each species.
+At the quadrat level we fit analagous models of extinction probability, percent cover increase/decrease, and quadrat colonization for each species.
+We describe the statistical models seprately since fitting the models required different approaches at the individual and quadrat levels.
+All models contain four climate covariate that we chose \emph{a priori}: fall through spring precipitation at \emph{t}-1 and \emph{t}-2 (ppt1 and ppt2, respectively) and mean spring temperature at \emph{t}-1 and \emph{t}-2 (TmeanSpr1 and TmeanSpr2, respectively), where \emph{t} is the observation year
 
-We used logistic regression to model survival probability:
+#### Vital rate models: individual level
+We used logistic regression to model survival probability ($S$) of genet $i$ from species $j$ in quadrat group $Q$ from time $t$ to $t+1$:
 
 \begin{equation}
-logit(s) = \beta_{0,t} + \beta_{s,t}x + \beta_{Q} + \beta_{c,1}\theta_{1,t}\\
-+ \cdots +  \beta_{c,i}\theta_{i,t} + \varepsilon_{t}
+\text{logit}(S_{ijQ,t}) = \gamma^{S}_{j,t} + \phi^{S}_{jQ} + \beta^{S}_{j,t}x_{ij,t} + \omega^{S}_{j}w_{ij,t} + \theta^{S}_{jk}C_{k,t} + \varepsilon^{S}_{t}
 \end{equation}
 
-where $x$ is the log of genet size (quadrat areal cover), $\beta_{0,t}$ is a year-specific intercept, $\beta_{s,t}$ is the year-specific slope parameter for size, $\beta_{Q}$ is the random effect of quadrat group location, $\theta$ is a matrix of $i$ climate effects over $t$ years, $\beta_{c,i}$ is the fixed parameter for the effect of the $i$th climate covariate, and $\varepsilon_{t}$ is the error term.
-At the quadrat level we did not fit random year effects for survival because those parameters did not reach convergence (see "Computing").
+where $x_{ij,t}$ is the log of genet size, $\gamma^{S}_{j,t}$ is a year-specific intercept, $\beta^{S}_{j,t}$ is the year-specific slope parameter for size, $\phi^{S}_{jQ}$ is the random effect of quadrat group location, and $\theta^{S}_{k}$ is the fixed parameter for the effect of the $k$th climate covariate at time $t$ ($C_{k,t}$).
+We include density-dependence by estimating the effect of crowding on the focal individual by other individuals of the same species.
+$\omega$ is the effect of crowding and $w_{t,Q}$ is the crowding experienced by the focal individual at time $t$ in quadrat $Q$.
 
-We modeled growth as gaussian process describing genet size (or quadrat cover) at time $t+1$ as a function of size at $t$ and climate covariates:
+We modeled growth as gaussian process describing genet size at time $t+1$ as a function of size at $t$ and climate covariates:
 
 \begin{equation}
-x_{t+1} = \beta_{0,t} + \beta_{s,t}x_{t} + \beta_{Q} + \beta_{c,1}\theta_{1,t}\\
-+ \cdots +  \beta_{c,i}\theta_{i,t} + \varepsilon_{t}
+x_{ijQ,t+1} = \gamma^{G}_{j,t} + \phi^{G}_{jQ} + \beta^{G}_{j,t}x_{ij,t} + \omega^{G}_{j}w_{ij,t} + \theta^{G}_{jk}C_{k,t}
 \end{equation}
 
-where $x$ is genet size and all other paramters are as described for the survival regression.
-For the quadrat-level approach we modeled growth as a process describing proportional cover within a quadrat at time $t+1$ as a function of proportional cover at time $t$ and climate covariates.
-Thus, instead of a gaussian process, with a normal likelihood, we modeled growth at the quadrat level as above but with a beta likelihood and a logit link to the linear predictors.
+where $x$ is log genet size and all other paramters are as described for the survival regression.
 
-While our approach for modeling survival and growth is similar at the individual and quadrat levels, the addition of new genets at the inidividual level or occupied sites at the quadrat level requires two separate approaches.
+Our data allows us to track new recruits, but we cannot assign a specific parent to new genets.
+So, for recruitment, we work at the quadrat level and model the number of new individuals of species $j$ in quadrat $q$ recruiting at time $t+1$ as a function of quadrat "effective cover" ($A'$) in the previous year ($t$).
+Effective cover is a mixture of observed cover ($A$) in the focal quadrat ($q$) and the mean cover across the entire group ($Q$) of quadrats in which $q$ is located:
 
-### Including climate covariates (and avoiding model selection)
-For both the individual-level IPM (IPM) and the quadrat-based IBM (QBM) we included four climate covariates in each vital rate regression: fall through spring precipitation at \emph{t}-1 and \emph{t}-2 (ppt1 and ppt2, respectively) and mean spring temperature at \emph{t}-1 and \emph{t}-2 (TmeanSpr1 and TmeanSpr2, respectively), where \emph{t} is the observation year.
-We included climate covariates as additive effects.
-We did not consider interactions among climate covariates or among climate covariates and plant size or proportional cover.
-Climate effects ($\beta_{c}$s) were modeled hierarchically in each vital rate regression so that each species specific climate effect is drawn from an interspecific climate effect distribution.
-This was necessary to achieve convergence at the quadrat level and we used the same approach at the individual level for uniformity.
+\begin{equation}
+A'_{jq,t} = p_{j}A_{jq,t} + (1-p_{j})\bar{A}_{jQ,t}
+\end{equation}
+
+where $p$ is a mixing fraction between 0 and 1 that is estimated within the model.
+
+We assume the number of individuals, $y$, recruiting at time $t+1$ follows a negative binomial distribution:
+
+\begin{equation}
+y_{jq,t+1} \sim \text{NegBin}(\lambda_{jq,t+1},\zeta)
+\end{equation}
+
+where $\lambda$ is the mean intensity and $\zeta$ is the size parameter. We define $\lambda$ as:
+
+\begin{equation}
+\lambda{jq,t+1} = A'_{jq,t}e^{(\gamma^{R}_{j,t} + \phi^{R}_{jQ} + \theta^{R}_{jk}C_{k,t} + \omega^{R}\sqrt{A'_{q,t}})}
+\end{equation}
+
+where $A'$ is effective cover ($\text{cm}^2$) of species $j$ in quadrat $q$ and all other terms are as in the survival and growth regressions.
+
+#### Vital rate models: quadrat level
+At the quadrat level we defined three vital rates:
+
+1. Probability of extirpation ($S$): the probability that, for a given species, a particular quadrat will go from non-zero cover at time $t$ to zero cover at time $t+1$.
+2. Cover change ($G$): the change in percent cover from time $t$ to $t+1$ for a given species within a particular quadrat.
+3. Probability of colonization ($R$): the probability that, for a give species, a particular quadrat will go from zero cover at time $t$ to non-zero cover at time $t+1$.
+
+We retain the abbreviations $S$, $G$, and $R$ from the analagous processes at the individual level.
+The vital rate models at the quadrat level are all based on quadrat proportional cover.
+Also, at the quadrat level we do not need to explicitly include a density dependent term.
+Since we are modeling proportional cover, we essentially get density-dependence for "free" when proportional cover is included as a covariate.
+That is, density-dependence emerges because quadrats with low cover gerenally increase in cover whereas quadrats with high cover generally decrease in cover.
+
+We modeled the probability of extirpation of species $j$ in quadrat $q$ from time $t$ to $t+1$ as:
+
+\begin{equation}
+\text{logit}(S_{jq,t}) = \gamma^{S}_{j} + \phi^{S}_{jQ} + \beta^{S}_{j}x_{jq,t} + \theta^{S}_{jk}C_{k,t}
+\end{equation}
+
+where all parameters are as in Eq.
+1 except that $x$ is now quadrat proportional cover.
+Note, however, that we do not include year random effects on the intercept $\gamma$ or the slope term $\beta$ for quadrat proportional cover.
+The quadrat data is inherently more sparse than the individual data from which it is aggregated, and this is especially evident when modeling rare events like extirpation and colonization.
+Thus, when we tried to fit random year effects, those terms did not converge.
+
+We modeled quadrat cover change ($G$) from time $t$ to $t+1$ as:
+
+\begin{equation}
+\text{logit}(x_{jq,t+1}) = \gamma^{G}_{j,t} + \phi^{G}_{jQ} + \beta^{G}_{j,t}x_{jq,t} + \theta^{S}_{jk}C_{k,t}
+\end{equation}
+
+where, in this case, we do include random year effects on the intercept $\gamma$ and the slope term $\beta$.
+For cover change we had enough data for those terms to converge.
+Note that our model for quadrat cover change uses a logit transformation to link the expected cover at $t+1$ ($x_{jq,t}$) to the linear predictors.
+We do so because during model fitting we use a beta likelihood since the data, proportional cover, is beta distributed.
+The beta likelihood requires shape ($\rho$) and rate ($\eta$) parameters that can be caclulated using moment-matching:
+
+\begin{align}
+\rho &= x_{jq,t+1}\tau_{j} \\
+\eta &= (1-x_{jq,t+1})\tau_{j}
+\end{align}
 
 ### Model fitting
-All of our analyses (model fitting and simulating) were conducted in Program R.
-We used the MCMC sampler in JAGS to estimate the posterior distributions of model parameters.
-We obtained posterior distributions for all model parameters from three parallel MCMC chains run for 50,000 iterations, after a 50,000 iteration burn in.
-We assessed convergence visually and using the Gelman diagnostic in the R package `coda.' Scale reduction factors for all parameters were less than 1.02, indicating convergence.
+All of our analyses (model fitting and simulating) were conducted in R [@Team2013].
+We used the MCMC sampler in JAGS [@Plummer2003] to estimate the posterior distributions of model parameters and the package 'r2jags' [@Su2012] to connect R to JAGS.
+We obtained posterior distributions for all model parameters from three parallel MCMC chains run for 50,000 iterations, after discarding an initial 50,000 iterations.
+We assessed convergence visually and using the @Gelman1992 diagnostic in the R package `coda' [@Plummer2006].
+Scale reduction factors for all parameters were less than 1.02, indicating convergence.
 For the purposes of introducing stochasticity in our population models, we saved the final 1,000 iterations from each chain for all parameters to be used as randomly drawn values during population simulation.
@@
 
-----------
-
 References

### 2014-12-19 manuscript/manuscript.tex
writing methods

@@
 difficult to scale up from the micro to meso-scales because the fitted
-parameters do not fully represent the spatial variation present scales
-beyond that at which the data are collected (Sæther et al. (2007)).
+parameters do not fully represent the spatial variation present at
+scales beyond that at which the data are collected (Sæther et al. 2007).
 Thus, our ability to use population models to predict the consequences
-of climate change is limited.
+of climate change is limited when we rely on individual-level data.
 
-Recently, Freckleton et al. (2011), building on work by Hastings and
-NAME, have proposed density-structured population models that focus on
-the transition of populations among discrete states, rather than the
-traditional approach of modeling the transitions of individuals. Such an
-approach could be extremely valuable because the data needed to
-parameterize density-structured population models is much easier, and
-less costly, to collect (Queenborough et al. year). For example, using a
-density-structured approach, one could build a population model using a
-time series of annual plot-based censuses of species percent cover.
-However, a major assumption of the density-structured approach is that
-the aggregate dynamics of the population observed at coarse spatial
-resolution faithfully represent, and correspond to, the fates of
-individual plants. In other words, using a density-structured approach
-requires a leap of faith that important covariates (e.g., climate
-variables) at the level of the individual are captured adequately at the
-population level. If we seek to forecast the impacts of climate change
-on plant populations, then clearly this assumption requires testing.
+In contrast to individual-level data, population-level data arising from
+large-scale census efforts
+
+Recently, Freckleton et al. (2011), building on work by Taylor and
+Hastings (2004), have proposed density-structured population models that
+focus on the transition of populations among discrete states, rather
+than the traditional approach of modeling the transitions of
+individuals. Such an approach could be extremely valuable because the
+data needed to parameterize density-structured population models is much
+easier, and less costly, to collect.
+
+For example, using a density-structured approach, one could build a
+population model using a time series of annual plot-based censuses of
+species percent cover. However, a major assumption of the
+density-structured approach is that the aggregate dynamics of the
+population observed at coarse spatial resolution faithfully represent,
+and correspond to, the fates of individual plants. In other words, using
+a density-structured approach requires a leap of faith that important
+covariates (e.g., climate variables) at the level of the individual are
+captured adequately at the population level. If we seek to forecast the
+impacts of climate change on plant populations, then clearly this
+assumption requires testing.
 
@@
 
+In this paper, we model populations based on two levels of data:
+individual and quadrat. The individual data is the ``raw'' data. For the
+quadrat level we data we simply sum individual areal cover for each
+quadrat by species. This is equivalent to a perfect census of quadrat
+percent cover, so we do not need to consider measurement error. Based on
+these two datasets we can compare population models built using
+individual level data and aggregated quadrat level data.
+
 \subsubsection{Stastical models of vital
@@
 
-The first step in building our single-species population models was to
-fit statistical models of vital rates at both levels of inference:
-individual-level and quadrat-level. Here we describe the general
-statistical approach at the individual-level, but the same approach
-applies at the quadrat-level. We modeled survival and growth as
-functions of genet size (quadrat cover in the quadrat-level models) and
-climate covariates (described in more detail below). We maintained a
-consistent random effects structure for both models that included three
-terms: (1) a random year effect on the intercept, (2) a random year
-effect on the coefficient for plant cover (either individual or quadrat
-level), and (3) a random effect of group (see Data set description) on
-the intercept. Vital rates for each species are modeled separately.
+At both levels of inference (individual and quadrat), the building
+blocks of our population models are vital rate regressions. For
+individual level data we fit models for survival, growth, and
+recruitment of new individuals for each species. At the quadrat level we
+fit analagous models of extinction probability, percent cover
+increase/decrease, and quadrat colonization for each species. We
+describe the statistical models seprately since fitting the models
+required different approaches at the individual and quadrat levels. All
+models contain four climate covariate that we chose \emph{a priori}:
+fall through spring precipitation at \emph{t}-1 and \emph{t}-2 (ppt1 and
+ppt2, respectively) and mean spring temperature at \emph{t}-1 and
+\emph{t}-2 (TmeanSpr1 and TmeanSpr2, respectively), where \emph{t} is
+the observation year
 
-We used logistic regression to model survival probability:
+\paragraph{Vital rate models: individual
+level}\label{vital-rate-models-individual-level}
+
+We used logistic regression to model survival probability ($S$) of genet
+$i$ from species $j$ in quadrat group $Q$ from time $t$ to $t+1$:
 
 \begin{equation}
-logit(s) = \beta_{0,t} + \beta_{s,t}x + \beta_{Q} + \beta_{c,1}\theta_{1,t}\\
-+ \cdots +  \beta_{c,i}\theta_{i,t} + \varepsilon_{t}
+\text{logit}(S_{ijQ,t}) = \gamma^{S}_{j,t} + \phi^{S}_{jQ} + \beta^{S}_{j,t}x_{ij,t} + \omega^{S}_{j}w_{ij,t} + \theta^{S}_{jk}C_{k,t} + \varepsilon^{S}_{t}
 \end{equation}
 
-where $x$ is the log of genet size (quadrat areal cover), $\beta_{0,t}$
-is a year-specific intercept, $\beta_{s,t}$ is the year-specific slope
-parameter for size, $\beta_{Q}$ is the random effect of quadrat group
-location, $\theta$ is a matrix of $i$ climate effects over $t$ years,
-$\beta_{c,i}$ is the fixed parameter for the effect of the $i$th climate
-covariate, and $\varepsilon_{t}$ is the error term. At the quadrat level
-we did not fit random year effects for survival because those parameters
-did not reach convergence (see ``Computing'').
+where $x_{ij,t}$ is the log of genet size, $\gamma^{S}_{j,t}$ is a
+year-specific intercept, $\beta^{S}_{j,t}$ is the year-specific slope
+parameter for size, $\phi^{S}_{jQ}$ is the random effect of quadrat
+group location, and $\theta^{S}_{k}$ is the fixed parameter for the
+effect of the $k$th climate covariate at time $t$ ($C_{k,t}$). We
+include density-dependence by estimating the effect of crowding on the
+focal individual by other individuals of the same species. $\omega$ is
+the effect of crowding and $w_{t,Q}$ is the crowding experienced by the
+focal individual at time $t$ in quadrat $Q$.
 
-We modeled growth as gaussian process describing genet size (or quadrat
-cover) at time $t+1$ as a function of size at $t$ and climate
-covariates:
+We modeled growth as gaussian process describing genet size at time
+$t+1$ as a function of size at $t$ and climate covariates:
 
 \begin{equation}
-x_{t+1} = \beta_{0,t} + \beta_{s,t}x_{t} + \beta_{Q} + \beta_{c,1}\theta_{1,t}\\
-+ \cdots +  \beta_{c,i}\theta_{i,t} + \varepsilon_{t}
+x_{ijQ,t+1} = \gamma^{G}_{j,t} + \phi^{G}_{jQ} + \beta^{G}_{j,t}x_{ij,t} + \omega^{G}_{j}w_{ij,t} + \theta^{G}_{jk}C_{k,t}
 \end{equation}
 
-where $x$ is genet size and all other paramters are as described for the
-survival regression. For the quadrat-level approach we modeled growth as
-a process describing proportional cover within a quadrat at time $t+1$
-as a function of proportional cover at time $t$ and climate covariates.
-Thus, instead of a gaussian process, with a normal likelihood, we
-modeled growth at the quadrat level as above but with a beta likelihood
-and a logit link to the linear predictors.
+where $x$ is log genet size and all other paramters are as described for
+the survival regression.
 
-While our approach for modeling survival and growth is similar at the
-individual and quadrat levels, the addition of new genets at the
-inidividual level or occupied sites at the quadrat level requires two
-separate approaches.
+Our data allows us to track new recruits, but we cannot assign a
+specific parent to new genets. So, for recruitment, we work at the
+quadrat level and model the number of new individuals of species $j$ in
+quadrat $q$ recruiting at time $t+1$ as a function of quadrat
+``effective cover'' ($A'$) in the previous year ($t$). Effective cover
+is a mixture of observed cover ($A$) in the focal quadrat ($q$) and the
+mean cover across the entire group ($Q$) of quadrats in which $q$ is
+located:
 
-\subsubsection{Including climate covariates (and avoiding model
-selection)}\label{including-climate-covariates-and-avoiding-model-selection}
+\begin{equation}
+A'_{jq,t} = p_{j}A_{jq,t} + (1-p_{j})\bar{A}_{jQ,t}
+\end{equation}
 
-For both the individual-level IPM (IPM) and the quadrat-based IBM (QBM)
-we included four climate covariates in each vital rate regression: fall
-through spring precipitation at \emph{t}-1 and \emph{t}-2 (ppt1 and
-ppt2, respectively) and mean spring temperature at \emph{t}-1 and
-\emph{t}-2 (TmeanSpr1 and TmeanSpr2, respectively), where \emph{t} is
-the observation year. We included climate covariates as additive
-effects. We did not consider interactions among climate covariates or
-among climate covariates and plant size or proportional cover. Climate
-effects ($\beta_{c}$s) were modeled hierarchically in each vital rate
-regression so that each species specific climate effect is drawn from an
-interspecific climate effect distribution. This was necessary to achieve
-convergence at the quadrat level and we used the same approach at the
-individual level for uniformity.
+where $p$ is a mixing fraction between 0 and 1 that is estimated within
+the model.
+
+We assume the number of individuals, $y$, recruiting at time $t+1$
+follows a negative binomial distribution:
+
+\begin{equation}
+y_{jq,t+1} \sim \text{NegBin}(\lambda_{jq,t+1},\zeta)
+\end{equation}
+
+where $\lambda$ is the mean intensity and $\zeta$ is the size parameter.
+We define $\lambda$ as:
+
+\begin{equation}
+\lambda{jq,t+1} = A'_{jq,t}e^{(\gamma^{R}_{j,t} + \phi^{R}_{jQ} + \theta^{R}_{jk}C_{k,t} + \omega^{R}\sqrt{A'_{q,t}})}
+\end{equation}
+
+where $A'$ is effective cover ($\text{cm}^2$) of species $j$ in quadrat
+$q$ and all other terms are as in the survival and growth regressions.
+
+\paragraph{Vital rate models: quadrat
+level}\label{vital-rate-models-quadrat-level}
+
+At the quadrat level we defined three vital rates:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\itemsep1pt\parskip0pt\parsep0pt
+\item
+  Probability of extirpation ($S$): the probability that, for a given
+  species, a particular quadrat will go from non-zero cover at time $t$
+  to zero cover at time $t+1$.
+\item
+  Cover change ($G$): the change in percent cover from time $t$ to $t+1$
+  for a given species within a particular quadrat.
+\item
+  Probability of colonization ($R$): the probability that, for a give
+  species, a particular quadrat will go from zero cover at time $t$ to
+  non-zero cover at time $t+1$.
+\end{enumerate}
+
+We retain the abbreviations $S$, $G$, and $R$ from the analagous
+processes at the individual level. The vital rate models at the quadrat
+level are all based on quadrat proportional cover. Also, at the quadrat
+level we do not need to explicitly include a density dependent term.
+Since we are modeling proportional cover, we essentially get
+density-dependence for ``free'' when proportional cover is included as a
+covariate. That is, density-dependence emerges because quadrats with low
+cover gerenally increase in cover whereas quadrats with high cover
+generally decrease in cover.
+
+We modeled the probability of extirpation of species $j$ in quadrat $q$
+from time $t$ to $t+1$ as:
+
+\begin{equation}
+\text{logit}(S_{jq,t}) = \gamma^{S}_{j} + \phi^{S}_{jQ} + \beta^{S}_{j}x_{jq,t} + \theta^{S}_{jk}C_{k,t}
+\end{equation}
+
+where all parameters are as in Eq. 1 except that $x$ is now quadrat
+proportional cover. Note, however, that we do not include year random
+effects on the intercept $\gamma$ or the slope term $\beta$ for quadrat
+proportional cover. The quadrat data is inherently more sparse than the
+individual data from which it is aggregated, and this is especially
+evident when modeling rare events like extirpation and colonization.
+Thus, when we tried to fit random year effects, those terms did not
+converge.
+
+We modeled quadrat cover change ($G$) from time $t$ to $t+1$ as:
+
+\begin{equation}
+\text{logit}(x_{jq,t+1}) = \gamma^{G}_{j,t} + \phi^{G}_{jQ} + \beta^{G}_{j,t}x_{jq,t} + \theta^{S}_{jk}C_{k,t}
+\end{equation}
+
+where, in this case, we do include random year effects on the intercept
+$\gamma$ and the slope term $\beta$. For cover change we had enough data
+for those terms to converge. Note that our model for quadrat cover
+change uses a logit transformation to link the expected cover at $t+1$
+($x_{jq,t}$) to the linear predictors. We do so because during model
+fitting we use a beta likelihood since the data, proportional cover, is
+beta distributed. The beta likelihood requires shape ($\rho$) and rate
+($\eta$) parameters that can be caclulated using moment-matching:
+
+\begin{align}
+\rho &= x_{jq,t+1}\tau_{j} \\
+\eta &= (1-x_{jq,t+1})\tau_{j}
+\end{align}
 
@@
 
-All of our analyses (model fitting and simulating) were conducted in
-Program R. We used the MCMC sampler in JAGS to estimate the posterior
-distributions of model parameters. We obtained posterior distributions
-for all model parameters from three parallel MCMC chains run for 50,000
-iterations, after a 50,000 iteration burn in. We assessed convergence
-visually and using the Gelman diagnostic in the R package `coda.' Scale
-reduction factors for all parameters were less than 1.02, indicating
-convergence. For the purposes of introducing stochasticity in our
-population models, we saved the final 1,000 iterations from each chain
-for all parameters to be used as randomly drawn values during population
-simulation.
+All of our analyses (model fitting and simulating) were conducted in R
+(Team 2013). We used the MCMC sampler in JAGS (Plummer 2003) to estimate
+the posterior distributions of model parameters and the package `r2jags'
+(Su and Yajima 2012) to connect R to JAGS. We obtained posterior
+distributions for all model parameters from three parallel MCMC chains
+run for 50,000 iterations, after discarding an initial 50,000
+iterations. We assessed convergence visually and using the Gelman and
+Rubin (1992) diagnostic in the R package `coda' (Plummer et al. 2006).
+Scale reduction factors for all parameters were less than 1.02,
+indicating convergence. For the purposes of introducing stochasticity in
+our population models, we saved the final 1,000 iterations from each
+chain for all parameters to be used as randomly drawn values during
+population simulation.
 
@@
 
-\begin{center}\rule{3in}{0.4pt}\end{center}
-
 \section{References}\label{references}
@@
 
+Gelman, A., and D. B. Rubin. 1992. Inference from Iterative Simulation
+Using Multiple Sequences.
+
+Plummer, M. 2003. JAGS: A Program for Analysis of Bayesian Graphical
+Models Using Gibbs Sampling. Pages 20--22 \emph{in} Proceedings of the
+3rd international workshop on distributed statistical computing (dSC
+2003). march.
+
+Plummer, M., N. Best, K. Cowles, and K. Vines. 2006. CODA: Convergence
+Diagnosis and Output Analysis for MCMC. R News 6:7--11.
+
 Queenborough, S. A., K. M. Burnet, W. J. Sutherland, A. R. Watkinson,
@@
 Evolution 2:289--302.
+
+Su, Y., and M. Yajima. 2012. R2jags: A Package for Running jags from R.
+http://CRAN. R-project. org/package= R2jags.
 
@@
 
+Taylor, C. M., and A. Hastings. 2004. Finding optimal control strategies
+for invasive species: a density-structured model for Spartina
+alterniflora. Journal of Applied Ecology 41:1049--1057.
+
+Team, R. 2013. R Development Core Team. R: A Language and Environment
+for Statistical Computing.
+
 \end{document}

### 2014-12-21 manuscript/manuscript.Rmd
updated ms with equations

@@
 We describe the statistical models seprately since fitting the models required different approaches at the individual and quadrat levels.
-All models contain four climate covariate that we chose \emph{a priori}: fall through spring precipitation at \emph{t}-1 and \emph{t}-2 (ppt1 and ppt2, respectively) and mean spring temperature at \emph{t}-1 and \emph{t}-2 (TmeanSpr1 and TmeanSpr2, respectively), where \emph{t} is the observation year
+All models contain four climate covariate that we chose \emph{a priori}: fall through spring precipitation at \emph{t}-1 and \emph{t}-2 (ppt1 and ppt2, respectively) and mean spring temperature at \emph{t}-1 and \emph{t}-2 (TmeanSpr1 and TmeanSpr2, respectively), where \emph{t} is the observation year.
+
+We fit all models using a hierarchical Bayesian approach, which we describe in more detail below.
+However, for each vital rate statistical model we also define the likelihood model we use.
+For the likelihood models, Y is always the relevant vector of observations (e.g., whether a  genet survived [1] or not [0] from year $t$ to $t+1$).
 
@@
 
-\begin{equation}
-\text{logit}(S_{ijQ,t}) = \gamma^{S}_{j,t} + \phi^{S}_{jQ} + \beta^{S}_{j,t}x_{ij,t} + \omega^{S}_{j}w_{ij,t} + \theta^{S}_{jk}C_{k,t} + \varepsilon^{S}_{t}
-\end{equation}
+\begin{align}
+\text{logit}(S_{ijQ,t}) &= \gamma^{S}_{j,t} + \phi^{S}_{jQ} + \beta^{S}_{j,t}x_{ij,t} + \omega^{S}_{j}w_{ij,t} + \theta^{S}_{jk}C_{k,t} + \varepsilon^{S}_{t} \\
+Y^{S}_{ijQ,t} &\sim \text{Bernoulli}(S_{ijQ,t})
+\end{align}
 
@@
 We include density-dependence by estimating the effect of crowding on the focal individual by other individuals of the same species.
-$\omega$ is the effect of crowding and $w_{t,Q}$ is the crowding experienced by the focal individual at time $t$ in quadrat $Q$.
+$\omega$ is the effect of crowding and $w_{t,Q}$ is the crowding experienced by the focal individual at time $t$ in quadrat group $Q$.
 
@@
 
-\begin{equation}
-x_{ijQ,t+1} = \gamma^{G}_{j,t} + \phi^{G}_{jQ} + \beta^{G}_{j,t}x_{ij,t} + \omega^{G}_{j}w_{ij,t} + \theta^{G}_{jk}C_{k,t}
-\end{equation}
+\begin{align}
+x_{ijQ,t+1} &= \gamma^{G}_{j,t} + \phi^{G}_{jQ} + \beta^{G}_{j,t}x_{ij,t} + \omega^{G}_{j}w_{ij,t} + \theta^{G}_{jk}C_{k,t} \\
+Y^{G}_{ijQ,t} &\sim \text{Normal}(x_{ijQ,t+1}, \sigma_{j})
+\end{align}
 
@@
 So, for recruitment, we work at the quadrat level and model the number of new individuals of species $j$ in quadrat $q$ recruiting at time $t+1$ as a function of quadrat "effective cover" ($A'$) in the previous year ($t$).
-Effective cover is a mixture of observed cover ($A$) in the focal quadrat ($q$) and the mean cover across the entire group ($Q$) of quadrats in which $q$ is located:
+Effective cover is a mixture of observed cover ($A$) in the focal quadrat ($q$) and the mean cover across the entire group ($\bar{A}$) of $Q$ quadrats in which $q$ is located:
 
@@
 
-We assume the number of individuals, $y$, recruiting at time $t+1$ follows a negative binomial distribution:
+We assume the number of individuals, $Y^{R}$, recruiting at time $t+1$ follows a negative binomial distribution:
 
 \begin{equation}
-y_{jq,t+1} \sim \text{NegBin}(\lambda_{jq,t+1},\zeta)
+Y^{R}_{jq,t+1} \sim \text{NegBin}(\lambda_{jq,t+1},\zeta)
 \end{equation}
@@
 \begin{equation}
-\lambda{jq,t+1} = A'_{jq,t}e^{(\gamma^{R}_{j,t} + \phi^{R}_{jQ} + \theta^{R}_{jk}C_{k,t} + \omega^{R}\sqrt{A'_{q,t}})}
+\lambda_{jq,t+1} = A'_{jq,t}e^{(\gamma^{R}_{j,t} + \phi^{R}_{jQ} + \theta^{R}_{jk}C_{k,t} + \omega^{R}\sqrt{A'_{q,t}})}
 \end{equation}
@@
 
-\begin{equation}
-\text{logit}(S_{jq,t}) = \gamma^{S}_{j} + \phi^{S}_{jQ} + \beta^{S}_{j}x_{jq,t} + \theta^{S}_{jk}C_{k,t}
-\end{equation}
+\begin{align}
+\text{logit}(S_{jq,t}) &= \gamma^{S}_{j} + \phi^{S}_{jQ} + \beta^{S}_{j}x_{jq,t} + \theta^{S}_{jk}C_{k,t} \\
+Y^{S}_{jq,t} &\sim \text{Bernoulli}(S_{jq,t})
+\end{align}
 
@@
 \begin{align}
-\rho &= x_{jq,t+1}\tau_{j} \\
-\eta &= (1-x_{jq,t+1})\tau_{j}
+\rho_{jq,t+1} &= x_{jq,t+1}\tau_{j} \\
+\eta_{jq,t+1} &= (1-x_{jq,t+1})\tau_{j}
+\end{align}
+
+with likelihood:
+
+\begin{equation}
+Y^{G}_{jq,t+1} \sim \text{Beta}(\rho_{jq,t+1}, \eta_{jq,t+1}).
+\end{equation}
+
+Finally, we modeled probability of colonization in quadrat $q$ by species $j$ from time $t$ to $t+1$ as:
+
+\begin{align}
+\text{logit}(R_{jq,t}) &= \gamma^{R}_{j} + \phi^{R}_{jQ} + \theta^{R}_{jk}C_{k,t} \\
+Y^{R}_{jq,t} &\sim \text{Bernoulli}(R_{jq,t}).
 \end{align}
@@
 ### Model fitting
+Our Bayesian approach to fitting the vital rate models required choosing appropriate priors for unknown parameters and deciding which, if any, of those prior should be hierarchical.
+We decided to fit models where all terms except climate covariates were fit by species, while the climate covariates were fit heierarchically where species-specific coefficients were drawn from a shared `global' coefficient distribution.
+We did so for two reasons: (1) the four focal species are all perrennial grasses that we expect to respond similarly to climate covariates, and (2) convergence of climate effects at the quadrat level was much easier to achieve when we modeled these terms hierarchically, allowing them to "share" statistical strenght via partial pooling [@Gelman2007].
+So, climate effects were modeled as:
+
+\begin{align}
+\theta_{jk} &\sim \text{Normal}(\bar{\theta_{k}}, \sigma_{k}) \\
+\bar{\theta_{k}} &\sim \text{Normal}(0, 1e-6)
+\end{align}
+
+We used uninformative priors for all unknown parameters, specifically:
+
+\begin{align}
+\boldsymbol{\gamma, \beta} &\sim \text{Normal}(0, 1e-6) \\
+\boldsymbol{\phi} &\sim \text{Normal}(0, \boldsymbol{\sigma_{\phi}}) \\
+\boldsymbol{\sigma_{\phi}} &\sim e(\text{Gamma}(2, 0.5)) \\
+\boldsymbol{\sigma_{\theta}, \sigma_{\gamma}, \sigma_{\beta}} &\sim \text{Gamma}(0.001, 0.001)
+\end{align}
+
 All of our analyses (model fitting and simulating) were conducted in R [@Team2013].
@@
 We assessed the statistical importance of the climate covariates included the final vital rate regressions by comparing the residual deviance of models with climate covariates and temporal random effects, climate covariates only, and temporal random effects only.
-When a model includes climate covariates, this comparison shows the relative contribution of the climate covariates in explaining the total interannual variability (@Adler2012).
+When a model includes climate covariates, this comparison shows the relative contribution of the climate covariates in explaining the total interannual variability [@Adler2012].
 

### 2014-12-21 manuscript/manuscript.tex
updated ms with equations

@@
 \emph{t}-2 (TmeanSpr1 and TmeanSpr2, respectively), where \emph{t} is
-the observation year
+the observation year.
+
+We fit all models using a hierarchical Bayesian approach, which we
+describe in more detail below. However, for each vital rate statistical
+model we also define the likelihood model we use. For the likelihood
+models, Y is always the relevant vector of observations (e.g., whether a
+genet survived {[}1{]} or not {[}0{]} from year $t$ to $t+1$).
 
@@
 the effect of crowding and $w_{t,Q}$ is the crowding experienced by the
-focal individual at time $t$ in quadrat $Q$.
+focal individual at time $t$ in quadrat group $Q$.
 
@@
 is a mixture of observed cover ($A$) in the focal quadrat ($q$) and the
-mean cover across the entire group ($Q$) of quadrats in which $q$ is
-located:
+mean cover across the entire group ($\bar{A}$) of $Q$ quadrats in which
+$q$ is located:
 
@@
 
-We assume the number of individuals, $y$, recruiting at time $t+1$
+We assume the number of individuals, $Y^{R}$, recruiting at time $t+1$
 follows a negative binomial distribution:
@@
 \begin{equation}
-y_{jq,t+1} \sim \text{NegBin}(\lambda_{jq,t+1},\zeta)
+Y^{R}_{jq,t+1} \sim \text{NegBin}(\lambda_{jq,t+1},\zeta)
 \end{equation}
@@
 \begin{align}
-\rho &= x_{jq,t+1}\tau_{j} \\
-\eta &= (1-x_{jq,t+1})\tau_{j}
+\rho_{jq,t+1} &= x_{jq,t+1}\tau_{j} \\
+\eta_{jq,t+1} &= (1-x_{jq,t+1})\tau_{j}
+\end{align}
+
+with likelihood:
+
+\begin{equation}
+Y^{G}_{jq,t+1} \sim \text{Beta}(\rho_{jq,t+1}, \eta_{jq,t+1}).
+\end{equation}
+
+Finally, we modeled probability of colonization in quadrat $q$ by
+species $j$ from time $t$ to $t+1$ as:
+
+\begin{align}
+\text{logit}(R_{jq,t}) &= \gamma^{R}_{j} + \phi^{R}_{jQ} + \theta^{R}_{jk}C_{k,t} \\
+Y^{R}_{jq,t} &\sim \text{Bernoulli}(R_{jq,t}).
 \end{align}
@@
 \subsubsection{Model fitting}\label{model-fitting}
+
+Our Bayesian approach to fitting the vital rate models required choosing
+appropriate priors for unknown parameters and deciding which, if any, of
+those prior should be hierarchical. We decided to fit models where all
+terms except climate covariates were fit by species, while the climate
+covariates were fit heierarchically where species-specific coefficients
+were drawn from a shared `global' coefficient distribution. We did so
+for two reasons: (1) the four focal species are all perrennial grasses
+that we expect to respond similarly to climate covariates, and (2)
+convergence of climate effects at the quadrat level was much easier to
+achieve when we modeled these terms hierarchically, allowing them to
+``share'' statistical strenght via partial pooling (Gelman and Hill
+2007). So, climate effects were modeled as:
+
+\begin{align}
+\theta_{jk} &\sim \text{Normal}(\bar{\theta_{k}}, \sigma_{k}) \\
+\bar{\theta_{k}} &\sim \text{Normal}(0, 1e-6)
+\end{align}
+
+We used uninformative priors for all unknown parameters, specifically:
+
+\begin{align}
+\boldsymbol{\gamma, \beta} &\sim \text{Normal}(0, 1e-6) \\
+\boldsymbol{\phi} &\sim \text{Normal}(0, \boldsymbol{\sigma_{\phi}}) \\
+\boldsymbol{\sigma_{\phi}} &\sim e(\text{Gamma}(2, 0.5)) \\
+\boldsymbol{\sigma_{\theta}, \sigma_{\gamma}, \sigma_{\beta}} &\sim \text{Gamma}(0.001, 0.001)
+\end{align}
 
@@
 contribution of the climate covariates in explaining the total
-interannual variability (Adler et al. (2012)).
+interannual variability (Adler et al. 2012).
 
@@
 
+Gelman, A., and J. Hill. 2007. Data analysis using regression and
+multilevel/hierarchical models. Page xxii, 625p.
+
 Gelman, A., and D. B. Rubin. 1992. Inference from Iterative Simulation

### 2014-12-21 manuscript/manuscript.Rmd
updated ms formatting

@@
 ---
-layout: 12pt,review
-linenumbers: true
-title: "Do we need detailed demographic data to forecast plant population responses to climate change?"
-author: 
-  - name: Andrew T. Tredennick
-    affiliation: usu
-    email: atredenn(at)gmail.com
-    footnote: Corresponding author
-  - name: Peter B. Adler
-    affiliation: usu
-address: 
-  - code: usu
-    address: | 
-      Department of Wildland Resources and the Ecology Center, 
-      Utah State University,
-      Logan, UT 84322, USA
-abstract: |
-  Forecasts of populations under future climate change are constrained by the spatial and temporal scales at which we can gather data.
-For plants, ecologists often require individual level data to build population models; data that is difficult and expensive to collect at spatial scales greater than one square meter.
-Models built using such data are difficult to scale-up to spatial scales relevant to land management decisions and policy intervention.
-A possible route forward are density-structured models based on coarse-scale census data, which are much easier to collect over large spatial extents.
-However, to use coarse-scale data in a climate change context requires the assumption that the climate signals in population dynamics identified using individual-level data can also be identified when individual-level data are averaged over.
-Here we use a longterm dataset to model population dynamics using individual-level and plot-level data.
-
-keywords:
-  - ecological forecasts
-  - population models
-  - integral projection model
-  - climate change
-  - grassland
-
+layout: 12pt
+header-includes:
+   - \usepackage{lineno}
+   - \linenumbers
+   - \usepackage{setspace}
+   - \doublespacing
 bibliography: ~/Dropbox/Bibliography/MicroMeso.bib
 csl: components/ecology.csl
-documentclass: components/elsarticle
 
@@
     fig_caption: true
-    template: components/elsarticle.latex
     keep_tex: true
+fontsize: 12pt
+geometry: margin=1in
 
 ---
+
+Do we need detailed demographic data to forecast population responses to climate change?
+========================================================================================
+
+### Andrew T. Tredennick and Peter B. Adler
+
+*Andrew T. Tredennick ([atredenn@gmail.com](mailto:atredenn@gmail.com)), Department of Wildland Resources and the Ecology Center, Utah State University, Logan, UT*
+
+*Peter B. Adler, Department of Wildland Resources and the Ecology Center, Utah State University, Logan, UT*
 
@@
 
+Abstract
+--------
+
 Introduction
-============
+------------
 Population models are important tools for predicting the impacts of environmental change on species.
@@
 Materials and Methods
-=====================
+---------------------
 ### Study site and data
@@
 19' N, 105 deg 48' W).
-The dataset is freely available on Ecological Archives (CITE), and interested readers should refer to the metadata therin for a complete description.
+The dataset is freely available on Ecological Archives [@Anderson2011], and interested readers should refer to the metadata therin for a complete description.
 The site is about 800 m above sea level and mean annual precipitation (1878-2009) is 334 mm, with most annual precipitaion falling from April through September (76).
-The site is grass dominated and, for the purposes of our study, we focus on the four most abundant graminoid species: \emph{Bouteloua gracilis}, \emph{Hesperostipa comata}, \emph{Pascopyrum smithii}, and \emph{Poa secunda}.
+The site is grass dominated and, for the purposes of our study, we focus on the four most abundant graminoid species: \emph{Bouteloua gracilis} (BOGR), \emph{Hesperostipa comata} (HECO), \emph{Pascopyrum smithii} (PASM), and \emph{Poa secunda} (POSE).
 
@@
 Results
-=======
+-------
 We assessed the statistical importance of the climate covariates included the final vital rate regressions by comparing the residual deviance of models with climate covariates and temporal random effects, climate covariates only, and temporal random effects only.
@@
 References
-==========
+----------

### 2014-12-21 manuscript/manuscript.tex
updated ms formatting

@@
-\documentclass[author-year, 12pt,review]{components/elsarticle} %review=doublespace preprint=single 5p=2 column
-%%% Begin My package additions %%%%%%%%%%%%%%%%%%%
-\usepackage[hyphens]{url}
-\usepackage{lineno} % add 
-  \linenumbers % turns line numbering on 
-\bibliographystyle{elsarticle-harv}
-\biboptions{sort&compress} % For natbib
-\usepackage{graphicx}
-\usepackage{booktabs} % book-quality tables
-%% Redefines the elsarticle footer
-\makeatletter
-\def\ps@pprintTitle{%
- \let\@oddhead\@empty
- \let\@evenhead\@empty
- \def\@oddfoot{\it \hfill\today}%
- \let\@evenfoot\@oddfoot}
-\makeatother
-
-% A modified page layout
-\textwidth 6.75in
-\oddsidemargin -0.15in
-\evensidemargin -0.15in
-\textheight 9in
-\topmargin -0.5in
-%%%%%%%%%%%%%%%% end my additions to header
-
+\documentclass[12pt,]{article}
 \usepackage[T1]{fontenc}
@@
 \else % if luatex or xelatex
+  \ifxetex
+    \usepackage{mathspec}
+    \usepackage{xltxtra,xunicode}
+  \else
     \usepackage{fontspec}
-  \ifxetex
-    \usepackage{xltxtra,xunicode}
   \fi
@@
 \IfFileExists{microtype.sty}{\usepackage{microtype}}{}
+\usepackage[margin=1in]{geometry}
 \usepackage{graphicx}
-% We will generate all images so they have a width \maxwidth. This means
-% that they will get their normal width if they fit onto the page, but
+% Redefine \includegraphics so that, unless explicit options are
+% given, the image width will not exceed the width of the page.
+% Images get their normal width if they fit onto the page, but
 % are scaled down if they would overflow the margins.
 \makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth
-\else\Gin@nat@width\fi}
+\def\ScaleIfNeeded{%
+  \ifdim\Gin@nat@width>\linewidth
+    \linewidth
+  \else
+    \Gin@nat@width
+  \fi
+}
 \makeatother
 \let\Oldincludegraphics\includegraphics
-\renewcommand{\includegraphics}[1]{\Oldincludegraphics[width=\maxwidth]{#1}}
+{%
+ \catcode`\@=11\relax%
+ \gdef\includegraphics{\@ifnextchar[{\Oldincludegraphics}{\Oldincludegraphics[width=\ScaleIfNeeded]}}%
+}%
 \ifxetex
@@
             pdfauthor={},
-            pdftitle={Do we need detailed demographic data to forecast plant population responses to climate change?},
+            pdftitle={},
             colorlinks=true,
+            citecolor=blue,
             urlcolor=blue,
@@
 \setcounter{secnumdepth}{0}
-% Pandoc toggle for numbering sections (defaults to be off)
-\setcounter{secnumdepth}{0}
-% Pandoc header
+
+\author{}
+\date{}
+\usepackage{lineno}
+\linenumbers
+\usepackage{setspace}
+\doublespacing
 
 \begin{document}
-\begin{frontmatter}
 
-  \title{Do we need detailed demographic data to forecast plant population
-responses to climate change?}
-    \author[usu]{Andrew T. Tredennick\corref{c1}}
-   \ead{atredenn(at)gmail.com} 
-   \cortext[c1]{Corresponding author}
-    \author[usu]{Peter B. Adler}
+\normalsize
 
-      \address[usu]{Department of Wildland Resources and the Ecology Center, Utah State
-University, Logan, UT 84322, USA}
+\section{Do we need detailed demographic data to forecast population
+responses to climate
+change?}\label{do-we-need-detailed-demographic-data-to-forecast-population-responses-to-climate-change}
 
-  \begin{abstract}
-  Forecasts of populations under future climate change are constrained by
-  the spatial and temporal scales at which we can gather data. For plants,
-  ecologists often require individual level data to build population
-  models; data that is difficult and expensive to collect at spatial
-  scales greater than one square meter. Models built using such data are
-  difficult to scale-up to spatial scales relevant to land management
-  decisions and policy intervention. A possible route forward are
-  density-structured models based on coarse-scale census data, which are
-  much easier to collect over large spatial extents. However, to use
-  coarse-scale data in a climate change context requires the assumption
-  that the climate signals in population dynamics identified using
-  individual-level data can also be identified when individual-level data
-  are averaged over. Here we use a longterm dataset to model population
-  dynamics using individual-level and plot-level data.
-  \end{abstract}
-   \begin{keyword} ecological forecasts \sep population models \sep integral projection model \sep climate change \sep grassland \sep \end{keyword}
- \end{frontmatter}
+\subsubsection{Andrew T. Tredennick and Peter B.
+Adler}\label{andrew-t.-tredennick-and-peter-b.-adler}
 
-\section{Introduction}\label{introduction}
+\emph{Andrew T. Tredennick
+(\href{mailto:atredenn@gmail.com}{\href{mailto:atredenn@gmail.com}{atredenn@gmail.com}}),
+Department of Wildland Resources and the Ecology Center, Utah State
+University, Logan, UT}
+
+\emph{Peter B. Adler, Department of Wildland Resources and the Ecology
+Center, Utah State University, Logan, UT}
+
+\subsection{Abstract}\label{abstract}
+
+\subsection{Introduction}\label{introduction}
 
@@
 
-\section{Materials and Methods}\label{materials-and-methods}
+\subsection{Materials and Methods}\label{materials-and-methods}
 
@@
 Miles City, Montana, USA (46 deg. 19' N, 105 deg 48' W). The dataset is
-freely available on Ecological Archives (CITE), and interested readers
-should refer to the metadata therin for a complete description. The site
-is about 800 m above sea level and mean annual precipitation (1878-2009)
-is 334 mm, with most annual precipitaion falling from April through
-September (76). The site is grass dominated and, for the purposes of our
-study, we focus on the four most abundant graminoid species:
-\emph{Bouteloua gracilis}, \emph{Hesperostipa comata},
-\emph{Pascopyrum smithii}, and \emph{Poa secunda}.
+freely available on Ecological Archives (Anderson et al. 2011), and
+interested readers should refer to the metadata therin for a complete
+description. The site is about 800 m above sea level and mean annual
+precipitation (1878-2009) is 334 mm, with most annual precipitaion
+falling from April through September (76). The site is grass dominated
+and, for the purposes of our study, we focus on the four most abundant
+graminoid species: \emph{Bouteloua gracilis} (BOGR),
+\emph{Hesperostipa comata} (HECO), \emph{Pascopyrum smithii} (PASM), and
+\emph{Poa secunda} (POSE).
 
@@
 
-\section{Results}\label{results}
+\subsection{Results}\label{results}
 
@@
 
-\section{References}\label{references}
+\subsection{References}\label{references}
 
@@
 interactions matter? Journal of Ecology 100:478--487.
+
+Anderson, J., L. Vermeire, and P. B. Adler. 2011. Fourteen years of
+mapped, permanent quadrats in a northern mixed prairie, USA. Ecology
+92:1703.
 

### 2014-12-23 montana/quadBM/vitalRateRegressions/growth/growthAllSpp_JAGS_ClimateTrack.R
added climate tracking script
Added a climate tracking script, a la Jim Clark. Probably won’t use it
for this project, but could be interesting to look at all species at
all sites and examine climate tracking…

@@ -1,0 +1,50 @@
+model{
+  #likelihood and process model
+  for(i in 1:nObs){
+    logit(gMu[i]) <- intYr[spp[i],yrs[i]] + intG[spp[i],grp[i]] + beta[spp[i],yrs[i]]*X[i] + temp1[spp[i]]*TmeanSpr1[i] + temp2[spp[i]]*TmeanSpr2[i] + rain1[spp[i]]*ppt1[i] + rain2[spp[i]]*ppt2[i]  
+    p[i] <- gMu[i] * tau[spp[i]]
+    q[i] <- (1 - gMu[i]) * tau[spp[i]]
+    C[i] ~ dbeta(p[i], q[i])
+  }
+  
+  #priors
+  for(s in 1:nSpp){
+    betaSpp[s] ~ dnorm(0, 1e-6)
+    temp1[s] ~ dnorm(temp1Mu, temp1Var)
+    temp2[s] ~ dnorm(temp2Mu, temp2Var)
+    rain1[s] ~ dnorm(rain1Mu, rain1Var)
+    rain2[s] ~ dnorm(rain2Mu, rain2Var)
+    intercept[s] ~ dnorm(0, 1e-6)
+    intVaryY[s] ~ dgamma(0.001, 0.001)
+    betaVar[s] ~ dgamma(0.001, 0.001)
+    t0[s] ~ dnorm(0, .01) 
+    tau[s] <- exp(t0[s])
+    intVarG[s] ~ dgamma(2, 0.5) 
+    for(y in 1:nYrs){
+      intYr[s,y] ~ dnorm(intercept[s], intVaryY[s])
+      beta[s,y] ~ dnorm(betaSpp[s], betaVar[s])
+    }
+    for(g in 1:nGrp){
+      intG[s,g] ~ dnorm(0, intVarG[s])
+    }
+  }
+  
+  temp1Mu ~ dnorm(0,1e-6)
+  temp2Mu ~ dnorm(0,1e-6)
+  rain1Mu ~ dnorm(1,1e-6)
+  rain2Mu ~ dnorm(0,1e-6)
+  temp1Var ~ dgamma(0.001, 0.001)
+  temp2Var ~ dgamma(0.001, 0.001)
+  rain1Var ~ dgamma(0.001, 0.001)
+  rain2Var ~ dgamma(0.001, 0.001)
+  
+  #Climate tracking derived quantities
+  for(i in 1:nObs){
[…]

### 2014-12-23 montana/quadBM/vitalRateRegressions/growth/growthClimTrack_MCMC.R

@@ -1,0 +1,121 @@
+#!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!#
+# Set working directory to location of this source file #
+#!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!#
+
+#clear everything, just to be safe 
+rm(list=ls(all=TRUE))
+
+#load libraries
+library(rjags)
+library(coda)
+library(ggplot2)
+library(plyr)
+library(reshape2)
+library(ggthemes)
+load.module("dic")
+
+#bring in data
+allD <- read.csv("../../../speciesData/quadAllCover.csv")
+allD <- allD[,2:ncol(allD)] #get rid of X ID column
+sppList <- as.character(unique(allD$Species))
+
+climD <- read.csv("../../../weather/Climate.csv")
+climD[3:6] <- scale(climD[3:6], center = TRUE, scale=TRUE)
+
+backD <- data.frame(climYear=NA,
+                    quad = NA,
+                    year= NA,
+                    totCover= NA,
+                    Species= NA,
+                    propCover= NA,
+                    lag.cover= NA,
+                    pptLag= NA,
+                    ppt1= NA,
+                    TmeanSpr1= NA,
+                    ppt2= NA,
+                    TmeanSpr2= NA,
+                    TmeanSum1= NA,
+                    TmeanSum2= NA,
+                    yearID= NA,
+                    group = NA,
+                    percCover = NA,
+                    percLagCover = NA)
+
+#loop through species and remake data frame
+for(spp in 1:length(sppList)){
+  doSpp <- sppList[spp]
+  sppD <- subset(allD, Species==doSpp)
+  
+  # create lag cover variable
+  tmp=sppD[,c("quad","year","totCover")]
+  tmp$year=tmp$year+1
+  names(tmp)[3]="lag.cover"
+  sppD=merge(sppD,tmp,all.x=T)
+  
+  # merge in climate data
+  sppD$climYear=sppD$year+1900-1  
+  sppD=merge(sppD,climD,by.x="climYear",by.y="year")
+  
+  #Growth observations
[…]

### 2014-12-29 manuscript/manuscript.Rmd
added climate tracking script
Added a climate tracking script, a la Jim Clark. Probably won’t use it
for this project, but could be interesting to look at all species at
all sites and examine climate tracking…
updated priors
new analysis update

@@
 
+*Keywords*: forecasting, climate change, grassland, integral projection model
+
 Introduction
@@
 The dataset is freely available on Ecological Archives [@Anderson2011], and interested readers should refer to the metadata therin for a complete description.
-The site is about 800 m above sea level and mean annual precipitation (1878-2009) is 334 mm, with most annual precipitaion falling from April through September (76).
+The site is about 800 m above sea level and mean annual precipitation (1878-2009) is 334 mm, with most annual precipitaion falling from April through September.
 The site is grass dominated and, for the purposes of our study, we focus on the four most abundant graminoid species: \emph{Bouteloua gracilis} (BOGR), \emph{Hesperostipa comata} (HECO), \emph{Pascopyrum smithii} (PASM), and \emph{Poa secunda} (POSE).
 
-From 1932 to 1945 individual plants were identified and mapped annualy in 44 1-m2 quadrats using a pantograph.
+From 1932 to 1945 individual plants were identified and mapped annualy in 44 1-$\text{m}^2$ quadrats using a pantograph.
 The quadrats were distributed in six pastures, each assigned a grazing treatment of light (1.24 ha/animal unit month), moderate (0.92 ha/aum), and heavy (0.76 ha/aum) stocking rates (two pastures per treatment).
@@
 \begin{align}
-\theta_{jk} &\sim \text{Normal}(\bar{\theta_{k}}, \sigma_{k}) \\
-\bar{\theta_{k}} &\sim \text{Normal}(0, 1e-6)
+\theta_{jk} &\sim \text{Normal}(\bar{\theta_{k}}, \sigma_{k})
 \end{align}
+
+where $\bar{\theta_{k}}$ is the interspecific effect of the $k$th climate covariate.
 
@@
 \begin{align}
-\boldsymbol{\gamma, \beta} &\sim \text{Normal}(0, 1e-6) \\
+\boldsymbol{\gamma, \beta, \bar{\theta}} &\sim \text{Normal}(0, 1e^{-6}) \\
 \boldsymbol{\phi} &\sim \text{Normal}(0, \boldsymbol{\sigma_{\phi}}) \\
-\boldsymbol{\sigma_{\phi}} &\sim e(\text{Gamma}(2, 0.5)) \\
-\boldsymbol{\sigma_{\theta}, \sigma_{\gamma}, \sigma_{\beta}} &\sim \text{Gamma}(0.001, 0.001)
+\boldsymbol{\sigma_{\phi}} &\sim e^{(\text{Gamma}(2, 0.5))} \\
+\boldsymbol{\sigma_{\theta}, \sigma_{\gamma}, \sigma_{\beta}, \tau, \zeta} &\sim \text{Gamma}(0.001, 0.001)
 \end{align}
@@
 
+We assessed the statistical importance of the climate covariates included the final vital rate regressions by comparing the residual deviance of models with climate covariates and temporal random effects, climate covariates only, and temporal random effects only.
+When a model includes climate covariates, this comparison shows the relative contribution of the climate covariates in explaining the total interannual variability [@Adler2012].
+
 ### Population models
 With the posterior distribution of the vital rate statistical models in hand, it is straightforward to simulate the population models.
+We used an Integral Projection Model (IPM) to model populations based on individual level data and an quadrat based version of an individually-based model (Quadrat-Based Model, QBM) to model populations based on quadrat level data.
+Both models take the general form:
+
+\begin{equation}
+N_{t+1} = S \times G + R.
+\end{equation}
+
+So, at each time step in a simulation, we use the survival regression to determine if each genet lives or not (if each quadrat remains occupied or not), the growth regression to determine size changes of surviving individuals (cover change of occupied quadrats), and the recruitment regression to determine the number of new recruits (if a quadrat is colonized or not).
+We first use one-step-ahead forecasts to assess each model's ability to reproduce observed cover changes.
+Then we use the models to analyze the effect of potential climate changes on population size over long time scales.
+
+We used random draws from the final 1,000 iterations from each of 3 MCMC chains to introduce stochasticity into our population models.
+At each time step, we first randomly selected climate covariates from one of the 14 observed years and also randomly drew one set of random year effects for each vital rate regression.
+Then, we drew the full parameter set (specific random year effects, climate effects, fixed effects) from a randomly selected MCMC iteration.
+We selected random year effects and the climate year independently.
 
@@
 -------
-We assessed the statistical importance of the climate covariates included the final vital rate regressions by comparing the residual deviance of models with climate covariates and temporal random effects, climate covariates only, and temporal random effects only.
-When a model includes climate covariates, this comparison shows the relative contribution of the climate covariates in explaining the total interannual variability [@Adler2012].
+### Climate effects on vital rates
+At the individual level we fit vital rate models that included climate covariates and random year effects (full model), just climate covariates (climate model), and no climate or random year effects (constant model).
+We used the metric from @Adler2012 to quantify the contribution of climate covariates to explaining interannual variability in vital rates based on comparisons of model deviance: (Climate model - Constant Model)/(Full model - Constant model).
+At the quadrat level, we could only fit random year effects for the growth regression.
+This is, in fact, one of the limitation of census-based data: local extirpation and colonization events are likely to be rare, making it difficult to fit complex statistical models.
+So, in this section we focus on growth only.
+
+Climate covariates improved the growth regressions at both levels, with a slghty greater contribution at the individual level (Figure 1).
+Parameter estimates were generally consistent in terms of sign and magnitude between the individual level and the quadrat level for each species and climate covariate (Figure 2).
+There is no clear pattern suggesting that parameters estimated using individual level data are more certain, as evidenced by credible interval widths, than estimates based on quadrat data.
+
+### Climate effects on population dynamics
+
+Discussion
+----------
+We sought to test the assumption that the sensitivities of plant populations to climate variables can be detected equally well using either individual level data or population level data.
+This is an important question to answer because population models are key tools for predicting the consequences of global climate change.
+However, they can be of limited use when built on data from a small subset of a population in space or time.
+If population level data (i.e., some aggregated form of individual level data) can be used to detect climate effects on population dynamics, then we would have a cheaper and easier option for data collection over relatively large temporal and spatial extents [_e.g._ @Freckleton2011].
+
+1. Climate detections
+2. Limitations:
+  * Sample size at quadrat level
+  * are these species inherently sensitive to climate (compare to Chu and Adler in review -- say why we chose Montana site -- most sensitive to climate variables)
+  * we know recruitment is important, and this is something that may be easily missed at quad level
+3. Implications
+
+References
+----------
+
+Figures
+-------
 
@@
   xlab("Species")+
-  facet_grid(X~.)+
+  facet_wrap(X~.)+
   coord_flip()+
@@
 
-References
-----------
+```{r figure_4, dependson="plot-options", fig.cap="Simulated changes species' median cover caused by a 1% increase in observed precipitation or temperature.
+We used the median of simulated results because the average was highly influence by some extreme outliers.", fig.width=4.5, fig.height=4, cache=FALSE, results='hide'}
+climBase <- readRDS("../montana/quadBM/simulations/climateChangeBase.rds")
+climPpt <- readRDS("../montana/quadBM/simulations/climateChangePrecip.rds")
+climTemp <- readRDS("../montana/quadBM/simulations/climateChangeTemperature.rds")
+climBase$sim <- rep("Observed", nrow(climBase)) 
+climPpt$sim <- rep("Precipitation", nrow(climBase)) 
+climTemp$sim <- rep("Temperature", nrow(climBase)) 
+climSimsQBM <- rbind(climBase, climPpt, climTemp)
+# ggplot(climSimsQBM)+
+#   geom_boxplot(aes(x=species, y=cover*100, fill=sim))+
+#   scale_y_continuous(limits=c(0,10))
+climSimsQBMavg <- ddply(climSimsQBM, .(sim, species), summarise,
+                        avgCover = median(cover))
+climSimQBMcast <- dcast(climSimsQBMavg, formula = species~sim)
+climSimQBMcast$diffP <- with(climSimQBMcast, (Precipitation*100)-(Observed*100)) 
+climSimQBMcast$diffT <- with(climSimQBMcast, (Temperature*100)-(Observed*100))
+climSimQBMplot <- melt(climSimQBMcast)
+ggplot(data=subset(climSimQBMplot, variable=="diffP"|variable=="diffT"))+
+  geom_bar(aes(x=species, y=value, fill=variable), stat="identity", position="dodge")+
+  scale_fill_manual(name="", labels=c("1% Precipitation Increase", "1% Temperature Increase"), values=c("grey25", "grey50"))+
+  theme_bw()+
+  theme(legend.position=c(0.6, 0.8))+
+  ylab("Cover Change (%)")+
+  xlab("Species")
+```

### 2014-12-29 manuscript/manuscript.tex
added climate tracking script
Added a climate tracking script, a la Jim Clark. Probably won’t use it
for this project, but could be interesting to look at all species at
all sites and examine climate tracking…
updated priors
new analysis update

@@
 
+\emph{Keywords}: forecasting, climate change, grassland, integral
+projection model
+
 \subsection{Introduction}\label{introduction}
@@
 precipitation (1878-2009) is 334 mm, with most annual precipitaion
-falling from April through September (76). The site is grass dominated
-and, for the purposes of our study, we focus on the four most abundant
+falling from April through September. The site is grass dominated and,
+for the purposes of our study, we focus on the four most abundant
 graminoid species: \emph{Bouteloua gracilis} (BOGR),
@@
 From 1932 to 1945 individual plants were identified and mapped annualy
-in 44 1-m2 quadrats using a pantograph. The quadrats were distributed in
-six pastures, each assigned a grazing treatment of light (1.24 ha/animal
-unit month), moderate (0.92 ha/aum), and heavy (0.76 ha/aum) stocking
-rates (two pastures per treatment). In this analysis we account for
-potential differences among the grazing treatments, but do not focus on
-grazing$\times$climate interactions. The annual maps of the quadrats
-were digitized and the fates of individual plants tracked and extracted
-using a computer program. Daily climate data, which we aggregated into
-climate variables of interest, are available for the duration of the
-data collection period (1932 - 1945) from the Miles City airport, Wiley
-Field, 9 km from the study site.
+in 44 1-$\text{m}^2$ quadrats using a pantograph. The quadrats were
+distributed in six pastures, each assigned a grazing treatment of light
+(1.24 ha/animal unit month), moderate (0.92 ha/aum), and heavy (0.76
+ha/aum) stocking rates (two pastures per treatment). In this analysis we
+account for potential differences among the grazing treatments, but do
+not focus on grazing$\times$climate interactions. The annual maps of the
+quadrats were digitized and the fates of individual plants tracked and
+extracted using a computer program. Daily climate data, which we
+aggregated into climate variables of interest, are available for the
+duration of the data collection period (1932 - 1945) from the Miles City
+airport, Wiley Field, 9 km from the study site.
 
@@
 \begin{align}
-\theta_{jk} &\sim \text{Normal}(\bar{\theta_{k}}, \sigma_{k}) \\
-\bar{\theta_{k}} &\sim \text{Normal}(0, 1e-6)
+\theta_{jk} &\sim \text{Normal}(\bar{\theta_{k}}, \sigma_{k})
 \end{align}
+
+where $\bar{\theta_{k}}$ is the interspecific effect of the $k$th
+climate covariate.
 
@@
 
-\subsubsection{Population models}\label{population-models}
-
-With the posterior distribution of the vital rate statistical models in
-hand, it is straightforward to simulate the population models.
-
-\subsection{Results}\label{results}
-
 We assessed the statistical importance of the climate covariates
@@
 interannual variability (Adler et al. 2012).
+
+\subsubsection{Population models}\label{population-models}
+
+With the posterior distribution of the vital rate statistical models in
+hand, it is straightforward to simulate the population models. We used
+an Integral Projection Model (IPM) to model populations based on
+individual level data and an quadrat based version of an
+individually-based model (Quadrat-Based Model, QBM) to model populations
+based on quadrat level data. Both models take the general form:
+
+\begin{equation}
+N_{t+1} = S \times G + R.
+\end{equation}
+
+So, at each time step in a simulation, we use the survival regression to
+determine if each genet lives or not (if each quadrat remains occupied
+or not), the growth regression to determine size changes of surviving
+individuals (cover change of occupied quadrats), and the recruitment
+regression to determine the number of new recruits (if a quadrat is
+colonized or not). We first use one-step-ahead forecasts to assess each
+model's ability to reproduce observed cover changes. Then we use the
+models to analyze the effect of potential climate changes on population
+size over long time scales.
+
+We used random draws from the final 1,000 iterations from each of 3 MCMC
+chains to introduce stochasticity into our population models. At each
+time step, we first randomly selected climate covariates from one of the
+14 observed years and also randomly drew one set of random year effects
+for each vital rate regression. Then, we drew the full parameter set
+(specific random year effects, climate effects, fixed effects) from a
+randomly selected MCMC iteration. We selected random year effects and
+the climate year independently.
+
+\subsection{Results}\label{results}
+
+\subsubsection{Climate effects on vital
+rates}\label{climate-effects-on-vital-rates}
+
+At the individual level we fit vital rate models that included climate
+covariates and random year effects (full model), just climate covariates
+(climate model), and no climate or random year effects (constant model).
+We used the metric from Adler et al. (2012) to quantify the contribution
+of climate covariates to explaining interannual variability in vital
+rates based on comparisons of model deviance: (Climate model - Constant
+Model)/(Full model - Constant model). At the quadrat level, we could
+only fit random year effects for the growth regression. This is, in
+fact, one of the limitation of census-based data: local extirpation and
+colonization events are likely to be rare, making it difficult to fit
+complex statistical models. So, in this section we focus on growth only.
+
+Climate covariates improved the growth regressions at both levels, with
+a slghty greater contribution at the individual level (Figure 1).
+Parameter estimates were generally consistent in terms of sign and
+magnitude between the individual level and the quadrat level for each
+species and climate covariate (Figure 2). There is no clear pattern
+suggesting that parameters estimated using individual level data are
+more certain, as evidenced by credible interval widths, than estimates
+based on quadrat data.
+
+\subsubsection{Climate effects on population
+dynamics}\label{climate-effects-on-population-dynamics}
+
+\subsection{Discussion}\label{discussion}
+
+We sought to test the assumption that the sensitivities of plant
+populations to climate variables can be detected equally well using
+either individual level data or population level data. This is an
+important question to answer because population models are key tools for
+predicting the consequences of global climate change. However, they can
+be of limited use when built on data from a small subset of a population
+in space or time. If population level data (i.e., some aggregated form
+of individual level data) can be used to detect climate effects on
+population dynamics, then we would have a cheaper and easier option for
+data collection over relatively large temporal and spatial extents
+(\emph{e.g.} Freckleton et al. 2011).
+
+\subsection{References}\label{references}
+
+\subsection{Figures}\label{figures}
 
@@
 
-\subsection{References}\label{references}
+\begin{figure}[htbp]
+\centering
+\includegraphics{components/figure/manuscript-figure_4.pdf}
+\caption{Simulated changes species' median cover caused by a 1\%
+increase in observed precipitation or temperature. We used the median of
+simulated results because the average was highly influence by some
+extreme outliers.}
+\end{figure}
 

### 2014-12-29 montana/ipm/vitalRateRegs/recruitment/import2ipm.R

@@ -1,0 +1,65 @@
+# ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ #
+# Test script to pull in rect parameters for IPM #
+# ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ #
+
+#Several arguments are required
+#' @param doYear  Specific climate year to pull year random effects
+#' @param mcDraw  A numeric scalar or vector for the row(s) of MCMC to draw parameters from 
+#' @param doSpp   A character scalar for the current speicies
+# @param group   A numeric scalar with group indicator
+
+library(reshape2)
+library(plyr)
+
+sppList=c("ARTR","HECO","POSE","PSSP") #in here for now; remove after testing
+
+####
+#### Read in full MCMC output and format as data frame
+####
+MCMC <- readRDS("recruitmentParamsMCMC.rds")
+prec2 <- melt(MCMC)
+prec2$Spp <- c(rep(rep(sppList, each=3000), times=13),
+                rep(rep(sppList, each=3000), times=6),
+                rep(rep(sppList, each=3000), times=13),
+                rep(rep(sppList, each=3000), times=5))
+prec2$Coef <- c(rep("beta", times=3000*4*13),
+                 rep("gInt", times=6*4*3000),
+                 rep("intYr", times=4*3000*13),
+                 rep("nb", times=4*3000),
+                 rep("rain1", times=4*3000),
+                 rep("rain2", times=4*3000),
+                 rep("temp1", times=4*3000),
+                 rep("temp2", times=4*3000))
+colnames(prec2)[1] <- "Iter"
+prec <- prec2[,c(1,3:5)]; rm(prec2)
+precAll <- subset(prec, Coef=="gInt"|Coef=="nb"|Coef=="rain1"|Coef=="rain2"|Coef=="temp1"|Coef=="temp2")
+precYrs <- subset(prec, Coef=="beta" | Coef=="intYr")
+years <- unique(allD$year)[2:14]+1900
+precYrs$Year <- c(rep(rep(years, each=3000), each=4),
+                   rep(rep(years, each=3000), each=4))
+
+####
+#### Now get subset defined by parameters; in a function
+####
+getRecCoefs <- function(doYear, mcDraw, doSpp){
+  #First do coefficients with random year effects
[…]

### 2014-12-29 montana/ipm/vitalRateRegs/survival/import2ipm.R

@@ -1,0 +1,65 @@
+# ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ #
+# Test script to pull in growth parameters for IPM #
+# ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ #
+
+#Several arguments are required
+#' @param doYear  Specific climate year to pull year random effects
+#' @param mcDraw  A numeric scalar or vector for the row(s) of MCMC to draw parameters from 
+#' @param doSpp   A character scalar for the current speicies
+# @param group   A numeric scalar with group indicator
+
+library(reshape2)
+library(plyr)
+
+sppList=c("ARTR","HECO","POSE","PSSP") #in here for now; remove after testing
+
+####
+#### Read in full MCMC output and format as data frame
+####
+MCMC <- readRDS("survivalParamsMCMC.rds")
+psurv2 <- melt(MCMC)
+psurv2$Spp <- c(rep(rep(sppList, each=3000), times=13),
+                rep(rep(sppList, each=3000), times=6),
+                rep(rep(sppList, each=3000), times=13),
+                rep(rep(sppList, each=3000), times=5))
+psurv2$Coef <- c(rep("beta", times=3000*4*13),
+                 rep("gInt", times=6*4*3000),
+                 rep("intYr", times=4*3000*13),
+                 rep("nb", times=4*3000),
+                 rep("rain1", times=4*3000),
+                 rep("rain2", times=4*3000),
+                 rep("temp1", times=4*3000),
+                 rep("temp2", times=4*3000))
+colnames(psurv2)[1] <- "Iter"
+psurv <- psurv2[,c(1,3:5)]; rm(psurv2)
+psurvAll <- subset(psurv, Coef=="gInt"|Coef=="nb"|Coef=="rain1"|Coef=="rain2"|Coef=="temp1"|Coef=="temp2")
+psurvYrs <- subset(psurv, Coef=="beta" | Coef=="intYr")
+years <- unique(allD$year)[2:14]+1900
+psurvYrs$Year <- c(rep(rep(years, each=3000), each=4),
+                   rep(rep(years, each=3000), each=4))
+
+####
+#### Now get subset defined by parameters; in a function
+####
+getSurvCoefs <- function(doYear, mcDraw, doSpp){
[…]

### 2014-12-29 montana/quadBM/simulations/PptTmpChangeQBM.R

@@ -22,7 +22,16 @@ sppList <- as.character(unique(allD$Species))
 
 #bring in climate data
 climD <- read.csv("../../weather/Climate.csv")
-climD[3:6] <- scale(climD[3:6], center = TRUE, scale = TRUE)
+climScale <- scale(climD[3:6], center = TRUE, scale = TRUE)
+climAvg <- apply(X = climD, MARGIN = 2, FUN = mean)
+climSD <- apply(X = climD, MARGIN = 2, FUN = sd)
+climD[c(3,5)] <- climD[c(3,5)]+(climD[c(3,5)]*0.01)
+climD[c(4,6)] <- climD[c(4,6)]+(climD[c(4,6)]*0.01)
+climD[3] <- (climD[3] - climAvg[3])/climSD[3]
+climD[4] <- (climD[4] - climAvg[4])/climSD[4]
+climD[5] <- (climD[5] - climAvg[5])/climSD[5]
+climD[6] <- (climD[6] - climAvg[6])/climSD[6]
+
 
 doSpp <- sppList[1]
 
@@ -146,71 +155,46 @@ colFunc <- function(pCol, N, climate, simsPerYear, doYear, sppSim){
 ####
 #### Run simulations -----------------------------------------------------
 ####
-outD <- data.frame(year=NA, cover=NA, sim=NA, species=NA)
+outD <- data.frame(variable=NA, cover=NA, species=NA)
 
 for(i in 1:length(sppList)){
   sppSim <- sppList[i]
-  nSim <- 1
-  yearsN <- length(unique(allD$year))
+  nSim <- 100
+  yearsN <- 100
   years <- unique(allD$year)+1900
   yearsID <- unique(allD$year)
   Nsave <- matrix(ncol=yearsN, nrow=nSim)
   sppD <- subset(allD, Species==sppSim)
   Nsave[,1] <- mean(subset(sppD, year==yearsID[1])$percCover)
   
-  for(sim in 1:nSim){
-    for(yr in 2:yearsN){
+  
+  for(yr in 2:yearsN){
+    for(sim in 1:nSim){
       N <- Nsave[sim,yr-1]
-      climate <- subset(climD, year==years[yr-1])[,c(3,5,4,6)]
+      climYr <- sample(climD$year,1)
+      climate <- subset(climD, year==climYr)[,c(3,5,4,6)]
+      doYear <- sample(years[2:length(years)], 1)
       
       ifelse(N[N>0],
[…]

### 2014-12-29 montana/quadBM/simulations/climChangeWrapper.R

@@ -1,0 +1,7 @@
+####
+#### Wrapper for climate change simulations of QBM
+####
+
+source("climateChangeQBM.R")
+source("precipitationChangeQBM.R")
+source("temperatureChangeQBM.R")

### 2014-12-29 montana/quadBM/simulations/climateChangeQBM.R

@@ -22,7 +22,14 @@ sppList <- as.character(unique(allD$Species))
 
 #bring in climate data
 climD <- read.csv("../../weather/Climate.csv")
-climD[3:6] <- scale(climD[3:6], center = TRUE, scale = TRUE)
+climScale <- scale(climD[3:6], center = TRUE, scale = TRUE)
+climAvg <- apply(X = climD, MARGIN = 2, FUN = mean)
+climSD <- apply(X = climD, MARGIN = 2, FUN = sd)
+climD[3] <- (climD[3] - climAvg[3])/climSD[3]
+climD[4] <- (climD[4] - climAvg[4])/climSD[4]
+climD[5] <- (climD[5] - climAvg[5])/climSD[5]
+climD[6] <- (climD[6] - climAvg[6])/climSD[6]
+
 
 doSpp <- sppList[1]
 
@@ -146,71 +153,46 @@ colFunc <- function(pCol, N, climate, simsPerYear, doYear, sppSim){
 ####
 #### Run simulations -----------------------------------------------------
 ####
-outD <- data.frame(year=NA, cover=NA, sim=NA, species=NA)
+outD <- data.frame(variable=NA, cover=NA, species=NA)
 
 for(i in 1:length(sppList)){
   sppSim <- sppList[i]
-  nSim <- 1
-  yearsN <- length(unique(allD$year))
+  nSim <- 100
+  yearsN <- 100
   years <- unique(allD$year)+1900
   yearsID <- unique(allD$year)
   Nsave <- matrix(ncol=yearsN, nrow=nSim)
   sppD <- subset(allD, Species==sppSim)
   Nsave[,1] <- mean(subset(sppD, year==yearsID[1])$percCover)
   
-  for(sim in 1:nSim){
-    for(yr in 2:yearsN){
+  
+  for(yr in 2:yearsN){
+    for(sim in 1:nSim){
       N <- Nsave[sim,yr-1]
-      climate <- subset(climD, year==years[yr-1])[,c(3,5,4,6)]
+      climYr <- sample(climD$year,1)
+      climate <- subset(climD, year==climYr)[,c(3,5,4,6)]
+      doYear <- sample(years[2:length(years)], 1)
       
       ifelse(N[N>0],
[…]

### 2014-12-29 montana/quadBM/simulations/precipitationChangeQBM.R

@@ -22,7 +22,16 @@ sppList <- as.character(unique(allD$Species))
 
 #bring in climate data
 climD <- read.csv("../../weather/Climate.csv")
-climD[3:6] <- scale(climD[3:6], center = TRUE, scale = TRUE)
+climScale <- scale(climD[3:6], center = TRUE, scale = TRUE)
+climAvg <- apply(X = climD, MARGIN = 2, FUN = mean)
+climSD <- apply(X = climD, MARGIN = 2, FUN = sd)
+climD[c(3,5)] <- climD[c(3,5)]+(climD[c(3,5)]*0.01)
+# climD[c(4,6)] <- climD[c(4,6)]+(climD[c(4,6)]*0.01)
+climD[3] <- (climD[3] - climAvg[3])/climSD[3]
+climD[4] <- (climD[4] - climAvg[4])/climSD[4]
+climD[5] <- (climD[5] - climAvg[5])/climSD[5]
+climD[6] <- (climD[6] - climAvg[6])/climSD[6]
+
 
 doSpp <- sppList[1]
 
@@ -146,71 +155,46 @@ colFunc <- function(pCol, N, climate, simsPerYear, doYear, sppSim){
 ####
 #### Run simulations -----------------------------------------------------
 ####
-outD <- data.frame(year=NA, cover=NA, sim=NA, species=NA)
+outD <- data.frame(variable=NA, cover=NA, species=NA)
 
 for(i in 1:length(sppList)){
   sppSim <- sppList[i]
-  nSim <- 1
-  yearsN <- length(unique(allD$year))
+  nSim <- 100
+  yearsN <- 100
   years <- unique(allD$year)+1900
   yearsID <- unique(allD$year)
   Nsave <- matrix(ncol=yearsN, nrow=nSim)
   sppD <- subset(allD, Species==sppSim)
   Nsave[,1] <- mean(subset(sppD, year==yearsID[1])$percCover)
   
-  for(sim in 1:nSim){
-    for(yr in 2:yearsN){
+  
+  for(yr in 2:yearsN){
+    for(sim in 1:nSim){
       N <- Nsave[sim,yr-1]
-      climate <- subset(climD, year==years[yr-1])[,c(3,5,4,6)]
+      climYr <- sample(climD$year,1)
+      climate <- subset(climD, year==climYr)[,c(3,5,4,6)]
+      doYear <- sample(years[2:length(years)], 1)
       
       ifelse(N[N>0],
[…]

### 2014-12-29 montana/quadBM/simulations/temperatureChangeQBM.R

@@ -22,7 +22,15 @@ sppList <- as.character(unique(allD$Species))
 
 #bring in climate data
 climD <- read.csv("../../weather/Climate.csv")
-climD[3:6] <- scale(climD[3:6], center = TRUE, scale = TRUE)
+climScale <- scale(climD[3:6], center = TRUE, scale = TRUE)
+climAvg <- apply(X = climD, MARGIN = 2, FUN = mean)
+climSD <- apply(X = climD, MARGIN = 2, FUN = sd)
+climD[c(4,6)] <- climD[c(4,6)]+(climD[c(4,6)]*0.01)
+climD[3] <- (climD[3] - climAvg[3])/climSD[3]
+climD[4] <- (climD[4] - climAvg[4])/climSD[4]
+climD[5] <- (climD[5] - climAvg[5])/climSD[5]
+climD[6] <- (climD[6] - climAvg[6])/climSD[6]
+
 
 doSpp <- sppList[1]
 
@@ -146,71 +154,46 @@ colFunc <- function(pCol, N, climate, simsPerYear, doYear, sppSim){
 ####
 #### Run simulations -----------------------------------------------------
 ####
-outD <- data.frame(year=NA, cover=NA, sim=NA, species=NA)
+outD <- data.frame(variable=NA, cover=NA, species=NA)
 
 for(i in 1:length(sppList)){
   sppSim <- sppList[i]
-  nSim <- 1
-  yearsN <- length(unique(allD$year))
+  nSim <- 100
+  yearsN <- 100
   years <- unique(allD$year)+1900
   yearsID <- unique(allD$year)
   Nsave <- matrix(ncol=yearsN, nrow=nSim)
   sppD <- subset(allD, Species==sppSim)
   Nsave[,1] <- mean(subset(sppD, year==yearsID[1])$percCover)
   
-  for(sim in 1:nSim){
-    for(yr in 2:yearsN){
+  
+  for(yr in 2:yearsN){
+    for(sim in 1:nSim){
       N <- Nsave[sim,yr-1]
-      climate <- subset(climD, year==years[yr-1])[,c(3,5,4,6)]
+      climYr <- sample(climD$year,1)
+      climate <- subset(climD, year==climYr)[,c(3,5,4,6)]
+      doYear <- sample(years[2:length(years)], 1)
       
       ifelse(N[N>0],
[…]

### 2015-01-05 manuscript/manuscript.Rmd
updated ms with climate sims

@@
 
-In contrast to individual-level data, population-level data arising from large-scale census efforts
-
-Recently, @Freckleton2011, building on work by @Taylor2004, have proposed density-structured population models that focus on the transition of populations among discrete states, rather than the traditional approach of modeling the transitions of individuals.
-Such an approach could be extremely valuable because the data needed to parameterize density-structured population models is much easier, and less costly, to collect.
-
-For example, using a density-structured approach, one could build a population model using a time series of annual plot-based censuses of species percent cover.
-However, a major assumption of the density-structured approach is that the aggregate dynamics of the population observed at coarse spatial resolution faithfully represent, and correspond to, the fates of individual plants.
-In other words, using a density-structured approach requires a leap of faith that important covariates (e.g., climate variables) at the level of the individual are captured adequately at the population level.
-If we seek to forecast the impacts of climate change on plant populations, then clearly this assumption requires testing.
+Aggregate measures of individual plant performance, such as those typically collected as part of large-scale census efforts, offer an alternative to detailed demographic data for modeling populations.
+Such population-level data will never match the precision of individual-level data,
 
@@
 ### Study site and data
-Our demographic data comes from the Fort Keogh Livestock and Range Research Laboratory in eastern Montana's northern mixed prairie near Miles City, Montana, USA (46 deg.
-19' N, 105 deg 48' W).
+Our demographic data comes from the Fort Keogh Livestock and Range Research Laboratory in eastern Montana's northern mixed prairie near Miles City, Montana, USA ($46^{\circ}$ 19' N, $105^{\circ}$ 48' W).
 The dataset is freely available on Ecological Archives [@Anderson2011], and interested readers should refer to the metadata therin for a complete description.
@@
 climTemp <- readRDS("../montana/quadBM/simulations/climateChangeTemperature.rds")
+climBoth <- readRDS("../montana/quadBM/simulations/climateChangePptTmp.rds")
 climBase$sim <- rep("Observed", nrow(climBase)) 
@@
 climTemp$sim <- rep("Temperature", nrow(climBase)) 
-climSimsQBM <- rbind(climBase, climPpt, climTemp)
+climBoth$sim <- rep("Both", nrow(climBase)) 
+climSimsQBM <- rbind(climBase, climPpt, climTemp, climBoth)
 # ggplot(climSimsQBM)+
@@
 climSimQBMcast$diffT <- with(climSimQBMcast, (Temperature*100)-(Observed*100))
+climSimQBMcast$diffB <- with(climSimQBMcast, (Both*100)-(Observed*100))
 climSimQBMplot <- melt(climSimQBMcast)
-ggplot(data=subset(climSimQBMplot, variable=="diffP"|variable=="diffT"))+
-  geom_bar(aes(x=species, y=value, fill=variable), stat="identity", position="dodge")+
-  scale_fill_manual(name="", labels=c("1% Precipitation Increase", "1% Temperature Increase"), values=c("grey25", "grey50"))+
+ggplot(data=subset(climSimQBMplot, variable=="diffP"|variable=="diffT"|variable=="diffB"))+
+  geom_bar(aes(x=species, y=value, fill=variable), stat="identity", position="dodge", color="white")+
+  scale_fill_manual(name="", labels=c("1% Precipitation Increase", "1% Temperature Increase", "1% Temp. & Ppt. Increase"), values=c("grey25", "grey50", "grey75"))+
   theme_bw()+

### 2015-01-05 manuscript/manuscript.tex
updated ms with climate sims

@@
 
-In contrast to individual-level data, population-level data arising from
-large-scale census efforts
-
-Recently, Freckleton et al. (2011), building on work by Taylor and
-Hastings (2004), have proposed density-structured population models that
-focus on the transition of populations among discrete states, rather
-than the traditional approach of modeling the transitions of
-individuals. Such an approach could be extremely valuable because the
-data needed to parameterize density-structured population models is much
-easier, and less costly, to collect.
-
-For example, using a density-structured approach, one could build a
-population model using a time series of annual plot-based censuses of
-species percent cover. However, a major assumption of the
-density-structured approach is that the aggregate dynamics of the
-population observed at coarse spatial resolution faithfully represent,
-and correspond to, the fates of individual plants. In other words, using
-a density-structured approach requires a leap of faith that important
-covariates (e.g., climate variables) at the level of the individual are
-captured adequately at the population level. If we seek to forecast the
-impacts of climate change on plant populations, then clearly this
-assumption requires testing.
+Aggregate measures of individual plant performance, such as those
+typically collected as part of large-scale census efforts, offer an
+alternative to detailed demographic data for modeling populations. Such
+population-level data will never match the precision of individual-level
+data,
 
@@
 Research Laboratory in eastern Montana's northern mixed prairie near
-Miles City, Montana, USA (46 deg. 19' N, 105 deg 48' W). The dataset is
-freely available on Ecological Archives (Anderson et al. 2011), and
-interested readers should refer to the metadata therin for a complete
-description. The site is about 800 m above sea level and mean annual
-precipitation (1878-2009) is 334 mm, with most annual precipitaion
-falling from April through September. The site is grass dominated and,
-for the purposes of our study, we focus on the four most abundant
-graminoid species: \emph{Bouteloua gracilis} (BOGR),
+Miles City, Montana, USA ($46^{\circ}$ 19' N, $105^{\circ}$ 48' W). The
+dataset is freely available on Ecological Archives (Anderson et al.
+2011), and interested readers should refer to the metadata therin for a
+complete description. The site is about 800 m above sea level and mean
+annual precipitation (1878-2009) is 334 mm, with most annual
+precipitaion falling from April through September. The site is grass
+dominated and, for the purposes of our study, we focus on the four most
+abundant graminoid species: \emph{Bouteloua gracilis} (BOGR),
 \emph{Hesperostipa comata} (HECO), \emph{Pascopyrum smithii} (PASM), and
@@
 
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\itemsep1pt\parskip0pt\parsep0pt
+\item
+  Climate detections
+\item
+  Limitations:
+\end{enumerate}
+
+\begin{itemize}
+\itemsep1pt\parskip0pt\parsep0pt
+\item
+  Sample size at quadrat level
+\item
+  are these species inherently sensitive to climate (compare to Chu and
+  Adler in review -- say why we chose Montana site -- most sensitive to
+  climate variables)
+\item
+  we know recruitment is important, and this is something that may be
+  easily missed at quad level
+\end{itemize}
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\setcounter{enumi}{2}
+\itemsep1pt\parskip0pt\parsep0pt
+\item
+  Implications
+\end{enumerate}
+
 \subsection{References}\label{references}
@@
 
-Taylor, C. M., and A. Hastings. 2004. Finding optimal control strategies
-for invasive species: a density-structured model for Spartina
-alterniflora. Journal of Applied Ecology 41:1049--1057.
-
 Team, R. 2013. R Development Core Team. R: A Language and Environment

### 2015-01-06 manuscript/manuscript.Rmd
updated with intro beef up

@@
 
-Aggregate measures of individual plant performance, such as those typically collected as part of large-scale census efforts, offer an alternative to detailed demographic data for modeling populations.
-Such population-level data will never match the precision of individual-level data,
+Aggregate measures of individual plant performance, such as those typically collected as part of large-scale census efforts, offer an alternative to detailed demographic data for modeling populations [@Clark2004; @Freckleton2011].
+Such population-level data will never match the precision of individual-level data, but it is more feasible to attain a broad coverage sample when collecting coarse-scale data.
+This presents a difficult trade-off: on the one hand, individual-level data leads to more reliable models; on the other hand, population-level data leads to models that will produce less precise predicitons but can be applied over greater spatial and temporal extents.
+If the goal of the model is to predict the consequences of climate change, then we argue that in the face of limited time and money population-level data are the best option.
+However, the performance of models based on population-level data have rarely been compared to that of models based on individual-level data.
+
+To date, relatively few studies have tried to model populations based on data other than detailed individual-level data.
+An important exception is an effort by @Taylor2004 to model the population growth rate of an invasive species to investigate the best strategies for invasion control.
+They used a "density-structured" model where the state variable is a discrete density state rather than a continuous density measure.
+Building on this work, @Freckleton2011 showed that density-structured models compare well to continuous models in theory, and @Queenborough2011 showed the application of such methods in a study on arable weeds.
+In particular, @Queenborough2011 provide emprical evidence that density-structured models are capable of reproducing population dynamics, even if some precision is lost when compared to fully continuous models.
+Thus, population models based on coarse, population-level data show promise for producing ecological forecasts at landscape and regional scales [@Queenborough2011].
+However, none of these models included environmental covariates.
+
+Basing population models on population- rather than individual-level data in a climate change context is hampered by the fact that it is individuals that respond to climate, not populations [@Clark2012].
+This fact puts us in uneasy proximity to an "ecological fallacy" where one deduces inference on the individual from statistical inference on the group [@PIANTADOSI1988].
 

### 2015-01-06 manuscript/manuscript.tex
updated with intro beef up

@@
 typically collected as part of large-scale census efforts, offer an
-alternative to detailed demographic data for modeling populations. Such
-population-level data will never match the precision of individual-level
-data,
+alternative to detailed demographic data for modeling populations (Clark
+and Bjørnstad 2004, Freckleton et al. 2011). Such population-level data
+will never match the precision of individual-level data, but it is more
+feasible to attain a broad coverage sample when collecting coarse-scale
+data. This presents a difficult trade-off: on the one hand,
+individual-level data leads to more reliable models; on the other hand,
+population-level data leads to models that will produce less precise
+predicitons but can be applied over greater spatial and temporal
+extents. If the goal of the model is to predict the consequences of
+climate change, then we argue that in the face of limited time and money
+population-level data are the best option. However, the performance of
+models based on population-level data have rarely been compared to that
+of models based on individual-level data.
+
+To date, relatively few studies have tried to model populations based on
+data other than detailed individual-level data. An important exception
+is an effort by Taylor and Hastings (2004) to model the population
+growth rate of an invasive species to investigate the best strategies
+for invasion control. They used a ``density-structured'' model where the
+state variable is a discrete density state rather than a continuous
+density measure. Building on this work, Freckleton et al. (2011) showed
+that density-structured models compare well to continuous models in
+theory, and Queenborough et al. (2011) showed the application of such
+methods in a study on arable weeds. In particular, Queenborough et al.
+(2011) provide emprical evidence that density-structured models are
+capable of reproducing population dynamics, even if some precision is
+lost when compared to fully continuous models. Thus, population models
+based on coarse, population-level data show promise for producing
+ecological forecasts at landscape and regional scales (Queenborough et
+al. 2011). However, none of these models included environmental
+covariates.
+
+Basing population models on population- rather than individual-level
+data in a climate change context is hampered by the fact that it is
+individuals that respond to climate, not populations (Clark et al.
+2012). This fact puts us in uneasy proximity to an ``ecological
+fallacy'' where one deduces inference on the individual from statistical
+inference on the group (Piantadosi et al. 1988).
 
@@
 
+Clark, J. S., and O. N. Bjørnstad. 2004. Population time series: Process
+variability, observation errors, missing values, lags, and hidden
+states. Ecology 85:3140--3150.
+
+Clark, J. S., D. M. Bell, M. Kwit, A. Stine, B. Vierra, and K. Zhu.
+2012. Individual-scale inference to anticipate climate-change
+vulnerability of biodiversity.
+
 Freckleton, R. P., W. J. Sutherland, A. R. Watkinson, and S. A.
@@
 Using Multiple Sequences.
+
+Piantadosi, S., D. P. Byar, and S. B. Green. 1988. The Ecological
+Fallacy. American Journal of Epidemiology 127:893--904.
 
@@
 
+Taylor, C. M., and A. Hastings. 2004. Finding optimal control strategies
+for invasive species: a density-structured model for Spartina
+alterniflora. Journal of Applied Ecology 41:1049--1057.
+
 Team, R. 2013. R Development Core Team. R: A Language and Environment

### 2015-01-06 manuscript/manuscript.Rmd
improved intro

@@
 Population models are important tools for predicting the impacts of environmental change on species.
-But reconciling the scales at which population models are parameterized and the scales at which environmental changes play out remains a challenge [@Freckleton2011; @Queenborough2011].
+But reconciling the scales at which population models are parameterized and the scales at which environmental changes play out remains a challenge [@Clark2010; @Freckleton2011; @Queenborough2011; @Clark2012].
 The major hurdle is that most population models, at least for plant species, are built using data from small, localized plots because parameterizing traditional population models requires tracking the fates of individuals.
@@
 
-Basing population models on population- rather than individual-level data in a climate change context is hampered by the fact that it is individuals that respond to climate, not populations [@Clark2012].
+Basing population models on aggregated individual-level data in a climate change context is hampered by the fact that it is individuals that respond to climate, not populations [@Clark2012].
 This fact puts us in uneasy proximity to an "ecological fallacy" where one deduces inference on the individual from statistical inference on the group [@PIANTADOSI1988].
+For example, individual plants may respond positively to precipitation but a negative trend is observed at the population level due to increased competition among plants as they grow larger and consume more resources.
+Thus, it is important to ask the question: Can aggregated data be used to detect climate signals of the same sign and magnitude as individual-level data?
+If not, then building population models with climate covariates on aggregated data will lead to incorrect forecasts.
+
+Here, we test the hypothesis (idea?) that statistical and population models based on aggregated data can detect climate signals as wells as models based on individual-level data.
+We use a unique demographic dataset that tracks the fates of individual plants from four species to build single-species population models, since those are often used tools for ecological forecasts and climate vulnerability assessments.
+We first fit population models with interannual variation in vital rates explained, in part, by climate covariates.
+We then perturb the climate covariates to test the sensitivities of species to climate change.
+By doing these analyses using both individual and aggregated forms of the same data, we provide a rigorous test of our hypothesis.
+We find that...
 
@@
 
-References
-----------
+\pagebreak{}
 
@@
 ```
+
+\pagebreak{}
+
+References
+----------

### 2015-01-06 manuscript/manuscript.tex
improved intro

@@
 population models are parameterized and the scales at which
-environmental changes play out remains a challenge (Freckleton et al.
-2011, Queenborough et al. 2011). The major hurdle is that most
-population models, at least for plant species, are built using data from
-small, localized plots because parameterizing traditional population
-models requires tracking the fates of individuals. These models are
-difficult to scale up from the micro to meso-scales because the fitted
-parameters do not fully represent the spatial variation present at
-scales beyond that at which the data are collected (Sæther et al. 2007).
-Thus, our ability to use population models to predict the consequences
-of climate change is limited when we rely on individual-level data.
+environmental changes play out remains a challenge (Clark et al. 2010,
+2012, Freckleton et al. 2011, Queenborough et al. 2011). The major
+hurdle is that most population models, at least for plant species, are
+built using data from small, localized plots because parameterizing
+traditional population models requires tracking the fates of
+individuals. These models are difficult to scale up from the micro to
+meso-scales because the fitted parameters do not fully represent the
+spatial variation present at scales beyond that at which the data are
+collected (Sæther et al. 2007). Thus, our ability to use population
+models to predict the consequences of climate change is limited when we
+rely on individual-level data.
 
@@
 
-Basing population models on population- rather than individual-level
-data in a climate change context is hampered by the fact that it is
-individuals that respond to climate, not populations (Clark et al.
-2012). This fact puts us in uneasy proximity to an ``ecological
-fallacy'' where one deduces inference on the individual from statistical
-inference on the group (Piantadosi et al. 1988).
+Basing population models on aggregated individual-level data in a
+climate change context is hampered by the fact that it is individuals
+that respond to climate, not populations (Clark et al. 2012). This fact
+puts us in uneasy proximity to an ``ecological fallacy'' where one
+deduces inference on the individual from statistical inference on the
+group (Piantadosi et al. 1988). For example, individual plants may
+respond positively to precipitation but a negative trend is observed at
+the population level due to increased competition among plants as they
+grow larger and consume more resources. Thus, it is important to ask the
+question: Can aggregated data be used to detect climate signals of the
+same sign and magnitude as individual-level data? If not, then building
+population models with climate covariates on aggregated data will lead
+to incorrect forecasts.
+
+Here, we test the hypothesis (idea?) that statistical and population
+models based on aggregated data can detect climate signals as wells as
+models based on individual-level data. We use a unique demographic
+dataset that tracks the fates of individual plants from four species to
+build single-species population models, since those are often used tools
+for ecological forecasts and climate vulnerability assessments. We first
+fit population models with interannual variation in vital rates
+explained, in part, by climate covariates. We then perturb the climate
+covariates to test the sensitivities of species to climate change. By
+doing these analyses using both individual and aggregated forms of the
+same data, we provide a rigorous test of our hypothesis. We find
+that\ldots{}
 
@@
 
-\subsection{References}\label{references}
+\pagebreak{}
 
@@
 
+\pagebreak{}
+
+\subsection{References}\label{references}
+
 Adler, P. B., H. J. Dalgleish, and S. P. Ellner. 2012. Forecasting plant
@@
 vulnerability of biodiversity.
+
+Clark, J. S., D. Bell, C. Chu, B. Courbaud, M. Dietze, M. Hersh, J.
+HilleRisLambers, I. Ibáñez, S. LaDeau, S. McMahon, J. Metcalf, J. Mohan,
+E. Moran, L. Pangle, S. Pearson, C. Salk, Z. Shen, D. Valle, and P.
+Wyckoff. 2010. High-dimensional coexistence based on individual
+variation: a synthesis of evidence. Ecological Monographs 80:569--608.
 

### 2015-01-07 manuscript/manuscript.Rmd
updated with supp mat

@@
 *Peter B. Adler, Department of Wildland Resources and the Ecology Center, Utah State University, Logan, UT*
+
+```{r libraris, include=FALSE}
+library(ggplot2)
+library(plyr)
+library(gridExtra)
+library(reshape2)
+```
 
@@
 -------
-
-```{r figure_1, dependson="plot-options", fig.cap="The proportion of interannual variability in  vital rates explained by the climate covariates.
-The contribution for growth is defined as: (Climate model - Constant Model)/(Full model - Constant model).
-The contribution for survival and colonization, where we could not estimate a full model with year random effects at the quadrat level, is defined as: (Constant Model - Climate Model)/Constant Model.", fig.width=7, fig.height=3, cache=FALSE, results='hide'}
-deviance <- data.frame(VitalRate = c("Growth", "Survival", "Colonization"),
-                       ConstantModel = rep(NA,3),
-                       ClimateModel = rep(NA,3),
-                       FullModel = rep(NA,3),
-                       ClimateContribution = rep(NA,3))
-deviance$FullModel[1] <- round(read.csv("../montana/quadBM/vitalRateRegressions/growth/growthDeviance.csv")[,2],2)
-deviance$ClimateModel[1] <- round(read.csv("../montana/quadBM/vitalRateRegressions/growth/growthDevianceCLIMATE.csv")[,2],2)
-deviance$ConstantModel[1] <- round(read.csv("../montana/quadBM/vitalRateRegressions/growth/growthDevianceCONSTANT.csv")[,2],2)
-deviance$ClimateContribution[1] <- with(deviance[1,], (ClimateModel-ConstantModel)/(FullModel-ConstantModel))
-deviance$ClimateModel[2] <- round(read.csv("../montana/quadBM/vitalRateRegressions/survival/survivalDeviance.csv")[,2],2)
-deviance$ConstantModel[2] <- round(read.csv("../montana/quadBM/vitalRateRegressions/survival/survivalDevianceCONSTANT.csv")[,2],2)
-deviance$ClimateModel[3] <- round(read.csv("../montana/quadBM/vitalRateRegressions/colonization/colonizationDeviance.csv")[,2],2)
-deviance$ConstantModel[3] <- round(read.csv("../montana/quadBM/vitalRateRegressions/colonization/colonizationDevianceCONSTANT.csv")[,2],2)
-deviance$ClimateContribution[2:3] <- with(deviance[2:3,], (ConstantModel-ClimateModel)/(ConstantModel))
-
-colnames(deviance) <- c("Vital Rate", "Constant model", "Climate model", "Full model", "Contribution of climate covariates")
-library(ggplot2)
-devGQ <- data.frame(VitalRate = c("Growth", "Survival", "Colonization"),
-                       ConstantModel = rep(NA,3),
-                       ClimateModel = rep(NA,3),
-                       FullModel = rep(NA,3),
-                       ClimateContribution = rep(NA,3))
-devGQ$FullModel[1] <- round(read.csv("../montana/ipm/vitalRateRegs/growth/growthDeviance.csv")[,2],2)
-devGQ$ClimateModel[1] <- round(read.csv("../montana/ipm/vitalRateRegs/growth/growthDevianceCLIMATE.csv")[,2],2)
-devGQ$ConstantModel[1] <- round(read.csv("../montana/ipm/vitalRateRegs/growth/growthDevianceCONSTANT.csv")[,2],2)
-devGQ$FullModel[2] <- round(read.csv("../montana/ipm/vitalRateRegs/survival/survivalDeviance.csv")[,2],2)
-devGQ$ClimateModel[2] <- round(read.csv("../montana/ipm/vitalRateRegs/survival/survivalDevianceCLIMATE.csv")[,2],2)
-devGQ$ConstantModel[2] <- round(read.csv("../montana/ipm/vitalRateRegs/survival/survivalDevianceCONSTANT.csv")[,2],2)
-devGQ$FullModel[3] <- round(read.csv("../montana/ipm/vitalRateRegs/recruitment/recruitmentDeviance.csv")[,2],2)
-devGQ$ClimateModel[3] <- round(read.csv("../montana/ipm/vitalRateRegs/recruitment/recruitmentDevianceCLIMATE.csv")[,2],2)
-devGQ$ConstantModel[3] <- round(read.csv("../montana/ipm/vitalRateRegs/recruitment/recruitmentDevianceCONSTANT.csv")[,2],2)
-devGQ$ClimateContribution[1] <- with(devGQ[1,], (ClimateModel-ConstantModel)/(FullModel-ConstantModel))
-devGQ$ClimateContribution[2] <- with(devGQ[2,], (ConstantModel-ClimateModel)/(ConstantModel))
-devGQ$ClimateContribution[3] <- with(devGQ[3,], (ConstantModel-ClimateModel)/(ConstantModel))
-
-devM <- devGQ[,c(1,5)]
-devM$Level <- rep("Individual", 3)
-tmpD <- deviance[,c(1,5)]
-tmpD$Level <- rep("Quadrat", 3)
-colnames(tmpD)[2] <- "value"
-colnames(tmpD)[1] <- "VitalRate"
-colnames(devM)[2] <- "value"
-plotD <- rbind(tmpD, devM)
-gClimCont1 <- ggplot(data=subset(plotD, Level=="Individual"), aes(x=VitalRate, y=value))+
-  geom_bar(stat="identity", position="dodge", fill="grey30")+
-  ylab("Contribution of Climate Effects")+
-  xlab("Vital Rate")+
-  ggtitle("Individual level")+
-  scale_y_continuous(limits=c(0, 0.5))+
-  scale_x_discrete(labels=c("Recruitment", "Growth", "Survival"))+
-  theme_bw()+
-  theme(axis.text.x = element_text(angle = 25, hjust = 1))
-gClimCont2 <- ggplot(data=subset(plotD, Level=="Quadrat"), aes(x=VitalRate, y=value))+
-  geom_bar(stat="identity", position="dodge", fill="grey30")+
-  ylab("Contribution of Climate Effects")+
-  xlab("Vital Rate")+
-  ggtitle("Quadrat level")+
-  scale_y_continuous(limits=c(0, 0.5))+
-  scale_x_discrete(labels=c("Colonization", "Cover change", "Extirpation"))+
-  theme_bw()+
-  theme(axis.text.x = element_text(angle = 25, hjust = 1))
-library(gridExtra)
-gClim <- grid.arrange(gClimCont1, gClimCont2, ncol=2, nrow=1)
-gClim
-
-```
-
-```{r figure_2, dependson="plot-options", fig.cap="Posterior means (vertical ticks), 75% credible intervals (heavy lines), and 95% credible intervals (light lines) of climate effects on growth at both levels of inferences.
-The dashed vertical line is at 0, indicating no effect.
-Horizontal line at 0 indicates perfect agreement between mean observed cover in that year and the model predictions.", fig.width=3, fig.height=7, cache=FALSE, results='hide'}
-gQC <- read.csv("../montana/quadBM/vitalRateRegressions/growth/growthQuants.csv")
-gQC <- gQC[129:144,]
-coefNames <- c("ppt1", "ppt2", "temp1", "temp2")
-gQC$X <- rep(coefNames, each=4)
-gQC$rate <- "growth"
-sQC <- read.csv("../montana/quadBM/vitalRateRegressions/survival/survivalQuants.csv")
-sQC <- sQC[33:48,]
-coefNames <- c("ppt1", "ppt2", "temp1", "temp2")
-sQC$X <- rep(coefNames, each=4)
-sQC$rate <- "survival"
-rQC <- read.csv("../montana/quadBM/vitalRateRegressions/colonization/colonizationQuants.csv")
-rQC <- rQC[29:44,]
-coefNames <- c("ppt1", "ppt2", "temp1", "temp2")
-rQC$X <- rep(coefNames, each=4)
-rQC$rate <- "colonization"
-plotQC <- rbind(gQC,sQC,rQC)
-gQ <- ggplot(plotQC, aes(x=species, y=X50., color=rate))+
-  geom_hline(aes(yintercept=0), linetype="dashed")+
-  geom_errorbar(aes(ymin=X2.5., ymax=X97.5.), width=0, position=position_dodge(width=0.5))+
-  geom_errorbar(aes(ymin=X25., ymax=X75.), width=0, size=1.25, position=position_dodge(width=0.5))+
-  geom_point(size=4, position=position_dodge(width=0.5), shape="|")+
-  ylab("Standardized Coefficient Value")+
-  xlab("Species")+
-  facet_grid(X~.)+
-  coord_flip()+
-  theme_bw()+
-  scale_color_manual(values=c("grey25", "steelblue", "slateblue"))+
-  guides(color=guide_legend(title=""))+
-  theme(legend.position="top")+
-  ggtitle("Quadrat level")
-
-gIC <- read.csv("../montana/ipm/vitalRateRegs/growth/growthQuants.csv")
-gIC <- gIC[133:148,]
-coefNames <- c("ppt1", "ppt2", "temp1", "temp2")
-gIC$X <- rep(coefNames, each=4)
-gIC$rate <- "growth"
-sIC <- read.csv("../montana/ipm/vitalRateRegs/survival/survivalQuants.csv")
-sIC <- sIC[133:148,]
-coefNames <- c("ppt1", "ppt2", "temp1", "temp2")
-sIC$X <- rep(coefNames, each=4)
-sIC$rate <- "survival"
-rIC <- read.csv("../montana/ipm/vitalRateRegs/recruitment/recruitmentQuants.csv")
-rIC <- rIC[81:96,]
-coefNames <- c("ppt1", "ppt2", "temp1", "temp2")
-rIC$X <- rep(coefNames, each=4)
-rIC$rate <- "colonization"
-plotIC <- rbind(gIC,sIC,rIC)
-gI <- ggplot(plotIC, aes(x=species, y=X50., color=rate))+
-  geom_hline(aes(yintercept=0), linetype="dashed")+
-  geom_errorbar(aes(ymin=X2.5., ymax=X97.5.), width=0, position=position_dodge(width=0.5))+
-  geom_errorbar(aes(ymin=X25., ymax=X75.), width=0, size=1.25, position=position_dodge(width=0.5))+
-  geom_point(size=4, position=position_dodge(width=0.5), shape="|")+
-  ylab("Standardized Coefficient Value")+
-  xlab("Species")+
-  facet_wrap(X~.)+
-  coord_flip()+
-  theme_bw()+
-  scale_color_manual(values=c("grey25", "steelblue", "slateblue"),
-                     labels=c("recruitment", "growth", "survival"))+
-#   guides(color=FALSE)+
-  guides(color=guide_legend(title=""))+
-  theme(legend.position="top")+
-  ggtitle("Individual level")
-
-####
-#### Plot just for growth
-####
-growthIC <- subset(plotIC, rate=="growth")
-growthQC <- subset(plotQC, rate=="growth")
-growthIC$Level <- "Individual"
-growthQC$Level <- "Quadrat"
-growthAll <- rbind(growthIC, growthQC)
-
-gGrowth <- ggplot(growthAll, aes(x=species, y=X50., color=Level))+
-  geom_hline(aes(yintercept=0), linetype="dashed")+
-  geom_errorbar(aes(ymin=X2.5., ymax=X97.5.), width=0, position=position_dodge(width=0.5))+
-  geom_errorbar(aes(ymin=X25., ymax=X75.), width=0, size=1.25, position=position_dodge(width=0.5))+
-  geom_point(size=4, position=position_dodge(width=0.5), shape="|")+
-  ylab("Standardized Coefficient Value")+
-  xlab("Species")+
-  facet_grid(X~.)+
-  coord_flip()+
-  theme_bw()+
-  scale_color_manual(values=c("grey25", "slateblue"))+
-#   guides(color=FALSE)+
-  guides(color=guide_legend(title=""))+
-  theme(legend.position="top")
-print(gGrowth)
-```
-
-```{r figure_3, dependson="plot-options", fig.cap="Boxplots of model residuals for one-step-ahead forecasts at each observation year.
+```{r figure_1, dependson="plot-options", fig.cap="Boxplots of model residuals for one-step-ahead forecasts at each observation year.
 Each one-step forecast was simulated `r nSims` times.
@@
 The models tend to underpredict and perform poorly when observed cover in a given year is a large deviant from the mean.", fig.width=5, fig.height=6, cache=FALSE, results='hide'}
-library(plyr)
-library(reshape2)
 resD <- readRDS("../montana/quadBM/simulations/quadBM_oneStep_Residuals.rds")
@@
 #   geom_line(data=diffD, aes(x=year, y=avgDiff, group=species), color="purple", alpha=0.5)+
-  stat_smooth(data=diffD, aes(x=year, y=avgDiff, group=species), color="lightblue", se = FALSE, size=1)+
+#   stat_smooth(data=diffD, aes(x=year, y=avgDiff, group=species), color="lightblue", se = FALSE, size=1)+
   geom_boxplot(aes(x=year, y=covChangeResiduals), outlier.shape = 1, width=0.3, outlier.size = 0.5)+
@@
 
-```{r figure_4, dependson="plot-options", fig.cap="Simulated changes species' median cover caused by a 1% increase in observed precipitation or temperature.
+```{r figure_2, dependson="plot-options", fig.cap="Simulated changes species' median cover caused by a 1% increase in observed precipitation or temperature.
 We used the median of simulated results because the average was highly influence by some extreme outliers.", fig.width=4.5, fig.height=4, cache=FALSE, results='hide'}

### 2015-01-07 manuscript/manuscript.tex
updated bibliography with correct R citation
updated with supp mat

@@
 All of our analyses (model fitting and simulating) were conducted in R
-(Team 2013). We used the MCMC sampler in JAGS (Plummer 2003) to estimate
-the posterior distributions of model parameters and the package `r2jags'
-(Su and Yajima 2012) to connect R to JAGS. We obtained posterior
-distributions for all model parameters from three parallel MCMC chains
-run for 50,000 iterations, after discarding an initial 50,000
-iterations. We assessed convergence visually and using the Gelman and
-Rubin (1992) diagnostic in the R package `coda' (Plummer et al. 2006).
-Scale reduction factors for all parameters were less than 1.02,
-indicating convergence. For the purposes of introducing stochasticity in
-our population models, we saved the final 1,000 iterations from each
-chain for all parameters to be used as randomly drawn values during
-population simulation.
+(R Development Core Team 2013). We used the MCMC sampler in JAGS
+(Plummer 2003) to estimate the posterior distributions of model
+parameters and the package `r2jags' (Su and Yajima 2012) to connect R to
+JAGS. We obtained posterior distributions for all model parameters from
+three parallel MCMC chains run for 50,000 iterations, after discarding
+an initial 50,000 iterations. We assessed convergence visually and using
+the Gelman and Rubin (1992) diagnostic in the R package `coda' (Plummer
+et al. 2006). Scale reduction factors for all parameters were less than
+1.02, indicating convergence. For the purposes of introducing
+stochasticity in our population models, we saved the final 1,000
+iterations from each chain for all parameters to be used as randomly
+drawn values during population simulation.
 
@@
 \includegraphics{components/figure/manuscript-figure_1.pdf}
-\caption{The proportion of interannual variability in vital rates
-explained by the climate covariates. The contribution for growth is
-defined as: (Climate model - Constant Model)/(Full model - Constant
-model). The contribution for survival and colonization, where we could
-not estimate a full model with year random effects at the quadrat level,
-is defined as: (Constant Model - Climate Model)/Constant Model.}
-\end{figure}
-
-\begin{figure}[htbp]
-\centering
-\includegraphics{components/figure/manuscript-figure_2.pdf}
-\caption{Posterior means (vertical ticks), 75\% credible intervals
-(heavy lines), and 95\% credible intervals (light lines) of climate
-effects on growth at both levels of inferences. The dashed vertical line
-is at 0, indicating no effect. Horizontal line at 0 indicates perfect
-agreement between mean observed cover in that year and the model
-predictions.}
-\end{figure}
-
-\begin{figure}[htbp]
-\centering
-\includegraphics{components/figure/manuscript-figure_3.pdf}
 \caption{Boxplots of model residuals for one-step-ahead forecasts at
@@
 \centering
-\includegraphics{components/figure/manuscript-figure_4.pdf}
+\includegraphics{components/figure/manuscript-figure_2.pdf}
 \caption{Simulated changes species' median cover caused by a 1\%
@@
 
+R Development Core Team. 2013. R Development Core Team. R: A Language
+and Environment for Statistical Computing.
+
 Su, Y., and M. Yajima. 2012. R2jags: A Package for Running jags from R.
@@
 
-Team, R. 2013. R Development Core Team. R: A Language and Environment
-for Statistical Computing.
-
 \end{document}

### 2015-01-07 manuscript/suppMat.Rmd
updated with supp mat

---
layout: 12pt
header-includes:
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{setspace}
   - \doublespacing
bibliography: ~/Dropbox/Bibliography/MicroMeso.bib
csl: components/ecology.csl

## rmarkdown render options
output:
  pdf_document:
    fig_caption: true
    keep_tex: true
fontsize: 12pt
geometry: margin=1in

---

Supplementary Information: Do we need detailed demographic data to forecast population responses to climate change?
========================================================================================

### Andrew T. Tredennick and Peter B. Adler

*Andrew T. Tredennick ([atredenn@gmail.com](mailto:atredenn@gmail.com)), Department of Wildland Resources and the Ecology Center, Utah State University, Logan, UT*

*Peter B. Adler, Department of Wildland Resources and the Ecology Center, Utah State University, Logan, UT*

```{r caching, include=FALSE}
library("methods")
library("knitr")
basename <- "manuscript"
opts_chunk$set(fig.path = paste("components/figure/", basename, "-", sep=""),
               cache.path = paste("components/cache/", basename, "/", sep=""))
opts_chunk$set(cache = 2)
opts_chunk$set(tidy=FALSE, warning=FALSE, message=FALSE, 
               comment = NA, verbose = TRUE, echo=FALSE)

# PDF-based figures
opts_chunk$set(dev='pdf')
```

```{r plot-options, message=FALSE, warning=FALSE, include=FALSE, echo=FALSE}
cbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", 
               "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

opts_chunk$set(dev='pdf', dev.args=c(version="1.3"))

```

```{r figure_S1, dependson="plot-options", fig.cap="The proportion of interannual variability in  vital rates explained by the climate covariates. The contribution for growth is defined as: (Climate model - Constant Model)/(Full model - Constant model). The contribution for survival and colonization, where we could not estimate a full model with year random effects at the quadrat level, is defined as: (Constant Model - Climate Model)/Constant Model.", fig.width=7, fig.height=3, cache=FALSE, results='hide'}
deviance <- data.frame(VitalRate = c("Growth", "Survival", "Colonization"),
                       ConstantModel = rep(NA,3),
                       ClimateModel = rep(NA,3),
                       FullModel = rep(NA,3),
                       ClimateContribution = rep(NA,3))
deviance$FullModel[1] <- round(read.csv("../montana/quadBM/vitalRateRegressions/growth/growthDeviance.csv")[,2],2)
deviance$ClimateModel[1] <- round(read.csv("../montana/quadBM/vitalRateRegressions/growth/growthDevianceCLIMATE.csv")[,2],2)
deviance$ConstantModel[1] <- round(read.csv("../montana/quadBM/vitalRateRegressions/growth/growthDevianceCONSTANT.csv")[,2],2)
deviance$ClimateContribution[1] <- with(deviance[1,], (ClimateModel-ConstantModel)/(FullModel-ConstantModel))
deviance$ClimateModel[2] <- round(read.csv("../montana/quadBM/vitalRateRegressions/survival/survivalDeviance.csv")[,2],2)
deviance$ConstantModel[2] <- round(read.csv("../montana/quadBM/vitalRateRegressions/survival/survivalDevianceCONSTANT.csv")[,2],2)
deviance$ClimateModel[3] <- round(read.csv("../montana/quadBM/vitalRateRegressions/colonization/colonizationDeviance.csv")[,2],2)
deviance$ConstantModel[3] <- round(read.csv("../montana/quadBM/vitalRateRegressions/colonization/colonizationDevianceCONSTANT.csv")[,2],2)
deviance$ClimateContribution[2:3] <- with(deviance[2:3,], (ConstantModel-ClimateModel)/(ConstantModel))

colnames(deviance) <- c("Vital Rate", "Constant model", "Climate model", "Full model", "Contribution of climate covariates")
library(ggplot2)
devGQ <- data.frame(VitalRate = c("Growth", "Survival", "Colonization"),
                       ConstantModel = rep(NA,3),
                       ClimateModel = rep(NA,3),
                       FullModel = rep(NA,3),
                       ClimateContribution = rep(NA,3))
devGQ$FullModel[1] <- round(read.csv("../montana/ipm/vitalRateRegs/growth/growthDeviance.csv")[,2],2)
devGQ$ClimateModel[1] <- round(read.csv("../montana/ipm/vitalRateRegs/growth/growthDevianceCLIMATE.csv")[,2],2)
devGQ$ConstantModel[1] <- round(read.csv("../montana/ipm/vitalRateRegs/growth/growthDevianceCONSTANT.csv")[,2],2)
devGQ$FullModel[2] <- round(read.csv("../montana/ipm/vitalRateRegs/survival/survivalDeviance.csv")[,2],2)
devGQ$ClimateModel[2] <- round(read.csv("../montana/ipm/vitalRateRegs/survival/survivalDevianceCLIMATE.csv")[,2],2)
devGQ$ConstantModel[2] <- round(read.csv("../montana/ipm/vitalRateRegs/survival/survivalDevianceCONSTANT.csv")[,2],2)
devGQ$FullModel[3] <- round(read.csv("../montana/ipm/vitalRateRegs/recruitment/recruitmentDeviance.csv")[,2],2)
devGQ$ClimateModel[3] <- round(read.csv("../montana/ipm/vitalRateRegs/recruitment/recruitmentDevianceCLIMATE.csv")[,2],2)
devGQ$ConstantModel[3] <- round(read.csv("../montana/ipm/vitalRateRegs/recruitment/recruitmentDevianceCONSTANT.csv")[,2],2)
devGQ$ClimateContribution[1] <- with(devGQ[1,], (ClimateModel-ConstantModel)/(FullModel-ConstantModel))
devGQ$ClimateContribution[2] <- with(devGQ[2,], (ConstantModel-ClimateModel)/(ConstantModel))
devGQ$ClimateContribution[3] <- with(devGQ[3,], (ConstantModel-ClimateModel)/(ConstantModel))

devM <- devGQ[,c(1,5)]
devM$Level <- rep("Individual", 3)
tmpD <- deviance[,c(1,5)]
tmpD$Level <- rep("Quadrat", 3)
colnames(tmpD)[2] <- "value"
colnames(tmpD)[1] <- "VitalRate"
colnames(devM)[2] <- "value"
plotD <- rbind(tmpD, devM)
gClimCont1 <- ggplot(data=subset(plotD, Level=="Individual"), aes(x=VitalRate, y=value))+
  geom_bar(stat="identity", position="dodge", fill="grey30")+
  ylab("Contribution of Climate Effects")+
  xlab("Vital Rate")+
  ggtitle("Individual level")+
  scale_y_continuous(limits=c(0, 0.5))+
  scale_x_discrete(labels=c("Recruitment", "Growth", "Survival"))+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 25, hjust = 1))
gClimCont2 <- ggplot(data=subset(plotD, Level=="Quadrat"), aes(x=VitalRate, y=value))+
  geom_bar(stat="identity", position="dodge", fill="grey30")+
  ylab("Contribution of Climate Effects")+
  xlab("Vital Rate")+
  ggtitle("Quadrat level")+
  scale_y_continuous(limits=c(0, 0.5))+
  scale_x_discrete(labels=c("Colonization", "Cover change", "Extirpation"))+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 25, hjust = 1))
library(gridExtra)
gClim <- grid.arrange(gClimCont1, gClimCont2, ncol=2, nrow=1)
gClim

```

```{r figure_S2, dependson="plot-options", fig.cap="Posterior means (vertical ticks), 75% credible intervals (heavy lines), and 95% credible intervals (light lines) of climate effects on growth at both levels of inferences. The dashed vertical line is at 0, indicating no effect.", fig.width=8, fig.height=7, cache=FALSE, results='hide'}
gQC <- read.csv("../montana/quadBM/vitalRateRegressions/growth/growthQuants.csv")
gQC <- gQC[129:144,]
coefNames <- c("ppt1", "ppt2", "temp1", "temp2")
gQC$X <- rep(coefNames, each=4)
gQC$rate <- "growth"
sQC <- read.csv("../montana/quadBM/vitalRateRegressions/survival/survivalQuants.csv")
sQC <- sQC[33:48,]
coefNames <- c("ppt1", "ppt2", "temp1", "temp2")
sQC$X <- rep(coefNames, each=4)
sQC$rate <- "survival"
rQC <- read.csv("../montana/quadBM/vitalRateRegressions/colonization/colonizationQuants.csv")
rQC <- rQC[29:44,]
coefNames <- c("ppt1", "ppt2", "temp1", "temp2")
rQC$X <- rep(coefNames, each=4)
rQC$rate <- "colonization"
plotQC <- rbind(gQC,sQC,rQC)
gQ <- ggplot(plotQC, aes(x=species, y=X50., color=rate))+
  geom_hline(aes(yintercept=0), linetype="dashed")+
  geom_errorbar(aes(ymin=X2.5., ymax=X97.5.), width=0, position=position_dodge(width=0.5))+
  geom_errorbar(aes(ymin=X25., ymax=X75.), width=0, size=1.25, position=position_dodge(width=0.5))+
  geom_point(size=4, position=position_dodge(width=0.5), shape="|")+
  ylab("Standardized Coefficient Value")+
  xlab("Species")+
  facet_grid(X~.)+
  coord_flip()+
  theme_bw()+
  scale_color_manual(values=c("grey25", "steelblue", "slateblue"))+
  guides(color=guide_legend(title=""))+
  theme(legend.position="top")+
  ggtitle("Quadrat level")

gIC <- read.csv("../montana/ipm/vitalRateRegs/growth/growthQuants.csv")
gIC <- gIC[133:148,]
coefNames <- c("ppt1", "ppt2", "temp1", "temp2")
gIC$X <- rep(coefNames, each=4)
gIC$rate <- "growth"
sIC <- read.csv("../montana/ipm/vitalRateRegs/survival/survivalQuants.csv")
sIC <- sIC[133:148,]
coefNames <- c("ppt1", "ppt2", "temp1", "temp2")
sIC$X <- rep(coefNames, each=4)
sIC$rate <- "survival"
rIC <- read.csv("../montana/ipm/vitalRateRegs/recruitment/recruitmentQuants.csv")
rIC <- rIC[81:96,]
coefNames <- c("ppt1", "ppt2", "temp1", "temp2")
rIC$X <- rep(coefNames, each=4)
rIC$rate <- "colonization"
plotIC <- rbind(gIC,sIC,rIC)
gI <- ggplot(plotIC, aes(x=species, y=X50., color=rate))+
  geom_hline(aes(yintercept=0), linetype="dashed")+
  geom_errorbar(aes(ymin=X2.5., ymax=X97.5.), width=0, position=position_dodge(width=0.5))+
  geom_errorbar(aes(ymin=X25., ymax=X75.), width=0, size=1.25, position=position_dodge(width=0.5))+
  geom_point(size=4, position=position_dodge(width=0.5), shape="|")+
  ylab("Standardized Coefficient Value")+
  xlab("Species")+
  facet_grid(X ~ .)+
  coord_flip()+
  theme_bw()+
  scale_color_manual(values=c("grey25", "steelblue", "slateblue"),
                     labels=c("recruitment", "growth", "survival"))+
#   guides(color=FALSE)+
  guides(color=guide_legend(title=""))+
  theme(legend.position="top")+
  ggtitle("Individual level")

####
#### Plot just for growth
####
growthIC <- subset(plotIC, rate=="growth")
growthQC <- subset(plotQC, rate=="growth")
growthIC$Level <- "Individual"
growthQC$Level <- "Quadrat"
growthAll <- rbind(growthIC, growthQC)

gGrowth <- ggplot(growthAll, aes(x=species, y=X50., color=Level))+
  geom_hline(aes(yintercept=0), linetype="dashed")+
  geom_errorbar(aes(ymin=X2.5., ymax=X97.5.), width=0, position=position_dodge(width=0.5))+
  geom_errorbar(aes(ymin=X25., ymax=X75.), width=0, size=1.25, position=position_dodge(width=0.5))+
  geom_point(size=4, position=position_dodge(width=0.5), shape="|")+
  ylab("Standardized Coefficient Value")+
  xlab("Species")+
  facet_grid(X~.)+
  coord_flip()+
  theme_bw()+
  scale_color_manual(values=c("grey25", "slateblue"))+
#   guides(color=FALSE)+
  guides(color=guide_legend(title=""))+
  theme(legend.position="top")
# print(gGrowth)

####
#### Print figure
####
gF2 <- grid.arrange(gQ, gI, ncol=2, nrow=1)
```

### 2015-01-07 manuscript/suppMat.tex
updated with supp mat

--- 2015-01-06 manuscript/manuscript.tex
@@
 
-\section{Do we need detailed demographic data to forecast population
-responses to climate
-change?}\label{do-we-need-detailed-demographic-data-to-forecast-population-responses-to-climate-change}
+\section{Supplementary Information: Do we need detailed demographic data
+to forecast population responses to climate
+change?}\label{supplementary-information-do-we-need-detailed-demographic-data-to-forecast-population-responses-to-climate-change}
 
@@
 
-\subsection{Abstract}\label{abstract}
-
-\emph{Keywords}: forecasting, climate change, grassland, integral
-projection model
-
-\subsection{Introduction}\label{introduction}
-
-Population models are important tools for predicting the impacts of
-environmental change on species. But reconciling the scales at which
-population models are parameterized and the scales at which
-environmental changes play out remains a challenge (Freckleton et al.
-2011, Queenborough et al. 2011). The major hurdle is that most
-population models, at least for plant species, are built using data from
-small, localized plots because parameterizing traditional population
-models requires tracking the fates of individuals. These models are
-difficult to scale up from the micro to meso-scales because the fitted
-parameters do not fully represent the spatial variation present at
-scales beyond that at which the data are collected (Sæther et al. 2007).
-Thus, our ability to use population models to predict the consequences
-of climate change is limited when we rely on individual-level data.
-
-Aggregate measures of individual plant performance, such as those
-typically collected as part of large-scale census efforts, offer an
-alternative to detailed demographic data for modeling populations (Clark
-and Bjørnstad 2004, Freckleton et al. 2011). Such population-level data
-will never match the precision of individual-level data, but it is more
-feasible to attain a broad coverage sample when collecting coarse-scale
-data. This presents a difficult trade-off: on the one hand,
-individual-level data leads to more reliable models; on the other hand,
-population-level data leads to models that will produce less precise
-predicitons but can be applied over greater spatial and temporal
-extents. If the goal of the model is to predict the consequences of
-climate change, then we argue that in the face of limited time and money
-population-level data are the best option. However, the performance of
-models based on population-level data have rarely been compared to that
-of models based on individual-level data.
-
-To date, relatively few studies have tried to model populations based on
-data other than detailed individual-level data. An important exception
-is an effort by Taylor and Hastings (2004) to model the population
-growth rate of an invasive species to investigate the best strategies
-for invasion control. They used a ``density-structured'' model where the
-state variable is a discrete density state rather than a continuous
-density measure. Building on this work, Freckleton et al. (2011) showed
-that density-structured models compare well to continuous models in
-theory, and Queenborough et al. (2011) showed the application of such
-methods in a study on arable weeds. In particular, Queenborough et al.
-(2011) provide emprical evidence that density-structured models are
-capable of reproducing population dynamics, even if some precision is
-lost when compared to fully continuous models. Thus, population models
-based on coarse, population-level data show promise for producing
-ecological forecasts at landscape and regional scales (Queenborough et
-al. 2011). However, none of these models included environmental
-covariates.
-
-Basing population models on population- rather than individual-level
-data in a climate change context is hampered by the fact that it is
-individuals that respond to climate, not populations (Clark et al.
-2012). This fact puts us in uneasy proximity to an ``ecological
-fallacy'' where one deduces inference on the individual from statistical
-inference on the group (Piantadosi et al. 1988).
-
-\subsection{Materials and Methods}\label{materials-and-methods}
-
-\subsubsection{Study site and data}\label{study-site-and-data}
-
-Our demographic data comes from the Fort Keogh Livestock and Range
-Research Laboratory in eastern Montana's northern mixed prairie near
-Miles City, Montana, USA ($46^{\circ}$ 19' N, $105^{\circ}$ 48' W). The
-dataset is freely available on Ecological Archives (Anderson et al.
-2011), and interested readers should refer to the metadata therin for a
-complete description. The site is about 800 m above sea level and mean
-annual precipitation (1878-2009) is 334 mm, with most annual
-precipitaion falling from April through September. The site is grass
-dominated and, for the purposes of our study, we focus on the four most
-abundant graminoid species: \emph{Bouteloua gracilis} (BOGR),
-\emph{Hesperostipa comata} (HECO), \emph{Pascopyrum smithii} (PASM), and
-\emph{Poa secunda} (POSE).
-
-From 1932 to 1945 individual plants were identified and mapped annualy
-in 44 1-$\text{m}^2$ quadrats using a pantograph. The quadrats were
-distributed in six pastures, each assigned a grazing treatment of light
-(1.24 ha/animal unit month), moderate (0.92 ha/aum), and heavy (0.76
-ha/aum) stocking rates (two pastures per treatment). In this analysis we
-account for potential differences among the grazing treatments, but do
-not focus on grazing$\times$climate interactions. The annual maps of the
-quadrats were digitized and the fates of individual plants tracked and
-extracted using a computer program. Daily climate data, which we
-aggregated into climate variables of interest, are available for the
-duration of the data collection period (1932 - 1945) from the Miles City
-airport, Wiley Field, 9 km from the study site.
-
-In this paper, we model populations based on two levels of data:
-individual and quadrat. The individual data is the ``raw'' data. For the
-quadrat level we data we simply sum individual areal cover for each
-quadrat by species. This is equivalent to a perfect census of quadrat
-percent cover, so we do not need to consider measurement error. Based on
-these two datasets we can compare population models built using
-individual level data and aggregated quadrat level data.
-
-\subsubsection{Stastical models of vital
-rates}\label{stastical-models-of-vital-rates}
-
-At both levels of inference (individual and quadrat), the building
-blocks of our population models are vital rate regressions. For
-individual level data we fit models for survival, growth, and
-recruitment of new individuals for each species. At the quadrat level we
-fit analagous models of extinction probability, percent cover
-increase/decrease, and quadrat colonization for each species. We
-describe the statistical models seprately since fitting the models
-required different approaches at the individual and quadrat levels. All
-models contain four climate covariate that we chose \emph{a priori}:
-fall through spring precipitation at \emph{t}-1 and \emph{t}-2 (ppt1 and
-ppt2, respectively) and mean spring temperature at \emph{t}-1 and
-\emph{t}-2 (TmeanSpr1 and TmeanSpr2, respectively), where \emph{t} is
-the observation year.
-
-We fit all models using a hierarchical Bayesian approach, which we
-describe in more detail below. However, for each vital rate statistical
-model we also define the likelihood model we use. For the likelihood
-models, Y is always the relevant vector of observations (e.g., whether a
-genet survived {[}1{]} or not {[}0{]} from year $t$ to $t+1$).
-
-\paragraph{Vital rate models: individual
-level}\label{vital-rate-models-individual-level}
-
-We used logistic regression to model survival probability ($S$) of genet
-$i$ from species $j$ in quadrat group $Q$ from time $t$ to $t+1$:
-
-\begin{align}
-\text{logit}(S_{ijQ,t}) &= \gamma^{S}_{j,t} + \phi^{S}_{jQ} + \beta^{S}_{j,t}x_{ij,t} + \omega^{S}_{j}w_{ij,t} + \theta^{S}_{jk}C_{k,t} + \varepsilon^{S}_{t} \\
-Y^{S}_{ijQ,t} &\sim \text{Bernoulli}(S_{ijQ,t})
-\end{align}
-
-where $x_{ij,t}$ is the log of genet size, $\gamma^{S}_{j,t}$ is a
-year-specific intercept, $\beta^{S}_{j,t}$ is the year-specific slope
-parameter for size, $\phi^{S}_{jQ}$ is the random effect of quadrat
-group location, and $\theta^{S}_{k}$ is the fixed parameter for the
-effect of the $k$th climate covariate at time $t$ ($C_{k,t}$). We
-include density-dependence by estimating the effect of crowding on the
-focal individual by other individuals of the same species. $\omega$ is
-the effect of crowding and $w_{t,Q}$ is the crowding experienced by the
-focal individual at time $t$ in quadrat group $Q$.
-
-We modeled growth as gaussian process describing genet size at time
-$t+1$ as a function of size at $t$ and climate covariates:
-
-\begin{align}
-x_{ijQ,t+1} &= \gamma^{G}_{j,t} + \phi^{G}_{jQ} + \beta^{G}_{j,t}x_{ij,t} + \omega^{G}_{j}w_{ij,t} + \theta^{G}_{jk}C_{k,t} \\
-Y^{G}_{ijQ,t} &\sim \text{Normal}(x_{ijQ,t+1}, \sigma_{j})
-\end{align}
-
-where $x$ is log genet size and all other paramters are as described for
-the survival regression.
-
-Our data allows us to track new recruits, but we cannot assign a
-specific parent to new genets. So, for recruitment, we work at the
-quadrat level and model the number of new individuals of species $j$ in
-quadrat $q$ recruiting at time $t+1$ as a function of quadrat
-``effective cover'' ($A'$) in the previous year ($t$). Effective cover
-is a mixture of observed cover ($A$) in the focal quadrat ($q$) and the
-mean cover across the entire group ($\bar{A}$) of $Q$ quadrats in which
-$q$ is located:
-
-\begin{equation}
-A'_{jq,t} = p_{j}A_{jq,t} + (1-p_{j})\bar{A}_{jQ,t}
-\end{equation}
-
-where $p$ is a mixing fraction between 0 and 1 that is estimated within
-the model.
-
-We assume the number of individuals, $Y^{R}$, recruiting at time $t+1$
-follows a negative binomial distribution:
-
-\begin{equation}
-Y^{R}_{jq,t+1} \sim \text{NegBin}(\lambda_{jq,t+1},\zeta)
-\end{equation}
-
-where $\lambda$ is the mean intensity and $\zeta$ is the size parameter.
-We define $\lambda$ as:
-
-\begin{equation}
-\lambda_{jq,t+1} = A'_{jq,t}e^{(\gamma^{R}_{j,t} + \phi^{R}_{jQ} + \theta^{R}_{jk}C_{k,t} + \omega^{R}\sqrt{A'_{q,t}})}
-\end{equation}
-
-where $A'$ is effective cover ($\text{cm}^2$) of species $j$ in quadrat
-$q$ and all other terms are as in the survival and growth regressions.
-
-\paragraph{Vital rate models: quadrat
-level}\label{vital-rate-models-quadrat-level}
-
-At the quadrat level we defined three vital rates:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\itemsep1pt\parskip0pt\parsep0pt
-\item
-  Probability of extirpation ($S$): the probability that, for a given
-  species, a particular quadrat will go from non-zero cover at time $t$
-  to zero cover at time $t+1$.
-\item
-  Cover change ($G$): the change in percent cover from time $t$ to $t+1$
-  for a given species within a particular quadrat.
-\item
-  Probability of colonization ($R$): the probability that, for a give
-  species, a particular quadrat will go from zero cover at time $t$ to
-  non-zero cover at time $t+1$.
-\end{enumerate}
-
-We retain the abbreviations $S$, $G$, and $R$ from the analagous
-processes at the individual level. The vital rate models at the quadrat
-level are all based on quadrat proportional cover. Also, at the quadrat
-level we do not need to explicitly include a density dependent term.
-Since we are modeling proportional cover, we essentially get
-density-dependence for ``free'' when proportional cover is included as a
-covariate. That is, density-dependence emerges because quadrats with low
-cover gerenally increase in cover whereas quadrats with high cover
-generally decrease in cover.
-
-We modeled the probability of extirpation of species $j$ in quadrat $q$
-from time $t$ to $t+1$ as:
-
-\begin{align}
-\text{logit}(S_{jq,t}) &= \gamma^{S}_{j} + \phi^{S}_{jQ} + \beta^{S}_{j}x_{jq,t} + \theta^{S}_{jk}C_{k,t} \\
-Y^{S}_{jq,t} &\sim \text{Bernoulli}(S_{jq,t})
-\end{align}
-
-where all parameters are as in Eq. 1 except that $x$ is now quadrat
-proportional cover. Note, however, that we do not include year random
-effects on the intercept $\gamma$ or the slope term $\beta$ for quadrat
-proportional cover. The quadrat data is inherently more sparse than the
-individual data from which it is aggregated, and this is especially
-evident when modeling rare events like extirpation and colonization.
-Thus, when we tried to fit random year effects, those terms did not
-converge.
-
-We modeled quadrat cover change ($G$) from time $t$ to $t+1$ as:
-
-\begin{equation}
-\text{logit}(x_{jq,t+1}) = \gamma^{G}_{j,t} + \phi^{G}_{jQ} + \beta^{G}_{j,t}x_{jq,t} + \theta^{S}_{jk}C_{k,t}
-\end{equation}
-
-where, in this case, we do include random year effects on the intercept
-$\gamma$ and the slope term $\beta$. For cover change we had enough data
-for those terms to converge. Note that our model for quadrat cover
-change uses a logit transformation to link the expected cover at $t+1$
-($x_{jq,t}$) to the linear predictors. We do so because during model
-fitting we use a beta likelihood since the data, proportional cover, is
-beta distributed. The beta likelihood requires shape ($\rho$) and rate
-($\eta$) parameters that can be caclulated using moment-matching:
-
-\begin{align}
-\rho_{jq,t+1} &= x_{jq,t+1}\tau_{j} \\
-\eta_{jq,t+1} &= (1-x_{jq,t+1})\tau_{j}
-\end{align}
-
-with likelihood:
-
-\begin{equation}
-Y^{G}_{jq,t+1} \sim \text{Beta}(\rho_{jq,t+1}, \eta_{jq,t+1}).
-\end{equation}
-
-Finally, we modeled probability of colonization in quadrat $q$ by
-species $j$ from time $t$ to $t+1$ as:
-
-\begin{align}
-\text{logit}(R_{jq,t}) &= \gamma^{R}_{j} + \phi^{R}_{jQ} + \theta^{R}_{jk}C_{k,t} \\
-Y^{R}_{jq,t} &\sim \text{Bernoulli}(R_{jq,t}).
-\end{align}
-
-\subsubsection{Model fitting}\label{model-fitting}
-
-Our Bayesian approach to fitting the vital rate models required choosing
-appropriate priors for unknown parameters and deciding which, if any, of
-those prior should be hierarchical. We decided to fit models where all
-terms except climate covariates were fit by species, while the climate
-covariates were fit heierarchically where species-specific coefficients
-were drawn from a shared `global' coefficient distribution. We did so
-for two reasons: (1) the four focal species are all perrennial grasses
-that we expect to respond similarly to climate covariates, and (2)
-convergence of climate effects at the quadrat level was much easier to
-achieve when we modeled these terms hierarchically, allowing them to
-``share'' statistical strenght via partial pooling (Gelman and Hill
-2007). So, climate effects were modeled as:
-
-\begin{align}
-\theta_{jk} &\sim \text{Normal}(\bar{\theta_{k}}, \sigma_{k})
-\end{align}
-
-where $\bar{\theta_{k}}$ is the interspecific effect of the $k$th
-climate covariate.
-
-We used uninformative priors for all unknown parameters, specifically:
-
-\begin{align}
-\boldsymbol{\gamma, \beta, \bar{\theta}} &\sim \text{Normal}(0, 1e^{-6}) \\
-\boldsymbol{\phi} &\sim \text{Normal}(0, \boldsymbol{\sigma_{\phi}}) \\
-\boldsymbol{\sigma_{\phi}} &\sim e^{(\text{Gamma}(2, 0.5))} \\
-\boldsymbol{\sigma_{\theta}, \sigma_{\gamma}, \sigma_{\beta}, \tau, \zeta} &\sim \text{Gamma}(0.001, 0.001)
-\end{align}
-
-All of our analyses (model fitting and simulating) were conducted in R
-(Team 2013). We used the MCMC sampler in JAGS (Plummer 2003) to estimate
-the posterior distributions of model parameters and the package `r2jags'
-(Su and Yajima 2012) to connect R to JAGS. We obtained posterior
-distributions for all model parameters from three parallel MCMC chains
-run for 50,000 iterations, after discarding an initial 50,000
-iterations. We assessed convergence visually and using the Gelman and
-Rubin (1992) diagnostic in the R package `coda' (Plummer et al. 2006).
-Scale reduction factors for all parameters were less than 1.02,
-indicating convergence. For the purposes of introducing stochasticity in
-our population models, we saved the final 1,000 iterations from each
-chain for all parameters to be used as randomly drawn values during
-population simulation.
-
-We assessed the statistical importance of the climate covariates
-included the final vital rate regressions by comparing the residual
-deviance of models with climate covariates and temporal random effects,
-climate covariates only, and temporal random effects only. When a model
-includes climate covariates, this comparison shows the relative
-contribution of the climate covariates in explaining the total
-interannual variability (Adler et al. 2012).
-
-\subsubsection{Population models}\label{population-models}
-
-With the posterior distribution of the vital rate statistical models in
-hand, it is straightforward to simulate the population models. We used
-an Integral Projection Model (IPM) to model populations based on
-individual level data and an quadrat based version of an
-individually-based model (Quadrat-Based Model, QBM) to model populations
-based on quadrat level data. Both models take the general form:
-
-\begin{equation}
-N_{t+1} = S \times G + R.
-\end{equation}
-
-So, at each time step in a simulation, we use the survival regression to
-determine if each genet lives or not (if each quadrat remains occupied
-or not), the growth regression to determine size changes of surviving
-individuals (cover change of occupied quadrats), and the recruitment
-regression to determine the number of new recruits (if a quadrat is
-colonized or not). We first use one-step-ahead forecasts to assess each
-model's ability to reproduce observed cover changes. Then we use the
-models to analyze the effect of potential climate changes on population
-size over long time scales.
-
-We used random draws from the final 1,000 iterations from each of 3 MCMC
-chains to introduce stochasticity into our population models. At each
-time step, we first randomly selected climate covariates from one of the
-14 observed years and also randomly drew one set of random year effects
-for each vital rate regression. Then, we drew the full parameter set
-(specific random year effects, climate effects, fixed effects) from a
-randomly selected MCMC iteration. We selected random year effects and
-the climate year independently.
-
-\subsection{Results}\label{results}
-
-\subsubsection{Climate effects on vital
-rates}\label{climate-effects-on-vital-rates}
-
-At the individual level we fit vital rate models that included climate
-covariates and random year effects (full model), just climate covariates
-(climate model), and no climate or random year effects (constant model).
-We used the metric from Adler et al. (2012) to quantify the contribution
-of climate covariates to explaining interannual variability in vital
-rates based on comparisons of model deviance: (Climate model - Constant
-Model)/(Full model - Constant model). At the quadrat level, we could
-only fit random year effects for the growth regression. This is, in
-fact, one of the limitation of census-based data: local extirpation and
-colonization events are likely to be rare, making it difficult to fit
-complex statistical models. So, in this section we focus on growth only.
-
-Climate covariates improved the growth regressions at both levels, with
-a slghty greater contribution at the individual level (Figure 1).
-Parameter estimates were generally consistent in terms of sign and
-magnitude between the individual level and the quadrat level for each
-species and climate covariate (Figure 2). There is no clear pattern
-suggesting that parameters estimated using individual level data are
-more certain, as evidenced by credible interval widths, than estimates
-based on quadrat data.
-
-\subsubsection{Climate effects on population
-dynamics}\label{climate-effects-on-population-dynamics}
-
-\subsection{Discussion}\label{discussion}
-
-We sought to test the assumption that the sensitivities of plant
-populations to climate variables can be detected equally well using
-either individual level data or population level data. This is an
-important question to answer because population models are key tools for
-predicting the consequences of global climate change. However, they can
-be of limited use when built on data from a small subset of a population
-in space or time. If population level data (i.e., some aggregated form
-of individual level data) can be used to detect climate effects on
-population dynamics, then we would have a cheaper and easier option for
-data collection over relatively large temporal and spatial extents
-(\emph{e.g.} Freckleton et al. 2011).
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\itemsep1pt\parskip0pt\parsep0pt
-\item
-  Climate detections
-\item
-  Limitations:
-\end{enumerate}
-
-\begin{itemize}
-\itemsep1pt\parskip0pt\parsep0pt
-\item
-  Sample size at quadrat level
-\item
-  are these species inherently sensitive to climate (compare to Chu and
-  Adler in review -- say why we chose Montana site -- most sensitive to
-  climate variables)
-\item
-  we know recruitment is important, and this is something that may be
-  easily missed at quad level
-\end{itemize}
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\setcounter{enumi}{2}
-\itemsep1pt\parskip0pt\parsep0pt
-\item
-  Implications
-\end{enumerate}
-
-\subsection{References}\label{references}
-
-\subsection{Figures}\label{figures}
-
 \begin{figure}[htbp]
 \centering
-\includegraphics{components/figure/manuscript-figure_1.pdf}
+\includegraphics{components/figure/manuscript-figure_S1.pdf}
 \caption{The proportion of interannual variability in vital rates
@@
 \centering
-\includegraphics{components/figure/manuscript-figure_2.pdf}
+\includegraphics{components/figure/manuscript-figure_S2.pdf}
 \caption{Posterior means (vertical ticks), 75\% credible intervals
@@
 effects on growth at both levels of inferences. The dashed vertical line
-is at 0, indicating no effect. Horizontal line at 0 indicates perfect
-agreement between mean observed cover in that year and the model
-predictions.}
+is at 0, indicating no effect.}
 \end{figure}
 
-\begin{figure}[htbp]
-\centering
-\includegraphics{components/figure/manuscript-figure_3.pdf}
-\caption{Boxplots of model residuals for one-step-ahead forecasts at
-each observation year. Each one-step forecast was simulated
-\texttt{r nSims} times. Note that the y-axes vary across panels. The
-light blue line shows the difference between the observed-year percent
-cover and the average cover observed across all years. The models tend
-to underpredict and perform poorly when observed cover in a given year
-is a large deviant from the mean.}
-\end{figure}
-
-\begin{figure}[htbp]
-\centering
-\includegraphics{components/figure/manuscript-figure_4.pdf}
-\caption{Simulated changes species' median cover caused by a 1\%
-increase in observed precipitation or temperature. We used the median of
-simulated results because the average was highly influence by some
-extreme outliers.}
-\end{figure}
-
-Adler, P. B., H. J. Dalgleish, and S. P. Ellner. 2012. Forecasting plant
-community impacts of climate variability and change: when do competitive
-interactions matter? Journal of Ecology 100:478--487.
-
-Anderson, J., L. Vermeire, and P. B. Adler. 2011. Fourteen years of
-mapped, permanent quadrats in a northern mixed prairie, USA. Ecology
-92:1703.
-
-Clark, J. S., and O. N. Bjørnstad. 2004. Population time series: Process
-variability, observation errors, missing values, lags, and hidden
-states. Ecology 85:3140--3150.
-
-Clark, J. S., D. M. Bell, M. Kwit, A. Stine, B. Vierra, and K. Zhu.
-2012. Individual-scale inference to anticipate climate-change
-vulnerability of biodiversity.
-
-Freckleton, R. P., W. J. Sutherland, A. R. Watkinson, and S. A.
-Queenborough. 2011. Density-structured models for plant population
-dynamics. American Naturalist 177:1--17.
-
-Gelman, A., and J. Hill. 2007. Data analysis using regression and
-multilevel/hierarchical models. Page xxii, 625p.
-
-Gelman, A., and D. B. Rubin. 1992. Inference from Iterative Simulation
-Using Multiple Sequences.
-
-Piantadosi, S., D. P. Byar, and S. B. Green. 1988. The Ecological
-Fallacy. American Journal of Epidemiology 127:893--904.
-
-Plummer, M. 2003. JAGS: A Program for Analysis of Bayesian Graphical
-Models Using Gibbs Sampling. Pages 20--22 \emph{in} Proceedings of the
-3rd international workshop on distributed statistical computing (dSC
-2003). march.
-
-Plummer, M., N. Best, K. Cowles, and K. Vines. 2006. CODA: Convergence
-Diagnosis and Output Analysis for MCMC. R News 6:7--11.
-
-Queenborough, S. A., K. M. Burnet, W. J. Sutherland, A. R. Watkinson,
-and R. P. Freckleton. 2011. From meso- to macroscale population
-dynamics: A new density-structured approach. Methods in Ecology and
-Evolution 2:289--302.
-
-Su, Y., and M. Yajima. 2012. R2jags: A Package for Running jags from R.
-http://CRAN. R-project. org/package= R2jags.
-
-Sæther, B. E., S. Engen, V. Grøtan, W. Fiedler, E. Matthysen, M. E.
-Visser, J. Wright, A. P. Møller, F. Adriaensen, H. {Van Balen}, D.
-Balmer, M. C. Mainwaring, R. H. McCleery, M. Pampus, and W. Winkel.
-2007. The extended Moran effect and large-scale synchronous fluctuations
-in the size of great tit and blue tit populations. Journal of Animal
-Ecology 76:315--325.
-
-Taylor, C. M., and A. Hastings. 2004. Finding optimal control strategies
-for invasive species: a density-structured model for Spartina
-alterniflora. Journal of Applied Ecology 41:1049--1057.
-
-Team, R. 2013. R Development Core Team. R: A Language and Environment
-for Statistical Computing.
-
 \end{document}

### 2015-01-07 manuscript/manuscript.Rmd
caption clean up

@@
 Each one-step forecast was simulated `r nSims` times.
-Note that the y-axes vary across panels.
-The light blue line shows the difference between the observed-year percent cover and the average cover observed across all years.
-The models tend to underpredict and perform poorly when observed cover in a given year is a large deviant from the mean.", fig.width=5, fig.height=6, cache=FALSE, results='hide'}
+Note that the y-axes vary across panels.", fig.width=5, fig.height=6, cache=FALSE, results='hide'}
 resD <- readRDS("../montana/quadBM/simulations/quadBM_oneStep_Residuals.rds")

### 2015-01-08 manuscript/manuscript.Rmd
updated intro

@@
 These models are difficult to scale up from the micro to meso-scales because the fitted parameters do not fully represent the spatial variation present at scales beyond that at which the data are collected [@Sæther2007].
+At the same time, most demographic data is collected over short time spans.
+For example, the most common study duration in the COMPADRE matrix population model database is 4 years and only a few exceed 10 years [@Salguero-Gomez2015].
+The constrained spatio-temporal extent of most demographic datasets reflects the difficulty of collecting such data, but those constraints limit our ability to extrapolate population models.
 Thus, our ability to use population models to predict the consequences of climate change is limited when we rely on individual-level data.
@@
 This presents a difficult trade-off: on the one hand, individual-level data leads to more reliable models; on the other hand, population-level data leads to models that will produce less precise predicitons but can be applied over greater spatial and temporal extents.
-If the goal of the model is to predict the consequences of climate change, then we argue that in the face of limited time and money population-level data are the best option.
-However, the performance of models based on population-level data have rarely been compared to that of models based on individual-level data.
+An open question is how well models based on population-level data compare to models based on individual-level data.
 
@@
 
-Here, we test the hypothesis (idea?) that statistical and population models based on aggregated data can detect climate signals as wells as models based on individual-level data.
-We use a unique demographic dataset that tracks the fates of individual plants from four species to build single-species population models, since those are often used tools for ecological forecasts and climate vulnerability assessments.
+Here, we test the assumption that statistical and population models based on aggregated data can detect climate signals as wells as models based on individual-level data.
+We use a unique demographic dataset that tracks the fates of individual plants from four species over 14 years to build single-species population models, since those are often used tools for ecological forecasts and climate vulnerability assessments.
 We first fit population models with interannual variation in vital rates explained, in part, by climate covariates.

### 2015-01-08 manuscript/manuscript.tex
updated intro

@@
 spatial variation present at scales beyond that at which the data are
-collected (Sæther et al. 2007). Thus, our ability to use population
-models to predict the consequences of climate change is limited when we
-rely on individual-level data.
+collected (Sæther et al. 2007). At the same time, most demographic data
+is collected over short time spans. For example, the most common study
+duration in the COMPADRE matrix population model database is 4 years and
+only a few exceed 10 years (Salguero-Gómez et al. 2015). The constrained
+spatio-temporal extent of most demographic datasets reflects the
+difficulty of collecting such data, but those constraints limit our
+ability to extrapolate population models. Thus, our ability to use
+population models to predict the consequences of climate change is
+limited when we rely on individual-level data.
 
@@
 predicitons but can be applied over greater spatial and temporal
-extents. If the goal of the model is to predict the consequences of
-climate change, then we argue that in the face of limited time and money
-population-level data are the best option. However, the performance of
-models based on population-level data have rarely been compared to that
-of models based on individual-level data.
+extents. An open question is how well models based on population-level
+data compare to models based on individual-level data.
 
@@
 
-Here, we test the hypothesis (idea?) that statistical and population
-models based on aggregated data can detect climate signals as wells as
-models based on individual-level data. We use a unique demographic
-dataset that tracks the fates of individual plants from four species to
+Here, we test the assumption that statistical and population models
+based on aggregated data can detect climate signals as wells as models
+based on individual-level data. We use a unique demographic dataset that
+tracks the fates of individual plants from four species over 14 years to
 build single-species population models, since those are often used tools
@@
 All of our analyses (model fitting and simulating) were conducted in R
-(R Development Core Team 2013). We used the MCMC sampler in JAGS
-(Plummer 2003) to estimate the posterior distributions of model
-parameters and the package `r2jags' (Su and Yajima 2012) to connect R to
-JAGS. We obtained posterior distributions for all model parameters from
-three parallel MCMC chains run for 50,000 iterations, after discarding
-an initial 50,000 iterations. We assessed convergence visually and using
-the Gelman and Rubin (1992) diagnostic in the R package `coda' (Plummer
-et al. 2006). Scale reduction factors for all parameters were less than
-1.02, indicating convergence. For the purposes of introducing
-stochasticity in our population models, we saved the final 1,000
-iterations from each chain for all parameters to be used as randomly
-drawn values during population simulation.
+(Team 2013). We used the MCMC sampler in JAGS (Plummer 2003) to estimate
+the posterior distributions of model parameters and the package `r2jags'
+(Su and Yajima 2012) to connect R to JAGS. We obtained posterior
+distributions for all model parameters from three parallel MCMC chains
+run for 50,000 iterations, after discarding an initial 50,000
+iterations. We assessed convergence visually and using the Gelman and
+Rubin (1992) diagnostic in the R package `coda' (Plummer et al. 2006).
+Scale reduction factors for all parameters were less than 1.02,
+indicating convergence. For the purposes of introducing stochasticity in
+our population models, we saved the final 1,000 iterations from each
+chain for all parameters to be used as randomly drawn values during
+population simulation.
 
@@
 each observation year. Each one-step forecast was simulated
-\texttt{r nSims} times. Note that the y-axes vary across panels. The
-light blue line shows the difference between the observed-year percent
-cover and the average cover observed across all years. The models tend
-to underpredict and perform poorly when observed cover in a given year
-is a large deviant from the mean.}
+\texttt{r nSims} times. Note that the y-axes vary across panels.}
 \end{figure}
@@
 
-R Development Core Team. 2013. R Development Core Team. R: A Language
-and Environment for Statistical Computing.
+Salguero-Gómez, R., O. R. Jones, C. R. Archer, Y. M. Buckley, J.
+Che-Castaldo, H. Caswell, D. Hodgson, A. Scheuerlein, D. A. Conde, E.
+Brinks, H. de Buhr, C. Farack, F. Gottschalk, A. Hartmann, A. Henning,
+G. Hoppe, G. Römer, J. Runge, T. Ruoff, J. Wille, S. Zeh, R. Davison, D.
+Vieregg, A. Baudisch, R. Altwegg, F. Colchero, M. Dong, H. de Kroon,
+J.-D. Lebreton, C. J. E. Metcalf, M. M. Neel, I. M. Parker, T. Takada,
+T. Valverde, L. A. Vélez-Espino, G. M. Wardle, M. Franco, and J. W.
+Vaupel. 2015. The compadrePlant Matrix Database: an open online
+repository for plant demography. Journal of Ecology 103:202--218.
 
@@
 
+Team, R. 2013. R Development Core Team. R: A Language and Environment
+for Statistical Computing.
+
 \end{document}

### 2015-01-08 montana/quadBM/simulations/oneStepForecastQBM.R

@@ -21,7 +21,7 @@
 rm(list=ls(all=TRUE))
 
 #Set number of sims per yearly transition
-NumberSimsPerYear <- 2
+NumberSimsPerYear <- 200
 
 ####
 #### Load libraries ----------------------------------
@@ -235,10 +235,10 @@ combD2$coverChangeObs <- with(combD2, percCover*100-lagCover*100)
 combD2$coverChangePred <- with(combD2, cover*100-lagCover*100)
 combD2$resids <- with(combD2, coverChangeObs - coverChangePred)
 resD <- combD2
-library(ggplot2)
-ggplot(data=resD, aes(x=year, y=resids))+
-  geom_boxplot()+
-  facet_grid(species~., scales = "free")
+# library(ggplot2)
+# ggplot(data=resD, aes(x=year, y=resids))+
+#   geom_boxplot()+
+#   facet_grid(species~., scales = "free")
 
 saveRDS(resD, file = "quadBM_oneStep_ResidualsFullModel.rds")
 

### 2015-01-09 manuscript/manuscript.Rmd
updated with correlation explanation

@@
 Such population-level data will never match the precision of individual-level data, but it is more feasible to attain a broad coverage sample when collecting coarse-scale data.
-This presents a difficult trade-off: on the one hand, individual-level data leads to more reliable models; on the other hand, population-level data leads to models that will produce less precise predicitons but can be applied over greater spatial and temporal extents.
+This presents a difficult trade-off: on the one hand, individual-level data leads to more reliable models; on the other hand, population-level data leads to models that will produce less precise predictions but can be applied over greater spatial and temporal extents.
 An open question is how well models based on population-level data compare to models based on individual-level data.
@@
 Building on this work, @Freckleton2011 showed that density-structured models compare well to continuous models in theory, and @Queenborough2011 showed the application of such methods in a study on arable weeds.
-In particular, @Queenborough2011 provide emprical evidence that density-structured models are capable of reproducing population dynamics, even if some precision is lost when compared to fully continuous models.
+In particular, @Queenborough2011 provide empirical evidence that density-structured models are capable of reproducing population dynamics, even if some precision is lost when compared to fully continuous models.
 Thus, population models based on coarse, population-level data show promise for producing ecological forecasts at landscape and regional scales [@Queenborough2011].
@@
 Our demographic data comes from the Fort Keogh Livestock and Range Research Laboratory in eastern Montana's northern mixed prairie near Miles City, Montana, USA ($46^{\circ}$ 19' N, $105^{\circ}$ 48' W).
-The dataset is freely available on Ecological Archives [@Anderson2011], and interested readers should refer to the metadata therin for a complete description.
-The site is about 800 m above sea level and mean annual precipitation (1878-2009) is 334 mm, with most annual precipitaion falling from April through September.
+The dataset is freely available on Ecological Archives [@Anderson2011], and interested readers should refer to the metadata therein for a complete description.
+The site is about 800 m above sea level and mean annual precipitation (1878-2009) is 334 mm, with most annual precipitation falling from April through September.
 The site is grass dominated and, for the purposes of our study, we focus on the four most abundant graminoid species: \emph{Bouteloua gracilis} (BOGR), \emph{Hesperostipa comata} (HECO), \emph{Pascopyrum smithii} (PASM), and \emph{Poa secunda} (POSE).
 
-From 1932 to 1945 individual plants were identified and mapped annualy in 44 1-$\text{m}^2$ quadrats using a pantograph.
+From 1932 to 1945 individual plants were identified and mapped annually in 44 1-$\text{m}^2$ quadrats using a pantograph.
 The quadrats were distributed in six pastures, each assigned a grazing treatment of light (1.24 ha/animal unit month), moderate (0.92 ha/aum), and heavy (0.76 ha/aum) stocking rates (two pastures per treatment).
@@
 For individual level data we fit models for survival, growth, and recruitment of new individuals for each species.
-At the quadrat level we fit analagous models of extinction probability, percent cover increase/decrease, and quadrat colonization for each species.
-We describe the statistical models seprately since fitting the models required different approaches at the individual and quadrat levels.
+At the quadrat level we fit analogous models of extinction probability, percent cover increase/decrease, and quadrat colonization for each species.
+We describe the statistical models separately since fitting the models required different approaches at the individual and quadrat levels.
 All models contain four climate covariate that we chose \emph{a priori}: fall through spring precipitation at \emph{t}-1 and \emph{t}-2 (ppt1 and ppt2, respectively) and mean spring temperature at \emph{t}-1 and \emph{t}-2 (TmeanSpr1 and TmeanSpr2, respectively), where \emph{t} is the observation year.
@@
 
-We modeled growth as gaussian process describing genet size at time $t+1$ as a function of size at $t$ and climate covariates:
+We modeled growth as Gaussian process describing genet size at time $t+1$ as a function of size at $t$ and climate covariates:
 
@@
 
-where $x$ is log genet size and all other paramters are as described for the survival regression.
+where $x$ is log genet size and all other parameters are as described for the survival regression.
 
@@
 
-We retain the abbreviations $S$, $G$, and $R$ from the analagous processes at the individual level.
+We retain the abbreviations $S$, $G$, and $R$ from the analogous processes at the individual level.
 The vital rate models at the quadrat level are all based on quadrat proportional cover.
@@
 Since we are modeling proportional cover, we essentially get density-dependence for "free" when proportional cover is included as a covariate.
-That is, density-dependence emerges because quadrats with low cover gerenally increase in cover whereas quadrats with high cover generally decrease in cover.
+That is, density-dependence emerges because quadrats with low cover generally increase in cover whereas quadrats with high cover generally decrease in cover.
 
@@
 We do so because during model fitting we use a beta likelihood since the data, proportional cover, is beta distributed.
-The beta likelihood requires shape ($\rho$) and rate ($\eta$) parameters that can be caclulated using moment-matching:
+The beta likelihood requires shape ($\rho$) and rate ($\eta$) parameters that can be calculated using moment-matching:
 
@@
 Our Bayesian approach to fitting the vital rate models required choosing appropriate priors for unknown parameters and deciding which, if any, of those prior should be hierarchical.
-We decided to fit models where all terms except climate covariates were fit by species, while the climate covariates were fit heierarchically where species-specific coefficients were drawn from a shared `global' coefficient distribution.
-We did so for two reasons: (1) the four focal species are all perrennial grasses that we expect to respond similarly to climate covariates, and (2) convergence of climate effects at the quadrat level was much easier to achieve when we modeled these terms hierarchically, allowing them to "share" statistical strenght via partial pooling [@Gelman2007].
+We decided to fit models where all terms except climate covariates were fit by species, while the climate covariates were fit hierarchically where species-specific coefficients were drawn from a shared `global' coefficient distribution.
+We did so for two reasons: (1) the four focal species are all perennial grasses that we expect to respond similarly to climate covariates, and (2) convergence of climate effects at the quadrat level was much easier to achieve when we modeled these terms hierarchically, allowing them to "share" statistical strength via partial pooling [@Gelman2007].
 So, climate effects were modeled as:
@@
 At each time step, we first randomly selected climate covariates from one of the 14 observed years and also randomly drew one set of random year effects for each vital rate regression.
+Climate years and random year effects were drawn independently because they were uncorrelated.
 Then, we drew the full parameter set (specific random year effects, climate effects, fixed effects) from a randomly selected MCMC iteration.
@@
 
-Climate covariates improved the growth regressions at both levels, with a slghty greater contribution at the individual level (Figure 1).
+Climate covariates improved the growth regressions at both levels, with a slightly greater contribution at the individual level (Figure 1).
 Parameter estimates were generally consistent in terms of sign and magnitude between the individual level and the quadrat level for each species and climate covariate (Figure 2).
@@
 
-1. Climate detections
+1. Climate detection
 2. Limitations:

### 2015-01-09 manuscript/manuscript.Rmd
new population model based on poisson likelihood

@@
 Note that the y-axes vary across panels.", fig.width=5, fig.height=6, cache=FALSE, results='hide'}
-resD <- readRDS("../montana/quadBM/simulations/quadBM_oneStep_Residuals.rds")
+resD <- readRDS("../montana/quadBM/simulations/quadBM_oneStep_ResidualsFullModel.rds")
 nSims <- length(unique(resD$sim))
@@
 ggplot(resD)+
-  geom_hline(aes(yintercept=0))+
+  geom_hline(aes(yintercept=0), linetype=2)+
 #   geom_line(data=diffD, aes(x=year, y=avgDiff, group=species), color="purple", alpha=0.5)+
 #   stat_smooth(data=diffD, aes(x=year, y=avgDiff, group=species), color="lightblue", se = FALSE, size=1)+
-  geom_boxplot(aes(x=year, y=covChangeResiduals), outlier.shape = 1, width=0.3, outlier.size = 0.5)+
+  geom_boxplot(aes(x=year, y=resids), outlier.shape = 1, width=0.3, outlier.size = 0.5)+
   facet_grid(species~., scales = "free_y")+

### 2015-01-09 montana/quadBM/simulations/oneStepForecastQBM_popVersion.R

@@ -21,7 +21,7 @@
 rm(list=ls(all=TRUE))
 
 #Set number of sims per yearly transition
-NumberSimsPerYear <- 1
+NumberSimsPerYear <- 2
 
 ####
 #### Load libraries ----------------------------------
@@ -46,43 +46,12 @@ climD <- read.csv("../../weather/Climate.csv")
 climD[3:6] <- scale(climD[3:6], center = TRUE, scale = TRUE)
 
 #load vital rate parameters
-pCol <- readRDS("../vitalRateRegressions/colonization/colonizationParamsMCMC.rds")
-pGrow <- readRDS("../vitalRateRegressions/growth/growthParamsMCMC.rds")
-pSurv <- readRDS("../vitalRateRegressions/survival/survivalParamsMCMC.rds")
+pGrow <- readRDS("../vitalRateRegressions/popGrowthParamsMCMC.rds")
 
 
 ####
 #### Organize parameter values ---------------------------------
 ####
-#colonization
-pCol2 <- melt(pCol)
-pCol2$Spp <- c(rep(rep(sppList, each=3000), times=6),
-               rep(rep(sppList, each=3000), times=5))
-pCol2$Coef <- c(rep("gInt", times=6*4*3000),
-                rep("int", times=4*3000),
-                rep("rain1", times=4*3000),
-                rep("rain2", times=4*3000),
-                rep("temp1", times=4*3000),
-                rep("temp2", times=4*3000))
-colnames(pCol2)[1] <- "Iter"
-pCol <- pCol2[,c(1,3:5)]; rm(pCol2)
-
-#survival
-pSurv2 <- melt(pSurv)
-pSurv2$Spp <- c(rep(rep(sppList, each=3000), times=1),
-                rep(rep(sppList, each=3000), times=6),
-               rep(rep(sppList, each=3000), times=5))
-pSurv2$Coef <- c(rep("beta", times=3000*4),
-                 rep("gInt", times=6*4*3000),
-                rep("int", times=4*3000),
-                rep("rain1", times=4*3000),
-                rep("rain2", times=4*3000),
-                rep("temp1", times=4*3000),
-                rep("temp2", times=4*3000))
-colnames(pSurv2)[1] <- "Iter"
-pSurv <- pSurv2[,c(1,3:5)]; rm(pSurv2)
-
-#growth
[…]

### 2015-01-09 montana/quadBM/vitalRateRegressions/popAbundanceModel.R

@@ -1,0 +1,125 @@
+#### ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ ####
+#### Script to fit population growth model to aggregated Montana individual-level data ####
+#### ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ ####
+
+#!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!#
+# Set working directory to location of this source file #
+#!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!#
+
+#clear everything, just to be safe 
+rm(list=ls(all=TRUE))
+
+#load libraries
+library(rjags)
+library(coda)
+load.module("dic")
+
+#bring in data
+allD <- read.csv("../../speciesData/quadAllCover.csv")
+allD <- allD[,2:ncol(allD)] #get rid of X ID column
+sppList <- as.character(unique(allD$Species))
+
+climD <- read.csv("../../weather/Climate.csv")
+climD[3:6] <- scale(climD[3:6], center = TRUE, scale = TRUE)
+
+backD <- data.frame(climYear=NA,
+                    quad = NA,
+                    year= NA,
+                    totCover= NA,
+                    Species= NA,
+                    propCover= NA,
+                    lag.cover= NA,
+                    pptLag= NA,
+                    ppt1= NA,
+                    TmeanSpr1= NA,
+                    ppt2= NA,
+                    TmeanSpr2= NA,
+                    TmeanSum1= NA,
+                    TmeanSum2= NA,
+                    yearID= NA,
+                    group = NA)
+
+#loop through species and remake data frame
+for(spp in 1:length(sppList)){
+  doSpp <- sppList[spp]
+  sppD <- subset(allD, Species==doSpp)
+  
+  # create lag cover variable
+  tmp=sppD[,c("quad","year","propCover")]
+  tmp$year=tmp$year+1
+  names(tmp)[3]="lag.cover"
+  sppD=merge(sppD,tmp,all.x=T)
+  
[…]

### 2015-01-09 montana/quadBM/vitalRateRegressions/popAbundanceModelJAGS.R

@@ -1,0 +1,38 @@
+model{
+  #likelihood and process model
+  for(i in 1:nObs){
+    log(lambda[i]) <- intYr[spp[i],yrs[i]] + intG[spp[i],grp[i]] + beta[spp[i],yrs[i]]*X[i] + temp1[spp[i]]*TmeanSpr1[i] + temp2[spp[i]]*TmeanSpr2[i] + rain1[spp[i]]*ppt1[i] + rain2[spp[i]]*ppt2[i]  
+    C[i] ~ dpois(lambda[i])
+  }
+  
+  #priors
+  for(s in 1:nSpp){
+    betaSpp[s] ~ dnorm(0, 1e-6)
+    temp1[s] ~ dnorm(temp1Mu, temp1Var)
+    temp2[s] ~ dnorm(temp2Mu, temp2Var)
+    rain1[s] ~ dnorm(rain1Mu, rain1Var)
+    rain2[s] ~ dnorm(rain2Mu, rain2Var)
+    intercept[s] ~ dnorm(0, 1e-6)
+    intVaryY[s] ~ dgamma(0.001, 0.001)
+    betaVar[s] ~ dgamma(0.001, 0.001)
+    t0[s] ~ dnorm(0, .01) 
+    tau[s] <- exp(t0[s])
+    intVarG[s] ~ dgamma(2, 0.5) 
+    for(y in 1:nYrs){
+      intYr[s,y] ~ dnorm(intercept[s], intVaryY[s])
+      beta[s,y] ~ dnorm(betaSpp[s], betaVar[s])
+    }
+    for(g in 1:nGrp){
+      intG[s,g] ~ dnorm(0, intVarG[s])
+    }
+  }
+  
+  temp1Mu ~ dnorm(0,1e-6)
+  temp2Mu ~ dnorm(0,1e-6)
+  rain1Mu ~ dnorm(0,1e-6)
+  rain2Mu ~ dnorm(0,1e-6)
+  temp1Var ~ dgamma(0.001, 0.001)
+  temp2Var ~ dgamma(0.001, 0.001)
+  rain1Var ~ dgamma(0.001, 0.001)
+  rain2Var ~ dgamma(0.001, 0.001)
+}

### 2015-01-11 manuscript/manuscript.Rmd
took our sure and col functions for probe

@@
 Note that the y-axes vary across panels.", fig.width=5, fig.height=6, cache=FALSE, results='hide'}
-resD <- readRDS("../montana/quadBM/simulations/quadBM_oneStep_ResidualsFullModel.rds")
+resD <- readRDS("../montana/quadBM/simulations/oneStepResultsFull.rds")
 nSims <- length(unique(resD$sim))
 diffD <- ddply(resD, c("year", "species"), summarise,
-               avgDiff = mean(yearDiff))
+               avgDiff = mean(resids),
+               sdDiff = sd(resids),
+               dnDiff = quantile(resids, probs = 0.025, names=FALSE),
+               upDiff = quantile(resids, probs = 0.975, names=FALSE))
+ggplot(diffD)+
+  geom_hline(aes(yintercept=0), linetype=2, color="grey50")+
+  geom_point(aes(x=year, y=avgDiff), size=3)+
+#   geom_errorbar(aes(x=year, ymax=avgDiff+sdDiff, ymin=avgDiff-sdDiff), width=0.1)+
+    geom_errorbar(aes(x=year, ymax=upDiff, ymin=dnDiff), width=0.1)+
+  facet_grid(species~., scales = "free_y")+
+  xlab("Year")+
+  ylab("Residuals (% cover difference)")+
+  theme_bw()+
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
 

### 2015-01-11 montana/quadBM/simulations/oneStepForecastQBM_beta.R

@@ -1,0 +1,292 @@
+######################################
+#### Quad-Based Model simulations ####
+######################################
+
+################################################
+#### Andrew Tredennick (atredenn@gmail.com) ####
+#### 1-7-2014                               ####
+################################################
+
+#!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!#
+# Set working directory to location of this source file #
+#!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!#
+
+## ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ ##
+## This is for one-step-ahead forecasting, by quadrat.                                         ##
+## So, initilize with specific quadrat cover, use model with group effect and project forward. ##
+## Do this N times for each yearly transition with different MCMC parameters.                  ##
+## ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ ##
+
+#clear everything, just to be safe 
+rm(list=ls(all=TRUE))
+
+#Set number of sims per yearly transition
+NumberSimsPerYear <- 2
+
+####
+#### Load libraries ----------------------------------
+####
+library(reshape2)
+library(plyr)
+library(ggplot2)
+library(ggthemes)
+library(gridExtra)
+
+####
+#### Bring in data -----------------------------------
+####
+#cover data
+allD <- read.csv("../../speciesData/quadAllCover.csv")
+allD <- allD[,2:ncol(allD)] #get rid of X ID column
+allD$percCover <- allD$totCover/10000
+sppList <- as.character(unique(allD$Species))
+
+#climate data
+climD <- read.csv("../../weather/Climate.csv")
+climD[3:6] <- scale(climD[3:6], center = TRUE, scale = TRUE)
+
+#load vital rate parameters
+pCol <- readRDS("../vitalRateRegressions/colonization/colonizationParamsMCMC.rds")
[…]

### 2015-01-11 montana/quadBM/vitalRateRegressions/popAbundanceModelJAGS_beta.R

@@ -1,0 +1,40 @@
+model{
+  #likelihood and process model
+  for(i in 1:nObs){
+    logit(lambda[i]) <- intYr[spp[i],yrs[i]] + intG[spp[i],grp[i]] + beta[spp[i],yrs[i]]*X[i] + temp1[spp[i]]*TmeanSpr1[i] + temp2[spp[i]]*TmeanSpr2[i] + rain1[spp[i]]*ppt1[i] + rain2[spp[i]]*ppt2[i]  
+    C[i] ~ dbeta(p[i], q[i])
+    p[i] <- lambda[i] * tau[spp[i]]
+    q[i] <- (1 - lambda[i]) * tau[spp[i]]
+  }
+  
+  #priors
+  for(s in 1:nSpp){
+    betaSpp[s] ~ dnorm(0, 1e-6)
+    temp1[s] ~ dnorm(temp1Mu, temp1Var)
+    temp2[s] ~ dnorm(temp2Mu, temp2Var)
+    rain1[s] ~ dnorm(rain1Mu, rain1Var)
+    rain2[s] ~ dnorm(rain2Mu, rain2Var)
+    intercept[s] ~ dnorm(0, 1e-6)
+    intVaryY[s] ~ dgamma(0.001, 0.001)
+    betaVar[s] ~ dgamma(0.001, 0.001)
+    t0[s] ~ dnorm(0, .01) 
+    tau[s] <- exp(t0[s])
+    intVarG[s] ~ dgamma(2, 0.5)
+    for(y in 1:nYrs){
+      intYr[s,y] ~ dnorm(intercept[s], intVaryY[s])
+      beta[s,y] ~ dnorm(betaSpp[s], betaVar[s])
+    }
+    for(g in 1:nGrp){
+      intG[s,g] ~ dnorm(0, intVarG[s])
+    }
+  }
+  
+  temp1Mu ~ dnorm(0,1e-6)
+  temp2Mu ~ dnorm(0,1e-6)
+  rain1Mu ~ dnorm(0,1e-6)
+  rain2Mu ~ dnorm(0,1e-6)
+  temp1Var ~ dgamma(0.001, 0.001)
+  temp2Var ~ dgamma(0.001, 0.001)
+  rain1Var ~ dgamma(0.001, 0.001)
+  rain2Var ~ dgamma(0.001, 0.001)
+}

### 2015-01-12 manuscript/manuscript.Rmd
updated figure

--- 2015-01-09 manuscript/manuscript.Rmd
@@
 Each one-step forecast was simulated `r nSims` times.
-Note that the y-axes vary across panels.", fig.width=5, fig.height=6, cache=FALSE, results='hide'}
-resD <- readRDS("../montana/quadBM/simulations/quadBM_oneStep_Residuals.rds")
-nSims <- length(unique(resD$sim))
-diffD <- ddply(resD, c("year", "species"), summarise,
-               avgDiff = mean(yearDiff))
-
-ggplot(resD)+
-  geom_hline(aes(yintercept=0))+
-#   geom_line(data=diffD, aes(x=year, y=avgDiff, group=species), color="purple", alpha=0.5)+
-#   stat_smooth(data=diffD, aes(x=year, y=avgDiff, group=species), color="lightblue", se = FALSE, size=1)+
-  geom_boxplot(aes(x=year, y=covChangeResiduals), outlier.shape = 1, width=0.3, outlier.size = 0.5)+
-  facet_grid(species~., scales = "free_y")+
+Note that the y-axes vary across panels.", fig.width=8, fig.height=8, cache=FALSE, results='hide'}
+resD <- readRDS("../montana/quadBM/simulations/quadBM_oneStep_ResidualsFullModel.rds")
+resDny <- readRDS("../montana/quadBM/simulations/quadBM_oneStep_ResidualsNoYear.rds")
+resD$simtype <- rep("full", nrow(resD))
+resDny$simtype <- rep("noyear", nrow(resDny))
+resDall <- rbind(resD, resDny)
+# nSims <- length(unique(resD$sim))
+diffD <- ddply(resDall, c("year", "species", "simtype"), summarise,
+               avgDiff = mean(resids, na.rm = TRUE),
+#                upDiff75 = quantile(resids, probs = 0.8775, names=FALSE, na.rm = TRUE),
+#                dnDiff75 = quantile(resids, probs = 0.1225, names=FALSE, na.rm = TRUE),
+               dnDiff95 = quantile(resids, probs = 0.025, names=FALSE, na.rm = TRUE),
+               upDiff95 = quantile(resids, probs = 0.975, names=FALSE, na.rm = TRUE))
+# ggplot(diffD, aes(color=simtype))+
+#   geom_hline(aes(yintercept=0), linetype=2, color="grey50")+
+#   geom_point(aes(x=year, y=avgDiff), size=3, position=position_dodge(width=0.5))+
+#   geom_errorbar(aes(x=year, ymax=upDiff95, ymin=dnDiff95), width=0.00001, position=position_dodge(width=0.5))+
+# #   geom_errorbar(aes(x=year, ymax=upDiff75, ymin=dnDiff75), size=1, width=0.00001, position=position_dodge(width=0.5))+
+#   facet_grid(species~., scales = "free_y")+
+#   xlab("Year")+
+#   ylab("Residuals (% cover difference)")+
+#   theme_bw()+
+#   theme(axis.text.x = element_text(angle = 45, hjust = 1))
+# 
+# ggplot(resDall)+
+#   geom_hline(aes(yintercept=0), linetype=2)+
+# #   geom_line(data=diffD, aes(x=year, y=avgDiff, group=species), color="purple", alpha=0.5)+
+# #   stat_smooth(data=diffD, aes(x=year, y=avgDiff, group=species), color="lightblue", se = FALSE, size=1)+
+#   geom_boxplot(aes(x=year, y=resids, fill=simtype), outlier.shape = 1, width=0.3, outlier.size = 0.5)+
+#   facet_grid(species~., scales = "free_y")+
+#   xlab("Year")+
+#   ylab("Residuals (% cover difference)")+
+#   theme_bw()+
+#   theme(axis.text.x = element_text(angle = 45, hjust = 1))
+p1 <- ggplot(subset(resDall, species=="BOGR"))+
+  geom_hline(aes(yintercept=0), linetype=2)+
+  geom_boxplot(aes(x=year, y=resids, fill=simtype), outlier.shape = 1, outlier.size = 0, alpha=0.75, color="grey25")+
   xlab("Year")+
@@
   theme_bw()+
-  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
+  scale_fill_manual(values=c("darkgoldenrod1", "steelblue4"))+
+  ggtitle(expression(italic("B. gracilis")))+
+  guides(fill=FALSE)
+p2 <- ggplot(subset(resDall, species=="HECO"))+
+  geom_hline(aes(yintercept=0), linetype=2)+
+  geom_boxplot(aes(x=year, y=resids, fill=simtype), outlier.shape = 1, outlier.size = 0, alpha=0.75, color="grey25")+
+  xlab("Year")+
+  ylab("Residuals (% cover difference)")+
+  theme_bw()+
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
+  scale_fill_manual(values=c("darkgoldenrod1", "steelblue4"), labels=c("Year effects", "No year effects"))+
+  ggtitle(expression(italic("H. comita")))+
+  coord_cartesian(ylim = c(-25, 5))+
+  guides(fill=guide_legend(title=""))+
+  theme(legend.position=c(0.7,0.2))
+p3 <- ggplot(subset(resDall, species=="PASM"))+
+  geom_hline(aes(yintercept=0), linetype=2)+
+  geom_boxplot(aes(x=year, y=resids, fill=simtype), outlier.shape = 1, outlier.size = 0, alpha=0.75, color="grey25")+
+  xlab("Year")+
+  ylab("Residuals (% cover difference)")+
+  theme_bw()+
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
+  scale_fill_manual(values=c("darkgoldenrod1", "steelblue4"))+
+  ggtitle(expression(italic("P. smithii")))+
+  guides(fill=FALSE)+
+  coord_cartesian(ylim = c(-2.5, 2.5))
+p4 <- ggplot(subset(resDall, species=="POSE"))+
+  geom_hline(aes(yintercept=0), linetype=2)+
+  geom_boxplot(aes(x=year, y=resids, fill=simtype), outlier.shape = 1, outlier.size = 0, alpha=0.75, color="grey25")+
+  xlab("Year")+
+  ylab("Residuals (% cover difference)")+
+  theme_bw()+
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
+  scale_fill_manual(values=c("darkgoldenrod1", "steelblue4"))+
+  ggtitle(expression(italic("P. secunda")))+
+  guides(fill=FALSE)+
+  coord_cartesian(ylim = c(-50, 10))
+pFinal <- grid.arrange(p1,p2,p3,p4, ncol=2, nrow=2)
+pFinal
+
 ```

### 2015-01-12 manuscript/manuscript.tex
updated figure

@@
 population-level data leads to models that will produce less precise
-predicitons but can be applied over greater spatial and temporal
+predictions but can be applied over greater spatial and temporal
 extents. An open question is how well models based on population-level
@@
 methods in a study on arable weeds. In particular, Queenborough et al.
-(2011) provide emprical evidence that density-structured models are
+(2011) provide empirical evidence that density-structured models are
 capable of reproducing population dynamics, even if some precision is
@@
 dataset is freely available on Ecological Archives (Anderson et al.
-2011), and interested readers should refer to the metadata therin for a
+2011), and interested readers should refer to the metadata therein for a
 complete description. The site is about 800 m above sea level and mean
 annual precipitation (1878-2009) is 334 mm, with most annual
-precipitaion falling from April through September. The site is grass
+precipitation falling from April through September. The site is grass
 dominated and, for the purposes of our study, we focus on the four most
@@
 
-From 1932 to 1945 individual plants were identified and mapped annualy
+From 1932 to 1945 individual plants were identified and mapped annually
 in 44 1-$\text{m}^2$ quadrats using a pantograph. The quadrats were
@@
 recruitment of new individuals for each species. At the quadrat level we
-fit analagous models of extinction probability, percent cover
+fit analogous models of extinction probability, percent cover
 increase/decrease, and quadrat colonization for each species. We
-describe the statistical models seprately since fitting the models
+describe the statistical models separately since fitting the models
 required different approaches at the individual and quadrat levels. All
@@
 
-We modeled growth as gaussian process describing genet size at time
+We modeled growth as Gaussian process describing genet size at time
 $t+1$ as a function of size at $t$ and climate covariates:
@@
 
-where $x$ is log genet size and all other paramters are as described for
-the survival regression.
+where $x$ is log genet size and all other parameters are as described
+for the survival regression.
 
@@
 
-We retain the abbreviations $S$, $G$, and $R$ from the analagous
+We retain the abbreviations $S$, $G$, and $R$ from the analogous
 processes at the individual level. The vital rate models at the quadrat
@@
 covariate. That is, density-dependence emerges because quadrats with low
-cover gerenally increase in cover whereas quadrats with high cover
+cover generally increase in cover whereas quadrats with high cover
 generally decrease in cover.
@@
 beta distributed. The beta likelihood requires shape ($\rho$) and rate
-($\eta$) parameters that can be caclulated using moment-matching:
+($\eta$) parameters that can be calculated using moment-matching:
 
@@
 terms except climate covariates were fit by species, while the climate
-covariates were fit heierarchically where species-specific coefficients
+covariates were fit hierarchically where species-specific coefficients
 were drawn from a shared `global' coefficient distribution. We did so
-for two reasons: (1) the four focal species are all perrennial grasses
+for two reasons: (1) the four focal species are all perennial grasses
 that we expect to respond similarly to climate covariates, and (2)
@@
 achieve when we modeled these terms hierarchically, allowing them to
-``share'' statistical strenght via partial pooling (Gelman and Hill
+``share'' statistical strength via partial pooling (Gelman and Hill
 2007). So, climate effects were modeled as:
@@
 14 observed years and also randomly drew one set of random year effects
-for each vital rate regression. Then, we drew the full parameter set
-(specific random year effects, climate effects, fixed effects) from a
-randomly selected MCMC iteration. We selected random year effects and
-the climate year independently.
+for each vital rate regression. Climate years and random year effects
+were drawn independently because they were uncorrelated. Then, we drew
+the full parameter set (specific random year effects, climate effects,
+fixed effects) from a randomly selected MCMC iteration. We selected
+random year effects and the climate year independently.
 
@@
 Climate covariates improved the growth regressions at both levels, with
-a slghty greater contribution at the individual level (Figure 1).
+a slightly greater contribution at the individual level (Figure 1).
 Parameter estimates were generally consistent in terms of sign and
@@
 \item
-  Climate detections
+  Climate detection
 \item

### 2015-01-12 manuscript/manuscript.Rmd
updated figure 1

@@
 -------
-```{r figure_1, dependson="plot-options", fig.cap="Boxplots of model residuals for one-step-ahead forecasts at each observation year.
-Each one-step forecast was simulated `r nSims` times.
+```{r figure_1, dependson="plot-options", fig.cap="Boxplots of model residuals for one-step-ahead forecasts at each observation year for each quadrat.
+Residuals represent the difference in observed and predicted cover change from year $t-1$ to year $t$.
+Each one-step forecast was simulated 100 times for each quadrat.
 Note that the y-axes vary across panels.", fig.width=8, fig.height=8, cache=FALSE, results='hide'}
@@
   scale_fill_manual(values=c("darkgoldenrod1", "steelblue4"))+
-  ggtitle(expression(italic("P. secunda")))+
+  ggtitle(expression(italic("Poa secunda")))+
   guides(fill=FALSE)+
@@
 pFinal <- grid.arrange(p1,p2,p3,p4, ncol=2, nrow=2)
-pFinal
+# pFinal
 

### 2015-01-12 manuscript/manuscript.tex
updated figure 1

@@
 \caption{Boxplots of model residuals for one-step-ahead forecasts at
-each observation year. Each one-step forecast was simulated
-\texttt{r nSims} times. Note that the y-axes vary across panels.}
+each observation year for each quadrat. Residuals represent the
+difference in observed and predicted cover change from year $t-1$ to
+year $t$. Each one-step forecast was simulated 100 times for each
+quadrat. Note that the y-axes vary across panels.}
 \end{figure}

### 2015-01-12 manuscript/manuscript.Rmd
updated fig colors

@@
 #   theme(axis.text.x = element_text(angle = 45, hjust = 1))
+myCols <- c("#237DA4", "#7ECAD0")
 p1 <- ggplot(subset(resDall, species=="BOGR"))+
   geom_hline(aes(yintercept=0), linetype=2)+
-  geom_boxplot(aes(x=year, y=resids, fill=simtype), outlier.shape = 1, outlier.size = 0, alpha=0.75, color="grey25")+
+  geom_boxplot(aes(x=year, y=resids, fill=simtype), outlier.shape = 1, outlier.size = 0, color="grey25")+
   xlab("Year")+
@@
   theme(axis.text.x = element_text(angle = 45, hjust = 1))+
-  scale_fill_manual(values=c("darkgoldenrod1", "steelblue4"))+
+  scale_fill_manual(values=myCols)+
   ggtitle(expression(italic("B. gracilis")))+
@@
   geom_hline(aes(yintercept=0), linetype=2)+
-  geom_boxplot(aes(x=year, y=resids, fill=simtype), outlier.shape = 1, outlier.size = 0, alpha=0.75, color="grey25")+
+  geom_boxplot(aes(x=year, y=resids, fill=simtype), outlier.shape = 1, outlier.size = 0, color="grey25")+
   xlab("Year")+
@@
   theme(axis.text.x = element_text(angle = 45, hjust = 1))+
-  scale_fill_manual(values=c("darkgoldenrod1", "steelblue4"), labels=c("Year effects", "No year effects"))+
+  scale_fill_manual(values=myCols, labels=c("Year effects", "No year effects"))+
   ggtitle(expression(italic("H. comita")))+
@@
   theme(axis.text.x = element_text(angle = 45, hjust = 1))+
-  scale_fill_manual(values=c("darkgoldenrod1", "steelblue4"))+
+  scale_fill_manual(values=myCols)+
   ggtitle(expression(italic("P. smithii")))+
@@
   theme(axis.text.x = element_text(angle = 45, hjust = 1))+
-  scale_fill_manual(values=c("darkgoldenrod1", "steelblue4"))+
+  scale_fill_manual(values=myCols)+
   ggtitle(expression(italic("Poa secunda")))+

### 2015-01-12 manuscript/manuscript.Rmd
updated figs

@@
 climSimQBMplot <- melt(climSimQBMcast)
+myCols2 <- c("#277BA8", "#7ABBBD", "#AED77A")
 ggplot(data=subset(climSimQBMplot, variable=="diffP"|variable=="diffT"|variable=="diffB"))+
   geom_bar(aes(x=species, y=value, fill=variable), stat="identity", position="dodge", color="white")+
-  scale_fill_manual(name="", labels=c("1% Precipitation Increase", "1% Temperature Increase", "1% Temp. & Ppt. Increase"), values=c("grey25", "grey50", "grey75"))+
+  scale_fill_manual(name="", labels=c("1% Precipitation Increase", "1% Temperature Increase", "1% Temp. & Ppt. Increase"), values=myCols2)+
   theme_bw()+

### 2015-01-13 manuscript/suppMat.Rmd
updated simulations

@@
 library("knitr")
+library(ggplot2)
+library(reshape2)
+library(plyr)
+library(gridExtra)
 basename <- "manuscript"
@@
 gQC <- read.csv("../montana/quadBM/vitalRateRegressions/growth/growthQuants.csv")
-gQC <- gQC[129:144,]
+gQC <- gQC[137:152,]
 coefNames <- c("ppt1", "ppt2", "temp1", "temp2")

### 2015-01-13 montana/quadBM/simulations/PptTmpChangeQBM.R

@@ -95,7 +95,7 @@ pGrowYrs <- subset(pGrow, Coef=="beta" | Coef=="intYr")
 years <- unique(allD$year)[2:14]+1900
 pGrowYrs$Year <- c(rep(rep(years, each=3000), each=4),
                    rep(rep(years, each=3000), each=4))
-
+histogram(subset(pGrowAll, Coef=="rain1")[,2])
 
 
 ####
@@ -162,8 +162,8 @@ outD <- data.frame(variable=NA, cover=NA, species=NA)
 
 for(i in 1:length(sppList)){
   sppSim <- sppList[i]
-  nSim <- 100
-  yearsN <- 100
+  nSim <- 10
+  yearsN <- 20
   years <- unique(allD$year)+1900
   yearsID <- unique(allD$year)
   Nsave <- matrix(ncol=yearsN, nrow=nSim)

### 2015-01-13 montana/quadBM/simulations/climateChangeQBM.R

@@ -198,5 +198,4 @@ for(i in 1:length(sppList)){
 #### Output
 ####
 outD <- outD[2:nrow(outD),]
-
-# saveRDS(outD, "climateChangeBase.rds")
+saveRDS(outD, "climateChangeBase.rds")

### 2015-01-13 montana/quadBM/vitalRateRegressions/growth/growthAllSppHierarchy_MCMC.R

@@ -83,8 +83,8 @@ X <- growD$percLagCover
 ####
 #### Run MCMC
 ####
-iterations <- 50000
-adapt <- 10000
+iterations <- 500
+adapt <- 100
 dataJ <- list(nGrp=nGrp, nYrs=nYrs, nObs=nObs, C=C, X=X, yrs=yrs, grp=grp,
               TmeanSpr1=TmeanSpr1, TmeanSpr2=TmeanSpr2, ppt1=ppt1, ppt2=ppt2, spp=spp, nSpp=nSpp)
 mod <- jags.model("growthAllSpp_JAGS.R", data=dataJ, n.chains=3, n.adapt=adapt)

### 2015-01-13 manuscript/manuscript.Rmd
update ms
small updates; see issues note

@@
   scale_fill_manual(values=myCols)+
+#   coord_cartesian(ylim = c(-50, 45))+
   ggtitle(expression(italic("B. gracilis")))+
@@
   ggtitle(expression(italic("H. comita")))+
-  coord_cartesian(ylim = c(-25, 5))+
+  coord_cartesian(ylim = c(-10, 5))+
   guides(fill=guide_legend(title=""))+
@@
 climSimQBMplot <- melt(climSimQBMcast)
+climSimQBMplot <- subset(climSimQBMplot, species!="POSE")
 myCols2 <- c("#277BA8", "#7ABBBD", "#AED77A")

### 2015-01-13 manuscript/suppMat.Rmd
small updates; see issues note

--- 2015-01-07 manuscript/suppMat.Rmd
@@
 The dashed vertical line is at 0, indicating no effect.", fig.width=8, fig.height=7, cache=FALSE, results='hide'}
-gQC <- read.csv("../montana/quadBM/vitalRateRegressions/growth/growthQuants.csv")
-gQC <- gQC[129:144,]
+gQCall <- read.csv("../montana/quadBM/vitalRateRegressions/growth/growthQuants.csv")
+# gQCall$diff <- gQCall$X97.5. - gQCall$X2.5.
+# ggplot(gQCall)+
+#   geom_histogram(aes(diff, fill=species))+
+#   facet_grid(species~.)
+gQC <- gQCall[137:152,]
 coefNames <- c("ppt1", "ppt2", "temp1", "temp2")
@@
 plotQC <- rbind(gQC,sQC,rQC)
-gQ <- ggplot(plotQC, aes(x=species, y=X50., color=rate))+
+gQ <- ggplot(plotQC, rate=="growth", aes(x=species, y=X50., color=rate))+
   geom_hline(aes(yintercept=0), linetype="dashed")+

### 2015-01-19 montana/quadBM/vitalRateRegressions/truncNormModel/growthAllSppHierarchy_MCMC.R
updated interactions

@@ -84,8 +84,8 @@ X <- log(growD$percLagCover)
 ####
 #### Run MCMC
 ####
-iterations <- 5000
-adapt <- 1000
+iterations <- 50000
+adapt <- 10000
 dataJ <- list(nGrp=nGrp, nYrs=nYrs, nObs=nObs, C=C, X=X, yrs=yrs, grp=grp,
               TmeanSpr1=TmeanSpr1, TmeanSpr2=TmeanSpr2, ppt1=ppt1, ppt2=ppt2, spp=spp, nSpp=nSpp)
 mod <- jags.model("growthAllSpp_JAGS.R", data=dataJ, n.chains=3, n.adapt=adapt)
@@ -98,33 +98,33 @@ out <- coda.samples(mod, c("intYr", "intercept", "beta", "betaSpp", "intG", "tem
 ####
 #### Check for convergence
 ####
-# gelmDiag <- gelman.diag(out)
+gelmDiag <- gelman.diag(out)
 # # heidel.diag(out)
 # # gelman.plot(out)
 # 
-# # pdf("growthOutPlots.pdf")
-# # plot(out, auto.layout=FALSE)
-# # dev.off()
+pdf("growthOutPlots.pdf")
+plot(out, auto.layout=FALSE)
+dev.off()
 # 
 # ####
 # #### Convert to dataframe for export and get other summaries
 # ####
-# outC <- rbind(out[[1]][(iterations-999):iterations,], 
-#               out[[2]][(iterations-999):iterations,], 
-#               out[[3]][(iterations-999):iterations,])
+outC <- rbind(out[[1]][(iterations-999):iterations,], 
+              out[[2]][(iterations-999):iterations,], 
+              out[[3]][(iterations-999):iterations,])
 # 
 outStat <- as.data.frame(summary(out)$stat)
-# outQuant <- as.data.frame(summary(out)$quantile)
-# outDeviance <- as.data.frame(summary(dic$deviance, mean)$stat)
+outQuant <- as.data.frame(summary(out)$quantile)
+outDeviance <- as.data.frame(summary(dic$deviance, mean)$stat)
 # 
 sppNames <- c(rep(sppList, 13+6+13+4+1+1+1))
 outStat$species <- sppNames
-# outQuant$species <- sppNames
+outQuant$species <- sppNames
 # 
-# saveRDS(outC, file = "growthParamsMCMC.rds")
[…]

### 2015-01-20 manuscript/manuscript.Rmd
updated interactions
new stats section for QBM

@@
 fontsize: 12pt
+mainfont: times
 geometry: margin=1in
@@
 #### Vital rate models: quadrat level
+The statistical approach used to model vital rates using aggregated data depends on the type of data collected.
+In our case, and as is often the case with census data, we have percent cover data (which can easily be transformed to proportion data, of course).
+We first considered fitting three vital rate models analagous to those we fit at the individual level: one for probability of extirpation within a quadrat (analagous to survival), one for cover change within a quadrat (analagous to growth), and one for probability of colonization within a quadrat (analagous to recruitment).
+However, within-quadrat extirpation and colonization events were rare in our time series (N=X and N=X, respectively).
+Given the broad spatial distribution of the quadrats we are studying, it is safe to assume that these events are in fact rare enough to be ignored for our purposes.
+So we constrained our statistical modeling of vital rates at the population level to change in percent cover within quadrats.
+For most of the remaining discussion of statistical modeling we refer to proportion data, which is simply percent data divided by 100.
+
+An obvious choice for fitting a linear model to proportion data is beta regression because the support of the beta distribution is [0,1], not including true zeros.
+However, when we used fitted model parameters from a beta regression in quadrat-based population model the simulated population tended toward 100% cover for all species.
+We therefore chose a more-constrained modeling approach based on a truncated log-normal likelihood.
+The model is
+
+\begin{align}
+x_{jq,t+1} &= \gamma^{G}_{j,t} + \phi^{G}_{jQ} + \beta^{G}_{j,t}x_{jq,t} + \theta^{S}_{jk}C_{k,t} \\
+Y^{G}_{jq,t+1} &\sim \text{LogNormal}(x_{jq,t+1}, \tau{j}) T(0,1)
+\end{align}
+
 At the quadrat level we defined three vital rates:

### 2015-01-21 manuscript/manuscript.Rmd
updated figure 1

@@
 Note that the y-axes vary across panels.", fig.width=8, fig.height=8, cache=FALSE, results='hide'}
-resD <- readRDS("../montana/quadBM/simulations/quadBM_oneStep_ResidualsFullModel.rds")
-resDny <- readRDS("../montana/quadBM/simulations/quadBM_oneStep_ResidualsNoYear.rds")
+resD <- readRDS("../montana/quadBM/simulations/oneStepResids_YearEffects.rds")
+resDny <- readRDS("../montana/quadBM/simulations/oneStepResids_NoYearEffects.rds")
 resD$simtype <- rep("full", nrow(resD))
@@
 resDall <- rbind(resD, resDny)
-# nSims <- length(unique(resD$sim))
-diffD <- ddply(resDall, c("year", "species", "simtype"), summarise,
-               avgDiff = mean(resids, na.rm = TRUE),
-#                upDiff75 = quantile(resids, probs = 0.8775, names=FALSE, na.rm = TRUE),
-#                dnDiff75 = quantile(resids, probs = 0.1225, names=FALSE, na.rm = TRUE),
-               dnDiff95 = quantile(resids, probs = 0.025, names=FALSE, na.rm = TRUE),
-               upDiff95 = quantile(resids, probs = 0.975, names=FALSE, na.rm = TRUE))
-# ggplot(diffD, aes(color=simtype))+
-#   geom_hline(aes(yintercept=0), linetype=2, color="grey50")+
-#   geom_point(aes(x=year, y=avgDiff), size=3, position=position_dodge(width=0.5))+
-#   geom_errorbar(aes(x=year, ymax=upDiff95, ymin=dnDiff95), width=0.00001, position=position_dodge(width=0.5))+
-# #   geom_errorbar(aes(x=year, ymax=upDiff75, ymin=dnDiff75), size=1, width=0.00001, position=position_dodge(width=0.5))+
-#   facet_grid(species~., scales = "free_y")+
+
+myCols <- c("#237DA4", "#7ECAD0")
+# ggplot(resDall, aes(x=resids, fill=simtype))+
+#   geom_histogram(alpha=0.5)+
+#   facet_grid(.~species, scales = "free_x")+
 #   xlab("Year")+
@@
 #   theme_bw()+
-#   theme(axis.text.x = element_text(angle = 45, hjust = 1))
+#   scale_fill_manual(values=myCols)
+
+# nSims <- length(unique(resD$sim))
+diffD <- ddply(resDall, c("year", "species", "simtype"), summarise,
+               avgDiff = median(resids, na.rm = TRUE),
+               upDiff75 = quantile(resids, probs = 0.8775, names=FALSE, na.rm = TRUE),
+               dnDiff75 = quantile(resids, probs = 0.1225, names=FALSE, na.rm = TRUE),
+               dnDiff95 = quantile(resids, probs = 0.025, names=FALSE, na.rm = TRUE),
+               upDiff95 = quantile(resids, probs = 0.975, names=FALSE, na.rm = TRUE))
+
+ggplot(diffD, aes(color=simtype))+
+  geom_hline(aes(yintercept=0), linetype=2, color="grey50")+
+  geom_point(aes(x=as.character(year), y=avgDiff), size=3, position=position_dodge(width=0.5))+
+  geom_errorbar(aes(x=as.character(year), ymax=upDiff95, ymin=dnDiff95), width=0.00001, position=position_dodge(width=0.5))+
+  geom_errorbar(aes(x=as.character(year), ymax=upDiff75, ymin=dnDiff75), size=1, width=0.00001, position=position_dodge(width=0.5))+
+  facet_grid(species~., scales = "free_y")+
+  xlab("Year")+
+  ylab("Residuals (% cover difference)")+
+  theme_bw()+
+  scale_color_manual(values=myCols)+
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
 # 
@@
 #   theme(axis.text.x = element_text(angle = 45, hjust = 1))
-myCols <- c("#237DA4", "#7ECAD0")
-p1 <- ggplot(subset(resDall, species=="BOGR"))+
-  geom_hline(aes(yintercept=0), linetype=2)+
-  geom_boxplot(aes(x=year, y=resids, fill=simtype), outlier.shape = 1, outlier.size = 0, color="grey25")+
-  xlab("Year")+
-  ylab("Residuals (% cover difference)")+
-  theme_bw()+
-  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
-  scale_fill_manual(values=myCols)+
-#   coord_cartesian(ylim = c(-50, 45))+
-  ggtitle(expression(italic("B. gracilis")))+
-  guides(fill=FALSE)
-p2 <- ggplot(subset(resDall, species=="HECO"))+
-  geom_hline(aes(yintercept=0), linetype=2)+
-  geom_boxplot(aes(x=year, y=resids, fill=simtype), outlier.shape = 1, outlier.size = 0, color="grey25")+
-  xlab("Year")+
-  ylab("Residuals (% cover difference)")+
-  theme_bw()+
-  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
-  scale_fill_manual(values=myCols, labels=c("Year effects", "No year effects"))+
-  ggtitle(expression(italic("H. comita")))+
-  coord_cartesian(ylim = c(-10, 5))+
-  guides(fill=guide_legend(title=""))+
-  theme(legend.position=c(0.7,0.2))
-p3 <- ggplot(subset(resDall, species=="PASM"))+
-  geom_hline(aes(yintercept=0), linetype=2)+
-  geom_boxplot(aes(x=year, y=resids, fill=simtype), outlier.shape = 1, outlier.size = 0, color="grey25")+
-  xlab("Year")+
-  ylab("Residuals (% cover difference)")+
-  theme_bw()+
-  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
-  scale_fill_manual(values=myCols)+
-  ggtitle(expression(italic("P. smithii")))+
-  guides(fill=FALSE)+
-  coord_cartesian(ylim = c(-2.5, 2.5))
-p4 <- ggplot(subset(resDall, species=="POSE"))+
-  geom_hline(aes(yintercept=0), linetype=2)+
-  geom_boxplot(aes(x=year, y=resids, fill=simtype), outlier.shape = 1, outlier.size = 0, color="grey25")+
-  xlab("Year")+
-  ylab("Residuals (% cover difference)")+
-  theme_bw()+
-  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
-  scale_fill_manual(values=myCols)+
-  ggtitle(expression(italic("Poa secunda")))+
-  guides(fill=FALSE)+
-  coord_cartesian(ylim = c(-50, 10))
-pFinal <- grid.arrange(p1,p2,p3,p4, ncol=2, nrow=2)
+# myCols <- c("#237DA4", "#7ECAD0")
+# p1 <- ggplot(subset(resDall, species=="BOGR"))+
+#   geom_hline(aes(yintercept=0), linetype=2)+
+#   geom_boxplot(aes(x=as.character(year), y=resids, fill=simtype), outlier.shape = 1, outlier.size = 0, color="grey25")+
+#   xlab("Year")+
+#   ylab("Residuals (% cover difference)")+
+#   theme_bw()+
+#   theme(axis.text.x = element_text(angle = 45, hjust = 1))+
+#   scale_fill_manual(values=myCols)+
+#   coord_cartesian(ylim = c(-10, 10))+
+#   ggtitle(expression(italic("B. gracilis")))+
+#   guides(fill=FALSE)
+# p2 <- ggplot(subset(resDall, species=="HECO"))+
+#   geom_hline(aes(yintercept=0), linetype=2)+
+#   geom_boxplot(aes(x=as.character(year), y=resids, fill=simtype), outlier.shape = 1, outlier.size = 0, color="grey25")+
+#   xlab("Year")+
+#   ylab("Residuals (% cover difference)")+
+#   theme_bw()+
+#   theme(axis.text.x = element_text(angle = 45, hjust = 1))+
+#   scale_fill_manual(values=myCols, labels=c("Year effects", "No year effects"))+
+#   ggtitle(expression(italic("H. comita")))+
+#   coord_cartesian(ylim = c(-0.5, 0.5))+
+#   guides(fill=guide_legend(title=""))+
+#   theme(legend.position=c(0.7,0.2))
+# p3 <- ggplot(subset(resDall, species=="PASM"))+
+#   geom_hline(aes(yintercept=0), linetype=2)+
+#   geom_boxplot(aes(x=as.character(year), y=resids, fill=simtype), outlier.shape = 1, outlier.size = 0, color="grey25")+
+#   xlab("Year")+
+#   ylab("Residuals (% cover difference)")+
+#   theme_bw()+
+#   theme(axis.text.x = element_text(angle = 45, hjust = 1))+
+#   scale_fill_manual(values=myCols)+
+#   ggtitle(expression(italic("P. smithii")))+
+#   guides(fill=FALSE)+
+#   coord_cartesian(ylim = c(-2.5, 2.5))
+# p4 <- ggplot(subset(resDall, species=="POSE"))+
+#   geom_hline(aes(yintercept=0), linetype=2)+
+#   geom_boxplot(aes(x=as.character(year), y=resids, fill=simtype), outlier.shape = 1, outlier.size = 0, color="grey25")+
+#   xlab("Year")+
+#   ylab("Residuals (% cover difference)")+
+#   theme_bw()+
+#   theme(axis.text.x = element_text(angle = 45, hjust = 1))+
+#   scale_fill_manual(values=myCols)+
+#   ggtitle(expression(italic("Poa secunda")))+
+#   guides(fill=FALSE)+
+#   coord_cartesian(ylim = c(-50, 10))
+# pFinal <- grid.arrange(p1,p2,p3,p4, ncol=2, nrow=2)
 # pFinal

### 2015-01-21 manuscript/manuscript.Rmd
just plot changes

@@
 # myCols <- c("#237DA4", "#7ECAD0")
-# p1 <- ggplot(subset(resDall, species=="BOGR"))+
-#   geom_hline(aes(yintercept=0), linetype=2)+
-#   geom_boxplot(aes(x=as.character(year), y=resids, fill=simtype), outlier.shape = 1, outlier.size = 0, color="grey25")+
-#   xlab("Year")+
-#   ylab("Residuals (% cover difference)")+
-#   theme_bw()+
-#   theme(axis.text.x = element_text(angle = 45, hjust = 1))+
-#   scale_fill_manual(values=myCols)+
-#   coord_cartesian(ylim = c(-10, 10))+
-#   ggtitle(expression(italic("B. gracilis")))+
-#   guides(fill=FALSE)
+p1 <- ggplot(subset(resDall, species=="BOGR"))+
+  geom_hline(aes(yintercept=0), linetype=2)+
+  geom_boxplot(aes(x=as.character(year), y=resids, fill=simtype), outlier.shape = 1, outlier.size = 0, color="grey25")+
+  xlab("Year")+
+  ylab("Residuals (% cover difference)")+
+  theme_bw()+
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
+  scale_fill_manual(values=myCols)+
+  coord_cartesian(ylim = c(-1, 1))+
+  ggtitle(expression(italic("B. gracilis")))+
+  guides(fill=FALSE)
 # p2 <- ggplot(subset(resDall, species=="HECO"))+

### 2015-01-22 manuscript/manuscript.Rmd
updated plot to remove 0 cover quads

--- 2015-01-20 manuscript/manuscript.Rmd
@@
 resDall <- rbind(resD, resDny)
+noZ <- which(resDall$lagCover!=0)
+resDall <- resDall[noZ,]
+
+myCols <- c("#237DA4", "#7ECAD0")
+# ggplot(resDall, aes(x=resids, fill=simtype))+
+#   geom_histogram(alpha=0.5)+
+#   facet_grid(.~species, scales = "free_x")+
+#   xlab("Year")+
+#   ylab("Residuals (% cover difference)")+
+#   theme_bw()+
+#   scale_fill_manual(values=myCols)
+
 # nSims <- length(unique(resD$sim))
-diffD <- ddply(resDall, c("year", "species", "simtype"), summarise,
-               avgDiff = mean(resids, na.rm = TRUE),
+# diffD <- ddply(resDall, c("year", "species", "simtype"), summarise,
+#                avgDiff = median(resids, na.rm = TRUE),
 #                upDiff75 = quantile(resids, probs = 0.8775, names=FALSE, na.rm = TRUE),
 #                dnDiff75 = quantile(resids, probs = 0.1225, names=FALSE, na.rm = TRUE),
-               dnDiff95 = quantile(resids, probs = 0.025, names=FALSE, na.rm = TRUE),
-               upDiff95 = quantile(resids, probs = 0.975, names=FALSE, na.rm = TRUE))
+#                dnDiff95 = quantile(resids, probs = 0.025, names=FALSE, na.rm = TRUE),
+#                upDiff95 = quantile(resids, probs = 0.975, names=FALSE, na.rm = TRUE))
+
 # ggplot(diffD, aes(color=simtype))+
 #   geom_hline(aes(yintercept=0), linetype=2, color="grey50")+
-#   geom_point(aes(x=year, y=avgDiff), size=3, position=position_dodge(width=0.5))+
-#   geom_errorbar(aes(x=year, ymax=upDiff95, ymin=dnDiff95), width=0.00001, position=position_dodge(width=0.5))+
-# #   geom_errorbar(aes(x=year, ymax=upDiff75, ymin=dnDiff75), size=1, width=0.00001, position=position_dodge(width=0.5))+
+#   geom_point(aes(x=as.character(year), y=avgDiff), size=3, position=position_dodge(width=0.5))+
+#   geom_errorbar(aes(x=as.character(year), ymax=upDiff95, ymin=dnDiff95), width=0.00001, position=position_dodge(width=0.5))+
+#   geom_errorbar(aes(x=as.character(year), ymax=upDiff75, ymin=dnDiff75), size=1, width=0.00001, position=position_dodge(width=0.5))+
 #   facet_grid(species~., scales = "free_y")+
@@
 #   theme_bw()+
+#   scale_color_manual(values=myCols)+
 #   theme(axis.text.x = element_text(angle = 45, hjust = 1))
@@
 #   theme(axis.text.x = element_text(angle = 45, hjust = 1))
-myCols <- c("#237DA4", "#7ECAD0")
+# myCols <- c("#237DA4", "#7ECAD0")
 p1 <- ggplot(subset(resDall, species=="BOGR"))+
   geom_hline(aes(yintercept=0), linetype=2)+
-  geom_boxplot(aes(x=year, y=resids, fill=simtype), outlier.shape = 1, outlier.size = 0, color="grey25")+
+  geom_boxplot(aes(x=as.character(year), y=resids, fill=simtype), outlier.shape = 1, outlier.size = 0, color="grey25")+
   xlab("Year")+
@@
   scale_fill_manual(values=myCols)+
-#   coord_cartesian(ylim = c(-50, 45))+
+  coord_cartesian(ylim = c(-50, 50))+
   ggtitle(expression(italic("B. gracilis")))+
@@
   geom_hline(aes(yintercept=0), linetype=2)+
-  geom_boxplot(aes(x=year, y=resids, fill=simtype), outlier.shape = 1, outlier.size = 0, color="grey25")+
+  geom_boxplot(aes(x=as.character(year), y=resids, fill=simtype), outlier.shape = 1, outlier.size = 0, color="grey25")+
   xlab("Year")+
@@
   guides(fill=guide_legend(title=""))+
-  theme(legend.position=c(0.7,0.2))
+  theme(legend.position=c(0.5,0.2))
 p3 <- ggplot(subset(resDall, species=="PASM"))+
   geom_hline(aes(yintercept=0), linetype=2)+
-  geom_boxplot(aes(x=year, y=resids, fill=simtype), outlier.shape = 1, outlier.size = 0, color="grey25")+
+  geom_boxplot(aes(x=as.character(year), y=resids, fill=simtype), outlier.shape = 1, outlier.size = 0, color="grey25")+
   xlab("Year")+
@@
   guides(fill=FALSE)+
-  coord_cartesian(ylim = c(-2.5, 2.5))
+  coord_cartesian(ylim = c(-1, 1))
 p4 <- ggplot(subset(resDall, species=="POSE"))+
   geom_hline(aes(yintercept=0), linetype=2)+
-  geom_boxplot(aes(x=year, y=resids, fill=simtype), outlier.shape = 1, outlier.size = 0, color="grey25")+
+  geom_boxplot(aes(x=as.character(year), y=resids, fill=simtype), outlier.shape = 1, outlier.size = 0, color="grey25")+
   xlab("Year")+
@@
   guides(fill=FALSE)+
-  coord_cartesian(ylim = c(-50, 10))
+  coord_cartesian(ylim = c(-10, 15))
 pFinal <- grid.arrange(p1,p2,p3,p4, ncol=2, nrow=2)
-# pFinal
+pFinal
 
@@
 We used the median of simulated results because the average was highly influence by some extreme outliers.", fig.width=4.5, fig.height=4, cache=FALSE, results='hide'}
-climBase <- readRDS("../montana/quadBM/simulations/climateChangeBase.rds")
-climPpt <- readRDS("../montana/quadBM/simulations/climateChangePrecip.rds")
-climTemp <- readRDS("../montana/quadBM/simulations/climateChangeTemperature.rds")
-climBoth <- readRDS("../montana/quadBM/simulations/climateChangePptTmp.rds")
-climBase$sim <- rep("Observed", nrow(climBase)) 
-climPpt$sim <- rep("Precipitation", nrow(climBase)) 
-climTemp$sim <- rep("Temperature", nrow(climBase)) 
-climBoth$sim <- rep("Both", nrow(climBase)) 
-climSimsQBM <- rbind(climBase, climPpt, climTemp, climBoth)
-# ggplot(climSimsQBM)+
-#   geom_boxplot(aes(x=species, y=cover*100, fill=sim))+
-#   scale_y_continuous(limits=c(0,10))
-climSimsQBMavg <- ddply(climSimsQBM, .(sim, species), summarise,
-                        avgCover = median(cover))
-climSimQBMcast <- dcast(climSimsQBMavg, formula = species~sim)
-climSimQBMcast$diffP <- with(climSimQBMcast, (Precipitation*100)-(Observed*100)) 
-climSimQBMcast$diffT <- with(climSimQBMcast, (Temperature*100)-(Observed*100))
-climSimQBMcast$diffB <- with(climSimQBMcast, (Both*100)-(Observed*100))
-climSimQBMplot <- melt(climSimQBMcast)
-climSimQBMplot <- subset(climSimQBMplot, species!="POSE")
-myCols2 <- c("#277BA8", "#7ABBBD", "#AED77A")
-ggplot(data=subset(climSimQBMplot, variable=="diffP"|variable=="diffT"|variable=="diffB"))+
-  geom_bar(aes(x=species, y=value, fill=variable), stat="identity", position="dodge", color="white")+
-  scale_fill_manual(name="", labels=c("1% Precipitation Increase", "1% Temperature Increase", "1% Temp. & Ppt. Increase"), values=myCols2)+
-  theme_bw()+
-  theme(legend.position=c(0.6, 0.8))+
-  ylab("Cover Change (%)")+
-  xlab("Species")
+# climBase <- readRDS("../montana/quadBM/simulations/climateChangeBase.rds")
+# climPpt <- readRDS("../montana/quadBM/simulations/climateChangePrecip.rds")
+# climTemp <- readRDS("../montana/quadBM/simulations/climateChangeTemperature.rds")
+# climBoth <- readRDS("../montana/quadBM/simulations/climateChangePptTmp.rds")
+# climBase$sim <- rep("Observed", nrow(climBase)) 
+# climPpt$sim <- rep("Precipitation", nrow(climBase)) 
+# climTemp$sim <- rep("Temperature", nrow(climBase)) 
+# climBoth$sim <- rep("Both", nrow(climBase)) 
+# climSimsQBM <- rbind(climBase, climPpt, climTemp, climBoth)
+# # ggplot(climSimsQBM)+
+# #   geom_boxplot(aes(x=species, y=cover*100, fill=sim))+
+# #   scale_y_continuous(limits=c(0,10))
+# climSimsQBMavg <- ddply(climSimsQBM, .(sim, species), summarise,
+#                         avgCover = median(cover))
+# climSimQBMcast <- dcast(climSimsQBMavg, formula = species~sim)
+# climSimQBMcast$diffP <- with(climSimQBMcast, (Precipitation*100)-(Observed*100)) 
+# climSimQBMcast$diffT <- with(climSimQBMcast, (Temperature*100)-(Observed*100))
+# climSimQBMcast$diffB <- with(climSimQBMcast, (Both*100)-(Observed*100))
+# climSimQBMplot <- melt(climSimQBMcast)
+# climSimQBMplot <- subset(climSimQBMplot, species!="POSE")
+# myCols2 <- c("#277BA8", "#7ABBBD", "#AED77A")
+# ggplot(data=subset(climSimQBMplot, variable=="diffP"|variable=="diffT"|variable=="diffB"))+
+#   geom_bar(aes(x=species, y=value, fill=variable), stat="identity", position="dodge", color="white")+
+#   scale_fill_manual(name="", labels=c("1% Precipitation Increase", "1% Temperature Increase", "1% Temp. & Ppt. Increase"), values=myCols2)+
+#   theme_bw()+
+#   theme(legend.position=c(0.6, 0.8))+
+#   ylab("Cover Change (%)")+
+#   xlab("Species")
 ```

### 2015-01-22 manuscript/manuscript.tex
updated plot to remove 0 cover quads

@@
   \newcommand{\euro}{€}
+    \setmainfont{times}
 \fi
@@
 
+The statistical approach used to model vital rates using aggregated data
+depends on the type of data collected. In our case, and as is often the
+case with census data, we have percent cover data (which can easily be
+transformed to proportion data, of course). We first considered fitting
+three vital rate models analagous to those we fit at the individual
+level: one for probability of extirpation within a quadrat (analagous to
+survival), one for cover change within a quadrat (analagous to growth),
+and one for probability of colonization within a quadrat (analagous to
+recruitment). However, within-quadrat extirpation and colonization
+events were rare in our time series (N=X and N=X, respectively). Given
+the broad spatial distribution of the quadrats we are studying, it is
+safe to assume that these events are in fact rare enough to be ignored
+for our purposes. So we constrained our statistical modeling of vital
+rates at the population level to change in percent cover within
+quadrats. For most of the remaining discussion of statistical modeling
+we refer to proportion data, which is simply percent data divided by
+100.
+
+An obvious choice for fitting a linear model to proportion data is beta
+regression because the support of the beta distribution is {[}0,1{]},
+not including true zeros. However, when we used fitted model parameters
+from a beta regression in quadrat-based population model the simulated
+population tended toward 100\% cover for all species. We therefore chose
+a more-constrained modeling approach based on a truncated log-normal
+likelihood. The model is
+
+\begin{align}
+x_{jq,t+1} &= \gamma^{G}_{j,t} + \phi^{G}_{jQ} + \beta^{G}_{j,t}x_{jq,t} + \theta^{S}_{jk}C_{k,t} \\
+Y^{G}_{jq,t+1} &\sim \text{LogNormal}(x_{jq,t+1}, \tau{j}) T(0,1)
+\end{align}
+
 At the quadrat level we defined three vital rates:
@@
 
-\begin{figure}[htbp]
-\centering
-\includegraphics{components/figure/manuscript-figure_2.pdf}
-\caption{Simulated changes species' median cover caused by a 1\%
-increase in observed precipitation or temperature. We used the median of
-simulated results because the average was highly influence by some
-extreme outliers.}
-\end{figure}
-
 \pagebreak{}

### 2015-01-22 manuscript/manuscript.Rmd
new climate change figure

@@
 We used the median of simulated results because the average was highly influence by some extreme outliers.", fig.width=4.5, fig.height=4, cache=FALSE, results='hide'}
-# climBase <- readRDS("../montana/quadBM/simulations/climateChangeBase.rds")
-# climPpt <- readRDS("../montana/quadBM/simulations/climateChangePrecip.rds")
-# climTemp <- readRDS("../montana/quadBM/simulations/climateChangeTemperature.rds")
-# climBoth <- readRDS("../montana/quadBM/simulations/climateChangePptTmp.rds")
-# climBase$sim <- rep("Observed", nrow(climBase)) 
-# climPpt$sim <- rep("Precipitation", nrow(climBase)) 
-# climTemp$sim <- rep("Temperature", nrow(climBase)) 
-# climBoth$sim <- rep("Both", nrow(climBase)) 
-# climSimsQBM <- rbind(climBase, climPpt, climTemp, climBoth)
-# # ggplot(climSimsQBM)+
-# #   geom_boxplot(aes(x=species, y=cover*100, fill=sim))+
-# #   scale_y_continuous(limits=c(0,10))
+climBase <- readRDS("../montana/quadBM/simulations/baselineSimulationQBM.rds")
+climPpt <- readRDS("../montana/quadBM/simulations/pptChangeSimulationQBM.rds")
+climTemp <- readRDS("../montana/quadBM/simulations/tempChangeSimulationQBM.rds")
+climBoth <- readRDS("../montana/quadBM/simulations/temp-pptChangeSimulationQBM.rds")
+climBase$sim <- rep("AObserved", nrow(climBase)) 
+climPpt$sim <- rep("BPrecipitation", nrow(climBase)) 
+climTemp$sim <- rep("CTemperature", nrow(climBase)) 
+climBoth$sim <- rep("DBoth", nrow(climBase)) 
+climSimsQBM <- rbind(climBase, climPpt, climTemp, climBoth)
+
+myCols2 <- c("grey45", "#277BA8", "#7ABBBD", "#AED77A")
+ggplot(climSimsQBM)+
+  geom_boxplot(aes(x=species, y=cover*100, fill=sim), outlier.size=0)+
+  theme_bw()+
+  coord_cartesian(ylim=c(0,50))+
+  scale_fill_manual(name="", labels=c("Observed", "1% Precipitation Increase", "1% Temperature Increase", "1% Temp. & Ppt. Increase"), values=myCols2)
+
 # climSimsQBMavg <- ddply(climSimsQBM, .(sim, species), summarise,
@@
 # climSimQBMplot <- subset(climSimQBMplot, species!="POSE")
-# myCols2 <- c("#277BA8", "#7ABBBD", "#AED77A")
+# 
 # ggplot(data=subset(climSimQBMplot, variable=="diffP"|variable=="diffT"|variable=="diffB"))+

### 2015-01-22 manuscript/manuscript.Rmd
updated ms with new figs and methods text

@@
 However, for each vital rate statistical model we also define the likelihood model we use.
-For the likelihood models, Y is always the relevant vector of observations (e.g., whether a  genet survived [1] or not [0] from year $t$ to $t+1$).
+For the likelihood models, \textbf{Y} is always the relevant vector of observations (e.g., whether a  genet survived [1] or not [0] from year $t$ to $t+1$).
 
@@
 We first considered fitting three vital rate models analagous to those we fit at the individual level: one for probability of extirpation within a quadrat (analagous to survival), one for cover change within a quadrat (analagous to growth), and one for probability of colonization within a quadrat (analagous to recruitment).
-However, within-quadrat extirpation and colonization events were rare in our time series (N=X and N=X, respectively).
+However, within-quadrat extirpation and colonization events were rare in our time series ($N=9$ and $N=10$, respectively).
 Given the broad spatial distribution of the quadrats we are studying, it is safe to assume that these events are in fact rare enough to be ignored for our purposes.
@@
 
-An obvious choice for fitting a linear model to proportion data is beta regression because the support of the beta distribution is [0,1], not including true zeros.
+An obvious choice for fitting a linear model to proportion data is beta regression because the support of the beta distribution is [0,1], not including true zeros or ones.
 However, when we used fitted model parameters from a beta regression in quadrat-based population model the simulated population tended toward 100% cover for all species.
-We therefore chose a more-constrained modeling approach based on a truncated log-normal likelihood.
-The model is
+We therefore chose a more constrained modeling approach based on a truncated log-normal likelihood.
+The model for quadrat cover change ($G$) from time $t$ to $t+1$
 
@@
 x_{jq,t+1} &= \gamma^{G}_{j,t} + \phi^{G}_{jQ} + \beta^{G}_{j,t}x_{jq,t} + \theta^{S}_{jk}C_{k,t} \\
-Y^{G}_{jq,t+1} &\sim \text{LogNormal}(x_{jq,t+1}, \tau{j}) T(0,1)
+Y^{G}_{jq,t+1} &\sim \text{LogNormal}(x_{jq,t+1}, \tau{j}) T[0,1]
 \end{align}
 
-At the quadrat level we defined three vital rates:
-
-1. Probability of extirpation ($S$): the probability that, for a given species, a particular quadrat will go from non-zero cover at time $t$ to zero cover at time $t+1$.
-2. Cover change ($G$): the change in percent cover from time $t$ to $t+1$ for a given species within a particular quadrat.
-3. Probability of colonization ($R$): the probability that, for a give species, a particular quadrat will go from zero cover at time $t$ to non-zero cover at time $t+1$.
-
-We retain the abbreviations $S$, $G$, and $R$ from the analogous processes at the individual level.
-The vital rate models at the quadrat level are all based on quadrat proportional cover.
-Also, at the quadrat level we do not need to explicitly include a density dependent term.
-Since we are modeling proportional cover, we essentially get density-dependence for "free" when proportional cover is included as a covariate.
-That is, density-dependence emerges because quadrats with low cover generally increase in cover whereas quadrats with high cover generally decrease in cover.
-
-We modeled the probability of extirpation of species $j$ in quadrat $q$ from time $t$ to $t+1$ as:
-
-\begin{align}
-\text{logit}(S_{jq,t}) &= \gamma^{S}_{j} + \phi^{S}_{jQ} + \beta^{S}_{j}x_{jq,t} + \theta^{S}_{jk}C_{k,t} \\
-Y^{S}_{jq,t} &\sim \text{Bernoulli}(S_{jq,t})
-\end{align}
-
-where all parameters are as in Eq.
-1 except that $x$ is now quadrat proportional cover.
-Note, however, that we do not include year random effects on the intercept $\gamma$ or the slope term $\beta$ for quadrat proportional cover.
-The quadrat data is inherently more sparse than the individual data from which it is aggregated, and this is especially evident when modeling rare events like extirpation and colonization.
-Thus, when we tried to fit random year effects, those terms did not converge.
-
-We modeled quadrat cover change ($G$) from time $t$ to $t+1$ as:
-
-\begin{equation}
-\text{logit}(x_{jq,t+1}) = \gamma^{G}_{j,t} + \phi^{G}_{jQ} + \beta^{G}_{j,t}x_{jq,t} + \theta^{S}_{jk}C_{k,t}
-\end{equation}
-
-where, in this case, we do include random year effects on the intercept $\gamma$ and the slope term $\beta$.
-For cover change we had enough data for those terms to converge.
-Note that our model for quadrat cover change uses a logit transformation to link the expected cover at $t+1$ ($x_{jq,t}$) to the linear predictors.
-We do so because during model fitting we use a beta likelihood since the data, proportional cover, is beta distributed.
-The beta likelihood requires shape ($\rho$) and rate ($\eta$) parameters that can be calculated using moment-matching:
-
-\begin{align}
-\rho_{jq,t+1} &= x_{jq,t+1}\tau_{j} \\
-\eta_{jq,t+1} &= (1-x_{jq,t+1})\tau_{j}
-\end{align}
-
-with likelihood:
-
-\begin{equation}
-Y^{G}_{jq,t+1} \sim \text{Beta}(\rho_{jq,t+1}, \eta_{jq,t+1}).
-\end{equation}
-
-Finally, we modeled probability of colonization in quadrat $q$ by species $j$ from time $t$ to $t+1$ as:
-
-\begin{align}
-\text{logit}(R_{jq,t}) &= \gamma^{R}_{j} + \phi^{R}_{jQ} + \theta^{R}_{jk}C_{k,t} \\
-Y^{R}_{jq,t} &\sim \text{Bernoulli}(R_{jq,t}).
-\end{align}
+where $x_{jq,t}$ is the log of species' $j$ proportional cover in quadrat $q$ at time $t$ and all other parameters are as in the individual-level growth model (Eq.
+#).
+The log normal likelihood includes a truncation ($T[0,1]) to ensure that predicted values do not exceed 100% cover.
 
 ### Model fitting
-Our Bayesian approach to fitting the vital rate models required choosing appropriate priors for unknown parameters and deciding which, if any, of those prior should be hierarchical.
-We decided to fit models where all terms except climate covariates were fit by species, while the climate covariates were fit hierarchically where species-specific coefficients were drawn from a shared `global' coefficient distribution.
+Our Bayesian approach to fitting the vital rate models required choosing appropriate priors for unknown parameters and deciding which, if any, of those priors should be hierarchical.
+We decided to fit models where all terms except climate covariates were fit by species, while the climate covariates were fit hierarchically with species-specific coefficients drawn from a shared `global' coefficient distribution.
 We did so for two reasons: (1) the four focal species are all perennial grasses that we expect to respond similarly to climate covariates, and (2) convergence of climate effects at the quadrat level was much easier to achieve when we modeled these terms hierarchically, allowing them to "share" statistical strength via partial pooling [@Gelman2007].
@@
 We assessed convergence visually and using the @Gelman1992 diagnostic in the R package `coda' [@Plummer2006].
-Scale reduction factors for all parameters were less than 1.02, indicating convergence.
-For the purposes of introducing stochasticity in our population models, we saved the final 1,000 iterations from each chain for all parameters to be used as randomly drawn values during population simulation.
-
-We assessed the statistical importance of the climate covariates included the final vital rate regressions by comparing the residual deviance of models with climate covariates and temporal random effects, climate covariates only, and temporal random effects only.
-When a model includes climate covariates, this comparison shows the relative contribution of the climate covariates in explaining the total interannual variability [@Adler2012].
+Scale reduction factors for all parameters were less than 1.01, indicating convergence.
+For the purposes of introducing stochasticity in our population models, we saved the final 1,000 iterations from each of the three MCMC chains for all parameters to be used as randomly drawn values during population simulation.
 
@@
 
-We used random draws from the final 1,000 iterations from each of 3 MCMC chains to introduce stochasticity into our population models.
+We used random draws from the final 1,000 iterations from each of three MCMC chains to introduce stochasticity into our population models.
 At each time step, we first randomly selected climate covariates from one of the 14 observed years and also randomly drew one set of random year effects for each vital rate regression.
@@
 -------
-### Climate effects on vital rates
-At the individual level we fit vital rate models that included climate covariates and random year effects (full model), just climate covariates (climate model), and no climate or random year effects (constant model).
-We used the metric from @Adler2012 to quantify the contribution of climate covariates to explaining interannual variability in vital rates based on comparisons of model deviance: (Climate model - Constant Model)/(Full model - Constant model).
-At the quadrat level, we could only fit random year effects for the growth regression.
-This is, in fact, one of the limitation of census-based data: local extirpation and colonization events are likely to be rare, making it difficult to fit complex statistical models.
-So, in this section we focus on growth only.
-
-Climate covariates improved the growth regressions at both levels, with a slightly greater contribution at the individual level (Figure 1).
-Parameter estimates were generally consistent in terms of sign and magnitude between the individual level and the quadrat level for each species and climate covariate (Figure 2).
-There is no clear pattern suggesting that parameters estimated using individual level data are more certain, as evidenced by credible interval widths, than estimates based on quadrat data.
-
-### Climate effects on population dynamics
 
@@
   guides(fill=FALSE)+
-  coord_cartesian(ylim = c(-1, 1))
+  coord_cartesian(ylim = c(-1.5, 1.5))
 p4 <- ggplot(subset(resDall, species=="POSE"))+
@@
 ```{r figure_2, dependson="plot-options", fig.cap="Simulated changes species' median cover caused by a 1% increase in observed precipitation or temperature.
-We used the median of simulated results because the average was highly influence by some extreme outliers.", fig.width=4.5, fig.height=4, cache=FALSE, results='hide'}
+We used the median of simulated results because the average was highly influence by some extreme outliers.", fig.width=6.5, fig.height=4, cache=FALSE, results='hide'}
 climBase <- readRDS("../montana/quadBM/simulations/baselineSimulationQBM.rds")

### 2015-01-22 manuscript/manuscript.tex
updated ms with new figs and methods text

@@
 model we also define the likelihood model we use. For the likelihood
-models, Y is always the relevant vector of observations (e.g., whether a
-genet survived {[}1{]} or not {[}0{]} from year $t$ to $t+1$).
+models, \textbf{Y} is always the relevant vector of observations (e.g.,
+whether a genet survived {[}1{]} or not {[}0{]} from year $t$ to $t+1$).
 
@@
 recruitment). However, within-quadrat extirpation and colonization
-events were rare in our time series (N=X and N=X, respectively). Given
-the broad spatial distribution of the quadrats we are studying, it is
-safe to assume that these events are in fact rare enough to be ignored
-for our purposes. So we constrained our statistical modeling of vital
-rates at the population level to change in percent cover within
+events were rare in our time series ($N=9$ and $N=10$, respectively).
+Given the broad spatial distribution of the quadrats we are studying, it
+is safe to assume that these events are in fact rare enough to be
+ignored for our purposes. So we constrained our statistical modeling of
+vital rates at the population level to change in percent cover within
 quadrats. For most of the remaining discussion of statistical modeling
@@
 regression because the support of the beta distribution is {[}0,1{]},
-not including true zeros. However, when we used fitted model parameters
-from a beta regression in quadrat-based population model the simulated
-population tended toward 100\% cover for all species. We therefore chose
-a more-constrained modeling approach based on a truncated log-normal
-likelihood. The model is
+not including true zeros or ones. However, when we used fitted model
+parameters from a beta regression in quadrat-based population model the
+simulated population tended toward 100\% cover for all species. We
+therefore chose a more constrained modeling approach based on a
+truncated log-normal likelihood. The model for quadrat cover change
+($G$) from time $t$ to $t+1$
 
@@
 x_{jq,t+1} &= \gamma^{G}_{j,t} + \phi^{G}_{jQ} + \beta^{G}_{j,t}x_{jq,t} + \theta^{S}_{jk}C_{k,t} \\
-Y^{G}_{jq,t+1} &\sim \text{LogNormal}(x_{jq,t+1}, \tau{j}) T(0,1)
+Y^{G}_{jq,t+1} &\sim \text{LogNormal}(x_{jq,t+1}, \tau{j}) T[0,1]
 \end{align}
 
-At the quadrat level we defined three vital rates:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\itemsep1pt\parskip0pt\parsep0pt
-\item
-  Probability of extirpation ($S$): the probability that, for a given
-  species, a particular quadrat will go from non-zero cover at time $t$
-  to zero cover at time $t+1$.
-\item
-  Cover change ($G$): the change in percent cover from time $t$ to $t+1$
-  for a given species within a particular quadrat.
-\item
-  Probability of colonization ($R$): the probability that, for a give
-  species, a particular quadrat will go from zero cover at time $t$ to
-  non-zero cover at time $t+1$.
-\end{enumerate}
-
-We retain the abbreviations $S$, $G$, and $R$ from the analogous
-processes at the individual level. The vital rate models at the quadrat
-level are all based on quadrat proportional cover. Also, at the quadrat
-level we do not need to explicitly include a density dependent term.
-Since we are modeling proportional cover, we essentially get
-density-dependence for ``free'' when proportional cover is included as a
-covariate. That is, density-dependence emerges because quadrats with low
-cover generally increase in cover whereas quadrats with high cover
-generally decrease in cover.
-
-We modeled the probability of extirpation of species $j$ in quadrat $q$
-from time $t$ to $t+1$ as:
-
-\begin{align}
-\text{logit}(S_{jq,t}) &= \gamma^{S}_{j} + \phi^{S}_{jQ} + \beta^{S}_{j}x_{jq,t} + \theta^{S}_{jk}C_{k,t} \\
-Y^{S}_{jq,t} &\sim \text{Bernoulli}(S_{jq,t})
-\end{align}
-
-where all parameters are as in Eq. 1 except that $x$ is now quadrat
-proportional cover. Note, however, that we do not include year random
-effects on the intercept $\gamma$ or the slope term $\beta$ for quadrat
-proportional cover. The quadrat data is inherently more sparse than the
-individual data from which it is aggregated, and this is especially
-evident when modeling rare events like extirpation and colonization.
-Thus, when we tried to fit random year effects, those terms did not
-converge.
-
-We modeled quadrat cover change ($G$) from time $t$ to $t+1$ as:
-
-\begin{equation}
-\text{logit}(x_{jq,t+1}) = \gamma^{G}_{j,t} + \phi^{G}_{jQ} + \beta^{G}_{j,t}x_{jq,t} + \theta^{S}_{jk}C_{k,t}
-\end{equation}
-
-where, in this case, we do include random year effects on the intercept
-$\gamma$ and the slope term $\beta$. For cover change we had enough data
-for those terms to converge. Note that our model for quadrat cover
-change uses a logit transformation to link the expected cover at $t+1$
-($x_{jq,t}$) to the linear predictors. We do so because during model
-fitting we use a beta likelihood since the data, proportional cover, is
-beta distributed. The beta likelihood requires shape ($\rho$) and rate
-($\eta$) parameters that can be calculated using moment-matching:
-
-\begin{align}
-\rho_{jq,t+1} &= x_{jq,t+1}\tau_{j} \\
-\eta_{jq,t+1} &= (1-x_{jq,t+1})\tau_{j}
-\end{align}
-
-with likelihood:
-
-\begin{equation}
-Y^{G}_{jq,t+1} \sim \text{Beta}(\rho_{jq,t+1}, \eta_{jq,t+1}).
-\end{equation}
-
-Finally, we modeled probability of colonization in quadrat $q$ by
-species $j$ from time $t$ to $t+1$ as:
-
-\begin{align}
-\text{logit}(R_{jq,t}) &= \gamma^{R}_{j} + \phi^{R}_{jQ} + \theta^{R}_{jk}C_{k,t} \\
-Y^{R}_{jq,t} &\sim \text{Bernoulli}(R_{jq,t}).
-\end{align}
+where $x_{jq,t}$ is the log of species' $j$ proportional cover in
+quadrat $q$ at time $t$ and all other parameters are as in the
+individual-level growth model (Eq. \#). The log normal likelihood
+includes a truncation (\$T{[}0,1{]}) to ensure that predicted values do
+not exceed 100\% cover.
 
@@
 appropriate priors for unknown parameters and deciding which, if any, of
-those prior should be hierarchical. We decided to fit models where all
+those priors should be hierarchical. We decided to fit models where all
 terms except climate covariates were fit by species, while the climate
-covariates were fit hierarchically where species-specific coefficients
-were drawn from a shared `global' coefficient distribution. We did so
-for two reasons: (1) the four focal species are all perennial grasses
-that we expect to respond similarly to climate covariates, and (2)
-convergence of climate effects at the quadrat level was much easier to
-achieve when we modeled these terms hierarchically, allowing them to
-``share'' statistical strength via partial pooling (Gelman and Hill
-2007). So, climate effects were modeled as:
+covariates were fit hierarchically with species-specific coefficients
+drawn from a shared `global' coefficient distribution. We did so for two
+reasons: (1) the four focal species are all perennial grasses that we
+expect to respond similarly to climate covariates, and (2) convergence
+of climate effects at the quadrat level was much easier to achieve when
+we modeled these terms hierarchically, allowing them to ``share''
+statistical strength via partial pooling (Gelman and Hill 2007). So,
+climate effects were modeled as:
 
@@
 Rubin (1992) diagnostic in the R package `coda' (Plummer et al. 2006).
-Scale reduction factors for all parameters were less than 1.02,
+Scale reduction factors for all parameters were less than 1.01,
 indicating convergence. For the purposes of introducing stochasticity in
-our population models, we saved the final 1,000 iterations from each
-chain for all parameters to be used as randomly drawn values during
-population simulation.
-
-We assessed the statistical importance of the climate covariates
-included the final vital rate regressions by comparing the residual
-deviance of models with climate covariates and temporal random effects,
-climate covariates only, and temporal random effects only. When a model
-includes climate covariates, this comparison shows the relative
-contribution of the climate covariates in explaining the total
-interannual variability (Adler et al. 2012).
+our population models, we saved the final 1,000 iterations from each of
+the three MCMC chains for all parameters to be used as randomly drawn
+values during population simulation.
 
@@
 
-We used random draws from the final 1,000 iterations from each of 3 MCMC
-chains to introduce stochasticity into our population models. At each
-time step, we first randomly selected climate covariates from one of the
-14 observed years and also randomly drew one set of random year effects
-for each vital rate regression. Climate years and random year effects
-were drawn independently because they were uncorrelated. Then, we drew
-the full parameter set (specific random year effects, climate effects,
-fixed effects) from a randomly selected MCMC iteration. We selected
-random year effects and the climate year independently.
+We used random draws from the final 1,000 iterations from each of three
+MCMC chains to introduce stochasticity into our population models. At
+each time step, we first randomly selected climate covariates from one
+of the 14 observed years and also randomly drew one set of random year
+effects for each vital rate regression. Climate years and random year
+effects were drawn independently because they were uncorrelated. Then,
+we drew the full parameter set (specific random year effects, climate
+effects, fixed effects) from a randomly selected MCMC iteration. We
+selected random year effects and the climate year independently.
 
 \subsection{Results}\label{results}
-
-\subsubsection{Climate effects on vital
-rates}\label{climate-effects-on-vital-rates}
-
-At the individual level we fit vital rate models that included climate
-covariates and random year effects (full model), just climate covariates
-(climate model), and no climate or random year effects (constant model).
-We used the metric from Adler et al. (2012) to quantify the contribution
-of climate covariates to explaining interannual variability in vital
-rates based on comparisons of model deviance: (Climate model - Constant
-Model)/(Full model - Constant model). At the quadrat level, we could
-only fit random year effects for the growth regression. This is, in
-fact, one of the limitation of census-based data: local extirpation and
-colonization events are likely to be rare, making it difficult to fit
-complex statistical models. So, in this section we focus on growth only.
-
-Climate covariates improved the growth regressions at both levels, with
-a slightly greater contribution at the individual level (Figure 1).
-Parameter estimates were generally consistent in terms of sign and
-magnitude between the individual level and the quadrat level for each
-species and climate covariate (Figure 2). There is no clear pattern
-suggesting that parameters estimated using individual level data are
-more certain, as evidenced by credible interval widths, than estimates
-based on quadrat data.
-
-\subsubsection{Climate effects on population
-dynamics}\label{climate-effects-on-population-dynamics}
 
@@
 
+\begin{figure}[htbp]
+\centering
+\includegraphics{components/figure/manuscript-figure_2.pdf}
+\caption{Simulated changes species' median cover caused by a 1\%
+increase in observed precipitation or temperature. We used the median of
+simulated results because the average was highly influence by some
+extreme outliers.}
+\end{figure}
+
 \pagebreak{}
@@
 \subsection{References}\label{references}
-
-Adler, P. B., H. J. Dalgleish, and S. P. Ellner. 2012. Forecasting plant
-community impacts of climate variability and change: when do competitive
-interactions matter? Journal of Ecology 100:478--487.
 

### 2015-02-25 manuscript/manuscript.Rmd
updated figures for climate change sim

@@
 
-```{r figure_2, dependson="plot-options", fig.cap="Simulated changes species' median cover caused by a 1% increase in observed precipitation or temperature.
-We used the median of simulated results because the average was highly influence by some extreme outliers.", fig.width=6.5, fig.height=4, cache=FALSE, results='hide'}
+```{r figure_2, dependson="plot-options", fig.cap="Changes in species' mean cover caused by a 1% increase in observed precipitation or temperature as predicted by the individual-based IPM and the aggregate-based QBM.
+Note that the change for POSE due to a precipitation increase predicted by the QBM is almost zero and so does not show up in the figure.", fig.width=6.5, fig.height=6, cache=FALSE, results='hide'}
 climBase <- readRDS("../montana/quadBM/simulations/baselineSimulationQBM.rds")
@@
 climSimsQBM <- rbind(climBase, climPpt, climTemp, climBoth)
+qbm_means <- ddply(climSimsQBM, .(species, sim), summarise,
+                   avg_cover = mean(cover))
+tmpid <- which(qbm_means$sim=="AObserved")
+diffs <- list()
+for(i in 1:length(tmpid)){
+  obs <- qbm_means[tmpid[i],"avg_cover"]
+  diffs[[i]] <- rep(obs,3) - qbm_means[(tmpid[i]+1):(tmpid[i]+3), "avg_cover"]
+}
+names(diffs) <- unique(qbm_means$species)
+diff_df <- melt(as.data.frame(diffs))
+diff_df$sim <- rep(c("pptChange", "tempChange", "temppptChange"),4)
+diff_df$type <- rep("QBM", nrow(diff_df))
+
+ipm_diffs <- readRDS("../montana/ipm/simulations/results/ipm_climatesims_percdiffs.rds")
+ipm_diffs$type <- rep("IPM", nrow(ipm_diffs))
+
+diff_df <- rbind(diff_df, ipm_diffs)
 
 myCols2 <- c("grey45", "#277BA8", "#7ABBBD", "#AED77A")
-ggplot(climSimsQBM)+
-  geom_boxplot(aes(x=species, y=cover*100, fill=sim), outlier.size=0)+
-  theme_bw()+
-  coord_cartesian(ylim=c(0,50))+
-  scale_fill_manual(name="", labels=c("Observed", "1% Precipitation Increase", "1% Temperature Increase", "1% Temp. & Ppt. Increase"), values=myCols2)
+# ggplot(climSimsQBM)+
+#   geom_boxplot(aes(x=species, y=cover*100, fill=sim), outlier.size=0)+
+#   theme_bw()+
+#   coord_cartesian(ylim=c(0,50))+
+#   scale_fill_manual(name="", labels=c("Observed", "1% Precipitation Increase", "1% Temperature Increase", "1% Temp. & Ppt. Increase"), values=myCols2)
+ggplot(diff_df, aes(x=sim, y=value, fill=type))+
+  geom_bar(stat="identity", position="dodge")+
+  geom_hline(aes(yintercept=0))+
+  scale_fill_manual(values = myCols2[c(2,3)], name="Model type")+
+  xlab("Climatic change")+
+  ylab("Cover change (%)")+
+  facet_wrap("variable", scales = "free")+
+  scale_x_discrete(labels=c("ppt", "temp", "ppt + tmp"))+
+  theme_bw()
+#   theme(axis.text.x = element_text(angle = 45, hjust = 1))
 

### 2015-02-25 manuscript/manuscript.tex
updated figures for climate change sim

@@
 \includegraphics{components/figure/manuscript-figure_2.pdf}
-\caption{Simulated changes species' median cover caused by a 1\%
-increase in observed precipitation or temperature. We used the median of
-simulated results because the average was highly influence by some
-extreme outliers.}
+\caption{Changes in species' mean cover caused by a 1\% increase in
+observed precipitation or temperature as predicted by the
+individual-based IPM and the aggregate-based QBM. Note that the change
+for POSE due to a precipitation increase predicted by the QBM is almost
+zero and so does not show up in the figure.}
 \end{figure}

### 2015-02-25 montana/ipm/simulations/one-step_ahead_wrapper.r

@@ -10,148 +10,150 @@
 root=ifelse(.Platform$OS.type=="windows","c:/repos","~/repos"); # modify as needed
 setwd(paste(root,"/MicroMesoForecast/montana/ipm/simulations",sep="")); # modify as needed 
 
-doSpp<-"HECO"
+# doSpp<-"BOGR"
 spp_list<-c("BOGR","HECO","PASM","POSE") # all Montana species
-reps<-1   # number of times to simulate each quadrat x year transition
+reps<-100   # number of times to simulate each quadrat x year transition
 nMCMC<-3000 # max number of MCMC iterations to draw parameters from
-outfile1<-paste(doSpp,"_sim_cover_1step_ahead_new.csv",sep="")
+
 #outfile2<-paste(doSpp,"_sim_density_1step_ahead.csv",sep="") # not implemented yet
 
 #--------------------------------------------------
 #  Set up simulation, Matrix size and initial vectors
 #--------------------------------------------------
-
-sppCode<-which(spp_list==doSpp)
-n_spp<-length(spp_list) # this is needed b/c all 4 spp parameters are imported at once
-
-Atotal<-10000 #area of quadrat in cm^2
-
-bigM=c(75,50,10,50)[sppCode]            #Set matrix dimension for each species
-maxSize=c(2500,120,25,100)[sppCode]    
-
-v=v.r=b.r=expv=Cr=WmatG=WmatS=NULL #a LIST for different species
-h=r.L=r.U=Ctot=NULL
-
-L=log(0.2)
-U=log(maxSize)*1.1     
+for(spp in 1:length(spp_list)){
+  doSpp <- spp_list[spp]
+  outfile1<-paste(doSpp,"_sim_cover_1step_ahead_new.csv",sep="")
+  sppCode<-which(spp_list==doSpp)
+  n_spp<-length(spp_list) # this is needed b/c all 4 spp parameters are imported at once
   
-# b---boundary points. Note: b chops up the size interval (L-U) into bigM-equal-sized portions.
-b = L+c(0:bigM)*(U-L)/bigM #log-transformed
-
-# mid points
-v = 0.5*(b[1:bigM]+b[2:(bigM+1)]) #log-transformed
-
-# step size for midpoint rule. (see equations 4 and 5 in Ellner and Rees (2006) Am Nat.)
-h = v[2]-v[1] #log-transformed 
-
[…]

### 2015-02-25 montana/ipm/simulations/results/clim_change_fig.R

@@ -7,6 +7,8 @@
 ####
 
 library(ggplot2)
+library(plyr)
+library(reshape2)
 
 
 files <- list.files()
@@ -26,3 +28,24 @@ all_sims <- all_sims[2:nrow(all_sims),]
 ggplot(all_sims)+
   geom_boxplot(aes(x=species, y=cover*100, fill=sim))+
   coord_cartesian(ylim = c(0,100))
+
+all_means <- ddply(all_sims, .(species, sim), summarise,
+                   avg_cover = mean(cover))
+tmpid <- which(all_means$sim=="observed")
+diffs <- list()
+for(i in 1:length(tmpid)){
+  obs <- all_means[tmpid[i],"avg_cover"]
+  diffs[[i]] <- rep(obs,3) - all_means[(tmpid[i]+1):(tmpid[i]+3), "avg_cover"]
+}
+names(diffs) <- unique(all_sims$species)
+diff_df <- melt(as.data.frame(diffs))
+diff_df$sim <- rep(c("pptChange", "tempChange", "temppptChange"),4)
+saveRDS(diff_df, "ipm_climatesims_percdiffs.rds")
+# myCols2 <- c("grey45", "#277BA8", "#7ABBBD", "#AED77A")
+# ggplot(diff_df, aes(x=variable, y=value, fill=sim))+
+#   geom_bar(stat="identity", position="dodge", color="white")+
+#   scale_fill_manual(values = myCols2[2:4])+
+#   xlab("Species")+
+#   ylab("Cover change (%)")+
+#   theme_bw()
+

### 2015-02-25 montana/ipm/vitalRateRegs/growth/modelVariance_postMCMC.R

@@ -53,6 +53,32 @@ crowd <- c(read.csv("BOGRgrowthCrowding.csv")[,2],
            read.csv("POSEgrowthCrowding.csv")[,2])
 growD$crowd <- crowd
 
+library(ggplot2)
+g1 <- ggplot(growD, aes(x=area.t0, y=area.t1))+
+  geom_point()+
+  stat_smooth(method="lm", se=FALSE)+
+  geom_abline(aes(intercept = 0, slope = 1), color="red", linetype=2)+
+  facet_wrap("species", scales = "free")+
+  ggtitle("Untransformed")
+g2 <- ggplot(growD, aes(x=log(area.t0), y=log(area.t1)))+
+  geom_point()+
+  stat_smooth(method="lm", se=FALSE)+
+  geom_abline(aes(intercept = 0, slope = 1), color="red", linetype=2)+
+  facet_wrap("species", scales = "free")+
+  ggtitle("Log ransformed")
+g3 <- ggplot(growD, aes(x=sqrt(area.t0), y=sqrt(area.t1)))+
+  geom_point()+
+  stat_smooth(method="lm", se=FALSE)+
+  geom_abline(aes(intercept = 0, slope = 1), color="red", linetype=2)+
+  facet_wrap("species", scales = "free")+
+  ggtitle("Sqrt transformed")
+pdf("transformations.pdf")
+print(g1)
+print(g2)
+print(g3)
+dev.off()
+
+
 ####
 #### Bring in mean MCMC-based parameters for predictions
 ####

### 2015-02-25 manuscript/manuscript.Rmd
updated figure to proportional change

--- 2015-01-22 manuscript/manuscript.Rmd
@@
 
-```{r figure_2, dependson="plot-options", fig.cap="Simulated changes species' median cover caused by a 1% increase in observed precipitation or temperature.
-We used the median of simulated results because the average was highly influence by some extreme outliers.", fig.width=6.5, fig.height=4, cache=FALSE, results='hide'}
+```{r figure_2, dependson="plot-options", fig.cap="Proportional change in species' mean cover caused by a 1% increase in observed precipitation, temperature, or both as predicted by the individual-based IPM and the aggregate-based QBM.
+Note that the change for POSE due to a precipitation increase predicted by the QBM is almost zero and so does not show up in the figure.
+Labels on x-axis refer to: 'ppt' = 1% increase in mean precipitation; 'temp' = 1% increase in mean temperature; 'ppt + temp' = 1% increase in mean precipitation and mean temperature.", fig.width=6.5, fig.height=6, cache=FALSE, results='hide'}
 climBase <- readRDS("../montana/quadBM/simulations/baselineSimulationQBM.rds")
@@
 climSimsQBM <- rbind(climBase, climPpt, climTemp, climBoth)
+qbm_means <- ddply(climSimsQBM, .(species, sim), summarise,
+                   avg_cover = mean(cover))
+tmpid <- which(qbm_means$sim=="AObserved")
+diffs <- list()
+for(i in 1:length(tmpid)){
+  obs <- qbm_means[tmpid[i],"avg_cover"]
+  diffs[[i]] <- (rep(obs,3) - qbm_means[(tmpid[i]+1):(tmpid[i]+3), "avg_cover"])/rep(obs,3)
+}
+names(diffs) <- unique(qbm_means$species)
+diff_df <- melt(as.data.frame(diffs))
+diff_df$sim <- rep(c("pptChange", "tempChange", "temppptChange"),4)
+diff_df$type <- rep("QBM", nrow(diff_df))
+
+ipm_diffs <- readRDS("../montana/ipm/simulations/results/ipm_climatesims_percdiffs.rds")
+ipm_diffs$type <- rep("IPM", nrow(ipm_diffs))
+
+diff_df <- rbind(diff_df, ipm_diffs)
 
 myCols2 <- c("grey45", "#277BA8", "#7ABBBD", "#AED77A")
-ggplot(climSimsQBM)+
-  geom_boxplot(aes(x=species, y=cover*100, fill=sim), outlier.size=0)+
-  theme_bw()+
-  coord_cartesian(ylim=c(0,50))+
-  scale_fill_manual(name="", labels=c("Observed", "1% Precipitation Increase", "1% Temperature Increase", "1% Temp. & Ppt. Increase"), values=myCols2)
+# ggplot(climSimsQBM)+
+#   geom_boxplot(aes(x=species, y=cover*100, fill=sim), outlier.size=0)+
+#   theme_bw()+
+#   coord_cartesian(ylim=c(0,50))+
+#   scale_fill_manual(name="", labels=c("Observed", "1% Precipitation Increase", "1% Temperature Increase", "1% Temp. & Ppt. Increase"), values=myCols2)
+ggplot(diff_df, aes(x=sim, y=value, fill=type))+
+  geom_bar(stat="identity", position="dodge")+
+  geom_hline(aes(yintercept=0))+
+  scale_fill_manual(values = myCols2[c(1,3)], name="Model type")+
+  xlab("Climatic change")+
+  ylab("Proportional change in percent cover")+
+  facet_wrap("variable", scales = "free")+
+  scale_x_discrete(labels=c("ppt", "temp", "ppt + tmp"))+
+  theme_bw()
+#   theme(axis.text.x = element_text(angle = 45, hjust = 1))
 

### 2015-02-25 manuscript/manuscript.tex
updated figure to proportional change

--- 2015-01-22 manuscript/manuscript.tex
@@
 \includegraphics{components/figure/manuscript-figure_2.pdf}
-\caption{Simulated changes species' median cover caused by a 1\%
-increase in observed precipitation or temperature. We used the median of
-simulated results because the average was highly influence by some
-extreme outliers.}
+\caption{Proportional change in species' mean cover caused by a 1\%
+increase in observed precipitation, temperature, or both as predicted by
+the individual-based IPM and the aggregate-based QBM. Note that the
+change for POSE due to a precipitation increase predicted by the QBM is
+almost zero and so does not show up in the figure. Labels on x-axis
+refer to: `ppt' = 1\% increase in mean precipitation; `temp' = 1\%
+increase in mean temperature; `ppt + temp' = 1\% increase in mean
+precipitation and mean temperature.}
 \end{figure}

### 2015-02-25 montana/ipm/simulations/results/clim_change_fig.R

@@ -35,7 +35,7 @@ tmpid <- which(all_means$sim=="observed")
 diffs <- list()
 for(i in 1:length(tmpid)){
   obs <- all_means[tmpid[i],"avg_cover"]
-  diffs[[i]] <- rep(obs,3) - all_means[(tmpid[i]+1):(tmpid[i]+3), "avg_cover"]
+  diffs[[i]] <- (rep(obs,3) - all_means[(tmpid[i]+1):(tmpid[i]+3), "avg_cover"])/rep(obs,3)
 }
 names(diffs) <- unique(all_sims$species)
 diff_df <- melt(as.data.frame(diffs))

### 2015-02-27 manuscript/manuscript.Rmd
updated ms with new figures from 1-steps

@@
 library(reshape2)
+library(xtable)
 ```
@@
 
+Tables
+------
+```{r table_1,results='asis',message=FALSE, comment=NA, echo=FALSE}
+# Make a table for average residual from each model and year effects structure
+ipm_onestep <- readRDS("../montana/ipm/simulations/one_step_results/ipm_one-step_forecasts_combined.rds")
+resD <- readRDS("../montana/quadBM/simulations/oneStepResids_YearEffects.rds")
+resDny <- readRDS("../montana/quadBM/simulations/oneStepResids_NoYearEffects.rds")
+resD$simtype <- rep("full", nrow(resD))
+resDny$simtype <- rep("noyear", nrow(resDny))
+resDall <- rbind(resD, resDny)
+noZ <- which(resDall$lagCover!=0)
+resDall <- resDall[noZ,]
+
+stats <- ddply(ipm_onestep, .(species, year), summarise,
+               mean_resid = mean(resid),
+               sd_resid = sd(resid),
+               min_resid = min(resid),
+               max_resid = max(resid))
+stats$model <- rep("IPM", nrow(stats))
+colnames(stats)[2] <- "simtype"
+stats2 <- ddply(resDall, .(species, simtype), summarise,
+               mean_resid = mean(resids),
+               sd_resid = sd(resids),
+               min_resid = min(resids),
+               max_resid = max(resids))
+stats2$model <- rep("QBM", nrow(stats2))
+stats$simtype <- stats2$simtype
+stats <- rbind(stats, stats2)
+stats <- stats[,c(ncol(stats),1:(ncol(stats)-1))]
+stats$simtype <- rep(c("Full", "Climate"),(nrow(stats)/2))
+colnames(stats) <- c("Model", "Species", "Year Structure", "Mean", "SD", "Min", "Max")
+print.xtable(xtable(stats, caption = "Table of residual summary statistics."),type="latex", comment=FALSE, include.rownames=FALSE)
+```
+
+\pagebreak{}
+
 Figures
@@
 
-```{r figure_2, dependson="plot-options", fig.cap="Proportional change in species' mean cover caused by a 1% increase in observed precipitation, temperature, or both as predicted by the individual-based IPM and the aggregate-based QBM.
+```{r figure_2, dependson="plot-options", fig.cap="One step forecasts from IPM version.",fig.width=8, fig.height=8, cache=FALSE, results='hide'}
+ipm_onestep <- readRDS("../montana/ipm/simulations/one_step_results/ipm_one-step_forecasts_combined.rds")
+ipm_onestep$t1 <- ipm_onestep$t1+1900
+p1 <- ggplot(subset(ipm_onestep, species=="BOGR"))+
+  geom_hline(aes(yintercept=0), linetype=2)+
+  geom_boxplot(aes(x=as.character(t1),y=resid,fill=year), outlier.shape = 1, outlier.size = 0, color="grey25")+
+  xlab("Year")+
+  ylab("Residuals (% cover difference)")+
+  theme_bw()+
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
+  scale_fill_manual(values=myCols)+
+  coord_cartesian(ylim = c(-50, 50))+
+  ggtitle(expression(italic("B. gracilis")))+
+  guides(fill=FALSE)
+p2 <- ggplot(subset(ipm_onestep, species=="HECO"))+
+  geom_hline(aes(yintercept=0), linetype=2)+
+  geom_boxplot(aes(x=as.character(t1),y=resid,fill=year), outlier.shape = 1, outlier.size = 0, color="grey25")+
+  xlab("Year")+
+  ylab("Residuals (% cover difference)")+
+  theme_bw()+
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
+  scale_fill_manual(values=myCols, labels=c("Year effects", "No year effects"))+
+  ggtitle(expression(italic("H. comita")))+
+  coord_cartesian(ylim = c(-10, 5))+
+  guides(fill=guide_legend(title=""))+
+  theme(legend.position=c(0.5,0.2))
+p3 <- ggplot(subset(ipm_onestep, species=="PASM"))+
+  geom_hline(aes(yintercept=0), linetype=2)+
+  geom_boxplot(aes(x=as.character(t1),y=resid,fill=year), outlier.shape = 1, outlier.size = 0, color="grey25")+
+  xlab("Year")+
+  ylab("Residuals (% cover difference)")+
+  theme_bw()+
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
+  scale_fill_manual(values=myCols)+
+  ggtitle(expression(italic("P. smithii")))+
+  guides(fill=FALSE)+
+  coord_cartesian(ylim = c(-1.5, 1.5))
+p4 <- ggplot(subset(ipm_onestep, species=="POSE"))+
+  geom_hline(aes(yintercept=0), linetype=2)+
+  geom_boxplot(aes(x=as.character(t1),y=resid,fill=year), outlier.shape = 1, outlier.size = 0, color="grey25")+
+  xlab("Year")+
+  ylab("Residuals (% cover difference)")+
+  theme_bw()+
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
+  scale_fill_manual(values=myCols)+
+  ggtitle(expression(italic("Poa secunda")))+
+  guides(fill=FALSE)+
+  coord_cartesian(ylim = c(-10, 15))
+pFinal <- grid.arrange(p1,p2,p3,p4, ncol=2, nrow=2)
+pFinal
+```
+
+```{r figure_3, dependson="plot-options", fig.cap="One step forecasts from IPM version.",fig.width=8, fig.height=8, cache=FALSE, results='hide'}
+ipm_onestep2 <- subset(ipm_onestep, year=="ayear effect")
+resDall2 <- subset(resDall, simtype=="full")
+ipm_qbm_onestep <- data.frame(year = c(ipm_onestep2$t1, resDall2$year),
+                              resids = c(ipm_onestep2$resid, resDall2$resids),
+                              model = c(rep("IPM", nrow(ipm_onestep2)),
+                                        rep("QBM", nrow(resDall2))),
+                              species = c(ipm_onestep2$species, resDall2$species))
+g1 <- ggplot(subset(ipm_qbm_onestep, species=="BOGR"))+
+  geom_hline(aes(yintercept=0), linetype=2)+
+  geom_boxplot(aes(x=as.character(year),y=resids,fill=model), outlier.shape = 1, outlier.size = 0, color="grey25")+
+  xlab("Year")+
+  ylab("Residuals (% cover difference)")+
+  theme_bw()+
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
+  scale_fill_manual(values=myCols)+
+  coord_cartesian(ylim = c(-50, 50))+
+  ggtitle(expression(italic("B. gracilis")))+
+  guides(fill=FALSE)
+g2 <- ggplot(subset(ipm_qbm_onestep, species=="HECO"))+
+  geom_hline(aes(yintercept=0), linetype=2)+
+  geom_boxplot(aes(x=as.character(year),y=resids,fill=model), outlier.shape = 1, outlier.size = 0, color="grey25")+
+  xlab("Year")+
+  ylab("Residuals (% cover difference)")+
+  theme_bw()+
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
+  scale_fill_manual(values=myCols)+
+  coord_cartesian(ylim = c(-10, 5))+
+  ggtitle(expression(italic("H. comata")))+
+  guides(fill=guide_legend(title=""))+
+  theme(legend.position=c(0.5,0.2))
+g3 <- ggplot(subset(ipm_qbm_onestep, species=="PASM"))+
+  geom_hline(aes(yintercept=0), linetype=2)+
+  geom_boxplot(aes(x=as.character(year),y=resids,fill=model), outlier.shape = 1, outlier.size = 0, color="grey25")+
+  xlab("Year")+
+  ylab("Residuals (% cover difference)")+
+  theme_bw()+
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
+  scale_fill_manual(values=myCols)+
+  ggtitle(expression(italic("P. smithii")))+
+  guides(fill=FALSE)+
+  coord_cartesian(ylim = c(-1.5, 1.5))
+g4 <- ggplot(subset(ipm_qbm_onestep, species=="POSE"))+
+  geom_hline(aes(yintercept=0), linetype=2)+
+  geom_boxplot(aes(x=as.character(year),y=resids,fill=model), outlier.shape = 1, outlier.size = 0, color="grey25")+
+  xlab("Year")+
+  ylab("Residuals (% cover difference)")+
+  theme_bw()+
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
+  scale_fill_manual(values=myCols)+
+  ggtitle(expression(italic("Poa secunda")))+
+  coord_cartesian(ylim = c(-10, 15))+
+  guides(fill=FALSE)
+pFinal <- grid.arrange(g1,g2,g3,g4, ncol=2, nrow=2)
+pFinal
+```
+
+```{r figure_4, dependson="plot-options", fig.cap="Proportional change in species' mean cover caused by a 1% increase in observed precipitation, temperature, or both as predicted by the individual-based IPM and the aggregate-based QBM.
 Note that the change for POSE due to a precipitation increase predicted by the QBM is almost zero and so does not show up in the figure.

### 2015-02-27 manuscript/manuscript.tex
updated ms with new figures from 1-steps

@@
 
+\subsection{Tables}\label{tables}
+
+\begin{table}[ht]
+\centering
+\begin{tabular}{lllrrrr}
+  \hline
+Model & Species & Year Structure & Mean & SD & Min & Max \\ 
+  \hline
+IPM & BOGR & Full & -0.24 & 5.78 & -46.52 & 43.05 \\ 
+  IPM & BOGR & Climate & -0.28 & 11.40 & -230.10 & 47.81 \\ 
+  IPM & HECO & Full & -0.29 & 0.95 & -10.99 & 4.33 \\ 
+  IPM & HECO & Climate & 0.14 & 1.05 & -8.65 & 6.67 \\ 
+  IPM & PASM & Full & -0.04 & 0.30 & -4.29 & 1.04 \\ 
+  IPM & PASM & Climate & 0.05 & 0.20 & -2.04 & 0.99 \\ 
+  IPM & POSE & Full & -0.58 & 1.65 & -30.73 & 7.34 \\ 
+  IPM & POSE & Climate & 0.23 & 1.51 & -6.08 & 14.32 \\ 
+  QBM & BOGR & Full & -2.06 & 12.45 & -81.48 & 52.88 \\ 
+  QBM & BOGR & Climate & -1.86 & 13.61 & -94.71 & 54.33 \\ 
+  QBM & HECO & Full & -0.25 & 2.28 & -68.53 & 7.84 \\ 
+  QBM & HECO & Climate & -0.47 & 3.49 & -94.57 & 7.91 \\ 
+  QBM & PASM & Full & -0.04 & 0.50 & -15.13 & 1.56 \\ 
+  QBM & PASM & Climate & -0.05 & 0.53 & -9.69 & 1.67 \\ 
+  QBM & POSE & Full & -0.05 & 2.07 & -58.34 & 13.97 \\ 
+  QBM & POSE & Climate & -0.01 & 2.47 & -73.14 & 15.18 \\ 
+   \hline
+\end{tabular}
+\caption{Table of residual summary statistics.} 
+\end{table}
+
+\pagebreak{}
+
 \subsection{Figures}\label{figures}
@@
 \includegraphics{components/figure/manuscript-figure_2.pdf}
+\caption{One step forecasts from IPM version.}
+\end{figure}
+
+\begin{figure}[htbp]
+\centering
+\includegraphics{components/figure/manuscript-figure_3.pdf}
+\caption{One step forecasts from IPM version.}
+\end{figure}
+
+\begin{figure}[htbp]
+\centering
+\includegraphics{components/figure/manuscript-figure_4.pdf}
 \caption{Proportional change in species' mean cover caused by a 1\%

### 2015-02-27 montana/ipm/simulations/one_step_results/check_1steps.R

@@ -12,28 +12,37 @@ for(i in 1:length(spp_list)){
 }
 all_d <- all_d[2:nrow(all_d),]
 all_d$resid <- with(all_d, (obs.cover.t1*100)-(cover.t1*100))
-all_d$year <- rep("year effect", nrow(all_d))
+all_d$year <- rep("ayear effect", nrow(all_d))
 
-# file <- "_sim_cover_1step_ahead_noYear.csv"
-# spp_list <- c("BOGR", "HECO", "PASM", "POSE")
-# all_d2 <- data.frame(quad=NA,t1=NA,t0=NA,rep=NA,cover.t0=NA,cover.t1=NA,obs.cover.t1=NA,species=NA)
-# for(i in 1:length(spp_list)){
-#   tmp_d <- read.csv(paste(spp_list[i],file,sep=""))
-#   tmp_d$species <- rep(spp_list[i],nrow(tmp_d))
-#   all_d2 <- rbind(all_d2, tmp_d)
-# }
-# all_d2 <- all_d2[2:nrow(all_d2),]
-# all_d2$resid <- with(all_d2, (obs.cover.t1*100)-(cover.t1*100))
-# all_d2$year <- rep("no year effect", nrow(all_d2))
-# 
-# all_d <- rbind(all_d, all_d2)
+file <- "_sim_cover_1step_ahead_noYear.csv"
+spp_list <- c("BOGR", "HECO", "PASM", "POSE")
+all_d2 <- data.frame(quad=NA,t1=NA,t0=NA,rep=NA,cover.t0=NA,cover.t1=NA,obs.cover.t1=NA,species=NA)
+for(i in 1:length(spp_list)){
+  tmp_d <- read.csv(paste(spp_list[i],file,sep=""))
+  tmp_d$species <- rep(spp_list[i],nrow(tmp_d))
+  all_d2 <- rbind(all_d2, tmp_d)
+}
+all_d2 <- all_d2[2:nrow(all_d2),]
+all_d2$resid <- with(all_d2, (obs.cover.t1*100)-(cover.t1*100))
+all_d2$year <- rep("no year effect", nrow(all_d2))
+
+all_d <- rbind(all_d, all_d2)
+
+saveRDS(all_d,"ipm_one-step_forecasts_combined.rds")
 
 library(ggplot2)
-ggplot(all_d,aes(x=t1,y=resid,group=t1))+
+myCols <- c("#237DA4", "#7ECAD0")
[…]

### 2015-02-27 montana/ipm/vitalRateRegs/plot_yearly_coefs.R

@@ -1,0 +1,58 @@
+######################################################
+#### Plot yearly intercepts and slopes to look for
+#### influential years by vital rate
+####
+
+library(ggplot2); library(plyr); library(reshape2)
+
+####
+#### Bring in data -----------------------
+####
+years <- c(1933:1945)
+grow_coef <- read.csv("./growth/growthQuants.csv")
+surv_coef <- read.csv("./survival/survivalQuants.csv")
+rec_coef <- read.csv("./recruitment/recruitmentQuants.csv")
+
+pdf(file = "yearly_coefs.pdf", onefile = TRUE, width = 8, height = 3)
+
+grow_coef <- grow_coef[-grep("betaSpp", x = grow_coef$X),]
+beta_id <- grep("beta", x = grow_coef$X)
+beta_grow <- grow_coef[beta_id,]
+beta_grow$year <- rep(years, each=4)
+ggplot(beta_grow, aes(x=as.character(year), y=X50., color=species, group=species))+
+  geom_line()+
+  geom_point()+
+  ggtitle("growth / size slope")
+int_id <- grep("intYr", x = grow_coef$X)
+int_grow <- grow_coef[int_id,]
+int_grow$year <- rep(years, each=4)
+ggplot(int_grow, aes(x=as.character(year), y=X50., color=species, group=species))+
+  geom_line()+
+  geom_point()+
+  ggtitle("growth / intercept")
+
+surv_coef <- surv_coef[-grep("betaSpp", x = surv_coef$X),]
+beta_id <- grep("beta", x = surv_coef$X)
+beta_surv <- surv_coef[beta_id,]
+beta_surv$year <- rep(years, each=4)
+ggplot(beta_surv, aes(x=as.character(year), y=X50., color=species, group=species))+
+  geom_line()+
+  geom_point()+
+  ggtitle("survival / size slope")
+int_id <- grep("intYr", x = surv_coef$X)
+int_surv <- surv_coef[int_id,]
+int_surv$year <- rep(years, each=4)
+ggplot(int_surv, aes(x=as.character(year), y=X50., color=species, group=species))+
+  geom_line()+
+  geom_point()+
+  ggtitle("survival / intercept")
+
+int_id <- grep("intYr", x = rec_coef$X)
+int_rec <- rec_coef[int_id,]
+int_rec$year <- rep(years, each=4)
+ggplot(int_rec, aes(x=as.character(year), y=X50., color=species, group=species))+
[…]

### 2015-03-05 manuscript/manuscript.Rmd
updated figures

@@
 stats <- ddply(ipm_onestep, .(species, year), summarise,
-               mean_resid = mean(resid),
+               mean_resid = mean(abs(resid)),
                sd_resid = sd(resid),
@@
 stats2 <- ddply(resDall, .(species, simtype), summarise,
-               mean_resid = mean(resids),
+               mean_resid = mean(abs(resids)),
                sd_resid = sd(resids),

### 2015-03-05 manuscript/manuscript.tex
updated figures

--- 2015-02-25 manuscript/manuscript.tex
@@
 
+\subsection{Tables}\label{tables}
+
+\begin{table}[ht]
+\centering
+\begin{tabular}{lllrrrr}
+  \hline
+Model & Species & Year Structure & Mean & SD & Min & Max \\ 
+  \hline
+IPM & BOGR & Full & 3.42 & 5.78 & -46.52 & 43.05 \\ 
+  IPM & BOGR & Climate & 5.77 & 11.40 & -230.10 & 47.81 \\ 
+  IPM & HECO & Full & 0.56 & 0.95 & -10.99 & 4.33 \\ 
+  IPM & HECO & Climate & 0.60 & 1.05 & -8.65 & 6.67 \\ 
+  IPM & PASM & Full & 0.17 & 0.30 & -4.29 & 1.04 \\ 
+  IPM & PASM & Climate & 0.14 & 0.20 & -2.04 & 0.99 \\ 
+  IPM & POSE & Full & 0.91 & 1.65 & -30.73 & 7.34 \\ 
+  IPM & POSE & Climate & 0.72 & 1.51 & -6.08 & 14.32 \\ 
+  QBM & BOGR & Full & 6.80 & 12.45 & -81.48 & 52.88 \\ 
+  QBM & BOGR & Climate & 7.69 & 13.61 & -94.71 & 54.33 \\ 
+  QBM & HECO & Full & 0.94 & 2.28 & -68.53 & 7.84 \\ 
+  QBM & HECO & Climate & 1.26 & 3.49 & -94.57 & 7.91 \\ 
+  QBM & PASM & Full & 0.26 & 0.50 & -15.13 & 1.56 \\ 
+  QBM & PASM & Climate & 0.27 & 0.53 & -9.69 & 1.67 \\ 
+  QBM & POSE & Full & 0.86 & 2.07 & -58.34 & 13.97 \\ 
+  QBM & POSE & Climate & 1.12 & 2.47 & -73.14 & 15.18 \\ 
+   \hline
+\end{tabular}
+\caption{Table of residual summary statistics.} 
+\end{table}
+
+\pagebreak{}
+
 \subsection{Figures}\label{figures}

### 2015-03-05 manuscript/suppMat.Rmd
updated figures

--- 2015-01-13 manuscript/suppMat.Rmd
@@
 
-```{r figure_S1, dependson="plot-options", fig.cap="The proportion of interannual variability in  vital rates explained by the climate covariates.
-The contribution for growth is defined as: (Climate model - Constant Model)/(Full model - Constant model).
-The contribution for survival and colonization, where we could not estimate a full model with year random effects at the quadrat level, is defined as: (Constant Model - Climate Model)/Constant Model.", fig.width=7, fig.height=3, cache=FALSE, results='hide'}
-deviance <- data.frame(VitalRate = c("Growth", "Survival", "Colonization"),
-                       ConstantModel = rep(NA,3),
-                       ClimateModel = rep(NA,3),
-                       FullModel = rep(NA,3),
-                       ClimateContribution = rep(NA,3))
-deviance$FullModel[1] <- round(read.csv("../montana/quadBM/vitalRateRegressions/growth/growthDeviance.csv")[,2],2)
-deviance$ClimateModel[1] <- round(read.csv("../montana/quadBM/vitalRateRegressions/growth/growthDevianceCLIMATE.csv")[,2],2)
-deviance$ConstantModel[1] <- round(read.csv("../montana/quadBM/vitalRateRegressions/growth/growthDevianceCONSTANT.csv")[,2],2)
-deviance$ClimateContribution[1] <- with(deviance[1,], (ClimateModel-ConstantModel)/(FullModel-ConstantModel))
-deviance$ClimateModel[2] <- round(read.csv("../montana/quadBM/vitalRateRegressions/survival/survivalDeviance.csv")[,2],2)
-deviance$ConstantModel[2] <- round(read.csv("../montana/quadBM/vitalRateRegressions/survival/survivalDevianceCONSTANT.csv")[,2],2)
-deviance$ClimateModel[3] <- round(read.csv("../montana/quadBM/vitalRateRegressions/colonization/colonizationDeviance.csv")[,2],2)
-deviance$ConstantModel[3] <- round(read.csv("../montana/quadBM/vitalRateRegressions/colonization/colonizationDevianceCONSTANT.csv")[,2],2)
-deviance$ClimateContribution[2:3] <- with(deviance[2:3,], (ConstantModel-ClimateModel)/(ConstantModel))
-
-colnames(deviance) <- c("Vital Rate", "Constant model", "Climate model", "Full model", "Contribution of climate covariates")
-library(ggplot2)
-devGQ <- data.frame(VitalRate = c("Growth", "Survival", "Colonization"),
-                       ConstantModel = rep(NA,3),
-                       ClimateModel = rep(NA,3),
-                       FullModel = rep(NA,3),
-                       ClimateContribution = rep(NA,3))
-devGQ$FullModel[1] <- round(read.csv("../montana/ipm/vitalRateRegs/growth/growthDeviance.csv")[,2],2)
-devGQ$ClimateModel[1] <- round(read.csv("../montana/ipm/vitalRateRegs/growth/growthDevianceCLIMATE.csv")[,2],2)
-devGQ$ConstantModel[1] <- round(read.csv("../montana/ipm/vitalRateRegs/growth/growthDevianceCONSTANT.csv")[,2],2)
-devGQ$FullModel[2] <- round(read.csv("../montana/ipm/vitalRateRegs/survival/survivalDeviance.csv")[,2],2)
-devGQ$ClimateModel[2] <- round(read.csv("../montana/ipm/vitalRateRegs/survival/survivalDevianceCLIMATE.csv")[,2],2)
-devGQ$ConstantModel[2] <- round(read.csv("../montana/ipm/vitalRateRegs/survival/survivalDevianceCONSTANT.csv")[,2],2)
-devGQ$FullModel[3] <- round(read.csv("../montana/ipm/vitalRateRegs/recruitment/recruitmentDeviance.csv")[,2],2)
-devGQ$ClimateModel[3] <- round(read.csv("../montana/ipm/vitalRateRegs/recruitment/recruitmentDevianceCLIMATE.csv")[,2],2)
-devGQ$ConstantModel[3] <- round(read.csv("../montana/ipm/vitalRateRegs/recruitment/recruitmentDevianceCONSTANT.csv")[,2],2)
-devGQ$ClimateContribution[1] <- with(devGQ[1,], (ClimateModel-ConstantModel)/(FullModel-ConstantModel))
-devGQ$ClimateContribution[2] <- with(devGQ[2,], (ConstantModel-ClimateModel)/(ConstantModel))
-devGQ$ClimateContribution[3] <- with(devGQ[3,], (ConstantModel-ClimateModel)/(ConstantModel))
-
-devM <- devGQ[,c(1,5)]
-devM$Level <- rep("Individual", 3)
-tmpD <- deviance[,c(1,5)]
-tmpD$Level <- rep("Quadrat", 3)
-colnames(tmpD)[2] <- "value"
-colnames(tmpD)[1] <- "VitalRate"
-colnames(devM)[2] <- "value"
-plotD <- rbind(tmpD, devM)
-gClimCont1 <- ggplot(data=subset(plotD, Level=="Individual"), aes(x=VitalRate, y=value))+
-  geom_bar(stat="identity", position="dodge", fill="grey30")+
-  ylab("Contribution of Climate Effects")+
-  xlab("Vital Rate")+
-  ggtitle("Individual level")+
-  scale_y_continuous(limits=c(0, 0.5))+
-  scale_x_discrete(labels=c("Recruitment", "Growth", "Survival"))+
-  theme_bw()+
-  theme(axis.text.x = element_text(angle = 25, hjust = 1))
-gClimCont2 <- ggplot(data=subset(plotD, Level=="Quadrat"), aes(x=VitalRate, y=value))+
-  geom_bar(stat="identity", position="dodge", fill="grey30")+
-  ylab("Contribution of Climate Effects")+
-  xlab("Vital Rate")+
-  ggtitle("Quadrat level")+
-  scale_y_continuous(limits=c(0, 0.5))+
-  scale_x_discrete(labels=c("Colonization", "Cover change", "Extirpation"))+
-  theme_bw()+
-  theme(axis.text.x = element_text(angle = 25, hjust = 1))
-library(gridExtra)
-gClim <- grid.arrange(gClimCont1, gClimCont2, ncol=2, nrow=1)
-gClim
-
-```
-
-```{r figure_S2, dependson="plot-options", fig.cap="Posterior means (vertical ticks), 75% credible intervals (heavy lines), and 95% credible intervals (light lines) of climate effects on growth at both levels of inferences.
+```{r figure_S1, dependson="plot-options", fig.cap="Posterior means (vertical ticks), 75% credible intervals (heavy lines), and 95% credible intervals (light lines) of climate effects on growth at both levels of inferences.
 The dashed vertical line is at 0, indicating no effect.", fig.width=8, fig.height=7, cache=FALSE, results='hide'}
-gQC <- read.csv("../montana/quadBM/vitalRateRegressions/growth/growthQuants.csv")
-gQC <- gQC[137:152,]
+qbm_quants <- read.csv("../montana/quadBM/vitalRateRegressions/truncNormModel/growthQuants.csv")
+qbm_quants <- qbm_quants[137:156,]
+qbm_quants <- qbm_quants[-c(9:12),]
 coefNames <- c("ppt1", "ppt2", "temp1", "temp2")
-gQC$X <- rep(coefNames, each=4)
-gQC$rate <- "growth"
-sQC <- read.csv("../montana/quadBM/vitalRateRegressions/survival/survivalQuants.csv")
-sQC <- sQC[33:48,]
-coefNames <- c("ppt1", "ppt2", "temp1", "temp2")
-sQC$X <- rep(coefNames, each=4)
-sQC$rate <- "survival"
-rQC <- read.csv("../montana/quadBM/vitalRateRegressions/colonization/colonizationQuants.csv")
-rQC <- rQC[29:44,]
-coefNames <- c("ppt1", "ppt2", "temp1", "temp2")
-rQC$X <- rep(coefNames, each=4)
-rQC$rate <- "colonization"
-plotQC <- rbind(gQC,sQC,rQC)
-gQ <- ggplot(plotQC, aes(x=species, y=X50., color=rate))+
+qbm_quants$X <- rep(coefNames, each=4)
+gQ <- ggplot(qbm_quants, aes(x=species, y=X50., group=species), color="grey25")+
   geom_hline(aes(yintercept=0), linetype="dashed")+
@@
   theme_bw()+
-  scale_color_manual(values=c("grey25", "steelblue", "slateblue"))+
   guides(color=guide_legend(title=""))+
@@
 gIC <- read.csv("../montana/ipm/vitalRateRegs/growth/growthQuants.csv")
-gIC <- gIC[133:148,]
+gIC <- gIC[141:156,]
 coefNames <- c("ppt1", "ppt2", "temp1", "temp2")
@@
 sIC <- read.csv("../montana/ipm/vitalRateRegs/survival/survivalQuants.csv")
-sIC <- sIC[133:148,]
+sIC <- sIC[141:156,]
 coefNames <- c("ppt1", "ppt2", "temp1", "temp2")
@@
 rIC <- read.csv("../montana/ipm/vitalRateRegs/recruitment/recruitmentQuants.csv")
-rIC <- rIC[81:96,]
+rIC <- rIC[85:100,]
 coefNames <- c("ppt1", "ppt2", "temp1", "temp2")
@@
 ####
-#### Plot just for growth
-####
-growthIC <- subset(plotIC, rate=="growth")
-growthQC <- subset(plotQC, rate=="growth")
-growthIC$Level <- "Individual"
-growthQC$Level <- "Quadrat"
-growthAll <- rbind(growthIC, growthQC)
-
-gGrowth <- ggplot(growthAll, aes(x=species, y=X50., color=Level))+
-  geom_hline(aes(yintercept=0), linetype="dashed")+
-  geom_errorbar(aes(ymin=X2.5., ymax=X97.5.), width=0, position=position_dodge(width=0.5))+
-  geom_errorbar(aes(ymin=X25., ymax=X75.), width=0, size=1.25, position=position_dodge(width=0.5))+
-  geom_point(size=4, position=position_dodge(width=0.5), shape="|")+
-  ylab("Standardized Coefficient Value")+
-  xlab("Species")+
-  facet_grid(X~.)+
-  coord_flip()+
-  theme_bw()+
-  scale_color_manual(values=c("grey25", "slateblue"))+
-#   guides(color=FALSE)+
-  guides(color=guide_legend(title=""))+
-  theme(legend.position="top")
-# print(gGrowth)
-
-####
 #### Print figure

### 2015-03-05 manuscript/suppMat.tex
updated figures

@@
 \includegraphics{components/figure/manuscript-figure_S1.pdf}
-\caption{The proportion of interannual variability in vital rates
-explained by the climate covariates. The contribution for growth is
-defined as: (Climate model - Constant Model)/(Full model - Constant
-model). The contribution for survival and colonization, where we could
-not estimate a full model with year random effects at the quadrat level,
-is defined as: (Constant Model - Climate Model)/Constant Model.}
-\end{figure}
-
-\begin{figure}[htbp]
-\centering
-\includegraphics{components/figure/manuscript-figure_S2.pdf}
 \caption{Posterior means (vertical ticks), 75\% credible intervals

### 2015-03-05 montana/quadBM/vitalRateRegressions/truncNormModel/growth_model_selection.R

@@ -77,25 +77,29 @@ growD <- backD[2:nrow(backD),]
 ####
 growD$interaction1 <- with(growD, ppt1*TmeanSpr1)
 growD$interaction2 <- with(growD, ppt2*TmeanSpr2)
+dic <- matrix(nrow=5, ncol=length(sppList))
 
-grow_now <- subset(growD, Species=="BOGR")
-X <- list(X1 = grow_now[,c("pptLag", "ppt1", "ppt2", "TmeanSpr1", "TmeanSpr2", "interaction1", "interaction2")],
-          X2 = grow_now[,c("pptLag", "ppt1", "ppt2", "TmeanSpr1", "TmeanSpr2", "interaction1")],
-          X3 = grow_now[,c("pptLag", "ppt1", "ppt2", "TmeanSpr1", "TmeanSpr2", "interaction2")],
-          X4 = grow_now[,c("pptLag", "ppt1", "ppt2", "TmeanSpr1", "TmeanSpr2")],
-          X5 = grow_now[,c("pptLag", "ppt1", "ppt2", "TmeanSpr1", "TmeanSpr2")])
-
-####
-#### Send to model fitting function
-####
-iters <- 100000
-dic <- numeric(length(X))
-for(i in 1:length(X)){
-  model_dic <- mont_qbm_jags(Y = grow_now$percCover, size = grow_now$percLagCover, X = X[[i]], 
-                             groups = as.numeric(as.factor(grow_now$group)), 
-                             species = as.numeric(as.factor(grow_now$Species)),
-                             years = (grow_now$year-32), iters=iters, adapt.iters=500, thins=5)
-  dic[i] <- sum(model_dic$deviance)+sum(model_dic$penalty)
+for(spp in 1:length(sppList)){
+  doSpp <- sppList[spp]
+  grow_now <- subset(growD, Species==doSpp)
+  X <- list(X1 = grow_now[,c("pptLag", "ppt1", "ppt2", "TmeanSpr1", "TmeanSpr2", "interaction1", "interaction2")],
+            X2 = grow_now[,c("pptLag", "ppt1", "ppt2", "TmeanSpr1", "TmeanSpr2", "interaction1")],
+            X3 = grow_now[,c("pptLag", "ppt1", "ppt2", "TmeanSpr1", "TmeanSpr2", "interaction2")],
+            X4 = grow_now[,c("pptLag", "ppt1", "ppt2", "TmeanSpr1", "TmeanSpr2")],
[…]

### 2015-03-06 manuscript/manuscript.Rmd
new figures

@@
 -------
-```{r figure_1, dependson="plot-options", fig.cap="Boxplots of model residuals for one-step-ahead forecasts at each observation year for each quadrat.
-Residuals represent the difference in observed and predicted cover change from year $t-1$ to year $t$.
-Each one-step forecast was simulated 100 times for each quadrat.
-Note that the y-axes vary across panels.", fig.width=8, fig.height=8, cache=FALSE, results='hide'}
-resD <- readRDS("../montana/quadBM/simulations/oneStepResids_YearEffects.rds")
-resDny <- readRDS("../montana/quadBM/simulations/oneStepResids_NoYearEffects.rds")
-resD$simtype <- rep("full", nrow(resD))
-resDny$simtype <- rep("noyear", nrow(resDny))
-resDall <- rbind(resD, resDny)
-noZ <- which(resDall$lagCover!=0)
-resDall <- resDall[noZ,]
+```{r figure_1,cache=FALSE, results='hide'}
+# resD <- readRDS("../montana/quadBM/simulations/oneStepResids_YearEffects.rds")
+# resDny <- readRDS("../montana/quadBM/simulations/oneStepResids_NoYearEffects.rds")
+# resD$simtype <- rep("full", nrow(resD))
+# resDny$simtype <- rep("noyear", nrow(resDny))
+# resDall <- rbind(resD, resDny)
+# noZ <- which(resDall$lagCover!=0)
+# resDall <- resDall[noZ,]
+# 
+# myCols <- c("#237DA4", "#7ECAD0")
+# # ggplot(resDall, aes(x=resids, fill=simtype))+
+# #   geom_histogram(alpha=0.5)+
+# #   facet_grid(.~species, scales = "free_x")+
+# #   xlab("Year")+
+# #   ylab("Residuals (% cover difference)")+
+# #   theme_bw()+
+# #   scale_fill_manual(values=myCols)
+# 
+# resAgg <- ddply(resDall, .(year, species, simtype), summarise,
+#                 avg_prediction = mean(cover)*100,
+#                 up_prediction = quantile(cover, probs = 0.875)*100,
+#                 down_prediction = quantile(cover, probs = 0.125)*100,
+#                 avg_obs = mean(propCover)*100)
+# resAgg$model <- rep("QBM", nrow(resAggipm))
 
-myCols <- c("#237DA4", "#7ECAD0")
-# ggplot(resDall, aes(x=resids, fill=simtype))+
-#   geom_histogram(alpha=0.5)+
-#   facet_grid(.~species, scales = "free_x")+
-#   xlab("Year")+
-#   ylab("Residuals (% cover difference)")+
-#   theme_bw()+
-#   scale_fill_manual(values=myCols)
-
+# ggplot(subset(resAgg))+
+#   geom_ribbon(aes(x=year, ymin=down_prediction, ymax=up_prediction, fill=simtype), alpha=0.3)+
+#   geom_line(aes(x=year, y=avg_prediction, color=simtype), size=1.5)+
+# #   geom_line(aes(x=year, y=up_prediction, color=simtype), linetype=2, alpha=0.5)+
+# #   geom_line(aes(x=year, y=down_prediction, color=simtype), linetype=2, alpha=0.5)+
+#   geom_point(aes(x=year, y=avg_obs), shape=1, size=3)+
+#   facet_grid(species~., scales="free")+
+#   scale_fill_manual(values=c(myCols[1], "darkorange"), labels=c("Year effects", "No year effects"))+
+#   scale_color_manual(values=c(myCols[1], "darkorange"))+
+#   guides(color=FALSE)
 # nSims <- length(unique(resD$sim))
@@
 # myCols <- c("#237DA4", "#7ECAD0")
-p1 <- ggplot(subset(resDall, species=="BOGR"))+
-  geom_hline(aes(yintercept=0), linetype=2)+
-  geom_boxplot(aes(x=as.character(year), y=resids, fill=simtype), outlier.shape = 1, outlier.size = 0, color="grey25")+
-  xlab("Year")+
-  ylab("Residuals (% cover difference)")+
-  theme_bw()+
-  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
-  scale_fill_manual(values=myCols)+
-  coord_cartesian(ylim = c(-50, 50))+
-  ggtitle(expression(italic("B. gracilis")))+
-  guides(fill=FALSE)
-p2 <- ggplot(subset(resDall, species=="HECO"))+
-  geom_hline(aes(yintercept=0), linetype=2)+
-  geom_boxplot(aes(x=as.character(year), y=resids, fill=simtype), outlier.shape = 1, outlier.size = 0, color="grey25")+
-  xlab("Year")+
-  ylab("Residuals (% cover difference)")+
-  theme_bw()+
-  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
-  scale_fill_manual(values=myCols, labels=c("Year effects", "No year effects"))+
-  ggtitle(expression(italic("H. comita")))+
-  coord_cartesian(ylim = c(-10, 5))+
-  guides(fill=guide_legend(title=""))+
-  theme(legend.position=c(0.5,0.2))
-p3 <- ggplot(subset(resDall, species=="PASM"))+
-  geom_hline(aes(yintercept=0), linetype=2)+
-  geom_boxplot(aes(x=as.character(year), y=resids, fill=simtype), outlier.shape = 1, outlier.size = 0, color="grey25")+
-  xlab("Year")+
-  ylab("Residuals (% cover difference)")+
-  theme_bw()+
-  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
-  scale_fill_manual(values=myCols)+
-  ggtitle(expression(italic("P. smithii")))+
-  guides(fill=FALSE)+
-  coord_cartesian(ylim = c(-1.5, 1.5))
-p4 <- ggplot(subset(resDall, species=="POSE"))+
-  geom_hline(aes(yintercept=0), linetype=2)+
-  geom_boxplot(aes(x=as.character(year), y=resids, fill=simtype), outlier.shape = 1, outlier.size = 0, color="grey25")+
-  xlab("Year")+
-  ylab("Residuals (% cover difference)")+
-  theme_bw()+
-  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
-  scale_fill_manual(values=myCols)+
-  ggtitle(expression(italic("Poa secunda")))+
-  guides(fill=FALSE)+
-  coord_cartesian(ylim = c(-10, 15))
-pFinal <- grid.arrange(p1,p2,p3,p4, ncol=2, nrow=2)
-pFinal
+# p1 <- ggplot(subset(resDall, species=="BOGR"))+
+#   geom_hline(aes(yintercept=0), linetype=2)+
+#   geom_boxplot(aes(x=as.character(year), y=resids, fill=simtype), outlier.shape = 1, outlier.size = 0, color="grey25")+
+#   xlab("Year")+
+#   ylab("Residuals (% cover difference)")+
+#   theme_bw()+
+#   theme(axis.text.x = element_text(angle = 45, hjust = 1))+
+#   scale_fill_manual(values=myCols)+
+#   coord_cartesian(ylim = c(-50, 50))+
+#   ggtitle(expression(italic("B. gracilis")))+
+#   guides(fill=FALSE)
+# p2 <- ggplot(subset(resDall, species=="HECO"))+
+#   geom_hline(aes(yintercept=0), linetype=2)+
+#   geom_boxplot(aes(x=as.character(year), y=resids, fill=simtype), outlier.shape = 1, outlier.size = 0, color="grey25")+
+#   xlab("Year")+
+#   ylab("Residuals (% cover difference)")+
+#   theme_bw()+
+#   theme(axis.text.x = element_text(angle = 45, hjust = 1))+
+#   scale_fill_manual(values=myCols, labels=c("Year effects", "No year effects"))+
+#   ggtitle(expression(italic("H. comita")))+
+#   coord_cartesian(ylim = c(-10, 5))+
+#   guides(fill=guide_legend(title=""))+
+#   theme(legend.position=c(0.5,0.2))
+# p3 <- ggplot(subset(resDall, species=="PASM"))+
+#   geom_hline(aes(yintercept=0), linetype=2)+
+#   geom_boxplot(aes(x=as.character(year), y=resids, fill=simtype), outlier.shape = 1, outlier.size = 0, color="grey25")+
+#   xlab("Year")+
+#   ylab("Residuals (% cover difference)")+
+#   theme_bw()+
+#   theme(axis.text.x = element_text(angle = 45, hjust = 1))+
+#   scale_fill_manual(values=myCols)+
+#   ggtitle(expression(italic("P. smithii")))+
+#   guides(fill=FALSE)+
+#   coord_cartesian(ylim = c(-1.5, 1.5))
+# p4 <- ggplot(subset(resDall, species=="POSE"))+
+#   geom_hline(aes(yintercept=0), linetype=2)+
+#   geom_boxplot(aes(x=as.character(year), y=resids, fill=simtype), outlier.shape = 1, outlier.size = 0, color="grey25")+
+#   xlab("Year")+
+#   ylab("Residuals (% cover difference)")+
+#   theme_bw()+
+#   theme(axis.text.x = element_text(angle = 45, hjust = 1))+
+#   scale_fill_manual(values=myCols)+
+#   ggtitle(expression(italic("Poa secunda")))+
+#   guides(fill=FALSE)+
+#   coord_cartesian(ylim = c(-10, 15))
+# pFinal <- grid.arrange(p1,p2,p3,p4, ncol=2, nrow=2)
+# pFinal
 
@@
 ```{r figure_2, dependson="plot-options", fig.cap="One step forecasts from IPM version.",fig.width=8, fig.height=8, cache=FALSE, results='hide'}
+resD <- readRDS("../montana/quadBM/simulations/oneStepResids_YearEffects.rds")
+resDny <- readRDS("../montana/quadBM/simulations/oneStepResids_NoYearEffects.rds")
+resD$simtype <- rep("full", nrow(resD))
+resDny$simtype <- rep("noyear", nrow(resDny))
+resDall <- rbind(resD, resDny)
+noZ <- which(resDall$lagCover!=0)
+resDall <- resDall[noZ,]
+
+myCols <- c("#237DA4", "#7ECAD0")
+# ggplot(resDall, aes(x=resids, fill=simtype))+
+#   geom_histogram(alpha=0.5)+
+#   facet_grid(.~species, scales = "free_x")+
+#   xlab("Year")+
+#   ylab("Residuals (% cover difference)")+
+#   theme_bw()+
+#   scale_fill_manual(values=myCols)
+
+resAgg <- ddply(resDall, .(year, species, simtype), summarise,
+                avg_prediction = mean(cover)*100,
+                up_prediction = quantile(cover, probs = 0.875)*100,
+                down_prediction = quantile(cover, probs = 0.125)*100,
+                avg_obs = mean(propCover)*100)
+resAgg$model <- rep("QBM", nrow(resAgg))
+
 ipm_onestep <- readRDS("../montana/ipm/simulations/one_step_results/ipm_one-step_forecasts_combined.rds")
 ipm_onestep$t1 <- ipm_onestep$t1+1900
-p1 <- ggplot(subset(ipm_onestep, species=="BOGR"))+
-  geom_hline(aes(yintercept=0), linetype=2)+
-  geom_boxplot(aes(x=as.character(t1),y=resid,fill=year), outlier.shape = 1, outlier.size = 0, color="grey25")+
-  xlab("Year")+
-  ylab("Residuals (% cover difference)")+
-  theme_bw()+
-  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
-  scale_fill_manual(values=myCols)+
-  coord_cartesian(ylim = c(-50, 50))+
-  ggtitle(expression(italic("B. gracilis")))+
-  guides(fill=FALSE)
-p2 <- ggplot(subset(ipm_onestep, species=="HECO"))+
-  geom_hline(aes(yintercept=0), linetype=2)+
-  geom_boxplot(aes(x=as.character(t1),y=resid,fill=year), outlier.shape = 1, outlier.size = 0, color="grey25")+
-  xlab("Year")+
-  ylab("Residuals (% cover difference)")+
-  theme_bw()+
-  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
-  scale_fill_manual(values=myCols, labels=c("Year effects", "No year effects"))+
-  ggtitle(expression(italic("H. comita")))+
-  coord_cartesian(ylim = c(-10, 5))+
-  guides(fill=guide_legend(title=""))+
-  theme(legend.position=c(0.5,0.2))
-p3 <- ggplot(subset(ipm_onestep, species=="PASM"))+
-  geom_hline(aes(yintercept=0), linetype=2)+
-  geom_boxplot(aes(x=as.character(t1),y=resid,fill=year), outlier.shape = 1, outlier.size = 0, color="grey25")+
-  xlab("Year")+
-  ylab("Residuals (% cover difference)")+
-  theme_bw()+
-  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
-  scale_fill_manual(values=myCols)+
-  ggtitle(expression(italic("P. smithii")))+
-  guides(fill=FALSE)+
-  coord_cartesian(ylim = c(-1.5, 1.5))
-p4 <- ggplot(subset(ipm_onestep, species=="POSE"))+
-  geom_hline(aes(yintercept=0), linetype=2)+
-  geom_boxplot(aes(x=as.character(t1),y=resid,fill=year), outlier.shape = 1, outlier.size = 0, color="grey25")+
-  xlab("Year")+
-  ylab("Residuals (% cover difference)")+
-  theme_bw()+
-  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
-  scale_fill_manual(values=myCols)+
-  ggtitle(expression(italic("Poa secunda")))+
-  guides(fill=FALSE)+
-  coord_cartesian(ylim = c(-10, 15))
-pFinal <- grid.arrange(p1,p2,p3,p4, ncol=2, nrow=2)
-pFinal
+
+resAggipm <- ddply(ipm_onestep, .(t1, species, year), summarise,
+                avg_prediction = mean(cover.t1)*100,
+                up_prediction = quantile(cover.t1, probs = 0.875)*100,
+                down_prediction = quantile(cover.t1, probs = 0.125)*100,
+                avg_obs = mean(obs.cover.t1)*100)
+colnames(resAggipm)[1:3] <- c("year", "species", "simtype")
+resAggipm$simtype[which(resAggipm$simtype=="ayear effect")] <- "full"
+resAggipm$simtype[which(resAggipm$simtype!="full")] <- "noyear"
+resAggipm$model <- rep("IPM", nrow(resAggipm))
+
+resAgg_both <- rbind(resAgg, resAggipm)
+
+ggplot(subset(resAgg_both))+
+  geom_ribbon(aes(x=year, ymin=down_prediction, ymax=up_prediction, fill=simtype), alpha=0.3)+
+  geom_line(aes(x=year, y=avg_prediction, color=simtype), size=1.5)+
+  geom_point(aes(x=year, y=avg_obs), shape=1, size=3)+
+  facet_grid(species~model, scales="free")+
+  scale_fill_manual(values=c(myCols[1], "darkorange"), labels=c("Year effects", "No year effects"))+
+  scale_color_manual(values=c(myCols[1], "darkorange"))+
+  guides(color=FALSE)
+
+ggplot(subset(resAgg_both))+
+  geom_ribbon(aes(x=year, ymin=down_prediction, ymax=up_prediction, fill=model), alpha=0.3)+
+  geom_line(aes(x=year, y=avg_prediction, color=model), size=1.5)+
+  geom_point(aes(x=year, y=avg_obs), shape=1, size=3)+
+  facet_grid(species~simtype, scales="free")+
+  scale_fill_manual(values=c(myCols[1], "darkorange"))+
+  scale_color_manual(values=c(myCols[1], "darkorange"))+
+  guides(color=FALSE)
+
+# p1 <- ggplot(subset(ipm_onestep, species=="BOGR"))+
+#   geom_hline(aes(yintercept=0), linetype=2)+
+#   geom_boxplot(aes(x=as.character(t1),y=resid,fill=year), outlier.shape = 1, outlier.size = 0, color="grey25")+
+#   xlab("Year")+
+#   ylab("Residuals (% cover difference)")+
+#   theme_bw()+
+#   theme(axis.text.x = element_text(angle = 45, hjust = 1))+
+#   scale_fill_manual(values=myCols)+
+#   coord_cartesian(ylim = c(-50, 50))+
+#   ggtitle(expression(italic("B. gracilis")))+
+#   guides(fill=FALSE)
+# p2 <- ggplot(subset(ipm_onestep, species=="HECO"))+
+#   geom_hline(aes(yintercept=0), linetype=2)+
+#   geom_boxplot(aes(x=as.character(t1),y=resid,fill=year), outlier.shape = 1, outlier.size = 0, color="grey25")+
+#   xlab("Year")+
+#   ylab("Residuals (% cover difference)")+
+#   theme_bw()+
+#   theme(axis.text.x = element_text(angle = 45, hjust = 1))+
+#   scale_fill_manual(values=myCols, labels=c("Year effects", "No year effects"))+
+#   ggtitle(expression(italic("H. comita")))+
+#   coord_cartesian(ylim = c(-10, 5))+
+#   guides(fill=guide_legend(title=""))+
+#   theme(legend.position=c(0.5,0.2))
+# p3 <- ggplot(subset(ipm_onestep, species=="PASM"))+
+#   geom_hline(aes(yintercept=0), linetype=2)+
+#   geom_boxplot(aes(x=as.character(t1),y=resid,fill=year), outlier.shape = 1, outlier.size = 0, color="grey25")+
+#   xlab("Year")+
+#   ylab("Residuals (% cover difference)")+
+#   theme_bw()+
+#   theme(axis.text.x = element_text(angle = 45, hjust = 1))+
+#   scale_fill_manual(values=myCols)+
+#   ggtitle(expression(italic("P. smithii")))+
+#   guides(fill=FALSE)+
+#   coord_cartesian(ylim = c(-1.5, 1.5))
+# p4 <- ggplot(subset(ipm_onestep, species=="POSE"))+
+#   geom_hline(aes(yintercept=0), linetype=2)+
+#   geom_boxplot(aes(x=as.character(t1),y=resid,fill=year), outlier.shape = 1, outlier.size = 0, color="grey25")+
+#   xlab("Year")+
+#   ylab("Residuals (% cover difference)")+
+#   theme_bw()+
+#   theme(axis.text.x = element_text(angle = 45, hjust = 1))+
+#   scale_fill_manual(values=myCols)+
+#   ggtitle(expression(italic("Poa secunda")))+
+#   guides(fill=FALSE)+
+#   coord_cartesian(ylim = c(-10, 15))
+# pFinal <- grid.arrange(p1,p2,p3,p4, ncol=2, nrow=2)
+# pFinal
 ```
 
-```{r figure_3, dependson="plot-options", fig.cap="One step forecasts from IPM version.",fig.width=8, fig.height=8, cache=FALSE, results='hide'}
+```{r figure_3, dependson="plot-options", fig.cap="One step forecasts from IPM version.",fig.width=8, fig.height=8, cache=FALSE, results='hide', eval=FALSE}
 ipm_onestep2 <- subset(ipm_onestep, year=="ayear effect")

### 2015-03-06 manuscript/manuscript.tex
new figures

@@
 \centering
-\includegraphics{components/figure/manuscript-figure_1.pdf}
-\caption{Boxplots of model residuals for one-step-ahead forecasts at
-each observation year for each quadrat. Residuals represent the
-difference in observed and predicted cover change from year $t-1$ to
-year $t$. Each one-step forecast was simulated 100 times for each
-quadrat. Note that the y-axes vary across panels.}
-\end{figure}
-
-\begin{figure}[htbp]
-\centering
 \includegraphics{components/figure/manuscript-figure_2.pdf}

### 2015-03-06 manuscript/suppMat.Rmd
updated for lag ppt

@@
 The dashed vertical line is at 0, indicating no effect.", fig.width=8, fig.height=7, cache=FALSE, results='hide'}
-qbm_quants <- read.csv("../montana/quadBM/vitalRateRegressions/truncNormModel/growthQuants.csv")
-qbm_quants <- qbm_quants[137:156,]
-qbm_quants <- qbm_quants[-c(9:12),]
-coefNames <- c("ppt1", "ppt2", "temp1", "temp2")
+qbm_quants <- read.csv("../montana/quadBM/vitalRateRegressions/truncNormModel/growthQuants_withLagPpt.csv")
+qbm_quants <- qbm_quants[137:160,]
+qbm_quants <- qbm_quants[-c(13:16),]
+coefNames <- c("ppt1", "ppt2", "lagppt", "temp1", "temp2")
 qbm_quants$X <- rep(coefNames, each=4)

### 2015-03-06 montana/quadBM/simulations/oneStepForecast_NoYearEffectsQBM_withLagPpt.R

@@ -1,0 +1,143 @@
+#Quad-Based Model simulations
+
+#!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!#
+# Set working directory to location of this source file #
+#!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!#
+
+#clear everything, just to be safe 
+rm(list=ls(all=TRUE))
+
+library(reshape2)
+library(plyr)
+library(ggplot2)
+library(ggthemes)
+library(gridExtra)
+library(EnvStats)
+NumberSimsPerYear <- 1
+
+#bring in data
+allD <- read.csv("../../speciesData/quadAllCover.csv")
+allD <- allD[,2:ncol(allD)] #get rid of X ID column
+allD$percCover <- allD$totCover/10000
+head(scale(allD$percCover, center=TRUE, scale=TRUE))
+sppList <- as.character(unique(allD$Species))
+
+#bring in climate data
+climD <- read.csv("../../weather/Climate.csv")
+climD[2:6] <- scale(climD[2:6], center = TRUE, scale = TRUE)
+
+#load vital rate parameters
+pGrow <- readRDS("../vitalRateRegressions/truncNormModel/growthParamsMCMC_withLagPpt.rds")
+pGrow2 <- melt(pGrow)
+pGrow2$Spp <- c(rep(rep(sppList, each=3000), times=13),
+                rep(rep(sppList, each=3000), times=1),
+                rep(rep(sppList, each=3000), times=6),
+                rep(rep(sppList, each=3000), times=13),
+                rep(rep(sppList, each=3000), times=7))
+pGrow2$Coef <- c(rep("beta", times=3000*4*13),
+                 rep("betaSpp", times=3000*4),
+                 rep("gInt", times=6*4*3000),
+                 rep("intYr", times=4*3000*13),
+                 rep("intercept", times=3000*4),
+                 rep("rain1", times=4*3000),
+                 rep("rain2", times=4*3000),
+                 rep("rainLag", times=4*3000),
+                 rep("tau", times=4*3000),
+                 rep("temp1", times=4*3000),
+                 rep("temp2", times=4*3000))
+colnames(pGrow2)[1] <- "Iter"
[…]

### 2015-03-12 manuscript/manuscript.Rmd
added model validation section

@@
 We used random draws from the final 1,000 iterations from each of three MCMC chains to introduce stochasticity into our population models.
-At each time step, we first randomly selected climate covariates from one of the 14 observed years and also randomly drew one set of random year effects for each vital rate regression.
-Climate years and random year effects were drawn independently because they were uncorrelated.
-Then, we drew the full parameter set (specific random year effects, climate effects, fixed effects) from a randomly selected MCMC iteration.
-We selected random year effects and the climate year independently.
+At each time step, we randomly selected climate covariates from one of the 14 observed years.
+Then, we drew the full parameter set (climate effects and density-dependence fixed effects) from a randomly selected MCMC iteration.
+Since our focus was on the contribution of climate covariates to population states, we set the random year effects and the random group effects to zero.
+
+### Model validation
+To test each model's ability to forecast the population state we used leave-one-year-out cross validation.
+For both levels of modeling, we fit the vital rate models using observations from all years except one, and then used those fitted parameters in the population models to perform a one-step-ahead forecast for the year whose observations were withheld from model fitting.
+We repeated this procedure for all 13 observation years.
+This model validation allowed us to compare accuracy and precision of the two modeling approaches (individual-level versus population-level).
 
@@
 -------
+### Vital rate statistical models
+
+### Model validation
+
+### Climate change simulations
 

### 2015-03-12 manuscript/manuscript.tex
added model validation section

@@
 MCMC chains to introduce stochasticity into our population models. At
-each time step, we first randomly selected climate covariates from one
-of the 14 observed years and also randomly drew one set of random year
-effects for each vital rate regression. Climate years and random year
-effects were drawn independently because they were uncorrelated. Then,
-we drew the full parameter set (specific random year effects, climate
-effects, fixed effects) from a randomly selected MCMC iteration. We
-selected random year effects and the climate year independently.
+each time step, we randomly selected climate covariates from one of the
+14 observed years. Then, we drew the full parameter set (climate effects
+and density-dependence fixed effects) from a randomly selected MCMC
+iteration. Since our focus was on the contribution of climate covariates
+to population states, we set the random year effects and the random
+group effects to zero.
+
+\subsubsection{Model validation}\label{model-validation}
+
+To test each model's ability to forecast the population state we used
+leave-one-year-out cross validation. For both levels of modeling, we fit
+the vital rate models using observations from all years except one, and
+then used those fitted parameters in the population models to perform a
+one-step-ahead forecast for the year whose observations were withheld
+from model fitting. We repeated this procedure for all 13 observation
+years. This model validation allowed us to compare accuracy and
+precision of the two modeling approaches (individual-level versus
+population-level).
 
 \subsection{Results}\label{results}
+
+\subsubsection{Vital rate statistical
+models}\label{vital-rate-statistical-models}
+
+\subsubsection{Model validation}\label{model-validation-1}
+
+\subsubsection{Climate change
+simulations}\label{climate-change-simulations}
 
@@
 
-\begin{figure}[htbp]
-\centering
-\includegraphics{components/figure/manuscript-figure_2.pdf}
-\caption{One step forecasts from IPM version.}
-\end{figure}
-
-\begin{figure}[htbp]
-\centering
-\includegraphics{components/figure/manuscript-figure_3.pdf}
-\caption{One step forecasts from IPM version.}
-\end{figure}
+\includegraphics{components/figure/manuscript-figure_21.pdf}
+\includegraphics{components/figure/manuscript-figure_22.pdf}
 

### 2015-04-02 manuscript/manuscript.Rmd
Fix typos.

@@
 For individual level data we fit models for survival, growth, and recruitment of new individuals for each species.
-At the quadrat level we fit analogous models of extinction probability, percent cover increase/decrease, and quadrat colonization for each species.
-We describe the statistical models separately since fitting the models required different approaches at the individual and quadrat levels.
-All models contain four climate covariate that we chose \emph{a priori}: fall through spring precipitation at \emph{t}-1 and \emph{t}-2 (ppt1 and ppt2, respectively) and mean spring temperature at \emph{t}-1 and \emph{t}-2 (TmeanSpr1 and TmeanSpr2, respectively), where \emph{t} is the observation year.
+At the quadrat level we fit a single regression model for population growth.
+We describe the statistical models separately since fitting the models required different approaches.
+All models contain five climate covariate that we chose \emph{a priori}: "water year" precipitation at \emph{t}-1 (lagppt); fall through spring precipitation at \emph{t}-1 and \emph{t}-2 (ppt1 and ppt2, respectively) and mean spring temperature at \emph{t}-1 and \emph{t}-2 (TmeanSpr1 and TmeanSpr2, respectively), where \emph{t} is the observation year.
 
@@
 
-#### Vital rate models: individual level
+#### Vital rate models at the individual level
 We used logistic regression to model survival probability ($S$) of genet $i$ from species $j$ in quadrat group $Q$ from time $t$ to $t+1$:
@@
 
-#### Vital rate models: quadrat level
+#### Population model at the quadrat level
 The statistical approach used to model vital rates using aggregated data depends on the type of data collected.
@@
 We first considered fitting three vital rate models analagous to those we fit at the individual level: one for probability of extirpation within a quadrat (analagous to survival), one for cover change within a quadrat (analagous to growth), and one for probability of colonization within a quadrat (analagous to recruitment).
-However, within-quadrat extirpation and colonization events were rare in our time series ($N=9$ and $N=10$, respectively).
+However, within-quadrat extirpation and colonization events were rare in our time series ($N=9$ and $N=10$, respectively across all species).
 Given the broad spatial distribution of the quadrats we are studying, it is safe to assume that these events are in fact rare enough to be ignored for our purposes.
 So we constrained our statistical modeling of vital rates at the population level to change in percent cover within quadrats.
-For most of the remaining discussion of statistical modeling we refer to proportion data, which is simply percent data divided by 100.
+For the remaining discussion of statistical modeling we refer to proportion data, which is simply percent data divided by 100.
 
 An obvious choice for fitting a linear model to proportion data is beta regression because the support of the beta distribution is [0,1], not including true zeros or ones.
-However, when we used fitted model parameters from a beta regression in quadrat-based population model the simulated population tended toward 100% cover for all species.
+However, when we used fitted model parameters from a beta regression in a quadrat-based population model the simulated population tended toward 100% cover for all species.
 We therefore chose a more constrained modeling approach based on a truncated log-normal likelihood.
-The model for quadrat cover change ($G$) from time $t$ to $t+1$
+The model for quadrat cover change ($G$) from time $t$ to $t+1$ is
 
@@
 #).
-The log normal likelihood includes a truncation ($T[0,1]) to ensure that predicted values do not exceed 100% cover.
+The log normal likelihood includes a truncation ($T$[0,1]) to ensure that predicted values do not exceed 100% cover.
 
@@
 All of our analyses (model fitting and simulating) were conducted in R [@Team2013].
-We used the MCMC sampler in JAGS [@Plummer2003] to estimate the posterior distributions of model parameters and the package 'r2jags' [@Su2012] to connect R to JAGS.
-We obtained posterior distributions for all model parameters from three parallel MCMC chains run for 50,000 iterations, after discarding an initial 50,000 iterations.
+We used the MCMC sampler in JAGS [@Plummer2003] to estimate the posterior distributions of model parameters and the package 'rjags' to connect R to JAGS.
+We obtained posterior distributions for all model parameters from three parallel MCMC chains run for 50,000 iterations, thinned by 50, after discarding an initial 50,000 iterations.
 We assessed convergence visually and using the @Gelman1992 diagnostic in the R package `coda' [@Plummer2006].

### 2015-05-20 manuscript/manuscript.Rmd
updates figure 4 in ms

@@
 Labels on x-axis refer to: 'ppt' = 1% increase in mean precipitation; 'temp' = 1% increase in mean temperature; 'ppt + temp' = 1% increase in mean precipitation and mean temperature.", fig.width=6.5, fig.height=6, cache=FALSE, results='hide'}
-climBase <- readRDS("../montana/quadBM/simulations/baselineSimulationQBM.rds")
-climPpt <- readRDS("../montana/quadBM/simulations/pptChangeSimulationQBM.rds")
-climTemp <- readRDS("../montana/quadBM/simulations/tempChangeSimulationQBM.rds")
-climBoth <- readRDS("../montana/quadBM/simulations/temp-pptChangeSimulationQBM.rds")
-climBase$sim <- rep("AObserved", nrow(climBase)) 
-climPpt$sim <- rep("BPrecipitation", nrow(climBase)) 
-climTemp$sim <- rep("CTemperature", nrow(climBase)) 
-climBoth$sim <- rep("DBoth", nrow(climBase)) 
-climSimsQBM <- rbind(climBase, climPpt, climTemp, climBoth)
-qbm_means <- ddply(climSimsQBM, .(species, sim), summarise,
-                   avg_cover = mean(cover))
-tmpid <- which(qbm_means$sim=="AObserved")
-diffs <- list()
-for(i in 1:length(tmpid)){
-  obs <- qbm_means[tmpid[i],"avg_cover"]
-  diffs[[i]] <- (rep(obs,3) - qbm_means[(tmpid[i]+1):(tmpid[i]+3), "avg_cover"])/rep(obs,3)
-}
-names(diffs) <- unique(qbm_means$species)
-diff_df <- melt(as.data.frame(diffs))
-diff_df$sim <- rep(c("pptChange", "tempChange", "temppptChange"),4)
-diff_df$type <- rep("QBM", nrow(diff_df))
-
+qbm_diffs <- readRDS("../montana/quadBM/simulations/results/qbm_climatesims_percdiffs.rds")
+qbm_diffs$type <- rep("QBM", nrow(qbm_diffs))
 ipm_diffs <- readRDS("../montana/ipm/simulations/results/ipm_climatesims_percdiffs.rds")
@@
 
-diff_df <- rbind(diff_df, ipm_diffs)
+diff_df <- rbind(qbm_diffs, ipm_diffs)
 

### 2015-05-21 manuscript/manuscript.Rmd
updates ms with error table

@@
 -------
-### Vital rate statistical models
+### Comparison of Forecast Models
 
-### Model validation
-
-### Climate change simulations
+### Forecasting Climate Change Impacts
 
@@
 ipm_onestep <- readRDS("../montana/ipm/simulations/one_step_results/ipm_one-step_forecasts_combined.rds")
-resD <- readRDS("../montana/quadBM/simulations/oneStepResids_YearEffects.rds")
-resDny <- readRDS("../montana/quadBM/simulations/oneStepResids_NoYearEffects.rds")
-resD$simtype <- rep("full", nrow(resD))
-resDny$simtype <- rep("noyear", nrow(resDny))
-resDall <- rbind(resD, resDny)
-noZ <- which(resDall$lagCover!=0)
-resDall <- resDall[noZ,]
+qbm_onestep <- readRDS("../montana/quadBM/simulations/validation_results/qbm_one-step_forecasts_combined.rds")
 
-stats <- ddply(ipm_onestep, .(species, year), summarise,
-               mean_resid = mean(abs(resid)),
-               sd_resid = sd(resid),
-               min_resid = min(resid),
-               max_resid = max(resid))
-stats$model <- rep("IPM", nrow(stats))
-colnames(stats)[2] <- "simtype"
-stats2 <- ddply(resDall, .(species, simtype), summarise,
-               mean_resid = mean(abs(resids)),
-               sd_resid = sd(resids),
-               min_resid = min(resids),
-               max_resid = max(resids))
-stats2$model <- rep("QBM", nrow(stats2))
-stats$simtype <- stats2$simtype
-stats <- rbind(stats, stats2)
-stats <- stats[,c(ncol(stats),1:(ncol(stats)-1))]
-stats$simtype <- rep(c("Full", "Climate"),(nrow(stats)/2))
-colnames(stats) <- c("Model", "Species", "Year Structure", "Mean", "SD", "Min", "Max")
-print.xtable(xtable(stats, caption = "Table of residual summary statistics."),type="latex", comment=FALSE, include.rownames=FALSE)
+#IPM
+tmpipm <- ddply(ipm_onestep, .(quad, t1, species), summarise,
+                mean_pred = mean(cover.t1),
+                var_pred = var(cover.t1),
+                mean_obs = mean(obs.cover.t1),
+                resid = mean(resid))
+ipmstats <- ddply(tmpipm, .(species), summarise,
+                rho = cor(mean_pred, mean_obs),
+                mae = mean(abs(resid)),
+                var = mean(var_pred),
+                mean_cover = mean(mean_obs*100))
+ipmstats$model <- "IPM"
+
+#QBM
+tmpqbm <- ddply(qbm_onestep, .(quad, year, species), summarise,
+                mean_pred = mean(predcov),
+                var_pred = var(predcov),
+                mean_obs = mean(obscov),
+                resid = mean(resid))
+qbmstats <- ddply(tmpqbm, .(species), summarise,
+                rho = cor(mean_pred, mean_obs),
+                mae = mean(abs(resid)),
+                var = mean(var_pred),
+                mean_cover = mean(mean_obs*100))
+qbmstats$model <- "QBM"
+
+stats_table <- rbind(ipmstats, qbmstats)
+stats_table <- stats_table[ order(stats_table[,"species"]), ]
+stats_table <- stats_table[,c("model", "species", "mae",
+                              "rho","mean_cover")]
+
+colnames(stats_table) <- c("Model", "Species", "MAE", "Accuracy", "Mean Obs. Cover")
+print.xtable(xtable(stats_table, caption = "Mean absolute error (MAE) and accuracy (Pearson's $\\rho$) for one-step-ahead forecast from both model types.
+Forecasts were made without random year effects; only climate covariates could explain year-to-year variation"),type="latex", comment=FALSE, include.rownames=FALSE)
 ```
@@
 ```{r figure_2, dependson="plot-options", fig.cap="One step forecasts from IPM version.",fig.width=8, fig.height=8, cache=FALSE, results='hide'}
-resD <- readRDS("../montana/quadBM/simulations/oneStepResids_YearEffects.rds")
-resDny <- readRDS("../montana/quadBM/simulations/oneStepResids_NoYearEffects.rds")
-resD$simtype <- rep("full", nrow(resD))
-resDny$simtype <- rep("noyear", nrow(resDny))
-resDall <- rbind(resD, resDny)
-noZ <- which(resDall$lagCover!=0)
-resDall <- resDall[noZ,]
-
-myCols <- c("#237DA4", "#7ECAD0")
-# ggplot(resDall, aes(x=resids, fill=simtype))+
-#   geom_histogram(alpha=0.5)+
-#   facet_grid(.~species, scales = "free_x")+
-#   xlab("Year")+
-#   ylab("Residuals (% cover difference)")+
-#   theme_bw()+
-#   scale_fill_manual(values=myCols)
-
-resAgg <- ddply(resDall, .(year, species, simtype), summarise,
-                avg_prediction = mean(cover)*100,
-                up_prediction = quantile(cover, probs = 0.875)*100,
-                down_prediction = quantile(cover, probs = 0.125)*100,
-                avg_obs = mean(propCover)*100)
-resAgg$model <- rep("QBM", nrow(resAgg))
-
-ipm_onestep <- readRDS("../montana/ipm/simulations/one_step_results/ipm_one-step_forecasts_combined.rds")
-ipm_onestep$t1 <- ipm_onestep$t1+1900
-
-resAggipm <- ddply(ipm_onestep, .(t1, species, year), summarise,
-                avg_prediction = mean(cover.t1)*100,
-                up_prediction = quantile(cover.t1, probs = 0.875)*100,
-                down_prediction = quantile(cover.t1, probs = 0.125)*100,
-                avg_obs = mean(obs.cover.t1)*100)
-colnames(resAggipm)[1:3] <- c("year", "species", "simtype")
-resAggipm$simtype[which(resAggipm$simtype=="ayear effect")] <- "full"
-resAggipm$simtype[which(resAggipm$simtype!="full")] <- "noyear"
-resAggipm$model <- rep("IPM", nrow(resAggipm))
-
-resAgg_both <- rbind(resAgg, resAggipm)
-
-ggplot(subset(resAgg_both))+
-  geom_ribbon(aes(x=year, ymin=down_prediction, ymax=up_prediction, fill=simtype), alpha=0.3)+
-  geom_line(aes(x=year, y=avg_prediction, color=simtype), size=1.5)+
-  geom_point(aes(x=year, y=avg_obs), shape=1, size=3)+
-  facet_grid(species~model, scales="free")+
-  scale_fill_manual(values=c(myCols[1], "darkorange"), labels=c("Year effects", "No year effects"))+
-  scale_color_manual(values=c(myCols[1], "darkorange"))+
-  guides(color=FALSE)
-
-ggplot(subset(resAgg_both))+
-  geom_ribbon(aes(x=year, ymin=down_prediction, ymax=up_prediction, fill=model), alpha=0.3)+
-  geom_line(aes(x=year, y=avg_prediction, color=model), size=1.5)+
-  geom_point(aes(x=year, y=avg_obs), shape=1, size=3)+
-  facet_grid(species~simtype, scales="free")+
-  scale_fill_manual(values=c(myCols[1], "darkorange"))+
-  scale_color_manual(values=c(myCols[1], "darkorange"))+
-  guides(color=FALSE)
+# resD <- readRDS("../montana/quadBM/simulations/oneStepResids_YearEffects.rds")
+# resDny <- readRDS("../montana/quadBM/simulations/oneStepResids_NoYearEffects.rds")
+# resD$simtype <- rep("full", nrow(resD))
+# resDny$simtype <- rep("noyear", nrow(resDny))
+# resDall <- rbind(resD, resDny)
+# noZ <- which(resDall$lagCover!=0)
+# resDall <- resDall[noZ,]
+# 
+# myCols <- c("#237DA4", "#7ECAD0")
+# # ggplot(resDall, aes(x=resids, fill=simtype))+
+# #   geom_histogram(alpha=0.5)+
+# #   facet_grid(.~species, scales = "free_x")+
+# #   xlab("Year")+
+# #   ylab("Residuals (% cover difference)")+
+# #   theme_bw()+
+# #   scale_fill_manual(values=myCols)
+# 
+# resAgg <- ddply(resDall, .(year, species, simtype), summarise,
+#                 avg_prediction = mean(cover)*100,
+#                 up_prediction = quantile(cover, probs = 0.875)*100,
+#                 down_prediction = quantile(cover, probs = 0.125)*100,
+#                 avg_obs = mean(propCover)*100)
+# resAgg$model <- rep("QBM", nrow(resAgg))
+# 
+# 
+# 
+# ipm_onestep <- readRDS("../montana/ipm/simulations/one_step_results/ipm_one-step_forecasts_combined.rds")
+# ipm_onestep$t1 <- ipm_onestep$t1+1900
+# 
+# resAggipm <- ddply(ipm_onestep, .(t1, species, year), summarise,
+#                 avg_prediction = mean(cover.t1)*100,
+#                 up_prediction = quantile(cover.t1, probs = 0.875)*100,
+#                 down_prediction = quantile(cover.t1, probs = 0.125)*100,
+#                 avg_obs = mean(obs.cover.t1)*100)
+# colnames(resAggipm)[1:3] <- c("year", "species", "simtype")
+# resAggipm$simtype[which(resAggipm$simtype=="ayear effect")] <- "full"
+# resAggipm$simtype[which(resAggipm$simtype!="full")] <- "noyear"
+# resAggipm$model <- rep("IPM", nrow(resAggipm))
+# 
+# resAgg_both <- rbind(resAgg, resAggipm)
+# 
+# ggplot(subset(resAgg_both))+
+#   geom_ribbon(aes(x=year, ymin=down_prediction, ymax=up_prediction, fill=simtype), alpha=0.3)+
+#   geom_line(aes(x=year, y=avg_prediction, color=simtype), size=1.5)+
+#   geom_point(aes(x=year, y=avg_obs), shape=1, size=3)+
+#   facet_grid(species~model, scales="free")+
+#   scale_fill_manual(values=c(myCols[1], "darkorange"), labels=c("Year effects", "No year effects"))+
+#   scale_color_manual(values=c(myCols[1], "darkorange"))+
+#   guides(color=FALSE)
+# 
+# ggplot(subset(resAgg_both))+
+#   geom_ribbon(aes(x=year, ymin=down_prediction, ymax=up_prediction, fill=model), alpha=0.3)+
+#   geom_line(aes(x=year, y=avg_prediction, color=model), size=1.5)+
+#   geom_point(aes(x=year, y=avg_obs), shape=1, size=3)+
+#   facet_grid(species~simtype, scales="free")+
+#   scale_fill_manual(values=c(myCols[1], "darkorange"))+
+#   scale_color_manual(values=c(myCols[1], "darkorange"))+
+#   guides(color=FALSE)
 
@@
 ```{r figure_3, dependson="plot-options", fig.cap="One step forecasts from IPM version.",fig.width=8, fig.height=8, cache=FALSE, results='hide', eval=FALSE}
-ipm_onestep2 <- subset(ipm_onestep, year=="ayear effect")
-resDall2 <- subset(resDall, simtype=="full")
-ipm_qbm_onestep <- data.frame(year = c(ipm_onestep2$t1, resDall2$year),
-                              resids = c(ipm_onestep2$resid, resDall2$resids),
-                              model = c(rep("IPM", nrow(ipm_onestep2)),
-                                        rep("QBM", nrow(resDall2))),
-                              species = c(ipm_onestep2$species, resDall2$species))
-g1 <- ggplot(subset(ipm_qbm_onestep, species=="BOGR"))+
-  geom_hline(aes(yintercept=0), linetype=2)+
-  geom_boxplot(aes(x=as.character(year),y=resids,fill=model), outlier.shape = 1, outlier.size = 0, color="grey25")+
-  xlab("Year")+
-  ylab("Residuals (% cover difference)")+
-  theme_bw()+
-  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
-  scale_fill_manual(values=myCols)+
-  coord_cartesian(ylim = c(-50, 50))+
-  ggtitle(expression(italic("B. gracilis")))+
-  guides(fill=FALSE)
-g2 <- ggplot(subset(ipm_qbm_onestep, species=="HECO"))+
-  geom_hline(aes(yintercept=0), linetype=2)+
-  geom_boxplot(aes(x=as.character(year),y=resids,fill=model), outlier.shape = 1, outlier.size = 0, color="grey25")+
-  xlab("Year")+
-  ylab("Residuals (% cover difference)")+
-  theme_bw()+
-  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
-  scale_fill_manual(values=myCols)+
-  coord_cartesian(ylim = c(-10, 5))+
-  ggtitle(expression(italic("H. comata")))+
-  guides(fill=guide_legend(title=""))+
-  theme(legend.position=c(0.5,0.2))
-g3 <- ggplot(subset(ipm_qbm_onestep, species=="PASM"))+
-  geom_hline(aes(yintercept=0), linetype=2)+
-  geom_boxplot(aes(x=as.character(year),y=resids,fill=model), outlier.shape = 1, outlier.size = 0, color="grey25")+
-  xlab("Year")+
-  ylab("Residuals (% cover difference)")+
-  theme_bw()+
-  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
-  scale_fill_manual(values=myCols)+
-  ggtitle(expression(italic("P. smithii")))+
-  guides(fill=FALSE)+
-  coord_cartesian(ylim = c(-1.5, 1.5))
-g4 <- ggplot(subset(ipm_qbm_onestep, species=="POSE"))+
-  geom_hline(aes(yintercept=0), linetype=2)+
-  geom_boxplot(aes(x=as.character(year),y=resids,fill=model), outlier.shape = 1, outlier.size = 0, color="grey25")+
-  xlab("Year")+
-  ylab("Residuals (% cover difference)")+
-  theme_bw()+
-  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
-  scale_fill_manual(values=myCols)+
-  ggtitle(expression(italic("Poa secunda")))+
-  coord_cartesian(ylim = c(-10, 15))+
-  guides(fill=FALSE)
-pFinal <- grid.arrange(g1,g2,g3,g4, ncol=2, nrow=2)
-pFinal
+# ipm_onestep2 <- subset(ipm_onestep, year=="ayear effect")
+# resDall2 <- subset(resDall, simtype=="full")
+# ipm_qbm_onestep <- data.frame(year = c(ipm_onestep2$t1, resDall2$year),
+#                               resids = c(ipm_onestep2$resid, resDall2$resids),
+#                               model = c(rep("IPM", nrow(ipm_onestep2)),
+#                                         rep("QBM", nrow(resDall2))),
+#                               species = c(ipm_onestep2$species, resDall2$species))
+# g1 <- ggplot(subset(ipm_qbm_onestep, species=="BOGR"))+
+#   geom_hline(aes(yintercept=0), linetype=2)+
+#   geom_boxplot(aes(x=as.character(year),y=resids,fill=model), outlier.shape = 1, outlier.size = 0, color="grey25")+
+#   xlab("Year")+
+#   ylab("Residuals (% cover difference)")+
+#   theme_bw()+
+#   theme(axis.text.x = element_text(angle = 45, hjust = 1))+
+#   scale_fill_manual(values=myCols)+
+#   coord_cartesian(ylim = c(-50, 50))+
+#   ggtitle(expression(italic("B. gracilis")))+
+#   guides(fill=FALSE)
+# g2 <- ggplot(subset(ipm_qbm_onestep, species=="HECO"))+
+#   geom_hline(aes(yintercept=0), linetype=2)+
+#   geom_boxplot(aes(x=as.character(year),y=resids,fill=model), outlier.shape = 1, outlier.size = 0, color="grey25")+
+#   xlab("Year")+
+#   ylab("Residuals (% cover difference)")+
+#   theme_bw()+
+#   theme(axis.text.x = element_text(angle = 45, hjust = 1))+
+#   scale_fill_manual(values=myCols)+
+#   coord_cartesian(ylim = c(-10, 5))+
+#   ggtitle(expression(italic("H. comata")))+
+#   guides(fill=guide_legend(title=""))+
+#   theme(legend.position=c(0.5,0.2))
+# g3 <- ggplot(subset(ipm_qbm_onestep, species=="PASM"))+
+#   geom_hline(aes(yintercept=0), linetype=2)+
+#   geom_boxplot(aes(x=as.character(year),y=resids,fill=model), outlier.shape = 1, outlier.size = 0, color="grey25")+
+#   xlab("Year")+
+#   ylab("Residuals (% cover difference)")+
+#   theme_bw()+
+#   theme(axis.text.x = element_text(angle = 45, hjust = 1))+
+#   scale_fill_manual(values=myCols)+
+#   ggtitle(expression(italic("P. smithii")))+
+#   guides(fill=FALSE)+
+#   coord_cartesian(ylim = c(-1.5, 1.5))
+# g4 <- ggplot(subset(ipm_qbm_onestep, species=="POSE"))+
+#   geom_hline(aes(yintercept=0), linetype=2)+
+#   geom_boxplot(aes(x=as.character(year),y=resids,fill=model), outlier.shape = 1, outlier.size = 0, color="grey25")+
+#   xlab("Year")+
+#   ylab("Residuals (% cover difference)")+
+#   theme_bw()+
+#   theme(axis.text.x = element_text(angle = 45, hjust = 1))+
+#   scale_fill_manual(values=myCols)+
+#   ggtitle(expression(italic("Poa secunda")))+
+#   coord_cartesian(ylim = c(-10, 15))+
+#   guides(fill=FALSE)
+# pFinal <- grid.arrange(g1,g2,g3,g4, ncol=2, nrow=2)
+# pFinal
 ```

### 2015-05-21 manuscript/manuscript.tex
updates ms with error table

@@
 recruitment of new individuals for each species. At the quadrat level we
-fit analogous models of extinction probability, percent cover
-increase/decrease, and quadrat colonization for each species. We
-describe the statistical models separately since fitting the models
-required different approaches at the individual and quadrat levels. All
-models contain four climate covariate that we chose \emph{a priori}:
-fall through spring precipitation at \emph{t}-1 and \emph{t}-2 (ppt1 and
-ppt2, respectively) and mean spring temperature at \emph{t}-1 and
-\emph{t}-2 (TmeanSpr1 and TmeanSpr2, respectively), where \emph{t} is
-the observation year.
+fit a single regression model for population growth. We describe the
+statistical models separately since fitting the models required
+different approaches. All models contain five climate covariate that we
+chose \emph{a priori}: ``water year'' precipitation at \emph{t}-1
+(lagppt); fall through spring precipitation at \emph{t}-1 and \emph{t}-2
+(ppt1 and ppt2, respectively) and mean spring temperature at \emph{t}-1
+and \emph{t}-2 (TmeanSpr1 and TmeanSpr2, respectively), where \emph{t}
+is the observation year.
 
@@
 
-\paragraph{Vital rate models: individual
-level}\label{vital-rate-models-individual-level}
+\paragraph{Vital rate models at the individual
+level}\label{vital-rate-models-at-the-individual-level}
 
@@
 
-\paragraph{Vital rate models: quadrat
-level}\label{vital-rate-models-quadrat-level}
+\paragraph{Population model at the quadrat
+level}\label{population-model-at-the-quadrat-level}
 
@@
 recruitment). However, within-quadrat extirpation and colonization
-events were rare in our time series ($N=9$ and $N=10$, respectively).
-Given the broad spatial distribution of the quadrats we are studying, it
-is safe to assume that these events are in fact rare enough to be
-ignored for our purposes. So we constrained our statistical modeling of
-vital rates at the population level to change in percent cover within
-quadrats. For most of the remaining discussion of statistical modeling
-we refer to proportion data, which is simply percent data divided by
-100.
+events were rare in our time series ($N=9$ and $N=10$, respectively
+across all species). Given the broad spatial distribution of the
+quadrats we are studying, it is safe to assume that these events are in
+fact rare enough to be ignored for our purposes. So we constrained our
+statistical modeling of vital rates at the population level to change in
+percent cover within quadrats. For the remaining discussion of
+statistical modeling we refer to proportion data, which is simply
+percent data divided by 100.
 
@@
 not including true zeros or ones. However, when we used fitted model
-parameters from a beta regression in quadrat-based population model the
-simulated population tended toward 100\% cover for all species. We
+parameters from a beta regression in a quadrat-based population model
+the simulated population tended toward 100\% cover for all species. We
 therefore chose a more constrained modeling approach based on a
 truncated log-normal likelihood. The model for quadrat cover change
-($G$) from time $t$ to $t+1$
+($G$) from time $t$ to $t+1$ is
 
@@
 individual-level growth model (Eq. \#). The log normal likelihood
-includes a truncation (\$T{[}0,1{]}) to ensure that predicted values do
+includes a truncation ($T${[}0,1{]}) to ensure that predicted values do
 not exceed 100\% cover.
@@
 (Team 2013). We used the MCMC sampler in JAGS (Plummer 2003) to estimate
-the posterior distributions of model parameters and the package `r2jags'
-(Su and Yajima 2012) to connect R to JAGS. We obtained posterior
-distributions for all model parameters from three parallel MCMC chains
-run for 50,000 iterations, after discarding an initial 50,000
-iterations. We assessed convergence visually and using the Gelman and
-Rubin (1992) diagnostic in the R package `coda' (Plummer et al. 2006).
-Scale reduction factors for all parameters were less than 1.01,
-indicating convergence. For the purposes of introducing stochasticity in
-our population models, we saved the final 1,000 iterations from each of
-the three MCMC chains for all parameters to be used as randomly drawn
-values during population simulation.
+the posterior distributions of model parameters and the package `rjags'
+to connect R to JAGS. We obtained posterior distributions for all model
+parameters from three parallel MCMC chains run for 50,000 iterations,
+thinned by 50, after discarding an initial 50,000 iterations. We
+assessed convergence visually and using the Gelman and Rubin (1992)
+diagnostic in the R package `coda' (Plummer et al. 2006). Scale
+reduction factors for all parameters were less than 1.01, indicating
+convergence. For the purposes of introducing stochasticity in our
+population models, we saved the final 1,000 iterations from each of the
+three MCMC chains for all parameters to be used as randomly drawn values
+during population simulation.
 
@@
 
-\subsubsection{Vital rate statistical
-models}\label{vital-rate-statistical-models}
+\subsubsection{Comparison of Forecast
+Models}\label{comparison-of-forecast-models}
 
-\subsubsection{Model validation}\label{model-validation-1}
-
-\subsubsection{Climate change
-simulations}\label{climate-change-simulations}
+\subsubsection{Forecasting Climate Change
+Impacts}\label{forecasting-climate-change-impacts}
 
@@
 \centering
-\begin{tabular}{lllrrrr}
+\begin{tabular}{llrrr}
   \hline
-Model & Species & Year Structure & Mean & SD & Min & Max \\ 
+Model & Species & MAE & Accuracy & Mean Obs. Cover \\ 
   \hline
-IPM & BOGR & Full & 3.42 & 5.78 & -46.52 & 43.05 \\ 
-  IPM & BOGR & Climate & 5.77 & 11.40 & -230.10 & 47.81 \\ 
-  IPM & HECO & Full & 0.56 & 0.95 & -10.99 & 4.33 \\ 
-  IPM & HECO & Climate & 0.60 & 1.05 & -8.65 & 6.67 \\ 
-  IPM & PASM & Full & 0.17 & 0.30 & -4.29 & 1.04 \\ 
-  IPM & PASM & Climate & 0.14 & 0.20 & -2.04 & 0.99 \\ 
-  IPM & POSE & Full & 0.91 & 1.65 & -30.73 & 7.34 \\ 
-  IPM & POSE & Climate & 0.72 & 1.51 & -6.08 & 14.32 \\ 
-  QBM & BOGR & Full & 6.80 & 12.45 & -81.48 & 52.88 \\ 
-  QBM & BOGR & Climate & 7.69 & 13.61 & -94.71 & 54.33 \\ 
-  QBM & HECO & Full & 0.94 & 2.28 & -68.53 & 7.84 \\ 
-  QBM & HECO & Climate & 1.26 & 3.49 & -94.57 & 7.91 \\ 
-  QBM & PASM & Full & 0.26 & 0.50 & -15.13 & 1.56 \\ 
-  QBM & PASM & Climate & 0.27 & 0.53 & -9.69 & 1.67 \\ 
-  QBM & POSE & Full & 0.86 & 2.07 & -58.34 & 13.97 \\ 
-  QBM & POSE & Climate & 1.12 & 2.47 & -73.14 & 15.18 \\ 
+IPM & BOGR & 4.77 & 0.90 & 9.16 \\ 
+  QBM & BOGR & 5.30 & 0.88 & 9.22 \\ 
+  IPM & HECO & 0.60 & 0.80 & 1.22 \\ 
+  QBM & HECO & 2.03 & 0.84 & 1.25 \\ 
+  IPM & PASM & 0.18 & 0.65 & 0.40 \\ 
+  QBM & PASM & 0.17 & 0.87 & 0.40 \\ 
+  IPM & POSE & 0.78 & 0.61 & 1.23 \\ 
+  QBM & POSE & 1.30 & 0.57 & 1.25 \\ 
    \hline
 \end{tabular}
-\caption{Table of residual summary statistics.} 
+\caption{Mean absolute error (MAE) and accuracy (Pearson's $\rho$) for one-step-ahead forecast from both model types.
+Forecasts were made without random year effects; only climate covariates could explain year-to-year variation}
 \end{table}
@@
 \subsection{Figures}\label{figures}
-
-\includegraphics{components/figure/manuscript-figure_21.pdf}
-\includegraphics{components/figure/manuscript-figure_22.pdf}
 
@@
 
-Su, Y., and M. Yajima. 2012. R2jags: A Package for Running jags from R.
-http://CRAN. R-project. org/package= R2jags.
-
 Sæther, B. E., S. Engen, V. Grøtan, W. Fiedler, E. Matthysen, M. E.

### 2015-05-21 montana/ipm/simulations/one_step_results/check_1steps.R

@@ -28,7 +28,7 @@ all_d$year <- rep("ayear effect", nrow(all_d))
 
 # all_d <- rbind(all_d, all_d2)
 
-# saveRDS(all_d,"ipm_one-step_forecasts_combined.rds")
+saveRDS(all_d,"ipm_one-step_forecasts_combined.rds")
 
 library(ggplot2)
 library(ggplot2)

### 2015-05-21 montana/quadBM/simulations/onestep_qbm.R

@@ -127,7 +127,7 @@ for(do_species in sppList){
   yearsID <- unique(sppD$year)
   
   output <- data.frame(year=NA, lagcov=NA, obscov=NA, predcov=NA, 
-                       group=NA, rep=NA, species=NA)
+                       group=NA, rep=NA, species=NA, quad=NA)
   for(do_year in 1:yearsN){
     yr_data <- subset(sppD, year==yearsID[do_year])
     quads <- as.numeric(as.factor(yr_data$group))
@@ -139,6 +139,7 @@ for(do_species in sppList){
       ntmp <- yr_data[i,"percLagCover"]
       gtmp <- yr_data[i,"group"]
       gnum <- as.numeric(subset(quadnames, group==gtmp)["groupnums"])
+      qtmp <- yr_data[i,"quad"]
       for(do_sim in 1:nSim){
         randchain <- sample(x = climeff$Chain, size = 1)
         randiter <- sample(x = climeff$Iteration, size = 1)
@@ -168,7 +169,7 @@ for(do_species in sppList){
         # Store in output dataframe
         tmpout <- data.frame(year=yearsID[do_year], lagcov=ntmp, 
                              obscov=yr_data[i,"percCover"], predcov=covertmp,
-                             group=gtmp, rep=do_sim, species=do_species)
+                             group=gtmp, rep=do_sim, species=do_species, quad=qtmp)
         output <- rbind(output, tmpout)
         print(paste("Done with rep", do_sim, "of observation", i, "for year", do_year))
       }#end quad-year rep loop

### 2015-05-21 montana/quadBM/simulations/validation_results/check_1steps.R

@@ -4,7 +4,7 @@
 
 spp_list <- c("BOGR", "HECO", "PASM", "POSE")
 all_d <- data.frame(year=NA, lagcov=NA, obscov=NA, predcov=NA, 
-                    group=NA, rep=NA, species=NA)
+                    group=NA, rep=NA, species=NA, quad=NA)
 for(i in 1:length(spp_list)){
   tmp_d <- readRDS(paste("onestep_", spp_list[i],".RDS",sep=""))
 #   tmp_d$species <- rep(spp_list[i],nrow(tmp_d))
@@ -12,6 +12,8 @@ for(i in 1:length(spp_list)){
 }
 all_d <- all_d[2:nrow(all_d),]
 all_d$resid <- with(all_d, (predcov*100)-(obscov*100))
+
+saveRDS(all_d,"qbm_one-step_forecasts_combined.rds")
 
 library(ggplot2)
 ggplot(all_d, aes(x=obscov, y=predcov))+

### 2015-05-21 manuscript/manuscript.Rmd
updates Table 1

@@
 
+mae_list <- list()
+mae_pvals <- list()
+species <- unique(qbmstats$species)
+for(spp in species){
+  tmp1 <- subset(tmpipm, species==spp)
+  tmp2 <- subset(tmpqbm, species==spp)
+  err1 <- as.data.frame(abs(tmp1$mean_obs - tmp1$mean_pred))
+  err2 <- as.data.frame(abs(tmp2$mean_obs - tmp2$mean_pred))
+  colnames(err1) <- "resid"
+  err1$model <- "ipm"
+  colnames(err2) <- "resid"
+  err2$model <- "qbm"
+  errd <-  rbind(err1, err2)
+  mae_ttest <- with(errd, t.test(resid~model, alternative = "less"))
+  mae_list[[spp]] <- mae_ttest
+  mae_pvals[[spp]] <- mae_ttest[["p.value"]]
+}
+
+mae_ps <- melt(mae_pvals)
+
 stats_table <- rbind(ipmstats, qbmstats)
@@
 
-colnames(stats_table) <- c("Model", "Species", "MAE", "Accuracy", "Mean Obs. Cover")
+colnames(stats_table) <- c("Species", "Model", "MAE", "Accuracy", "Mean Obs. Cover")
 print.xtable(xtable(stats_table, caption = "Mean absolute error (MAE) and accuracy (Pearson's $\\rho$) for one-step-ahead forecast from both model types.
-Forecasts were made without random year effects; only climate covariates could explain year-to-year variation"),type="latex", comment=FALSE, include.rownames=FALSE)
+Forecasts were made without random year effects; only climate covariates could explain year-to-year variation."),type="latex", comment=FALSE, include.rownames=FALSE)
 ```

### 2015-05-21 manuscript/manuscript.tex
updates Table 1

@@
   \hline
-Model & Species & MAE & Accuracy & Mean Obs. Cover \\ 
+Species & Model & MAE & Accuracy & Mean Obs. Cover \\ 
   \hline
@@
 \caption{Mean absolute error (MAE) and accuracy (Pearson's $\rho$) for one-step-ahead forecast from both model types.
-Forecasts were made without random year effects; only climate covariates could explain year-to-year variation}
+Forecasts were made without random year effects; only climate covariates could explain year-to-year variation.}
 \end{table}

### 2015-05-21 manuscript/manuscript.Rmd
updates figures in ms

@@
 library(xtable)
+library(ggmcmc)
 ```
@@
 stats_table <- stats_table[ order(stats_table[,"species"]), ]
-stats_table <- stats_table[,c("model", "species", "mae",
+stats_table <- stats_table[,c("species", "model", "mae",
                               "rho","mean_cover")]
+# comment          <- list()
+# comment$pos      <- list()
+# comment$pos[[1]] <- c(nrow(stats_table))
+# comment$command  <- c(paste("\\hline \n",  # we`ll replace all default hlines with this and the ones below
+#                             "The IPM MAE is significantly lower for HECO ($p = 3.3\times10^{-10}$) and POSE ($p = 0.00013$. MAEs are statisticially similar between models for BOGR and PASM.   \n",
+#                             sep = ""))
 
@@
 print.xtable(xtable(stats_table, caption = "Mean absolute error (MAE) and accuracy (Pearson's $\\rho$) for one-step-ahead forecast from both model types.
-Forecasts were made without random year effects; only climate covariates could explain year-to-year variation."),type="latex", comment=FALSE, include.rownames=FALSE)
+Forecasts were made without random year effects; only climate covariates could explain year-to-year variation."),type="latex", comment=FALSE,
+             include.rownames=FALSE, caption.placement="top")
 ```
+The IPM MAE is significantly lower for HECO ($p = 3.3 \times 10^{-10}$) and POSE ($p = 0.00013$). MAEs are statisticially similar between models for BOGR and PASM.
 
@@
 
+```{r ipm_climeffs, cache=FALSE, results='hide', include=FALSE}
+source("../montana/ipm/vitalRateRegs/climate_effects_plots.R")
+```
+
+![Standardized regression coefficients for climate effects in the vital rate models for the IPM.
+Shown are the median (circles), 75% confidence intervals (thick lines), and 95% confidence intervals (thin lines).
+Left panels (black) are for the growth regressions; middle panels (blue) are for the survival regressions; right panels (orange) are for the recruitment regressions.](components/figure/ipm_climeffs.pdf)
+
+```{r quad_climeffs, cache=FALSE, results='hide', include=FALSE}
+source("../montana/quadBM/vitalRateRegressions/truncNormModel/climate_effects_plot.R")
+```
+
+![Standardized regression coefficients for climate effects in the vital rate models for the QBM.
+Shown are the median (circles), 75% confidence intervals (thick lines), and 95% confidence intervals (thin lines).](components/figure/qbm_climeffs.pdf)
+
 ```{r figure_4, dependson="plot-options", fig.cap="Proportional change in species' mean cover caused by a 1% increase in observed precipitation, temperature, or both as predicted by the individual-based IPM and the aggregate-based QBM.

### 2015-05-21 manuscript/manuscript.tex
updates figures in ms

@@
 \documentclass[12pt,]{article}
-\usepackage[T1]{fontenc}
 \usepackage{lmodern}
@@
 \usepackage{fixltx2e} % provides \textsubscript
-% use upquote if available, for straight quotes in verbatim environments
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
 \ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
+  \usepackage[T1]{fontenc}
   \usepackage[utf8]{inputenc}
@@
 \fi
+% use upquote if available, for straight quotes in verbatim environments
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
 % use microtype if available
-\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
+\IfFileExists{microtype.sty}{%
+\usepackage{microtype}
+\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
+}{}
 \usepackage[margin=1in]{geometry}
 \usepackage{graphicx}
-% Redefine \includegraphics so that, unless explicit options are
-% given, the image width will not exceed the width of the page.
-% Images get their normal width if they fit onto the page, but
-% are scaled down if they would overflow the margins.
 \makeatletter
-\def\ScaleIfNeeded{%
-  \ifdim\Gin@nat@width>\linewidth
-    \linewidth
-  \else
-    \Gin@nat@width
-  \fi
-}
+\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
+\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
 \makeatother
-\let\Oldincludegraphics\includegraphics
-{%
- \catcode`\@=11\relax%
- \gdef\includegraphics{\@ifnextchar[{\Oldincludegraphics}{\Oldincludegraphics[width=\ScaleIfNeeded]}}%
-}%
+% Scale images if necessary, so that they will not overflow the page
+% margins by default, and it is still possible to overwrite the defaults
+% using explicit options in \includegraphics[width, height, ...]{}
+\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
 \ifxetex
@@
 
+%%% Use protect on footnotes to avoid problems with footnotes in titles
+\let\rmarkdownfootnote\footnote%
+\def\footnote{\protect\rmarkdownfootnote}
+
+%%% Change title format to be more compact
+\usepackage{titling}
+\setlength{\droptitle}{-2em}
+  \title{}
+  \pretitle{\vspace{\droptitle}}
+  \posttitle{}
   \author{}
+  \preauthor{}\postauthor{}
   \date{}
+  \predate{}\postdate{}
+
 \usepackage{lineno}
@@
 
-\normalsize
+\maketitle
 
@@
 \emph{Andrew T. Tredennick
-(\href{mailto:atredenn@gmail.com}{\href{mailto:atredenn@gmail.com}{atredenn@gmail.com}}),
+(\href{mailto:atredenn@gmail.com}{\href{mailto:atredenn@gmail.com}{\nolinkurl{atredenn@gmail.com}}}),
 Department of Wildland Resources and the Ecology Center, Utah State
@@
 Research Laboratory in eastern Montana's northern mixed prairie near
-Miles City, Montana, USA ($46^{\circ}$ 19' N, $105^{\circ}$ 48' W). The
-dataset is freely available on Ecological Archives (Anderson et al.
+Miles City, Montana, USA (\(46^{\circ}\) 19' N, \(105^{\circ}\) 48' W).
+The dataset is freely available on Ecological Archives (Anderson et al.
 2011), and interested readers should refer to the metadata therein for a
@@
 From 1932 to 1945 individual plants were identified and mapped annually
-in 44 1-$\text{m}^2$ quadrats using a pantograph. The quadrats were
+in 44 1-\(\text{m}^2\) quadrats using a pantograph. The quadrats were
 distributed in six pastures, each assigned a grazing treatment of light
@@
 account for potential differences among the grazing treatments, but do
-not focus on grazing$\times$climate interactions. The annual maps of the
-quadrats were digitized and the fates of individual plants tracked and
-extracted using a computer program. Daily climate data, which we
+not focus on grazing\(\times\)climate interactions. The annual maps of
+the quadrats were digitized and the fates of individual plants tracked
+and extracted using a computer program. Daily climate data, which we
 aggregated into climate variables of interest, are available for the
@@
 models, \textbf{Y} is always the relevant vector of observations (e.g.,
-whether a genet survived {[}1{]} or not {[}0{]} from year $t$ to $t+1$).
+whether a genet survived {[}1{]} or not {[}0{]} from year \(t\) to
+\(t+1\)).
 
@@
 
-We used logistic regression to model survival probability ($S$) of genet
-$i$ from species $j$ in quadrat group $Q$ from time $t$ to $t+1$:
+We used logistic regression to model survival probability (\(S\)) of
+genet \(i\) from species \(j\) in quadrat group \(Q\) from time \(t\) to
+\(t+1\):
 
@@
 
-where $x_{ij,t}$ is the log of genet size, $\gamma^{S}_{j,t}$ is a
-year-specific intercept, $\beta^{S}_{j,t}$ is the year-specific slope
-parameter for size, $\phi^{S}_{jQ}$ is the random effect of quadrat
-group location, and $\theta^{S}_{k}$ is the fixed parameter for the
-effect of the $k$th climate covariate at time $t$ ($C_{k,t}$). We
+where \(x_{ij,t}\) is the log of genet size, \(\gamma^{S}_{j,t}\) is a
+year-specific intercept, \(\beta^{S}_{j,t}\) is the year-specific slope
+parameter for size, \(\phi^{S}_{jQ}\) is the random effect of quadrat
+group location, and \(\theta^{S}_{k}\) is the fixed parameter for the
+effect of the \(k\)th climate covariate at time \(t\) (\(C_{k,t}\)). We
 include density-dependence by estimating the effect of crowding on the
-focal individual by other individuals of the same species. $\omega$ is
-the effect of crowding and $w_{t,Q}$ is the crowding experienced by the
-focal individual at time $t$ in quadrat group $Q$.
+focal individual by other individuals of the same species. \(\omega\) is
+the effect of crowding and \(w_{t,Q}\) is the crowding experienced by
+the focal individual at time \(t\) in quadrat group \(Q\).
 
 We modeled growth as Gaussian process describing genet size at time
-$t+1$ as a function of size at $t$ and climate covariates:
+\(t+1\) as a function of size at \(t\) and climate covariates:
 
@@
 
-where $x$ is log genet size and all other parameters are as described
+where \(x\) is log genet size and all other parameters are as described
 for the survival regression.
@@
 specific parent to new genets. So, for recruitment, we work at the
-quadrat level and model the number of new individuals of species $j$ in
-quadrat $q$ recruiting at time $t+1$ as a function of quadrat
-``effective cover'' ($A'$) in the previous year ($t$). Effective cover
-is a mixture of observed cover ($A$) in the focal quadrat ($q$) and the
-mean cover across the entire group ($\bar{A}$) of $Q$ quadrats in which
-$q$ is located:
+quadrat level and model the number of new individuals of species \(j\)
+in quadrat \(q\) recruiting at time \(t+1\) as a function of quadrat
+``effective cover'' (\(A'\)) in the previous year (\(t\)). Effective
+cover is a mixture of observed cover (\(A\)) in the focal quadrat
+(\(q\)) and the mean cover across the entire group (\(\bar{A}\)) of
+\(Q\) quadrats in which \(q\) is located:
 
@@
 
-where $p$ is a mixing fraction between 0 and 1 that is estimated within
-the model.
+where \(p\) is a mixing fraction between 0 and 1 that is estimated
+within the model.
 
-We assume the number of individuals, $Y^{R}$, recruiting at time $t+1$
-follows a negative binomial distribution:
+We assume the number of individuals, \(Y^{R}\), recruiting at time
+\(t+1\) follows a negative binomial distribution:
 
@@
 
-where $\lambda$ is the mean intensity and $\zeta$ is the size parameter.
-We define $\lambda$ as:
+where \(\lambda\) is the mean intensity and \(\zeta\) is the size
+parameter. We define \(\lambda\) as:
 
@@
 
-where $A'$ is effective cover ($\text{cm}^2$) of species $j$ in quadrat
-$q$ and all other terms are as in the survival and growth regressions.
+where \(A'\) is effective cover (\(\text{cm}^2\)) of species \(j\) in
+quadrat \(q\) and all other terms are as in the survival and growth
+regressions.
 
@@
 recruitment). However, within-quadrat extirpation and colonization
-events were rare in our time series ($N=9$ and $N=10$, respectively
+events were rare in our time series (\(N=9\) and \(N=10\), respectively
 across all species). Given the broad spatial distribution of the
@@
 truncated log-normal likelihood. The model for quadrat cover change
-($G$) from time $t$ to $t+1$ is
+(\(G\)) from time \(t\) to \(t+1\) is
 
@@
 
-where $x_{jq,t}$ is the log of species' $j$ proportional cover in
-quadrat $q$ at time $t$ and all other parameters are as in the
+where \(x_{jq,t}\) is the log of species' \(j\) proportional cover in
+quadrat \(q\) at time \(t\) and all other parameters are as in the
 individual-level growth model (Eq. \#). The log normal likelihood
-includes a truncation ($T${[}0,1{]}) to ensure that predicted values do
-not exceed 100\% cover.
+includes a truncation (\(T\){[}0,1{]}) to ensure that predicted values
+do not exceed 100\% cover.
 
@@
 
-where $\bar{\theta_{k}}$ is the interspecific effect of the $k$th
+where \(\bar{\theta_{k}}\) is the interspecific effect of the \(k\)th
 climate covariate.
@@
 \centering
+\caption{Mean absolute error (MAE) and accuracy (Pearson's $\rho$) for one-step-ahead forecast from both model types.
+Forecasts were made without random year effects; only climate covariates could explain year-to-year variation.}
 \begin{tabular}{llrrr}
@@
   \hline
-IPM & BOGR & 4.77 & 0.90 & 9.16 \\ 
-  QBM & BOGR & 5.30 & 0.88 & 9.22 \\ 
-  IPM & HECO & 0.60 & 0.80 & 1.22 \\ 
-  QBM & HECO & 2.03 & 0.84 & 1.25 \\ 
-  IPM & PASM & 0.18 & 0.65 & 0.40 \\ 
-  QBM & PASM & 0.17 & 0.87 & 0.40 \\ 
-  IPM & POSE & 0.78 & 0.61 & 1.23 \\ 
-  QBM & POSE & 1.30 & 0.57 & 1.25 \\ 
+BOGR & IPM & 4.77 & 0.90 & 9.16 \\ 
+  BOGR & QBM & 5.30 & 0.88 & 9.22 \\ 
+  HECO & IPM & 0.60 & 0.80 & 1.22 \\ 
+  HECO & QBM & 2.03 & 0.84 & 1.25 \\ 
+  PASM & IPM & 0.18 & 0.65 & 0.40 \\ 
+  PASM & QBM & 0.17 & 0.87 & 0.40 \\ 
+  POSE & IPM & 0.78 & 0.61 & 1.23 \\ 
+  POSE & QBM & 1.30 & 0.57 & 1.25 \\ 
    \hline
 \end{tabular}
-\caption{Mean absolute error (MAE) and accuracy (Pearson's $\rho$) for one-step-ahead forecast from both model types.
-Forecasts were made without random year effects; only climate covariates could explain year-to-year variation.}
 \end{table}
+
+The IPM MAE is significantly lower for HECO
+(\(p = 3.3 \times 10^{-10}\)) and POSE (\(p = 0.00013\)). MAEs are
+statisticially similar between models for BOGR and PASM
 
@@
 \subsection{Figures}\label{figures}
+
+\begin{figure}[htbp]
+\centering
+\includegraphics{components/figure/ipm_climeffs.pdf}
+\caption{Standardized regression coefficients for climate effects in the
+vital rate models for the IPM. Shown are the median (circles), 75\%
+confidence intervals (thick lines), and 95\% confidence intervals (thin
+lines). Left panels (black) are for the growth regressions; middle
+panels (blue) are for the survival regressions; right panels (orange)
+are for the recruitment regressions.}
+\end{figure}
+
+\begin{figure}[htbp]
+\centering
+\includegraphics{components/figure/qbm_climeffs.pdf}
+\caption{Standardized regression coefficients for climate effects in the
+vital rate models for the QBM. Shown are the median (circles), 75\%
+confidence intervals (thick lines), and 95\% confidence intervals (thin
+lines).}
+\end{figure}
 
@@
 
-\subsection{References}\label{references}
+\subsection*{References}\label{references}
+\addcontentsline{toc}{subsection}{References}
 

### 2015-05-21 montana/ipm/vitalRateRegs/climate_effects_plots.R

@@ -3,7 +3,9 @@
 library(ggmcmc)
 library(gridExtra)
 
-allfiles <- list.files("../../speciesData/")
+setwd("~/Repos/MicroMesoForecast/manuscript")
+
+allfiles <- list.files("../montana/speciesData/")
 removals <- grep("csv", allfiles)
 species_list <- allfiles[-removals]
 
@@ -36,7 +38,7 @@ make_plot <- function(data, col="black"){
 ##  Growth
 glist <- list()
 for(spp in species_list){
-  tmp_data <- readRDS(paste("./growth/growth_stanmcmc_", spp, ".RDS", sep=""))
+  tmp_data <- readRDS(paste("../montana/ipm/vitalRateRegs/growth/growth_stanmcmc_", spp, ".RDS", sep=""))
   tmp_clim <- tmp_data[grep("b2", tmp_data[,"Parameter"]), ]
   tmp_clim[,"Parameter"] <- rep(clim_covs, each=3000)
   tmp_agg <- ddply(tmp_clim, .(Parameter), summarise,
@@ -58,7 +60,7 @@ for(i in 1:12){
   params[i] <- paste("b2[",i,"]", sep="")
 }
 for(spp in species_list){
-  tmp_data <- readRDS(paste("./survival/survival_stanmcmc_", spp, ".RDS", sep=""))
+  tmp_data <- readRDS(paste("../montana/ipm/vitalRateRegs/survival/survival_stanmcmc_", spp, ".RDS", sep=""))
   keeps <- which(tmp_data$Parameter %in% params)
   tmp_clim <- tmp_data[keeps, ]
   tmp_clim[,"Parameter"] <- rep(clim_covs, each=3000)
@@ -81,7 +83,7 @@ for(i in 1:7){
   params[i] <- paste("b2[",i,"]", sep="")
 }
 for(spp in species_list){
-  tmp_data <- readRDS(paste("./recruitment/recruitment_stanmcmc_", spp, ".RDS", sep=""))
+  tmp_data <- readRDS(paste("../montana/ipm/vitalRateRegs/recruitment/recruitment_stanmcmc_", spp, ".RDS", sep=""))
   keeps <- which(tmp_data$Parameter %in% params)
   tmp_clim <- tmp_data[keeps, ]
   tmp_clim[,"Parameter"] <- rep(clim_covs[1:7], each=3000)
@@ -103,10 +105,16 @@ for(spp in species_list){
 #                          rlist[[1]], rlist[[2]], rlist[[3]], rlist[[4]],
[…]

### 2015-05-21 montana/quadBM/vitalRateRegressions/truncNormModel/climate_effects_plot.R

@@ -1,0 +1,48 @@
+##  Script to plot climate effects from the quad model
+
+library(ggmcmc)
+library(gridExtra)
+
+setwd("~/Repos/MicroMesoForecast/manuscript")
+
+allfiles <- list.files("../montana/speciesData/")
+removals <- grep("csv", allfiles)
+species_list <- allfiles[-removals]
+
+clim_covs <- c("pptLag", "ppt1", "ppt2", "TmeanSpr1", "TmeanSpr2",
+               "ppt1xTmeanSpr1", "ppt2xTmeanSpr2",
+               "coverXpptLag", "coverXppt1", "coverXppt2",
+               "coverXTmeanSpr1", "coverXTmeanSpr2")
+
+## Plot function
+make_plot <- function(data, col="black"){
+  gtmp <- ggplot(data, aes(x = med, y = Parameter)) + 
+    geom_vline(aes(xintercept=0), linetype=2)+
+    geom_point(size = 3, color=col) + 
+    geom_segment(aes(x = Low, xend = High, yend = Parameter), size = 1.5, color=col) + 
+    geom_segment(aes(x = low, xend = high, yend = Parameter), size = 0.5, color=col) + 
+    xlab("HPD") +
+    ylab("Parameter")+
+    ggtitle(spp)
+}
+
+
+glist <- list()
+for(spp in species_list){
+  tmp_data <- readRDS(paste("../montana/quadBM/vitalRateRegressions/truncNormModel/popgrowth_stanmcmc_", spp, ".RDS", sep=""))
+  tmp_clim <- tmp_data[grep("b2", tmp_data[,"Parameter"]), ]
+  tmp_clim[,"Parameter"] <- rep(clim_covs, each=3000)
+  tmp_agg <- ddply(tmp_clim, .(Parameter), summarise,
+                   med = median(value),
+                   high = quantile(value, 0.975),
+                   High = quantile(value, 0.875),
+                   low = quantile(value, 0.025),
+                   Low = quantile(value, 0.125))
+  glist[[spp]] <- make_plot(tmp_agg, col="black")
+}
+grid.arrange(glist[[1]], glist[[2]], glist[[3]], glist[[4]],
+                       ncol=2, nrow=2)
+g <- arrangeGrob(glist[[1]], glist[[2]], glist[[3]], glist[[4]],
+                 ncol=2, nrow=2)
+ggsave(file="components/figure/qbm_climeffs.pdf", g, width = 7, height = 5)
+

### 2015-05-27 manuscript/manuscript.Rmd
ms updates

@@
 Then, we drew the full parameter set (climate effects and density-dependence fixed effects) from a randomly selected MCMC iteration.
+Using this approach, rather than simply using coefficient point estimates, ensures that relatively unimportant climate covariates (those that broadly overlap 0) have little effect on the simulation results.
 Since our focus was on the contribution of climate covariates to population states, we set the random year effects and the random group effects to zero.
@@
 ### Comparison of Forecast Models
+We performed one-step-ahead forecasts to compare the two models (IPM vs.
+QBM).
+We did not use the random year effects for these forecasts.
+Thus, beyond the effect of genet size or quadrat cover, only the climate covariates could affect expect plant cover.
+The models had similar accuracy ($\rho$, correlation between observations and mean prediction from 100 forecasts) across all species (Table 1).
+However, the IPM had significantly lower overall error (MAE, mean absolute error) for two species (*H.
+comita* and *P.
+smithii*), and in no case did the QBM significantly outperform the IPM (Table 1).
 
 ### Forecasting Climate Change Impacts
+Equilibrium cover simulated from the models was sensitive to climate perturbations, but the IPM and QBM produced inconsistent results (Figure 3).
+For *B.
+gracilis*, the IPM predicted a modest increase in cover with a 1% increase in the mean of precipitation or a 1% increase in the mean of temperature, the compounding effect of both being a 20% increase in cover.
+This reflects the relatively strong effects of precipitation and climate on *B.
+gracilis* genet growth and recruitment (Figure 1).
+The QBM also predicted increased *B.
+gracilis* cover with a precipitation increase, but increasing temperature decreased equilibrium cover (Figure 3).
+
+The IPM and QBM produced consistent predictions for *H.
+comita* under increased precipitation and when both precipitation and temperature were increased (Figure 3).
+However, the IPM predicted more modest changes than the QBM, and for a temperature increase the two models differed: the IPM predicted an increase in cover while the QBM predicted the opposite.
 
@@
 
-1. Climate detection
-2. Limitations:
-  * Sample size at quadrat level
-  * are these species inherently sensitive to climate (compare to Chu and Adler in review -- say why we chose Montana site -- most sensitive to climate variables)
-  * we know recruitment is important, and this is something that may be easily missed at quad level
-3. Implications
+It is perhaps startling that two different types of models fit using the same data produce drastically different forecasts (Figure 3).
 

### 2015-05-27 manuscript/manuscript.tex
ms updates

--- 2015-05-21 manuscript/manuscript.tex
@@
 and density-dependence fixed effects) from a randomly selected MCMC
-iteration. Since our focus was on the contribution of climate covariates
+iteration. Using this approach, rather than simply using coefficient
+point estimates, ensures that relatively unimportant climate covariates
+(those that broadly overlap 0) have little effect on the simulation
+results. Since our focus was on the contribution of climate covariates
 to population states, we set the random year effects and the random
@@
 
+We performed one-step-ahead forecasts to compare the two models (IPM
+vs.~QBM). We did not use the random year effects for these forecasts.
+Thus, beyond the effect of genet size or quadrat cover, only the climate
+covariates could affect expect plant cover. The models had similar
+accuracy ($\rho$, correlation between observations and mean prediction
+from 100 forecasts) across all species (Table 1). However, the IPM had
+significantly lower overall error (MAE, mean absolute error) for two
+species (\emph{H. comita} and \emph{P. smithii}), and in no case did the
+QBM significantly outperform the IPM (Table 1).
+
 \subsubsection{Forecasting Climate Change
 Impacts}\label{forecasting-climate-change-impacts}
+
+Equilibrium cover simulated from the models was sensitive to climate
+perturbations, but the IPM and QBM produced inconsistent results (Figure
+3). For \emph{B. gracilis}, the IPM predicted a modest increase in cover
+with a 1\% increase in the mean of precipitation or a 1\% increase in
+the mean of temperature, the compounding effect of both being a 20\%
+increase in cover. This reflects the relatively strong effects of
+precipitation and climate on \emph{B. gracilis} genet growth and
+recruitment (Figure 1). The QBM also predicted increased \emph{B.
+gracilis} cover with a precipitation increase, but increasing
+temperature decreased equilibrium cover (Figure 3).
+
+The IPM and QBM produced consistent predictions for \emph{H. comita}
+under increased precipitation and when both precipitation and
+temperature were increased (Figure 3). However, the IPM predicted more
+modest changes than the QBM, and for a temperature increase the two
+models differed: the IPM predicted an increase in cover while the QBM
+predicted the opposite.
 
@@
 (\emph{e.g.} Freckleton et al. 2011).
+
+It is perhaps startling that two different types of models fit using the
+same data produce drastically different forecasts (Figure 3).
 
@@
   \hline
-IPM & BOGR & 4.77 & 0.90 & 9.16 \\ 
-  QBM & BOGR & 5.30 & 0.88 & 9.22 \\ 
-  IPM & HECO & 0.60 & 0.80 & 1.22 \\ 
-  QBM & HECO & 2.03 & 0.84 & 1.25 \\ 
-  IPM & PASM & 0.18 & 0.65 & 0.40 \\ 
-  QBM & PASM & 0.17 & 0.87 & 0.40 \\ 
-  IPM & POSE & 0.78 & 0.61 & 1.23 \\ 
-  QBM & POSE & 1.30 & 0.57 & 1.25 \\ 
+BOGR & IPM & 4.77 & 0.90 & 9.16 \\ 
+  BOGR & QBM & 5.30 & 0.88 & 9.22 \\ 
+  HECO & IPM & 0.60 & 0.80 & 1.22 \\ 
+  HECO & QBM & 2.03 & 0.84 & 1.25 \\ 
+  PASM & IPM & 0.18 & 0.65 & 0.40 \\ 
+  PASM & QBM & 0.17 & 0.87 & 0.40 \\ 
+  POSE & IPM & 0.78 & 0.61 & 1.23 \\ 
+  POSE & QBM & 1.30 & 0.57 & 1.25 \\ 
    \hline
 \end{tabular}
-\caption{Mean absolute error (MAE) and accuracy (Pearson's $\rho$) for one-step-ahead forecast from both model types.
-Forecasts were made without random year effects; only climate covariates could explain year-to-year variation.}
 \end{table}
+
+The IPM MAE is significantly lower for HECO ($p = 3.3 \times 10^{-10}$)
+and POSE ($p = 0.00013$). MAEs are statisticially similar between models
+for BOGR and PASM.
 

### 2015-05-28 manuscript/manuscript.Rmd
updates ms and adds climate sensitivity figure script

@@
 -------
-```{r figure_1,cache=FALSE, results='hide'}
-# resD <- readRDS("../montana/quadBM/simulations/oneStepResids_YearEffects.rds")
-# resDny <- readRDS("../montana/quadBM/simulations/oneStepResids_NoYearEffects.rds")
-# resD$simtype <- rep("full", nrow(resD))
-# resDny$simtype <- rep("noyear", nrow(resDny))
-# resDall <- rbind(resD, resDny)
-# noZ <- which(resDall$lagCover!=0)
-# resDall <- resDall[noZ,]
-# 
-# myCols <- c("#237DA4", "#7ECAD0")
-# # ggplot(resDall, aes(x=resids, fill=simtype))+
-# #   geom_histogram(alpha=0.5)+
-# #   facet_grid(.~species, scales = "free_x")+
-# #   xlab("Year")+
-# #   ylab("Residuals (% cover difference)")+
-# #   theme_bw()+
-# #   scale_fill_manual(values=myCols)
-# 
-# resAgg <- ddply(resDall, .(year, species, simtype), summarise,
-#                 avg_prediction = mean(cover)*100,
-#                 up_prediction = quantile(cover, probs = 0.875)*100,
-#                 down_prediction = quantile(cover, probs = 0.125)*100,
-#                 avg_obs = mean(propCover)*100)
-# resAgg$model <- rep("QBM", nrow(resAggipm))
 
-# ggplot(subset(resAgg))+
-#   geom_ribbon(aes(x=year, ymin=down_prediction, ymax=up_prediction, fill=simtype), alpha=0.3)+
-#   geom_line(aes(x=year, y=avg_prediction, color=simtype), size=1.5)+
-# #   geom_line(aes(x=year, y=up_prediction, color=simtype), linetype=2, alpha=0.5)+
-# #   geom_line(aes(x=year, y=down_prediction, color=simtype), linetype=2, alpha=0.5)+
-#   geom_point(aes(x=year, y=avg_obs), shape=1, size=3)+
-#   facet_grid(species~., scales="free")+
-#   scale_fill_manual(values=c(myCols[1], "darkorange"), labels=c("Year effects", "No year effects"))+
-#   scale_color_manual(values=c(myCols[1], "darkorange"))+
-#   guides(color=FALSE)
-# nSims <- length(unique(resD$sim))
-# diffD <- ddply(resDall, c("year", "species", "simtype"), summarise,
-#                avgDiff = median(resids, na.rm = TRUE),
-#                upDiff75 = quantile(resids, probs = 0.8775, names=FALSE, na.rm = TRUE),
-#                dnDiff75 = quantile(resids, probs = 0.1225, names=FALSE, na.rm = TRUE),
-#                dnDiff95 = quantile(resids, probs = 0.025, names=FALSE, na.rm = TRUE),
-#                upDiff95 = quantile(resids, probs = 0.975, names=FALSE, na.rm = TRUE))
-
-# ggplot(diffD, aes(color=simtype))+
-#   geom_hline(aes(yintercept=0), linetype=2, color="grey50")+
-#   geom_point(aes(x=as.character(year), y=avgDiff), size=3, position=position_dodge(width=0.5))+
-#   geom_errorbar(aes(x=as.character(year), ymax=upDiff95, ymin=dnDiff95), width=0.00001, position=position_dodge(width=0.5))+
-#   geom_errorbar(aes(x=as.character(year), ymax=upDiff75, ymin=dnDiff75), size=1, width=0.00001, position=position_dodge(width=0.5))+
-#   facet_grid(species~., scales = "free_y")+
-#   xlab("Year")+
-#   ylab("Residuals (% cover difference)")+
-#   theme_bw()+
-#   scale_color_manual(values=myCols)+
-#   theme(axis.text.x = element_text(angle = 45, hjust = 1))
-# 
-# ggplot(resDall)+
-#   geom_hline(aes(yintercept=0), linetype=2)+
-# #   geom_line(data=diffD, aes(x=year, y=avgDiff, group=species), color="purple", alpha=0.5)+
-# #   stat_smooth(data=diffD, aes(x=year, y=avgDiff, group=species), color="lightblue", se = FALSE, size=1)+
-#   geom_boxplot(aes(x=year, y=resids, fill=simtype), outlier.shape = 1, width=0.3, outlier.size = 0.5)+
-#   facet_grid(species~., scales = "free_y")+
-#   xlab("Year")+
-#   ylab("Residuals (% cover difference)")+
-#   theme_bw()+
-#   theme(axis.text.x = element_text(angle = 45, hjust = 1))
-# myCols <- c("#237DA4", "#7ECAD0")
-# p1 <- ggplot(subset(resDall, species=="BOGR"))+
-#   geom_hline(aes(yintercept=0), linetype=2)+
-#   geom_boxplot(aes(x=as.character(year), y=resids, fill=simtype), outlier.shape = 1, outlier.size = 0, color="grey25")+
-#   xlab("Year")+
-#   ylab("Residuals (% cover difference)")+
-#   theme_bw()+
-#   theme(axis.text.x = element_text(angle = 45, hjust = 1))+
-#   scale_fill_manual(values=myCols)+
-#   coord_cartesian(ylim = c(-50, 50))+
-#   ggtitle(expression(italic("B. gracilis")))+
-#   guides(fill=FALSE)
-# p2 <- ggplot(subset(resDall, species=="HECO"))+
-#   geom_hline(aes(yintercept=0), linetype=2)+
-#   geom_boxplot(aes(x=as.character(year), y=resids, fill=simtype), outlier.shape = 1, outlier.size = 0, color="grey25")+
-#   xlab("Year")+
-#   ylab("Residuals (% cover difference)")+
-#   theme_bw()+
-#   theme(axis.text.x = element_text(angle = 45, hjust = 1))+
-#   scale_fill_manual(values=myCols, labels=c("Year effects", "No year effects"))+
-#   ggtitle(expression(italic("H. comita")))+
-#   coord_cartesian(ylim = c(-10, 5))+
-#   guides(fill=guide_legend(title=""))+
-#   theme(legend.position=c(0.5,0.2))
-# p3 <- ggplot(subset(resDall, species=="PASM"))+
-#   geom_hline(aes(yintercept=0), linetype=2)+
-#   geom_boxplot(aes(x=as.character(year), y=resids, fill=simtype), outlier.shape = 1, outlier.size = 0, color="grey25")+
-#   xlab("Year")+
-#   ylab("Residuals (% cover difference)")+
-#   theme_bw()+
-#   theme(axis.text.x = element_text(angle = 45, hjust = 1))+
-#   scale_fill_manual(values=myCols)+
-#   ggtitle(expression(italic("P. smithii")))+
-#   guides(fill=FALSE)+
-#   coord_cartesian(ylim = c(-1.5, 1.5))
-# p4 <- ggplot(subset(resDall, species=="POSE"))+
-#   geom_hline(aes(yintercept=0), linetype=2)+
-#   geom_boxplot(aes(x=as.character(year), y=resids, fill=simtype), outlier.shape = 1, outlier.size = 0, color="grey25")+
-#   xlab("Year")+
-#   ylab("Residuals (% cover difference)")+
-#   theme_bw()+
-#   theme(axis.text.x = element_text(angle = 45, hjust = 1))+
-#   scale_fill_manual(values=myCols)+
-#   ggtitle(expression(italic("Poa secunda")))+
-#   guides(fill=FALSE)+
-#   coord_cartesian(ylim = c(-10, 15))
-# pFinal <- grid.arrange(p1,p2,p3,p4, ncol=2, nrow=2)
-# pFinal
-
+```{r ipm_climeffs, cache=FALSE, results='hide', include=FALSE}
+# source("../montana/ipm/vitalRateRegs/climate_effects_plots.R")
+# ![Standardized regression coefficients for climate effects in the vital rate models for the IPM.
+Shown are the median (circles), 75% confidence intervals (thick lines), and 95% confidence intervals (thin lines).
+Left panels (black) are for the growth regressions; middle panels (blue) are for the survival regressions; right panels (orange) are for the recruitment regressions.](components/figure/ipm_climeffs.pdf)
 ```
 
-```{r figure_2, dependson="plot-options", fig.cap="One step forecasts from IPM version.",fig.width=8, fig.height=8, cache=FALSE, results='hide'}
-# resD <- readRDS("../montana/quadBM/simulations/oneStepResids_YearEffects.rds")
-# resDny <- readRDS("../montana/quadBM/simulations/oneStepResids_NoYearEffects.rds")
-# resD$simtype <- rep("full", nrow(resD))
-# resDny$simtype <- rep("noyear", nrow(resDny))
-# resDall <- rbind(resD, resDny)
-# noZ <- which(resDall$lagCover!=0)
-# resDall <- resDall[noZ,]
-# 
-# myCols <- c("#237DA4", "#7ECAD0")
-# # ggplot(resDall, aes(x=resids, fill=simtype))+
-# #   geom_histogram(alpha=0.5)+
-# #   facet_grid(.~species, scales = "free_x")+
-# #   xlab("Year")+
-# #   ylab("Residuals (% cover difference)")+
-# #   theme_bw()+
-# #   scale_fill_manual(values=myCols)
-# 
-# resAgg <- ddply(resDall, .(year, species, simtype), summarise,
-#                 avg_prediction = mean(cover)*100,
-#                 up_prediction = quantile(cover, probs = 0.875)*100,
-#                 down_prediction = quantile(cover, probs = 0.125)*100,
-#                 avg_obs = mean(propCover)*100)
-# resAgg$model <- rep("QBM", nrow(resAgg))
-# 
-# 
-# 
-# ipm_onestep <- readRDS("../montana/ipm/simulations/one_step_results/ipm_one-step_forecasts_combined.rds")
-# ipm_onestep$t1 <- ipm_onestep$t1+1900
-# 
-# resAggipm <- ddply(ipm_onestep, .(t1, species, year), summarise,
-#                 avg_prediction = mean(cover.t1)*100,
-#                 up_prediction = quantile(cover.t1, probs = 0.875)*100,
-#                 down_prediction = quantile(cover.t1, probs = 0.125)*100,
-#                 avg_obs = mean(obs.cover.t1)*100)
-# colnames(resAggipm)[1:3] <- c("year", "species", "simtype")
-# resAggipm$simtype[which(resAggipm$simtype=="ayear effect")] <- "full"
-# resAggipm$simtype[which(resAggipm$simtype!="full")] <- "noyear"
-# resAggipm$model <- rep("IPM", nrow(resAggipm))
-# 
-# resAgg_both <- rbind(resAgg, resAggipm)
-# 
-# ggplot(subset(resAgg_both))+
-#   geom_ribbon(aes(x=year, ymin=down_prediction, ymax=up_prediction, fill=simtype), alpha=0.3)+
-#   geom_line(aes(x=year, y=avg_prediction, color=simtype), size=1.5)+
-#   geom_point(aes(x=year, y=avg_obs), shape=1, size=3)+
-#   facet_grid(species~model, scales="free")+
-#   scale_fill_manual(values=c(myCols[1], "darkorange"), labels=c("Year effects", "No year effects"))+
-#   scale_color_manual(values=c(myCols[1], "darkorange"))+
-#   guides(color=FALSE)
-# 
-# ggplot(subset(resAgg_both))+
-#   geom_ribbon(aes(x=year, ymin=down_prediction, ymax=up_prediction, fill=model), alpha=0.3)+
-#   geom_line(aes(x=year, y=avg_prediction, color=model), size=1.5)+
-#   geom_point(aes(x=year, y=avg_obs), shape=1, size=3)+
-#   facet_grid(species~simtype, scales="free")+
-#   scale_fill_manual(values=c(myCols[1], "darkorange"))+
-#   scale_color_manual(values=c(myCols[1], "darkorange"))+
-#   guides(color=FALSE)
-
-# p1 <- ggplot(subset(ipm_onestep, species=="BOGR"))+
-#   geom_hline(aes(yintercept=0), linetype=2)+
-#   geom_boxplot(aes(x=as.character(t1),y=resid,fill=year), outlier.shape = 1, outlier.size = 0, color="grey25")+
-#   xlab("Year")+
-#   ylab("Residuals (% cover difference)")+
-#   theme_bw()+
-#   theme(axis.text.x = element_text(angle = 45, hjust = 1))+
-#   scale_fill_manual(values=myCols)+
-#   coord_cartesian(ylim = c(-50, 50))+
-#   ggtitle(expression(italic("B. gracilis")))+
-#   guides(fill=FALSE)
-# p2 <- ggplot(subset(ipm_onestep, species=="HECO"))+
-#   geom_hline(aes(yintercept=0), linetype=2)+
-#   geom_boxplot(aes(x=as.character(t1),y=resid,fill=year), outlier.shape = 1, outlier.size = 0, color="grey25")+
-#   xlab("Year")+
-#   ylab("Residuals (% cover difference)")+
-#   theme_bw()+
-#   theme(axis.text.x = element_text(angle = 45, hjust = 1))+
-#   scale_fill_manual(values=myCols, labels=c("Year effects", "No year effects"))+
-#   ggtitle(expression(italic("H. comita")))+
-#   coord_cartesian(ylim = c(-10, 5))+
-#   guides(fill=guide_legend(title=""))+
-#   theme(legend.position=c(0.5,0.2))
-# p3 <- ggplot(subset(ipm_onestep, species=="PASM"))+
-#   geom_hline(aes(yintercept=0), linetype=2)+
-#   geom_boxplot(aes(x=as.character(t1),y=resid,fill=year), outlier.shape = 1, outlier.size = 0, color="grey25")+
-#   xlab("Year")+
-#   ylab("Residuals (% cover difference)")+
-#   theme_bw()+
-#   theme(axis.text.x = element_text(angle = 45, hjust = 1))+
-#   scale_fill_manual(values=myCols)+
-#   ggtitle(expression(italic("P. smithii")))+
-#   guides(fill=FALSE)+
-#   coord_cartesian(ylim = c(-1.5, 1.5))
-# p4 <- ggplot(subset(ipm_onestep, species=="POSE"))+
-#   geom_hline(aes(yintercept=0), linetype=2)+
-#   geom_boxplot(aes(x=as.character(t1),y=resid,fill=year), outlier.shape = 1, outlier.size = 0, color="grey25")+
-#   xlab("Year")+
-#   ylab("Residuals (% cover difference)")+
-#   theme_bw()+
-#   theme(axis.text.x = element_text(angle = 45, hjust = 1))+
-#   scale_fill_manual(values=myCols)+
-#   ggtitle(expression(italic("Poa secunda")))+
-#   guides(fill=FALSE)+
-#   coord_cartesian(ylim = c(-10, 15))
-# pFinal <- grid.arrange(p1,p2,p3,p4, ncol=2, nrow=2)
-# pFinal
+```{r quad_climeffs, cache=FALSE, results='hide', include=FALSE}
+# source("../montana/quadBM/vitalRateRegressions/truncNormModel/climate_effects_plot.R")
+# ![Standardized regression coefficients for climate effects in the vital rate models for the QBM.
+Shown are the median (circles), 75% confidence intervals (thick lines), and 95% confidence intervals (thin lines).](components/figure/qbm_climeffs.pdf)
 ```
 
-```{r figure_3, dependson="plot-options", fig.cap="One step forecasts from IPM version.",fig.width=8, fig.height=8, cache=FALSE, results='hide', eval=FALSE}
-# ipm_onestep2 <- subset(ipm_onestep, year=="ayear effect")
-# resDall2 <- subset(resDall, simtype=="full")
-# ipm_qbm_onestep <- data.frame(year = c(ipm_onestep2$t1, resDall2$year),
-#                               resids = c(ipm_onestep2$resid, resDall2$resids),
-#                               model = c(rep("IPM", nrow(ipm_onestep2)),
-#                                         rep("QBM", nrow(resDall2))),
-#                               species = c(ipm_onestep2$species, resDall2$species))
-# g1 <- ggplot(subset(ipm_qbm_onestep, species=="BOGR"))+
-#   geom_hline(aes(yintercept=0), linetype=2)+
-#   geom_boxplot(aes(x=as.character(year),y=resids,fill=model), outlier.shape = 1, outlier.size = 0, color="grey25")+
-#   xlab("Year")+
-#   ylab("Residuals (% cover difference)")+
-#   theme_bw()+
-#   theme(axis.text.x = element_text(angle = 45, hjust = 1))+
-#   scale_fill_manual(values=myCols)+
-#   coord_cartesian(ylim = c(-50, 50))+
-#   ggtitle(expression(italic("B. gracilis")))+
-#   guides(fill=FALSE)
-# g2 <- ggplot(subset(ipm_qbm_onestep, species=="HECO"))+
-#   geom_hline(aes(yintercept=0), linetype=2)+
-#   geom_boxplot(aes(x=as.character(year),y=resids,fill=model), outlier.shape = 1, outlier.size = 0, color="grey25")+
-#   xlab("Year")+
-#   ylab("Residuals (% cover difference)")+
-#   theme_bw()+
-#   theme(axis.text.x = element_text(angle = 45, hjust = 1))+
-#   scale_fill_manual(values=myCols)+
-#   coord_cartesian(ylim = c(-10, 5))+
-#   ggtitle(expression(italic("H. comata")))+
-#   guides(fill=guide_legend(title=""))+
-#   theme(legend.position=c(0.5,0.2))
-# g3 <- ggplot(subset(ipm_qbm_onestep, species=="PASM"))+
-#   geom_hline(aes(yintercept=0), linetype=2)+
-#   geom_boxplot(aes(x=as.character(year),y=resids,fill=model), outlier.shape = 1, outlier.size = 0, color="grey25")+
-#   xlab("Year")+
-#   ylab("Residuals (% cover difference)")+
-#   theme_bw()+
-#   theme(axis.text.x = element_text(angle = 45, hjust = 1))+
-#   scale_fill_manual(values=myCols)+
-#   ggtitle(expression(italic("P. smithii")))+
-#   guides(fill=FALSE)+
-#   coord_cartesian(ylim = c(-1.5, 1.5))
-# g4 <- ggplot(subset(ipm_qbm_onestep, species=="POSE"))+
-#   geom_hline(aes(yintercept=0), linetype=2)+
-#   geom_boxplot(aes(x=as.character(year),y=resids,fill=model), outlier.shape = 1, outlier.size = 0, color="grey25")+
-#   xlab("Year")+
-#   ylab("Residuals (% cover difference)")+
-#   theme_bw()+
-#   theme(axis.text.x = element_text(angle = 45, hjust = 1))+
-#   scale_fill_manual(values=myCols)+
-#   ggtitle(expression(italic("Poa secunda")))+
-#   coord_cartesian(ylim = c(-10, 15))+
-#   guides(fill=FALSE)
-# pFinal <- grid.arrange(g1,g2,g3,g4, ncol=2, nrow=2)
-# pFinal
+```{r figure_1, dependson="plot-options", fig.cap="Sensitivity of equilibrium cover to removal of climate effects from each vital rate regression.", fig.width=4.5, fig.height=6, cache=FALSE, results='hide'}
+setwd("../montana/ipm/simulations/sensitivity_results/")
+files <- list.files()
+cover_files <- files[grep("cover", files)]
+spp_id <- substr(cover_files, 1, 4)
+clim_id <- rep(c("allClim", "noPpt", "noTemp",
+                 "noPpt", "noTemp",
+                 "noPpt", "noTemp"), 4)
+vital_id <- rep(c("All", "Growth", "Growth",
+                  "Survival", "Survival",
+                  "Recruitment", "Recruitment"), 4)
+num_files <- length(cover_files)
+all_sims <- data.frame(time=NA, cover=NA, species=NA, sim=NA, vital=NA)
+for(i in 1:num_files){
+  tmp <- read.csv(cover_files[i])
+  tmp$species <- rep(spp_id[i], nrow(tmp))
+  tmp$sim <- rep(clim_id[i], nrow(tmp))
+  tmp$vital <- rep(vital_id[i], nrow(tmp))
+  all_sims <- rbind(all_sims, tmp)
+}
+all_sims <- all_sims[2:nrow(all_sims),]
+all_clim_sims <- subset(all_sims, vital=="All")
+
+species <- unique(all_sims$species)
+vital_adds <- c("Growth", "Survival", "Recruitment") 
+out_controls <- data.frame(time=NA, cover=NA, species=NA, sim=NA, vital=NA)
+for(spp in species){
+  tmp <- subset(all_clim_sims, species==spp)
+  for(vital in vital_adds){
+    tmp$vital <- vital
+    out_controls <- rbind(out_controls, tmp)
+  }
+}
+out_controls <- out_controls[2:nrow(out_controls),]
+
+all_noclim_sims <- subset(all_sims, vital!="All")
+final <- rbind(out_controls, all_noclim_sims)
+
+equilibrium_cover <- ddply(final, .(species, sim, vital), summarise,
+                           eq_cover = median(cover))
+
+mean_cover <- ddply(subset(equilibrium_cover, sim=="allClim"), .(species), summarise,
+                    value = mean(eq_cover))
+
+myCols2 <- c("#277BA8", "#7ABBBD", "#AED77A")
+ggplot(equilibrium_cover, aes(x=sim, y=eq_cover*100, 
+                              linetype=vital, group=vital, shape=vital))+
+  geom_hline(data=mean_cover, aes(yintercept=value*100), linetype=2, color="grey45")+
+  geom_line()+
+  geom_point(size=4)+
+  facet_grid(species~., scales = "free")+
+  scale_linetype_manual(values=c(1,2,3), name="Vital Rate")+
+  scale_shape_manual(values=c(19,17,15), name="Vital Rate")+
+  ylab("Equilibrium Cover (%)")+
+  xlab("Simulation")+
+  theme_bw()
+setwd("~/Repos/MicroMesoForecast/manuscript")
 ```
 
-```{r ipm_climeffs, cache=FALSE, results='hide', include=FALSE}
-source("../montana/ipm/vitalRateRegs/climate_effects_plots.R")
-```
-
-![Standardized regression coefficients for climate effects in the vital rate models for the IPM.
-Shown are the median (circles), 75% confidence intervals (thick lines), and 95% confidence intervals (thin lines).
-Left panels (black) are for the growth regressions; middle panels (blue) are for the survival regressions; right panels (orange) are for the recruitment regressions.](components/figure/ipm_climeffs.pdf)
-
-```{r quad_climeffs, cache=FALSE, results='hide', include=FALSE}
-source("../montana/quadBM/vitalRateRegressions/truncNormModel/climate_effects_plot.R")
-```
-
-![Standardized regression coefficients for climate effects in the vital rate models for the QBM.
-Shown are the median (circles), 75% confidence intervals (thick lines), and 95% confidence intervals (thin lines).](components/figure/qbm_climeffs.pdf)
-
-```{r figure_4, dependson="plot-options", fig.cap="Proportional change in species' mean cover caused by a 1% increase in observed precipitation, temperature, or both as predicted by the individual-based IPM and the aggregate-based QBM.
+```{r figure_2, dependson="plot-options", fig.cap="Proportional change in species' mean cover caused by a 1% increase in observed precipitation, temperature, or both as predicted by the individual-based IPM and the aggregate-based QBM.
 Note that the change for POSE due to a precipitation increase predicted by the QBM is almost zero and so does not show up in the figure.

### 2015-05-28 manuscript/manuscript.tex
updates ms and adds climate sensitivity figure script

@@
 
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\itemsep1pt\parskip0pt\parsep0pt
-\item
-  Climate detection
-\item
-  Limitations:
-\end{enumerate}
-
-\begin{itemize}
-\itemsep1pt\parskip0pt\parsep0pt
-\item
-  Sample size at quadrat level
-\item
-  are these species inherently sensitive to climate (compare to Chu and
-  Adler in review -- say why we chose Montana site -- most sensitive to
-  climate variables)
-\item
-  we know recruitment is important, and this is something that may be
-  easily missed at quad level
-\end{itemize}
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\setcounter{enumi}{2}
-\itemsep1pt\parskip0pt\parsep0pt
-\item
-  Implications
-\end{enumerate}
-
 \pagebreak{}
@@
 \centering
-\includegraphics{components/figure/ipm_climeffs.pdf}
-\caption{Standardized regression coefficients for climate effects in the
-vital rate models for the IPM. Shown are the median (circles), 75\%
-confidence intervals (thick lines), and 95\% confidence intervals (thin
-lines). Left panels (black) are for the growth regressions; middle
-panels (blue) are for the survival regressions; right panels (orange)
-are for the recruitment regressions.}
+\includegraphics{components/figure/manuscript-figure_1.pdf}
+\caption{Sensitivity of equilibrium cover to removal of climate effects
+from each vital rate regression.}
 \end{figure}
@@
 \centering
-\includegraphics{components/figure/qbm_climeffs.pdf}
-\caption{Standardized regression coefficients for climate effects in the
-vital rate models for the QBM. Shown are the median (circles), 75\%
-confidence intervals (thick lines), and 95\% confidence intervals (thin
-lines).}
-\end{figure}
-
-\begin{figure}[htbp]
-\centering
-\includegraphics{components/figure/manuscript-figure_4.pdf}
+\includegraphics{components/figure/manuscript-figure_2.pdf}
 \caption{Proportional change in species' mean cover caused by a 1\%

### 2015-05-28 montana/ipm/simulations/sensitivity_results/clim_sensitivity_fig.R

@@ -30,20 +30,38 @@ for(i in 1:num_files){
   all_sims <- rbind(all_sims, tmp)
 }
 all_sims <- all_sims[2:nrow(all_sims),]
+all_clim_sims <- subset(all_sims, vital=="All")
 
-ggplot(all_sims)+
-  geom_boxplot(aes(x=species, y=cover, fill=sim))
-#   coord_cartesian(ylim = c(0,100))
+species <- unique(all_sims$species)
+vital_adds <- c("Growth", "Survival", "Recruitment") 
+out_controls <- data.frame(time=NA, cover=NA, species=NA, sim=NA, vital=NA)
+for(spp in species){
+  tmp <- subset(all_clim_sims, species==spp)
+  for(vital in vital_adds){
+    tmp$vital <- vital
+    out_controls <- rbind(out_controls, tmp)
+  }
+}
+out_controls <- out_controls[2:nrow(out_controls),]
 
-all_means <- ddply(all_sims, .(species, sim), summarise,
-                   avg_cover = median(cover))
+all_noclim_sims <- subset(all_sims, vital!="All")
+final <- rbind(out_controls, all_noclim_sims)
 
-myCols2 <- c("grey45", "#277BA8", "#7ABBBD", "#AED77A")
-ggplot(diff_df, aes(x=sim, y=value, fill=sim))+
-  geom_bar(stat="identity", position="dodge", color="white")+
-  geom_hline(aes(yintercept=0))+
-  scale_fill_manual(values = myCols2[2:4])+
-  xlab("Species")+
-  ylab("Cover change (%)")+
+equilibrium_cover <- ddply(final, .(species, sim, vital), summarise,
+                           eq_cover = median(cover))
+
+mean_cover <- ddply(subset(equilibrium_cover, sim=="allClim"), .(species), summarise,
+                    value = mean(eq_cover))
+
+myCols2 <- c("#277BA8", "#7ABBBD", "#AED77A")
+ggplot(equilibrium_cover, aes(x=sim, y=eq_cover*100, 
+                              linetype=vital, group=vital, shape=vital))+
+  geom_hline(data=mean_cover, aes(yintercept=value*100), linetype=2, color="grey45")+
+  geom_line()+
+  geom_point(size=4)+
+  facet_grid(species~., scales = "free")+
+  scale_linetype_manual(values=c(1,2,3), name="Vital Rate")+
[…]

### 2015-05-28 manuscript/manuscript.Rmd
updates ms with new climate sensitivity figure

@@
 spp_id <- substr(cover_files, 1, 4)
-clim_id <- rep(c("allClim", "noPpt", "noTemp",
-                 "noPpt", "noTemp",
-                 "noPpt", "noTemp"), 4)
+clim_id <- rep(c("noClim", "Ppt", "Temp",
+                 "Ppt", "Temp",
+                 "Ppt", "Temp"), 4)
 vital_id <- rep(c("All", "Growth", "Growth",
@@
 
-mean_cover <- ddply(subset(equilibrium_cover, sim=="allClim"), .(species), summarise,
+mean_cover <- ddply(subset(equilibrium_cover, sim=="noClim"), .(species), summarise,
                     value = mean(eq_cover))

### 2015-05-28 manuscript/manuscript.Rmd
updates Table 1 with 90th percentile distance

@@
                 mean_obs = mean(obs.cover.t1),
-                resid = mean(resid))
+                resid = mean(resid),
+                quant_dist = quantile(cover.t1*100, probs = 0.95) - quantile(cover.t1, probs = 0.05))
 ipmstats <- ddply(tmpipm, .(species), summarise,
@@
                 var = mean(var_pred),
-                mean_cover = mean(mean_obs*100))
+                mean_cover = mean(mean_obs*100),
+                quant_dist = mean(quant_dist))
 ipmstats$model <- "IPM"
@@
                 mean_obs = mean(obscov),
-                resid = mean(resid))
+                resid = mean(resid),
+                quant_dist = quantile(predcov*100, probs = 0.95) - quantile(predcov, probs = 0.05))
 qbmstats <- ddply(tmpqbm, .(species), summarise,
@@
                 var = mean(var_pred),
-                mean_cover = mean(mean_obs*100))
+                mean_cover = mean(mean_obs*100),
+                quant_dist = mean(quant_dist))
 qbmstats$model <- "QBM"
@@
 stats_table <- stats_table[,c("species", "model", "mae",
-                              "rho","mean_cover")]
+                              "quant_dist","mean_cover")]
 # comment          <- list()
@@
 
-colnames(stats_table) <- c("Species", "Model", "MAE", "Accuracy", "Mean Obs. Cover")
+colnames(stats_table) <- c("Species", "Model", "MAE", "90% Distance", "Mean Obs. Cover")
 print.xtable(xtable(stats_table, caption = "Mean absolute error (MAE) and accuracy (Pearson's $\\rho$) for one-step-ahead forecast from both model types.
-Forecasts were made without random year effects; only climate covariates could explain year-to-year variation."),type="latex", comment=FALSE,
+Forecasts were made without random year effects; only climate covariates could explain year-to-year variation.
+90 Distance refers to the average distance between the upper and lower 90 percentiles of the 100 predicted values for each quadrat-year combination."),type="latex", comment=FALSE,
              include.rownames=FALSE, caption.placement="top")

### 2015-05-28 manuscript/manuscript.tex
updates Table 1 with 90th percentile distance

@@
 \caption{Mean absolute error (MAE) and accuracy (Pearson's $\rho$) for one-step-ahead forecast from both model types.
-Forecasts were made without random year effects; only climate covariates could explain year-to-year variation.}
+Forecasts were made without random year effects; only climate covariates could explain year-to-year variation.
+90 Distance refers to the average distance between the upper and lower 90 percentiles of the 100 predicted values for each quadrat-year combination.}
 \begin{tabular}{llrrr}
   \hline
-Species & Model & MAE & Accuracy & Mean Obs. Cover \\ 
+Species & Model & MAE & 90\% Distance & Mean Obs. Cover \\ 
   \hline
-BOGR & IPM & 4.77 & 0.90 & 9.16 \\ 
-  BOGR & QBM & 5.30 & 0.88 & 9.22 \\ 
-  HECO & IPM & 0.60 & 0.80 & 1.22 \\ 
-  HECO & QBM & 2.03 & 0.84 & 1.25 \\ 
-  PASM & IPM & 0.18 & 0.65 & 0.40 \\ 
-  PASM & QBM & 0.17 & 0.87 & 0.40 \\ 
-  POSE & IPM & 0.78 & 0.61 & 1.23 \\ 
-  POSE & QBM & 1.30 & 0.57 & 1.25 \\ 
+BOGR & IPM & 4.77 & 29.33 & 9.16 \\ 
+  BOGR & QBM & 5.30 & 44.63 & 9.22 \\ 
+  HECO & IPM & 0.60 & 2.82 & 1.22 \\ 
+  HECO & QBM & 2.03 & 12.05 & 1.25 \\ 
+  PASM & IPM & 0.18 & 0.49 & 0.40 \\ 
+  PASM & QBM & 0.17 & 1.34 & 0.40 \\ 
+  POSE & IPM & 0.78 & 2.16 & 1.23 \\ 
+  POSE & QBM & 1.30 & 7.04 & 1.25 \\ 
    \hline

### 2015-05-28 manuscript/manuscript.Rmd
updates MCMC setion with STAN talk

@@
 Our Bayesian approach to fitting the vital rate models required choosing appropriate priors for unknown parameters and deciding which, if any, of those priors should be hierarchical.
-We decided to fit models where all terms except climate covariates were fit by species, while the climate covariates were fit hierarchically with species-specific coefficients drawn from a shared `global' coefficient distribution.
-We did so for two reasons: (1) the four focal species are all perennial grasses that we expect to respond similarly to climate covariates, and (2) convergence of climate effects at the quadrat level was much easier to achieve when we modeled these terms hierarchically, allowing them to "share" statistical strength via partial pooling [@Gelman2007].
-So, climate effects were modeled as:
-
-\begin{align}
-\theta_{jk} &\sim \text{Normal}(\bar{\theta_{k}}, \sigma_{k})
-\end{align}
-
-where $\bar{\theta_{k}}$ is the interspecific effect of the $k$th climate covariate.
-
-We used uninformative priors for all unknown parameters, specifically:
-
-\begin{align}
-\boldsymbol{\gamma, \beta, \bar{\theta}} &\sim \text{Normal}(0, 1e^{-6}) \\
-\boldsymbol{\phi} &\sim \text{Normal}(0, \boldsymbol{\sigma_{\phi}}) \\
-\boldsymbol{\sigma_{\phi}} &\sim e^{(\text{Gamma}(2, 0.5))} \\
-\boldsymbol{\sigma_{\theta}, \sigma_{\gamma}, \sigma_{\beta}, \tau, \zeta} &\sim \text{Gamma}(0.001, 0.001)
-\end{align}
+We decided to fit models where all terms were fit by species.
+Within a species, we fit yearly size effects and yearly intercepts hierarchically where year-specific coefficients were drawn from global distributions representing the mean size effect and intercept.
+We used flat, uninformative priors for all unknown parameters (Appendix X).
 
 All of our analyses (model fitting and simulating) were conducted in R [@Team2013].
-We used the MCMC sampler in JAGS [@Plummer2003] to estimate the posterior distributions of model parameters and the package 'rjags' to connect R to JAGS.
-We obtained posterior distributions for all model parameters from three parallel MCMC chains run for 50,000 iterations, thinned by 50, after discarding an initial 50,000 iterations.
+We used the `No-U-Turn` MCMC sampler in STAN to estimate the posterior distributions of model parameters using the package 'rstan'.
+We obtained posterior distributions for all model parameters from three parallel MCMC chains run for 1,000 iterations after discarding an initial 1,000 iterations.
+We recignize such short MCMC chains may surprise some more familiar with other MCMC samplers.
 We assessed convergence visually and using the @Gelman1992 diagnostic in the R package `coda' [@Plummer2006].

### 2015-05-28 manuscript/manuscript.Rmd
updates ms with Stan talk

@@
 
-All of our analyses (model fitting and simulating) were conducted in R [@Team2013].
-We used the `No-U-Turn` MCMC sampler in STAN to estimate the posterior distributions of model parameters using the package 'rstan'.
+All of our analyses (model fitting and simulating) were conducted in R [@R2013].
+We used the 'No-U-Turn' MCMC sampler in Stan [@stan2014]  to estimate the posterior distributions of model parameters using the package 'rstan' [@rstan2014].
 We obtained posterior distributions for all model parameters from three parallel MCMC chains run for 1,000 iterations after discarding an initial 1,000 iterations.
-We recignize such short MCMC chains may surprise some more familiar with other MCMC samplers.
-We assessed convergence visually and using the @Gelman1992 diagnostic in the R package `coda' [@Plummer2006].
-Scale reduction factors for all parameters were less than 1.01, indicating convergence.
+We recignize such short MCMC chains may surprise those more familiar with other MCMC samplers (i.e.
+JAGS or WinBUGS), but the Stan sampler is exceptionally efficient, which reduces the number of iterations needed to achieve convergence.
+We assessed convergence visually and made sure scale reduction factors for all parameters were less than 1.01.
 For the purposes of introducing stochasticity in our population models, we saved the final 1,000 iterations from each of the three MCMC chains for all parameters to be used as randomly drawn values during population simulation.
+Importantly, this step alleviates the need to reduce model parameters since sampling from the full parameter space in the MCMC ensures that if a parameter broadly overlaps zero, on average the effect in the population models will also be near zero.
 

### 2015-05-28 manuscript/manuscript.tex
updates ms with Stan talk

@@
 those priors should be hierarchical. We decided to fit models where all
-terms except climate covariates were fit by species, while the climate
-covariates were fit hierarchically with species-specific coefficients
-drawn from a shared `global' coefficient distribution. We did so for two
-reasons: (1) the four focal species are all perennial grasses that we
-expect to respond similarly to climate covariates, and (2) convergence
-of climate effects at the quadrat level was much easier to achieve when
-we modeled these terms hierarchically, allowing them to ``share''
-statistical strength via partial pooling (Gelman and Hill 2007). So,
-climate effects were modeled as:
-
-\begin{align}
-\theta_{jk} &\sim \text{Normal}(\bar{\theta_{k}}, \sigma_{k})
-\end{align}
-
-where $\bar{\theta_{k}}$ is the interspecific effect of the $k$th
-climate covariate.
-
-We used uninformative priors for all unknown parameters, specifically:
-
-\begin{align}
-\boldsymbol{\gamma, \beta, \bar{\theta}} &\sim \text{Normal}(0, 1e^{-6}) \\
-\boldsymbol{\phi} &\sim \text{Normal}(0, \boldsymbol{\sigma_{\phi}}) \\
-\boldsymbol{\sigma_{\phi}} &\sim e^{(\text{Gamma}(2, 0.5))} \\
-\boldsymbol{\sigma_{\theta}, \sigma_{\gamma}, \sigma_{\beta}, \tau, \zeta} &\sim \text{Gamma}(0.001, 0.001)
-\end{align}
+terms were fit by species. Within a species, we fit yearly size effects
+and yearly intercepts hierarchically where year-specific coefficients
+were drawn from global distributions representing the mean size effect
+and intercept. We used flat, uninformative priors for all unknown
+parameters (Appendix X).
 
 All of our analyses (model fitting and simulating) were conducted in R
-(Team 2013). We used the MCMC sampler in JAGS (Plummer 2003) to estimate
-the posterior distributions of model parameters and the package `rjags'
-to connect R to JAGS. We obtained posterior distributions for all model
-parameters from three parallel MCMC chains run for 50,000 iterations,
-thinned by 50, after discarding an initial 50,000 iterations. We
-assessed convergence visually and using the Gelman and Rubin (1992)
-diagnostic in the R package `coda' (Plummer et al. 2006). Scale
-reduction factors for all parameters were less than 1.01, indicating
-convergence. For the purposes of introducing stochasticity in our
-population models, we saved the final 1,000 iterations from each of the
-three MCMC chains for all parameters to be used as randomly drawn values
-during population simulation.
+(R Core Development Team 2013). We used the `No-U-Turn' MCMC sampler in
+Stan (Stan Development Team 2014a) to estimate the posterior
+distributions of model parameters using the package `rstan' (Stan
+Development Team 2014b). We obtained posterior distributions for all
+model parameters from three parallel MCMC chains run for 1,000
+iterations after discarding an initial 1,000 iterations. We recignize
+such short MCMC chains may surprise those more familiar with other MCMC
+samplers (i.e.~JAGS or WinBUGS), but the Stan sampler is exceptionally
+efficient, which reduces the number of iterations needed to achieve
+convergence. We assessed convergence visually and made sure scale
+reduction factors for all parameters were less than 1.01. For the
+purposes of introducing stochasticity in our population models, we saved
+the final 1,000 iterations from each of the three MCMC chains for all
+parameters to be used as randomly drawn values during population
+simulation. Importantly, this step alleviates the need to reduce model
+parameters since sampling from the full parameter space in the MCMC
+ensures that if a parameter broadly overlaps zero, on average the effect
+in the population models will also be near zero.
 
@@
 
-Gelman, A., and J. Hill. 2007. Data analysis using regression and
-multilevel/hierarchical models. Page xxii, 625p.
-
-Gelman, A., and D. B. Rubin. 1992. Inference from Iterative Simulation
-Using Multiple Sequences.
-
 Piantadosi, S., D. P. Byar, and S. B. Green. 1988. The Ecological
 Fallacy. American Journal of Epidemiology 127:893--904.
-
-Plummer, M. 2003. JAGS: A Program for Analysis of Bayesian Graphical
-Models Using Gibbs Sampling. Pages 20--22 \emph{in} Proceedings of the
-3rd international workshop on distributed statistical computing (dSC
-2003). march.
-
-Plummer, M., N. Best, K. Cowles, and K. Vines. 2006. CODA: Convergence
-Diagnosis and Output Analysis for MCMC. R News 6:7--11.
 
@@
 Evolution 2:289--302.
+
+R Core Development Team. 2013. R: A language and environment for
+statistical computing.
 
@@
 repository for plant demography. Journal of Ecology 103:202--218.
+
+Stan Development Team. 2014a. Stan: A C++ Library for Probability and
+Sampling, Version 2.5.0.
+
+Stan Development Team. 2014b. Rstan: the R interface to Stan, Version
+2.5.0.
 

### 2015-06-05 manuscript/manuscript.tex
updates ms with climate sensitivity figure

@@
 covariates could affect expect plant cover. The models had similar
-accuracy ($\rho$, correlation between observations and mean prediction
+accuracy (\(\rho\), correlation between observations and mean prediction
 from 100 forecasts) across all species (Table 1). However, the IPM had
@@
 
-The IPM MAE is significantly lower for HECO ($p = 3.3 \times 10^{-10}$)
-and POSE ($p = 0.00013$). MAEs are statisticially similar between models
-for BOGR and PASM.
+The IPM MAE is significantly lower for HECO
+(\(p = 3.3 \times 10^{-10}\)) and POSE (\(p = 0.00013\)). MAEs are
+statisticially similar between models for BOGR and PASM.
 

### 2015-06-06 manuscript/manuscript.Rmd
updates ms fig 2 to show differences from mean cover, matches better with sensitivity analysis

@@
 Note that the change for POSE due to a precipitation increase predicted by the QBM is almost zero and so does not show up in the figure.
-Labels on x-axis refer to: 'ppt' = 1% increase in mean precipitation; 'temp' = 1% increase in mean temperature; 'ppt + temp' = 1% increase in mean precipitation and mean temperature.", fig.width=6.5, fig.height=6, cache=FALSE, results='hide'}
+Labels on x-axis refer to: 'ppt' = 1% increase in mean precipitation; 'temp' = 1% increase in mean temperature; 'ppt + temp' = 1% increase in mean precipitation and mean temperature.", fig.width=8.5, fig.height=3, cache=FALSE, results='hide'}
 qbm_diffs <- readRDS("../montana/quadBM/simulations/results/qbm_climatesims_percdiffs.rds")
@@
   ylab("Proportional change in percent cover")+
-  facet_wrap("variable", scales = "free")+
+  facet_grid(.~variable)+
   scale_x_discrete(labels=c("ppt", "temp", "ppt + tmp"))+
   theme_bw()
-#   theme(axis.text.x = element_text(angle = 45, hjust = 1))
+```
 
-# climSimsQBMavg <- ddply(climSimsQBM, .(sim, species), summarise,
-#                         avgCover = median(cover))
-# climSimQBMcast <- dcast(climSimsQBMavg, formula = species~sim)
-# climSimQBMcast$diffP <- with(climSimQBMcast, (Precipitation*100)-(Observed*100)) 
-# climSimQBMcast$diffT <- with(climSimQBMcast, (Temperature*100)-(Observed*100))
-# climSimQBMcast$diffB <- with(climSimQBMcast, (Both*100)-(Observed*100))
-# climSimQBMplot <- melt(climSimQBMcast)
-# climSimQBMplot <- subset(climSimQBMplot, species!="POSE")
-# 
-# ggplot(data=subset(climSimQBMplot, variable=="diffP"|variable=="diffT"|variable=="diffB"))+
-#   geom_bar(aes(x=species, y=value, fill=variable), stat="identity", position="dodge", color="white")+
-#   scale_fill_manual(name="", labels=c("1% Precipitation Increase", "1% Temperature Increase", "1% Temp. & Ppt. Increase"), values=myCols2)+
-#   theme_bw()+
-#   theme(legend.position=c(0.6, 0.8))+
-#   ylab("Cover Change (%)")+
-#   xlab("Species")
-```
+\pagebreak{}
+
+.
 

### 2015-06-06 montana/ipm/simulations/results/clim_change_fig.R

@@ -30,7 +30,7 @@ ggplot(all_sims)+
 #   coord_cartesian(ylim = c(0,100))
 
 all_means <- ddply(all_sims, .(species, sim), summarise,
-                   avg_cover = median(cover))
+                   avg_cover = mean(cover))
 tmpid <- which(all_means$sim=="observed")
 diffs <- list()
 for(i in 1:length(tmpid)){

### 2015-06-06 manuscript/manuscript.Rmd
updates ms with comparison results section

@@
 x_{jq,t+1} &= \gamma^{G}_{j,t} + \phi^{G}_{jQ} + \beta^{G}_{j,t}x_{jq,t} + \theta^{S}_{jk}C_{k,t} \\
-Y^{G}_{jq,t+1} &\sim \text{LogNormal}(x_{jq,t+1}, \tau{j}) T[0,1]
+Y^{G}_{jq,t+1} &\sim \text{LogNormal}(x_{jq,t+1}, \tau{j}) \text{T}[0,1]
 \end{align}
@@
 #).
-The log normal likelihood includes a truncation ($T$[0,1]) to ensure that predicted values do not exceed 100% cover.
+The log normal likelihood includes a truncation (T[0,1]) to ensure that predicted values do not exceed 100% cover.
 
@@
 We assessed convergence visually and made sure scale reduction factors for all parameters were less than 1.01.
-For the purposes of introducing stochasticity in our population models, we saved the final 1,000 iterations from each of the three MCMC chains for all parameters to be used as randomly drawn values during population simulation.
-Importantly, this step alleviates the need to reduce model parameters since sampling from the full parameter space in the MCMC ensures that if a parameter broadly overlaps zero, on average the effect in the population models will also be near zero.
+For the purposes of including parameter uncertaintyy in our population models, we saved the final 1,000 iterations from each of the three MCMC chains for all parameters to be used as randomly drawn values during population simulation.
+This step alleviates the need to reduce model parameters by model selection since sampling from the full parameter space in the MCMC ensures that if a parameter broadly overlaps zero, on average the effect in the population models will also be near zero.
 
@@
 Thus, beyond the effect of genet size or quadrat cover, only the climate covariates could affect expect plant cover.
-The models had similar accuracy ($\rho$, correlation between observations and mean prediction from 100 forecasts) across all species (Table 1).
-However, the IPM had significantly lower overall error (MAE, mean absolute error) for two species (*H.
-comita* and *P.
-smithii*), and in no case did the QBM significantly outperform the IPM (Table 1).
+The IPM had significantly lower overall error (MAE, mean absolute error) for two species (*H.
+comata* and *P.
+smithii*; Table 1).
+In no case did the QBM significantly outperform the IPM (Table 1).
+As a metric of model uncertainty, we calculated the euclidean distance between the upper and lower 90% quantiles for each model's forecasts.
+The IPM always outperformed the QBM in this respect (Table 1), with consistently lower distances between the 90% quantiles.
+In general the IPM outperformed the QBM because it had (i) lower MAE for two species, (ii) statistically similar MAE for the other two species, and (iii) considerably more precise forecasts for all species.
+
+### Vital rate sensitivities to climate (IPM)
 
@@
 The IPM and QBM produced consistent predictions for *H.
-comita* under increased precipitation and when both precipitation and temperature were increased (Figure 3).
+comata* under increased precipitation and when both precipitation and temperature were increased (Figure 3).
 However, the IPM predicted more modest changes than the QBM, and for a temperature increase the two models differed: the IPM predicted an increase in cover while the QBM predicted the opposite.

### 2015-06-09 manuscript/manuscript.tex
updates ms fig 2 to show differences from mean cover, matches better with sensitivity analysis

--- 2015-05-28 manuscript/manuscript.tex
@@
 individual-level growth model (Eq. \#). The log normal likelihood
-includes a truncation ($T${[}0,1{]}) to ensure that predicted values do
+includes a truncation (T{[}0,1{]}) to ensure that predicted values do
 not exceed 100\% cover.
@@
 reduction factors for all parameters were less than 1.01. For the
-purposes of introducing stochasticity in our population models, we saved
-the final 1,000 iterations from each of the three MCMC chains for all
-parameters to be used as randomly drawn values during population
-simulation. Importantly, this step alleviates the need to reduce model
-parameters since sampling from the full parameter space in the MCMC
+purposes of including parameter uncertaintyy in our population models,
+we saved the final 1,000 iterations from each of the three MCMC chains
+for all parameters to be used as randomly drawn values during population
+simulation. This step alleviates the need to reduce model parameters by
+model selection since sampling from the full parameter space in the MCMC
 ensures that if a parameter broadly overlaps zero, on average the effect
@@
 Thus, beyond the effect of genet size or quadrat cover, only the climate
-covariates could affect expect plant cover. The models had similar
-accuracy ($\rho$, correlation between observations and mean prediction
-from 100 forecasts) across all species (Table 1). However, the IPM had
-significantly lower overall error (MAE, mean absolute error) for two
-species (\emph{H. comita} and \emph{P. smithii}), and in no case did the
-QBM significantly outperform the IPM (Table 1).
+covariates could affect expect plant cover. The IPM had significantly
+lower overall error (MAE, mean absolute error) for two species (\emph{H.
+comata} and \emph{P. smithii}; Table 1). In no case did the QBM
+significantly outperform the IPM (Table 1). As a metric of model
+uncertainty, we calculated the euclidean distance between the upper and
+lower 90\% quantiles for each model's forecasts. The IPM always
+outperformed the QBM in this respect (Table 1), with consistently lower
+distances between the 90\% quantiles. In general the IPM outperformed
+the QBM because it had (i) lower MAE for two species, (ii) statistically
+similar MAE for the other two species, and (iii) considerably more
+precise forecasts for all species.
 
@@
 
-The IPM and QBM produced consistent predictions for \emph{H. comita}
+The IPM and QBM produced consistent predictions for \emph{H. comata}
 under increased precipitation and when both precipitation and
@@
 
+.
+
+\pagebreak{}
+
 \subsection{References}\label{references}

### 2015-06-09 manuscript/manuscript.Rmd
updates ms with new figure 1

@@
 
-*Andrew T. Tredennick ([atredenn@gmail.com](mailto:atredenn@gmail.com)), Department of Wildland Resources and the Ecology Center, Utah State University, Logan, UT*
+*Andrew T. Tredennick[^corrauth], Department of Wildland Resources and the Ecology Center, Utah State University, Logan, UT*
 
 *Peter B. Adler, Department of Wildland Resources and the Ecology Center, Utah State University, Logan, UT*
+
+[^corrauth]: Corresponding author: [atredenn@gmail.com](mailto:atredenn@gmail.com)
 
@@
 --------
+The ability of population models to skillfully forecast future states under climate change is constrained by the limited spatial and temporal extent of demographic data.
+This data is often limited because it is difficult and costly to collect.
+An alternative is to rely on aggregate, population-level data that is easier and less costly to collect.
+Doing so requires assuming that population-level data accurately represents the aggregate response of the individuals that actually respond to weather.
+We tested this assumption using population models of four Montana grassland species fit using individual and aggregated forms of the same data.
+We fit population models with interannual variation in vital rates explained, in part, by climate covariates and then perturbed observed climate to compare model forecasts.
+Population models based on individual-level demographic data outperformed the models based on aggregate-level data in terms of accuracy and precision.
+The two model types produced inconsistent forecasts when we perturbed climate.
+Thus, it seems that, at least for the species in this location, demographic data is necessary to pick up key climate drivers of vital rates like survival that are not well resolved in aggregated data.
+On a pessimistic note, our work shows that even with detailed demographic data, model forecasts are extremely uncertain.
+It is becoming clear the long time series (over two decades) are necessary to produce meaningful forecasts.
 
-*Keywords*: forecasting, climate change, grassland, integral projection model
+*Keywords*: forecasting, climate change, grassland, integral projection model, population model
 
@@
 We then perturb the climate covariates to test the sensitivities of species to climate change.
-By doing these analyses using both individual and aggregated forms of the same data, we provide a rigorous test of our hypothesis.
-We find that...
+By doing these analyses using both individual and aggregated forms of the same data we can directly compare the two types of models.
+
+In general, we find that population models based on detailed demographic data are more accurate and precise than models based on aggregated data.
+Both types of models are able to detect climate signals, as evidenced by the sensitivity of simulated equilbrium plant cover under a perturbed climate scenario.
+But the two types of models produce inconsistent forecasts, in some cases producing completely opposing predictions.
+This leads us to conclude that, at least for these species at this location, detailed demographic data is necessary to detect the "right" climate signal.
+A worrying caveat to our work is that forecasts from both models were very uncertain.
+It seems that even 14 years worth of demographic data is not enough to produce meaningful forecasts when model uncertainty is explicitly considered.
 
@@
 Our demographic data comes from the Fort Keogh Livestock and Range Research Laboratory in eastern Montana's northern mixed prairie near Miles City, Montana, USA ($46^{\circ}$ 19' N, $105^{\circ}$ 48' W).
-The dataset is freely available on Ecological Archives [@Anderson2011], and interested readers should refer to the metadata therein for a complete description.
+The dataset is freely available on Ecological Archives[^link] [@Anderson2011] , and interested readers should refer to the metadata therein for a complete description.
 The site is about 800 m above sea level and mean annual precipitation (1878-2009) is 334 mm, with most annual precipitation falling from April through September.
 The site is grass dominated and, for the purposes of our study, we focus on the four most abundant graminoid species: \emph{Bouteloua gracilis} (BOGR), \emph{Hesperostipa comata} (HECO), \emph{Pascopyrum smithii} (PASM), and \emph{Poa secunda} (POSE).
+
+[^link]: http://esapubs.org/archive/ecol/E092/143/
 
@@
 
-In this paper, we model populations based on two levels of data: individual and quadrat.
+In this paper, we model populations based on two levels of data: individual and quadrat (Figure 1).
 The individual data is the "raw" data.
@@
 -------
+```{r fig.cap="Work flow of the data aggregation, model fitting, and population simulating.", fig.width=5, fig.height=8}
+library(png)
+library(grid)
+img <- readPNG("components/figure/micromeso_flow.png")
+grid.raster(img)
+```
 

### 2015-06-09 manuscript/manuscript.tex
updates ms with new figure 1

@@
 
-\emph{Andrew T. Tredennick
-(\href{mailto:atredenn@gmail.com}{\href{mailto:atredenn@gmail.com}{atredenn@gmail.com}}),
+\emph{Andrew T. Tredennick\footnote{Corresponding author:
+  \href{mailto:atredenn@gmail.com}{\href{mailto:atredenn@gmail.com}{atredenn@gmail.com}}},
 Department of Wildland Resources and the Ecology Center, Utah State
@@
 
+The ability of population models to skillfully forecast future states
+under climate change is constrained by the limited spatial and temporal
+extent of demographic data. This data is often limited because it is
+difficult and costly to collect. An alternative is to rely on aggregate,
+population-level data that is easier and less costly to collect. Doing
+so requires assuming that population-level data accurately represents
+the aggregate response of the individuals that actually respond to
+weather. We tested this assumption using population models of four
+Montana grassland species fit using individual and aggregated forms of
+the same data. We fit population models with interannual variation in
+vital rates explained, in part, by climate covariates and then perturbed
+observed climate to compare model forecasts. Population models based on
+individual-level demographic data outperformed the models based on
+aggregate-level data in terms of accuracy and precision. The two model
+types produced inconsistent forecasts when we perturbed climate. Thus,
+it seems that, at least for the species in this location, demographic
+data is necessary to pick up key climate drivers of vital rates like
+survival that are not well resolved in aggregated data. On a pessimistic
+note, our work shows that even with detailed demographic data, model
+forecasts are extremely uncertain. It is becoming clear the long time
+series (over two decades) are necessary to produce meaningful forecasts.
+
 \emph{Keywords}: forecasting, climate change, grassland, integral
-projection model
+projection model, population model
 
@@
 doing these analyses using both individual and aggregated forms of the
-same data, we provide a rigorous test of our hypothesis. We find
-that\ldots{}
+same data we can directly compare the two types of models.
+
+In general, we find that population models based on detailed demographic
+data are more accurate and precise than models based on aggregated data.
+Both types of models are able to detect climate signals, as evidenced by
+the sensitivity of simulated equilbrium plant cover under a perturbed
+climate scenario. But the two types of models produce inconsistent
+forecasts, in some cases producing completely opposing predictions. This
+leads us to conclude that, at least for these species at this location,
+detailed demographic data is necessary to detect the ``right'' climate
+signal. A worrying caveat to our work is that forecasts from both models
+were very uncertain. It seems that even 14 years worth of demographic
+data is not enough to produce meaningful forecasts when model
+uncertainty is explicitly considered.
 
@@
 Miles City, Montana, USA ($46^{\circ}$ 19' N, $105^{\circ}$ 48' W). The
-dataset is freely available on Ecological Archives (Anderson et al.
-2011), and interested readers should refer to the metadata therein for a
-complete description. The site is about 800 m above sea level and mean
-annual precipitation (1878-2009) is 334 mm, with most annual
-precipitation falling from April through September. The site is grass
-dominated and, for the purposes of our study, we focus on the four most
-abundant graminoid species: \emph{Bouteloua gracilis} (BOGR),
-\emph{Hesperostipa comata} (HECO), \emph{Pascopyrum smithii} (PASM), and
-\emph{Poa secunda} (POSE).
+dataset is freely available on Ecological Archives\footnote{\url{http://esapubs.org/archive/ecol/E092/143/}}
+(Anderson et al. 2011) , and interested readers should refer to the
+metadata therein for a complete description. The site is about 800 m
+above sea level and mean annual precipitation (1878-2009) is 334 mm,
+with most annual precipitation falling from April through September. The
+site is grass dominated and, for the purposes of our study, we focus on
+the four most abundant graminoid species: \emph{Bouteloua gracilis}
+(BOGR), \emph{Hesperostipa comata} (HECO), \emph{Pascopyrum smithii}
+(PASM), and \emph{Poa secunda} (POSE).
 
@@
 In this paper, we model populations based on two levels of data:
-individual and quadrat. The individual data is the ``raw'' data. For the
-quadrat level we data we simply sum individual areal cover for each
-quadrat by species. This is equivalent to a perfect census of quadrat
-percent cover, so we do not need to consider measurement error. Based on
-these two datasets we can compare population models built using
+individual and quadrat (Figure 1). The individual data is the ``raw''
+data. For the quadrat level we data we simply sum individual areal cover
+for each quadrat by species. This is equivalent to a perfect census of
+quadrat percent cover, so we do not need to consider measurement error.
+Based on these two datasets we can compare population models built using
 individual level data and aggregated quadrat level data.
@@
 
+\subsubsection{Vital rate sensitivities to climate
+(IPM)}\label{vital-rate-sensitivities-to-climate-ipm}
+
 \subsubsection{Forecasting Climate Change
@@
 \subsection{Figures}\label{figures}
+
+\begin{figure}[htbp]
+\centering
+\includegraphics{components/figure/manuscript-unnamed-chunk-1.pdf}
+\caption{Work flow of the data aggregation, model fitting, and
+population simulating.}
+\end{figure}
 

### 2015-06-09 manuscript/manuscript.Rmd
updates ms with IPM description

@@
 
-Do we need detailed demographic data to forecast population responses to climate change?
+Do we need detailed demographic data to forecast population responses to climate change? Probably.
 ========================================================================================
 
-### Andrew T. Tredennick and Peter B. Adler
+### Andrew T. Tredennick[^corrauth] and Peter B. Adler
 
-*Andrew T. Tredennick[^corrauth], Department of Wildland Resources and the Ecology Center, Utah State University, Logan, UT*
+*Andrew T. Tredennick, Department of Wildland Resources and the Ecology Center, Utah State University, Logan, UT*
 
@@
 
+All R code and data necessary to replicate our analysis is archived on GitHub as release 1.0 (http://github.com/atredennick/MicroMesoForecast/releases).
+That stable release will remain static as a record of this analysis, but subsequent versions may appear if we update this work.
+
 ### Stastical models of vital rates
@@
 We assessed convergence visually and made sure scale reduction factors for all parameters were less than 1.01.
-For the purposes of including parameter uncertaintyy in our population models, we saved the final 1,000 iterations from each of the three MCMC chains for all parameters to be used as randomly drawn values during population simulation.
+For the purposes of including parameter uncertainty in our population models, we saved the final 1,000 iterations from each of the three MCMC chains for all parameters to be used as randomly drawn values during population simulation.
 This step alleviates the need to reduce model parameters by model selection since sampling from the full parameter space in the MCMC ensures that if a parameter broadly overlaps zero, on average the effect in the population models will also be near zero.
@@
 We used an Integral Projection Model (IPM) to model populations based on individual level data and an quadrat based version of an individually-based model (Quadrat-Based Model, QBM) to model populations based on quadrat level data.
-Both models take the general form:
+We describe each in turn.
+
+#### Integral projection model
+We use an environmentally stochastic IPM [@Rees2009] that includes the random year effects and the climate covariates from the vital rate statistical models.
+But note that we can, and do for some simulations, ignore the random year effects so that only the climate effects can drive interannual variation.
+Our IPM follows the specification of @Chu2015 where the population of species _j_ is a density function $n(u_{j},t)$ giving the density of sized-_u_ genets at time _t_.
+Genet size is on the natural log scale, so that $n(u_{j},t)du$ is the number of genets whose area (on the arithmetic scale) is between $e^{u_{j}}$ and $e^{u_{j}+du}$.
+So, the density function for any size _v_ at time $t+1$ is
 
 \begin{equation}
-N_{t+1} = S \times G + R.
+n(v_{j},t+1) = \int_{L_{j}}^{U_{j}} k_{j}(v_{j},u_{j},\bar{\bold{w_{j}}}(u_{j}))n(u_{j},t)
 \end{equation}
 
-So, at each time step in a simulation, we use the survival regression to determine if each genet lives or not (if each quadrat remains occupied or not), the growth regression to determine size changes of surviving individuals (cover change of occupied quadrats), and the recruitment regression to determine the number of new recruits (if a quadrat is colonized or not).
-We first use one-step-ahead forecasts to assess each model's ability to reproduce observed cover changes.
-Then we use the models to analyze the effect of potential climate changes on population size over long time scales.
+where $k_{j}(v_{j},u_{j},\bar{\bold{w_{j}}})$ is the population kernal that describes all possible transitions from size $u$ to $v$ and $\bar{\bold{w_{j}}}$ is a vector of estimates of average crowding experienced from all other species by a genet of size $u_j$ and species $j$.
+The integral is evaluated over all possible sizes between predefined lower (_L_) and upper (_U_) size limits that extend beyond the range of observed genet sizes.
+
+The population kernal is defined as the joint contributions of survival (_S_), growth (_G_), and recruitment (_R_):
+
+\begin{equation}
+k_{j}(v_{j},u_{j},\bar{\bold{w_{j}}}) = S_j(u_j, \bar{\bold{w_{j}}}(u_{j}))G_j(v_{j},u_{j},\bar{\bold{w_{j}}}(u_{j})) + R_j(v_{j},u_{j},\bar{\bold{w_{j}}}),
+\end{equation}
+
+which, said plainly, means we are calculating growth (_G_) for individuals that survive (_S_) from time _t_ to _t+1_ and adding in newly recruited (_R_) individuals of an average sized one-year-old genet for the focal species.
+Our stastical model for recruitment (_R_, described above) returns the number of new recruit produced per quadrat.
+Following previous work [@Adler2012; @Chu2015], we assume that fecundity increases linearly with size ($R_j(v_{j},u_{j},\bar{\bold{w_{j}}}) = e^{u_j}R_j(v_{j},\bar{\bold{w_{j}}})$) to incorporate the recruitment function in the spatially-implicit IPM.
 
 We used random draws from the final 1,000 iterations from each of three MCMC chains to introduce stochasticity into our population models.
+We thinned these chains by 10 to reduce potential autocorrelation, leaving us with 300 samples for each parameter.
 At each time step, we randomly selected climate covariates from one of the 14 observed years.
@@
 Since our focus was on the contribution of climate covariates to population states, we set the random year effects and the random group effects to zero.
+
+#### Quad-based model
+Our quad-based model (QBM) perfectly mirrors its statistical description (Eq. #). We use the same approach for drawing parameter values as described for the IPM.
 

### 2015-06-09 manuscript/manuscript.tex
updates ms with IPM description

@@
 \section{Do we need detailed demographic data to forecast population
-responses to climate
-change?}\label{do-we-need-detailed-demographic-data-to-forecast-population-responses-to-climate-change}
+responses to climate change?
+Probably.}\label{do-we-need-detailed-demographic-data-to-forecast-population-responses-to-climate-change-probably.}
 
-\subsubsection{Andrew T. Tredennick and Peter B.
-Adler}\label{andrew-t.-tredennick-and-peter-b.-adler}
+\subsubsection[Andrew T. Tredennick and Peter B. Adler]{Andrew T.
+Tredennick\footnote{Corresponding author:
+  \href{mailto:atredenn@gmail.com}{\href{mailto:atredenn@gmail.com}{atredenn@gmail.com}}}
+and Peter B.
+Adler}\label{andrew-t.-tredennickcorrauth-and-peter-b.-adler}
 
-\emph{Andrew T. Tredennick\footnote{Corresponding author:
-  \href{mailto:atredenn@gmail.com}{\href{mailto:atredenn@gmail.com}{atredenn@gmail.com}}},
-Department of Wildland Resources and the Ecology Center, Utah State
-University, Logan, UT}
+\emph{Andrew T. Tredennick, Department of Wildland Resources and the
+Ecology Center, Utah State University, Logan, UT}
 
@@
 
+All R code and data necessary to replicate our analysis is archived on
+GitHub as release 1.0
+(\url{http://github.com/atredennick/MicroMesoForecast/releases}). That
+stable release will remain static as a record of this analysis, but
+subsequent versions may appear if we update this work.
+
 \subsubsection{Stastical models of vital
@@
 reduction factors for all parameters were less than 1.01. For the
-purposes of including parameter uncertaintyy in our population models,
-we saved the final 1,000 iterations from each of the three MCMC chains
-for all parameters to be used as randomly drawn values during population
+purposes of including parameter uncertainty in our population models, we
+saved the final 1,000 iterations from each of the three MCMC chains for
+all parameters to be used as randomly drawn values during population
 simulation. This step alleviates the need to reduce model parameters by
@@
 individually-based model (Quadrat-Based Model, QBM) to model populations
-based on quadrat level data. Both models take the general form:
+based on quadrat level data. We describe each in turn.
+
+\paragraph{Integral projection model}\label{integral-projection-model}
+
+We use an environmentally stochastic IPM (Rees and Ellner 2009) that
+includes the random year effects and the climate covariates from the
+vital rate statistical models. But note that we can, and do for some
+simulations, ignore the random year effects so that only the climate
+effects can drive interannual variation. Our IPM follows the
+specification of Chu and Adler (2015) where the population of species
+\emph{j} is a density function $n(u_{j},t)$ giving the density of
+sized-\emph{u} genets at time \emph{t}. Genet size is on the natural log
+scale, so that $n(u_{j},t)du$ is the number of genets whose area (on the
+arithmetic scale) is between $e^{u_{j}}$ and $e^{u_{j}+du}$. So, the
+density function for any size \emph{v} at time $t+1$ is
 
 \begin{equation}
-N_{t+1} = S \times G + R.
+n(v_{j},t+1) = \int_{L_{j}}^{U_{j}} k_{j}(v_{j},u_{j},\bar{\bold{w_{j}}}(u_{j}))n(u_{j},t)
 \end{equation}
 
-So, at each time step in a simulation, we use the survival regression to
-determine if each genet lives or not (if each quadrat remains occupied
-or not), the growth regression to determine size changes of surviving
-individuals (cover change of occupied quadrats), and the recruitment
-regression to determine the number of new recruits (if a quadrat is
-colonized or not). We first use one-step-ahead forecasts to assess each
-model's ability to reproduce observed cover changes. Then we use the
-models to analyze the effect of potential climate changes on population
-size over long time scales.
+where $k_{j}(v_{j},u_{j},\bar{\bold{w_{j}}})$ is the population kernal
+that describes all possible transitions from size $u$ to $v$ and
+$\bar{\bold{w_{j}}}$ is a vector of estimates of average crowding
+experienced from all other species by a genet of size $u_j$ and species
+$j$. The integral is evaluated over all possible sizes between
+predefined lower (\emph{L}) and upper (\emph{U}) size limits that extend
+beyond the range of observed genet sizes.
+
+The population kernal is defined as the joint contributions of survival
+(\emph{S}), growth (\emph{G}), and recruitment (\emph{R}):
+
+\begin{equation}
+k_{j}(v_{j},u_{j},\bar{\bold{w_{j}}}) = S_j(u_j, \bar{\bold{w_{j}}}(u_{j}))G_j(v_{j},u_{j},\bar{\bold{w_{j}}}(u_{j})) + R_j(v_{j},u_{j},\bar{\bold{w_{j}}}),
+\end{equation}
+
+which, said plainly, means we are calculating growth (\emph{G}) for
+individuals that survive (\emph{S}) from time \emph{t} to \emph{t+1} and
+adding in newly recruited (\emph{R}) individuals of an average sized
+one-year-old genet for the focal species. Our stastical model for
+recruitment (\emph{R}, described above) returns the number of new
+recruit produced per quadrat. Following previous work (Adler et al.
+2012, Chu and Adler 2015), we assume that fecundity increases linearly
+with size
+($R_j(v_{j},u_{j},\bar{\bold{w_{j}}}) = e^{u_j}R_j(v_{j},\bar{\bold{w_{j}}})$)
+to incorporate the recruitment function in the spatially-implicit IPM.
 
 We used random draws from the final 1,000 iterations from each of three
-MCMC chains to introduce stochasticity into our population models. At
-each time step, we randomly selected climate covariates from one of the
-14 observed years. Then, we drew the full parameter set (climate effects
-and density-dependence fixed effects) from a randomly selected MCMC
-iteration. Using this approach, rather than simply using coefficient
-point estimates, ensures that relatively unimportant climate covariates
-(those that broadly overlap 0) have little effect on the simulation
-results. Since our focus was on the contribution of climate covariates
-to population states, we set the random year effects and the random
-group effects to zero.
+MCMC chains to introduce stochasticity into our population models. We
+thinned these chains by 10 to reduce potential autocorrelation, leaving
+us with 300 samples for each parameter. At each time step, we randomly
+selected climate covariates from one of the 14 observed years. Then, we
+drew the full parameter set (climate effects and density-dependence
+fixed effects) from a randomly selected MCMC iteration. Using this
+approach, rather than simply using coefficient point estimates, ensures
+that relatively unimportant climate covariates (those that broadly
+overlap 0) have little effect on the simulation results. Since our focus
+was on the contribution of climate covariates to population states, we
+set the random year effects and the random group effects to zero.
+
+\paragraph{Quad-based model}\label{quad-based-model}
+
+Our quad-based model (QBM) perfectly mirrors its statistical description
+(Eq. \#). We use the same approach for drawing parameter values as
+described for the IPM.
 
@@
 
+Adler, P. B., H. J. Dalgleish, and S. P. Ellner. 2012. Forecasting plant
+community impacts of climate variability and change: when do competitive
+interactions matter? Journal of Ecology 100:478--487.
+
 Anderson, J., L. Vermeire, and P. B. Adler. 2011. Fourteen years of
@@
 92:1703.
+
+Chu, C., and P. B. Adler. 2015. Large niche differences emerge at the
+recruitment stage to stabilize grassland coexistence. Ecological
+Monographs.
 
@@
 
+Rees, M., and S. P. Ellner. 2009. Integral projection models for
+populations in temporally varying environments. Ecological Monographs
+79:575--594.
+
 Salguero-Gómez, R., O. R. Jones, C. R. Archer, Y. M. Buckley, J.

### 2015-06-10 manuscript/manuscript.Rmd
updates climate perturbation results

@@
 spp_id <- substr(cover_files, 1, 4)
-clim_id <- rep(c("noClim", "Ppt", "Temp",
-                 "Ppt", "Temp",
-                 "Ppt", "Temp"), 4)
-vital_id <- rep(c("All", "Growth", "Growth",
-                  "Survival", "Survival",
-                  "Recruitment", "Recruitment"), 4)
+clim_id <- rep(c("allClim", "Ppt", "ZPptTemp", "Temp",
+                 "Ppt", "ZPptTemp", "Temp", 
+                 "Ppt", "ZPptTemp","Temp"), 4)
+vital_id <- rep(c("All", "Growth", "Growth", "Growth",
+                  "Survival", "Survival", "Survival",
+                  "Recruitment", "Recruitment", "Recruitment"), 4)
 num_files <- length(cover_files)
@@
 
-mean_cover <- ddply(subset(equilibrium_cover, sim=="noClim"), .(species), summarise,
+mean_cover <- ddply(subset(equilibrium_cover, sim=="allClim"), .(species), summarise,
                     value = mean(eq_cover))

### 2015-06-10 manuscript/manuscript.tex
updates ms with results

@@
 and \emph{t}-2 (TmeanSpr1 and TmeanSpr2, respectively), where \emph{t}
-is the observation year.
+is the observation year. We also include interactions among same-year
+climate covariates (e.g., ppt1 $\times$ TmeansSpr1) and climate $\times$
+size interactions.
 
@@
 
+\begin{table}[ht]
+\centering
+\caption{Percent change from equilibrium (temporal median of simulated plant cover) at observed climate relative to a 1\% climate change as forecast by each model type.} 
+\begin{tabular}{llrr}
+  \hline
+Species & Simulation & IPM Change (\%) & QBM Change (\%) \\ 
+  \hline
+BOGR & ppt & 0.92 & 19.43 \\ 
+  BOGR & temp & 12.50 & -93.56 \\ 
+  BOGR & ppt+temp & 25.06 & -27.88 \\ 
+  HECO & ppt & -6.65 & -9.40 \\ 
+  HECO & temp & -6.46 & -10.63 \\ 
+  HECO & ppt+temp & -25.72 & -14.27 \\ 
+  PASM & ppt & -8.53 & 0.25 \\ 
+  PASM & temp & 8.03 & 8.06 \\ 
+  PASM & ppt+temp & 8.74 & 1.30 \\ 
+  POSE & ppt & 1.71 & -8.99 \\ 
+  POSE & temp & 20.06 & -16.80 \\ 
+  POSE & ppt+temp & 11.06 & 4.04 \\ 
+   \hline
+\end{tabular}
+\end{table}
+
+\pagebreak{}
+
 \subsection{Figures}\label{figures}

### 2015-06-10 manuscript/manuscript.Rmd
updates sensitivity results

@@
 All models contain five climate covariate that we chose \emph{a priori}: "water year" precipitation at \emph{t}-1 (lagppt); fall through spring precipitation at \emph{t}-1 and \emph{t}-2 (ppt1 and ppt2, respectively) and mean spring temperature at \emph{t}-1 and \emph{t}-2 (TmeanSpr1 and TmeanSpr2, respectively), where \emph{t} is the observation year.
+We also include interactions among same-year climate covariates (e.g., ppt1 $\times$ TmeansSpr1) and climate $\times$ size interactions.
 
@@
 
-### Vital rate sensitivities to climate (IPM)
+### Forecasting Climate Change Impacts
+Simulated equilibrium cover simulated from the models was sensitive to climate perturbations, but the IPM and QBM produced inconsistent results (Table 2).
+For the IPM, proportional changes range from about 1-25% for *B.
+gracilis*, 7-25% for *H.
+comata*, 8% for *P.
+smithii*, and 2-20% for *P.
+secunda*.
+For the QBM, proportional changes due to climate perturbation ranged from 20-100% for *B.
+gracilis*, 10-14% for *H.
+comata*, 0.25-8% for *P.
+smithii*, and 4-17% for *P.
+secunda*.
+However, the forecasts were extremely uncertain (Figure 1).
+The IPM produced more certain forecasts than the QBM, as expected based on the validation results (Table 1).
+But even for the IPM forecasts, predicted percent changes in median equilibrium cover ranged from -200% to 200% (Figure 1).
+Forecast uncertainty did not vary appreciably among species.
 
-### Forecasting Climate Change Impacts
-Equilibrium cover simulated from the models was sensitive to climate perturbations, but the IPM and QBM produced inconsistent results (Figure 3).
+The population responses of species as forecast by the IPM is a function of the combined effects of individual vital rate sensitivities to climate and the contribution of each vital to population dynamics.
+For *B.
+gracilis*, all vital rates (survival, growth, and rectruitment) were positively affected by increases in precipitation and temperature (Figure 2).
+Intuitively, this leads to an overall increase in *B.
+gracilis* cover (Table 2).
+
+Precipitation increase resulted in all vital rates contributing to an increase in simulated cover for *H.
+comata*.
+But there is a small negative effect of temperature on recruitment.
+THIS IS WEIRD!!!
+CHECK SIMULATIONS.
+
+*P smithii* responded negatively to an increase in precipitation, temperature, or both regardless of the vital rate through which perturbed climate was acting.
+The effects were small, however, leading to relatively minor population consequences under climate change that did not vary much across scenarios (Table 2).
+
 For *B.
@@
 
-It is perhaps startling that two different types of models fit using the same data produce drastically different forecasts (Figure 3).
+### Forecasting the future, and the future of forecasting
+Our goal was not make any explicit forecast for the future state of these populations based on predicted climate change.
+But our results highlight the state of affairs in ecology when it comes to forecasting the impacts of climate change.
+The analysis we conducted here could be considered, with some exceptions of course, at the forefront of ecological forecasting in terms of the statistical approach employed (hierarchical Bayesian), the type of population model we used (stochastic IPM with parameter uncertainty), and the amount of data we had at our disposal (14 years of individual-level data).
+Yet, model predictions proved so uncertain that any forecast, when bounded with uncertainty, would be at best not useful and at worst meaningless.
+
+Something about fitting the models...then cite Britta's paper: 20-25 years needed!
 
@@
 \pagebreak{}
+```{r table_2,results='asis',message=FALSE, comment=NA, echo=FALSE}
+qbm_diffs <- readRDS("../montana/quadBM/simulations/results/qbm_climatesims_logdiffs.RDS")
+qbm_diffs$type <- rep("QBM", nrow(qbm_diffs))
+ipm_diffs <- readRDS("../montana/ipm/simulations/results/ipm_climatesims_logdiffs.RDS")
+ipm_diffs$type <- rep("IPM", nrow(ipm_diffs))
+
+diff_df <- rbind(qbm_diffs, ipm_diffs)
+med_cover_perc <- as.data.frame(diff_df$med_cover*100)
+med_cover_perc$model <- diff_df$type 
+med_cover_perc$sim <- diff_df$variable
+med_cover_perc$species <- diff_df$species
+colnames(med_cover_perc)[1] <- "cover"
+medcast <- dcast(med_cover_perc, species+sim~model, value.var = "cover")
+colnames(medcast) <- c("Species", "Simulation", "IPM Change (%)", "QBM Change (%)")
+simnames <- c("ppt", "temp", "ppt+temp")
+medcast$Simulation <- rep(simnames, 4)
+print.xtable(xtable(medcast, caption = "Percent change from equilibrium (temporal median of simulated plant cover) at observed climate relative to a 1\\% climate change as forecast by each model type."),type="latex", comment=FALSE,
+             include.rownames=FALSE, caption.placement="top")
+```
+
+\pagebreak{}
 
@@
 -------
-```{r fig.cap="Work flow of the data aggregation, model fitting, and population simulating.", fig.width=5, fig.height=8}
+```{r fig.cap="Work flow of the data aggregation, model fitting, and population simulating.", fig.height=6}
 library(png)
@@
 mean_cover <- ddply(subset(equilibrium_cover, sim=="allClim"), .(species), summarise,
-                    value = mean(eq_cover))
+                    value = median(eq_cover))
 
@@
   xlab("Simulation")+
-  theme_bw()
+  scale_x_discrete(labels=c("obs", "ppt", "temp", "ppt + tmp"))+
+  theme_bw()+
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
 setwd("~/Repos/MicroMesoForecast/manuscript")
@@
 Labels on x-axis refer to: 'ppt' = 1% increase in mean precipitation; 'temp' = 1% increase in mean temperature; 'ppt + temp' = 1% increase in mean precipitation and mean temperature.", fig.width=8.5, fig.height=3, cache=FALSE, results='hide'}
-qbm_diffs <- readRDS("../montana/quadBM/simulations/results/qbm_climatesims_percdiffs.rds")
+qbm_diffs <- readRDS("../montana/quadBM/simulations/results/qbm_climatesims_logdiffs.RDS")
 qbm_diffs$type <- rep("QBM", nrow(qbm_diffs))
-ipm_diffs <- readRDS("../montana/ipm/simulations/results/ipm_climatesims_percdiffs.rds")
+ipm_diffs <- readRDS("../montana/ipm/simulations/results/ipm_climatesims_logdiffs.RDS")
 ipm_diffs$type <- rep("IPM", nrow(ipm_diffs))
@@
 myCols2 <- c("grey45", "#277BA8", "#7ABBBD", "#AED77A")
-# ggplot(climSimsQBM)+
-#   geom_boxplot(aes(x=species, y=cover*100, fill=sim), outlier.size=0)+
-#   theme_bw()+
-#   coord_cartesian(ylim=c(0,50))+
-#   scale_fill_manual(name="", labels=c("Observed", "1% Precipitation Increase", "1% Temperature Increase", "1% Temp. & Ppt. Increase"), values=myCols2)
-ggplot(diff_df, aes(x=sim, y=value, fill=type))+
-  geom_bar(stat="identity", position="dodge")+
-  geom_hline(aes(yintercept=0))+
-  scale_fill_manual(values = myCols2[c(1,3)], name="Model type")+
+dgd = position_dodge(width = 0.60)
+ggplot(diff_df)+
+  geom_hline(aes(yintercept=0), linetype=2)+
+  geom_errorbar(aes(x=variable, ymin=lo_cover, ymax=up_cover, group=type),
+                position=dgd, width=0.25, color="grey45")+
+  geom_point(aes(x=variable, y=med_cover, shape=type),
+             position=dgd, size=5, color="white")+
+  geom_point(aes(x=variable, y=med_cover, shape=type),
+             position=dgd, size=3)+
+  scale_shape_manual(values = c(19,17), name="Model type")+
   xlab("Climatic change")+
   ylab("Proportional change in percent cover")+
-  facet_grid(.~variable)+
+  facet_grid(.~species)+
   scale_x_discrete(labels=c("ppt", "temp", "ppt + tmp"))+

### 2015-06-10 manuscript/manuscript.tex
updates sensitivity results

@@
 
-\subsubsection{Vital rate sensitivities to climate
-(IPM)}\label{vital-rate-sensitivities-to-climate-ipm}
-
 \subsubsection{Forecasting Climate Change
@@
 
-Equilibrium cover simulated from the models was sensitive to climate
-perturbations, but the IPM and QBM produced inconsistent results (Figure
-3). For \emph{B. gracilis}, the IPM predicted a modest increase in cover
+Simulated equilibrium cover simulated from the models was sensitive to
+climate perturbations, but the IPM and QBM produced inconsistent results
+(Table 2). For the IPM, proportional changes range from about 1-25\% for
+\emph{B. gracilis}, 7-25\% for \emph{H. comata}, 8\% for \emph{P.
+smithii}, and 2-20\% for \emph{P. secunda}. For the QBM, proportional
+changes due to climate perturbation ranged from 20-100\% for \emph{B.
+gracilis}, 10-14\% for \emph{H. comata}, 0.25-8\% for \emph{P. smithii},
+and 4-17\% for \emph{P. secunda}. However, the forecasts were extremely
+uncertain (Figure 1). The IPM produced more certain forecasts than the
+QBM, as expected based on the validation results (Table 1). But even for
+the IPM forecasts, predicted percent changes in median equilibrium cover
+ranged from -200\% to 200\% (Figure 1). Forecast uncertainty did not
+vary appreciably among species.
+
+The population responses of species as forecast by the IPM is a function
+of the combined effects of individual vital rate sensitivities to
+climate and the contribution of each vital to population dynamics. For
+\emph{B. gracilis}, all vital rates (survival, growth, and rectruitment)
+were positively affected by increases in precipitation and temperature
+(Figure 2). Intuitively, this leads to an overall increase in \emph{B.
+gracilis} cover (Table 2).
+
+Precipitation increase resulted in all vital rates contributing to an
+increase in simulated cover for \emph{H. comata}. But there is a small
+negative effect of temperature on recruitment. THIS IS WEIRD!!! CHECK
+SIMULATIONS.
+
+\emph{P smithii} responded negatively to an increase in precipitation,
+temperature, or both regardless of the vital rate through which
+perturbed climate was acting. The effects were small, however, leading
+to relatively minor population consequences under climate change that
+did not vary much across scenarios (Table 2).
+
+For \emph{B. gracilis}, the IPM predicted a modest increase in cover
 with a 1\% increase in the mean of precipitation or a 1\% increase in
@@
 
-It is perhaps startling that two different types of models fit using the
-same data produce drastically different forecasts (Figure 3).
+\subsubsection{Forecasting the future, and the future of
+forecasting}\label{forecasting-the-future-and-the-future-of-forecasting}
+
+Our goal was not make any explicit forecast for the future state of
+these populations based on predicted climate change. But our results
+highlight the state of affairs in ecology when it comes to forecasting
+the impacts of climate change. The analysis we conducted here could be
+considered, with some exceptions of course, at the forefront of
+ecological forecasting in terms of the statistical approach employed
+(hierarchical Bayesian), the type of population model we used
+(stochastic IPM with parameter uncertainty), and the amount of data we
+had at our disposal (14 years of individual-level data).
 

### 2015-06-11 manuscript/manuscript.Rmd
updates ms with new climate change figure

@@
 Labels on x-axis refer to: 'ppt' = 1% increase in mean precipitation; 'temp' = 1% increase in mean temperature; 'ppt + temp' = 1% increase in mean precipitation and mean temperature.", fig.width=8.5, fig.height=3, cache=FALSE, results='hide'}
-qbm_diffs <- readRDS("../montana/quadBM/simulations/results/qbm_climatesims_logdiffs.RDS")
-qbm_diffs$type <- rep("QBM", nrow(qbm_diffs))
 ipm_diffs <- readRDS("../montana/ipm/simulations/results/ipm_climatesims_logdiffs.RDS")
-ipm_diffs$type <- rep("IPM", nrow(ipm_diffs))
+ipm_diffs$model <- rep("IPM", nrow(ipm_diffs))
+ipm_diffs$type <- rep("paramunc", nrow(ipm_diffs))
+ipm_diffs2 <- readRDS("../montana/ipm/simulations/results/ipm_climatesims_logdiffs_mean.RDS")
+ipm_diffs2$model <- rep("IPM", nrow(ipm_diffs))
+ipm_diffs2$type <- rep("parammean", nrow(ipm_diffs))
+ipm_diffs$med_cover <- ipm_diffs2$med_cover
+ipm_diffs <- rbind(ipm_diffs, ipm_diffs2)
 
-diff_df <- rbind(qbm_diffs, ipm_diffs)
-
-myCols2 <- c("grey45", "#277BA8", "#7ABBBD", "#AED77A")
-dgd = position_dodge(width = 0.60)
-ggplot(diff_df)+
+ggplot(ipm_diffs)+
   geom_hline(aes(yintercept=0), linetype=2)+
-  geom_errorbar(aes(x=variable, ymin=lo_cover, ymax=up_cover, group=type),
+  geom_errorbar(aes(x=species, ymin=lo_cover, ymax=up_cover, group=variable),
                 position=dgd, width=0.25, color="grey45")+
-  geom_point(aes(x=variable, y=med_cover, shape=type),
+  geom_point(aes(x=species, y=med_cover, shape=variable),
              position=dgd, size=5, color="white")+
-  geom_point(aes(x=variable, y=med_cover, shape=type),
+  geom_point(aes(x=species, y=med_cover, shape=variable),
              position=dgd, size=3)+
-  scale_shape_manual(values = c(19,17), name="Model type")+
+#   scale_shape_manual(values = c(19,17), name="Model type")+
   xlab("Climatic change")+
   ylab("Proportional change in percent cover")+
-  facet_grid(.~species)+
-  scale_x_discrete(labels=c("ppt", "temp", "ppt + tmp"))+
+  facet_wrap("type",scales = "free")+
+#   scale_x_discrete(labels=c("ppt", "temp", "ppt + tmp"))+
   theme_bw()
+
+# qbm_diffs <- readRDS("../montana/quadBM/simulations/results/qbm_climatesims_logdiffs.RDS")
+# qbm_diffs$type <- rep("QBM", nrow(qbm_diffs))
+# ipm_diffs <- readRDS("../montana/ipm/simulations/results/ipm_climatesims_logdiffs.RDS")
+# ipm_diffs$type <- rep("IPM", nrow(ipm_diffs))
+# 
+# diff_df <- rbind(qbm_diffs, ipm_diffs)
+# 
+# myCols2 <- c("grey45", "#277BA8", "#7ABBBD", "#AED77A")
+# dgd = position_dodge(width = 0.60)
+# ggplot(diff_df)+
+#   geom_hline(aes(yintercept=0), linetype=2)+
+#   geom_errorbar(aes(x=variable, ymin=lo_cover, ymax=up_cover, group=type),
+#                 position=dgd, width=0.25, color="grey45")+
+#   geom_point(aes(x=variable, y=med_cover, shape=type),
+#              position=dgd, size=5, color="white")+
+#   geom_point(aes(x=variable, y=med_cover, shape=type),
+#              position=dgd, size=3)+
+#   scale_shape_manual(values = c(19,17), name="Model type")+
+#   xlab("Climatic change")+
+#   ylab("Proportional change in percent cover")+
+#   facet_grid(.~species)+
+#   scale_x_discrete(labels=c("ppt", "temp", "ppt + tmp"))+
+#   theme_bw()
 ```

### 2015-06-11 manuscript/manuscript.Rmd
updates for sensitivity figure

@@
 spp_id <- substr(cover_files, 1, 4)
-clim_id <- rep(c("allClim", "Ppt", "ZPptTemp", "Temp",
-                 "Ppt", "ZPptTemp", "Temp", 
-                 "Ppt", "ZPptTemp","Temp"), 4)
-vital_id <- rep(c("All", "Growth", "Growth", "Growth",
-                  "Survival", "Survival", "Survival",
-                  "Recruitment", "Recruitment", "Recruitment"), 4)
+# clim_id <-substr(cover_files, 16, )
+# clim_id <- rep(c("Ppt", "PptTemp", "Temp",
+#                  "Ppt", "ZPptTemp", "Temp", 
+#                  "Ppt", "ZPptTemp","Temp"), 4)
+# vital_id <- rep(c("All", "Growth", "Growth", "Growth",
+#                   "Survival", "Survival", "Survival",
+#                   "Recruitment", "Recruitment", "Recruitment"), 4)
 num_files <- length(cover_files)
@@
   tmp <- read.csv(cover_files[i])
+  tmpid <- substr(cover_files[i], 16, nchar(cover_files[i])-4)
+  tmpid <- strsplit(tmpid, "(?!^)(?=[[:upper:]])", perl=T)
+  if(length(tmpid[[1]])==3){
+    ttt <- paste(tmpid[[1]][[1]], tmpid[[1]][[2]], sep="")
+    tmpid[[1]] <- c(ttt, tmpid[[1]][3])
+  }
   tmp$species <- rep(spp_id[i], nrow(tmp))
-  tmp$sim <- rep(clim_id[i], nrow(tmp))
-  tmp$vital <- rep(vital_id[i], nrow(tmp))
+  tmp$sim <- rep(tmpid[[1]][2], nrow(tmp))
+  tmp$vital <- rep(tmpid[[1]][1], nrow(tmp))
   all_sims <- rbind(all_sims, tmp)

### 2015-06-12 manuscript/manuscript.tex
adds updated climate sensitivity figure for all vital rate combos

@@
 (stochastic IPM with parameter uncertainty), and the amount of data we
-had at our disposal (14 years of individual-level data).
+had at our disposal (14 years of individual-level data). Yet, model
+predictions proved so uncertain that any forecast, when bounded with
+uncertainty, would be at best not useful and at worst meaningless.
+
+Something about fitting the models\ldots{}then cite Britta's paper:
+20-25 years needed!
 
@@
 from each vital rate regression.}
-\end{figure}
-
-\begin{figure}[htbp]
-\centering
-\includegraphics{components/figure/manuscript-figure_2.pdf}
-\caption{Proportional change in species' mean cover caused by a 1\%
-increase in observed precipitation, temperature, or both as predicted by
-the individual-based IPM and the aggregate-based QBM. Note that the
-change for POSE due to a precipitation increase predicted by the QBM is
-almost zero and so does not show up in the figure. Labels on x-axis
-refer to: `ppt' = 1\% increase in mean precipitation; `temp' = 1\%
-increase in mean temperature; `ppt + temp' = 1\% increase in mean
-precipitation and mean temperature.}
 \end{figure}

### 2015-06-12 manuscript/manuscript.Rmd

--- 2015-06-11 manuscript/manuscript.Rmd
@@
 
-```{r figure_1, dependson="plot-options", fig.cap="Sensitivity of equilibrium cover to removal of climate effects from each vital rate regression.", fig.width=4.5, fig.height=6, cache=FALSE, results='hide'}
+```{r figure_1, dependson="plot-options", fig.cap="Sensitivity of equilibrium cover to a 1% increase in precipitation (+ppt), temperature (+temp), or both (+ppt&temp) applied to individual and combined vital rates.
+For example, the points associated with "G" show the median cover from IPM simulations where a climate perturbation is applied only to the growth regression climate covariates.", fig.width=4.5, fig.height=6, cache=FALSE, results='hide'}
 setwd("../montana/ipm/simulations/sensitivity_results/")
@@
 spp_id <- substr(cover_files, 1, 4)
-clim_id <- rep(c("allClim", "Ppt", "ZPptTemp", "Temp",
-                 "Ppt", "ZPptTemp", "Temp", 
-                 "Ppt", "ZPptTemp","Temp"), 4)
-vital_id <- rep(c("All", "Growth", "Growth", "Growth",
-                  "Survival", "Survival", "Survival",
-                  "Recruitment", "Recruitment", "Recruitment"), 4)
 num_files <- length(cover_files)
-all_sims <- data.frame(time=NA, cover=NA, species=NA, sim=NA, vital=NA)
+all_sims <- data.frame(time=NA, cover=NA, species=NA, climsim=NA, vital=NA)
 for(i in 1:num_files){
@@
   tmp$species <- rep(spp_id[i], nrow(tmp))
-  tmp$sim <- rep(clim_id[i], nrow(tmp))
-  tmp$vital <- rep(vital_id[i], nrow(tmp))
   all_sims <- rbind(all_sims, tmp)
@@
 all_sims <- all_sims[2:nrow(all_sims),]
-all_clim_sims <- subset(all_sims, vital=="All")
 
+# Rename vital rates with codes
+vital_names <- unique(all_sims$vital)
+vital_codes <- c("A", "G", "GR", "GS", "R", "S" , "SR") 
+vitals <- data.frame(names=vital_names,code=vital_codes)
+for(vnow in vital_names){
+  ids <- which(all_sims[,"vital"]==vnow)
+  idcode <- which(vitals[,"names"]==vnow)
+  all_sims[ids,"vital"] <- as.character(vitals[idcode,"code"])
+}
+
+# Rep the 'all' simulation for each vital rate for plotting
+all_clim_sims <- subset(all_sims, vital=="A")
 species <- unique(all_sims$species)
-vital_adds <- c("Growth", "Survival", "Recruitment") 
-out_controls <- data.frame(time=NA, cover=NA, species=NA, sim=NA, vital=NA)
+out_controls <- data.frame(time=NA, cover=NA, species=NA, climsim=NA, vital=NA)
 for(spp in species){
   tmp <- subset(all_clim_sims, species==spp)
-  for(vital in vital_adds){
+  for(vital in vital_codes){
     tmp$vital <- vital
@@
 
-all_noclim_sims <- subset(all_sims, vital!="All")
+# Combine the datasets
+all_noclim_sims <- subset(all_sims, vital!="all")
 final <- rbind(out_controls, all_noclim_sims)
 
-equilibrium_cover <- ddply(final, .(species, sim, vital), summarise,
+# Calculate statistics for plotting
+equilibrium_cover <- ddply(final, .(species, climsim, vital), summarise,
                            eq_cover = median(cover))
+mean_cover <- ddply(subset(equilibrium_cover, climsim=="all"), .(species), summarise,
+                    value = mean(eq_cover))
 
-mean_cover <- ddply(subset(equilibrium_cover, sim=="allClim"), .(species), summarise,
-                    value = median(eq_cover))
-
-myCols2 <- c("#277BA8", "#7ABBBD", "#AED77A")
-ggplot(equilibrium_cover, aes(x=sim, y=eq_cover*100, 
-                              linetype=vital, group=vital, shape=vital))+
+equilibrium_cover <- subset(equilibrium_cover, vital!="A")
+bothid <- which(equilibrium_cover[,"climsim"]=="ppttemp")
+equilibrium_cover[bothid,"climsim"] <- "zppttemp"
+# myCols2 <- c("#277BA8", "#7ABBBD", "#AED77A")
+ggplot(equilibrium_cover, aes(x=climsim, y=eq_cover*100, 
+                              shape=vital, group=vital, linetype=vital))+
   geom_hline(data=mean_cover, aes(yintercept=value*100), linetype=2, color="grey45")+
@@
   facet_grid(species~., scales = "free")+
-  scale_linetype_manual(values=c(1,2,3), name="Vital Rate")+
-  scale_shape_manual(values=c(19,17,15), name="Vital Rate")+
+  scale_linetype_discrete(name="Vital Rate")+
+  scale_shape_discrete(name="Vital Rate")+
+  scale_x_discrete(labels=c("baseline", "+ppt", "+temp", "+ppt&temp"))+
   ylab("Equilibrium Cover (%)")+
   xlab("Simulation")+
-  scale_x_discrete(labels=c("obs", "ppt", "temp", "ppt + tmp"))+
   theme_bw()+
@@
 
-ggplot(ipm_diffs)+
-  geom_hline(aes(yintercept=0), linetype=2)+
-  geom_errorbar(aes(x=species, ymin=lo_cover, ymax=up_cover, group=variable),
-                position=dgd, width=0.25, color="grey45")+
-  geom_point(aes(x=species, y=med_cover, shape=variable),
-             position=dgd, size=5, color="white")+
-  geom_point(aes(x=species, y=med_cover, shape=variable),
-             position=dgd, size=3)+
-#   scale_shape_manual(values = c(19,17), name="Model type")+
-  xlab("Climatic change")+
-  ylab("Proportional change in percent cover")+
-  facet_wrap("type",scales = "free")+
-#   scale_x_discrete(labels=c("ppt", "temp", "ppt + tmp"))+
-  theme_bw()
+# ggplot(ipm_diffs)+
+#   geom_hline(aes(yintercept=0), linetype=2)+
+#   geom_errorbar(aes(x=species, ymin=lo_cover, ymax=up_cover, group=variable),
+#                 position=dgd, width=0.25, color="grey45")+
+#   geom_point(aes(x=species, y=med_cover, shape=variable),
+#              position=dgd, size=5, color="white")+
+#   geom_point(aes(x=species, y=med_cover, shape=variable),
+#              position=dgd, size=3)+
+# #   scale_shape_manual(values = c(19,17), name="Model type")+
+#   xlab("Climatic change")+
+#   ylab("Proportional change in percent cover")+
+#   facet_wrap("type",scales = "free")+
+# #   scale_x_discrete(labels=c("ppt", "temp", "ppt + tmp"))+
+#   theme_bw()
 

### 2015-06-12 analysis/quadBM/TableFigureScripts/ModelSelectionDIC/DICrankings_Appendix.Rnw
cleaning and moving files

--- 2014-12-05 manuscript/ms_MicroMeso.Rnw
@@
-\documentclass[12pt]{article}
+\documentclass[10pt]{article}
 \usepackage{graphics}
@@
 
-\title{\LARGE Micro- versus meso-scale population models for ecological forecasting under climate change}
+\title{\LARGE Appendix for: Micro- versus meso-scale population models for ecological forecasting under climate change}
 \author[]{\large Andrew T. Tredennick\footnote{\emph{email:} atredenn@gmail.com}}
@@
 
-\begin{abstract}
-\noindent Forecasts of populations under future climate change are constrained by the spatial and temporal scales at which we can gather data.
-For plants, ecologists often require individual level data to build population models; data that is difficult and expensive to collect at spatial scales greater than one square meter.
-Models built using such data are difficult to scale-up to spatial scales relevant to land management decisions and policy intervention.
-A possible route forward are density-structured models based on coarse-scale census data, which are much easier to collect over large spatial extents.
-However, to use coarse-scale data in a climate change context requires the assumption that the climate signals in population dynamics identified using individual-level data can also be identified when individual-level data are averaged over.
-Here we use a longterm dataset to model population dynamics using individual-level and plot-level data.
-\end{abstract}
+This is an appendix showing tables of top-ranked (by DIC) models for each vital rate regression for each species.
+Models were fit using INLA in R.
+Formulas have the response variable on the left-hand-side of the `~' with covariates on the right.
+When interaction terms are present the main effects are implied.
 
-\noindent{}\textbf{Keywords:} population model, forecasting, integral projection model, etc.
-
-\section{Introduction}
-Producing skillfull forecasts at spatial scales relevant to global environmental changes that play out at the landscape and regional scale has proven challenging for population ecologists.
-The major hurdle is that most population models, at least for plant species, are built using data from small, localized plots because parameterizing vital rate regressions requires tracking the fates of individuals.
-These models are difficult to scale up from the micro to meso-scales because the fitted parameters do not fully represent the spatial variation present at spatial scales beyond that at which the data are collected.
-
-Recently, Freckelton et al., building on work by Hastings and NAME, have proposed density-structured population models that focus on the transition of populations among discrete states, rather than the traditional approach of modeling the transitions of individuals.
-Such an approach could be extremely valuable because the data needed to parameterize density-structured population models is much easier, and less costly, to collect (Queenborough et al.
-year).
-For example, using a density-structured approach, one could build a population model using a time series of annual plot-based censuses of species percent cover.
-However, a major assumption of the density-structured approach is that the aggregate dynamics of the population observed at coarse spatial resolution faithfully represent, and correspond to, the fates of individual plants.
-In other words, using a density-structured approach requires a leap of faith that important covariates (e.g., climate variables) at the level of the individual are captured adequately at the population level.
-If we seek to forecast the impacts of climate change on plant populations, then clearly this assumption requires testing.
-
-\section{Materials and Methods}
-\subsection{Study site and data}
-Our demographic data comes from the Fort Keogh Livestock and Range Research Laboratory in eastern Montana's northern mixed prairie near Miles City, Montana, USA (46 deg.
-19' N, 105 deg 48' W).
-The dataset is freely available on Ecological Archives (CITE), and interested readers should refer to the metadata therin for a complete description.
-The site is about 800 m above sea level and mean annual precipitation (1878-2009) is 334 mm, with most annual precipitaion falling from April through September (76).
-The site is grass dominated and, for the purposes of our study, we focus on the four most abundant graminoid species: \emph{Bouteloua gracilis}, \emph{Hesperostipa comata}, \emph{Pascopyrum smithii}, and \emph{Poa secunda}.
-
-From 1932 to 1945 individual plants were identified and mapped annualy in 44 1-m2 quadrats using a pantograph.
-The quadrats were distributed in six pastures, each assigned a grazing treatment of light (1.24 ha/animal unit month), moderate (0.92 ha/aum), and heavy (0.76 ha/aum) stocking rates (two pastures per treatment).
-In this analysis we account for potential differences among the grazing treatments, but do not focus on grazing$\times$climate interactions.
-The annual maps of the quadrats were digitized and the fates of individual plants tracked and extracted using a computer program.
-Daily climate data, which we aggregated into climate variables of interest, are available for the duration of the data collection period (1932 - 1945) from the Miles City airport, Wiley Field, 9 km from the study site.
-
-\subsection{Stastical models of vital rates}
-The first step in building our single-species population models was to fit statistical models of vital rates at both levels of inference: individual-level and quadrat-level.
-Here we describe the general statistical approach at the individual-level, but the same approach applies at the quadrat-level.
-We modeled survival and growth as functions of genet size (quadrat cover in the quadrat-level models) and climate covariates (described in more detail below).
-We maintained a consistent random effects structure for both models that included three terms: (1) a random year effect on the intercept, (2) a random year effect on the coefficient for plant cover (either individual or quadrat level), and (3) a random effect of group (see Data set description) on the intercept.
-Vital rates for each species are modeled separately.
-
-We used logistic regression to model survival probability:
-
-\begin{equation}
-logit(s) = \beta_{0,t} + \beta_{s,t}x + \beta_{Q} + \beta_{c,1}\theta_{1,t}\\
-+ \cdots +  \beta_{c,i}\theta_{i,t} + \varepsilon_{t}
-\end{equation}
-
-\noindent where $x$ is the log of genet size (or log of quadrat areal cover), $\beta_{0,t}$ is a year-specific intercept, $\beta_{s,t}$ is the year-specific slope parameter for size, $\beta_{Q}$ is the random effect of quadrat group location, $\theta$ is a matrix of $i$ climate effects over $t$ years, $\beta_{c,i}$ is the fixed parameter for the effect of the $i$th climate covariate, and $\varepsilon_{t}$ is the error term.
-At the quadrat level we did not fit random year effects for survival because those parameters did not reach convergence (see "Computing").
-
-We modeled growth as gaussian process describing genet size (or quadrat cover) at time $t+1$ as a function of size at $t$ and climate covariates:
-
-\begin{equation}
-x_{t+1} = \beta_{0,t} + \beta_{s,t}x_{t} + \beta_{Q} + \beta_{c,1}\theta_{1,t}\\
-+ \cdots +  \beta_{c,i}\theta_{i,t} + \varepsilon_{t}
-\end{equation}
-
-\noindent where $x$ is genet size and all other paramters are as described for the survival regression.
-For the quadrat-level approach we modeled growth as a process describing proportional cover within a quadrat at time $t+1$ as a function of proportional cover at time $t$ and climate covariates.
-Thus, instead of a gaussian process, with a normal likelihood, we modeled growth at the quadrat level as above but with a beta likelihood and a logit link to the linear predictors.
-
-While our approach for modeling survival and growth is similar at the individual and quadrat levels, the addition of new genets at the inidividual level or occupied sites at the quadrat level requires two separate approaches.
-
-\subsection{Including climate covariates (and avoiding model selection)}
-For both the individual-level IPM (IPM) and the quadrat-based IBM (QBM) we included four climate covariates in each vital rate regression: fall through spring precipitation at \emph{t}-1 and \emph{t}-2 (ppt1 and ppt2, respectively) and mean spring temperature at \emph{t}-1 and \emph{t}-2 (TmeanSpr1 and TmeanSpr2, respectively), where \emph{t} is the observation year.
-We included climate covariates as additive effects.
-We did not consider interactions among climate covariates or among climate covariates and plant size or proportional cover.
-Climate effects ($\beta_{c}$s) were modeled hierarchically in each vital rate regression so that each species specific climate effect is drawn from an interspecific climate effect distribution.
-This was necessary to achieve convergence at the quadrat level and we used the same approach at the individual level for uniformity.
-
-\subsection{Model fitting}
-All of our analyses (model fitting and simulating) were conducted in Program R.
-We used the MCMC sampler in JAGS to estimate the posterior distributions of model parameters.
-We obtained posterior distributions for all model parameters from three parallel MCMC chains run for 50,000 iterations, after a 50,000 iteration burn in.
-We assessed convergence visually and using the Gelman diagnostic in the R package `coda.' Scale reduction factors for all parameters were less than 1.02, indicating convergence.
-For the purposes of introducing stochasticity in our population models, we saved the final 1,000 iterations from each chain for all parameters to be used as randomly drawn values during population simulation.
-
-\subsection{Population models}
-With the posterior distribution of the vital rate statistical models in hand, it is straightforward to simulate the population models.
-
-\section{Results}
-We assessed the statistical importance of the climate covariates included the final vital rate regressions by comparing the residual deviance of models with climate covariates and temporal random effects, climate covariates only, and temporal random effects only.
-When a model includes climate covariates, this comparison shows the relative contribution of the climate covariates in explaining the total interannual variability (Adler et al.
-2012).
-
-<<echo=FALSE,results=tex, label=GetQuadDeviance>>=
+\newcolumntype{Z}{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}X}
+<<echo=FALSE, results=tex>>=
 library(xtable)
-deviance <- data.frame(VitalRate = c("Growth", "Survival", "Colonization"),
-                       ConstantModel = rep(NA,3),
-                       ClimateModel = rep(NA,3),
-                       FullModel = rep(NA,3),
-                       ClimateContribution = rep(NA,3))
-deviance$FullModel[1] <- round(read.csv("../montana/quadBM/vitalRateRegressions/growth/growthDeviance.csv")[,2],2)
-deviance$ClimateModel[1] <- round(read.csv("../montana/quadBM/vitalRateRegressions/growth/growthDevianceCLIMATE.csv")[,2],2)
-deviance$ConstantModel[1] <- round(read.csv("../montana/quadBM/vitalRateRegressions/growth/growthDevianceCONSTANT.csv")[,2],2)
-deviance$ClimateContribution[1] <- with(deviance[1,], (ClimateModel-ConstantModel)/(FullModel-ConstantModel))
-deviance$ClimateModel[2] <- round(read.csv("../montana/quadBM/vitalRateRegressions/survival/survivalDeviance.csv")[,2],2)
-deviance$ConstantModel[2] <- round(read.csv("../montana/quadBM/vitalRateRegressions/survival/survivalDevianceCONSTANT.csv")[,2],2)
-deviance$ClimateModel[3] <- round(read.csv("../montana/quadBM/vitalRateRegressions/colonization/colonizationDeviance.csv")[,2],2)
-deviance$ConstantModel[3] <- round(read.csv("../montana/quadBM/vitalRateRegressions/colonization/colonizationDevianceCONSTANT.csv")[,2],2)
-deviance$ClimateContribution[2:3] <- with(deviance[2:3,], (ConstantModel-ClimateModel)/(ConstantModel))
-
-colnames(deviance) <- c("Vital Rate", "Constant model", "Climate model", "Full model", "Contribution of climate covariates")
-
-# print(xtable(deviance, align=c("l", "l", "l", "l", "l", "X"), 
-#              caption="The proportion of interannual variability in  vital rates explained by the climate covariates.
-The contribution for growth is defined as: (Climate model - Constant Model)/(Full model - Constant model).
-The contribution for survival and colonization, where we could not estimate a full model with year random effects, is defined as: (Constant Model - Climate Model)/Constant Model.
-The values are model deviance estimated from the `coda' package in R."),
-#       tabular.environment="tabularx", caption.placement="top",
-#       width="\\textwidth", include.rownames=FALSE)
+source("DIC_Table.R")
+colnames(tabG)[3] <- "$\\Delta$DIC"
+print(xtable(tabG[,2:4], align=c("l", "l", "l", "Z"),
+      caption="Growth models."),
+      tabular.environment="tabularx", caption.placement="top",
+      width="\\textwidth", include.rownames=FALSE, sanitize.colnames.function=function(x){x})
 @
 
-<<echo=FALSE, include=FALSE, fig=TRUE, label=DevFig, width=5, height=3, results=hide>>=
-library(ggplot2)
-devGQ <- data.frame(VitalRate = c("Growth", "Survival", "Colonization"),
-                       ConstantModel = rep(NA,3),
-                       ClimateModel = rep(NA,3),
-                       FullModel = rep(NA,3),
-                       ClimateContribution = rep(NA,3))
-devGQ$FullModel[1] <- round(read.csv("../montana/ipm/vitalRateRegs/growth/growthDeviance.csv")[,2],2)
-devGQ$ClimateModel[1] <- round(read.csv("../montana/ipm/vitalRateRegs/growth/growthDevianceCLIMATE.csv")[,2],2)
-devGQ$ConstantModel[1] <- round(read.csv("../montana/ipm/vitalRateRegs/growth/growthDevianceCONSTANT.csv")[,2],2)
-devGQ$FullModel[2] <- round(read.csv("../montana/ipm/vitalRateRegs/survival/survivalDeviance.csv")[,2],2)
-devGQ$ClimateModel[2] <- round(read.csv("../montana/ipm/vitalRateRegs/survival/survivalDevianceCLIMATE.csv")[,2],2)
-devGQ$ConstantModel[2] <- round(read.csv("../montana/ipm/vitalRateRegs/survival/survivalDevianceCONSTANT.csv")[,2],2)
-devGQ$FullModel[3] <- round(read.csv("../montana/ipm/vitalRateRegs/recruitment/recruitmentDeviance.csv")[,2],2)
-devGQ$ClimateModel[3] <- round(read.csv("../montana/ipm/vitalRateRegs/recruitment/recruitmentDevianceCLIMATE.csv")[,2],2)
-devGQ$ConstantModel[3] <- round(read.csv("../montana/ipm/vitalRateRegs/recruitment/recruitmentDevianceCONSTANT.csv")[,2],2)
-devGQ$ClimateContribution[1] <- with(devGQ[1,], (ClimateModel-ConstantModel)/(FullModel-ConstantModel))
-devGQ$ClimateContribution[2] <- with(devGQ[2,], (ConstantModel-ClimateModel)/(ConstantModel))
-devGQ$ClimateContribution[3] <- with(devGQ[3,], (ConstantModel-ClimateModel)/(ConstantModel))
-
-devM <- devGQ[,c(1,5)]
-devM$Level <- rep("Individual", 3)
-tmpD <- deviance[,c(1,5)]
-tmpD$Level <- rep("Quadrat", 3)
-colnames(tmpD)[2] <- "value"
-colnames(tmpD)[1] <- "VitalRate"
-colnames(devM)[2] <- "value"
-plotD <- rbind(tmpD, devM)
-gClimCont <- ggplot(data=plotD, aes(x=VitalRate, y=value, fill=Level))+
-  geom_bar(stat="identity", position="dodge")+
-  scale_fill_manual(values=c("steelblue", "darkorange"))+
-  ylab("Contribution of Climate Effects")+
-  xlab("Vital Rate")+
-  theme_bw()
-gClimCont
+<<echo=FALSE, results=tex>>=
+library(xtable)
+source("DIC_Table.R")
+colnames(tabS)[3] <- "$\\Delta$DIC"
+print(xtable(tabS[,2:4], align=c("l", "l", "l", "Z"),
+      caption="Survival models."),
+      tabular.environment="tabularx", caption.placement="top",
+      width="\\textwidth", include.rownames=FALSE, sanitize.colnames.function=function(x){x})
 @
 
-<<echo=FALSE, include=FALSE, fig=TRUE, label=CoefFig, width=8.5, height=10, results=hide>>=
-gQC <- read.csv("../montana/quadBM/vitalRateRegressions/growth/growthQuants.csv")
-gQC <- gQC[129:144,]
-coefNames <- c("ppt1", "ppt2", "temp1", "temp2")
-gQC$X <- rep(coefNames, each=4)
-gQC$rate <- "growth"
-sQC <- read.csv("../montana/quadBM/vitalRateRegressions/survival/survivalQuants.csv")
-sQC <- sQC[33:48,]
-coefNames <- c("ppt1", "ppt2", "temp1", "temp2")
-sQC$X <- rep(coefNames, each=4)
-sQC$rate <- "survival"
-rQC <- read.csv("../montana/quadBM/vitalRateRegressions/colonization/colonizationQuants.csv")
-rQC <- rQC[29:44,]
-coefNames <- c("ppt1", "ppt2", "temp1", "temp2")
-rQC$X <- rep(coefNames, each=4)
-rQC$rate <- "colonization"
-plotQC <- rbind(gQC,sQC,rQC)
-gQ <- ggplot(plotQC, aes(x=species, y=X50., color=rate))+
-  geom_hline(aes(yintercept=0), linetype="dashed")+
-  geom_errorbar(aes(ymin=X2.5., ymax=X97.5.), width=0, position=position_dodge(width=0.5))+
-  geom_errorbar(aes(ymin=X25., ymax=X75.), width=0, size=1.25, position=position_dodge(width=0.5))+
-  geom_point(size=4, position=position_dodge(width=0.5), shape="|")+
-  ylab("Standardized Coefficient Value")+
-  xlab("Species")+
-  facet_grid(X~.)+
-  coord_flip()+
-  theme_bw()+
-  scale_color_manual(values=c("grey25", "steelblue", "slateblue"))+
-  guides(color=guide_legend(title=""))+
-  theme(legend.position="top")+
-  ggtitle("Quadrat level")
-
-gIC <- read.csv("../montana/ipm/vitalRateRegs/growth/growthQuants.csv")
-gIC <- gIC[133:148,]
-coefNames <- c("ppt1", "ppt2", "temp1", "temp2")
-gIC$X <- rep(coefNames, each=4)
-gIC$rate <- "growth"
-sIC <- read.csv("../montana/ipm/vitalRateRegs/survival/survivalQuants.csv")
-sIC <- sIC[133:148,]
-coefNames <- c("ppt1", "ppt2", "temp1", "temp2")
-sIC$X <- rep(coefNames, each=4)
-sIC$rate <- "survival"
-rIC <- read.csv("../montana/ipm/vitalRateRegs/recruitment/recruitmentQuants.csv")
-rIC <- rIC[81:96,]
-coefNames <- c("ppt1", "ppt2", "temp1", "temp2")
-rIC$X <- rep(coefNames, each=4)
-rIC$rate <- "colonization"
-plotIC <- rbind(gIC,sIC,rIC)
-gI <- ggplot(plotIC, aes(x=species, y=X50., color=rate))+
-  geom_hline(aes(yintercept=0), linetype="dashed")+
-  geom_errorbar(aes(ymin=X2.5., ymax=X97.5.), width=0, position=position_dodge(width=0.5))+
-  geom_errorbar(aes(ymin=X25., ymax=X75.), width=0, size=1.25, position=position_dodge(width=0.5))+
-  geom_point(size=4, position=position_dodge(width=0.5), shape="|")+
-  ylab("Standardized Coefficient Value")+
-  xlab("Species")+
-  facet_grid(X~.)+
-  coord_flip()+
-  theme_bw()+
-  scale_color_manual(values=c("grey25", "steelblue", "slateblue"),
-                     labels=c("recruitment", "growth", "survival"))+
-#   guides(color=FALSE)+
-  guides(color=guide_legend(title=""))+
-  theme(legend.position="top")+
-  ggtitle("Individual level")
-
-library(gridExtra)
-g <- grid.arrange(gQ, gI, ncol=2, nrow=1)
-g
+<<echo=FALSE, results=tex>>=
+library(xtable)
+source("DIC_Table.R")
+colnames(tabC)[3] <- "$\\Delta$DIC"
+print(xtable(tabC[,2:4], align=c("l", "l", "l", "Z"),
+      caption="Colonization models."),
+      tabular.environment="tabularx", caption.placement="top",
+      width="\\textwidth", include.rownames=FALSE, sanitize.colnames.function=function(x){x})
 @
 
-\begin{figure}[th!]
-\begin{center}
-\includegraphics[ width=0.5\textwidth ]{Fig-DevFig}
-\end{center}
-\caption{The proportion of interannual variability in  vital rates explained by the climate covariates.
-The contribution for growth is defined as: (Climate model - Constant Model)/(Full model - Constant model).
-The contribution for survival and colonization, where we could not estimate a full model with year random effects at the quadrat level, is defined as: (Constant Model - Climate Model)/Constant Model.}
-\label{fig:DevFig}
-\end{figure}
-
-\begin{figure}[th!]
-\begin{center}
-\includegraphics[ width=\textwidth] {Fig-CoefFig}
-\end{center}
-\caption{Posterior means (vertical ticks), 75\% credible intervals (heavy lines), and 95\% credible intervals (light lines) of climate effects in each vital rate statistical model.
-The dashed vertical line is at 0, indicating no effect.}
-\label{fig:CoefFig}
-\end{figure}
-
 \end{document}

### 2015-06-12 montana/ipm/simulations/results/clim_change_fig.R

@@ -1,81 +1,0 @@
-##################################################
-#### Script for figure of climate change results
-#### Transfer to manuscript.rmd after complete
-####
-#### Andrew Tredennick
-#### 2-11-2015
-####
-
-library(ggplot2)
-library(plyr)
-library(reshape2)
-
-
-files <- list.files()
-cover_files <- files[grep("cover", files)]
-cover_files <- cover_files[-grep("mean", cover_files)]
-spp_id <- substr(cover_files, 1, 4)
-clim_id <- rep(c("observed", "pptChange", "tempChange", "temppptChange"), 4)
-num_files <- length(cover_files)
-all_sims <- data.frame(time=NA, cover=NA, species=NA, sim=NA)
-for(i in 1:num_files){
-  tmp <- read.csv(cover_files[i])
-#   print(nrow(tmp))
-  tmp$species <- rep(spp_id[i], nrow(tmp))
-  tmp$sim <- rep(clim_id[i], nrow(tmp))
-  all_sims <- rbind(all_sims, tmp)
-}
-all_sims <- all_sims[2:nrow(all_sims),]
-
-ggplot(all_sims)+
-  geom_boxplot(aes(x=species, y=log(cover), fill=sim))
-#   coord_cartesian(ylim = c(0,100))
-tmp_sims <- subset(all_sims, cover < 1)
-ggplot(tmp_sims, aes(x=time, y=cover*100, color=sim))+
-  geom_line()+
-  facet_wrap("species", scales="free")
-
-simcols <- dcast(all_sims, species+time~sim, value.var = "cover")
-obstmp <- simcols$observed
-pertstmp <- simcols[,c("pptChange", "tempChange", "temppptChange")]
-difffxn <- function(X){(log(X)-log(obstmp))}
-tmpout <- as.data.frame(apply(pertstmp, MARGIN = 2, difffxn))
-tmpout$species <- simcols$species
-tmpm <- melt(tmpout, id.vars = "species")
-all_means <- ddply(tmpm, .(species, variable), summarise,
-                   avg_cover = mean(value),
-                   med_cover = median(value),
-                   sd_cover = sd(value),
-                   up_cover = quantile(value, probs=0.875),
-                   lo_cover = quantile(value, probs=0.125))
-dgd = position_dodge(width = 0.60)
-ggplot(all_means)+
[…]

### 2015-06-12 montana/ipm/simulations/results/clim_change_fig_meanparams.R

@@ -1,81 +1,0 @@
-##################################################
-#### Script for figure of climate change results
-#### Transfer to manuscript.rmd after complete
-####
-#### Andrew Tredennick
-#### 2-11-2015
-####
-
-library(ggplot2)
-library(plyr)
-library(reshape2)
-
-
-files <- list.files()
-cover_files <- files[grep("cover", files)]
-cover_files <- cover_files[grep("mean", cover_files)]
-spp_id <- substr(cover_files, 1, 4)
-clim_id <- rep(c("observed", "pptChange", "tempChange", "temppptChange"), 4)
-num_files <- length(cover_files)
-all_sims <- data.frame(time=NA, cover=NA, species=NA, sim=NA)
-for(i in 1:num_files){
-  tmp <- read.csv(cover_files[i])
-#   print(nrow(tmp))
-  tmp$species <- rep(spp_id[i], nrow(tmp))
-  tmp$sim <- rep(clim_id[i], nrow(tmp))
-  all_sims <- rbind(all_sims, tmp)
-}
-all_sims <- all_sims[2:nrow(all_sims),]
-
-ggplot(all_sims)+
-  geom_boxplot(aes(x=species, y=log(cover), fill=sim))
-#   coord_cartesian(ylim = c(0,100))
-tmp_sims <- subset(all_sims, cover < 1)
-ggplot(tmp_sims, aes(x=time, y=cover*100, color=sim))+
-  geom_line()+
-  facet_wrap("species", scales="free")
-
-simcols <- dcast(all_sims, species+time~sim, value.var = "cover")
-obstmp <- simcols$observed
-pertstmp <- simcols[,c("pptChange", "tempChange", "temppptChange")]
-difffxn <- function(X){(log(X)-log(obstmp))}
-tmpout <- as.data.frame(apply(pertstmp, MARGIN = 2, difffxn))
-tmpout$species <- simcols$species
-tmpm <- melt(tmpout, id.vars = "species")
-all_means <- ddply(tmpm, .(species, variable), summarise,
-                   avg_cover = mean(value),
-                   med_cover = median(value),
-                   sd_cover = sd(value),
-                   up_cover = quantile(value, probs=0.875),
-                   lo_cover = quantile(value, probs=0.125))
-dgd = position_dodge(width = 0.60)
-ggplot(all_means)+
[…]

### 2015-06-15 manuscript/manuscript.Rmd
updates ms with loyo one step forecasts for IPM

@@
    - \doublespacing
+   - \setlength\parindent{24pt}
 bibliography: ~/Dropbox/Bibliography/MicroMeso.bib
@@
 ```{r table_1,results='asis',message=FALSE, comment=NA, echo=FALSE}
-# Make a table for average residual from each model and year effects structure
-ipm_onestep <- readRDS("../montana/ipm/simulations/one_step_results/ipm_one-step_forecasts_combined.rds")
-qbm_onestep <- readRDS("../montana/quadBM/simulations/validation_results/qbm_one-step_forecasts_combined.rds")
+####
+##  Make a table for average residual from each model and year effects structure
+####
 
-#IPM
-tmpipm <- ddply(ipm_onestep, .(quad, t1, species), summarise,
-                mean_pred = mean(cover.t1),
-                var_pred = var(cover.t1),
-                mean_obs = mean(obs.cover.t1),
-                resid = mean(resid),
+##  IPM results
+# Read in data
+ipm_onestep <- readRDS("../analysis/ipm/simulations/results/one_step_validation/ipm_loyo_forecasts_combined.rds")
+# Remove predictions >1 for fair comparison to [0,1] truncated quad model
+resets <- which(ipm_onestep[,"cover.t1"]>1)
+ipm_onestep[resets, "cover.t1"] <- 1
+# Average predictions over quad-year reps
+avg_ipm <- ddply(ipm_onestep, .(species, quad, t1), summarise,
+                 avg_prediction = median(cover.t1),
+                 observation = mean(cover.t0),
                  quant_dist = quantile(cover.t1*100, probs = 0.95) - quantile(cover.t1, probs = 0.05))
-ipmstats <- ddply(tmpipm, .(species), summarise,
-                rho = cor(mean_pred, mean_obs),
-                mae = mean(abs(resid)),
-                var = mean(var_pred),
-                mean_cover = mean(mean_obs*100),
+# Calculate error
+avg_ipm$error <- avg_ipm$observation*100 - avg_ipm$avg_prediction*100
+# Aggregate over species
+ipmstats <- ddply(avg_ipm, .(species), summarise,
+                  mae = mean(abs(error)),
+                  mean_cover = mean(observation*100),
                   quant_dist = mean(quant_dist))
@@
 
-#QBM
-tmpqbm <- ddply(qbm_onestep, .(quad, year, species), summarise,
-                mean_pred = mean(predcov),
-                var_pred = var(predcov),
-                mean_obs = mean(obscov),
-                resid = mean(resid),
+##  QBM results
+# Read in data
+qbm_onestep <- readRDS("../analysis/quadBM/simulations/validation_results/qbm_one-step_forecasts_combined.rds")
+# Average predictions over quad-year reps
+avg_qbm <- ddply(qbm_onestep, .(species, quad, year), summarise,
+                 avg_prediction = median(predcov),
+                 observation = mean(obscov),
                  quant_dist = quantile(predcov*100, probs = 0.95) - quantile(predcov, probs = 0.05))
-qbmstats <- ddply(tmpqbm, .(species), summarise,
-                rho = cor(mean_pred, mean_obs),
-                mae = mean(abs(resid)),
-                var = mean(var_pred),
-                mean_cover = mean(mean_obs*100),
+# Calculate error
+avg_qbm$error <- avg_qbm$observation*100 - avg_qbm$avg_prediction*100
+# Aggregate over species
+qbmstats <- ddply(avg_qbm, .(species), summarise,
+                  mae = mean(abs(error)),
+                  mean_cover = mean(observation*100),
                   quant_dist = mean(quant_dist))

### 2015-06-15 analysis/ipm/simulations/results/one_step_validation/check_1steps.R

@@ -11,10 +11,12 @@ for(i in 1:length(spp_list)){
   all_d <- rbind(all_d, tmp_d)
 }
 all_d <- all_d[2:nrow(all_d),]
-all_d$resid <- with(all_d, (cover.t1*100)-(obs.cover.t1*100))
-all_d$year <- rep("ayear effect", nrow(all_d))
-resets <- which(all_d[,"cover.t1"]>1)
-all_d[resets, "cover.t1"] <- 1
+saveRDS(all_d,"ipm_loyo_forecasts_combined.rds")
+
+# all_d$resid <- with(all_d, (cover.t1*100)-(obs.cover.t1*100))
+# all_d$year <- rep("ayear effect", nrow(all_d))
+# resets <- which(all_d[,"cover.t1"]>1)
+# all_d[resets, "cover.t1"] <- 1
 
 # file <- "_sim_cover_1step_ahead_noYear.csv"
 # spp_list <- c("BOGR", "HECO", "PASM", "POSE")
@@ -30,33 +32,33 @@ all_d[resets, "cover.t1"] <- 1
 
 # all_d <- rbind(all_d, all_d2)
 
-saveRDS(all_d,"ipm_loyo_forecasts_combined.rds")
 
-library(ggplot2)
-library(ggplot2)
-ggplot(all_d, aes(x=obs.cover.t1, y=cover.t1))+
-  geom_point()+
-  geom_abline(aes(intercept=0, slope=1), color="red")+
-  facet_wrap("species", scales="free")
 # 
-# myCols <- c("#237DA4", "#7ECAD0")
-# ggplot(all_d,aes(x=as.character(t1),y=resid))+
-#   geom_boxplot(outlier.size = 0)+
-#   facet_wrap("species", scale="free") #+
-#   scale_fill_manual(values=myCols, labels=c("Year effect", "No year effect"))+
-#   theme_bw()
-
-#Calculate average error by species
-library(plyr)
-avgd <- ddply(all_d, .(species, t1), summarise,
-              avg_prediction = median(cover.t1),
-              observation = mean(cover.t0))
-avgd$error <- avgd$observation*100 - avgd$avg_prediction*100
-
-stats <- ddply(avgd, .(species), summarise,
-               mean_abs_error = mean(abs(error)),
-               mean_cover = mean(observation*100))
-stats
-
-
-
+# library(ggplot2)
+# library(ggplot2)
+# ggplot(all_d, aes(x=obs.cover.t1, y=cover.t1))+
+#   geom_point()+
[…]

### 2015-06-15 manuscript/manuscript.Rmd
adds QBM loyo simulation results and updates ms Table 1 with those results

@@
    - \doublespacing
-   - \setlength\parindent{24pt}
 bibliography: ~/Dropbox/Bibliography/MicroMeso.bib
@@
 
+Our analysis offers good news and bad news.
+The good news is that for two out of our four focal species, the QBM faired no worse than the demographically detailed IPM in terms of mean absolute (Table 1).
+Thus, it seems that aggregate level data can provide forecasts of similar skill as individual level data.
+
 ### Forecasting the future, and the future of forecasting
@@
 avg_qbm <- ddply(qbm_onestep, .(species, quad, year), summarise,
-                 avg_prediction = median(predcov),
-                 observation = mean(obscov),
-                 quant_dist = quantile(predcov*100, probs = 0.95) - quantile(predcov, probs = 0.05))
+                 avg_prediction = median(cover.t1),
+                 observation = mean(cover.t0),
+                 quant_dist = quantile(cover.t1*100, probs = 0.95) - quantile(cover.t1, probs = 0.05))
 # Calculate error
@@
 
-mae_list <- list()
-mae_pvals <- list()
-species <- unique(qbmstats$species)
-for(spp in species){
-  tmp1 <- subset(tmpipm, species==spp)
-  tmp2 <- subset(tmpqbm, species==spp)
-  err1 <- as.data.frame(abs(tmp1$mean_obs - tmp1$mean_pred))
-  err2 <- as.data.frame(abs(tmp2$mean_obs - tmp2$mean_pred))
-  colnames(err1) <- "resid"
-  err1$model <- "ipm"
-  colnames(err2) <- "resid"
-  err2$model <- "qbm"
-  errd <-  rbind(err1, err2)
-  mae_ttest <- with(errd, t.test(resid~model, alternative = "less"))
-  mae_list[[spp]] <- mae_ttest
-  mae_pvals[[spp]] <- mae_ttest[["p.value"]]
-}
-
-mae_ps <- melt(mae_pvals)
+# mae_list <- list()
+# mae_pvals <- list()
+# species <- unique(qbmstats$species)
+# for(spp in species){
+#   tmp1 <- subset(tmpipm, species==spp)
+#   tmp2 <- subset(tmpqbm, species==spp)
+#   err1 <- as.data.frame(abs(tmp1$mean_obs - tmp1$mean_pred))
+#   err2 <- as.data.frame(abs(tmp2$mean_obs - tmp2$mean_pred))
+#   colnames(err1) <- "resid"
+#   err1$model <- "ipm"
+#   colnames(err2) <- "resid"
+#   err2$model <- "qbm"
+#   errd <-  rbind(err1, err2)
+#   mae_ttest <- with(errd, t.test(resid~model, alternative = "less"))
+#   mae_list[[spp]] <- mae_ttest
+#   mae_pvals[[spp]] <- mae_ttest[["p.value"]]
+# }
+# 
+# mae_ps <- melt(mae_pvals)
 
@@
 ```{r table_2,results='asis',message=FALSE, comment=NA, echo=FALSE}
-qbm_diffs <- readRDS("../montana/quadBM/simulations/results/qbm_climatesims_logdiffs.RDS")
-qbm_diffs$type <- rep("QBM", nrow(qbm_diffs))
-ipm_diffs <- readRDS("../montana/ipm/simulations/results/ipm_climatesims_logdiffs.RDS")
-ipm_diffs$type <- rep("IPM", nrow(ipm_diffs))
-
-diff_df <- rbind(qbm_diffs, ipm_diffs)
-med_cover_perc <- as.data.frame(diff_df$med_cover*100)
-med_cover_perc$model <- diff_df$type 
-med_cover_perc$sim <- diff_df$variable
-med_cover_perc$species <- diff_df$species
-colnames(med_cover_perc)[1] <- "cover"
-medcast <- dcast(med_cover_perc, species+sim~model, value.var = "cover")
-colnames(medcast) <- c("Species", "Simulation", "IPM Change (%)", "QBM Change (%)")
-simnames <- c("ppt", "temp", "ppt+temp")
-medcast$Simulation <- rep(simnames, 4)
-print.xtable(xtable(medcast, caption = "Percent change from equilibrium (temporal median of simulated plant cover) at observed climate relative to a 1\\% climate change as forecast by each model type."),type="latex", comment=FALSE,
-             include.rownames=FALSE, caption.placement="top")
+# qbm_diffs <- readRDS("../montana/quadBM/simulations/results/qbm_climatesims_logdiffs.RDS")
+# qbm_diffs$type <- rep("QBM", nrow(qbm_diffs))
+# ipm_diffs <- readRDS("../montana/ipm/simulations/results/ipm_climatesims_logdiffs.RDS")
+# ipm_diffs$type <- rep("IPM", nrow(ipm_diffs))
+# 
+# diff_df <- rbind(qbm_diffs, ipm_diffs)
+# med_cover_perc <- as.data.frame(diff_df$med_cover*100)
+# med_cover_perc$model <- diff_df$type 
+# med_cover_perc$sim <- diff_df$variable
+# med_cover_perc$species <- diff_df$species
+# colnames(med_cover_perc)[1] <- "cover"
+# medcast <- dcast(med_cover_perc, species+sim~model, value.var = "cover")
+# colnames(medcast) <- c("Species", "Simulation", "IPM Change (%)", "QBM Change (%)")
+# simnames <- c("ppt", "temp", "ppt+temp")
+# medcast$Simulation <- rep(simnames, 4)
+# print.xtable(xtable(medcast, caption = "Percent change from equilibrium (temporal median of simulated plant cover) at observed climate relative to a 1\\% climate change as forecast by each model type."),type="latex", comment=FALSE,
+#              include.rownames=FALSE, caption.placement="top")
 ```
@@
 -------
-```{r fig.cap="Work flow of the data aggregation, model fitting, and population simulating.", fig.height=6}
+```{r figure_1,dependson="plot-options", fig.cap="Work flow of the data aggregation, model fitting, and population simulating.", fig.height=6}
 library(png)
@@
 
-```{r figure_1, dependson="plot-options", fig.cap="Sensitivity of equilibrium cover to a 1% increase in precipitation (+ppt), temperature (+temp), or both (+ppt&temp) applied to individual and combined vital rates.
-For example, the points associated with "G" show the median cover from IPM simulations where a climate perturbation is applied only to the growth regression climate covariates.", fig.width=4.5, fig.height=6, cache=FALSE, results='hide'}
-setwd("../montana/ipm/simulations/sensitivity_results/")
+```{r figure_2, dependson="plot-options", fig.cap="Sensitivity of equilibrium cover to a 1% increase in precipitation (+ppt), temperature (+temp), or both (+ppt&temp) applied to individual and combined vital rates.
+For example, the points associated with G show the median cover from IPM simulations where a climate perturbation is applied only to the growth regression climate covariates.", fig.width=4.5, fig.height=6, cache=FALSE, results='hide'}
+setwd("../analysis/ipm/simulations/results/climate_sensitivity/")
 files <- list.files()
@@
   theme(axis.text.x = element_text(angle = 45, hjust = 1))
+
+covertest <- subset(equilibrium_cover, climsim=="all")
+colnames(covertest)[4] <- "obs_cover"
+out <- merge(subset(equilibrium_cover, climsim!="all"), covertest[,c("species","vital","obs_cover")])
+out$propdiff <- with(out, log(eq_cover*100)-log(obs_cover*100))
 setwd("~/Repos/MicroMesoForecast/manuscript")
@@
 
-```{r figure_2, dependson="plot-options", fig.cap="Proportional change in species' mean cover caused by a 1% increase in observed precipitation, temperature, or both as predicted by the individual-based IPM and the aggregate-based QBM.
+```{r figure_3, dependson="plot-options", fig.cap="Proportional change in species' mean cover caused by a 1% increase in observed precipitation, temperature, or both as predicted by the individual-based IPM and the aggregate-based QBM.
 Note that the change for POSE due to a precipitation increase predicted by the QBM is almost zero and so does not show up in the figure.
 Labels on x-axis refer to: 'ppt' = 1% increase in mean precipitation; 'temp' = 1% increase in mean temperature; 'ppt + temp' = 1% increase in mean precipitation and mean temperature.", fig.width=8.5, fig.height=3, cache=FALSE, results='hide'}
-ipm_diffs <- readRDS("../montana/ipm/simulations/results/ipm_climatesims_logdiffs.RDS")
-ipm_diffs$model <- rep("IPM", nrow(ipm_diffs))
-ipm_diffs$type <- rep("paramunc", nrow(ipm_diffs))
-ipm_diffs2 <- readRDS("../montana/ipm/simulations/results/ipm_climatesims_logdiffs_mean.RDS")
-ipm_diffs2$model <- rep("IPM", nrow(ipm_diffs))
-ipm_diffs2$type <- rep("parammean", nrow(ipm_diffs))
-ipm_diffs$med_cover <- ipm_diffs2$med_cover
-ipm_diffs <- rbind(ipm_diffs, ipm_diffs2)
+# ipm_diffs <- readRDS("../montana/ipm/simulations/results/ipm_climatesims_logdiffs.RDS")
+# ipm_diffs$model <- rep("IPM", nrow(ipm_diffs))
+# ipm_diffs$type <- rep("paramunc", nrow(ipm_diffs))
+# ipm_diffs2 <- readRDS("../montana/ipm/simulations/results/ipm_climatesims_logdiffs_mean.RDS")
+# ipm_diffs2$model <- rep("IPM", nrow(ipm_diffs))
+# ipm_diffs2$type <- rep("parammean", nrow(ipm_diffs))
+# ipm_diffs$med_cover <- ipm_diffs2$med_cover
+# ipm_diffs <- rbind(ipm_diffs, ipm_diffs2)
 

### 2015-06-15 manuscript/manuscript.tex
adds QBM loyo simulation results and updates ms Table 1 with those results

@@
 \doublespacing
+\setlength\parindent{24pt}
 
@@
   \hline
-BOGR & IPM & 4.77 & 29.33 & 9.16 \\ 
-  BOGR & QBM & 5.30 & 44.63 & 9.22 \\ 
-  HECO & IPM & 0.60 & 2.82 & 1.22 \\ 
-  HECO & QBM & 2.03 & 12.05 & 1.25 \\ 
-  PASM & IPM & 0.18 & 0.49 & 0.40 \\ 
-  PASM & QBM & 0.17 & 1.34 & 0.40 \\ 
-  POSE & IPM & 0.78 & 2.16 & 1.23 \\ 
-  POSE & QBM & 1.30 & 7.04 & 1.25 \\ 
+BOGR & IPM & 12.18 & 38.52 & 9.43 \\ 
+  BOGR & QBM & 19.66 & 56.50 & 9.26 \\ 
+  HECO & IPM & 1.22 & 6.47 & 1.15 \\ 
+  HECO & QBM & 12.35 & 41.11 & 1.18 \\ 
+  PASM & IPM & 0.19 & 1.65 & 0.42 \\ 
+  PASM & QBM & 0.55 & 7.78 & 0.42 \\ 
+  POSE & IPM & 1.37 & 7.64 & 1.25 \\ 
+  POSE & QBM & 1.79 & 40.59 & 1.27 \\ 
    \hline
@@
 
-\begin{table}[ht]
-\centering
-\caption{Percent change from equilibrium (temporal median of simulated plant cover) at observed climate relative to a 1\% climate change as forecast by each model type.} 
-\begin{tabular}{llrr}
-  \hline
-Species & Simulation & IPM Change (\%) & QBM Change (\%) \\ 
-  \hline
-BOGR & ppt & 0.92 & 19.43 \\ 
-  BOGR & temp & 12.50 & -93.56 \\ 
-  BOGR & ppt+temp & 25.06 & -27.88 \\ 
-  HECO & ppt & -6.65 & -9.40 \\ 
-  HECO & temp & -6.46 & -10.63 \\ 
-  HECO & ppt+temp & -25.72 & -14.27 \\ 
-  PASM & ppt & -8.53 & 0.25 \\ 
-  PASM & temp & 8.03 & 8.06 \\ 
-  PASM & ppt+temp & 8.74 & 1.30 \\ 
-  POSE & ppt & 1.71 & -8.99 \\ 
-  POSE & temp & 20.06 & -16.80 \\ 
-  POSE & ppt+temp & 11.06 & 4.04 \\ 
-   \hline
-\end{tabular}
-\end{table}
-
 \pagebreak{}
@@
 \centering
-\includegraphics{components/figure/manuscript-unnamed-chunk-1.pdf}
+\includegraphics{components/figure/manuscript-figure_1.pdf}
 \caption{Work flow of the data aggregation, model fitting, and
@@
 \centering
-\includegraphics{components/figure/manuscript-figure_1.pdf}
-\caption{Sensitivity of equilibrium cover to removal of climate effects
-from each vital rate regression.}
+\includegraphics{components/figure/manuscript-figure_2.pdf}
+\caption{Sensitivity of equilibrium cover to a 1\% increase in
+precipitation (+ppt), temperature (+temp), or both (+ppt\&temp) applied
+to individual and combined vital rates. For example, the points
+associated with G show the median cover from IPM simulations where a
+climate perturbation is applied only to the growth regression climate
+covariates.}
 \end{figure}

### 2015-06-15 analysis/quadBM/simulations/loyo_validation_qbm.R

@@ -206,8 +206,9 @@ for(do_species in sppList){
     colnames(startNs)[1:2] <- c("quad", "cover.t0")
     
     outsaves <- merge(newNs, startNs)
+    outall <- rbind(outall, outsaves)
   }#end year loop
-  outall <- rbind(outall, outsaves)
+  outall <- outall[2:nrow(outall),]
   outname <- paste(do_species,"_sim_cover_1step_ahead_year.RDS", sep="")
   saveRDS(outall, paste("./validation_results/", outname, sep=""))
 }#end species loop

### 2015-06-15 manuscript/manuscript.Rmd
updates ms with new cliamte change figure

@@
 
-All R code and data necessary to replicate our analysis is archived on GitHub as release 1.0 (http://github.com/atredennick/MicroMesoForecast/releases).
+All R code and data necessary to reproduce our analysis is archived on GitHub as release 1.0 (http://github.com/atredennick/MicroMesoForecast/releases).
 That stable release will remain static as a record of this analysis, but subsequent versions may appear if we update this work.
@@
 # Read in data
-qbm_onestep <- readRDS("../analysis/quadBM/simulations/validation_results/qbm_one-step_forecasts_combined.rds")
+qbm_onestep <- readRDS("../analysis/quadBM/simulations/results/one_step_validation/qbm_one-step_forecasts_combined.rds")
 # Average predictions over quad-year reps
@@
 Forecasts were made without random year effects; only climate covariates could explain year-to-year variation.
-90 Distance refers to the average distance between the upper and lower 90 percentiles of the 100 predicted values for each quadrat-year combination."),type="latex", comment=FALSE,
+90\\% Distance refers to the average distance between the upper and lower 90th percentiles of the 100 predicted values for each quadrat-year combination."),type="latex", comment=FALSE,
              include.rownames=FALSE, caption.placement="top")
@@
 library(grid)
-img <- readPNG("components/figure/micromeso_flow.png")
+img <- readPNG("components/figure/micromeso_flowchart.png")
 grid.raster(img)
@@
 
-```{r figure_3, dependson="plot-options", fig.cap="Proportional change in species' mean cover caused by a 1% increase in observed precipitation, temperature, or both as predicted by the individual-based IPM and the aggregate-based QBM.
-Note that the change for POSE due to a precipitation increase predicted by the QBM is almost zero and so does not show up in the figure.
-Labels on x-axis refer to: 'ppt' = 1% increase in mean precipitation; 'temp' = 1% increase in mean temperature; 'ppt + temp' = 1% increase in mean precipitation and mean temperature.", fig.width=8.5, fig.height=3, cache=FALSE, results='hide'}
-# ipm_diffs <- readRDS("../montana/ipm/simulations/results/ipm_climatesims_logdiffs.RDS")
-# ipm_diffs$model <- rep("IPM", nrow(ipm_diffs))
-# ipm_diffs$type <- rep("paramunc", nrow(ipm_diffs))
-# ipm_diffs2 <- readRDS("../montana/ipm/simulations/results/ipm_climatesims_logdiffs_mean.RDS")
-# ipm_diffs2$model <- rep("IPM", nrow(ipm_diffs))
-# ipm_diffs2$type <- rep("parammean", nrow(ipm_diffs))
-# ipm_diffs$med_cover <- ipm_diffs2$med_cover
-# ipm_diffs <- rbind(ipm_diffs, ipm_diffs2)
+```{r figure_3, dependson="plot-options", fig.cap="Proportional change in species' mean cover caused by a 1% increase in observed precipitation (+ppt), temperature (+temp), or both (+ppt&temp) as predicted by the individual-based IPM and the aggregate-based QBM.
+Top panels show the mean predicted proportional change in cover; lower panels show the 90% quantiles of predicted proportional cover change.
+Mean predictions and their associated uncertainties are shown in separate panels for visual clarity.", fig.width=8.5, fig.height=6, cache=FALSE, results='hide'}
+setwd("~/Repos/MicroMesoForecast/analysis/")
 
-# ggplot(ipm_diffs)+
-#   geom_hline(aes(yintercept=0), linetype=2)+
-#   geom_errorbar(aes(x=species, ymin=lo_cover, ymax=up_cover, group=variable),
-#                 position=dgd, width=0.25, color="grey45")+
-#   geom_point(aes(x=species, y=med_cover, shape=variable),
-#              position=dgd, size=5, color="white")+
-#   geom_point(aes(x=species, y=med_cover, shape=variable),
-#              position=dgd, size=3)+
-# #   scale_shape_manual(values = c(19,17), name="Model type")+
-#   xlab("Climatic change")+
-#   ylab("Proportional change in percent cover")+
-#   facet_wrap("type",scales = "free")+
-# #   scale_x_discrete(labels=c("ppt", "temp", "ppt + tmp"))+
-#   theme_bw()
+####
+##  Mean parameter runs
+####
+meanparam_ipm <- data.frame(cover=NA, species=NA, 
+                             climsim=NA, model=NA)
+##  IPM 
+path <- "./ipm/simulations/results/climchange_meanparams/"
+ipmfiles_mean <- list.files(path)
+ipmfiles_mean <- ipmfiles_mean[grep("cover", ipmfiles_mean)]
+for(file in ipmfiles_mean){
+  tmp <- read.csv(paste(path,file,sep=""))
+  tmp$model <- "IPM"
+  meanparam_ipm <- rbind(meanparam_ipm, tmp[,c("cover", "species", "climsim", "model")])
+}
+meanparam_ipm <- meanparam_ipm[2:nrow(meanparam_ipm),]
 
-# qbm_diffs <- readRDS("../montana/quadBM/simulations/results/qbm_climatesims_logdiffs.RDS")
-# qbm_diffs$type <- rep("QBM", nrow(qbm_diffs))
-# ipm_diffs <- readRDS("../montana/ipm/simulations/results/ipm_climatesims_logdiffs.RDS")
-# ipm_diffs$type <- rep("IPM", nrow(ipm_diffs))
-# 
-# diff_df <- rbind(qbm_diffs, ipm_diffs)
-# 
-# myCols2 <- c("grey45", "#277BA8", "#7ABBBD", "#AED77A")
-# dgd = position_dodge(width = 0.60)
-# ggplot(diff_df)+
-#   geom_hline(aes(yintercept=0), linetype=2)+
-#   geom_errorbar(aes(x=variable, ymin=lo_cover, ymax=up_cover, group=type),
-#                 position=dgd, width=0.25, color="grey45")+
-#   geom_point(aes(x=variable, y=med_cover, shape=type),
-#              position=dgd, size=5, color="white")+
-#   geom_point(aes(x=variable, y=med_cover, shape=type),
-#              position=dgd, size=3)+
-#   scale_shape_manual(values = c(19,17), name="Model type")+
-#   xlab("Climatic change")+
-#   ylab("Proportional change in percent cover")+
-#   facet_grid(.~species)+
-#   scale_x_discrete(labels=c("ppt", "temp", "ppt + tmp"))+
-#   theme_bw()
+newd <- data.frame(cover=NA, species=NA, 
+                   climsim=NA, model=NA, obscover=NA)
+for(dospp in unique(meanparam_ipm$species)){
+  tmp <- subset(meanparam_ipm, species==dospp)
+  obscover <- subset(tmp, climsim=="obs")["cover"]
+  tmp2 <- subset(tmp, climsim!="obs")
+  tmp2$obscover <- rep(obscover$cover, times = length(unique(tmp2$climsim)))
+  newd <- rbind(newd, tmp2)
+}
+ipm_meansims <- newd[2:nrow(newd),]
+ipm_meansims$propdiff <- with(ipm_meansims, log(cover*100)-log(obscover*100))
+
+##  QBM
+meanparam_qbm <- data.frame(cover=NA, species=NA, 
+                             climsim=NA, model=NA)
+path <- "./quadBM/simulations/results/climatechange_meanparams/"
+qbmfiles_mean <- list.files(path)
+for(file in qbmfiles_mean){
+  tmp <- readRDS(paste(path,file,sep=""))
+  tmp$model <- "QBM"
+  meanparam_qbm <- rbind(meanparam_qbm, tmp[,c("cover", "species", "climsim", "model")])
+}
+meanparam_qbm <- meanparam_qbm[2:nrow(meanparam_qbm),]
+
+newd <- data.frame(cover=NA, species=NA, 
+                   climsim=NA, model=NA, obscover=NA)
+for(dospp in unique(meanparam_qbm$species)){
+  tmp <- subset(meanparam_qbm, species==dospp)
+  obscover <- subset(tmp, climsim=="obs")["cover"]
+  tmp2 <- subset(tmp, climsim!="obs")
+  tmp2$obscover <- rep(obscover$cover, times = length(unique(tmp2$climsim)))
+  newd <- rbind(newd, tmp2)
+}
+qbm_meansims <- newd[2:nrow(newd),]
+qbm_meansims$propdiff <- with(qbm_meansims, log(cover*100)-log(obscover*100))
+
+##  Merge IPM and QBM mean parameter results
+bothmodel_outs <- rbind(ipm_meansims, qbm_meansims)
+meanparam_agg <- ddply(bothmodel_outs, .(species, climsim, model), summarise,
+                       meandiff = mean(propdiff),
+                       hidiff = quantile(propdiff, probs = 0.95),
+                       lodiff = quantile(propdiff, probs = 0.05))
+dgd = position_dodge(width = 0.9)
+meanplot <- ggplot(meanparam_agg)+
+  geom_bar(aes(x=climsim, y=meandiff, fill=model),
+             position=dgd, stat="identity")+
+  theme_bw()+
+  scale_fill_manual(values=c("black", "grey45"), name="Model")+
+  facet_wrap("species", scales="free", nrow=1)+
+  scale_x_discrete(labels=c("+ppt", "+temp", "+ppt&temp"))+
+  ylab("Mean proportional difference")+
+  xlab("Climate simulation")+
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+
+dgd = position_dodge(width = 0.6)
+uncertainplot <- ggplot(meanparam_agg)+
+  geom_errorbar(aes(x=climsim, ymax=hidiff, ymin=lodiff, group=model, linetype=model),
+                position=dgd, width=0.25)+
+  theme_bw()+
+  scale_linetype_discrete(name="Model")+
+  facet_wrap("species", nrow=1, scales="free")+
+  scale_x_discrete(labels=c("+ppt", "+temp", "+ppt&temp"))+
+  ylab("Proportional difference")+
+  xlab("Climate simulation")+
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+
+meanparamplot <- grid.arrange(meanplot, uncertainplot, ncol=1, nrow=2)
+print(meanparamplot)
 ```
-
-\pagebreak{}
-
-.
 

### 2015-06-15 manuscript/manuscript.tex
updates ms with new cliamte change figure

@@
 \doublespacing
-\setlength\parindent{24pt}
 
@@
 Tredennick\footnote{Corresponding author:
-  \href{mailto:atredenn@gmail.com}{\href{mailto:atredenn@gmail.com}{atredenn@gmail.com}}}
+  \href{mailto:atredenn@gmail.com}{\href{mailto:atredenn@gmail.com}{\nolinkurl{atredenn@gmail.com}}}}
 and Peter B.
@@
 Research Laboratory in eastern Montana's northern mixed prairie near
-Miles City, Montana, USA ($46^{\circ}$ 19' N, $105^{\circ}$ 48' W). The
-dataset is freely available on Ecological Archives\footnote{\url{http://esapubs.org/archive/ecol/E092/143/}}
+Miles City, Montana, USA (\(46^{\circ}\) 19' N, \(105^{\circ}\) 48' W).
+The dataset is freely available on Ecological Archives\footnote{\url{http://esapubs.org/archive/ecol/E092/143/}}
 (Anderson et al. 2011) , and interested readers should refer to the
@@
 is the observation year. We also include interactions among same-year
-climate covariates (e.g., ppt1 $\times$ TmeansSpr1) and climate $\times$
-size interactions.
+climate covariates (e.g., ppt1 \(\times\) TmeansSpr1) and climate
+\(\times\) size interactions.
 
@@
 
-where $x_{jq,t}$ is the log of species' $j$ proportional cover in
-quadrat $q$ at time $t$ and all other parameters are as in the
+where \(x_{jq,t}\) is the log of species' \(j\) proportional cover in
+quadrat \(q\) at time \(t\) and all other parameters are as in the
 individual-level growth model (Eq. \#). The log normal likelihood
@@
 specification of Chu and Adler (2015) where the population of species
-\emph{j} is a density function $n(u_{j},t)$ giving the density of
+\emph{j} is a density function \(n(u_{j},t)\) giving the density of
 sized-\emph{u} genets at time \emph{t}. Genet size is on the natural log
-scale, so that $n(u_{j},t)du$ is the number of genets whose area (on the
-arithmetic scale) is between $e^{u_{j}}$ and $e^{u_{j}+du}$. So, the
-density function for any size \emph{v} at time $t+1$ is
+scale, so that \(n(u_{j},t)du\) is the number of genets whose area (on
+the arithmetic scale) is between \(e^{u_{j}}\) and \(e^{u_{j}+du}\). So,
+the density function for any size \emph{v} at time \(t+1\) is
 
@@
 
-where $k_{j}(v_{j},u_{j},\bar{\bold{w_{j}}})$ is the population kernal
-that describes all possible transitions from size $u$ to $v$ and
-$\bar{\bold{w_{j}}}$ is a vector of estimates of average crowding
-experienced from all other species by a genet of size $u_j$ and species
-$j$. The integral is evaluated over all possible sizes between
+where \(k_{j}(v_{j},u_{j},\bar{\bold{w_{j}}})\) is the population kernal
+that describes all possible transitions from size \(u\) to \(v\) and
+\(\bar{\bold{w_{j}}}\) is a vector of estimates of average crowding
+experienced from all other species by a genet of size \(u_j\) and
+species \(j\). The integral is evaluated over all possible sizes between
 predefined lower (\emph{L}) and upper (\emph{U}) size limits that extend
@@
 with size
-($R_j(v_{j},u_{j},\bar{\bold{w_{j}}}) = e^{u_j}R_j(v_{j},\bar{\bold{w_{j}}})$)
+(\(R_j(v_{j},u_{j},\bar{\bold{w_{j}}}) = e^{u_j}R_j(v_{j},\bar{\bold{w_{j}}})\))
 to incorporate the recruitment function in the spatially-implicit IPM.
@@
 
+Our analysis offers good news and bad news. The good news is that for
+two out of our four focal species, the QBM faired no worse than the
+demographically detailed IPM in terms of mean absolute (Table 1). Thus,
+it seems that aggregate level data can provide forecasts of similar
+skill as individual level data.
+
 \subsubsection{Forecasting the future, and the future of
@@
 Forecasts were made without random year effects; only climate covariates could explain year-to-year variation.
-90 Distance refers to the average distance between the upper and lower 90 percentiles of the 100 predicted values for each quadrat-year combination.}
+90\% Distance refers to the average distance between the upper and lower 90th percentiles of the 100 predicted values for each quadrat-year combination.}
 \begin{tabular}{llrrr}
@@
 
-\pagebreak{}
-
-.
+\begin{figure}[htbp]
+\centering
+\includegraphics{components/figure/manuscript-figure_3.pdf}
+\caption{Proportional change in species' mean cover caused by a 1\%
+increase in observed precipitation (+ppt), temperature (+temp), or both
+(+ppt\&temp) as predicted by the individual-based IPM and the
+aggregate-based QBM. Top panels show the mean predicted proportional
+change in cover; lower panels show the 90\% quantiles of predicted
+proportional cover change. Mean predictions and their associated
+uncertainties are shown in separate panels for visual clarity.}
+\end{figure}
 

### 2015-06-16 manuscript/manuscript.Rmd
updates ms with MAE statistics and more text on validation methods

@@
 
-Do we need detailed demographic data to forecast population responses to climate change? Probably.
+Microscale models outperform mesoscale models when forecasting climate change impacts on plant populations
 ========================================================================================
@@
 We used random draws from the final 1,000 iterations from each of three MCMC chains to introduce stochasticity into our population models.
-We thinned these chains by 10 to reduce potential autocorrelation, leaving us with 300 samples for each parameter.
 At each time step, we randomly selected climate covariates from one of the 14 observed years.
@@
 ### Model validation
-To test each model's ability to forecast the population state we used leave-one-year-out cross validation.
+To test each model's ability to forecast the population state we used out of sample predictions using leave-one-year-out cross validation.
 For both levels of modeling, we fit the vital rate models using observations from all years except one, and then used those fitted parameters in the population models to perform a one-step-ahead forecast for the year whose observations were withheld from model fitting.
-We repeated this procedure for all 13 observation years.
+Within each observation year, several quadrats are sampled.
+So we made predictions for each observed quadrat in the focal year initialized with cover the previous year.
+Since we were making quadrat specific predictions we incorporated the group effect on the intercept for both models.
+We repeated this procedure for all 13 observation years, making 100 one-step-ahead forecasts for each quadrat-year combination with parameter uncertainty included via randomdrawd from the MCMC chain as described above.
+Random year effects were set to zero since year effects cannot be assigned to unobserved years.
+
 This model validation allowed us to compare accuracy and precision of the two modeling approaches (individual-level versus population-level).
+We first calculated the median predicted cover across the 100 simulations for each quadrat-year and then calculated the absolute error as the difference between the observed cover for a given quadrat-year and the median prediction.
+To arrive at mean absolute error (MAE), we then averaged the absolute error within each species across the quadrat-year specific errors.
+We use MAE as our measure of accuracy.
+To measure precision calculated the distance between the upper and lower 90th quantiles of the 100 predictions as a measure of forecast precision and averaged this value over quadrat-years for each species.
+
+### Testing sensitivity to climate covariates
+Our main goal in this paper is to see if models based on aggregate level data are as sensitive to climate as models based on individual level data.
+So, with our fitted and validated models in hand, we ran simulations for each model type (IPM and QBM) under four climate perturbation scenarios: (1) observed climate, (2) precipitation increased by 1%, (3) temperature increased by 1%, and (4) precipitation and temperature increased by 1%.
+We ran this simulations for 2,500 time steps, enough to estimate equilibrium cover after discarding an initial 500 time steps as burn-in.
+Each simulation was run under two parameter scenarios: (1) using mean parameter estimates and (2) using randomly drawn parameters from the MCMC chain.
+We use (1) to detect the overall sensitivity of equilibrium cover to climate, and we use (2) to show the impact of model uncertainty on forecast precision.
+
+As an effort to identify potential discrepencies between IPM and QBM forecasts, we also ran simulations designed to quantify the sensitivities of individual and combined vital rates to climate for the IPM.
+Specifically, we ran simulations for the above climate scenarios, but applied the perturbed climate covariates to survival, growth, and recruitment vital rates individually and in pairwise combinations.
+This allows us to isolate the vital rate(s) most sensitive to climate.
+For this analysis, we used mean parameter estimates to reduce the sources of uncertainty in the sensitivity estimates.
 
@@
 ### Comparison of Forecast Models
-We performed one-step-ahead forecasts to compare the two models (IPM vs.
-QBM).
-We did not use the random year effects for these forecasts.
-Thus, beyond the effect of genet size or quadrat cover, only the climate covariates could affect expect plant cover.
 The IPM had significantly lower overall error (MAE, mean absolute error) for two species (*H.
@@
 In no case did the QBM significantly outperform the IPM (Table 1).
-As a metric of model uncertainty, we calculated the euclidean distance between the upper and lower 90% quantiles for each model's forecasts.
-The IPM always outperformed the QBM in this respect (Table 1), with consistently lower distances between the 90% quantiles.
+The IPM was consistently more precise than the QBM, with lower distances between the 90% quantiles across all species (Table 1).
 In general the IPM outperformed the QBM because it had (i) lower MAE for two species, (ii) statistically similar MAE for the other two species, and (iii) considerably more precise forecasts for all species.
@@
 ### Forecasting Climate Change Impacts
-Simulated equilibrium cover simulated from the models was sensitive to climate perturbations, but the IPM and QBM produced inconsistent results (Table 2).
-For the IPM, proportional changes range from about 1-25% for *B.
-gracilis*, 7-25% for *H.
-comata*, 8% for *P.
-smithii*, and 2-20% for *P.
-secunda*.
-For the QBM, proportional changes due to climate perturbation ranged from 20-100% for *B.
-gracilis*, 10-14% for *H.
-comata*, 0.25-8% for *P.
-smithii*, and 4-17% for *P.
-secunda*.
-However, the forecasts were extremely uncertain (Figure 1).
-The IPM produced more certain forecasts than the QBM, as expected based on the validation results (Table 1).
-But even for the IPM forecasts, predicted percent changes in median equilibrium cover ranged from -200% to 200% (Figure 1).
-Forecast uncertainty did not vary appreciably among species.
-
-The population responses of species as forecast by the IPM is a function of the combined effects of individual vital rate sensitivities to climate and the contribution of each vital to population dynamics.
-For *B.
-gracilis*, all vital rates (survival, growth, and rectruitment) were positively affected by increases in precipitation and temperature (Figure 2).
-Intuitively, this leads to an overall increase in *B.
-gracilis* cover (Table 2).
-
-Precipitation increase resulted in all vital rates contributing to an increase in simulated cover for *H.
-comata*.
-But there is a small negative effect of temperature on recruitment.
-THIS IS WEIRD!!!
-CHECK SIMULATIONS.
-
-*P smithii* responded negatively to an increase in precipitation, temperature, or both regardless of the vital rate through which perturbed climate was acting.
-The effects were small, however, leading to relatively minor population consequences under climate change that did not vary much across scenarios (Table 2).
-
-For *B.
-gracilis*, the IPM predicted a modest increase in cover with a 1% increase in the mean of precipitation or a 1% increase in the mean of temperature, the compounding effect of both being a 20% increase in cover.
-This reflects the relatively strong effects of precipitation and climate on *B.
-gracilis* genet growth and recruitment (Figure 1).
-The QBM also predicted increased *B.
-gracilis* cover with a precipitation increase, but increasing temperature decreased equilibrium cover (Figure 3).
-
-The IPM and QBM produced consistent predictions for *H.
-comata* under increased precipitation and when both precipitation and temperature were increased (Figure 3).
-However, the IPM predicted more modest changes than the QBM, and for a temperature increase the two models differed: the IPM predicted an increase in cover while the QBM predicted the opposite.
 
@@
 
-# mae_list <- list()
-# mae_pvals <- list()
-# species <- unique(qbmstats$species)
-# for(spp in species){
-#   tmp1 <- subset(tmpipm, species==spp)
-#   tmp2 <- subset(tmpqbm, species==spp)
-#   err1 <- as.data.frame(abs(tmp1$mean_obs - tmp1$mean_pred))
-#   err2 <- as.data.frame(abs(tmp2$mean_obs - tmp2$mean_pred))
-#   colnames(err1) <- "resid"
-#   err1$model <- "ipm"
-#   colnames(err2) <- "resid"
-#   err2$model <- "qbm"
-#   errd <-  rbind(err1, err2)
-#   mae_ttest <- with(errd, t.test(resid~model, alternative = "less"))
-#   mae_list[[spp]] <- mae_ttest
-#   mae_pvals[[spp]] <- mae_ttest[["p.value"]]
-# }
-# 
-# mae_ps <- melt(mae_pvals)
+mae_list <- list()
+mae_pvals <- list()
+species <- unique(qbmstats$species)
+for(spp in species){
+  tmp1 <- subset(avg_ipm, species==spp)
+  tmp2 <- subset(avg_qbm, species==spp)
+  err1 <- as.data.frame(abs(tmp1$observation - tmp1$avg_prediction))
+  err2 <- as.data.frame(abs(tmp2$observation - tmp2$avg_prediction))
+  colnames(err1) <- "resid"
+  err1$model <- "ipm"
+  colnames(err2) <- "resid"
+  err2$model <- "qbm"
+  errd <-  rbind(err1, err2)
+  mae_ttest <- with(errd, t.test(resid~model, alternative = "less"))
+  mae_list[[spp]] <- mae_ttest
+  mae_pvals[[spp]] <- mae_ttest[["p.value"]]
+}
+
+mae_ps <- melt(mae_pvals)
 
@@
                               "quant_dist","mean_cover")]
-# comment          <- list()
-# comment$pos      <- list()
-# comment$pos[[1]] <- c(nrow(stats_table))
-# comment$command  <- c(paste("\\hline \n",  # we`ll replace all default hlines with this and the ones below
-#                             "The IPM MAE is significantly lower for HECO ($p = 3.3\times10^{-10}$) and POSE ($p = 0.00013$. MAEs are statisticially similar between models for BOGR and PASM.   \n",
-#                             sep = ""))
 
 colnames(stats_table) <- c("Species", "Model", "MAE", "90% Distance", "Mean Obs. Cover")
-print.xtable(xtable(stats_table, caption = "Mean absolute error (MAE) and accuracy (Pearson's $\\rho$) for one-step-ahead forecast from both model types.
+print.xtable(xtable(stats_table, caption = "Accuracy (mean absoluet error, MAE) and precision (90\\% Distance) of out of sample predictions.
 Forecasts were made without random year effects; only climate covariates could explain year-to-year variation.
@@
 ```
-The IPM MAE is significantly lower for HECO ($p = 3.3 \times 10^{-10}$) and POSE ($p = 0.00013$). MAEs are statisticially similar between models for BOGR and PASM.
+*NOTES*: The IPM MAE is significantly lower at $\alpha=0.05$ for *B.
+gracilis* (*P* = `r mae_ps[1,"value"]`), *H.
+comata* (*P* = `r mae_ps[2,"value"]`), and *P.
+smithii* (*P* = `r mae_ps[3,"value"]`).
+MAEs are statisticially similar between models for *P.
+secunda* (*P* = `r mae_ps[4,"value"]`).
 

### 2015-06-16 manuscript/manuscript.tex
updates ms with MAE statistics and more text on validation methods

--- 2015-06-15 manuscript/manuscript.tex
@@
 
-\section{Do we need detailed demographic data to forecast population
-responses to climate change?
-Probably.}\label{do-we-need-detailed-demographic-data-to-forecast-population-responses-to-climate-change-probably.}
+\section{Microscale models outperform mesoscale models when forecasting
+climate change impacts on plant
+populations}\label{microscale-models-outperform-mesoscale-models-when-forecasting-climate-change-impacts-on-plant-populations}
 
@@
 
-All R code and data necessary to replicate our analysis is archived on
+All R code and data necessary to reproduce our analysis is archived on
 GitHub as release 1.0
@@
 We used random draws from the final 1,000 iterations from each of three
-MCMC chains to introduce stochasticity into our population models. We
-thinned these chains by 10 to reduce potential autocorrelation, leaving
-us with 300 samples for each parameter. At each time step, we randomly
-selected climate covariates from one of the 14 observed years. Then, we
-drew the full parameter set (climate effects and density-dependence
-fixed effects) from a randomly selected MCMC iteration. Using this
-approach, rather than simply using coefficient point estimates, ensures
-that relatively unimportant climate covariates (those that broadly
-overlap 0) have little effect on the simulation results. Since our focus
-was on the contribution of climate covariates to population states, we
-set the random year effects and the random group effects to zero.
+MCMC chains to introduce stochasticity into our population models. At
+each time step, we randomly selected climate covariates from one of the
+14 observed years. Then, we drew the full parameter set (climate effects
+and density-dependence fixed effects) from a randomly selected MCMC
+iteration. Using this approach, rather than simply using coefficient
+point estimates, ensures that relatively unimportant climate covariates
+(those that broadly overlap 0) have little effect on the simulation
+results. Since our focus was on the contribution of climate covariates
+to population states, we set the random year effects and the random
+group effects to zero.
 
@@
 To test each model's ability to forecast the population state we used
-leave-one-year-out cross validation. For both levels of modeling, we fit
-the vital rate models using observations from all years except one, and
-then used those fitted parameters in the population models to perform a
-one-step-ahead forecast for the year whose observations were withheld
-from model fitting. We repeated this procedure for all 13 observation
-years. This model validation allowed us to compare accuracy and
-precision of the two modeling approaches (individual-level versus
-population-level).
+out of sample predictions using leave-one-year-out cross validation. For
+both levels of modeling, we fit the vital rate models using observations
+from all years except one, and then used those fitted parameters in the
+population models to perform a one-step-ahead forecast for the year
+whose observations were withheld from model fitting. Within each
+observation year, several quadrats are sampled. So we made predictions
+for each observed quadrat in the focal year initialized with cover the
+previous year. Since we were making quadrat specific predictions we
+incorporated the group effect on the intercept for both models. We
+repeated this procedure for all 13 observation years, making 100
+one-step-ahead forecasts for each quadrat-year combination with
+parameter uncertainty included via randomdrawd from the MCMC chain as
+described above. Random year effects were set to zero since year effects
+cannot be assigned to unobserved years.
+
+This model validation allowed us to compare accuracy and precision of
+the two modeling approaches (individual-level versus population-level).
+We first calculated the median predicted cover across the 100
+simulations for each quadrat-year and then calculated the absolute error
+as the difference between the observed cover for a given quadrat-year
+and the median prediction. To arrive at mean absolute error (MAE), we
+then averaged the absolute error within each species across the
+quadrat-year specific errors. We use MAE as our measure of accuracy. To
+measure precision calculated the distance between the upper and lower
+90th quantiles of the 100 predictions as a measure of forecast precision
+and averaged this value over quadrat-years for each species.
+
+\subsubsection{Testing sensitivity to climate
+covariates}\label{testing-sensitivity-to-climate-covariates}
+
+Our main goal in this paper is to see if models based on aggregate level
+data are as sensitive to climate as models based on individual level
+data. So, with our fitted and validated models in hand, we ran
+simulations for each model type (IPM and QBM) under four climate
+perturbation scenarios: (1) observed climate, (2) precipitation
+increased by 1\%, (3) temperature increased by 1\%, and (4)
+precipitation and temperature increased by 1\%. We ran this simulations
+for 2,500 time steps, enough to estimate equilibrium cover after
+discarding an initial 500 time steps as burn-in. Each simulation was run
+under two parameter scenarios: (1) using mean parameter estimates and
+(2) using randomly drawn parameters from the MCMC chain. We use (1) to
+detect the overall sensitivity of equilibrium cover to climate, and we
+use (2) to show the impact of model uncertainty on forecast precision.
+
+As an effort to identify potential discrepencies between IPM and QBM
+forecasts, we also ran simulations designed to quantify the
+sensitivities of individual and combined vital rates to climate for the
+IPM. Specifically, we ran simulations for the above climate scenarios,
+but applied the perturbed climate covariates to survival, growth, and
+recruitment vital rates individually and in pairwise combinations. This
+allows us to isolate the vital rate(s) most sensitive to climate. For
+this analysis, we used mean parameter estimates to reduce the sources of
+uncertainty in the sensitivity estimates.
 
@@
 
-We performed one-step-ahead forecasts to compare the two models (IPM
-vs.~QBM). We did not use the random year effects for these forecasts.
-Thus, beyond the effect of genet size or quadrat cover, only the climate
-covariates could affect expect plant cover. The IPM had significantly
-lower overall error (MAE, mean absolute error) for two species (\emph{H.
-comata} and \emph{P. smithii}; Table 1). In no case did the QBM
-significantly outperform the IPM (Table 1). As a metric of model
-uncertainty, we calculated the euclidean distance between the upper and
-lower 90\% quantiles for each model's forecasts. The IPM always
-outperformed the QBM in this respect (Table 1), with consistently lower
-distances between the 90\% quantiles. In general the IPM outperformed
-the QBM because it had (i) lower MAE for two species, (ii) statistically
-similar MAE for the other two species, and (iii) considerably more
-precise forecasts for all species.
+The IPM had significantly lower overall error (MAE, mean absolute error)
+for two species (\emph{H. comata} and \emph{P. smithii}; Table 1). In no
+case did the QBM significantly outperform the IPM (Table 1). The IPM was
+consistently more precise than the QBM, with lower distances between the
+90\% quantiles across all species (Table 1). In general the IPM
+outperformed the QBM because it had (i) lower MAE for two species, (ii)
+statistically similar MAE for the other two species, and (iii)
+considerably more precise forecasts for all species.
 
@@
 Impacts}\label{forecasting-climate-change-impacts}
-
-Simulated equilibrium cover simulated from the models was sensitive to
-climate perturbations, but the IPM and QBM produced inconsistent results
-(Table 2). For the IPM, proportional changes range from about 1-25\% for
-\emph{B. gracilis}, 7-25\% for \emph{H. comata}, 8\% for \emph{P.
-smithii}, and 2-20\% for \emph{P. secunda}. For the QBM, proportional
-changes due to climate perturbation ranged from 20-100\% for \emph{B.
-gracilis}, 10-14\% for \emph{H. comata}, 0.25-8\% for \emph{P. smithii},
-and 4-17\% for \emph{P. secunda}. However, the forecasts were extremely
-uncertain (Figure 1). The IPM produced more certain forecasts than the
-QBM, as expected based on the validation results (Table 1). But even for
-the IPM forecasts, predicted percent changes in median equilibrium cover
-ranged from -200\% to 200\% (Figure 1). Forecast uncertainty did not
-vary appreciably among species.
-
-The population responses of species as forecast by the IPM is a function
-of the combined effects of individual vital rate sensitivities to
-climate and the contribution of each vital to population dynamics. For
-\emph{B. gracilis}, all vital rates (survival, growth, and rectruitment)
-were positively affected by increases in precipitation and temperature
-(Figure 2). Intuitively, this leads to an overall increase in \emph{B.
-gracilis} cover (Table 2).
-
-Precipitation increase resulted in all vital rates contributing to an
-increase in simulated cover for \emph{H. comata}. But there is a small
-negative effect of temperature on recruitment. THIS IS WEIRD!!! CHECK
-SIMULATIONS.
-
-\emph{P smithii} responded negatively to an increase in precipitation,
-temperature, or both regardless of the vital rate through which
-perturbed climate was acting. The effects were small, however, leading
-to relatively minor population consequences under climate change that
-did not vary much across scenarios (Table 2).
-
-For \emph{B. gracilis}, the IPM predicted a modest increase in cover
-with a 1\% increase in the mean of precipitation or a 1\% increase in
-the mean of temperature, the compounding effect of both being a 20\%
-increase in cover. This reflects the relatively strong effects of
-precipitation and climate on \emph{B. gracilis} genet growth and
-recruitment (Figure 1). The QBM also predicted increased \emph{B.
-gracilis} cover with a precipitation increase, but increasing
-temperature decreased equilibrium cover (Figure 3).
-
-The IPM and QBM produced consistent predictions for \emph{H. comata}
-under increased precipitation and when both precipitation and
-temperature were increased (Figure 3). However, the IPM predicted more
-modest changes than the QBM, and for a temperature increase the two
-models differed: the IPM predicted an increase in cover while the QBM
-predicted the opposite.
 
@@
 (\emph{e.g.} Freckleton et al. 2011).
+
+Our analysis offers good news and bad news. The good news is that for
+two out of our four focal species, the QBM faired no worse than the
+demographically detailed IPM in terms of mean absolute (Table 1). Thus,
+it seems that aggregate level data can provide forecasts of similar
+skill as individual level data.
 
@@
 \centering
-\caption{Mean absolute error (MAE) and accuracy (Pearson's $\rho$) for one-step-ahead forecast from both model types.
+\caption{Accuracy (mean absoluet error, MAE) and precision (90\% Distance) of out of sample predictions.
 Forecasts were made without random year effects; only climate covariates could explain year-to-year variation.
-90 Distance refers to the average distance between the upper and lower 90 percentiles of the 100 predicted values for each quadrat-year combination.}
+90\% Distance refers to the average distance between the upper and lower 90th percentiles of the 100 predicted values for each quadrat-year combination.}
 \begin{tabular}{llrrr}
@@
 
-The IPM MAE is significantly lower for HECO ($p = 3.3 \times 10^{-10}$)
-and POSE ($p = 0.00013$). MAEs are statisticially similar between models
-for BOGR and PASM.
+\emph{NOTES}: The IPM MAE is significantly lower at $\alpha=0.05$ for
+\emph{B. gracilis} (\emph{P} = 0.0012), \emph{H. comata} (\emph{P} =
+4.0586 × 10-8), and \emph{P. smithii} (\emph{P} = 3.183 × 10-5). MAEs
+are statisticially similar between models for \emph{P. secunda}
+(\emph{P} = 0.0922).
 

### 2015-06-16 manuscript/manuscript.Rmd
fixes a few typos

@@
 ### Model validation
-To test each model's ability to forecast the population state we used out of sample predictions using leave-one-year-out cross validation.
+To test each model's ability to forecast the population state we made out of sample predictions using leave-one-year-out cross validation.
 For both levels of modeling, we fit the vital rate models using observations from all years except one, and then used those fitted parameters in the population models to perform a one-step-ahead forecast for the year whose observations were withheld from model fitting.
@@
 We use MAE as our measure of accuracy.
-To measure precision calculated the distance between the upper and lower 90th quantiles of the 100 predictions as a measure of forecast precision and averaged this value over quadrat-years for each species.
+To measure precision we calculated the distance between the upper and lower 90th quantiles of the 100 predictions and averaged this value over quadrat-years for each species.
 
@@
 So, with our fitted and validated models in hand, we ran simulations for each model type (IPM and QBM) under four climate perturbation scenarios: (1) observed climate, (2) precipitation increased by 1%, (3) temperature increased by 1%, and (4) precipitation and temperature increased by 1%.
-We ran this simulations for 2,500 time steps, enough to estimate equilibrium cover after discarding an initial 500 time steps as burn-in.
+We ran the simulations for 2,500 time steps, enough to estimate equilibrium cover after discarding an initial 500 time steps as burn-in.
 Each simulation was run under two parameter scenarios: (1) using mean parameter estimates and (2) using randomly drawn parameters from the MCMC chain.
@@
 -------
-### Comparison of Forecast Models
-The IPM had significantly lower overall error (MAE, mean absolute error) for two species (*H.
-comata* and *P.
+### Comparison of forecast models
+The IPM had significantly lower overall error (MAE, mean absolute error) for three species (*B.
+gracilis*, *H.
+comata*, *P.
 smithii*; Table 1).
@@
 The IPM was consistently more precise than the QBM, with lower distances between the 90% quantiles across all species (Table 1).
-In general the IPM outperformed the QBM because it had (i) lower MAE for two species, (ii) statistically similar MAE for the other two species, and (iii) considerably more precise forecasts for all species.
+In general the IPM outperformed the QBM because it had (1) lower MAE for three of the four species, (2) statistically similar MAE for the one other species, and (3) considerably more precise forecasts for all species.
 
-### Forecasting Climate Change Impacts
+### Sensitivity of models to climate
 

### 2015-06-16 manuscript/manuscript.tex
fixes a few typos

@@
 
-To test each model's ability to forecast the population state we used
+To test each model's ability to forecast the population state we made
 out of sample predictions using leave-one-year-out cross validation. For
@@
 quadrat-year specific errors. We use MAE as our measure of accuracy. To
-measure precision calculated the distance between the upper and lower
-90th quantiles of the 100 predictions as a measure of forecast precision
-and averaged this value over quadrat-years for each species.
+measure precision we calculated the distance between the upper and lower
+90th quantiles of the 100 predictions and averaged this value over
+quadrat-years for each species.
 
@@
 
-\subsubsection{Comparison of Forecast
-Models}\label{comparison-of-forecast-models}
+\subsubsection{Comparison of forecast
+models}\label{comparison-of-forecast-models}
 
 The IPM had significantly lower overall error (MAE, mean absolute error)
-for two species (\emph{H. comata} and \emph{P. smithii}; Table 1). In no
-case did the QBM significantly outperform the IPM (Table 1). The IPM was
-consistently more precise than the QBM, with lower distances between the
-90\% quantiles across all species (Table 1). In general the IPM
-outperformed the QBM because it had (i) lower MAE for two species, (ii)
-statistically similar MAE for the other two species, and (iii)
-considerably more precise forecasts for all species.
+for three species (\emph{B. gracilis}, \emph{H. comata}, \emph{P.
+smithii}; Table 1). In no case did the QBM significantly outperform the
+IPM (Table 1). The IPM was consistently more precise than the QBM, with
+lower distances between the 90\% quantiles across all species (Table 1).
+In general the IPM outperformed the QBM because it had (1) lower MAE for
+three of the four species, (2) statistically similar MAE for the one
+other species, and (3) considerably more precise forecasts for all
+species.
 
-\subsubsection{Forecasting Climate Change
-Impacts}\label{forecasting-climate-change-impacts}
+\subsubsection{Sensitivity of models to
+climate}\label{sensitivity-of-models-to-climate}
 

### 2015-06-16 manuscript/manuscript.Rmd
fixes some more typos

@@
 
+Acknowledgments
+---------------
+This work was funded by the National Science Foundation through a Postdoctoral Research Fellowship in Biology to ATT (DBI-1400370) and a CAREER award to PBA (DEB-1054040).
+We thank the original mappers of the permanent quadrats in Montana and the digitizers in the Adler lab, without whom this work would not have been possible.
+Informal conversations with Stephen Ellner, Giles Hooker, Robin Snyder, and a series of meetings among the Adler and Weecology labs at USU sharpened our thinking.
+
 \pagebreak{}
@@
 colnames(stats_table) <- c("Species", "Model", "MAE", "90% Distance", "Mean Obs. Cover")
-print.xtable(xtable(stats_table, caption = "Accuracy (mean absoluet error, MAE) and precision (90\\% Distance) of out of sample predictions.
+print.xtable(xtable(stats_table, caption = "Accuracy (mean absolute error, MAE) and precision (90\\% Distance) of out of sample predictions.
 Forecasts were made without random year effects; only climate covariates could explain year-to-year variation.

### 2015-06-16 manuscript/manuscript.tex
fixes some more typos

@@
 increased by 1\%, (3) temperature increased by 1\%, and (4)
-precipitation and temperature increased by 1\%. We ran this simulations
+precipitation and temperature increased by 1\%. We ran the simulations
 for 2,500 time steps, enough to estimate equilibrium cover after
@@
 20-25 years needed!
+
+\subsection{Acknowledgments}\label{acknowledgments}
+
+This work was funded by the National Science Foundation through a
+Postdoctoral Research Fellowship in Biology to ATT (DBI-1400370) and a
+CAREER award to PBA (DEB-1054040). We thank the original mappers of the
+permanent quadrats in Montana and the digitizers in the Adler lab,
+without whom this work would not have been possible. Informal
+conversations with Stephen Ellner, Giles Hooker, Robin Snyder, and a
+series of meetings among the Adler and Weecology labs at USU sharpened
+our thinking.
 

### 2015-06-16 manuscript/manuscript.Rmd
adds to discussion

@@
 ----------
-We sought to test the assumption that the sensitivities of plant populations to climate variables can be detected equally well using either individual level data or population level data.
-This is an important question to answer because population models are key tools for predicting the consequences of global climate change.
-However, they can be of limited use when built on data from a small subset of a population in space or time.
-If population level data (i.e., some aggregated form of individual level data) can be used to detect climate effects on population dynamics, then we would have a cheaper and easier option for data collection over relatively large temporal and spatial extents [_e.g._ @Freckleton2011].
+Demographic models are costly to parameterize, yet remain the most widely-used tools for studying plant population dynamics (e.g., @Salguero-Gomez2015).
+New density-structured approaches sacrifice some precision at the local scale for better spatial coverage that is virtually unattainable, due to time and funding constraints, using traditional individual-based data [@Freckleton2011; @Queenborough2011].
+These approaches are appealing because traditional population models suffer from constrained parameter space -- meaning forecasts quickly extend beyond the parameter space used to fit the model.
+Density-structured approaches rely on easy-to-collect census data (e.g., percent cover in $1\text{m}^2$ quadrats) so sampling populations across large spatial extents would be possible, thus covering more potential parameter space.
 
-Our analysis offers good news and bad news.
-The good news is that for two out of our four focal species, the QBM faired no worse than the demographically detailed IPM in terms of mean absolute (Table 1).
-Thus, it seems that aggregate level data can provide forecasts of similar skill as individual level data.
+Estimating parameter variability in space and time is critical when using population models to forecast responses to directional changes in exogoneous drivers like climate [@Clark2011; @Vanderwel2013; @Wolkovich2014a; @Petchey2015].
+Compared to demographic models, density-structured modeling approaches may provide an efficient route toward estimating such uncertainty.
+However, population models based on aggregated inidividual data have yet to be tested with climate drivers.
+Since it is individuals that respond to weather, rather than populations, models based on a aggregated data may produce incorrect forecasts [@Clark2012].
+Thus, we sought to compare climate sensitivities of individual and population based models of four grassland species populations.
 

### 2015-06-16 manuscript/manuscript.tex
adds to discussion

@@
 
-We use an environmentally stochastic IPM (Rees and Ellner 2009) that
+We use an environmentally stochastic IPM (Mark and Ellner 2009) that
 includes the random year effects and the climate covariates from the
@@
 
-We sought to test the assumption that the sensitivities of plant
-populations to climate variables can be detected equally well using
-either individual level data or population level data. This is an
-important question to answer because population models are key tools for
-predicting the consequences of global climate change. However, they can
-be of limited use when built on data from a small subset of a population
-in space or time. If population level data (i.e., some aggregated form
-of individual level data) can be used to detect climate effects on
-population dynamics, then we would have a cheaper and easier option for
-data collection over relatively large temporal and spatial extents
-(\emph{e.g.} Freckleton et al. 2011).
+Demographic models are costly to parameterize, yet remain the most
+widely-used tools for studying plant population dynamics (e.g.,
+Salguero-Gómez et al. (2015)). New density-structured approaches
+sacrifice some precision at the local scale for better spatial coverage
+that is virtually unattainable, due to time and funding constraints,
+using traditional individual-based data (Freckleton et al. 2011,
+Queenborough et al. 2011). These approaches are appealing because
+traditional population models suffer from constrained parameter space --
+meaning forecasts quickly extend beyond the parameter space used to fit
+the model. Density-structured approaches rely on easy-to-collect census
+data (e.g., percent cover in \(1\text{m}^2\) quadrats) so sampling
+populations across large spatial extents would be possible, thus
+covering more potential parameter space.
 
-Our analysis offers good news and bad news. The good news is that for
-two out of our four focal species, the QBM faired no worse than the
-demographically detailed IPM in terms of mean absolute (Table 1). Thus,
-it seems that aggregate level data can provide forecasts of similar
-skill as individual level data.
+Estimating parameter variability in space and time is critical when
+using population models to forecast responses to directional changes in
+exogoneous drivers like climate (Clark et al. 2011, Vanderwel et al.
+2013, Wolkovich et al. 2014, Petchey et al. 2015). Compared to
+demographic models, density-structured modeling approaches may provide
+an efficient route toward estimating such uncertainty. However,
+population models based on aggregated inidividual data have yet to be
+tested with climate drivers. Since it is individuals that respond to
+weather, rather than populations, models based on a aggregated data may
+produce incorrect forecasts (Clark et al. 2012). Thus, we sought to
+compare climate sensitivities of individual and population based models
+of four grassland species populations.
 
@@
 \centering
-\caption{Accuracy (mean absoluet error, MAE) and precision (90\% Distance) of out of sample predictions.
+\caption{Accuracy (mean absolute error, MAE) and precision (90\% Distance) of out of sample predictions.
 Forecasts were made without random year effects; only climate covariates could explain year-to-year variation.
@@
 
-\emph{NOTES}: The IPM MAE is significantly lower at $\alpha=0.05$ for
+\emph{NOTES}: The IPM MAE is significantly lower at \(\alpha=0.05\) for
 \emph{B. gracilis} (\emph{P} = 0.0012), \emph{H. comata} (\emph{P} =
@@
 
+Clark, J. S., D. M. Bell, M. H. Hersh, and L. Nichols. 2011. Climate
+change vulnerability of forest biodiversity: Climate and competition
+tracking of demographic rates. Global Change Biology 17:1834--1849.
+
 Clark, J. S., D. M. Bell, M. Kwit, A. Stine, B. Vierra, and K. Zhu.
@@
 
+Mark, R., and S. P. Ellner. 2009. Integral projection models for
+populations in temporally varying environments. Ecological Monographs
+79:575--594.
+
+Petchey, O. L., M. Pontarp, T. M. Massie, S. Kéfi, A. Ozgul, M.
+Weilenmann, G. M. Palamara, F. Altermatt, B. Matthews, J. M. Levine, D.
+Z. Childs, B. J. McGill, M. E. Schaepman, B. Schmid, P. Spaak, A. P.
+Beckerman, F. Pennekamp, and I. S. Pearse. 2015. The ecological forecast
+horizon, and examples of its uses and determinants. Ecology Letters
+18:597--611.
+
 Piantadosi, S., D. P. Byar, and S. B. Green. 1988. The Ecological
@@
 statistical computing.
-
-Rees, M., and S. P. Ellner. 2009. Integral projection models for
-populations in temporally varying environments. Ecological Monographs
-79:575--594.
 
@@
 
+Vanderwel, M. C., V. S. Lyutsarev, and D. W. Purves. 2013.
+Climate-related variation in mortality and recruitment determine
+regional forest-type distributions. Global Ecology and Biogeography
+22:1192--1203.
+
+Wolkovich, E. M., B. I. Cook, K. K. McLauchlan, and T. J. Davies. 2014.
+Temporal ecology in the Anthropocene. Ecology Letters 17:1365--1379.
+
 \end{document}

### 2015-06-16 manuscript/manuscript.Rmd
adds results section; completed draft

@@
 
-In this paper, we model populations based on two levels of data: individual and quadrat (Figure 1).
+In this paper, we model populations based on two levels of data: individual and quadrat (Fig.
+1).
 The individual data is the "raw" data.
@@
 ### Sensitivity of models to climate
+```{r get_results, echo=FALSE, eval=TRUE}
+setwd("~/Repos/MicroMesoForecast/analysis/")
+
+####
+##  Mean parameter runs
+####
+meanparam_ipm <- data.frame(cover=NA, species=NA, 
+                             climsim=NA, model=NA)
+##  IPM 
+path <- "./ipm/simulations/results/climchange_meanparams/"
+ipmfiles_mean <- list.files(path)
+ipmfiles_mean <- ipmfiles_mean[grep("cover", ipmfiles_mean)]
+for(file in ipmfiles_mean){
+  tmp <- read.csv(paste(path,file,sep=""))
+  tmp$model <- "IPM"
+  meanparam_ipm <- rbind(meanparam_ipm, tmp[,c("cover", "species", "climsim", "model")])
+}
+meanparam_ipm <- meanparam_ipm[2:nrow(meanparam_ipm),]
+
+newd <- data.frame(cover=NA, species=NA, 
+                   climsim=NA, model=NA, obscover=NA)
+for(dospp in unique(meanparam_ipm$species)){
+  tmp <- subset(meanparam_ipm, species==dospp)
+  obscover <- subset(tmp, climsim=="obs")["cover"]
+  tmp2 <- subset(tmp, climsim!="obs")
+  tmp2$obscover <- rep(obscover$cover, times = length(unique(tmp2$climsim)))
+  newd <- rbind(newd, tmp2)
+}
+ipm_meansims <- newd[2:nrow(newd),]
+ipm_meansims$propdiff <- with(ipm_meansims, log(cover*100)-log(obscover*100))
+
+##  QBM
+meanparam_qbm <- data.frame(cover=NA, species=NA, 
+                             climsim=NA, model=NA)
+path <- "./quadBM/simulations/results/climatechange_meanparams/"
+qbmfiles_mean <- list.files(path)
+for(file in qbmfiles_mean){
+  tmp <- readRDS(paste(path,file,sep=""))
+  tmp$model <- "QBM"
+  meanparam_qbm <- rbind(meanparam_qbm, tmp[,c("cover", "species", "climsim", "model")])
+}
+meanparam_qbm <- meanparam_qbm[2:nrow(meanparam_qbm),]
+
+newd <- data.frame(cover=NA, species=NA, 
+                   climsim=NA, model=NA, obscover=NA)
+for(dospp in unique(meanparam_qbm$species)){
+  tmp <- subset(meanparam_qbm, species==dospp)
+  obscover <- subset(tmp, climsim=="obs")["cover"]
+  tmp2 <- subset(tmp, climsim!="obs")
+  tmp2$obscover <- rep(obscover$cover, times = length(unique(tmp2$climsim)))
+  newd <- rbind(newd, tmp2)
+}
+qbm_meansims <- newd[2:nrow(newd),]
+qbm_meansims$propdiff <- with(qbm_meansims, log(cover*100)-log(obscover*100))
+
+##  Merge IPM and QBM mean parameter results
+bothmodel_outs <- rbind(ipm_meansims, qbm_meansims)
+meanparam_agg <- ddply(bothmodel_outs, .(species, climsim, model), summarise,
+                       meandiff = mean(propdiff),
+                       hidiff = quantile(propdiff, probs = 0.95),
+                       lodiff = quantile(propdiff, probs = 0.05))
+setwd("~/Repos/MicroMesoForecast/manuscript")
+```
+
+Equilibrium cover from both models was sensitive to climate (Fig.
+2a-d).
+The IPM projected percent changes in equilibrium cover from `r round(min(subset(meanparam_agg, species=="BOGR" & model=="IPM")["meandiff"])*100)` - `r round(max(subset(meanparam_agg, species=="BOGR" & model=="IPM")["meandiff"])*100)`% for *B.
+gracilis*, `r round(min(subset(meanparam_agg, species=="HECO" & model=="IPM")["meandiff"])*100)` - `r round(max(subset(meanparam_agg, species=="HECO" & model=="IPM")["meandiff"])*100)`% for *H.
+comata*, `r round(min(subset(meanparam_agg, species=="PASM" & model=="IPM")["meandiff"])*100)` - `r round(max(subset(meanparam_agg, species=="PASM" & model=="IPM")["meandiff"])*100)`% for *P.
+smithii*, and `r round(min(subset(meanparam_agg, species=="POSE" & model=="IPM")["meandiff"])*100)` - `r round(max(subset(meanparam_agg, species=="POSE" & model=="IPM")["meandiff"])*100)`% for *P.
+secunda*.
+The QBM projected opposite and greater percent changes in equilibrium cover for *B.
+gracilis* (`r round(min(subset(meanparam_agg, species=="BOGR" & model=="QBM")["meandiff"])*100)` - `r round(max(subset(meanparam_agg, species=="BOGR" & model=="QBM")["meandiff"])*100)`%) and *H.
+comata* (`r round(min(subset(meanparam_agg, species=="HECO" & model=="QBM")["meandiff"])*100)` - `r round(max(subset(meanparam_agg, species=="HECO" & model=="QBM")["meandiff"])*100)`%; Fig.
+2a-b).
+For *P.
+smithii*, the QBM projected opposite changes in equilibrium cover than the IPM, but of similar magnitude (`r round(min(subset(meanparam_agg, species=="PASM" & model=="QBM")["meandiff"])*100)` - `r round(max(subset(meanparam_agg, species=="PASM" & model=="QBM")["meandiff"])*100)`%; Fig.
+2c).
+*P.
+secunda* was the only species that the IPM and QBM made projections of the same sign and somewhat similar magnitude (Fig.
+2d).
+As expected based on model validation (Table 1), IPM projections were more uncertain than QBM projections for all species and all climate change scenarios (Fig.
+2e-h).
+
+The response of a population to climate change is a result of the aggregate effects of climate on individual vital rates.
+Since the IPM approach relies on vital rate regressions, we were able to quantify the sensitivity of each vital rate in isolation and in pairwise combinations.
+Species showed similar trends (Fig.
+3).
+Growth was the most sensitive vital rate for all species, showing a negative response to increased precipitation, and stronger positive response to increased temperature, and a mostly positive response when both climate factors are increased (Fig.
+3).
+*B.
+gracilis* survival rates were sensitive to temperature, resulting in an increase in plant cover under increased temperature (Fig.
+3a).
+In isolation, recruitment and survival were insensitive to climate factors for *H.
+comata* (Fig.
+3b).
+Survival and recruitment of *P.
+smithii* were both sensitive, negatively, to temperature and precipitation (Fig.
+3c).
+*P.
+secunda* equilibrium cover was sensitive to the climate effects on survival and recruitment, showing a negative effect on both vital rates for increased precipition, but a strong positive effect on survival with increased temperature (Fig.
+3d).
+The climate impact of recruitment on equilibrium cover was negative for precipitation and temperature increases (Fig.
+3d).
 
@@
 
-```{r figure_2, dependson="plot-options", fig.cap="Sensitivity of equilibrium cover to a 1% increase in precipitation (+ppt), temperature (+temp), or both (+ppt&temp) applied to individual and combined vital rates.
-For example, the points associated with G show the median cover from IPM simulations where a climate perturbation is applied only to the growth regression climate covariates.", fig.width=4.5, fig.height=6, cache=FALSE, results='hide'}
-setwd("../analysis/ipm/simulations/results/climate_sensitivity/")
-files <- list.files()
-cover_files <- files[grep("cover", files)]
-spp_id <- substr(cover_files, 1, 4)
-num_files <- length(cover_files)
-all_sims <- data.frame(time=NA, cover=NA, species=NA, climsim=NA, vital=NA)
-for(i in 1:num_files){
-  tmp <- read.csv(cover_files[i])
-  tmp$species <- rep(spp_id[i], nrow(tmp))
-  all_sims <- rbind(all_sims, tmp)
-}
-all_sims <- all_sims[2:nrow(all_sims),]
-
-# Rename vital rates with codes
-vital_names <- unique(all_sims$vital)
-vital_codes <- c("A", "G", "GR", "GS", "R", "S" , "SR") 
-vitals <- data.frame(names=vital_names,code=vital_codes)
-for(vnow in vital_names){
-  ids <- which(all_sims[,"vital"]==vnow)
-  idcode <- which(vitals[,"names"]==vnow)
-  all_sims[ids,"vital"] <- as.character(vitals[idcode,"code"])
-}
-
-# Rep the 'all' simulation for each vital rate for plotting
-all_clim_sims <- subset(all_sims, vital=="A")
-species <- unique(all_sims$species)
-out_controls <- data.frame(time=NA, cover=NA, species=NA, climsim=NA, vital=NA)
-for(spp in species){
-  tmp <- subset(all_clim_sims, species==spp)
-  for(vital in vital_codes){
-    tmp$vital <- vital
-    out_controls <- rbind(out_controls, tmp)
-  }
-}
-out_controls <- out_controls[2:nrow(out_controls),]
-
-# Combine the datasets
-all_noclim_sims <- subset(all_sims, vital!="all")
-final <- rbind(out_controls, all_noclim_sims)
-
-# Calculate statistics for plotting
-equilibrium_cover <- ddply(final, .(species, climsim, vital), summarise,
-                           eq_cover = median(cover))
-mean_cover <- ddply(subset(equilibrium_cover, climsim=="all"), .(species), summarise,
-                    value = mean(eq_cover))
-
-equilibrium_cover <- subset(equilibrium_cover, vital!="A")
-bothid <- which(equilibrium_cover[,"climsim"]=="ppttemp")
-equilibrium_cover[bothid,"climsim"] <- "zppttemp"
-# myCols2 <- c("#277BA8", "#7ABBBD", "#AED77A")
-ggplot(equilibrium_cover, aes(x=climsim, y=eq_cover*100, 
-                              shape=vital, group=vital, linetype=vital))+
-  geom_hline(data=mean_cover, aes(yintercept=value*100), linetype=2, color="grey45")+
-  geom_line()+
-  geom_point(size=4)+
-  facet_grid(species~., scales = "free")+
-  scale_linetype_discrete(name="Vital Rate")+
-  scale_shape_discrete(name="Vital Rate")+
-  scale_x_discrete(labels=c("baseline", "+ppt", "+temp", "+ppt&temp"))+
-  ylab("Equilibrium Cover (%)")+
-  xlab("Simulation")+
-  theme_bw()+
-  theme(axis.text.x = element_text(angle = 45, hjust = 1))
-
-covertest <- subset(equilibrium_cover, climsim=="all")
-colnames(covertest)[4] <- "obs_cover"
-out <- merge(subset(equilibrium_cover, climsim!="all"), covertest[,c("species","vital","obs_cover")])
-out$propdiff <- with(out, log(eq_cover*100)-log(obs_cover*100))
-setwd("~/Repos/MicroMesoForecast/manuscript")
-```
-
-```{r figure_3, dependson="plot-options", fig.cap="Proportional change in species' mean cover caused by a 1% increase in observed precipitation (+ppt), temperature (+temp), or both (+ppt&temp) as predicted by the individual-based IPM and the aggregate-based QBM.
+```{r figure_2, dependson="plot-options", fig.cap="Proportional change in species' mean cover caused by a 1% increase in observed precipitation (+ppt), temperature (+temp), or both (+ppt&temp) as predicted by the individual-based IPM and the aggregate-based QBM.
 Top panels show the mean predicted proportional change in cover; lower panels show the 90% quantiles of predicted proportional cover change.
@@
 
+```{r figure_3, dependson="plot-options", fig.cap="Sensitivity of equilibrium cover to a 1% increase in precipitation (+ppt), temperature (+temp), or both (+ppt&temp) applied to individual and combined vital rates.
+For example, the points associated with G show the median cover from IPM simulations where a climate perturbation is applied only to the growth regression climate covariates.", fig.width=4.5, fig.height=6, cache=FALSE, results='hide'}
+setwd("../analysis/ipm/simulations/results/climate_sensitivity/")
+files <- list.files()
+cover_files <- files[grep("cover", files)]
+spp_id <- substr(cover_files, 1, 4)
+num_files <- length(cover_files)
+all_sims <- data.frame(time=NA, cover=NA, species=NA, climsim=NA, vital=NA)
+for(i in 1:num_files){
+  tmp <- read.csv(cover_files[i])
+  tmp$species <- rep(spp_id[i], nrow(tmp))
+  all_sims <- rbind(all_sims, tmp)
+}
+all_sims <- all_sims[2:nrow(all_sims),]
+
+# Rename vital rates with codes
+vital_names <- unique(all_sims$vital)
+vital_codes <- c("A", "G", "GR", "GS", "R", "S" , "SR") 
+vitals <- data.frame(names=vital_names,code=vital_codes)
+for(vnow in vital_names){
+  ids <- which(all_sims[,"vital"]==vnow)
+  idcode <- which(vitals[,"names"]==vnow)
+  all_sims[ids,"vital"] <- as.character(vitals[idcode,"code"])
+}
+
+# Rep the 'all' simulation for each vital rate for plotting
+all_clim_sims <- subset(all_sims, vital=="A")
+species <- unique(all_sims$species)
+out_controls <- data.frame(time=NA, cover=NA, species=NA, climsim=NA, vital=NA)
+for(spp in species){
+  tmp <- subset(all_clim_sims, species==spp)
+  for(vital in vital_codes){
+    tmp$vital <- vital
+    out_controls <- rbind(out_controls, tmp)
+  }
+}
+out_controls <- out_controls[2:nrow(out_controls),]
+
+# Combine the datasets
+all_noclim_sims <- subset(all_sims, vital!="all")
+final <- rbind(out_controls, all_noclim_sims)
+
+# Calculate statistics for plotting
+equilibrium_cover <- ddply(final, .(species, climsim, vital), summarise,
+                           eq_cover = median(cover))
+mean_cover <- ddply(subset(equilibrium_cover, climsim=="all"), .(species), summarise,
+                    value = mean(eq_cover))
+
+equilibrium_cover <- subset(equilibrium_cover, vital!="A")
+bothid <- which(equilibrium_cover[,"climsim"]=="ppttemp")
+equilibrium_cover[bothid,"climsim"] <- "zppttemp"
+# myCols2 <- c("#277BA8", "#7ABBBD", "#AED77A")
+ggplot(equilibrium_cover, aes(x=climsim, y=eq_cover*100, 
+                              shape=vital, group=vital, linetype=vital))+
+  geom_hline(data=mean_cover, aes(yintercept=value*100), linetype=2, color="grey45")+
+  geom_line()+
+  geom_point(size=4)+
+  facet_grid(species~., scales = "free")+
+  scale_linetype_discrete(name="Vital Rate")+
+  scale_shape_discrete(name="Vital Rate")+
+  scale_x_discrete(labels=c("baseline", "+ppt", "+temp", "+ppt&temp"))+
+  ylab("Equilibrium Cover (%)")+
+  xlab("Simulation")+
+  theme_bw()+
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+
+covertest <- subset(equilibrium_cover, climsim=="all")
+colnames(covertest)[4] <- "obs_cover"
+out <- merge(subset(equilibrium_cover, climsim!="all"), covertest[,c("species","vital","obs_cover")])
+out$propdiff <- with(out, log(eq_cover*100)-log(obs_cover*100))
+setwd("~/Repos/MicroMesoForecast/manuscript")
+```
+
 \pagebreak{}

### 2015-06-16 manuscript/manuscript.tex
adds results section; completed draft

--- 2015-06-16 manuscript/manuscript.tex
@@
 In this paper, we model populations based on two levels of data:
-individual and quadrat (Figure 1). The individual data is the ``raw''
+individual and quadrat (Fig. 1). The individual data is the ``raw''
 data. For the quadrat level we data we simply sum individual areal cover
@@
 
+Equilibrium cover from both models was sensitive to climate (Fig. 2a-d).
+The IPM projected percent changes in equilibrium cover from -3 - 8\% for
+\emph{B. gracilis}, -4 - 3\% for \emph{H. comata}, -15 - 9\% for
+\emph{P. smithii}, and -17 - 53\% for \emph{P. secunda}. The QBM
+projected opposite and greater percent changes in equilibrium cover for
+\emph{B. gracilis} (-63 - 30\%) and \emph{H. comata} (-50 - -18\%; Fig.
+2a-b). For \emph{P. smithii}, the QBM projected opposite changes in
+equilibrium cover than the IPM, but of similar magnitude (-5 - 6\%; Fig.
+2C). \emph{P. secunda} was the only species that the IPM and QBM made
+projections of the same sign and somewhat similar magnitude (Fig. 2d).
+As expected based on model validation (Table 1), IPM projections were
+more uncertain than QBM projections for all species and all climate
+change scenarios (Fig. 2e-h).
+
+The response of a population to climate change is a result of the
+aggregate effects of climate on individual vital rates. Since the IPM
+approach relies on vital rate regressions, we were able to quantify the
+sensitivity of each vital rate in isolation and in pairwise
+combinations. Species showed similar trends (Fig. 3). Growth was the
+most sensitive vital rate for all species, showing a negative response
+to increased precipitation, and stronger positive response to increased
+temperature, and a mostly positive response when both climate factors
+are increased (Fig. 3). \emph{B. gracilis} survival rates were sensitive
+to temperature, resulting in an increase in plant cover under increased
+temperature (Fig. 3a). In isolation, recruitment and survival were
+insensitive to climate factors for \emph{H. comata} (Fig. 3b). Survival
+and recruitment of \emph{P. smithii} were both sensitive, negatively, to
+temperature and precipitation (Fig. 3c). \emph{P. secunda} equilibrium
+cover was sensitive to the climate effects on survival and recruitment,
+showing a negative effect on both vital rates for increased precipition,
+but a strong positive effect on survival with increased temperature
+(Fig. 3d). The climate impact of recruitment on equilibrium cover was
+negative for precipitation and temperature increases (Fig. 3d).
+
 \subsection{Discussion}\label{discussion}
 
-We sought to test the assumption that the sensitivities of plant
-populations to climate variables can be detected equally well using
-either individual level data or population level data. This is an
-important question to answer because population models are key tools for
-predicting the consequences of global climate change. However, they can
-be of limited use when built on data from a small subset of a population
-in space or time. If population level data (i.e., some aggregated form
-of individual level data) can be used to detect climate effects on
-population dynamics, then we would have a cheaper and easier option for
-data collection over relatively large temporal and spatial extents
-(\emph{e.g.} Freckleton et al. 2011).
+Demographic models are costly to parameterize, yet remain the most
+widely-used tools for studying plant population dynamics (e.g.,
+Salguero-Gómez et al. (2015)). New density-structured approaches
+sacrifice some precision at the local scale for better spatial coverage
+that is virtually unattainable, due to time and funding constraints,
+using traditional individual-based data (Freckleton et al. 2011,
+Queenborough et al. 2011). These approaches are appealing because
+traditional population models suffer from constrained parameter space --
+meaning forecasts quickly extend beyond the parameter space used to fit
+the model. Density-structured approaches rely on easy-to-collect census
+data (e.g., percent cover in $1\text{m}^2$ quadrats) so sampling
+populations across large spatial extents would be possible, thus
+covering more potential parameter space.
 
-Our analysis offers good news and bad news. The good news is that for
-two out of our four focal species, the QBM faired no worse than the
-demographically detailed IPM in terms of mean absolute (Table 1). Thus,
-it seems that aggregate level data can provide forecasts of similar
-skill as individual level data.
+Estimating parameter variability in space and time is critical when
+using population models to forecast responses to directional changes in
+exogoneous drivers like climate ({\textbf{???}}, {\textbf{???}},
+{\textbf{???}}, {\textbf{???}}). Compared to demographic models,
+density-structured modeling approaches may provide an efficient route
+toward estimating such uncertainty. However, population models based on
+aggregated inidividual data have yet to be tested with climate drivers.
+Since it is individuals that respond to weather, rather than
+populations, models based on a aggregated data may produce incorrect
+forecasts (Clark et al. 2012). Thus, we sought to compare climate
+sensitivities of individual and population based models of four
+grassland species populations.
 
@@
 \includegraphics{components/figure/manuscript-figure_2.pdf}
-\caption{Sensitivity of equilibrium cover to a 1\% increase in
-precipitation (+ppt), temperature (+temp), or both (+ppt\&temp) applied
-to individual and combined vital rates. For example, the points
-associated with G show the median cover from IPM simulations where a
-climate perturbation is applied only to the growth regression climate
-covariates.}
-\end{figure}
-
-\begin{figure}[htbp]
-\centering
-\includegraphics{components/figure/manuscript-figure_3.pdf}
 \caption{Proportional change in species' mean cover caused by a 1\%
@@
 uncertainties are shown in separate panels for visual clarity.}
+\end{figure}
+
+\begin{figure}[htbp]
+\centering
+\includegraphics{components/figure/manuscript-figure_3.pdf}
+\caption{Sensitivity of equilibrium cover to a 1\% increase in
+precipitation (+ppt), temperature (+temp), or both (+ppt\&temp) applied
+to individual and combined vital rates. For example, the points
+associated with G show the median cover from IPM simulations where a
+climate perturbation is applied only to the growth regression climate
+covariates.}
 \end{figure}
@@
 2012. Individual-scale inference to anticipate climate-change
-vulnerability of biodiversity.
+vulnerability of biodiversity. Philosophical Transactions of the Royal
+Society B: Biological Sciences 367:236--246.
 

### 2015-06-17 manuscript/manuscript.Rmd
fixes references

@@
 
-Estimating parameter variability in space and time is critical when using population models to forecast responses to directional changes in exogoneous drivers like climate [@Clark2011; @Vanderwel2013; @Wolkovich2014a; @Petchey2015].
+Estimating parameter variability in space and time is critical when using population models to forecast responses to directional changes in exogoneous drivers like climate [@Clark2011; @Vanderwel2013; @Wolkovich2014; @Petchey2015].
 Compared to demographic models, density-structured modeling approaches may provide an efficient route toward estimating such uncertainty.

### 2015-06-17 manuscript/manuscript.tex
fixes references

@@
 equilibrium cover than the IPM, but of similar magnitude (-5 - 6\%; Fig.
-2C). \emph{P. secunda} was the only species that the IPM and QBM made
+2c). \emph{P. secunda} was the only species that the IPM and QBM made
 projections of the same sign and somewhat similar magnitude (Fig. 2d).
@@
 using population models to forecast responses to directional changes in
-exogoneous drivers like climate ({\textbf{???}}, {\textbf{???}},
-{\textbf{???}}, {\textbf{???}}). Compared to demographic models,
-density-structured modeling approaches may provide an efficient route
-toward estimating such uncertainty. However, population models based on
-aggregated inidividual data have yet to be tested with climate drivers.
-Since it is individuals that respond to weather, rather than
-populations, models based on a aggregated data may produce incorrect
-forecasts (Clark et al. 2012). Thus, we sought to compare climate
-sensitivities of individual and population based models of four
-grassland species populations.
+exogoneous drivers like climate (Clark et al. 2011, Vanderwel et al.
+2013, Wolkovich et al. 2014, Petchey et al. 2015). Compared to
+demographic models, density-structured modeling approaches may provide
+an efficient route toward estimating such uncertainty. However,
+population models based on aggregated inidividual data have yet to be
+tested with climate drivers. Since it is individuals that respond to
+weather, rather than populations, models based on a aggregated data may
+produce incorrect forecasts (Clark et al. 2012). Thus, we sought to
+compare climate sensitivities of individual and population based models
+of four grassland species populations.
 
@@
 dynamics. American Naturalist 177:1--17.
+
+Petchey, O. L., M. Pontarp, T. M. Massie, S. Kéfi, A. Ozgul, M.
+Weilenmann, G. M. Palamara, F. Altermatt, B. Matthews, J. M. Levine, D.
+Z. Childs, B. J. McGill, M. E. Schaepman, B. Schmid, P. Spaak, A. P.
+Beckerman, F. Pennekamp, and I. S. Pearse. 2015. The ecological forecast
+horizon, and examples of its uses and determinants. Ecology Letters
+18:597--611.
 

### 2015-06-17 manuscript/manuscript.Rmd
updates ms discussion

@@
 fontsize: 12pt
-mainfont: times
 geometry: margin=1in
@@
 
-Microscale models outperform mesoscale models when forecasting climate change impacts on plant populations
+Do we need detailed demographic data to forecast the impacts of climate change on populations?
 ========================================================================================
@@
 ----------
-Demographic models are costly to parameterize, yet remain the most widely-used tools for studying plant population dynamics (e.g., @Salguero-Gomez2015).
-New density-structured approaches sacrifice some precision at the local scale for better spatial coverage that is virtually unattainable, due to time and funding constraints, using traditional individual-based data [@Freckleton2011; @Queenborough2011].
-These approaches are appealing because traditional population models suffer from constrained parameter space -- meaning forecasts quickly extend beyond the parameter space used to fit the model.
-Density-structured approaches rely on easy-to-collect census data (e.g., percent cover in $1\text{m}^2$ quadrats) so sampling populations across large spatial extents would be possible, thus covering more potential parameter space.
+Perhaps the greatest challenge for ecology in the 21st century is to forecast the impacts of environmental change [@Clark2001].
+To do so requires sophisticated modeling approaches that fully account for uncertainty and variability in the ecological process and associated parameters.
+This requires large amounts of data collected over large spatio-temporal extents.
+State-of-the-science modeling techniques cannot overcome data limitations.
+Such is the case for many population models.
 
-Estimating parameter variability in space and time is critical when using population models to forecast responses to directional changes in exogoneous drivers like climate [@Clark2011; @Vanderwel2013; @Wolkovich2014; @Petchey2015].
-Compared to demographic models, density-structured modeling approaches may provide an efficient route toward estimating such uncertainty.
-However, population models based on aggregated inidividual data have yet to be tested with climate drivers.
-Since it is individuals that respond to weather, rather than populations, models based on a aggregated data may produce incorrect forecasts [@Clark2012].
-Thus, we sought to compare climate sensitivities of individual and population based models of four grassland species populations.
+As a potential remedy to the "data dearth" problem, @Queenborough2011 and @Freckleton2011, building on work by @Taylor2004, advocate a "density-structured" modeling approach.
+Such models do not require individual level demographic data.
+Indeed, @Queenborough2011 provide empirical evidence that density-structured models adequately described population dynamics.
+The results from density-structured models are not as precise as those from traditional population models, but the loss in precision is traded off with a gain in data.
+The study by @Queenborough2011 included data from 500 fields (4 hectares each) in 49 farms, all collected by two people in 6 weeks.
+This is far more data from a far greater spatial extent than possible if measuring individual plant demography (in a world of limited time and money, at least).
+The appeal of density-structured approaches is clear.
+
+However, at their core, density-structured models rely on individual level data aggregated to a population level metric (e.g., density classes or percent cover).
+This creates a potential problem if such models are to be used in a climate change context because inidividuals respond to climate, not populations [@Clark2012].
+Are models based on population level metrics as sensitive to climate as models based on individual level metrics?
+Do these two types of models produce consistent forecasts?
+Do we need detailed demographic data to forecast the impacts of climate change?
+These are the questions we sought to answer here.
+
+Using individual and aggregated forms of the same dataset, we were able to directly compare a traditional demographic modeling approach to a population model based on aggregated data.
+Our quad-based model (QBM) is based on percent cover data and so is in the spirit of density-structured models.
+In terms of each model's forecasting ability, the IPM outperformed the QBM (Table 1).
+This is unsurprising since we expected to lose some precision at the aggregated level.
+However, the underwhelming performance of the QBM could call in the question forecasts that differ from the IPM.
+
+Indeed, when we perturbed climate factors the QBM made forecasts completely contradictory to those of the IPM for three of our four study species (Fig.
+2a-d).
+In a perfect world, the QBM would have made forecasts of at least the same direction as the IPM.
+If that had been the case we could conclude that aggregate level models could prove useful for forecasting climate change impacts on populations.
+Unfortunately, this was not the case.
+
+Given the superior ability of the IPM to predict out of sample observations (Table 1), we have no choice but to conclude it is the superior model.
+Following that logic, we can only assume that, at least contingent on the data in hand, the IPM is producing the correct forecasts to climate perturbations.
+The QBM failed to match IPM forecasts, implying that detailed demographic data may be necessary to accurately detect climate signals that are utlimately important at the population level.
+This result further confirms related work on the importance of individual variability on population level responses to exogenous drivers (CITATIONS).
+
+This result is not as dire as it seems when we consider model uncerainty.
+Yes, the QBM failed to match IPM forecasts under perturbed climate, but the uncertainties around those forecasts were so great (Fig.
+2e-h) that we cannot say that the two model's forecasts were different with any level of confidence.
+This offers hope that with "enough" population level data one will be able to reduce parameter uncertainty and thus model forecasts.
+This remains an open question.
 
@@
 Something about fitting the models...then cite Britta's paper: 20-25 years needed!
+
+Conclusions
+-----------
+
+This work is not a critique of density-structured population models.
+Instead, we consider our results a cautionary tale.
+In some cases and for certain species, population models based on aggregated data may prove useful and unbiased.
+But, as our work shows, that is not a foregone conclusion when considering the impact of exogenous drivers like climate on population states.
+Thus, forecasts from aggregate level models should be viewed with caution and should never be unaccompianed by uncertainty.
+As we have shown, uncertainty around forecasts was large for both model types.
+Which leads us to our most pessimistic conclusion: even with 14 years of detailed demographic data and sophisticated modeling techniques we failed to produce forecasts with any level of acceptable uncertainty.
+In our view, uncertainty of climate change related forecasts can be reduced by (1) longer time series and (2) climate covariates that match the scale of inference (e.g., plot rather than landscape level climate/weather metrics).
+Do we need detailed demographic data to forecast the impacts of climate change on populations?
+Probably.
 

### 2015-06-17 manuscript/manuscript.tex
updates ms discussion

--- 2015-06-16 manuscript/manuscript.tex
@@
   \newcommand{\euro}{€}
-    \setmainfont{times}
 \fi
@@
 
-\section{Microscale models outperform mesoscale models when forecasting
-climate change impacts on plant
-populations}\label{microscale-models-outperform-mesoscale-models-when-forecasting-climate-change-impacts-on-plant-populations}
+\section{Do we need detailed demographic data to forecast the impacts of
+climate change on
+populations?}\label{do-we-need-detailed-demographic-data-to-forecast-the-impacts-of-climate-change-on-populations}
 
@@
 
-Demographic models are costly to parameterize, yet remain the most
-widely-used tools for studying plant population dynamics (e.g.,
-Salguero-Gómez et al. (2015)). New density-structured approaches
-sacrifice some precision at the local scale for better spatial coverage
-that is virtually unattainable, due to time and funding constraints,
-using traditional individual-based data (Freckleton et al. 2011,
-Queenborough et al. 2011). These approaches are appealing because
-traditional population models suffer from constrained parameter space --
-meaning forecasts quickly extend beyond the parameter space used to fit
-the model. Density-structured approaches rely on easy-to-collect census
-data (e.g., percent cover in $1\text{m}^2$ quadrats) so sampling
-populations across large spatial extents would be possible, thus
-covering more potential parameter space.
+Perhaps the greatest challenge for ecology in the 21st century is to
+forecast the impacts of environmental change (Clark et al. 2001). To do
+so requires sophisticated modeling approaches that fully account for
+uncertainty and variability in the ecological process and associated
+parameters. This requires large amounts of data collected over large
+spatio-temporal extents. State-of-the-science modeling techniques cannot
+overcome data limitations. Such is the case for many population models.
 
-Estimating parameter variability in space and time is critical when
-using population models to forecast responses to directional changes in
-exogoneous drivers like climate ({\textbf{???}}, {\textbf{???}},
-{\textbf{???}}, {\textbf{???}}). Compared to demographic models,
-density-structured modeling approaches may provide an efficient route
-toward estimating such uncertainty. However, population models based on
-aggregated inidividual data have yet to be tested with climate drivers.
-Since it is individuals that respond to weather, rather than
-populations, models based on a aggregated data may produce incorrect
-forecasts (Clark et al. 2012). Thus, we sought to compare climate
-sensitivities of individual and population based models of four
-grassland species populations.
+As a potential remedy to the ``data dearth'' problem, Queenborough et
+al. (2011) and Freckleton et al. (2011), building on work by Taylor and
+Hastings (2004), advocate a ``density-structured'' modeling approach.
+Such models do not require individual level demographic data. Indeed,
+Queenborough et al. (2011) provide empirical evidence that
+density-structured models adequately described population dynamics. The
+results from density-structured models are not as precise as those from
+traditional population models, but the loss in precision is traded off
+with a gain in data. The study by Queenborough et al. (2011) included
+data from 500 fields (4 hectares each) in 49 farms, all collected by two
+people in 6 weeks. This is far more data from a far greater spatial
+extent than possible if measuring individual plant demography (in a
+world of limited time and money, at least). The appeal of
+density-structured approaches is clear.
+
+However, at their core, density-structured models rely on individual
+level data aggregated to a population level metric (e.g., density
+classes or percent cover). This creates a potential problem if such
+models are to be used in a climate change context because inidividuals
+respond to climate, not populations (Clark et al. 2012). Are models
+based on population level metrics as sensitive to climate as models
+based on individual level metrics? Do these two types of models produce
+consistent forecasts? Do we need detailed demographic data to forecast
+the impacts of climate change? These are the questions we sought to
+answer here.
+
+Using individual and aggregated forms of the same dataset, we were able
+to directly compare a traditional demographic modeling approach to a
+population model based on aggregated data. Our quad-based model (QBM) is
+based on percent cover data and so is in the spirit of
+density-structured models. In terms of each model's forecasting ability,
+the IPM outperformed the QBM (Table 1). This is unsurprising since we
+expected to lose some precision at the aggregated level. However, the
+underwhelming performance of the QBM could call in the question
+forecasts that differ from the IPM.
+
+Indeed, when we perturbed climate factors the QBM made forecasts
+completely contradictory to those of the IPM for three of our four study
+species (Fig. 2a-d). In a perfect world, the QBM would have made
+forecasts of at least the same direction as the IPM. If that had been
+the case we could conclude that aggregate level models could prove
+useful for forecasting climate change impacts on populations.
+Unfortunately, this was not the case.
+
+Given the superior ability of the IPM to predict out of sample
+observations (Table 1), we have no choice but to conclude it is the
+superior model. Following that logic, we can only assume that, at least
+contingent on the data in hand, the IPM is producing the correct
+forecasts to climate perturbations. The QBM failed to match IPM
+forecasts, implying that detailed demographic data may be necessary to
+accurately detect climate signals that are utlimately important at the
+population level. This result further confirms related work on the
+importance of individual variability on population level responses to
+exogenous drivers (CITATIONS).
+
+This result is not as dire as it seems when we consider model
+uncerainty. Yes, the QBM failed to match IPM forecasts under perturbed
+climate, but the uncertainties around those forecasts were so great
+(Fig. 2e-h) that we cannot say that the two model's forecasts were
+different with any level of confidence. This offers hope that with
+``enough'' population level data one will be able to reduce parameter
+uncertainty and thus model forecasts. This remains an open question.
 
@@
 20-25 years needed!
+
+\subsection{Conclusions}\label{conclusions}
+
+This work is not a critique of density-structured population models.
+Instead, we consider our results a cautionary tale. In some cases and
+for certain species, population models based on aggregated data may
+prove useful and unbiased. But, as our work shows, that is not a
+foregone conclusion when considering the impact of exogenous drivers
+like climate on population states. Thus, forecasts from aggregate level
+models should be viewed with caution and should never be unaccompianed
+by uncertainty. As we have shown, uncertainty around forecasts was large
+for both model types. Which leads us to our most pessimistic conclusion:
+even with 14 years of detailed demographic data and sophisticated
+modeling techniques we failed to produce forecasts with any level of
+acceptable uncertainty. In our view, uncertainty of climate change
+related forecasts can be reduced by (1) longer time series and (2)
+climate covariates that match the scale of inference (e.g., plot rather
+than landscape level climate/weather metrics). Do we need detailed
+demographic data to forecast the impacts of climate change on
+populations? Probably.
 
@@
 
+Clark, J. S., S. R. Carpenter, M. Barber, S. Collins, A. Dobson, J. A.
+Foley, D. M. Lodge, M. Pascual, R. Pielke, W. Pizer, C. Pringle, W. V.
+Reid, K. A. Rose, O. Sala, W. H. Schlesinger, D. H. Wall, and D. Wear.
+2001. Ecological forecasts: an emerging imperative. Science (New York,
+N.Y.) 293:657--660.
+
 Freckleton, R. P., W. J. Sutherland, A. R. Watkinson, and S. A.

### 2015-06-17 manuscript/manuscript.Rmd
adds vital rate section to discussion

@@
 
+### The IPM and QBM produced inconsistent forecasts
 Using individual and aggregated forms of the same dataset, we were able to directly compare a traditional demographic modeling approach to a population model based on aggregated data.
@@
 This remains an open question.
+
+### The role of vital rate climate dependence
+Our intuition was that species with strong climate-dependence on vital rates not well resolved at the aggregate level would result in different forecasts from the two models.
+For example, survival is very size dependent: smaller individuals have a higher probability of death [@Chu2014].
+At the same time, a single small individual contributes relatively little to percent cover estimates at the plot scale.
+So, if survival of individuals was positively impacted by temperature increases, for example, we would expect to detect this signal in the individual level data but not in the aggregate level data.
 

### 2015-06-17 manuscript/manuscript.tex
adds vital rate section to discussion

@@
 
+\subsubsection{The IPM and QBM produced inconsistent
+forecasts}\label{the-ipm-and-qbm-produced-inconsistent-forecasts}
+
 Using individual and aggregated forms of the same dataset, we were able
@@
 uncertainty and thus model forecasts. This remains an open question.
+
+\subsubsection{The role of vital rate climate
+dependence}\label{the-role-of-vital-rate-climate-dependence}
+
+Our intuition was that species with strong climate-dependence on vital
+rates not well resolved at the aggregate level would result in different
+forecasts from the two models. For example, survival is very size
+dependent: smaller individuals have a higher probability of death (Chu
+and Adler 2014). At the same time, a single small individual contributes
+relatively little to percent cover estimates at the plot scale. So, if
+survival of individuals was positively impacted by temperature
+increases, for example, we would expect to detect this signal in the
+individual level data but not in the aggregate level data. However, the
+species with recruitment and survival rates most sensitive to climate
+factors (\emph{P. secunda}; Fig. 3d) was also the only species for which
+the IPM and QBM made consistent forecasts (Fig. 2d).
+
+Individual growth was the most sensitive vital rate to climate factors
+for all species (Fig. 3). Individual plant growth, if survival of large
+plants is high, likely plays the biggest role in year-to-year percent
+cover at the plot level. Therefore, we would naively expect forecasts
+from the two models to be consistent. Since this was not the case, and
+other vital rates were relatively insensitive to
 
@@
 
+Chu, C., and P. B. Adler. 2014. When should plant population models
+include age structure? Journal of Ecology 102:531--543.
+
 Chu, C., and P. B. Adler. 2015. Large niche differences emerge at the

### 2015-06-18 manuscript/manuscript.Rmd
updates ms discussion

@@
 
-Do we need detailed demographic data to forecast the impacts of climate change on populations?
+Do we need detailed demographic data to forecast the impacts of climate change on plant populations?
 ========================================================================================
@@
 The site is about 800 m above sea level and mean annual precipitation (1878-2009) is 334 mm, with most annual precipitation falling from April through September.
-The site is grass dominated and, for the purposes of our study, we focus on the four most abundant graminoid species: \emph{Bouteloua gracilis} (BOGR), \emph{Hesperostipa comata} (HECO), \emph{Pascopyrum smithii} (PASM), and \emph{Poa secunda} (POSE).
+The site is grass dominated and, for the purposes of our study, we focus on the four most abundant graminoid species: \emph{Bouteloua gracilis} (BOGR), \emph{Hesperostipa comata} (HECO), \emph{Pascopyrum smithii} (PASM), and \emph{Poa secunda} (POSE) (Fig.
+1).
 
@@
 In this paper, we model populations based on two levels of data: individual and quadrat (Fig.
-1).
+2).
 The individual data is the "raw" data.
@@
 Equilibrium cover from both models was sensitive to climate (Fig.
-2a-d).
+3a-d).
 The IPM projected percent changes in equilibrium cover from `r round(min(subset(meanparam_agg, species=="BOGR" & model=="IPM")["meandiff"])*100)` - `r round(max(subset(meanparam_agg, species=="BOGR" & model=="IPM")["meandiff"])*100)`% for *B.
@@
 comata* (`r round(min(subset(meanparam_agg, species=="HECO" & model=="QBM")["meandiff"])*100)` - `r round(max(subset(meanparam_agg, species=="HECO" & model=="QBM")["meandiff"])*100)`%; Fig.
-2a-b).
+3a-b).
 For *P.
 smithii*, the QBM projected opposite changes in equilibrium cover than the IPM, but of similar magnitude (`r round(min(subset(meanparam_agg, species=="PASM" & model=="QBM")["meandiff"])*100)` - `r round(max(subset(meanparam_agg, species=="PASM" & model=="QBM")["meandiff"])*100)`%; Fig.
-2c).
+3c).
 *P.
 secunda* was the only species that the IPM and QBM made projections of the same sign and somewhat similar magnitude (Fig.
-2d).
-As expected based on model validation (Table 1), IPM projections were more uncertain than QBM projections for all species and all climate change scenarios (Fig.
-2e-h).
+3d).
+As expected based on model validation (Table 1), QBM projections were more uncertain than IPM projections for all species except *P.
+smithiii* (Fig.
+3e-h).
 
@@
 Species showed similar trends (Fig.
-3).
+4).
 Growth was the most sensitive vital rate for all species, showing a negative response to increased precipitation, and stronger positive response to increased temperature, and a mostly positive response when both climate factors are increased (Fig.
-3).
+4).
 *B.
 gracilis* survival rates were sensitive to temperature, resulting in an increase in plant cover under increased temperature (Fig.
-3a).
+4a).
 In isolation, recruitment and survival were insensitive to climate factors for *H.
 comata* (Fig.
-3b).
+4b).
 Survival and recruitment of *P.
 smithii* were both sensitive, negatively, to temperature and precipitation (Fig.
-3c).
+4c).
 *P.
 secunda* equilibrium cover was sensitive to the climate effects on survival and recruitment, showing a negative effect on both vital rates for increased precipition, but a strong positive effect on survival with increased temperature (Fig.
-3d).
+4d).
 The climate impact of recruitment on equilibrium cover was negative for precipitation and temperature increases (Fig.
-3d).
+4d).
 
@@
 As a potential remedy to the "data dearth" problem, @Queenborough2011 and @Freckleton2011, building on work by @Taylor2004, advocate a "density-structured" modeling approach.
-Such models do not require individual level demographic data.
-Indeed, @Queenborough2011 provide empirical evidence that density-structured models adequately described population dynamics.
+Such models do not require individual level demographic data and can adequately describe population dynamics [@Queenborough2011].
 The results from density-structured models are not as precise as those from traditional population models, but the loss in precision is traded off with a gain in data.
@@
 Indeed, when we perturbed climate factors the QBM made forecasts completely contradictory to those of the IPM for three of our four study species (Fig.
-2a-d).
+3a-d).
 In a perfect world, the QBM would have made forecasts of at least the same direction as the IPM.
@@
 Yes, the QBM failed to match IPM forecasts under perturbed climate, but the uncertainties around those forecasts were so great (Fig.
-2e-h) that we cannot say that the two model's forecasts were different with any level of confidence.
+3e-h) that we cannot say that the two model's forecasts were different with any level of confidence.
 This offers hope that with "enough" population level data one will be able to reduce parameter uncertainty and thus model forecasts.
@@
 So, if survival of individuals was positively impacted by temperature increases, for example, we would expect to detect this signal in the individual level data but not in the aggregate level data.
+**PETER: do you think this is the right direction to go with this?**
 
@@
 Yet, model predictions proved so uncertain that any forecast, when bounded with uncertainty, would be at best not useful and at worst meaningless.
+How might we improve on this state of affairs?
 
-Something about fitting the models...then cite Britta's paper: 20-25 years needed!
+First, forecasts could be improved by matching the spatial scale of predictor variables with the spatial scale of observations.
+One of the major limitations of the models we fit here is that the climate data are at a much larger scale than the plot and individual level observations of plant cover and size.
+Climate covariates only vary by year, with no spatial variability within years.
+Thus, even if we fit models to individual level data, we are missing the key interaction point between weather and individual plants.
+**PETER: any good citations here for decent population models fit with fine-scale climate/environmental data??**
+
+Second, accurately detecting climate signals will take even longer time series.
+Recent theoretical work on detecting climate signals in noisy data suggests that even advanced approaches to parameter fitting like LASSO and random forest require 20-25 year time series (Teller et al., in review).
 
@@
 -------
-```{r figure_1,dependson="plot-options", fig.cap="Work flow of the data aggregation, model fitting, and population simulating.", fig.height=6}
+```{r figure_1, dependson="plot-options", fig.cap="Time series of average percent cover of our four focal species: *Bouteloua gracilis* (BOGR), *Hesperostipa comata* (HECO), *Pascopyrum smithii* (PASM), and *Poa secunda* (POSE).", fig.height=4}
+obs_data <- read.csv("../analysis/speciesData/quadAllCover.csv")
+#remove zeros for averaging
+obs_data <- obs_data[which(obs_data$propCover>0),]
+avgobs <- ddply(obs_data, .(year, Species), summarise,
+                mean_cover = mean(propCover))
+ggplot(avgobs, aes(x=(1900+year), y=mean_cover*100, shape=Species, linetype=Species))+
+  geom_line()+
+  geom_point(size=5, color="white")+
+  geom_point(size=3)+
+  ylab("Mean cover (%)")+
+  xlab("Year")+
+  theme_bw()
+```
+
+```{r figure_2,dependson="plot-options", fig.cap="Work flow of the data aggregation, model fitting, and population simulating.", fig.height=6}
 library(png)
@@
 
-```{r figure_2, dependson="plot-options", fig.cap="Proportional change in species' mean cover caused by a 1% increase in observed precipitation (+ppt), temperature (+temp), or both (+ppt&temp) as predicted by the individual-based IPM and the aggregate-based QBM.
+```{r figure_3, dependson="plot-options", fig.cap="Proportional change in species' mean cover caused by a 1% increase in observed precipitation (+ppt), temperature (+temp), or both (+ppt&temp) as predicted by the individual-based IPM and the aggregate-based QBM.
 Top panels show the mean predicted proportional change in cover; lower panels show the 90% quantiles of predicted proportional cover change.
@@
 
+####
+##  Vary parameter runs
+####
+meanparam_ipm <- data.frame(cover=NA, species=NA, 
+                             climsim=NA, model=NA)
+##  IPM 
+path <- "./ipm/simulations/results/climchange_varyparams/"
+ipmfiles_mean <- list.files(path)
+ipmfiles_mean <- ipmfiles_mean[grep("cover", ipmfiles_mean)]
+for(file in ipmfiles_mean){
+  tmp <- read.csv(paste(path,file,sep=""))
+  tmp$model <- "IPM"
+  meanparam_ipm <- rbind(meanparam_ipm, tmp[,c("cover", "species", "climsim", "model")])
+}
+meanparam_ipm <- meanparam_ipm[2:nrow(meanparam_ipm),]
+
+newd <- data.frame(cover=NA, species=NA, 
+                   climsim=NA, model=NA, obscover=NA)
+for(dospp in unique(meanparam_ipm$species)){
+  tmp <- subset(meanparam_ipm, species==dospp)
+  obscover <- subset(tmp, climsim=="obs")["cover"]
+  tmp2 <- subset(tmp, climsim!="obs")
+  tmp2$obscover <- rep(obscover$cover, times = length(unique(tmp2$climsim)))
+  newd <- rbind(newd, tmp2)
+}
+ipm_meansims <- newd[2:nrow(newd),]
+ipm_meansims$propdiff <- with(ipm_meansims, log(cover*100)-log(obscover*100))
+# ggplot(subset(ipm_meansims, species=="PASM"))+
+#   geom_histogram(aes(x=cover))+
+#   facet_wrap("climsim", scales = "free")+
+#   scale_x_continuous(limits=c(0,0.05))
+##  QBM
+meanparam_qbm <- data.frame(cover=NA, species=NA, 
+                             climsim=NA, model=NA)
+path <- "./quadBM/simulations/results/climatechange_varyparams/"
+qbmfiles_mean <- list.files(path)
+for(file in qbmfiles_mean){
+  tmp <- readRDS(paste(path,file,sep=""))
+  tmp$model <- "QBM"
+  meanparam_qbm <- rbind(meanparam_qbm, tmp[,c("cover", "species", "climsim", "model")])
+}
+meanparam_qbm <- meanparam_qbm[2:nrow(meanparam_qbm),]
+
+newd <- data.frame(cover=NA, species=NA, 
+                   climsim=NA, model=NA, obscover=NA)
+for(dospp in unique(meanparam_qbm$species)){
+  tmp <- subset(meanparam_qbm, species==dospp)
+  obscover <- subset(tmp, climsim=="obs")["cover"]
+  tmp2 <- subset(tmp, climsim!="obs")
+  tmp2$obscover <- rep(obscover$cover, times = length(unique(tmp2$climsim)))
+  newd <- rbind(newd, tmp2)
+}
+qbm_meansims <- newd[2:nrow(newd),]
+qbm_meansims$propdiff <- with(qbm_meansims, log(cover*100)-log(obscover*100))
+
+##  Merge IPM and QBM mean parameter results
+bothmodel_outs <- rbind(ipm_meansims, qbm_meansims)
+meanparam_agg <- ddply(bothmodel_outs, .(species, climsim, model), summarise,
+                       meandiff = mean(propdiff),
+                       hidiff = quantile(propdiff, probs = 0.95),
+                       lodiff = quantile(propdiff, probs = 0.05))
 dgd = position_dodge(width = 0.6)
@@
 
-```{r figure_3, dependson="plot-options", fig.cap="Sensitivity of equilibrium cover to a 1% increase in precipitation (+ppt), temperature (+temp), or both (+ppt&temp) applied to individual and combined vital rates.
+```{r figure_4, dependson="plot-options", fig.cap="Sensitivity of equilibrium cover to a 1% increase in precipitation (+ppt), temperature (+temp), or both (+ppt&temp) applied to individual and combined vital rates.
 For example, the points associated with G show the median cover from IPM simulations where a climate perturbation is applied only to the growth regression climate covariates.", fig.width=4.5, fig.height=6, cache=FALSE, results='hide'}

### 2015-06-18 manuscript/manuscript.tex
updates ms discussion

@@
 
-\normalsize
+\maketitle
 
 \section{Do we need detailed demographic data to forecast the impacts of
-climate change on
-populations?}\label{do-we-need-detailed-demographic-data-to-forecast-the-impacts-of-climate-change-on-populations}
+climate change on plant
+populations?}\label{do-we-need-detailed-demographic-data-to-forecast-the-impacts-of-climate-change-on-plant-populations}
 
@@
 (BOGR), \emph{Hesperostipa comata} (HECO), \emph{Pascopyrum smithii}
-(PASM), and \emph{Poa secunda} (POSE).
+(PASM), and \emph{Poa secunda} (POSE) (Fig. 1).
 
 From 1932 to 1945 individual plants were identified and mapped annually
-in 44 1-$\text{m}^2$ quadrats using a pantograph. The quadrats were
+in 44 1-\(\text{m}^2\) quadrats using a pantograph. The quadrats were
 distributed in six pastures, each assigned a grazing treatment of light
@@
 In this paper, we model populations based on two levels of data:
-individual and quadrat (Fig. 1). The individual data is the ``raw''
+individual and quadrat (Fig. 2). The individual data is the ``raw''
 data. For the quadrat level we data we simply sum individual areal cover
@@
 
-Equilibrium cover from both models was sensitive to climate (Fig. 2a-d).
+Equilibrium cover from both models was sensitive to climate (Fig. 3a-d).
 The IPM projected percent changes in equilibrium cover from -3 - 8\% for
@@
 \emph{B. gracilis} (-63 - 30\%) and \emph{H. comata} (-50 - -18\%; Fig.
-2a-b). For \emph{P. smithii}, the QBM projected opposite changes in
+3a-b). For \emph{P. smithii}, the QBM projected opposite changes in
 equilibrium cover than the IPM, but of similar magnitude (-5 - 6\%; Fig.
-2c). \emph{P. secunda} was the only species that the IPM and QBM made
-projections of the same sign and somewhat similar magnitude (Fig. 2d).
-As expected based on model validation (Table 1), IPM projections were
-more uncertain than QBM projections for all species and all climate
-change scenarios (Fig. 2e-h).
+3c). \emph{P. secunda} was the only species that the IPM and QBM made
+projections of the same sign and somewhat similar magnitude (Fig. 3d).
+As expected based on model validation (Table 1), QBM projections were
+more uncertain than IPM projections for all species except \emph{P.
+smithiii} (Fig. 3e-h).
 
@@
 sensitivity of each vital rate in isolation and in pairwise
-combinations. Species showed similar trends (Fig. 3). Growth was the
+combinations. Species showed similar trends (Fig. 4). Growth was the
 most sensitive vital rate for all species, showing a negative response
@@
 temperature, and a mostly positive response when both climate factors
-are increased (Fig. 3). \emph{B. gracilis} survival rates were sensitive
+are increased (Fig. 4). \emph{B. gracilis} survival rates were sensitive
 to temperature, resulting in an increase in plant cover under increased
-temperature (Fig. 3a). In isolation, recruitment and survival were
-insensitive to climate factors for \emph{H. comata} (Fig. 3b). Survival
+temperature (Fig. 4a). In isolation, recruitment and survival were
+insensitive to climate factors for \emph{H. comata} (Fig. 4b). Survival
 and recruitment of \emph{P. smithii} were both sensitive, negatively, to
-temperature and precipitation (Fig. 3c). \emph{P. secunda} equilibrium
+temperature and precipitation (Fig. 4c). \emph{P. secunda} equilibrium
 cover was sensitive to the climate effects on survival and recruitment,
@@
 but a strong positive effect on survival with increased temperature
-(Fig. 3d). The climate impact of recruitment on equilibrium cover was
-negative for precipitation and temperature increases (Fig. 3d).
+(Fig. 4d). The climate impact of recruitment on equilibrium cover was
+negative for precipitation and temperature increases (Fig. 4d).
 
@@
 Hastings (2004), advocate a ``density-structured'' modeling approach.
-Such models do not require individual level demographic data. Indeed,
-Queenborough et al. (2011) provide empirical evidence that
-density-structured models adequately described population dynamics. The
+Such models do not require individual level demographic data and can
+adequately describe population dynamics (Queenborough et al. 2011). The
 results from density-structured models are not as precise as those from
@@
 completely contradictory to those of the IPM for three of our four study
-species (Fig. 2a-d). In a perfect world, the QBM would have made
+species (Fig. 3a-d). In a perfect world, the QBM would have made
 forecasts of at least the same direction as the IPM. If that had been
@@
 climate, but the uncertainties around those forecasts were so great
-(Fig. 2e-h) that we cannot say that the two model's forecasts were
+(Fig. 3e-h) that we cannot say that the two model's forecasts were
 different with any level of confidence. This offers hope that with
@@
 increases, for example, we would expect to detect this signal in the
-individual level data but not in the aggregate level data. However, the
-species with recruitment and survival rates most sensitive to climate
-factors (\emph{P. secunda}; Fig. 3d) was also the only species for which
-the IPM and QBM made consistent forecasts (Fig. 2d).
-
-Individual growth was the most sensitive vital rate to climate factors
-for all species (Fig. 3). Individual plant growth, if survival of large
-plants is high, likely plays the biggest role in year-to-year percent
-cover at the plot level. Therefore, we would naively expect forecasts
-from the two models to be consistent. Since this was not the case, and
-other vital rates were relatively insensitive to
+individual level data but not in the aggregate level data.
+\textbf{PETER: do you think this is the right direction to go with
+this?}
 
@@
 predictions proved so uncertain that any forecast, when bounded with
-uncertainty, would be at best not useful and at worst meaningless.
+uncertainty, would be at best not useful and at worst meaningless. How
+might we improve on this state of affairs?
 
-Something about fitting the models\ldots{}then cite Britta's paper:
-20-25 years needed!
+First, forecasts could be improved by matching the spatial scale of
+predictor variables with the spatial scale of observations. One of the
+major limitations of the models we fit here is that the climate data are
+at a much larger scale than the plot and individual level observations
+of plant cover and size. Climate covariates only vary by year, with no
+spatial variability within years. Thus, even if we fit models to
+individual level data, we are missing the key interaction point between
+weather and individual plants. \textbf{PETER: any good citations here
+for decent population models fit with fine-scale climate/environmental
+data??}
+
+Second, accurately detecting climate signals will take even longer time
+series. Recent theoretical work on detecting climate signals in noisy
+data suggests that even advanced approaches to parameter fitting like
+LASSO and random forest require 20-25 year time series (Teller et al.,
+in review).
 
@@
 \includegraphics{components/figure/manuscript-figure_1.pdf}
+\caption{Time series of average percent cover of our four focal species:
+\emph{Bouteloua gracilis} (BOGR), \emph{Hesperostipa comata} (HECO),
+\emph{Pascopyrum smithii} (PASM), and \emph{Poa secunda} (POSE).}
+\end{figure}
+
+\begin{figure}[htbp]
+\centering
+\includegraphics{components/figure/manuscript-figure_2.pdf}
 \caption{Work flow of the data aggregation, model fitting, and
@@
 \centering
-\includegraphics{components/figure/manuscript-figure_2.pdf}
+\includegraphics{components/figure/manuscript-figure_3.pdf}
 \caption{Proportional change in species' mean cover caused by a 1\%
@@
 \centering
-\includegraphics{components/figure/manuscript-figure_3.pdf}
+\includegraphics{components/figure/manuscript-figure_4.pdf}
 \caption{Sensitivity of equilibrium cover to a 1\% increase in

### 2015-06-26 manuscript/manuscript.Rmd
updates fig 1 with multipanel

@@
 -------
-```{r figure_1, dependson="plot-options", fig.cap="Time series of average percent cover of our four focal species: *Bouteloua gracilis* (BOGR), *Hesperostipa comata* (HECO), *Pascopyrum smithii* (PASM), and *Poa secunda* (POSE).", fig.height=4}
+```{r figure_1, dependson="plot-options", fig.cap="Time series of average percent cover over all quadrats for our four focal species: *Bouteloua gracilis* (BOGR), *Hesperostipa comata* (HECO), *Pascopyrum smithii* (PASM), and *Poa secunda* (POSE).
+Light grey lines show trajectories of individual quadrats.
+Note the different y-axis scales across panels.", fig.height=3, fig.width=8.5}
 obs_data <- read.csv("../analysis/speciesData/quadAllCover.csv")
@@
                 mean_cover = mean(propCover))
-ggplot(avgobs, aes(x=(1900+year), y=mean_cover*100, shape=Species, linetype=Species))+
-  geom_line()+
-  geom_point(size=5, color="white")+
-  geom_point(size=3)+
+
+ggplot()+
+  geom_line(data = obs_data, aes(x=(1900+year), y=propCover*100, group=quad), 
+            color="grey", size=0.25)+
+  geom_line(data=avgobs, aes(x=(1900+year), y=mean_cover*100))+
+  geom_point(data=avgobs, aes(x=(1900+year), y=mean_cover*100),size=3)+
   ylab("Mean cover (%)")+
   xlab("Year")+
-  theme_bw()
+  theme_bw()+
+  facet_wrap("Species", ncol=4, nrow=1, scales = "free")+
+  guides(shape=FALSE, linetype=FALSE)
 ```

### 2015-06-26 manuscript/manuscript.tex
updates fig 1 with multipanel

--- 2015-06-17 manuscript/manuscript.tex
@@
 \section{Do we need detailed demographic data to forecast the impacts of
-climate change on
-populations?}\label{do-we-need-detailed-demographic-data-to-forecast-the-impacts-of-climate-change-on-populations}
+climate change on plant
+populations?}\label{do-we-need-detailed-demographic-data-to-forecast-the-impacts-of-climate-change-on-plant-populations}
 
@@
 (BOGR), \emph{Hesperostipa comata} (HECO), \emph{Pascopyrum smithii}
-(PASM), and \emph{Poa secunda} (POSE).
+(PASM), and \emph{Poa secunda} (POSE) (Fig. 1).
 
@@
 \includegraphics{components/figure/manuscript-figure_1.pdf}
+\caption{Time series of average percent cover over all quadrats for our
+four focal species: \emph{Bouteloua gracilis} (BOGR), \emph{Hesperostipa
+comata} (HECO), \emph{Pascopyrum smithii} (PASM), and \emph{Poa secunda}
+(POSE).}
+\end{figure}
+
+\begin{figure}[htbp]
+\centering
+\includegraphics{components/figure/manuscript-figure_2.pdf}
 \caption{Work flow of the data aggregation, model fitting, and

### 2015-06-29 manuscript/manuscript.Rmd
updates figures in ms with parameter vary runs

@@
 
+An important caveat to the above conclusion is that our comparison of the IPM and the QBM
+
 This result is not as dire as it seems when we consider model uncerainty.
@@
 ####
-meanparam_ipm <- data.frame(cover=NA, species=NA, 
-                             climsim=NA, model=NA)
-##  IPM 
-path <- "./ipm/simulations/results/climchange_meanparams/"
-ipmfiles_mean <- list.files(path)
-ipmfiles_mean <- ipmfiles_mean[grep("cover", ipmfiles_mean)]
-for(file in ipmfiles_mean){
-  tmp <- read.csv(paste(path,file,sep=""))
-  tmp$model <- "IPM"
-  meanparam_ipm <- rbind(meanparam_ipm, tmp[,c("cover", "species", "climsim", "model")])
-}
-meanparam_ipm <- meanparam_ipm[2:nrow(meanparam_ipm),]
-
-newd <- data.frame(cover=NA, species=NA, 
-                   climsim=NA, model=NA, obscover=NA)
-for(dospp in unique(meanparam_ipm$species)){
-  tmp <- subset(meanparam_ipm, species==dospp)
-  obscover <- subset(tmp, climsim=="obs")["cover"]
-  tmp2 <- subset(tmp, climsim!="obs")
-  tmp2$obscover <- rep(obscover$cover, times = length(unique(tmp2$climsim)))
-  newd <- rbind(newd, tmp2)
-}
-ipm_meansims <- newd[2:nrow(newd),]
-ipm_meansims$propdiff <- with(ipm_meansims, log(cover*100)-log(obscover*100))
-
-##  QBM
-meanparam_qbm <- data.frame(cover=NA, species=NA, 
-                             climsim=NA, model=NA)
-path <- "./quadBM/simulations/results/climatechange_meanparams/"
-qbmfiles_mean <- list.files(path)
-for(file in qbmfiles_mean){
-  tmp <- readRDS(paste(path,file,sep=""))
-  tmp$model <- "QBM"
-  meanparam_qbm <- rbind(meanparam_qbm, tmp[,c("cover", "species", "climsim", "model")])
-}
-meanparam_qbm <- meanparam_qbm[2:nrow(meanparam_qbm),]
-
-newd <- data.frame(cover=NA, species=NA, 
-                   climsim=NA, model=NA, obscover=NA)
-for(dospp in unique(meanparam_qbm$species)){
-  tmp <- subset(meanparam_qbm, species==dospp)
-  obscover <- subset(tmp, climsim=="obs")["cover"]
-  tmp2 <- subset(tmp, climsim!="obs")
-  tmp2$obscover <- rep(obscover$cover, times = length(unique(tmp2$climsim)))
-  newd <- rbind(newd, tmp2)
-}
-qbm_meansims <- newd[2:nrow(newd),]
-qbm_meansims$propdiff <- with(qbm_meansims, log(cover*100)-log(obscover*100))
-
-##  Merge IPM and QBM mean parameter results
-bothmodel_outs <- rbind(ipm_meansims, qbm_meansims)
-meanparam_agg <- ddply(bothmodel_outs, .(species, climsim, model), summarise,
-                       meandiff = mean(propdiff),
-                       hidiff = quantile(propdiff, probs = 0.95),
-                       lodiff = quantile(propdiff, probs = 0.05))
-dgd = position_dodge(width = 0.9)
-meanplot <- ggplot(meanparam_agg)+
-  geom_bar(aes(x=climsim, y=meandiff, fill=model),
-             position=dgd, stat="identity")+
-  theme_bw()+
-  scale_fill_manual(values=c("black", "grey45"), name="Model")+
-  facet_wrap("species", scales="free", nrow=1)+
-  scale_x_discrete(labels=c("+ppt", "+temp", "+ppt&temp"))+
-  ylab("Mean proportional difference")+
-  xlab("Climate simulation")+
-  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+# meanparam_ipm <- data.frame(cover=NA, species=NA, 
+#                              climsim=NA, model=NA)
+# ##  IPM 
+# path <- "./ipm/simulations/results/climchange_meanparams/"
+# ipmfiles_mean <- list.files(path)
+# ipmfiles_mean <- ipmfiles_mean[grep("cover", ipmfiles_mean)]
+# for(file in ipmfiles_mean){
+#   tmp <- read.csv(paste(path,file,sep=""))
+#   tmp$model <- "IPM"
+#   meanparam_ipm <- rbind(meanparam_ipm, tmp[,c("cover", "species", "climsim", "model")])
+# }
+# meanparam_ipm <- meanparam_ipm[2:nrow(meanparam_ipm),]
+# 
+# newd <- data.frame(cover=NA, species=NA, 
+#                    climsim=NA, model=NA, obscover=NA)
+# for(dospp in unique(meanparam_ipm$species)){
+#   tmp <- subset(meanparam_ipm, species==dospp)
+#   obscover <- subset(tmp, climsim=="obs")["cover"]
+#   tmp2 <- subset(tmp, climsim!="obs")
+#   tmp2$obscover <- rep(obscover$cover, times = length(unique(tmp2$climsim)))
+#   newd <- rbind(newd, tmp2)
+# }
+# ipm_meansims <- newd[2:nrow(newd),]
+# ipm_meansims$propdiff <- with(ipm_meansims, log(cover*100)-log(obscover*100))
+# 
+# ##  QBM
+# meanparam_qbm <- data.frame(cover=NA, species=NA, 
+#                              climsim=NA, model=NA)
+# path <- "./quadBM/simulations/results/climatechange_meanparams/"
+# qbmfiles_mean <- list.files(path)
+# for(file in qbmfiles_mean){
+#   tmp <- readRDS(paste(path,file,sep=""))
+#   tmp$model <- "QBM"
+#   meanparam_qbm <- rbind(meanparam_qbm, tmp[,c("cover", "species", "climsim", "model")])
+# }
+# meanparam_qbm <- meanparam_qbm[2:nrow(meanparam_qbm),]
+# 
+# newd <- data.frame(cover=NA, species=NA, 
+#                    climsim=NA, model=NA, obscover=NA)
+# for(dospp in unique(meanparam_qbm$species)){
+#   tmp <- subset(meanparam_qbm, species==dospp)
+#   obscover <- subset(tmp, climsim=="obs")["cover"]
+#   tmp2 <- subset(tmp, climsim!="obs")
+#   tmp2$obscover <- rep(obscover$cover, times = length(unique(tmp2$climsim)))
+#   newd <- rbind(newd, tmp2)
+# }
+# qbm_meansims <- newd[2:nrow(newd),]
+# qbm_meansims$propdiff <- with(qbm_meansims, log(cover*100)-log(obscover*100))
+# 
+# ##  Merge IPM and QBM mean parameter results
+# bothmodel_outs <- rbind(ipm_meansims, qbm_meansims)
+# meanparam_agg <- ddply(bothmodel_outs, .(species, climsim, model), summarise,
+#                        meandiff = mean(propdiff),
+#                        hidiff = quantile(propdiff, probs = 0.95),
+#                        lodiff = quantile(propdiff, probs = 0.05))
+# dgd = position_dodge(width = 0.9)
+# meanplot <- ggplot(meanparam_agg)+
+#   geom_bar(aes(x=climsim, y=meandiff, fill=model),
+#              position=dgd, stat="identity")+
+#   theme_bw()+
+#   scale_fill_manual(values=c("black", "grey45"), name="Model")+
+#   facet_wrap("species", scales="free", nrow=1)+
+#   scale_x_discrete(labels=c("+ppt", "+temp", "+ppt&temp"))+
+#   ylab("Mean proportional difference")+
+#   xlab("Climate simulation")+
+#   theme(axis.text.x = element_text(angle = 45, hjust = 1))
+#
 
@@
 bothmodel_outs <- rbind(ipm_meansims, qbm_meansims)
+
 meanparam_agg <- ddply(bothmodel_outs, .(species, climsim, model), summarise,
@@
   ylab("Proportional difference")+
+  xlab("Climate simulation")+
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+
+dgd = position_dodge(width = 0.9)
+meanplot <- ggplot(meanparam_agg)+
+  geom_bar(aes(x=climsim, y=meandiff, fill=model),
+             position=dgd, stat="identity")+
+  theme_bw()+
+  scale_fill_manual(values=c("black", "grey45"), name="Model")+
+  facet_wrap("species", scales="free", nrow=1)+
+  scale_x_discrete(labels=c("+ppt", "+temp", "+ppt&temp"))+
+  ylab("Mean proportional difference")+
   xlab("Climate simulation")+
@@
 
+```{r figure_5, dependson="plot-options", fig.cap="Correlations between QBM and IPM estimates of climate effects.
+The QBM does not have multiple vital rates, so its values are repeated across panels within each species.", fig.width=6, fig.height=6, cache=FALSE, results='hide'}
+##  Figure script to look at correlations between IPM and QBM
+##    climate effect estimates.
+
+library(ggmcmc)
+library(gridExtra)
+library(ggplot2)
+
+allfiles <- list.files("../analysis/speciesData/")
+removals <- grep("csv", allfiles)
+species_list <- allfiles[-removals]
+
+clim_covs <- c("pptLag", "ppt1", "ppt2", "TmeanSpr1", "TmeanSpr2",
+               "ppt1xTmeanSpr1", "ppt2xTmeanSpr2",
+               "sizeXpptLag", "sizeXppt1", "sizeXppt2",
+               "sizeXTmeanSpr1", "sizeXTmeanSpr2")
+clim_mains <- c("pptLag", "ppt1", "ppt2", "TmeanSpr1", "TmeanSpr2",
+                "ppt1xTmeanSpr1", "ppt2xTmeanSpr2")
+
+####
+##  Get mean growth climate effects
+####
+growth_parms <- data.frame(species=NA, model=NA, parameter=NA, value=NA)
+for(spp in species_list){
+  tmp_data <- readRDS(paste("../analysis/ipm/vitalRateRegs/growth/growth_stanmcmc_", spp, ".RDS", sep=""))
+  tmp_clim <- tmp_data[grep("b2", tmp_data[,"Parameter"]), ]
+  tmp_clim[,"Parameter"] <- rep(clim_covs, each=3000)
+  tmp_agg <- ddply(tmp_clim, .(Parameter), summarise,
+                   value = median(value))
+  tmp_agg2 <- tmp_agg[which(tmp_agg$Parameter %in% clim_mains), ]
+  tmp_agg2$species <- spp
+  tmp_agg2$model <- "ipm"
+  colnames(tmp_agg2) <- tolower(colnames(tmp_agg2))
+  growth_parms <- rbind(growth_parms,
+                        tmp_agg2[,c("species", "model", "parameter", "value")])
+}
+
+####
+##  Get mean survival climate effects
+####
+surv_parms <- data.frame(species=NA, model=NA, parameter=NA, value=NA)
+params <- character(length = 12)
+for(i in 1:12){
+  params[i] <- paste("b2[",i,"]", sep="")
+}
+for(spp in species_list){
+  tmp_data <- readRDS(paste("../analysis/ipm/vitalRateRegs/survival/survival_stanmcmc_", spp, ".RDS", sep=""))
+  keeps <- which(tmp_data$Parameter %in% params)
+  tmp_clim <- tmp_data[keeps, ]
+  tmp_clim[,"Parameter"] <- rep(clim_covs, each=3000)
+  tmp_agg <- ddply(tmp_clim, .(Parameter), summarise,
+                   value = median(value))
+  tmp_agg2 <- tmp_agg[which(tmp_agg$Parameter %in% clim_mains), ]
+  tmp_agg2$species <- spp
+  tmp_agg2$model <- "ipm"
+  colnames(tmp_agg2) <- tolower(colnames(tmp_agg2))
+  surv_parms <- rbind(surv_parms,
+                        tmp_agg2[,c("species", "model", "parameter", "value")])
+}
+
+####
+##  Get mean recruitment climate effects
+####
+rec_parms <- data.frame(species=NA, model=NA, parameter=NA, value=NA)
+params <- character(length = 12)
+for(i in 1:12){
+  params[i] <- paste("b2[",i,"]", sep="")
+}
+for(spp in species_list){
+  tmp_data <- readRDS(paste("../analysis/ipm/vitalRateRegs/recruitment/recruitment_stanmcmc_", spp, ".RDS", sep=""))
+  keeps <- which(tmp_data$Parameter %in% params)
+  tmp_clim <- tmp_data[keeps, ]
+  tmp_clim[,"Parameter"] <- rep(clim_mains, each=3000)
+  tmp_agg <- ddply(tmp_clim, .(Parameter), summarise,
+                   value = median(value))
+  tmp_agg2 <- tmp_agg[which(tmp_agg$Parameter %in% clim_mains), ]
+  tmp_agg2$species <- spp
+  tmp_agg2$model <- "ipm"
+  colnames(tmp_agg2) <- tolower(colnames(tmp_agg2))
+  rec_parms <- rbind(rec_parms,
+                      tmp_agg2[,c("species", "model", "parameter", "value")])
+}
+
+####
+##  Get QBM growth climate estimates
+####
+clim_covs2 <- c("pptLag", "ppt1", "ppt2", "TmeanSpr1", "TmeanSpr2",
+               "ppt1xTmeanSpr1", "ppt2xTmeanSpr2",
+               "coverXpptLag", "coverXppt1", "coverXppt2",
+               "coverXTmeanSpr1", "coverXTmeanSpr2")
+
+qbm_parms <- data.frame(species=NA, model=NA, parameter=NA, qbmvalue=NA)
+params <- character(length = 12)
+for(i in 1:12){
+  params[i] <- paste("b2[",i,"]", sep="")
+}
+for(spp in species_list){
+  tmp_data <- readRDS(paste("../analysis/quadBM/vitalRateRegressions/truncNormModel/popgrowth_stanmcmc_", spp, ".RDS", sep=""))
+  keeps <- which(tmp_data$Parameter %in% params)
+  tmp_clim <- tmp_data[keeps, ]
+  tmp_clim[,"Parameter"] <- rep(clim_covs2, each=3000)
+  tmp_agg <- ddply(tmp_clim, .(Parameter), summarise,
+                   qbmvalue = median(value))
+  tmp_agg2 <- tmp_agg[which(tmp_agg$Parameter %in% clim_mains), ]
+  tmp_agg2$species <- spp
+  tmp_agg2$model <- "qbm"
+  colnames(tmp_agg2) <- tolower(colnames(tmp_agg2))
+  qbm_parms <- rbind(qbm_parms,
+                     tmp_agg2[,c("species", "model", "parameter", "qbmvalue")])
+}
+
+####
+##  Format data, combine, and plot
+####
+growth_parms <- growth_parms[2:nrow(growth_parms), ]
+growth_parms$vitalrate <- "growth"
+surv_parms <- surv_parms[2:nrow(surv_parms), ]
+surv_parms$vitalrate <- "surv"
+rec_parms <- rec_parms[2:nrow(rec_parms), ]
+rec_parms$vitalrate <- "rec"
+ipm_parms <- rbind(growth_parms, surv_parms, rec_parms)
+
+qbm_parms <- qbm_parms[2:nrow(qbm_parms), ]
+qbm_parms$vitalrate <- "growth"
+vitals <- c("surv", "rec")
+qbmalone <- qbm_parms
+for(dov in vitals){
+  tmp <- qbmalone
+  tmp$vitalrate <- dov
+  qbm_parms <- rbind(qbm_parms, tmp)
+}
+
+ipm_parms <- ipm_parms[,c("species","parameter","value","vitalrate")]
+qbm_parms <- qbm_parms[,c("species","parameter","qbmvalue","vitalrate")]
+all_ests <- merge(ipm_parms, qbm_parms, by = c("species", "parameter", "vitalrate"))
+
+# Calculate correlation for each group
+cors <- ddply(all_ests, c("species", "vitalrate"), summarise, 
+              cor = round(cor(value, qbmvalue), 2))
+
+ggplot(all_ests, aes(x=qbmvalue, y=value))+
+  geom_point(size=3)+
+  geom_smooth(method="lm", color="black", fill="grey")+
+  facet_grid(species~vitalrate)+
+  geom_text(data=cors, aes(label=paste("r = ", cor, sep="")), x=-1, y=1, size=4)+
+  xlab("QBM Estimate")+
+  ylab("IPM Estimate")+
+  theme_bw()
+
+```
+
 \pagebreak{}

### 2015-06-29 manuscript/manuscript.tex
updates figures in ms with parameter vary runs

@@
 
+An important caveat to the above conclusion is that our comparison of
+the IPM and the QBM
+
 This result is not as dire as it seems when we consider model
@@
 comata} (HECO), \emph{Pascopyrum smithii} (PASM), and \emph{Poa secunda}
-(POSE).}
+(POSE). Light grey lines show trajectories of individual quadrats. Note
+the different y-axis scales across panels.}
 \end{figure}
@@
 covariates.}
+\end{figure}
+
+\begin{figure}[htbp]
+\centering
+\includegraphics{components/figure/manuscript-figure_5.pdf}
+\caption{Correlations between QBM and IPM estimates of climate effects.
+The QBM does not have multiple vital rates, so its values are repeated
+across panels within each species.}
 \end{figure}

### 2015-06-29 FigureScripts/clim_eff_correlations.R

@@ -1,0 +1,152 @@
+##  Figure script to look at correlations between IPM and QBM
+##    climate effect estimates.
+
+library(ggmcmc)
+library(gridExtra)
+library(ggplot2)
+
+allfiles <- list.files("../analysis/speciesData/")
+removals <- grep("csv", allfiles)
+species_list <- allfiles[-removals]
+
+clim_covs <- c("pptLag", "ppt1", "ppt2", "TmeanSpr1", "TmeanSpr2",
+               "ppt1xTmeanSpr1", "ppt2xTmeanSpr2",
+               "sizeXpptLag", "sizeXppt1", "sizeXppt2",
+               "sizeXTmeanSpr1", "sizeXTmeanSpr2")
+clim_mains <- c("pptLag", "ppt1", "ppt2", "TmeanSpr1", "TmeanSpr2",
+                "ppt1xTmeanSpr1", "ppt2xTmeanSpr2")
+
+
+####
+##  Get mean growth climate effects
+####
+growth_parms <- data.frame(species=NA, model=NA, parameter=NA, value=NA)
+for(spp in species_list){
+  tmp_data <- readRDS(paste("../analysis/ipm/vitalRateRegs/growth/growth_stanmcmc_", spp, ".RDS", sep=""))
+  tmp_clim <- tmp_data[grep("b2", tmp_data[,"Parameter"]), ]
+  tmp_clim[,"Parameter"] <- rep(clim_covs, each=3000)
+  tmp_agg <- ddply(tmp_clim, .(Parameter), summarise,
+                   value = median(value))
+  tmp_agg2 <- tmp_agg[which(tmp_agg$Parameter %in% clim_mains), ]
+  tmp_agg2$species <- spp
+  tmp_agg2$model <- "ipm"
+  colnames(tmp_agg2) <- tolower(colnames(tmp_agg2))
+  growth_parms <- rbind(growth_parms,
+                        tmp_agg2[,c("species", "model", "parameter", "value")])
+}
+
+####
+##  Get mean survival climate effects
+####
+surv_parms <- data.frame(species=NA, model=NA, parameter=NA, value=NA)
+params <- character(length = 12)
+for(i in 1:12){
+  params[i] <- paste("b2[",i,"]", sep="")
+}
+for(spp in species_list){
[…]

### 2015-06-30 manuscript/manuscript.tex
unceratinty test notes added

--- 2015-06-18 manuscript/manuscript.tex
@@
 \includegraphics{components/figure/manuscript-figure_1.pdf}
-\caption{Time series of average percent cover of our four focal species:
-\emph{Bouteloua gracilis} (BOGR), \emph{Hesperostipa comata} (HECO),
-\emph{Pascopyrum smithii} (PASM), and \emph{Poa secunda} (POSE).}
+\caption{Time series of average percent cover over all quadrats for our
+four focal species: \emph{Bouteloua gracilis} (BOGR), \emph{Hesperostipa
+comata} (HECO), \emph{Pascopyrum smithii} (PASM), and \emph{Poa secunda}
+(POSE). Light grey lines show trajectories of individual quadrats. Note
+the different y-axis scales across panels.}
 \end{figure}

### 2015-07-06 manuscript/manuscript.Rmd
updates with new quad exclusion figure

@@
 Informal conversations with Stephen Ellner, Giles Hooker, Robin Snyder, and a series of meetings among the Adler and Weecology labs at USU sharpened our thinking.
+Compute, storage and other resources from the Division of Research Computing in the Office of Research and Graduate Studies at Utah State University are gratefully acknowledged.
 
@@
 secunda* (*P* = `r mae_ps[4,"value"]`).
+*P* values are highly sensitive to sample size, so not entirely appropriate in simulation exercises where we control the samples size.
+But, for our purposes they serve as relatively unbiased comparison metrics.
 
@@
 ```{r figure_3, dependson="plot-options", fig.cap="Proportional change in species' mean cover caused by a 1% increase in observed precipitation (+ppt), temperature (+temp), or both (+ppt&temp) as predicted by the individual-based IPM and the aggregate-based QBM.
-Top panels show the mean predicted proportional change in cover; lower panels show the 90% quantiles of predicted proportional cover change.
-Mean predictions and their associated uncertainties are shown in separate panels for visual clarity.", fig.width=8.5, fig.height=6, cache=FALSE, results='hide'}
+Top panels show the mean predicted proportional change in cover.
+Lower panels show the sensitivity of equilibrium cover simulated from the IPM to each climate scenario applied to individual and combined vital rates.
+For example, the points associated with G show the median cover from IPM simulations where a climate perturbation is applied only to the growth regression climate covariates.", fig.width=8.5, fig.height=5, cache=FALSE, results='hide'}
 setwd("~/Repos/MicroMesoForecast/analysis/")
@@
 ####
-# meanparam_ipm <- data.frame(cover=NA, species=NA, 
-#                              climsim=NA, model=NA)
-# ##  IPM 
-# path <- "./ipm/simulations/results/climchange_meanparams/"
-# ipmfiles_mean <- list.files(path)
-# ipmfiles_mean <- ipmfiles_mean[grep("cover", ipmfiles_mean)]
-# for(file in ipmfiles_mean){
-#   tmp <- read.csv(paste(path,file,sep=""))
-#   tmp$model <- "IPM"
-#   meanparam_ipm <- rbind(meanparam_ipm, tmp[,c("cover", "species", "climsim", "model")])
-# }
-# meanparam_ipm <- meanparam_ipm[2:nrow(meanparam_ipm),]
-# 
-# newd <- data.frame(cover=NA, species=NA, 
-#                    climsim=NA, model=NA, obscover=NA)
-# for(dospp in unique(meanparam_ipm$species)){
-#   tmp <- subset(meanparam_ipm, species==dospp)
-#   obscover <- subset(tmp, climsim=="obs")["cover"]
-#   tmp2 <- subset(tmp, climsim!="obs")
-#   tmp2$obscover <- rep(obscover$cover, times = length(unique(tmp2$climsim)))
-#   newd <- rbind(newd, tmp2)
-# }
-# ipm_meansims <- newd[2:nrow(newd),]
-# ipm_meansims$propdiff <- with(ipm_meansims, log(cover*100)-log(obscover*100))
-# 
-# ##  QBM
-# meanparam_qbm <- data.frame(cover=NA, species=NA, 
-#                              climsim=NA, model=NA)
-# path <- "./quadBM/simulations/results/climatechange_meanparams/"
-# qbmfiles_mean <- list.files(path)
-# for(file in qbmfiles_mean){
-#   tmp <- readRDS(paste(path,file,sep=""))
-#   tmp$model <- "QBM"
-#   meanparam_qbm <- rbind(meanparam_qbm, tmp[,c("cover", "species", "climsim", "model")])
-# }
-# meanparam_qbm <- meanparam_qbm[2:nrow(meanparam_qbm),]
-# 
-# newd <- data.frame(cover=NA, species=NA, 
-#                    climsim=NA, model=NA, obscover=NA)
-# for(dospp in unique(meanparam_qbm$species)){
-#   tmp <- subset(meanparam_qbm, species==dospp)
-#   obscover <- subset(tmp, climsim=="obs")["cover"]
-#   tmp2 <- subset(tmp, climsim!="obs")
-#   tmp2$obscover <- rep(obscover$cover, times = length(unique(tmp2$climsim)))
-#   newd <- rbind(newd, tmp2)
-# }
-# qbm_meansims <- newd[2:nrow(newd),]
-# qbm_meansims$propdiff <- with(qbm_meansims, log(cover*100)-log(obscover*100))
-# 
-# ##  Merge IPM and QBM mean parameter results
-# bothmodel_outs <- rbind(ipm_meansims, qbm_meansims)
-# meanparam_agg <- ddply(bothmodel_outs, .(species, climsim, model), summarise,
-#                        meandiff = mean(propdiff),
-#                        hidiff = quantile(propdiff, probs = 0.95),
-#                        lodiff = quantile(propdiff, probs = 0.05))
-# dgd = position_dodge(width = 0.9)
-# meanplot <- ggplot(meanparam_agg)+
-#   geom_bar(aes(x=climsim, y=meandiff, fill=model),
-#              position=dgd, stat="identity")+
-#   theme_bw()+
-#   scale_fill_manual(values=c("black", "grey45"), name="Model")+
-#   facet_wrap("species", scales="free", nrow=1)+
-#   scale_x_discrete(labels=c("+ppt", "+temp", "+ppt&temp"))+
-#   ylab("Mean proportional difference")+
-#   xlab("Climate simulation")+
-#   theme(axis.text.x = element_text(angle = 45, hjust = 1))
-#
-
+meanparam_ipm <- data.frame(cover=NA, species=NA, 
+                             climsim=NA, model=NA)
+##  IPM 
+path <- "./ipm/simulations/results/climchange_meanparams/"
+ipmfiles_mean <- list.files(path)
+ipmfiles_mean <- ipmfiles_mean[grep("cover", ipmfiles_mean)]
+for(file in ipmfiles_mean){
+  tmp <- read.csv(paste(path,file,sep=""))
+  tmp$model <- "IPM"
+  meanparam_ipm <- rbind(meanparam_ipm, tmp[,c("cover", "species", "climsim", "model")])
+}
+meanparam_ipm <- meanparam_ipm[2:nrow(meanparam_ipm),]
+
+newd <- data.frame(cover=NA, species=NA, 
+                   climsim=NA, model=NA, obscover=NA)
+for(dospp in unique(meanparam_ipm$species)){
+  tmp <- subset(meanparam_ipm, species==dospp)
+  obscover <- subset(tmp, climsim=="obs")["cover"]
+  tmp2 <- subset(tmp, climsim!="obs")
+  tmp2$obscover <- rep(obscover$cover, times = length(unique(tmp2$climsim)))
+  newd <- rbind(newd, tmp2)
+}
+ipm_meansims <- newd[2:nrow(newd),]
+ipm_meansims$propdiff <- with(ipm_meansims, log(cover*100)-log(obscover*100))
+
+##  QBM
+meanparam_qbm <- data.frame(cover=NA, species=NA, 
+                             climsim=NA, model=NA)
+path <- "./quadBM/simulations/results/climatechange_meanparams/"
+qbmfiles_mean <- list.files(path)
+for(file in qbmfiles_mean){
+  tmp <- readRDS(paste(path,file,sep=""))
+  tmp$model <- "QBM"
+  meanparam_qbm <- rbind(meanparam_qbm, tmp[,c("cover", "species", "climsim", "model")])
+}
+meanparam_qbm <- meanparam_qbm[2:nrow(meanparam_qbm),]
+
+newd <- data.frame(cover=NA, species=NA, 
+                   climsim=NA, model=NA, obscover=NA)
+for(dospp in unique(meanparam_qbm$species)){
+  tmp <- subset(meanparam_qbm, species==dospp)
+  obscover <- subset(tmp, climsim=="obs")["cover"]
+  tmp2 <- subset(tmp, climsim!="obs")
+  tmp2$obscover <- rep(obscover$cover, times = length(unique(tmp2$climsim)))
+  newd <- rbind(newd, tmp2)
+}
+qbm_meansims <- newd[2:nrow(newd),]
+qbm_meansims$propdiff <- with(qbm_meansims, log(cover*100)-log(obscover*100))
+
+##  Merge IPM and QBM mean parameter results
+bothmodel_outs <- rbind(ipm_meansims, qbm_meansims)
+meanparam_agg <- ddply(bothmodel_outs, .(species, climsim, model), summarise,
+                       meandiff = mean(propdiff),
+                       hidiff = quantile(propdiff, probs = 0.95),
+                       lodiff = quantile(propdiff, probs = 0.05))
+dgd = position_dodge(width = 0.9)
+meanplot <- ggplot(meanparam_agg)+
+  geom_bar(aes(x=climsim, y=meandiff, fill=model),
+             position=dgd, stat="identity")+
+  theme_bw()+
+  scale_fill_manual(values=c("black", "grey45"), name="Model")+
+  facet_wrap("species", scales="free", nrow=1)+
+  scale_x_discrete(labels=c("+ppt", "+temp", "+ppt&temp"))+
+  ylab("Mean proportional difference")+
+  xlab("Climate simulation")+
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+
+####
+####  Make IPM sensitivity plot for lower panels
+####
+setwd("../analysis/ipm/simulations/results/climate_sensitivity/")
+files <- list.files()
+cover_files <- files[grep("cover", files)]
+spp_id <- substr(cover_files, 1, 4)
+num_files <- length(cover_files)
+all_sims <- data.frame(time=NA, cover=NA, species=NA, climsim=NA, vital=NA)
+for(i in 1:num_files){
+  tmp <- read.csv(cover_files[i])
+  tmp$species <- rep(spp_id[i], nrow(tmp))
+  all_sims <- rbind(all_sims, tmp)
+}
+all_sims <- all_sims[2:nrow(all_sims),]
+
+# Rename vital rates with codes
+vital_names <- unique(all_sims$vital)
+vital_codes <- c("A", "G", "GR", "GS", "R", "S" , "SR") 
+vitals <- data.frame(names=vital_names,code=vital_codes)
+for(vnow in vital_names){
+  ids <- which(all_sims[,"vital"]==vnow)
+  idcode <- which(vitals[,"names"]==vnow)
+  all_sims[ids,"vital"] <- as.character(vitals[idcode,"code"])
+}
+
+# Rep the 'all' simulation for each vital rate for plotting
+all_clim_sims <- subset(all_sims, vital=="A")
+species <- unique(all_sims$species)
+out_controls <- data.frame(time=NA, cover=NA, species=NA, climsim=NA, vital=NA)
+for(spp in species){
+  tmp <- subset(all_clim_sims, species==spp)
+  for(vital in vital_codes){
+    tmp$vital <- vital
+    out_controls <- rbind(out_controls, tmp)
+  }
+}
+out_controls <- out_controls[2:nrow(out_controls),]
+
+# Combine the datasets
+all_noclim_sims <- subset(all_sims, vital!="all")
+final <- rbind(out_controls, all_noclim_sims)
+
+# Calculate statistics for plotting
+equilibrium_cover <- ddply(final, .(species, climsim, vital), summarise,
+                           eq_cover = median(cover))
+mean_cover <- ddply(subset(equilibrium_cover, climsim=="all"), .(species), summarise,
+                    value = mean(eq_cover))
+
+equilibrium_cover <- subset(equilibrium_cover, vital!="A")
+bothid <- which(equilibrium_cover[,"climsim"]=="ppttemp")
+equilibrium_cover[bothid,"climsim"] <- "zppttemp"
+# myCols2 <- c("#277BA8", "#7ABBBD", "#AED77A")
+sensplot <- ggplot(equilibrium_cover, aes(x=climsim, y=eq_cover*100, 
+                              shape=vital, group=vital, linetype=vital))+
+  geom_hline(data=mean_cover, aes(yintercept=value*100), linetype=2, color="grey45")+
+  geom_line()+
+  geom_point(size=4)+
+  facet_wrap("species", scales = "free", nrow=1)+
+  scale_linetype_discrete(name="Vital Rate")+
+  scale_shape_discrete(name="Vital Rate")+
+  scale_x_discrete(labels=c("baseline", "+ppt", "+temp", "+ppt&temp"))+
+  ylab("Equilibrium Cover (%)")+
+  xlab("Simulation")+
+  theme_bw()+
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+
+covertest <- subset(equilibrium_cover, climsim=="all")
+colnames(covertest)[4] <- "obs_cover"
+out <- merge(subset(equilibrium_cover, climsim!="all"), covertest[,c("species","vital","obs_cover")])
+out$propdiff <- with(out, log(eq_cover*100)-log(obs_cover*100))
+
+meanparamplot <- grid.arrange(meanplot, sensplot, ncol=1, nrow=2)
+print(meanparamplot)
+
+setwd("~/Repos/MicroMesoForecast/manuscript")
+```
+
+```{r figure_4, dependson="plot-options", fig.cap="Equilibrium cover and 90% quantiles around the mean prediction when model and parameter uncertainty are propogated through the simulation phase.
+Climate simulations are as in Figure 3.", fig.width=4.5, fig.height=7, cache=FALSE, results='hide'}
 ####
@@
 ####
+setwd("~/Repos/MicroMesoForecast/analysis/")
 meanparam_ipm <- data.frame(cover=NA, species=NA, 
@@
 dgd = position_dodge(width = 0.6)
-uncertainplot <- ggplot(meanparam_agg)+
+ggplot(meanparam_agg)+
   geom_errorbar(aes(x=climsim, ymax=hidiff, ymin=lodiff, group=model, linetype=model),
+                position=dgd, width=0.25)+
+  geom_point(aes(x=climsim, y=meandiff, group=model, shape=model),
                 position=dgd, width=0.25)+
@@
   scale_linetype_discrete(name="Model")+
-  facet_wrap("species", nrow=1, scales="free")+
+  scale_shape_discrete(name="Model")+
+  facet_wrap("species", ncol=1, scales="free")+
   scale_x_discrete(labels=c("+ppt", "+temp", "+ppt&temp"))+
   ylab("Proportional difference")+
-  xlab("Climate simulation")+
-  theme(axis.text.x = element_text(angle = 45, hjust = 1))
-
-dgd = position_dodge(width = 0.9)
-meanplot <- ggplot(meanparam_agg)+
-  geom_bar(aes(x=climsim, y=meandiff, fill=model),
-             position=dgd, stat="identity")+
-  theme_bw()+
-  scale_fill_manual(values=c("black", "grey45"), name="Model")+
-  facet_wrap("species", scales="free", nrow=1)+
-  scale_x_discrete(labels=c("+ppt", "+temp", "+ppt&temp"))+
-  ylab("Mean proportional difference")+
-  xlab("Climate simulation")+
-  theme(axis.text.x = element_text(angle = 45, hjust = 1))
-
-meanparamplot <- grid.arrange(meanplot, uncertainplot, ncol=1, nrow=2)
-print(meanparamplot)
-```
-
-```{r figure_4, dependson="plot-options", fig.cap="Sensitivity of equilibrium cover to a 1% increase in precipitation (+ppt), temperature (+temp), or both (+ppt&temp) applied to individual and combined vital rates.
-For example, the points associated with G show the median cover from IPM simulations where a climate perturbation is applied only to the growth regression climate covariates.", fig.width=4.5, fig.height=6, cache=FALSE, results='hide'}
-setwd("../analysis/ipm/simulations/results/climate_sensitivity/")
-files <- list.files()
-cover_files <- files[grep("cover", files)]
-spp_id <- substr(cover_files, 1, 4)
-num_files <- length(cover_files)
-all_sims <- data.frame(time=NA, cover=NA, species=NA, climsim=NA, vital=NA)
-for(i in 1:num_files){
-  tmp <- read.csv(cover_files[i])
-  tmp$species <- rep(spp_id[i], nrow(tmp))
-  all_sims <- rbind(all_sims, tmp)
-}
-all_sims <- all_sims[2:nrow(all_sims),]
-
-# Rename vital rates with codes
-vital_names <- unique(all_sims$vital)
-vital_codes <- c("A", "G", "GR", "GS", "R", "S" , "SR") 
-vitals <- data.frame(names=vital_names,code=vital_codes)
-for(vnow in vital_names){
-  ids <- which(all_sims[,"vital"]==vnow)
-  idcode <- which(vitals[,"names"]==vnow)
-  all_sims[ids,"vital"] <- as.character(vitals[idcode,"code"])
-}
-
-# Rep the 'all' simulation for each vital rate for plotting
-all_clim_sims <- subset(all_sims, vital=="A")
-species <- unique(all_sims$species)
-out_controls <- data.frame(time=NA, cover=NA, species=NA, climsim=NA, vital=NA)
-for(spp in species){
-  tmp <- subset(all_clim_sims, species==spp)
-  for(vital in vital_codes){
-    tmp$vital <- vital
-    out_controls <- rbind(out_controls, tmp)
-  }
-}
-out_controls <- out_controls[2:nrow(out_controls),]
-
-# Combine the datasets
-all_noclim_sims <- subset(all_sims, vital!="all")
-final <- rbind(out_controls, all_noclim_sims)
-
-# Calculate statistics for plotting
-equilibrium_cover <- ddply(final, .(species, climsim, vital), summarise,
-                           eq_cover = median(cover))
-mean_cover <- ddply(subset(equilibrium_cover, climsim=="all"), .(species), summarise,
-                    value = mean(eq_cover))
-
-equilibrium_cover <- subset(equilibrium_cover, vital!="A")
-bothid <- which(equilibrium_cover[,"climsim"]=="ppttemp")
-equilibrium_cover[bothid,"climsim"] <- "zppttemp"
-# myCols2 <- c("#277BA8", "#7ABBBD", "#AED77A")
-ggplot(equilibrium_cover, aes(x=climsim, y=eq_cover*100, 
-                              shape=vital, group=vital, linetype=vital))+
-  geom_hline(data=mean_cover, aes(yintercept=value*100), linetype=2, color="grey45")+
-  geom_line()+
-  geom_point(size=4)+
-  facet_grid(species~., scales = "free")+
-  scale_linetype_discrete(name="Vital Rate")+
-  scale_shape_discrete(name="Vital Rate")+
-  scale_x_discrete(labels=c("baseline", "+ppt", "+temp", "+ppt&temp"))+
-  ylab("Equilibrium Cover (%)")+
-  xlab("Simulation")+
-  theme_bw()+
-  theme(axis.text.x = element_text(angle = 45, hjust = 1))
-
-covertest <- subset(equilibrium_cover, climsim=="all")
-colnames(covertest)[4] <- "obs_cover"
-out <- merge(subset(equilibrium_cover, climsim!="all"), covertest[,c("species","vital","obs_cover")])
-out$propdiff <- with(out, log(eq_cover*100)-log(obs_cover*100))
-setwd("~/Repos/MicroMesoForecast/manuscript")
+  xlab("Climate simulation")
+#   theme(axis.text.x = element_text(angle = 45, hjust = 1))
+setwd("~/Repos/MicroMesoForecast/manuscript/")
 ```
@@
 ```{r figure_5, dependson="plot-options", fig.cap="Correlations between QBM and IPM estimates of climate effects.
+We ignore sizeXclimate interactions since these are not directly comparable across model types.
 The QBM does not have multiple vital rates, so its values are repeated across panels within each species.", fig.width=6, fig.height=6, cache=FALSE, results='hide'}
@@
 
+setwd("~/Repos/MicroMesoForecast/manuscript/")
+```
+
+```{r figure_6, dependson="plot-options", fig.cap="Effect of quadrat sample size on the precision (standard deviation) of main climate effect estimates in the QBM.
+Increasing the number of quadrats results in diminishing returns in terms of parameter certainty.
+Light dashed lines show individual climate effects at five quadrat sample sizes.
+Thick dark lines are inverse gaussian fits showing the mean effect of increasing quadrat sample size on parameter precision.", fig.width=3, fig.height=7.5, cache=FALSE, results='hide'}
+library(mgcv)
+
+setwd("../analysis/quadBM/vitalRateRegressions/exclude_quad_test")
+
+####
+####  Read in model fits and summarize results
+####
+source("popgrowth_read_data.R")
+spp <- unique(growD_all$Species)
+num_quads <- numeric(length(spp))
+for(i in 1:length(spp)){
+  tmp <- subset(growD_all, Species==spp[i])
+  num_quads[i] <- length(unique(tmp$quad))
+}
+num_quads <- as.data.frame(num_quads)
+num_quads$species <- spp
+
+all_files <- list.files("./results/")
+all_fits <- data.frame(Parameter=NA, stddev=NA, id=NA, 
+                       rmsquad=NA, rep=NA, species=NA)
+for(i in 1:length(all_files)){
+  tmp <- all_files[i]
+  spp <- substring(tmp, 1, 4)
+  qsrm1 <- unlist(strsplit(tmp, "[_]"))[2]
+  qsrm <- gsub("numout", "", qsrm1)
+  rep1 <- unlist(strsplit(tmp, "[_]"))[3]
+  rep2 <- gsub(".RDS","",rep1)
+  rep <- gsub("rep", "", rep2)
+
+  long <- readRDS(paste("./results/",tmp, sep=""))
+  short <- ddply(long, .(Parameter), summarise,
+                 stddev = sd(value))
+  climeff <- short[grep("b2", short$Parameter),]
+  climeff$id <- substr(climeff$Parameter, 4, length(climeff$Parameter))
+  climeff$id <- as.numeric(unlist(strsplit(climeff$id, split=']')))  
+  climeff <- climeff[with(climeff, order(id)),]
+  climeff$rmsquad <- qsrm
+  climeff$rep <- rep
+  climeff$species <- spp
+  all_fits <- rbind(all_fits, climeff)
+} # end file loop
+
+all_fits <- all_fits[2:nrow(all_fits),]
+
+####
+####  Read in full fits
+####
+dir <- "../truncNormModel/"
+files <- list.files(dir)
+files <- files[grep(".RDS", files)]
+all_alls <- data.frame(Parameter=NA, stddev=NA, id=NA, 
+                       rmsquad=NA, rep=NA, species=NA)
+for(i in 1:length(files)){
+  tmp <- files[i]
+  spp1 <- unlist(strsplit(tmp, "[_]"))[3]
+  spp <- gsub(".RDS","",spp1)
+
+  long <- readRDS(paste(dir,tmp, sep=""))
+  short <- ddply(long, .(Parameter), summarise,
+                 stddev = sd(value))
+  climeff <- short[grep("b2", short$Parameter),]
+  climeff$id <- substr(climeff$Parameter, 4, length(climeff$Parameter))
+  climeff$id <- as.numeric(unlist(strsplit(climeff$id, split=']')))  
+  climeff <- climeff[with(climeff, order(id)),]
+  climeff$rmsquad <- 0
+  climeff$rep <- 1
+  climeff$species <- spp
+  all_alls <- rbind(all_alls, climeff)
+}
+all_alls <- all_alls[2:nrow(all_alls),]
+
+####
+####  Aggregate results by species and rep
+####
+all_fits <- rbind(all_fits, all_alls)
+agg_fits <- ddply(all_fits, .(Parameter, rmsquad, species, id), summarise,
+                  avg_stddev = mean(stddev))
+agg_fits2 <- agg_fits[which(agg_fits$id %in% c(1:5)),]
+agg_fits2 <- merge(agg_fits2, num_quads, by="species")
+agg_fits2$quadsfit <- with(agg_fits2, num_quads-as.numeric(rmsquad))
+
+####
+####  Fit model (need to do by species)
+####
+modBOGR <- gam(avg_stddev~quadsfit, data=subset(agg_fits2, species=="BOGR"),
+               family=quasi(link="1/mu^2"))
+modHECO <- gam(avg_stddev~quadsfit, data=subset(agg_fits2, species=="HECO"),
+               family=quasi(link="1/mu^2"))
+modPASM <- gam(avg_stddev~quadsfit, data=subset(agg_fits2, species=="PASM"),
+               family=quasi(link="1/mu^2"))
+modPOSE <- gam(avg_stddev~quadsfit, data=subset(agg_fits2, species=="POSE"),
+               family=quasi(link="1/mu^2"))
+summary(modBOGR)
+summary(modHECO)
+summary(modPASM)
+summary(modPOSE)
+
+####
+#### Plot
+####
+ggplot(agg_fits2, aes(x=quadsfit, y=avg_stddev))+
+  geom_line(aes(group=Parameter), alpha=0.3, linetype=2)+
+  geom_point(aes(group=Parameter), alpha=0.3, size=3)+
+#   stat_smooth(method="loess", size=2, color="black", se=FALSE)+
+  geom_smooth(method="glm", formula=y~x, family=quasi(link="1/mu^2"),
+              size=2, color="black", se=FALSE)+
+  facet_wrap("species", scales = "free", ncol=1)+
+  theme_bw()+
+  xlab("Number of quadrats fit")+
+  ylab("Standard deviation of coefficient")
+
 ```

### 2015-07-06 manuscript/manuscript.tex
updates with new quad exclusion figure

--- 2015-06-29 manuscript/manuscript.tex
@@
 series of meetings among the Adler and Weecology labs at USU sharpened
-our thinking.
+our thinking. Compute, storage and other resources from the Division of
+Research Computing in the Office of Research and Graduate Studies at
+Utah State University are gratefully acknowledged.
 
@@
 are statisticially similar between models for \emph{P. secunda}
-(\emph{P} = 0.0922).
+(\emph{P} = 0.0922). \emph{P} values are highly sensitive to sample
+size, so not entirely appropriate in simulation exercises where we
+control the samples size. But, for our purposes they serve as relatively
+unbiased comparison metrics.
 
@@
 aggregate-based QBM. Top panels show the mean predicted proportional
-change in cover; lower panels show the 90\% quantiles of predicted
-proportional cover change. Mean predictions and their associated
-uncertainties are shown in separate panels for visual clarity.}
+change in cover. Lower panels show the sensitivity of equilibrium cover
+simulated from the IPM to each climate scenario applied to individual
+and combined vital rates. For example, the points associated with G show
+the median cover from IPM simulations where a climate perturbation is
+applied only to the growth regression climate covariates.}
 \end{figure}
@@
 \includegraphics{components/figure/manuscript-figure_4.pdf}
-\caption{Sensitivity of equilibrium cover to a 1\% increase in
-precipitation (+ppt), temperature (+temp), or both (+ppt\&temp) applied
-to individual and combined vital rates. For example, the points
-associated with G show the median cover from IPM simulations where a
-climate perturbation is applied only to the growth regression climate
-covariates.}
+\caption{Equilibrium cover and 90\% quantiles around the mean prediction
+when model and parameter uncertainty are propogated through the
+simulation phase. Climate simulations are as in Figure 3.}
 \end{figure}
@@
 \caption{Correlations between QBM and IPM estimates of climate effects.
-The QBM does not have multiple vital rates, so its values are repeated
-across panels within each species.}
+We ignore sizeXclimate interactions since these are not directly
+comparable across model types. The QBM does not have multiple vital
+rates, so its values are repeated across panels within each species.}
+\end{figure}
+
+\begin{figure}[htbp]
+\centering
+\includegraphics{components/figure/manuscript-figure_6.pdf}
+\caption{Effect of quadrat sample size on the precision (standard
+deviation) of main climate effect estimates in the QBM. Increasing the
+number of quadrats results in diminishing returns in terms of parameter
+certainty. Light dashed lines show individual climate effects at five
+quadrat sample sizes. Thick dark lines are inverse gaussian fits showing
+the mean effect of increasing quadrat sample size on parameter
+precision.}
 \end{figure}

### 2015-07-06 manuscript/manuscript.Rmd
draft 2 of ms

@@
 \begin{align}
-\text{logit}(S_{ijQ,t}) &= \gamma^{S}_{j,t} + \phi^{S}_{jQ} + \beta^{S}_{j,t}x_{ij,t} + \omega^{S}_{j}w_{ij,t} + \theta^{S}_{jk}C_{k,t} + \varepsilon^{S}_{t} \\
+\text{logit}(S_{ijQ,t}) &= \gamma^{S}_{j,t} + \phi^{S}_{jQ} + \beta^{S}_{j,t}x_{ij,t} + \omega^{S}_{j}w_{ij,t} + \theta^{S}_{jk}C_{k,t} \\
 Y^{S}_{ijQ,t} &\sim \text{Bernoulli}(S_{ijQ,t})
@@
 where $x_{jq,t}$ is the log of species' $j$ proportional cover in quadrat $q$ at time $t$ and all other parameters are as in the individual-level growth model (Eq.
-#).
+3).
 The log normal likelihood includes a truncation (T[0,1]) to ensure that predicted values do not exceed 100% cover.
@@
 #### Quad-based model
-Our quad-based model (QBM) perfectly mirrors its statistical description (Eq. #). We use the same approach for drawing parameter values as described for the IPM.
+Our quad-based model (QBM) perfectly mirrors its statistical description (Eq. 8-9). We use the same approach for drawing parameter values as described for the IPM.
 
@@
 3d).
-As expected based on model validation (Table 1), QBM projections were more uncertain than IPM projections for all species except *P.
-smithiii* (Fig.
-3e-h).
 
@@
 Species showed similar trends (Fig.
-4).
+3e-h).
 Growth was the most sensitive vital rate for all species, showing a negative response to increased precipitation, and stronger positive response to increased temperature, and a mostly positive response when both climate factors are increased (Fig.
-4).
+3e-h).
 *B.
 gracilis* survival rates were sensitive to temperature, resulting in an increase in plant cover under increased temperature (Fig.
-4a).
+3e).
 In isolation, recruitment and survival were insensitive to climate factors for *H.
 comata* (Fig.
-4b).
+3f).
 Survival and recruitment of *P.
 smithii* were both sensitive, negatively, to temperature and precipitation (Fig.
-4c).
+3g).
 *P.
 secunda* equilibrium cover was sensitive to the climate effects on survival and recruitment, showing a negative effect on both vital rates for increased precipition, but a strong positive effect on survival with increased temperature (Fig.
-4d).
+3h).
 The climate impact of recruitment on equilibrium cover was negative for precipitation and temperature increases (Fig.
-4d).
+3h).
+
+Forecasts based on 1% climate changes were extremely uncertain when we considered model error and parameter uncertainty (Fig.
+4).
+As expected based on model validation (Table 1), QBM projections were more uncertain than IPM projections for all species except *P.
+smithiii* (Fig.
+4).
 
@@
 ----------
-Perhaps the greatest challenge for ecology in the 21st century is to forecast the impacts of environmental change [@Clark2001].
-To do so requires sophisticated modeling approaches that fully account for uncertainty and variability in the ecological process and associated parameters.
+Perhaps the greatest challenge for ecology in the 21st century is to forecast the impacts of environmental change [@Clark2001; @Petchey2015].
+To do so requires sophisticated modeling approaches that fully account for uncertainty and variability in the ecological process and associated parameters [@Luo2011a].
 This requires large amounts of data collected over large spatio-temporal extents.
@@
 This is unsurprising since we expected to lose some precision at the aggregated level.
-However, the underwhelming performance of the QBM could call in the question forecasts that differ from the IPM.
+However, the underwhelming performance of the QBM could call into the question forecasts that differ from the IPM.
 
@@
 The QBM failed to match IPM forecasts, implying that detailed demographic data may be necessary to accurately detect climate signals that are utlimately important at the population level.
-This result further confirms related work on the importance of individual variability on population level responses to exogenous drivers (CITATIONS).
-
-An important caveat to the above conclusion is that our comparison of the IPM and the QBM
-
-This result is not as dire as it seems when we consider model uncerainty.
-Yes, the QBM failed to match IPM forecasts under perturbed climate, but the uncertainties around those forecasts were so great (Fig.
-3e-h) that we cannot say that the two model's forecasts were different with any level of confidence.
-This offers hope that with "enough" population level data one will be able to reduce parameter uncertainty and thus model forecasts.
-This remains an open question.
+This result further confirms related work on the importance of individual variability on population level responses to exogenous drivers [@Clark2011; @Clark2011b; @Clark2012; @Galvan2014].
 
 ### The role of vital rate climate dependence
+We can think of two reasons why the IPM outperformed the QBM.
+First, the quadrat level data has a much reduced sample size compared to the individual level data.
+In an ideal world we would have compared the IPM and QBM using data collected over the same amount of person hours, not just the same number of quadrats.
+Then the sample size of the quadrat level data would be much greater and carry more statistical power.
+To address this limitation in our work we fit the QBM statistical model (Eq.
+8-9) with different numbers of quadrats to see the effect of sample size on the precision of climate effect estimates.
+It appears that including additional quadrats leads to rapidly diminishing returns in terms of parameter precision (Fig.
+5).
+Thus, while sample size surely plays some role, we do not think it is the main driver of the difference between the IPM and QBM.
+
+The second reason the IPM could have outperformed the QBM is that the population level model is in fact missing important climate effects that act on individual vital rates, rather than population growth.
 Our intuition was that species with strong climate-dependence on vital rates not well resolved at the aggregate level would result in different forecasts from the two models.
@@
 So, if survival of individuals was positively impacted by temperature increases, for example, we would expect to detect this signal in the individual level data but not in the aggregate level data.
-**PETER: do you think this is the right direction to go with this?**
+To see if this is the case we can regress climate effects from each vital rate statistical model at the individual level against the same climate effects from the QBM statistical model (Fig.
+6).
+
+In general, the QBM climate effects are most correlated with climate effects from the growth model at the individual level (Fig.
+6).
+In no case does the QBM statistical model have strong correlations across all three vital rates (Fig.
+6).
+Thus, for each species the QBM is "missing" climate signals associated with at least one vital rate.
+This has large impacts on predictions of long term population dynamics, as seen in our equilibrium simulations (Fig.
+3a-d).
 
@@
 The analysis we conducted here could be considered, with some exceptions of course, at the forefront of ecological forecasting in terms of the statistical approach employed (hierarchical Bayesian), the type of population model we used (stochastic IPM with parameter uncertainty), and the amount of data we had at our disposal (14 years of individual-level data).
-Yet, model predictions proved so uncertain that any forecast, when bounded with uncertainty, would be at best not useful and at worst meaningless.
+Yet, model predictions proved so uncertain that any forecast, when bounded with model and parameter uncertainty, would be at best not useful and at worst meaningless.
 How might we improve on this state of affairs?
@@
 Climate covariates only vary by year, with no spatial variability within years.
-Thus, even if we fit models to individual level data, we are missing the key interaction point between weather and individual plants.
-**PETER: any good citations here for decent population models fit with fine-scale climate/environmental data??**
+Thus, even if we fit models to individual level data, we are missing the key interaction point between weather and individual plants [@Clark2011b].
 
 Second, accurately detecting climate signals will take even longer time series.
-Recent theoretical work on detecting climate signals in noisy data suggests that even advanced approaches to parameter fitting like LASSO and random forest require 20-25 year time series (Teller et al., in review).
+Recent theoretical work on detecting climate signals in noisy data suggests that even advanced approaches to parameter fitting like LASSO, functional linear models (splines), and random forest require 20-25 year time series (Teller et al., in review).
+Alternatively, as we suggest above, Teller et al.
+(in review) also find that matching the scale of the response and predictors improves estimate precision.
+
+Third, ecologists as a community need to get serious about reporting uncertainty.
+There is a strong culture around explicitly considering model uncertainty, but parameter uncertainty is often ignored.
+In some cases this is because the easiest statistical methods do no make propogating parameter uncertainty easy.
+But even using Bayesian approaches that allow straightforward integration of model fitting and forecasting [@Hooten2014; @Hobbs2015] is not simple when using modeling approaches like integral projection models that separate the model fitting and simulation stages [@Rees2009].
+However, as we have done here, it is still possible to include parameter uncertainty by drawing parameter values from MCMC iterations, taking care to draw all parameters from the same chain and iteration to account for their correlations.
+Only by being honest about our forecasts cen we begin to produce better ones.
 
@@
 -----------
-
 This work is not a critique of density-structured population models.
-Instead, we consider our results a cautionary tale.
 In some cases and for certain species, population models based on aggregated data may prove useful and unbiased.
-But, as our work shows, that is not a foregone conclusion when considering the impact of exogenous drivers like climate on population states.
+However, our work here is the first comparison of population models based on individual and aggregated data in a climate change context.
+Our results confirm theoretical arguments [@Clark2011b] and empirical evidence [@Clark2011; @Clark2012] that individual responses are critical to predicting species' responses to climate change.
 Thus, forecasts from aggregate level models should be viewed with caution and should never be unaccompianed by uncertainty.
-As we have shown, uncertainty around forecasts was large for both model types.
+
+Our results also offer a cautionary tale because uncertainty around forecasts was large for both model types.
 Which leads us to our most pessimistic conclusion: even with 14 years of detailed demographic data and sophisticated modeling techniques we failed to produce forecasts with any level of acceptable uncertainty.
 In our view, uncertainty of climate change related forecasts can be reduced by (1) longer time series and (2) climate covariates that match the scale of inference (e.g., plot rather than landscape level climate/weather metrics).
+Still, given the poor performance of the quad-based model, it seems there is no short cut to producing accurate and precise population forecasts.
 Do we need detailed demographic data to forecast the impacts of climate change on populations?
@@
 
-```{r figure_3, dependson="plot-options", fig.cap="Proportional change in species' mean cover caused by a 1% increase in observed precipitation (+ppt), temperature (+temp), or both (+ppt&temp) as predicted by the individual-based IPM and the aggregate-based QBM.
+```{r figure_3, dependson="plot-options", fig.cap="Proportional change in species' mean cover caused by a 1% increase in observed precipitation (+ppt), temperature (+temp), or both (+ppt&temp) as predicted by the individual-based IPM and the aggregate-based QBM using mean parameter values.
 Top panels show the mean predicted proportional change in cover.
 Lower panels show the sensitivity of equilibrium cover simulated from the IPM to each climate scenario applied to individual and combined vital rates.
-For example, the points associated with G show the median cover from IPM simulations where a climate perturbation is applied only to the growth regression climate covariates.", fig.width=8.5, fig.height=5, cache=FALSE, results='hide'}
+For example, the points associated with G show the median cover from IPM simulations where a climate perturbation is applied only to the growth regression climate covariates.
+These simulations also use mean parameter values for clarity.", fig.width=8.5, fig.height=5, cache=FALSE, results='hide'}
 setwd("~/Repos/MicroMesoForecast/analysis/")
@@
 
-```{r figure_4, dependson="plot-options", fig.cap="Equilibrium cover and 90% quantiles around the mean prediction when model and parameter uncertainty are propogated through the simulation phase.
+```{r figure_4, dependson="plot-options", fig.cap="Equilibrium cover and 90% quantiles around the mean prediction when model error and parameter uncertainty are propogated through the simulation phase.
 Climate simulations are as in Figure 3.", fig.width=4.5, fig.height=7, cache=FALSE, results='hide'}
@@
 
-```{r figure_5, dependson="plot-options", fig.cap="Correlations between QBM and IPM estimates of climate effects.
+```{r figure_5, dependson="plot-options", fig.cap="Effect of quadrat sample size on the precision (standard deviation) of main climate effect estimates in the QBM.
+Increasing the number of quadrats results in diminishing returns in terms of parameter certainty.
+Light dashed lines show individual climate effects at five quadrat sample sizes.
+Thick dark lines are inverse gaussian fits showing the mean effect of increasing quadrat sample size on parameter precision.", fig.width=3, fig.height=7.5, cache=FALSE, results='hide'}
+library(mgcv)
+
+setwd("../analysis/quadBM/vitalRateRegressions/exclude_quad_test")
+
+####
+####  Read in model fits and summarize results
+####
+source("popgrowth_read_data.R")
+spp <- unique(growD_all$Species)
+num_quads <- numeric(length(spp))
+for(i in 1:length(spp)){
+  tmp <- subset(growD_all, Species==spp[i])
+  num_quads[i] <- length(unique(tmp$quad))
+}
+num_quads <- as.data.frame(num_quads)
+num_quads$species <- spp
+
+all_files <- list.files("./results/")
+all_fits <- data.frame(Parameter=NA, stddev=NA, id=NA, 
+                       rmsquad=NA, rep=NA, species=NA)
+for(i in 1:length(all_files)){
+  tmp <- all_files[i]
+  spp <- substring(tmp, 1, 4)
+  qsrm1 <- unlist(strsplit(tmp, "[_]"))[2]
+  qsrm <- gsub("numout", "", qsrm1)
+  rep1 <- unlist(strsplit(tmp, "[_]"))[3]
+  rep2 <- gsub(".RDS","",rep1)
+  rep <- gsub("rep", "", rep2)
+
+  long <- readRDS(paste("./results/",tmp, sep=""))
+  short <- ddply(long, .(Parameter), summarise,
+                 stddev = sd(value))
+  climeff <- short[grep("b2", short$Parameter),]
+  climeff$id <- substr(climeff$Parameter, 4, length(climeff$Parameter))
+  climeff$id <- as.numeric(unlist(strsplit(climeff$id, split=']')))  
+  climeff <- climeff[with(climeff, order(id)),]
+  climeff$rmsquad <- qsrm
+  climeff$rep <- rep
+  climeff$species <- spp
+  all_fits <- rbind(all_fits, climeff)
+} # end file loop
+
+all_fits <- all_fits[2:nrow(all_fits),]
+
+####
+####  Read in full fits
+####
+dir <- "../truncNormModel/"
+files <- list.files(dir)
+files <- files[grep(".RDS", files)]
+all_alls <- data.frame(Parameter=NA, stddev=NA, id=NA, 
+                       rmsquad=NA, rep=NA, species=NA)
+for(i in 1:length(files)){
+  tmp <- files[i]
+  spp1 <- unlist(strsplit(tmp, "[_]"))[3]
+  spp <- gsub(".RDS","",spp1)
+
+  long <- readRDS(paste(dir,tmp, sep=""))
+  short <- ddply(long, .(Parameter), summarise,
+                 stddev = sd(value))
+  climeff <- short[grep("b2", short$Parameter),]
+  climeff$id <- substr(climeff$Parameter, 4, length(climeff$Parameter))
+  climeff$id <- as.numeric(unlist(strsplit(climeff$id, split=']')))  
+  climeff <- climeff[with(climeff, order(id)),]
+  climeff$rmsquad <- 0
+  climeff$rep <- 1
+  climeff$species <- spp
+  all_alls <- rbind(all_alls, climeff)
+}
+all_alls <- all_alls[2:nrow(all_alls),]
+
+####
+####  Aggregate results by species and rep
+####
+all_fits <- rbind(all_fits, all_alls)
+agg_fits <- ddply(all_fits, .(Parameter, rmsquad, species, id), summarise,
+                  avg_stddev = mean(stddev))
+agg_fits2 <- agg_fits[which(agg_fits$id %in% c(1:5)),]
+agg_fits2 <- merge(agg_fits2, num_quads, by="species")
+agg_fits2$quadsfit <- with(agg_fits2, num_quads-as.numeric(rmsquad))
+
+####
+####  Fit model (need to do by species)
+####
+modBOGR <- gam(avg_stddev~quadsfit, data=subset(agg_fits2, species=="BOGR"),
+               family=quasi(link="1/mu^2"))
+modHECO <- gam(avg_stddev~quadsfit, data=subset(agg_fits2, species=="HECO"),
+               family=quasi(link="1/mu^2"))
+modPASM <- gam(avg_stddev~quadsfit, data=subset(agg_fits2, species=="PASM"),
+               family=quasi(link="1/mu^2"))
+modPOSE <- gam(avg_stddev~quadsfit, data=subset(agg_fits2, species=="POSE"),
+               family=quasi(link="1/mu^2"))
+summary(modBOGR)
+summary(modHECO)
+summary(modPASM)
+summary(modPOSE)
+
+####
+#### Plot
+####
+ggplot(agg_fits2, aes(x=quadsfit, y=avg_stddev))+
+  geom_line(aes(group=Parameter), alpha=0.3, linetype=2)+
+  geom_point(aes(group=Parameter), alpha=0.3, size=3)+
+#   stat_smooth(method="loess", size=2, color="black", se=FALSE)+
+  geom_smooth(method="glm", formula=y~x, family=quasi(link="1/mu^2"),
+              size=2, color="black", se=FALSE)+
+  facet_wrap("species", scales = "free", ncol=1)+
+  theme_bw()+
+  xlab("Number of quadrats fit")+
+  ylab("Standard deviation of coefficient")
+
+```
+
+```{r figure_6, dependson="plot-options", fig.cap="Correlations between QBM and IPM estimates of climate effects.
 We ignore sizeXclimate interactions since these are not directly comparable across model types.
@@
 
-```{r figure_6, dependson="plot-options", fig.cap="Effect of quadrat sample size on the precision (standard deviation) of main climate effect estimates in the QBM.
-Increasing the number of quadrats results in diminishing returns in terms of parameter certainty.
-Light dashed lines show individual climate effects at five quadrat sample sizes.
-Thick dark lines are inverse gaussian fits showing the mean effect of increasing quadrat sample size on parameter precision.", fig.width=3, fig.height=7.5, cache=FALSE, results='hide'}
-library(mgcv)
-
-setwd("../analysis/quadBM/vitalRateRegressions/exclude_quad_test")
-
-####
-####  Read in model fits and summarize results
-####
-source("popgrowth_read_data.R")
-spp <- unique(growD_all$Species)
-num_quads <- numeric(length(spp))
-for(i in 1:length(spp)){
-  tmp <- subset(growD_all, Species==spp[i])
-  num_quads[i] <- length(unique(tmp$quad))
-}
-num_quads <- as.data.frame(num_quads)
-num_quads$species <- spp
-
-all_files <- list.files("./results/")
-all_fits <- data.frame(Parameter=NA, stddev=NA, id=NA, 
-                       rmsquad=NA, rep=NA, species=NA)
-for(i in 1:length(all_files)){
-  tmp <- all_files[i]
-  spp <- substring(tmp, 1, 4)
-  qsrm1 <- unlist(strsplit(tmp, "[_]"))[2]
-  qsrm <- gsub("numout", "", qsrm1)
-  rep1 <- unlist(strsplit(tmp, "[_]"))[3]
-  rep2 <- gsub(".RDS","",rep1)
-  rep <- gsub("rep", "", rep2)
-
-  long <- readRDS(paste("./results/",tmp, sep=""))
-  short <- ddply(long, .(Parameter), summarise,
-                 stddev = sd(value))
-  climeff <- short[grep("b2", short$Parameter),]
-  climeff$id <- substr(climeff$Parameter, 4, length(climeff$Parameter))
-  climeff$id <- as.numeric(unlist(strsplit(climeff$id, split=']')))  
-  climeff <- climeff[with(climeff, order(id)),]
-  climeff$rmsquad <- qsrm
-  climeff$rep <- rep
-  climeff$species <- spp
-  all_fits <- rbind(all_fits, climeff)
-} # end file loop
-
-all_fits <- all_fits[2:nrow(all_fits),]
-
-####
-####  Read in full fits
-####
-dir <- "../truncNormModel/"
-files <- list.files(dir)
-files <- files[grep(".RDS", files)]
-all_alls <- data.frame(Parameter=NA, stddev=NA, id=NA, 
-                       rmsquad=NA, rep=NA, species=NA)
-for(i in 1:length(files)){
-  tmp <- files[i]
-  spp1 <- unlist(strsplit(tmp, "[_]"))[3]
-  spp <- gsub(".RDS","",spp1)
-
-  long <- readRDS(paste(dir,tmp, sep=""))
-  short <- ddply(long, .(Parameter), summarise,
-                 stddev = sd(value))
-  climeff <- short[grep("b2", short$Parameter),]
-  climeff$id <- substr(climeff$Parameter, 4, length(climeff$Parameter))
-  climeff$id <- as.numeric(unlist(strsplit(climeff$id, split=']')))  
-  climeff <- climeff[with(climeff, order(id)),]
-  climeff$rmsquad <- 0
-  climeff$rep <- 1
-  climeff$species <- spp
-  all_alls <- rbind(all_alls, climeff)
-}
-all_alls <- all_alls[2:nrow(all_alls),]
-
-####
-####  Aggregate results by species and rep
-####
-all_fits <- rbind(all_fits, all_alls)
-agg_fits <- ddply(all_fits, .(Parameter, rmsquad, species, id), summarise,
-                  avg_stddev = mean(stddev))
-agg_fits2 <- agg_fits[which(agg_fits$id %in% c(1:5)),]
-agg_fits2 <- merge(agg_fits2, num_quads, by="species")
-agg_fits2$quadsfit <- with(agg_fits2, num_quads-as.numeric(rmsquad))
-
-####
-####  Fit model (need to do by species)
-####
-modBOGR <- gam(avg_stddev~quadsfit, data=subset(agg_fits2, species=="BOGR"),
-               family=quasi(link="1/mu^2"))
-modHECO <- gam(avg_stddev~quadsfit, data=subset(agg_fits2, species=="HECO"),
-               family=quasi(link="1/mu^2"))
-modPASM <- gam(avg_stddev~quadsfit, data=subset(agg_fits2, species=="PASM"),
-               family=quasi(link="1/mu^2"))
-modPOSE <- gam(avg_stddev~quadsfit, data=subset(agg_fits2, species=="POSE"),
-               family=quasi(link="1/mu^2"))
-summary(modBOGR)
-summary(modHECO)
-summary(modPASM)
-summary(modPOSE)
-
-####
-#### Plot
-####
-ggplot(agg_fits2, aes(x=quadsfit, y=avg_stddev))+
-  geom_line(aes(group=Parameter), alpha=0.3, linetype=2)+
-  geom_point(aes(group=Parameter), alpha=0.3, size=3)+
-#   stat_smooth(method="loess", size=2, color="black", se=FALSE)+
-  geom_smooth(method="glm", formula=y~x, family=quasi(link="1/mu^2"),
-              size=2, color="black", se=FALSE)+
-  facet_wrap("species", scales = "free", ncol=1)+
-  theme_bw()+
-  xlab("Number of quadrats fit")+
-  ylab("Standard deviation of coefficient")
-
-```
-
 \pagebreak{}

### 2015-07-06 manuscript/manuscript.tex
draft 2 of ms

@@
 quadrat $q$ at time $t$ and all other parameters are as in the
-individual-level growth model (Eq. \#). The log normal likelihood
+individual-level growth model (Eq. 3). The log normal likelihood
 includes a truncation (T{[}0,1{]}) to ensure that predicted values do
@@
 Our quad-based model (QBM) perfectly mirrors its statistical description
-(Eq. \#). We use the same approach for drawing parameter values as
+(Eq. 8-9). We use the same approach for drawing parameter values as
 described for the IPM.
@@
 projections of the same sign and somewhat similar magnitude (Fig. 3d).
-As expected based on model validation (Table 1), QBM projections were
-more uncertain than IPM projections for all species except \emph{P.
-smithiii} (Fig. 3e-h).
 
@@
 sensitivity of each vital rate in isolation and in pairwise
-combinations. Species showed similar trends (Fig. 4). Growth was the
+combinations. Species showed similar trends (Fig. 3e-h). Growth was the
 most sensitive vital rate for all species, showing a negative response
@@
 temperature, and a mostly positive response when both climate factors
-are increased (Fig. 4). \emph{B. gracilis} survival rates were sensitive
-to temperature, resulting in an increase in plant cover under increased
-temperature (Fig. 4a). In isolation, recruitment and survival were
-insensitive to climate factors for \emph{H. comata} (Fig. 4b). Survival
-and recruitment of \emph{P. smithii} were both sensitive, negatively, to
-temperature and precipitation (Fig. 4c). \emph{P. secunda} equilibrium
-cover was sensitive to the climate effects on survival and recruitment,
-showing a negative effect on both vital rates for increased precipition,
-but a strong positive effect on survival with increased temperature
-(Fig. 4d). The climate impact of recruitment on equilibrium cover was
-negative for precipitation and temperature increases (Fig. 4d).
+are increased (Fig. 3e-h). \emph{B. gracilis} survival rates were
+sensitive to temperature, resulting in an increase in plant cover under
+increased temperature (Fig. 3e). In isolation, recruitment and survival
+were insensitive to climate factors for \emph{H. comata} (Fig. 3f).
+Survival and recruitment of \emph{P. smithii} were both sensitive,
+negatively, to temperature and precipitation (Fig. 3g). \emph{P.
+secunda} equilibrium cover was sensitive to the climate effects on
+survival and recruitment, showing a negative effect on both vital rates
+for increased precipition, but a strong positive effect on survival with
+increased temperature (Fig. 3h). The climate impact of recruitment on
+equilibrium cover was negative for precipitation and temperature
+increases (Fig. 3h).
+
+Forecasts based on 1\% climate changes were extremely uncertain when we
+considered model error and parameter uncertainty (Fig. 4). As expected
+based on model validation (Table 1), QBM projections were more uncertain
+than IPM projections for all species except \emph{P. smithiii} (Fig. 4).
 
@@
 Perhaps the greatest challenge for ecology in the 21st century is to
-forecast the impacts of environmental change (Clark et al. 2001). To do
-so requires sophisticated modeling approaches that fully account for
-uncertainty and variability in the ecological process and associated
-parameters. This requires large amounts of data collected over large
-spatio-temporal extents. State-of-the-science modeling techniques cannot
-overcome data limitations. Such is the case for many population models.
+forecast the impacts of environmental change (Clark et al. 2001, Petchey
+et al. 2015). To do so requires sophisticated modeling approaches that
+fully account for uncertainty and variability in the ecological process
+and associated parameters (Luo et al. 2011). This requires large amounts
+of data collected over large spatio-temporal extents.
+State-of-the-science modeling techniques cannot overcome data
+limitations. Such is the case for many population models.
 
@@
 expected to lose some precision at the aggregated level. However, the
-underwhelming performance of the QBM could call in the question
+underwhelming performance of the QBM could call into the question
 forecasts that differ from the IPM.
@@
 importance of individual variability on population level responses to
-exogenous drivers (CITATIONS).
-
-An important caveat to the above conclusion is that our comparison of
-the IPM and the QBM
-
-This result is not as dire as it seems when we consider model
-uncerainty. Yes, the QBM failed to match IPM forecasts under perturbed
-climate, but the uncertainties around those forecasts were so great
-(Fig. 3e-h) that we cannot say that the two model's forecasts were
-different with any level of confidence. This offers hope that with
-``enough'' population level data one will be able to reduce parameter
-uncertainty and thus model forecasts. This remains an open question.
+exogenous drivers (Clark et al. 2011a, 2011b, 2012, Galván et al. 2014).
 
@@
 
-Our intuition was that species with strong climate-dependence on vital
-rates not well resolved at the aggregate level would result in different
+We can think of two reasons why the IPM outperformed the QBM. First, the
+quadrat level data has a much reduced sample size compared to the
+individual level data. In an ideal world we would have compared the IPM
+and QBM using data collected over the same amount of person hours, not
+just the same number of quadrats. Then the sample size of the quadrat
+level data would be much greater and carry more statistical power. To
+address this limitation in our work we fit the QBM statistical model
+(Eq. 8-9) with different numbers of quadrats to see the effect of sample
+size on the precision of climate effect estimates. It appears that
+including additional quadrats leads to rapidly diminishing returns in
+terms of parameter precision (Fig. 5). Thus, while sample size surely
+plays some role, we do not think it is the main driver of the difference
+between the IPM and QBM.
+
+The second reason the IPM could have outperformed the QBM is that the
+population level model is in fact missing important climate effects that
+act on individual vital rates, rather than population growth. Our
+intuition was that species with strong climate-dependence on vital rates
+not well resolved at the aggregate level would result in different
 forecasts from the two models. For example, survival is very size
@@
 increases, for example, we would expect to detect this signal in the
-individual level data but not in the aggregate level data.
-\textbf{PETER: do you think this is the right direction to go with
-this?}
+individual level data but not in the aggregate level data. To see if
+this is the case we can regress climate effects from each vital rate
+statistical model at the individual level against the same climate
+effects from the QBM statistical model (Fig. 6).
+
+In general, the QBM climate effects are most correlated with climate
+effects from the growth model at the individual level (Fig. 6). In no
+case does the QBM statistical model have strong correlations across all
+three vital rates (Fig. 6). Thus, for each species the QBM is
+``missing'' climate signals associated with at least one vital rate.
+This has large impacts on predictions of long term population dynamics,
+as seen in our equilibrium simulations (Fig. 3a-d).
 
@@
 predictions proved so uncertain that any forecast, when bounded with
-uncertainty, would be at best not useful and at worst meaningless. How
-might we improve on this state of affairs?
+model and parameter uncertainty, would be at best not useful and at
+worst meaningless. How might we improve on this state of affairs?
 
@@
 individual level data, we are missing the key interaction point between
-weather and individual plants. \textbf{PETER: any good citations here
-for decent population models fit with fine-scale climate/environmental
-data??}
+weather and individual plants (Clark et al. 2011b).
 
@@
 data suggests that even advanced approaches to parameter fitting like
-LASSO and random forest require 20-25 year time series (Teller et al.,
-in review).
+LASSO, functional linear models (splines), and random forest require
+20-25 year time series (Teller et al., in review). Alternatively, as we
+suggest above, Teller et al. (in review) also find that matching the
+scale of the response and predictors improves estimate precision.
+
+Third, ecologists as a community need to get serious about reporting
+uncertainty. There is a strong culture around explicitly considering
+model uncertainty, but parameter uncertainty is often ignored. In some
+cases this is because the easiest statistical methods do no make
+propogating parameter uncertainty easy. But even using Bayesian
+approaches that allow straightforward integration of model fitting and
+forecasting (Hooten and Hobbs 2014, Hobbs and Hooten 2015) is not simple
+when using modeling approaches like integral projection models that
+separate the model fitting and simulation stages (Rees and Ellner 2009).
+However, as we have done here, it is still possible to include parameter
+uncertainty by drawing parameter values from MCMC iterations, taking
+care to draw all parameters from the same chain and iteration to account
+for their correlations. Only by being honest about our forecasts cen we
+begin to produce better ones.
 
@@
 
-This work is not a critique of density-structured population models.
-Instead, we consider our results a cautionary tale. In some cases and
-for certain species, population models based on aggregated data may
-prove useful and unbiased. But, as our work shows, that is not a
-foregone conclusion when considering the impact of exogenous drivers
-like climate on population states. Thus, forecasts from aggregate level
-models should be viewed with caution and should never be unaccompianed
-by uncertainty. As we have shown, uncertainty around forecasts was large
-for both model types. Which leads us to our most pessimistic conclusion:
-even with 14 years of detailed demographic data and sophisticated
-modeling techniques we failed to produce forecasts with any level of
-acceptable uncertainty. In our view, uncertainty of climate change
-related forecasts can be reduced by (1) longer time series and (2)
-climate covariates that match the scale of inference (e.g., plot rather
-than landscape level climate/weather metrics). Do we need detailed
-demographic data to forecast the impacts of climate change on
-populations? Probably.
+This work is not a critique of density-structured population models. In
+some cases and for certain species, population models based on
+aggregated data may prove useful and unbiased. However, our work here is
+the first comparison of population models based on individual and
+aggregated data in a climate change context. Our results confirm
+theoretical arguments (Clark et al. 2011b) and empirical evidence (Clark
+et al. 2011a, 2012) that individual responses are critical to predicting
+species' responses to climate change. Thus, forecasts from aggregate
+level models should be viewed with caution and should never be
+unaccompianed by uncertainty.
+
+Our results also offer a cautionary tale because uncertainty around
+forecasts was large for both model types. Which leads us to our most
+pessimistic conclusion: even with 14 years of detailed demographic data
+and sophisticated modeling techniques we failed to produce forecasts
+with any level of acceptable uncertainty. In our view, uncertainty of
+climate change related forecasts can be reduced by (1) longer time
+series and (2) climate covariates that match the scale of inference
+(e.g., plot rather than landscape level climate/weather metrics). Still,
+given the poor performance of the quad-based model, it seems there is no
+short cut to producing accurate and precise population forecasts. Do we
+need detailed demographic data to forecast the impacts of climate change
+on populations? Probably.
 
@@
 (+ppt\&temp) as predicted by the individual-based IPM and the
-aggregate-based QBM. Top panels show the mean predicted proportional
-change in cover. Lower panels show the sensitivity of equilibrium cover
-simulated from the IPM to each climate scenario applied to individual
-and combined vital rates. For example, the points associated with G show
-the median cover from IPM simulations where a climate perturbation is
-applied only to the growth regression climate covariates.}
+aggregate-based QBM using mean parameter values. Top panels show the
+mean predicted proportional change in cover. Lower panels show the
+sensitivity of equilibrium cover simulated from the IPM to each climate
+scenario applied to individual and combined vital rates. For example,
+the points associated with G show the median cover from IPM simulations
+where a climate perturbation is applied only to the growth regression
+climate covariates. These simulations also use mean parameter values for
+clarity.}
 \end{figure}
@@
 \caption{Equilibrium cover and 90\% quantiles around the mean prediction
-when model and parameter uncertainty are propogated through the
+when model error and parameter uncertainty are propogated through the
 simulation phase. Climate simulations are as in Figure 3.}
@@
 \includegraphics{components/figure/manuscript-figure_5.pdf}
-\caption{Correlations between QBM and IPM estimates of climate effects.
-We ignore sizeXclimate interactions since these are not directly
-comparable across model types. The QBM does not have multiple vital
-rates, so its values are repeated across panels within each species.}
-\end{figure}
-
-\begin{figure}[htbp]
-\centering
-\includegraphics{components/figure/manuscript-figure_6.pdf}
 \caption{Effect of quadrat sample size on the precision (standard
@@
 precision.}
+\end{figure}
+
+\begin{figure}[htbp]
+\centering
+\includegraphics{components/figure/manuscript-figure_6.pdf}
+\caption{Correlations between QBM and IPM estimates of climate effects.
+We ignore sizeXclimate interactions since these are not directly
+comparable across model types. The QBM does not have multiple vital
+rates, so its values are repeated across panels within each species.}
 \end{figure}
@@
 
+Clark, J. S., D. M. Bell, M. H. Hersh, and L. Nichols. 2011a. Climate
+change vulnerability of forest biodiversity: Climate and competition
+tracking of demographic rates. Global Change Biology 17:1834--1849.
+
+Clark, J. S., D. M. Bell, M. H. Hersh, M. C. Kwit, E. Moran, C. Salk, A.
+Stine, D. Valle, and K. Zhu. 2011b. Individual-scale variation,
+species-scale differences: Inference needed to understand diversity.
+
 Clark, J. S., D. M. Bell, M. Kwit, A. Stine, B. Vierra, and K. Zhu.
@@
 dynamics. American Naturalist 177:1--17.
+
+Galván, J. D., J. J. Camarero, and E. Gutiérrez. 2014. Seeing the trees
+for the forest: Drivers of individual growth responses to climate in
+Pinus uncinata mountain forests. Journal of Ecology 102:1244--1257.
+
+Hobbs, N. T., and M. B. Hooten. 2015. Bayesian Models: A Statistical
+Primer for Ecologists. Princeton University PressPrinceton.
+
+Hooten, M. B., and N. T. Hobbs. 2014. A guide to Bayesian model
+selection for ecologists. Ecological Monographs 85:3--28.
+
+Luo, Y., K. Ogle, C. Tucker, S. Fei, C. Gao, S. LaDeau, J. S. Clark, and
+D. S. Schimel. 2011. Ecological forecasting and data assimilation in a
+data-rich era. Ecological Applications 21:1429--1442.
+
+Petchey, O. L., M. Pontarp, T. M. Massie, S. Kéfi, A. Ozgul, M.
+Weilenmann, G. M. Palamara, F. Altermatt, B. Matthews, J. M. Levine, D.
+Z. Childs, B. J. McGill, M. E. Schaepman, B. Schmid, P. Spaak, A. P.
+Beckerman, F. Pennekamp, and I. S. Pearse. 2015. The ecological forecast
+horizon, and examples of its uses and determinants. Ecology Letters
+18:597--611.
 

### 2015-07-07 manuscript/appA.Rmd
adds appendix A draft

---
layout: 12pt
header-includes:
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{setspace}
   - \doublespacing
bibliography: ~/Dropbox/Bibliography/MicroMeso.bib
csl: components/ecology.csl

## rmarkdown render options
output:
  pdf_document:
    fig_caption: true
    keep_tex: true
fontsize: 12pt
geometry: margin=1in

---

Appendix A: Detailed model descriptions
=======================================
### Tredennick and Adler

Here we describe the statistical models in complete detail, including full description of the Bayesian models and Stan code for each model. We write each model as three parts: (1) a data model (likelihood), (2) a process model, and (3) a parameter model. We then combine these three models to write the posterior and joint distributions as a single model statement. Following the full model expression we provide the Stan code that directly corresponds the statement of posterior and joint distributions. In our expressions of the posterior and joint distributions we use general probability notation where [.] represents some unspecified distribution. We do this for clarity of model presentation. For each model we include a table of prior distributions for necessary parameters. Also for clarity, we avoid subscripting species (the *j* subscripts in the main text).

\newpage{}

Individual level models
-----------------------
### Survival
Data, process, and parameter models.

\begin{align}
&\mathtt{data} \qquad &\textbf{y} \sim \text{Bernoulli}(\boldsymbol{\mu}) \\
&\mathtt{process} \qquad &\text{logit}(\mu_{iQ,t}) = \gamma_{t} + \phi_{Q} + \beta_{t}x_{ij,t} + \omega w_{i,t} + \theta_{k}C_{k,t} \\
&\mathtt{parameters} \qquad &\gamma_{t} \sim \text{Normal}(\tilde{\gamma}, \sigma_\gamma^2)\\
& \qquad & \phi_{Q} \sim \text{Normal}(0,\sigma_Q^2) \\
& \qquad & \beta_{t} \sim \text{Normal}(\tilde{\beta}, \sigma_\beta^2) \\
& \qquad & \omega \sim \text{Normal}(0, 0.0001) \\
& \qquad & \boldsymbol{\theta} \sim \text{Normal}(0, 0.0001)
\end{align}

Full expression of posterior and joint distributions.
\begin{align}
[\boldsymbol{\gamma}, \tilde{\gamma}, \boldsymbol{\phi}, \boldsymbol{\beta}, \tilde{\beta}, \omega, \boldsymbol{\theta}, \sigma_Q^2, \sigma_\beta^2] &\propto \\
&\prod_{t=1}^T \prod_{i=1}^n [y_{iQ,t} | \gamma_{t}, \phi_{Q}, \beta_{t}, \omega, \boldsymbol{\theta}] [\gamma_{t} | \tilde{\gamma}] [\beta_{t} | \tilde{\beta}, \sigma_\beta^2] \times \\
&\prod_{Q=1}^{Q_{tot}} [\phi_{Q} | \sigma_Q^2] \times \\
&[\boldsymbol{\theta}] [\omega] [\tilde{\gamma}] [\tilde{\beta}] [\sigma_Q^2] [\sigma_\beta^2]
\end{align}

Stan code for model.
```{r survival_stan, echo=TRUE, eval=FALSE}
data{
  int<lower=0> N; // observations
  int<lower=0> Yrs; // years
  int<lower=0> yid[N]; // year id
  int<lower=0> Covs; // climate covariates
  int<lower=0> G; // groups
  int<lower=0> gid[N]; // group id
  int<lower=0,upper=1> Y[N]; // observation vector
  matrix[N,Covs] C; // climate matrix
  vector[N] X; // size vector
  matrix[N,2] W; // crowding matrix
}
parameters{
  real a_mu;
  vector[Yrs] a;
  real b1_mu;
  vector[Yrs] b1;
  vector[Covs] b2;
  vector[2] w;
  vector[G] gint;
  real<lower=0> sig_a;
  real<lower=0> sig_b1;
  real<lower=0> sig_G;
}
transformed parameters{
  real mu[N];
  vector[N] climEff;
  vector[N] crowdEff;
  climEff <- C*b2;
  crowdEff <- W*w;
  for(n in 1:N){
    mu[n] <- inv_logit(a[yid[n]] + gint[gid[n]] + b1[yid[n]]*X[n] + 
                         crowdEff[n] + climEff[n]);
  }
}
model{
  // Priors
  a_mu ~ uniform(-300,300);
  w ~ uniform(-100,100);
  b1_mu ~ uniform(-100,100);
  sig_a ~ cauchy(0,5);
  sig_b1 ~ cauchy(0,5);
  sig_G ~ cauchy(0,5);
  gint ~ normal(0, sig_G);
  b2 ~ uniform(-10,10);
  a ~ normal(a_mu, sig_a);
  b1 ~ normal(b1_mu, sig_b1);

  // Likelihood
  Y ~ binomial(1,mu);
}
```

### 2015-07-08 manuscript/manuscript.Rmd
updates ms with new abstract and adds crowdXsize interactions in equations

@@
 --------
-The ability of population models to skillfully forecast future states under climate change is constrained by the limited spatial and temporal extent of demographic data.
-This data is often limited because it is difficult and costly to collect.
-An alternative is to rely on aggregate, population-level data that is easier and less costly to collect.
-Doing so requires assuming that population-level data accurately represents the aggregate response of the individuals that actually respond to weather.
-We tested this assumption using population models of four Montana grassland species fit using individual and aggregated forms of the same data.
-We fit population models with interannual variation in vital rates explained, in part, by climate covariates and then perturbed observed climate to compare model forecasts.
-Population models based on individual-level demographic data outperformed the models based on aggregate-level data in terms of accuracy and precision.
-The two model types produced inconsistent forecasts when we perturbed climate.
-Thus, it seems that, at least for the species in this location, demographic data is necessary to pick up key climate drivers of vital rates like survival that are not well resolved in aggregated data.
-On a pessimistic note, our work shows that even with detailed demographic data, model forecasts are extremely uncertain.
-It is becoming clear the long time series (over two decades) are necessary to produce meaningful forecasts.
+Forecasting future states of populations has taken on new urgency as the rate of climate change increases.
+Traditional plant population models have limited utility in this regard because they are based on detailed demographic data from small, localized plots.
+These models are difficult to scale up to spatial scales relevant to land managers that require such forecasts to make decisions.
+To overcome the data limitations of traditional population models, some have proposed population models based on population level, rather than individual level, data that is much easier to collect over broad spatial scales.
+Using such models violates a central assumption of ecology: individuals respond to weather, not populations to climate.
+Here, we test whether this assumption is important when forecasting climate change impacts on four perennial grass species in a semi-arid Montana grassland.
+We parameterized two population models, one based on inidividual level data with three vital rates and one on an aggregated version of the same data (percent cover), with climate covariates and compared their accuracy, precision, and sensitivity to climate.
+The individual level model was more accurate and precise than the aggregate level model when predicting out of sample observations.
+The aggregate level model made countervailing forecasts to 1% climate changes when compared to the individual level model, indicating the aggregate level model produces incorrect forecasts.
+When comparing climate effects from both models, the aggregate level model tends to "miss" important climate effects from at least one vital rate for each species.
+It appears there is no short cut to forecasting climate change impacts on plant populations — detailed demographic data is essential.
+But, forecasts were very uncertain, so we advocate for a focus on new methods to collect demographic data more efficiently across environmental gradients in space and time.
 
@@
 We also include interactions among same-year climate covariates (e.g., ppt1 $\times$ TmeansSpr1) and climate $\times$ size interactions.
+Climate $\times$ size interactions are for climate main effects only, that is we do not include interactions among size and interacting climate effects.
 
@@
 \begin{align}
-\text{logit}(S_{ijQ,t}) &= \gamma^{S}_{j,t} + \phi^{S}_{jQ} + \beta^{S}_{j,t}x_{ij,t} + \omega^{S}_{j}w_{ij,t} + \theta^{S}_{jk}C_{k,t} \\
+\text{logit}(S_{ijQ,t}) &= \gamma^{S}_{j,t} + \phi^{S}_{jQ} + \beta^{S}_{j,t}x_{ij,t} + \omega^{S}_{j}w_{ij,t} + \nu^{S}_{j}w_{ij,t}x_{ij,t} + \theta^{S}_{jk}C_{k,t} \\
 Y^{S}_{ijQ,t} &\sim \text{Bernoulli}(S_{ijQ,t})
@@
 where $x_{ij,t}$ is the log of genet size, $\gamma^{S}_{j,t}$ is a year-specific intercept, $\beta^{S}_{j,t}$ is the year-specific slope parameter for size, $\phi^{S}_{jQ}$ is the random effect of quadrat group location, and $\theta^{S}_{k}$ is the fixed parameter for the effect of the $k$th climate covariate at time $t$ ($C_{k,t}$).
+Note that the vector of climate covariates (**C**) includes climate variable interactions and climate$\times$size interactions.
 We include density-dependence by estimating the effect of crowding on the focal individual by other individuals of the same species.
 $\omega$ is the effect of crowding and $w_{t,Q}$ is the crowding experienced by the focal individual at time $t$ in quadrat group $Q$.
+We include a size$\times$crowding interaction effect ($\nu^{S}$).
 
@@
 \begin{align}
-x_{ijQ,t+1} &= \gamma^{G}_{j,t} + \phi^{G}_{jQ} + \beta^{G}_{j,t}x_{ij,t} + \omega^{G}_{j}w_{ij,t} + \theta^{G}_{jk}C_{k,t} \\
+x_{ijQ,t+1} &= \gamma^{G}_{j,t} + \phi^{G}_{jQ} + \beta^{G}_{j,t}x_{ij,t} + \omega^{G}_{j}w_{ij,t} + \nu^{S}_{j}w_{ij,t}x_{ij,t} + \theta^{G}_{jk}C_{k,t} \\
 Y^{G}_{ijQ,t} &\sim \text{Normal}(x_{ijQ,t+1}, \sigma_{j})
@@
 3).
+Again, note that the climate covariate vector (**C**) includes the climate$\times$cover interaction.
 The log normal likelihood includes a truncation (T[0,1]) to ensure that predicted values do not exceed 100% cover.
@@
 In some cases and for certain species, population models based on aggregated data may prove useful and unbiased.
-However, our work here is the first comparison of population models based on individual and aggregated data in a climate change context.
+However, our work here is the first comparison, to our knowledge, of population models based on individual and aggregated data in a climate change context.
 Our results confirm theoretical arguments [@Clark2011b] and empirical evidence [@Clark2011; @Clark2012] that individual responses are critical to predicting species' responses to climate change.
 Thus, forecasts from aggregate level models should be viewed with caution and should never be unaccompianed by uncertainty.
+Given the importance of demographic data and its current difficulty to collect, we advocate for research on modern methods to collect demographic data more efficiently across environmental gradients in space and time.
+For example, recent advances in object recognition algorithms [@Liu2014] provide new opportunities to speed up data collection and digital processing.
 

### 2015-07-08 manuscript/manuscript.tex
updates ms with new abstract and adds crowdXsize interactions in equations

@@
 
+\newpage{}
+
 \subsection{Abstract}\label{abstract}
 
-The ability of population models to skillfully forecast future states
-under climate change is constrained by the limited spatial and temporal
-extent of demographic data. This data is often limited because it is
-difficult and costly to collect. An alternative is to rely on aggregate,
-population-level data that is easier and less costly to collect. Doing
-so requires assuming that population-level data accurately represents
-the aggregate response of the individuals that actually respond to
-weather. We tested this assumption using population models of four
-Montana grassland species fit using individual and aggregated forms of
-the same data. We fit population models with interannual variation in
-vital rates explained, in part, by climate covariates and then perturbed
-observed climate to compare model forecasts. Population models based on
-individual-level demographic data outperformed the models based on
-aggregate-level data in terms of accuracy and precision. The two model
-types produced inconsistent forecasts when we perturbed climate. Thus,
-it seems that, at least for the species in this location, demographic
-data is necessary to pick up key climate drivers of vital rates like
-survival that are not well resolved in aggregated data. On a pessimistic
-note, our work shows that even with detailed demographic data, model
-forecasts are extremely uncertain. It is becoming clear the long time
-series (over two decades) are necessary to produce meaningful forecasts.
+Forecasting future states of populations has taken on new urgency as the
+rate of climate change increases. Traditional plant population models
+have limited utility in this regard because they are based on detailed
+demographic data from small, localized plots. These models are difficult
+to scale up to spatial scales relevant to land managers that require
+such forecasts to make decisions. To overcome the data limitations of
+traditional population models, some have proposed population models
+based on population level, rather than individual level, data that is
+much easier to collect over broad spatial scales. Using such models
+violates a central assumption of ecology: individuals respond to
+weather, not populations to climate. Here, we test whether this
+assumption is important when forecasting climate change impacts on four
+perennial grass species in a semi-arid Montana grassland. We
+parameterized two population models, one based on inidividual level data
+with three vital rates and one on an aggregated version of the same data
+(percent cover), with climate covariates and compared their accuracy,
+precision, and sensitivity to climate. The individual level model was
+more accurate and precise than the aggregate level model when predicting
+out of sample observations. The aggregate level model made
+countervailing forecasts to 1\% climate changes when compared to the
+individual level model, indicating the aggregate level model produces
+incorrect forecasts. When comparing climate effects from both models,
+the aggregate level model tends to ``miss'' important climate effects
+from at least one vital rate for each species. It appears there is no
+short cut to forecasting climate change impacts on plant populations ---
+detailed demographic data is essential. But, forecasts were very
+uncertain, so we advocate for a focus on new methods to collect
+demographic data more efficiently across environmental gradients in
+space and time.
 

### 2015-07-08 manuscript/appA.Rmd
updates appendix with IPM model dscriptions

@@
    - \usepackage{setspace}
-   - \doublespacing
 bibliography: ~/Dropbox/Bibliography/MicroMeso.bib
@@
 =======================================
-### Tredennick and Adler
-
 Here we describe the statistical models in complete detail, including full description of the Bayesian models and Stan code for each model.
@@
 For each model we include a table of prior distributions for necessary parameters.
-Also for clarity, we avoid subscripting species (the *j* subscripts in the main text).
+Also for clarity, we avoid subscripting species (the *j* subscripts) after initial description of each model.
 
-\newpage{}
+Survival (IPM)
+--------------
+We used logistic regression to model survival probability ($S$) of genet $i$ from species $j$ in quadrat group $Q$ from time $t$ to $t+1$:
 
-Individual level models
------------------------
-### Survival
-Data, process, and parameter models.
+\begin{align}
+\text{logit}(S_{ijQ,t}) &= \gamma^{S}_{j,t} + \phi^{S}_{jQ} + \beta^{S}_{j,t}x_{ij,t} + \omega^{S}_{j}w_{ij,t} + \nu^{S}_{j}w_{ij,t}x_{ij,t} + \theta^{S}_{jk}C_{k,t} \\
+Y^{S}_{ijQ,t} &\sim \text{Bernoulli}(S_{ijQ,t})
+\end{align}
+
+where $x_{ij,t}$ is the log of genet size, $\gamma^{S}_{j,t}$ is a year-specific intercept, $\beta^{S}_{j,t}$ is the year-specific slope parameter for size, $\phi^{S}_{jQ}$ is the random effect of quadrat group location, and $\theta^{S}_{k}$ is the fixed parameter for the effect of the $k$th climate covariate at time $t$ ($C_{k,t}$).
+Note that the vector of climate covariates (**C**) includes climate variable interactions and climate$\times$size interactions.
+We include density-dependence by estimating the effect of crowding on the focal individual by other individuals of the same species.
+$\omega$ is the effect of crowding and $w_{t,Q}$ is the crowding experienced by the focal individual at time $t$ in quadrat group $Q$.
+We include a size$\times$crowding interaction effect ($\nu^{S}$).
+
+**Data, process, and parameter models**
 
@@
 &\mathtt{data} \qquad &\textbf{y} \sim \text{Bernoulli}(\boldsymbol{\mu}) \\
-&\mathtt{process} \qquad &\text{logit}(\mu_{iQ,t}) = \gamma_{t} + \phi_{Q} + \beta_{t}x_{ij,t} + \omega w_{i,t} + \theta_{k}C_{k,t} \\
+&\mathtt{process} \qquad &\text{logit}(\mu_{iQ,t}) = \gamma_{t} + \phi_{Q} + \beta_{t}x_{i,t} + \omega w_{i,t} + \nu w_{i,t}x_{i,t} + \theta_{k}C_{k,t} \\
 &\mathtt{parameters} \qquad &\gamma_{t} \sim \text{Normal}(\tilde{\gamma}, \sigma_\gamma^2)\\
@@
 & \qquad & \beta_{t} \sim \text{Normal}(\tilde{\beta}, \sigma_\beta^2) \\
-& \qquad & \omega \sim \text{Normal}(0, 0.0001) \\
-& \qquad & \boldsymbol{\theta} \sim \text{Normal}(0, 0.0001)
+& \qquad & \omega \sim \text{Uniform}(-100, 100) \\
+& \qquad & \nu \sim \text{Uniform}(-100, 100) \\
+& \qquad & \boldsymbol{\theta} \sim \text{Uniform}(-10, 10) \\
+& \qquad & \tilde{\gamma} \sim \text{Uniform}(-300, 300) \\
+& \qquad & \tilde{\beta} \sim \text{Uniform}(-100, 100) \\
+& \qquad & \sigma_Q^2 \sim \text{Cauchy}(0, 5) \\
+& \qquad & \sigma_\gamma^2 \sim \text{Cauchy}(0, 5) \\
+& \qquad & \sigma_\beta^2 \sim \text{Cauchy}(0, 5)
 \end{align}
 
-Full expression of posterior and joint distributions.
+**Full expression of posterior and joint distributions**
 \begin{align}
@@
 
-Stan code for model.
+**Stan code for model**
 ```{r survival_stan, echo=TRUE, eval=FALSE}
@@
 ```
+
+Growth (IPM)
+------------
+We modeled growth as Gaussian process describing genet size at time $t+1$ as a function of size at $t$ and climate covariates:
+
+\begin{align}
+x_{ijQ,t+1} &= \gamma^{G}_{j,t} + \phi^{G}_{jQ} + \beta^{G}_{j,t}x_{ij,t} + \omega^{G}_{j}w_{ij,t} + \nu^{S}_{j}w_{ij,t}x_{ij,t} + \theta^{G}_{jk}C_{k,t} \\
+Y^{G}_{ijQ,t} &\sim \text{Normal}(x_{ijQ,t+1}, \sigma_{j})
+\end{align}
+
+where $x$ is log genet size and all other parameters are as described for the survival regression.
+
+**Data, process, and parameter models.**
+\begin{align}
+&\mathtt{data} \qquad &\textbf{y} \sim \text{Normal}(\boldsymbol{\mu}, \boldsymbol{\sigma}^2) \\
+&\mathtt{process} \qquad &\mu_{iQ,t+1} = \gamma_{t} + \phi_{Q} + \beta_{t}x_{i,t} + \omega w_{i,t} + \nu w_{i,t}x_{i,t} + \theta_{k}C_{k,t} \\
+& \qquad & \sigma^2_{iQ,t} = ae^{b\mu_{iQ,t+1}} \\
+&\mathtt{parameters} \qquad &\gamma_{t} \sim \text{Normal}(\tilde{\gamma}, \sigma_\gamma^2)\\
+& \qquad & \phi_{Q} \sim \text{Normal}(0,\sigma_Q^2) \\
+& \qquad & \beta_{t} \sim \text{Normal}(\tilde{\beta}, \sigma_\beta^2) \\
+& \qquad & \omega \sim \text{Normal}(0, 0.001) \\
+& \qquad & \nu \sim \text{Normal}(0, 0.001) \\
+& \qquad & \boldsymbol{\theta} \sim \text{Normal}(0, 0.001) \\
+& \qquad & \tilde{\gamma} \sim \text{Normal}(0, 0.001) \\
+& \qquad & \tilde{\beta} \sim \text{Normal}(0, 0.001) \\
+& \qquad & a \sim \text{Normal}(0, 1000) \\
+& \qquad & b \sim \text{Normal}(0, 1000) \\
+& \qquad & \sigma_Q^2 \sim \text{Uniform}(0, 1000) \\
+& \qquad & \sigma_\gamma^2 \sim \text{Uniform}(0, 1000) \\
+& \qquad & \sigma_\beta^2 \sim \text{Uniform}(0, 1000)
+\end{align}
+
+**Full expression of posterior and joint distributions**
+\begin{align}
+[\boldsymbol{\gamma}, \tilde{\gamma}, \boldsymbol{\phi}, \boldsymbol{\beta}, \tilde{\beta}, \omega, \boldsymbol{\theta}, \sigma^2, \sigma_Q^2, \sigma_\beta^2] &\propto \\
+&\prod_{t=1}^T \prod_{i=1}^n [y_{iQ,t} | \gamma_{t}, \phi_{Q}, \beta_{t}, \omega, \boldsymbol{\theta}, \sigma^2] [\gamma_{t} | \tilde{\gamma}] [\beta_{t} | \tilde{\beta}, \sigma_\beta^2] \times \\
+&\prod_{Q=1}^{Q_{tot}} [\phi_{Q} | \sigma_Q^2] \times \\
+&[\boldsymbol{\theta}] [\omega] [\tilde{\gamma}] [\tilde{\beta}] [\sigma_Q^2] [\sigma_\beta^2]
+\end{align}
+
+**Stan code for model**
+```{r grow_stan, echo=TRUE, eval=FALSE}
+data{
+  int<lower=0> N; // observations
+  int<lower=0> Yrs; // years
+  int<lower=0> yid[N]; // year id
+  int<lower=0> Covs; // climate covariates
+  int<lower=0> G; // groups
+  int<lower=0> gid[N]; // group id
+  vector[N] Y; // observation vector
+  matrix[N,Covs] C; // climate matrix
+  vector[N] X; // size vector
+  matrix[N,2] W; // crowding matrix
+}
+parameters{
+  real a_mu;
+  vector[Yrs] a;
+  real b1_mu;
+  vector[Yrs] b1;
+  vector[Covs] b2;
+  vector[2] w;
+  real gint[G];
+  real tau;
+  real tauSize;
+  real<lower=0> sig_a;
+  real<lower=0> sig_b1;
+  real<lower=0> sig_G;
+}
+transformed parameters{
+  real mu[N];
+  real<lower=0> sigma[N];
+  vector[N] climEff;
+  vector[N] crowdEff;
+  climEff <- C*b2;
+  crowdEff <- W*w;
+  for(n in 1:N){
+    mu[n] <- a[yid[n]] + gint[gid[n]] + b1[yid[n]]*X[n] + crowdEff[n] + climEff[n];
+    sigma[n] <- sqrt((fmax(tau*exp(tauSize*mu[n]), 0.0000001)));  
+  }
+}
+model{
+  // Priors
+  a_mu ~ normal(0,1000);
+  w ~ normal(0,1000);
+  b1_mu ~ normal(0,1000);
+  tau ~ normal(0,1000);
+  tauSize ~ normal(0,1000);
+  sig_a ~ uniform(0,1000);
+  sig_b1 ~ uniform(0,1000);
+  sig_G ~ uniform(0,1000);
+  for(g in 1:G)
+      gint[g] ~ normal(0, sig_G);
+  for(c in 1:Covs)
+    b2[c] ~ normal(0,1000);
+  for(y in 1:Yrs){
+    a[y] ~ normal(a_mu, sig_a);
+    b1[y] ~ normal(b1_mu, sig_b1);
+  }
+
+  // Likelihood
+  Y ~ normal(mu, sigma);
+}
+```
+
+Recruitment (IPM)
+-----------------
+Our data allows us to track new recruits, but we cannot assign a specific parent to new genets.
+So, for recruitment, we work at the quadrat level and model the number of new individuals of species $j$ in quadrat $q$ recruiting at time $t+1$ as a function of quadrat "effective cover" ($A'$) in the previous year ($t$).
+Effective cover is a mixture of observed cover ($A$) in the focal quadrat ($q$) and the mean cover across the entire group ($\bar{A}$) of $Q$ quadrats in which $q$ is located:
+
+\begin{equation}
+A'_{jq,t} = p_{j}A_{jq,t} + (1-p_{j})\bar{A}_{jQ,t}
+\end{equation}
+
+where $p$ is a mixing fraction between 0 and 1 that is estimated within the model.
+
+We assume the number of individuals, $Y^{R}$, recruiting at time $t+1$ follows a negative binomial distribution:
+
+\begin{equation}
+Y^{R}_{jq,t+1} \sim \text{NegBin}(\lambda_{jq,t+1},\zeta)
+\end{equation}
+
+where $\lambda$ is the mean intensity and $\zeta$ is the size parameter. We define $\lambda$ as:
+
+\begin{equation}
+\lambda_{jq,t+1} = A'_{jq,t}e^{(\gamma^{R}_{j,t} + \phi^{R}_{jQ} + \theta^{R}_{jk}C_{k,t} + \omega^{R}\sqrt{A'_{q,t}})}
+\end{equation}
+
+where $A'$ is effective cover ($\text{cm}^2$) of species $j$ in quadrat $q$ and all other terms are as in the survival and growth regressions.
+
+**Data, process, and parameter models**
+\begin{align}
+&\mathtt{data} \qquad &\textbf{y} \sim \text{NegBin}(\boldsymbol{\lambda},\zeta) \\
+&\mathtt{process} \qquad &\lambda_{jq,t+1} = A'_{q,t}e^{(\gamma_{t} + \phi_{Q} + \theta_{k}C_{k,t} + \omega\sqrt{A'_{q,t}})} \\
+&\mathtt{parameters} \qquad &\gamma_{t} \sim \text{Normal}(\tilde{\gamma}, \sigma_\gamma^2)\\
+& \qquad & \phi_{Q} \sim \text{Normal}(0,\sigma_Q^2) \\
+& \qquad & \omega \sim \text{Uniform}(-100, 100) \\
+& \qquad & \boldsymbol{\theta} \sim \text{Uniform}(-10, 10) \\
+& \qquad & \tilde{\gamma} \sim \text{Uniform}(-300, 300) \\
+& \qquad & \sigma_Q^2 \sim \text{Cauchy}(0, 5) \\
+& \qquad & \sigma_\gamma^2 \sim \text{Cauchy}(0, 5)
+\end{align}
+
+**Full expression of posterior and joint distributions**
+\begin{align}
+[\boldsymbol{\gamma}, \tilde{\gamma}, \boldsymbol{\phi}, \boldsymbol{\beta}, \tilde{\beta}, \omega, \boldsymbol{\theta}, \sigma_Q^2, \sigma_\beta^2] &\propto \\
+&\prod_{t=1}^T \prod_{i=1}^n [y_{iQ,t} | \gamma_{t}, \phi_{Q}, \beta_{t}, \omega, \boldsymbol{\theta}] [\gamma_{t} | \tilde{\gamma}] [\beta_{t} | \tilde{\beta}, \sigma_\beta^2] \times \\
+&\prod_{Q=1}^{Q_{tot}} [\phi_{Q} | \sigma_Q^2] \times \\
+&[\boldsymbol{\theta}] [\omega] [\tilde{\gamma}] [\tilde{\beta}] [\sigma_Q^2] [\sigma_\beta^2]
+\end{align}
+
+**Stan code for model**
+```{r rec_stan, echo=TRUE, eval=FALSE}
+data{
+  int<lower=0> N; // observations
+  int<lower=0> Yrs; // years
+  int<lower=0> yid[N]; // year id
+  int<lower=0> Covs; // climate covariates
+  int<lower=0> G; // groups
+  int<lower=0> gid[N]; // group id
+  int<lower=0> Y[N]; // observation vector
+  matrix[N,Covs] C; // climate matrix
+  vector[N] parents1; // crowding vector
+  vector[N] parents2; // crowding vector
+}
+parameters{
+  real a_mu;
+  vector[Yrs] a;
+  vector[Covs] b2;
+  real dd;
+  real gint[G];
+  real<lower=0> sig_a;
+  real<lower=0> theta;
+  real<lower=0> sig_G;
+  real<lower=0, upper=1> u;
+}
+transformed parameters{
+  real mu[N];
+  vector[N] climEff;
+  vector[N] trueP1;
+  vector[N] trueP2;
+  vector[N] lambda;
+  vector[N] q;
+  climEff <- C*b2;
+  for(n in 1:N){
+    trueP1[n] <- parents1[n]*u + parents2[n]*(1-u);
+    trueP2[n] <- sqrt(trueP1[n]);
+    mu[n] <- exp(a[yid[n]] + gint[gid[n]] + dd*trueP2[n] + climEff[n]);
+    lambda[n] <- trueP1[n]*mu[n];
+    q[n] <- lambda[n]*theta;
+  }
+}
+model{
+  // Priors
+  u ~ uniform(0,1);
+  theta ~ uniform(0,100);
+  a_mu ~ normal(0,1000);
+  dd ~ uniform(-100,100);
+  sig_a ~ cauchy(0,5);
+  sig_G ~ cauchy(0,5);
+  for(g in 1:G)
+      gint[g] ~ normal(0, sig_G);
+  for(c in 1:Covs)
+    b2 ~ uniform(-100,100);
+  for(y in 1:Yrs){
+    a[y] ~ normal(a_mu, sig_a);
+  }
+
+  // Likelihood
+  Y ~ neg_binomial_2(q, theta);
+}
+```

### 2015-07-08 manuscript/manuscript.Rmd
updates appendix with IPM model dscriptions

@@
 
+These results lead us to conclude that detailed demographic data is necessary to forecast climate change impacts on plant populations.
+This is unwelcome news since this data is difficult to collect and the models built on such data are of little use to land managers that make decisions at scales beyond that of traditional population models [@Queenborough2011].
+While density-structured approaches may fail when climate covariates are considered, there are other alternatives.
+For example, @Clark2011a use Forest Inventory and Analysis (FIA) data to parameterize a population model with multiple vital rates and climate dependence.
+Another example are distributed efforts like PlantPopNet (http://plantago.plantpopnet.com) that will allow researchers to estimate variation around climate responses for widespread species.
+Lastly, we foresee new approaches on the horizon that leverage photo/video of plots and advanced object recognition algorithms to streamline plant mapping and digitizing efforts.
+
 ### Forecasting the future, and the future of forecasting
@@
 In some cases this is because the easiest statistical methods do no make propogating parameter uncertainty easy.
-But even using Bayesian approaches that allow straightforward integration of model fitting and forecasting [@Hooten2014; @Hobbs2015] is not simple when using modeling approaches like integral projection models that separate the model fitting and simulation stages [@Rees2009].
+Even using Bayesian approaches that allow straightforward integration of model fitting and forecasting [@Hobbs2015] is not simple when using modeling approaches like integral projection models that separate the model fitting and simulation stages [@Rees2009].
 However, as we have done here, it is still possible to include parameter uncertainty by drawing parameter values from MCMC iterations, taking care to draw all parameters from the same chain and iteration to account for their correlations.
-Only by being honest about our forecasts cen we begin to produce better ones.
+Only by being honest about our forecasts can we begin to produce better ones.
 

### 2015-07-08 manuscript/manuscript.tex
updates appendix with IPM model dscriptions

@@
 
-\newpage{}
-
 \subsection{Abstract}\label{abstract}
@@
 climate covariates (e.g., ppt1 $\times$ TmeansSpr1) and climate $\times$
-size interactions.
+size interactions. Climate $\times$ size interactions are for climate
+main effects only, that is we do not include interactions among size and
+interacting climate effects.
 
@@
 group location, and $\theta^{S}_{k}$ is the fixed parameter for the
-effect of the $k$th climate covariate at time $t$ ($C_{k,t}$). We
-include density-dependence by estimating the effect of crowding on the
-focal individual by other individuals of the same species. $\omega$ is
-the effect of crowding and $w_{t,Q}$ is the crowding experienced by the
-focal individual at time $t$ in quadrat group $Q$.
+effect of the $k$th climate covariate at time $t$ ($C_{k,t}$). Note that
+the vector of climate covariates (\textbf{C}) includes climate variable
+interactions and climate$\times$size interactions. We include
+density-dependence by estimating the effect of crowding on the focal
+individual by other individuals of the same species. $\omega$ is the
+effect of crowding and $w_{t,Q}$ is the crowding experienced by the
+focal individual at time $t$ in quadrat group $Q$. We include a
+size$\times$crowding interaction effect ($\nu^{S}$).
 
@@
 quadrat $q$ at time $t$ and all other parameters are as in the
-individual-level growth model (Eq. 3). The log normal likelihood
-includes a truncation (T{[}0,1{]}) to ensure that predicted values do
-not exceed 100\% cover.
+individual-level growth model (Eq. 3). Again, note that the climate
+covariate vector (\textbf{C}) includes the climate$\times$cover
+interaction. The log normal likelihood includes a truncation
+(T{[}0,1{]}) to ensure that predicted values do not exceed 100\% cover.
 
@@
 
+These results lead us to conclude that detailed demographic data is
+necessary to forecast climate change impacts on plant populations. This
+is unwelcome news since this data is difficult to collect and the models
+built on such data are of little use to land managers that make
+decisions at scales beyond that of traditional population models
+(Queenborough et al. 2011). While density-structured approaches may fail
+when climate covariates are considered, there are other alternatives.
+For example, ({\textbf{???}}) use Forest Inventory and Analysis (FIA)
+data to parameterize a population model with multiple vital rates and
+climate dependence. Another example are distributed efforts like
+PlantPopNet (\url{http://plantago.plantpopnet.com}) that will allow
+researchers to estimate variation around climate responses for
+widespread species. Lastly, we foresee new approaches on the horizon
+that leverage photo/video of plots and advanced object recognition
+algorithms to streamline plant mapping and digitizing efforts.
+
 \subsubsection{Forecasting the future, and the future of
@@
 cases this is because the easiest statistical methods do no make
-propogating parameter uncertainty easy. But even using Bayesian
-approaches that allow straightforward integration of model fitting and
-forecasting (Hooten and Hobbs 2014, Hobbs and Hooten 2015) is not simple
-when using modeling approaches like integral projection models that
-separate the model fitting and simulation stages (Rees and Ellner 2009).
-However, as we have done here, it is still possible to include parameter
-uncertainty by drawing parameter values from MCMC iterations, taking
-care to draw all parameters from the same chain and iteration to account
-for their correlations. Only by being honest about our forecasts cen we
-begin to produce better ones.
+propogating parameter uncertainty easy. Even using Bayesian approaches
+that allow straightforward integration of model fitting and forecasting
+(Hobbs and Hooten 2015) is not simple when using modeling approaches
+like integral projection models that separate the model fitting and
+simulation stages (Rees and Ellner 2009). However, as we have done here,
+it is still possible to include parameter uncertainty by drawing
+parameter values from MCMC iterations, taking care to draw all
+parameters from the same chain and iteration to account for their
+correlations. Only by being honest about our forecasts can we begin to
+produce better ones.
 
@@
 aggregated data may prove useful and unbiased. However, our work here is
-the first comparison of population models based on individual and
-aggregated data in a climate change context. Our results confirm
-theoretical arguments (Clark et al. 2011b) and empirical evidence (Clark
-et al. 2011a, 2012) that individual responses are critical to predicting
-species' responses to climate change. Thus, forecasts from aggregate
-level models should be viewed with caution and should never be
-unaccompianed by uncertainty.
+the first comparison, to our knowledge, of population models based on
+individual and aggregated data in a climate change context. Our results
+confirm theoretical arguments (Clark et al. 2011b) and empirical
+evidence (Clark et al. 2011a, 2012) that individual responses are
+critical to predicting species' responses to climate change. Thus,
+forecasts from aggregate level models should be viewed with caution and
+should never be unaccompianed by uncertainty. Given the importance of
+demographic data and its current difficulty to collect, we advocate for
+research on modern methods to collect demographic data more efficiently
+across environmental gradients in space and time. For example, recent
+advances in object recognition algorithms (Liu et al. 2014) provide new
+opportunities to speed up data collection and digital processing.
 
@@
 
-Hooten, M. B., and N. T. Hobbs. 2014. A guide to Bayesian model
-selection for ecologists. Ecological Monographs 85:3--28.
+Liu, Y., Y. Jang, W. Woo, and T.-K. Kim. 2014. Video-Based Object
+Recognition Using Novel Set-of-Sets Representations.
 

### 2015-07-09 manuscript/appB.Rmd
adds appendix B for statistical model results

---
layout: 12pt
header-includes:
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{setspace}
bibliography: ~/Dropbox/Bibliography/MicroMeso.bib
csl: components/ecology.csl

## rmarkdown render options
output:
  pdf_document:
    fig_caption: true
    keep_tex: true
fontsize: 12pt
geometry: margin=1in

---

```{r lib_reqs, echo=FALSE, message=FALSE, warning=FALSE}
library(xtable)
library(plyr)
library(reshape2)
```

```{r agg_fxn, echo=FALSE, message=FALSE, warning=FALSE}
agg_fxn <- function(mcmc_frame){
  out <- ddply(mcmc_frame, .(Parameter), summarise,
               mean_value = mean(value),
               sd_value = sd(value),
               up_BCI = quantile(value, 0.975),
               lo_BCI = quantile(value, 0.025))
  return(out)
}
spp_list <- c("BOGR", "HECO", "PASM", "POSE")
spp_name <- c("B. gracilis.", "H. comata.", "P. smithii.", "P. secunda.")
```

Appendix B: Statistical results for model fits
===============================================
Here we provide parameter estimates (mean, standard deviations, upper and lower 95% Bayesian credible intervals) from each vital rate model for the integral projection model and for the quadrat based model. We provide parameter-to-notation keys below for clarity.

Parameter to model notation keys
--------------------------------

Table:  Definition of parameters and model notation for IPM survival and growth models.

| Parameter | Definition | Notation in statistical results tables |
| --------- | ---------- | -------------------------------------- |
| $\gamma_{t}$ | intercept for year *t* | ``a[t]`` where ``t`` is the numeric year |
| $\tilde{\gamma}$ | global intercept | ``a_mu`` |
| $\phi_{Q}$ | random effect of quadrat group *Q* | ``gint[Q]`` where ``Q`` is the quadrat group |
| $\beta_{t}$ | size effect for year *t* | ``b1[t]`` where ``t`` is the numeric year |
| $\tilde{\beta}$ | global size effect | ``b1_mu`` |
| $\gamma$ | effect of crowding | ``w[1]`` |
| $\nu$ | crowding $\times$ size effect | ``w[2]`` |
| $\theta_{k}$ | effect of climate covariate *k* | ``b2[k]`` where ``k`` is the climate effect |
| *a* | first parameter for growth variance | ``tau`` |
| *b* | second parameter for growth variance | ``tauSize`` |

Table:  Definition of parameters and model notation for IPM recruitment model.

| Parameter | Definition | Notation in statistical results tables |
| --------- | ---------- | -------------------------------------- |
| $\gamma_{t}$ | intercept for year *t* | ``a[t]`` where ``t`` is the numeric year |
| $\tilde{\gamma}$ | global intercept | ``a_mu`` |
| $\phi_{Q}$ | random effect of quadrat group *Q* | ``gint[Q]`` where ``Q`` is the quadrat group |
| $\gamma$ | effect of plot cover | ``dd`` |
| $\theta_{k}$ | effect of climate covariate *k* | ``b2[k]`` where ``k`` is the climate effect |
| *p* | mixing fraction for effective plot cover | ``u`` |
| $\zeta$ | size parameter for negative binimial likelihood | ``theta`` |

Table:  Definition of parameters and model notation for quadrat based model.

| Parameter | Definition | Notation in statistical results tables |
| --------- | ---------- | -------------------------------------- |
| $\gamma_{t}$ | intercept for year *t* | ``a[t]`` where ``t`` is the numeric year |
| $\tilde{\gamma}$ | global intercept | ``a_mu`` |
| $\phi_{Q}$ | random effect of quadrat group *Q* | ``gint[Q]`` where ``Q`` is the quadrat group |
| $\beta_{t}$ | cover effect for year *t* | ``b1[t]`` where ``t`` is the numeric year |
| $\tilde{\beta}$ | global cover effect | ``b1_mu`` |
| $\theta_{k}$ | effect of climate covariate *k* | ``b2[k]`` where ``k`` is the climate effect |
| $\tau$ | model variance in log normal likelihood | ``tau`` |

Table: Climate effect key.

| Integer ID (*k*) | Climate covariate |
| ---------------- | ----------------- |
| 1 | pptLag |
| 2 | ppt1 |
| 3 | ppt2 |
| 4 | TmeanSpr1 |
| 5 | TmeanSpr2 |
| 6 | ppt1 $\times$ TmeanSpr1 |
| 7 | ppt2 $\times$ TmeanSpr2 |
| 8 | pptLag $\times$ size |
| 9 | ppt1 $\times$ size |
| 10 | ppt2 $\times$ size |
| 11 | TmeanSpr1 $\times$ size |
| 12 | TmeanSpr2 $\times$ size |

Results from statistical models
-------------------------------
*Tables begin on next page...*

```{r surv_params, echo=FALSE, results='asis', message=FALSE, warning=FALSE, comment=NA}
for(i in 1:length(spp_list)){
  file <- paste("../analysis/ipm/vitalRateRegs/survival/survival_stanmcmc_",
              spp_list[i], ".RDS",
              sep="")
  survmcmc <- readRDS(file)
  aggsurv <- agg_fxn(survmcmc)
  cap <- paste("Statistical results from IPM survival model for", spp_name[i])
  print(xtable(aggsurv, caption = cap), comment=FALSE, include.rownames=FALSE, 
        size="\\footnotesize", caption.placement = "top")
}
```

```{r growth_params, echo=FALSE, results='asis', message=FALSE, warning=FALSE, comment=NA}
for(i in 1:length(spp_list)){
  file <- paste("../analysis/ipm/vitalRateRegs/growth/growth_stanmcmc_",
              spp_list[i], ".RDS",
              sep="")
  survmcmc <- readRDS(file)
  aggsurv <- agg_fxn(survmcmc)
  cap <- paste("Statistical results from IPM growth model for", spp_name[i])
  print(xtable(aggsurv, caption = cap), comment=FALSE, include.rownames=FALSE, 
        size="\\footnotesize", caption.placement = "top")
}
```

```{r rec_params, echo=FALSE, results='asis', message=FALSE, warning=FALSE, comment=NA}
for(i in 1:length(spp_list)){
  file <- paste("../analysis/ipm/vitalRateRegs/recruitment/recruitment_stanmcmc_",
              spp_list[i], ".RDS",
              sep="")
  survmcmc <- readRDS(file)
  aggsurv <- agg_fxn(survmcmc)
  cap <- paste("Statistical results from IPM recruitment model for", spp_name[i])
  print(xtable(aggsurv, caption = cap), comment=FALSE, include.rownames=FALSE, 
        size="\\footnotesize", caption.placement = "top")
}
```

```{r qbm_params, echo=FALSE, results='asis', message=FALSE, warning=FALSE, comment=NA}
for(i in 1:length(spp_list)){
  file <- paste("../analysis/quadBM/vitalRateRegressions/truncNormModel/popgrowth_stanmcmc_",
              spp_list[i], ".RDS",
              sep="")
  survmcmc <- readRDS(file)
  aggsurv <- agg_fxn(survmcmc)
  cap <- paste("Statistical results from QBM population model for", spp_name[i])
  print(xtable(aggsurv, caption = cap), comment=FALSE, include.rownames=FALSE, 
        size="\\footnotesize", caption.placement = "top")
}
```

### 2015-07-09 manuscript/manuscript.Rmd
updates ms with new references

@@
 
-where $k_{j}(v_{j},u_{j},\bar{\bold{w_{j}}})$ is the population kernal that describes all possible transitions from size $u$ to $v$ and $\bar{\bold{w_{j}}}$ is a vector of estimates of average crowding experienced from all other species by a genet of size $u_j$ and species $j$.
+where $k_{j}(v_{j},u_{j},\bar{\bold{w_{j}}})$ is the population kernel that describes all possible transitions from size $u$ to $v$ and $\bar{\bold{w_{j}}}$ is a vector of estimates of average crowding experienced from all other species by a genet of size $u_j$ and species $j$.
 The integral is evaluated over all possible sizes between predefined lower (_L_) and upper (_U_) size limits that extend beyond the range of observed genet sizes.
 
-The population kernal is defined as the joint contributions of survival (_S_), growth (_G_), and recruitment (_R_):
+The population kernel is defined as the joint contributions of survival (_S_), growth (_G_), and recruitment (_R_):
 
@@
 3a-d).
-The IPM projected percent changes in equilibrium cover from `r round(min(subset(meanparam_agg, species=="BOGR" & model=="IPM")["meandiff"])*100)` - `r round(max(subset(meanparam_agg, species=="BOGR" & model=="IPM")["meandiff"])*100)`% for *B.
-gracilis*, `r round(min(subset(meanparam_agg, species=="HECO" & model=="IPM")["meandiff"])*100)` - `r round(max(subset(meanparam_agg, species=="HECO" & model=="IPM")["meandiff"])*100)`% for *H.
-comata*, `r round(min(subset(meanparam_agg, species=="PASM" & model=="IPM")["meandiff"])*100)` - `r round(max(subset(meanparam_agg, species=="PASM" & model=="IPM")["meandiff"])*100)`% for *P.
-smithii*, and `r round(min(subset(meanparam_agg, species=="POSE" & model=="IPM")["meandiff"])*100)` - `r round(max(subset(meanparam_agg, species=="POSE" & model=="IPM")["meandiff"])*100)`% for *P.
+The IPM projected percent changes in equilibrium cover from `r round(min(subset(meanparam_agg, species=="BOGR" & model=="IPM")["meandiff"])*100)` to `r round(max(subset(meanparam_agg, species=="BOGR" & model=="IPM")["meandiff"])*100)`% for *B.
+gracilis*, `r round(min(subset(meanparam_agg, species=="HECO" & model=="IPM")["meandiff"])*100)` to `r round(max(subset(meanparam_agg, species=="HECO" & model=="IPM")["meandiff"])*100)`% for *H.
+comata*, `r round(min(subset(meanparam_agg, species=="PASM" & model=="IPM")["meandiff"])*100)` to `r round(max(subset(meanparam_agg, species=="PASM" & model=="IPM")["meandiff"])*100)`% for *P.
+smithii*, and `r round(min(subset(meanparam_agg, species=="POSE" & model=="IPM")["meandiff"])*100)` to `r round(max(subset(meanparam_agg, species=="POSE" & model=="IPM")["meandiff"])*100)`% for *P.
 secunda*.
 The QBM projected opposite and greater percent changes in equilibrium cover for *B.
-gracilis* (`r round(min(subset(meanparam_agg, species=="BOGR" & model=="QBM")["meandiff"])*100)` - `r round(max(subset(meanparam_agg, species=="BOGR" & model=="QBM")["meandiff"])*100)`%) and *H.
-comata* (`r round(min(subset(meanparam_agg, species=="HECO" & model=="QBM")["meandiff"])*100)` - `r round(max(subset(meanparam_agg, species=="HECO" & model=="QBM")["meandiff"])*100)`%; Fig.
+gracilis* (`r round(min(subset(meanparam_agg, species=="BOGR" & model=="QBM")["meandiff"])*100)` to `r round(max(subset(meanparam_agg, species=="BOGR" & model=="QBM")["meandiff"])*100)`%) and *H.
+comata* (`r round(min(subset(meanparam_agg, species=="HECO" & model=="QBM")["meandiff"])*100)` to `r round(max(subset(meanparam_agg, species=="HECO" & model=="QBM")["meandiff"])*100)`%; Fig.
 3a-b).
 For *P.
-smithii*, the QBM projected opposite changes in equilibrium cover than the IPM, but of similar magnitude (`r round(min(subset(meanparam_agg, species=="PASM" & model=="QBM")["meandiff"])*100)` - `r round(max(subset(meanparam_agg, species=="PASM" & model=="QBM")["meandiff"])*100)`%; Fig.
+smithii*, the QBM projected opposite changes in equilibrium cover than the IPM, but of similar magnitude (`r round(min(subset(meanparam_agg, species=="PASM" & model=="QBM")["meandiff"])*100)` to `r round(max(subset(meanparam_agg, species=="PASM" & model=="QBM")["meandiff"])*100)`%; Fig.
 3c).
 *P.
-secunda* was the only species that the IPM and QBM made projections of the same sign and somewhat similar magnitude (Fig.
+secunda* was the only species that the IPM and QBM made projections of the same sign and somewhat similar magnitude (`r round(min(subset(meanparam_agg, species=="POSE" & model=="QBM")["meandiff"])*100)` to `r round(max(subset(meanparam_agg, species=="POSE" & model=="QBM")["meandiff"])*100)`%; Fig.
 3d).
@@
 3h).
+At least two of three vital rates were sensitive to climate for each species (Fig.
+3).
 
@@
 While density-structured approaches may fail when climate covariates are considered, there are other alternatives.
-For example, @Clark2011a use Forest Inventory and Analysis (FIA) data to parameterize a population model with multiple vital rates and climate dependence.
-Another example are distributed efforts like PlantPopNet (http://plantago.plantpopnet.com) that will allow researchers to estimate variation around climate responses for widespread species.
-Lastly, we foresee new approaches on the horizon that leverage photo/video of plots and advanced object recognition algorithms to streamline plant mapping and digitizing efforts.
+For example, @Clark2011 use Forest Inventory and Analysis (FIA) data to parameterize a population model with multiple vital rates and climate dependence.
+Another example are distributed efforts like PlantPopNet (http://plantago.plantpopnet.com) that will allow researchers to estimate variation around climate responses for widespread species by taking advantage of spatial variation in climate [e.g.
+@Doak2010].
+Lastly, we foresee new approaches on the horizon that leverage photo/video of plots and advanced object recognition algorithms [e.g.
+@Liu2014] to streamline plant mapping and digitizing efforts.
 
 ### Forecasting the future, and the future of forecasting
-Our goal was not make any explicit forecast for the future state of these populations based on predicted climate change.
+Our goal was not to make any explicit forecast for the future state of these populations based on predicted climate change.
 But our results highlight the state of affairs in ecology when it comes to forecasting the impacts of climate change.
-The analysis we conducted here could be considered, with some exceptions of course, at the forefront of ecological forecasting in terms of the statistical approach employed (hierarchical Bayesian), the type of population model we used (stochastic IPM with parameter uncertainty), and the amount of data we had at our disposal (14 years of individual-level data).
+The analysis we conducted here could be considered, with some exceptions of course, at the forefront of ecological forecasting in terms of the statistical approach employed (hierarchical Bayesian), the type of population model we used (stochastic IPM with parameter uncertainty), and the amount of high quality data we had at our disposal (14 years of individual level data).
 Yet, model predictions proved so uncertain that any forecast, when bounded with model and parameter uncertainty, would be at best not useful and at worst meaningless.
@@
 First, forecasts could be improved by matching the spatial scale of predictor variables with the spatial scale of observations.
-One of the major limitations of the models we fit here is that the climate data are at a much larger scale than the plot and individual level observations of plant cover and size.
+One of the major limitations of the models we fit here is that the climate data are at a much larger scale than the individual level observations of plant size.
 Climate covariates only vary by year, with no spatial variability within years.
-Thus, even if we fit models to individual level data, we are missing the key interaction point between weather and individual plants [@Clark2011b].
+Thus, even if we fit models to individual level data, we are missing the key interaction point between weather and individual plants [@Clark2011b] because all observations share the same climate covariates.
+Demographic studies should be designed with at least plot level measurements of climate related variables (e.g., soil moisture).
 
 Second, accurately detecting climate signals will take even longer time series.
-Recent theoretical work on detecting climate signals in noisy data suggests that even advanced approaches to parameter fitting like LASSO, functional linear models (splines), and random forest require 20-25 year time series (Teller et al., in review).
+Recent theoretical work on detecting climate signals in noisy data suggests that even advanced approaches to parameter fitting like LASSO, functional linear models (splines), and Random Forest models require 20-25 year time series (Teller et al., in review).
 Alternatively, as we suggest above, Teller et al.
@@
 There is a strong culture around explicitly considering model uncertainty, but parameter uncertainty is often ignored.
-In some cases this is because the easiest statistical methods do no make propogating parameter uncertainty easy.
-Even using Bayesian approaches that allow straightforward integration of model fitting and forecasting [@Hobbs2015] is not simple when using modeling approaches like integral projection models that separate the model fitting and simulation stages [@Rees2009].
+In some cases this is because the easiest statistical methods do no make propogating parameter uncertainty a straighforward task.
+Even Bayesian approaches that allow integration of model fitting and forecasting [@Hobbs2015] are not simple when using modeling approaches like integral projection models that separate the model fitting and simulation stages [@Rees2009].
 However, as we have done here, it is still possible to include parameter uncertainty by drawing parameter values from MCMC iterations, taking care to draw all parameters from the same chain and iteration to account for their correlations.
@@
 Given the importance of demographic data and its current difficulty to collect, we advocate for research on modern methods to collect demographic data more efficiently across environmental gradients in space and time.
-For example, recent advances in object recognition algorithms [@Liu2014] provide new opportunities to speed up data collection and digital processing.
 
@@
 We thank the original mappers of the permanent quadrats in Montana and the digitizers in the Adler lab, without whom this work would not have been possible.
-Informal conversations with Stephen Ellner, Giles Hooker, Robin Snyder, and a series of meetings among the Adler and Weecology labs at USU sharpened our thinking.
+Informal conversations with Stephen Ellner, Giles Hooker, Robin Snyder, and a series of meetings between the Adler and Weecology labs at USU sharpened our thinking.
 Compute, storage and other resources from the Division of Research Computing in the Office of Research and Graduate Studies at Utah State University are gratefully acknowledged.

### 2015-07-09 manuscript/manuscript.tex
updates ms with new references

@@
 
-where $k_{j}(v_{j},u_{j},\bar{\bold{w_{j}}})$ is the population kernal
+where $k_{j}(v_{j},u_{j},\bar{\bold{w_{j}}})$ is the population kernel
 that describes all possible transitions from size $u$ to $v$ and
@@
 
-The population kernal is defined as the joint contributions of survival
+The population kernel is defined as the joint contributions of survival
 (\emph{S}), growth (\emph{G}), and recruitment (\emph{R}):
@@
 Equilibrium cover from both models was sensitive to climate (Fig. 3a-d).
-The IPM projected percent changes in equilibrium cover from -3 - 8\% for
-\emph{B. gracilis}, -4 - 3\% for \emph{H. comata}, -15 - 9\% for
-\emph{P. smithii}, and -17 - 53\% for \emph{P. secunda}. The QBM
+The IPM projected percent changes in equilibrium cover from -3 to 8\%
+for \emph{B. gracilis}, -4 to 3\% for \emph{H. comata}, -15 to 9\% for
+\emph{P. smithii}, and -17 to 53\% for \emph{P. secunda}. The QBM
 projected opposite and greater percent changes in equilibrium cover for
-\emph{B. gracilis} (-63 - 30\%) and \emph{H. comata} (-50 - -18\%; Fig.
-3a-b). For \emph{P. smithii}, the QBM projected opposite changes in
-equilibrium cover than the IPM, but of similar magnitude (-5 - 6\%; Fig.
-3c). \emph{P. secunda} was the only species that the IPM and QBM made
-projections of the same sign and somewhat similar magnitude (Fig. 3d).
+\emph{B. gracilis} (-63 to 30\%) and \emph{H. comata} (-50 to -18\%;
+Fig. 3a-b). For \emph{P. smithii}, the QBM projected opposite changes in
+equilibrium cover than the IPM, but of similar magnitude (-5 to 6\%;
+Fig. 3c). \emph{P. secunda} was the only species that the IPM and QBM
+made projections of the same sign and somewhat similar magnitude (-20 to
+14\%; Fig. 3d).
 
@@
 equilibrium cover was negative for precipitation and temperature
-increases (Fig. 3h).
+increases (Fig. 3h). At least two of three vital rates were sensitive to
+climate for each species (Fig. 3).
 
@@
 when climate covariates are considered, there are other alternatives.
-For example, ({\textbf{???}}) use Forest Inventory and Analysis (FIA)
-data to parameterize a population model with multiple vital rates and
-climate dependence. Another example are distributed efforts like
+For example, Clark et al. (2011a) use Forest Inventory and Analysis
+(FIA) data to parameterize a population model with multiple vital rates
+and climate dependence. Another example are distributed efforts like
 PlantPopNet (\url{http://plantago.plantpopnet.com}) that will allow
 researchers to estimate variation around climate responses for
-widespread species. Lastly, we foresee new approaches on the horizon
-that leverage photo/video of plots and advanced object recognition
-algorithms to streamline plant mapping and digitizing efforts.
+widespread species by taking advantage of spatial variation in climate
+(e.g. Doak and Morris 2010). Lastly, we foresee new approaches on the
+horizon that leverage photo/video of plots and advanced object
+recognition algorithms (e.g. Liu et al. 2014) to streamline plant
+mapping and digitizing efforts.
 
@@
 
-Our goal was not make any explicit forecast for the future state of
+Our goal was not to make any explicit forecast for the future state of
 these populations based on predicted climate change. But our results
@@
 (hierarchical Bayesian), the type of population model we used
-(stochastic IPM with parameter uncertainty), and the amount of data we
-had at our disposal (14 years of individual-level data). Yet, model
-predictions proved so uncertain that any forecast, when bounded with
-model and parameter uncertainty, would be at best not useful and at
-worst meaningless. How might we improve on this state of affairs?
+(stochastic IPM with parameter uncertainty), and the amount of high
+quality data we had at our disposal (14 years of individual level data).
+Yet, model predictions proved so uncertain that any forecast, when
+bounded with model and parameter uncertainty, would be at best not
+useful and at worst meaningless. How might we improve on this state of
+affairs?
 
@@
 major limitations of the models we fit here is that the climate data are
-at a much larger scale than the plot and individual level observations
-of plant cover and size. Climate covariates only vary by year, with no
-spatial variability within years. Thus, even if we fit models to
-individual level data, we are missing the key interaction point between
-weather and individual plants (Clark et al. 2011b).
+at a much larger scale than the individual level observations of plant
+size. Climate covariates only vary by year, with no spatial variability
+within years. Thus, even if we fit models to individual level data, we
+are missing the key interaction point between weather and individual
+plants (Clark et al. 2011b) because all observations share the same
+climate covariates. Demographic studies should be designed with at least
+plot level measurements of climate related variables (e.g., soil
+moisture).
 
@@
 data suggests that even advanced approaches to parameter fitting like
-LASSO, functional linear models (splines), and random forest require
-20-25 year time series (Teller et al., in review). Alternatively, as we
-suggest above, Teller et al. (in review) also find that matching the
-scale of the response and predictors improves estimate precision.
+LASSO, functional linear models (splines), and Random Forest models
+require 20-25 year time series (Teller et al., in review).
+Alternatively, as we suggest above, Teller et al. (in review) also find
+that matching the scale of the response and predictors improves estimate
+precision.
 
@@
 cases this is because the easiest statistical methods do no make
-propogating parameter uncertainty easy. Even using Bayesian approaches
-that allow straightforward integration of model fitting and forecasting
-(Hobbs and Hooten 2015) is not simple when using modeling approaches
+propogating parameter uncertainty a straighforward task. Even Bayesian
+approaches that allow integration of model fitting and forecasting
+(Hobbs and Hooten 2015) are not simple when using modeling approaches
 like integral projection models that separate the model fitting and
@@
 research on modern methods to collect demographic data more efficiently
-across environmental gradients in space and time. For example, recent
-advances in object recognition algorithms (Liu et al. 2014) provide new
-opportunities to speed up data collection and digital processing.
+across environmental gradients in space and time.
 
@@
 conversations with Stephen Ellner, Giles Hooker, Robin Snyder, and a
-series of meetings among the Adler and Weecology labs at USU sharpened
+series of meetings between the Adler and Weecology labs at USU sharpened
 our thinking. Compute, storage and other resources from the Division of
@@
 
+Doak, D. F., and W. F. Morris. 2010. Demographic compensation and
+tipping points in climate-induced range shifts. Nature 467:959--962.
+
 Freckleton, R. P., W. J. Sutherland, A. R. Watkinson, and S. A.

### 2015-07-09 manuscript/appB.Rmd
switches order of up and lo BCI to lo then up for tables

@@
                sd_value = sd(value),
-               up_BCI = quantile(value, 0.975),
-               lo_BCI = quantile(value, 0.025))
+               lo_BCI = quantile(value, 0.025),
+               up_BCI = quantile(value, 0.975))
   return(out)

### 2015-07-10 manuscript/appA.Rmd
adds QBM full model description and STAN code

@@
 Also for clarity, we avoid subscripting species (the *j* subscripts) after initial description of each model.
+
+Parameter to model notation keys
+--------------------------------
+
+Table:  Definition of parameters and model notation for IPM survival and growth models.
+
+| Parameter | Definition | Notation in statistical results tables |
+| --------- | ---------- | -------------------------------------- |
+| $\gamma_{t}$ | intercept for year *t* | ``a[t]`` where ``t`` is the numeric year |
+| $\tilde{\gamma}$ | global intercept | ``a_mu`` |
+| $\phi_{Q}$ | random effect of quadrat group *Q* | ``gint[Q]`` where ``Q`` is the quadrat group |
+| $\beta_{t}$ | size effect for year *t* | ``b1[t]`` where ``t`` is the numeric year |
+| $\tilde{\beta}$ | global size effect | ``b1_mu`` |
+| $\gamma$ | effect of crowding | ``w[1]`` |
+| $\nu$ | crowding $\times$ size effect | ``w[2]`` |
+| $\theta_{k}$ | effect of climate covariate *k* | ``b2[k]`` where ``k`` is the climate effect |
+| *a* | first parameter for growth variance | ``tau`` |
+| *b* | second parameter for growth variance | ``tauSize`` |
+
+Table:  Definition of parameters and model notation for IPM recruitment model.
+
+| Parameter | Definition | Notation in statistical results tables |
+| --------- | ---------- | -------------------------------------- |
+| $\gamma_{t}$ | intercept for year *t* | ``a[t]`` where ``t`` is the numeric year |
+| $\tilde{\gamma}$ | global intercept | ``a_mu`` |
+| $\phi_{Q}$ | random effect of quadrat group *Q* | ``gint[Q]`` where ``Q`` is the quadrat group |
+| $\gamma$ | effect of plot cover | ``dd`` |
+| $\theta_{k}$ | effect of climate covariate *k* | ``b2[k]`` where ``k`` is the climate effect |
+| *p* | mixing fraction for effective plot cover | ``u`` |
+| $\zeta$ | size parameter for negative binimial likelihood | ``theta`` |
+
+Table:  Definition of parameters and model notation for quadrat based model.
+
+| Parameter | Definition | Notation in statistical results tables |
+| --------- | ---------- | -------------------------------------- |
+| $\gamma_{t}$ | intercept for year *t* | ``a[t]`` where ``t`` is the numeric year |
+| $\tilde{\gamma}$ | global intercept | ``a_mu`` |
+| $\phi_{Q}$ | random effect of quadrat group *Q* | ``gint[Q]`` where ``Q`` is the quadrat group |
+| $\beta_{t}$ | cover effect for year *t* | ``b1[t]`` where ``t`` is the numeric year |
+| $\tilde{\beta}$ | global cover effect | ``b1_mu`` |
+| $\theta_{k}$ | effect of climate covariate *k* | ``b2[k]`` where ``k`` is the climate effect |
+| $\tau$ | model variance in log normal likelihood | ``tau`` |
+
+Table: Climate effect key.
+
+| Integer ID (*k*) | Climate covariate |
+| ---------------- | ----------------- |
+| 1 | pptLag |
+| 2 | ppt1 |
+| 3 | ppt2 |
+| 4 | TmeanSpr1 |
+| 5 | TmeanSpr2 |
+| 6 | ppt1 $\times$ TmeanSpr1 |
+| 7 | ppt2 $\times$ TmeanSpr2 |
+| 8 | pptLag $\times$ size |
+| 9 | ppt1 $\times$ size |
+| 10 | ppt2 $\times$ size |
+| 11 | TmeanSpr1 $\times$ size |
+| 12 | TmeanSpr2 $\times$ size |
 
@@
 \begin{align}
-x_{ijQ,t+1} &= \gamma^{G}_{j,t} + \phi^{G}_{jQ} + \beta^{G}_{j,t}x_{ij,t} + \omega^{G}_{j}w_{ij,t} + \nu^{S}_{j}w_{ij,t}x_{ij,t} + \theta^{G}_{jk}C_{k,t} \\
+x_{ijQ,t+1} &= \gamma^{G}_{j,t} + \phi^{G}_{jQ} + \beta^{G}_{j,t}x_{ij,t} + \omega^{G}_{j}w_{ij,t} + \nu^{G}_{j}w_{ij,t}x_{ij,t} + \theta^{G}_{jk}C_{k,t} \\
 Y^{G}_{ijQ,t} &\sim \text{Normal}(x_{ijQ,t+1}, \sigma_{j})
@@
 & \qquad & \tilde{\beta} \sim \text{Normal}(0, 0.001) \\
-& \qquad & a \sim \text{Normal}(0, 1000) \\
-& \qquad & b \sim \text{Normal}(0, 1000) \\
+& \qquad & a \sim \text{Normal}(0, 0.001) \\
+& \qquad & b \sim \text{Normal}(0, 0.001) \\
 & \qquad & \sigma_Q^2 \sim \text{Uniform}(0, 1000) \\
@@
 \begin{align}
-[\boldsymbol{\gamma}, \tilde{\gamma}, \boldsymbol{\phi}, \boldsymbol{\beta}, \tilde{\beta}, \omega, \boldsymbol{\theta}, \sigma^2, \sigma_Q^2, \sigma_\beta^2] &\propto \\
-&\prod_{t=1}^T \prod_{i=1}^n [y_{iQ,t} | \gamma_{t}, \phi_{Q}, \beta_{t}, \omega, \boldsymbol{\theta}, \sigma^2] [\gamma_{t} | \tilde{\gamma}] [\beta_{t} | \tilde{\beta}, \sigma_\beta^2] \times \\
+[\boldsymbol{\gamma}, \tilde{\gamma}, \boldsymbol{\phi}, \boldsymbol{\beta}, \tilde{\beta}, \omega, \nu, \boldsymbol{\theta}, a, b, \sigma_Q^2, \sigma_\beta^2] &\propto \\
+&\prod_{t=1}^T \prod_{i=1}^n [y_{iQ,t} | \gamma_{t}, \phi_{Q}, \beta_{t}, \omega, \nu, \boldsymbol{\theta}, a, b] [\gamma_{t} | \tilde{\gamma}] [\beta_{t} | \tilde{\beta}, \sigma_\beta^2] \times \\
 &\prod_{Q=1}^{Q_{tot}} [\phi_{Q} | \sigma_Q^2] \times \\
-&[\boldsymbol{\theta}] [\omega] [\tilde{\gamma}] [\tilde{\beta}] [\sigma_Q^2] [\sigma_\beta^2]
+&[\boldsymbol{\theta}] [\omega] [\nu] [\tilde{\gamma}] [\tilde{\beta}] [\sigma_Q^2] [\sigma_\beta^2] [a] [b]
 \end{align}
@@
 & \qquad & \sigma_Q^2 \sim \text{Cauchy}(0, 5) \\
-& \qquad & \sigma_\gamma^2 \sim \text{Cauchy}(0, 5)
+& \qquad & \sigma_\gamma^2 \sim \text{Cauchy}(0, 5) \\
+& \qquad & \zeta \sim \text{Uniform}(0, 100) \\
+& \qquad & u \sim \text{Uniform}(0, 1)
 \end{align}
@@
 \begin{align}
-[\boldsymbol{\gamma}, \tilde{\gamma}, \boldsymbol{\phi}, \boldsymbol{\beta}, \tilde{\beta}, \omega, \boldsymbol{\theta}, \sigma_Q^2, \sigma_\beta^2] &\propto \\
-&\prod_{t=1}^T \prod_{i=1}^n [y_{iQ,t} | \gamma_{t}, \phi_{Q}, \beta_{t}, \omega, \boldsymbol{\theta}] [\gamma_{t} | \tilde{\gamma}] [\beta_{t} | \tilde{\beta}, \sigma_\beta^2] \times \\
-&\prod_{Q=1}^{Q_{tot}} [\phi_{Q} | \sigma_Q^2] \times \\
-&[\boldsymbol{\theta}] [\omega] [\tilde{\gamma}] [\tilde{\beta}] [\sigma_Q^2] [\sigma_\beta^2]
+[\boldsymbol{\gamma}, \tilde{\gamma}, \boldsymbol{\phi}, \omega, \boldsymbol{\theta}, \sigma_Q^2, \zeta, u] &\propto \\
+&\prod_{t=1}^T \prod_{i=1}^n [y_{iQ,t} | \gamma_{t}, \phi_{Q}, \omega, \boldsymbol{\theta}, \zeta, u] [\gamma_{t} | \tilde{\gamma}] \times \\
+&\prod_{Q=1}^{Q_{tot}} [\phi_{Q} | \sigma_Q^2] [\boldsymbol{\theta}] [\omega] [\tilde{\gamma}] [\sigma_Q^2] [\zeta] [u]
 \end{align}
@@
 ```
+
+Quadrat based model (QBM)
+-------------------------
+The model for quadrat cover ($\mu$) change from time $t$ to $t+1$ is
+
+\begin{align}
+\mu_{jq,t+1} &= \gamma_{j,t} + \phi_{jQ} + \beta_{j,t}x_{jq,t} + \theta_{jk}C_{k,t} \\
+Y_{jq,t+1} &\sim \text{LogNormal}(\mu_{jq,t+1}, \tau{j}) \text{T}[0,1]
+\end{align}
+
+where $x_{jq,t}$ is the log of species' $j$ proportional cover in quadrat $q$ at time $t$ and all other parameters are as in the individual-level growth model (Eq.
+3).
+Again, note that the climate covariate vector (**C**) includes the climate$\times$cover interaction.
+The log normal likelihood includes a truncation (T[0,1]) to ensure that predicted values do not exceed 100% cover.
+
+**Data, process, and parameter models.**
+\begin{align}
+&\mathtt{data} \qquad &\textbf{y} \sim \text{LogNormal}(\boldsymbol{\mu}, \boldsymbol{\tau}) \text{T}[0,1] \\
+&\mathtt{process} \qquad &\mu_{iQ,t+1} = \gamma_{t} + \phi_{Q} + \beta_{t}x_{i,t} + \theta_{k}C_{k,t} \\
+&\mathtt{parameters} \qquad &\gamma_{t} \sim \text{Normal}(\tilde{\gamma}, \sigma_\gamma^2)\\
+& \qquad & \phi_{Q} \sim \text{Normal}(0,\sigma_Q^2) \\
+& \qquad & \beta_{t} \sim \text{Normal}(\tilde{\beta}, \sigma_\beta^2) \\
+& \qquad & \boldsymbol{\theta} \sim \text{Uniform}(-10, 10) \\
+& \qquad & \tilde{\gamma} \sim \text{Uniform}(-300, 300) \\
+& \qquad & \tilde{\beta} \sim \text{Uniform}(-100, 100) \\
+& \qquad & \sigma_Q^2 \sim \text{Cauchy}(0, 5) \\
+& \qquad & \sigma_\gamma^2 \sim \text{Cauchy}(0, 5) \\
+& \qquad & \sigma_\beta^2 \sim \text{Cauchy}(0, 5) \\
+& \qquad & \tau \sim \text{Cauchy}(0, 5)
+\end{align}
+
+**Full expression of posterior and joint distributions**
+\begin{align}
+[\boldsymbol{\gamma}, \tilde{\gamma}, \boldsymbol{\phi}, \boldsymbol{\beta}, \tilde{\beta}, \boldsymbol{\theta}, \sigma_Q^2, \sigma_\beta^2, \tau] &\propto \\
+&\prod_{t=1}^T \prod_{i=1}^n [y_{iQ,t} | \gamma_{t}, \phi_{Q}, \beta_{t}, \boldsymbol{\theta}, \tau] [\gamma_{t} | \tilde{\gamma}] [\beta_{t} | \tilde{\beta}, \sigma_\beta^2] \times \\
+&\prod_{Q=1}^{Q_{tot}} [\phi_{Q} | \sigma_Q^2] \times \\
+&[\boldsymbol{\theta}] [\omega] [\tilde{\gamma}] [\tilde{\beta}] [\sigma_Q^2] [\sigma_\beta^2] [\tau]
+\end{align}
+
+**Stan code for model**
+```{r qbm_stan, echo=TRUE, eval=FALSE}
+data{
+  int<lower=0> N; // observations
+  int<lower=0> Yrs; // years
+  int<lower=0> yid[N]; // year id
+  int<lower=0> Covs; // climate covariates
+  int<lower=0> G; // groups
+  int<lower=0> gid[N]; // group id
+  real<lower=0,upper=1> Y[N]; // observation vector
+  matrix[N,Covs] C; // climate matrix
+  vector[N] X; // size vector
+}
+parameters{
+  real a_mu;
+  vector[Yrs] a;
+  real b1_mu;
+  vector[Yrs] b1;
+  vector[Covs] b2;
+  vector[G] gint;
+  real<lower=0> sig_a;
+  real<lower=0> sig_b1;
+  real<lower=0> sig_G;
+  real<lower=0> tau;
+}
+transformed parameters{
+  real mu[N];
+  vector[N] climEff;
+  climEff <- C*b2;
+  for(n in 1:N)
+    mu[n] <- a[yid[n]] + gint[gid[n]] + b1[yid[n]]*X[n] + climEff[n];
+}
+model{
+  // Priors
+  a_mu ~ uniform(-300,300);
+  b1_mu ~ uniform(-100,100);
+  sig_a ~ cauchy(0,5);
+  sig_b1 ~ cauchy(0,5);
+  sig_G ~ cauchy(0,5);
+  gint ~ normal(0, sig_G);
+  b2 ~ uniform(-10,10);
+  a ~ normal(a_mu, sig_a);
+  b1 ~ normal(b1_mu, sig_b1);
+  tau ~ cauchy(0,5);
+
+  //Likelihood
+  Y ~ lognormal(mu, tau);
+}
+```

### 2015-07-12 manuscript/manuscript.Rmd
adds section on SDMs to ms intro

@@
 
+That population models do not scale easily explains, in part, why species distribution models (SDMs) have become a major tool for land managers and conservation planners.
+SDMs typically rely on easy-to-collect presence/absence data (but see...) that allow researchers to cover large spatial extents (e.g., ).
+Likewise, predictors in SDMs tend to be remotely sensed products.
+Thus, it is relatively straightforward to parameterize and project SDMs at landscape and regional scales, the very scales at which land managers work.
+However, the limitations of SDMs are well known (cite some review).
+SOME MORE STUFF...
+It seems that what land managers really want are population models at the landscape and regional scale, if only the ecological community could provide them.
+
 Aggregate measures of individual plant performance, such as those typically collected as part of large-scale census efforts, offer an alternative to detailed demographic data for modeling populations [@Clark2004; @Freckleton2011].
@@
 
-All R code and data necessary to reproduce our analysis is archived on GitHub as release 1.0 (http://github.com/atredennick/MicroMesoForecast/releases).
+All R code and data necessary to reproduce our analysis is archived on GitHub as release v1.0[^revnote] (http://github.com/atredennick/MicroMesoForecast/releases).
 That stable release will remain static as a record of this analysis, but subsequent versions may appear if we update this work.
+
+[^revnote]: *Note to reviewers*: so that v1.0 will be associated with the published version of the manuscript, we have released v0.1 to be associated with this review version.
 
@@
 
-We fit all models using a hierarchical Bayesian approach, which we describe in more detail below.
-However, for each vital rate statistical model we also define the likelihood model we use.
-For the likelihood models, \textbf{Y} is always the relevant vector of observations (e.g., whether a  genet survived [1] or not [0] from year $t$ to $t+1$).
+We fit all models using a hierarchical Bayesian approach.
+The models are fully descibed in Appendix A, so here we focus on the main process and the mode likelihood.
+For the likelihood models, $\textbf{y}^X$ is always the relevant vector of observations for vital rate *X* ($X = S, G, R, or P for survival, growth, recruitment, or population growth) For example, $\textbf{y}^S$ is a vector of 0s and 1s indicating whether a genet surives from *t* to *t+1*, or not.
 
@@
 \text{logit}(S_{ijQ,t}) &= \gamma^{S}_{j,t} + \phi^{S}_{jQ} + \beta^{S}_{j,t}x_{ij,t} + \omega^{S}_{j}w_{ij,t} + \nu^{S}_{j}w_{ij,t}x_{ij,t} + \theta^{S}_{jk}C_{k,t} \\
-Y^{S}_{ijQ,t} &\sim \text{Bernoulli}(S_{ijQ,t})
+y^{S}_{ijQ,t} &\sim \text{Bernoulli}(S_{ijQ,t})
 \end{align}
@@
 x_{ijQ,t+1} &= \gamma^{G}_{j,t} + \phi^{G}_{jQ} + \beta^{G}_{j,t}x_{ij,t} + \omega^{G}_{j}w_{ij,t} + \nu^{S}_{j}w_{ij,t}x_{ij,t} + \theta^{G}_{jk}C_{k,t} \\
-Y^{G}_{ijQ,t} &\sim \text{Normal}(x_{ijQ,t+1}, \sigma_{j})
+y^{G}_{ijQ,t} &\sim \text{Normal}(x_{ijQ,t+1}, \sigma_{j})
 \end{align}
@@
 \begin{equation}
-Y^{R}_{jq,t+1} \sim \text{NegBin}(\lambda_{jq,t+1},\zeta)
+y^{R}_{jq,t+1} \sim \text{NegBin}(\lambda_{jq,t+1},\zeta)
 \end{equation}
@@
 \begin{align}
-x_{jq,t+1} &= \gamma^{G}_{j,t} + \phi^{G}_{jQ} + \beta^{G}_{j,t}x_{jq,t} + \theta^{S}_{jk}C_{k,t} \\
-Y^{G}_{jq,t+1} &\sim \text{LogNormal}(x_{jq,t+1}, \tau{j}) \text{T}[0,1]
+x_{jq,t+1} &= \gamma^{P}_{j,t} + \phi^{P}_{jQ} + \beta^{P}_{j,t}x_{jq,t} + \theta^{P}_{jk}C_{k,t} \\
+y^{P}_{jq,t+1} &\sim \text{LogNormal}(x_{jq,t+1}, \tau{j}) \text{T}[0,1]
 \end{align}
@@
 Within a species, we fit yearly size effects and yearly intercepts hierarchically where year-specific coefficients were drawn from global distributions representing the mean size effect and intercept.
-We used flat, uninformative priors for all unknown parameters (Appendix X).
+We used uninformative priors (Appendix A).
 
@@
 This step alleviates the need to reduce model parameters by model selection since sampling from the full parameter space in the MCMC ensures that if a parameter broadly overlaps zero, on average the effect in the population models will also be near zero.
+We report the posterior mean, standard deviation, and 95% Bayesian Credible Intervals for every parameter of each model for each species in Appendix B.
 
@@
 #### Quad-based model
-Our quad-based model (QBM) perfectly mirrors its statistical description (Eq. 8-9). We use the same approach for drawing parameter values as described for the IPM.
+Our quad-based model (QBM) perfectly mirrors its statistical description (Eqs. 8-9). We use the same approach for drawing parameter values as described for the IPM.
 
@@
 We ignore sizeXclimate interactions since these are not directly comparable across model types.
-The QBM does not have multiple vital rates, so its values are repeated across panels within each species.", fig.width=6, fig.height=6, cache=FALSE, results='hide'}
+The QBM does not have multiple vital rates, so its values are repeated across panels within each species.
+Across top panels, 'growth' = growth regression, 'rec' = recruitment regression, 'surv' = survival regression.", fig.width=6, fig.height=6, cache=FALSE, results='hide'}
 ##  Figure script to look at correlations between IPM and QBM
@@
   geom_point(size=3)+
-  geom_smooth(method="lm", color="black", fill="grey")+
+  geom_smooth(method="lm", color="black", fill="grey", se=FALSE)+
   facet_grid(species~vitalrate)+

### 2015-07-12 manuscript/manuscript.tex
adds section on SDMs to ms intro

@@
 
+That population models do not scale easily explains, in part, why
+species distribution models (SDMs) have become a major tool for land
+managers and conservation planners. SDMs typically rely on
+easy-to-collect presence/absence data (but see\ldots{}) that allow
+researchers to cover large spatial extents (e.g., ). Likewise,
+predictors in SDMs tend to be remotely sensed products. Thus, it is
+relatively straightforward to parameterize and project SDMs at landscape
+and regional scales, the very scales at which land managers work.
+However, the limitations of SDMs are well known (cite some review). SOME
+MORE STUFF\ldots{} It seems that what land managers really want are
+population models at the landscape and regional scale, if only the
+ecological community could provide them.
+
 Aggregate measures of individual plant performance, such as those
@@
 All R code and data necessary to reproduce our analysis is archived on
-GitHub as release 1.0
+GitHub as release v1.0\footnote{\emph{Note to reviewers}: so that v1.0
+  will be associated with the published version of the manuscript, we
+  have released v0.1 to be associated with this review version.}
 (\url{http://github.com/atredennick/MicroMesoForecast/releases}). That
@@
 
-We fit all models using a hierarchical Bayesian approach, which we
-describe in more detail below. However, for each vital rate statistical
-model we also define the likelihood model we use. For the likelihood
-models, \textbf{Y} is always the relevant vector of observations (e.g.,
-whether a genet survived {[}1{]} or not {[}0{]} from year $t$ to $t+1$).
+We fit all models using a hierarchical Bayesian approach. The models are
+fully descibed in Appendix A, so here we focus on the main process and
+the mode likelihood. For the likelihood models, $\textbf{y}^X$ is always
+the relevant vector of observations for vital rate \emph{X}
+($X = S, G, R, or P for survival, growth, recruitment, or population growth) For example, $\textbf{y}\^{}S\$
+is a vector of 0s and 1s indicating whether a genet surives from
+\emph{t} to \emph{t+1}, or not.
 
@@
 were drawn from global distributions representing the mean size effect
-and intercept. We used flat, uninformative priors for all unknown
-parameters (Appendix X).
+and intercept. We used uninformative priors (Appendix A).
 
@@
 ensures that if a parameter broadly overlaps zero, on average the effect
-in the population models will also be near zero.
+in the population models will also be near zero. We report the posterior
+mean, standard deviation, and 95\% Bayesian Credible Intervals for every
+parameter of each model for each species in Appendix B.
 
@@
 Our quad-based model (QBM) perfectly mirrors its statistical description
-(Eq. 8-9). We use the same approach for drawing parameter values as
+(Eqs. 8-9). We use the same approach for drawing parameter values as
 described for the IPM.
@@
 comparable across model types. The QBM does not have multiple vital
-rates, so its values are repeated across panels within each species.}
+rates, so its values are repeated across panels within each species.
+Across top panels, `growth' = growth regression, `rec' = recruitment
+regression, `surv' = survival regression.}
 \end{figure}

### 2015-07-13 esa_talk/esa2015.Rmd
adds to SDM ms section; adds esa presentation draft

--- 2015-07-08 manuscript/manuscript.Rmd
@@
 ---
-layout: 12pt
-header-includes:
-   - \usepackage{lineno}
-   - \linenumbers
-   - \usepackage{setspace}
-   - \doublespacing
-bibliography: ~/Dropbox/Bibliography/MicroMeso.bib
-csl: components/ecology.csl
-
-## rmarkdown render options
-output:
-  pdf_document:
-    fig_caption: true
-    keep_tex: true
-fontsize: 12pt
-geometry: margin=1in
-
+title: "Do we need detailed demographic data to forecast the impacts of climate change on plant populations?"
+author: "Andrew Tredennick and Peter Adler"
+date: " "
+output: beamer_presentation
 ---
 
-Do we need detailed demographic data to forecast the impacts of climate change on plant populations?
-========================================================================================
+## This is hard...
 
-### Andrew T. Tredennick[^corrauth] and Peter B. Adler
+\begin{center}
+{\includegraphics{idaho_field.pdf}}
+\end{center}
 
-*Andrew T. Tredennick, Department of Wildland Resources and the Ecology Center, Utah State University, Logan, UT*
+## This is hard...
 
-*Peter B. Adler, Department of Wildland Resources and the Ecology Center, Utah State University, Logan, UT*
+\begin{center}
+{\includegraphics{idaho_field.pdf}}
+\end{center}
 
-[^corrauth]: Corresponding author: [atredenn@gmail.com](mailto:atredenn@gmail.com)
+## Density-structured population models
 
-```{r libraris, include=FALSE}
+\begin{center}
+{\includegraphics{freck.png}}
+\end{center}
+
+## Ecological fallacy
+
+"...individuals respond to weather, not populations to climate."
+
+-- Clark 2011 (Ecology Letters)
+
+## Data
+
+```{r, echo=FALSE, include=TRUE, warning=FALSE, message=FALSE, fig.height=6}
+library(plyr)
+library(reshape2)
 library(ggplot2)
-library(plyr)
-library(gridExtra)
-library(reshape2)
-library(xtable)
-library(ggmcmc)
+obs_data <- read.csv("../analysis/speciesData/quadAllCover.csv")
+#remove zeros for averaging
+obs_data <- obs_data[which(obs_data$propCover>0),]
+avgobs <- ddply(obs_data, .(year, Species), summarise,
+                mean_cover = mean(propCover))
+
+ggplot()+
+  geom_line(data = obs_data, aes(x=(1900+year), y=propCover*100, group=quad), 
+            color="purple", size=0.25, alpha=0.25)+
+  geom_line(data=avgobs, aes(x=(1900+year), y=mean_cover*100))+
+  geom_point(data=avgobs, aes(x=(1900+year), y=mean_cover*100),size=3)+
+  ylab("Mean cover (%)")+
+  xlab("Year")+
+  theme_bw()+
+  facet_wrap("Species", ncol=2, nrow=2, scales = "free")+
+  guides(shape=FALSE, linetype=FALSE)
 ```
 
-```{r caching, include=FALSE}
-library("methods")
-library("knitr")
-basename <- "manuscript"
-opts_chunk$set(fig.path = paste("components/figure/", basename, "-", sep=""),
-               cache.path = paste("components/cache/", basename, "/", sep=""))
-opts_chunk$set(cache = 2)
-opts_chunk$set(tidy=FALSE, warning=FALSE, message=FALSE, 
-               comment = NA, verbose = TRUE, echo=FALSE)
+## Data
 
-# PDF-based figures
-opts_chunk$set(dev='pdf')
-```
+\begin{center}
+{\includegraphics{micromeso_flowchart.png}}
+\end{center}
 
-```{r plot-options, message=FALSE, warning=FALSE, include=FALSE, echo=FALSE}
-cbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", 
-               "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
+## Models
 
-opts_chunk$set(dev='pdf', dev.args=c(version="1.3"))
+1. Integral projection model (IPM)
+    + Survival: $p(S) = f(x_{t} + climate + density dependence)$
+    + Growth: $x_{t+1} = f(x_{t} + climate + density dependence)$
+    + Recruitment: $n_{t+1} = f(parents + climate)$
 
-```
+2. Quadrat based model (QBM)
+    + Population growth: $c_{t+1} = f(c_{t} + climate)$
 
-Abstract
---------
-Forecasting future states of populations has taken on new urgency as the rate of climate change increases.
-Traditional plant population models have limited utility in this regard because they are based on detailed demographic data from small, localized plots.
-These models are difficult to scale up to spatial scales relevant to land managers that require such forecasts to make decisions.
-To overcome the data limitations of traditional population models, some have proposed population models based on population level, rather than individual level, data that is much easier to collect over broad spatial scales.
-Using such models violates a central assumption of ecology: individuals respond to weather, not populations to climate.
-Here, we test whether this assumption is important when forecasting climate change impacts on four perennial grass species in a semi-arid Montana grassland.
-We parameterized two population models, one based on inidividual level data with three vital rates and one on an aggregated version of the same data (percent cover), with climate covariates and compared their accuracy, precision, and sensitivity to climate.
-The individual level model was more accurate and precise than the aggregate level model when predicting out of sample observations.
-The aggregate level model made countervailing forecasts to 1% climate changes when compared to the individual level model, indicating the aggregate level model produces incorrect forecasts.
-When comparing climate effects from both models, the aggregate level model tends to "miss" important climate effects from at least one vital rate for each species.
-It appears there is no short cut to forecasting climate change impacts on plant populations — detailed demographic data is essential.
-But, forecasts were very uncertain, so we advocate for a focus on new methods to collect demographic data more efficiently across environmental gradients in space and time.
+##  Model performance
 
-*Keywords*: forecasting, climate change, grassland, integral projection model, population model
-
-Introduction
-------------
-Population models are important tools for predicting the impacts of environmental change on species.
-But reconciling the scales at which population models are parameterized and the scales at which environmental changes play out remains a challenge [@Clark2010; @Freckleton2011; @Queenborough2011; @Clark2012].
-The major hurdle is that most population models, at least for plant species, are built using data from small, localized plots because parameterizing traditional population models requires tracking the fates of individuals.
-These models are difficult to scale up from the micro to meso-scales because the fitted parameters do not fully represent the spatial variation present at scales beyond that at which the data are collected [@Sæther2007].
-At the same time, most demographic data is collected over short time spans.
-For example, the most common study duration in the COMPADRE matrix population model database is 4 years and only a few exceed 10 years [@Salguero-Gomez2015].
-The constrained spatio-temporal extent of most demographic datasets reflects the difficulty of collecting such data, but those constraints limit our ability to extrapolate population models.
-Thus, our ability to use population models to predict the consequences of climate change is limited when we rely on individual-level data.
-
-Aggregate measures of individual plant performance, such as those typically collected as part of large-scale census efforts, offer an alternative to detailed demographic data for modeling populations [@Clark2004; @Freckleton2011].
-Such population-level data will never match the precision of individual-level data, but it is more feasible to attain a broad coverage sample when collecting coarse-scale data.
-This presents a difficult trade-off: on the one hand, individual-level data leads to more reliable models; on the other hand, population-level data leads to models that will produce less precise predictions but can be applied over greater spatial and temporal extents.
-An open question is how well models based on population-level data compare to models based on individual-level data.
-
-To date, relatively few studies have tried to model populations based on data other than detailed individual-level data.
-An important exception is an effort by @Taylor2004 to model the population growth rate of an invasive species to investigate the best strategies for invasion control.
-They used a "density-structured" model where the state variable is a discrete density state rather than a continuous density measure.
-Building on this work, @Freckleton2011 showed that density-structured models compare well to continuous models in theory, and @Queenborough2011 showed the application of such methods in a study on arable weeds.
-In particular, @Queenborough2011 provide empirical evidence that density-structured models are capable of reproducing population dynamics, even if some precision is lost when compared to fully continuous models.
-Thus, population models based on coarse, population-level data show promise for producing ecological forecasts at landscape and regional scales [@Queenborough2011].
-However, none of these models included environmental covariates.
-
-Basing population models on aggregated individual-level data in a climate change context is hampered by the fact that it is individuals that respond to climate, not populations [@Clark2012].
-This fact puts us in uneasy proximity to an "ecological fallacy" where one deduces inference on the individual from statistical inference on the group [@PIANTADOSI1988].
-For example, individual plants may respond positively to precipitation but a negative trend is observed at the population level due to increased competition among plants as they grow larger and consume more resources.
-Thus, it is important to ask the question: Can aggregated data be used to detect climate signals of the same sign and magnitude as individual-level data?
-If not, then building population models with climate covariates on aggregated data will lead to incorrect forecasts.
-
-Here, we test the assumption that statistical and population models based on aggregated data can detect climate signals as wells as models based on individual-level data.
-We use a unique demographic dataset that tracks the fates of individual plants from four species over 14 years to build single-species population models, since those are often used tools for ecological forecasts and climate vulnerability assessments.
-We first fit population models with interannual variation in vital rates explained, in part, by climate covariates.
-We then perturb the climate covariates to test the sensitivities of species to climate change.
-By doing these analyses using both individual and aggregated forms of the same data we can directly compare the two types of models.
-
-In general, we find that population models based on detailed demographic data are more accurate and precise than models based on aggregated data.
-Both types of models are able to detect climate signals, as evidenced by the sensitivity of simulated equilbrium plant cover under a perturbed climate scenario.
-But the two types of models produce inconsistent forecasts, in some cases producing completely opposing predictions.
-This leads us to conclude that, at least for these species at this location, detailed demographic data is necessary to detect the "right" climate signal.
-A worrying caveat to our work is that forecasts from both models were very uncertain.
-It seems that even 14 years worth of demographic data is not enough to produce meaningful forecasts when model uncertainty is explicitly considered.
-
-Materials and Methods
----------------------
-### Study site and data
-Our demographic data comes from the Fort Keogh Livestock and Range Research Laboratory in eastern Montana's northern mixed prairie near Miles City, Montana, USA ($46^{\circ}$ 19' N, $105^{\circ}$ 48' W).
-The dataset is freely available on Ecological Archives[^link] [@Anderson2011] , and interested readers should refer to the metadata therein for a complete description.
-The site is about 800 m above sea level and mean annual precipitation (1878-2009) is 334 mm, with most annual precipitation falling from April through September.
-The site is grass dominated and, for the purposes of our study, we focus on the four most abundant graminoid species: \emph{Bouteloua gracilis} (BOGR), \emph{Hesperostipa comata} (HECO), \emph{Pascopyrum smithii} (PASM), and \emph{Poa secunda} (POSE) (Fig.
-1).
-
-[^link]: http://esapubs.org/archive/ecol/E092/143/
-
-From 1932 to 1945 individual plants were identified and mapped annually in 44 1-$\text{m}^2$ quadrats using a pantograph.
-The quadrats were distributed in six pastures, each assigned a grazing treatment of light (1.24 ha/animal unit month), moderate (0.92 ha/aum), and heavy (0.76 ha/aum) stocking rates (two pastures per treatment).
-In this analysis we account for potential differences among the grazing treatments, but do not focus on grazing$\times$climate interactions.
-The annual maps of the quadrats were digitized and the fates of individual plants tracked and extracted using a computer program.
-Daily climate data, which we aggregated into climate variables of interest, are available for the duration of the data collection period (1932 - 1945) from the Miles City airport, Wiley Field, 9 km from the study site.
-
-In this paper, we model populations based on two levels of data: individual and quadrat (Fig.
-2).
-The individual data is the "raw" data.
-For the quadrat level we data we simply sum individual areal cover for each quadrat by species.
-This is equivalent to a perfect census of quadrat percent cover, so we do not need to consider measurement error.
-Based on these two datasets we can compare population models built using individual level data and aggregated quadrat level data.
-
-All R code and data necessary to reproduce our analysis is archived on GitHub as release 1.0 (http://github.com/atredennick/MicroMesoForecast/releases).
-That stable release will remain static as a record of this analysis, but subsequent versions may appear if we update this work.
-
-### Stastical models of vital rates
-At both levels of inference (individual and quadrat), the building blocks of our population models are vital rate regressions.
-For individual level data we fit models for survival, growth, and recruitment of new individuals for each species.
-At the quadrat level we fit a single regression model for population growth.
-We describe the statistical models separately since fitting the models required different approaches.
-All models contain five climate covariate that we chose \emph{a priori}: "water year" precipitation at \emph{t}-1 (lagppt); fall through spring precipitation at \emph{t}-1 and \emph{t}-2 (ppt1 and ppt2, respectively) and mean spring temperature at \emph{t}-1 and \emph{t}-2 (TmeanSpr1 and TmeanSpr2, respectively), where \emph{t} is the observation year.
-We also include interactions among same-year climate covariates (e.g., ppt1 $\times$ TmeansSpr1) and climate $\times$ size interactions.
-Climate $\times$ size interactions are for climate main effects only, that is we do not include interactions among size and interacting climate effects.
-
-We fit all models using a hierarchical Bayesian approach, which we describe in more detail below.
-However, for each vital rate statistical model we also define the likelihood model we use.
-For the likelihood models, \textbf{Y} is always the relevant vector of observations (e.g., whether a  genet survived [1] or not [0] from year $t$ to $t+1$).
-
-#### Vital rate models at the individual level
-We used logistic regression to model survival probability ($S$) of genet $i$ from species $j$ in quadrat group $Q$ from time $t$ to $t+1$:
-
-\begin{align}
-\text{logit}(S_{ijQ,t}) &= \gamma^{S}_{j,t} + \phi^{S}_{jQ} + \beta^{S}_{j,t}x_{ij,t} + \omega^{S}_{j}w_{ij,t} + \nu^{S}_{j}w_{ij,t}x_{ij,t} + \theta^{S}_{jk}C_{k,t} \\
-Y^{S}_{ijQ,t} &\sim \text{Bernoulli}(S_{ijQ,t})
-\end{align}
-
-where $x_{ij,t}$ is the log of genet size, $\gamma^{S}_{j,t}$ is a year-specific intercept, $\beta^{S}_{j,t}$ is the year-specific slope parameter for size, $\phi^{S}_{jQ}$ is the random effect of quadrat group location, and $\theta^{S}_{k}$ is the fixed parameter for the effect of the $k$th climate covariate at time $t$ ($C_{k,t}$).
-Note that the vector of climate covariates (**C**) includes climate variable interactions and climate$\times$size interactions.
-We include density-dependence by estimating the effect of crowding on the focal individual by other individuals of the same species.
-$\omega$ is the effect of crowding and $w_{t,Q}$ is the crowding experienced by the focal individual at time $t$ in quadrat group $Q$.
-We include a size$\times$crowding interaction effect ($\nu^{S}$).
-
-We modeled growth as Gaussian process describing genet size at time $t+1$ as a function of size at $t$ and climate covariates:
-
-\begin{align}
-x_{ijQ,t+1} &= \gamma^{G}_{j,t} + \phi^{G}_{jQ} + \beta^{G}_{j,t}x_{ij,t} + \omega^{G}_{j}w_{ij,t} + \nu^{S}_{j}w_{ij,t}x_{ij,t} + \theta^{G}_{jk}C_{k,t} \\
-Y^{G}_{ijQ,t} &\sim \text{Normal}(x_{ijQ,t+1}, \sigma_{j})
-\end{align}
-
-where $x$ is log genet size and all other parameters are as described for the survival regression.
-
-Our data allows us to track new recruits, but we cannot assign a specific parent to new genets.
-So, for recruitment, we work at the quadrat level and model the number of new individuals of species $j$ in quadrat $q$ recruiting at time $t+1$ as a function of quadrat "effective cover" ($A'$) in the previous year ($t$).
-Effective cover is a mixture of observed cover ($A$) in the focal quadrat ($q$) and the mean cover across the entire group ($\bar{A}$) of $Q$ quadrats in which $q$ is located:
-
-\begin{equation}
-A'_{jq,t} = p_{j}A_{jq,t} + (1-p_{j})\bar{A}_{jQ,t}
-\end{equation}
-
-where $p$ is a mixing fraction between 0 and 1 that is estimated within the model.
-
-We assume the number of individuals, $Y^{R}$, recruiting at time $t+1$ follows a negative binomial distribution:
-
-\begin{equation}
-Y^{R}_{jq,t+1} \sim \text{NegBin}(\lambda_{jq,t+1},\zeta)
-\end{equation}
-
-where $\lambda$ is the mean intensity and $\zeta$ is the size parameter. We define $\lambda$ as:
-
-\begin{equation}
-\lambda_{jq,t+1} = A'_{jq,t}e^{(\gamma^{R}_{j,t} + \phi^{R}_{jQ} + \theta^{R}_{jk}C_{k,t} + \omega^{R}\sqrt{A'_{q,t}})}
-\end{equation}
-
-where $A'$ is effective cover ($\text{cm}^2$) of species $j$ in quadrat $q$ and all other terms are as in the survival and growth regressions.
-
-#### Population model at the quadrat level
-The statistical approach used to model vital rates using aggregated data depends on the type of data collected.
-In our case, and as is often the case with census data, we have percent cover data (which can easily be transformed to proportion data, of course).
-We first considered fitting three vital rate models analagous to those we fit at the individual level: one for probability of extirpation within a quadrat (analagous to survival), one for cover change within a quadrat (analagous to growth), and one for probability of colonization within a quadrat (analagous to recruitment).
-However, within-quadrat extirpation and colonization events were rare in our time series ($N=9$ and $N=10$, respectively across all species).
-Given the broad spatial distribution of the quadrats we are studying, it is safe to assume that these events are in fact rare enough to be ignored for our purposes.
-So we constrained our statistical modeling of vital rates at the population level to change in percent cover within quadrats.
-For the remaining discussion of statistical modeling we refer to proportion data, which is simply percent data divided by 100.
-
-An obvious choice for fitting a linear model to proportion data is beta regression because the support of the beta distribution is [0,1], not including true zeros or ones.
-However, when we used fitted model parameters from a beta regression in a quadrat-based population model the simulated population tended toward 100% cover for all species.
-We therefore chose a more constrained modeling approach based on a truncated log-normal likelihood.
-The model for quadrat cover change ($G$) from time $t$ to $t+1$ is
-
-\begin{align}
-x_{jq,t+1} &= \gamma^{G}_{j,t} + \phi^{G}_{jQ} + \beta^{G}_{j,t}x_{jq,t} + \theta^{S}_{jk}C_{k,t} \\
-Y^{G}_{jq,t+1} &\sim \text{LogNormal}(x_{jq,t+1}, \tau{j}) \text{T}[0,1]
-\end{align}
-
-where $x_{jq,t}$ is the log of species' $j$ proportional cover in quadrat $q$ at time $t$ and all other parameters are as in the individual-level growth model (Eq.
-3).
-Again, note that the climate covariate vector (**C**) includes the climate$\times$cover interaction.
-The log normal likelihood includes a truncation (T[0,1]) to ensure that predicted values do not exceed 100% cover.
-
-### Model fitting
-Our Bayesian approach to fitting the vital rate models required choosing appropriate priors for unknown parameters and deciding which, if any, of those priors should be hierarchical.
-We decided to fit models where all terms were fit by species.
-Within a species, we fit yearly size effects and yearly intercepts hierarchically where year-specific coefficients were drawn from global distributions representing the mean size effect and intercept.
-We used flat, uninformative priors for all unknown parameters (Appendix X).
-
-All of our analyses (model fitting and simulating) were conducted in R [@R2013].
-We used the 'No-U-Turn' MCMC sampler in Stan [@stan2014]  to estimate the posterior distributions of model parameters using the package 'rstan' [@rstan2014].
-We obtained posterior distributions for all model parameters from three parallel MCMC chains run for 1,000 iterations after discarding an initial 1,000 iterations.
-We recignize such short MCMC chains may surprise those more familiar with other MCMC samplers (i.e.
-JAGS or WinBUGS), but the Stan sampler is exceptionally efficient, which reduces the number of iterations needed to achieve convergence.
-We assessed convergence visually and made sure scale reduction factors for all parameters were less than 1.01.
-For the purposes of including parameter uncertainty in our population models, we saved the final 1,000 iterations from each of the three MCMC chains for all parameters to be used as randomly drawn values during population simulation.
-This step alleviates the need to reduce model parameters by model selection since sampling from the full parameter space in the MCMC ensures that if a parameter broadly overlaps zero, on average the effect in the population models will also be near zero.
-
-### Population models
-With the posterior distribution of the vital rate statistical models in hand, it is straightforward to simulate the population models.
-We used an Integral Projection Model (IPM) to model populations based on individual level data and an quadrat based version of an individually-based model (Quadrat-Based Model, QBM) to model populations based on quadrat level data.
-We describe each in turn.
-
-#### Integral projection model
-We use an environmentally stochastic IPM [@Rees2009] that includes the random year effects and the climate covariates from the vital rate statistical models.
-But note that we can, and do for some simulations, ignore the random year effects so that only the climate effects can drive interannual variation.
-Our IPM follows the specification of @Chu2015 where the population of species _j_ is a density function $n(u_{j},t)$ giving the density of sized-_u_ genets at time _t_.
-Genet size is on the natural log scale, so that $n(u_{j},t)du$ is the number of genets whose area (on the arithmetic scale) is between $e^{u_{j}}$ and $e^{u_{j}+du}$.
-So, the density function for any size _v_ at time $t+1$ is
-
-\begin{equation}
-n(v_{j},t+1) = \int_{L_{j}}^{U_{j}} k_{j}(v_{j},u_{j},\bar{\bold{w_{j}}}(u_{j}))n(u_{j},t)
-\end{equation}
-
-where $k_{j}(v_{j},u_{j},\bar{\bold{w_{j}}})$ is the population kernal that describes all possible transitions from size $u$ to $v$ and $\bar{\bold{w_{j}}}$ is a vector of estimates of average crowding experienced from all other species by a genet of size $u_j$ and species $j$.
-The integral is evaluated over all possible sizes between predefined lower (_L_) and upper (_U_) size limits that extend beyond the range of observed genet sizes.
-
-The population kernal is defined as the joint contributions of survival (_S_), growth (_G_), and recruitment (_R_):
-
-\begin{equation}
-k_{j}(v_{j},u_{j},\bar{\bold{w_{j}}}) = S_j(u_j, \bar{\bold{w_{j}}}(u_{j}))G_j(v_{j},u_{j},\bar{\bold{w_{j}}}(u_{j})) + R_j(v_{j},u_{j},\bar{\bold{w_{j}}}),
-\end{equation}
-
-which, said plainly, means we are calculating growth (_G_) for individuals that survive (_S_) from time _t_ to _t+1_ and adding in newly recruited (_R_) individuals of an average sized one-year-old genet for the focal species.
-Our stastical model for recruitment (_R_, described above) returns the number of new recruit produced per quadrat.
-Following previous work [@Adler2012; @Chu2015], we assume that fecundity increases linearly with size ($R_j(v_{j},u_{j},\bar{\bold{w_{j}}}) = e^{u_j}R_j(v_{j},\bar{\bold{w_{j}}})$) to incorporate the recruitment function in the spatially-implicit IPM.
-
-We used random draws from the final 1,000 iterations from each of three MCMC chains to introduce stochasticity into our population models.
-At each time step, we randomly selected climate covariates from one of the 14 observed years.
-Then, we drew the full parameter set (climate effects and density-dependence fixed effects) from a randomly selected MCMC iteration.
-Using this approach, rather than simply using coefficient point estimates, ensures that relatively unimportant climate covariates (those that broadly overlap 0) have little effect on the simulation results.
-Since our focus was on the contribution of climate covariates to population states, we set the random year effects and the random group effects to zero.
-
-#### Quad-based model
-Our quad-based model (QBM) perfectly mirrors its statistical description (Eq. 8-9). We use the same approach for drawing parameter values as described for the IPM.
-
-### Model validation
-To test each model's ability to forecast the population state we made out of sample predictions using leave-one-year-out cross validation.
-For both levels of modeling, we fit the vital rate models using observations from all years except one, and then used those fitted parameters in the population models to perform a one-step-ahead forecast for the year whose observations were withheld from model fitting.
-Within each observation year, several quadrats are sampled.
-So we made predictions for each observed quadrat in the focal year initialized with cover the previous year.
-Since we were making quadrat specific predictions we incorporated the group effect on the intercept for both models.
-We repeated this procedure for all 13 observation years, making 100 one-step-ahead forecasts for each quadrat-year combination with parameter uncertainty included via randomdrawd from the MCMC chain as described above.
-Random year effects were set to zero since year effects cannot be assigned to unobserved years.
-
-This model validation allowed us to compare accuracy and precision of the two modeling approaches (individual-level versus population-level).
-We first calculated the median predicted cover across the 100 simulations for each quadrat-year and then calculated the absolute error as the difference between the observed cover for a given quadrat-year and the median prediction.
-To arrive at mean absolute error (MAE), we then averaged the absolute error within each species across the quadrat-year specific errors.
-We use MAE as our measure of accuracy.
-To measure precision we calculated the distance between the upper and lower 90th quantiles of the 100 predictions and averaged this value over quadrat-years for each species.
-
-### Testing sensitivity to climate covariates
-Our main goal in this paper is to see if models based on aggregate level data are as sensitive to climate as models based on individual level data.
-So, with our fitted and validated models in hand, we ran simulations for each model type (IPM and QBM) under four climate perturbation scenarios: (1) observed climate, (2) precipitation increased by 1%, (3) temperature increased by 1%, and (4) precipitation and temperature increased by 1%.
-We ran the simulations for 2,500 time steps, enough to estimate equilibrium cover after discarding an initial 500 time steps as burn-in.
-Each simulation was run under two parameter scenarios: (1) using mean parameter estimates and (2) using randomly drawn parameters from the MCMC chain.
-We use (1) to detect the overall sensitivity of equilibrium cover to climate, and we use (2) to show the impact of model uncertainty on forecast precision.
-
-As an effort to identify potential discrepencies between IPM and QBM forecasts, we also ran simulations designed to quantify the sensitivities of individual and combined vital rates to climate for the IPM.
-Specifically, we ran simulations for the above climate scenarios, but applied the perturbed climate covariates to survival, growth, and recruitment vital rates individually and in pairwise combinations.
-This allows us to isolate the vital rate(s) most sensitive to climate.
-For this analysis, we used mean parameter estimates to reduce the sources of uncertainty in the sensitivity estimates.
-
-Results
--------
-### Comparison of forecast models
-The IPM had significantly lower overall error (MAE, mean absolute error) for three species (*B.
-gracilis*, *H.
-comata*, *P.
-smithii*; Table 1).
-In no case did the QBM significantly outperform the IPM (Table 1).
-The IPM was consistently more precise than the QBM, with lower distances between the 90% quantiles across all species (Table 1).
-In general the IPM outperformed the QBM because it had (1) lower MAE for three of the four species, (2) statistically similar MAE for the one other species, and (3) considerably more precise forecasts for all species.
-
-### Sensitivity of models to climate
-```{r get_results, echo=FALSE, eval=TRUE}
-setwd("~/Repos/MicroMesoForecast/analysis/")
-
-####
-##  Mean parameter runs
-####
-meanparam_ipm <- data.frame(cover=NA, species=NA, 
-                             climsim=NA, model=NA)
-##  IPM 
-path <- "./ipm/simulations/results/climchange_meanparams/"
-ipmfiles_mean <- list.files(path)
-ipmfiles_mean <- ipmfiles_mean[grep("cover", ipmfiles_mean)]
-for(file in ipmfiles_mean){
-  tmp <- read.csv(paste(path,file,sep=""))
-  tmp$model <- "IPM"
-  meanparam_ipm <- rbind(meanparam_ipm, tmp[,c("cover", "species", "climsim", "model")])
-}
-meanparam_ipm <- meanparam_ipm[2:nrow(meanparam_ipm),]
-
-newd <- data.frame(cover=NA, species=NA, 
-                   climsim=NA, model=NA, obscover=NA)
-for(dospp in unique(meanparam_ipm$species)){
-  tmp <- subset(meanparam_ipm, species==dospp)
-  obscover <- subset(tmp, climsim=="obs")["cover"]
-  tmp2 <- subset(tmp, climsim!="obs")
-  tmp2$obscover <- rep(obscover$cover, times = length(unique(tmp2$climsim)))
-  newd <- rbind(newd, tmp2)
-}
-ipm_meansims <- newd[2:nrow(newd),]
-ipm_meansims$propdiff <- with(ipm_meansims, log(cover*100)-log(obscover*100))
-
-##  QBM
-meanparam_qbm <- data.frame(cover=NA, species=NA, 
-                             climsim=NA, model=NA)
-path <- "./quadBM/simulations/results/climatechange_meanparams/"
-qbmfiles_mean <- list.files(path)
-for(file in qbmfiles_mean){
-  tmp <- readRDS(paste(path,file,sep=""))
-  tmp$model <- "QBM"
-  meanparam_qbm <- rbind(meanparam_qbm, tmp[,c("cover", "species", "climsim", "model")])
-}
-meanparam_qbm <- meanparam_qbm[2:nrow(meanparam_qbm),]
-
-newd <- data.frame(cover=NA, species=NA, 
-                   climsim=NA, model=NA, obscover=NA)
-for(dospp in unique(meanparam_qbm$species)){
-  tmp <- subset(meanparam_qbm, species==dospp)
-  obscover <- subset(tmp, climsim=="obs")["cover"]
-  tmp2 <- subset(tmp, climsim!="obs")
-  tmp2$obscover <- rep(obscover$cover, times = length(unique(tmp2$climsim)))
-  newd <- rbind(newd, tmp2)
-}
-qbm_meansims <- newd[2:nrow(newd),]
-qbm_meansims$propdiff <- with(qbm_meansims, log(cover*100)-log(obscover*100))
-
-##  Merge IPM and QBM mean parameter results
-bothmodel_outs <- rbind(ipm_meansims, qbm_meansims)
-meanparam_agg <- ddply(bothmodel_outs, .(species, climsim, model), summarise,
-                       meandiff = mean(propdiff),
-                       hidiff = quantile(propdiff, probs = 0.95),
-                       lodiff = quantile(propdiff, probs = 0.05))
-setwd("~/Repos/MicroMesoForecast/manuscript")
-```
-
-Equilibrium cover from both models was sensitive to climate (Fig.
-3a-d).
-The IPM projected percent changes in equilibrium cover from `r round(min(subset(meanparam_agg, species=="BOGR" & model=="IPM")["meandiff"])*100)` - `r round(max(subset(meanparam_agg, species=="BOGR" & model=="IPM")["meandiff"])*100)`% for *B.
-gracilis*, `r round(min(subset(meanparam_agg, species=="HECO" & model=="IPM")["meandiff"])*100)` - `r round(max(subset(meanparam_agg, species=="HECO" & model=="IPM")["meandiff"])*100)`% for *H.
-comata*, `r round(min(subset(meanparam_agg, species=="PASM" & model=="IPM")["meandiff"])*100)` - `r round(max(subset(meanparam_agg, species=="PASM" & model=="IPM")["meandiff"])*100)`% for *P.
-smithii*, and `r round(min(subset(meanparam_agg, species=="POSE" & model=="IPM")["meandiff"])*100)` - `r round(max(subset(meanparam_agg, species=="POSE" & model=="IPM")["meandiff"])*100)`% for *P.
-secunda*.
-The QBM projected opposite and greater percent changes in equilibrium cover for *B.
-gracilis* (`r round(min(subset(meanparam_agg, species=="BOGR" & model=="QBM")["meandiff"])*100)` - `r round(max(subset(meanparam_agg, species=="BOGR" & model=="QBM")["meandiff"])*100)`%) and *H.
-comata* (`r round(min(subset(meanparam_agg, species=="HECO" & model=="QBM")["meandiff"])*100)` - `r round(max(subset(meanparam_agg, species=="HECO" & model=="QBM")["meandiff"])*100)`%; Fig.
-3a-b).
-For *P.
-smithii*, the QBM projected opposite changes in equilibrium cover than the IPM, but of similar magnitude (`r round(min(subset(meanparam_agg, species=="PASM" & model=="QBM")["meandiff"])*100)` - `r round(max(subset(meanparam_agg, species=="PASM" & model=="QBM")["meandiff"])*100)`%; Fig.
-3c).
-*P.
-secunda* was the only species that the IPM and QBM made projections of the same sign and somewhat similar magnitude (Fig.
-3d).
-
-The response of a population to climate change is a result of the aggregate effects of climate on individual vital rates.
-Since the IPM approach relies on vital rate regressions, we were able to quantify the sensitivity of each vital rate in isolation and in pairwise combinations.
-Species showed similar trends (Fig.
-3e-h).
-Growth was the most sensitive vital rate for all species, showing a negative response to increased precipitation, and stronger positive response to increased temperature, and a mostly positive response when both climate factors are increased (Fig.
-3e-h).
-*B.
-gracilis* survival rates were sensitive to temperature, resulting in an increase in plant cover under increased temperature (Fig.
-3e).
-In isolation, recruitment and survival were insensitive to climate factors for *H.
-comata* (Fig.
-3f).
-Survival and recruitment of *P.
-smithii* were both sensitive, negatively, to temperature and precipitation (Fig.
-3g).
-*P.
-secunda* equilibrium cover was sensitive to the climate effects on survival and recruitment, showing a negative effect on both vital rates for increased precipition, but a strong positive effect on survival with increased temperature (Fig.
-3h).
-The climate impact of recruitment on equilibrium cover was negative for precipitation and temperature increases (Fig.
-3h).
-
-Forecasts based on 1% climate changes were extremely uncertain when we considered model error and parameter uncertainty (Fig.
-4).
-As expected based on model validation (Table 1), QBM projections were more uncertain than IPM projections for all species except *P.
-smithiii* (Fig.
-4).
-
-Discussion
-----------
-Perhaps the greatest challenge for ecology in the 21st century is to forecast the impacts of environmental change [@Clark2001; @Petchey2015].
-To do so requires sophisticated modeling approaches that fully account for uncertainty and variability in the ecological process and associated parameters [@Luo2011a].
-This requires large amounts of data collected over large spatio-temporal extents.
-State-of-the-science modeling techniques cannot overcome data limitations.
-Such is the case for many population models.
-
-As a potential remedy to the "data dearth" problem, @Queenborough2011 and @Freckleton2011, building on work by @Taylor2004, advocate a "density-structured" modeling approach.
-Such models do not require individual level demographic data and can adequately describe population dynamics [@Queenborough2011].
-The results from density-structured models are not as precise as those from traditional population models, but the loss in precision is traded off with a gain in data.
-The study by @Queenborough2011 included data from 500 fields (4 hectares each) in 49 farms, all collected by two people in 6 weeks.
-This is far more data from a far greater spatial extent than possible if measuring individual plant demography (in a world of limited time and money, at least).
-The appeal of density-structured approaches is clear.
-
-However, at their core, density-structured models rely on individual level data aggregated to a population level metric (e.g., density classes or percent cover).
-This creates a potential problem if such models are to be used in a climate change context because inidividuals respond to climate, not populations [@Clark2012].
-Are models based on population level metrics as sensitive to climate as models based on individual level metrics?
-Do these two types of models produce consistent forecasts?
-Do we need detailed demographic data to forecast the impacts of climate change?
-These are the questions we sought to answer here.
-
-### The IPM and QBM produced inconsistent forecasts
-Using individual and aggregated forms of the same dataset, we were able to directly compare a traditional demographic modeling approach to a population model based on aggregated data.
-Our quad-based model (QBM) is based on percent cover data and so is in the spirit of density-structured models.
-In terms of each model's forecasting ability, the IPM outperformed the QBM (Table 1).
-This is unsurprising since we expected to lose some precision at the aggregated level.
-However, the underwhelming performance of the QBM could call into the question forecasts that differ from the IPM.
-
-Indeed, when we perturbed climate factors the QBM made forecasts completely contradictory to those of the IPM for three of our four study species (Fig.
-3a-d).
-In a perfect world, the QBM would have made forecasts of at least the same direction as the IPM.
-If that had been the case we could conclude that aggregate level models could prove useful for forecasting climate change impacts on populations.
-Unfortunately, this was not the case.
-
-Given the superior ability of the IPM to predict out of sample observations (Table 1), we have no choice but to conclude it is the superior model.
-Following that logic, we can only assume that, at least contingent on the data in hand, the IPM is producing the correct forecasts to climate perturbations.
-The QBM failed to match IPM forecasts, implying that detailed demographic data may be necessary to accurately detect climate signals that are utlimately important at the population level.
-This result further confirms related work on the importance of individual variability on population level responses to exogenous drivers [@Clark2011; @Clark2011b; @Clark2012; @Galvan2014].
-
-### The role of vital rate climate dependence
-We can think of two reasons why the IPM outperformed the QBM.
-First, the quadrat level data has a much reduced sample size compared to the individual level data.
-In an ideal world we would have compared the IPM and QBM using data collected over the same amount of person hours, not just the same number of quadrats.
-Then the sample size of the quadrat level data would be much greater and carry more statistical power.
-To address this limitation in our work we fit the QBM statistical model (Eq.
-8-9) with different numbers of quadrats to see the effect of sample size on the precision of climate effect estimates.
-It appears that including additional quadrats leads to rapidly diminishing returns in terms of parameter precision (Fig.
-5).
-Thus, while sample size surely plays some role, we do not think it is the main driver of the difference between the IPM and QBM.
-
-The second reason the IPM could have outperformed the QBM is that the population level model is in fact missing important climate effects that act on individual vital rates, rather than population growth.
-Our intuition was that species with strong climate-dependence on vital rates not well resolved at the aggregate level would result in different forecasts from the two models.
-For example, survival is very size dependent: smaller individuals have a higher probability of death [@Chu2014].
-At the same time, a single small individual contributes relatively little to percent cover estimates at the plot scale.
-So, if survival of individuals was positively impacted by temperature increases, for example, we would expect to detect this signal in the individual level data but not in the aggregate level data.
-To see if this is the case we can regress climate effects from each vital rate statistical model at the individual level against the same climate effects from the QBM statistical model (Fig.
-6).
-
-In general, the QBM climate effects are most correlated with climate effects from the growth model at the individual level (Fig.
-6).
-In no case does the QBM statistical model have strong correlations across all three vital rates (Fig.
-6).
-Thus, for each species the QBM is "missing" climate signals associated with at least one vital rate.
-This has large impacts on predictions of long term population dynamics, as seen in our equilibrium simulations (Fig.
-3a-d).
-
-These results lead us to conclude that detailed demographic data is necessary to forecast climate change impacts on plant populations.
-This is unwelcome news since this data is difficult to collect and the models built on such data are of little use to land managers that make decisions at scales beyond that of traditional population models [@Queenborough2011].
-While density-structured approaches may fail when climate covariates are considered, there are other alternatives.
-For example, @Clark2011a use Forest Inventory and Analysis (FIA) data to parameterize a population model with multiple vital rates and climate dependence.
-Another example are distributed efforts like PlantPopNet (http://plantago.plantpopnet.com) that will allow researchers to estimate variation around climate responses for widespread species.
-Lastly, we foresee new approaches on the horizon that leverage photo/video of plots and advanced object recognition algorithms to streamline plant mapping and digitizing efforts.
-
-### Forecasting the future, and the future of forecasting
-Our goal was not make any explicit forecast for the future state of these populations based on predicted climate change.
-But our results highlight the state of affairs in ecology when it comes to forecasting the impacts of climate change.
-The analysis we conducted here could be considered, with some exceptions of course, at the forefront of ecological forecasting in terms of the statistical approach employed (hierarchical Bayesian), the type of population model we used (stochastic IPM with parameter uncertainty), and the amount of data we had at our disposal (14 years of individual-level data).
-Yet, model predictions proved so uncertain that any forecast, when bounded with model and parameter uncertainty, would be at best not useful and at worst meaningless.
-How might we improve on this state of affairs?
-
-First, forecasts could be improved by matching the spatial scale of predictor variables with the spatial scale of observations.
-One of the major limitations of the models we fit here is that the climate data are at a much larger scale than the plot and individual level observations of plant cover and size.
-Climate covariates only vary by year, with no spatial variability within years.
-Thus, even if we fit models to individual level data, we are missing the key interaction point between weather and individual plants [@Clark2011b].
-
-Second, accurately detecting climate signals will take even longer time series.
-Recent theoretical work on detecting climate signals in noisy data suggests that even advanced approaches to parameter fitting like LASSO, functional linear models (splines), and random forest require 20-25 year time series (Teller et al., in review).
-Alternatively, as we suggest above, Teller et al.
-(in review) also find that matching the scale of the response and predictors improves estimate precision.
-
-Third, ecologists as a community need to get serious about reporting uncertainty.
-There is a strong culture around explicitly considering model uncertainty, but parameter uncertainty is often ignored.
-In some cases this is because the easiest statistical methods do no make propogating parameter uncertainty easy.
-Even using Bayesian approaches that allow straightforward integration of model fitting and forecasting [@Hobbs2015] is not simple when using modeling approaches like integral projection models that separate the model fitting and simulation stages [@Rees2009].
-However, as we have done here, it is still possible to include parameter uncertainty by drawing parameter values from MCMC iterations, taking care to draw all parameters from the same chain and iteration to account for their correlations.
-Only by being honest about our forecasts can we begin to produce better ones.
-
-Conclusions
------------
-This work is not a critique of density-structured population models.
-In some cases and for certain species, population models based on aggregated data may prove useful and unbiased.
-However, our work here is the first comparison, to our knowledge, of population models based on individual and aggregated data in a climate change context.
-Our results confirm theoretical arguments [@Clark2011b] and empirical evidence [@Clark2011; @Clark2012] that individual responses are critical to predicting species' responses to climate change.
-Thus, forecasts from aggregate level models should be viewed with caution and should never be unaccompianed by uncertainty.
-Given the importance of demographic data and its current difficulty to collect, we advocate for research on modern methods to collect demographic data more efficiently across environmental gradients in space and time.
-For example, recent advances in object recognition algorithms [@Liu2014] provide new opportunities to speed up data collection and digital processing.
-
-Our results also offer a cautionary tale because uncertainty around forecasts was large for both model types.
-Which leads us to our most pessimistic conclusion: even with 14 years of detailed demographic data and sophisticated modeling techniques we failed to produce forecasts with any level of acceptable uncertainty.
-In our view, uncertainty of climate change related forecasts can be reduced by (1) longer time series and (2) climate covariates that match the scale of inference (e.g., plot rather than landscape level climate/weather metrics).
-Still, given the poor performance of the quad-based model, it seems there is no short cut to producing accurate and precise population forecasts.
-Do we need detailed demographic data to forecast the impacts of climate change on populations?
-Probably.
-
-Acknowledgments
----------------
-This work was funded by the National Science Foundation through a Postdoctoral Research Fellowship in Biology to ATT (DBI-1400370) and a CAREER award to PBA (DEB-1054040).
-We thank the original mappers of the permanent quadrats in Montana and the digitizers in the Adler lab, without whom this work would not have been possible.
-Informal conversations with Stephen Ellner, Giles Hooker, Robin Snyder, and a series of meetings among the Adler and Weecology labs at USU sharpened our thinking.
-Compute, storage and other resources from the Division of Research Computing in the Office of Research and Graduate Studies at Utah State University are gratefully acknowledged.
-
-\pagebreak{}
-
-Tables
-------
-```{r table_1,results='asis',message=FALSE, comment=NA, echo=FALSE}
+```{r, echo=FALSE, include=TRUE, warning=FALSE, message=FALSE, fig.height=4}
 ####
@@
                               "quant_dist","mean_cover")]
+colnames(stats_table)[3:4] <- c("mean absolute error", "quantile distance")
+plotd <- melt(stats_table, id.vars = c("species", "model"))
 
-colnames(stats_table) <- c("Species", "Model", "MAE", "90% Distance", "Mean Obs. Cover")
-print.xtable(xtable(stats_table, caption = "Accuracy (mean absolute error, MAE) and precision (90\\% Distance) of out of sample predictions.
-Forecasts were made without random year effects; only climate covariates could explain year-to-year variation.
-90\\% Distance refers to the average distance between the upper and lower 90th percentiles of the 100 predicted values for each quadrat-year combination."),type="latex", comment=FALSE,
-             include.rownames=FALSE, caption.placement="top")
+ggplot(subset(plotd, variable!="mean_cover"), aes(x=species, y=value, fill=model))+
+  geom_bar(stat="identity", position="dodge")+
+  facet_wrap("variable", nrow=1, scales="free")+
+  scale_fill_manual(values = c("darkorange", "steelblue"))
 ```
-*NOTES*: The IPM MAE is significantly lower at $\alpha=0.05$ for *B.
-gracilis* (*P* = `r mae_ps[1,"value"]`), *H.
-comata* (*P* = `r mae_ps[2,"value"]`), and *P.
-smithii* (*P* = `r mae_ps[3,"value"]`).
-MAEs are statisticially similar between models for *P.
-secunda* (*P* = `r mae_ps[4,"value"]`).
-*P* values are highly sensitive to sample size, so not entirely appropriate in simulation exercises where we control the samples size.
-But, for our purposes they serve as relatively unbiased comparison metrics.
-
-\pagebreak{}
-```{r table_2,results='asis',message=FALSE, comment=NA, echo=FALSE}
-# qbm_diffs <- readRDS("../montana/quadBM/simulations/results/qbm_climatesims_logdiffs.RDS")
-# qbm_diffs$type <- rep("QBM", nrow(qbm_diffs))
-# ipm_diffs <- readRDS("../montana/ipm/simulations/results/ipm_climatesims_logdiffs.RDS")
-# ipm_diffs$type <- rep("IPM", nrow(ipm_diffs))
-# 
-# diff_df <- rbind(qbm_diffs, ipm_diffs)
-# med_cover_perc <- as.data.frame(diff_df$med_cover*100)
-# med_cover_perc$model <- diff_df$type 
-# med_cover_perc$sim <- diff_df$variable
-# med_cover_perc$species <- diff_df$species
-# colnames(med_cover_perc)[1] <- "cover"
-# medcast <- dcast(med_cover_perc, species+sim~model, value.var = "cover")
-# colnames(medcast) <- c("Species", "Simulation", "IPM Change (%)", "QBM Change (%)")
-# simnames <- c("ppt", "temp", "ppt+temp")
-# medcast$Simulation <- rep(simnames, 4)
-# print.xtable(xtable(medcast, caption = "Percent change from equilibrium (temporal median of simulated plant cover) at observed climate relative to a 1\\% climate change as forecast by each model type."),type="latex", comment=FALSE,
-#              include.rownames=FALSE, caption.placement="top")
-```
-
-\pagebreak{}
-
-Figures
--------
-```{r figure_1, dependson="plot-options", fig.cap="Time series of average percent cover over all quadrats for our four focal species: *Bouteloua gracilis* (BOGR), *Hesperostipa comata* (HECO), *Pascopyrum smithii* (PASM), and *Poa secunda* (POSE).
-Light grey lines show trajectories of individual quadrats.
-Note the different y-axis scales across panels.", fig.height=3, fig.width=8.5}
-obs_data <- read.csv("../analysis/speciesData/quadAllCover.csv")
-#remove zeros for averaging
-obs_data <- obs_data[which(obs_data$propCover>0),]
-avgobs <- ddply(obs_data, .(year, Species), summarise,
-                mean_cover = mean(propCover))
-
-ggplot()+
-  geom_line(data = obs_data, aes(x=(1900+year), y=propCover*100, group=quad), 
-            color="grey", size=0.25)+
-  geom_line(data=avgobs, aes(x=(1900+year), y=mean_cover*100))+
-  geom_point(data=avgobs, aes(x=(1900+year), y=mean_cover*100),size=3)+
-  ylab("Mean cover (%)")+
-  xlab("Year")+
-  theme_bw()+
-  facet_wrap("Species", ncol=4, nrow=1, scales = "free")+
-  guides(shape=FALSE, linetype=FALSE)
-```
-
-```{r figure_2,dependson="plot-options", fig.cap="Work flow of the data aggregation, model fitting, and population simulating.", fig.height=6}
-library(png)
-library(grid)
-img <- readPNG("components/figure/micromeso_flowchart.png")
-grid.raster(img)
-```
-
-```{r ipm_climeffs, cache=FALSE, results='hide', include=FALSE}
-# source("../montana/ipm/vitalRateRegs/climate_effects_plots.R")
-# ![Standardized regression coefficients for climate effects in the vital rate models for the IPM.
-Shown are the median (circles), 75% confidence intervals (thick lines), and 95% confidence intervals (thin lines).
-Left panels (black) are for the growth regressions; middle panels (blue) are for the survival regressions; right panels (orange) are for the recruitment regressions.](components/figure/ipm_climeffs.pdf)
-```
-
-```{r quad_climeffs, cache=FALSE, results='hide', include=FALSE}
-# source("../montana/quadBM/vitalRateRegressions/truncNormModel/climate_effects_plot.R")
-# ![Standardized regression coefficients for climate effects in the vital rate models for the QBM.
-Shown are the median (circles), 75% confidence intervals (thick lines), and 95% confidence intervals (thin lines).](components/figure/qbm_climeffs.pdf)
-```
-
-```{r figure_3, dependson="plot-options", fig.cap="Proportional change in species' mean cover caused by a 1% increase in observed precipitation (+ppt), temperature (+temp), or both (+ppt&temp) as predicted by the individual-based IPM and the aggregate-based QBM using mean parameter values.
-Top panels show the mean predicted proportional change in cover.
-Lower panels show the sensitivity of equilibrium cover simulated from the IPM to each climate scenario applied to individual and combined vital rates.
-For example, the points associated with G show the median cover from IPM simulations where a climate perturbation is applied only to the growth regression climate covariates.
-These simulations also use mean parameter values for clarity.", fig.width=8.5, fig.height=5, cache=FALSE, results='hide'}
-setwd("~/Repos/MicroMesoForecast/analysis/")
-
-####
-##  Mean parameter runs
-####
-meanparam_ipm <- data.frame(cover=NA, species=NA, 
-                             climsim=NA, model=NA)
-##  IPM 
-path <- "./ipm/simulations/results/climchange_meanparams/"
-ipmfiles_mean <- list.files(path)
-ipmfiles_mean <- ipmfiles_mean[grep("cover", ipmfiles_mean)]
-for(file in ipmfiles_mean){
-  tmp <- read.csv(paste(path,file,sep=""))
-  tmp$model <- "IPM"
-  meanparam_ipm <- rbind(meanparam_ipm, tmp[,c("cover", "species", "climsim", "model")])
-}
-meanparam_ipm <- meanparam_ipm[2:nrow(meanparam_ipm),]
-
-newd <- data.frame(cover=NA, species=NA, 
-                   climsim=NA, model=NA, obscover=NA)
-for(dospp in unique(meanparam_ipm$species)){
-  tmp <- subset(meanparam_ipm, species==dospp)
-  obscover <- subset(tmp, climsim=="obs")["cover"]
-  tmp2 <- subset(tmp, climsim!="obs")
-  tmp2$obscover <- rep(obscover$cover, times = length(unique(tmp2$climsim)))
-  newd <- rbind(newd, tmp2)
-}
-ipm_meansims <- newd[2:nrow(newd),]
-ipm_meansims$propdiff <- with(ipm_meansims, log(cover*100)-log(obscover*100))
-
-##  QBM
-meanparam_qbm <- data.frame(cover=NA, species=NA, 
-                             climsim=NA, model=NA)
-path <- "./quadBM/simulations/results/climatechange_meanparams/"
-qbmfiles_mean <- list.files(path)
-for(file in qbmfiles_mean){
-  tmp <- readRDS(paste(path,file,sep=""))
-  tmp$model <- "QBM"
-  meanparam_qbm <- rbind(meanparam_qbm, tmp[,c("cover", "species", "climsim", "model")])
-}
-meanparam_qbm <- meanparam_qbm[2:nrow(meanparam_qbm),]
-
-newd <- data.frame(cover=NA, species=NA, 
-                   climsim=NA, model=NA, obscover=NA)
-for(dospp in unique(meanparam_qbm$species)){
-  tmp <- subset(meanparam_qbm, species==dospp)
-  obscover <- subset(tmp, climsim=="obs")["cover"]
-  tmp2 <- subset(tmp, climsim!="obs")
-  tmp2$obscover <- rep(obscover$cover, times = length(unique(tmp2$climsim)))
-  newd <- rbind(newd, tmp2)
-}
-qbm_meansims <- newd[2:nrow(newd),]
-qbm_meansims$propdiff <- with(qbm_meansims, log(cover*100)-log(obscover*100))
-
-##  Merge IPM and QBM mean parameter results
-bothmodel_outs <- rbind(ipm_meansims, qbm_meansims)
-meanparam_agg <- ddply(bothmodel_outs, .(species, climsim, model), summarise,
-                       meandiff = mean(propdiff),
-                       hidiff = quantile(propdiff, probs = 0.95),
-                       lodiff = quantile(propdiff, probs = 0.05))
-dgd = position_dodge(width = 0.9)
-meanplot <- ggplot(meanparam_agg)+
-  geom_bar(aes(x=climsim, y=meandiff, fill=model),
-             position=dgd, stat="identity")+
-  theme_bw()+
-  scale_fill_manual(values=c("black", "grey45"), name="Model")+
-  facet_wrap("species", scales="free", nrow=1)+
-  scale_x_discrete(labels=c("+ppt", "+temp", "+ppt&temp"))+
-  ylab("Mean proportional difference")+
-  xlab("Climate simulation")+
-  theme(axis.text.x = element_text(angle = 45, hjust = 1))
-
-####
-####  Make IPM sensitivity plot for lower panels
-####
-setwd("../analysis/ipm/simulations/results/climate_sensitivity/")
-files <- list.files()
-cover_files <- files[grep("cover", files)]
-spp_id <- substr(cover_files, 1, 4)
-num_files <- length(cover_files)
-all_sims <- data.frame(time=NA, cover=NA, species=NA, climsim=NA, vital=NA)
-for(i in 1:num_files){
-  tmp <- read.csv(cover_files[i])
-  tmp$species <- rep(spp_id[i], nrow(tmp))
-  all_sims <- rbind(all_sims, tmp)
-}
-all_sims <- all_sims[2:nrow(all_sims),]
-
-# Rename vital rates with codes
-vital_names <- unique(all_sims$vital)
-vital_codes <- c("A", "G", "GR", "GS", "R", "S" , "SR") 
-vitals <- data.frame(names=vital_names,code=vital_codes)
-for(vnow in vital_names){
-  ids <- which(all_sims[,"vital"]==vnow)
-  idcode <- which(vitals[,"names"]==vnow)
-  all_sims[ids,"vital"] <- as.character(vitals[idcode,"code"])
-}
-
-# Rep the 'all' simulation for each vital rate for plotting
-all_clim_sims <- subset(all_sims, vital=="A")
-species <- unique(all_sims$species)
-out_controls <- data.frame(time=NA, cover=NA, species=NA, climsim=NA, vital=NA)
-for(spp in species){
-  tmp <- subset(all_clim_sims, species==spp)
-  for(vital in vital_codes){
-    tmp$vital <- vital
-    out_controls <- rbind(out_controls, tmp)
-  }
-}
-out_controls <- out_controls[2:nrow(out_controls),]
-
-# Combine the datasets
-all_noclim_sims <- subset(all_sims, vital!="all")
-final <- rbind(out_controls, all_noclim_sims)
-
-# Calculate statistics for plotting
-equilibrium_cover <- ddply(final, .(species, climsim, vital), summarise,
-                           eq_cover = median(cover))
-mean_cover <- ddply(subset(equilibrium_cover, climsim=="all"), .(species), summarise,
-                    value = mean(eq_cover))
-
-equilibrium_cover <- subset(equilibrium_cover, vital!="A")
-bothid <- which(equilibrium_cover[,"climsim"]=="ppttemp")
-equilibrium_cover[bothid,"climsim"] <- "zppttemp"
-# myCols2 <- c("#277BA8", "#7ABBBD", "#AED77A")
-sensplot <- ggplot(equilibrium_cover, aes(x=climsim, y=eq_cover*100, 
-                              shape=vital, group=vital, linetype=vital))+
-  geom_hline(data=mean_cover, aes(yintercept=value*100), linetype=2, color="grey45")+
-  geom_line()+
-  geom_point(size=4)+
-  facet_wrap("species", scales = "free", nrow=1)+
-  scale_linetype_discrete(name="Vital Rate")+
-  scale_shape_discrete(name="Vital Rate")+
-  scale_x_discrete(labels=c("baseline", "+ppt", "+temp", "+ppt&temp"))+
-  ylab("Equilibrium Cover (%)")+
-  xlab("Simulation")+
-  theme_bw()+
-  theme(axis.text.x = element_text(angle = 45, hjust = 1))
-
-covertest <- subset(equilibrium_cover, climsim=="all")
-colnames(covertest)[4] <- "obs_cover"
-out <- merge(subset(equilibrium_cover, climsim!="all"), covertest[,c("species","vital","obs_cover")])
-out$propdiff <- with(out, log(eq_cover*100)-log(obs_cover*100))
-
-meanparamplot <- grid.arrange(meanplot, sensplot, ncol=1, nrow=2)
-print(meanparamplot)
-
-setwd("~/Repos/MicroMesoForecast/manuscript")
-```
-
-```{r figure_4, dependson="plot-options", fig.cap="Equilibrium cover and 90% quantiles around the mean prediction when model error and parameter uncertainty are propogated through the simulation phase.
-Climate simulations are as in Figure 3.", fig.width=4.5, fig.height=7, cache=FALSE, results='hide'}
-####
-##  Vary parameter runs
-####
-setwd("~/Repos/MicroMesoForecast/analysis/")
-meanparam_ipm <- data.frame(cover=NA, species=NA, 
-                             climsim=NA, model=NA)
-##  IPM 
-path <- "./ipm/simulations/results/climchange_varyparams/"
-ipmfiles_mean <- list.files(path)
-ipmfiles_mean <- ipmfiles_mean[grep("cover", ipmfiles_mean)]
-for(file in ipmfiles_mean){
-  tmp <- read.csv(paste(path,file,sep=""))
-  tmp$model <- "IPM"
-  meanparam_ipm <- rbind(meanparam_ipm, tmp[,c("cover", "species", "climsim", "model")])
-}
-meanparam_ipm <- meanparam_ipm[2:nrow(meanparam_ipm),]
-
-newd <- data.frame(cover=NA, species=NA, 
-                   climsim=NA, model=NA, obscover=NA)
-for(dospp in unique(meanparam_ipm$species)){
-  tmp <- subset(meanparam_ipm, species==dospp)
-  obscover <- subset(tmp, climsim=="obs")["cover"]
-  tmp2 <- subset(tmp, climsim!="obs")
-  tmp2$obscover <- rep(obscover$cover, times = length(unique(tmp2$climsim)))
-  newd <- rbind(newd, tmp2)
-}
-ipm_meansims <- newd[2:nrow(newd),]
-ipm_meansims$propdiff <- with(ipm_meansims, log(cover*100)-log(obscover*100))
-# ggplot(subset(ipm_meansims, species=="PASM"))+
-#   geom_histogram(aes(x=cover))+
-#   facet_wrap("climsim", scales = "free")+
-#   scale_x_continuous(limits=c(0,0.05))
-##  QBM
-meanparam_qbm <- data.frame(cover=NA, species=NA, 
-                             climsim=NA, model=NA)
-path <- "./quadBM/simulations/results/climatechange_varyparams/"
-qbmfiles_mean <- list.files(path)
-for(file in qbmfiles_mean){
-  tmp <- readRDS(paste(path,file,sep=""))
-  tmp$model <- "QBM"
-  meanparam_qbm <- rbind(meanparam_qbm, tmp[,c("cover", "species", "climsim", "model")])
-}
-meanparam_qbm <- meanparam_qbm[2:nrow(meanparam_qbm),]
-
-newd <- data.frame(cover=NA, species=NA, 
-                   climsim=NA, model=NA, obscover=NA)
-for(dospp in unique(meanparam_qbm$species)){
-  tmp <- subset(meanparam_qbm, species==dospp)
-  obscover <- subset(tmp, climsim=="obs")["cover"]
-  tmp2 <- subset(tmp, climsim!="obs")
-  tmp2$obscover <- rep(obscover$cover, times = length(unique(tmp2$climsim)))
-  newd <- rbind(newd, tmp2)
-}
-qbm_meansims <- newd[2:nrow(newd),]
-qbm_meansims$propdiff <- with(qbm_meansims, log(cover*100)-log(obscover*100))
-
-##  Merge IPM and QBM mean parameter results
-bothmodel_outs <- rbind(ipm_meansims, qbm_meansims)
-
-meanparam_agg <- ddply(bothmodel_outs, .(species, climsim, model), summarise,
-                       meandiff = mean(propdiff),
-                       hidiff = quantile(propdiff, probs = 0.95),
-                       lodiff = quantile(propdiff, probs = 0.05))
-dgd = position_dodge(width = 0.6)
-ggplot(meanparam_agg)+
-  geom_errorbar(aes(x=climsim, ymax=hidiff, ymin=lodiff, group=model, linetype=model),
-                position=dgd, width=0.25)+
-  geom_point(aes(x=climsim, y=meandiff, group=model, shape=model),
-                position=dgd, width=0.25)+
-  theme_bw()+
-  scale_linetype_discrete(name="Model")+
-  scale_shape_discrete(name="Model")+
-  facet_wrap("species", ncol=1, scales="free")+
-  scale_x_discrete(labels=c("+ppt", "+temp", "+ppt&temp"))+
-  ylab("Proportional difference")+
-  xlab("Climate simulation")
-#   theme(axis.text.x = element_text(angle = 45, hjust = 1))
-setwd("~/Repos/MicroMesoForecast/manuscript/")
-```
-
-```{r figure_5, dependson="plot-options", fig.cap="Effect of quadrat sample size on the precision (standard deviation) of main climate effect estimates in the QBM.
-Increasing the number of quadrats results in diminishing returns in terms of parameter certainty.
-Light dashed lines show individual climate effects at five quadrat sample sizes.
-Thick dark lines are inverse gaussian fits showing the mean effect of increasing quadrat sample size on parameter precision.", fig.width=3, fig.height=7.5, cache=FALSE, results='hide'}
-library(mgcv)
-
-setwd("../analysis/quadBM/vitalRateRegressions/exclude_quad_test")
-
-####
-####  Read in model fits and summarize results
-####
-source("popgrowth_read_data.R")
-spp <- unique(growD_all$Species)
-num_quads <- numeric(length(spp))
-for(i in 1:length(spp)){
-  tmp <- subset(growD_all, Species==spp[i])
-  num_quads[i] <- length(unique(tmp$quad))
-}
-num_quads <- as.data.frame(num_quads)
-num_quads$species <- spp
-
-all_files <- list.files("./results/")
-all_fits <- data.frame(Parameter=NA, stddev=NA, id=NA, 
-                       rmsquad=NA, rep=NA, species=NA)
-for(i in 1:length(all_files)){
-  tmp <- all_files[i]
-  spp <- substring(tmp, 1, 4)
-  qsrm1 <- unlist(strsplit(tmp, "[_]"))[2]
-  qsrm <- gsub("numout", "", qsrm1)
-  rep1 <- unlist(strsplit(tmp, "[_]"))[3]
-  rep2 <- gsub(".RDS","",rep1)
-  rep <- gsub("rep", "", rep2)
-
-  long <- readRDS(paste("./results/",tmp, sep=""))
-  short <- ddply(long, .(Parameter), summarise,
-                 stddev = sd(value))
-  climeff <- short[grep("b2", short$Parameter),]
-  climeff$id <- substr(climeff$Parameter, 4, length(climeff$Parameter))
-  climeff$id <- as.numeric(unlist(strsplit(climeff$id, split=']')))  
-  climeff <- climeff[with(climeff, order(id)),]
-  climeff$rmsquad <- qsrm
-  climeff$rep <- rep
-  climeff$species <- spp
-  all_fits <- rbind(all_fits, climeff)
-} # end file loop
-
-all_fits <- all_fits[2:nrow(all_fits),]
-
-####
-####  Read in full fits
-####
-dir <- "../truncNormModel/"
-files <- list.files(dir)
-files <- files[grep(".RDS", files)]
-all_alls <- data.frame(Parameter=NA, stddev=NA, id=NA, 
-                       rmsquad=NA, rep=NA, species=NA)
-for(i in 1:length(files)){
-  tmp <- files[i]
-  spp1 <- unlist(strsplit(tmp, "[_]"))[3]
-  spp <- gsub(".RDS","",spp1)
-
-  long <- readRDS(paste(dir,tmp, sep=""))
-  short <- ddply(long, .(Parameter), summarise,
-                 stddev = sd(value))
-  climeff <- short[grep("b2", short$Parameter),]
-  climeff$id <- substr(climeff$Parameter, 4, length(climeff$Parameter))
-  climeff$id <- as.numeric(unlist(strsplit(climeff$id, split=']')))  
-  climeff <- climeff[with(climeff, order(id)),]
-  climeff$rmsquad <- 0
-  climeff$rep <- 1
-  climeff$species <- spp
-  all_alls <- rbind(all_alls, climeff)
-}
-all_alls <- all_alls[2:nrow(all_alls),]
-
-####
-####  Aggregate results by species and rep
-####
-all_fits <- rbind(all_fits, all_alls)
-agg_fits <- ddply(all_fits, .(Parameter, rmsquad, species, id), summarise,
-                  avg_stddev = mean(stddev))
-agg_fits2 <- agg_fits[which(agg_fits$id %in% c(1:5)),]
-agg_fits2 <- merge(agg_fits2, num_quads, by="species")
-agg_fits2$quadsfit <- with(agg_fits2, num_quads-as.numeric(rmsquad))
-
-####
-####  Fit model (need to do by species)
-####
-modBOGR <- gam(avg_stddev~quadsfit, data=subset(agg_fits2, species=="BOGR"),
-               family=quasi(link="1/mu^2"))
-modHECO <- gam(avg_stddev~quadsfit, data=subset(agg_fits2, species=="HECO"),
-               family=quasi(link="1/mu^2"))
-modPASM <- gam(avg_stddev~quadsfit, data=subset(agg_fits2, species=="PASM"),
-               family=quasi(link="1/mu^2"))
-modPOSE <- gam(avg_stddev~quadsfit, data=subset(agg_fits2, species=="POSE"),
-               family=quasi(link="1/mu^2"))
-summary(modBOGR)
-summary(modHECO)
-summary(modPASM)
-summary(modPOSE)
-
-####
-#### Plot
-####
-ggplot(agg_fits2, aes(x=quadsfit, y=avg_stddev))+
-  geom_line(aes(group=Parameter), alpha=0.3, linetype=2)+
-  geom_point(aes(group=Parameter), alpha=0.3, size=3)+
-#   stat_smooth(method="loess", size=2, color="black", se=FALSE)+
-  geom_smooth(method="glm", formula=y~x, family=quasi(link="1/mu^2"),
-              size=2, color="black", se=FALSE)+
-  facet_wrap("species", scales = "free", ncol=1)+
-  theme_bw()+
-  xlab("Number of quadrats fit")+
-  ylab("Standard deviation of coefficient")
-
-```
-
-```{r figure_6, dependson="plot-options", fig.cap="Correlations between QBM and IPM estimates of climate effects.
-We ignore sizeXclimate interactions since these are not directly comparable across model types.
-The QBM does not have multiple vital rates, so its values are repeated across panels within each species.", fig.width=6, fig.height=6, cache=FALSE, results='hide'}
-##  Figure script to look at correlations between IPM and QBM
-##    climate effect estimates.
-
-library(ggmcmc)
-library(gridExtra)
-library(ggplot2)
-
-allfiles <- list.files("../analysis/speciesData/")
-removals <- grep("csv", allfiles)
-species_list <- allfiles[-removals]
-
-clim_covs <- c("pptLag", "ppt1", "ppt2", "TmeanSpr1", "TmeanSpr2",
-               "ppt1xTmeanSpr1", "ppt2xTmeanSpr2",
-               "sizeXpptLag", "sizeXppt1", "sizeXppt2",
-               "sizeXTmeanSpr1", "sizeXTmeanSpr2")
-clim_mains <- c("pptLag", "ppt1", "ppt2", "TmeanSpr1", "TmeanSpr2",
-                "ppt1xTmeanSpr1", "ppt2xTmeanSpr2")
-
-####
-##  Get mean growth climate effects
-####
-growth_parms <- data.frame(species=NA, model=NA, parameter=NA, value=NA)
-for(spp in species_list){
-  tmp_data <- readRDS(paste("../analysis/ipm/vitalRateRegs/growth/growth_stanmcmc_", spp, ".RDS", sep=""))
-  tmp_clim <- tmp_data[grep("b2", tmp_data[,"Parameter"]), ]
-  tmp_clim[,"Parameter"] <- rep(clim_covs, each=3000)
-  tmp_agg <- ddply(tmp_clim, .(Parameter), summarise,
-                   value = median(value))
-  tmp_agg2 <- tmp_agg[which(tmp_agg$Parameter %in% clim_mains), ]
-  tmp_agg2$species <- spp
-  tmp_agg2$model <- "ipm"
-  colnames(tmp_agg2) <- tolower(colnames(tmp_agg2))
-  growth_parms <- rbind(growth_parms,
-                        tmp_agg2[,c("species", "model", "parameter", "value")])
-}
-
-####
-##  Get mean survival climate effects
-####
-surv_parms <- data.frame(species=NA, model=NA, parameter=NA, value=NA)
-params <- character(length = 12)
-for(i in 1:12){
-  params[i] <- paste("b2[",i,"]", sep="")
-}
-for(spp in species_list){
-  tmp_data <- readRDS(paste("../analysis/ipm/vitalRateRegs/survival/survival_stanmcmc_", spp, ".RDS", sep=""))
-  keeps <- which(tmp_data$Parameter %in% params)
-  tmp_clim <- tmp_data[keeps, ]
-  tmp_clim[,"Parameter"] <- rep(clim_covs, each=3000)
-  tmp_agg <- ddply(tmp_clim, .(Parameter), summarise,
-                   value = median(value))
-  tmp_agg2 <- tmp_agg[which(tmp_agg$Parameter %in% clim_mains), ]
-  tmp_agg2$species <- spp
-  tmp_agg2$model <- "ipm"
-  colnames(tmp_agg2) <- tolower(colnames(tmp_agg2))
-  surv_parms <- rbind(surv_parms,
-                        tmp_agg2[,c("species", "model", "parameter", "value")])
-}
-
-####
-##  Get mean recruitment climate effects
-####
-rec_parms <- data.frame(species=NA, model=NA, parameter=NA, value=NA)
-params <- character(length = 12)
-for(i in 1:12){
-  params[i] <- paste("b2[",i,"]", sep="")
-}
-for(spp in species_list){
-  tmp_data <- readRDS(paste("../analysis/ipm/vitalRateRegs/recruitment/recruitment_stanmcmc_", spp, ".RDS", sep=""))
-  keeps <- which(tmp_data$Parameter %in% params)
-  tmp_clim <- tmp_data[keeps, ]
-  tmp_clim[,"Parameter"] <- rep(clim_mains, each=3000)
-  tmp_agg <- ddply(tmp_clim, .(Parameter), summarise,
-                   value = median(value))
-  tmp_agg2 <- tmp_agg[which(tmp_agg$Parameter %in% clim_mains), ]
-  tmp_agg2$species <- spp
-  tmp_agg2$model <- "ipm"
-  colnames(tmp_agg2) <- tolower(colnames(tmp_agg2))
-  rec_parms <- rbind(rec_parms,
-                      tmp_agg2[,c("species", "model", "parameter", "value")])
-}
-
-####
-##  Get QBM growth climate estimates
-####
-clim_covs2 <- c("pptLag", "ppt1", "ppt2", "TmeanSpr1", "TmeanSpr2",
-               "ppt1xTmeanSpr1", "ppt2xTmeanSpr2",
-               "coverXpptLag", "coverXppt1", "coverXppt2",
-               "coverXTmeanSpr1", "coverXTmeanSpr2")
-
-qbm_parms <- data.frame(species=NA, model=NA, parameter=NA, qbmvalue=NA)
-params <- character(length = 12)
-for(i in 1:12){
-  params[i] <- paste("b2[",i,"]", sep="")
-}
-for(spp in species_list){
-  tmp_data <- readRDS(paste("../analysis/quadBM/vitalRateRegressions/truncNormModel/popgrowth_stanmcmc_", spp, ".RDS", sep=""))
-  keeps <- which(tmp_data$Parameter %in% params)
-  tmp_clim <- tmp_data[keeps, ]
-  tmp_clim[,"Parameter"] <- rep(clim_covs2, each=3000)
-  tmp_agg <- ddply(tmp_clim, .(Parameter), summarise,
-                   qbmvalue = median(value))
-  tmp_agg2 <- tmp_agg[which(tmp_agg$Parameter %in% clim_mains), ]
-  tmp_agg2$species <- spp
-  tmp_agg2$model <- "qbm"
-  colnames(tmp_agg2) <- tolower(colnames(tmp_agg2))
-  qbm_parms <- rbind(qbm_parms,
-                     tmp_agg2[,c("species", "model", "parameter", "qbmvalue")])
-}
-
-####
-##  Format data, combine, and plot
-####
-growth_parms <- growth_parms[2:nrow(growth_parms), ]
-growth_parms$vitalrate <- "growth"
-surv_parms <- surv_parms[2:nrow(surv_parms), ]
-surv_parms$vitalrate <- "surv"
-rec_parms <- rec_parms[2:nrow(rec_parms), ]
-rec_parms$vitalrate <- "rec"
-ipm_parms <- rbind(growth_parms, surv_parms, rec_parms)
-
-qbm_parms <- qbm_parms[2:nrow(qbm_parms), ]
-qbm_parms$vitalrate <- "growth"
-vitals <- c("surv", "rec")
-qbmalone <- qbm_parms
-for(dov in vitals){
-  tmp <- qbmalone
-  tmp$vitalrate <- dov
-  qbm_parms <- rbind(qbm_parms, tmp)
-}
-
-ipm_parms <- ipm_parms[,c("species","parameter","value","vitalrate")]
-qbm_parms <- qbm_parms[,c("species","parameter","qbmvalue","vitalrate")]
-all_ests <- merge(ipm_parms, qbm_parms, by = c("species", "parameter", "vitalrate"))
-
-# Calculate correlation for each group
-cors <- ddply(all_ests, c("species", "vitalrate"), summarise, 
-              cor = round(cor(value, qbmvalue), 2))
-
-ggplot(all_ests, aes(x=qbmvalue, y=value))+
-  geom_point(size=3)+
-  geom_smooth(method="lm", color="black", fill="grey")+
-  facet_grid(species~vitalrate)+
-  geom_text(data=cors, aes(label=paste("r = ", cor, sep="")), x=-1, y=1, size=4)+
-  xlab("QBM Estimate")+
-  ylab("IPM Estimate")+
-  theme_bw()
-
-setwd("~/Repos/MicroMesoForecast/manuscript/")
-```
-
-\pagebreak{}
-
-References
-----------

### 2015-07-13 manuscript/manuscript.Rmd
adds to SDM ms section; adds esa presentation draft

@@
 
-That population models do not scale easily explains, in part, why species distribution models (SDMs) have become a major tool for land managers and conservation planners.
-SDMs typically rely on easy-to-collect presence/absence data (but see...) that allow researchers to cover large spatial extents (e.g., ).
+That population models do not scale easily explains, in part, why species distribution models (SDMs) have become a major tool for land managers and conservation planners [see @Guisan2005 for a review].
+SDMs typically rely on easy-to-collect presence/absence data [but see @Clark2014 for new methods] that allow researchers to cover large spatial extents [e.g., @Maiorano2013].
 Likewise, predictors in SDMs tend to be remotely sensed products.
 Thus, it is relatively straightforward to parameterize and project SDMs at landscape and regional scales, the very scales at which land managers work.
-However, the limitations of SDMs are well known (cite some review).
-SOME MORE STUFF...
+However, the limitations of SDMs are well known [@Elith2009].
 It seems that what land managers really want are population models at the landscape and regional scale, if only the ecological community could provide them.
@@
 
-In general, the QBM climate effects are most correlated with climate effects from the growth model at the individual level (Fig.
+In general, the QBM climate effects are most correlated with climate effects from the growth regression at the individual level (Fig.
 6).
@@
 3a-d).
+The inability of the QBM to separate the sometimes countervailing effects of climate on survival, growth, and recruitment (Fig.
+3e-h) results in inaccurate (Fig.
+3a-d) and unprecise (Fig.
+4) forecasts.
+The QBM statistical model struggles to explain variation due to climate variables because they can have positive and negative impacts on different vital rates.
+When this is the case, as it is for all our species to varying degrees (Fig.
+3e-h), statistical models of aggregated population responses will fail.
 
@@
 In some cases and for certain species, population models based on aggregated data may prove useful and unbiased.
-However, our work here is the first comparison, to our knowledge, of population models based on individual and aggregated data in a climate change context.
+However, our work here is the first comparison, to our knowledge, of population models based on individual and aggregated forms of the same data in a climate change context.
 Our results confirm theoretical arguments [@Clark2011b] and empirical evidence [@Clark2011; @Clark2012] that individual responses are critical to predicting species' responses to climate change.
@@
 
-```{r figure_6, dependson="plot-options", fig.cap="Correlations between QBM and IPM estimates of climate effects.
+```{r figure_6, dependson="plot-options", fig.cap="Correlations (r) between QBM and IPM estimates of climate effects.
 We ignore sizeXclimate interactions since these are not directly comparable across model types.

### 2015-07-13 manuscript/manuscript.tex
adds to SDM ms section; adds esa presentation draft

@@
 species distribution models (SDMs) have become a major tool for land
-managers and conservation planners. SDMs typically rely on
-easy-to-collect presence/absence data (but see\ldots{}) that allow
-researchers to cover large spatial extents (e.g., ). Likewise,
+managers and conservation planners (see Guisan and Thuiller 2005 for a
+review). SDMs typically rely on easy-to-collect presence/absence data
+(but see Clark et al. 2014 for new methods) that allow researchers to
+cover large spatial extents (e.g., Maiorano et al. 2013). Likewise,
 predictors in SDMs tend to be remotely sensed products. Thus, it is
@@
 and regional scales, the very scales at which land managers work.
-However, the limitations of SDMs are well known (cite some review). SOME
-MORE STUFF\ldots{} It seems that what land managers really want are
-population models at the landscape and regional scale, if only the
-ecological community could provide them.
+However, the limitations of SDMs are well known (Elith and Leathwick
+2009). It seems that what land managers really want are population
+models at the landscape and regional scale, if only the ecological
+community could provide them.
 
@@
 In general, the QBM climate effects are most correlated with climate
-effects from the growth model at the individual level (Fig. 6). In no
-case does the QBM statistical model have strong correlations across all
-three vital rates (Fig. 6). Thus, for each species the QBM is
+effects from the growth regression at the individual level (Fig. 6). In
+no case does the QBM statistical model have strong correlations across
+all three vital rates (Fig. 6). Thus, for each species the QBM is
 ``missing'' climate signals associated with at least one vital rate.
 This has large impacts on predictions of long term population dynamics,
-as seen in our equilibrium simulations (Fig. 3a-d).
+as seen in our equilibrium simulations (Fig. 3a-d). The inability of the
+QBM to separate the sometimes countervailing effects of climate on
+survival, growth, and recruitment (Fig. 3e-h) results in inaccurate
+(Fig. 3a-d) and unprecise (Fig. 4) forecasts. The QBM statistical model
+struggles to explain variation due to climate variables because they can
+have positive and negative impacts on different vital rates. When this
+is the case, as it is for all our species to varying degrees (Fig.
+3e-h), statistical models of aggregated population responses will fail.
 
@@
 the first comparison, to our knowledge, of population models based on
-individual and aggregated data in a climate change context. Our results
-confirm theoretical arguments (Clark et al. 2011b) and empirical
-evidence (Clark et al. 2011a, 2012) that individual responses are
-critical to predicting species' responses to climate change. Thus,
-forecasts from aggregate level models should be viewed with caution and
-should never be unaccompianed by uncertainty. Given the importance of
-demographic data and its current difficulty to collect, we advocate for
-research on modern methods to collect demographic data more efficiently
-across environmental gradients in space and time.
+individual and aggregated forms of the same data in a climate change
+context. Our results confirm theoretical arguments (Clark et al. 2011b)
+and empirical evidence (Clark et al. 2011a, 2012) that individual
+responses are critical to predicting species' responses to climate
+change. Thus, forecasts from aggregate level models should be viewed
+with caution and should never be unaccompianed by uncertainty. Given the
+importance of demographic data and its current difficulty to collect, we
+advocate for research on modern methods to collect demographic data more
+efficiently across environmental gradients in space and time.
 
@@
 \includegraphics{components/figure/manuscript-figure_6.pdf}
-\caption{Correlations between QBM and IPM estimates of climate effects.
-We ignore sizeXclimate interactions since these are not directly
-comparable across model types. The QBM does not have multiple vital
-rates, so its values are repeated across panels within each species.
-Across top panels, `growth' = growth regression, `rec' = recruitment
-regression, `surv' = survival regression.}
+\caption{Correlations (r) between QBM and IPM estimates of climate
+effects. We ignore sizeXclimate interactions since these are not
+directly comparable across model types. The QBM does not have multiple
+vital rates, so its values are repeated across panels within each
+species. Across top panels, `growth' = growth regression, `rec' =
+recruitment regression, `surv' = survival regression.}
 \end{figure}
@@
 
+Clark, J. S., A. E. Gelfand, C. W. Woodall, and K. Zhu. 2014. More than
+the sum of the parts: Forest climate response from joint species
+distribution models. Ecological Applications 24:990--999.
+
 Doak, D. F., and W. F. Morris. 2010. Demographic compensation and
 tipping points in climate-induced range shifts. Nature 467:959--962.
+
+Elith, J., and J. R. Leathwick. 2009. Species Distribution Models:
+Ecological Explanation and Prediction Across Space and Time.
 
@@
 
+Guisan, A., and W. Thuiller. 2005. Predicting species distribution:
+Offering more than simple habitat models.
+
 Hobbs, N. T., and M. B. Hooten. 2015. Bayesian Models: A Statistical
@@
 data-rich era. Ecological Applications 21:1429--1442.
+
+Maiorano, L., R. Cheddadi, N. E. Zimmermann, L. Pellissier, B.
+Petitpierre, J. Pottier, H. Laborde, B. I. Hurdu, P. B. Pearman, A.
+Psomas, J. S. Singarayer, O. Broennimann, P. Vittoz, A. Dubuis, M. E.
+Edwards, H. A. Binney, and A. Guisan. 2013. Building the niche through
+time: using 13,000 years of data to predict the effects of climate
+change on three tree species in Europe. Global Ecology and Biogeography
+22:302--317.
 

### 2015-07-13 manuscript/manuscript.Rmd
small update to ms

@@
 
-Do we need detailed demographic data to forecast the impacts of climate change on plant populations?
-========================================================================================
-
-### Andrew T. Tredennick[^corrauth] and Peter B. Adler
-
-*Andrew T. Tredennick, Department of Wildland Resources and the Ecology Center, Utah State University, Logan, UT*
-
-*Peter B. Adler, Department of Wildland Resources and the Ecology Center, Utah State University, Logan, UT*
-
-[^corrauth]: Corresponding author: [atredenn@gmail.com](mailto:atredenn@gmail.com)
+\begin{centering}
+\textbf{\large{Do we need detailed demographic data to forecast the impacts of climate change on plant populations?}}
+
+\textsc{\small{Andrew T. Tredennick\footnote{E-mail: atredenn@gmail.com} and Peter B. Adler}}
+
+\textit{\small{Department of Wildland Resources and the Ecology Center, 5230 Old Main Hill, Utah State University, Logan, Utah 84322-5230 USA}}
+
+\end{centering}
 
@@
 Here, we test whether this assumption is important when forecasting climate change impacts on four perennial grass species in a semi-arid Montana grassland.
-We parameterized two population models, one based on inidividual level data with three vital rates and one on an aggregated version of the same data (percent cover), with climate covariates and compared their accuracy, precision, and sensitivity to climate.
+We parameterized two population models, one based on inidividual level data with three vital rates and one on an aggregated version of the same data (percent cover), and compared their accuracy, precision, and sensitivity to climate.
 The individual level model was more accurate and precise than the aggregate level model when predicting out of sample observations.
@@
 
-*Keywords*: forecasting, climate change, grassland, integral projection model, population model
+*Key words: forecasting, climate change, grassland, integral projection model, population model*
 

### 2015-07-13 manuscript/manuscript.tex
small update to ms

@@
 
-\section{Do we need detailed demographic data to forecast the impacts of
-climate change on plant
-populations?}\label{do-we-need-detailed-demographic-data-to-forecast-the-impacts-of-climate-change-on-plant-populations}
-
-\subsubsection[Andrew T. Tredennick and Peter B. Adler]{Andrew T.
-Tredennick\footnote{Corresponding author:
-  \href{mailto:atredenn@gmail.com}{\href{mailto:atredenn@gmail.com}{atredenn@gmail.com}}}
-and Peter B.
-Adler}\label{andrew-t.-tredennickcorrauth-and-peter-b.-adler}
-
-\emph{Andrew T. Tredennick, Department of Wildland Resources and the
-Ecology Center, Utah State University, Logan, UT}
-
-\emph{Peter B. Adler, Department of Wildland Resources and the Ecology
-Center, Utah State University, Logan, UT}
+\begin{centering}
+\textbf{\large{Do we need detailed demographic data to forecast the impacts of climate change on plant populations?}}
+
+\textsc{\small{Andrew T. Tredennick\footnote{E-mail: atredenn@gmail.com} and Peter B. Adler}}
+
+\textit{\small{Department of Wildland Resources and the Ecology Center, 5230 Old Main Hill, Utah State University, Logan, Utah 84322-5230 USA}}
+
+\end{centering}
 
@@
 with three vital rates and one on an aggregated version of the same data
-(percent cover), with climate covariates and compared their accuracy,
-precision, and sensitivity to climate. The individual level model was
-more accurate and precise than the aggregate level model when predicting
-out of sample observations. The aggregate level model made
-countervailing forecasts to 1\% climate changes when compared to the
-individual level model, indicating the aggregate level model produces
-incorrect forecasts. When comparing climate effects from both models,
-the aggregate level model tends to ``miss'' important climate effects
-from at least one vital rate for each species. It appears there is no
-short cut to forecasting climate change impacts on plant populations ---
-detailed demographic data is essential. But, forecasts were very
-uncertain, so we advocate for a focus on new methods to collect
-demographic data more efficiently across environmental gradients in
-space and time.
-
-\emph{Keywords}: forecasting, climate change, grassland, integral
-projection model, population model
+(percent cover), and compared their accuracy, precision, and sensitivity
+to climate. The individual level model was more accurate and precise
+than the aggregate level model when predicting out of sample
+observations. The aggregate level model made countervailing forecasts to
+1\% climate changes when compared to the individual level model,
+indicating the aggregate level model produces incorrect forecasts. When
+comparing climate effects from both models, the aggregate level model
+tends to ``miss'' important climate effects from at least one vital rate
+for each species. It appears there is no short cut to forecasting
+climate change impacts on plant populations --- detailed demographic
+data is essential. But, forecasts were very uncertain, so we advocate
+for a focus on new methods to collect demographic data more efficiently
+across environmental gradients in space and time.
+
+\emph{Key words: forecasting, climate change, grassland, integral
+projection model, population model}
 

### 2015-07-14 manuscript/manuscript.Rmd

@@
 ------------
-Population models are important tools for predicting the impacts of environmental change on species.
+Population models are important tools for predicting the impacts of environmental change on species persistence and abundance.
 But reconciling the scales at which population models are parameterized and the scales at which environmental changes play out remains a challenge [@Clark2010; @Freckleton2011; @Queenborough2011; @Clark2012].
-The major hurdle is that most population models, at least for plant species, are built using data from small, localized plots because parameterizing traditional population models requires tracking the fates of individuals.
-These models are difficult to scale up from the micro to meso-scales because the fitted parameters do not fully represent the spatial variation present at scales beyond that at which the data are collected [@Sæther2007].
-At the same time, most demographic data is collected over short time spans.
-For example, the most common study duration in the COMPADRE matrix population model database is 4 years and only a few exceed 10 years [@Salguero-Gomez2015].
-The constrained spatio-temporal extent of most demographic datasets reflects the difficulty of collecting such data, but those constraints limit our ability to extrapolate population models.
-Thus, our ability to use population models to predict the consequences of climate change is limited when we rely on individual-level data.
-
-That population models do not scale easily explains, in part, why species distribution models (SDMs) have become a major tool for land managers and conservation planners [see @Guisan2005 for a review].
-SDMs typically rely on easy-to-collect presence/absence data [but see @Clark2014 for new methods] that allow researchers to cover large spatial extents [e.g., @Maiorano2013].
-Likewise, predictors in SDMs tend to be remotely sensed products.
-Thus, it is relatively straightforward to parameterize and project SDMs at landscape and regional scales, the very scales at which land managers work.
+The major hurdle is that most population models are built using data from a single study site because collecting those data, which involves tracking the fates of individuals plants, is so difficult.
+The resulting models cannot be applied to the landscape and regional scales relevant to decision-makers without information about how the fitted parameters respond to spatial variation in biotic and abiotic drivers [@Sæther2007].
+The temporal coverage of demographic datasets is no better.
+[I think the temporal extent is a separate issue, might muddy the waters to bring it up here...] For example, the most common study duration in the COMPADRE matrix population model database is 4 years and only a few studies exceed 10 years [@Salguero-Gomez2015].
+The limited spatio-temporal extent of individual-level demographic datasets constrains our ability to use population models to address applied questions about the consequences of climate change.
+
+The inability of many population models to address landscape-scale problems may explain why land managers and conservation planners have embraced species distribution models (SDMs) [see @Guisan2005 for a review].
+SDMs typically rely on easy-to-collect presence/absence data [but see @Clark2014 for new methods] and remotely-sensed environmental covariates that allow researchers to model large spatial extents [e.g., @Maiorano2013].
+Thus, it is relatively straightforward to parameterize and project SDMs over landscapes and regions, the scales at which many land-use decisions are made.
 However, the limitations of SDMs are well known [@Elith2009].
-It seems that what land managers really want are population models at the landscape and regional scale, if only the ecological community could provide them.
-
-Aggregate measures of individual plant performance, such as those typically collected as part of large-scale census efforts, offer an alternative to detailed demographic data for modeling populations [@Clark2004; @Freckleton2011].
-Such population-level data will never match the precision of individual-level data, but it is more feasible to attain a broad coverage sample when collecting coarse-scale data.
-This presents a difficult trade-off: on the one hand, individual-level data leads to more reliable models; on the other hand, population-level data leads to models that will produce less precise predictions but can be applied over greater spatial and temporal extents.
-An open question is how well models based on population-level data compare to models based on individual-level data.
-
-To date, relatively few studies have tried to model populations based on data other than detailed individual-level data.
-An important exception is an effort by @Taylor2004 to model the population growth rate of an invasive species to investigate the best strategies for invasion control.
+Ideally, researchers would provide managers with landscape-scale population models, combing the extent of SDMs with information about dynamics and species abundances.
+
+Aggregate measures of population status, rather than individual performance, offer an intriguing alternative for modeling populations [@Clark2004; @Freckleton2011].
+Population-level data cannot provide inference about
+#will never match the precision of individual-level data, IS THIS TRUE? WHAT DOES PRECISION MEAN?
+demographic mechanisms, but it might be sufficient for modeling future population states, especially since 
+it is feasible to collect across broad spatial extents.
+The choice between individual and population-level data involves a difficult trade-off: while individual-level data leads to more mechanistic models, population-level data leads to models that can be applied over greater spatial and temporal extents.
+An open question is how much forecasting skill is lost when we build models based on population rather than individual-level data.
+
+To date, most empirical population modelers have relied on individual-level data, with few attempts to exploit population-level data.
+An important exception was an effort by @Taylor2004 to model the population growth rate of an invasive species to identify the best strategies for invasion control.
 They used a "density-structured" model where the state variable is a discrete density state rather than a continuous density measure.
-Building on this work, @Freckleton2011 showed that density-structured models compare well to continuous models in theory, and @Queenborough2011 showed the application of such methods in a study on arable weeds.
+Building on this work, @Freckleton2011 showed that density-structured models compare well to continuous models in theory, and @Queenborough2011 demonstrated the application of such methods in a study on arable weeds.
 In particular, @Queenborough2011 provide empirical evidence that density-structured models are capable of reproducing population dynamics, even if some precision is lost when compared to fully continuous models.
@@
 
-Basing population models on aggregated individual-level data in a climate change context is hampered by the fact that it is individuals that respond to climate, not populations [@Clark2012].
-This fact puts us in uneasy proximity to an "ecological fallacy" where one deduces inference on the individual from statistical inference on the group [@PIANTADOSI1988].
+Addressing climate change questions with models fit to population-level data is potentially problematic.
+It is individuals, not populations, that respond to climate variables [@Clark2012].
+Ignoring this fact puts us in uneasy proximity to an "ecological fallacy," where inference about the individual relies on statistical inference on the group [@PIANTADOSI1988].
 For example, individual plants may respond positively to precipitation but a negative trend is observed at the population level due to increased competition among plants as they grow larger and consume more resources.
-Thus, it is important to ask the question: Can aggregated data be used to detect climate signals of the same sign and magnitude as individual-level data?
-If not, then building population models with climate covariates on aggregated data will lead to incorrect forecasts.
-
-Here, we test the assumption that statistical and population models based on aggregated data can detect climate signals as wells as models based on individual-level data.
-We use a unique demographic dataset that tracks the fates of individual plants from four species over 14 years to build single-species population models, since those are often used tools for ecological forecasts and climate vulnerability assessments.
-We first fit population models with interannual variation in vital rates explained, in part, by climate covariates.
-We then perturb the climate covariates to test the sensitivities of species to climate change.
-By doing these analyses using both individual and aggregated forms of the same data we can directly compare the two types of models.
-
-In general, we find that population models based on detailed demographic data are more accurate and precise than models based on aggregated data.
-Both types of models are able to detect climate signals, as evidenced by the sensitivity of simulated equilbrium plant cover under a perturbed climate scenario.
-But the two types of models produce inconsistent forecasts, in some cases producing completely opposing predictions.
-This leads us to conclude that, at least for these species at this location, detailed demographic data is necessary to detect the "right" climate signal.
-A worrying caveat to our work is that forecasts from both models were very uncertain.
-It seems that even 14 years worth of demographic data is not enough to produce meaningful forecasts when model uncertainty is explicitly considered.
+# I think you need a better example. This example makes the population level approach look more sensible than the
+# individual level approach. Better to introduce the ideas we focus on in Discussion--growth vs survival vs 
+# fecundity, with different signals at different time scales...? 
+If population-level data cannot detect important impacts of climate drivers on populations, then population models built with such data will make poor forecasts.
+
+Here, we ask whether statistical and population models based on aggregated, population-level data can detect climate signals as wells as models based on individual-level data.
+We used a unique demographic dataset that tracks the fates of individual plants from four species over 14 years to build two kinds of single-species population models, traditional models using individual growth, survival, and recruitment data and alternative models based on basal cover.
+In both models, interannual variation is explained, in part, by climate covariates.
+We then performed simulations to quantify the sensitivities of species' cover to small perturbations in average precipitation and temperature.
+We found that population models based on detailed demographic data are more accurate and precise than models based on aggregated data.
+While both types of models detected climate signals, they produced inconsistent forecasts, in some cases even predicting opposing qualitative responses to a climate perturbation.
+For these species at this location, detailed demographic data appears necessary to make accurate forecasts.
+A worrying caveat to our work is that forecasts from both models were very uncertain when we considered full process and parameter uncertainty.
+Even 14 years worth of demographic data may not be sufficient to make useful forecasts.
 
@@
 ### Study site and data
-Our demographic data comes from the Fort Keogh Livestock and Range Research Laboratory in eastern Montana's northern mixed prairie near Miles City, Montana, USA ($46^{\circ}$ 19' N, $105^{\circ}$ 48' W).
-The dataset is freely available on Ecological Archives[^link] [@Anderson2011] , and interested readers should refer to the metadata therein for a complete description.
+Our demographic data come from the Fort Keogh Livestock and Range Research Laboratory in eastern Montana's northern mixed prairie near Miles City, Montana, USA ($46^{\circ}$ 19' N, $105^{\circ}$ 48' W).
+The dataset is freely available on Ecological Archives[^link] [@Anderson2011] , and interested readers should refer to the metadata for a complete description.
 The site is about 800 m above sea level and mean annual precipitation (1878-2009) is 334 mm, with most annual precipitation falling from April through September.
-The site is grass dominated and, for the purposes of our study, we focus on the four most abundant graminoid species: \emph{Bouteloua gracilis} (BOGR), \emph{Hesperostipa comata} (HECO), \emph{Pascopyrum smithii} (PASM), and \emph{Poa secunda} (POSE) (Fig.
+The community is grass-dominated and we focused on the four most abundant grass species: \emph{Bouteloua gracilis} (BOGR), \emph{Hesperostipa comata} (HECO), \emph{Pascopyrum smithii} (PASM), and \emph{Poa secunda} (POSE) (Fig.
 1).
@@
 In this analysis we account for potential differences among the grazing treatments, but do not focus on grazing$\times$climate interactions.
-The annual maps of the quadrats were digitized and the fates of individual plants tracked and extracted using a computer program.
+The annual maps of the quadrats were digitized and the fates of individual plants tracked and extracted using a computer program [cite Lauenroth and Adler 2008, Chu et al.
+2014].
 Daily climate data, which we aggregated into climate variables of interest, are available for the duration of the data collection period (1932 - 1945) from the Miles City airport, Wiley Field, 9 km from the study site.
@@
 The individual data is the "raw" data.
-For the quadrat level we data we simply sum individual areal cover for each quadrat by species.
+For the quadrat level we data we simply sum individual basal cover for each quadrat by species.
 This is equivalent to a perfect census of quadrat percent cover, so we do not need to consider measurement error.
@@
 
+# might be better to say that we have quantified measurement error and it doesn't seem large enough to
+# influence our results (Chu and Adler 2014)
+
 All R code and data necessary to reproduce our analysis is archived on GitHub as release v1.0[^revnote] (http://github.com/atredennick/MicroMesoForecast/releases).
@@
 
+# You prefer GitHub to a Data Dryad package? Dryad would be more dependable long term...?
+
 ### Stastical models of vital rates
 At both levels of inference (individual and quadrat), the building blocks of our population models are vital rate regressions.
-For individual level data we fit models for survival, growth, and recruitment of new individuals for each species.
-At the quadrat level we fit a single regression model for population growth.
+For individual-level data we fit models for survival, growth, and recruitment for each species.
+At the quadrat-level we fit a single regression model for population growth.
 We describe the statistical models separately since fitting the models required different approaches.
@@
 
+# let's talk about timing of "spring". Miles City gets a lot of moisture in May and June...
+
 We fit all models using a hierarchical Bayesian approach.
-The models are fully descibed in Appendix A, so here we focus on the main process and the mode likelihood.
-For the likelihood models, $\textbf{y}^X$ is always the relevant vector of observations for vital rate *X* ($X = S, G, R, or P for survival, growth, recruitment, or population growth) For example, $\textbf{y}^S$ is a vector of 0s and 1s indicating whether a genet surives from *t* to *t+1*, or not.
+The models are fully descibed in Appendix A, so here we focus on the main process and the model likelihood.
+For the likelihood models, $\textbf{y}^X$ is always the relevant vector of observations for vital rate *X* ($X = S, G, R, or P for survival, growth, recruitment, or population growth).
+For example, $\textbf{y}^S$ is a vector of 0s and 1s indicating whether a genet survives from *t* to *t+1*, or not.
 
@@
 The statistical approach used to model vital rates using aggregated data depends on the type of data collected.
-In our case, and as is often the case with census data, we have percent cover data (which can easily be transformed to proportion data, of course).
+In our case, and as is often the case with census data, we have percent cover data (which can easily be transformed to proportion data).
 We first considered fitting three vital rate models analagous to those we fit at the individual level: one for probability of extirpation within a quadrat (analagous to survival), one for cover change within a quadrat (analagous to growth), and one for probability of colonization within a quadrat (analagous to recruitment).
-However, within-quadrat extirpation and colonization events were rare in our time series ($N=9$ and $N=10$, respectively across all species).
+However, within-quadrat extirpation and colonization events were rare in our time series ($N=9$ and $N=10$, respectively, across all species).
 Given the broad spatial distribution of the quadrats we are studying, it is safe to assume that these events are in fact rare enough to be ignored for our purposes.
@@
 An obvious choice for fitting a linear model to proportion data is beta regression because the support of the beta distribution is [0,1], not including true zeros or ones.
-However, when we used fitted model parameters from a beta regression in a quadrat-based population model the simulated population tended toward 100% cover for all species.
+However, when we used fitted model parameters from a beta regression in a quadrat-based population model, the simulated population tended toward 100% cover for all species.
 We therefore chose a more constrained modeling approach based on a truncated log-normal likelihood.
@@
 We assessed convergence visually and made sure scale reduction factors for all parameters were less than 1.01.
-For the purposes of including parameter uncertainty in our population models, we saved the final 1,000 iterations from each of the three MCMC chains for all parameters to be used as randomly drawn values during population simulation.
+For the purposes of including parameter uncertainty in our population models, we saved the final 1,000 iterations from each of the three MCMC chains to be used as randomly drawn values during population simulation.
 This step alleviates the need to reduce model parameters by model selection since sampling from the full parameter space in the MCMC ensures that if a parameter broadly overlaps zero, on average the effect in the population models will also be near zero.
@@
 With the posterior distribution of the vital rate statistical models in hand, it is straightforward to simulate the population models.
-We used an Integral Projection Model (IPM) to model populations based on individual level data and an quadrat based version of an individually-based model (Quadrat-Based Model, QBM) to model populations based on quadrat level data.
+We used an Integral Projection Model (IPM) to model populations based on individual level data [cite Ellner and Rees 2006] and an quadrat based version of an individually-based model (Quadrat-Based Model, QBM) to model populations based on quadrat level data.
 We describe each in turn.
@@
 We use an environmentally stochastic IPM [@Rees2009] that includes the random year effects and the climate covariates from the vital rate statistical models.
-But note that we can, and do for some simulations, ignore the random year effects so that only the climate effects can drive interannual variation.
+However, for some simulations, we ignore the random year effects so that only the climate effects drive interannual variation.
 Our IPM follows the specification of @Chu2015 where the population of species _j_ is a density function $n(u_{j},t)$ giving the density of sized-_u_ genets at time _t_.
@@
 
-which, said plainly, means we are calculating growth (_G_) for individuals that survive (_S_) from time _t_ to _t+1_ and adding in newly recruited (_R_) individuals of an average sized one-year-old genet for the focal species.
+which means we are calculating growth (_G_) for individuals that survive (_S_) from time _t_ to _t+1_ and adding in newly recruited (_R_) individuals of an average sized one-year-old genet for the focal species.
 Our stastical model for recruitment (_R_, described above) returns the number of new recruit produced per quadrat.
 Following previous work [@Adler2012; @Chu2015], we assume that fecundity increases linearly with size ($R_j(v_{j},u_{j},\bar{\bold{w_{j}}}) = e^{u_j}R_j(v_{j},\bar{\bold{w_{j}}})$) to incorporate the recruitment function in the spatially-implicit IPM.
+[You probably should say something about spatial structure and cite Adler et al.
+2010 EcoLetts for the "no overlap rule"]
 
@@
 #### Quad-based model
-Our quad-based model (QBM) perfectly mirrors its statistical description (Eqs. 8-9). We use the same approach for drawing parameter values as described for the IPM.
+Our quad-based model (QBM) perfectly mirrors [I think you need to say a little more about how you simulate it] its statistical description (Eqs.
+8-9).
+We use the same approach for drawing parameter values as described for the IPM.
 
 ### Model validation
-To test each model's ability to forecast the population state we made out of sample predictions using leave-one-year-out cross validation.
+To test each model's ability to forecast population state, we made out-of-sample predictions using leave-one-year-out cross validation.
 For both levels of modeling, we fit the vital rate models using observations from all years except one, and then used those fitted parameters in the population models to perform a one-step-ahead forecast for the year whose observations were withheld from model fitting.
-Within each observation year, several quadrats are sampled.
-So we made predictions for each observed quadrat in the focal year initialized with cover the previous year.
-Since we were making quadrat specific predictions we incorporated the group effect on the intercept for both models.
-We repeated this procedure for all 13 observation years, making 100 one-step-ahead forecasts for each quadrat-year combination with parameter uncertainty included via randomdrawd from the MCMC chain as described above.
+Within each observation year, several quadrats were sampled.
+So we made predictions for each observed quadrat in the focal year, initializing each simulation with cover in the quadrat the previous year.
+Since we were making quadrat-specific predictions, we incorporated the group effect on the intercept for both models.
+We repeated this procedure for all 13 observation years, making 100 one-step-ahead forecasts for each quadrat-year combination with parameter uncertainty included via random draw from the MCMC chain as described above.
 Random year effects were set to zero since year effects cannot be assigned to unobserved years.
 
-This model validation allowed us to compare accuracy and precision of the two modeling approaches (individual-level versus population-level).
-We first calculated the median predicted cover across the 100 simulations for each quadrat-year and then calculated the absolute error as the difference between the observed cover for a given quadrat-year and the median prediction.
+This model validation allowed us to compare accuracy and precision of the two modeling approaches (IPM versus QBM).
+We first calculated the median predicted cover across the 100 simulations for each quadrat-year and then calculated the absolute error as the absolute value of the difference between the observed cover for a given quadrat-year and the median prediction.
 To arrive at mean absolute error (MAE), we then averaged the absolute error within each species across the quadrat-year specific errors.
@@
 ### Testing sensitivity to climate covariates
-Our main goal in this paper is to see if models based on aggregate level data are as sensitive to climate as models based on individual level data.
+Our main goal in this paper is to see if models based on aggregated data are as sensitive to climate covariates as models based on individual level data.
 So, with our fitted and validated models in hand, we ran simulations for each model type (IPM and QBM) under four climate perturbation scenarios: (1) observed climate, (2) precipitation increased by 1%, (3) temperature increased by 1%, and (4) precipitation and temperature increased by 1%.
@@
 As an effort to identify potential discrepencies between IPM and QBM forecasts, we also ran simulations designed to quantify the sensitivities of individual and combined vital rates to climate for the IPM.
-Specifically, we ran simulations for the above climate scenarios, but applied the perturbed climate covariates to survival, growth, and recruitment vital rates individually and in pairwise combinations.
-This allows us to isolate the vital rate(s) most sensitive to climate.
+Specifically, we ran simulations for the above climate scenarios, but applied the perturbed climate covariates to survival, growth, or recruitment vital rates individually and in pairwise combinations.
+This allowed us to isolate the vital rate(s) most sensitive to climate.
 For this analysis, we used mean parameter estimates to reduce the sources of uncertainty in the sensitivity estimates.
@@
 ----------
+
+MOVE FIRST TWO PARAGRAPHS TO INTRO. MOVE FIG 5 and 6 TO METHODS AND RESULTS
+
 Perhaps the greatest challenge for ecology in the 21st century is to forecast the impacts of environmental change [@Clark2001; @Petchey2015].
@@
 Yet, model predictions proved so uncertain that any forecast, when bounded with model and parameter uncertainty, would be at best not useful and at worst meaningless.
+For all species, the 90% quantiles of predicted changes in populatoin size overlapped zero; we cannot even predict whether a change in precipitation or temperature will cause populations to increase or decrease.
 How might we improve on this state of affairs?

### 2015-07-15 manuscript/appC.Rmd
adds appC for recruitment uncertainty IPM sims

---
layout: 12pt
header-includes:
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{setspace}
bibliography: ~/Dropbox/Bibliography/MicroMeso.bib
csl: components/ecology.csl

## rmarkdown render options
output:
  pdf_document:
    fig_caption: true
    keep_tex: true
fontsize: 12pt
geometry: margin=1in

---

\renewcommand\thefigure{C\arabic{figure}}

Appendix C: Effect of recruitment on IPM uncertainty
=====================================================
In Fig. 4 of the main text where we show the uncertainty around our models' forecasts to a 1% change in climate covariates, the obvious outlier is *Pascopyrum smithii* (PASM). The uncertainty around PASM forecasts from the integral projection model is orders of magnitude larger than for any other species, and is even larger than the uncertainty around the quad-based model forecasts. It turns out that the recruitment parameters we estimate lead to overcompensating density-dependence that is amplified when those parameters are allowed to vary. When we allow all parameters to vary except those associated with recruitment, uncertainty around PASM decreases to levels similar to those associated with other species (Fig. C1). So detailed demographic data is not a cure-all, and can sometimes lead to interesting dynamics when parameter uncertainty is considered.

```{r plot_it, echo=FALSE, message=FALSE, warning=FALSE, include=TRUE, fig.width=5, fig.cap="Equilibrium cover and 90% quantiles around the mean prediction from the IPM when model error and parameter uncertainty are propogated through the simulation phase for all vital rates except recruitment."}
library(plyr)
library(ggplot2)
setwd("~/Repos/MicroMesoForecast/analysis/")
meanparam_ipm <- data.frame(cover=NA, species=NA, 
                             climsim=NA, model=NA)
##  IPM 
path <- "./ipm/simulations/results/cc_rectest/"
ipmfiles_mean <- list.files(path)
ipmfiles_mean <- ipmfiles_mean[grep("cover", ipmfiles_mean)]
for(file in ipmfiles_mean){
  tmp <- read.csv(paste(path,file,sep=""))
  tmp$model <- "IPM"
  meanparam_ipm <- rbind(meanparam_ipm, tmp[,c("cover", "species", "climsim", "model")])
}
meanparam_ipm <- meanparam_ipm[2:nrow(meanparam_ipm),]

newd <- data.frame(cover=NA, species=NA, 
                   climsim=NA, model=NA, obscover=NA)
for(dospp in unique(meanparam_ipm$species)){
  tmp <- subset(meanparam_ipm, species==dospp)
  obscover <- subset(tmp, climsim=="obs")["cover"]
  tmp2 <- subset(tmp, climsim!="obs")
  tmp2$obscover <- rep(obscover$cover, times = length(unique(tmp2$climsim)))
  newd <- rbind(newd, tmp2)
}
ipm_meansims <- newd[2:nrow(newd),]
ipm_meansims$propdiff <- with(ipm_meansims, log(cover*100)-log(obscover*100))

meanparam_agg <- ddply(ipm_meansims, .(species, climsim, model), summarise,
                       meandiff = mean(propdiff),
                       hidiff = quantile(propdiff, probs = 0.95),
                       lodiff = quantile(propdiff, probs = 0.05))
dgd = position_dodge(width = 0.6)
ggplot(meanparam_agg, aes(color=species))+
  geom_errorbar(aes(x=climsim, ymax=hidiff, ymin=lodiff),
                position=dgd, width=0.25)+
  geom_point(aes(x=climsim, y=meandiff),
                position=dgd, width=0.25)+
  theme_bw()+
#   facet_wrap("species", ncol=1, scales="free")+
  scale_x_discrete(labels=c("+ppt", "+temp", "+ppt&temp"))+
  ylab("Proportional difference")+
  xlab("Climate simulation")
```

### 2015-07-15 manuscript/appC.Rmd
updates colors in appC figure

@@
 
+```{r caching, include=FALSE}
+library("methods")
+library("knitr")
+basename <- "appC"
+opts_chunk$set(fig.path = paste("appC/figure/", basename, "-", sep=""),
+               cache.path = paste("appC/cache/", basename, "/", sep=""))
+opts_chunk$set(cache = 2)
+opts_chunk$set(tidy=FALSE, warning=FALSE, message=FALSE, 
+               comment = NA, verbose = TRUE, echo=FALSE)
+
+# PDF-based figures
+opts_chunk$set(dev='pdf')
+```
+
 Appendix C: Effect of recruitment on IPM uncertainty
@@
 
-```{r plot_it, echo=FALSE, message=FALSE, warning=FALSE, include=TRUE, fig.width=5, fig.cap="Equilibrium cover and 90% quantiles around the mean prediction from the IPM when model error and parameter uncertainty are propogated through the simulation phase for all vital rates except recruitment."}
+```{r plot-options, message=FALSE, warning=FALSE, include=FALSE, echo=FALSE}
+cbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", 
+               "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
+
+opts_chunk$set(dev='pdf', dev.args=c(version="1.3"))
+
+```
+
+```{r plot_it, echo=FALSE, message=FALSE, warning=FALSE, include=TRUE, fig.width=5, fig.cap="Equilibrium cover and 90% quantiles around the mean prediction from the IPM when model error and parameter uncertainty are propogated through the simulation phase for all vital rates except recruitment.", dependson="plot-options"}
 library(plyr)
@@
   ylab("Proportional difference")+
-  xlab("Climate simulation")
+  xlab("Climate simulation")+
+  scale_color_manual(values=cbPalette[1:4])
 ```

### 2015-07-15 manuscript/manuscript.Rmd
incorporates some of Peter's suggestions

@@
    - \doublespacing
+   - \usepackage{todonotes}
 bibliography: ~/Dropbox/Bibliography/MicroMeso.bib
@@
 ------------
+
+Perhaps the greatest challenge for ecology in the 21st century is to forecast the impacts of environmental change [@Clark2001; @Petchey2015].
+To do so requires sophisticated modeling approaches that fully account for uncertainty and variability in the ecological process and associated parameters [@Luo2011a].
+This requires large amounts of data collected over large spatio-temporal extents.
+State-of-the-science modeling techniques cannot overcome data limitations.
+Such is the case for many population models.
+
 Population models are important tools for predicting the impacts of environmental change on species persistence and abundance.
@@
 Aggregate measures of population status, rather than individual performance, offer an intriguing alternative for modeling populations [@Clark2004; @Freckleton2011].
-Population-level data cannot provide inference about
-#will never match the precision of individual-level data, IS THIS TRUE? WHAT DOES PRECISION MEAN?
-demographic mechanisms, but it might be sufficient for modeling future population states, especially since 
-it is feasible to collect across broad spatial extents.
+Population-level data cannot provide inference about demographic mechanisms, but it might be sufficient for modeling future population states, especially since it is feasible to collect across broad spatial extents [e.g., @Queenborough2011].
 The choice between individual and population-level data involves a difficult trade-off: while individual-level data leads to more mechanistic models, population-level data leads to models that can be applied over greater spatial and temporal extents.
@@
 They used a "density-structured" model where the state variable is a discrete density state rather than a continuous density measure.
-Building on this work, @Freckleton2011 showed that density-structured models compare well to continuous models in theory, and @Queenborough2011 demonstrated the application of such methods in a study on arable weeds.
+Such models do not require individual level demographic data and can adequately describe population dynamics.
+
+Building on @Taylor2004, @Freckleton2011 showed that density-structured models compare well to continuous models in theory, and @Queenborough2011 demonstrated the application of such methods in a study on arable weeds.
 In particular, @Queenborough2011 provide empirical evidence that density-structured models are capable of reproducing population dynamics, even if some precision is lost when compared to fully continuous models.
-Thus, population models based on coarse, population-level data show promise for producing ecological forecasts at landscape and regional scales [@Queenborough2011].
+The study by @Queenborough2011 included data from 500 fields (4 hectares each) in 49 farms, all collected by two people in 6 weeks.
+This is far more data from a far greater spatial extent than possible if measuring individual plant demography (in a world of limited time and money, at least).
+The appeal of density-structured approaches is clear.
 However, none of these models included environmental covariates.
@@
 It is individuals, not populations, that respond to climate variables [@Clark2012].
-Ignoring this fact puts us in uneasy proximity to an "ecological fallacy," where inference about the individual relies on statistical inference on the group [@PIANTADOSI1988].
-For example, individual plants may respond positively to precipitation but a negative trend is observed at the population level due to increased competition among plants as they grow larger and consume more resources.
-# I think you need a better example. This example makes the population level approach look more sensible than the
-# individual level approach. Better to introduce the ideas we focus on in Discussion--growth vs survival vs 
-# fecundity, with different signals at different time scales...? 
+Ignoring this fact puts us in uneasy proximity to an "ecological fallacy", where inference about the individual relies on statistical inference on the group [@PIANTADOSI1988].
+Population growth (or decline) is an epiphenomenon of demographic processes like survival, growth, and recruitment that occur at the level of individual plants.
+Climate can affect each demographic process in unique, potentially opossing, ways [@Dalgleish2011].
+These unique climate responses may difficult resolve in statistical models based on population-level data where demographic processes are not identifiable.
 If population-level data cannot detect important impacts of climate drivers on populations, then population models built with such data will make poor forecasts.
@@
 We found that population models based on detailed demographic data are more accurate and precise than models based on aggregated data.
-While both types of models detected climate signals, they produced inconsistent forecasts, in some cases even predicting opposing qualitative responses to a climate perturbation.
+Our results suggest the popuation-level model is less accurate and precise because important demographic climate signals go undetected.
 For these species at this location, detailed demographic data appears necessary to make accurate forecasts.
@@
 In this analysis we account for potential differences among the grazing treatments, but do not focus on grazing$\times$climate interactions.
-The annual maps of the quadrats were digitized and the fates of individual plants tracked and extracted using a computer program [cite Lauenroth and Adler 2008, Chu et al.
-2014].
+The annual maps of the quadrats were digitized and the fates of individual plants tracked and extracted using a computer program [@Lauenroth2008; @Chu2014a].
 Daily climate data, which we aggregated into climate variables of interest, are available for the duration of the data collection period (1932 - 1945) from the Miles City airport, Wiley Field, 9 km from the study site.
@@
 For the quadrat level we data we simply sum individual basal cover for each quadrat by species.
-This is equivalent to a perfect census of quadrat percent cover, so we do not need to consider measurement error.
+This is equivalent to a near-perfect census of quadrat percent cover because previous analysis shows that measurement error at the individual level is small [@Chu2014].
 Based on these two datasets we can compare population models built using individual level data and aggregated quadrat level data.
-
-# might be better to say that we have quantified measurement error and it doesn't seem large enough to
-# influence our results (Chu and Adler 2014)
 
@@
 That stable release will remain static as a record of this analysis, but subsequent versions may appear if we update this work.
+We have also deposited the v1.0 release on Dryad (*link here after acceptance*).
 
 [^revnote]: *Note to reviewers*: so that v1.0 will be associated with the published version of the manuscript, we have released v0.1 to be associated with this review version.
-
-# You prefer GitHub to a Data Dryad package? Dryad would be more dependable long term...?
 
@@
 
-# let's talk about timing of "spring". Miles City gets a lot of moisture in May and June...
+\todo{let's talk about timing of "spring". Miles City gets a lot of moisture in May and June...}
 
@@
 The models are fully descibed in Appendix A, so here we focus on the main process and the model likelihood.
-For the likelihood models, $\textbf{y}^X$ is always the relevant vector of observations for vital rate *X* ($X = S, G, R, or P for survival, growth, recruitment, or population growth).
+For the likelihood models, $\textbf{y}^X$ is always the relevant vector of observations for vital rate *X* ($X = S, G, R, or P$ for survival, growth, recruitment, or population growth).
 For example, $\textbf{y}^S$ is a vector of 0s and 1s indicating whether a genet survives from *t* to *t+1*, or not.
@@
 Following previous work [@Adler2012; @Chu2015], we assume that fecundity increases linearly with size ($R_j(v_{j},u_{j},\bar{\bold{w_{j}}}) = e^{u_j}R_j(v_{j},\bar{\bold{w_{j}}})$) to incorporate the recruitment function in the spatially-implicit IPM.
-[You probably should say something about spatial structure and cite Adler et al.
-2010 EcoLetts for the "no overlap rule"]
+\todo{Add spatial structure mention.}
 
@@
 #### Quad-based model
-Our quad-based model (QBM) perfectly mirrors [I think you need to say a little more about how you simulate it] its statistical description (Eqs.
+Our quad-based model (QBM) perfectly mirrors its statistical description (Eqs.
 8-9).
 We use the same approach for drawing parameter values as described for the IPM.
+After drawing the appropriate parameter set, we calculate the mean response (population cover at *t+1* = $x_{t+1}$) according to Eq.
+8.
+We then make a random draw from a [0,1] truncated lognormal distribution with mean equal to $x_{t+1}$ from Eq.
+8 and the variance estimate from the fitted model.
+We can then iterate the model forward by drawing a new parameter set (unique to climate year and MCMC iteration) at each timestep.
 
@@
 For this analysis, we used mean parameter estimates to reduce the sources of uncertainty in the sensitivity estimates.
+
+Our *a priori* expectation is that the IPM will produce more accurate and precise forecasts.
+This could be due to differences in sample sizes leading to larger parameter uncertainty for the QBM, or due to the QBM climate effects being weakly associated with one or more vital rate climate effects at the individual level.
+To assess the impact of sample size on QBM parameter uncertainty we refit the QBM statistical model (Eqs.
+8-9) after removing sets of 2, 5, 10, and 15 quadrats.
+We fit 10 models at each level of quadrat removal (2, 5, 10, 15 quadrats), removing a different randomly selected set of quadrats for each fit.
+We calculated the standard deviation of climate main effects (pptlag, ppt1, ppt2, TmeanSpr1, and TmeanSpr2) for each model and averaged those over replicates within each set of quadrat removals.
+This allowed us to regress sample size against parameter uncertainty.
+
+To see if the QBM climate effects are correlated, or not, with climate effects for each vital rate model for the IPM, we simply regressed the QBM climate coefficients against each vital rate model's climate coefficients and calculate Pearson's $\rho$.
+Strong correlations indicate the QBM is capable of detecting climate effects associated with individual vital rates.
+A weak correlation indicates the QBM "misses" the climate effect on a particular vital rate.
 
@@
 Since the IPM approach relies on vital rate regressions, we were able to quantify the sensitivity of each vital rate in isolation and in pairwise combinations.
-Species showed similar trends (Fig.
+Across all species, climate covariates can have opposing effects on different vital rates (Fig.
 3e-h).
+
 Growth was the most sensitive vital rate for all species, showing a negative response to increased precipitation, and stronger positive response to increased temperature, and a mostly positive response when both climate factors are increased (Fig.
@@
 4).
+IPM forecasts for *P.
+smithiii* were very uncertain due to overcompensating density dependence and nonlinearities recruitment that get amplified by parameter uncertainty (Appendx C).
+Neither model was capable of making forecasts of proportional cover change distinguishable from 0 (i.e., no change in population state) when we included model error and parameter uncertainty.
+
+### Sources of uncertainty in the QBM
+Sample size had a relatively weak effect on QBM climate parameter uncertainty after the number of quadrats exceeded about 10 (Fig.
+5).
+Inverse-gaussian fits show that increasing sample size beyond the number of quadrats we used will results in deminishing returns in terms of parameter certainty (Fig.
+5).
+
+Climate effects estimated from the QBM are most correlated with climate effects from the growth regression at the individual level (Fig.
+6).
+In no case does the QBM statistical model have strong correlations across all three vital rates (Fig.
+6).
+QBM climate effects were most weakly correlated with those from individual-level recruitment models for *B.
+gracilis*, *H.
+comata*, and *P.
+secunda* (Fig.
+6a,b,d).
+For *P.
+smithii*, QBM climate effects showed no correlation with the survival model effects (Fig.
+6c).
 
@@
 ----------
-
-MOVE FIRST TWO PARAGRAPHS TO INTRO. MOVE FIG 5 and 6 TO METHODS AND RESULTS
-
-Perhaps the greatest challenge for ecology in the 21st century is to forecast the impacts of environmental change [@Clark2001; @Petchey2015].
-To do so requires sophisticated modeling approaches that fully account for uncertainty and variability in the ecological process and associated parameters [@Luo2011a].
-This requires large amounts of data collected over large spatio-temporal extents.
-State-of-the-science modeling techniques cannot overcome data limitations.
-Such is the case for many population models.
-
-As a potential remedy to the "data dearth" problem, @Queenborough2011 and @Freckleton2011, building on work by @Taylor2004, advocate a "density-structured" modeling approach.
-Such models do not require individual level demographic data and can adequately describe population dynamics [@Queenborough2011].
-The results from density-structured models are not as precise as those from traditional population models, but the loss in precision is traded off with a gain in data.
-The study by @Queenborough2011 included data from 500 fields (4 hectares each) in 49 farms, all collected by two people in 6 weeks.
-This is far more data from a far greater spatial extent than possible if measuring individual plant demography (in a world of limited time and money, at least).
-The appeal of density-structured approaches is clear.
-
-However, at their core, density-structured models rely on individual level data aggregated to a population level metric (e.g., density classes or percent cover).
+Population models built using individual-level data allow inference on demographic processes, but their ability to forecast future population states is limited by the spatial extent of the data.
+Population-level data, like percent cover or discrete density states, are much easier to collect of broad spatial extents, so populations built using such data offer an appealing alternative to traditional population models [@Queenborough2011].
+However, at their core, density-structured models rely on individual level data aggregated to a population level metric.
 This creates a potential problem if such models are to be used in a climate change context because inidividuals respond to climate, not populations [@Clark2012].
@@
 ### The IPM and QBM produced inconsistent forecasts
-Using individual and aggregated forms of the same dataset, we were able to directly compare a traditional demographic modeling approach to a population model based on aggregated data.
+Using individual-level and population-level forms of the same dataset, we were able to directly compare a traditional demographic modeling approach to a population model based on percent cover data.
 Our quad-based model (QBM) is based on percent cover data and so is in the spirit of density-structured models.

### 2015-07-15 manuscript/manuscript.tex
incorporates some of Peter's suggestions

@@
 \doublespacing
+\usepackage{todonotes}
 
@@
 Population models are important tools for predicting the impacts of
-environmental change on species. But reconciling the scales at which
-population models are parameterized and the scales at which
-environmental changes play out remains a challenge (Clark et al. 2010,
-2012, Freckleton et al. 2011, Queenborough et al. 2011). The major
-hurdle is that most population models, at least for plant species, are
-built using data from small, localized plots because parameterizing
-traditional population models requires tracking the fates of
-individuals. These models are difficult to scale up from the micro to
-meso-scales because the fitted parameters do not fully represent the
-spatial variation present at scales beyond that at which the data are
-collected (Sæther et al. 2007). At the same time, most demographic data
-is collected over short time spans. For example, the most common study
-duration in the COMPADRE matrix population model database is 4 years and
-only a few exceed 10 years (Salguero-Gómez et al. 2015). The constrained
-spatio-temporal extent of most demographic datasets reflects the
-difficulty of collecting such data, but those constraints limit our
-ability to extrapolate population models. Thus, our ability to use
-population models to predict the consequences of climate change is
-limited when we rely on individual-level data.
-
-That population models do not scale easily explains, in part, why
-species distribution models (SDMs) have become a major tool for land
-managers and conservation planners (see Guisan and Thuiller 2005 for a
-review). SDMs typically rely on easy-to-collect presence/absence data
-(but see Clark et al. 2014 for new methods) that allow researchers to
-cover large spatial extents (e.g., Maiorano et al. 2013). Likewise,
-predictors in SDMs tend to be remotely sensed products. Thus, it is
-relatively straightforward to parameterize and project SDMs at landscape
-and regional scales, the very scales at which land managers work.
-However, the limitations of SDMs are well known (Elith and Leathwick
-2009). It seems that what land managers really want are population
-models at the landscape and regional scale, if only the ecological
-community could provide them.
-
-Aggregate measures of individual plant performance, such as those
-typically collected as part of large-scale census efforts, offer an
-alternative to detailed demographic data for modeling populations (Clark
-and Bjørnstad 2004, Freckleton et al. 2011). Such population-level data
-will never match the precision of individual-level data, but it is more
-feasible to attain a broad coverage sample when collecting coarse-scale
-data. This presents a difficult trade-off: on the one hand,
-individual-level data leads to more reliable models; on the other hand,
-population-level data leads to models that will produce less precise
-predictions but can be applied over greater spatial and temporal
-extents. An open question is how well models based on population-level
-data compare to models based on individual-level data.
-
-To date, relatively few studies have tried to model populations based on
-data other than detailed individual-level data. An important exception
-is an effort by Taylor and Hastings (2004) to model the population
-growth rate of an invasive species to investigate the best strategies
-for invasion control. They used a ``density-structured'' model where the
-state variable is a discrete density state rather than a continuous
-density measure. Building on this work, Freckleton et al. (2011) showed
-that density-structured models compare well to continuous models in
-theory, and Queenborough et al. (2011) showed the application of such
-methods in a study on arable weeds. In particular, Queenborough et al.
-(2011) provide empirical evidence that density-structured models are
-capable of reproducing population dynamics, even if some precision is
-lost when compared to fully continuous models. Thus, population models
-based on coarse, population-level data show promise for producing
-ecological forecasts at landscape and regional scales (Queenborough et
-al. 2011). However, none of these models included environmental
-covariates.
-
-Basing population models on aggregated individual-level data in a
-climate change context is hampered by the fact that it is individuals
-that respond to climate, not populations (Clark et al. 2012). This fact
-puts us in uneasy proximity to an ``ecological fallacy'' where one
-deduces inference on the individual from statistical inference on the
-group (Piantadosi et al. 1988). For example, individual plants may
-respond positively to precipitation but a negative trend is observed at
-the population level due to increased competition among plants as they
-grow larger and consume more resources. Thus, it is important to ask the
-question: Can aggregated data be used to detect climate signals of the
-same sign and magnitude as individual-level data? If not, then building
-population models with climate covariates on aggregated data will lead
-to incorrect forecasts.
-
-Here, we test the assumption that statistical and population models
-based on aggregated data can detect climate signals as wells as models
-based on individual-level data. We use a unique demographic dataset that
-tracks the fates of individual plants from four species over 14 years to
-build single-species population models, since those are often used tools
-for ecological forecasts and climate vulnerability assessments. We first
-fit population models with interannual variation in vital rates
-explained, in part, by climate covariates. We then perturb the climate
-covariates to test the sensitivities of species to climate change. By
-doing these analyses using both individual and aggregated forms of the
-same data we can directly compare the two types of models.
-
-In general, we find that population models based on detailed demographic
-data are more accurate and precise than models based on aggregated data.
-Both types of models are able to detect climate signals, as evidenced by
-the sensitivity of simulated equilbrium plant cover under a perturbed
-climate scenario. But the two types of models produce inconsistent
-forecasts, in some cases producing completely opposing predictions. This
-leads us to conclude that, at least for these species at this location,
-detailed demographic data is necessary to detect the ``right'' climate
-signal. A worrying caveat to our work is that forecasts from both models
-were very uncertain. It seems that even 14 years worth of demographic
-data is not enough to produce meaningful forecasts when model
-uncertainty is explicitly considered.
+environmental change on species persistence and abundance. But
+reconciling the scales at which population models are parameterized and
+the scales at which environmental changes play out remains a challenge
+(Clark et al. 2010, 2012, Freckleton et al. 2011, Queenborough et al.
+2011). The major hurdle is that most population models are built using
+data from a single study site because collecting those data, which
+involves tracking the fates of individuals plants, is so difficult. The
+resulting models cannot be applied to the landscape and regional scales
+relevant to decision-makers without information about how the fitted
+parameters respond to spatial variation in biotic and abiotic drivers
+(Sæther et al. 2007). The temporal coverage of demographic datasets is
+no better. {[}I think the temporal extent is a separate issue, might
+muddy the waters to bring it up here\ldots{}{]} For example, the most
+common study duration in the COMPADRE matrix population model database
+is 4 years and only a few studies exceed 10 years (Salguero-Gómez et al.
+2015). The limited spatio-temporal extent of individual-level
+demographic datasets constrains our ability to use population models to
+address applied questions about the consequences of climate change.
+
+The inability of many population models to address landscape-scale
+problems may explain why land managers and conservation planners have
+embraced species distribution models (SDMs) (see Guisan and Thuiller
+2005 for a review). SDMs typically rely on easy-to-collect
+presence/absence data (but see Clark et al. 2014 for new methods) and
+remotely-sensed environmental covariates that allow researchers to model
+large spatial extents (e.g., Maiorano et al. 2013). Thus, it is
+relatively straightforward to parameterize and project SDMs over
+landscapes and regions, the scales at which many land-use decisions are
+made. However, the limitations of SDMs are well known (Elith and
+Leathwick 2009). Ideally, researchers would provide managers with
+landscape-scale population models, combing the extent of SDMs with
+information about dynamics and species abundances.
+
+Aggregate measures of population status, rather than individual
+performance, offer an intriguing alternative for modeling populations
+(Clark and Bjørnstad 2004, Freckleton et al. 2011). Population-level
+data cannot provide inference about demographic mechanisms, but it might
+be sufficient for modeling future population states, especially since it
+is feasible to collect across broad spatial extents (e.g., Queenborough
+et al. 2011). The choice between individual and population-level data
+involves a difficult trade-off: while individual-level data leads to
+more mechanistic models, population-level data leads to models that can
+be applied over greater spatial and temporal extents. An open question
+is how much forecasting skill is lost when we build models based on
+population rather than individual-level data.
+
+To date, most empirical population modelers have relied on
+individual-level data, with few attempts to exploit population-level
+data. An important exception was an effort by Taylor and Hastings (2004)
+to model the population growth rate of an invasive species to identify
+the best strategies for invasion control. They used a
+``density-structured'' model where the state variable is a discrete
+density state rather than a continuous density measure. Building on this
+work, Freckleton et al. (2011) showed that density-structured models
+compare well to continuous models in theory, and Queenborough et al.
+(2011) demonstrated the application of such methods in a study on arable
+weeds. In particular, Queenborough et al. (2011) provide empirical
+evidence that density-structured models are capable of reproducing
+population dynamics, even if some precision is lost when compared to
+fully continuous models. Thus, population models based on coarse,
+population-level data show promise for producing ecological forecasts at
+landscape and regional scales (Queenborough et al. 2011). However, none
+of these models included environmental covariates.
+
+Addressing climate change questions with models fit to population-level
+data is potentially problematic. It is individuals, not populations,
+that respond to climate variables (Clark et al. 2012). Ignoring this
+fact puts us in uneasy proximity to an ``ecological fallacy'', where
+inference about the individual relies on statistical inference on the
+group (Piantadosi et al. 1988). Population growth (or decline) is an
+epiphenomenon of demographic processes like survival, growth, and
+recruitment that occur at the level of individual plants. Climate can
+affect each demographic process in unique, potentially opossing, ways
+(Dalgleish et al. 2011). These unique climate responses may difficult
+resolve in statistical models based on population-level data where
+demographic processes are not identifiable. If population-level data
+cannot detect important impacts of climate drivers on populations, then
+population models built with such data will make poor forecasts.
+
+Here, we ask whether statistical and population models based on
+aggregated, population-level data can detect climate signals as wells as
+models based on individual-level data. We used a unique demographic
+dataset that tracks the fates of individual plants from four species
+over 14 years to build two kinds of single-species population models,
+traditional models using individual growth, survival, and recruitment
+data and alternative models based on basal cover. In both models,
+interannual variation is explained, in part, by climate covariates. We
+then performed simulations to quantify the sensitivities of species'
+cover to small perturbations in average precipitation and temperature.
+We found that population models based on detailed demographic data are
+more accurate and precise than models based on aggregated data. Our
+results suggest the popuation-level model is less accurate and precise
+because important demographic climate signals go undetected. For these
+species at this location, detailed demographic data appears necessary to
+make accurate forecasts. A worrying caveat to our work is that forecasts
+from both models were very uncertain when we considered full process and
+parameter uncertainty. Even 14 years worth of demographic data may not
+be sufficient to make useful forecasts.
 
@@
 
-Our demographic data comes from the Fort Keogh Livestock and Range
+Our demographic data come from the Fort Keogh Livestock and Range
 Research Laboratory in eastern Montana's northern mixed prairie near
@@
 (Anderson et al. 2011) , and interested readers should refer to the
-metadata therein for a complete description. The site is about 800 m
-above sea level and mean annual precipitation (1878-2009) is 334 mm,
-with most annual precipitation falling from April through September. The
-site is grass dominated and, for the purposes of our study, we focus on
-the four most abundant graminoid species: \emph{Bouteloua gracilis}
-(BOGR), \emph{Hesperostipa comata} (HECO), \emph{Pascopyrum smithii}
-(PASM), and \emph{Poa secunda} (POSE) (Fig. 1).
+metadata for a complete description. The site is about 800 m above sea
+level and mean annual precipitation (1878-2009) is 334 mm, with most
+annual precipitation falling from April through September. The community
+is grass-dominated and we focused on the four most abundant grass
+species: \emph{Bouteloua gracilis} (BOGR), \emph{Hesperostipa comata}
+(HECO), \emph{Pascopyrum smithii} (PASM), and \emph{Poa secunda} (POSE)
+(Fig. 1).
 
@@
 quadrats were digitized and the fates of individual plants tracked and
-extracted using a computer program. Daily climate data, which we
-aggregated into climate variables of interest, are available for the
-duration of the data collection period (1932 - 1945) from the Miles City
-airport, Wiley Field, 9 km from the study site.
+extracted using a computer program (Lauenroth and Adler 2008, Chu et al.
+2014). Daily climate data, which we aggregated into climate variables of
+interest, are available for the duration of the data collection period
+(1932 - 1945) from the Miles City airport, Wiley Field, 9 km from the
+study site.
 
@@
 individual and quadrat (Fig. 2). The individual data is the ``raw''
-data. For the quadrat level we data we simply sum individual areal cover
-for each quadrat by species. This is equivalent to a perfect census of
-quadrat percent cover, so we do not need to consider measurement error.
+data. For the quadrat level we data we simply sum individual basal cover
+for each quadrat by species. This is equivalent to a near-perfect census
+of quadrat percent cover because previous analysis shows that
+measurement error at the individual level is small (Chu and Adler 2014).
 Based on these two datasets we can compare population models built using
@@
 stable release will remain static as a record of this analysis, but
-subsequent versions may appear if we update this work.
+subsequent versions may appear if we update this work. We have also
+deposited the v1.0 release on Dryad (\emph{link here after acceptance}).
 
@@
 blocks of our population models are vital rate regressions. For
-individual level data we fit models for survival, growth, and
-recruitment of new individuals for each species. At the quadrat level we
-fit a single regression model for population growth. We describe the
-statistical models separately since fitting the models required
-different approaches. All models contain five climate covariate that we
-chose \emph{a priori}: ``water year'' precipitation at \emph{t}-1
-(lagppt); fall through spring precipitation at \emph{t}-1 and \emph{t}-2
-(ppt1 and ppt2, respectively) and mean spring temperature at \emph{t}-1
-and \emph{t}-2 (TmeanSpr1 and TmeanSpr2, respectively), where \emph{t}
-is the observation year. We also include interactions among same-year
+individual-level data we fit models for survival, growth, and
+recruitment for each species. At the quadrat-level we fit a single
+regression model for population growth. We describe the statistical
+models separately since fitting the models required different
+approaches. All models contain five climate covariate that we chose
+\emph{a priori}: ``water year'' precipitation at \emph{t}-1 (lagppt);
+fall through spring precipitation at \emph{t}-1 and \emph{t}-2 (ppt1 and
+ppt2, respectively) and mean spring temperature at \emph{t}-1 and
+\emph{t}-2 (TmeanSpr1 and TmeanSpr2, respectively), where \emph{t} is
+the observation year. We also include interactions among same-year
 climate covariates (e.g., ppt1 $\times$ TmeansSpr1) and climate $\times$
@@
 
+\todo{let's talk about timing of "spring". Miles City gets a lot of moisture in May and June...}
+
 We fit all models using a hierarchical Bayesian approach. The models are
 fully descibed in Appendix A, so here we focus on the main process and
-the mode likelihood. For the likelihood models, $\textbf{y}^X$ is always
-the relevant vector of observations for vital rate \emph{X}
-($X = S, G, R, or P for survival, growth, recruitment, or population growth) For example, $\textbf{y}\^{}S\$
-is a vector of 0s and 1s indicating whether a genet surives from
+the model likelihood. For the likelihood models, $\textbf{y}^X$ is
+always the relevant vector of observations for vital rate \emph{X}
+($X = S, G, R, or P for survival, growth, recruitment, or population growth). For example, $\textbf{y}\^{}S\$
+is a vector of 0s and 1s indicating whether a genet survives from
 \emph{t} to \emph{t+1}, or not.
@@
 case with census data, we have percent cover data (which can easily be
-transformed to proportion data, of course). We first considered fitting
-three vital rate models analagous to those we fit at the individual
-level: one for probability of extirpation within a quadrat (analagous to
-survival), one for cover change within a quadrat (analagous to growth),
-and one for probability of colonization within a quadrat (analagous to
-recruitment). However, within-quadrat extirpation and colonization
-events were rare in our time series ($N=9$ and $N=10$, respectively
-across all species). Given the broad spatial distribution of the
-quadrats we are studying, it is safe to assume that these events are in
-fact rare enough to be ignored for our purposes. So we constrained our
-statistical modeling of vital rates at the population level to change in
-percent cover within quadrats. For the remaining discussion of
-statistical modeling we refer to proportion data, which is simply
-percent data divided by 100.
+transformed to proportion data). We first considered fitting three vital
+rate models analagous to those we fit at the individual level: one for
+probability of extirpation within a quadrat (analagous to survival), one
+for cover change within a quadrat (analagous to growth), and one for
+probability of colonization within a quadrat (analagous to recruitment).
+However, within-quadrat extirpation and colonization events were rare in
+our time series ($N=9$ and $N=10$, respectively, across all species).
+Given the broad spatial distribution of the quadrats we are studying, it
+is safe to assume that these events are in fact rare enough to be
+ignored for our purposes. So we constrained our statistical modeling of
+vital rates at the population level to change in percent cover within
+quadrats. For the remaining discussion of statistical modeling we refer
+to proportion data, which is simply percent data divided by 100.
 
@@
 not including true zeros or ones. However, when we used fitted model
-parameters from a beta regression in a quadrat-based population model
+parameters from a beta regression in a quadrat-based population model,
 the simulated population tended toward 100\% cover for all species. We
@@
 purposes of including parameter uncertainty in our population models, we
-saved the final 1,000 iterations from each of the three MCMC chains for
-all parameters to be used as randomly drawn values during population
-simulation. This step alleviates the need to reduce model parameters by
-model selection since sampling from the full parameter space in the MCMC
-ensures that if a parameter broadly overlaps zero, on average the effect
-in the population models will also be near zero. We report the posterior
-mean, standard deviation, and 95\% Bayesian Credible Intervals for every
-parameter of each model for each species in Appendix B.
+saved the final 1,000 iterations from each of the three MCMC chains to
+be used as randomly drawn values during population simulation. This step
+alleviates the need to reduce model parameters by model selection since
+sampling from the full parameter space in the MCMC ensures that if a
+parameter broadly overlaps zero, on average the effect in the population
+models will also be near zero. We report the posterior mean, standard
+deviation, and 95\% Bayesian Credible Intervals for every parameter of
+each model for each species in Appendix B.
 
@@
 an Integral Projection Model (IPM) to model populations based on
-individual level data and an quadrat based version of an
-individually-based model (Quadrat-Based Model, QBM) to model populations
-based on quadrat level data. We describe each in turn.
+individual level data {[}cite Ellner and Rees 2006{]} and an quadrat
+based version of an individually-based model (Quadrat-Based Model, QBM)
+to model populations based on quadrat level data. We describe each in
+turn.
 
@@
 includes the random year effects and the climate covariates from the
-vital rate statistical models. But note that we can, and do for some
-simulations, ignore the random year effects so that only the climate
-effects can drive interannual variation. Our IPM follows the
-specification of Chu and Adler (2015) where the population of species
-\emph{j} is a density function $n(u_{j},t)$ giving the density of
-sized-\emph{u} genets at time \emph{t}. Genet size is on the natural log
-scale, so that $n(u_{j},t)du$ is the number of genets whose area (on the
-arithmetic scale) is between $e^{u_{j}}$ and $e^{u_{j}+du}$. So, the
-density function for any size \emph{v} at time $t+1$ is
+vital rate statistical models. However, for some simulations, we ignore
+the random year effects so that only the climate effects drive
+interannual variation. Our IPM follows the specification of Chu and
+Adler (2015) where the population of species \emph{j} is a density
+function $n(u_{j},t)$ giving the density of sized-\emph{u} genets at
+time \emph{t}. Genet size is on the natural log scale, so that
+$n(u_{j},t)du$ is the number of genets whose area (on the arithmetic
+scale) is between $e^{u_{j}}$ and $e^{u_{j}+du}$. So, the density
+function for any size \emph{v} at time $t+1$ is
 
@@
 
-which, said plainly, means we are calculating growth (\emph{G}) for
-individuals that survive (\emph{S}) from time \emph{t} to \emph{t+1} and
-adding in newly recruited (\emph{R}) individuals of an average sized
-one-year-old genet for the focal species. Our stastical model for
-recruitment (\emph{R}, described above) returns the number of new
-recruit produced per quadrat. Following previous work (Adler et al.
-2012, Chu and Adler 2015), we assume that fecundity increases linearly
-with size
+which means we are calculating growth (\emph{G}) for individuals that
+survive (\emph{S}) from time \emph{t} to \emph{t+1} and adding in newly
+recruited (\emph{R}) individuals of an average sized one-year-old genet
+for the focal species. Our stastical model for recruitment (\emph{R},
+described above) returns the number of new recruit produced per quadrat.
+Following previous work (Adler et al. 2012, Chu and Adler 2015), we
+assume that fecundity increases linearly with size
 ($R_j(v_{j},u_{j},\bar{\bold{w_{j}}}) = e^{u_j}R_j(v_{j},\bar{\bold{w_{j}}})$)
 to incorporate the recruitment function in the spatially-implicit IPM.
+{[}You probably should say something about spatial structure and cite
+Adler et al. 2010 EcoLetts for the ``no overlap rule''{]}
 
@@
 
-Our quad-based model (QBM) perfectly mirrors its statistical description
+Our quad-based model (QBM) perfectly mirrors {[}I think you need to say
+a little more about how you simulate it{]} its statistical description
 (Eqs. 8-9). We use the same approach for drawing parameter values as
@@
 
-To test each model's ability to forecast the population state we made
-out of sample predictions using leave-one-year-out cross validation. For
+To test each model's ability to forecast population state, we made
+out-of-sample predictions using leave-one-year-out cross validation. For
 both levels of modeling, we fit the vital rate models using observations
@@
 whose observations were withheld from model fitting. Within each
-observation year, several quadrats are sampled. So we made predictions
-for each observed quadrat in the focal year initialized with cover the
-previous year. Since we were making quadrat specific predictions we
-incorporated the group effect on the intercept for both models. We
-repeated this procedure for all 13 observation years, making 100
-one-step-ahead forecasts for each quadrat-year combination with
-parameter uncertainty included via randomdrawd from the MCMC chain as
-described above. Random year effects were set to zero since year effects
-cannot be assigned to unobserved years.
+observation year, several quadrats were sampled. So we made predictions
+for each observed quadrat in the focal year, initializing each
+simulation with cover in the quadrat the previous year. Since we were
+making quadrat-specific predictions, we incorporated the group effect on
+the intercept for both models. We repeated this procedure for all 13
+observation years, making 100 one-step-ahead forecasts for each
+quadrat-year combination with parameter uncertainty included via random
+draw from the MCMC chain as described above. Random year effects were
+set to zero since year effects cannot be assigned to unobserved years.
 
 This model validation allowed us to compare accuracy and precision of
-the two modeling approaches (individual-level versus population-level).
-We first calculated the median predicted cover across the 100
-simulations for each quadrat-year and then calculated the absolute error
-as the difference between the observed cover for a given quadrat-year
-and the median prediction. To arrive at mean absolute error (MAE), we
-then averaged the absolute error within each species across the
-quadrat-year specific errors. We use MAE as our measure of accuracy. To
-measure precision we calculated the distance between the upper and lower
-90th quantiles of the 100 predictions and averaged this value over
+the two modeling approaches (IPM versus QBM). We first calculated the
+median predicted cover across the 100 simulations for each quadrat-year
+and then calculated the absolute error as the absolute value of the
+difference between the observed cover for a given quadrat-year and the
+median prediction. To arrive at mean absolute error (MAE), we then
+averaged the absolute error within each species across the quadrat-year
+specific errors. We use MAE as our measure of accuracy. To measure
+precision we calculated the distance between the upper and lower 90th
+quantiles of the 100 predictions and averaged this value over
 quadrat-years for each species.
@@
 
-Our main goal in this paper is to see if models based on aggregate level
-data are as sensitive to climate as models based on individual level
-data. So, with our fitted and validated models in hand, we ran
+Our main goal in this paper is to see if models based on aggregated data
+are as sensitive to climate covariates as models based on individual
+level data. So, with our fitted and validated models in hand, we ran
 simulations for each model type (IPM and QBM) under four climate
@@
 IPM. Specifically, we ran simulations for the above climate scenarios,
-but applied the perturbed climate covariates to survival, growth, and
+but applied the perturbed climate covariates to survival, growth, or
 recruitment vital rates individually and in pairwise combinations. This
-allows us to isolate the vital rate(s) most sensitive to climate. For
+allowed us to isolate the vital rate(s) most sensitive to climate. For
 this analysis, we used mean parameter estimates to reduce the sources of
@@
 
+MOVE FIRST TWO PARAGRAPHS TO INTRO. MOVE FIG 5 and 6 TO METHODS AND
+RESULTS
+
 Perhaps the greatest challenge for ecology in the 21st century is to
@@
 bounded with model and parameter uncertainty, would be at best not
-useful and at worst meaningless. How might we improve on this state of
-affairs?
+useful and at worst meaningless. For all species, the 90\% quantiles of
+predicted changes in populatoin size overlapped zero; we cannot even
+predict whether a change in precipitation or temperature will cause
+populations to increase or decrease. How might we improve on this state
+of affairs?
 
@@
 
+Chu, C., K. M. Havstad, N. Kaplan, W. K. Lauenroth, M. P. McClaran, D.
+P. Peters, L. T. Vermeire, and P. B. Adler. 2014. Life form influences
+survivorship patterns for 109 herbaceous perennials from six semi-arid
+ecosystems. Journal of Vegetation Science 25:947--954.
+
 Clark, J. S., and O. N. Bjørnstad. 2004. Population time series: Process
@@
 
+Dalgleish, H. J., D. N. Koons, M. B. Hooten, C. A. Moffet, and P. B.
+Adler. 2011. Climate influences the demography of three dominant
+sagebrush steppe plants. Ecology 92:75--85.
+
 Doak, D. F., and W. F. Morris. 2010. Demographic compensation and
@@
 Primer for Ecologists. Princeton University PressPrinceton.
+
+Lauenroth, W. K., and P. B. Adler. 2008. Demography of perennial
+grassland plants: Survival, life expectancy and life span. Journal of
+Ecology 96:1023--1032.
 

### 2015-07-16 manuscript/manuscript.Rmd
more ms updates; moves discussion opener to intro

@@
 We describe the statistical models separately since fitting the models required different approaches.
-All models contain five climate covariate that we chose \emph{a priori}: "water year" precipitation at \emph{t}-1 (lagppt); fall through spring precipitation at \emph{t}-1 and \emph{t}-2 (ppt1 and ppt2, respectively) and mean spring temperature at \emph{t}-1 and \emph{t}-2 (TmeanSpr1 and TmeanSpr2, respectively), where \emph{t} is the observation year.
+All models contain five climate covariates that we chose \emph{a priori}: "water year" precipitation at \emph{t}-1 (lagppt); April through June precipitation at \emph{t}-1 and \emph{t}-2 (ppt1 and ppt2, respectively) and April through June temperature at \emph{t}-1 and \emph{t}-2 (TmeanSpr1 and TmeanSpr2, respectively), where \emph{t} is the observation year.
 We also include interactions among same-year climate covariates (e.g., ppt1 $\times$ TmeansSpr1) and climate $\times$ size interactions.
 Climate $\times$ size interactions are for climate main effects only, that is we do not include interactions among size and interacting climate effects.
-
-\todo{let's talk about timing of "spring". Miles City gets a lot of moisture in May and June...}
 

### 2015-07-16 manuscript/manuscript.tex
more ms updates; moves discussion opener to intro

@@
 
+Perhaps the greatest challenge for ecology in the 21st century is to
+forecast the impacts of environmental change (Clark et al. 2001, Petchey
+et al. 2015). To do so requires sophisticated modeling approaches that
+fully account for uncertainty and variability in the ecological process
+and associated parameters (Luo et al. 2011). This requires large amounts
+of data collected over large spatio-temporal extents.
+State-of-the-science modeling techniques cannot overcome data
+limitations. Such is the case for many population models.
+
 Population models are important tools for predicting the impacts of
@@
 ``density-structured'' model where the state variable is a discrete
-density state rather than a continuous density measure. Building on this
-work, Freckleton et al. (2011) showed that density-structured models
-compare well to continuous models in theory, and Queenborough et al.
-(2011) demonstrated the application of such methods in a study on arable
-weeds. In particular, Queenborough et al. (2011) provide empirical
-evidence that density-structured models are capable of reproducing
-population dynamics, even if some precision is lost when compared to
-fully continuous models. Thus, population models based on coarse,
-population-level data show promise for producing ecological forecasts at
-landscape and regional scales (Queenborough et al. 2011). However, none
-of these models included environmental covariates.
+density state rather than a continuous density measure. Such models do
+not require individual level demographic data and can adequately
+describe population dynamics.
+
+Building on Taylor and Hastings (2004), Freckleton et al. (2011) showed
+that density-structured models compare well to continuous models in
+theory, and Queenborough et al. (2011) demonstrated the application of
+such methods in a study on arable weeds. In particular, Queenborough et
+al. (2011) provide empirical evidence that density-structured models are
+capable of reproducing population dynamics, even if some precision is
+lost when compared to fully continuous models. The study by Queenborough
+et al. (2011) included data from 500 fields (4 hectares each) in 49
+farms, all collected by two people in 6 weeks. This is far more data
+from a far greater spatial extent than possible if measuring individual
+plant demography (in a world of limited time and money, at least). The
+appeal of density-structured approaches is clear. However, none of these
+models included environmental covariates.
 
@@
 models separately since fitting the models required different
-approaches. All models contain five climate covariate that we chose
+approaches. All models contain five climate covariates that we chose
 \emph{a priori}: ``water year'' precipitation at \emph{t}-1 (lagppt);
-fall through spring precipitation at \emph{t}-1 and \emph{t}-2 (ppt1 and
-ppt2, respectively) and mean spring temperature at \emph{t}-1 and
+April through June precipitation at \emph{t}-1 and \emph{t}-2 (ppt1 and
+ppt2, respectively) and April through June temperature at \emph{t}-1 and
 \emph{t}-2 (TmeanSpr1 and TmeanSpr2, respectively), where \emph{t} is
@@
 
-\todo{let's talk about timing of "spring". Miles City gets a lot of moisture in May and June...}
-
 We fit all models using a hierarchical Bayesian approach. The models are
@@
 always the relevant vector of observations for vital rate \emph{X}
-($X = S, G, R, or P for survival, growth, recruitment, or population growth). For example, $\textbf{y}\^{}S\$
-is a vector of 0s and 1s indicating whether a genet survives from
-\emph{t} to \emph{t+1}, or not.
+($X = S, G, R, or P$ for survival, growth, recruitment, or population
+growth). For example, $\textbf{y}^S$ is a vector of 0s and 1s indicating
+whether a genet survives from \emph{t} to \emph{t+1}, or not.
 
@@
 to incorporate the recruitment function in the spatially-implicit IPM.
-{[}You probably should say something about spatial structure and cite
-Adler et al. 2010 EcoLetts for the ``no overlap rule''{]}
+\todo{Add spatial structure mention.}
 
@@
 
-Our quad-based model (QBM) perfectly mirrors {[}I think you need to say
-a little more about how you simulate it{]} its statistical description
+Our quad-based model (QBM) perfectly mirrors its statistical description
 (Eqs. 8-9). We use the same approach for drawing parameter values as
-described for the IPM.
+described for the IPM. After drawing the appropriate parameter set, we
+calculate the mean response (population cover at \emph{t+1} = $x_{t+1}$)
+according to Eq. 8. We then make a random draw from a {[}0,1{]}
+truncated lognormal distribution with mean equal to $x_{t+1}$ from Eq. 8
+and the variance estimate from the fitted model. We can then iterate the
+model forward by drawing a new parameter set (unique to climate year and
+MCMC iteration) at each timestep.
 
@@
 
+Our \emph{a priori} expectation is that the IPM will produce more
+accurate and precise forecasts. This could be due to differences in
+sample sizes leading to larger parameter uncertainty for the QBM, or due
+to the QBM climate effects being weakly associated with one or more
+vital rate climate effects at the individual level. To assess the impact
+of sample size on QBM parameter uncertainty we refit the QBM statistical
+model (Eqs. 8-9) after removing sets of 2, 5, 10, and 15 quadrats. We
+fit 10 models at each level of quadrat removal (2, 5, 10, 15 quadrats),
+removing a different randomly selected set of quadrats for each fit. We
+calculated the standard deviation of climate main effects (pptlag, ppt1,
+ppt2, TmeanSpr1, and TmeanSpr2) for each model and averaged those over
+replicates within each set of quadrat removals. This allowed us to
+regress sample size against parameter uncertainty.
+
+To see if the QBM climate effects are correlated, or not, with climate
+effects for each vital rate model for the IPM, we simply regressed the
+QBM climate coefficients against each vital rate model's climate
+coefficients and calculate Pearson's $\rho$. Strong correlations
+indicate the QBM is capable of detecting climate effects associated with
+individual vital rates. A weak correlation indicates the QBM ``misses''
+the climate effect on a particular vital rate.
+
 \subsection{Results}\label{results}
@@
 sensitivity of each vital rate in isolation and in pairwise
-combinations. Species showed similar trends (Fig. 3e-h). Growth was the
-most sensitive vital rate for all species, showing a negative response
-to increased precipitation, and stronger positive response to increased
-temperature, and a mostly positive response when both climate factors
-are increased (Fig. 3e-h). \emph{B. gracilis} survival rates were
-sensitive to temperature, resulting in an increase in plant cover under
-increased temperature (Fig. 3e). In isolation, recruitment and survival
-were insensitive to climate factors for \emph{H. comata} (Fig. 3f).
-Survival and recruitment of \emph{P. smithii} were both sensitive,
-negatively, to temperature and precipitation (Fig. 3g). \emph{P.
-secunda} equilibrium cover was sensitive to the climate effects on
-survival and recruitment, showing a negative effect on both vital rates
-for increased precipition, but a strong positive effect on survival with
-increased temperature (Fig. 3h). The climate impact of recruitment on
-equilibrium cover was negative for precipitation and temperature
-increases (Fig. 3h). At least two of three vital rates were sensitive to
-climate for each species (Fig. 3).
+combinations. Across all species, climate covariates can have opposing
+effects on different vital rates (Fig. 3e-h).
+
+Growth was the most sensitive vital rate for all species, showing a
+negative response to increased precipitation, and stronger positive
+response to increased temperature, and a mostly positive response when
+both climate factors are increased (Fig. 3e-h). \emph{B. gracilis}
+survival rates were sensitive to temperature, resulting in an increase
+in plant cover under increased temperature (Fig. 3e). In isolation,
+recruitment and survival were insensitive to climate factors for
+\emph{H. comata} (Fig. 3f). Survival and recruitment of \emph{P.
+smithii} were both sensitive, negatively, to temperature and
+precipitation (Fig. 3g). \emph{P. secunda} equilibrium cover was
+sensitive to the climate effects on survival and recruitment, showing a
+negative effect on both vital rates for increased precipition, but a
+strong positive effect on survival with increased temperature (Fig. 3h).
+The climate impact of recruitment on equilibrium cover was negative for
+precipitation and temperature increases (Fig. 3h). At least two of three
+vital rates were sensitive to climate for each species (Fig. 3).
 
@@
 than IPM projections for all species except \emph{P. smithiii} (Fig. 4).
+IPM forecasts for \emph{P. smithiii} were very uncertain due to
+overcompensating density dependence and nonlinearities recruitment that
+get amplified by parameter uncertainty (Appendx C). Neither model was
+capable of making forecasts of proportional cover change distinguishable
+from 0 (i.e., no change in population state) when we included model
+error and parameter uncertainty.
+
+\subsubsection{Sources of uncertainty in the
+QBM}\label{sources-of-uncertainty-in-the-qbm}
+
+Sample size had a relatively weak effect on QBM climate parameter
+uncertainty after the number of quadrats exceeded about 10 (Fig. 5).
+Inverse-gaussian fits show that increasing sample size beyond the number
+of quadrats we used will results in deminishing returns in terms of
+parameter certainty (Fig. 5).
+
+Climate effects estimated from the QBM are most correlated with climate
+effects from the growth regression at the individual level (Fig. 6). In
+no case does the QBM statistical model have strong correlations across
+all three vital rates (Fig. 6). QBM climate effects were most weakly
+correlated with those from individual-level recruitment models for
+\emph{B. gracilis}, \emph{H. comata}, and \emph{P. secunda} (Fig.
+6a,b,d). For \emph{P. smithii}, QBM climate effects showed no
+correlation with the survival model effects (Fig. 6c).
 
@@
 
-MOVE FIRST TWO PARAGRAPHS TO INTRO. MOVE FIG 5 and 6 TO METHODS AND
-RESULTS
-
-Perhaps the greatest challenge for ecology in the 21st century is to
-forecast the impacts of environmental change (Clark et al. 2001, Petchey
-et al. 2015). To do so requires sophisticated modeling approaches that
-fully account for uncertainty and variability in the ecological process
-and associated parameters (Luo et al. 2011). This requires large amounts
-of data collected over large spatio-temporal extents.
-State-of-the-science modeling techniques cannot overcome data
-limitations. Such is the case for many population models.
-
-As a potential remedy to the ``data dearth'' problem, Queenborough et
-al. (2011) and Freckleton et al. (2011), building on work by Taylor and
-Hastings (2004), advocate a ``density-structured'' modeling approach.
-Such models do not require individual level demographic data and can
-adequately describe population dynamics (Queenborough et al. 2011). The
-results from density-structured models are not as precise as those from
-traditional population models, but the loss in precision is traded off
-with a gain in data. The study by Queenborough et al. (2011) included
-data from 500 fields (4 hectares each) in 49 farms, all collected by two
-people in 6 weeks. This is far more data from a far greater spatial
-extent than possible if measuring individual plant demography (in a
-world of limited time and money, at least). The appeal of
-density-structured approaches is clear.
-
-However, at their core, density-structured models rely on individual
-level data aggregated to a population level metric (e.g., density
-classes or percent cover). This creates a potential problem if such
-models are to be used in a climate change context because inidividuals
-respond to climate, not populations (Clark et al. 2012). Are models
-based on population level metrics as sensitive to climate as models
-based on individual level metrics? Do these two types of models produce
-consistent forecasts? Do we need detailed demographic data to forecast
-the impacts of climate change? These are the questions we sought to
-answer here.
+Population models built using individual-level data allow inference on
+demographic processes, but their ability to forecast future population
+states is limited by the spatial extent of the data. Population-level
+data, like percent cover or discrete density states, are much easier to
+collect of broad spatial extents, so populations built using such data
+offer an appealing alternative to traditional population models
+(Queenborough et al. 2011). However, at their core, density-structured
+models rely on individual level data aggregated to a population level
+metric. This creates a potential problem if such models are to be used
+in a climate change context because inidividuals respond to climate, not
+populations (Clark et al. 2012). Are models based on population level
+metrics as sensitive to climate as models based on individual level
+metrics? Do these two types of models produce consistent forecasts? Do
+we need detailed demographic data to forecast the impacts of climate
+change? These are the questions we sought to answer here.
 
@@
 
-Using individual and aggregated forms of the same dataset, we were able
-to directly compare a traditional demographic modeling approach to a
-population model based on aggregated data. Our quad-based model (QBM) is
-based on percent cover data and so is in the spirit of
-density-structured models. In terms of each model's forecasting ability,
-the IPM outperformed the QBM (Table 1). This is unsurprising since we
-expected to lose some precision at the aggregated level. However, the
-underwhelming performance of the QBM could call into the question
-forecasts that differ from the IPM.
+Using individual-level and population-level forms of the same dataset,
+we were able to directly compare a traditional demographic modeling
+approach to a population model based on percent cover data. Our
+quad-based model (QBM) is based on percent cover data and so is in the
+spirit of density-structured models. In terms of each model's
+forecasting ability, the IPM outperformed the QBM (Table 1). This is
+unsurprising since we expected to lose some precision at the aggregated
+level. However, the underwhelming performance of the QBM could call into
+the question forecasts that differ from the IPM.
 

### 2015-07-17 manuscript/manuscript.Rmd
updates ms with new structure

@@
 Here, we test whether this assumption is important when forecasting climate change impacts on four perennial grass species in a semi-arid Montana grassland.
-We parameterized two population models, one based on inidividual level data with three vital rates and one on an aggregated version of the same data (percent cover), and compared their accuracy, precision, and sensitivity to climate.
-The individual level model was more accurate and precise than the aggregate level model when predicting out of sample observations.
-The aggregate level model made countervailing forecasts to 1% climate changes when compared to the individual level model, indicating the aggregate level model produces incorrect forecasts.
-When comparing climate effects from both models, the aggregate level model tends to "miss" important climate effects from at least one vital rate for each species.
+We parameterized two population models, one based on inidividual-level data with three vital rates and one on population-level data from the same dataset (percent cover of quadrats), and compared their accuracy, precision, and sensitivity to climate.
+The individual level model was more accurate and precise than the aggregate level model when predicting out-of-sample observations.
+When comparing climate effects from both models, the population-level model tends to "miss" important climate effects from at least one vital rate for each species.
+We also find that increasing the sample size at the population-level would not necessarily reduce forecast certainty; meaning the only way to reduce uncertainty is to capture unique climate dependence on individual vital rates.
 It appears there is no short cut to forecasting climate change impacts on plant populations — detailed demographic data is essential.
-But, forecasts were very uncertain, so we advocate for a focus on new methods to collect demographic data more efficiently across environmental gradients in space and time.
+However, forecasts from our individual-level model were very uncertain, so we advocate for a focus on new methods to collect demographic data more efficiently across environmental gradients in space and time.
 
@@
 The resulting models cannot be applied to the landscape and regional scales relevant to decision-makers without information about how the fitted parameters respond to spatial variation in biotic and abiotic drivers [@Sæther2007].
-The temporal coverage of demographic datasets is no better.
-[I think the temporal extent is a separate issue, might muddy the waters to bring it up here...] For example, the most common study duration in the COMPADRE matrix population model database is 4 years and only a few studies exceed 10 years [@Salguero-Gomez2015].
-The limited spatio-temporal extent of individual-level demographic datasets constrains our ability to use population models to address applied questions about the consequences of climate change.
+The limited spatial extent of individual-level demographic datasets constrains our ability to use population models to address applied questions about the consequences of climate change.
 
@@
 Thus, it is relatively straightforward to parameterize and project SDMs over landscapes and regions, the scales at which many land-use decisions are made.
-However, the limitations of SDMs are well known [@Elith2009].
+However, the limitations of SDMs are well known [@Pearson2003; @Elith2009; @Araujo2012].
 Ideally, researchers would provide managers with landscape-scale population models, combing the extent of SDMs with information about dynamics and species abundances.
@@
 
-We assume the number of individuals, $Y^{R}$, recruiting at time $t+1$ follows a negative binomial distribution:
+We assume the number of individuals, $y^{R}$, recruiting at time $t+1$ follows a negative binomial distribution:
 
@@
 Each simulation was run under two parameter scenarios: (1) using mean parameter estimates and (2) using randomly drawn parameters from the MCMC chain.
-We use (1) to detect the overall sensitivity of equilibrium cover to climate, and we use (2) to show the impact of model uncertainty on forecast precision.
+We use (1) to detect the overall sensitivity of equilibrium cover to climate, and we use (2) to show the impact of model and parameter uncertainty on forecast precision.
 
@@
 
-Our *a priori* expectation is that the IPM will produce more accurate and precise forecasts.
+Our expectation is that the IPM will produce more accurate and precise forecasts.
 This could be due to differences in sample sizes leading to larger parameter uncertainty for the QBM, or due to the QBM climate effects being weakly associated with one or more vital rate climate effects at the individual level.
@@
 ### Sensitivity of models to climate
-```{r get_results, echo=FALSE, eval=TRUE}
+The response of a population to climate change is a result of the aggregate effects of climate on individual vital rates.
+Since the IPM approach relies on vital rate regressions, we were able to quantify the sensitivity of each vital rate in isolation and in pairwise combinations.
+Across all species, climate covariates can have opposing effects on different vital rates (Fig.
+3).
+Growth was the most sensitive vital rate for all species, showing a negative response to increased precipitation, and stronger positive response to increased temperature, and a mostly positive response when both climate factors are increased (Fig.
+3).
+*B.
+gracilis* survival rates were sensitive to temperature, resulting in an increase in plant cover under increased temperature (Fig.
+3a).
+In isolation, recruitment and survival were insensitive to climate factors for *H.
+comata* (Fig.
+3b).
+Survival and recruitment of *P.
+smithii* were both sensitive, negatively, to temperature and precipitation (Fig.
+3c).
+*P.
+secunda* equilibrium cover was sensitive to the climate effects on survival and recruitment, showing a negative effect on both vital rates for increased precipition, but a strong positive effect on survival with increased temperature (Fig.
+3d).
+The climate impact of recruitment on equilibrium cover was negative for precipitation and temperature increases (Fig.
+3d).
+At least two of three vital rates were sensitive to climate for each species (Fig.
+3).
+
+### Sources of uncertainty in the QBM
+Sample size had a relatively weak effect on QBM climate parameter uncertainty after the number of quadrats exceeded about 10 (Fig.
+5).
+Inverse-gaussian fits show that increasing sample size beyond the number of quadrats we used will results in deminishing returns in terms of parameter certainty (Fig.
+5).
+
+Climate effects estimated from the QBM are most correlated with climate effects from the growth regression at the individual level (Fig.
+6).
+In no case does the QBM statistical model have strong correlations across all three vital rates (Fig.
+6).
+QBM climate effects were most weakly correlated with those from individual-level recruitment models for *B.
+gracilis*, *H.
+comata*, and *P.
+secunda* (Fig.
+6a,b,d).
+For *P.
+smithii*, QBM climate effects showed no correlation with the survival model effects (Fig.
+6c).
+
+### Model forecasts
+Forecasts based on 1% climate changes were extremely uncertain when we considered model error and parameter uncertainty (Fig.
+6).
+As expected based on model validation (Table 1), QBM projections were more uncertain than IPM projections for all species except *P.
+smithiii* (Fig.
+6).
+IPM forecasts for *P.
+smithiii* were very uncertain due to overcompensating density dependence and nonlinearities recruitment that get amplified by parameter uncertainty (Appendx C).
+Neither model was capable of making forecasts of proportional cover change distinguishable from 0 (i.e., no change in population state) when we included model error and parameter uncertainty.
+
+```{r get_results, echo=FALSE, eval=FALSE}
 setwd("~/Repos/MicroMesoForecast/analysis/")
@@
 
+```{r comment_out, echo=FALSE, eval=FALSE}
 Equilibrium cover from both models was sensitive to climate (Fig.
@@
 3d).
-
-The response of a population to climate change is a result of the aggregate effects of climate on individual vital rates.
-Since the IPM approach relies on vital rate regressions, we were able to quantify the sensitivity of each vital rate in isolation and in pairwise combinations.
-Across all species, climate covariates can have opposing effects on different vital rates (Fig.
-3e-h).
-
-Growth was the most sensitive vital rate for all species, showing a negative response to increased precipitation, and stronger positive response to increased temperature, and a mostly positive response when both climate factors are increased (Fig.
-3e-h).
-*B.
-gracilis* survival rates were sensitive to temperature, resulting in an increase in plant cover under increased temperature (Fig.
-3e).
-In isolation, recruitment and survival were insensitive to climate factors for *H.
-comata* (Fig.
-3f).
-Survival and recruitment of *P.
-smithii* were both sensitive, negatively, to temperature and precipitation (Fig.
-3g).
-*P.
-secunda* equilibrium cover was sensitive to the climate effects on survival and recruitment, showing a negative effect on both vital rates for increased precipition, but a strong positive effect on survival with increased temperature (Fig.
-3h).
-The climate impact of recruitment on equilibrium cover was negative for precipitation and temperature increases (Fig.
-3h).
-At least two of three vital rates were sensitive to climate for each species (Fig.
-3).
-
-Forecasts based on 1% climate changes were extremely uncertain when we considered model error and parameter uncertainty (Fig.
-4).
-As expected based on model validation (Table 1), QBM projections were more uncertain than IPM projections for all species except *P.
-smithiii* (Fig.
-4).
-IPM forecasts for *P.
-smithiii* were very uncertain due to overcompensating density dependence and nonlinearities recruitment that get amplified by parameter uncertainty (Appendx C).
-Neither model was capable of making forecasts of proportional cover change distinguishable from 0 (i.e., no change in population state) when we included model error and parameter uncertainty.
-
-### Sources of uncertainty in the QBM
-Sample size had a relatively weak effect on QBM climate parameter uncertainty after the number of quadrats exceeded about 10 (Fig.
-5).
-Inverse-gaussian fits show that increasing sample size beyond the number of quadrats we used will results in deminishing returns in terms of parameter certainty (Fig.
-5).
-
-Climate effects estimated from the QBM are most correlated with climate effects from the growth regression at the individual level (Fig.
-6).
-In no case does the QBM statistical model have strong correlations across all three vital rates (Fig.
-6).
-QBM climate effects were most weakly correlated with those from individual-level recruitment models for *B.
-gracilis*, *H.
-comata*, and *P.
-secunda* (Fig.
-6a,b,d).
-For *P.
-smithii*, QBM climate effects showed no correlation with the survival model effects (Fig.
-6c).
+```
 
@@
 
-### The IPM and QBM produced inconsistent forecasts
 Using individual-level and population-level forms of the same dataset, we were able to directly compare a traditional demographic modeling approach to a population model based on percent cover data.
 Our quad-based model (QBM) is based on percent cover data and so is in the spirit of density-structured models.
-In terms of each model's forecasting ability, the IPM outperformed the QBM (Table 1).
-This is unsurprising since we expected to lose some precision at the aggregated level.
-However, the underwhelming performance of the QBM could call into the question forecasts that differ from the IPM.
-
-Indeed, when we perturbed climate factors the QBM made forecasts completely contradictory to those of the IPM for three of our four study species (Fig.
-3a-d).
-In a perfect world, the QBM would have made forecasts of at least the same direction as the IPM.
-If that had been the case we could conclude that aggregate level models could prove useful for forecasting climate change impacts on populations.
-Unfortunately, this was not the case.
-
-Given the superior ability of the IPM to predict out of sample observations (Table 1), we have no choice but to conclude it is the superior model.
-Following that logic, we can only assume that, at least contingent on the data in hand, the IPM is producing the correct forecasts to climate perturbations.
-The QBM failed to match IPM forecasts, implying that detailed demographic data may be necessary to accurately detect climate signals that are utlimately important at the population level.
-This result further confirms related work on the importance of individual variability on population level responses to exogenous drivers [@Clark2011; @Clark2011b; @Clark2012; @Galvan2014].
-
-### The role of vital rate climate dependence
-We can think of two reasons why the IPM outperformed the QBM.
-First, the quadrat level data has a much reduced sample size compared to the individual level data.
-In an ideal world we would have compared the IPM and QBM using data collected over the same amount of person hours, not just the same number of quadrats.
-Then the sample size of the quadrat level data would be much greater and carry more statistical power.
-To address this limitation in our work we fit the QBM statistical model (Eq.
-8-9) with different numbers of quadrats to see the effect of sample size on the precision of climate effect estimates.
-It appears that including additional quadrats leads to rapidly diminishing returns in terms of parameter precision (Fig.
+In terms of each model's forecasting ability, the IPM outperformed the QBM: the IPM was more accurate and precise than the QBM (Table 1).
+As we suggest in the introduction, this could be due to differences in sample size or the effect of averaging over climate responses unique to individual-level vital rates.
+
+We found no evidence that increasing sample size at the quadrat level would lead to higher precision of climate coefficient estimates (Fig.
+4).
+We did, however, find evidence that the QBM statistical model failed to identify climate dependence for some vital rates (Fig.
 5).
-Thus, while sample size surely plays some role, we do not think it is the main driver of the difference between the IPM and QBM.
-
-The second reason the IPM could have outperformed the QBM is that the population level model is in fact missing important climate effects that act on individual vital rates, rather than population growth.
-Our intuition was that species with strong climate-dependence on vital rates not well resolved at the aggregate level would result in different forecasts from the two models.
-For example, survival is very size dependent: smaller individuals have a higher probability of death [@Chu2014].
-At the same time, a single small individual contributes relatively little to percent cover estimates at the plot scale.
-So, if survival of individuals was positively impacted by temperature increases, for example, we would expect to detect this signal in the individual level data but not in the aggregate level data.
-To see if this is the case we can regress climate effects from each vital rate statistical model at the individual level against the same climate effects from the QBM statistical model (Fig.
-6).
-
-In general, the QBM climate effects are most correlated with climate effects from the growth regression at the individual level (Fig.
-6).
-In no case does the QBM statistical model have strong correlations across all three vital rates (Fig.
-6).
+For no species were climate effects from the QBM strongly correlated with all three vital rates (Fig.
+5).
+@Freckleton2011 acknowledge that averaging over complex stage dependence will lead to poorly specified models.
+This is analagous to our situation, but instead of averaging over complex life histories, we are averaging over complex climate dependence.
+
 Thus, for each species the QBM is "missing" climate signals associated with at least one vital rate.
-This has large impacts on predictions of long term population dynamics, as seen in our equilibrium simulations (Fig.
-3a-d).
-The inability of the QBM to separate the sometimes countervailing effects of climate on survival, growth, and recruitment (Fig.
-3e-h) results in inaccurate (Fig.
-3a-d) and unprecise (Fig.
-4) forecasts.
-The QBM statistical model struggles to explain variation due to climate variables because they can have positive and negative impacts on different vital rates.
+This leads to inaccurate and imprecise forecasts because the QBM statistical model struggles to explain variation due to climate variables that have positive and negative impacts on different vital rates.
 When this is the case, as it is for all our species to varying degrees (Fig.
-3e-h), statistical models of aggregated population responses will fail.
-
-These results lead us to conclude that detailed demographic data is necessary to forecast climate change impacts on plant populations.
+3), statistical models of population-level data will fail.
+This result further confirms related work on the importance of individual-level data to forecast population responses to exogenous drivers [@Clark2011; @Clark2011b; @Clark2012; @Galvan2014].
+
+These results lead us to conclude that detailed demographic data is necessary to forecast climate change impacts on plant populations when vital rates have unique climate responses.
 This is unwelcome news since this data is difficult to collect and the models built on such data are of little use to land managers that make decisions at scales beyond that of traditional population models [@Queenborough2011].
@@
 
-### Forecasting the future, and the future of forecasting
+What does all this mean for model forecasts?
+The answer is more nuanced than one might expect.
+The typical approach in ecology is to use point estimates of model parameters to project populations forward according to the specified model; usually allowing for some variability around the determinstic process (e.g., @Battin2007; @Jenouvrier2009; @Adler2012).
+If we follow tradition and calculate the mean response to climate change over many iterations of the model with only model error and temporal variation included, the IPM and the QBM produce opposing forecasts for three of four species (Fig.
+D1).
+However, if we introduce parameter uncertainty, the forecasts are actually indistinguishable (Fig.
+6).
+IPM forecasts are generally more precise, as expected from model validation (Table 1).
+The real story is not that the QBM produces "wrong" forecasts, but that it produces highly uncertain forecasts.
+Our sample size analysis reveals that this uncertainty is not easily overcome by including more data (Fig.
+4).
+The only way to reduce uncertainty is to model these populations based on individual-level data where climate dependence can be properly identified (Fig.
+5).
+Yet, even the IPM forecasts were highly variable (Fig.
+6).
+
 Our goal was not to make any explicit forecast for the future state of these populations based on predicted climate change.
@@
 Yet, model predictions proved so uncertain that any forecast, when bounded with model and parameter uncertainty, would be at best not useful and at worst meaningless.
-For all species, the 90% quantiles of predicted changes in populatoin size overlapped zero; we cannot even predict whether a change in precipitation or temperature will cause populations to increase or decrease.
+For all species, the 90% quantiles of predicted changes in population size overlapped zero; we cannot even predict whether a change in precipitation or temperature will cause populations to increase or decrease.
 How might we improve on this state of affairs?
@@
 Climate covariates only vary by year, with no spatial variability within years.
-Thus, even if we fit models to individual level data, we are missing the key interaction point between weather and individual plants [@Clark2011b] because all observations share the same climate covariates.
+Thus, even if we fit models to individual-level data, we are missing the key interaction point between weather and individual plants [@Clark2011b] because all observations share the same climate covariates.
 Demographic studies should be designed with at least plot level measurements of climate related variables (e.g., soil moisture).
@@
 However, as we have done here, it is still possible to include parameter uncertainty by drawing parameter values from MCMC iterations, taking care to draw all parameters from the same chain and iteration to account for their correlations.
-Only by being honest about our forecasts can we begin to produce better ones.
+Only by being honest about our forecasts can we begin to produce better ones, and forecasts reported without parameter error are disingenuous.
+In many cases, ignoring parameter error is justifiable when investigating basic demographic mechanisms, but it is never justifiable if forecasts are being made.
 
@@
 This work is not a critique of density-structured population models.
-In some cases and for certain species, population models based on aggregated data may prove useful and unbiased.
+In some cases and for certain species where climate dependence is weak or concentrated within a single vital rate, population models based on aggregated data may prove useful and unbiased.
 However, our work here is the first comparison, to our knowledge, of population models based on individual and aggregated forms of the same data in a climate change context.
 Our results confirm theoretical arguments [@Clark2011b] and empirical evidence [@Clark2011; @Clark2012] that individual responses are critical to predicting species' responses to climate change.
-Thus, forecasts from aggregate level models should be viewed with caution and should never be unaccompianed by uncertainty.
+Thus, forecasts from population-level models should be viewed with caution and should never be unaccompianed by uncertainty.
 Given the importance of demographic data and its current difficulty to collect, we advocate for research on modern methods to collect demographic data more efficiently across environmental gradients in space and time.
@@
 
-```{r figure_3, dependson="plot-options", fig.cap="Proportional change in species' mean cover caused by a 1% increase in observed precipitation (+ppt), temperature (+temp), or both (+ppt&temp) as predicted by the individual-based IPM and the aggregate-based QBM using mean parameter values.
+```{r figure_3, dependson="plot-options", fig.cap="Sensitivity of equilibrium cover simulated from the IPM to each climate scenario applied to individual and combined vital rates.
+For example, the points associated with G show the median cover from IPM simulations where a climate perturbation is applied only to the growth regression climate covariates.
+These simulations use mean parameter values for clarity.", fig.width=4.5, fig.height=8, cache=FALSE, results='hide'}
+setwd("~/Repos/MicroMesoForecast/analysis/")
+#Proportional change in species' mean cover caused by a 1% increase in observed precipitation (+ppt), temperature (+temp), or both (+ppt&temp) as predicted by the individual-based IPM and the aggregate-based QBM using mean parameter values.
 Top panels show the mean predicted proportional change in cover.
-Lower panels show the sensitivity of equilibrium cover simulated from the IPM to each climate scenario applied to individual and combined vital rates.
-For example, the points associated with G show the median cover from IPM simulations where a climate perturbation is applied only to the growth regression climate covariates.
-These simulations also use mean parameter values for clarity.", fig.width=8.5, fig.height=5, cache=FALSE, results='hide'}
-setwd("~/Repos/MicroMesoForecast/analysis/")
-
+Lower panels show the
 ####
@@
 ####
-meanparam_ipm <- data.frame(cover=NA, species=NA, 
-                             climsim=NA, model=NA)
-##  IPM 
-path <- "./ipm/simulations/results/climchange_meanparams/"
-ipmfiles_mean <- list.files(path)
-ipmfiles_mean <- ipmfiles_mean[grep("cover", ipmfiles_mean)]
-for(file in ipmfiles_mean){
-  tmp <- read.csv(paste(path,file,sep=""))
-  tmp$model <- "IPM"
-  meanparam_ipm <- rbind(meanparam_ipm, tmp[,c("cover", "species", "climsim", "model")])
-}
-meanparam_ipm <- meanparam_ipm[2:nrow(meanparam_ipm),]
-
-newd <- data.frame(cover=NA, species=NA, 
-                   climsim=NA, model=NA, obscover=NA)
-for(dospp in unique(meanparam_ipm$species)){
-  tmp <- subset(meanparam_ipm, species==dospp)
-  obscover <- subset(tmp, climsim=="obs")["cover"]
-  tmp2 <- subset(tmp, climsim!="obs")
-  tmp2$obscover <- rep(obscover$cover, times = length(unique(tmp2$climsim)))
-  newd <- rbind(newd, tmp2)
-}
-ipm_meansims <- newd[2:nrow(newd),]
-ipm_meansims$propdiff <- with(ipm_meansims, log(cover*100)-log(obscover*100))
-
-##  QBM
-meanparam_qbm <- data.frame(cover=NA, species=NA, 
-                             climsim=NA, model=NA)
-path <- "./quadBM/simulations/results/climatechange_meanparams/"
-qbmfiles_mean <- list.files(path)
-for(file in qbmfiles_mean){
-  tmp <- readRDS(paste(path,file,sep=""))
-  tmp$model <- "QBM"
-  meanparam_qbm <- rbind(meanparam_qbm, tmp[,c("cover", "species", "climsim", "model")])
-}
-meanparam_qbm <- meanparam_qbm[2:nrow(meanparam_qbm),]
-
-newd <- data.frame(cover=NA, species=NA, 
-                   climsim=NA, model=NA, obscover=NA)
-for(dospp in unique(meanparam_qbm$species)){
-  tmp <- subset(meanparam_qbm, species==dospp)
-  obscover <- subset(tmp, climsim=="obs")["cover"]
-  tmp2 <- subset(tmp, climsim!="obs")
-  tmp2$obscover <- rep(obscover$cover, times = length(unique(tmp2$climsim)))
-  newd <- rbind(newd, tmp2)
-}
-qbm_meansims <- newd[2:nrow(newd),]
-qbm_meansims$propdiff <- with(qbm_meansims, log(cover*100)-log(obscover*100))
-
-##  Merge IPM and QBM mean parameter results
-bothmodel_outs <- rbind(ipm_meansims, qbm_meansims)
-meanparam_agg <- ddply(bothmodel_outs, .(species, climsim, model), summarise,
-                       meandiff = mean(propdiff),
-                       hidiff = quantile(propdiff, probs = 0.95),
-                       lodiff = quantile(propdiff, probs = 0.05))
-dgd = position_dodge(width = 0.9)
-meanplot <- ggplot(meanparam_agg)+
-  geom_bar(aes(x=climsim, y=meandiff, fill=model),
-             position=dgd, stat="identity")+
-  theme_bw()+
-  scale_fill_manual(values=c("black", "grey45"), name="Model")+
-  facet_wrap("species", scales="free", nrow=1)+
-  scale_x_discrete(labels=c("+ppt", "+temp", "+ppt&temp"))+
-  ylab("Mean proportional difference")+
-  xlab("Climate simulation")+
-  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+# meanparam_ipm <- data.frame(cover=NA, species=NA, 
+#                              climsim=NA, model=NA)
+# ##  IPM 
+# path <- "./ipm/simulations/results/climchange_meanparams/"
+# ipmfiles_mean <- list.files(path)
+# ipmfiles_mean <- ipmfiles_mean[grep("cover", ipmfiles_mean)]
+# for(file in ipmfiles_mean){
+#   tmp <- read.csv(paste(path,file,sep=""))
+#   tmp$model <- "IPM"
+#   meanparam_ipm <- rbind(meanparam_ipm, tmp[,c("cover", "species", "climsim", "model")])
+# }
+# meanparam_ipm <- meanparam_ipm[2:nrow(meanparam_ipm),]
+# 
+# newd <- data.frame(cover=NA, species=NA, 
+#                    climsim=NA, model=NA, obscover=NA)
+# for(dospp in unique(meanparam_ipm$species)){
+#   tmp <- subset(meanparam_ipm, species==dospp)
+#   obscover <- subset(tmp, climsim=="obs")["cover"]
+#   tmp2 <- subset(tmp, climsim!="obs")
+#   tmp2$obscover <- rep(obscover$cover, times = length(unique(tmp2$climsim)))
+#   newd <- rbind(newd, tmp2)
+# }
+# ipm_meansims <- newd[2:nrow(newd),]
+# ipm_meansims$propdiff <- with(ipm_meansims, log(cover*100)-log(obscover*100))
+# 
+# ##  QBM
+# meanparam_qbm <- data.frame(cover=NA, species=NA, 
+#                              climsim=NA, model=NA)
+# path <- "./quadBM/simulations/results/climatechange_meanparams/"
+# qbmfiles_mean <- list.files(path)
+# for(file in qbmfiles_mean){
+#   tmp <- readRDS(paste(path,file,sep=""))
+#   tmp$model <- "QBM"
+#   meanparam_qbm <- rbind(meanparam_qbm, tmp[,c("cover", "species", "climsim", "model")])
+# }
+# meanparam_qbm <- meanparam_qbm[2:nrow(meanparam_qbm),]
+# 
+# newd <- data.frame(cover=NA, species=NA, 
+#                    climsim=NA, model=NA, obscover=NA)
+# for(dospp in unique(meanparam_qbm$species)){
+#   tmp <- subset(meanparam_qbm, species==dospp)
+#   obscover <- subset(tmp, climsim=="obs")["cover"]
+#   tmp2 <- subset(tmp, climsim!="obs")
+#   tmp2$obscover <- rep(obscover$cover, times = length(unique(tmp2$climsim)))
+#   newd <- rbind(newd, tmp2)
+# }
+# qbm_meansims <- newd[2:nrow(newd),]
+# qbm_meansims$propdiff <- with(qbm_meansims, log(cover*100)-log(obscover*100))
+# 
+# ##  Merge IPM and QBM mean parameter results
+# bothmodel_outs <- rbind(ipm_meansims, qbm_meansims)
+# meanparam_agg <- ddply(bothmodel_outs, .(species, climsim, model), summarise,
+#                        meandiff = mean(propdiff),
+#                        hidiff = quantile(propdiff, probs = 0.95),
+#                        lodiff = quantile(propdiff, probs = 0.05))
+# dgd = position_dodge(width = 0.9)
+# meanplot <- ggplot(meanparam_agg)+
+#   geom_bar(aes(x=climsim, y=meandiff, fill=model),
+#              position=dgd, stat="identity")+
+#   theme_bw()+
+#   scale_fill_manual(values=c("black", "grey45"), name="Model")+
+#   facet_wrap("species", scales="free", nrow=1)+
+#   scale_x_discrete(labels=c("+ppt", "+temp", "+ppt&temp"))+
+#   ylab("Mean proportional difference")+
+#   xlab("Climate simulation")+
+#   theme(axis.text.x = element_text(angle = 45, hjust = 1))
 
@@
   geom_point(size=4)+
-  facet_wrap("species", scales = "free", nrow=1)+
+  facet_wrap("species", scales = "free", ncol=1)+
   scale_linetype_discrete(name="Vital Rate")+
@@
 
-meanparamplot <- grid.arrange(meanplot, sensplot, ncol=1, nrow=2)
-print(meanparamplot)
+# meanparamplot <- grid.arrange(meanplot, sensplot, ncol=1, nrow=2)
+# print(meanparamplot)
+print(sensplot)
 
@@
 
-```{r figure_4, dependson="plot-options", fig.cap="Equilibrium cover and 90% quantiles around the mean prediction when model error and parameter uncertainty are propogated through the simulation phase.
+```{r figure_4, dependson="plot-options", fig.cap="Effect of quadrat sample size on the precision (standard deviation) of main climate effect estimates in the QBM.
+Increasing the number of quadrats results in diminishing returns in terms of parameter certainty.
+Light dashed lines show individual climate effects at five quadrat sample sizes.
+Thick dark lines are inverse gaussian fits showing the mean effect of increasing quadrat sample size on parameter precision.", fig.width=3, fig.height=7.5, cache=FALSE, results='hide'}
+library(mgcv)
+
+setwd("../analysis/quadBM/vitalRateRegressions/exclude_quad_test")
+
+####
+####  Read in model fits and summarize results
+####
+source("popgrowth_read_data.R")
+spp <- unique(growD_all$Species)
+num_quads <- numeric(length(spp))
+for(i in 1:length(spp)){
+  tmp <- subset(growD_all, Species==spp[i])
+  num_quads[i] <- length(unique(tmp$quad))
+}
+num_quads <- as.data.frame(num_quads)
+num_quads$species <- spp
+
+all_files <- list.files("./results/")
+all_fits <- data.frame(Parameter=NA, stddev=NA, id=NA, 
+                       rmsquad=NA, rep=NA, species=NA)
+for(i in 1:length(all_files)){
+  tmp <- all_files[i]
+  spp <- substring(tmp, 1, 4)
+  qsrm1 <- unlist(strsplit(tmp, "[_]"))[2]
+  qsrm <- gsub("numout", "", qsrm1)
+  rep1 <- unlist(strsplit(tmp, "[_]"))[3]
+  rep2 <- gsub(".RDS","",rep1)
+  rep <- gsub("rep", "", rep2)
+
+  long <- readRDS(paste("./results/",tmp, sep=""))
+  short <- ddply(long, .(Parameter), summarise,
+                 stddev = sd(value))
+  climeff <- short[grep("b2", short$Parameter),]
+  climeff$id <- substr(climeff$Parameter, 4, length(climeff$Parameter))
+  climeff$id <- as.numeric(unlist(strsplit(climeff$id, split=']')))  
+  climeff <- climeff[with(climeff, order(id)),]
+  climeff$rmsquad <- qsrm
+  climeff$rep <- rep
+  climeff$species <- spp
+  all_fits <- rbind(all_fits, climeff)
+} # end file loop
+
+all_fits <- all_fits[2:nrow(all_fits),]
+
+####
+####  Read in full fits
+####
+dir <- "../truncNormModel/"
+files <- list.files(dir)
+files <- files[grep(".RDS", files)]
+all_alls <- data.frame(Parameter=NA, stddev=NA, id=NA, 
+                       rmsquad=NA, rep=NA, species=NA)
+for(i in 1:length(files)){
+  tmp <- files[i]
+  spp1 <- unlist(strsplit(tmp, "[_]"))[3]
+  spp <- gsub(".RDS","",spp1)
+
+  long <- readRDS(paste(dir,tmp, sep=""))
+  short <- ddply(long, .(Parameter), summarise,
+                 stddev = sd(value))
+  climeff <- short[grep("b2", short$Parameter),]
+  climeff$id <- substr(climeff$Parameter, 4, length(climeff$Parameter))
+  climeff$id <- as.numeric(unlist(strsplit(climeff$id, split=']')))  
+  climeff <- climeff[with(climeff, order(id)),]
+  climeff$rmsquad <- 0
+  climeff$rep <- 1
+  climeff$species <- spp
+  all_alls <- rbind(all_alls, climeff)
+}
+all_alls <- all_alls[2:nrow(all_alls),]
+
+####
+####  Aggregate results by species and rep
+####
+all_fits <- rbind(all_fits, all_alls)
+agg_fits <- ddply(all_fits, .(Parameter, rmsquad, species, id), summarise,
+                  avg_stddev = mean(stddev))
+agg_fits2 <- agg_fits[which(agg_fits$id %in% c(1:5)),]
+agg_fits2 <- merge(agg_fits2, num_quads, by="species")
+agg_fits2$quadsfit <- with(agg_fits2, num_quads-as.numeric(rmsquad))
+
+####
+####  Fit model (need to do by species)
+####
+modBOGR <- gam(avg_stddev~quadsfit, data=subset(agg_fits2, species=="BOGR"),
+               family=quasi(link="1/mu^2"))
+modHECO <- gam(avg_stddev~quadsfit, data=subset(agg_fits2, species=="HECO"),
+               family=quasi(link="1/mu^2"))
+modPASM <- gam(avg_stddev~quadsfit, data=subset(agg_fits2, species=="PASM"),
+               family=quasi(link="1/mu^2"))
+modPOSE <- gam(avg_stddev~quadsfit, data=subset(agg_fits2, species=="POSE"),
+               family=quasi(link="1/mu^2"))
+summary(modBOGR)
+summary(modHECO)
+summary(modPASM)
+summary(modPOSE)
+
+####
+#### Plot
+####
+ggplot(agg_fits2, aes(x=quadsfit, y=avg_stddev))+
+  geom_line(aes(group=Parameter), alpha=0.3, linetype=2)+
+  geom_point(aes(group=Parameter), alpha=0.3, size=3)+
+#   stat_smooth(method="loess", size=2, color="black", se=FALSE)+
+  geom_smooth(method="glm", formula=y~x, family=quasi(link="1/mu^2"),
+              size=2, color="black", se=FALSE)+
+  facet_wrap("species", scales = "free", ncol=1)+
+  theme_bw()+
+  xlab("Number of quadrats fit")+
+  ylab("Standard deviation of coefficient")
+
+```
+
+```{r figure_5, dependson="plot-options", fig.cap="Correlations (r) between QBM and IPM estimates of climate effects.
+We ignore sizeXclimate interactions since these are not directly comparable across model types.
+The QBM does not have multiple vital rates, so its values are repeated across panels within each species.
+Across top panels, 'growth' = growth regression, 'rec' = recruitment regression, 'surv' = survival regression.", fig.width=6, fig.height=6, cache=FALSE, results='hide'}
+##  Figure script to look at correlations between IPM and QBM
+##    climate effect estimates.
+
+library(ggmcmc)
+library(gridExtra)
+library(ggplot2)
+
+allfiles <- list.files("../analysis/speciesData/")
+removals <- grep("csv", allfiles)
+species_list <- allfiles[-removals]
+
+clim_covs <- c("pptLag", "ppt1", "ppt2", "TmeanSpr1", "TmeanSpr2",
+               "ppt1xTmeanSpr1", "ppt2xTmeanSpr2",
+               "sizeXpptLag", "sizeXppt1", "sizeXppt2",
+               "sizeXTmeanSpr1", "sizeXTmeanSpr2")
+clim_mains <- c("pptLag", "ppt1", "ppt2", "TmeanSpr1", "TmeanSpr2",
+                "ppt1xTmeanSpr1", "ppt2xTmeanSpr2")
+
+####
+##  Get mean growth climate effects
+####
+growth_parms <- data.frame(species=NA, model=NA, parameter=NA, value=NA)
+for(spp in species_list){
+  tmp_data <- readRDS(paste("../analysis/ipm/vitalRateRegs/growth/growth_stanmcmc_", spp, ".RDS", sep=""))
+  tmp_clim <- tmp_data[grep("b2", tmp_data[,"Parameter"]), ]
+  tmp_clim[,"Parameter"] <- rep(clim_covs, each=3000)
+  tmp_agg <- ddply(tmp_clim, .(Parameter), summarise,
+                   value = median(value))
+  tmp_agg2 <- tmp_agg[which(tmp_agg$Parameter %in% clim_mains), ]
+  tmp_agg2$species <- spp
+  tmp_agg2$model <- "ipm"
+  colnames(tmp_agg2) <- tolower(colnames(tmp_agg2))
+  growth_parms <- rbind(growth_parms,
+                        tmp_agg2[,c("species", "model", "parameter", "value")])
+}
+
+####
+##  Get mean survival climate effects
+####
+surv_parms <- data.frame(species=NA, model=NA, parameter=NA, value=NA)
+params <- character(length = 12)
+for(i in 1:12){
+  params[i] <- paste("b2[",i,"]", sep="")
+}
+for(spp in species_list){
+  tmp_data <- readRDS(paste("../analysis/ipm/vitalRateRegs/survival/survival_stanmcmc_", spp, ".RDS", sep=""))
+  keeps <- which(tmp_data$Parameter %in% params)
+  tmp_clim <- tmp_data[keeps, ]
+  tmp_clim[,"Parameter"] <- rep(clim_covs, each=3000)
+  tmp_agg <- ddply(tmp_clim, .(Parameter), summarise,
+                   value = median(value))
+  tmp_agg2 <- tmp_agg[which(tmp_agg$Parameter %in% clim_mains), ]
+  tmp_agg2$species <- spp
+  tmp_agg2$model <- "ipm"
+  colnames(tmp_agg2) <- tolower(colnames(tmp_agg2))
+  surv_parms <- rbind(surv_parms,
+                        tmp_agg2[,c("species", "model", "parameter", "value")])
+}
+
+####
+##  Get mean recruitment climate effects
+####
+rec_parms <- data.frame(species=NA, model=NA, parameter=NA, value=NA)
+params <- character(length = 12)
+for(i in 1:12){
+  params[i] <- paste("b2[",i,"]", sep="")
+}
+for(spp in species_list){
+  tmp_data <- readRDS(paste("../analysis/ipm/vitalRateRegs/recruitment/recruitment_stanmcmc_", spp, ".RDS", sep=""))
+  keeps <- which(tmp_data$Parameter %in% params)
+  tmp_clim <- tmp_data[keeps, ]
+  tmp_clim[,"Parameter"] <- rep(clim_mains, each=3000)
+  tmp_agg <- ddply(tmp_clim, .(Parameter), summarise,
+                   value = median(value))
+  tmp_agg2 <- tmp_agg[which(tmp_agg$Parameter %in% clim_mains), ]
+  tmp_agg2$species <- spp
+  tmp_agg2$model <- "ipm"
+  colnames(tmp_agg2) <- tolower(colnames(tmp_agg2))
+  rec_parms <- rbind(rec_parms,
+                      tmp_agg2[,c("species", "model", "parameter", "value")])
+}
+
+####
+##  Get QBM growth climate estimates
+####
+clim_covs2 <- c("pptLag", "ppt1", "ppt2", "TmeanSpr1", "TmeanSpr2",
+               "ppt1xTmeanSpr1", "ppt2xTmeanSpr2",
+               "coverXpptLag", "coverXppt1", "coverXppt2",
+               "coverXTmeanSpr1", "coverXTmeanSpr2")
+
+qbm_parms <- data.frame(species=NA, model=NA, parameter=NA, qbmvalue=NA)
+params <- character(length = 12)
+for(i in 1:12){
+  params[i] <- paste("b2[",i,"]", sep="")
+}
+for(spp in species_list){
+  tmp_data <- readRDS(paste("../analysis/quadBM/vitalRateRegressions/truncNormModel/popgrowth_stanmcmc_", spp, ".RDS", sep=""))
+  keeps <- which(tmp_data$Parameter %in% params)
+  tmp_clim <- tmp_data[keeps, ]
+  tmp_clim[,"Parameter"] <- rep(clim_covs2, each=3000)
+  tmp_agg <- ddply(tmp_clim, .(Parameter), summarise,
+                   qbmvalue = median(value))
+  tmp_agg2 <- tmp_agg[which(tmp_agg$Parameter %in% clim_mains), ]
+  tmp_agg2$species <- spp
+  tmp_agg2$model <- "qbm"
+  colnames(tmp_agg2) <- tolower(colnames(tmp_agg2))
+  qbm_parms <- rbind(qbm_parms,
+                     tmp_agg2[,c("species", "model", "parameter", "qbmvalue")])
+}
+
+####
+##  Format data, combine, and plot
+####
+growth_parms <- growth_parms[2:nrow(growth_parms), ]
+growth_parms$vitalrate <- "growth"
+surv_parms <- surv_parms[2:nrow(surv_parms), ]
+surv_parms$vitalrate <- "surv"
+rec_parms <- rec_parms[2:nrow(rec_parms), ]
+rec_parms$vitalrate <- "rec"
+ipm_parms <- rbind(growth_parms, surv_parms, rec_parms)
+
+qbm_parms <- qbm_parms[2:nrow(qbm_parms), ]
+qbm_parms$vitalrate <- "growth"
+vitals <- c("surv", "rec")
+qbmalone <- qbm_parms
+for(dov in vitals){
+  tmp <- qbmalone
+  tmp$vitalrate <- dov
+  qbm_parms <- rbind(qbm_parms, tmp)
+}
+
+ipm_parms <- ipm_parms[,c("species","parameter","value","vitalrate")]
+qbm_parms <- qbm_parms[,c("species","parameter","qbmvalue","vitalrate")]
+all_ests <- merge(ipm_parms, qbm_parms, by = c("species", "parameter", "vitalrate"))
+
+# Calculate correlation for each group
+cors <- ddply(all_ests, c("species", "vitalrate"), summarise, 
+              cor = round(cor(value, qbmvalue), 2))
+
+ggplot(all_ests, aes(x=qbmvalue, y=value))+
+  geom_point(size=3)+
+  geom_smooth(method="lm", color="black", fill="grey", se=FALSE)+
+  facet_grid(species~vitalrate)+
+  geom_text(data=cors, aes(label=paste("r = ", cor, sep="")), x=-1, y=1, size=4)+
+  xlab("QBM Estimate")+
+  ylab("IPM Estimate")+
+  theme_bw()
+
+setwd("~/Repos/MicroMesoForecast/manuscript/")
+```
+
+```{r figure_6, dependson="plot-options", fig.cap="Equilibrium cover and 90% quantiles around the mean prediction when model error and parameter uncertainty are propogated through the simulation phase.
 Climate simulations are as in Figure 3.", fig.width=4.5, fig.height=7, cache=FALSE, results='hide'}
@@
 
-```{r figure_5, dependson="plot-options", fig.cap="Effect of quadrat sample size on the precision (standard deviation) of main climate effect estimates in the QBM.
-Increasing the number of quadrats results in diminishing returns in terms of parameter certainty.
-Light dashed lines show individual climate effects at five quadrat sample sizes.
-Thick dark lines are inverse gaussian fits showing the mean effect of increasing quadrat sample size on parameter precision.", fig.width=3, fig.height=7.5, cache=FALSE, results='hide'}
-library(mgcv)
-
-setwd("../analysis/quadBM/vitalRateRegressions/exclude_quad_test")
-
-####
-####  Read in model fits and summarize results
-####
-source("popgrowth_read_data.R")
-spp <- unique(growD_all$Species)
-num_quads <- numeric(length(spp))
-for(i in 1:length(spp)){
-  tmp <- subset(growD_all, Species==spp[i])
-  num_quads[i] <- length(unique(tmp$quad))
-}
-num_quads <- as.data.frame(num_quads)
-num_quads$species <- spp
-
-all_files <- list.files("./results/")
-all_fits <- data.frame(Parameter=NA, stddev=NA, id=NA, 
-                       rmsquad=NA, rep=NA, species=NA)
-for(i in 1:length(all_files)){
-  tmp <- all_files[i]
-  spp <- substring(tmp, 1, 4)
-  qsrm1 <- unlist(strsplit(tmp, "[_]"))[2]
-  qsrm <- gsub("numout", "", qsrm1)
-  rep1 <- unlist(strsplit(tmp, "[_]"))[3]
-  rep2 <- gsub(".RDS","",rep1)
-  rep <- gsub("rep", "", rep2)
-
-  long <- readRDS(paste("./results/",tmp, sep=""))
-  short <- ddply(long, .(Parameter), summarise,
-                 stddev = sd(value))
-  climeff <- short[grep("b2", short$Parameter),]
-  climeff$id <- substr(climeff$Parameter, 4, length(climeff$Parameter))
-  climeff$id <- as.numeric(unlist(strsplit(climeff$id, split=']')))  
-  climeff <- climeff[with(climeff, order(id)),]
-  climeff$rmsquad <- qsrm
-  climeff$rep <- rep
-  climeff$species <- spp
-  all_fits <- rbind(all_fits, climeff)
-} # end file loop
-
-all_fits <- all_fits[2:nrow(all_fits),]
-
-####
-####  Read in full fits
-####
-dir <- "../truncNormModel/"
-files <- list.files(dir)
-files <- files[grep(".RDS", files)]
-all_alls <- data.frame(Parameter=NA, stddev=NA, id=NA, 
-                       rmsquad=NA, rep=NA, species=NA)
-for(i in 1:length(files)){
-  tmp <- files[i]
-  spp1 <- unlist(strsplit(tmp, "[_]"))[3]
-  spp <- gsub(".RDS","",spp1)
-
-  long <- readRDS(paste(dir,tmp, sep=""))
-  short <- ddply(long, .(Parameter), summarise,
-                 stddev = sd(value))
-  climeff <- short[grep("b2", short$Parameter),]
-  climeff$id <- substr(climeff$Parameter, 4, length(climeff$Parameter))
-  climeff$id <- as.numeric(unlist(strsplit(climeff$id, split=']')))  
-  climeff <- climeff[with(climeff, order(id)),]
-  climeff$rmsquad <- 0
-  climeff$rep <- 1
-  climeff$species <- spp
-  all_alls <- rbind(all_alls, climeff)
-}
-all_alls <- all_alls[2:nrow(all_alls),]
-
-####
-####  Aggregate results by species and rep
-####
-all_fits <- rbind(all_fits, all_alls)
-agg_fits <- ddply(all_fits, .(Parameter, rmsquad, species, id), summarise,
-                  avg_stddev = mean(stddev))
-agg_fits2 <- agg_fits[which(agg_fits$id %in% c(1:5)),]
-agg_fits2 <- merge(agg_fits2, num_quads, by="species")
-agg_fits2$quadsfit <- with(agg_fits2, num_quads-as.numeric(rmsquad))
-
-####
-####  Fit model (need to do by species)
-####
-modBOGR <- gam(avg_stddev~quadsfit, data=subset(agg_fits2, species=="BOGR"),
-               family=quasi(link="1/mu^2"))
-modHECO <- gam(avg_stddev~quadsfit, data=subset(agg_fits2, species=="HECO"),
-               family=quasi(link="1/mu^2"))
-modPASM <- gam(avg_stddev~quadsfit, data=subset(agg_fits2, species=="PASM"),
-               family=quasi(link="1/mu^2"))
-modPOSE <- gam(avg_stddev~quadsfit, data=subset(agg_fits2, species=="POSE"),
-               family=quasi(link="1/mu^2"))
-summary(modBOGR)
-summary(modHECO)
-summary(modPASM)
-summary(modPOSE)
-
-####
-#### Plot
-####
-ggplot(agg_fits2, aes(x=quadsfit, y=avg_stddev))+
-  geom_line(aes(group=Parameter), alpha=0.3, linetype=2)+
-  geom_point(aes(group=Parameter), alpha=0.3, size=3)+
-#   stat_smooth(method="loess", size=2, color="black", se=FALSE)+
-  geom_smooth(method="glm", formula=y~x, family=quasi(link="1/mu^2"),
-              size=2, color="black", se=FALSE)+
-  facet_wrap("species", scales = "free", ncol=1)+
-  theme_bw()+
-  xlab("Number of quadrats fit")+
-  ylab("Standard deviation of coefficient")
-
-```
-
-```{r figure_6, dependson="plot-options", fig.cap="Correlations (r) between QBM and IPM estimates of climate effects.
-We ignore sizeXclimate interactions since these are not directly comparable across model types.
-The QBM does not have multiple vital rates, so its values are repeated across panels within each species.
-Across top panels, 'growth' = growth regression, 'rec' = recruitment regression, 'surv' = survival regression.", fig.width=6, fig.height=6, cache=FALSE, results='hide'}
-##  Figure script to look at correlations between IPM and QBM
-##    climate effect estimates.
-
-library(ggmcmc)
-library(gridExtra)
-library(ggplot2)
-
-allfiles <- list.files("../analysis/speciesData/")
-removals <- grep("csv", allfiles)
-species_list <- allfiles[-removals]
-
-clim_covs <- c("pptLag", "ppt1", "ppt2", "TmeanSpr1", "TmeanSpr2",
-               "ppt1xTmeanSpr1", "ppt2xTmeanSpr2",
-               "sizeXpptLag", "sizeXppt1", "sizeXppt2",
-               "sizeXTmeanSpr1", "sizeXTmeanSpr2")
-clim_mains <- c("pptLag", "ppt1", "ppt2", "TmeanSpr1", "TmeanSpr2",
-                "ppt1xTmeanSpr1", "ppt2xTmeanSpr2")
-
-####
-##  Get mean growth climate effects
-####
-growth_parms <- data.frame(species=NA, model=NA, parameter=NA, value=NA)
-for(spp in species_list){
-  tmp_data <- readRDS(paste("../analysis/ipm/vitalRateRegs/growth/growth_stanmcmc_", spp, ".RDS", sep=""))
-  tmp_clim <- tmp_data[grep("b2", tmp_data[,"Parameter"]), ]
-  tmp_clim[,"Parameter"] <- rep(clim_covs, each=3000)
-  tmp_agg <- ddply(tmp_clim, .(Parameter), summarise,
-                   value = median(value))
-  tmp_agg2 <- tmp_agg[which(tmp_agg$Parameter %in% clim_mains), ]
-  tmp_agg2$species <- spp
-  tmp_agg2$model <- "ipm"
-  colnames(tmp_agg2) <- tolower(colnames(tmp_agg2))
-  growth_parms <- rbind(growth_parms,
-                        tmp_agg2[,c("species", "model", "parameter", "value")])
-}
-
-####
-##  Get mean survival climate effects
-####
-surv_parms <- data.frame(species=NA, model=NA, parameter=NA, value=NA)
-params <- character(length = 12)
-for(i in 1:12){
-  params[i] <- paste("b2[",i,"]", sep="")
-}
-for(spp in species_list){
-  tmp_data <- readRDS(paste("../analysis/ipm/vitalRateRegs/survival/survival_stanmcmc_", spp, ".RDS", sep=""))
-  keeps <- which(tmp_data$Parameter %in% params)
-  tmp_clim <- tmp_data[keeps, ]
-  tmp_clim[,"Parameter"] <- rep(clim_covs, each=3000)
-  tmp_agg <- ddply(tmp_clim, .(Parameter), summarise,
-                   value = median(value))
-  tmp_agg2 <- tmp_agg[which(tmp_agg$Parameter %in% clim_mains), ]
-  tmp_agg2$species <- spp
-  tmp_agg2$model <- "ipm"
-  colnames(tmp_agg2) <- tolower(colnames(tmp_agg2))
-  surv_parms <- rbind(surv_parms,
-                        tmp_agg2[,c("species", "model", "parameter", "value")])
-}
-
-####
-##  Get mean recruitment climate effects
-####
-rec_parms <- data.frame(species=NA, model=NA, parameter=NA, value=NA)
-params <- character(length = 12)
-for(i in 1:12){
-  params[i] <- paste("b2[",i,"]", sep="")
-}
-for(spp in species_list){
-  tmp_data <- readRDS(paste("../analysis/ipm/vitalRateRegs/recruitment/recruitment_stanmcmc_", spp, ".RDS", sep=""))
-  keeps <- which(tmp_data$Parameter %in% params)
-  tmp_clim <- tmp_data[keeps, ]
-  tmp_clim[,"Parameter"] <- rep(clim_mains, each=3000)
-  tmp_agg <- ddply(tmp_clim, .(Parameter), summarise,
-                   value = median(value))
-  tmp_agg2 <- tmp_agg[which(tmp_agg$Parameter %in% clim_mains), ]
-  tmp_agg2$species <- spp
-  tmp_agg2$model <- "ipm"
-  colnames(tmp_agg2) <- tolower(colnames(tmp_agg2))
-  rec_parms <- rbind(rec_parms,
-                      tmp_agg2[,c("species", "model", "parameter", "value")])
-}
-
-####
-##  Get QBM growth climate estimates
-####
-clim_covs2 <- c("pptLag", "ppt1", "ppt2", "TmeanSpr1", "TmeanSpr2",
-               "ppt1xTmeanSpr1", "ppt2xTmeanSpr2",
-               "coverXpptLag", "coverXppt1", "coverXppt2",
-               "coverXTmeanSpr1", "coverXTmeanSpr2")
-
-qbm_parms <- data.frame(species=NA, model=NA, parameter=NA, qbmvalue=NA)
-params <- character(length = 12)
-for(i in 1:12){
-  params[i] <- paste("b2[",i,"]", sep="")
-}
-for(spp in species_list){
-  tmp_data <- readRDS(paste("../analysis/quadBM/vitalRateRegressions/truncNormModel/popgrowth_stanmcmc_", spp, ".RDS", sep=""))
-  keeps <- which(tmp_data$Parameter %in% params)
-  tmp_clim <- tmp_data[keeps, ]
-  tmp_clim[,"Parameter"] <- rep(clim_covs2, each=3000)
-  tmp_agg <- ddply(tmp_clim, .(Parameter), summarise,
-                   qbmvalue = median(value))
-  tmp_agg2 <- tmp_agg[which(tmp_agg$Parameter %in% clim_mains), ]
-  tmp_agg2$species <- spp
-  tmp_agg2$model <- "qbm"
-  colnames(tmp_agg2) <- tolower(colnames(tmp_agg2))
-  qbm_parms <- rbind(qbm_parms,
-                     tmp_agg2[,c("species", "model", "parameter", "qbmvalue")])
-}
-
-####
-##  Format data, combine, and plot
-####
-growth_parms <- growth_parms[2:nrow(growth_parms), ]
-growth_parms$vitalrate <- "growth"
-surv_parms <- surv_parms[2:nrow(surv_parms), ]
-surv_parms$vitalrate <- "surv"
-rec_parms <- rec_parms[2:nrow(rec_parms), ]
-rec_parms$vitalrate <- "rec"
-ipm_parms <- rbind(growth_parms, surv_parms, rec_parms)
-
-qbm_parms <- qbm_parms[2:nrow(qbm_parms), ]
-qbm_parms$vitalrate <- "growth"
-vitals <- c("surv", "rec")
-qbmalone <- qbm_parms
-for(dov in vitals){
-  tmp <- qbmalone
-  tmp$vitalrate <- dov
-  qbm_parms <- rbind(qbm_parms, tmp)
-}
-
-ipm_parms <- ipm_parms[,c("species","parameter","value","vitalrate")]
-qbm_parms <- qbm_parms[,c("species","parameter","qbmvalue","vitalrate")]
-all_ests <- merge(ipm_parms, qbm_parms, by = c("species", "parameter", "vitalrate"))
-
-# Calculate correlation for each group
-cors <- ddply(all_ests, c("species", "vitalrate"), summarise, 
-              cor = round(cor(value, qbmvalue), 2))
-
-ggplot(all_ests, aes(x=qbmvalue, y=value))+
-  geom_point(size=3)+
-  geom_smooth(method="lm", color="black", fill="grey", se=FALSE)+
-  facet_grid(species~vitalrate)+
-  geom_text(data=cors, aes(label=paste("r = ", cor, sep="")), x=-1, y=1, size=4)+
-  xlab("QBM Estimate")+
-  ylab("IPM Estimate")+
-  theme_bw()
-
-setwd("~/Repos/MicroMesoForecast/manuscript/")
-```
-
 \pagebreak{}

### 2015-07-17 manuscript/manuscript.tex
updates ms with new structure

@@
 perennial grass species in a semi-arid Montana grassland. We
-parameterized two population models, one based on inidividual level data
-with three vital rates and one on an aggregated version of the same data
-(percent cover), and compared their accuracy, precision, and sensitivity
-to climate. The individual level model was more accurate and precise
-than the aggregate level model when predicting out of sample
-observations. The aggregate level model made countervailing forecasts to
-1\% climate changes when compared to the individual level model,
-indicating the aggregate level model produces incorrect forecasts. When
-comparing climate effects from both models, the aggregate level model
-tends to ``miss'' important climate effects from at least one vital rate
-for each species. It appears there is no short cut to forecasting
-climate change impacts on plant populations --- detailed demographic
-data is essential. But, forecasts were very uncertain, so we advocate
-for a focus on new methods to collect demographic data more efficiently
-across environmental gradients in space and time.
+parameterized two population models, one based on inidividual-level data
+with three vital rates and one on population-level data from the same
+dataset (percent cover of quadrats), and compared their accuracy,
+precision, and sensitivity to climate. The individual level model was
+more accurate and precise than the aggregate level model when predicting
+out-of-sample observations. When comparing climate effects from both
+models, the population-level model tends to ``miss'' important climate
+effects from at least one vital rate for each species. We also find that
+increasing the sample size at the population-level would not necessarily
+reduce forecast certainty; meaning the only way to reduce uncertainty is
+to capture unique climate dependence on individual vital rates. It
+appears there is no short cut to forecasting climate change impacts on
+plant populations --- detailed demographic data is essential. However,
+forecasts from our individual-level model were very uncertain, so we
+advocate for a focus on new methods to collect demographic data more
+efficiently across environmental gradients in space and time.
 
@@
 parameters respond to spatial variation in biotic and abiotic drivers
-(Sæther et al. 2007). The temporal coverage of demographic datasets is
-no better. {[}I think the temporal extent is a separate issue, might
-muddy the waters to bring it up here\ldots{}{]} For example, the most
-common study duration in the COMPADRE matrix population model database
-is 4 years and only a few studies exceed 10 years (Salguero-Gómez et al.
-2015). The limited spatio-temporal extent of individual-level
+(Sæther et al. 2007). The limited spatial extent of individual-level
 demographic datasets constrains our ability to use population models to
@@
 landscapes and regions, the scales at which many land-use decisions are
-made. However, the limitations of SDMs are well known (Elith and
-Leathwick 2009). Ideally, researchers would provide managers with
-landscape-scale population models, combing the extent of SDMs with
-information about dynamics and species abundances.
+made. However, the limitations of SDMs are well known (Pearson and
+Dawson 2003, Elith and Leathwick 2009, Araújo and Peterson 2012).
+Ideally, researchers would provide managers with landscape-scale
+population models, combing the extent of SDMs with information about
+dynamics and species abundances.
 
@@
 
-We assume the number of individuals, $Y^{R}$, recruiting at time $t+1$
+We assume the number of individuals, $y^{R}$, recruiting at time $t+1$
 follows a negative binomial distribution:
@@
 detect the overall sensitivity of equilibrium cover to climate, and we
-use (2) to show the impact of model uncertainty on forecast precision.
+use (2) to show the impact of model and parameter uncertainty on
+forecast precision.
 
@@
 
-Our \emph{a priori} expectation is that the IPM will produce more
-accurate and precise forecasts. This could be due to differences in
-sample sizes leading to larger parameter uncertainty for the QBM, or due
-to the QBM climate effects being weakly associated with one or more
-vital rate climate effects at the individual level. To assess the impact
-of sample size on QBM parameter uncertainty we refit the QBM statistical
-model (Eqs. 8-9) after removing sets of 2, 5, 10, and 15 quadrats. We
-fit 10 models at each level of quadrat removal (2, 5, 10, 15 quadrats),
-removing a different randomly selected set of quadrats for each fit. We
-calculated the standard deviation of climate main effects (pptlag, ppt1,
-ppt2, TmeanSpr1, and TmeanSpr2) for each model and averaged those over
+Our expectation is that the IPM will produce more accurate and precise
+forecasts. This could be due to differences in sample sizes leading to
+larger parameter uncertainty for the QBM, or due to the QBM climate
+effects being weakly associated with one or more vital rate climate
+effects at the individual level. To assess the impact of sample size on
+QBM parameter uncertainty we refit the QBM statistical model (Eqs. 8-9)
+after removing sets of 2, 5, 10, and 15 quadrats. We fit 10 models at
+each level of quadrat removal (2, 5, 10, 15 quadrats), removing a
+different randomly selected set of quadrats for each fit. We calculated
+the standard deviation of climate main effects (pptlag, ppt1, ppt2,
+TmeanSpr1, and TmeanSpr2) for each model and averaged those over
 replicates within each set of quadrat removals. This allowed us to
@@
 
-Equilibrium cover from both models was sensitive to climate (Fig. 3a-d).
-The IPM projected percent changes in equilibrium cover from -3 to 8\%
-for \emph{B. gracilis}, -4 to 3\% for \emph{H. comata}, -15 to 9\% for
-\emph{P. smithii}, and -17 to 53\% for \emph{P. secunda}. The QBM
-projected opposite and greater percent changes in equilibrium cover for
-\emph{B. gracilis} (-63 to 30\%) and \emph{H. comata} (-50 to -18\%;
-Fig. 3a-b). For \emph{P. smithii}, the QBM projected opposite changes in
-equilibrium cover than the IPM, but of similar magnitude (-5 to 6\%;
-Fig. 3c). \emph{P. secunda} was the only species that the IPM and QBM
-made projections of the same sign and somewhat similar magnitude (-20 to
-14\%; Fig. 3d).
-
 The response of a population to climate change is a result of the
@@
 combinations. Across all species, climate covariates can have opposing
-effects on different vital rates (Fig. 3e-h).
-
-Growth was the most sensitive vital rate for all species, showing a
-negative response to increased precipitation, and stronger positive
-response to increased temperature, and a mostly positive response when
-both climate factors are increased (Fig. 3e-h). \emph{B. gracilis}
-survival rates were sensitive to temperature, resulting in an increase
-in plant cover under increased temperature (Fig. 3e). In isolation,
-recruitment and survival were insensitive to climate factors for
-\emph{H. comata} (Fig. 3f). Survival and recruitment of \emph{P.
-smithii} were both sensitive, negatively, to temperature and
-precipitation (Fig. 3g). \emph{P. secunda} equilibrium cover was
-sensitive to the climate effects on survival and recruitment, showing a
-negative effect on both vital rates for increased precipition, but a
-strong positive effect on survival with increased temperature (Fig. 3h).
-The climate impact of recruitment on equilibrium cover was negative for
-precipitation and temperature increases (Fig. 3h). At least two of three
-vital rates were sensitive to climate for each species (Fig. 3).
-
-Forecasts based on 1\% climate changes were extremely uncertain when we
-considered model error and parameter uncertainty (Fig. 4). As expected
-based on model validation (Table 1), QBM projections were more uncertain
-than IPM projections for all species except \emph{P. smithiii} (Fig. 4).
-IPM forecasts for \emph{P. smithiii} were very uncertain due to
-overcompensating density dependence and nonlinearities recruitment that
-get amplified by parameter uncertainty (Appendx C). Neither model was
-capable of making forecasts of proportional cover change distinguishable
-from 0 (i.e., no change in population state) when we included model
-error and parameter uncertainty.
+effects on different vital rates (Fig. 3). Growth was the most sensitive
+vital rate for all species, showing a negative response to increased
+precipitation, and stronger positive response to increased temperature,
+and a mostly positive response when both climate factors are increased
+(Fig. 3). \emph{B. gracilis} survival rates were sensitive to
+temperature, resulting in an increase in plant cover under increased
+temperature (Fig. 3a). In isolation, recruitment and survival were
+insensitive to climate factors for \emph{H. comata} (Fig. 3b). Survival
+and recruitment of \emph{P. smithii} were both sensitive, negatively, to
+temperature and precipitation (Fig. 3c). \emph{P. secunda} equilibrium
+cover was sensitive to the climate effects on survival and recruitment,
+showing a negative effect on both vital rates for increased precipition,
+but a strong positive effect on survival with increased temperature
+(Fig. 3d). The climate impact of recruitment on equilibrium cover was
+negative for precipitation and temperature increases (Fig. 3d). At least
+two of three vital rates were sensitive to climate for each species
+(Fig. 3).
 
@@
 correlation with the survival model effects (Fig. 6c).
+
+\subsubsection{Model forecasts}\label{model-forecasts}
+
+Forecasts based on 1\% climate changes were extremely uncertain when we
+considered model error and parameter uncertainty (Fig. 6). As expected
+based on model validation (Table 1), QBM projections were more uncertain
+than IPM projections for all species except \emph{P. smithiii} (Fig. 6).
+IPM forecasts for \emph{P. smithiii} were very uncertain due to
+overcompensating density dependence and nonlinearities recruitment that
+get amplified by parameter uncertainty (Appendx C). Neither model was
+capable of making forecasts of proportional cover change distinguishable
+from 0 (i.e., no change in population state) when we included model
+error and parameter uncertainty.
 
@@
 
-\subsubsection{The IPM and QBM produced inconsistent
-forecasts}\label{the-ipm-and-qbm-produced-inconsistent-forecasts}
-
 Using individual-level and population-level forms of the same dataset,
@@
 spirit of density-structured models. In terms of each model's
-forecasting ability, the IPM outperformed the QBM (Table 1). This is
-unsurprising since we expected to lose some precision at the aggregated
-level. However, the underwhelming performance of the QBM could call into
-the question forecasts that differ from the IPM.
-
-Indeed, when we perturbed climate factors the QBM made forecasts
-completely contradictory to those of the IPM for three of our four study
-species (Fig. 3a-d). In a perfect world, the QBM would have made
-forecasts of at least the same direction as the IPM. If that had been
-the case we could conclude that aggregate level models could prove
-useful for forecasting climate change impacts on populations.
-Unfortunately, this was not the case.
-
-Given the superior ability of the IPM to predict out of sample
-observations (Table 1), we have no choice but to conclude it is the
-superior model. Following that logic, we can only assume that, at least
-contingent on the data in hand, the IPM is producing the correct
-forecasts to climate perturbations. The QBM failed to match IPM
-forecasts, implying that detailed demographic data may be necessary to
-accurately detect climate signals that are utlimately important at the
-population level. This result further confirms related work on the
-importance of individual variability on population level responses to
-exogenous drivers (Clark et al. 2011a, 2011b, 2012, Galván et al. 2014).
-
-\subsubsection{The role of vital rate climate
-dependence}\label{the-role-of-vital-rate-climate-dependence}
-
-We can think of two reasons why the IPM outperformed the QBM. First, the
-quadrat level data has a much reduced sample size compared to the
-individual level data. In an ideal world we would have compared the IPM
-and QBM using data collected over the same amount of person hours, not
-just the same number of quadrats. Then the sample size of the quadrat
-level data would be much greater and carry more statistical power. To
-address this limitation in our work we fit the QBM statistical model
-(Eq. 8-9) with different numbers of quadrats to see the effect of sample
-size on the precision of climate effect estimates. It appears that
-including additional quadrats leads to rapidly diminishing returns in
-terms of parameter precision (Fig. 5). Thus, while sample size surely
-plays some role, we do not think it is the main driver of the difference
-between the IPM and QBM.
-
-The second reason the IPM could have outperformed the QBM is that the
-population level model is in fact missing important climate effects that
-act on individual vital rates, rather than population growth. Our
-intuition was that species with strong climate-dependence on vital rates
-not well resolved at the aggregate level would result in different
-forecasts from the two models. For example, survival is very size
-dependent: smaller individuals have a higher probability of death (Chu
-and Adler 2014). At the same time, a single small individual contributes
-relatively little to percent cover estimates at the plot scale. So, if
-survival of individuals was positively impacted by temperature
-increases, for example, we would expect to detect this signal in the
-individual level data but not in the aggregate level data. To see if
-this is the case we can regress climate effects from each vital rate
-statistical model at the individual level against the same climate
-effects from the QBM statistical model (Fig. 6).
-
-In general, the QBM climate effects are most correlated with climate
-effects from the growth regression at the individual level (Fig. 6). In
-no case does the QBM statistical model have strong correlations across
-all three vital rates (Fig. 6). Thus, for each species the QBM is
-``missing'' climate signals associated with at least one vital rate.
-This has large impacts on predictions of long term population dynamics,
-as seen in our equilibrium simulations (Fig. 3a-d). The inability of the
-QBM to separate the sometimes countervailing effects of climate on
-survival, growth, and recruitment (Fig. 3e-h) results in inaccurate
-(Fig. 3a-d) and unprecise (Fig. 4) forecasts. The QBM statistical model
-struggles to explain variation due to climate variables because they can
-have positive and negative impacts on different vital rates. When this
-is the case, as it is for all our species to varying degrees (Fig.
-3e-h), statistical models of aggregated population responses will fail.
+forecasting ability, the IPM outperformed the QBM: the IPM was more
+accurate and precise than the QBM (Table 1). As we suggest in the
+introduction, this could be due to differences in sample size or the
+effect of averaging over climate responses unique to individual-level
+vital rates.
+
+We found no evidence that increasing sample size at the quadrat level
+would lead to higher precision of climate coefficient estimates (Fig.
+4). We did, however, find evidence that the QBM statistical model failed
+to identify climate dependence for some vital rates (Fig. 5). For no
+species were climate effects from the QBM strongly correlated with all
+three vital rates (Fig. 5). Freckleton et al. (2011) acknowledge that
+averaging over complex stage dependence will lead to poorly specified
+models. This is analagous to our situation, but instead of averaging
+over complex life histories, we are averaging over complex climate
+dependence.
+
+Thus, for each species the QBM is ``missing'' climate signals associated
+with at least one vital rate. This leads to inaccurate and imprecise
+forecasts because the QBM statistical model struggles to explain
+variation due to climate variables that have positive and negative
+impacts on different vital rates. When this is the case, as it is for
+all our species to varying degrees (Fig. 3), statistical models of
+population-level data will fail. This result further confirms related
+work on the importance of individual-level data to forecast population
+responses to exogenous drivers (Clark et al. 2011a, 2011b, 2012, Galván
+et al. 2014).
 
 These results lead us to conclude that detailed demographic data is
-necessary to forecast climate change impacts on plant populations. This
-is unwelcome news since this data is difficult to collect and the models
-built on such data are of little use to land managers that make
-decisions at scales beyond that of traditional population models
-(Queenborough et al. 2011). While density-structured approaches may fail
-when climate covariates are considered, there are other alternatives.
-For example, Clark et al. (2011a) use Forest Inventory and Analysis
-(FIA) data to parameterize a population model with multiple vital rates
-and climate dependence. Another example are distributed efforts like
-PlantPopNet (\url{http://plantago.plantpopnet.com}) that will allow
-researchers to estimate variation around climate responses for
-widespread species by taking advantage of spatial variation in climate
-(e.g. Doak and Morris 2010). Lastly, we foresee new approaches on the
-horizon that leverage photo/video of plots and advanced object
-recognition algorithms (e.g. Liu et al. 2014) to streamline plant
-mapping and digitizing efforts.
-
-\subsubsection{Forecasting the future, and the future of
-forecasting}\label{forecasting-the-future-and-the-future-of-forecasting}
+necessary to forecast climate change impacts on plant populations when
+vital rates have unique climate responses. This is unwelcome news since
+this data is difficult to collect and the models built on such data are
+of little use to land managers that make decisions at scales beyond that
+of traditional population models (Queenborough et al. 2011). While
+density-structured approaches may fail when climate covariates are
+considered, there are other alternatives. For example, Clark et al.
+(2011a) use Forest Inventory and Analysis (FIA) data to parameterize a
+population model with multiple vital rates and climate dependence.
+Another example are distributed efforts like PlantPopNet
+(\url{http://plantago.plantpopnet.com}) that will allow researchers to
+estimate variation around climate responses for widespread species by
+taking advantage of spatial variation in climate (e.g. Doak and Morris
+2010). Lastly, we foresee new approaches on the horizon that leverage
+photo/video of plots and advanced object recognition algorithms (e.g.
+Liu et al. 2014) to streamline plant mapping and digitizing efforts.
+
+What does all this mean for model forecasts? The answer is more nuanced
+than one might expect. The typical approach in ecology is to use point
+estimates of model parameters to project populations forward according
+to the specified model; usually allowing for some variability around the
+determinstic process (e.g., Battin et al. (2007); Jenouvrier et al.
+(2009); Adler et al. (2012)). If we follow tradition and calculate the
+mean response to climate change over many iterations of the model with
+only model error and temporal variation included, the IPM and the QBM
+produce opposing forecasts for three of four species (Fig. D1). However,
+if we introduce parameter uncertainty, the forecasts are actually
+indistinguishable (Fig. 6). IPM forecasts are generally more precise, as
+expected from model validation (Table 1). The real story is not that the
+QBM produces ``wrong'' forecasts, but that it produces highly uncertain
+forecasts. Our sample size analysis reveals that this uncertainty is not
+easily overcome by including more data (Fig. 4). The only way to reduce
+uncertainty is to model these populations based on individual-level data
+where climate dependence can be properly identified (Fig. 5). Yet, even
+the IPM forecasts were highly variable (Fig. 6).
 
@@
 useful and at worst meaningless. For all species, the 90\% quantiles of
-predicted changes in populatoin size overlapped zero; we cannot even
+predicted changes in population size overlapped zero; we cannot even
 predict whether a change in precipitation or temperature will cause
@@
 size. Climate covariates only vary by year, with no spatial variability
-within years. Thus, even if we fit models to individual level data, we
+within years. Thus, even if we fit models to individual-level data, we
 are missing the key interaction point between weather and individual
@@
 correlations. Only by being honest about our forecasts can we begin to
-produce better ones.
+produce better ones, and forecasts reported without parameter error are
+disingenuous. In many cases, ignoring parameter error is justifiable
+when investigating basic demographic mechanisms, but it is never
+justifiable if forecasts are being made.
 
@@
 This work is not a critique of density-structured population models. In
-some cases and for certain species, population models based on
+some cases and for certain species where climate dependence is weak or
+concentrated within a single vital rate, population models based on
 aggregated data may prove useful and unbiased. However, our work here is
@@
 responses are critical to predicting species' responses to climate
-change. Thus, forecasts from aggregate level models should be viewed
+change. Thus, forecasts from population-level models should be viewed
 with caution and should never be unaccompianed by uncertainty. Given the
@@
 \includegraphics{components/figure/manuscript-figure_3.pdf}
-\caption{Proportional change in species' mean cover caused by a 1\%
-increase in observed precipitation (+ppt), temperature (+temp), or both
-(+ppt\&temp) as predicted by the individual-based IPM and the
-aggregate-based QBM using mean parameter values. Top panels show the
-mean predicted proportional change in cover. Lower panels show the
-sensitivity of equilibrium cover simulated from the IPM to each climate
-scenario applied to individual and combined vital rates. For example,
-the points associated with G show the median cover from IPM simulations
-where a climate perturbation is applied only to the growth regression
-climate covariates. These simulations also use mean parameter values for
-clarity.}
+\caption{Sensitivity of equilibrium cover simulated from the IPM to each
+climate scenario applied to individual and combined vital rates. For
+example, the points associated with G show the median cover from IPM
+simulations where a climate perturbation is applied only to the growth
+regression climate covariates. These simulations use mean parameter
+values for clarity.}
 \end{figure}
@@
 \includegraphics{components/figure/manuscript-figure_4.pdf}
-\caption{Equilibrium cover and 90\% quantiles around the mean prediction
-when model error and parameter uncertainty are propogated through the
-simulation phase. Climate simulations are as in Figure 3.}
-\end{figure}
-
-\begin{figure}[htbp]
-\centering
-\includegraphics{components/figure/manuscript-figure_5.pdf}
 \caption{Effect of quadrat sample size on the precision (standard
@@
 \centering
-\includegraphics{components/figure/manuscript-figure_6.pdf}
+\includegraphics{components/figure/manuscript-figure_5.pdf}
 \caption{Correlations (r) between QBM and IPM estimates of climate
@@
 
+\begin{figure}[htbp]
+\centering
+\includegraphics{components/figure/manuscript-figure_6.pdf}
+\caption{Equilibrium cover and 90\% quantiles around the mean prediction
+when model error and parameter uncertainty are propogated through the
+simulation phase. Climate simulations are as in Figure 3.}
+\end{figure}
+
 \pagebreak{}
@@
 92:1703.
+
+Araújo, M. B., and A. T. Peterson. 2012. Uses and misuses of bioclimatic
+envelope modeling. Ecology 93:1527--1539.
+
+Battin, J., M. W. Wiley, M. H. Ruckelshaus, R. N. Palmer, E. Korb, K. K.
+Bartz, and H. Imaki. 2007. Projected impacts of climate change on salmon
+habitat restoration. Proceedings of the National Academy of Sciences of
+the United States of America 104:6720--6725.
 
@@
 Primer for Ecologists. Princeton University PressPrinceton.
+
+Jenouvrier, S., H. Caswell, C. Barbraud, M. Holland, J. Stroeve, and H.
+Weimerskirch. 2009. Demographic models and IPCC climate projections
+predict the decline of an emperor penguin population. Proceedings of the
+National Academy of Sciences of the United States of America
+106:1844--1847.
 
@@
 
+Pearson, R. G., and T. P. Dawson. 2003. Predicting the impacts of
+climate change on the distribution of species: Are bioclimate envelope
+models useful? Global Ecology and Biogeography 12:361--371.
+
 Petchey, O. L., M. Pontarp, T. M. Massie, S. Kéfi, A. Ozgul, M.
@@
 
-Salguero-Gómez, R., O. R. Jones, C. R. Archer, Y. M. Buckley, J.
-Che-Castaldo, H. Caswell, D. Hodgson, A. Scheuerlein, D. A. Conde, E.
-Brinks, H. de Buhr, C. Farack, F. Gottschalk, A. Hartmann, A. Henning,
-G. Hoppe, G. Römer, J. Runge, T. Ruoff, J. Wille, S. Zeh, R. Davison, D.
-Vieregg, A. Baudisch, R. Altwegg, F. Colchero, M. Dong, H. de Kroon,
-J.-D. Lebreton, C. J. E. Metcalf, M. M. Neel, I. M. Parker, T. Takada,
-T. Valverde, L. A. Vélez-Espino, G. M. Wardle, M. Franco, and J. W.
-Vaupel. 2015. The compadrePlant Matrix Database: an open online
-repository for plant demography. Journal of Ecology 103:202--218.
-
 Stan Development Team. 2014a. Stan: A C++ Library for Probability and

### 2015-07-17 manuscript/appD.Rmd
adds appendix D for model without parameter uncertainty

---
layout: 12pt
header-includes:
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{setspace}
bibliography: ~/Dropbox/Bibliography/MicroMeso.bib
csl: components/ecology.csl

## rmarkdown render options
output:
  pdf_document:
    fig_caption: true
    keep_tex: true
fontsize: 12pt
geometry: margin=1in

---

\renewcommand\thefigure{D\arabic{figure}}

```{r caching, include=FALSE}
library("methods")
library("knitr")
basename <- "appD"
opts_chunk$set(fig.path = paste("appD/figure/", basename, "-", sep=""),
               cache.path = paste("appD/cache/", basename, "/", sep=""))
opts_chunk$set(cache = 2)
opts_chunk$set(tidy=FALSE, warning=FALSE, message=FALSE, 
               comment = NA, verbose = TRUE, echo=FALSE)

# PDF-based figures
opts_chunk$set(dev='pdf')
```

Appendix D: Model forecasts without parameter uncertainty
=========================================================
In the main text we show results from model forecasts where we included full model and parameter uncertainty (Fig. 6). However, the typical approach in ecology is to use point estimates of model parameters and only include model error. For example, to predict population size *N* at time *t+1* one could do the following:

\begin{align}
\mu_{t+1} &= f(N_{t}) \\
N_{t+1} &\sim \text{Normal}(\mu_{t+1}, \sigma^2)
\end{align}

where $f(N_t)$ is some determinstic process (e.g., a linear regression model) and $\sigma^2$ is the model error. Simulating populations in such a way ignores variation in the parameters of $f(N_t)$. When we take this approach to making forecasts of equilibrium cover under 1% climate changes the IPM and QBM make opposing predictions if we simply look at the mean projection (Fig. D1). For both models, precision is also artificially high (Fig. D2 and D3), since we know that parameter uncertainty reduces forecast precision considerably (Fig. 6 in the main text). All three presentations of model results (mean projection, mean projection with model error, and forecasts with full uncertainty) can be found in the literature, which is why we present them here as well. However, as we state in the main text, true forecasts should only be presented as we did in Fig. 6 of the main text.

```{r plot-options, message=FALSE, warning=FALSE, include=FALSE, echo=FALSE}
cbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", 
               "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

opts_chunk$set(dev='pdf', dev.args=c(version="1.3"))

```

```{r figure_d1, echo=FALSE, message=FALSE, warning=FALSE, include=TRUE, fig.width=8.5, fig.height=4, fig.cap="Projected proportional changes in equilibrium cover caused by a 1% increase in observed precipitation (+ppt), temperature (+temp), or both (+ppt&temp) as predicted by the individual-level IPM and the population-level QBM using mean parameter values.", dependson="plot-options"}
library(plyr)
library(ggplot2)
library(gridExtra)
####
##  Mean parameter runs
####
setwd("~/Repos/MicroMesoForecast/analysis/")
meanparam_ipm <- data.frame(cover=NA, species=NA, 
                             climsim=NA, model=NA)
##  IPM 
path <- "./ipm/simulations/results/climchange_meanparams/"
ipmfiles_mean <- list.files(path)
ipmfiles_mean <- ipmfiles_mean[grep("cover", ipmfiles_mean)]
for(file in ipmfiles_mean){
  tmp <- read.csv(paste(path,file,sep=""))
  tmp$model <- "IPM"
  meanparam_ipm <- rbind(meanparam_ipm, tmp[,c("cover", "species", "climsim", "model")])
}
meanparam_ipm <- meanparam_ipm[2:nrow(meanparam_ipm),]

newd <- data.frame(cover=NA, species=NA, 
                   climsim=NA, model=NA, obscover=NA)
for(dospp in unique(meanparam_ipm$species)){
  tmp <- subset(meanparam_ipm, species==dospp)
  obscover <- subset(tmp, climsim=="obs")["cover"]
  tmp2 <- subset(tmp, climsim!="obs")
  tmp2$obscover <- rep(obscover$cover, times = length(unique(tmp2$climsim)))
  newd <- rbind(newd, tmp2)
}
ipm_meansims <- newd[2:nrow(newd),]
ipm_meansims$propdiff <- with(ipm_meansims, log(cover*100)-log(obscover*100))

##  QBM
meanparam_qbm <- data.frame(cover=NA, species=NA, 
                             climsim=NA, model=NA)
path <- "./quadBM/simulations/results/climatechange_meanparams/"
qbmfiles_mean <- list.files(path)
for(file in qbmfiles_mean){
  tmp <- readRDS(paste(path,file,sep=""))
  tmp$model <- "QBM"
  meanparam_qbm <- rbind(meanparam_qbm, tmp[,c("cover", "species", "climsim", "model")])
}
meanparam_qbm <- meanparam_qbm[2:nrow(meanparam_qbm),]

newd <- data.frame(cover=NA, species=NA, 
                   climsim=NA, model=NA, obscover=NA)
for(dospp in unique(meanparam_qbm$species)){
  tmp <- subset(meanparam_qbm, species==dospp)
  obscover <- subset(tmp, climsim=="obs")["cover"]
  tmp2 <- subset(tmp, climsim!="obs")
  tmp2$obscover <- rep(obscover$cover, times = length(unique(tmp2$climsim)))
  newd <- rbind(newd, tmp2)
}
qbm_meansims <- newd[2:nrow(newd),]
qbm_meansims$propdiff <- with(qbm_meansims, log(cover*100)-log(obscover*100))

##  Merge IPM and QBM mean parameter results
bothmodel_outs <- rbind(ipm_meansims, qbm_meansims)
meanparam_agg <- ddply(bothmodel_outs, .(species, climsim, model), summarise,
                       meandiff = mean(propdiff),
                       hidiff = quantile(propdiff, probs = 0.95),
                       lodiff = quantile(propdiff, probs = 0.05))
dgd = position_dodge(width = 0.9)
meanplot <- ggplot(meanparam_agg)+
  geom_bar(aes(x=climsim, y=meandiff, fill=model),
             position=dgd, stat="identity")+
  theme_bw()+
  scale_fill_manual(values=c("black", "grey45"), name="Model")+
  facet_wrap("species", scales="free", nrow=1)+
  scale_x_discrete(labels=c("+ppt", "+temp", "+ppt&temp"))+
  ylab("Mean proportional difference")+
  xlab("Climate simulation")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(meanplot)
```

```{r figure_d2, echo=FALSE, message=FALSE, warning=FALSE, include=TRUE, fig.width=8.5, fig.height=4, fig.cap="Projected mean (points) proportional changes in equilibrium cover and associated 90% quantiles (errorbars) caused by a 1% increase in observed precipitation (+ppt), temperature (+temp), or both (+ppt&temp) as predicted by the individual-level IPM using mean parameter values. Dashed line indicates proportional change of 0.", dependson="plot-options"}

errorplot1 <- ggplot(subset(meanparam_agg, model=="IPM"))+
  geom_point(aes(x=climsim, y=meandiff),
             position=dgd, stat="identity")+
  geom_errorbar(aes(x=climsim, ymax=hidiff, ymin=lodiff),
                position=dgd, width=0.25)+
  geom_hline(aes(yintercept=0), linetype=2)+
  theme_bw()+
  facet_wrap("species", scales="free", nrow=1)+
  scale_x_discrete(labels=c("+ppt", "+temp", "+ppt&temp"))+
  ylab("Mean proportional difference")+
  xlab("Climate simulation")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
print(errorplot1)
# print(meanplot)
```

```{r figure_d3, echo=FALSE, message=FALSE, warning=FALSE, include=TRUE, fig.width=8.5, fig.height=4, fig.cap="Projected mean (points) proportional changes in equilibrium cover and associated 90% quantiles (errorbars) caused by a 1% increase in observed precipitation (+ppt), temperature (+temp), or both (+ppt&temp) as predicted by the population-level QBM using mean parameter values. Dashed line indicates proportional change of 0.", dependson="plot-options"}

errorplot2 <- ggplot(subset(meanparam_agg, model=="QBM"))+
  geom_point(aes(x=climsim, y=meandiff),
             position=dgd, stat="identity")+
  geom_errorbar(aes(x=climsim, ymax=hidiff, ymin=lodiff),
                position=dgd, width=0.25)+
  geom_hline(aes(yintercept=0), linetype=2)+
  theme_bw()+
  facet_wrap("species", scales="free", nrow=1)+
  scale_x_discrete(labels=c("+ppt", "+temp", "+ppt&temp"))+
  ylab("Mean proportional difference")+
  xlab("Climate simulation")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
print(errorplot2)
# print(meanplot)
```

### 2015-07-20 manuscript/appC.Rmd
adds more on the pasm recruitment booms

@@
 =====================================================
+Seems to be due to the climate effects.
+Are they really more uncertain than the other species -- or is it more that it is such low cover so boom years are more influential?
+I think it is because average recruits are very high relative to other species, so then tack on occassional big years due to climate draws, and you get booms with subsequent low survival busts.
+
 In Fig.
@@
 The uncertainty around PASM forecasts from the integral projection model is orders of magnitude larger than for any other species, and is even larger than the uncertainty around the quad-based model forecasts.
-It turns out that the recruitment parameters we estimate lead to overcompensating density-dependence that is amplified when those parameters are allowed to vary.
+It turns out that the estimated dispersion parameter for the negative binomial distribution is much higher for PASM than for other species (Appendix B).
+This is because PASM has higher recruitment rates than the other species and the relationship between number of recruits and quadrat cover (see main text) is weak.
+So, when we draw parameters from the MCMC chain for our simulations, even larger values for the dispersion parameter are possible.
+Thus, there can be very large "boom" years where recruitment is high.
 When we allow all parameters to vary except those associated with recruitment, uncertainty around PASM decreases to levels similar to those associated with other species (Fig.
 C1).
-So detailed demographic data is not a cure-all, and can sometimes lead to interesting dynamics when parameter uncertainty is considered.
+We also performed simulations where all recruitment parameters except the dispersion parameter were allowed to vary.
+Results from those simulations show the effect of the dispersion parameter on boom and bust years (Fig.
+C2).
 

### 2015-07-20 manuscript/manuscript.Rmd
adds more on the pasm recruitment booms

@@
 IPM forecasts for *P.
-smithiii* were very uncertain due to overcompensating density dependence and nonlinearities recruitment that get amplified by parameter uncertainty (Appendx C).
+smithiii* were very uncertain due to a large dispersion parameter ($\zeta$ in Eq.
+6) that can lead to boom years with very large recruitment and subsequent busts due to low survival of so many young plants (Appendx C).
 Neither model was capable of making forecasts of proportional cover change distinguishable from 0 (i.e., no change in population state) when we included model error and parameter uncertainty.

### 2015-07-20 manuscript/manuscript.tex
adds more on the pasm recruitment booms

@@
 than IPM projections for all species except \emph{P. smithiii} (Fig. 6).
-IPM forecasts for \emph{P. smithiii} were very uncertain due to
-overcompensating density dependence and nonlinearities recruitment that
-get amplified by parameter uncertainty (Appendx C). Neither model was
-capable of making forecasts of proportional cover change distinguishable
-from 0 (i.e., no change in population state) when we included model
-error and parameter uncertainty.
+IPM forecasts for \emph{P. smithiii} were very uncertain due to a large
+dispersion parameter ($\zeta$ in Eq. 6) that can lead to boom years with
+very large recruitment and subsequent busts due to low survival of so
+many young plants (Appendx C). Neither model was capable of making
+forecasts of proportional cover change distinguishable from 0 (i.e., no
+change in population state) when we included model error and parameter
+uncertainty.
 

### 2015-07-20 analysis/ipm/simulations/supp_sims/pasm_recdispersion_test.R

@@ -1,0 +1,491 @@
+###############################################################
+#### Wrapper for running climate change scenarios by species
+####
+#### Andrew Tredennick
+#### 2-11-2014
+####
+
+root=ifelse(.Platform$OS.type=="windows","c:/repos","~/repos"); # modify as needed
+setwd(paste(root,"/MicroMesoForecast/analysis/ipm/simulations",sep="")); # modify as needed 
+
+# Set some global parameters for the simulations
+tlimit<-500  ## number of years to simulate
+burn.in<-0    # years to cut before calculations
+spp_list <- c("BOGR","HECO","PASM","POSE")
+n_spp <- length(spp_list)
+
+# Get climate and random year effect sequences
+yrSave <- readRDS("../../random_year_effects_sequence.rds")
+climYr <- readRDS("../../climate_year_sequence.rds")
+# paramSeq <- readRDS("../../random_chain_iter_sequence.RDS")
+
+# get calendar years
+raw_data <- read.csv("../../speciesData/quadAllCover.csv")
+years <- unique(raw_data$year)
+years <- years[1:(length(years)-1)] #lop off 1945 since no climate for that year
+
+# get alphas values (needed to calculate neighborhood crowding)
+alpha_grow <- read.csv("../../alpha_list_growth.csv")
+alpha_surv <- read.csv("../../alpha_list_survival.csv")
+
+# Get mean recruitment dispersion parameter
+source("../vitalRateRegs/recruitment/import2ipm_means.R", echo = FALSE)
+Rpars_tmp <- getRecCoefs(1,NA)
+clim_mean <- Rpars_tmp$clim
+
+# Source functions
+source("vital_rate_ipm_functions.R", echo = FALSE)
+source("../vitalRateRegs/survival/import2ipm.R", echo = FALSE)
+source("../vitalRateRegs/growth/import2ipm.R", echo = FALSE)
+source("../vitalRateRegs/recruitment/import2ipm.R", echo = FALSE)
+
+
+####
+#### Simulation 1. Observed climate sequence -------------------------------------
+####
+# Set climate for observed climate run
+clim_data <- read.csv("../../weather/Climate.csv")
+clim_data <- clim_data[,c("year", "pptLag","ppt1","ppt2","TmeanSpr1","TmeanSpr2")] # subset and reorder to match regression param import
[…]

### 2015-07-25 esa_talk/esa2015.Rmd
updates figure 2 MAE

@@
 ggplot(subset(plotd, variable!="mean_cover"), aes(x=species, y=value, fill=model))+
-  geom_bar(stat="identity", position="dodge")+
+  geom_bar(stat="identity", position="dodge", color="white")+
   facet_wrap("variable", nrow=1, scales="free")+
-  scale_fill_manual(values = c("darkorange", "steelblue"))
+  scale_fill_manual(values = c("darkorange", "steelblue"))+
+  theme_bw()
 ```

### 2015-07-31 manuscript/appE.Rmd
adds new results from upping the perturbation percentages for climate

---
layout: 12pt
header-includes:
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{setspace}
bibliography: ~/Dropbox/Bibliography/MicroMeso.bib
csl: components/ecology.csl

## rmarkdown render options
output:
  pdf_document:
    fig_caption: true
    keep_tex: true
fontsize: 12pt
geometry: margin=1in

---

\renewcommand\thefigure{E\arabic{figure}}

```{r caching, include=FALSE}
library("methods")
library("knitr")
basename <- "appE"
opts_chunk$set(fig.path = paste("appE/figure/", basename, "-", sep=""),
               cache.path = paste("appE/cache/", basename, "/", sep=""))
opts_chunk$set(cache = 2)
opts_chunk$set(tidy=FALSE, warning=FALSE, message=FALSE, 
               comment = NA, verbose = TRUE, echo=FALSE)

# PDF-based figures
opts_chunk$set(dev='pdf')
```

Appendix E: Model forecasts under larger climate perturbations
==============================================================
In the main text we show results from model forecasts where we perturbed climate covariates by 1% (Fig. 6). Given the uncertainty in these forecasts, we conducted further simulations where we increased climate covariates by 10%, 20%, and 30% to see how large of a perturbation is necessary to produce forecasts that do not overlap zero. Here we show the results.

```{r plot-options, message=FALSE, warning=FALSE, include=FALSE, echo=FALSE}
cbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", 
               "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

opts_chunk$set(dev='pdf', dev.args=c(version="1.3"))

```

```{r figure_e1, echo=FALSE, message=FALSE, warning=FALSE, include=TRUE, fig.width=8.5, fig.height=4, fig.cap="Projected proportional changes in equilibrium cover caused by 10%, 20%, and 30% increases in observed precipitation (+ppt), temperature (+temp), or both (+ppt&temp) as predicted by the individual-level IPM.", dependson="plot-options"}
library(plyr)
library(ggplot2)
library(gridExtra)
####
##  Mean parameter runs
####
setwd("~/Repos/MicroMesoForecast/analysis/ipm/simulations/results/climchange_varyparams/")
meanparam_ipm <- data.frame(cover=NA, species=NA, 
                             climsim=NA, pertamount=NA)
##  IPM 
ipmfiles_mean <- list.files()
ipmfiles_mean <- ipmfiles_mean[grep("cover", ipmfiles_mean)]
toget <- grep("10", ipmfiles_mean)
ipmnows <- ipmfiles_mean[toget]
for(file in ipmnows){
  tmp <- read.csv(file)
  tmp$pertamount <- "10%"
  meanparam_ipm <- rbind(meanparam_ipm, tmp[,c("cover", "species", "climsim", "pertamount")])
}
ipm10 <- meanparam_ipm[2:nrow(meanparam_ipm),]

# plot(c(1:2001),subset(ipm20, species=="HECO" & climsim=="temp")["cover"], type="l")

meanparam_ipm <- data.frame(cover=NA, species=NA, 
                             climsim=NA, pertamount=NA)
toget <- grep("20", ipmfiles_mean)
ipmnows <- ipmfiles_mean[toget]
for(file in ipmnows){
  tmp <- read.csv(file)
  tmp$pertamount <- "20%"
  meanparam_ipm <- rbind(meanparam_ipm, tmp[,c("cover", "species", "climsim", "pertamount")])
}
ipm20 <- meanparam_ipm[2:nrow(meanparam_ipm),]

meanparam_ipm <- data.frame(cover=NA, species=NA, 
                             climsim=NA, pertamount=NA)
toget <- grep("30", ipmfiles_mean)
ipmnows <- ipmfiles_mean[toget]
for(file in ipmnows){
  tmp <- read.csv(file)
  tmp$pertamount <- "30%"
  meanparam_ipm <- rbind(meanparam_ipm, tmp[,c("cover", "species", "climsim", "pertamount")])
}
ipm30 <- meanparam_ipm[2:nrow(meanparam_ipm),]

pertd <- rbind(ipm10, ipm20, ipm30)
pertd[which(pertd$cover>1),"cover"] <- 1

obsfiles <- ipmfiles_mean[grep("obsClimate",ipmfiles_mean)]
obsd <- data.frame(cover=NA, species=NA, 
                   climsim=NA, pertamount=NA)
for(file in obsfiles){
  tmp <- read.csv(file)
  tmp$pertamount <- "0%"
  obsd <- rbind(obsd, tmp[,c("cover", "species", "climsim", "pertamount")])
}
obsd <- obsd[2:nrow(obsd),]

alld <- rbind(obsd, pertd)

newd <- data.frame(cover=NA, species=NA, 
                   climsim=NA, pertamount=NA, obscover=NA)
for(dospp in unique(alld$species)){
  tmp <- subset(alld, species==dospp)
  obscover <- subset(tmp, climsim=="obs")["cover"]
  tmp2 <- subset(tmp, climsim!="obs")
  tmp2$obscover <- rep(rep(obscover$cover, times = length(unique(tmp2$climsim))), times=3)
  newd <- rbind(newd, tmp2)
}
ipm_meansims <- newd[2:nrow(newd),]
ipm_meansims$propdiff <- with(ipm_meansims, log(cover*100)-log(obscover*100))

meanparam_agg <- ddply(ipm_meansims, .(species, climsim, pertamount), summarise,
                       meandiff = mean(propdiff),
                       hidiff = quantile(propdiff, probs = 0.95),
                       lodiff = quantile(propdiff, probs = 0.05))
dgd = position_dodge(width = 0.9)
meanplot <- ggplot(meanparam_agg)+
  geom_hline(aes(yintercept=0), linetype=2)+
  geom_point(aes(x=pertamount, y=meandiff),
             position=dgd, stat="identity")+
  geom_errorbar(aes(x=pertamount, ymax=hidiff, ymin=lodiff),
                position=dgd, width=0.25)+
  theme_bw()+
#   scale_fill_manual(values=c("black", "grey45"), name="PertAmount")+
  facet_wrap(species~climsim, scales="free", nrow=4, ncol=3)+
  scale_x_discrete(labels=c("+ppt", "+temp", "+ppt&temp"))+
  ylab("Mean proportional difference")+
  xlab("Climate simulation")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(meanplot)
```

### 2015-07-31 manuscript/manuscript.Rmd
adds new results from upping the perturbation percentages for climate

--- 2015-07-17 manuscript/manuscript.Rmd
@@
 Forecasting future states of populations has taken on new urgency as the rate of climate change increases.
-Traditional plant population models have limited utility in this regard because they are based on detailed demographic data from small, localized plots.
-These models are difficult to scale up to spatial scales relevant to land managers that require such forecasts to make decisions.
-To overcome the data limitations of traditional population models, some have proposed population models based on population level, rather than individual level, data that is much easier to collect over broad spatial scales.
-Using such models violates a central assumption of ecology: individuals respond to weather, not populations to climate.
-Here, we test whether this assumption is important when forecasting climate change impacts on four perennial grass species in a semi-arid Montana grassland.
-We parameterized two population models, one based on inidividual-level data with three vital rates and one on population-level data from the same dataset (percent cover of quadrats), and compared their accuracy, precision, and sensitivity to climate.
+Traditional plant population models, typically built using detailed demographic observations from one study site, can address climate change impacts at one location, but are difficult to scale up to the landscape and regional scales relevant to management decisions.
+One proposal to overcome the limitations of traditional population models is to build models using population-level data that is much easier to collect over broad spatial scales than individual-level data.
+However, such models ignore the fact that climate drives population growth through its influence on individual performance.
+Here, we test the consequences of aggregating individual responses when forecasting climate change impacts on four perennial grass species in a semi-arid Montana grassland.
+We parameterized two population models, one based on individual-level data (survival, growth and recruitment) and one on population-level data from the same dataset (percent cover of quadrats), and compared their accuracy, precision, and sensitivity to climate.
 The individual level model was more accurate and precise than the aggregate level model when predicting out-of-sample observations.
 When comparing climate effects from both models, the population-level model tends to "miss" important climate effects from at least one vital rate for each species.
-We also find that increasing the sample size at the population-level would not necessarily reduce forecast certainty; meaning the only way to reduce uncertainty is to capture unique climate dependence on individual vital rates.
-It appears there is no short cut to forecasting climate change impacts on plant populations — detailed demographic data is essential.
-However, forecasts from our individual-level model were very uncertain, so we advocate for a focus on new methods to collect demographic data more efficiently across environmental gradients in space and time.
+Increasing the sample size at the population-level would not necessarily reduce forecast uncertainty; the way to reduce uncertainty is to capture unique climate dependence of individual vital rates.
+Our analysis indicates that there is no short cut to forecasting climate change impacts on plant populations — detailed demographic data is essential.
+Despite the superiority of the individual-level model, the forecasts it generated still were too uncertain to be useful for decision-makers.
+We need new methods to collect demographic data more efficiently across environmental gradients in space and time.
 
@@
 Perhaps the greatest challenge for ecology in the 21st century is to forecast the impacts of environmental change [@Clark2001; @Petchey2015].
-To do so requires sophisticated modeling approaches that fully account for uncertainty and variability in the ecological process and associated parameters [@Luo2011a].
-This requires large amounts of data collected over large spatio-temporal extents.
-State-of-the-science modeling techniques cannot overcome data limitations.
-Such is the case for many population models.
+Forecasts require sophisticated modeling approaches that fully account for uncertainty and variability in both ecological process and model parameters [@Luo2011a].
+This requires large amounts of data collected over broad spatio-temporal extents, and state-of-the-science modeling techniques cannot overcome data limitations.
+Population models are a case in point.
 
@@
 But reconciling the scales at which population models are parameterized and the scales at which environmental changes play out remains a challenge [@Clark2010; @Freckleton2011; @Queenborough2011; @Clark2012].
-The major hurdle is that most population models are built using data from a single study site because collecting those data, which involves tracking the fates of individuals plants, is so difficult.
+The problem is that most population models are built using data from a single study site because collecting those data, which involves tracking the fates of individuals plants, is so difficult.
 The resulting models cannot be applied to the landscape and regional scales relevant to decision-makers without information about how the fitted parameters respond to spatial variation in biotic and abiotic drivers [@Sæther2007].
@@
 
-The inability of many population models to address landscape-scale problems may explain why land managers and conservation planners have embraced species distribution models (SDMs) [see @Guisan2005 for a review].
+The inability of most population models to address landscape-scale problems may explain why land managers and conservation planners have embraced species distribution models (SDMs) [see @Guisan2005 for a review].
 SDMs typically rely on easy-to-collect presence/absence data [but see @Clark2014 for new methods] and remotely-sensed environmental covariates that allow researchers to model large spatial extents [e.g., @Maiorano2013].
@@
 However, the limitations of SDMs are well known [@Pearson2003; @Elith2009; @Araujo2012].
-Ideally, researchers would provide managers with landscape-scale population models, combing the extent of SDMs with information about dynamics and species abundances.
+Ideally, researchers would provide managers with landscape-scale population models, combining the extent of SDMs with information about dynamics and species abundances.
 
@@
 The study by @Queenborough2011 included data from 500 fields (4 hectares each) in 49 farms, all collected by two people in 6 weeks.
-This is far more data from a far greater spatial extent than possible if measuring individual plant demography (in a world of limited time and money, at least).
+This is far more data from a far greater spatial extent than would be possible if measuring individual plant demography (in a world of limited time and money).
 The appeal of density-structured approaches is clear.
@@
 It is individuals, not populations, that respond to climate variables [@Clark2012].
-Ignoring this fact puts us in uneasy proximity to an "ecological fallacy", where inference about the individual relies on statistical inference on the group [@PIANTADOSI1988].
-Population growth (or decline) is an epiphenomenon of demographic processes like survival, growth, and recruitment that occur at the level of individual plants.
+Ignoring this fact amounts to an "ecological fallacy", where inference about the individual relies on statistical inference on the group [@PIANTADOSI1988].
+Population growth (or decline) is the outcome of demographic processes like survival, growth, and recruitment that occur at the level of individual plants.
 Climate can affect each demographic process in unique, potentially opossing, ways [@Dalgleish2011].
-These unique climate responses may difficult resolve in statistical models based on population-level data where demographic processes are not identifiable.
+These unique climate responses may be difficult to resolve in statistical models based on population-level data where demographic processes are not identifiable.
 If population-level data cannot detect important impacts of climate drivers on populations, then population models built with such data will make poor forecasts.
 
-Here, we ask whether statistical and population models based on aggregated, population-level data can detect climate signals as wells as models based on individual-level data.
+Here, we ask whether statistical and population models based on aggregated, population-level data can predict climate change impacts as wells as models based on individual-level data.
 We used a unique demographic dataset that tracks the fates of individual plants from four species over 14 years to build two kinds of single-species population models, traditional models using individual growth, survival, and recruitment data and alternative models based on basal cover.
@@
 We then performed simulations to quantify the sensitivities of species' cover to small perturbations in average precipitation and temperature.
-We found that population models based on detailed demographic data are more accurate and precise than models based on aggregated data.
-Our results suggest the popuation-level model is less accurate and precise because important demographic climate signals go undetected.
-For these species at this location, detailed demographic data appears necessary to make accurate forecasts.
-A worrying caveat to our work is that forecasts from both models were very uncertain when we considered full process and parameter uncertainty.
-Even 14 years worth of demographic data may not be sufficient to make useful forecasts.
+Individual level models were clearly better, but, unfortunately, still too uncertain to inform management decisions.
 
@@
 The annual maps of the quadrats were digitized and the fates of individual plants tracked and extracted using a computer program [@Lauenroth2008; @Chu2014a].
-Daily climate data, which we aggregated into climate variables of interest, are available for the duration of the data collection period (1932 - 1945) from the Miles City airport, Wiley Field, 9 km from the study site.
+Daily climate data are available for the duration of the data collection period (1932 - 1945) from the Miles City airport, Wiley Field, 9 km from the study site.
 
@@
 At both levels of inference (individual and quadrat), the building blocks of our population models are vital rate regressions.
-For individual-level data we fit models for survival, growth, and recruitment for each species.
-At the quadrat-level we fit a single regression model for population growth.
+For individual-level data, we fit models for survival, growth, and recruitment for each species.
+At the quadrat-level, we fit a single regression model for population growth.
 We describe the statistical models separately since fitting the models required different approaches.
@@
 We also include interactions among same-year climate covariates (e.g., ppt1 $\times$ TmeansSpr1) and climate $\times$ size interactions.
-Climate $\times$ size interactions are for climate main effects only, that is we do not include interactions among size and interacting climate effects.
+Climate $\times$ size interactions are for climate main effects only; we do not include interactions between size and pairs of interacting climate effects.
 
@@
 
-We modeled growth as Gaussian process describing genet size at time $t+1$ as a function of size at $t$ and climate covariates:
+We modeled growth as a Gaussian process describing genet size at time $t+1$ as a function of size at $t$ and climate covariates:
 
@@
 x_{ijQ,t+1} &= \gamma^{G}_{j,t} + \phi^{G}_{jQ} + \beta^{G}_{j,t}x_{ij,t} + \omega^{G}_{j}w_{ij,t} + \nu^{S}_{j}w_{ij,t}x_{ij,t} + \theta^{G}_{jk}C_{k,t} \\
-y^{G}_{ijQ,t} &\sim \text{Normal}(x_{ijQ,t+1}, \sigma_{j})
+y^{G}_{ijQ,t} &\sim \text{Normal}(x_{ijQ,t+1}, \varepsilon_{ij,t})
 \end{align}
@@
 where $x$ is log genet size and all other parameters are as described for the survival regression.
+We capture non-constant error variance in growth by modeling the variance around the growth regression ($\varepsilon$) as a nonlinear function of predicted genet size:
+
+\begin{align}
+\varepsilon_{ij,t} = a e^{b x_{ijQ,t+1}}
+\end{align}
 
@@
 x_{jq,t+1} &= \gamma^{P}_{j,t} + \phi^{P}_{jQ} + \beta^{P}_{j,t}x_{jq,t} + \theta^{P}_{jk}C_{k,t} \\
-y^{P}_{jq,t+1} &\sim \text{LogNormal}(x_{jq,t+1}, \tau{j}) \text{T}[0,1]
+y^{P}_{jq,t+1} &\sim \text{LogNormal}(x_{jq,t+1}, \tau_{j}) \text{T}[0,1]
 \end{align}
@@
 We obtained posterior distributions for all model parameters from three parallel MCMC chains run for 1,000 iterations after discarding an initial 1,000 iterations.
-We recignize such short MCMC chains may surprise those more familiar with other MCMC samplers (i.e.
+We recognize such short MCMC chains may surprise those more familiar with other MCMC samplers (i.e.
 JAGS or WinBUGS), but the Stan sampler is exceptionally efficient, which reduces the number of iterations needed to achieve convergence.
@@
 With the posterior distribution of the vital rate statistical models in hand, it is straightforward to simulate the population models.
-We used an Integral Projection Model (IPM) to model populations based on individual level data [cite Ellner and Rees 2006] and an quadrat based version of an individually-based model (Quadrat-Based Model, QBM) to model populations based on quadrat level data.
+We used an Integral Projection Model (IPM) to model populations based on individual level data [@Ellner2006] and a quadrat-based version of an individually-based model (Quadrat-Based Model, QBM) to model populations based on quadrat level data.
 We describe each in turn.
@@
 #### Integral projection model
-We use an environmentally stochastic IPM [@Rees2009] that includes the random year effects and the climate covariates from the vital rate statistical models.
-However, for some simulations, we ignore the random year effects so that only the climate effects drive interannual variation.
+We use an environmentally stochastic IPM [@Rees2009] that includes the climate covariates from the vital rate statistical models.
+In all simulations we ignore the random year effects so that interannual variation is driven solely by climate.
+We fit the random year effects in the vital rate regressions to avoid over-attributing variation to climate covariates.
 Our IPM follows the specification of @Chu2015 where the population of species _j_ is a density function $n(u_{j},t)$ giving the density of sized-_u_ genets at time _t_.
 Genet size is on the natural log scale, so that $n(u_{j},t)du$ is the number of genets whose area (on the arithmetic scale) is between $e^{u_{j}}$ and $e^{u_{j}+du}$.
-So, the density function for any size _v_ at time $t+1$ is
+The density function for any size _v_ at time $t+1$ is
 
@@
 
+Since the IPM is spatially-implicit, we cannot calculate neighborhood crowding for specific genets ($w_{ij}$).
+Instead, we use an approximation ($\bar{\bold{w_{j}}}$) that captures the essential features of neighborhood interactions [@Adler2010].
+We also use a 'no-overlap' rule for conspecific genets to approximate the overdispersion of large genets in space [@Adler2010; @Adler2012; @Chu2015].
+
 The population kernel is defined as the joint contributions of survival (_S_), growth (_G_), and recruitment (_R_):
@@
 Following previous work [@Adler2012; @Chu2015], we assume that fecundity increases linearly with size ($R_j(v_{j},u_{j},\bar{\bold{w_{j}}}) = e^{u_j}R_j(v_{j},\bar{\bold{w_{j}}})$) to incorporate the recruitment function in the spatially-implicit IPM.
-\todo{Add spatial structure mention.}
 
@@
 Our quad-based model (QBM) perfectly mirrors its statistical description (Eqs.
-8-9).
+9-10).
 We use the same approach for drawing parameter values as described for the IPM.
 After drawing the appropriate parameter set, we calculate the mean response (population cover at *t+1* = $x_{t+1}$) according to Eq.
-8.
+9.
 We then make a random draw from a [0,1] truncated lognormal distribution with mean equal to $x_{t+1}$ from Eq.
-8 and the variance estimate from the fitted model.
-We can then iterate the model forward by drawing a new parameter set (unique to climate year and MCMC iteration) at each timestep.
+9 and the variance estimate from the fitted model.
+We can then project the model forward by drawing a new parameter set (unique to climate year and MCMC iteration) at each timestep.
+As with the IPM, random year effects are ignored for all simulations.
 
@@
 ### Testing sensitivity to climate covariates
-Our main goal in this paper is to see if models based on aggregated data are as sensitive to climate covariates as models based on individual level data.
+To see if models based on aggregated data are as sensitive to climate covariates as models based on individual level data.
 So, with our fitted and validated models in hand, we ran simulations for each model type (IPM and QBM) under four climate perturbation scenarios: (1) observed climate, (2) precipitation increased by 1%, (3) temperature increased by 1%, and (4) precipitation and temperature increased by 1%.
@@
 Our expectation is that the IPM will produce more accurate and precise forecasts.
-This could be due to differences in sample sizes leading to larger parameter uncertainty for the QBM, or due to the QBM climate effects being weakly associated with one or more vital rate climate effects at the individual level.
+This could be due to (1) the smaller  sample size of the quadrat level data sets compared to the individual level data sets, leading to larger parameter uncertainty for the QBM, or (2) the QBM climate effects being weakly associated with one or more vital rate climate effects at the individual level.
 To assess the impact of sample size on QBM parameter uncertainty we refit the QBM statistical model (Eqs.
-8-9) after removing sets of 2, 5, 10, and 15 quadrats.
+9-10) after removing sets of 2, 5, 10, and 15 quadrats.
 We fit 10 models at each level of quadrat removal (2, 5, 10, 15 quadrats), removing a different randomly selected set of quadrats for each fit.
-We calculated the standard deviation of climate main effects (pptlag, ppt1, ppt2, TmeanSpr1, and TmeanSpr2) for each model and averaged those over replicates within each set of quadrat removals.
+We calculated the standard deviation of climate main effects (pptLag, ppt1, ppt2, TmeanSpr1, and TmeanSpr2) for each model and averaged those over replicates within each set of quadrat removals.
 This allowed us to regress sample size against parameter uncertainty.
 
-To see if the QBM climate effects are correlated, or not, with climate effects for each vital rate model for the IPM, we simply regressed the QBM climate coefficients against each vital rate model's climate coefficients and calculate Pearson's $\rho$.
+To see if the QBM climate effects are correlated, or not, with climate effects for each vital rate model for the IPM, we simply regressed the QBM climate coefficients against each vital rate model's climate coefficients and calculated Pearson's $\rho$.
 Strong correlations indicate the QBM is capable of detecting climate effects associated with individual vital rates.
@@
 3d).
-The climate impact of recruitment on equilibrium cover was negative for precipitation and temperature increases (Fig.
+Equilibrium cover responded negatively when increased precipitation and temperature affect recruitment  (Fig.
 3d).
@@
 ### Sources of uncertainty in the QBM
-Sample size had a relatively weak effect on QBM climate parameter uncertainty after the number of quadrats exceeded about 10 (Fig.
+Sample size had a relatively weak effect on QBM climate parameter uncertainty after the number of quadrats used in fitting exceeded about 10 (Fig.
 5).
-Inverse-gaussian fits show that increasing sample size beyond the number of quadrats we used will results in deminishing returns in terms of parameter certainty (Fig.
+Inverse-gaussian fits show that increasing sample size beyond the number of quadrats we used would result in diminishing returns in terms of parameter certainty (Fig.
 5).
@@
 Forecasts based on 1% climate changes were extremely uncertain when we considered model error and parameter uncertainty (Fig.
-6).
+6; simulations with mean parameters are in Appendix D for comparison).
 As expected based on model validation (Table 1), QBM projections were more uncertain than IPM projections for all species except *P.
@@
 IPM forecasts for *P.
-smithiii* were very uncertain due to overcompensating density dependence and nonlinearities recruitment that get amplified by parameter uncertainty (Appendx C).
-Neither model was capable of making forecasts of proportional cover change distinguishable from 0 (i.e., no change in population state) when we included model error and parameter uncertainty.
+smithiii* were very uncertain due to a very high instrinsic rate of recruitment combined with uncertainty in climate coefficients which lead to high recruitment boom years and subsequent busts when young plants suffer high mortality (Appendx C).
+When we included model error and parameter uncertainty, forecast changes in proportional cover always spanned a wide range of negative to positive values.
+In other words, neither model could predict whether a climate perturbation would increase or decrease equilibrium population size.
 
@@
 ```{r comment_out, echo=FALSE, eval=FALSE}
-Equilibrium cover from both models was sensitive to climate (Fig.
+# Equilibrium cover from both models was sensitive to climate (Fig.
 3a-d).
@@
 ----------
-Population models built using individual-level data allow inference on demographic processes, but their ability to forecast future population states is limited by the spatial extent of the data.
-Population-level data, like percent cover or discrete density states, are much easier to collect of broad spatial extents, so populations built using such data offer an appealing alternative to traditional population models [@Queenborough2011].
-However, at their core, density-structured models rely on individual level data aggregated to a population level metric.
-This creates a potential problem if such models are to be used in a climate change context because inidividuals respond to climate, not populations [@Clark2012].
+Population models built using individual-level data allow inference on demographic processes, but they can only forecast future population states across the (typically limited) spatial extent of the observations.
+Population-level data are much easier to collect across broad spatial extents, so models built using such data offer an appealing alternative to traditional population models [@Queenborough2011].
+However, density-structured models rely on the aggregation of individual level data.
+This creates a potential problem if such models are to be used in a climate change context because it is individuals, not populations, which respond to climate [@Clark2012].
 Are models based on population level metrics as sensitive to climate as models based on individual level metrics?
@@
 Do we need detailed demographic data to forecast the impacts of climate change?
-These are the questions we sought to answer here.
-
-Using individual-level and population-level forms of the same dataset, we were able to directly compare a traditional demographic modeling approach to a population model based on percent cover data.
-Our quad-based model (QBM) is based on percent cover data and so is in the spirit of density-structured models.
-In terms of each model's forecasting ability, the IPM outperformed the QBM: the IPM was more accurate and precise than the QBM (Table 1).
-As we suggest in the introduction, this could be due to differences in sample size or the effect of averaging over climate responses unique to individual-level vital rates.
-
-We found no evidence that increasing sample size at the quadrat level would lead to higher precision of climate coefficient estimates (Fig.
+
+### The importance of demographic data
+
+Our comparison of a traditional, demographic population model (the IPM) with a model inspired by density-structured models (the QBM) showed that the IPM outperformed the QBM: the IPM was more accurate and precise than the QBM in out-of-sample cross validation (Table 1).
+The superiority of the IPM could reflect either differences in sample size or the effect of averaging over unique effects of climate on each individual-level vital rate.
+Although increasing sample size of quadrat percent-cover observations would be easy to do in the field, we found little evidence that it would lead to higher precision of climate coefficient estimates (Fig.
 4).
@@
 
-Thus, for each species the QBM is "missing" climate signals associated with at least one vital rate.
+Our interpretation is that the QBM is "missing" climate signals associated with at least one vital rate for each species.
 This leads to inaccurate and imprecise forecasts because the QBM statistical model struggles to explain variation due to climate variables that have positive and negative impacts on different vital rates.
@@
 3), statistical models of population-level data will fail.
-This result further confirms related work on the importance of individual-level data to forecast population responses to exogenous drivers [@Clark2011; @Clark2011b; @Clark2012; @Galvan2014].
-
-These results lead us to conclude that detailed demographic data is necessary to forecast climate change impacts on plant populations when vital rates have unique climate responses.
-This is unwelcome news since this data is difficult to collect and the models built on such data are of little use to land managers that make decisions at scales beyond that of traditional population models [@Queenborough2011].
-While density-structured approaches may fail when climate covariates are considered, there are other alternatives.
+Our result is consistent with related work on the importance of individual-level data to forecast population responses to exogenous drivers [@Clark2011; @Clark2011b; @Clark2012; @Galvan2014].
+
+Detailed demographic data appears to be necessary to forecast climate change impacts on plant populations when vital rates have unique climate responses.
+How then can we build models to make forecasts for the landscape and regional scales beyond the scope of traditional population models [@Queenborough2011]?
+There are alternatives to density-structured models.
 For example, @Clark2011 use Forest Inventory and Analysis (FIA) data to parameterize a population model with multiple vital rates and climate dependence.
-Another example are distributed efforts like PlantPopNet (http://plantago.plantpopnet.com) that will allow researchers to estimate variation around climate responses for widespread species by taking advantage of spatial variation in climate [e.g.
+Distributed efforts such as PlantPopNet (http://plantago.plantpopnet.com) will allow researchers to estimate variation around climate responses for widespread species by taking advantage of spatial variation in climate [e.g.
 @Doak2010].
-Lastly, we foresee new approaches on the horizon that leverage photo/video of plots and advanced object recognition algorithms [e.g.
-@Liu2014] to streamline plant mapping and digitizing efforts.
-
-What does all this mean for model forecasts?
-The answer is more nuanced than one might expect.
-The typical approach in ecology is to use point estimates of model parameters to project populations forward according to the specified model; usually allowing for some variability around the determinstic process (e.g., @Battin2007; @Jenouvrier2009; @Adler2012).
-If we follow tradition and calculate the mean response to climate change over many iterations of the model with only model error and temporal variation included, the IPM and the QBM produce opposing forecasts for three of four species (Fig.
+Finally, new approaches on the horizon that leverage photo/video of plots and advanced object recognition algorithms [e.g.
+@Liu2014] will increase the efficiency of plant mapping and digitizing efforts.
+
+### The challenge of uncertainty
+
+An important, but unexpected, result of our analysis was the great uncertainty in forecasts, even for our best model.
+The typical approach in ecology is to use point estimates of model parameters to project populations forward according to the specified model, usually allowing for some variability around the determinstic process [e.g.
+@Battin2007; @Jenouvrier2009; @Adler2012].
+If we follow tradition and calculate the mean response to climate perturbation with only model error and interannual variation included, the IPM and the QBM produce opposing forecasts for three of four species (Fig.
 D1).
+It would be tempting to interpret this inconsistency as further evidence for the superiority of the IPM.
 However, if we introduce parameter uncertainty, the forecasts are actually indistinguishable (Fig.
-6).
-IPM forecasts are generally more precise, as expected from model validation (Table 1).
-The real story is not that the QBM produces "wrong" forecasts, but that it produces highly uncertain forecasts.
-Our sample size analysis reveals that this uncertainty is not easily overcome by including more data (Fig.
-4).
-The only way to reduce uncertainty is to model these populations based on individual-level data where climate dependence can be properly identified (Fig.
-5).
-Yet, even the IPM forecasts were highly variable (Fig.
-6).
-
-Our goal was not to make any explicit forecast for the future state of these populations based on predicted climate change.
-But our results highlight the state of affairs in ecology when it comes to forecasting the impacts of climate change.
-The analysis we conducted here could be considered, with some exceptions of course, at the forefront of ecological forecasting in terms of the statistical approach employed (hierarchical Bayesian), the type of population model we used (stochastic IPM with parameter uncertainty), and the amount of high quality data we had at our disposal (14 years of individual level data).
+6), though the IPM projections are generally more precise (consistent with our cross-validation results).
+The real story is that both models produce highly uncertain forecasts.
+For all species, the 90% quantiles of predicted changes in population size overlapped zero; we cannot even predict whether a change in precipitation or temperature will cause populations to increase or decrease.
+\todo{Add in sentence on larger perturbations.}
+
+Our results highlight the state of affairs in ecology when it comes to forecasting the impacts of climate change.
+The analysis we conducted here could be considered at the forefront of ecological forecasting with repsect to the statistical approach employed (hierarchical Bayesian), the type of population model we used (density-dependent, stochastic IPM with parameter uncertainty), and the amount of high quality data we had at our disposal (14 years of individual level data).
 Yet, model predictions proved so uncertain that any forecast, when bounded with model and parameter uncertainty, would be at best not useful and at worst meaningless.
-For all species, the 90% quantiles of predicted changes in population size overlapped zero; we cannot even predict whether a change in precipitation or temperature will cause populations to increase or decrease.
+
 How might we improve on this state of affairs?
-
 First, forecasts could be improved by matching the spatial scale of predictor variables with the spatial scale of observations.
@@
 Thus, even if we fit models to individual-level data, we are missing the key interaction point between weather and individual plants [@Clark2011b] because all observations share the same climate covariates.
-Demographic studies should be designed with at least plot level measurements of climate related variables (e.g., soil moisture).
-
+Demographic studies should be designed with at least plot-level measurements of climate related variables (e.g., soil moisture).
 Second, accurately detecting climate signals will take even longer time series.
-Recent theoretical work on detecting climate signals in noisy data suggests that even advanced approaches to parameter fitting like LASSO, functional linear models (splines), and Random Forest models require 20-25 year time series (Teller et al., in review).
-Alternatively, as we suggest above, Teller et al.
-(in review) also find that matching the scale of the response and predictors improves estimate precision.
-
-Third, ecologists as a community need to get serious about reporting uncertainty.
-There is a strong culture around explicitly considering model uncertainty, but parameter uncertainty is often ignored.
-In some cases this is because the easiest statistical methods do no make propogating parameter uncertainty a straighforward task.
-Even Bayesian approaches that allow integration of model fitting and forecasting [@Hobbs2015] are not simple when using modeling approaches like integral projection models that separate the model fitting and simulation stages [@Rees2009].
+Recent theoretical work on detecting climate signals in noisy data suggests that even advanced approaches to parameter fitting require 20-25 year time series (Teller et al., in review).
+Third, ecologists need a stronger commitment to reporting uncertainty.
+Although most modeling studies explicitly consider model uncertainty, parameter uncertainty is often ignored.
+In some cases this is because the most convenient statistical methods make it difficult to propogate parameter uncertainty.
+Yet even Bayesian approaches that allow integration of model fitting and forecasting [@Hobbs2015] are not simple when using modeling approaches like integral projection models that separate the model fitting and simulation stages [@Rees2009].
 However, as we have done here, it is still possible to include parameter uncertainty by drawing parameter values from MCMC iterations, taking care to draw all parameters from the same chain and iteration to account for their correlations.
 Only by being honest about our forecasts can we begin to produce better ones, and forecasts reported without parameter error are disingenuous.
-In many cases, ignoring parameter error is justifiable when investigating basic demographic mechanisms, but it is never justifiable if forecasts are being made.
+In many cases, ignoring parameter error is justifiable when investigating basic demographic mechanisms, but it is not justifiable when forecasting is the goal.
 
@@
 This work is not a critique of density-structured population models.
-In some cases and for certain species where climate dependence is weak or concentrated within a single vital rate, population models based on aggregated data may prove useful and unbiased.
-However, our work here is the first comparison, to our knowledge, of population models based on individual and aggregated forms of the same data in a climate change context.
-Our results confirm theoretical arguments [@Clark2011b] and empirical evidence [@Clark2011; @Clark2012] that individual responses are critical to predicting species' responses to climate change.
-Thus, forecasts from population-level models should be viewed with caution and should never be unaccompianed by uncertainty.
-Given the importance of demographic data and its current difficulty to collect, we advocate for research on modern methods to collect demographic data more efficiently across environmental gradients in space and time.
-
-Our results also offer a cautionary tale because uncertainty around forecasts was large for both model types.
-Which leads us to our most pessimistic conclusion: even with 14 years of detailed demographic data and sophisticated modeling techniques we failed to produce forecasts with any level of acceptable uncertainty.
-In our view, uncertainty of climate change related forecasts can be reduced by (1) longer time series and (2) climate covariates that match the scale of inference (e.g., plot rather than landscape level climate/weather metrics).
-Still, given the poor performance of the quad-based model, it seems there is no short cut to producing accurate and precise population forecasts.
+We are confident that density-structured models will prove to be a valuable tool for many applications.
+However, our analysis represents the first comparison, to our knowledge, of population models based on individual and aggregated forms of the same data in a climate change context.
+Our results confirm theoretical arguments [@Clark2011b] and empirical evidence [@Clark2011; @Clark2012] that individual responses are critical for predicting species' responses to climate change.
+It seems there is no short cut to producing accurate and precise population forecasts.
 Do we need detailed demographic data to forecast the impacts of climate change on populations?
 Probably.
+Given the importance of demographic data and its current collection cost, we need modern methods to collect demographic data more efficiently across environmental gradients in space and time.
+
+Our results also offer a cautionary tale because forecast uncertainty was large for both model types.
+Even with 14 years of detailed demographic data and sophisticated modeling techniques, we failed to produce forecasts with any level of acceptable uncertainty.
+Uncertainty of climate change related forecasts can be reduced by (1) longer time series and (2) climate covariates that match the scale of inference (e.g., plot rather than landscape level climate/weather metrics).
 
@@
 ipmfiles_mean <- ipmfiles_mean[grep("cover", ipmfiles_mean)]
+torm <- c(grep("10", ipmfiles_mean), grep("20", ipmfiles_mean), grep("30", ipmfiles_mean))
+ipmfiles_mean <- ipmfiles_mean[-torm]
 for(file in ipmfiles_mean){

### 2015-07-31 manuscript/manuscript.tex
adds new results from upping the perturbation percentages for climate

--- 2015-07-17 manuscript/manuscript.tex
@@
 Forecasting future states of populations has taken on new urgency as the
-rate of climate change increases. Traditional plant population models
-have limited utility in this regard because they are based on detailed
-demographic data from small, localized plots. These models are difficult
-to scale up to spatial scales relevant to land managers that require
-such forecasts to make decisions. To overcome the data limitations of
-traditional population models, some have proposed population models
-based on population level, rather than individual level, data that is
-much easier to collect over broad spatial scales. Using such models
-violates a central assumption of ecology: individuals respond to
-weather, not populations to climate. Here, we test whether this
-assumption is important when forecasting climate change impacts on four
-perennial grass species in a semi-arid Montana grassland. We
-parameterized two population models, one based on inidividual-level data
-with three vital rates and one on population-level data from the same
-dataset (percent cover of quadrats), and compared their accuracy,
-precision, and sensitivity to climate. The individual level model was
-more accurate and precise than the aggregate level model when predicting
-out-of-sample observations. When comparing climate effects from both
-models, the population-level model tends to ``miss'' important climate
-effects from at least one vital rate for each species. We also find that
-increasing the sample size at the population-level would not necessarily
-reduce forecast certainty; meaning the only way to reduce uncertainty is
-to capture unique climate dependence on individual vital rates. It
-appears there is no short cut to forecasting climate change impacts on
-plant populations --- detailed demographic data is essential. However,
-forecasts from our individual-level model were very uncertain, so we
-advocate for a focus on new methods to collect demographic data more
-efficiently across environmental gradients in space and time.
+rate of climate change increases. Traditional plant population models,
+typically built using detailed demographic observations from one study
+site, can address climate change impacts at one location, but are
+difficult to scale up to the landscape and regional scales relevant to
+management decisions. One proposal to overcome the limitations of
+traditional population models is to build models using population-level
+data that is much easier to collect over broad spatial scales than
+individual-level data. However, such models ignore the fact that climate
+drives population growth through its influence on individual
+performance. Here, we test the consequences of aggregating individual
+responses when forecasting climate change impacts on four perennial
+grass species in a semi-arid Montana grassland. We parameterized two
+population models, one based on individual-level data (survival, growth
+and recruitment) and one on population-level data from the same dataset
+(percent cover of quadrats), and compared their accuracy, precision, and
+sensitivity to climate. The individual level model was more accurate and
+precise than the aggregate level model when predicting out-of-sample
+observations. When comparing climate effects from both models, the
+population-level model tends to ``miss'' important climate effects from
+at least one vital rate for each species. Increasing the sample size at
+the population-level would not necessarily reduce forecast uncertainty;
+the way to reduce uncertainty is to capture unique climate dependence of
+individual vital rates. Our analysis indicates that there is no short
+cut to forecasting climate change impacts on plant populations ---
+detailed demographic data is essential. Despite the superiority of the
+individual-level model, the forecasts it generated still were too
+uncertain to be useful for decision-makers. We need new methods to
+collect demographic data more efficiently across environmental gradients
+in space and time.
 
@@
 forecast the impacts of environmental change (Clark et al. 2001, Petchey
-et al. 2015). To do so requires sophisticated modeling approaches that
-fully account for uncertainty and variability in the ecological process
-and associated parameters (Luo et al. 2011). This requires large amounts
-of data collected over large spatio-temporal extents.
-State-of-the-science modeling techniques cannot overcome data
-limitations. Such is the case for many population models.
+et al. 2015). Forecasts require sophisticated modeling approaches that
+fully account for uncertainty and variability in both ecological process
+and model parameters (Luo et al. 2011). This requires large amounts of
+data collected over broad spatio-temporal extents, and
+state-of-the-science modeling techniques cannot overcome data
+limitations. Population models are a case in point.
 
@@
 (Clark et al. 2010, 2012, Freckleton et al. 2011, Queenborough et al.
-2011). The major hurdle is that most population models are built using
-data from a single study site because collecting those data, which
-involves tracking the fates of individuals plants, is so difficult. The
-resulting models cannot be applied to the landscape and regional scales
-relevant to decision-makers without information about how the fitted
-parameters respond to spatial variation in biotic and abiotic drivers
-(Sæther et al. 2007). The limited spatial extent of individual-level
-demographic datasets constrains our ability to use population models to
-address applied questions about the consequences of climate change.
-
-The inability of many population models to address landscape-scale
+2011). The problem is that most population models are built using data
+from a single study site because collecting those data, which involves
+tracking the fates of individuals plants, is so difficult. The resulting
+models cannot be applied to the landscape and regional scales relevant
+to decision-makers without information about how the fitted parameters
+respond to spatial variation in biotic and abiotic drivers (Sæther et
+al. 2007). The limited spatial extent of individual-level demographic
+datasets constrains our ability to use population models to address
+applied questions about the consequences of climate change.
+
+The inability of most population models to address landscape-scale
 problems may explain why land managers and conservation planners have
@@
 Ideally, researchers would provide managers with landscape-scale
-population models, combing the extent of SDMs with information about
+population models, combining the extent of SDMs with information about
 dynamics and species abundances.
@@
 farms, all collected by two people in 6 weeks. This is far more data
-from a far greater spatial extent than possible if measuring individual
-plant demography (in a world of limited time and money, at least). The
+from a far greater spatial extent than would be possible if measuring
+individual plant demography (in a world of limited time and money). The
 appeal of density-structured approaches is clear. However, none of these
@@
 that respond to climate variables (Clark et al. 2012). Ignoring this
-fact puts us in uneasy proximity to an ``ecological fallacy'', where
-inference about the individual relies on statistical inference on the
-group (Piantadosi et al. 1988). Population growth (or decline) is an
-epiphenomenon of demographic processes like survival, growth, and
-recruitment that occur at the level of individual plants. Climate can
-affect each demographic process in unique, potentially opossing, ways
-(Dalgleish et al. 2011). These unique climate responses may difficult
-resolve in statistical models based on population-level data where
-demographic processes are not identifiable. If population-level data
-cannot detect important impacts of climate drivers on populations, then
-population models built with such data will make poor forecasts.
+fact amounts to an ``ecological fallacy'', where inference about the
+individual relies on statistical inference on the group (Piantadosi et
+al. 1988). Population growth (or decline) is the outcome of demographic
+processes like survival, growth, and recruitment that occur at the level
+of individual plants. Climate can affect each demographic process in
+unique, potentially opossing, ways (Dalgleish et al. 2011). These unique
+climate responses may be difficult to resolve in statistical models
+based on population-level data where demographic processes are not
+identifiable. If population-level data cannot detect important impacts
+of climate drivers on populations, then population models built with
+such data will make poor forecasts.
 
 Here, we ask whether statistical and population models based on
-aggregated, population-level data can detect climate signals as wells as
-models based on individual-level data. We used a unique demographic
-dataset that tracks the fates of individual plants from four species
-over 14 years to build two kinds of single-species population models,
-traditional models using individual growth, survival, and recruitment
-data and alternative models based on basal cover. In both models,
-interannual variation is explained, in part, by climate covariates. We
-then performed simulations to quantify the sensitivities of species'
-cover to small perturbations in average precipitation and temperature.
-We found that population models based on detailed demographic data are
-more accurate and precise than models based on aggregated data. Our
-results suggest the popuation-level model is less accurate and precise
-because important demographic climate signals go undetected. For these
-species at this location, detailed demographic data appears necessary to
-make accurate forecasts. A worrying caveat to our work is that forecasts
-from both models were very uncertain when we considered full process and
-parameter uncertainty. Even 14 years worth of demographic data may not
-be sufficient to make useful forecasts.
+aggregated, population-level data can predict climate change impacts as
+wells as models based on individual-level data. We used a unique
+demographic dataset that tracks the fates of individual plants from four
+species over 14 years to build two kinds of single-species population
+models, traditional models using individual growth, survival, and
+recruitment data and alternative models based on basal cover. In both
+models, interannual variation is explained, in part, by climate
+covariates. We then performed simulations to quantify the sensitivities
+of species' cover to small perturbations in average precipitation and
+temperature. Individual level models were clearly better, but,
+unfortunately, still too uncertain to inform management decisions.
 
@@
 extracted using a computer program (Lauenroth and Adler 2008, Chu et al.
-2014). Daily climate data, which we aggregated into climate variables of
-interest, are available for the duration of the data collection period
-(1932 - 1945) from the Miles City airport, Wiley Field, 9 km from the
-study site.
+2014). Daily climate data are available for the duration of the data
+collection period (1932 - 1945) from the Miles City airport, Wiley
+Field, 9 km from the study site.
 
@@
 blocks of our population models are vital rate regressions. For
-individual-level data we fit models for survival, growth, and
-recruitment for each species. At the quadrat-level we fit a single
+individual-level data, we fit models for survival, growth, and
+recruitment for each species. At the quadrat-level, we fit a single
 regression model for population growth. We describe the statistical
@@
 size interactions. Climate $\times$ size interactions are for climate
-main effects only, that is we do not include interactions among size and
-interacting climate effects.
+main effects only; we do not include interactions between size and pairs
+of interacting climate effects.
 
@@
 
-We modeled growth as Gaussian process describing genet size at time
+We modeled growth as a Gaussian process describing genet size at time
 $t+1$ as a function of size at $t$ and climate covariates:
@@
 where $x$ is log genet size and all other parameters are as described
-for the survival regression.
+for the survival regression. We capture non-constant error variance in
+growth by modeling the variance around the growth regression
+($\varepsilon$) as a nonlinear function of predicted genet size:
+
+\begin{align}
+\varepsilon_{ij,t} = a e^{b x_{ijQ,t+1}}
+\end{align}
 
@@
 model parameters from three parallel MCMC chains run for 1,000
-iterations after discarding an initial 1,000 iterations. We recignize
+iterations after discarding an initial 1,000 iterations. We recognize
 such short MCMC chains may surprise those more familiar with other MCMC
@@
 an Integral Projection Model (IPM) to model populations based on
-individual level data {[}cite Ellner and Rees 2006{]} and an quadrat
-based version of an individually-based model (Quadrat-Based Model, QBM)
-to model populations based on quadrat level data. We describe each in
-turn.
+individual level data (Ellner and Rees 2006) and a quadrat-based version
+of an individually-based model (Quadrat-Based Model, QBM) to model
+populations based on quadrat level data. We describe each in turn.
 
@@
 We use an environmentally stochastic IPM (Rees and Ellner 2009) that
-includes the random year effects and the climate covariates from the
-vital rate statistical models. However, for some simulations, we ignore
-the random year effects so that only the climate effects drive
-interannual variation. Our IPM follows the specification of Chu and
-Adler (2015) where the population of species \emph{j} is a density
-function $n(u_{j},t)$ giving the density of sized-\emph{u} genets at
-time \emph{t}. Genet size is on the natural log scale, so that
-$n(u_{j},t)du$ is the number of genets whose area (on the arithmetic
-scale) is between $e^{u_{j}}$ and $e^{u_{j}+du}$. So, the density
-function for any size \emph{v} at time $t+1$ is
+includes the climate covariates from the vital rate statistical models.
+In all simulations we ignore the random year effects so that interannual
+variation is driven solely by climate. We fit the random year effects in
+the vital rate regressions to avoid over-attributing variation to
+climate covariates. Our IPM follows the specification of Chu and Adler
+(2015) where the population of species \emph{j} is a density function
+$n(u_{j},t)$ giving the density of sized-\emph{u} genets at time
+\emph{t}. Genet size is on the natural log scale, so that $n(u_{j},t)du$
+is the number of genets whose area (on the arithmetic scale) is between
+$e^{u_{j}}$ and $e^{u_{j}+du}$. The density function for any size
+\emph{v} at time $t+1$ is
 
@@
 beyond the range of observed genet sizes.
+
+Since the IPM is spatially-implicit, we cannot calculate neighborhood
+crowding for specific genets ($w_{ij}$). Instead, we use an
+approximation ($\bar{\bold{w_{j}}}$) that captures the essential
+features of neighborhood interactions (Adler et al. 2010). We also use a
+`no-overlap' rule for conspecific genets to approximate the
+overdispersion of large genets in space (Adler et al. 2010, 2012, Chu
+and Adler 2015).
 
@@
 to incorporate the recruitment function in the spatially-implicit IPM.
-\todo{Add spatial structure mention.}
 
@@
 Our quad-based model (QBM) perfectly mirrors its statistical description
-(Eqs. 8-9). We use the same approach for drawing parameter values as
+(Eqs. 9-10). We use the same approach for drawing parameter values as
 described for the IPM. After drawing the appropriate parameter set, we
 calculate the mean response (population cover at \emph{t+1} = $x_{t+1}$)
-according to Eq. 8. We then make a random draw from a {[}0,1{]}
-truncated lognormal distribution with mean equal to $x_{t+1}$ from Eq. 8
-and the variance estimate from the fitted model. We can then iterate the
+according to Eq. 9. We then make a random draw from a {[}0,1{]}
+truncated lognormal distribution with mean equal to $x_{t+1}$ from Eq. 9
+and the variance estimate from the fitted model. We can then project the
 model forward by drawing a new parameter set (unique to climate year and
-MCMC iteration) at each timestep.
+MCMC iteration) at each timestep. As with the IPM, random year effects
+are ignored for all simulations.
 
@@
 
-Our main goal in this paper is to see if models based on aggregated data
-are as sensitive to climate covariates as models based on individual
-level data. So, with our fitted and validated models in hand, we ran
-simulations for each model type (IPM and QBM) under four climate
-perturbation scenarios: (1) observed climate, (2) precipitation
-increased by 1\%, (3) temperature increased by 1\%, and (4)
-precipitation and temperature increased by 1\%. We ran the simulations
-for 2,500 time steps, enough to estimate equilibrium cover after
-discarding an initial 500 time steps as burn-in. Each simulation was run
-under two parameter scenarios: (1) using mean parameter estimates and
-(2) using randomly drawn parameters from the MCMC chain. We use (1) to
-detect the overall sensitivity of equilibrium cover to climate, and we
-use (2) to show the impact of model and parameter uncertainty on
-forecast precision.
+To see if models based on aggregated data are as sensitive to climate
+covariates as models based on individual level data. So, with our fitted
+and validated models in hand, we ran simulations for each model type
+(IPM and QBM) under four climate perturbation scenarios: (1) observed
+climate, (2) precipitation increased by 1\%, (3) temperature increased
+by 1\%, and (4) precipitation and temperature increased by 1\%. We ran
+the simulations for 2,500 time steps, enough to estimate equilibrium
+cover after discarding an initial 500 time steps as burn-in. Each
+simulation was run under two parameter scenarios: (1) using mean
+parameter estimates and (2) using randomly drawn parameters from the
+MCMC chain. We use (1) to detect the overall sensitivity of equilibrium
+cover to climate, and we use (2) to show the impact of model and
+parameter uncertainty on forecast precision.
 
@@
 Our expectation is that the IPM will produce more accurate and precise
-forecasts. This could be due to differences in sample sizes leading to
-larger parameter uncertainty for the QBM, or due to the QBM climate
-effects being weakly associated with one or more vital rate climate
-effects at the individual level. To assess the impact of sample size on
-QBM parameter uncertainty we refit the QBM statistical model (Eqs. 8-9)
-after removing sets of 2, 5, 10, and 15 quadrats. We fit 10 models at
-each level of quadrat removal (2, 5, 10, 15 quadrats), removing a
-different randomly selected set of quadrats for each fit. We calculated
-the standard deviation of climate main effects (pptlag, ppt1, ppt2,
-TmeanSpr1, and TmeanSpr2) for each model and averaged those over
+forecasts. This could be due to (1) the smaller sample size of the
+quadrat level data sets compared to the individual level data sets,
+leading to larger parameter uncertainty for the QBM, or (2) the QBM
+climate effects being weakly associated with one or more vital rate
+climate effects at the individual level. To assess the impact of sample
+size on QBM parameter uncertainty we refit the QBM statistical model
+(Eqs. 9-10) after removing sets of 2, 5, 10, and 15 quadrats. We fit 10
+models at each level of quadrat removal (2, 5, 10, 15 quadrats),
+removing a different randomly selected set of quadrats for each fit. We
+calculated the standard deviation of climate main effects (pptLag, ppt1,
+ppt2, TmeanSpr1, and TmeanSpr2) for each model and averaged those over
 replicates within each set of quadrat removals. This allowed us to
@@
 QBM climate coefficients against each vital rate model's climate
-coefficients and calculate Pearson's $\rho$. Strong correlations
+coefficients and calculated Pearson's $\rho$. Strong correlations
 indicate the QBM is capable of detecting climate effects associated with
@@
 but a strong positive effect on survival with increased temperature
-(Fig. 3d). The climate impact of recruitment on equilibrium cover was
-negative for precipitation and temperature increases (Fig. 3d). At least
-two of three vital rates were sensitive to climate for each species
-(Fig. 3).
+(Fig. 3d). Equilibrium cover responded negatively when increased
+precipitation and temperature affect recruitment (Fig. 3d). At least two
+of three vital rates were sensitive to climate for each species (Fig.
+3).
 
@@
 Sample size had a relatively weak effect on QBM climate parameter
-uncertainty after the number of quadrats exceeded about 10 (Fig. 5).
-Inverse-gaussian fits show that increasing sample size beyond the number
-of quadrats we used will results in deminishing returns in terms of
-parameter certainty (Fig. 5).
+uncertainty after the number of quadrats used in fitting exceeded about
+10 (Fig. 5). Inverse-gaussian fits show that increasing sample size
+beyond the number of quadrats we used would result in diminishing
+returns in terms of parameter certainty (Fig. 5).
 
@@
 Forecasts based on 1\% climate changes were extremely uncertain when we
-considered model error and parameter uncertainty (Fig. 6). As expected
+considered model error and parameter uncertainty (Fig. 6; simulations
+with mean parameters are in Appendix D for comparison). As expected
 based on model validation (Table 1), QBM projections were more uncertain
 than IPM projections for all species except \emph{P. smithiii} (Fig. 6).
-IPM forecasts for \emph{P. smithiii} were very uncertain due to
-overcompensating density dependence and nonlinearities recruitment that
-get amplified by parameter uncertainty (Appendx C). Neither model was
-capable of making forecasts of proportional cover change distinguishable
-from 0 (i.e., no change in population state) when we included model
-error and parameter uncertainty.
+IPM forecasts for \emph{P. smithiii} were very uncertain due to a very
+high instrinsic rate of recruitment combined with uncertainty in climate
+coefficients which lead to high recruitment boom years and subsequent
+busts when young plants suffer high mortality (Appendx C). When we
+included model error and parameter uncertainty, forecast changes in
+proportional cover always spanned a wide range of negative to positive
+values. In other words, neither model could predict whether a climate
+perturbation would increase or decrease equilibrium population size.
 
@@
 Population models built using individual-level data allow inference on
-demographic processes, but their ability to forecast future population
-states is limited by the spatial extent of the data. Population-level
-data, like percent cover or discrete density states, are much easier to
-collect of broad spatial extents, so populations built using such data
-offer an appealing alternative to traditional population models
-(Queenborough et al. 2011). However, at their core, density-structured
-models rely on individual level data aggregated to a population level
-metric. This creates a potential problem if such models are to be used
-in a climate change context because inidividuals respond to climate, not
-populations (Clark et al. 2012). Are models based on population level
-metrics as sensitive to climate as models based on individual level
-metrics? Do these two types of models produce consistent forecasts? Do
-we need detailed demographic data to forecast the impacts of climate
-change? These are the questions we sought to answer here.
-
-Using individual-level and population-level forms of the same dataset,
-we were able to directly compare a traditional demographic modeling
-approach to a population model based on percent cover data. Our
-quad-based model (QBM) is based on percent cover data and so is in the
-spirit of density-structured models. In terms of each model's
-forecasting ability, the IPM outperformed the QBM: the IPM was more
-accurate and precise than the QBM (Table 1). As we suggest in the
-introduction, this could be due to differences in sample size or the
-effect of averaging over climate responses unique to individual-level
-vital rates.
-
-We found no evidence that increasing sample size at the quadrat level
-would lead to higher precision of climate coefficient estimates (Fig.
-4). We did, however, find evidence that the QBM statistical model failed
-to identify climate dependence for some vital rates (Fig. 5). For no
-species were climate effects from the QBM strongly correlated with all
-three vital rates (Fig. 5). Freckleton et al. (2011) acknowledge that
-averaging over complex stage dependence will lead to poorly specified
-models. This is analagous to our situation, but instead of averaging
-over complex life histories, we are averaging over complex climate
-dependence.
-
-Thus, for each species the QBM is ``missing'' climate signals associated
-with at least one vital rate. This leads to inaccurate and imprecise
-forecasts because the QBM statistical model struggles to explain
-variation due to climate variables that have positive and negative
-impacts on different vital rates. When this is the case, as it is for
-all our species to varying degrees (Fig. 3), statistical models of
-population-level data will fail. This result further confirms related
-work on the importance of individual-level data to forecast population
-responses to exogenous drivers (Clark et al. 2011a, 2011b, 2012, Galván
-et al. 2014).
-
-These results lead us to conclude that detailed demographic data is
-necessary to forecast climate change impacts on plant populations when
-vital rates have unique climate responses. This is unwelcome news since
-this data is difficult to collect and the models built on such data are
-of little use to land managers that make decisions at scales beyond that
-of traditional population models (Queenborough et al. 2011). While
-density-structured approaches may fail when climate covariates are
-considered, there are other alternatives. For example, Clark et al.
-(2011a) use Forest Inventory and Analysis (FIA) data to parameterize a
-population model with multiple vital rates and climate dependence.
-Another example are distributed efforts like PlantPopNet
-(\url{http://plantago.plantpopnet.com}) that will allow researchers to
-estimate variation around climate responses for widespread species by
-taking advantage of spatial variation in climate (e.g. Doak and Morris
-2010). Lastly, we foresee new approaches on the horizon that leverage
-photo/video of plots and advanced object recognition algorithms (e.g.
-Liu et al. 2014) to streamline plant mapping and digitizing efforts.
-
-What does all this mean for model forecasts? The answer is more nuanced
-than one might expect. The typical approach in ecology is to use point
-estimates of model parameters to project populations forward according
-to the specified model; usually allowing for some variability around the
-determinstic process (e.g., Battin et al. (2007); Jenouvrier et al.
-(2009); Adler et al. (2012)). If we follow tradition and calculate the
-mean response to climate change over many iterations of the model with
-only model error and temporal variation included, the IPM and the QBM
-produce opposing forecasts for three of four species (Fig. D1). However,
-if we introduce parameter uncertainty, the forecasts are actually
-indistinguishable (Fig. 6). IPM forecasts are generally more precise, as
-expected from model validation (Table 1). The real story is not that the
-QBM produces ``wrong'' forecasts, but that it produces highly uncertain
-forecasts. Our sample size analysis reveals that this uncertainty is not
-easily overcome by including more data (Fig. 4). The only way to reduce
-uncertainty is to model these populations based on individual-level data
-where climate dependence can be properly identified (Fig. 5). Yet, even
-the IPM forecasts were highly variable (Fig. 6).
-
-Our goal was not to make any explicit forecast for the future state of
-these populations based on predicted climate change. But our results
-highlight the state of affairs in ecology when it comes to forecasting
-the impacts of climate change. The analysis we conducted here could be
-considered, with some exceptions of course, at the forefront of
-ecological forecasting in terms of the statistical approach employed
-(hierarchical Bayesian), the type of population model we used
-(stochastic IPM with parameter uncertainty), and the amount of high
-quality data we had at our disposal (14 years of individual level data).
-Yet, model predictions proved so uncertain that any forecast, when
-bounded with model and parameter uncertainty, would be at best not
-useful and at worst meaningless. For all species, the 90\% quantiles of
+demographic processes, but they can only forecast future population
+states across the (typically limited) spatial extent of the
+observations. Population-level data are much easier to collect across
+broad spatial extents, so models built using such data offer an
+appealing alternative to traditional population models (Queenborough et
+al. 2011). However, density-structured models rely on the aggregation of
+individual level data. This creates a potential problem if such models
+are to be used in a climate change context because it is individuals,
+not populations, which respond to climate (Clark et al. 2012). Are
+models based on population level metrics as sensitive to climate as
+models based on individual level metrics? Do these two types of models
+produce consistent forecasts? Do we need detailed demographic data to
+forecast the impacts of climate change?
+
+\subsubsection{The importance of demographic
+data}\label{the-importance-of-demographic-data}
+
+Our comparison of a traditional, demographic population model (the IPM)
+with a model inspired by density-structured models (the QBM) showed that
+the IPM outperformed the QBM: the IPM was more accurate and precise than
+the QBM in out-of-sample cross validation (Table 1). The superiority of
+the IPM could reflect either differences in sample size or the effect of
+averaging over unique effects of climate on each individual-level vital
+rate. Although increasing sample size of quadrat percent-cover
+observations would be easy to do in the field, we found little evidence
+that it would lead to higher precision of climate coefficient estimates
+(Fig. 4). We did, however, find evidence that the QBM statistical model
+failed to identify climate dependence for some vital rates (Fig. 5). For
+no species were climate effects from the QBM strongly correlated with
+all three vital rates (Fig. 5). Freckleton et al. (2011) acknowledge
+that averaging over complex stage dependence will lead to poorly
+specified models. This is analagous to our situation, but instead of
+averaging over complex life histories, we are averaging over complex
+climate dependence.
+
+Our interpretation is that the QBM is ``missing'' climate signals
+associated with at least one vital rate for each species. This leads to
+inaccurate and imprecise forecasts because the QBM statistical model
+struggles to explain variation due to climate variables that have
+positive and negative impacts on different vital rates. When this is the
+case, as it is for all our species to varying degrees (Fig. 3),
+statistical models of population-level data will fail. Our result is
+consistent with related work on the importance of individual-level data
+to forecast population responses to exogenous drivers (Clark et al.
+2011a, 2011b, 2012, Galván et al. 2014).
+
+Detailed demographic data appears to be necessary to forecast climate
+change impacts on plant populations when vital rates have unique climate
+responses. How then can we build models to make forecasts for the
+landscape and regional scales beyond the scope of traditional population
+models (Queenborough et al. 2011)? There are alternatives to
+density-structured models. For example, Clark et al. (2011a) use Forest
+Inventory and Analysis (FIA) data to parameterize a population model
+with multiple vital rates and climate dependence. Distributed efforts
+such as PlantPopNet (\url{http://plantago.plantpopnet.com}) will allow
+researchers to estimate variation around climate responses for
+widespread species by taking advantage of spatial variation in climate
+(e.g. Doak and Morris 2010). Finally, new approaches on the horizon that
+leverage photo/video of plots and advanced object recognition algorithms
+(e.g. Liu et al. 2014) will increase the efficiency of plant mapping and
+digitizing efforts.
+
+\subsubsection{The challenge of
+uncertainty}\label{the-challenge-of-uncertainty}
+
+An important, but unexpected, result of our analysis was the great
+uncertainty in forecasts, even for our best model. The typical approach
+in ecology is to use point estimates of model parameters to project
+populations forward according to the specified model, usually allowing
+for some variability around the determinstic process (e.g. Battin et al.
+2007, Jenouvrier et al. 2009, Adler et al. 2012). If we follow tradition
+and calculate the mean response to climate perturbation with only model
+error and interannual variation included, the IPM and the QBM produce
+opposing forecasts for three of four species (Fig. D1). It would be
+tempting to interpret this inconsistency as further evidence for the
+superiority of the IPM. However, if we introduce parameter uncertainty,
+the forecasts are actually indistinguishable (Fig. 6), though the IPM
+projections are generally more precise (consistent with our
+cross-validation results). The real story is that both models produce
+highly uncertain forecasts. For all species, the 90\% quantiles of
 predicted changes in population size overlapped zero; we cannot even
 predict whether a change in precipitation or temperature will cause
-populations to increase or decrease. How might we improve on this state
-of affairs?
-
-First, forecasts could be improved by matching the spatial scale of
-predictor variables with the spatial scale of observations. One of the
-major limitations of the models we fit here is that the climate data are
-at a much larger scale than the individual level observations of plant
-size. Climate covariates only vary by year, with no spatial variability
-within years. Thus, even if we fit models to individual-level data, we
-are missing the key interaction point between weather and individual
-plants (Clark et al. 2011b) because all observations share the same
-climate covariates. Demographic studies should be designed with at least
-plot level measurements of climate related variables (e.g., soil
-moisture).
-
-Second, accurately detecting climate signals will take even longer time
-series. Recent theoretical work on detecting climate signals in noisy
-data suggests that even advanced approaches to parameter fitting like
-LASSO, functional linear models (splines), and Random Forest models
-require 20-25 year time series (Teller et al., in review).
-Alternatively, as we suggest above, Teller et al. (in review) also find
-that matching the scale of the response and predictors improves estimate
-precision.
-
-Third, ecologists as a community need to get serious about reporting
-uncertainty. There is a strong culture around explicitly considering
-model uncertainty, but parameter uncertainty is often ignored. In some
-cases this is because the easiest statistical methods do no make
-propogating parameter uncertainty a straighforward task. Even Bayesian
-approaches that allow integration of model fitting and forecasting
-(Hobbs and Hooten 2015) are not simple when using modeling approaches
-like integral projection models that separate the model fitting and
-simulation stages (Rees and Ellner 2009). However, as we have done here,
-it is still possible to include parameter uncertainty by drawing
-parameter values from MCMC iterations, taking care to draw all
-parameters from the same chain and iteration to account for their
-correlations. Only by being honest about our forecasts can we begin to
-produce better ones, and forecasts reported without parameter error are
-disingenuous. In many cases, ignoring parameter error is justifiable
-when investigating basic demographic mechanisms, but it is never
-justifiable if forecasts are being made.
+populations to increase or decrease.
+\todo{Add in sentence on larger perturbations.}
+
+Our results highlight the state of affairs in ecology when it comes to
+forecasting the impacts of climate change. The analysis we conducted
+here could be considered at the forefront of ecological forecasting with
+repsect to the statistical approach employed (hierarchical Bayesian),
+the type of population model we used (density-dependent, stochastic IPM
+with parameter uncertainty), and the amount of high quality data we had
+at our disposal (14 years of individual level data). Yet, model
+predictions proved so uncertain that any forecast, when bounded with
+model and parameter uncertainty, would be at best not useful and at
+worst meaningless.
+
+How might we improve on this state of affairs? First, forecasts could be
+improved by matching the spatial scale of predictor variables with the
+spatial scale of observations. One of the major limitations of the
+models we fit here is that the climate data are at a much larger scale
+than the individual level observations of plant size. Climate covariates
+only vary by year, with no spatial variability within years. Thus, even
+if we fit models to individual-level data, we are missing the key
+interaction point between weather and individual plants (Clark et al.
+2011b) because all observations share the same climate covariates.
+Demographic studies should be designed with at least plot-level
+measurements of climate related variables (e.g., soil moisture). Second,
+accurately detecting climate signals will take even longer time series.
+Recent theoretical work on detecting climate signals in noisy data
+suggests that even advanced approaches to parameter fitting require
+20-25 year time series (Teller et al., in review). Third, ecologists
+need a stronger commitment to reporting uncertainty. Although most
+modeling studies explicitly consider model uncertainty, parameter
+uncertainty is often ignored. In some cases this is because the most
+convenient statistical methods make it difficult to propogate parameter
+uncertainty. Yet even Bayesian approaches that allow integration of
+model fitting and forecasting (Hobbs and Hooten 2015) are not simple
+when using modeling approaches like integral projection models that
+separate the model fitting and simulation stages (Rees and Ellner 2009).
+However, as we have done here, it is still possible to include parameter
+uncertainty by drawing parameter values from MCMC iterations, taking
+care to draw all parameters from the same chain and iteration to account
+for their correlations. Only by being honest about our forecasts can we
+begin to produce better ones, and forecasts reported without parameter
+error are disingenuous. In many cases, ignoring parameter error is
+justifiable when investigating basic demographic mechanisms, but it is
+not justifiable when forecasting is the goal.
 
@@
 
-This work is not a critique of density-structured population models. In
-some cases and for certain species where climate dependence is weak or
-concentrated within a single vital rate, population models based on
-aggregated data may prove useful and unbiased. However, our work here is
-the first comparison, to our knowledge, of population models based on
-individual and aggregated forms of the same data in a climate change
-context. Our results confirm theoretical arguments (Clark et al. 2011b)
-and empirical evidence (Clark et al. 2011a, 2012) that individual
-responses are critical to predicting species' responses to climate
-change. Thus, forecasts from population-level models should be viewed
-with caution and should never be unaccompianed by uncertainty. Given the
-importance of demographic data and its current difficulty to collect, we
-advocate for research on modern methods to collect demographic data more
-efficiently across environmental gradients in space and time.
-
-Our results also offer a cautionary tale because uncertainty around
-forecasts was large for both model types. Which leads us to our most
-pessimistic conclusion: even with 14 years of detailed demographic data
-and sophisticated modeling techniques we failed to produce forecasts
-with any level of acceptable uncertainty. In our view, uncertainty of
-climate change related forecasts can be reduced by (1) longer time
+This work is not a critique of density-structured population models. We
+are confident that density-structured models will prove to be a valuable
+tool for many applications. However, our analysis represents the first
+comparison, to our knowledge, of population models based on individual
+and aggregated forms of the same data in a climate change context. Our
+results confirm theoretical arguments (Clark et al. 2011b) and empirical
+evidence (Clark et al. 2011a, 2012) that individual responses are
+critical for predicting species' responses to climate change. It seems
+there is no short cut to producing accurate and precise population
+forecasts. Do we need detailed demographic data to forecast the impacts
+of climate change on populations? Probably. Given the importance of
+demographic data and its current collection cost, we need modern methods
+to collect demographic data more efficiently across environmental
+gradients in space and time.
+
+Our results also offer a cautionary tale because forecast uncertainty
+was large for both model types. Even with 14 years of detailed
+demographic data and sophisticated modeling techniques, we failed to
+produce forecasts with any level of acceptable uncertainty. Uncertainty
+of climate change related forecasts can be reduced by (1) longer time
 series and (2) climate covariates that match the scale of inference
-(e.g., plot rather than landscape level climate/weather metrics). Still,
-given the poor performance of the quad-based model, it seems there is no
-short cut to producing accurate and precise population forecasts. Do we
-need detailed demographic data to forecast the impacts of climate change
-on populations? Probably.
+(e.g., plot rather than landscape level climate/weather metrics).
 
@@
 
+Adler, P. B., S. P. Ellner, and J. M. Levine. 2010. Coexistence of
+perennial plants: An embarrassment of niches.
+
 Anderson, J., L. Vermeire, and P. B. Adler. 2011. Fourteen years of
@@
 Ecological Explanation and Prediction Across Space and Time.
+
+Ellner, S. P., and M. Rees. 2006. Integral projection models for species
+with complex demography. The American naturalist 167:410--428.
 

### 2015-07-31 manuscript/manuscript.Rmd
fixes typo

@@
 ### Testing sensitivity to climate covariates
-To see if models based on aggregated data are as sensitive to climate covariates as models based on individual level data.
-So, with our fitted and validated models in hand, we ran simulations for each model type (IPM and QBM) under four climate perturbation scenarios: (1) observed climate, (2) precipitation increased by 1%, (3) temperature increased by 1%, and (4) precipitation and temperature increased by 1%.
+With our fitted and validated models in hand, we ran simulations for each model type (IPM and QBM) under four climate perturbation scenarios: (1) observed climate, (2) precipitation increased by 1%, (3) temperature increased by 1%, and (4) precipitation and temperature increased by 1%.
 We ran the simulations for 2,500 time steps, enough to estimate equilibrium cover after discarding an initial 500 time steps as burn-in.

### 2015-08-21 manuscript/manuscript.Rmd
updates ms with Britta's comments

@@
 The individual level model was more accurate and precise than the aggregate level model when predicting out-of-sample observations.
-When comparing climate effects from both models, the population-level model tends to "miss" important climate effects from at least one vital rate for each species.
+When comparing climate effects from both models, the population-level model tends to miss important climate effects from at least one vital rate for each species.
 Increasing the sample size at the population-level would not necessarily reduce forecast uncertainty; the way to reduce uncertainty is to capture unique climate dependence of individual vital rates.
-Our analysis indicates that there is no short cut to forecasting climate change impacts on plant populations — detailed demographic data is essential.
+Our analysis indicates that there is no shortcut to forecasting climate change impacts on plant populations — detailed demographic data is essential.
 Despite the superiority of the individual-level model, the forecasts it generated still were too uncertain to be useful for decision-makers.
@@
 Perhaps the greatest challenge for ecology in the 21st century is to forecast the impacts of environmental change [@Clark2001; @Petchey2015].
-Forecasts require sophisticated modeling approaches that fully account for uncertainty and variability in both ecological process and model parameters [@Luo2011a].
+Forecasts require sophisticated modeling approaches that fully account for uncertainty and variability in both ecological process and model parameters [@Luo2011a; but see @Perretti2013a for an argument against modeling the ecological process].
 This requires large amounts of data collected over broad spatio-temporal extents, and state-of-the-science modeling techniques cannot overcome data limitations.
@@
 However, the limitations of SDMs are well known [@Pearson2003; @Elith2009; @Araujo2012].
-Ideally, researchers would provide managers with landscape-scale population models, combining the extent of SDMs with information about dynamics and species abundances.
+Ideally, researchers would provide managers with landscape-scale population models, combining the extent of SDMs with information about dynamics and species abundances [@Schurr2012; @Merow2014].
 
@@
 
-To date, most empirical population modelers have relied on individual-level data, with few attempts to exploit population-level data.
+To date, most empirical population modelers have relied on individual-level data, with few attempts to capitalize on population-level measures.
 An important exception was an effort by @Taylor2004 to model the population growth rate of an invasive species to identify the best strategies for invasion control.
@@
 These unique climate responses may be difficult to resolve in statistical models based on population-level data where demographic processes are not identifiable.
-If population-level data cannot detect important impacts of climate drivers on populations, then population models built with such data will make poor forecasts.
-
-Here, we ask whether statistical and population models based on aggregated, population-level data can predict climate change impacts as wells as models based on individual-level data.
+If important climate drivers are missed because they are averaged over in population-level data, then population models built with such data will make uninformative or unreliable forecasts when confronted with novel climate covariates.
+
+Here, we ask whether statistical and population models based on aggregated, population-level data can predict climate change impacts as well as models based on individual-level data.
 We used a unique demographic dataset that tracks the fates of individual plants from four species over 14 years to build two kinds of single-species population models, traditional models using individual growth, survival, and recruitment data and alternative models based on basal cover.
@@
 For all species, the 90% quantiles of predicted changes in population size overlapped zero; we cannot even predict whether a change in precipitation or temperature will cause populations to increase or decrease.
-\todo{Add in sentence on larger perturbations.}
+This result held when we tried perturbing climate by 10% and 20% as well.
 
@@
 Second, accurately detecting climate signals will take even longer time series.
-Recent theoretical work on detecting climate signals in noisy data suggests that even advanced approaches to parameter fitting require 20-25 year time series (Teller et al., in review).
+Recent theoretical work on detecting climate signals in noisy data suggests that even advanced approaches to parameter fitting require 20-25 year time series [@Tellerinpress].
 Third, ecologists need a stronger commitment to reporting uncertainty.
@@
 Informal conversations with Stephen Ellner, Giles Hooker, Robin Snyder, and a series of meetings between the Adler and Weecology labs at USU sharpened our thinking.
+Brittany Teller and two anonymous reviewers provided comments that improved our manuscript.
 Compute, storage and other resources from the Division of Research Computing in the Office of Research and Graduate Studies at Utah State University are gratefully acknowledged.

### 2015-08-21 manuscript/manuscript.tex
updates ms with Britta's comments

@@
 observations. When comparing climate effects from both models, the
-population-level model tends to ``miss'' important climate effects from
-at least one vital rate for each species. Increasing the sample size at
-the population-level would not necessarily reduce forecast uncertainty;
-the way to reduce uncertainty is to capture unique climate dependence of
-individual vital rates. Our analysis indicates that there is no short
-cut to forecasting climate change impacts on plant populations ---
-detailed demographic data is essential. Despite the superiority of the
+population-level model tends to miss important climate effects from at
+least one vital rate for each species. Increasing the sample size at the
+population-level would not necessarily reduce forecast uncertainty; the
+way to reduce uncertainty is to capture unique climate dependence of
+individual vital rates. Our analysis indicates that there is no shortcut
+to forecasting climate change impacts on plant populations --- detailed
+demographic data is essential. Despite the superiority of the
 individual-level model, the forecasts it generated still were too
@@
 fully account for uncertainty and variability in both ecological process
-and model parameters (Luo et al. 2011). This requires large amounts of
-data collected over broad spatio-temporal extents, and
+and model parameters (Luo et al. 2011, but see Perretti et al. 2013 for
+an argument against modeling the ecological process). This requires
+large amounts of data collected over broad spatio-temporal extents, and
 state-of-the-science modeling techniques cannot overcome data
@@
 population models, combining the extent of SDMs with information about
-dynamics and species abundances.
+dynamics and species abundances (Schurr et al. 2012, Merow et al. 2014).
 
@@
 To date, most empirical population modelers have relied on
-individual-level data, with few attempts to exploit population-level
-data. An important exception was an effort by Taylor and Hastings (2004)
-to model the population growth rate of an invasive species to identify
-the best strategies for invasion control. They used a
-``density-structured'' model where the state variable is a discrete
-density state rather than a continuous density measure. Such models do
-not require individual level demographic data and can adequately
-describe population dynamics.
+individual-level data, with few attempts to capitalize on
+population-level measures. An important exception was an effort by
+Taylor and Hastings (2004) to model the population growth rate of an
+invasive species to identify the best strategies for invasion control.
+They used a ``density-structured'' model where the state variable is a
+discrete density state rather than a continuous density measure. Such
+models do not require individual level demographic data and can
+adequately describe population dynamics.
 
@@
 based on population-level data where demographic processes are not
-identifiable. If population-level data cannot detect important impacts
-of climate drivers on populations, then population models built with
-such data will make poor forecasts.
+identifiable. If important climate drivers are missed because they are
+averaged over in population-level data, then population models built
+with such data will make uninformative or unreliable forecasts when
+confronted with novel climate covariates.
 
@@
 aggregated, population-level data can predict climate change impacts as
-wells as models based on individual-level data. We used a unique
+well as models based on individual-level data. We used a unique
 demographic dataset that tracks the fates of individual plants from four
@@
 
-We use an environmentally stochastic IPM (Rees and Ellner 2009) that
+We use an environmentally stochastic IPM (Mark and Ellner 2009) that
 includes the climate covariates from the vital rate statistical models.
@@
 
-To see if models based on aggregated data are as sensitive to climate
-covariates as models based on individual level data. So, with our fitted
-and validated models in hand, we ran simulations for each model type
-(IPM and QBM) under four climate perturbation scenarios: (1) observed
-climate, (2) precipitation increased by 1\%, (3) temperature increased
-by 1\%, and (4) precipitation and temperature increased by 1\%. We ran
-the simulations for 2,500 time steps, enough to estimate equilibrium
-cover after discarding an initial 500 time steps as burn-in. Each
-simulation was run under two parameter scenarios: (1) using mean
-parameter estimates and (2) using randomly drawn parameters from the
-MCMC chain. We use (1) to detect the overall sensitivity of equilibrium
-cover to climate, and we use (2) to show the impact of model and
-parameter uncertainty on forecast precision.
+With our fitted and validated models in hand, we ran simulations for
+each model type (IPM and QBM) under four climate perturbation scenarios:
+(1) observed climate, (2) precipitation increased by 1\%, (3)
+temperature increased by 1\%, and (4) precipitation and temperature
+increased by 1\%. We ran the simulations for 2,500 time steps, enough to
+estimate equilibrium cover after discarding an initial 500 time steps as
+burn-in. Each simulation was run under two parameter scenarios: (1)
+using mean parameter estimates and (2) using randomly drawn parameters
+from the MCMC chain. We use (1) to detect the overall sensitivity of
+equilibrium cover to climate, and we use (2) to show the impact of model
+and parameter uncertainty on forecast precision.
 
@@
 predict whether a change in precipitation or temperature will cause
-populations to increase or decrease.
-\todo{Add in sentence on larger perturbations.}
+populations to increase or decrease. This result held when we tried
+perturbing climate by 10\% and 20\% as well.
 
@@
 suggests that even advanced approaches to parameter fitting require
-20-25 year time series (Teller et al., in review). Third, ecologists
-need a stronger commitment to reporting uncertainty. Although most
-modeling studies explicitly consider model uncertainty, parameter
-uncertainty is often ignored. In some cases this is because the most
-convenient statistical methods make it difficult to propogate parameter
+20-25 year time series (Teller et al. n.d.). Third, ecologists need a
+stronger commitment to reporting uncertainty. Although most modeling
+studies explicitly consider model uncertainty, parameter uncertainty is
+often ignored. In some cases this is because the most convenient
+statistical methods make it difficult to propogate parameter
 uncertainty. Yet even Bayesian approaches that allow integration of
@@
 when using modeling approaches like integral projection models that
-separate the model fitting and simulation stages (Rees and Ellner 2009).
+separate the model fitting and simulation stages (Mark and Ellner 2009).
 However, as we have done here, it is still possible to include parameter
@@
 series of meetings between the Adler and Weecology labs at USU sharpened
-our thinking. Compute, storage and other resources from the Division of
-Research Computing in the Office of Research and Graduate Studies at
-Utah State University are gratefully acknowledged.
+our thinking. Brittany Teller and two anonymous reviewers provided
+comments that improved our manuscript. Compute, storage and other
+resources from the Division of Research Computing in the Office of
+Research and Graduate Studies at Utah State University are gratefully
+acknowledged.
 
@@
 
+Mark, R., and S. P. Ellner. 2009. Integral projection models for
+populations in temporally varying environments. Ecological Monographs
+79:575--594.
+
+Merow, C., A. M. Latimer, A. M. Wilson, S. M. McMahon, A. G. Rebelo, and
+J. A. Silander. 2014. On using integral projection models to generate
+demographically driven predictions of species' distributions:
+development and validation using sparse data. Ecography 37:1167--1183.
+
 Pearson, R. G., and T. P. Dawson. 2003. Predicting the impacts of
@@
 models useful? Global Ecology and Biogeography 12:361--371.
+
+Perretti, C. T., S. B. Munch, and G. Sugihara. 2013. Model-free
+forecasting outperforms the correct mechanistic model for simulated and
+experimental data. Proceedings of the National Academy of Sciences of
+the United States of America 110:5253--5257.
 
@@
 
-Rees, M., and S. P. Ellner. 2009. Integral projection models for
-populations in temporally varying environments. Ecological Monographs
-79:575--594.
+Schurr, F. M., J. Pagel, J. S. Cabral, J. Groeneveld, O. Bykova,
+O'HaraR. B., F. Hartig, W. D. Kissling, H. P. Linder, G. F. Midgley, B.
+Schröder, A. Singer, and N. E. Zimmermann. 2012. How to understand
+species' niches and range dynamics: A demographic research agenda for
+biogeography. Journal of Biogeography 39:2146--2162.
 
@@
 
+Teller, B. J., P. B. Adler, C. B. Edwards, G. Hooker, R. E. Snyder, and
+S. P. Ellner. (n.d.). Linking demography with drivers: climate and
+competition. Methods in Ecology and Evolution.
+
 \end{document}

### 2015-08-22 manuscript/appC.Rmd
small appendix updates to match ms

--- 2015-07-15 manuscript/appC.Rmd
@@
 The uncertainty around PASM forecasts from the integral projection model is orders of magnitude larger than for any other species, and is even larger than the uncertainty around the quad-based model forecasts.
-It turns out that the recruitment parameters we estimate lead to overcompensating density-dependence that is amplified when those parameters are allowed to vary.
+It turns out that the estimated dispersion parameter for the negative binomial distribution is much higher for PASM than for other species (Appendix B).
+This is because PASM has higher recruitment rates than the other species and the relationship between number of recruits and quadrat cover (see main text) is weak.
+So, when we draw parameters from the MCMC chain for our simulations, even larger values for the dispersion parameter are possible.
+Thus, there can be very large "boom" years where recruitment is high.
 When we allow all parameters to vary except those associated with recruitment, uncertainty around PASM decreases to levels similar to those associated with other species (Fig.
 C1).
-So detailed demographic data is not a cure-all, and can sometimes lead to interesting dynamics when parameter uncertainty is considered.
 
@@
 
-```{r plot_it, echo=FALSE, message=FALSE, warning=FALSE, include=TRUE, fig.width=5, fig.cap="Equilibrium cover and 90% quantiles around the mean prediction from the IPM when model error and parameter uncertainty are propogated through the simulation phase for all vital rates except recruitment.", dependson="plot-options"}
+```{r plot_it, echo=FALSE, message=FALSE, warning=FALSE, include=TRUE, fig.width=5, fig.height=3, fig.cap="Equilibrium cover and 90% quantiles around the mean prediction from the IPM when model error and parameter uncertainty are propogated through the simulation phase for all vital rates except recruitment.", dependson="plot-options"}
 library(plyr)

### 2015-08-28 manuscript/manuscript.Rmd
small appendix updates to match ms
updates ms with PBA edits

@@
 \begin{centering}
-\textbf{\large{Do we need detailed demographic data to forecast the impacts of climate change on plant populations?}}
+\textbf{\large{Do we need demographic data to forecast population responses to climate change?}}
 
@@
 --------
-Forecasting future states of populations has taken on new urgency as the rate of climate change increases.
-Traditional plant population models, typically built using detailed demographic observations from one study site, can address climate change impacts at one location, but are difficult to scale up to the landscape and regional scales relevant to management decisions.
-One proposal to overcome the limitations of traditional population models is to build models using population-level data that is much easier to collect over broad spatial scales than individual-level data.
+Rapid climate change has generated growing interest in forecasts of future population trajectories.
+Traditional population models, typically built using detailed demographic observations from one study site, can address climate change impacts at one location, but are difficult to scale up to the landscape and regional scales relevant to management decisions.
+An alternative is to build models using population-level data that is much easier to collect over broad spatial scales than individual-level data.
 However, such models ignore the fact that climate drives population growth through its influence on individual performance.
-Here, we test the consequences of aggregating individual responses when forecasting climate change impacts on four perennial grass species in a semi-arid Montana grassland.
-We parameterized two population models, one based on individual-level data (survival, growth and recruitment) and one on population-level data from the same dataset (percent cover of quadrats), and compared their accuracy, precision, and sensitivity to climate.
-The individual level model was more accurate and precise than the aggregate level model when predicting out-of-sample observations.
-When comparing climate effects from both models, the population-level model tends to miss important climate effects from at least one vital rate for each species.
+Here, we test the consequences of aggregating individual responses when forecasting climate change impacts on four perennial grass species in a semi-arid grassland in Montana, USA.
+We parameterized two population models based on the same dataset, one based on individual-level data (survival, growth and recruitment) and one on population-level data (percent cover), and compared their accuracy, precision, and sensitivity to climate variables.
+The individual-level model was more accurate and precise than the aggregated model when predicting out-of-sample observations.
+When comparing climate effects from both models, the population-level model missed important climate effects from at least one vital rate for each species.
 Increasing the sample size at the population-level would not necessarily reduce forecast uncertainty; the way to reduce uncertainty is to capture unique climate dependence of individual vital rates.
@@
 Despite the superiority of the individual-level model, the forecasts it generated still were too uncertain to be useful for decision-makers.
-We need new methods to collect demographic data more efficiently across environmental gradients in space and time.
+We need new methods to collect demographic data efficiently across environmental gradients in space and time.
 
@@
 Forecasts require sophisticated modeling approaches that fully account for uncertainty and variability in both ecological process and model parameters [@Luo2011a; but see @Perretti2013a for an argument against modeling the ecological process].
-This requires large amounts of data collected over broad spatio-temporal extents, and state-of-the-science modeling techniques cannot overcome data limitations.
-Population models are a case in point.
-
-Population models are important tools for predicting the impacts of environmental change on species persistence and abundance.
+The increasing statistical sophistication of population models [@Rees2009] makes them promising tools for predicting the impacts of environmental change on species persistence and abundance.
 But reconciling the scales at which population models are parameterized and the scales at which environmental changes play out remains a challenge [@Clark2010; @Freckleton2011; @Queenborough2011; @Clark2012].
 The problem is that most population models are built using data from a single study site because collecting those data, which involves tracking the fates of individuals plants, is so difficult.
-The resulting models cannot be applied to the landscape and regional scales relevant to decision-makers without information about how the fitted parameters respond to spatial variation in biotic and abiotic drivers [@Sæther2007].
+The resulting models cannot be applied to the landscape and regional scales relevant to decision-making without information about how the fitted parameters respond to spatial variation in biotic and abiotic drivers [@Sæther2007].
 The limited spatial extent of individual-level demographic datasets constrains our ability to use population models to address applied questions about the consequences of climate change.
@@
 SDMs typically rely on easy-to-collect presence/absence data [but see @Clark2014 for new methods] and remotely-sensed environmental covariates that allow researchers to model large spatial extents [e.g., @Maiorano2013].
-Thus, it is relatively straightforward to parameterize and project SDMs over landscapes and regions, the scales at which many land-use decisions are made.
+Thus, it is relatively straightforward to parameterize and project SDMs over landscapes and regions.
 However, the limitations of SDMs are well known [@Pearson2003; @Elith2009; @Araujo2012].
@@
 Aggregate measures of population status, rather than individual performance, offer an intriguing alternative for modeling populations [@Clark2004; @Freckleton2011].
-Population-level data cannot provide inference about demographic mechanisms, but it might be sufficient for modeling future population states, especially since it is feasible to collect across broad spatial extents [e.g., @Queenborough2011].
+Population-level data cannot provide inference about demographic mechanisms, but might be sufficient for modeling future population states, especially since such data are feasible to collect across broad spatial extents [e.g., @Queenborough2011].
 The choice between individual and population-level data involves a difficult trade-off: while individual-level data leads to more mechanistic models, population-level data leads to models that can be applied over greater spatial and temporal extents.
@@
 They used a "density-structured" model where the state variable is a discrete density state rather than a continuous density measure.
-Such models do not require individual level demographic data and can adequately describe population dynamics.
-
-Building on @Taylor2004, @Freckleton2011 showed that density-structured models compare well to continuous models in theory, and @Queenborough2011 demonstrated the application of such methods in a study on arable weeds.
-In particular, @Queenborough2011 provide empirical evidence that density-structured models are capable of reproducing population dynamics, even if some precision is lost when compared to fully continuous models.
-The study by @Queenborough2011 included data from 500 fields (4 hectares each) in 49 farms, all collected by two people in 6 weeks.
-This is far more data from a far greater spatial extent than would be possible if measuring individual plant demography (in a world of limited time and money).
+Such models do not require individual-level demographic data and can adequately describe population dynamics.
+Building on @Taylor2004, @Freckleton2011 showed that density-structured models compare well to continuous models in theory, and @Queenborough2011 provide empirical evidence that density-structured models are capable of reproducing population dynamics at landscape spatial scales, even if some precision is lost when compared to fully continuous models.
 The appeal of density-structured approaches is clear.
@@
 Ignoring this fact amounts to an "ecological fallacy", where inference about the individual relies on statistical inference on the group [@PIANTADOSI1988].
-Population growth (or decline) is the outcome of demographic processes like survival, growth, and recruitment that occur at the level of individual plants.
+Population growth (or decline) is the outcome of demographic processes such as survival, growth, and recruitment that occur at the level of individual plants.
 Climate can affect each demographic process in unique, potentially opossing, ways [@Dalgleish2011].
 These unique climate responses may be difficult to resolve in statistical models based on population-level data where demographic processes are not identifiable.
-If important climate drivers are missed because they are averaged over in population-level data, then population models built with such data will make uninformative or unreliable forecasts when confronted with novel climate covariates.
-
-Here, we ask whether statistical and population models based on aggregated, population-level data can predict climate change impacts as well as models based on individual-level data.
+If important climate effects are missed because of the aggregation inherent in in population-level data, then population models built with such data will make uninformative or unreliable forecasts.
+
+Here, we compare the forecasting skill of statistical and population models based on aggregated, population-level data with models based on individual-level data.
 We used a unique demographic dataset that tracks the fates of individual plants from four species over 14 years to build two kinds of single-species population models, traditional models using individual growth, survival, and recruitment data and alternative models based on basal cover.
 In both models, interannual variation is explained, in part, by climate covariates.
+We first quantify forecasting skill using cross-validation.
 We then performed simulations to quantify the sensitivities of species' cover to small perturbations in average precipitation and temperature.
-Individual level models were clearly better, but, unfortunately, still too uncertain to inform management decisions.
+Based on the cross-validation results, predictions of individual level models were clearly better, but, unfortunately, still too uncertain to inform management decisions.
 
@@
 
-In this paper, we model populations based on two levels of data: individual and quadrat (Fig.
+We modeled each grass population based on two levels of data: individual and quadrat (Fig.
 2).
 The individual data is the "raw" data.
-For the quadrat level we data we simply sum individual basal cover for each quadrat by species.
-This is equivalent to a near-perfect census of quadrat percent cover because previous analysis shows that measurement error at the individual level is small [@Chu2014].
-Based on these two datasets we can compare population models built using individual level data and aggregated quadrat level data.
+For the quadrat-level we data we simply sum individual basal cover for each quadrat by species.
+This is equivalent to a near-perfect census of quadrat percent cover because previous analysis shows that measurement error at the individual-level is small [@Chu2014].
+Based on these two datasets we can compare population models built using individual-level data and aggregated, quadrat-level data.
 
@@
 At both levels of inference (individual and quadrat), the building blocks of our population models are vital rate regressions.
-For individual-level data, we fit models for survival, growth, and recruitment for each species.
+For individual-level data, we fit regressions for survival, growth, and recruitment for each species.
 At the quadrat-level, we fit a single regression model for population growth.
@@
 For the likelihood models, $\textbf{y}^X$ is always the relevant vector of observations for vital rate *X* ($X = S, G, R, or P$ for survival, growth, recruitment, or population growth).
-For example, $\textbf{y}^S$ is a vector of 0s and 1s indicating whether a genet survives from *t* to *t+1*, or not.
+For example, $\textbf{y}^S$ is a vector of 0's and 1's indicating whether a genet survives from *t* to *t+1*, or not.
 
@@
 Our data allows us to track new recruits, but we cannot assign a specific parent to new genets.
-So, for recruitment, we work at the quadrat level and model the number of new individuals of species $j$ in quadrat $q$ recruiting at time $t+1$ as a function of quadrat "effective cover" ($A'$) in the previous year ($t$).
+THerefore, we model recruitment at the quadrat level: the number of new individuals of species $j$ in quadrat $q$ recruiting at time $t+1$ as a function of quadrat "effective cover" ($A'$) in the previous year ($t$).
 Effective cover is a mixture of observed cover ($A$) in the focal quadrat ($q$) and the mean cover across the entire group ($\bar{A}$) of $Q$ quadrats in which $q$ is located:
@@
 #### Population model at the quadrat level
-The statistical approach used to model vital rates using aggregated data depends on the type of data collected.
-In our case, and as is often the case with census data, we have percent cover data (which can easily be transformed to proportion data).
+The statistical approach used to model aggregated data depends on the type of data collected.
+We have percent cover data, which can easily be transformed to proportion data.
 We first considered fitting three vital rate models analagous to those we fit at the individual level: one for probability of extirpation within a quadrat (analagous to survival), one for cover change within a quadrat (analagous to growth), and one for probability of colonization within a quadrat (analagous to recruitment).
@@
 So we constrained our statistical modeling of vital rates at the population level to change in percent cover within quadrats.
-For the remaining discussion of statistical modeling we refer to proportion data, which is simply percent data divided by 100.
+For the remaining discussion of statistical modeling, we refer to proportion data, which is simply percent cover divided by 100.
 
@@
 We therefore chose a more constrained modeling approach based on a truncated log-normal likelihood.
-The model for quadrat cover change ($G$) from time $t$ to $t+1$ is
+The model for quadrat cover change from time $t$ to $t+1$ is
 
@@
 Within a species, we fit yearly size effects and yearly intercepts hierarchically where year-specific coefficients were drawn from global distributions representing the mean size effect and intercept.
-We used uninformative priors (Appendix A).
+Quadrat random effects were also fit hierarchically, with quadrat offsets being drawn from distributions with mean zero and a shared variance term (independent Gaussian priors, Appendix A).
+Climate effects were not modeled hierarchically, and each was given a diffuse prior distribution.
+We used standard diffuse priors for all unknown parameters (Appendix A).
 
@@
 We obtained posterior distributions for all model parameters from three parallel MCMC chains run for 1,000 iterations after discarding an initial 1,000 iterations.
-We recognize such short MCMC chains may surprise those more familiar with other MCMC samplers (i.e.
+Such short MCMC chains may surprise readers more familiar with other MCMC samplers (i.e.
 JAGS or WinBUGS), but the Stan sampler is exceptionally efficient, which reduces the number of iterations needed to achieve convergence.
@@
 With the posterior distribution of the vital rate statistical models in hand, it is straightforward to simulate the population models.
-We used an Integral Projection Model (IPM) to model populations based on individual level data [@Ellner2006] and a quadrat-based version of an individually-based model (Quadrat-Based Model, QBM) to model populations based on quadrat level data.
+We used an Integral Projection Model (IPM) to model populations based on individual-level data [@Ellner2006] and a quadrat-based version of an individually-based model (Quadrat-Based Model, QBM) to model populations based on quadrat-level data.
 We describe each in turn.
@@
 #### Integral projection model
-We use an environmentally stochastic IPM [@Rees2009] that includes the climate covariates from the vital rate statistical models.
+We use a stochastic IPM [@Rees2009] that includes the climate covariates from the vital rate statistical models.
 In all simulations we ignore the random year effects so that interannual variation is driven solely by climate.
@@
 \begin{equation}
-n(v_{j},t+1) = \int_{L_{j}}^{U_{j}} k_{j}(v_{j},u_{j},\bar{\bold{w_{j}}}(u_{j}))n(u_{j},t)
+n(v_{j},t+1) = \int_{L_{j}}^{U_{j}} k_{j}(v_{j},u_{j},\bar{w_{j}}(u_{j}))n(u_{j},t)
 \end{equation}
 
-where $k_{j}(v_{j},u_{j},\bar{\bold{w_{j}}})$ is the population kernel that describes all possible transitions from size $u$ to $v$ and $\bar{\bold{w_{j}}}$ is a vector of estimates of average crowding experienced from all other species by a genet of size $u_j$ and species $j$.
+where $k_{j}(v_{j},u_{j},\bar{w_{j}})$ is the population kernel that describes all possible transitions from size $u$ to $v$ and $\bar{w_{j}}$ is a scalar representing the average intraspecific crowding experienced by a genet of size $u_j$ and species $j$.
 The integral is evaluated over all possible sizes between predefined lower (_L_) and upper (_U_) size limits that extend beyond the range of observed genet sizes.
@@
 Since the IPM is spatially-implicit, we cannot calculate neighborhood crowding for specific genets ($w_{ij}$).
-Instead, we use an approximation ($\bar{\bold{w_{j}}}$) that captures the essential features of neighborhood interactions [@Adler2010].
-We also use a 'no-overlap' rule for conspecific genets to approximate the overdispersion of large genets in space [@Adler2010; @Adler2012; @Chu2015].
+Instead, we use an approximation ($\bar{w_{j}}$) that captures the essential features of neighborhood interactions [@Adler2010].
+This approximation relies on a 'no-overlap' rule for conspecific genets to approximate the overdispersion of large genets in space [@Adler2010].
 
@@
 \begin{equation}
-k_{j}(v_{j},u_{j},\bar{\bold{w_{j}}}) = S_j(u_j, \bar{\bold{w_{j}}}(u_{j}))G_j(v_{j},u_{j},\bar{\bold{w_{j}}}(u_{j})) + R_j(v_{j},u_{j},\bar{\bold{w_{j}}}),
+k_{j}(v_{j},u_{j},\bar{w_{j}}) = S_j(u_j, \bar{w_{j}}(u_{j}))G_j(v_{j},u_{j},\bar{w_{j}}(u_{j})) + R_j(v_{j},u_{j},\bar{w_{j}}),
 \end{equation}
@@
 Our stastical model for recruitment (_R_, described above) returns the number of new recruit produced per quadrat.
-Following previous work [@Adler2012; @Chu2015], we assume that fecundity increases linearly with size ($R_j(v_{j},u_{j},\bar{\bold{w_{j}}}) = e^{u_j}R_j(v_{j},\bar{\bold{w_{j}}})$) to incorporate the recruitment function in the spatially-implicit IPM.
+Following previous work [@Adler2012; @Chu2015], we assume that fecundity increases linearly with size ($R_j(v_{j},u_{j},\bar{w_{j}}) = e^{u_j}R_j(v_{j},\bar{w_{j}})$) to incorporate the recruitment function in the spatially-implicit IPM.
 
@@
 Then, we drew the full parameter set (climate effects and density-dependence fixed effects) from a randomly selected MCMC iteration.
-Using this approach, rather than simply using coefficient point estimates, ensures that relatively unimportant climate covariates (those that broadly overlap 0) have little effect on the simulation results.
+Using this approach, rather than simply using coefficient point estimates, captures the effect of parameter uncertainty.
+Relatively unimportant climate covariates (those that broadly overlap 0) will have little effect on the mean of the simulation results, but can contribute to their variation.
 Since our focus was on the contribution of climate covariates to population states, we set the random year effects and the random group effects to zero.
@@
 #### Quad-based model
-Our quad-based model (QBM) perfectly mirrors its statistical description (Eqs.
+To simulate our quad-based model (QBM), we simply iterate the quadrat-level statistical model (Eqs.
 9-10).
@@
 Within each observation year, several quadrats were sampled.
-So we made predictions for each observed quadrat in the focal year, initializing each simulation with cover in the quadrat the previous year.
-Since we were making quadrat-specific predictions, we incorporated the group effect on the intercept for both models.
+We made predictions for each observed quadrat in the focal year, initializing each simulation with cover in the quadrat the previous year.
+Since we were making quadrat-specific predictions, we incorporated the group random effect on the intercept for both models.
 We repeated this procedure for all 13 observation years, making 100 one-step-ahead forecasts for each quadrat-year combination with parameter uncertainty included via random draw from the MCMC chain as described above.
@@
 
-This model validation allowed us to compare accuracy and precision of the two modeling approaches (IPM versus QBM).
+This cross-validation procedure allowed us to compare accuracy and precision of the two modeling approaches (IPM versus QBM).
 We first calculated the median predicted cover across the 100 simulations for each quadrat-year and then calculated the absolute error as the absolute value of the difference between the observed cover for a given quadrat-year and the median prediction.
@@
 
-Our expectation is that the IPM will produce more accurate and precise forecasts.
-This could be due to (1) the smaller  sample size of the quadrat level data sets compared to the individual level data sets, leading to larger parameter uncertainty for the QBM, or (2) the QBM climate effects being weakly associated with one or more vital rate climate effects at the individual level.
+We expected the IPM to produce more accurate and precise forecasts due to either (1) the smaller sample size of the quadrat level data sets compared to the individual level data sets, leading to larger parameter uncertainty for the QBM, or (2) the QBM climate effects being weakly associated with one or more vital rate climate effects at the individual level.
 To assess the impact of sample size on QBM parameter uncertainty we refit the QBM statistical model (Eqs.
@@
 We calculated the standard deviation of climate main effects (pptLag, ppt1, ppt2, TmeanSpr1, and TmeanSpr2) for each model and averaged those over replicates within each set of quadrat removals.
-This allowed us to regress sample size against parameter uncertainty.
-
-To see if the QBM climate effects are correlated, or not, with climate effects for each vital rate model for the IPM, we simply regressed the QBM climate coefficients against each vital rate model's climate coefficients and calculated Pearson's $\rho$.
+This allowed us to regress parameter uncertainty against sample size.
+
+To deterime if the QBM climate effects are correlated with climate effects for each vital rate model in the IPM, we simply regressed the QBM climate coefficients against each vital rate model's climate coefficients and calculated Pearson's $\rho$.
 Strong correlations indicate the QBM is capable of detecting climate effects associated with individual vital rates.
@@
 ### Comparison of forecast models
-The IPM had significantly lower overall error (MAE, mean absolute error) for three species (*B.
-gracilis*, *H.
-comata*, *P.
-smithii*; Table 1).
+The IPM had lower overall error (MAE, mean absolute error) compared to the QBM for three species (Table 1).
+The IPM MAE is significantly lower at $\alpha=0.05$ for *B.
+gracilis* (*P* = `r mae_ps[1,"value"]`), *H.
+comata* (*P* = `r mae_ps[2,"value"]`), and *P.
+smithii* (*P* = `r mae_ps[3,"value"]`).
+MAEs are statisticially similar between models for *P.
+secunda* (*P* = `r mae_ps[4,"value"]`).
+*P* values are highly sensitive to sample size, so not entirely appropriate in simulation exercises where we control the samples size.
+But, for our purposes they serve as relatively unbiased comparison metrics.
 In no case did the QBM significantly outperform the IPM (Table 1).
@@
 ```{r get_results, echo=FALSE, eval=FALSE}
-setwd("~/Repos/MicroMesoForecast/analysis/")
-
-####
-##  Mean parameter runs
-####
-meanparam_ipm <- data.frame(cover=NA, species=NA, 
-                             climsim=NA, model=NA)
-##  IPM 
-path <- "./ipm/simulations/results/climchange_meanparams/"
-ipmfiles_mean <- list.files(path)
-ipmfiles_mean <- ipmfiles_mean[grep("cover", ipmfiles_mean)]
-for(file in ipmfiles_mean){
-  tmp <- read.csv(paste(path,file,sep=""))
-  tmp$model <- "IPM"
-  meanparam_ipm <- rbind(meanparam_ipm, tmp[,c("cover", "species", "climsim", "model")])
-}
-meanparam_ipm <- meanparam_ipm[2:nrow(meanparam_ipm),]
-
-newd <- data.frame(cover=NA, species=NA, 
-                   climsim=NA, model=NA, obscover=NA)
-for(dospp in unique(meanparam_ipm$species)){
-  tmp <- subset(meanparam_ipm, species==dospp)
-  obscover <- subset(tmp, climsim=="obs")["cover"]
-  tmp2 <- subset(tmp, climsim!="obs")
-  tmp2$obscover <- rep(obscover$cover, times = length(unique(tmp2$climsim)))
-  newd <- rbind(newd, tmp2)
-}
-ipm_meansims <- newd[2:nrow(newd),]
-ipm_meansims$propdiff <- with(ipm_meansims, log(cover*100)-log(obscover*100))
-
-##  QBM
-meanparam_qbm <- data.frame(cover=NA, species=NA, 
-                             climsim=NA, model=NA)
-path <- "./quadBM/simulations/results/climatechange_meanparams/"
-qbmfiles_mean <- list.files(path)
-for(file in qbmfiles_mean){
-  tmp <- readRDS(paste(path,file,sep=""))
-  tmp$model <- "QBM"
-  meanparam_qbm <- rbind(meanparam_qbm, tmp[,c("cover", "species", "climsim", "model")])
-}
-meanparam_qbm <- meanparam_qbm[2:nrow(meanparam_qbm),]
-
-newd <- data.frame(cover=NA, species=NA, 
-                   climsim=NA, model=NA, obscover=NA)
-for(dospp in unique(meanparam_qbm$species)){
-  tmp <- subset(meanparam_qbm, species==dospp)
-  obscover <- subset(tmp, climsim=="obs")["cover"]
-  tmp2 <- subset(tmp, climsim!="obs")
-  tmp2$obscover <- rep(obscover$cover, times = length(unique(tmp2$climsim)))
-  newd <- rbind(newd, tmp2)
-}
-qbm_meansims <- newd[2:nrow(newd),]
-qbm_meansims$propdiff <- with(qbm_meansims, log(cover*100)-log(obscover*100))
-
-##  Merge IPM and QBM mean parameter results
-bothmodel_outs <- rbind(ipm_meansims, qbm_meansims)
-meanparam_agg <- ddply(bothmodel_outs, .(species, climsim, model), summarise,
-                       meandiff = mean(propdiff),
-                       hidiff = quantile(propdiff, probs = 0.95),
-                       lodiff = quantile(propdiff, probs = 0.05))
-setwd("~/Repos/MicroMesoForecast/manuscript")
+# setwd("~/Repos/MicroMesoForecast/analysis/")
+# 
+# ####
+# ##  Mean parameter runs
+# ####
+# meanparam_ipm <- data.frame(cover=NA, species=NA, 
+#                              climsim=NA, model=NA)
+# ##  IPM 
+# path <- "./ipm/simulations/results/climchange_meanparams/"
+# ipmfiles_mean <- list.files(path)
+# ipmfiles_mean <- ipmfiles_mean[grep("cover", ipmfiles_mean)]
+# for(file in ipmfiles_mean){
+#   tmp <- read.csv(paste(path,file,sep=""))
+#   tmp$model <- "IPM"
+#   meanparam_ipm <- rbind(meanparam_ipm, tmp[,c("cover", "species", "climsim", "model")])
+# }
+# meanparam_ipm <- meanparam_ipm[2:nrow(meanparam_ipm),]
+# 
+# newd <- data.frame(cover=NA, species=NA, 
+#                    climsim=NA, model=NA, obscover=NA)
+# for(dospp in unique(meanparam_ipm$species)){
+#   tmp <- subset(meanparam_ipm, species==dospp)
+#   obscover <- subset(tmp, climsim=="obs")["cover"]
+#   tmp2 <- subset(tmp, climsim!="obs")
+#   tmp2$obscover <- rep(obscover$cover, times = length(unique(tmp2$climsim)))
+#   newd <- rbind(newd, tmp2)
+# }
+# ipm_meansims <- newd[2:nrow(newd),]
+# ipm_meansims$propdiff <- with(ipm_meansims, log(cover*100)-log(obscover*100))
+# 
+# ##  QBM
+# meanparam_qbm <- data.frame(cover=NA, species=NA, 
+#                              climsim=NA, model=NA)
+# path <- "./quadBM/simulations/results/climatechange_meanparams/"
+# qbmfiles_mean <- list.files(path)
+# for(file in qbmfiles_mean){
+#   tmp <- readRDS(paste(path,file,sep=""))
+#   tmp$model <- "QBM"
+#   meanparam_qbm <- rbind(meanparam_qbm, tmp[,c("cover", "species", "climsim", "model")])
+# }
+# meanparam_qbm <- meanparam_qbm[2:nrow(meanparam_qbm),]
+# 
+# newd <- data.frame(cover=NA, species=NA, 
+#                    climsim=NA, model=NA, obscover=NA)
+# for(dospp in unique(meanparam_qbm$species)){
+#   tmp <- subset(meanparam_qbm, species==dospp)
+#   obscover <- subset(tmp, climsim=="obs")["cover"]
+#   tmp2 <- subset(tmp, climsim!="obs")
+#   tmp2$obscover <- rep(obscover$cover, times = length(unique(tmp2$climsim)))
+#   newd <- rbind(newd, tmp2)
+# }
+# qbm_meansims <- newd[2:nrow(newd),]
+# qbm_meansims$propdiff <- with(qbm_meansims, log(cover*100)-log(obscover*100))
+# 
+# ##  Merge IPM and QBM mean parameter results
+# bothmodel_outs <- rbind(ipm_meansims, qbm_meansims)
+# meanparam_agg <- ddply(bothmodel_outs, .(species, climsim, model), summarise,
+#                        meandiff = mean(propdiff),
+#                        hidiff = quantile(propdiff, probs = 0.95),
+#                        lodiff = quantile(propdiff, probs = 0.05))
+# setwd("~/Repos/MicroMesoForecast/manuscript")
 ```
@@
 Population-level data are much easier to collect across broad spatial extents, so models built using such data offer an appealing alternative to traditional population models [@Queenborough2011].
-However, density-structured models rely on the aggregation of individual level data.
+However, density-structured models rely on the aggregation of individual-level data.
 This creates a potential problem if such models are to be used in a climate change context because it is individuals, not populations, which respond to climate [@Clark2012].
-Are models based on population level metrics as sensitive to climate as models based on individual level metrics?
+Are models based on population-level metrics as sensitive to climate as models based on individual-level metrics?
 Do these two types of models produce consistent forecasts?
@@
 4).
+
 We did, however, find evidence that the QBM statistical model failed to identify climate dependence for some vital rates (Fig.
@@
 This is analagous to our situation, but instead of averaging over complex life histories, we are averaging over complex climate dependence.
+Though our work here focused on plant species, this finding is applicable to any species with vital rates that respond uniquely to weather/climate.
 
@@
 When this is the case, as it is for all our species to varying degrees (Fig.
-3), statistical models of population-level data will fail.
+3), forecasts from models based on population-level data will fail.
 Our result is consistent with related work on the importance of individual-level data to forecast population responses to exogenous drivers [@Clark2011; @Clark2011b; @Clark2012; @Galvan2014].
@@
 Our results highlight the state of affairs in ecology when it comes to forecasting the impacts of climate change.
-The analysis we conducted here could be considered at the forefront of ecological forecasting with repsect to the statistical approach employed (hierarchical Bayesian), the type of population model we used (density-dependent, stochastic IPM with parameter uncertainty), and the amount of high quality data we had at our disposal (14 years of individual level data).
-Yet, model predictions proved so uncertain that any forecast, when bounded with model and parameter uncertainty, would be at best not useful and at worst meaningless.
+The analysis we conducted here could be considered at the forefront of ecological forecasting with respect to the statistical approach employed (hierarchical Bayesian), the type of population model we used (density-dependent, stochastic IPM with parameter uncertainty), and the amount of high quality data we had at our disposal (14 years of individual-level data).
+Yet, model predictions proved so uncertain that any forecast, when bounded with model and parameter uncertainty, would be uninformative.
 
@@
 First, forecasts could be improved by matching the spatial scale of predictor variables with the spatial scale of observations.
-One of the major limitations of the models we fit here is that the climate data are at a much larger scale than the individual level observations of plant size.
+One of the major limitations of the models we fit here is that the climate data are collected at a larger scale than the individual-level observations of plant size.
 Climate covariates only vary by year, with no spatial variability within years.
@@
 Only by being honest about our forecasts can we begin to produce better ones, and forecasts reported without parameter error are disingenuous.
-In many cases, ignoring parameter error is justifiable when investigating basic demographic mechanisms, but it is not justifiable when forecasting is the goal.
+Ignoring parameter error may be justifiable when the goal is investigating basic processes, but it is indefensible when forecasting is the goal.
 
@@
 Our results confirm theoretical arguments [@Clark2011b] and empirical evidence [@Clark2011; @Clark2012] that individual responses are critical for predicting species' responses to climate change.
-It seems there is no short cut to producing accurate and precise population forecasts.
-Do we need detailed demographic data to forecast the impacts of climate change on populations?
-Probably.
+It seems there is no short cut to producing accurate and precise population forecasts: we need detailed demographic data to forecast the impacts of climate change on populations.
 Given the importance of demographic data and its current collection cost, we need modern methods to collect demographic data more efficiently across environmental gradients in space and time.
@@
 Our results also offer a cautionary tale because forecast uncertainty was large for both model types.
-Even with 14 years of detailed demographic data and sophisticated modeling techniques, we failed to produce forecasts with any level of acceptable uncertainty.
-Uncertainty of climate change related forecasts can be reduced by (1) longer time series and (2) climate covariates that match the scale of inference (e.g., plot rather than landscape level climate/weather metrics).
+Even with 14 years of detailed demographic data and sophisticated modeling techniques, our projections contained too much uncertainty to be informative.
+Uncertainty in demographic responses to climate can be reduced by collecting (1) longer time series and (2) climate covariates that match the scale of inference (e.g., plot rather than landscape level climate/weather metrics).
 
@@
 Informal conversations with Stephen Ellner, Giles Hooker, Robin Snyder, and a series of meetings between the Adler and Weecology labs at USU sharpened our thinking.
-Brittany Teller and two anonymous reviewers provided comments that improved our manuscript.
+Brittany Teller provided comments that improved our manuscript.
 Compute, storage and other resources from the Division of Research Computing in the Office of Research and Graduate Studies at Utah State University are gratefully acknowledged.
@@
 ```
-*NOTES*: The IPM MAE is significantly lower at $\alpha=0.05$ for *B.
-gracilis* (*P* = `r mae_ps[1,"value"]`), *H.
-comata* (*P* = `r mae_ps[2,"value"]`), and *P.
-smithii* (*P* = `r mae_ps[3,"value"]`).
-MAEs are statisticially similar between models for *P.
-secunda* (*P* = `r mae_ps[4,"value"]`).
-*P* values are highly sensitive to sample size, so not entirely appropriate in simulation exercises where we control the samples size.
-But, for our purposes they serve as relatively unbiased comparison metrics.
-
-\pagebreak{}
-```{r table_2,results='asis',message=FALSE, comment=NA, echo=FALSE}
-# qbm_diffs <- readRDS("../montana/quadBM/simulations/results/qbm_climatesims_logdiffs.RDS")
-# qbm_diffs$type <- rep("QBM", nrow(qbm_diffs))
-# ipm_diffs <- readRDS("../montana/ipm/simulations/results/ipm_climatesims_logdiffs.RDS")
-# ipm_diffs$type <- rep("IPM", nrow(ipm_diffs))
-# 
-# diff_df <- rbind(qbm_diffs, ipm_diffs)
-# med_cover_perc <- as.data.frame(diff_df$med_cover*100)
-# med_cover_perc$model <- diff_df$type 
-# med_cover_perc$sim <- diff_df$variable
-# med_cover_perc$species <- diff_df$species
-# colnames(med_cover_perc)[1] <- "cover"
-# medcast <- dcast(med_cover_perc, species+sim~model, value.var = "cover")
-# colnames(medcast) <- c("Species", "Simulation", "IPM Change (%)", "QBM Change (%)")
-# simnames <- c("ppt", "temp", "ppt+temp")
-# medcast$Simulation <- rep(simnames, 4)
-# print.xtable(xtable(medcast, caption = "Percent change from equilibrium (temporal median of simulated plant cover) at observed climate relative to a 1\\% climate change as forecast by each model type."),type="latex", comment=FALSE,
-#              include.rownames=FALSE, caption.placement="top")
-```
 
@@
 grid.raster(img)
-```
-
-```{r ipm_climeffs, cache=FALSE, results='hide', include=FALSE}
-# source("../montana/ipm/vitalRateRegs/climate_effects_plots.R")
-# ![Standardized regression coefficients for climate effects in the vital rate models for the IPM.
-Shown are the median (circles), 75% confidence intervals (thick lines), and 95% confidence intervals (thin lines).
-Left panels (black) are for the growth regressions; middle panels (blue) are for the survival regressions; right panels (orange) are for the recruitment regressions.](components/figure/ipm_climeffs.pdf)
-```
-
-```{r quad_climeffs, cache=FALSE, results='hide', include=FALSE}
-# source("../montana/quadBM/vitalRateRegressions/truncNormModel/climate_effects_plot.R")
-# ![Standardized regression coefficients for climate effects in the vital rate models for the QBM.
-Shown are the median (circles), 75% confidence intervals (thick lines), and 95% confidence intervals (thin lines).](components/figure/qbm_climeffs.pdf)
 ```
@@
 
-```{r figure_6, dependson="plot-options", fig.cap="Equilibrium cover and 90% quantiles around the mean prediction when model error and parameter uncertainty are propogated through the simulation phase.
+```{r figure_6, dependson="plot-options", fig.cap="Mean (points) and 90% quantiles (errorbars) for the proportional difference between baseline simulations (using observed climate) and the climate pertubation simulation on the x-axis.
+We calculated proportional difference as log(perturbed climate cover) - log(observed climate cover), where "perturbed" and "observed" refer to the climate time series used to drive interannual variation in the simulations.
+Model error and parameter uncertainty were propagated through the simulation phase.
 Climate simulations are as in Figure 3.", fig.width=4.5, fig.height=7, cache=FALSE, results='hide'}

### 2015-08-28 manuscript/appB.Rmd
small edits from PBA on appendices

@@
 -------------------------------
-*Tables begin on next page...*
+In all tables, "mean_value" is the mean parameter estimate, "sd_value" is the standard deviation of the estimate from the MCMC, "lo_BCI" is the lower limit of the 95% Bayesian Credible Interval, and "up_BCI" is the upper limit of the 95% Bayesian Credible Interval
 

### 2015-08-28 manuscript/manuscript.Rmd
small edits from PBA on appendices

@@
 -------
+```{r pvalues, echo=FALSE, eval=TRUE}
+##  IPM results
+# Read in data
+setwd("/Users/atredenn/Repos/MicroMesoForecast/manuscript/")
+ipm_onestep <- readRDS("../analysis/ipm/simulations/results/one_step_validation/ipm_loyo_forecasts_combined.rds")
+# Remove predictions >1 for fair comparison to [0,1] truncated quad model
+resets <- which(ipm_onestep[,"cover.t1"]>1)
+ipm_onestep[resets, "cover.t1"] <- 1
+# Average predictions over quad-year reps
+avg_ipm <- ddply(ipm_onestep, .(species, quad, t1), summarise,
+                 avg_prediction = median(cover.t1),
+                 observation = mean(cover.t0),
+                 quant_dist = quantile(cover.t1*100, probs = 0.95) - quantile(cover.t1, probs = 0.05))
+# Calculate error
+avg_ipm$error <- avg_ipm$observation*100 - avg_ipm$avg_prediction*100
+# Aggregate over species
+ipmstats <- ddply(avg_ipm, .(species), summarise,
+                  mae = mean(abs(error)),
+                  mean_cover = mean(observation*100),
+                  quant_dist = mean(quant_dist))
+ipmstats$model <- "IPM"
+
+##  QBM results
+# Read in data
+qbm_onestep <- readRDS("../analysis/quadBM/simulations/results/one_step_validation/qbm_one-step_forecasts_combined.rds")
+# Average predictions over quad-year reps
+avg_qbm <- ddply(qbm_onestep, .(species, quad, year), summarise,
+                 avg_prediction = median(cover.t1),
+                 observation = mean(cover.t0),
+                 quant_dist = quantile(cover.t1*100, probs = 0.95) - quantile(cover.t1, probs = 0.05))
+# Calculate error
+avg_qbm$error <- avg_qbm$observation*100 - avg_qbm$avg_prediction*100
+# Aggregate over species
+qbmstats <- ddply(avg_qbm, .(species), summarise,
+                  mae = mean(abs(error)),
+                  mean_cover = mean(observation*100),
+                  quant_dist = mean(quant_dist))
+qbmstats$model <- "QBM"
+
+mae_list <- list()
+mae_pvals <- list()
+species <- unique(qbmstats$species)
+for(spp in species){
+  tmp1 <- subset(avg_ipm, species==spp)
+  tmp2 <- subset(avg_qbm, species==spp)
+  err1 <- as.data.frame(abs(tmp1$observation - tmp1$avg_prediction))
+  err2 <- as.data.frame(abs(tmp2$observation - tmp2$avg_prediction))
+  colnames(err1) <- "resid"
+  err1$model <- "ipm"
+  colnames(err2) <- "resid"
+  err2$model <- "qbm"
+  errd <-  rbind(err1, err2)
+  mae_ttest <- with(errd, t.test(resid~model, alternative = "less"))
+  mae_list[[spp]] <- mae_ttest
+  mae_pvals[[spp]] <- mae_ttest[["p.value"]]
+}
+
+mae_ps <- melt(mae_pvals)
+```
+
 ### Comparison of forecast models
@@
 ```{r get_results, echo=FALSE, eval=FALSE}
-# setwd("~/Repos/MicroMesoForecast/analysis/")
-# 
-# ####
-# ##  Mean parameter runs
-# ####
-# meanparam_ipm <- data.frame(cover=NA, species=NA, 
-#                              climsim=NA, model=NA)
-# ##  IPM 
-# path <- "./ipm/simulations/results/climchange_meanparams/"
-# ipmfiles_mean <- list.files(path)
-# ipmfiles_mean <- ipmfiles_mean[grep("cover", ipmfiles_mean)]
-# for(file in ipmfiles_mean){
-#   tmp <- read.csv(paste(path,file,sep=""))
-#   tmp$model <- "IPM"
-#   meanparam_ipm <- rbind(meanparam_ipm, tmp[,c("cover", "species", "climsim", "model")])
-# }
-# meanparam_ipm <- meanparam_ipm[2:nrow(meanparam_ipm),]
-# 
-# newd <- data.frame(cover=NA, species=NA, 
-#                    climsim=NA, model=NA, obscover=NA)
-# for(dospp in unique(meanparam_ipm$species)){
-#   tmp <- subset(meanparam_ipm, species==dospp)
-#   obscover <- subset(tmp, climsim=="obs")["cover"]
-#   tmp2 <- subset(tmp, climsim!="obs")
-#   tmp2$obscover <- rep(obscover$cover, times = length(unique(tmp2$climsim)))
-#   newd <- rbind(newd, tmp2)
-# }
-# ipm_meansims <- newd[2:nrow(newd),]
-# ipm_meansims$propdiff <- with(ipm_meansims, log(cover*100)-log(obscover*100))
-# 
-# ##  QBM
-# meanparam_qbm <- data.frame(cover=NA, species=NA, 
-#                              climsim=NA, model=NA)
-# path <- "./quadBM/simulations/results/climatechange_meanparams/"
-# qbmfiles_mean <- list.files(path)
-# for(file in qbmfiles_mean){
-#   tmp <- readRDS(paste(path,file,sep=""))
-#   tmp$model <- "QBM"
-#   meanparam_qbm <- rbind(meanparam_qbm, tmp[,c("cover", "species", "climsim", "model")])
-# }
-# meanparam_qbm <- meanparam_qbm[2:nrow(meanparam_qbm),]
-# 
-# newd <- data.frame(cover=NA, species=NA, 
-#                    climsim=NA, model=NA, obscover=NA)
-# for(dospp in unique(meanparam_qbm$species)){
-#   tmp <- subset(meanparam_qbm, species==dospp)
-#   obscover <- subset(tmp, climsim=="obs")["cover"]
-#   tmp2 <- subset(tmp, climsim!="obs")
-#   tmp2$obscover <- rep(obscover$cover, times = length(unique(tmp2$climsim)))
-#   newd <- rbind(newd, tmp2)
-# }
-# qbm_meansims <- newd[2:nrow(newd),]
-# qbm_meansims$propdiff <- with(qbm_meansims, log(cover*100)-log(obscover*100))
-# 
-# ##  Merge IPM and QBM mean parameter results
-# bothmodel_outs <- rbind(ipm_meansims, qbm_meansims)
-# meanparam_agg <- ddply(bothmodel_outs, .(species, climsim, model), summarise,
-#                        meandiff = mean(propdiff),
-#                        hidiff = quantile(propdiff, probs = 0.95),
-#                        lodiff = quantile(propdiff, probs = 0.05))
-# setwd("~/Repos/MicroMesoForecast/manuscript")
+setwd("~/Repos/MicroMesoForecast/analysis/")
+
+####
+##  Mean parameter runs
+####
+meanparam_ipm <- data.frame(cover=NA, species=NA, 
+                             climsim=NA, model=NA)
+##  IPM 
+path <- "./ipm/simulations/results/climchange_meanparams/"
+ipmfiles_mean <- list.files(path)
+ipmfiles_mean <- ipmfiles_mean[grep("cover", ipmfiles_mean)]
+for(file in ipmfiles_mean){
+  tmp <- read.csv(paste(path,file,sep=""))
+  tmp$model <- "IPM"
+  meanparam_ipm <- rbind(meanparam_ipm, tmp[,c("cover", "species", "climsim", "model")])
+}
+meanparam_ipm <- meanparam_ipm[2:nrow(meanparam_ipm),]
+
+newd <- data.frame(cover=NA, species=NA, 
+                   climsim=NA, model=NA, obscover=NA)
+for(dospp in unique(meanparam_ipm$species)){
+  tmp <- subset(meanparam_ipm, species==dospp)
+  obscover <- subset(tmp, climsim=="obs")["cover"]
+  tmp2 <- subset(tmp, climsim!="obs")
+  tmp2$obscover <- rep(obscover$cover, times = length(unique(tmp2$climsim)))
+  newd <- rbind(newd, tmp2)
+}
+ipm_meansims <- newd[2:nrow(newd),]
+ipm_meansims$propdiff <- with(ipm_meansims, log(cover*100)-log(obscover*100))
+
+##  QBM
+meanparam_qbm <- data.frame(cover=NA, species=NA, 
+                             climsim=NA, model=NA)
+path <- "./quadBM/simulations/results/climatechange_meanparams/"
+qbmfiles_mean <- list.files(path)
+for(file in qbmfiles_mean){
+  tmp <- readRDS(paste(path,file,sep=""))
+  tmp$model <- "QBM"
+  meanparam_qbm <- rbind(meanparam_qbm, tmp[,c("cover", "species", "climsim", "model")])
+}
+meanparam_qbm <- meanparam_qbm[2:nrow(meanparam_qbm),]
+
+newd <- data.frame(cover=NA, species=NA, 
+                   climsim=NA, model=NA, obscover=NA)
+for(dospp in unique(meanparam_qbm$species)){
+  tmp <- subset(meanparam_qbm, species==dospp)
+  obscover <- subset(tmp, climsim=="obs")["cover"]
+  tmp2 <- subset(tmp, climsim!="obs")
+  tmp2$obscover <- rep(obscover$cover, times = length(unique(tmp2$climsim)))
+  newd <- rbind(newd, tmp2)
+}
+qbm_meansims <- newd[2:nrow(newd),]
+qbm_meansims$propdiff <- with(qbm_meansims, log(cover*100)-log(obscover*100))
+
+##  Merge IPM and QBM mean parameter results
+bothmodel_outs <- rbind(ipm_meansims, qbm_meansims)
+meanparam_agg <- ddply(bothmodel_outs, .(species, climsim, model), summarise,
+                       meandiff = mean(propdiff),
+                       hidiff = quantile(propdiff, probs = 0.95),
+                       lodiff = quantile(propdiff, probs = 0.05))
+setwd("~/Repos/MicroMesoForecast/manuscript")
 ```
@@
 # Read in data
+setwd("/Users/atredenn/Repos/MicroMesoForecast/manuscript/")
 ipm_onestep <- readRDS("../analysis/ipm/simulations/results/one_step_validation/ipm_loyo_forecasts_combined.rds")
@@
 For example, the points associated with G show the median cover from IPM simulations where a climate perturbation is applied only to the growth regression climate covariates.
-These simulations use mean parameter values for clarity.", fig.width=4.5, fig.height=8, cache=FALSE, results='hide'}
+These simulations use mean parameter values for clarity.", fig.width=8, fig.height=6, cache=FALSE, results='hide'}
 setwd("~/Repos/MicroMesoForecast/analysis/")
-#Proportional change in species' mean cover caused by a 1% increase in observed precipitation (+ppt), temperature (+temp), or both (+ppt&temp) as predicted by the individual-based IPM and the aggregate-based QBM using mean parameter values.
-Top panels show the mean predicted proportional change in cover.
-Lower panels show the
-####
-##  Mean parameter runs
-####
-# meanparam_ipm <- data.frame(cover=NA, species=NA, 
-#                              climsim=NA, model=NA)
-# ##  IPM 
-# path <- "./ipm/simulations/results/climchange_meanparams/"
-# ipmfiles_mean <- list.files(path)
-# ipmfiles_mean <- ipmfiles_mean[grep("cover", ipmfiles_mean)]
-# for(file in ipmfiles_mean){
-#   tmp <- read.csv(paste(path,file,sep=""))
-#   tmp$model <- "IPM"
-#   meanparam_ipm <- rbind(meanparam_ipm, tmp[,c("cover", "species", "climsim", "model")])
-# }
-# meanparam_ipm <- meanparam_ipm[2:nrow(meanparam_ipm),]
-# 
-# newd <- data.frame(cover=NA, species=NA, 
-#                    climsim=NA, model=NA, obscover=NA)
-# for(dospp in unique(meanparam_ipm$species)){
-#   tmp <- subset(meanparam_ipm, species==dospp)
-#   obscover <- subset(tmp, climsim=="obs")["cover"]
-#   tmp2 <- subset(tmp, climsim!="obs")
-#   tmp2$obscover <- rep(obscover$cover, times = length(unique(tmp2$climsim)))
-#   newd <- rbind(newd, tmp2)
-# }
-# ipm_meansims <- newd[2:nrow(newd),]
-# ipm_meansims$propdiff <- with(ipm_meansims, log(cover*100)-log(obscover*100))
-# 
-# ##  QBM
-# meanparam_qbm <- data.frame(cover=NA, species=NA, 
-#                              climsim=NA, model=NA)
-# path <- "./quadBM/simulations/results/climatechange_meanparams/"
-# qbmfiles_mean <- list.files(path)
-# for(file in qbmfiles_mean){
-#   tmp <- readRDS(paste(path,file,sep=""))
-#   tmp$model <- "QBM"
-#   meanparam_qbm <- rbind(meanparam_qbm, tmp[,c("cover", "species", "climsim", "model")])
-# }
-# meanparam_qbm <- meanparam_qbm[2:nrow(meanparam_qbm),]
-# 
-# newd <- data.frame(cover=NA, species=NA, 
-#                    climsim=NA, model=NA, obscover=NA)
-# for(dospp in unique(meanparam_qbm$species)){
-#   tmp <- subset(meanparam_qbm, species==dospp)
-#   obscover <- subset(tmp, climsim=="obs")["cover"]
-#   tmp2 <- subset(tmp, climsim!="obs")
-#   tmp2$obscover <- rep(obscover$cover, times = length(unique(tmp2$climsim)))
-#   newd <- rbind(newd, tmp2)
-# }
-# qbm_meansims <- newd[2:nrow(newd),]
-# qbm_meansims$propdiff <- with(qbm_meansims, log(cover*100)-log(obscover*100))
-# 
-# ##  Merge IPM and QBM mean parameter results
-# bothmodel_outs <- rbind(ipm_meansims, qbm_meansims)
-# meanparam_agg <- ddply(bothmodel_outs, .(species, climsim, model), summarise,
-#                        meandiff = mean(propdiff),
-#                        hidiff = quantile(propdiff, probs = 0.95),
-#                        lodiff = quantile(propdiff, probs = 0.05))
-# dgd = position_dodge(width = 0.9)
-# meanplot <- ggplot(meanparam_agg)+
-#   geom_bar(aes(x=climsim, y=meandiff, fill=model),
-#              position=dgd, stat="identity")+
-#   theme_bw()+
-#   scale_fill_manual(values=c("black", "grey45"), name="Model")+
-#   facet_wrap("species", scales="free", nrow=1)+
-#   scale_x_discrete(labels=c("+ppt", "+temp", "+ppt&temp"))+
-#   ylab("Mean proportional difference")+
-#   xlab("Climate simulation")+
-#   theme(axis.text.x = element_text(angle = 45, hjust = 1))
 
@@
   geom_point(size=4)+
-  facet_wrap("species", scales = "free", ncol=1)+
+  facet_wrap("species", scales = "free", ncol=2)+
   scale_linetype_discrete(name="Vital Rate")+
@@
 ```{r figure_6, dependson="plot-options", fig.cap="Mean (points) and 90% quantiles (errorbars) for the proportional difference between baseline simulations (using observed climate) and the climate pertubation simulation on the x-axis.
-We calculated proportional difference as log(perturbed climate cover) - log(observed climate cover), where "perturbed" and "observed" refer to the climate time series used to drive interannual variation in the simulations.
+We calculated proportional difference as log(perturbed climate cover) - log(observed climate cover), where 'perturbed' and 'observed' refer to the climate time series used to drive interannual variation in the simulations.
 Model error and parameter uncertainty were propagated through the simulation phase.

### 2015-08-28 manuscript/manuscript.tex
small appendix updates to match ms
small edits from PBA on appendices

@@
 
-Forecasting future states of populations has taken on new urgency as the
-rate of climate change increases. Traditional plant population models,
-typically built using detailed demographic observations from one study
-site, can address climate change impacts at one location, but are
-difficult to scale up to the landscape and regional scales relevant to
-management decisions. One proposal to overcome the limitations of
-traditional population models is to build models using population-level
-data that is much easier to collect over broad spatial scales than
+Rapid climate change has generated growing interest in forecasts of
+future population trajectories. Traditional population models, typically
+built using detailed demographic observations from one study site, can
+address climate change impacts at one location, but are difficult to
+scale up to the landscape and regional scales relevant to management
+decisions. An alternative is to build models using population-level data
+that is much easier to collect over broad spatial scales than
 individual-level data. However, such models ignore the fact that climate
@@
 responses when forecasting climate change impacts on four perennial
-grass species in a semi-arid Montana grassland. We parameterized two
-population models, one based on individual-level data (survival, growth
-and recruitment) and one on population-level data from the same dataset
-(percent cover of quadrats), and compared their accuracy, precision, and
-sensitivity to climate. The individual level model was more accurate and
-precise than the aggregate level model when predicting out-of-sample
-observations. When comparing climate effects from both models, the
-population-level model tends to miss important climate effects from at
-least one vital rate for each species. Increasing the sample size at the
-population-level would not necessarily reduce forecast uncertainty; the
-way to reduce uncertainty is to capture unique climate dependence of
-individual vital rates. Our analysis indicates that there is no shortcut
-to forecasting climate change impacts on plant populations --- detailed
-demographic data is essential. Despite the superiority of the
-individual-level model, the forecasts it generated still were too
-uncertain to be useful for decision-makers. We need new methods to
-collect demographic data more efficiently across environmental gradients
-in space and time.
+grass species in a semi-arid grassland in Montana, USA. We parameterized
+two population models based on the same dataset, one based on
+individual-level data (survival, growth and recruitment) and one on
+population-level data (percent cover), and compared their accuracy,
+precision, and sensitivity to climate variables. The individual-level
+model was more accurate and precise than the aggregated model when
+predicting out-of-sample observations. When comparing climate effects
+from both models, the population-level model missed important climate
+effects from at least one vital rate for each species. Increasing the
+sample size at the population-level would not necessarily reduce
+forecast uncertainty; the way to reduce uncertainty is to capture unique
+climate dependence of individual vital rates. Our analysis indicates
+that there is no shortcut to forecasting climate change impacts on plant
+populations --- detailed demographic data is essential. Despite the
+superiority of the individual-level model, the forecasts it generated
+still were too uncertain to be useful for decision-makers. We need new
+methods to collect demographic data efficiently across environmental
+gradients in space and time.
 
@@
 and model parameters (Luo et al. 2011, but see Perretti et al. 2013 for
-an argument against modeling the ecological process). This requires
-large amounts of data collected over broad spatio-temporal extents, and
-state-of-the-science modeling techniques cannot overcome data
-limitations. Population models are a case in point.
-
-Population models are important tools for predicting the impacts of
-environmental change on species persistence and abundance. But
-reconciling the scales at which population models are parameterized and
-the scales at which environmental changes play out remains a challenge
-(Clark et al. 2010, 2012, Freckleton et al. 2011, Queenborough et al.
-2011). The problem is that most population models are built using data
-from a single study site because collecting those data, which involves
-tracking the fates of individuals plants, is so difficult. The resulting
-models cannot be applied to the landscape and regional scales relevant
-to decision-makers without information about how the fitted parameters
-respond to spatial variation in biotic and abiotic drivers (Sæther et
-al. 2007). The limited spatial extent of individual-level demographic
-datasets constrains our ability to use population models to address
-applied questions about the consequences of climate change.
+an argument against modeling the ecological process). The increasing
+statistical sophistication of population models (Rees and Ellner 2009)
+makes them promising tools for predicting the impacts of environmental
+change on species persistence and abundance. But reconciling the scales
+at which population models are parameterized and the scales at which
+environmental changes play out remains a challenge (Clark et al. 2010,
+2012, Freckleton et al. 2011, Queenborough et al. 2011). The problem is
+that most population models are built using data from a single study
+site because collecting those data, which involves tracking the fates of
+individuals plants, is so difficult. The resulting models cannot be
+applied to the landscape and regional scales relevant to decision-making
+without information about how the fitted parameters respond to spatial
+variation in biotic and abiotic drivers (Sæther et al. 2007). The
+limited spatial extent of individual-level demographic datasets
+constrains our ability to use population models to address applied
+questions about the consequences of climate change.
 
@@
 relatively straightforward to parameterize and project SDMs over
-landscapes and regions, the scales at which many land-use decisions are
-made. However, the limitations of SDMs are well known (Pearson and
-Dawson 2003, Elith and Leathwick 2009, Araújo and Peterson 2012).
-Ideally, researchers would provide managers with landscape-scale
+landscapes and regions. However, the limitations of SDMs are well known
+(Pearson and Dawson 2003, Elith and Leathwick 2009, Araújo and Peterson
+2012). Ideally, researchers would provide managers with landscape-scale
 population models, combining the extent of SDMs with information about
@@
 (Clark and Bjørnstad 2004, Freckleton et al. 2011). Population-level
-data cannot provide inference about demographic mechanisms, but it might
-be sufficient for modeling future population states, especially since it
-is feasible to collect across broad spatial extents (e.g., Queenborough
-et al. 2011). The choice between individual and population-level data
-involves a difficult trade-off: while individual-level data leads to
-more mechanistic models, population-level data leads to models that can
-be applied over greater spatial and temporal extents. An open question
-is how much forecasting skill is lost when we build models based on
-population rather than individual-level data.
+data cannot provide inference about demographic mechanisms, but might be
+sufficient for modeling future population states, especially since such
+data are feasible to collect across broad spatial extents (e.g.,
+Queenborough et al. 2011). The choice between individual and
+population-level data involves a difficult trade-off: while
+individual-level data leads to more mechanistic models, population-level
+data leads to models that can be applied over greater spatial and
+temporal extents. An open question is how much forecasting skill is lost
+when we build models based on population rather than individual-level
+data.
 
@@
 discrete density state rather than a continuous density measure. Such
-models do not require individual level demographic data and can
-adequately describe population dynamics.
-
-Building on Taylor and Hastings (2004), Freckleton et al. (2011) showed
-that density-structured models compare well to continuous models in
-theory, and Queenborough et al. (2011) demonstrated the application of
-such methods in a study on arable weeds. In particular, Queenborough et
-al. (2011) provide empirical evidence that density-structured models are
-capable of reproducing population dynamics, even if some precision is
-lost when compared to fully continuous models. The study by Queenborough
-et al. (2011) included data from 500 fields (4 hectares each) in 49
-farms, all collected by two people in 6 weeks. This is far more data
-from a far greater spatial extent than would be possible if measuring
-individual plant demography (in a world of limited time and money). The
-appeal of density-structured approaches is clear. However, none of these
-models included environmental covariates.
+models do not require individual-level demographic data and can
+adequately describe population dynamics. Building on Taylor and Hastings
+(2004), Freckleton et al. (2011) showed that density-structured models
+compare well to continuous models in theory, and Queenborough et al.
+(2011) provide empirical evidence that density-structured models are
+capable of reproducing population dynamics at landscape spatial scales,
+even if some precision is lost when compared to fully continuous models.
+The appeal of density-structured approaches is clear. However, none of
+these models included environmental covariates.
 
@@
 al. 1988). Population growth (or decline) is the outcome of demographic
-processes like survival, growth, and recruitment that occur at the level
-of individual plants. Climate can affect each demographic process in
-unique, potentially opossing, ways (Dalgleish et al. 2011). These unique
-climate responses may be difficult to resolve in statistical models
-based on population-level data where demographic processes are not
-identifiable. If important climate drivers are missed because they are
-averaged over in population-level data, then population models built
-with such data will make uninformative or unreliable forecasts when
-confronted with novel climate covariates.
-
-Here, we ask whether statistical and population models based on
-aggregated, population-level data can predict climate change impacts as
-well as models based on individual-level data. We used a unique
-demographic dataset that tracks the fates of individual plants from four
-species over 14 years to build two kinds of single-species population
-models, traditional models using individual growth, survival, and
-recruitment data and alternative models based on basal cover. In both
-models, interannual variation is explained, in part, by climate
-covariates. We then performed simulations to quantify the sensitivities
-of species' cover to small perturbations in average precipitation and
-temperature. Individual level models were clearly better, but,
+processes such as survival, growth, and recruitment that occur at the
+level of individual plants. Climate can affect each demographic process
+in unique, potentially opossing, ways (Dalgleish et al. 2011). These
+unique climate responses may be difficult to resolve in statistical
+models based on population-level data where demographic processes are
+not identifiable. If important climate effects are missed because of the
+aggregation inherent in in population-level data, then population models
+built with such data will make uninformative or unreliable forecasts.
+
+Here, we compare the forecasting skill of statistical and population
+models based on aggregated, population-level data with models based on
+individual-level data. We used a unique demographic dataset that tracks
+the fates of individual plants from four species over 14 years to build
+two kinds of single-species population models, traditional models using
+individual growth, survival, and recruitment data and alternative models
+based on basal cover. In both models, interannual variation is
+explained, in part, by climate covariates. We first quantify forecasting
+skill using cross-validation. We then performed simulations to quantify
+the sensitivities of species' cover to small perturbations in average
+precipitation and temperature. Based on the cross-validation results,
+predictions of individual level models were clearly better, but,
 unfortunately, still too uncertain to inform management decisions.
@@
 
-In this paper, we model populations based on two levels of data:
-individual and quadrat (Fig. 2). The individual data is the ``raw''
-data. For the quadrat level we data we simply sum individual basal cover
-for each quadrat by species. This is equivalent to a near-perfect census
-of quadrat percent cover because previous analysis shows that
-measurement error at the individual level is small (Chu and Adler 2014).
-Based on these two datasets we can compare population models built using
-individual level data and aggregated quadrat level data.
+We modeled each grass population based on two levels of data: individual
+and quadrat (Fig. 2). The individual data is the ``raw'' data. For the
+quadrat-level we data we simply sum individual basal cover for each
+quadrat by species. This is equivalent to a near-perfect census of
+quadrat percent cover because previous analysis shows that measurement
+error at the individual-level is small (Chu and Adler 2014). Based on
+these two datasets we can compare population models built using
+individual-level data and aggregated, quadrat-level data.
 
@@
 blocks of our population models are vital rate regressions. For
-individual-level data, we fit models for survival, growth, and
+individual-level data, we fit regressions for survival, growth, and
 recruitment for each species. At the quadrat-level, we fit a single
@@
 ($X = S, G, R, or P$ for survival, growth, recruitment, or population
-growth). For example, $\textbf{y}^S$ is a vector of 0s and 1s indicating
-whether a genet survives from \emph{t} to \emph{t+1}, or not.
+growth). For example, $\textbf{y}^S$ is a vector of 0's and 1's
+indicating whether a genet survives from \emph{t} to \emph{t+1}, or not.
 
@@
 Our data allows us to track new recruits, but we cannot assign a
-specific parent to new genets. So, for recruitment, we work at the
-quadrat level and model the number of new individuals of species $j$ in
-quadrat $q$ recruiting at time $t+1$ as a function of quadrat
-``effective cover'' ($A'$) in the previous year ($t$). Effective cover
-is a mixture of observed cover ($A$) in the focal quadrat ($q$) and the
-mean cover across the entire group ($\bar{A}$) of $Q$ quadrats in which
-$q$ is located:
+specific parent to new genets. THerefore, we model recruitment at the
+quadrat level: the number of new individuals of species $j$ in quadrat
+$q$ recruiting at time $t+1$ as a function of quadrat ``effective
+cover'' ($A'$) in the previous year ($t$). Effective cover is a mixture
+of observed cover ($A$) in the focal quadrat ($q$) and the mean cover
+across the entire group ($\bar{A}$) of $Q$ quadrats in which $q$ is
+located:
 
@@
 
-The statistical approach used to model vital rates using aggregated data
-depends on the type of data collected. In our case, and as is often the
-case with census data, we have percent cover data (which can easily be
-transformed to proportion data). We first considered fitting three vital
+The statistical approach used to model aggregated data depends on the
+type of data collected. We have percent cover data, which can easily be
+transformed to proportion data. We first considered fitting three vital
 rate models analagous to those we fit at the individual level: one for
@@
 vital rates at the population level to change in percent cover within
-quadrats. For the remaining discussion of statistical modeling we refer
-to proportion data, which is simply percent data divided by 100.
+quadrats. For the remaining discussion of statistical modeling, we refer
+to proportion data, which is simply percent cover divided by 100.
 
@@
 therefore chose a more constrained modeling approach based on a
-truncated log-normal likelihood. The model for quadrat cover change
-($G$) from time $t$ to $t+1$ is
+truncated log-normal likelihood. The model for quadrat cover change from
+time $t$ to $t+1$ is
 
@@
 were drawn from global distributions representing the mean size effect
-and intercept. We used uninformative priors (Appendix A).
+and intercept. Quadrat random effects were also fit hierarchically, with
+quadrat offsets being drawn from distributions with mean zero and a
+shared variance term (independent Gaussian priors, Appendix A). Climate
+effects were not modeled hierarchically, and each was given a diffuse
+prior distribution. We used standard diffuse priors for all unknown
+parameters (Appendix A).
 
@@
 model parameters from three parallel MCMC chains run for 1,000
-iterations after discarding an initial 1,000 iterations. We recognize
-such short MCMC chains may surprise those more familiar with other MCMC
-samplers (i.e.~JAGS or WinBUGS), but the Stan sampler is exceptionally
-efficient, which reduces the number of iterations needed to achieve
-convergence. We assessed convergence visually and made sure scale
-reduction factors for all parameters were less than 1.01. For the
-purposes of including parameter uncertainty in our population models, we
-saved the final 1,000 iterations from each of the three MCMC chains to
-be used as randomly drawn values during population simulation. This step
-alleviates the need to reduce model parameters by model selection since
-sampling from the full parameter space in the MCMC ensures that if a
-parameter broadly overlaps zero, on average the effect in the population
-models will also be near zero. We report the posterior mean, standard
-deviation, and 95\% Bayesian Credible Intervals for every parameter of
-each model for each species in Appendix B.
+iterations after discarding an initial 1,000 iterations. Such short MCMC
+chains may surprise readers more familiar with other MCMC samplers
+(i.e.~JAGS or WinBUGS), but the Stan sampler is exceptionally efficient,
+which reduces the number of iterations needed to achieve convergence. We
+assessed convergence visually and made sure scale reduction factors for
+all parameters were less than 1.01. For the purposes of including
+parameter uncertainty in our population models, we saved the final 1,000
+iterations from each of the three MCMC chains to be used as randomly
+drawn values during population simulation. This step alleviates the need
+to reduce model parameters by model selection since sampling from the
+full parameter space in the MCMC ensures that if a parameter broadly
+overlaps zero, on average the effect in the population models will also
+be near zero. We report the posterior mean, standard deviation, and 95\%
+Bayesian Credible Intervals for every parameter of each model for each
+species in Appendix B.
 
@@
 an Integral Projection Model (IPM) to model populations based on
-individual level data (Ellner and Rees 2006) and a quadrat-based version
+individual-level data (Ellner and Rees 2006) and a quadrat-based version
 of an individually-based model (Quadrat-Based Model, QBM) to model
-populations based on quadrat level data. We describe each in turn.
+populations based on quadrat-level data. We describe each in turn.
 
@@
 
-We use an environmentally stochastic IPM (Mark and Ellner 2009) that
-includes the climate covariates from the vital rate statistical models.
-In all simulations we ignore the random year effects so that interannual
-variation is driven solely by climate. We fit the random year effects in
-the vital rate regressions to avoid over-attributing variation to
-climate covariates. Our IPM follows the specification of Chu and Adler
-(2015) where the population of species \emph{j} is a density function
-$n(u_{j},t)$ giving the density of sized-\emph{u} genets at time
-\emph{t}. Genet size is on the natural log scale, so that $n(u_{j},t)du$
-is the number of genets whose area (on the arithmetic scale) is between
-$e^{u_{j}}$ and $e^{u_{j}+du}$. The density function for any size
-\emph{v} at time $t+1$ is
+We use a stochastic IPM (Rees and Ellner 2009) that includes the climate
+covariates from the vital rate statistical models. In all simulations we
+ignore the random year effects so that interannual variation is driven
+solely by climate. We fit the random year effects in the vital rate
+regressions to avoid over-attributing variation to climate covariates.
+Our IPM follows the specification of Chu and Adler (2015) where the
+population of species \emph{j} is a density function $n(u_{j},t)$ giving
+the density of sized-\emph{u} genets at time \emph{t}. Genet size is on
+the natural log scale, so that $n(u_{j},t)du$ is the number of genets
+whose area (on the arithmetic scale) is between $e^{u_{j}}$ and
+$e^{u_{j}+du}$. The density function for any size \emph{v} at time $t+1$
+is
 
 \begin{equation}
-n(v_{j},t+1) = \int_{L_{j}}^{U_{j}} k_{j}(v_{j},u_{j},\bar{\bold{w_{j}}}(u_{j}))n(u_{j},t)
+n(v_{j},t+1) = \int_{L_{j}}^{U_{j}} k_{j}(v_{j},u_{j},\bar{w_{j}}(u_{j}))n(u_{j},t)
 \end{equation}
 
-where $k_{j}(v_{j},u_{j},\bar{\bold{w_{j}}})$ is the population kernel
-that describes all possible transitions from size $u$ to $v$ and
-$\bar{\bold{w_{j}}}$ is a vector of estimates of average crowding
-experienced from all other species by a genet of size $u_j$ and species
-$j$. The integral is evaluated over all possible sizes between
-predefined lower (\emph{L}) and upper (\emph{U}) size limits that extend
-beyond the range of observed genet sizes.
+where $k_{j}(v_{j},u_{j},\bar{w_{j}})$ is the population kernel that
+describes all possible transitions from size $u$ to $v$ and
+$\bar{w_{j}}$ is a scalar representing the average intraspecific
+crowding experienced by a genet of size $u_j$ and species $j$. The
+integral is evaluated over all possible sizes between predefined lower
+(\emph{L}) and upper (\emph{U}) size limits that extend beyond the range
+of observed genet sizes.
 
@@
 crowding for specific genets ($w_{ij}$). Instead, we use an
-approximation ($\bar{\bold{w_{j}}}$) that captures the essential
-features of neighborhood interactions (Adler et al. 2010). We also use a
-`no-overlap' rule for conspecific genets to approximate the
-overdispersion of large genets in space (Adler et al. 2010, 2012, Chu
-and Adler 2015).
+approximation ($\bar{w_{j}}$) that captures the essential features of
+neighborhood interactions (Adler et al. 2010). This approximation relies
+on a `no-overlap' rule for conspecific genets to approximate the
+overdispersion of large genets in space (Adler et al. 2010).
 
@@
 assume that fecundity increases linearly with size
-($R_j(v_{j},u_{j},\bar{\bold{w_{j}}}) = e^{u_j}R_j(v_{j},\bar{\bold{w_{j}}})$)
-to incorporate the recruitment function in the spatially-implicit IPM.
+($R_j(v_{j},u_{j},\bar{w_{j}}) = e^{u_j}R_j(v_{j},\bar{w_{j}})$) to
+incorporate the recruitment function in the spatially-implicit IPM.
 
@@
 iteration. Using this approach, rather than simply using coefficient
-point estimates, ensures that relatively unimportant climate covariates
-(those that broadly overlap 0) have little effect on the simulation
-results. Since our focus was on the contribution of climate covariates
-to population states, we set the random year effects and the random
-group effects to zero.
+point estimates, captures the effect of parameter uncertainty.
+Relatively unimportant climate covariates (those that broadly overlap 0)
+will have little effect on the mean of the simulation results, but can
+contribute to their variation. Since our focus was on the contribution
+of climate covariates to population states, we set the random year
+effects and the random group effects to zero.
 
@@
 
-Our quad-based model (QBM) perfectly mirrors its statistical description
-(Eqs. 9-10). We use the same approach for drawing parameter values as
-described for the IPM. After drawing the appropriate parameter set, we
-calculate the mean response (population cover at \emph{t+1} = $x_{t+1}$)
-according to Eq. 9. We then make a random draw from a {[}0,1{]}
-truncated lognormal distribution with mean equal to $x_{t+1}$ from Eq. 9
-and the variance estimate from the fitted model. We can then project the
-model forward by drawing a new parameter set (unique to climate year and
-MCMC iteration) at each timestep. As with the IPM, random year effects
-are ignored for all simulations.
+To simulate our quad-based model (QBM), we simply iterate the
+quadrat-level statistical model (Eqs. 9-10). We use the same approach
+for drawing parameter values as described for the IPM. After drawing the
+appropriate parameter set, we calculate the mean response (population
+cover at \emph{t+1} = $x_{t+1}$) according to Eq. 9. We then make a
+random draw from a {[}0,1{]} truncated lognormal distribution with mean
+equal to $x_{t+1}$ from Eq. 9 and the variance estimate from the fitted
+model. We can then project the model forward by drawing a new parameter
+set (unique to climate year and MCMC iteration) at each timestep. As
+with the IPM, random year effects are ignored for all simulations.
 
@@
 whose observations were withheld from model fitting. Within each
-observation year, several quadrats were sampled. So we made predictions
-for each observed quadrat in the focal year, initializing each
-simulation with cover in the quadrat the previous year. Since we were
-making quadrat-specific predictions, we incorporated the group effect on
+observation year, several quadrats were sampled. We made predictions for
+each observed quadrat in the focal year, initializing each simulation
+with cover in the quadrat the previous year. Since we were making
+quadrat-specific predictions, we incorporated the group random effect on
 the intercept for both models. We repeated this procedure for all 13
@@
 
-This model validation allowed us to compare accuracy and precision of
-the two modeling approaches (IPM versus QBM). We first calculated the
-median predicted cover across the 100 simulations for each quadrat-year
-and then calculated the absolute error as the absolute value of the
-difference between the observed cover for a given quadrat-year and the
-median prediction. To arrive at mean absolute error (MAE), we then
-averaged the absolute error within each species across the quadrat-year
-specific errors. We use MAE as our measure of accuracy. To measure
-precision we calculated the distance between the upper and lower 90th
-quantiles of the 100 predictions and averaged this value over
+This cross-validation procedure allowed us to compare accuracy and
+precision of the two modeling approaches (IPM versus QBM). We first
+calculated the median predicted cover across the 100 simulations for
+each quadrat-year and then calculated the absolute error as the absolute
+value of the difference between the observed cover for a given
+quadrat-year and the median prediction. To arrive at mean absolute error
+(MAE), we then averaged the absolute error within each species across
+the quadrat-year specific errors. We use MAE as our measure of accuracy.
+To measure precision we calculated the distance between the upper and
+lower 90th quantiles of the 100 predictions and averaged this value over
 quadrat-years for each species.
@@
 
-Our expectation is that the IPM will produce more accurate and precise
-forecasts. This could be due to (1) the smaller sample size of the
-quadrat level data sets compared to the individual level data sets,
-leading to larger parameter uncertainty for the QBM, or (2) the QBM
-climate effects being weakly associated with one or more vital rate
-climate effects at the individual level. To assess the impact of sample
-size on QBM parameter uncertainty we refit the QBM statistical model
-(Eqs. 9-10) after removing sets of 2, 5, 10, and 15 quadrats. We fit 10
-models at each level of quadrat removal (2, 5, 10, 15 quadrats),
-removing a different randomly selected set of quadrats for each fit. We
-calculated the standard deviation of climate main effects (pptLag, ppt1,
-ppt2, TmeanSpr1, and TmeanSpr2) for each model and averaged those over
-replicates within each set of quadrat removals. This allowed us to
-regress sample size against parameter uncertainty.
-
-To see if the QBM climate effects are correlated, or not, with climate
-effects for each vital rate model for the IPM, we simply regressed the
+We expected the IPM to produce more accurate and precise forecasts due
+to either (1) the smaller sample size of the quadrat level data sets
+compared to the individual level data sets, leading to larger parameter
+uncertainty for the QBM, or (2) the QBM climate effects being weakly
+associated with one or more vital rate climate effects at the individual
+level. To assess the impact of sample size on QBM parameter uncertainty
+we refit the QBM statistical model (Eqs. 9-10) after removing sets of 2,
+5, 10, and 15 quadrats. We fit 10 models at each level of quadrat
+removal (2, 5, 10, 15 quadrats), removing a different randomly selected
+set of quadrats for each fit. We calculated the standard deviation of
+climate main effects (pptLag, ppt1, ppt2, TmeanSpr1, and TmeanSpr2) for
+each model and averaged those over replicates within each set of quadrat
+removals. This allowed us to regress parameter uncertainty against
+sample size.
+
+To deterime if the QBM climate effects are correlated with climate
+effects for each vital rate model in the IPM, we simply regressed the
 QBM climate coefficients against each vital rate model's climate
@@
 
-The IPM had significantly lower overall error (MAE, mean absolute error)
-for three species (\emph{B. gracilis}, \emph{H. comata}, \emph{P.
-smithii}; Table 1). In no case did the QBM significantly outperform the
-IPM (Table 1). The IPM was consistently more precise than the QBM, with
-lower distances between the 90\% quantiles across all species (Table 1).
-In general the IPM outperformed the QBM because it had (1) lower MAE for
-three of the four species, (2) statistically similar MAE for the one
-other species, and (3) considerably more precise forecasts for all
-species.
+The IPM had lower overall error (MAE, mean absolute error) compared to
+the QBM for three species (Table 1). The IPM MAE is significantly lower
+at $\alpha=0.05$ for \emph{B. gracilis} (\emph{P} = 0.0012), \emph{H.
+comata} (\emph{P} = 4.0586 × 10-8), and \emph{P. smithii} (\emph{P} =
+3.183 × 10-5). MAEs are statisticially similar between models for
+\emph{P. secunda} (\emph{P} = 0.0922). \emph{P} values are highly
+sensitive to sample size, so not entirely appropriate in simulation
+exercises where we control the samples size. But, for our purposes they
+serve as relatively unbiased comparison metrics. In no case did the QBM
+significantly outperform the IPM (Table 1). The IPM was consistently
+more precise than the QBM, with lower distances between the 90\%
+quantiles across all species (Table 1). In general the IPM outperformed
+the QBM because it had (1) lower MAE for three of the four species, (2)
+statistically similar MAE for the one other species, and (3)
+considerably more precise forecasts for all species.
 
@@
 al. 2011). However, density-structured models rely on the aggregation of
-individual level data. This creates a potential problem if such models
+individual-level data. This creates a potential problem if such models
 are to be used in a climate change context because it is individuals,
 not populations, which respond to climate (Clark et al. 2012). Are
-models based on population level metrics as sensitive to climate as
-models based on individual level metrics? Do these two types of models
+models based on population-level metrics as sensitive to climate as
+models based on individual-level metrics? Do these two types of models
 produce consistent forecasts? Do we need detailed demographic data to
@@
 that it would lead to higher precision of climate coefficient estimates
-(Fig. 4). We did, however, find evidence that the QBM statistical model
-failed to identify climate dependence for some vital rates (Fig. 5). For
-no species were climate effects from the QBM strongly correlated with
-all three vital rates (Fig. 5). Freckleton et al. (2011) acknowledge
-that averaging over complex stage dependence will lead to poorly
-specified models. This is analagous to our situation, but instead of
-averaging over complex life histories, we are averaging over complex
-climate dependence.
+(Fig. 4).
+
+We did, however, find evidence that the QBM statistical model failed to
+identify climate dependence for some vital rates (Fig. 5). For no
+species were climate effects from the QBM strongly correlated with all
+three vital rates (Fig. 5). Freckleton et al. (2011) acknowledge that
+averaging over complex stage dependence will lead to poorly specified
+models. This is analagous to our situation, but instead of averaging
+over complex life histories, we are averaging over complex climate
+dependence. Though our work here focused on plant species, this finding
+is applicable to any species with vital rates that respond uniquely to
+weather/climate.
 
@@
 case, as it is for all our species to varying degrees (Fig. 3),
-statistical models of population-level data will fail. Our result is
-consistent with related work on the importance of individual-level data
-to forecast population responses to exogenous drivers (Clark et al.
-2011a, 2011b, 2012, Galván et al. 2014).
+forecasts from models based on population-level data will fail. Our
+result is consistent with related work on the importance of
+individual-level data to forecast population responses to exogenous
+drivers (Clark et al. 2011a, 2011b, 2012, Galván et al. 2014).
 
@@
 here could be considered at the forefront of ecological forecasting with
-repsect to the statistical approach employed (hierarchical Bayesian),
+respect to the statistical approach employed (hierarchical Bayesian),
 the type of population model we used (density-dependent, stochastic IPM
 with parameter uncertainty), and the amount of high quality data we had
-at our disposal (14 years of individual level data). Yet, model
+at our disposal (14 years of individual-level data). Yet, model
 predictions proved so uncertain that any forecast, when bounded with
-model and parameter uncertainty, would be at best not useful and at
-worst meaningless.
+model and parameter uncertainty, would be uninformative.
 
@@
 spatial scale of observations. One of the major limitations of the
-models we fit here is that the climate data are at a much larger scale
-than the individual level observations of plant size. Climate covariates
-only vary by year, with no spatial variability within years. Thus, even
-if we fit models to individual-level data, we are missing the key
-interaction point between weather and individual plants (Clark et al.
-2011b) because all observations share the same climate covariates.
+models we fit here is that the climate data are collected at a larger
+scale than the individual-level observations of plant size. Climate
+covariates only vary by year, with no spatial variability within years.
+Thus, even if we fit models to individual-level data, we are missing the
+key interaction point between weather and individual plants (Clark et
+al. 2011b) because all observations share the same climate covariates.
 Demographic studies should be designed with at least plot-level
@@
 when using modeling approaches like integral projection models that
-separate the model fitting and simulation stages (Mark and Ellner 2009).
+separate the model fitting and simulation stages (Rees and Ellner 2009).
 However, as we have done here, it is still possible to include parameter
@@
 begin to produce better ones, and forecasts reported without parameter
-error are disingenuous. In many cases, ignoring parameter error is
-justifiable when investigating basic demographic mechanisms, but it is
-not justifiable when forecasting is the goal.
+error are disingenuous. Ignoring parameter error may be justifiable when
+the goal is investigating basic processes, but it is indefensible when
+forecasting is the goal.
 
@@
 there is no short cut to producing accurate and precise population
-forecasts. Do we need detailed demographic data to forecast the impacts
-of climate change on populations? Probably. Given the importance of
-demographic data and its current collection cost, we need modern methods
-to collect demographic data more efficiently across environmental
-gradients in space and time.
+forecasts: we need detailed demographic data to forecast the impacts of
+climate change on populations. Given the importance of demographic data
+and its current collection cost, we need modern methods to collect
+demographic data more efficiently across environmental gradients in
+space and time.
 
@@
 was large for both model types. Even with 14 years of detailed
-demographic data and sophisticated modeling techniques, we failed to
-produce forecasts with any level of acceptable uncertainty. Uncertainty
-of climate change related forecasts can be reduced by (1) longer time
-series and (2) climate covariates that match the scale of inference
+demographic data and sophisticated modeling techniques, our projections
+contained too much uncertainty to be informative. Uncertainty in
+demographic responses to climate can be reduced by collecting (1) longer
+time series and (2) climate covariates that match the scale of inference
 (e.g., plot rather than landscape level climate/weather metrics).
@@
 series of meetings between the Adler and Weecology labs at USU sharpened
-our thinking. Brittany Teller and two anonymous reviewers provided
-comments that improved our manuscript. Compute, storage and other
-resources from the Division of Research Computing in the Office of
-Research and Graduate Studies at Utah State University are gratefully
-acknowledged.
+our thinking. Brittany Teller provided comments that improved our
+manuscript. Compute, storage and other resources from the Division of
+Research Computing in the Office of Research and Graduate Studies at
+Utah State University are gratefully acknowledged.
 
@@
 
-\emph{NOTES}: The IPM MAE is significantly lower at $\alpha=0.05$ for
-\emph{B. gracilis} (\emph{P} = 0.0012), \emph{H. comata} (\emph{P} =
-4.0586 × 10-8), and \emph{P. smithii} (\emph{P} = 3.183 × 10-5). MAEs
-are statisticially similar between models for \emph{P. secunda}
-(\emph{P} = 0.0922). \emph{P} values are highly sensitive to sample
-size, so not entirely appropriate in simulation exercises where we
-control the samples size. But, for our purposes they serve as relatively
-unbiased comparison metrics.
-
-\pagebreak{}
-
 \pagebreak{}
@@
 \includegraphics{components/figure/manuscript-figure_6.pdf}
-\caption{Equilibrium cover and 90\% quantiles around the mean prediction
-when model error and parameter uncertainty are propogated through the
-simulation phase. Climate simulations are as in Figure 3.}
+\caption{Mean (points) and 90\% quantiles (errorbars) for the
+proportional difference between baseline simulations (using observed
+climate) and the climate pertubation simulation on the x-axis. We
+calculated proportional difference as log(perturbed climate cover) -
+log(observed climate cover), where `perturbed' and `observed' refer to
+the climate time series used to drive interannual variation in the
+simulations. Model error and parameter uncertainty were propagated
+through the simulation phase. Climate simulations are as in Figure 3.}
 \end{figure}
@@
 
-Mark, R., and S. P. Ellner. 2009. Integral projection models for
-populations in temporally varying environments. Ecological Monographs
-79:575--594.
-
 Merow, C., A. M. Latimer, A. M. Wilson, S. M. McMahon, A. G. Rebelo, and
@@
 
+Rees, M., and S. P. Ellner. 2009. Integral projection models for
+populations in temporally varying environments. Ecological Monographs
+79:575--594.
+
 Schurr, F. M., J. Pagel, J. S. Cabral, J. Groeneveld, O. Bykova,

### 2015-08-31 manuscript/manuscript.Rmd
updates page numbering for ESA siwtch around

@@
    - \usepackage{todonotes}
+   - \usepackage[document]{ragged2e}
+   - \pagewiselinenumbers
 bibliography: ~/Dropbox/Bibliography/MicroMeso.bib
@@
 
-\begin{centering}
-\textbf{\large{Do we need demographic data to forecast population responses to climate change?}}
-
-\textsc{\small{Andrew T. Tredennick\footnote{E-mail: atredenn@gmail.com} and Peter B. Adler}}
-
-\textit{\small{Department of Wildland Resources and the Ecology Center, 5230 Old Main Hill, Utah State University, Logan, Utah 84322-5230 USA}}
-
-\end{centering}
+### Title
+Do we need demographic data to forecast population responses to climate change?
+
+### Authors 
+1. Andrew T. Tredennick (corresponding author: atredenn@gmail.com) 
+2. Peter B. Adler (peter.adler@usu.edu)
+
+### Affiliations
+1. Department of Wildland Resources and the Ecology Center, 5230 Old Main Hill, Utah State University, Logan, Utah 84322-5230 USA
+2. Department of Wildland Resources and the Ecology Center, 5230 Old Main Hill, Utah State University, Logan, Utah 84322-5230 USA
+
+\newpage{}
 
@@
 *Key words: forecasting, climate change, grassland, integral projection model, population model*
+
+\newpage{}
 
@@
 
+\setcounter{page}{29}
+
 Tables
@@
 
+Figure Legends
+--------------
+Figure 1.
+Time series of average percent cover over all quadrats for our four focal species: *Bouteloua gracilis* (BOGR), *Hesperostipa comata* (HECO), *Pascopyrum smithii* (PASM), and *Poa secunda* (POSE).
+Light grey lines show trajectories of individual quadrats.
+Note the different y-axis scales across panels.
+
+Figure 2. Work flow of the data aggregation, model fitting, and population simulating.
+
+Figure 3.
+Sensitivity of equilibrium cover simulated from the IPM to each climate scenario applied to individual and combined vital rates.
+For example, the points associated with G show the median cover from IPM simulations where a climate perturbation is applied only to the growth regression climate covariates.
+These simulations use mean parameter values for clarity.
+
+Figure 4.
+Effect of quadrat sample size on the precision (standard deviation) of main climate effect estimates in the QBM.
+Increasing the number of quadrats results in diminishing returns in terms of parameter certainty.
+Light dashed lines show individual climate effects at five quadrat sample sizes.
+Thick dark lines are inverse gaussian fits showing the mean effect of increasing quadrat sample size on parameter precision.
+
+Figure 5.
+Correlations (r) between QBM and IPM estimates of climate effects.
+We ignore sizeXclimate interactions since these are not directly comparable across model types.
+The QBM does not have multiple vital rates, so its values are repeated across panels within each species.
+Across top panels, 'growth' = growth regression, 'rec' = recruitment regression, 'surv' = survival regression.
+
+Figure 6.
+Mean (points) and 90% quantiles (errorbars) for the proportional difference between baseline simulations (using observed climate) and the climate pertubation simulation on the x-axis.
+We calculated proportional difference as log(perturbed climate cover) - log(observed climate cover), where 'perturbed' and 'observed' refer to the climate time series used to drive interannual variation in the simulations.
+Model error and parameter uncertainty were propagated through the simulation phase.
+Climate simulations are as in Figure 3.
+
+\pagebreak{}
+
 Figures
@@
 
+\setcounter{page}{24}
+
 References

### 2015-08-31 manuscript/manuscript.tex
updates page numbering for ESA siwtch around

@@
 \usepackage{todonotes}
+\usepackage[document]{ragged2e}
+\pagewiselinenumbers
 
@@
 
-\normalsize
-
-\begin{centering}
-\textbf{\large{Do we need demographic data to forecast population responses to climate change?}}
-
-\textsc{\small{Andrew T. Tredennick\footnote{E-mail: atredenn@gmail.com} and Peter B. Adler}}
-
-\textit{\small{Department of Wildland Resources and the Ecology Center, 5230 Old Main Hill, Utah State University, Logan, Utah 84322-5230 USA}}
-
-\end{centering}
+\maketitle
+
+\subsubsection{Title}\label{title}
+
+Do we need demographic data to forecast population responses to climate
+change?
+
+\subsubsection{Authors}\label{authors}
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\itemsep1pt\parskip0pt\parsep0pt
+\item
+  Andrew T. Tredennick (corresponding author:
+  \href{mailto:atredenn@gmail.com}{\nolinkurl{atredenn@gmail.com}})
+\item
+  Peter B. Adler
+  (\href{mailto:peter.adler@usu.edu}{\nolinkurl{peter.adler@usu.edu}})
+\end{enumerate}
+
+\subsubsection{Affiliations}\label{affiliations}
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\itemsep1pt\parskip0pt\parsep0pt
+\item
+  Department of Wildland Resources and the Ecology Center, 5230 Old Main
+  Hill, Utah State University, Logan, Utah 84322-5230 USA
+\item
+  Department of Wildland Resources and the Ecology Center, 5230 Old Main
+  Hill, Utah State University, Logan, Utah 84322-5230 USA
+\end{enumerate}
+
+\newpage{}
 
@@
 
+\newpage{}
+
 \subsection{Introduction}\label{introduction}
@@
 account for potential differences among the grazing treatments, but do
-not focus on grazing$\times$climate interactions. The annual maps of the
-quadrats were digitized and the fates of individual plants tracked and
-extracted using a computer program (Lauenroth and Adler 2008, Chu et al.
-2014). Daily climate data are available for the duration of the data
+not focus on grazing\(\times\)climate interactions. The annual maps of
+the quadrats were digitized and the fates of individual plants tracked
+and extracted using a computer program (Lauenroth and Adler 2008, Chu et
+al. 2014). Daily climate data are available for the duration of the data
 collection period (1932 - 1945) from the Miles City airport, Wiley
@@
 the observation year. We also include interactions among same-year
-climate covariates (e.g., ppt1 $\times$ TmeansSpr1) and climate $\times$
-size interactions. Climate $\times$ size interactions are for climate
-main effects only; we do not include interactions between size and pairs
-of interacting climate effects.
+climate covariates (e.g., ppt1 \(\times\) TmeansSpr1) and climate
+\(\times\) size interactions. Climate \(\times\) size interactions are
+for climate main effects only; we do not include interactions between
+size and pairs of interacting climate effects.
 
@@
 fully descibed in Appendix A, so here we focus on the main process and
-the model likelihood. For the likelihood models, $\textbf{y}^X$ is
+the model likelihood. For the likelihood models, \(\textbf{y}^X\) is
 always the relevant vector of observations for vital rate \emph{X}
-($X = S, G, R, or P$ for survival, growth, recruitment, or population
-growth). For example, $\textbf{y}^S$ is a vector of 0's and 1's
+(\(X = S, G, R, or P\) for survival, growth, recruitment, or population
+growth). For example, \(\textbf{y}^S\) is a vector of 0's and 1's
 indicating whether a genet survives from \emph{t} to \emph{t+1}, or not.
@@
 
-where $x_{ij,t}$ is the log of genet size, $\gamma^{S}_{j,t}$ is a
-year-specific intercept, $\beta^{S}_{j,t}$ is the year-specific slope
-parameter for size, $\phi^{S}_{jQ}$ is the random effect of quadrat
-group location, and $\theta^{S}_{k}$ is the fixed parameter for the
-effect of the $k$th climate covariate at time $t$ ($C_{k,t}$). Note that
-the vector of climate covariates (\textbf{C}) includes climate variable
-interactions and climate$\times$size interactions. We include
+where \(x_{ij,t}\) is the log of genet size, \(\gamma^{S}_{j,t}\) is a
+year-specific intercept, \(\beta^{S}_{j,t}\) is the year-specific slope
+parameter for size, \(\phi^{S}_{jQ}\) is the random effect of quadrat
+group location, and \(\theta^{S}_{k}\) is the fixed parameter for the
+effect of the \(k\)th climate covariate at time \(t\) (\(C_{k,t}\)).
+Note that the vector of climate covariates (\textbf{C}) includes climate
+variable interactions and climate\(\times\)size interactions. We include
 density-dependence by estimating the effect of crowding on the focal
-individual by other individuals of the same species. $\omega$ is the
-effect of crowding and $w_{t,Q}$ is the crowding experienced by the
-focal individual at time $t$ in quadrat group $Q$. We include a
-size$\times$crowding interaction effect ($\nu^{S}$).
+individual by other individuals of the same species. \(\omega\) is the
+effect of crowding and \(w_{t,Q}\) is the crowding experienced by the
+focal individual at time \(t\) in quadrat group \(Q\). We include a
+size\(\times\)crowding interaction effect (\(\nu^{S}\)).
 
 We modeled growth as a Gaussian process describing genet size at time
-$t+1$ as a function of size at $t$ and climate covariates:
+\(t+1\) as a function of size at \(t\) and climate covariates:
 
@@
 
-where $x$ is log genet size and all other parameters are as described
+where \(x\) is log genet size and all other parameters are as described
 for the survival regression. We capture non-constant error variance in
 growth by modeling the variance around the growth regression
-($\varepsilon$) as a nonlinear function of predicted genet size:
+(\(\varepsilon\)) as a nonlinear function of predicted genet size:
 
@@
 specific parent to new genets. THerefore, we model recruitment at the
-quadrat level: the number of new individuals of species $j$ in quadrat
-$q$ recruiting at time $t+1$ as a function of quadrat ``effective
-cover'' ($A'$) in the previous year ($t$). Effective cover is a mixture
-of observed cover ($A$) in the focal quadrat ($q$) and the mean cover
-across the entire group ($\bar{A}$) of $Q$ quadrats in which $q$ is
-located:
+quadrat level: the number of new individuals of species \(j\) in quadrat
+\(q\) recruiting at time \(t+1\) as a function of quadrat ``effective
+cover'' (\(A'\)) in the previous year (\(t\)). Effective cover is a
+mixture of observed cover (\(A\)) in the focal quadrat (\(q\)) and the
+mean cover across the entire group (\(\bar{A}\)) of \(Q\) quadrats in
+which \(q\) is located:
 
@@
 
-where $p$ is a mixing fraction between 0 and 1 that is estimated within
-the model.
-
-We assume the number of individuals, $y^{R}$, recruiting at time $t+1$
-follows a negative binomial distribution:
+where \(p\) is a mixing fraction between 0 and 1 that is estimated
+within the model.
+
+We assume the number of individuals, \(y^{R}\), recruiting at time
+\(t+1\) follows a negative binomial distribution:
 
@@
 However, within-quadrat extirpation and colonization events were rare in
-our time series ($N=9$ and $N=10$, respectively, across all species).
-Given the broad spatial distribution of the quadrats we are studying, it
-is safe to assume that these events are in fact rare enough to be
-ignored for our purposes. So we constrained our statistical modeling of
-vital rates at the population level to change in percent cover within
-quadrats. For the remaining discussion of statistical modeling, we refer
-to proportion data, which is simply percent cover divided by 100.
+our time series (\(N=9\) and \(N=10\), respectively, across all
+species). Given the broad spatial distribution of the quadrats we are
+studying, it is safe to assume that these events are in fact rare enough
+to be ignored for our purposes. So we constrained our statistical
+modeling of vital rates at the population level to change in percent
+cover within quadrats. For the remaining discussion of statistical
+modeling, we refer to proportion data, which is simply percent cover
+divided by 100.
 
@@
 truncated log-normal likelihood. The model for quadrat cover change from
-time $t$ to $t+1$ is
+time \(t\) to \(t+1\) is
 
@@
 
-where $x_{jq,t}$ is the log of species' $j$ proportional cover in
-quadrat $q$ at time $t$ and all other parameters are as in the
+where \(x_{jq,t}\) is the log of species' \(j\) proportional cover in
+quadrat \(q\) at time \(t\) and all other parameters are as in the
 individual-level growth model (Eq. 3). Again, note that the climate
-covariate vector (\textbf{C}) includes the climate$\times$cover
+covariate vector (\textbf{C}) includes the climate\(\times\)cover
 interaction. The log normal likelihood includes a truncation
@@
 Our IPM follows the specification of Chu and Adler (2015) where the
-population of species \emph{j} is a density function $n(u_{j},t)$ giving
-the density of sized-\emph{u} genets at time \emph{t}. Genet size is on
-the natural log scale, so that $n(u_{j},t)du$ is the number of genets
-whose area (on the arithmetic scale) is between $e^{u_{j}}$ and
-$e^{u_{j}+du}$. The density function for any size \emph{v} at time $t+1$
-is
+population of species \emph{j} is a density function \(n(u_{j},t)\)
+giving the density of sized-\emph{u} genets at time \emph{t}. Genet size
+is on the natural log scale, so that \(n(u_{j},t)du\) is the number of
+genets whose area (on the arithmetic scale) is between \(e^{u_{j}}\) and
+\(e^{u_{j}+du}\). The density function for any size \emph{v} at time
+\(t+1\) is
 
@@
 
-where $k_{j}(v_{j},u_{j},\bar{w_{j}})$ is the population kernel that
-describes all possible transitions from size $u$ to $v$ and
-$\bar{w_{j}}$ is a scalar representing the average intraspecific
-crowding experienced by a genet of size $u_j$ and species $j$. The
+where \(k_{j}(v_{j},u_{j},\bar{w_{j}})\) is the population kernel that
+describes all possible transitions from size \(u\) to \(v\) and
+\(\bar{w_{j}}\) is a scalar representing the average intraspecific
+crowding experienced by a genet of size \(u_j\) and species \(j\). The
 integral is evaluated over all possible sizes between predefined lower
@@
 Since the IPM is spatially-implicit, we cannot calculate neighborhood
-crowding for specific genets ($w_{ij}$). Instead, we use an
-approximation ($\bar{w_{j}}$) that captures the essential features of
+crowding for specific genets (\(w_{ij}\)). Instead, we use an
+approximation (\(\bar{w_{j}}\)) that captures the essential features of
 neighborhood interactions (Adler et al. 2010). This approximation relies
@@
 assume that fecundity increases linearly with size
-($R_j(v_{j},u_{j},\bar{w_{j}}) = e^{u_j}R_j(v_{j},\bar{w_{j}})$) to
+(\(R_j(v_{j},u_{j},\bar{w_{j}}) = e^{u_j}R_j(v_{j},\bar{w_{j}})\)) to
 incorporate the recruitment function in the spatially-implicit IPM.
@@
 appropriate parameter set, we calculate the mean response (population
-cover at \emph{t+1} = $x_{t+1}$) according to Eq. 9. We then make a
+cover at \emph{t+1} = \(x_{t+1}\)) according to Eq. 9. We then make a
 random draw from a {[}0,1{]} truncated lognormal distribution with mean
-equal to $x_{t+1}$ from Eq. 9 and the variance estimate from the fitted
-model. We can then project the model forward by drawing a new parameter
-set (unique to climate year and MCMC iteration) at each timestep. As
-with the IPM, random year effects are ignored for all simulations.
+equal to \(x_{t+1}\) from Eq. 9 and the variance estimate from the
+fitted model. We can then project the model forward by drawing a new
+parameter set (unique to climate year and MCMC iteration) at each
+timestep. As with the IPM, random year effects are ignored for all
+simulations.
 
@@
 QBM climate coefficients against each vital rate model's climate
-coefficients and calculated Pearson's $\rho$. Strong correlations
+coefficients and calculated Pearson's \(\rho\). Strong correlations
 indicate the QBM is capable of detecting climate effects associated with
@@
 the QBM for three species (Table 1). The IPM MAE is significantly lower
-at $\alpha=0.05$ for \emph{B. gracilis} (\emph{P} = 0.0012), \emph{H.
+at \(\alpha=0.05\) for \emph{B. gracilis} (\emph{P} = 0.0012), \emph{H.
 comata} (\emph{P} = 4.0586 × 10-8), and \emph{P. smithii} (\emph{P} =
@@
 
+\setcounter{page}{29}
+
 \subsection{Tables}\label{tables}
@@
 
+\subsection{Figure Legends}\label{figure-legends}
+
+Figure 1. Time series of average percent cover over all quadrats for our
+four focal species: \emph{Bouteloua gracilis} (BOGR), \emph{Hesperostipa
+comata} (HECO), \emph{Pascopyrum smithii} (PASM), and \emph{Poa secunda}
+(POSE). Light grey lines show trajectories of individual quadrats. Note
+the different y-axis scales across panels.
+
+Figure 2. Work flow of the data aggregation, model fitting, and
+population simulating.
+
+Figure 3. Sensitivity of equilibrium cover simulated from the IPM to
+each climate scenario applied to individual and combined vital rates.
+For example, the points associated with G show the median cover from IPM
+simulations where a climate perturbation is applied only to the growth
+regression climate covariates. These simulations use mean parameter
+values for clarity.
+
+Figure 4. Effect of quadrat sample size on the precision (standard
+deviation) of main climate effect estimates in the QBM. Increasing the
+number of quadrats results in diminishing returns in terms of parameter
+certainty. Light dashed lines show individual climate effects at five
+quadrat sample sizes. Thick dark lines are inverse gaussian fits showing
+the mean effect of increasing quadrat sample size on parameter
+precision.
+
+Figure 5. Correlations (r) between QBM and IPM estimates of climate
+effects. We ignore sizeXclimate interactions since these are not
+directly comparable across model types. The QBM does not have multiple
+vital rates, so its values are repeated across panels within each
+species. Across top panels, `growth' = growth regression, `rec' =
+recruitment regression, `surv' = survival regression.
+
+Figure 6. Mean (points) and 90\% quantiles (errorbars) for the
+proportional difference between baseline simulations (using observed
+climate) and the climate pertubation simulation on the x-axis. We
+calculated proportional difference as log(perturbed climate cover) -
+log(observed climate cover), where `perturbed' and `observed' refer to
+the climate time series used to drive interannual variation in the
+simulations. Model error and parameter uncertainty were propagated
+through the simulation phase. Climate simulations are as in Figure 3.
+
+\pagebreak{}
+
 \subsection{Figures}\label{figures}
@@
 
-\subsection{References}\label{references}
+\setcounter{page}{24}
+
+\subsection*{References}\label{references}
+\addcontentsline{toc}{subsection}{References}
 
@@
 Hobbs, N. T., and M. B. Hooten. 2015. Bayesian Models: A Statistical
-Primer for Ecologists. Princeton University PressPrinceton.
+Primer for Ecologists. Princeton University Press, Princeton.
 
@@
 
-Schurr, F. M., J. Pagel, J. S. Cabral, J. Groeneveld, O. Bykova,
-O'HaraR. B., F. Hartig, W. D. Kissling, H. P. Linder, G. F. Midgley, B.
+Schurr, F. M., J. Pagel, J. S. Cabral, J. Groeneveld, O. Bykova, R. B.
+O'Hara, F. Hartig, W. D. Kissling, H. P. Linder, G. F. Midgley, B.
 Schröder, A. Singer, and N. E. Zimmermann. 2012. How to understand

### 2015-09-04 README.Rmd
updates README

--- 2015-08-28 manuscript/manuscript.Rmd
@@
 ---
-layout: 12pt
-header-includes:
-   - \usepackage{lineno}
-   - \linenumbers
-   - \usepackage{setspace}
-   - \doublespacing
-   - \usepackage{todonotes}
-bibliography: ~/Dropbox/Bibliography/MicroMeso.bib
-csl: components/ecology.csl
+title: "Read Me"
+date: "April 20, 2015"
+output: html_document
+---
+This is a 'read me' file for data and code associated with "Do we need demographic data to forecast population responses to climate change?" by Andrew T.
+Tredennick and Peter B.
+Adler (in review).
+The bioRxiv preprint can be found [here](http://biorxiv.org/content/early/2015/08/30/025742).
 
-## rmarkdown render options
-output:
-  pdf_document:
-    fig_caption: true
-    keep_tex: true
-fontsize: 12pt
-geometry: margin=1in
-
----
-
-\begin{centering}
-\textbf{\large{Do we need demographic data to forecast population responses to climate change?}}
-
-\textsc{\small{Andrew T. Tredennick\footnote{E-mail: atredenn@gmail.com} and Peter B. Adler}}
-
-\textit{\small{Department of Wildland Resources and the Ecology Center, 5230 Old Main Hill, Utah State University, Logan, Utah 84322-5230 USA}}
-
-\end{centering}
-
-```{r libraris, include=FALSE}
-library(ggplot2)
-library(plyr)
-library(gridExtra)
-library(reshape2)
-library(xtable)
-library(ggmcmc)
-```
-
-```{r caching, include=FALSE}
-library("methods")
-library("knitr")
-basename <- "manuscript"
-opts_chunk$set(fig.path = paste("components/figure/", basename, "-", sep=""),
-               cache.path = paste("components/cache/", basename, "/", sep=""))
-opts_chunk$set(cache = 2)
-opts_chunk$set(tidy=FALSE, warning=FALSE, message=FALSE, 
-               comment = NA, verbose = TRUE, echo=FALSE)
-
-# PDF-based figures
-opts_chunk$set(dev='pdf')
-```
-
-```{r plot-options, message=FALSE, warning=FALSE, include=FALSE, echo=FALSE}
-cbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", 
-               "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
-
-opts_chunk$set(dev='pdf', dev.args=c(version="1.3"))
-
-```
-
-Abstract
---------
-Rapid climate change has generated growing interest in forecasts of future population trajectories.
+### Paper abstract
+>Rapid climate change has generated growing interest in forecasts of future population trajectories.
 Traditional population models, typically built using detailed demographic observations from one study site, can address climate change impacts at one location, but are difficult to scale up to the landscape and regional scales relevant to management decisions.
@@
 
-*Key words: forecasting, climate change, grassland, integral projection model, population model*
+Funding for this work was provided by the National Science Foundation through a Postdoctoral Research Fellowship in Biology to Andrew ([DBI-1400370](http://www.nsf.gov/awardsearch/showAward?AWD_ID=1400370&HistoricalAwards=false)) and a CAREER award to Peter ([DEB-1054040](http://www.nsf.gov/awardsearch/showAward?AWD_ID=1054040&HistoricalAwards=false)).
 
-Introduction
-------------
+_Send questions to_: Andrew Tredennick (atredenn@gmail.com)
 
-Perhaps the greatest challenge for ecology in the 21st century is to forecast the impacts of environmental change [@Clark2001; @Petchey2015].
-Forecasts require sophisticated modeling approaches that fully account for uncertainty and variability in both ecological process and model parameters [@Luo2011a; but see @Perretti2013a for an argument against modeling the ecological process].
-The increasing statistical sophistication of population models [@Rees2009] makes them promising tools for predicting the impacts of environmental change on species persistence and abundance.
-But reconciling the scales at which population models are parameterized and the scales at which environmental changes play out remains a challenge [@Clark2010; @Freckleton2011; @Queenborough2011; @Clark2012].
-The problem is that most population models are built using data from a single study site because collecting those data, which involves tracking the fates of individuals plants, is so difficult.
-The resulting models cannot be applied to the landscape and regional scales relevant to decision-making without information about how the fitted parameters respond to spatial variation in biotic and abiotic drivers [@Sæther2007].
-The limited spatial extent of individual-level demographic datasets constrains our ability to use population models to address applied questions about the consequences of climate change.
+General Information
+-------------------
+This directory contains all of the data and R code necessary to reproduce the analysis and figures from Tredennick and Adler 201x.
+The ``analysis/`` directory holds all code and data, while the ``\manuscript`` directory contains a R Markdown file (``manuscript.Rmd``) with paper text and the code to produce figures embedded.
 
-The inability of most population models to address landscape-scale problems may explain why land managers and conservation planners have embraced species distribution models (SDMs) [see @Guisan2005 for a review].
-SDMs typically rely on easy-to-collect presence/absence data [but see @Clark2014 for new methods] and remotely-sensed environmental covariates that allow researchers to model large spatial extents [e.g., @Maiorano2013].
-Thus, it is relatively straightforward to parameterize and project SDMs over landscapes and regions.
-However, the limitations of SDMs are well known [@Pearson2003; @Elith2009; @Araujo2012].
-Ideally, researchers would provide managers with landscape-scale population models, combining the extent of SDMs with information about dynamics and species abundances [@Schurr2012; @Merow2014].
+File Structure
+--------------
+Here is the file structure assuming the root directory is ``ssde_harvest_dryad/``:
 
-Aggregate measures of population status, rather than individual performance, offer an intriguing alternative for modeling populations [@Clark2004; @Freckleton2011].
-Population-level data cannot provide inference about demographic mechanisms, but might be sufficient for modeling future population states, especially since such data are feasible to collect across broad spatial extents [e.g., @Queenborough2011].
-The choice between individual and population-level data involves a difficult trade-off: while individual-level data leads to more mechanistic models, population-level data leads to models that can be applied over greater spatial and temporal extents.
-An open question is how much forecasting skill is lost when we build models based on population rather than individual-level data.
+### analysis/
+* ``ipm/`` -- contains R code for all analyses and figure production. 
+    + ``sumlations/`` -- 
+    + ``vitalRateRegs/`` -- 
+* ``alpha_list_growth.csv`` -- CSV file with the alpha values used to define the spatial influence of genets on one another for the growth regressions (see [Chu and Adler 2015](http://www.esajournals.org/doi/abs/10.1890/14-1741.1))
+* ``alpha_list_survival.csv`` -- CSV file with the alpha values used to define the spatial influence of genets on one another for the survival regressions (see [Chu and Adler 2015](http://www.esajournals.org/doi/abs/10.1890/14-1741.1))
+* ``get_chainiter_sequence.R`` -- R script to make a random sequence chain and iteration draws so that all simulations use the same sequence of parameter values.
+* ``get_year_sequence.R`` -- R script to make a random sequence year draws so that all simulations use the same sequence of climate years.
 
-To date, most empirical population modelers have relied on individual-level data, with few attempts to capitalize on population-level measures.
-An important exception was an effort by @Taylor2004 to model the population growth rate of an invasive species to identify the best strategies for invasion control.
-They used a "density-structured" model where the state variable is a discrete density state rather than a continuous density measure.
-Such models do not require individual-level demographic data and can adequately describe population dynamics.
-Building on @Taylor2004, @Freckleton2011 showed that density-structured models compare well to continuous models in theory, and @Queenborough2011 provide empirical evidence that density-structured models are capable of reproducing population dynamics at landscape spatial scales, even if some precision is lost when compared to fully continuous models.
-The appeal of density-structured approaches is clear.
-However, none of these models included environmental covariates.
-
-Addressing climate change questions with models fit to population-level data is potentially problematic.
-It is individuals, not populations, that respond to climate variables [@Clark2012].
-Ignoring this fact amounts to an "ecological fallacy", where inference about the individual relies on statistical inference on the group [@PIANTADOSI1988].
-Population growth (or decline) is the outcome of demographic processes such as survival, growth, and recruitment that occur at the level of individual plants.
-Climate can affect each demographic process in unique, potentially opossing, ways [@Dalgleish2011].
-These unique climate responses may be difficult to resolve in statistical models based on population-level data where demographic processes are not identifiable.
-If important climate effects are missed because of the aggregation inherent in in population-level data, then population models built with such data will make uninformative or unreliable forecasts.
-
-Here, we compare the forecasting skill of statistical and population models based on aggregated, population-level data with models based on individual-level data.
-We used a unique demographic dataset that tracks the fates of individual plants from four species over 14 years to build two kinds of single-species population models, traditional models using individual growth, survival, and recruitment data and alternative models based on basal cover.
-In both models, interannual variation is explained, in part, by climate covariates.
-We first quantify forecasting skill using cross-validation.
-We then performed simulations to quantify the sensitivities of species' cover to small perturbations in average precipitation and temperature.
-Based on the cross-validation results, predictions of individual level models were clearly better, but, unfortunately, still too uncertain to inform management decisions.
-
-Materials and Methods
----------------------
-### Study site and data
-Our demographic data come from the Fort Keogh Livestock and Range Research Laboratory in eastern Montana's northern mixed prairie near Miles City, Montana, USA ($46^{\circ}$ 19' N, $105^{\circ}$ 48' W).
-The dataset is freely available on Ecological Archives[^link] [@Anderson2011] , and interested readers should refer to the metadata for a complete description.
-The site is about 800 m above sea level and mean annual precipitation (1878-2009) is 334 mm, with most annual precipitation falling from April through September.
-The community is grass-dominated and we focused on the four most abundant grass species: \emph{Bouteloua gracilis} (BOGR), \emph{Hesperostipa comata} (HECO), \emph{Pascopyrum smithii} (PASM), and \emph{Poa secunda} (POSE) (Fig.
-1).
-
-[^link]: http://esapubs.org/archive/ecol/E092/143/
-
-From 1932 to 1945 individual plants were identified and mapped annually in 44 1-$\text{m}^2$ quadrats using a pantograph.
-The quadrats were distributed in six pastures, each assigned a grazing treatment of light (1.24 ha/animal unit month), moderate (0.92 ha/aum), and heavy (0.76 ha/aum) stocking rates (two pastures per treatment).
-In this analysis we account for potential differences among the grazing treatments, but do not focus on grazing$\times$climate interactions.
-The annual maps of the quadrats were digitized and the fates of individual plants tracked and extracted using a computer program [@Lauenroth2008; @Chu2014a].
-Daily climate data are available for the duration of the data collection period (1932 - 1945) from the Miles City airport, Wiley Field, 9 km from the study site.
-
-We modeled each grass population based on two levels of data: individual and quadrat (Fig.
-2).
-The individual data is the "raw" data.
-For the quadrat-level we data we simply sum individual basal cover for each quadrat by species.
-This is equivalent to a near-perfect census of quadrat percent cover because previous analysis shows that measurement error at the individual-level is small [@Chu2014].
-Based on these two datasets we can compare population models built using individual-level data and aggregated, quadrat-level data.
-
-All R code and data necessary to reproduce our analysis is archived on GitHub as release v1.0[^revnote] (http://github.com/atredennick/MicroMesoForecast/releases).
-That stable release will remain static as a record of this analysis, but subsequent versions may appear if we update this work.
-We have also deposited the v1.0 release on Dryad (*link here after acceptance*).
-
-[^revnote]: *Note to reviewers*: so that v1.0 will be associated with the published version of the manuscript, we have released v0.1 to be associated with this review version.
-
-### Stastical models of vital rates
-At both levels of inference (individual and quadrat), the building blocks of our population models are vital rate regressions.
-For individual-level data, we fit regressions for survival, growth, and recruitment for each species.
-At the quadrat-level, we fit a single regression model for population growth.
-We describe the statistical models separately since fitting the models required different approaches.
-All models contain five climate covariates that we chose \emph{a priori}: "water year" precipitation at \emph{t}-1 (lagppt); April through June precipitation at \emph{t}-1 and \emph{t}-2 (ppt1 and ppt2, respectively) and April through June temperature at \emph{t}-1 and \emph{t}-2 (TmeanSpr1 and TmeanSpr2, respectively), where \emph{t} is the observation year.
-We also include interactions among same-year climate covariates (e.g., ppt1 $\times$ TmeansSpr1) and climate $\times$ size interactions.
-Climate $\times$ size interactions are for climate main effects only; we do not include interactions between size and pairs of interacting climate effects.
-
-We fit all models using a hierarchical Bayesian approach.
-The models are fully descibed in Appendix A, so here we focus on the main process and the model likelihood.
-For the likelihood models, $\textbf{y}^X$ is always the relevant vector of observations for vital rate *X* ($X = S, G, R, or P$ for survival, growth, recruitment, or population growth).
-For example, $\textbf{y}^S$ is a vector of 0's and 1's indicating whether a genet survives from *t* to *t+1*, or not.
-
-#### Vital rate models at the individual level
-We used logistic regression to model survival probability ($S$) of genet $i$ from species $j$ in quadrat group $Q$ from time $t$ to $t+1$:
-
-\begin{align}
-\text{logit}(S_{ijQ,t}) &= \gamma^{S}_{j,t} + \phi^{S}_{jQ} + \beta^{S}_{j,t}x_{ij,t} + \omega^{S}_{j}w_{ij,t} + \nu^{S}_{j}w_{ij,t}x_{ij,t} + \theta^{S}_{jk}C_{k,t} \\
-y^{S}_{ijQ,t} &\sim \text{Bernoulli}(S_{ijQ,t})
-\end{align}
-
-where $x_{ij,t}$ is the log of genet size, $\gamma^{S}_{j,t}$ is a year-specific intercept, $\beta^{S}_{j,t}$ is the year-specific slope parameter for size, $\phi^{S}_{jQ}$ is the random effect of quadrat group location, and $\theta^{S}_{k}$ is the fixed parameter for the effect of the $k$th climate covariate at time $t$ ($C_{k,t}$).
-Note that the vector of climate covariates (**C**) includes climate variable interactions and climate$\times$size interactions.
-We include density-dependence by estimating the effect of crowding on the focal individual by other individuals of the same species.
-$\omega$ is the effect of crowding and $w_{t,Q}$ is the crowding experienced by the focal individual at time $t$ in quadrat group $Q$.
-We include a size$\times$crowding interaction effect ($\nu^{S}$).
-
-We modeled growth as a Gaussian process describing genet size at time $t+1$ as a function of size at $t$ and climate covariates:
-
-\begin{align}
-x_{ijQ,t+1} &= \gamma^{G}_{j,t} + \phi^{G}_{jQ} + \beta^{G}_{j,t}x_{ij,t} + \omega^{G}_{j}w_{ij,t} + \nu^{S}_{j}w_{ij,t}x_{ij,t} + \theta^{G}_{jk}C_{k,t} \\
-y^{G}_{ijQ,t} &\sim \text{Normal}(x_{ijQ,t+1}, \varepsilon_{ij,t})
-\end{align}
-
-where $x$ is log genet size and all other parameters are as described for the survival regression.
-We capture non-constant error variance in growth by modeling the variance around the growth regression ($\varepsilon$) as a nonlinear function of predicted genet size:
-
-\begin{align}
-\varepsilon_{ij,t} = a e^{b x_{ijQ,t+1}}
-\end{align}
-
-Our data allows us to track new recruits, but we cannot assign a specific parent to new genets.
-THerefore, we model recruitment at the quadrat level: the number of new individuals of species $j$ in quadrat $q$ recruiting at time $t+1$ as a function of quadrat "effective cover" ($A'$) in the previous year ($t$).
-Effective cover is a mixture of observed cover ($A$) in the focal quadrat ($q$) and the mean cover across the entire group ($\bar{A}$) of $Q$ quadrats in which $q$ is located:
-
-\begin{equation}
-A'_{jq,t} = p_{j}A_{jq,t} + (1-p_{j})\bar{A}_{jQ,t}
-\end{equation}
-
-where $p$ is a mixing fraction between 0 and 1 that is estimated within the model.
-
-We assume the number of individuals, $y^{R}$, recruiting at time $t+1$ follows a negative binomial distribution:
-
-\begin{equation}
-y^{R}_{jq,t+1} \sim \text{NegBin}(\lambda_{jq,t+1},\zeta)
-\end{equation}
-
-where $\lambda$ is the mean intensity and $\zeta$ is the size parameter. We define $\lambda$ as:
-
-\begin{equation}
-\lambda_{jq,t+1} = A'_{jq,t}e^{(\gamma^{R}_{j,t} + \phi^{R}_{jQ} + \theta^{R}_{jk}C_{k,t} + \omega^{R}\sqrt{A'_{q,t}})}
-\end{equation}
-
-where $A'$ is effective cover ($\text{cm}^2$) of species $j$ in quadrat $q$ and all other terms are as in the survival and growth regressions.
-
-#### Population model at the quadrat level
-The statistical approach used to model aggregated data depends on the type of data collected.
-We have percent cover data, which can easily be transformed to proportion data.
-We first considered fitting three vital rate models analagous to those we fit at the individual level: one for probability of extirpation within a quadrat (analagous to survival), one for cover change within a quadrat (analagous to growth), and one for probability of colonization within a quadrat (analagous to recruitment).
-However, within-quadrat extirpation and colonization events were rare in our time series ($N=9$ and $N=10$, respectively, across all species).
-Given the broad spatial distribution of the quadrats we are studying, it is safe to assume that these events are in fact rare enough to be ignored for our purposes.
-So we constrained our statistical modeling of vital rates at the population level to change in percent cover within quadrats.
-For the remaining discussion of statistical modeling, we refer to proportion data, which is simply percent cover divided by 100.
-
-An obvious choice for fitting a linear model to proportion data is beta regression because the support of the beta distribution is [0,1], not including true zeros or ones.
-However, when we used fitted model parameters from a beta regression in a quadrat-based population model, the simulated population tended toward 100% cover for all species.
-We therefore chose a more constrained modeling approach based on a truncated log-normal likelihood.
-The model for quadrat cover change from time $t$ to $t+1$ is
-
-\begin{align}
-x_{jq,t+1} &= \gamma^{P}_{j,t} + \phi^{P}_{jQ} + \beta^{P}_{j,t}x_{jq,t} + \theta^{P}_{jk}C_{k,t} \\
-y^{P}_{jq,t+1} &\sim \text{LogNormal}(x_{jq,t+1}, \tau_{j}) \text{T}[0,1]
-\end{align}
-
-where $x_{jq,t}$ is the log of species' $j$ proportional cover in quadrat $q$ at time $t$ and all other parameters are as in the individual-level growth model (Eq.
-3).
-Again, note that the climate covariate vector (**C**) includes the climate$\times$cover interaction.
-The log normal likelihood includes a truncation (T[0,1]) to ensure that predicted values do not exceed 100% cover.
-
-### Model fitting
-Our Bayesian approach to fitting the vital rate models required choosing appropriate priors for unknown parameters and deciding which, if any, of those priors should be hierarchical.
-We decided to fit models where all terms were fit by species.
-Within a species, we fit yearly size effects and yearly intercepts hierarchically where year-specific coefficients were drawn from global distributions representing the mean size effect and intercept.
-Quadrat random effects were also fit hierarchically, with quadrat offsets being drawn from distributions with mean zero and a shared variance term (independent Gaussian priors, Appendix A).
-Climate effects were not modeled hierarchically, and each was given a diffuse prior distribution.
-We used standard diffuse priors for all unknown parameters (Appendix A).
-
-All of our analyses (model fitting and simulating) were conducted in R [@R2013].
-We used the 'No-U-Turn' MCMC sampler in Stan [@stan2014]  to estimate the posterior distributions of model parameters using the package 'rstan' [@rstan2014].
-We obtained posterior distributions for all model parameters from three parallel MCMC chains run for 1,000 iterations after discarding an initial 1,000 iterations.
-Such short MCMC chains may surprise readers more familiar with other MCMC samplers (i.e.
-JAGS or WinBUGS), but the Stan sampler is exceptionally efficient, which reduces the number of iterations needed to achieve convergence.
-We assessed convergence visually and made sure scale reduction factors for all parameters were less than 1.01.
-For the purposes of including parameter uncertainty in our population models, we saved the final 1,000 iterations from each of the three MCMC chains to be used as randomly drawn values during population simulation.
-This step alleviates the need to reduce model parameters by model selection since sampling from the full parameter space in the MCMC ensures that if a parameter broadly overlaps zero, on average the effect in the population models will also be near zero.
-We report the posterior mean, standard deviation, and 95% Bayesian Credible Intervals for every parameter of each model for each species in Appendix B.
-
-### Population models
-With the posterior distribution of the vital rate statistical models in hand, it is straightforward to simulate the population models.
-We used an Integral Projection Model (IPM) to model populations based on individual-level data [@Ellner2006] and a quadrat-based version of an individually-based model (Quadrat-Based Model, QBM) to model populations based on quadrat-level data.
-We describe each in turn.
-
-#### Integral projection model
-We use a stochastic IPM [@Rees2009] that includes the climate covariates from the vital rate statistical models.
-In all simulations we ignore the random year effects so that interannual variation is driven solely by climate.
-We fit the random year effects in the vital rate regressions to avoid over-attributing variation to climate covariates.
-Our IPM follows the specification of @Chu2015 where the population of species _j_ is a density function $n(u_{j},t)$ giving the density of sized-_u_ genets at time _t_.
-Genet size is on the natural log scale, so that $n(u_{j},t)du$ is the number of genets whose area (on the arithmetic scale) is between $e^{u_{j}}$ and $e^{u_{j}+du}$.
-The density function for any size _v_ at time $t+1$ is
-
-\begin{equation}
-n(v_{j},t+1) = \int_{L_{j}}^{U_{j}} k_{j}(v_{j},u_{j},\bar{w_{j}}(u_{j}))n(u_{j},t)
-\end{equation}
-
-where $k_{j}(v_{j},u_{j},\bar{w_{j}})$ is the population kernel that describes all possible transitions from size $u$ to $v$ and $\bar{w_{j}}$ is a scalar representing the average intraspecific crowding experienced by a genet of size $u_j$ and species $j$.
-The integral is evaluated over all possible sizes between predefined lower (_L_) and upper (_U_) size limits that extend beyond the range of observed genet sizes.
-
-Since the IPM is spatially-implicit, we cannot calculate neighborhood crowding for specific genets ($w_{ij}$).
-Instead, we use an approximation ($\bar{w_{j}}$) that captures the essential features of neighborhood interactions [@Adler2010].
-This approximation relies on a 'no-overlap' rule for conspecific genets to approximate the overdispersion of large genets in space [@Adler2010].
-
-The population kernel is defined as the joint contributions of survival (_S_), growth (_G_), and recruitment (_R_):
-
-\begin{equation}
-k_{j}(v_{j},u_{j},\bar{w_{j}}) = S_j(u_j, \bar{w_{j}}(u_{j}))G_j(v_{j},u_{j},\bar{w_{j}}(u_{j})) + R_j(v_{j},u_{j},\bar{w_{j}}),
-\end{equation}
-
-which means we are calculating growth (_G_) for individuals that survive (_S_) from time _t_ to _t+1_ and adding in newly recruited (_R_) individuals of an average sized one-year-old genet for the focal species.
-Our stastical model for recruitment (_R_, described above) returns the number of new recruit produced per quadrat.
-Following previous work [@Adler2012; @Chu2015], we assume that fecundity increases linearly with size ($R_j(v_{j},u_{j},\bar{w_{j}}) = e^{u_j}R_j(v_{j},\bar{w_{j}})$) to incorporate the recruitment function in the spatially-implicit IPM.
-
-We used random draws from the final 1,000 iterations from each of three MCMC chains to introduce stochasticity into our population models.
-At each time step, we randomly selected climate covariates from one of the 14 observed years.
-Then, we drew the full parameter set (climate effects and density-dependence fixed effects) from a randomly selected MCMC iteration.
-Using this approach, rather than simply using coefficient point estimates, captures the effect of parameter uncertainty.
-Relatively unimportant climate covariates (those that broadly overlap 0) will have little effect on the mean of the simulation results, but can contribute to their variation.
-Since our focus was on the contribution of climate covariates to population states, we set the random year effects and the random group effects to zero.
-
-#### Quad-based model
-To simulate our quad-based model (QBM), we simply iterate the quadrat-level statistical model (Eqs.
-9-10).
-We use the same approach for drawing parameter values as described for the IPM.
-After drawing the appropriate parameter set, we calculate the mean response (population cover at *t+1* = $x_{t+1}$) according to Eq.
-9.
-We then make a random draw from a [0,1] truncated lognormal distribution with mean equal to $x_{t+1}$ from Eq.
-9 and the variance estimate from the fitted model.
-We can then project the model forward by drawing a new parameter set (unique to climate year and MCMC iteration) at each timestep.
-As with the IPM, random year effects are ignored for all simulations.
-
-### Model validation
-To test each model's ability to forecast population state, we made out-of-sample predictions using leave-one-year-out cross validation.
-For both levels of modeling, we fit the vital rate models using observations from all years except one, and then used those fitted parameters in the population models to perform a one-step-ahead forecast for the year whose observations were withheld from model fitting.
-Within each observation year, several quadrats were sampled.
-We made predictions for each observed quadrat in the focal year, initializing each simulation with cover in the quadrat the previous year.
-Since we were making quadrat-specific predictions, we incorporated the group random effect on the intercept for both models.
-We repeated this procedure for all 13 observation years, making 100 one-step-ahead forecasts for each quadrat-year combination with parameter uncertainty included via random draw from the MCMC chain as described above.
-Random year effects were set to zero since year effects cannot be assigned to unobserved years.
-
-This cross-validation procedure allowed us to compare accuracy and precision of the two modeling approaches (IPM versus QBM).
-We first calculated the median predicted cover across the 100 simulations for each quadrat-year and then calculated the absolute error as the absolute value of the difference between the observed cover for a given quadrat-year and the median prediction.
-To arrive at mean absolute error (MAE), we then averaged the absolute error within each species across the quadrat-year specific errors.
-We use MAE as our measure of accuracy.
-To measure precision we calculated the distance between the upper and lower 90th quantiles of the 100 predictions and averaged this value over quadrat-years for each species.
-
-### Testing sensitivity to climate covariates
-With our fitted and validated models in hand, we ran simulations for each model type (IPM and QBM) under four climate perturbation scenarios: (1) observed climate, (2) precipitation increased by 1%, (3) temperature increased by 1%, and (4) precipitation and temperature increased by 1%.
-We ran the simulations for 2,500 time steps, enough to estimate equilibrium cover after discarding an initial 500 time steps as burn-in.
-Each simulation was run under two parameter scenarios: (1) using mean parameter estimates and (2) using randomly drawn parameters from the MCMC chain.
-We use (1) to detect the overall sensitivity of equilibrium cover to climate, and we use (2) to show the impact of model and parameter uncertainty on forecast precision.
-
-As an effort to identify potential discrepencies between IPM and QBM forecasts, we also ran simulations designed to quantify the sensitivities of individual and combined vital rates to climate for the IPM.
-Specifically, we ran simulations for the above climate scenarios, but applied the perturbed climate covariates to survival, growth, or recruitment vital rates individually and in pairwise combinations.
-This allowed us to isolate the vital rate(s) most sensitive to climate.
-For this analysis, we used mean parameter estimates to reduce the sources of uncertainty in the sensitivity estimates.
-
-We expected the IPM to produce more accurate and precise forecasts due to either (1) the smaller sample size of the quadrat level data sets compared to the individual level data sets, leading to larger parameter uncertainty for the QBM, or (2) the QBM climate effects being weakly associated with one or more vital rate climate effects at the individual level.
-To assess the impact of sample size on QBM parameter uncertainty we refit the QBM statistical model (Eqs.
-9-10) after removing sets of 2, 5, 10, and 15 quadrats.
-We fit 10 models at each level of quadrat removal (2, 5, 10, 15 quadrats), removing a different randomly selected set of quadrats for each fit.
-We calculated the standard deviation of climate main effects (pptLag, ppt1, ppt2, TmeanSpr1, and TmeanSpr2) for each model and averaged those over replicates within each set of quadrat removals.
-This allowed us to regress parameter uncertainty against sample size.
-
-To deterime if the QBM climate effects are correlated with climate effects for each vital rate model in the IPM, we simply regressed the QBM climate coefficients against each vital rate model's climate coefficients and calculated Pearson's $\rho$.
-Strong correlations indicate the QBM is capable of detecting climate effects associated with individual vital rates.
-A weak correlation indicates the QBM "misses" the climate effect on a particular vital rate.
-
-Results
--------
-```{r pvalues, echo=FALSE, eval=TRUE}
-##  IPM results
-# Read in data
-setwd("/Users/atredenn/Repos/MicroMesoForecast/manuscript/")
-ipm_onestep <- readRDS("../analysis/ipm/simulations/results/one_step_validation/ipm_loyo_forecasts_combined.rds")
-# Remove predictions >1 for fair comparison to [0,1] truncated quad model
-resets <- which(ipm_onestep[,"cover.t1"]>1)
-ipm_onestep[resets, "cover.t1"] <- 1
-# Average predictions over quad-year reps
-avg_ipm <- ddply(ipm_onestep, .(species, quad, t1), summarise,
-                 avg_prediction = median(cover.t1),
-                 observation = mean(cover.t0),
-                 quant_dist = quantile(cover.t1*100, probs = 0.95) - quantile(cover.t1, probs = 0.05))
-# Calculate error
-avg_ipm$error <- avg_ipm$observation*100 - avg_ipm$avg_prediction*100
-# Aggregate over species
-ipmstats <- ddply(avg_ipm, .(species), summarise,
-                  mae = mean(abs(error)),
-                  mean_cover = mean(observation*100),
-                  quant_dist = mean(quant_dist))
-ipmstats$model <- "IPM"
-
-##  QBM results
-# Read in data
-qbm_onestep <- readRDS("../analysis/quadBM/simulations/results/one_step_validation/qbm_one-step_forecasts_combined.rds")
-# Average predictions over quad-year reps
-avg_qbm <- ddply(qbm_onestep, .(species, quad, year), summarise,
-                 avg_prediction = median(cover.t1),
-                 observation = mean(cover.t0),
-                 quant_dist = quantile(cover.t1*100, probs = 0.95) - quantile(cover.t1, probs = 0.05))
-# Calculate error
-avg_qbm$error <- avg_qbm$observation*100 - avg_qbm$avg_prediction*100
-# Aggregate over species
-qbmstats <- ddply(avg_qbm, .(species), summarise,
-                  mae = mean(abs(error)),
-                  mean_cover = mean(observation*100),
-                  quant_dist = mean(quant_dist))
-qbmstats$model <- "QBM"
-
-mae_list <- list()
-mae_pvals <- list()
-species <- unique(qbmstats$species)
-for(spp in species){
-  tmp1 <- subset(avg_ipm, species==spp)
-  tmp2 <- subset(avg_qbm, species==spp)
-  err1 <- as.data.frame(abs(tmp1$observation - tmp1$avg_prediction))
-  err2 <- as.data.frame(abs(tmp2$observation - tmp2$avg_prediction))
-  colnames(err1) <- "resid"
-  err1$model <- "ipm"
-  colnames(err2) <- "resid"
-  err2$model <- "qbm"
-  errd <-  rbind(err1, err2)
-  mae_ttest <- with(errd, t.test(resid~model, alternative = "less"))
-  mae_list[[spp]] <- mae_ttest
-  mae_pvals[[spp]] <- mae_ttest[["p.value"]]
-}
-
-mae_ps <- melt(mae_pvals)
-```
-
-### Comparison of forecast models
-The IPM had lower overall error (MAE, mean absolute error) compared to the QBM for three species (Table 1).
-The IPM MAE is significantly lower at $\alpha=0.05$ for *B.
-gracilis* (*P* = `r mae_ps[1,"value"]`), *H.
-comata* (*P* = `r mae_ps[2,"value"]`), and *P.
-smithii* (*P* = `r mae_ps[3,"value"]`).
-MAEs are statisticially similar between models for *P.
-secunda* (*P* = `r mae_ps[4,"value"]`).
-*P* values are highly sensitive to sample size, so not entirely appropriate in simulation exercises where we control the samples size.
-But, for our purposes they serve as relatively unbiased comparison metrics.
-In no case did the QBM significantly outperform the IPM (Table 1).
-The IPM was consistently more precise than the QBM, with lower distances between the 90% quantiles across all species (Table 1).
-In general the IPM outperformed the QBM because it had (1) lower MAE for three of the four species, (2) statistically similar MAE for the one other species, and (3) considerably more precise forecasts for all species.
-
-### Sensitivity of models to climate
-The response of a population to climate change is a result of the aggregate effects of climate on individual vital rates.
-Since the IPM approach relies on vital rate regressions, we were able to quantify the sensitivity of each vital rate in isolation and in pairwise combinations.
-Across all species, climate covariates can have opposing effects on different vital rates (Fig.
-3).
-Growth was the most sensitive vital rate for all species, showing a negative response to increased precipitation, and stronger positive response to increased temperature, and a mostly positive response when both climate factors are increased (Fig.
-3).
-*B.
-gracilis* survival rates were sensitive to temperature, resulting in an increase in plant cover under increased temperature (Fig.
-3a).
-In isolation, recruitment and survival were insensitive to climate factors for *H.
-comata* (Fig.
-3b).
-Survival and recruitment of *P.
-smithii* were both sensitive, negatively, to temperature and precipitation (Fig.
-3c).
-*P.
-secunda* equilibrium cover was sensitive to the climate effects on survival and recruitment, showing a negative effect on both vital rates for increased precipition, but a strong positive effect on survival with increased temperature (Fig.
-3d).
-Equilibrium cover responded negatively when increased precipitation and temperature affect recruitment  (Fig.
-3d).
-At least two of three vital rates were sensitive to climate for each species (Fig.
-3).
-
-### Sources of uncertainty in the QBM
-Sample size had a relatively weak effect on QBM climate parameter uncertainty after the number of quadrats used in fitting exceeded about 10 (Fig.
-5).
-Inverse-gaussian fits show that increasing sample size beyond the number of quadrats we used would result in diminishing returns in terms of parameter certainty (Fig.
-5).
-
-Climate effects estimated from the QBM are most correlated with climate effects from the growth regression at the individual level (Fig.
-6).
-In no case does the QBM statistical model have strong correlations across all three vital rates (Fig.
-6).
-QBM climate effects were most weakly correlated with those from individual-level recruitment models for *B.
-gracilis*, *H.
-comata*, and *P.
-secunda* (Fig.
-6a,b,d).
-For *P.
-smithii*, QBM climate effects showed no correlation with the survival model effects (Fig.
-6c).
-
-### Model forecasts
-Forecasts based on 1% climate changes were extremely uncertain when we considered model error and parameter uncertainty (Fig.
-6; simulations with mean parameters are in Appendix D for comparison).
-As expected based on model validation (Table 1), QBM projections were more uncertain than IPM projections for all species except *P.
-smithiii* (Fig.
-6).
-IPM forecasts for *P.
-smithiii* were very uncertain due to a very high instrinsic rate of recruitment combined with uncertainty in climate coefficients which lead to high recruitment boom years and subsequent busts when young plants suffer high mortality (Appendx C).
-When we included model error and parameter uncertainty, forecast changes in proportional cover always spanned a wide range of negative to positive values.
-In other words, neither model could predict whether a climate perturbation would increase or decrease equilibrium population size.
-
-```{r get_results, echo=FALSE, eval=FALSE}
-setwd("~/Repos/MicroMesoForecast/analysis/")
-
-####
-##  Mean parameter runs
-####
-meanparam_ipm <- data.frame(cover=NA, species=NA, 
-                             climsim=NA, model=NA)
-##  IPM 
-path <- "./ipm/simulations/results/climchange_meanparams/"
-ipmfiles_mean <- list.files(path)
-ipmfiles_mean <- ipmfiles_mean[grep("cover", ipmfiles_mean)]
-for(file in ipmfiles_mean){
-  tmp <- read.csv(paste(path,file,sep=""))
-  tmp$model <- "IPM"
-  meanparam_ipm <- rbind(meanparam_ipm, tmp[,c("cover", "species", "climsim", "model")])
-}
-meanparam_ipm <- meanparam_ipm[2:nrow(meanparam_ipm),]
-
-newd <- data.frame(cover=NA, species=NA, 
-                   climsim=NA, model=NA, obscover=NA)
-for(dospp in unique(meanparam_ipm$species)){
-  tmp <- subset(meanparam_ipm, species==dospp)
-  obscover <- subset(tmp, climsim=="obs")["cover"]
-  tmp2 <- subset(tmp, climsim!="obs")
-  tmp2$obscover <- rep(obscover$cover, times = length(unique(tmp2$climsim)))
-  newd <- rbind(newd, tmp2)
-}
-ipm_meansims <- newd[2:nrow(newd),]
-ipm_meansims$propdiff <- with(ipm_meansims, log(cover*100)-log(obscover*100))
-
-##  QBM
-meanparam_qbm <- data.frame(cover=NA, species=NA, 
-                             climsim=NA, model=NA)
-path <- "./quadBM/simulations/results/climatechange_meanparams/"
-qbmfiles_mean <- list.files(path)
-for(file in qbmfiles_mean){
-  tmp <- readRDS(paste(path,file,sep=""))
-  tmp$model <- "QBM"
-  meanparam_qbm <- rbind(meanparam_qbm, tmp[,c("cover", "species", "climsim", "model")])
-}
-meanparam_qbm <- meanparam_qbm[2:nrow(meanparam_qbm),]
-
-newd <- data.frame(cover=NA, species=NA, 
-                   climsim=NA, model=NA, obscover=NA)
-for(dospp in unique(meanparam_qbm$species)){
-  tmp <- subset(meanparam_qbm, species==dospp)
-  obscover <- subset(tmp, climsim=="obs")["cover"]
-  tmp2 <- subset(tmp, climsim!="obs")
-  tmp2$obscover <- rep(obscover$cover, times = length(unique(tmp2$climsim)))
-  newd <- rbind(newd, tmp2)
-}
-qbm_meansims <- newd[2:nrow(newd),]
-qbm_meansims$propdiff <- with(qbm_meansims, log(cover*100)-log(obscover*100))
-
-##  Merge IPM and QBM mean parameter results
-bothmodel_outs <- rbind(ipm_meansims, qbm_meansims)
-meanparam_agg <- ddply(bothmodel_outs, .(species, climsim, model), summarise,
-                       meandiff = mean(propdiff),
-                       hidiff = quantile(propdiff, probs = 0.95),
-                       lodiff = quantile(propdiff, probs = 0.05))
-setwd("~/Repos/MicroMesoForecast/manuscript")
-```
-
-```{r comment_out, echo=FALSE, eval=FALSE}
-# Equilibrium cover from both models was sensitive to climate (Fig.
-3a-d).
-The IPM projected percent changes in equilibrium cover from `r round(min(subset(meanparam_agg, species=="BOGR" & model=="IPM")["meandiff"])*100)` to `r round(max(subset(meanparam_agg, species=="BOGR" & model=="IPM")["meandiff"])*100)`% for *B.
-gracilis*, `r round(min(subset(meanparam_agg, species=="HECO" & model=="IPM")["meandiff"])*100)` to `r round(max(subset(meanparam_agg, species=="HECO" & model=="IPM")["meandiff"])*100)`% for *H.
-comata*, `r round(min(subset(meanparam_agg, species=="PASM" & model=="IPM")["meandiff"])*100)` to `r round(max(subset(meanparam_agg, species=="PASM" & model=="IPM")["meandiff"])*100)`% for *P.
-smithii*, and `r round(min(subset(meanparam_agg, species=="POSE" & model=="IPM")["meandiff"])*100)` to `r round(max(subset(meanparam_agg, species=="POSE" & model=="IPM")["meandiff"])*100)`% for *P.
-secunda*.
-The QBM projected opposite and greater percent changes in equilibrium cover for *B.
-gracilis* (`r round(min(subset(meanparam_agg, species=="BOGR" & model=="QBM")["meandiff"])*100)` to `r round(max(subset(meanparam_agg, species=="BOGR" & model=="QBM")["meandiff"])*100)`%) and *H.
-comata* (`r round(min(subset(meanparam_agg, species=="HECO" & model=="QBM")["meandiff"])*100)` to `r round(max(subset(meanparam_agg, species=="HECO" & model=="QBM")["meandiff"])*100)`%; Fig.
-3a-b).
-For *P.
-smithii*, the QBM projected opposite changes in equilibrium cover than the IPM, but of similar magnitude (`r round(min(subset(meanparam_agg, species=="PASM" & model=="QBM")["meandiff"])*100)` to `r round(max(subset(meanparam_agg, species=="PASM" & model=="QBM")["meandiff"])*100)`%; Fig.
-3c).
-*P.
-secunda* was the only species that the IPM and QBM made projections of the same sign and somewhat similar magnitude (`r round(min(subset(meanparam_agg, species=="POSE" & model=="QBM")["meandiff"])*100)` to `r round(max(subset(meanparam_agg, species=="POSE" & model=="QBM")["meandiff"])*100)`%; Fig.
-3d).
-```
-
-Discussion
-----------
-Population models built using individual-level data allow inference on demographic processes, but they can only forecast future population states across the (typically limited) spatial extent of the observations.
-Population-level data are much easier to collect across broad spatial extents, so models built using such data offer an appealing alternative to traditional population models [@Queenborough2011].
-However, density-structured models rely on the aggregation of individual-level data.
-This creates a potential problem if such models are to be used in a climate change context because it is individuals, not populations, which respond to climate [@Clark2012].
-Are models based on population-level metrics as sensitive to climate as models based on individual-level metrics?
-Do these two types of models produce consistent forecasts?
-Do we need detailed demographic data to forecast the impacts of climate change?
-
-### The importance of demographic data
-
-Our comparison of a traditional, demographic population model (the IPM) with a model inspired by density-structured models (the QBM) showed that the IPM outperformed the QBM: the IPM was more accurate and precise than the QBM in out-of-sample cross validation (Table 1).
-The superiority of the IPM could reflect either differences in sample size or the effect of averaging over unique effects of climate on each individual-level vital rate.
-Although increasing sample size of quadrat percent-cover observations would be easy to do in the field, we found little evidence that it would lead to higher precision of climate coefficient estimates (Fig.
-4).
-
-We did, however, find evidence that the QBM statistical model failed to identify climate dependence for some vital rates (Fig.
-5).
-For no species were climate effects from the QBM strongly correlated with all three vital rates (Fig.
-5).
-@Freckleton2011 acknowledge that averaging over complex stage dependence will lead to poorly specified models.
-This is analagous to our situation, but instead of averaging over complex life histories, we are averaging over complex climate dependence.
-Though our work here focused on plant species, this finding is applicable to any species with vital rates that respond uniquely to weather/climate.
-
-Our interpretation is that the QBM is "missing" climate signals associated with at least one vital rate for each species.
-This leads to inaccurate and imprecise forecasts because the QBM statistical model struggles to explain variation due to climate variables that have positive and negative impacts on different vital rates.
-When this is the case, as it is for all our species to varying degrees (Fig.
-3), forecasts from models based on population-level data will fail.
-Our result is consistent with related work on the importance of individual-level data to forecast population responses to exogenous drivers [@Clark2011; @Clark2011b; @Clark2012; @Galvan2014].
-
-Detailed demographic data appears to be necessary to forecast climate change impacts on plant populations when vital rates have unique climate responses.
-How then can we build models to make forecasts for the landscape and regional scales beyond the scope of traditional population models [@Queenborough2011]?
-There are alternatives to density-structured models.
-For example, @Clark2011 use Forest Inventory and Analysis (FIA) data to parameterize a population model with multiple vital rates and climate dependence.
-Distributed efforts such as PlantPopNet (http://plantago.plantpopnet.com) will allow researchers to estimate variation around climate responses for widespread species by taking advantage of spatial variation in climate [e.g.
-@Doak2010].
-Finally, new approaches on the horizon that leverage photo/video of plots and advanced object recognition algorithms [e.g.
-@Liu2014] will increase the efficiency of plant mapping and digitizing efforts.
-
-### The challenge of uncertainty
-
-An important, but unexpected, result of our analysis was the great uncertainty in forecasts, even for our best model.
-The typical approach in ecology is to use point estimates of model parameters to project populations forward according to the specified model, usually allowing for some variability around the determinstic process [e.g.
-@Battin2007; @Jenouvrier2009; @Adler2012].
-If we follow tradition and calculate the mean response to climate perturbation with only model error and interannual variation included, the IPM and the QBM produce opposing forecasts for three of four species (Fig.
-D1).
-It would be tempting to interpret this inconsistency as further evidence for the superiority of the IPM.
-However, if we introduce parameter uncertainty, the forecasts are actually indistinguishable (Fig.
-6), though the IPM projections are generally more precise (consistent with our cross-validation results).
-The real story is that both models produce highly uncertain forecasts.
-For all species, the 90% quantiles of predicted changes in population size overlapped zero; we cannot even predict whether a change in precipitation or temperature will cause populations to increase or decrease.
-This result held when we tried perturbing climate by 10% and 20% as well.
-
-Our results highlight the state of affairs in ecology when it comes to forecasting the impacts of climate change.
-The analysis we conducted here could be considered at the forefront of ecological forecasting with respect to the statistical approach employed (hierarchical Bayesian), the type of population model we used (density-dependent, stochastic IPM with parameter uncertainty), and the amount of high quality data we had at our disposal (14 years of individual-level data).
-Yet, model predictions proved so uncertain that any forecast, when bounded with model and parameter uncertainty, would be uninformative.
-
-How might we improve on this state of affairs?
-First, forecasts could be improved by matching the spatial scale of predictor variables with the spatial scale of observations.
-One of the major limitations of the models we fit here is that the climate data are collected at a larger scale than the individual-level observations of plant size.
-Climate covariates only vary by year, with no spatial variability within years.
-Thus, even if we fit models to individual-level data, we are missing the key interaction point between weather and individual plants [@Clark2011b] because all observations share the same climate covariates.
-Demographic studies should be designed with at least plot-level measurements of climate related variables (e.g., soil moisture).
-Second, accurately detecting climate signals will take even longer time series.
-Recent theoretical work on detecting climate signals in noisy data suggests that even advanced approaches to parameter fitting require 20-25 year time series [@Tellerinpress].
-Third, ecologists need a stronger commitment to reporting uncertainty.
-Although most modeling studies explicitly consider model uncertainty, parameter uncertainty is often ignored.
-In some cases this is because the most convenient statistical methods make it difficult to propogate parameter uncertainty.
-Yet even Bayesian approaches that allow integration of model fitting and forecasting [@Hobbs2015] are not simple when using modeling approaches like integral projection models that separate the model fitting and simulation stages [@Rees2009].
-However, as we have done here, it is still possible to include parameter uncertainty by drawing parameter values from MCMC iterations, taking care to draw all parameters from the same chain and iteration to account for their correlations.
-Only by being honest about our forecasts can we begin to produce better ones, and forecasts reported without parameter error are disingenuous.
-Ignoring parameter error may be justifiable when the goal is investigating basic processes, but it is indefensible when forecasting is the goal.
-
-Conclusions
------------
-This work is not a critique of density-structured population models.
-We are confident that density-structured models will prove to be a valuable tool for many applications.
-However, our analysis represents the first comparison, to our knowledge, of population models based on individual and aggregated forms of the same data in a climate change context.
-Our results confirm theoretical arguments [@Clark2011b] and empirical evidence [@Clark2011; @Clark2012] that individual responses are critical for predicting species' responses to climate change.
-It seems there is no short cut to producing accurate and precise population forecasts: we need detailed demographic data to forecast the impacts of climate change on populations.
-Given the importance of demographic data and its current collection cost, we need modern methods to collect demographic data more efficiently across environmental gradients in space and time.
-
-Our results also offer a cautionary tale because forecast uncertainty was large for both model types.
-Even with 14 years of detailed demographic data and sophisticated modeling techniques, our projections contained too much uncertainty to be informative.
-Uncertainty in demographic responses to climate can be reduced by collecting (1) longer time series and (2) climate covariates that match the scale of inference (e.g., plot rather than landscape level climate/weather metrics).
-
-Acknowledgments
----------------
-This work was funded by the National Science Foundation through a Postdoctoral Research Fellowship in Biology to ATT (DBI-1400370) and a CAREER award to PBA (DEB-1054040).
-We thank the original mappers of the permanent quadrats in Montana and the digitizers in the Adler lab, without whom this work would not have been possible.
-Informal conversations with Stephen Ellner, Giles Hooker, Robin Snyder, and a series of meetings between the Adler and Weecology labs at USU sharpened our thinking.
-Brittany Teller provided comments that improved our manuscript.
-Compute, storage and other resources from the Division of Research Computing in the Office of Research and Graduate Studies at Utah State University are gratefully acknowledged.
-
-\pagebreak{}
-
-Tables
-------
-```{r table_1,results='asis',message=FALSE, comment=NA, echo=FALSE}
-####
-##  Make a table for average residual from each model and year effects structure
-####
-
-##  IPM results
-# Read in data
-setwd("/Users/atredenn/Repos/MicroMesoForecast/manuscript/")
-ipm_onestep <- readRDS("../analysis/ipm/simulations/results/one_step_validation/ipm_loyo_forecasts_combined.rds")
-# Remove predictions >1 for fair comparison to [0,1] truncated quad model
-resets <- which(ipm_onestep[,"cover.t1"]>1)
-ipm_onestep[resets, "cover.t1"] <- 1
-# Average predictions over quad-year reps
-avg_ipm <- ddply(ipm_onestep, .(species, quad, t1), summarise,
-                 avg_prediction = median(cover.t1),
-                 observation = mean(cover.t0),
-                 quant_dist = quantile(cover.t1*100, probs = 0.95) - quantile(cover.t1, probs = 0.05))
-# Calculate error
-avg_ipm$error <- avg_ipm$observation*100 - avg_ipm$avg_prediction*100
-# Aggregate over species
-ipmstats <- ddply(avg_ipm, .(species), summarise,
-                  mae = mean(abs(error)),
-                  mean_cover = mean(observation*100),
-                  quant_dist = mean(quant_dist))
-ipmstats$model <- "IPM"
-
-##  QBM results
-# Read in data
-qbm_onestep <- readRDS("../analysis/quadBM/simulations/results/one_step_validation/qbm_one-step_forecasts_combined.rds")
-# Average predictions over quad-year reps
-avg_qbm <- ddply(qbm_onestep, .(species, quad, year), summarise,
-                 avg_prediction = median(cover.t1),
-                 observation = mean(cover.t0),
-                 quant_dist = quantile(cover.t1*100, probs = 0.95) - quantile(cover.t1, probs = 0.05))
-# Calculate error
-avg_qbm$error <- avg_qbm$observation*100 - avg_qbm$avg_prediction*100
-# Aggregate over species
-qbmstats <- ddply(avg_qbm, .(species), summarise,
-                  mae = mean(abs(error)),
-                  mean_cover = mean(observation*100),
-                  quant_dist = mean(quant_dist))
-qbmstats$model <- "QBM"
-
-mae_list <- list()
-mae_pvals <- list()
-species <- unique(qbmstats$species)
-for(spp in species){
-  tmp1 <- subset(avg_ipm, species==spp)
-  tmp2 <- subset(avg_qbm, species==spp)
-  err1 <- as.data.frame(abs(tmp1$observation - tmp1$avg_prediction))
-  err2 <- as.data.frame(abs(tmp2$observation - tmp2$avg_prediction))
-  colnames(err1) <- "resid"
-  err1$model <- "ipm"
-  colnames(err2) <- "resid"
-  err2$model <- "qbm"
-  errd <-  rbind(err1, err2)
-  mae_ttest <- with(errd, t.test(resid~model, alternative = "less"))
-  mae_list[[spp]] <- mae_ttest
-  mae_pvals[[spp]] <- mae_ttest[["p.value"]]
-}
-
-mae_ps <- melt(mae_pvals)
-
-stats_table <- rbind(ipmstats, qbmstats)
-stats_table <- stats_table[ order(stats_table[,"species"]), ]
-stats_table <- stats_table[,c("species", "model", "mae",
-                              "quant_dist","mean_cover")]
-
-colnames(stats_table) <- c("Species", "Model", "MAE", "90% Distance", "Mean Obs. Cover")
-print.xtable(xtable(stats_table, caption = "Accuracy (mean absolute error, MAE) and precision (90\\% Distance) of out of sample predictions.
-Forecasts were made without random year effects; only climate covariates could explain year-to-year variation.
-90\\% Distance refers to the average distance between the upper and lower 90th percentiles of the 100 predicted values for each quadrat-year combination."),type="latex", comment=FALSE,
-             include.rownames=FALSE, caption.placement="top")
-```
-
-\pagebreak{}
-
-Figures
--------
-```{r figure_1, dependson="plot-options", fig.cap="Time series of average percent cover over all quadrats for our four focal species: *Bouteloua gracilis* (BOGR), *Hesperostipa comata* (HECO), *Pascopyrum smithii* (PASM), and *Poa secunda* (POSE).
-Light grey lines show trajectories of individual quadrats.
-Note the different y-axis scales across panels.", fig.height=3, fig.width=8.5}
-obs_data <- read.csv("../analysis/speciesData/quadAllCover.csv")
-#remove zeros for averaging
-obs_data <- obs_data[which(obs_data$propCover>0),]
-avgobs <- ddply(obs_data, .(year, Species), summarise,
-                mean_cover = mean(propCover))
-
-ggplot()+
-  geom_line(data = obs_data, aes(x=(1900+year), y=propCover*100, group=quad), 
-            color="grey", size=0.25)+
-  geom_line(data=avgobs, aes(x=(1900+year), y=mean_cover*100))+
-  geom_point(data=avgobs, aes(x=(1900+year), y=mean_cover*100),size=3)+
-  ylab("Mean cover (%)")+
-  xlab("Year")+
-  theme_bw()+
-  facet_wrap("Species", ncol=4, nrow=1, scales = "free")+
-  guides(shape=FALSE, linetype=FALSE)
-```
-
-```{r figure_2,dependson="plot-options", fig.cap="Work flow of the data aggregation, model fitting, and population simulating.", fig.height=6}
-library(png)
-library(grid)
-img <- readPNG("components/figure/micromeso_flowchart.png")
-grid.raster(img)
-```
-
-```{r figure_3, dependson="plot-options", fig.cap="Sensitivity of equilibrium cover simulated from the IPM to each climate scenario applied to individual and combined vital rates.
-For example, the points associated with G show the median cover from IPM simulations where a climate perturbation is applied only to the growth regression climate covariates.
-These simulations use mean parameter values for clarity.", fig.width=8, fig.height=6, cache=FALSE, results='hide'}
-setwd("~/Repos/MicroMesoForecast/analysis/")
-
-####
-####  Make IPM sensitivity plot for lower panels
-####
-setwd("../analysis/ipm/simulations/results/climate_sensitivity/")
-files <- list.files()
-cover_files <- files[grep("cover", files)]
-spp_id <- substr(cover_files, 1, 4)
-num_files <- length(cover_files)
-all_sims <- data.frame(time=NA, cover=NA, species=NA, climsim=NA, vital=NA)
-for(i in 1:num_files){
-  tmp <- read.csv(cover_files[i])
-  tmp$species <- rep(spp_id[i], nrow(tmp))
-  all_sims <- rbind(all_sims, tmp)
-}
-all_sims <- all_sims[2:nrow(all_sims),]
-
-# Rename vital rates with codes
-vital_names <- unique(all_sims$vital)
-vital_codes <- c("A", "G", "GR", "GS", "R", "S" , "SR") 
-vitals <- data.frame(names=vital_names,code=vital_codes)
-for(vnow in vital_names){
-  ids <- which(all_sims[,"vital"]==vnow)
-  idcode <- which(vitals[,"names"]==vnow)
-  all_sims[ids,"vital"] <- as.character(vitals[idcode,"code"])
-}
-
-# Rep the 'all' simulation for each vital rate for plotting
-all_clim_sims <- subset(all_sims, vital=="A")
-species <- unique(all_sims$species)
-out_controls <- data.frame(time=NA, cover=NA, species=NA, climsim=NA, vital=NA)
-for(spp in species){
-  tmp <- subset(all_clim_sims, species==spp)
-  for(vital in vital_codes){
-    tmp$vital <- vital
-    out_controls <- rbind(out_controls, tmp)
-  }
-}
-out_controls <- out_controls[2:nrow(out_controls),]
-
-# Combine the datasets
-all_noclim_sims <- subset(all_sims, vital!="all")
-final <- rbind(out_controls, all_noclim_sims)
-
-# Calculate statistics for plotting
-equilibrium_cover <- ddply(final, .(species, climsim, vital), summarise,
-                           eq_cover = median(cover))
-mean_cover <- ddply(subset(equilibrium_cover, climsim=="all"), .(species), summarise,
-                    value = mean(eq_cover))
-
-equilibrium_cover <- subset(equilibrium_cover, vital!="A")
-bothid <- which(equilibrium_cover[,"climsim"]=="ppttemp")
-equilibrium_cover[bothid,"climsim"] <- "zppttemp"
-# myCols2 <- c("#277BA8", "#7ABBBD", "#AED77A")
-sensplot <- ggplot(equilibrium_cover, aes(x=climsim, y=eq_cover*100, 
-                              shape=vital, group=vital, linetype=vital))+
-  geom_hline(data=mean_cover, aes(yintercept=value*100), linetype=2, color="grey45")+
-  geom_line()+
-  geom_point(size=4)+
-  facet_wrap("species", scales = "free", ncol=2)+
-  scale_linetype_discrete(name="Vital Rate")+
-  scale_shape_discrete(name="Vital Rate")+
-  scale_x_discrete(labels=c("baseline", "+ppt", "+temp", "+ppt&temp"))+
-  ylab("Equilibrium Cover (%)")+
-  xlab("Simulation")+
-  theme_bw()+
-  theme(axis.text.x = element_text(angle = 45, hjust = 1))
-
-covertest <- subset(equilibrium_cover, climsim=="all")
-colnames(covertest)[4] <- "obs_cover"
-out <- merge(subset(equilibrium_cover, climsim!="all"), covertest[,c("species","vital","obs_cover")])
-out$propdiff <- with(out, log(eq_cover*100)-log(obs_cover*100))
-
-# meanparamplot <- grid.arrange(meanplot, sensplot, ncol=1, nrow=2)
-# print(meanparamplot)
-print(sensplot)
-
-setwd("~/Repos/MicroMesoForecast/manuscript")
-```
-
-```{r figure_4, dependson="plot-options", fig.cap="Effect of quadrat sample size on the precision (standard deviation) of main climate effect estimates in the QBM.
-Increasing the number of quadrats results in diminishing returns in terms of parameter certainty.
-Light dashed lines show individual climate effects at five quadrat sample sizes.
-Thick dark lines are inverse gaussian fits showing the mean effect of increasing quadrat sample size on parameter precision.", fig.width=3, fig.height=7.5, cache=FALSE, results='hide'}
-library(mgcv)
-
-setwd("../analysis/quadBM/vitalRateRegressions/exclude_quad_test")
-
-####
-####  Read in model fits and summarize results
-####
-source("popgrowth_read_data.R")
-spp <- unique(growD_all$Species)
-num_quads <- numeric(length(spp))
-for(i in 1:length(spp)){
-  tmp <- subset(growD_all, Species==spp[i])
-  num_quads[i] <- length(unique(tmp$quad))
-}
-num_quads <- as.data.frame(num_quads)
-num_quads$species <- spp
-
-all_files <- list.files("./results/")
-all_fits <- data.frame(Parameter=NA, stddev=NA, id=NA, 
-                       rmsquad=NA, rep=NA, species=NA)
-for(i in 1:length(all_files)){
-  tmp <- all_files[i]
-  spp <- substring(tmp, 1, 4)
-  qsrm1 <- unlist(strsplit(tmp, "[_]"))[2]
-  qsrm <- gsub("numout", "", qsrm1)
-  rep1 <- unlist(strsplit(tmp, "[_]"))[3]
-  rep2 <- gsub(".RDS","",rep1)
-  rep <- gsub("rep", "", rep2)
-
-  long <- readRDS(paste("./results/",tmp, sep=""))
-  short <- ddply(long, .(Parameter), summarise,
-                 stddev = sd(value))
-  climeff <- short[grep("b2", short$Parameter),]
-  climeff$id <- substr(climeff$Parameter, 4, length(climeff$Parameter))
-  climeff$id <- as.numeric(unlist(strsplit(climeff$id, split=']')))  
-  climeff <- climeff[with(climeff, order(id)),]
-  climeff$rmsquad <- qsrm
-  climeff$rep <- rep
-  climeff$species <- spp
-  all_fits <- rbind(all_fits, climeff)
-} # end file loop
-
-all_fits <- all_fits[2:nrow(all_fits),]
-
-####
-####  Read in full fits
-####
-dir <- "../truncNormModel/"
-files <- list.files(dir)
-files <- files[grep(".RDS", files)]
-all_alls <- data.frame(Parameter=NA, stddev=NA, id=NA, 
-                       rmsquad=NA, rep=NA, species=NA)
-for(i in 1:length(files)){
-  tmp <- files[i]
-  spp1 <- unlist(strsplit(tmp, "[_]"))[3]
-  spp <- gsub(".RDS","",spp1)
-
-  long <- readRDS(paste(dir,tmp, sep=""))
-  short <- ddply(long, .(Parameter), summarise,
-                 stddev = sd(value))
-  climeff <- short[grep("b2", short$Parameter),]
-  climeff$id <- substr(climeff$Parameter, 4, length(climeff$Parameter))
-  climeff$id <- as.numeric(unlist(strsplit(climeff$id, split=']')))  
-  climeff <- climeff[with(climeff, order(id)),]
-  climeff$rmsquad <- 0
-  climeff$rep <- 1
-  climeff$species <- spp
-  all_alls <- rbind(all_alls, climeff)
-}
-all_alls <- all_alls[2:nrow(all_alls),]
-
-####
-####  Aggregate results by species and rep
-####
-all_fits <- rbind(all_fits, all_alls)
-agg_fits <- ddply(all_fits, .(Parameter, rmsquad, species, id), summarise,
-                  avg_stddev = mean(stddev))
-agg_fits2 <- agg_fits[which(agg_fits$id %in% c(1:5)),]
-agg_fits2 <- merge(agg_fits2, num_quads, by="species")
-agg_fits2$quadsfit <- with(agg_fits2, num_quads-as.numeric(rmsquad))
-
-####
-####  Fit model (need to do by species)
-####
-modBOGR <- gam(avg_stddev~quadsfit, data=subset(agg_fits2, species=="BOGR"),
-               family=quasi(link="1/mu^2"))
-modHECO <- gam(avg_stddev~quadsfit, data=subset(agg_fits2, species=="HECO"),
-               family=quasi(link="1/mu^2"))
-modPASM <- gam(avg_stddev~quadsfit, data=subset(agg_fits2, species=="PASM"),
-               family=quasi(link="1/mu^2"))
-modPOSE <- gam(avg_stddev~quadsfit, data=subset(agg_fits2, species=="POSE"),
-               family=quasi(link="1/mu^2"))
-summary(modBOGR)
-summary(modHECO)
-summary(modPASM)
-summary(modPOSE)
-
-####
-#### Plot
-####
-ggplot(agg_fits2, aes(x=quadsfit, y=avg_stddev))+
-  geom_line(aes(group=Parameter), alpha=0.3, linetype=2)+
-  geom_point(aes(group=Parameter), alpha=0.3, size=3)+
-#   stat_smooth(method="loess", size=2, color="black", se=FALSE)+
-  geom_smooth(method="glm", formula=y~x, family=quasi(link="1/mu^2"),
-              size=2, color="black", se=FALSE)+
-  facet_wrap("species", scales = "free", ncol=1)+
-  theme_bw()+
-  xlab("Number of quadrats fit")+
-  ylab("Standard deviation of coefficient")
-
-```
-
-```{r figure_5, dependson="plot-options", fig.cap="Correlations (r) between QBM and IPM estimates of climate effects.
-We ignore sizeXclimate interactions since these are not directly comparable across model types.
-The QBM does not have multiple vital rates, so its values are repeated across panels within each species.
-Across top panels, 'growth' = growth regression, 'rec' = recruitment regression, 'surv' = survival regression.", fig.width=6, fig.height=6, cache=FALSE, results='hide'}
-##  Figure script to look at correlations between IPM and QBM
-##    climate effect estimates.
-
-library(ggmcmc)
-library(gridExtra)
-library(ggplot2)
-
-allfiles <- list.files("../analysis/speciesData/")
-removals <- grep("csv", allfiles)
-species_list <- allfiles[-removals]
-
-clim_covs <- c("pptLag", "ppt1", "ppt2", "TmeanSpr1", "TmeanSpr2",
-               "ppt1xTmeanSpr1", "ppt2xTmeanSpr2",
-               "sizeXpptLag", "sizeXppt1", "sizeXppt2",
-               "sizeXTmeanSpr1", "sizeXTmeanSpr2")
-clim_mains <- c("pptLag", "ppt1", "ppt2", "TmeanSpr1", "TmeanSpr2",
-                "ppt1xTmeanSpr1", "ppt2xTmeanSpr2")
-
-####
-##  Get mean growth climate effects
-####
-growth_parms <- data.frame(species=NA, model=NA, parameter=NA, value=NA)
-for(spp in species_list){
-  tmp_data <- readRDS(paste("../analysis/ipm/vitalRateRegs/growth/growth_stanmcmc_", spp, ".RDS", sep=""))
-  tmp_clim <- tmp_data[grep("b2", tmp_data[,"Parameter"]), ]
-  tmp_clim[,"Parameter"] <- rep(clim_covs, each=3000)
-  tmp_agg <- ddply(tmp_clim, .(Parameter), summarise,
-                   value = median(value))
-  tmp_agg2 <- tmp_agg[which(tmp_agg$Parameter %in% clim_mains), ]
-  tmp_agg2$species <- spp
-  tmp_agg2$model <- "ipm"
-  colnames(tmp_agg2) <- tolower(colnames(tmp_agg2))
-  growth_parms <- rbind(growth_parms,
-                        tmp_agg2[,c("species", "model", "parameter", "value")])
-}
-
-####
-##  Get mean survival climate effects
-####
-surv_parms <- data.frame(species=NA, model=NA, parameter=NA, value=NA)
-params <- character(length = 12)
-for(i in 1:12){
-  params[i] <- paste("b2[",i,"]", sep="")
-}
-for(spp in species_list){
-  tmp_data <- readRDS(paste("../analysis/ipm/vitalRateRegs/survival/survival_stanmcmc_", spp, ".RDS", sep=""))
-  keeps <- which(tmp_data$Parameter %in% params)
-  tmp_clim <- tmp_data[keeps, ]
-  tmp_clim[,"Parameter"] <- rep(clim_covs, each=3000)
-  tmp_agg <- ddply(tmp_clim, .(Parameter), summarise,
-                   value = median(value))
-  tmp_agg2 <- tmp_agg[which(tmp_agg$Parameter %in% clim_mains), ]
-  tmp_agg2$species <- spp
-  tmp_agg2$model <- "ipm"
-  colnames(tmp_agg2) <- tolower(colnames(tmp_agg2))
-  surv_parms <- rbind(surv_parms,
-                        tmp_agg2[,c("species", "model", "parameter", "value")])
-}
-
-####
-##  Get mean recruitment climate effects
-####
-rec_parms <- data.frame(species=NA, model=NA, parameter=NA, value=NA)
-params <- character(length = 12)
-for(i in 1:12){
-  params[i] <- paste("b2[",i,"]", sep="")
-}
-for(spp in species_list){
-  tmp_data <- readRDS(paste("../analysis/ipm/vitalRateRegs/recruitment/recruitment_stanmcmc_", spp, ".RDS", sep=""))
-  keeps <- which(tmp_data$Parameter %in% params)
-  tmp_clim <- tmp_data[keeps, ]
-  tmp_clim[,"Parameter"] <- rep(clim_mains, each=3000)
-  tmp_agg <- ddply(tmp_clim, .(Parameter), summarise,
-                   value = median(value))
-  tmp_agg2 <- tmp_agg[which(tmp_agg$Parameter %in% clim_mains), ]
-  tmp_agg2$species <- spp
-  tmp_agg2$model <- "ipm"
-  colnames(tmp_agg2) <- tolower(colnames(tmp_agg2))
-  rec_parms <- rbind(rec_parms,
-                      tmp_agg2[,c("species", "model", "parameter", "value")])
-}
-
-####
-##  Get QBM growth climate estimates
-####
-clim_covs2 <- c("pptLag", "ppt1", "ppt2", "TmeanSpr1", "TmeanSpr2",
-               "ppt1xTmeanSpr1", "ppt2xTmeanSpr2",
-               "coverXpptLag", "coverXppt1", "coverXppt2",
-               "coverXTmeanSpr1", "coverXTmeanSpr2")
-
-qbm_parms <- data.frame(species=NA, model=NA, parameter=NA, qbmvalue=NA)
-params <- character(length = 12)
-for(i in 1:12){
-  params[i] <- paste("b2[",i,"]", sep="")
-}
-for(spp in species_list){
-  tmp_data <- readRDS(paste("../analysis/quadBM/vitalRateRegressions/truncNormModel/popgrowth_stanmcmc_", spp, ".RDS", sep=""))
-  keeps <- which(tmp_data$Parameter %in% params)
-  tmp_clim <- tmp_data[keeps, ]
-  tmp_clim[,"Parameter"] <- rep(clim_covs2, each=3000)
-  tmp_agg <- ddply(tmp_clim, .(Parameter), summarise,
-                   qbmvalue = median(value))
-  tmp_agg2 <- tmp_agg[which(tmp_agg$Parameter %in% clim_mains), ]
-  tmp_agg2$species <- spp
-  tmp_agg2$model <- "qbm"
-  colnames(tmp_agg2) <- tolower(colnames(tmp_agg2))
-  qbm_parms <- rbind(qbm_parms,
-                     tmp_agg2[,c("species", "model", "parameter", "qbmvalue")])
-}
-
-####
-##  Format data, combine, and plot
-####
-growth_parms <- growth_parms[2:nrow(growth_parms), ]
-growth_parms$vitalrate <- "growth"
-surv_parms <- surv_parms[2:nrow(surv_parms), ]
-surv_parms$vitalrate <- "surv"
-rec_parms <- rec_parms[2:nrow(rec_parms), ]
-rec_parms$vitalrate <- "rec"
-ipm_parms <- rbind(growth_parms, surv_parms, rec_parms)
-
-qbm_parms <- qbm_parms[2:nrow(qbm_parms), ]
-qbm_parms$vitalrate <- "growth"
-vitals <- c("surv", "rec")
-qbmalone <- qbm_parms
-for(dov in vitals){
-  tmp <- qbmalone
-  tmp$vitalrate <- dov
-  qbm_parms <- rbind(qbm_parms, tmp)
-}
-
-ipm_parms <- ipm_parms[,c("species","parameter","value","vitalrate")]
-qbm_parms <- qbm_parms[,c("species","parameter","qbmvalue","vitalrate")]
-all_ests <- merge(ipm_parms, qbm_parms, by = c("species", "parameter", "vitalrate"))
-
-# Calculate correlation for each group
-cors <- ddply(all_ests, c("species", "vitalrate"), summarise, 
-              cor = round(cor(value, qbmvalue), 2))
-
-ggplot(all_ests, aes(x=qbmvalue, y=value))+
-  geom_point(size=3)+
-  geom_smooth(method="lm", color="black", fill="grey", se=FALSE)+
-  facet_grid(species~vitalrate)+
-  geom_text(data=cors, aes(label=paste("r = ", cor, sep="")), x=-1, y=1, size=4)+
-  xlab("QBM Estimate")+
-  ylab("IPM Estimate")+
-  theme_bw()
-
-setwd("~/Repos/MicroMesoForecast/manuscript/")
-```
-
-```{r figure_6, dependson="plot-options", fig.cap="Mean (points) and 90% quantiles (errorbars) for the proportional difference between baseline simulations (using observed climate) and the climate pertubation simulation on the x-axis.
-We calculated proportional difference as log(perturbed climate cover) - log(observed climate cover), where 'perturbed' and 'observed' refer to the climate time series used to drive interannual variation in the simulations.
-Model error and parameter uncertainty were propagated through the simulation phase.
-Climate simulations are as in Figure 3.", fig.width=4.5, fig.height=7, cache=FALSE, results='hide'}
-####
-##  Vary parameter runs
-####
-setwd("~/Repos/MicroMesoForecast/analysis/")
-meanparam_ipm <- data.frame(cover=NA, species=NA, 
-                             climsim=NA, model=NA)
-##  IPM 
-path <- "./ipm/simulations/results/climchange_varyparams/"
-ipmfiles_mean <- list.files(path)
-ipmfiles_mean <- ipmfiles_mean[grep("cover", ipmfiles_mean)]
-torm <- c(grep("10", ipmfiles_mean), grep("20", ipmfiles_mean), grep("30", ipmfiles_mean))
-ipmfiles_mean <- ipmfiles_mean[-torm]
-for(file in ipmfiles_mean){
-  tmp <- read.csv(paste(path,file,sep=""))
-  tmp$model <- "IPM"
-  meanparam_ipm <- rbind(meanparam_ipm, tmp[,c("cover", "species", "climsim", "model")])
-}
-meanparam_ipm <- meanparam_ipm[2:nrow(meanparam_ipm),]
-
-newd <- data.frame(cover=NA, species=NA, 
-                   climsim=NA, model=NA, obscover=NA)
-for(dospp in unique(meanparam_ipm$species)){
-  tmp <- subset(meanparam_ipm, species==dospp)
-  obscover <- subset(tmp, climsim=="obs")["cover"]
-  tmp2 <- subset(tmp, climsim!="obs")
-  tmp2$obscover <- rep(obscover$cover, times = length(unique(tmp2$climsim)))
-  newd <- rbind(newd, tmp2)
-}
-ipm_meansims <- newd[2:nrow(newd),]
-ipm_meansims$propdiff <- with(ipm_meansims, log(cover*100)-log(obscover*100))
-# ggplot(subset(ipm_meansims, species=="PASM"))+
-#   geom_histogram(aes(x=cover))+
-#   facet_wrap("climsim", scales = "free")+
-#   scale_x_continuous(limits=c(0,0.05))
-##  QBM
-meanparam_qbm <- data.frame(cover=NA, species=NA, 
-                             climsim=NA, model=NA)
-path <- "./quadBM/simulations/results/climatechange_varyparams/"
-qbmfiles_mean <- list.files(path)
-for(file in qbmfiles_mean){
-  tmp <- readRDS(paste(path,file,sep=""))
-  tmp$model <- "QBM"
-  meanparam_qbm <- rbind(meanparam_qbm, tmp[,c("cover", "species", "climsim", "model")])
-}
-meanparam_qbm <- meanparam_qbm[2:nrow(meanparam_qbm),]
-
-newd <- data.frame(cover=NA, species=NA, 
-                   climsim=NA, model=NA, obscover=NA)
-for(dospp in unique(meanparam_qbm$species)){
-  tmp <- subset(meanparam_qbm, species==dospp)
-  obscover <- subset(tmp, climsim=="obs")["cover"]
-  tmp2 <- subset(tmp, climsim!="obs")
-  tmp2$obscover <- rep(obscover$cover, times = length(unique(tmp2$climsim)))
-  newd <- rbind(newd, tmp2)
-}
-qbm_meansims <- newd[2:nrow(newd),]
-qbm_meansims$propdiff <- with(qbm_meansims, log(cover*100)-log(obscover*100))
-
-##  Merge IPM and QBM mean parameter results
-bothmodel_outs <- rbind(ipm_meansims, qbm_meansims)
-
-meanparam_agg <- ddply(bothmodel_outs, .(species, climsim, model), summarise,
-                       meandiff = mean(propdiff),
-                       hidiff = quantile(propdiff, probs = 0.95),
-                       lodiff = quantile(propdiff, probs = 0.05))
-dgd = position_dodge(width = 0.6)
-ggplot(meanparam_agg)+
-  geom_errorbar(aes(x=climsim, ymax=hidiff, ymin=lodiff, group=model, linetype=model),
-                position=dgd, width=0.25)+
-  geom_point(aes(x=climsim, y=meandiff, group=model, shape=model),
-                position=dgd, width=0.25)+
-  theme_bw()+
-  scale_linetype_discrete(name="Model")+
-  scale_shape_discrete(name="Model")+
-  facet_wrap("species", ncol=1, scales="free")+
-  scale_x_discrete(labels=c("+ppt", "+temp", "+ppt&temp"))+
-  ylab("Proportional difference")+
-  xlab("Climate simulation")
-#   theme(axis.text.x = element_text(angle = 45, hjust = 1))
-setwd("~/Repos/MicroMesoForecast/manuscript/")
-```
-
-\pagebreak{}
-
-References
-----------
+### manuscript/
+*

### 2015-09-04 README.Rmd
another update to README

@@
     + ``sumlations/`` -- 
+        - ``another`` --
     + ``vitalRateRegs/`` -- 

### 2015-12-03 manuscript/manuscript.Rmd
tests more spp for QBM

@@
 
-Abstract
---------
+Summary
+-------
+1.
 Rapid climate change has generated growing interest in forecasts of future population trajectories.
@@
 However, such models ignore the fact that climate drives population growth through its influence on individual performance.
-Here, we test the consequences of aggregating individual responses when forecasting climate change impacts on four perennial grass species in a semi-arid grassland in Montana, USA.
-We parameterized two population models based on the same dataset, one based on individual-level data (survival, growth and recruitment) and one on population-level data (percent cover), and compared their accuracy, precision, and sensitivity to climate variables.
+2.
+We test the consequences of aggregating individual responses when forecasting climate change impacts on two perennial grass species (*Pascopyrum smithii* and *Sprobolis flexuosus*).
+We parameterized two population models for each species, one based on individual-level data (survival, growth and recruitment) and one on population-level data (percent cover), and compared their accuracy, precision, and sensitivity to climate variables.
+For both models we used Bayesian ridge regression to identify the optimal predictive model in terms of climate covariate strengths.
+3.
 The individual-level model was more accurate and precise than the aggregated model when predicting out-of-sample observations.
@@
 Increasing the sample size at the population-level would not necessarily reduce forecast uncertainty; the way to reduce uncertainty is to capture unique climate dependence of individual vital rates.
+4.
 Our analysis indicates that there is no shortcut to forecasting climate change impacts on plant populations — detailed demographic data is essential.
@@
 ### Comparison of forecast models
-The IPM had lower overall error (MAE, mean absolute error) compared to the QBM for three species (Table 1).
-The IPM MAE is significantly lower at $\alpha=0.05$ for *B.
-gracilis* (*P* = `r mae_ps[1,"value"]`), *H.
-comata* (*P* = `r mae_ps[2,"value"]`), and *P.
-smithii* (*P* = `r mae_ps[3,"value"]`).
-MAEs are statisticially similar between models for *P.
-secunda* (*P* = `r mae_ps[4,"value"]`).
-*P* values are highly sensitive to sample size, so not entirely appropriate in simulation exercises where we control the samples size.
-But, for our purposes they serve as relatively unbiased comparison metrics.
-In no case did the QBM significantly outperform the IPM (Table 1).
-The IPM was consistently more precise than the QBM, with lower distances between the 90% quantiles across all species (Table 1).
-In general the IPM outperformed the QBM because it had (1) lower MAE for three of the four species, (2) statistically similar MAE for the one other species, and (3) considerably more precise forecasts for all species.
 
@@
 ------
-```{r table_1,results='asis',message=FALSE, comment=NA, echo=FALSE}
+```{r table_1,results='asis',message=FALSE, comment=NA, echo=FALSE, eval=F}
 ####
@@
 Light grey lines show trajectories of individual quadrats.
-Note the different y-axis scales across panels.", fig.height=3, fig.width=8.5}
+Note the different y-axis scales across panels.", fig.height=3, fig.width=8.5, eval=F}
 obs_data <- read.csv("../analysis/speciesData/quadAllCover.csv")
@@
 
-```{r figure_2,dependson="plot-options", fig.cap="Work flow of the data aggregation, model fitting, and population simulating.", fig.height=6}
+```{r figure_2,dependson="plot-options", fig.cap="Work flow of the data aggregation, model fitting, and population simulating.", fig.height=6, eval=F}
 library(png)
@@
 For example, the points associated with G show the median cover from IPM simulations where a climate perturbation is applied only to the growth regression climate covariates.
-These simulations use mean parameter values for clarity.", fig.width=8, fig.height=6, cache=FALSE, results='hide'}
+These simulations use mean parameter values for clarity.", fig.width=8, fig.height=6, cache=FALSE, results='hide', eval=F}
 setwd("~/Repos/MicroMesoForecast/analysis/")
@@
 Light dashed lines show individual climate effects at five quadrat sample sizes.
-Thick dark lines are inverse gaussian fits showing the mean effect of increasing quadrat sample size on parameter precision.", fig.width=3, fig.height=7.5, cache=FALSE, results='hide'}
+Thick dark lines are inverse gaussian fits showing the mean effect of increasing quadrat sample size on parameter precision.", fig.width=3, fig.height=7.5, cache=FALSE, results='hide', eval=F}
 library(mgcv)
@@
 The QBM does not have multiple vital rates, so its values are repeated across panels within each species.
-Across top panels, 'growth' = growth regression, 'rec' = recruitment regression, 'surv' = survival regression.", fig.width=6, fig.height=6, cache=FALSE, results='hide'}
+Across top panels, 'growth' = growth regression, 'rec' = recruitment regression, 'surv' = survival regression.", fig.width=6, fig.height=6, cache=FALSE, results='hide', eval=F}
 ##  Figure script to look at correlations between IPM and QBM
@@
 Model error and parameter uncertainty were propagated through the simulation phase.
-Climate simulations are as in Figure 3.", fig.width=4.5, fig.height=7, cache=FALSE, results='hide'}
+Climate simulations are as in Figure 3.", fig.width=4.5, fig.height=7, cache=FALSE, results='hide', eval=F}
 ####

### 2015-12-03 manuscript/manuscript.tex
tests more spp for QBM

@@
 \usepackage[margin=1in]{geometry}
-\usepackage{graphicx}
-\makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
-\makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
 \ifxetex
@@
 \usepackage{titling}
+
+% Create subtitle command for use in maketitle
+\newcommand{\subtitle}[1]{
+  \posttitle{
+    \begin{center}\large#1\end{center}
+    }
+}
+
 \setlength{\droptitle}{-2em}
@@
 
-\subsection{Abstract}\label{abstract}
-
+\subsection{Summary}\label{summary}
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\itemsep1pt\parskip0pt\parsep0pt
+\item
   Rapid climate change has generated growing interest in forecasts of
-future population trajectories. Traditional population models, typically
-built using detailed demographic observations from one study site, can
-address climate change impacts at one location, but are difficult to
-scale up to the landscape and regional scales relevant to management
-decisions. An alternative is to build models using population-level data
-that is much easier to collect over broad spatial scales than
-individual-level data. However, such models ignore the fact that climate
-drives population growth through its influence on individual
-performance. Here, we test the consequences of aggregating individual
-responses when forecasting climate change impacts on four perennial
-grass species in a semi-arid grassland in Montana, USA. We parameterized
-two population models based on the same dataset, one based on
+  future population trajectories. Traditional population models,
+  typically built using detailed demographic observations from one study
+  site, can address climate change impacts at one location, but are
+  difficult to scale up to the landscape and regional scales relevant to
+  management decisions. An alternative is to build models using
+  population-level data that is much easier to collect over broad
+  spatial scales than individual-level data. However, such models ignore
+  the fact that climate drives population growth through its influence
+  on individual performance.
+\item
+  We test the consequences of aggregating individual responses when
+  forecasting climate change impacts on two perennial grass species
+  (\emph{Pascopyrum smithii} and \emph{Sprobolis flexuosus}). We
+  parameterized two population models for each species, one based on
   individual-level data (survival, growth and recruitment) and one on
   population-level data (percent cover), and compared their accuracy,
-precision, and sensitivity to climate variables. The individual-level
-model was more accurate and precise than the aggregated model when
-predicting out-of-sample observations. When comparing climate effects
-from both models, the population-level model missed important climate
-effects from at least one vital rate for each species. Increasing the
-sample size at the population-level would not necessarily reduce
-forecast uncertainty; the way to reduce uncertainty is to capture unique
-climate dependence of individual vital rates. Our analysis indicates
-that there is no shortcut to forecasting climate change impacts on plant
-populations --- detailed demographic data is essential. Despite the
-superiority of the individual-level model, the forecasts it generated
-still were too uncertain to be useful for decision-makers. We need new
-methods to collect demographic data efficiently across environmental
-gradients in space and time.
+  precision, and sensitivity to climate variables. For both models we
+  used Bayesian ridge regression to identify the optimal predictive
+  model in terms of climate covariate strengths.
+\item
+  The individual-level model was more accurate and precise than the
+  aggregated model when predicting out-of-sample observations. When
+  comparing climate effects from both models, the population-level model
+  missed important climate effects from at least one vital rate for each
+  species. Increasing the sample size at the population-level would not
+  necessarily reduce forecast uncertainty; the way to reduce uncertainty
+  is to capture unique climate dependence of individual vital rates.
+\item
+  Our analysis indicates that there is no shortcut to forecasting
+  climate change impacts on plant populations --- detailed demographic
+  data is essential. Despite the superiority of the individual-level
+  model, the forecasts it generated still were too uncertain to be
+  useful for decision-makers. We need new methods to collect demographic
+  data efficiently across environmental gradients in space and time.
+\end{enumerate}
 
@@
 without information about how the fitted parameters respond to spatial
-variation in biotic and abiotic drivers (Sæther et al. 2007). The
+variation in biotic and abiotic drivers (S{æ}ther et al. 2007). The
 limited spatial extent of individual-level demographic datasets
@@
 landscapes and regions. However, the limitations of SDMs are well known
-(Pearson and Dawson 2003, Elith and Leathwick 2009, Araújo and Peterson
-2012). Ideally, researchers would provide managers with landscape-scale
-population models, combining the extent of SDMs with information about
-dynamics and species abundances (Schurr et al. 2012, Merow et al. 2014).
+(Pearson and Dawson 2003, Elith and Leathwick 2009, Ara{ú}jo and
+Peterson 2012). Ideally, researchers would provide managers with
+landscape-scale population models, combining the extent of SDMs with
+information about dynamics and species abundances (Schurr et al. 2012,
+Merow et al. 2014).
 
@@
 performance, offer an intriguing alternative for modeling populations
-(Clark and Bjørnstad 2004, Freckleton et al. 2011). Population-level
+(Clark and Bj{ø}rnstad 2004, Freckleton et al. 2011). Population-level
 data cannot provide inference about demographic mechanisms, but might be
@@
 models}\label{comparison-of-forecast-models}
-
-The IPM had lower overall error (MAE, mean absolute error) compared to
-the QBM for three species (Table 1). The IPM MAE is significantly lower
-at \(\alpha=0.05\) for \emph{B. gracilis} (\emph{P} = 0.0012), \emph{H.
-comata} (\emph{P} = 4.0586 × 10-8), and \emph{P. smithii} (\emph{P} =
-3.183 × 10-5). MAEs are statisticially similar between models for
-\emph{P. secunda} (\emph{P} = 0.0922). \emph{P} values are highly
-sensitive to sample size, so not entirely appropriate in simulation
-exercises where we control the samples size. But, for our purposes they
-serve as relatively unbiased comparison metrics. In no case did the QBM
-significantly outperform the IPM (Table 1). The IPM was consistently
-more precise than the QBM, with lower distances between the 90\%
-quantiles across all species (Table 1). In general the IPM outperformed
-the QBM because it had (1) lower MAE for three of the four species, (2)
-statistically similar MAE for the one other species, and (3)
-considerably more precise forecasts for all species.
 
@@
 individual-level data to forecast population responses to exogenous
-drivers (Clark et al. 2011a, 2011b, 2012, Galván et al. 2014).
+drivers (Clark et al. 2011a, 2011b, 2012, Galv{á}n et al. 2014).
 
@@
 
-\begin{table}[ht]
-\centering
-\caption{Accuracy (mean absolute error, MAE) and precision (90\% Distance) of out of sample predictions.
-Forecasts were made without random year effects; only climate covariates could explain year-to-year variation.
-90\% Distance refers to the average distance between the upper and lower 90th percentiles of the 100 predicted values for each quadrat-year combination.}
-\begin{tabular}{llrrr}
-  \hline
-Species & Model & MAE & 90\% Distance & Mean Obs. Cover \\ 
-  \hline
-BOGR & IPM & 12.18 & 38.52 & 9.43 \\ 
-  BOGR & QBM & 19.66 & 56.50 & 9.26 \\ 
-  HECO & IPM & 1.22 & 6.47 & 1.15 \\ 
-  HECO & QBM & 12.35 & 41.11 & 1.18 \\ 
-  PASM & IPM & 0.19 & 1.65 & 0.42 \\ 
-  PASM & QBM & 0.55 & 7.78 & 0.42 \\ 
-  POSE & IPM & 1.37 & 7.64 & 1.25 \\ 
-  POSE & QBM & 1.79 & 40.59 & 1.27 \\ 
-   \hline
-\end{tabular}
-\end{table}
-
 \pagebreak{}
@@
 
-\begin{figure}[htbp]
-\centering
-\includegraphics{components/figure/manuscript-figure_1.pdf}
-\caption{Time series of average percent cover over all quadrats for our
-four focal species: \emph{Bouteloua gracilis} (BOGR), \emph{Hesperostipa
-comata} (HECO), \emph{Pascopyrum smithii} (PASM), and \emph{Poa secunda}
-(POSE). Light grey lines show trajectories of individual quadrats. Note
-the different y-axis scales across panels.}
-\end{figure}
-
-\begin{figure}[htbp]
-\centering
-\includegraphics{components/figure/manuscript-figure_2.pdf}
-\caption{Work flow of the data aggregation, model fitting, and
-population simulating.}
-\end{figure}
-
-\begin{figure}[htbp]
-\centering
-\includegraphics{components/figure/manuscript-figure_3.pdf}
-\caption{Sensitivity of equilibrium cover simulated from the IPM to each
-climate scenario applied to individual and combined vital rates. For
-example, the points associated with G show the median cover from IPM
-simulations where a climate perturbation is applied only to the growth
-regression climate covariates. These simulations use mean parameter
-values for clarity.}
-\end{figure}
-
-\begin{figure}[htbp]
-\centering
-\includegraphics{components/figure/manuscript-figure_4.pdf}
-\caption{Effect of quadrat sample size on the precision (standard
-deviation) of main climate effect estimates in the QBM. Increasing the
-number of quadrats results in diminishing returns in terms of parameter
-certainty. Light dashed lines show individual climate effects at five
-quadrat sample sizes. Thick dark lines are inverse gaussian fits showing
-the mean effect of increasing quadrat sample size on parameter
-precision.}
-\end{figure}
-
-\begin{figure}[htbp]
-\centering
-\includegraphics{components/figure/manuscript-figure_5.pdf}
-\caption{Correlations (r) between QBM and IPM estimates of climate
-effects. We ignore sizeXclimate interactions since these are not
-directly comparable across model types. The QBM does not have multiple
-vital rates, so its values are repeated across panels within each
-species. Across top panels, `growth' = growth regression, `rec' =
-recruitment regression, `surv' = survival regression.}
-\end{figure}
-
-\begin{figure}[htbp]
-\centering
-\includegraphics{components/figure/manuscript-figure_6.pdf}
-\caption{Mean (points) and 90\% quantiles (errorbars) for the
-proportional difference between baseline simulations (using observed
-climate) and the climate pertubation simulation on the x-axis. We
-calculated proportional difference as log(perturbed climate cover) -
-log(observed climate cover), where `perturbed' and `observed' refer to
-the climate time series used to drive interannual variation in the
-simulations. Model error and parameter uncertainty were propagated
-through the simulation phase. Climate simulations are as in Figure 3.}
-\end{figure}
-
 \pagebreak{}
@@
 
-Araújo, M. B., and A. T. Peterson. 2012. Uses and misuses of bioclimatic
-envelope modeling. Ecology 93:1527--1539.
+Ara{ú}jo, M. B., and A. T. Peterson. 2012. Uses and misuses of
+bioclimatic envelope modeling. Ecology 93:1527--1539.
 
@@
 recruitment stage to stabilize grassland coexistence. Ecological
-Monographs.
+Monographs 85:373--392.
 
@@
 
-Clark, J. S., and O. N. Bjørnstad. 2004. Population time series: Process
-variability, observation errors, missing values, lags, and hidden
-states. Ecology 85:3140--3150.
+Clark, J. S., and O. N. Bj{ø}rnstad. 2004. Population time series:
+Process variability, observation errors, missing values, lags, and
+hidden states. Ecology 85:3140--3150.
 
@@
 Clark, J. S., D. Bell, C. Chu, B. Courbaud, M. Dietze, M. Hersh, J.
-HilleRisLambers, I. Ibáñez, S. LaDeau, S. McMahon, J. Metcalf, J. Mohan,
-E. Moran, L. Pangle, S. Pearson, C. Salk, Z. Shen, D. Valle, and P.
-Wyckoff. 2010. High-dimensional coexistence based on individual
+HilleRisLambers, I. Ib{á}{ñ}ez, S. LaDeau, S. McMahon, J. Metcalf, J.
+Mohan, E. Moran, L. Pangle, S. Pearson, C. Salk, Z. Shen, D. Valle, and
+P. Wyckoff. 2010. High-dimensional coexistence based on individual
 variation: a synthesis of evidence. Ecological Monographs 80:569--608.
@@
 
-Galván, J. D., J. J. Camarero, and E. Gutiérrez. 2014. Seeing the trees
-for the forest: Drivers of individual growth responses to climate in
-Pinus uncinata mountain forests. Journal of Ecology 102:1244--1257.
+Galv{á}n, J. D., J. J. Camarero, and E. Guti{é}rrez. 2014. Seeing the
+trees for the forest: Drivers of individual growth responses to climate
+in Pinus uncinata mountain forests. Journal of Ecology 102:1244--1257.
 
@@
 
-Petchey, O. L., M. Pontarp, T. M. Massie, S. Kéfi, A. Ozgul, M.
+Petchey, O. L., M. Pontarp, T. M. Massie, S. K{é}fi, A. Ozgul, M.
 Weilenmann, G. M. Palamara, F. Altermatt, B. Matthews, J. M. Levine, D.
@@
 O'Hara, F. Hartig, W. D. Kissling, H. P. Linder, G. F. Midgley, B.
-Schröder, A. Singer, and N. E. Zimmermann. 2012. How to understand
+Schr{ö}der, A. Singer, and N. E. Zimmermann. 2012. How to understand
 species' niches and range dynamics: A demographic research agenda for
@@
 
-Sæther, B. E., S. Engen, V. Grøtan, W. Fiedler, E. Matthysen, M. E.
-Visser, J. Wright, A. P. Møller, F. Adriaensen, H. {Van Balen}, D.
+S{æ}ther, B. E., S. Engen, V. Gr{ø}tan, W. Fiedler, E. Matthysen, M. E.
+Visser, J. Wright, A. P. M{ø}ller, F. Adriaensen, H. {Van Balen}, D.
 Balmer, M. C. Mainwaring, R. H. McCleery, M. Pampus, and W. Winkel.

### 2015-12-03 analysis/quadBM/popgrowth_fits/popgrowth_spfl.R

@@ -1,0 +1,230 @@
+## Script to estimate parameters for the quad-based model using STAN
+
+library(rstan)
+library(parallel)
+library(ggmcmc)
+library(snowfall)
+library(rlecuyer)
+source("../../waic_fxns.R")
+
+##  Read in data
+#bring in data
+allD <- read.csv("../../nmData/SPFL/quadratCover.csv")
+allD <- subset(allD, year<41)
+# allD <- allD[,2:ncol(allD)] #get rid of X ID column
+sppList <- "SPFL"
+
+climD <- read.csv("../../nmData/Climate.csv")
+climD[2:6] <- scale(climD[2:6], center = TRUE, scale = TRUE)
+
+backD <- data.frame(climYear=NA,
+                    quad = NA,
+                    year= NA,
+                    totCover= NA,
+                    lag.cover= NA,
+                    pptLag= NA,
+                    ppt1= NA,
+                    TmeanSum1= NA,
+                    ppt2= NA,
+                    TmeanSum2= NA,
+                    yearID= NA,
+                    group = NA,
+                    percCover = NA,
+                    percLagCover = NA)
+
+sppD <- allD
+
+# create lag cover variable
+tmp=sppD[,c("quad","year","totCover")]
+tmp$year=tmp$year+1
+names(tmp)[3]="lag.cover"
+sppD=merge(sppD,tmp,all.x=T)
+
+# merge in climate data
+sppD$climYear=sppD$year+1900-1  
+sppD=merge(sppD,climD,by.x="climYear",by.y="year")
+
+#Growth observations
+growD <- subset(sppD,lag.cover>0 & totCover>0)
+growD$yearID <- growD$year #for random year offset on intercept
+growD$group <- substring(growD$quad, 1, 1)
+growD$percCover <- growD$totCover/10000
+growD$percLagCover <- growD$lag.cover/10000
+backD <- rbind(backD, growD)
+
+growD_all <- backD[2:nrow(backD),]
+
+
+growD <- growD_all
+clim_covs <- growD[,c("pptLag", "ppt1", "ppt2", "TmeanSum1", "TmeanSum2")]
[…]

### 2016-03-12 manuscript/manuscript.Rmd
updates ms with JECOL style abstract

@@
 Traditional population models, typically built using detailed demographic observations from one study site, can address climate change impacts at one location, but are difficult to scale up to the landscape and regional scales relevant to management decisions.
+2.
 An alternative is to build models using population-level data that is much easier to collect over broad spatial scales than individual-level data.
 However, such models ignore the fact that climate drives population growth through its influence on individual performance.
-2.
-We test the consequences of aggregating individual responses when forecasting climate change impacts on two perennial grass species (*Pascopyrum smithii* and *Sprobolis flexuosus*).
+3.
+We test the consequences of aggregating individual responses when forecasting climate change impacts on four perennial grass species.
 We parameterized two population models for each species, one based on individual-level data (survival, growth and recruitment) and one on population-level data (percent cover), and compared their accuracy, precision, and sensitivity to climate variables.
-For both models we used Bayesian ridge regression to identify the optimal predictive model in terms of climate covariate strengths.
-3.
+For both model types we used Bayesian ridge regression to identify the optimal predictive model in terms of climate covariate strengths.
+4.
 The individual-level model was more accurate and precise than the aggregated model when predicting out-of-sample observations.
@@
 Increasing the sample size at the population-level would not necessarily reduce forecast uncertainty; the way to reduce uncertainty is to capture unique climate dependence of individual vital rates.
-4.
-Our analysis indicates that there is no shortcut to forecasting climate change impacts on plant populations — detailed demographic data is essential.
+5.
+*Synthesis* Our analysis indicates that there is no shortcut to forecasting climate change impacts on plant populations — detailed demographic data is essential.
 Despite the superiority of the individual-level model, the forecasts it generated still were too uncertain to be useful for decision-makers.

### 2016-03-12 manuscript/manuscript.tex
updates ms with JECOL style abstract

@@
   difficult to scale up to the landscape and regional scales relevant to
-  management decisions. An alternative is to build models using
-  population-level data that is much easier to collect over broad
-  spatial scales than individual-level data. However, such models ignore
-  the fact that climate drives population growth through its influence
-  on individual performance.
+  management decisions.
+\item
+  An alternative is to build models using population-level data that is
+  much easier to collect over broad spatial scales than individual-level
+  data. However, such models ignore the fact that climate drives
+  population growth through its influence on individual performance.
 \item
   We test the consequences of aggregating individual responses when
-  forecasting climate change impacts on two perennial grass species
-  (\emph{Pascopyrum smithii} and \emph{Sprobolis flexuosus}). We
+  forecasting climate change impacts on four perennial grass species. We
   parameterized two population models for each species, one based on
@@
   population-level data (percent cover), and compared their accuracy,
-  precision, and sensitivity to climate variables. For both models we
-  used Bayesian ridge regression to identify the optimal predictive
+  precision, and sensitivity to climate variables. For both model types
+  we used Bayesian ridge regression to identify the optimal predictive
   model in terms of climate covariate strengths.
@@
 \item
-  Our analysis indicates that there is no shortcut to forecasting
-  climate change impacts on plant populations --- detailed demographic
-  data is essential. Despite the superiority of the individual-level
-  model, the forecasts it generated still were too uncertain to be
-  useful for decision-makers. We need new methods to collect demographic
-  data efficiently across environmental gradients in space and time.
+  \emph{Synthesis} Our analysis indicates that there is no shortcut to
+  forecasting climate change impacts on plant populations --- detailed
+  demographic data is essential. Despite the superiority of the
+  individual-level model, the forecasts it generated still were too
+  uncertain to be useful for decision-makers. We need new methods to
+  collect demographic data efficiently across environmental gradients in
+  space and time.
 \end{enumerate}
@@
 without information about how the fitted parameters respond to spatial
-variation in biotic and abiotic drivers (S{æ}ther et al. 2007). The
+variation in biotic and abiotic drivers (Sæther et al. 2007). The
 limited spatial extent of individual-level demographic datasets
@@
 landscapes and regions. However, the limitations of SDMs are well known
-(Pearson and Dawson 2003, Elith and Leathwick 2009, Ara{ú}jo and
-Peterson 2012). Ideally, researchers would provide managers with
-landscape-scale population models, combining the extent of SDMs with
-information about dynamics and species abundances (Schurr et al. 2012,
-Merow et al. 2014).
+(Pearson and Dawson 2003, Elith and Leathwick 2009, Araújo and Peterson
+2012). Ideally, researchers would provide managers with landscape-scale
+population models, combining the extent of SDMs with information about
+dynamics and species abundances (Schurr et al. 2012, Merow et al. 2014).
 
@@
 performance, offer an intriguing alternative for modeling populations
-(Clark and Bj{ø}rnstad 2004, Freckleton et al. 2011). Population-level
+(Clark and Bjørnstad 2004, Freckleton et al. 2011). Population-level
 data cannot provide inference about demographic mechanisms, but might be
@@
 All of our analyses (model fitting and simulating) were conducted in R
-(R Core Development Team 2013). We used the `No-U-Turn' MCMC sampler in
-Stan (Stan Development Team 2014a) to estimate the posterior
-distributions of model parameters using the package `rstan' (Stan
-Development Team 2014b). We obtained posterior distributions for all
-model parameters from three parallel MCMC chains run for 1,000
-iterations after discarding an initial 1,000 iterations. Such short MCMC
-chains may surprise readers more familiar with other MCMC samplers
-(i.e.~JAGS or WinBUGS), but the Stan sampler is exceptionally efficient,
-which reduces the number of iterations needed to achieve convergence. We
-assessed convergence visually and made sure scale reduction factors for
-all parameters were less than 1.01. For the purposes of including
-parameter uncertainty in our population models, we saved the final 1,000
-iterations from each of the three MCMC chains to be used as randomly
-drawn values during population simulation. This step alleviates the need
-to reduce model parameters by model selection since sampling from the
-full parameter space in the MCMC ensures that if a parameter broadly
-overlaps zero, on average the effect in the population models will also
-be near zero. We report the posterior mean, standard deviation, and 95\%
-Bayesian Credible Intervals for every parameter of each model for each
-species in Appendix B.
+(R Core Team 2013). We used the `No-U-Turn' MCMC sampler in Stan (Stan
+Development Team 2014a) to estimate the posterior distributions of model
+parameters using the package `rstan' (Stan Development Team 2014b). We
+obtained posterior distributions for all model parameters from three
+parallel MCMC chains run for 1,000 iterations after discarding an
+initial 1,000 iterations. Such short MCMC chains may surprise readers
+more familiar with other MCMC samplers (i.e.~JAGS or WinBUGS), but the
+Stan sampler is exceptionally efficient, which reduces the number of
+iterations needed to achieve convergence. We assessed convergence
+visually and made sure scale reduction factors for all parameters were
+less than 1.01. For the purposes of including parameter uncertainty in
+our population models, we saved the final 1,000 iterations from each of
+the three MCMC chains to be used as randomly drawn values during
+population simulation. This step alleviates the need to reduce model
+parameters by model selection since sampling from the full parameter
+space in the MCMC ensures that if a parameter broadly overlaps zero, on
+average the effect in the population models will also be near zero. We
+report the posterior mean, standard deviation, and 95\% Bayesian
+Credible Intervals for every parameter of each model for each species in
+Appendix B.
 
@@
 individual-level data to forecast population responses to exogenous
-drivers (Clark et al. 2011a, 2011b, 2012, Galv{á}n et al. 2014).
+drivers (Clark et al. 2011a, 2011b, 2012, Galván et al. 2014).
 
@@
 Adler, P. B., S. P. Ellner, and J. M. Levine. 2010. Coexistence of
-perennial plants: An embarrassment of niches.
+perennial plants: An embarrassment of niches. Ecology Letters
+13:1019--1029.
 
@@
 
-Ara{ú}jo, M. B., and A. T. Peterson. 2012. Uses and misuses of
-bioclimatic envelope modeling. Ecology 93:1527--1539.
+Araújo, M. B., and A. T. Peterson. 2012. Uses and misuses of bioclimatic
+envelope modeling. Ecology 93:1527--1539.
 
@@
 
-Clark, J. S., and O. N. Bj{ø}rnstad. 2004. Population time series:
-Process variability, observation errors, missing values, lags, and
-hidden states. Ecology 85:3140--3150.
+Clark, J. S., and O. N. Bjørnstad. 2004. Population time series: Process
+variability, observation errors, missing values, lags, and hidden
+states. Ecology 85:3140--3150.
 
@@
 Clark, J. S., D. Bell, C. Chu, B. Courbaud, M. Dietze, M. Hersh, J.
-HilleRisLambers, I. Ib{á}{ñ}ez, S. LaDeau, S. McMahon, J. Metcalf, J.
-Mohan, E. Moran, L. Pangle, S. Pearson, C. Salk, Z. Shen, D. Valle, and
-P. Wyckoff. 2010. High-dimensional coexistence based on individual
+HilleRisLambers, I. Ibáñez, S. LaDeau, S. McMahon, J. Metcalf, J. Mohan,
+E. Moran, L. Pangle, S. Pearson, C. Salk, Z. Shen, D. Valle, and P.
+Wyckoff. 2010. High-dimensional coexistence based on individual
 variation: a synthesis of evidence. Ecological Monographs 80:569--608.
@@
 
-Galv{á}n, J. D., J. J. Camarero, and E. Guti{é}rrez. 2014. Seeing the
-trees for the forest: Drivers of individual growth responses to climate
-in Pinus uncinata mountain forests. Journal of Ecology 102:1244--1257.
+Galván, J. D., J. J. Camarero, and E. Gutiérrez. 2014. Seeing the trees
+for the forest: Drivers of individual growth responses to climate in
+Pinus uncinata mountain forests. Journal of Ecology 102:1244--1257.
 
@@
 
-Petchey, O. L., M. Pontarp, T. M. Massie, S. K{é}fi, A. Ozgul, M.
+Petchey, O. L., M. Pontarp, T. M. Massie, S. Kéfi, A. Ozgul, M.
 Weilenmann, G. M. Palamara, F. Altermatt, B. Matthews, J. M. Levine, D.
@@
 
-R Core Development Team. 2013. R: A language and environment for
-statistical computing.
+R Core Team. 2013. R: A language and environment for statistical
+computing.
 
@@
 O'Hara, F. Hartig, W. D. Kissling, H. P. Linder, G. F. Midgley, B.
-Schr{ö}der, A. Singer, and N. E. Zimmermann. 2012. How to understand
+Schröder, A. Singer, and N. E. Zimmermann. 2012. How to understand
 species' niches and range dynamics: A demographic research agenda for
@@
 
-S{æ}ther, B. E., S. Engen, V. Gr{ø}tan, W. Fiedler, E. Matthysen, M. E.
-Visser, J. Wright, A. P. M{ø}ller, F. Adriaensen, H. {Van Balen}, D.
+Sæther, B. E., S. Engen, V. Grøtan, W. Fiedler, E. Matthysen, M. E.
+Visser, J. Wright, A. P. Møller, F. Adriaensen, H. {Van Balen}, D.
 Balmer, M. C. Mainwaring, R. H. McCleery, M. Pampus, and W. Winkel.

### 2016-03-12 analysis/quadBM/popgrowth_fits/popgrowth_fit.R

@@ -1,0 +1,159 @@
+## Script to estimate parameters for the quad-based model using STAN
+
+
+####
+####  Load Libraries and Subroutines
+####
+library(rstan)
+library(plyr)
+library(reshape2)
+library(ggmcmc)
+library(matrixStats)
+library(parallel)
+
+
+
+##  Read in data
+allD <- read.csv("../../speciesData/quadAllCover.csv")
+allD <- allD[,2:ncol(allD)] #get rid of X ID column
+sppList <- as.character(unique(allD$Species))
+climD <- read.csv("../../weather/Climate.csv")
+clim_vars <- c("pptLag", "ppt1", "ppt2", "TmeanSpr1", "TmeanSpr2")
+
+backD <- data.frame(climYear=NA,
+                    quad = NA,
+                    year= NA,
+                    totCover= NA,
+                    Species= NA,
+                    propCover= NA,
+                    lag.cover= NA,
+                    pptLag= NA,
+                    ppt1= NA,
+                    TmeanSpr1= NA,
+                    ppt2= NA,
+                    TmeanSpr2= NA,
+                    TmeanSum1= NA,
+                    TmeanSum2= NA,
+                    yearID= NA,
+                    group = NA,
+                    percCover = NA,
+                    percLagCover = NA)
+
+#loop through species and remake data frame
+for(spp in 1:length(sppList)){
+  doSpp <- sppList[spp]
+  sppD <- subset(allD, Species==doSpp)
+  
+  # create lag cover variable
+  tmp=sppD[,c("quad","year","totCover")]
+  tmp$year=tmp$year+1
+  names(tmp)[3]="lag.cover"
+  sppD=merge(sppD,tmp,all.x=T)
+  
+  # merge in climate data
+  sppD$climYear=sppD$year+1900-1  
+  sppD=merge(sppD,climD,by.x="climYear",by.y="year")
+  
+  #Growth observations
+  growD <- subset(sppD,lag.cover>0 & totCover>0)
+  growD$yearID <- growD$year #for random year offset on intercept
+  growD$group <- substring(growD$quad, 1, 1)
[…]

### 2016-03-12 analysis/quadBM/simulations/noClimChangeQBM_varyparams.R

@@ -21,12 +21,11 @@ library(ggthemes)
 library(gridExtra)
 library(EnvStats)
 
-#bring in data
+##  Read in data
 allD <- read.csv("../../speciesData/quadAllCover.csv")
 allD <- allD[,2:ncol(allD)] #get rid of X ID column
-allD$percCover <- allD$totCover/10000
-# head(scale(allD$percCover, center=TRUE, scale=TRUE))
 sppList <- as.character(unique(allD$Species))
+climD <- read.csv("../../weather/Climate.csv")
 
 #perturb climate data
 climD <- read.csv("../../weather/Climate.csv")
@@ -34,16 +33,60 @@ climScale <- scale(climD[2:6], center = TRUE, scale = TRUE)
 climAvg <- apply(X = climD, MARGIN = 2, FUN = mean)
 climSD <- apply(X = climD, MARGIN = 2, FUN = sd)
 
-# pptVars=grep("ppt",names(climD))
-# tmp1=perc_change*colMeans(climD)
-# tmp1=matrix(tmp1,NROW(climD),length(tmp1),byrow=T)
-# climD[,pptVars]=climD[,pptVars]+tmp1[,pptVars]
-
 climD[2] <- (climD[2] - climAvg[2])/climSD[2]
 climD[3] <- (climD[3] - climAvg[3])/climSD[3]
 climD[4] <- (climD[4] - climAvg[4])/climSD[4]
 climD[5] <- (climD[5] - climAvg[5])/climSD[5]
 climD[6] <- (climD[6] - climAvg[6])/climSD[6]
+
+
+clim_vars <- c("pptLag", "ppt1", "ppt2", "TmeanSpr1", "TmeanSpr2")
+
+backD <- data.frame(climYear=NA,
+                    quad = NA,
+                    year= NA,
+                    totCover= NA,
+                    Species= NA,
+                    propCover= NA,
+                    lag.cover= NA,
+                    pptLag= NA,
+                    ppt1= NA,
+                    TmeanSpr1= NA,
+                    ppt2= NA,
+                    TmeanSpr2= NA,
+                    TmeanSum1= NA,
+                    TmeanSum2= NA,
+                    yearID= NA,
+                    group = NA,
+                    percCover = NA,
+                    percLagCover = NA)
+
+#loop through species and remake data frame
+for(spp in 1:length(sppList)){
+  doSpp <- sppList[spp]
[…]

### 2016-03-23 manuscript/manuscript.Rmd
adds lppd stuff to ms and formats for JEcol

--- 2015-12-03 manuscript/manuscript.Rmd
@@
    - \usepackage{setspace}
+   - \usepackage{times}
+   - \usepackage{sectsty}
    - \doublespacing
@@
    - \usepackage[document]{ragged2e}
-   - \pagewiselinenumbers
 bibliography: ~/Dropbox/Bibliography/MicroMeso.bib
@@
     fig_caption: true
-    keep_tex: true
+    keep_tex: false
 fontsize: 12pt
 geometry: margin=1in
+linkcolor: black
+urlcolor: black
 
 ---
-
-### Title
-Do we need demographic data to forecast population responses to climate change?
-
-### Authors 
-1. Andrew T. Tredennick (corresponding author: atredenn@gmail.com) 
-2. Peter B. Adler (peter.adler@usu.edu)
-
-### Affiliations
-1. Department of Wildland Resources and the Ecology Center, 5230 Old Main Hill, Utah State University, Logan, Utah 84322-5230 USA
-2. Department of Wildland Resources and the Ecology Center, 5230 Old Main Hill, Utah State University, Logan, Utah 84322-5230 USA
-
-\newpage{}
+\textsf{\Large{\textbf{Do we need demographic data to forecast population \\ responses to climate change?}}}
+
+\renewcommand*{\thefootnote}{\fnsymbol{footnote}}
+
+\vspace{1em}
+
+\textsf{\normalsize{\textbf{Andrew T. Tredennick\textsuperscript{1}\footnote{Corresponding author: atredenn@gmail.com}, Mevin B. Hooten\textsuperscript{2}, and Peter B. Adler\textsuperscript{1}}}}
+
+\vspace{1em}
+
+\textsf{\textit{\small{\textsuperscript{1}Department of Wildland Resources and the Ecology Center, 5230 Old Main Hill, Utah State University, Logan, Utah 84322, USA; \textsuperscript{2}U.S.
+Geological Survey, Colorado Cooperative Fish and Wildlife Research Unit, Department of Fish, Wildlife and Conservation Biology, Colorado State University, Fort Collins, CO 80523-1403, USA}}}
+
+\allsectionsfont{\normalfont\sffamily\bfseries}
+\renewcommand*{\thefootnote}{\arabic{footnote}}
 
@@
 -------
-1.
-Rapid climate change has generated growing interest in forecasts of future population trajectories.
+1. Rapid climate change has generated growing interest in forecasts of future population trajectories.
 Traditional population models, typically built using detailed demographic observations from one study site, can address climate change impacts at one location, but are difficult to scale up to the landscape and regional scales relevant to management decisions.
-An alternative is to build models using population-level data that is much easier to collect over broad spatial scales than individual-level data.
+An alternative is to build models using population-level data that are much easier to collect over broad spatial scales than individual-level data.
 However, such models ignore the fact that climate drives population growth through its influence on individual performance.
-2.
-We test the consequences of aggregating individual responses when forecasting climate change impacts on two perennial grass species (*Pascopyrum smithii* and *Sprobolis flexuosus*).
+
+2. Here, we test the consequences of aggregating individual responses when forecasting climate change impacts on four perennial grass species in a semi-arid grassland in Montana, USA.
 We parameterized two population models for each species, one based on individual-level data (survival, growth and recruitment) and one on population-level data (percent cover), and compared their accuracy, precision, and sensitivity to climate variables.
 For both models we used Bayesian ridge regression to identify the optimal predictive model in terms of climate covariate strengths.
-3.
-The individual-level model was more accurate and precise than the aggregated model when predicting out-of-sample observations.
+
+3. The individual-level model was more accurate and precise than the aggregated model when predicting out-of-sample observations.
 When comparing climate effects from both models, the population-level model missed important climate effects from at least one vital rate for each species.
 Increasing the sample size at the population-level would not necessarily reduce forecast uncertainty; the way to reduce uncertainty is to capture unique climate dependence of individual vital rates.
-4.
-Our analysis indicates that there is no shortcut to forecasting climate change impacts on plant populations — detailed demographic data is essential.
+
+4. *Synthesis.* Our analysis indicates that there is no shortcut to forecasting climate change impacts on plant populations — detailed demographic data are essential.
 Despite the superiority of the individual-level model, the forecasts it generated still were too uncertain to be useful for decision-makers.
@@
 
-*Key words: forecasting, climate change, grassland, integral projection model, population model*
-
-\newpage{}
+\textsf{\textbf{Key-words:}} forecasting, climate change, grassland, integral projection model, population model
 
@@
 
-### Model fitting
+### Model fitting and stastical regularization
+####  Model fitting
 Our Bayesian approach to fitting the vital rate models required choosing appropriate priors for unknown parameters and deciding which, if any, of those priors should be hierarchical.
@@
 We report the posterior mean, standard deviation, and 95% Bayesian Credible Intervals for every parameter of each model for each species in Appendix B.
+
+#### Statistical regularization: Bayesian ridge regression
+Our objective is to model the response of our focal grass species to interannual variation in climate, even if those responses are weak.
+Therefore, we avoid selecting among models with all possible combinations of climate covariates, and instead use Bayesian ridge regression to regulate, or constrain, the posterior distributions of each climate covariate [@Hooten2015].
+Ridge regression is a specific application of statistical regularization that seeks to optimize model generality by trading off bias and variance.
+As the name implies, statistical regularization involves the use of a regulator that constrains an optimization.
+The natural regulator in a Bayesian application is the prior on the coefficient of interest.
+Each of our statistical models includes the effects of climate covariates via the term $\textbf{c}'\boldsymbol{\theta}$ with prior $\boldsymbol{\theta} \sim \text{Normal}(\boldsymbol{\mu}_\theta, \sigma_\theta^2)$.
+Since we standardized all climate covariates, including interaction terms, to have mean zero and variance one we can set $\boldsymbol{\mu}_\theta=0$, letting $\sigma_\theta^2$ serve as the regulator that can shrink covariates toward zero -- the smaller the prior variance, the more $\theta$ is shrunk toward zero, and the stronger the penalty [@Hooten2015].
+
+To find the optimal penalty (i.e., optimal value of the hyperparameter $\sigma_\theta^2$), we fit each statistical model with a range of values for $\sigma_\theta^2$ and compared predictive scores from leave-one-year-out cross-validation.
+We performed the grid search over 24 evenly-spaced values of $\sigma_\theta^2$, ranging from $\sigma_\theta^2 = 0.01$ to $\sigma_\theta^2 = 2.25$.
+For each statistical model and each species we fit $13\times24=312$ models (13 years to leave out for cross-validation and 24 values of $\sigma_\theta^2$) -- a total of 4992 models.
+We calculated the log pointwise predictive density (*lppd*) to score each model's ability to predict the left-out data.
+Thus, for training data $y_{train}$ and held-out data $y_{hold}$ at a given value of  $\sigma_\theta^2$ across all MCMC samples $s=1,2,...,S$ and all hold outs of data from year *t* to year *T*, *lppd* is
+
+\begin{equation}
+\text{lppd}_{\text{CV}} = \sum_{t=1}^T \text{log}_e \int [y_{t,hold}|\theta][\theta|y_{train}]d\theta,
+\end{equation}
+
+and computed as
+
+\begin{equation}
+\sum_{t=1}^T \text{log}_e \left(\frac{1}{S} \sum^S_{s=1}(y_{t,hold}|\theta_{ts})  \right).
+\end{equation}
+
+We chose the optimal prior variance for each species-statistical model combination as the one that produced the highest *lppd* and then fit each species-statistical model combination using the full data set for each species and the optimal prior variance.
+In some cases *lppd* was highest at the highest (most vague) value of prior variance, in which case we explored a slightly expanded range of prior variances (e.g., $\sigma_\theta^2$>2.25).
 

### 2016-03-23 analysis/ipm/vitalRateRegs/PASM_reducedmodel/growth/get_plot_lppd.R

@@ -20,7 +20,7 @@ library(ggplot2)
 ####
 ####  Loop Through *.RDS Files and Load LPPD Results
 ####
-do_spp <- "BOGR"
+do_spp <- "POSE"
 hpc_dir <- "/Volumes/A02046115/growth_oos/"
 all_files <- list.files(hpc_dir)
 lppd_files <- all_files[grep(".RDS", all_files)]

### 2016-03-24 manuscript/manuscript.Rmd
udpates ms fig 1 and formatting

@@
 
+\rule{\textwidth}{1pt}
+
 Summary
 -------
-1. Rapid climate change has generated growing interest in forecasts of future population trajectories.
+**1.** Rapid climate change has generated growing interest in forecasts of future population trajectories.
 Traditional population models, typically built using detailed demographic observations from one study site, can address climate change impacts at one location, but are difficult to scale up to the landscape and regional scales relevant to management decisions.
-An alternative is to build models using population-level data that are much easier to collect over broad spatial scales than individual-level data.
+
+**2.** An alternative is to build models using population-level data that are much easier to collect over broad spatial scales than individual-level data.
 However, such models ignore the fact that climate drives population growth through its influence on individual performance.
 
-2. Here, we test the consequences of aggregating individual responses when forecasting climate change impacts on four perennial grass species in a semi-arid grassland in Montana, USA.
+**3.** Here, we test the consequences of aggregating individual responses when forecasting climate change impacts on four perennial grass species in a semi-arid grassland in Montana, USA.
 We parameterized two population models for each species, one based on individual-level data (survival, growth and recruitment) and one on population-level data (percent cover), and compared their accuracy, precision, and sensitivity to climate variables.
@@
 
-3. The individual-level model was more accurate and precise than the aggregated model when predicting out-of-sample observations.
+**4.** The individual-level model was more accurate and precise than the aggregated model when predicting out-of-sample observations.
 When comparing climate effects from both models, the population-level model missed important climate effects from at least one vital rate for each species.
@@
 
-4. *Synthesis.* Our analysis indicates that there is no shortcut to forecasting climate change impacts on plant populations — detailed demographic data are essential.
+**5.** *Synthesis.* Our analysis indicates that there is no shortcut to forecasting climate change impacts on plant populations — detailed demographic data are essential.
 Despite the superiority of the individual-level model, the forecasts it generated still were too uncertain to be useful for decision-makers.
@@
 
+Data accessibility
+------------------
+Chart-quadrat data available from *Ecological Archives* at http://esapubs.org/archive/ecol/E092/143/.
+All code to reproduce our work is available on Dryad (link) and GitHub (https://github.com/atredennick/MicroMesoForecast).
+
 \pagebreak{}
@@
 --------------
-Figure 1.
-Time series of average percent cover over all quadrats for our four focal species: *Bouteloua gracilis* (BOGR), *Hesperostipa comata* (HECO), *Pascopyrum smithii* (PASM), and *Poa secunda* (POSE).
+**Figure 1.** Time series of average percent cover over all quadrats for our four focal species: *Bouteloua gracilis* (BOGR), *Hesperostipa comata* (HECO), *Pascopyrum smithii* (PASM), and *Poa secunda* (POSE).
 Light grey lines show trajectories of individual quadrats.
@@
 Light grey lines show trajectories of individual quadrats.
-Note the different y-axis scales across panels.", fig.height=3, fig.width=8.5, eval=F}
+Note the different y-axis scales across panels.", fig.height=2, fig.width=8.5, eval=T}
+library(ggthemes)
 obs_data <- read.csv("../analysis/speciesData/quadAllCover.csv")
@@
   xlab("Year")+
-  theme_bw()+
+  theme_few()+
   facet_wrap("Species", ncol=4, nrow=1, scales = "free")+

### 2016-03-24 manuscript/manuscript.Rmd
updates figures

@@
 
-\textsf{\normalsize{\textbf{Andrew T. Tredennick\textsuperscript{1}\footnote{Corresponding author: atredenn@gmail.com}, Mevin B. Hooten\textsuperscript{2}, and Peter B. Adler\textsuperscript{1}}}}
+\textsf{\normalsize{\textbf{Andrew T.
+Tredennick\textsuperscript{1}\footnote{Corresponding author: E-mail: atredenn@gmail.com}, Mevin B.
+Hooten\textsuperscript{2}, and Peter B.
+Adler\textsuperscript{1}}}}
 
@@
 Second, accurately detecting climate signals will take even longer time series.
-Recent theoretical work on detecting climate signals in noisy data suggests that even advanced approaches to parameter fitting require 20-25 year time series [@Tellerinpress].
+Recent theoretical work on detecting climate signals in noisy data suggests that even advanced approaches to parameter fitting require 20-25 year time series [@Teller2016].
 Third, ecologists need a stronger commitment to reporting uncertainty.
@@
 
-```{r figure_2,dependson="plot-options", fig.cap="Work flow of the data aggregation, model fitting, and population simulating.", fig.height=6, eval=F}
+\newpage{}
+
+```{r figure_2,dependson="plot-options", fig.cap="Work flow of the data aggregation, model fitting, and population simulating.", fig.height=4, eval=T}
 library(png)
@@
 ```
+
+\newpage{}
 
@@
 
-\setcounter{page}{24}
+\singlespace{}
 

### 2016-03-24 manuscript/manuscript.Rmd
re-formats for Ecology

@@
    - \usepackage{setspace}
-   - \usepackage{times}
    - \usepackage{sectsty}
@@
 ---
-\textsf{\Large{\textbf{Do we need demographic data to forecast population \\ responses to climate change?}}}
-
 \renewcommand*{\thefootnote}{\fnsymbol{footnote}}
 
-\vspace{1em}
-
-\textsf{\normalsize{\textbf{Andrew T.
-Tredennick\textsuperscript{1}\footnote{Corresponding author: E-mail: atredenn@gmail.com}, Mevin B.
-Hooten\textsuperscript{2}, and Peter B.
-Adler\textsuperscript{1}}}}
-
-\vspace{1em}
-
-\textsf{\textit{\small{\textsuperscript{1}Department of Wildland Resources and the Ecology Center, 5230 Old Main Hill, Utah State University, Logan, Utah 84322, USA; \textsuperscript{2}U.S.
-Geological Survey, Colorado Cooperative Fish and Wildlife Research Unit, Department of Fish, Wildlife and Conservation Biology, Colorado State University, Fort Collins, CO 80523-1403, USA}}}
-
+\begin{centering}
+
+\textbf{\Large{Do we need demographic data to forecast population responses to climate change?}}
+
+\textsc{\normalsize{Andrew T. Tredennick\footnote{Corresponding author: atredenn@gmail.com}\textsuperscript{1}, Mevin B. Hooten\textsuperscript{2,3,4}, and Peter B. Adler\textsuperscript{1}}}
+
+\textit{\small{\textsuperscript{1}Department of Wildland Resources and the Ecology Center, 5230 Old Main Hill, Utah State University, Logan, Utah 84322 USA}} \\
+\textit{\small{\textsuperscript{2}U.S. Geological Survey, Colorado Cooperative Fish and Wildlife Research Unit, Colorado State University, Fort Collins, CO 80523 USA}} \\
+\textit{\small{\textsuperscript{3}Department of Fish, Wildlife, and Conservation Biology, Colorado State University, Fort Collins, CO 80523 USA}} \\
+\textit{\small{\textsuperscript{4}Department of Statistics, Colorado State University, Fort Collins, CO 80523 USA}}
+
+\end{centering}
+
+\renewcommand*{\thefootnote}{\arabic{footnote}}
+\setcounter{footnote}{0}
+
+<!-- For J. Ecol
 \allsectionsfont{\normalfont\sffamily\bfseries}
-\renewcommand*{\thefootnote}{\arabic{footnote}}
+-->
 
@@
 
-\rule{\textwidth}{1pt}
-
-Summary
--------
-**1.** Rapid climate change has generated growing interest in forecasts of future population trajectories.
+\begin{abstract}
+Rapid climate change has generated growing interest in forecasts of future population trajectories.
 Traditional population models, typically built using detailed demographic observations from one study site, can address climate change impacts at one location, but are difficult to scale up to the landscape and regional scales relevant to management decisions.
-
-**2.** An alternative is to build models using population-level data that are much easier to collect over broad spatial scales than individual-level data.
+An alternative is to build models using population-level data that are much easier to collect over broad spatial scales than individual-level data.
 However, such models ignore the fact that climate drives population growth through its influence on individual performance. 
-
-**3.** Here, we test the consequences of aggregating individual responses when forecasting climate change impacts on four perennial grass species in a semi-arid grassland in Montana, USA.
+Here, we test the consequences of aggregating individual responses when forecasting climate change impacts on four perennial grass species in a semi-arid grassland in Montana, USA.
 We parameterized two population models for each species, one based on individual-level data (survival, growth and recruitment) and one on population-level data (percent cover), and compared their accuracy, precision, and sensitivity to climate variables.
 For both models we used Bayesian ridge regression to identify the optimal predictive model in terms of climate covariate strengths. 
-
-**4.** The individual-level model was more accurate and precise than the aggregated model when predicting out-of-sample observations.
+The individual-level model was more accurate and precise than the aggregated model when predicting out-of-sample observations.
 When comparing climate effects from both models, the population-level model missed important climate effects from at least one vital rate for each species.
 Increasing the sample size at the population-level would not necessarily reduce forecast uncertainty; the way to reduce uncertainty is to capture unique climate dependence of individual vital rates. 
-
-**5.** *Synthesis.* Our analysis indicates that there is no shortcut to forecasting climate change impacts on plant populations — detailed demographic data are essential.
+Our analysis indicates that there is no shortcut to forecasting climate change impacts on plant populations — detailed demographic data are essential.
 Despite the superiority of the individual-level model, the forecasts it generated still were too uncertain to be useful for decision-makers.
@@
 
-\textsf{\textbf{Key-words:}} forecasting, climate change, grassland, integral projection model, population model
+\textit{Key words: forecasting, climate change, grassland, integral projection model, population model}
+
+\end{abstract}
 

### 2016-03-29 manuscript/appC.tex
updates collated qbm 1-step oos

--- 2015-03-05 manuscript/suppMat.tex
@@
 \usepackage{setspace}
-\doublespacing
 
@@
 
-\section{Supplementary Information: Do we need detailed demographic data
-to forecast population responses to climate
-change?}\label{supplementary-information-do-we-need-detailed-demographic-data-to-forecast-population-responses-to-climate-change}
+\renewcommand\thefigure{C\arabic{figure}}
 
-\subsubsection{Andrew T. Tredennick and Peter B.
-Adler}\label{andrew-t.-tredennick-and-peter-b.-adler}
+\section{Appendix C: Effect of recruitment on IPM
+uncertainty}\label{appendix-c-effect-of-recruitment-on-ipm-uncertainty}
 
-\emph{Andrew T. Tredennick
-(\href{mailto:atredenn@gmail.com}{\href{mailto:atredenn@gmail.com}{atredenn@gmail.com}}),
-Department of Wildland Resources and the Ecology Center, Utah State
-University, Logan, UT}
-
-\emph{Peter B. Adler, Department of Wildland Resources and the Ecology
-Center, Utah State University, Logan, UT}
+In Fig. 4 of the main text where we show the uncertainty around our
+models' forecasts to a 1\% change in climate covariates, the obvious
+outlier is \emph{Pascopyrum smithii} (PASM). The uncertainty around PASM
+forecasts from the integral projection model is orders of magnitude
+larger than for any other species, and is even larger than the
+uncertainty around the quad-based model forecasts. It turns out that the
+estimated dispersion parameter for the negative binomial distribution is
+much higher for PASM than for other species (Appendix B). This is
+because PASM has higher recruitment rates than the other species and the
+relationship between number of recruits and quadrat cover (see main
+text) is weak. So, when we draw parameters from the MCMC chain for our
+simulations, even larger values for the dispersion parameter are
+possible. Thus, there can be very large ``boom'' years where recruitment
+is high. When we allow all parameters to vary except those associated
+with recruitment, uncertainty around PASM decreases to levels similar to
+those associated with other species (Fig. C1).
 
@@
 \centering
-\includegraphics{components/figure/manuscript-figure_S1.pdf}
-\caption{Posterior means (vertical ticks), 75\% credible intervals
-(heavy lines), and 95\% credible intervals (light lines) of climate
-effects on growth at both levels of inferences. The dashed vertical line
-is at 0, indicating no effect.}
+\includegraphics{appC/figure/appC-plot_it.pdf}
+\caption{Mean (points) and 90\% quantiles (errorbars) for the
+proportional difference between baseline simulations (using observed
+climate) and the climate pertubation simulation on the x-axis. We
+calculated proportional difference as log(perturbed climate cover) -
+log(observed climate cover), where `perturbed' and `observed' refer to
+the climate time series used to drive interannual variation in the
+simulations. Model error and parameter uncertainty were propagated
+through the simulation phase for all vital rates except recruitment.}
 \end{figure}

### 2016-03-29 manuscript/appD.tex
updates collated qbm 1-step oos

--- 2015-03-05 manuscript/suppMat.tex
@@
 
-\section{Supplementary Information: Do we need detailed demographic data
-to forecast population responses to climate
-change?}\label{supplementary-information-do-we-need-detailed-demographic-data-to-forecast-population-responses-to-climate-change}
+\renewcommand\thefigure{D\arabic{figure}}
 
-\subsubsection{Andrew T. Tredennick and Peter B.
-Adler}\label{andrew-t.-tredennick-and-peter-b.-adler}
+\section{Appendix D: Model forecasts without parameter
+uncertainty}\label{appendix-d-model-forecasts-without-parameter-uncertainty}
 
-\emph{Andrew T. Tredennick
-(\href{mailto:atredenn@gmail.com}{\href{mailto:atredenn@gmail.com}{atredenn@gmail.com}}),
-Department of Wildland Resources and the Ecology Center, Utah State
-University, Logan, UT}
+In the main text we show results from model forecasts where we included
+full model and parameter uncertainty (Fig. 6). However, the typical
+approach in ecology is to use point estimates of model parameters and
+only include model error. For example, to predict population size
+\emph{N} at time \emph{t+1} one could do the following:
 
-\emph{Peter B. Adler, Department of Wildland Resources and the Ecology
-Center, Utah State University, Logan, UT}
+\begin{align}
+\mu_{t+1} &= f(N_{t}) \\
+N_{t+1} &\sim \text{Normal}(\mu_{t+1}, \sigma^2)
+\end{align}
+
+where $f(N_t)$ is some determinstic process (e.g., a linear regression
+model) and $\sigma^2$ is the model error. Simulating populations in such
+a way ignores variation in the parameters of $f(N_t)$. When we take this
+approach to making forecasts of equilibrium cover under 1\% climate
+perturbations the IPM and QBM make opposing predictions if we simply
+look at the mean projection (Fig. D1). For both models, precision is
+also artificially high (Fig. D2 and D3), since we know that parameter
+uncertainty reduces forecast precision considerably (Fig. 6 in the main
+text). All three presentations of model results (mean projection, mean
+projection with model error, and forecasts with full uncertainty) can be
+found in the literature, which is why we present them here as well.
+However, as we state in the main text, true forecasts should only be
+presented as we did in Fig. 6 of the main text.
 
@@
 \centering
-\includegraphics{components/figure/manuscript-figure_S1.pdf}
-\caption{Posterior means (vertical ticks), 75\% credible intervals
-(heavy lines), and 95\% credible intervals (light lines) of climate
-effects on growth at both levels of inferences. The dashed vertical line
-is at 0, indicating no effect.}
+\includegraphics{appD/figure/appD-figure_d1.pdf}
+\caption{Projected proportional changes in equilibrium cover caused by a
+1\% increase in observed precipitation (+ppt), temperature (+temp), or
+both (+ppt\&temp) as predicted by the individual-level IPM and the
+population-level QBM using mean parameter values.}
+\end{figure}
+
+\begin{figure}[htbp]
+\centering
+\includegraphics{appD/figure/appD-figure_d2.pdf}
+\caption{Projected mean (points) proportional changes in equilibrium
+cover and associated 90\% quantiles (errorbars) caused by a 1\% increase
+in observed precipitation (+ppt), temperature (+temp), or both
+(+ppt\&temp) as predicted by the individual-level IPM using mean
+parameter values. Dashed line indicates proportional change of 0.}
+\end{figure}
+
+\begin{figure}[htbp]
+\centering
+\includegraphics{appD/figure/appD-figure_d3.pdf}
+\caption{Projected mean (points) proportional changes in equilibrium
+cover and associated 90\% quantiles (errorbars) caused by a 1\% increase
+in observed precipitation (+ppt), temperature (+temp), or both
+(+ppt\&temp) as predicted by the population-level QBM using mean
+parameter values. Dashed line indicates proportional change of 0.}
 \end{figure}

### 2016-04-02 analysis/ipm/vitalRateRegs/regularization/growth/README.Rmd
file cleanup; adds HPC readmes; adds stan files for regressions

### README for growth ridge-regression scripts

*The files in this directory are meant to be run an a High Performance Computing system, NOT on a PC.*

* The ``*Crowding.csv`` files contain information on the crowding experienced by a focal genet at a specific time.

* The ``Climate.csv`` file contains aggregated climate covarites for each year.

* Species-specific subdirectories contain the longitudinal demographic data for genet growth.

* There are three ``R`` files:
    + ``growth_oos_launcher.R``: reads in demographic growth data and crowding data; compiles Stan model for MCMC fitting; calls ``growth_fxns.R`` to fit training data and calculate the log pointwise predictive density (*lppd*) for the model predicting out-of-sample data. Out-of-sample data is a left-out year from the data.
    + ``growth_fxns.R``: contains functions to call Stan models given a particular left-out year and prior standard deviation for climate covariates.
    + ``waic_fxns.R``: function for calculating various metrics of model predictive ability from Vehtari and Gelman (2014; http://www.stat.columbia.edu/~gelman/research/unpublished/waic_stan.pdf). We are interested in the summed (total) log pointwise densities returned by the function.

* The ``.stan`` file contains a Stan model for fitting the regression via ‘No-U-Turn’ Hamiltonian Monte Carlo and to predict out-of-sample data.

* The ``.sh`` file is a bach script for submitting an array job to the Utah State University HPC system using SLURM commands. We run each combination of prior standard deviation and left-out year in parallel.

### 2016-04-02 analysis/ipm/vitalRateRegs/regularization/recruitment/README.Rmd
file cleanup; adds HPC readmes; adds stan files for regressions

--- 2016-04-02 analysis/ipm/vitalRateRegs/regularization/growth/README.Rmd
@@
-### README for growth ridge-regression scripts
+### README for recruitment ridge-regression scripts
 
@@
 
-* The ``*Crowding.csv`` files contain information on the crowding experienced by a focal genet at a specific time.
-
 * The ``Climate.csv`` file contains aggregated climate covarites for each year.
 
-* Species-specific subdirectories contain the longitudinal demographic data for genet growth.
+* Species-specific subdirectories contain the longitudinal demographic data for number of recruits in each quadrat-year.
 
 * There are three ``R`` files:
-    + ``growth_oos_launcher.R``: reads in demographic growth data and crowding data; compiles Stan model for MCMC fitting; calls ``growth_fxns.R`` to fit training data and calculate the log pointwise predictive density (*lppd*) for the model predicting out-of-sample data.
+    + ``recruitment_oos_launcher.R``: reads in demographic growth data; compiles Stan model for MCMC fitting; calls ``recruitment_fxns.R`` to fit training data and calculate the log pointwise predictive density (*lppd*) for the model predicting out-of-sample data.
 Out-of-sample data is a left-out year from the data.
-    + ``growth_fxns.R``: contains functions to call Stan models given a particular left-out year and prior standard deviation for climate covariates.
+    + ``recruitment_fxns.R``: contains functions to call Stan models given a particular left-out year and prior standard deviation for climate covariates.
     + ``waic_fxns.R``: function for calculating various metrics of model predictive ability from Vehtari and Gelman (2014; http://www.stat.columbia.edu/~gelman/research/unpublished/waic_stan.pdf).

### 2016-04-02 analysis/ipm/vitalRateRegs/regularization/survival/README.Rmd
file cleanup; adds HPC readmes; adds stan files for regressions

--- 2016-04-02 analysis/ipm/vitalRateRegs/regularization/growth/README.Rmd
@@
-### README for growth ridge-regression scripts
+### README for survival ridge-regression scripts
 
@@
 
-* Species-specific subdirectories contain the longitudinal demographic data for genet growth.
+* Species-specific subdirectories contain the longitudinal demographic data for genet survival.
 
 * There are three ``R`` files:
-    + ``growth_oos_launcher.R``: reads in demographic growth data and crowding data; compiles Stan model for MCMC fitting; calls ``growth_fxns.R`` to fit training data and calculate the log pointwise predictive density (*lppd*) for the model predicting out-of-sample data.
+    + ``survival_oos_launcher.R``: reads in demographic survival data and crowding data; compiles Stan model for MCMC fitting; calls ``survival_fxns.R`` to fit training data and calculate the log pointwise predictive density (*lppd*) for the model predicting out-of-sample data.
 Out-of-sample data is a left-out year from the data.
-    + ``growth_fxns.R``: contains functions to call Stan models given a particular left-out year and prior standard deviation for climate covariates.
+    + ``survival_fxns.R``: contains functions to call Stan models given a particular left-out year and prior standard deviation for climate covariates.
     + ``waic_fxns.R``: function for calculating various metrics of model predictive ability from Vehtari and Gelman (2014; http://www.stat.columbia.edu/~gelman/research/unpublished/waic_stan.pdf).

### 2016-04-02 analysis/ipm/vitalRateRegs/growth/growthAllSpp_STAN.R

@@ -66,86 +66,17 @@ crowd <- rbind(c1,c2,c3,c4)
 # Merge crowding and growth data
 growD_all <- merge(growD, crowd, by=c("species", "X"))
 
-model_string <- "
-data{
-  int<lower=0> N; // observations
-  int<lower=0> Yrs; // years
-  int<lower=0> yid[N]; // year id
-  int<lower=0> Covs; // climate covariates
-  int<lower=0> G; // groups
-  int<lower=0> gid[N]; // group id
-  vector[N] Y; // observation vector
-  matrix[N,Covs] C; // climate matrix
-  vector[N] X; // size vector
-  matrix[N,2] W; // crowding matrix
-  real tauclim; // prior stdev
-}
-parameters{
-  real a_mu;
-  vector[Yrs] a;
-  real b1_mu;
-  vector[Yrs] b1;
-  vector[Covs] b2;
-  vector[2] w;
-  real gint[G];
-  real tau;
-  real tauSize;
-  real<lower=0> sig_a;
-  real<lower=0> sig_b1;
-  real<lower=0> sig_G;
-}
-transformed parameters{
-  real mu[N];
-  real<lower=0> sigma[N];
-  vector[N] climEff;
-  vector[N] crowdEff;
-  climEff <- C*b2;
-  crowdEff <- W*w;
-  for(n in 1:N){
-    mu[n] <- a[yid[n]] + gint[gid[n]] + b1[yid[n]]*X[n] + crowdEff[n] + climEff[n];
-    sigma[n] <- sqrt((fmax(tau*exp(tauSize*mu[n]), 0.0000001)));  
-  }
-}
-model{
-  // Priors
-  a_mu ~ normal(0,100);
-  w ~ normal(0,10);
-  b1_mu ~ normal(0,100);
-  tau ~ normal(0,100);
-  tauSize ~ normal(0,100);
-  sig_a ~ cauchy(0,2);
-  sig_b1 ~ cauchy(0,2);
-  sig_G ~ cauchy(0,2);
-  b2 ~ normal(0, tauclim);
-  for(g in 1:G)
-    gint[g] ~ normal(0, sig_G);
-  for(y in 1:Yrs){
-    a[y] ~ normal(a_mu, sig_a);
-    b1[y] ~ normal(b1_mu, sig_b1);
-  }
-
-  // Likelihood
-  Y ~ normal(mu, sigma);
-}
-"
-
 ## Compile model outside of loop
-growD <- subset(growD_all, species==sppList[1])
[…]

### 2016-04-02 analysis/ipm/vitalRateRegs/regularization/get_plot_lppd.R

@@ -9,6 +9,13 @@
 
 rm(list=ls())
 
+
+##  Set Species and Vital Rate
+do_spp <- "PASM"
+hpc_dir <- "/Volumes/A02046115/surv_oos/" # for fetching lppd RDS files
+vital_name <- "Survival" # for plotting
+
+
 ####
 ####  Load Libraries
 ####
@@ -17,11 +24,10 @@ library(reshape2)
 library(ggplot2)
 
 
+
 ####
 ####  Loop Through *.RDS Files and Load LPPD Results
 ####
-do_spp <- "POSE"
-hpc_dir <- "/Volumes/A02046115/rec_oos/"
 all_files <- list.files(hpc_dir)
 lppd_files <- all_files[grep(".RDS", all_files)]
 lppd_files <- lppd_files[grep(do_spp, lppd_files)]
@@ -34,7 +40,7 @@ for(i in 1:length(lppd_files)){
   lppd_mat[i,2] <- as.numeric(readRDS(paste0(hpc_dir, tmp)))
 }
 lppd_mat <- as.data.frame(lppd_mat[order(lppd_mat[,1]),])
-lppd_mat <- lppd_mat[complete.cases(lppd_mat),] 
+
 
 
 ####
@@ -73,7 +79,7 @@ score_cv_vec <- apply(lppd_cast,1,sum)
 plot_df <- data.frame(betavar = sd_vec^2,
                       lppd = score_cv_vec)
 
-opt_var <- plot_df[which(plot_df$lppd==max(plot_df$lppd, na.rm = T)), "betavar"]
+opt_var <- plot_df[which(plot_df$lppd==max(plot_df$lppd)), "betavar"]
 
 ggplot(data=plot_df, aes(x=betavar, y=lppd))+
   geom_line(size=1)+
@@ -82,6 +88,6 @@ ggplot(data=plot_df, aes(x=betavar, y=lppd))+
   xlab(bquote(sigma[beta]^2))+
   ylab("Log Predictive Score (lppd)")+
   theme_bw()+
-  ggtitle(paste(do_spp,"Recruitment"))
+  ggtitle(paste(do_spp,vital_name))
 
 

### 2016-04-02 analysis/ipm/vitalRateRegs/regularization/recruitment/import2ipm.R

@@ -1,112 +1,0 @@
-
-library(reshape2)
-library(plyr)
-fitthin <- data.frame(Iteration=NA, Chain=NA, Parameter=NA,
-                      value=NA, keep=NA, species=NA)
-for(ispp in spp_list){
-  fitlong <- readRDS(paste("../vitalRateRegs/recruitment/recruitment_stanmcmc_", ispp, ".RDS", sep=""))
-  fitlong$keep <- "no"
-  keepseq <- seq(from = 1, to = nrow(fitlong), by = 1)
-  fitlong[keepseq,"keep"] <- "yes"
-  tmp <- subset(fitlong, keep=="yes")
-  tmp$species <- ispp
-  fitthin <- rbind(fitthin,tmp)
-}
-fitthin <- fitthin[2:nrow(fitthin),]
-
-##  Break up MCMC into regression components
-# Climate effects
-climeff_rec <- fitthin[grep("b2", fitthin$Parameter),]
-
-# Yearly intercepts
-intercept_rec <- fitthin[grep("a", fitthin$Parameter),]
-intercept_rec <- subset(intercept_rec, Parameter!="a_mu")
-intercept_rec <- subset(intercept_rec, Parameter!="tau")
-intercept_rec$yearid <- substr(intercept_rec$Parameter, 3, length(intercept_rec$Parameter))
-intercept_rec$yearid <- unlist(strsplit(intercept_rec$yearid, split=']'))
-
-# Mean intercept
-interceptmu_rec <- fitthin[grep("a_mu", fitthin$Parameter),]
-
-# Group effects
-group_rec <- fitthin[grep("gint", fitthin$Parameter),]
-group_rec$groupid <- substr(group_rec$Parameter, 6, length(group_rec$Parameter))
-group_rec$groupid <- unlist(strsplit(group_rec$groupid, split=']'))
-
-# Get parent density-dependent effects
-densdep_rec <- fitthin[grep("dd", fitthin$Parameter),]
-
-# Get mixing fraction u
-mixfrac_rec <- fitthin[grep("u", fitthin$Parameter),]
-mixfrac_rec <- subset(mixfrac_rec, Parameter!="a_mu")
-
-# Get theta
-theta_rec <- fitthin[grep("theta", fitthin$Parameter),]
-
-# Get rid of big objects
-rm(list = c("tmp","fitthin","fitlong"))
-
-##  Define function to format survival coefficients
-getRecCoefs <- function(doYear, groupnum){
[…]

### 2016-04-02 analysis/ipm/vitalRateRegs/regularization/recruitment/recruitmentPASM_indvar_STAN.R

@@ -1,354 +1,0 @@
-#!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!#
-# Set working directory to location of this source file #
-#!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!#
-
-#clear everything, just to be safe 
-rm(list=ls(all=TRUE))
-
-#load libraries
-library(rstan)
-library(parallel)
-library(reshape2)
-
-## Set do_year for validation from command line prompt
-args <- commandArgs(trailingOnly = F)
-myargument <- args[length(args)]
-myargument <- sub("-","",myargument)
-do_species <- as.numeric(myargument)
-
-sppList=sort(c("BOGR","HECO","PASM","POSE"))
-do_species <- 2
-
-####
-#### Read in data by species and make one long data frame -------------
-####
-for(spp in 1:length(sppList)){
-  doSpp <- sppList[spp]
-  
-  if(doSpp == "BOGR"){
-    sppD <- read.csv(paste("../../../../speciesData/", doSpp, "/edited/recArea.csv", sep=""))
-    sppD$species <- doSpp 
-  }else{
-    sppD <- read.csv(paste("../../../../speciesData/", doSpp, "/recArea.csv", sep=""))
-    sppD$species <- doSpp 
-  }
-  
-  sppD$Group <- as.factor(substr(sppD$quad,1,1)) #add by Chengjin
-  sppD <- sppD[,c("quad","year","NRquad","totParea","Group")]
-  names(sppD)[3] <- paste("R.",sppList[spp],sep="")
-  names(sppD)[4] <- paste("cov.",sppList[spp],sep="")
-  if(spp==1){
-    D <- sppD
-  }else{
-    D <- merge(D,sppD,all=T)
-  }
-}
-D[is.na(D)]=0  # replace missing values 
-
-# calculate mean cover by group and year
-tmpD <- D[,c("quad","year","Group",paste("cov.",sppList,sep=""))]
-tmpD <- aggregate(tmpD[,4:NCOL(tmpD)],
-                  by=list("year"=tmpD$year,"Group"=tmpD$Group),FUN=mean)
-names(tmpD)[3:NCOL(tmpD)] <- paste("Gcov.",sppList,sep="")
-D <- merge(D,tmpD,all.x=T)
-
-
-
-
-###if using square transform, there is no need of the following code
[…]

### 2016-04-02 analysis/ipm/vitalRateRegs/regularization/recruitment/recruitment_reg_cv.R

@@ -1,353 +1,0 @@
-##
-##  R Script for Recruitment Ricker Model (for IPM)
-##
-##  Includes Model Selection via Bayesian Regularization
-##
-##  Author:       Andrew Tredennick
-##  Email:        atredenn@gmail.com
-##  Date created: 12-15-2015
-##
-##  NOTICE: 
-##  As is, this script will take ~20 minutes to run.
-##  Production runs (more MCMC iterations) required use of 
-##  Utah State University's High Performance Computing facility.
-##  See hpc_scripts/ directory for fully parallelized versions.
-##
-
-
-# Clear the workspace
-rm(list=ls())
-
-
-
-####
-####  Load libraries and subroutines
-####
-library(rstan); library(plyr)
-library(reshape2); library(ggmcmc)
-library(parallel); library(rlecuyer)
-library(snowfall)
-source("../../../../waic_fxns.R")
-
-
-
-####
-####  Load Climate Covariates and Survival Data
-####
-sppList=sort(c("BOGR","HECO","PASM","POSE"))
-do_species <- 4
-
-for(spp in 1:length(sppList)){
-  doSpp <- sppList[spp]
-  
-  if(doSpp == "BOGR"){
-    sppD <- read.csv(paste("../../../../speciesData/", doSpp, "/edited/recArea.csv", sep=""))
-    sppD$species <- doSpp 
-  }else{
-    sppD <- read.csv(paste("../../../../speciesData/", doSpp, "/recArea.csv", sep=""))
-    sppD$species <- doSpp 
-  }
-  
-  sppD$Group <- as.factor(substr(sppD$quad,1,1)) #add by Chengjin
-  sppD <- sppD[,c("quad","year","NRquad","totParea","Group")]
-  names(sppD)[3] <- paste("R.",sppList[spp],sep="")
-  names(sppD)[4] <- paste("cov.",sppList[spp],sep="")
-  if(spp==1){
-    D <- sppD
-  }else{
-    D <- merge(D,sppD,all=T)
-  }
-}
-D[is.na(D)]=0  # replace missing values 
-
-# calculate mean cover by group and year
-tmpD <- D[,c("quad","year","Group",paste("cov.",sppList,sep=""))]
-tmpD <- aggregate(tmpD[,4:NCOL(tmpD)],
[…]

### 2016-04-02 analysis/ipm/vitalRateRegs/regularization/survival/get_plot_lppd.R

@@ -1,86 +1,0 @@
-##
-##  R Script to load and collate CV Regularization Results
-##  from HPC user directory.
-##
-##  Author: Andrew Tredennick
-##  Email: atredenn@gmail.com
-##  Date created: 12-7-2015
-##
-
-rm(list=ls())
-
-####
-####  Load Libraries
-####
-library(plyr)
-library(reshape2)
-library(ggplot2)
-
-
-do_spp <- "PASM"
-
-####
-####  Loop Through *.RDS Files and Load LPPD Results
-####
-hpc_dir <- "/Volumes/A02046115/surv_oos/"
-all_files <- list.files(hpc_dir)
-lppd_files <- all_files[grep(".RDS", all_files)]
-lppd_files <- lppd_files[grep(do_spp, lppd_files)]
-lppd_mat <- matrix(NA, ncol=2, nrow=length(lppd_files))
-for(i in 1:length(lppd_files)){
-  tmp <- lppd_files[i]
-  tmpsplit <- unlist(strsplit(tmp, "[.]"))
-  tmp_grid_num <- unlist(strsplit(tmpsplit[1], "_"))[5]
-  lppd_mat[i,1] <- as.numeric(tmp_grid_num)
-  lppd_mat[i,2] <- as.numeric(readRDS(paste0(hpc_dir, tmp)))
-}
-lppd_mat <- as.data.frame(lppd_mat[order(lppd_mat[,1]),])
-
-
-
-####
-####  Recreate Sdev*CV Grid
-####
-n.beta <- 24
-sd_vec <- seq(0.1,1.5,length.out = n.beta)
-K <- 13 # number of years
-cv.s2.grid <- expand.grid(1:n.beta,1:K)
-n.grid <- dim(cv.s2.grid)[1]
-fold.idx.mat <- matrix(1:13,ncol=K)
-grid_df <- as.data.frame(cv.s2.grid)
-grid_df$id <- rownames(grid_df)
-
-
-##  Find out which grid ids didn't run on HPC; i.e., are missing
-"%w/o%" <- function(x, y) x[!x %in% y] # x without y
-allgrids <- c(1:n.grid)
-rungrids <- lppd_mat[,1]
-missedgrids <- allgrids %w/o% rungrids
-paste0(missedgrids,collapse=",")
-if(length(rungrids)+length(missedgrids)!=n.grid){ 
-  stop("number of grids not matching up") }
-if(length(missedgrids)>0){ 
[…]

### 2016-04-02 analysis/ipm/vitalRateRegs/regularization/survival/import2ipm.R

@@ -1,119 +1,0 @@
-# ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ #
-# Test script to pull in growth parameters for IPM #
-# ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ #
-
-#Several arguments are required
-#' @param doYear  Specific climate year to pull year random effects (19xx)
-#' @param mcDraw  A numeric scalar or vector for the row(s) of MCMC to draw parameters from 
-#' @param doSpp   A character scalar for the current speicies
-#' @param group   A numeric scalar with group indicator
-
-library(reshape2)
-library(plyr)
-
-fitthin <- data.frame(Iteration=NA, Chain=NA, Parameter=NA,
-                      value=NA, keep=NA, species=NA)
-for(ispp in spp_list){
-  fitlong <- readRDS(paste("../vitalRateRegs/survival/survival_stanmcmc_", ispp, ".RDS", sep=""))
-  fitlong$keep <- "no"
-  keepseq <- seq(from = 1, to = nrow(fitlong), by = 1)
-  fitlong[keepseq,"keep"] <- "yes"
-  tmp <- subset(fitlong, keep=="yes")
-  tmp$species <- ispp
-  fitthin <- rbind(fitthin,tmp)
-}
-fitthin <- fitthin[2:nrow(fitthin),]
-
-##  Break up MCMC into regression components
-# Climate effects
-climeff_surv <- fitthin[grep("b2", fitthin$Parameter),]
-
-# Yearly cover (size) effects
-coveff_surv <- fitthin[grep(glob2rx("b1[*]"), fitthin$Parameter),]
-coveff_surv$yearid <- substr(coveff_surv$Parameter, 4, length(coveff_surv$Parameter))
-coveff_surv$yearid <- unlist(strsplit(coveff_surv$yearid, split=']'))
-
-# Mean cover effect
-covermu_surv <- fitthin[grep("b1_mu", fitthin$Parameter),]
-
-# Yearly intercepts
-intercept_surv <- fitthin[grep("a", fitthin$Parameter),]
-intercept_surv <- subset(intercept_surv, Parameter!="a_mu")
-intercept_surv <- subset(intercept_surv, Parameter!="tau")
-intercept_surv$yearid <- substr(intercept_surv$Parameter, 3, length(intercept_surv$Parameter))
[…]

### 2016-04-02 analysis/ipm/vitalRateRegs/regularization/survival/survivalPASM_STAN.R

@@ -1,145 +1,0 @@
-##  The script takes 'do_species' as a command line prompt. So, e.g.,
-##    run as: "R CMD BATCH -1 survivalAllSpp_STAN.R" for species 1.
-
-#clear everything, just to be safe 
-rm(list=ls(all=TRUE))
-
-#load libraries
-library(rstan)
-library(parallel)
-
-## Set do_year for validation from command line prompt
-args <- commandArgs(trailingOnly = F)
-myargument <- args[length(args)]
-myargument <- sub("-","",myargument)
-do_species <- as.numeric(myargument)
-
-sppList=sort(c("BOGR","HECO","PASM","POSE"))
-do_species <- 3
-
-####
-#### Read in data by species and make one long data frame -------------
-####
-outD <- data.frame(X=NA,
-                   quad=NA,
-                   year=NA,
-                   trackID=NA,
-                   area=NA,
-                   survives=NA,
-                   age=NA,
-                   distEdgeMin=NA,
-                   allEdge=NA,
-                   species=NA)
-
-data_path <- "../../../../speciesData/" #on local machine
-# data_path <- "speciesData/" #on HPC server
-
-for(spp in 1:length(sppList)){
-  doSpp <- sppList[spp]
-  
-  if(doSpp == "BOGR"){
-    sppD <- read.csv(paste(data_path, doSpp, "/edited/survD.csv", sep=""))
-    sppD$species <- doSpp 
-  }else{
-    sppD <- read.csv(paste(data_path, doSpp, "/survD.csv", sep=""))
-    sppD$species <- doSpp 
-  }
-  outD <- rbind(outD, sppD)
-}
-
-survD <- outD[2:nrow(outD),]
-
-climD <- read.csv("../../../../weather/Climate.csv") #on local machine
-# climD <- read.csv("Climate.csv") #on HPC server
-clim_vars <- c("pptLag", "ppt1", "ppt2", "TmeanSpr1", "TmeanSpr2")
-climD[,clim_vars] <- scale(climD[,clim_vars], center = TRUE, scale = TRUE)
-climD$year <- climD$year-1900
-survD <- merge(survD,climD)
[…]

### 2016-04-02 analysis/ipm/vitalRateRegs/regularization/survival/survival_reg_cv.R

@@ -1,312 +1,0 @@
-##
-##  R Script for survival logistic model (for IPM)
-##
-##  Includes Model Selection via Bayesian Regularization
-##
-##  Author:       Andrew Tredennick
-##  Email:        atredenn@gmail.com
-##  Date created: 12-2-2015
-##
-##  NOTICE: 
-##  As is, this script will take ~5 hours to run.
-##  Production runs (more MCMC iterations) required use of 
-##  Utah State University's High Performance Computing facility.
-##
-
-
-# Clear the workspace
-rm(list=ls())
-
-
-####
-####  Load libraries and subroutines
-####
-library(rstan); library(plyr)
-library(reshape2); library(ggmcmc)
-library(parallel); library(rlecuyer)
-library(snowfall)
-source("../../../../waic_fxns.R")
-
-
-
-####
-####  Load Climate Covariates and Survival Data
-####
-sppList <- sort(c("BOGR","HECO","PASM","POSE"))
-do_species <- 3 # for PASM, for testing
-
-##  Read in data by species and make one long data frame
-outD <- data.frame(X=NA,
-                   quad=NA,
-                   year=NA,
-                   trackID=NA,
-                   area=NA,
-                   survives=NA,
-                   age=NA,
-                   distEdgeMin=NA,
-                   allEdge=NA,
-                   species=NA)
-
-data_path <- "../../../../speciesData/" #on local machine
-# data_path <- "speciesData/" #on HPC server
-
-for(spp in 1:length(sppList)){
-  doSpp <- sppList[spp]
-  
-  if(doSpp == "BOGR"){
-    sppD <- read.csv(paste(data_path, doSpp, "/edited/survD.csv", sep=""))
-    sppD$species <- doSpp 
-  }else{
-    sppD <- read.csv(paste(data_path, doSpp, "/survD.csv", sep=""))
-    sppD$species <- doSpp 
-  }
-  outD <- rbind(outD, sppD)
-}
-
-survD <- outD[2:nrow(outD),]
-
-##  Read in climate data
[…]

### 2016-04-02 analysis/ipm/vitalRateRegs/regularization/waic_fxns.R

@@ -1,24 +1,0 @@
-waic <- function(stanfit){
-  log_lik <- extract (stanfit, "log_lik")$log_lik
-  dim(log_lik) <- if (length(dim(log_lik))==1) c(length(log_lik),1) else
-    c(dim(log_lik)[1], prod(dim(log_lik)[2:length(dim(log_lik))]))
-  S <- nrow(log_lik)
-  n <- ncol(log_lik)
-  lpd <- log(colMeans(exp(log_lik)))
-  p_waic <- colVars(log_lik)
-  elpd_waic <- lpd - p_waic
-  waic <- -2*elpd_waic
-  loo_weights_raw <- 1/exp(log_lik-max(log_lik))
-  loo_weights_normalized <- loo_weights_raw/
-    matrix(colMeans(loo_weights_raw),nrow=S,ncol=n,byrow=TRUE)
-  loo_weights_regularized <- pmin (loo_weights_normalized, sqrt(S))
-  elpd_loo <- log(colMeans(exp(log_lik)*loo_weights_regularized)/
-                    colMeans(loo_weights_regularized))
-  p_loo <- lpd - elpd_loo
-  pointwise <- cbind(waic,lpd,p_waic,elpd_waic,p_loo,elpd_loo)
-  total <- colSums(pointwise)
-  se <- sqrt(n*colVars(pointwise))
-  return(list(waic=total["waic"], elpd_waic=total["elpd_waic"],
-              p_waic=total["p_waic"], elpd_loo=total["elpd_loo"], p_loo=total["p_loo"],
-              pointwise=pointwise, total=total, se=se))
-}

### 2016-04-02 analysis/ipm/vitalRateRegs/survival/import2ipm.R

@@ -15,7 +15,7 @@ library(ggmcmc)
 fitthin <- data.frame(Iteration=NA, Chain=NA, Parameter=NA,
                       value=NA, keep=NA, species=NA)
 for(ispp in spp_list){
-  fitlong <- ggs(readRDS(paste("../vitalRateRegs/survival/survival_stanmcmc_", ispp, ".RDS", sep="")))
+  fitlong <- readRDS(paste("../vitalRateRegs/survival/survival_stanmcmc_", ispp, ".RDS", sep=""))
   fitlong$keep <- "no"
   keepseq <- seq(from = 1, to = nrow(fitlong), by = 1)
   fitlong[keepseq,"keep"] <- "yes"

### 2016-04-02 analysis/ipm/vitalRateRegs/survival/import2ipm_means.R

@@ -5,7 +5,7 @@ library(ggmcmc)
 
 fitthin <- data.frame(Parameter=NA, value=NA, species=NA)
 for(ispp in spp_list){
-  fitlong <- ggs(readRDS(paste("../vitalRateRegs/survival/survival_stanmcmc_", ispp, ".RDS", sep="")))
+  fitlong <- readRDS(paste("../vitalRateRegs/survival/survival_stanmcmc_", ispp, ".RDS", sep=""))
   tmp <- ddply(fitlong, .(Parameter), summarise,
                value = mean(value))
   tmp$species <- ispp

### 2016-04-02 analysis/ipm/vitalRateRegs/survival/survivalAllSpp_STAN.R

@@ -18,7 +18,7 @@ myargument <- args[length(args)]
 myargument <- sub("-","",myargument)
 do_species <- as.numeric(myargument)
 sppList=sort(c("BOGR","HECO","PASM","POSE"))
-do_species <- 1
+do_species <- 1 # for testing; remove for HPC runs
 
 ####
 #### Read in data by species and make one long data frame -------------
@@ -74,78 +74,17 @@ colnames(crowd) <- c("W", "X", "species")
 # Merge crowding and growth data
 survD_all <- merge(survD, crowd, by=c("species", "X"))
 
-
-model_string <- "
-data{
-  int<lower=0> N; // observations
-  int<lower=0> Yrs; // years
-  int<lower=0> yid[N]; // year id
-  int<lower=0> Covs; // climate covariates
-  int<lower=0> G; // groups
-  int<lower=0> gid[N]; // group id
-  int<lower=0,upper=1> Y[N]; // observation vector
-  matrix[N,Covs] C; // climate matrix
-  vector[N] X; // size vector
-  matrix[N,2] W; // crowding matrix
-  real tauclim; // climate prior std dev
-}
-parameters{
-  real a_mu;
-  vector[Yrs] a;
-  real b1_mu;
-  vector[Yrs] b1;
-  vector[Covs] b2;
-  vector[2] w;
-  vector[G] gint;
-  real<lower=0> sig_a;
-  real<lower=0> sig_b1;
-  real<lower=0> sig_G;
-}
-transformed parameters{
-  real mu[N];
-  vector[N] climEff;
-  vector[N] crowdEff;
-  climEff <- C*b2;
-  crowdEff <- W*w;
-  for(n in 1:N){
-    mu[n] <- inv_logit(a[yid[n]] + gint[gid[n]] + b1[yid[n]]*X[n] + crowdEff[n] + climEff[n]);
-  }
-}
-model{
-  // Priors
-  a_mu ~ normal(0,100);
-  w ~ normal(0,100);
-  b1_mu ~ normal(0,100);
-  sig_a ~ cauchy(0,5);
-  sig_b1 ~ cauchy(0,5);
-  sig_G ~ cauchy(0,5);
-  gint ~ normal(0, sig_G);
-  b2 ~ normal(0, tauclim);
-  a ~ normal(a_mu, sig_a);
-  b1 ~ normal(b1_mu, sig_b1);
-
-  // Likelihood
-  Y ~ binomial(1,mu);
-}
-"
-
 ## Compile model 
[…]

### 2016-04-04 analysis/ipm/vitalRateRegs/regularization/README.Rmd
new README for regularization directory

### README for ridge-regression scripts

*The files in these sub-directories are meant to be run an a High Performance Computing system, NOT on a PC.*

There are directories for each vital rate (growth, survival, and recruitment) that include scripts for running the ridge-regression fits on a High Performance Computing system.

The ``get_plot_lppd.R`` file is script for fetching ridge-regression results (specifically, the log pointwise predictive density from cross-validation model fits) and plotting the lppd versus the prior variance assigned to climate effects.

### 2016-04-04 analysis/ipm/vitalRateRegs/README.Rmd
adds vital rate regression dir overview README

### README for vital rate regression directories

This directory contains sub-directories for fitting vital rate regression models for each vital rate (``growth/``, ``survival/``, and ``recruitment/``), a subdirectory containing scripts for running cross-validation and ridge regression for each vital rate (``regularization/``), and a subdirectory containing scripts for leave-one-year-out cross-validation for each vital rate regression (``validation/``). There is also a ``cache/`` directory containing old versions of scripts.

There are two R scripts:

* ``checkParmCorrs.R`` -- calculates correlations between fitted climate covariates and random year effects

* ``climate_effects_plot.R`` -- plots posterior means and 95% credible intervals for each climate effect from the MCMC

### 2016-04-04 analysis/ipm/vitalRateRegs/growth/README.Rmd
adds growth regression dir README

### README for growth regression scripts

* The ``*Crowding.csv`` files contain information on the crowding experienced by a focal genet at a specific time.

* There are four ``R`` files:
    + ``crowdingCalculation.R``: calculates the amount crowding experience by a focal genet in a quadrat-year. Results are saved in ``*Crowding.csv`` files to be imported for growth regression fitting script.
    + ``growthAllSpp_STAN.R``: script to read-in and format data, and then fit genet growth regression using Stan (see ``growth.stan`` file for model code).
    + ``import2ipm_means.R``: collates MCMC results and defines function for importing posterior mean model coefficients for IPM simulations.
    + ``import2ipm.R``: collates MCMC results and defines function for importing ranomly selected model coefficients from the full MCMC for IPM simulations with variable parameters.

* The ``growth.stan`` file contains a Stan model for fitting the genet growth regression via ‘No-U-Turn’ Hamiltonian Monte Carlo.

* The ``*.RDS`` files contain the MCMC samples from model fitting for each species.

* The ``cache/`` directory contains old versions of R scripts.

### 2016-04-04 analysis/ipm/vitalRateRegs/recruitment/README.Rmd
adds survival and recreuitment dir READMEs

--- 2016-04-04 analysis/ipm/vitalRateRegs/growth/README.Rmd
@@
-### README for growth regression scripts
+### README for recruitment regression scripts
 
@@
 * There are four ``R`` files:
-    + ``crowdingCalculation.R``: calculates the amount crowding experience by a focal genet in a quadrat-year.
-Results are saved in ``*Crowding.csv`` files to be imported for growth regression fitting script.
-    + ``growthAllSpp_STAN.R``: script to read-in and format data, and then fit genet growth regression using Stan (see ``growth.stan`` file for model code).
+    + ``recruitmentAllSpp_STAN.R``: script to read-in and format data, and then fit genet recruitment regression using Stan (see ``recruitment.stan`` file for model code).
     + ``import2ipm_means.R``: collates MCMC results and defines function for importing posterior mean model coefficients for IPM simulations.
@@
 
-* The ``growth.stan`` file contains a Stan model for fitting the genet growth regression via ‘No-U-Turn’ Hamiltonian Monte Carlo.
+* The ``recruitment.stan`` file contains a Stan model for fitting the genet recruitment regression via ‘No-U-Turn’ Hamiltonian Monte Carlo.
 
 * The ``*.RDS`` files contain the MCMC samples from model fitting for each species.
-
-* The ``cache/`` directory contains old versions of R scripts.

### 2016-04-04 analysis/ipm/vitalRateRegs/survival/README.Rmd
adds survival and recreuitment dir READMEs

--- 2016-04-04 analysis/ipm/vitalRateRegs/growth/README.Rmd
@@
-### README for growth regression scripts
+### README for survival regression scripts
 
@@
     + ``crowdingCalculation.R``: calculates the amount crowding experience by a focal genet in a quadrat-year.
-Results are saved in ``*Crowding.csv`` files to be imported for growth regression fitting script.
-    + ``growthAllSpp_STAN.R``: script to read-in and format data, and then fit genet growth regression using Stan (see ``growth.stan`` file for model code).
+Results are saved in ``*Crowding.csv`` files to be imported for survival regression fitting script.
+    + ``survivalAllSpp_STAN.R``: script to read-in and format data, and then fit genet survival regression using Stan (see ``survival.stan`` file for model code).
     + ``import2ipm_means.R``: collates MCMC results and defines function for importing posterior mean model coefficients for IPM simulations.
@@
 
-* The ``growth.stan`` file contains a Stan model for fitting the genet growth regression via ‘No-U-Turn’ Hamiltonian Monte Carlo.
+* The ``survival.stan`` file contains a Stan model for fitting the genet survival regression via ‘No-U-Turn’ Hamiltonian Monte Carlo.
 

### 2016-04-07 manuscript/manuscript.Rmd
fixes typo
adds lppd figure and collation script
file clean up; reformatted and rewrote equations in ms

@@
 
+ \renewcommand{\abstractname}{}
+
 \begin{abstract}
-Rapid climate change has generated growing interest in forecasts of future population trajectories.
+\textit{Abstract.} \hspace{1em} Rapid climate change has generated growing interest in forecasts of future population trajectories.
 Traditional population models, typically built using detailed demographic observations from one study site, can address climate change impacts at one location, but are difficult to scale up to the landscape and regional scales relevant to management decisions.
@@
 The annual maps of the quadrats were digitized and the fates of individual plants tracked and extracted using a computer program [@Lauenroth2008; @Chu2014a].
+The permanent quadrats have not been relocated, but their distribution in six different pastures means the data represent a broad spatial distribution for the study area.
 Daily climate data are available for the duration of the data collection period (1932 - 1945) from the Miles City airport, Wiley Field, 9 km from the study site.
@@
 We also include interactions among same-year climate covariates (e.g., ppt1 $\times$ TmeansSpr1) and climate $\times$ size interactions.
-Climate $\times$ size interactions are for climate main effects only; we do not include interactions between size and pairs of interacting climate effects.
 
@@
 The models are fully descibed in Appendix A, so here we focus on the main process and the model likelihood.
-For the likelihood models, $\textbf{y}^X$ is always the relevant vector of observations for vital rate *X* ($X = S, G, R, or P$ for survival, growth, recruitment, or population growth).
+For the likelihood models, $\textbf{y}^X$ is always the relevant vector of observations for vital rate *X* (*X* = *S*, *G*, *R*, or *P*) for survival, growth, recruitment, or population growth).
 For example, $\textbf{y}^S$ is a vector of 0's and 1's indicating whether a genet survives from *t* to *t+1*, or not.
+All model parameters are species-specific, but we omit subscripts for species in model descriptions below to reduce visual clutter.
 
 #### Vital rate models at the individual level
-We used logistic regression to model survival probability ($S$) of genet $i$ from species $j$ in quadrat group $Q$ from time $t$ to $t+1$:
-
+We used logistic regression to model survival probability ($s$) of a genet:
+
+\vspace{-3em}
 \begin{align}
-\text{logit}(S_{ijQ,t}) &= \gamma^{S}_{j,t} + \phi^{S}_{jQ} + \beta^{S}_{j,t}x_{ij,t} + \omega^{S}_{j}w_{ij,t} + \nu^{S}_{j}w_{ij,t}x_{ij,t} + \theta^{S}_{jk}C_{k,t} \\
-y^{S}_{ijQ,t} &\sim \text{Bernoulli}(S_{ijQ,t})
+\textbf{y}^{S} &\sim \text{Bernoulli}(\textbf{s}) \\
+\text{logit}[s(x,\textbf{z}_t,w)] &= \beta_{0,t} + \beta_{s,t}x + \beta_Q + \textbf{z}'_t \boldsymbol{\beta}_c + \beta_{d,1} w + \beta_{d,2} (xw)
 \end{align}
-
-where $x_{ij,t}$ is the log of genet size, $\gamma^{S}_{j,t}$ is a year-specific intercept, $\beta^{S}_{j,t}$ is the year-specific slope parameter for size, $\phi^{S}_{jQ}$ is the random effect of quadrat group location, and $\theta^{S}_{k}$ is the fixed parameter for the effect of the $k$th climate covariate at time $t$ ($C_{k,t}$).
-Note that the vector of climate covariates (**C**) includes climate variable interactions and climate$\times$size interactions.
-We include density-dependence by estimating the effect of crowding on the focal individual by other individuals of the same species.
-$\omega$ is the effect of crowding and $w_{t,Q}$ is the crowding experienced by the focal individual at time $t$ in quadrat group $Q$.
-We include a size$\times$crowding interaction effect ($\nu^{S}$).
+\vspace{-3em}
+
+where $x$ is the log of genet basal area, $\beta_{0,t}$ is a year specific intercept intercept, $\beta_Q$ is the random effect of quadrat group location, $\beta_{s,t}$ is the year-specific slope parameter for size, **z** is a vector of *i* climate covariates specific to year *t*, $\boldsymbol{\beta}_c$ is a vector of fixed climate effects of length *i*, $\beta_{d,1}$ is the effect of intraspecific crowding experienced by the focal genet (*w*), and $\beta_{d,2}$ is a size by crowding ($xw$) interaction effect.
 
@@
 
+\vspace{-3em}
 \begin{align}
-x_{ijQ,t+1} &= \gamma^{G}_{j,t} + \phi^{G}_{jQ} + \beta^{G}_{j,t}x_{ij,t} + \omega^{G}_{j}w_{ij,t} + \nu^{S}_{j}w_{ij,t}x_{ij,t} + \theta^{G}_{jk}C_{k,t} \\
-y^{G}_{ijQ,t} &\sim \text{Normal}(x_{ijQ,t+1}, \varepsilon_{ij,t})
+\textbf{y}^G &\sim \text{Normal}(\boldsymbol{\mu}, \sigma^2_{x,t}) \\
+\mu(x,\textbf{z}_t,w) &= \beta_{0,t} + \beta_{s,t}x + \beta_Q + \textbf{z}'_t \boldsymbol{\beta}_c + \beta_{d,1} w + \beta_{d,2} (xw)
 \end{align}
-
-where $x$ is log genet size and all other parameters are as described for the survival regression.
-We capture non-constant error variance in growth by modeling the variance around the growth regression ($\varepsilon$) as a nonlinear function of predicted genet size:
-
+\vspace{-3em}
+
+where $\mu(x,\textbf{z}_t,w)$ is log of predicted genet size at time *t* + 1, and all other parameters are as described for the survival regression.
+We capture non-constant error variance in growth by modeling the variance around the growth regression ($\sigma^2_{x,t}$) as a nonlinear function of predicted genet size:
+
+\vspace{-3em}
 \begin{align}
-\varepsilon_{ij,t} = a e^{b x_{ijQ,t+1}}
+\sigma^2_{x,t} = a \, \text{exp}[b \times \mu(x,\textbf{z}_t,w)]
 \end{align}
+\vspace{-3em}
+
+where $\mu(x,\textbf{z}_t,w)$ is log of predicted genet size predicted from the growth regression (Eq. 4), and *a* and *b* are constants.
 
 Our data allows us to track new recruits, but we cannot assign a specific parent to new genets. 
-THerefore, we model recruitment at the quadrat level: the number of new individuals of species $j$ in quadrat $q$ recruiting at time $t+1$ as a function of quadrat "effective cover" ($A'$) in the previous year ($t$).
-Effective cover is a mixture of observed cover ($A$) in the focal quadrat ($q$) and the mean cover across the entire group ($\bar{A}$) of $Q$ quadrats in which $q$ is located:
-
-\begin{equation}
-A'_{jq,t} = p_{j}A_{jq,t} + (1-p_{j})\bar{A}_{jQ,t}
-\end{equation}
+Therefore, we model recruitment at the quadrat level.
+We assume the number of individuals, $y^{R}$, recruiting at time $t+1$ in quadrat q follows a negative binomial distribution:
+
+\vspace{-3em}
+\begin{align}
+y^{R}_{q,t+1} \sim \text{NegBin}(\lambda_{q,t+1},\phi)
+\end{align}
+\vspace{-3em}
+
+where $\lambda$ is the mean intensity and $\phi$ is the size parameter. We define $\lambda$ as a function of quadrat composition and climate in the previous year:
+
+\vspace{-3em}
+\begin{align}
+\lambda_{q,t+1} = C'_{q,t} \, \text{exp}\left(\beta_{0,t} + \beta_Q + \textbf{z}'_t \boldsymbol{\beta}_c + \beta_{d}\sqrt{C'_{q,t}}\right)
+\end{align}
+\vspace{-3em}
+
+where $C'$ is effective cover ($\text{cm}^2$) of the focal species in quadrat $q$, and all other terms are as in the survival and growth regressions.
+Effective cover is a mixture of observed cover ($C$) in the focal quadrat ($q$) and the mean cover across the entire group ($\bar{C}$) of $Q$ quadrats in which $q$ is located:
+
+\vspace{-3em}
+\begin{align}
+C'_{q,t} = pC_{q,t} + (1-p)\bar{C}_{Q,t}
+\end{align}
+\vspace{-3em}
 
 where $p$ is a mixing fraction between 0 and 1 that is estimated within the model.
-
-We assume the number of individuals, $y^{R}$, recruiting at time $t+1$ follows a negative binomial distribution:
-
-\begin{equation}
-y^{R}_{jq,t+1} \sim \text{NegBin}(\lambda_{jq,t+1},\zeta)
-\end{equation}
-
-where $\lambda$ is the mean intensity and $\zeta$ is the size parameter. We define $\lambda$ as:
-
-\begin{equation}
-\lambda_{jq,t+1} = A'_{jq,t}e^{(\gamma^{R}_{j,t} + \phi^{R}_{jQ} + \theta^{R}_{jk}C_{k,t} + \omega^{R}\sqrt{A'_{q,t}})}
-\end{equation}
-
-where $A'$ is effective cover ($\text{cm}^2$) of species $j$ in quadrat $q$ and all other terms are as in the survival and growth regressions.
 
@@
 So we constrained our statistical modeling of vital rates at the population level to change in percent cover within quadrats.
-For the remaining discussion of statistical modeling, we refer to proportion data, which is simply percent cover divided by 100.
+For the remaining discussion of statistical modeling, we refer to proportion data, which is simply the percent cover divided by 100.
 
@@
 However, when we used fitted model parameters from a beta regression in a quadrat-based population model, the simulated population tended toward 100% cover for all species.
-We therefore chose a more constrained modeling approach based on a truncated log-normal likelihood.
+We therefore chose a modeling approach based on a truncated log-normal likelihood.
 The model for quadrat cover change from time $t$ to $t+1$ is
 
+\vspace{-3em}
 \begin{align}
-x_{jq,t+1} &= \gamma^{P}_{j,t} + \phi^{P}_{jQ} + \beta^{P}_{j,t}x_{jq,t} + \theta^{P}_{jk}C_{k,t} \\
-y^{P}_{jq,t+1} &\sim \text{LogNormal}(x_{jq,t+1}, \tau_{j}) \text{T}[0,1]
+\textbf{y}^{P} &\sim \text{LogNormal}(\mu(x,\textbf{z}_t), \sigma^2) \text{T}[0,1] \\
+\mu(x,\textbf{z}_t) &= \beta_{0,t} + \beta_{s,t}x + \beta_Q + \textbf{z}'_t \boldsymbol{\beta}_c
 \end{align}
-
-where $x_{jq,t}$ is the log of species' $j$ proportional cover in quadrat $q$ at time $t$ and all other parameters are as in the individual-level growth model (Eq.
-3).
-Again, note that the climate covariate vector (**C**) includes the climate$\times$cover interaction.
+\vspace{-3em}
+
+where $\mu(x,\textbf{z}_t)$ is the log of proportional cover in quadrat $q$ at time $t+1$, and all other parameters are as in the individual-level growth model (Eq.
+4) except that *x* now represents log of proportional cover.
 The log normal likelihood includes a truncation (T[0,1]) to ensure that predicted values do not exceed 100% cover.
@@
 The natural regulator in a Bayesian application is the prior on the coefficient of interest.
-Each of our statistical models includes the effects of climate covariates via the term $\textbf{c}'\boldsymbol{\theta}$ with prior $\boldsymbol{\theta} \sim \text{Normal}(\boldsymbol{\mu}_\theta, \sigma_\theta^2)$.
-Since we standardized all climate covariates, including interaction terms, to have mean zero and variance one we can set $\boldsymbol{\mu}_\theta=0$, letting $\sigma_\theta^2$ serve as the regulator that can shrink covariates toward zero -- the smaller the prior variance, the more $\theta$ is shrunk toward zero, and the stronger the penalty [@Hooten2015].
-
-To find the optimal penalty (i.e., optimal value of the hyperparameter $\sigma_\theta^2$), we fit each statistical model with a range of values for $\sigma_\theta^2$ and compared predictive scores from leave-one-year-out cross-validation.
-We performed the grid search over 24 evenly-spaced values of $\sigma_\theta^2$, ranging from $\sigma_\theta^2 = 0.01$ to $\sigma_\theta^2 = 2.25$.
-For each statistical model and each species we fit $13\times24=312$ models (13 years to leave out for cross-validation and 24 values of $\sigma_\theta^2$) -- a total of 4992 models.
+Each of our statistical models includes the effects of climate covariates via the term $\textbf{z}'_t \boldsymbol{\beta}_c$ with prior $\boldsymbol{\beta}_c \sim \text{Normal}(\boldsymbol{\mu}_{\beta_c}, \sigma_{\beta_c}^2)$.
+Since we standardized all climate covariates to have mean zero and variance one, we can set $\boldsymbol{\mu}_{\beta_c} = 0$ and let $\sigma_{\beta_c}^2$ serve as the regulator that can shrink covariates toward zero -- the smaller the prior variance, the more the posteriors of $\boldsymbol{\beta}_c$ are shrunk toward zero, and the stronger the penalty [@Hooten2015].
+
+To find the optimal penalty (i.e., optimal value of the hyperparameter $\sigma_{\beta_c}^2$), we fit each statistical model with a range of values for $\sigma_{\beta_c}^2$ and compared predictive scores from leave-one-year-out cross-validation.
+We performed the grid search over 24 values of $\sigma_{\beta_c}^2$, ranging from $\sigma_{\beta_c}^2 = 0.01$ to $\sigma_{\beta_c}^2 = 2.25$.
+For each statistical model and each species we fit $13\times24=312$ models (13 years to leave out for cross-validation and 24 values of $\sigma_{\beta_c}^2$) -- a total of 4,992 models.
 We calculated the log pointwise predictive density (*lppd*) to score each model's ability to predict the left-out data.
-Thus, for training data $y_{train}$ and held-out data $y_{hold}$ at a given value of  $\sigma_\theta^2$ across all MCMC samples $s=1,2,...,S$ and all hold outs of data from year *t* to year *T*, *lppd* is
+Thus, for training data $y_{train}$ and held-out data $y_{hold}$ at a given value of  $\sigma_\theta^2$ across all MCMC samples $s=1,2,...,S$ and all hold outs of data from year *t* to year *T*, and letting $\theta$ represent all unknowns, *lppd* is
 
@@
 We chose the optimal prior variance for each species-statistical model combination as the one that produced the highest *lppd* and then fit each species-statistical model combination using the full data set for each species and the optimal prior variance.
-In some cases *lppd* was highest at the highest (most vague) value of prior variance, in which case we explored a slightly expanded range of prior variances (e.g., $\sigma_\theta^2$>2.25).
 
@@
 We use the same approach for drawing parameter values as described for the IPM.
-After drawing the appropriate parameter set, we calculate the mean response (population cover at *t+1* = $x_{t+1}$) according to Eq.
-9.
-We then make a random draw from a [0,1] truncated lognormal distribution with mean equal to $x_{t+1}$ from Eq.
-9 and the variance estimate from the fitted model.
+After drawing the appropriate parameter set, we calculate the mean response (log cover at *t+1* = $\mu_{t+1}$) according to Eq.
+10.
+We then make a random draw from a [0,1] truncated lognormal distribution with mean equal to $\mu_{t+1}$ from Eq.
+10 and the variance estimate from the fitted model.
 We can then project the model forward by drawing a new parameter set (unique to climate year and MCMC iteration) at each timestep.

### 2016-04-11 manuscript/manuscript.Rmd
gets spasing right around equations
adds spp biology and quadrat and record counts to methods:

@@
 
-\textsc{\normalsize{Andrew T. Tredennick\footnote{Corresponding author: atredenn@gmail.com}\textsuperscript{1}, Mevin B. Hooten\textsuperscript{2,3,4}, and Peter B. Adler\textsuperscript{1}}}
+\textsc{\normalsize{Andrew T. Tredennick\footnote{Corresponding author: atredenn@gmail.com}\textsuperscript{1}, Peter B. Adler\textsuperscript{1}, and Mevin B. Hooten\textsuperscript{2,3,4}}}
 
@@
 Our demographic data come from the Fort Keogh Livestock and Range Research Laboratory in eastern Montana's northern mixed prairie near Miles City, Montana, USA ($46^{\circ}$ 19' N, $105^{\circ}$ 48' W).
-The dataset is freely available on Ecological Archives[^link] [@Anderson2011] , and interested readers should refer to the metadata for a complete description.
+The dataset is freely available on Ecological Archives[^link] [@Anderson2011], and interested readers should refer to the metadata for a complete description.
 The site is about 800 m above sea level and mean annual precipitation (1878-2009) is 334 mm, with most annual precipitation falling from April through September.
-The community is grass-dominated and we focused on the four most abundant grass species: \emph{Bouteloua gracilis} (BOGR), \emph{Hesperostipa comata} (HECO), \emph{Pascopyrum smithii} (PASM), and \emph{Poa secunda} (POSE) (Fig.
+The community is grass-dominated, and we focused on the four most abundant grass species: \emph{Bouteloua gracilis} (BOGR), \emph{Hesperostipa comata} (HECO), \emph{Pascopyrum smithii} (PASM), and \emph{Poa secunda} (POSE) (Fig.
 1).
+*B.
+gracilis* is a warm-season perennial grass, whereas *H.
+comata*, *P.
+smithii*, and *P.
+secunda* are cool-season perennial grasses.
+All species typically begin growth in the early spring, reach maximum growth and flower in early to mid summer (May-June), and disperse seed in mid to late summer (July-September).
 
@@
 Based on these two datasets we can compare population models built using individual-level data and aggregated, quadrat-level data.
+At the individual level we explicitly model three vital rates: growth (13 year-to-year transitions, 29 quadrats, 18,730 records), survival (13 years, 33 quadrats, 29,353 records), and recruitment (13 years, 33 quadrats, 304 records).
+At the quadrat level we model population growth as change in percent cover of quadrats with non-zero cover in year *t* and in year *t-1* (13 year-to-year transitions, 19 quadrats, 784 records).
+For modeling population growth at the quadrat level we ignore within-quadrat extirpation and colonization events because they are very rare in our time series ($N=9$ and $N=10$, respectively, across all species).
+Given the broad spatial distribution of the quadrats we are studying, it is safe to assume that these events are in fact rare enough to be ignored for our purposes.
 
@@
 We describe the statistical models separately since fitting the models required different approaches.
-All models contain five climate covariates that we chose \emph{a priori}: "water year" precipitation at \emph{t}-1 (lagppt); April through June precipitation at \emph{t}-1 and \emph{t}-2 (ppt1 and ppt2, respectively) and April through June temperature at \emph{t}-1 and \emph{t}-2 (TmeanSpr1 and TmeanSpr2, respectively), where \emph{t} is the observation year.
-We also include interactions among same-year climate covariates (e.g., ppt1 $\times$ TmeansSpr1) and climate $\times$ size interactions.
+All models contain five climate covariates that we chose \emph{a priori}: "water year" precipitation at \emph{t}-2 (lagppt); April through June precipitation at \emph{t}-1 and \emph{t} (ppt1 and ppt2, respectively) and April through June temperature at \emph{t}-1 and \emph{t} (TmeanSpr1 and TmeanSpr2, respectively), where \emph{t}-1 to *t* is the transition of interest.
+We also include interactions among same-year climate covariates (e.g., ppt1 $\times$ TmeansSpr1).
 
@@
 We have percent cover data, which can easily be transformed to proportion data.
-We first considered fitting three vital rate models analagous to those we fit at the individual level: one for probability of extirpation within a quadrat (analagous to survival), one for cover change within a quadrat (analagous to growth), and one for probability of colonization within a quadrat (analagous to recruitment).
-However, within-quadrat extirpation and colonization events were rare in our time series ($N=9$ and $N=10$, respectively, across all species).
-Given the broad spatial distribution of the quadrats we are studying, it is safe to assume that these events are in fact rare enough to be ignored for our purposes.
-So we constrained our statistical modeling of vital rates at the population level to change in percent cover within quadrats.
-For the remaining discussion of statistical modeling, we refer to proportion data, which is simply the percent cover divided by 100.
-
 An obvious choice for fitting a linear model to proportion data is beta regression because the support of the beta distribution is [0,1], not including true zeros or ones.
@@
 
-\begin{equation}
+\vspace{-3em}
+\begin{align}
 \text{lppd}_{\text{CV}} = \sum_{t=1}^T \text{log}_e \int [y_{t,hold}|\theta][\theta|y_{train}]d\theta,
-\end{equation}
+\end{align}
+\vspace{-3em}
 
@@
 
-\begin{equation}
+\vspace{-3em}
+\begin{align}
 \sum_{t=1}^T \text{log}_e \left(\frac{1}{S} \sum^S_{s=1}(y_{t,hold}|\theta_{ts})  \right).
-\end{equation}
+\end{align}
+\vspace{-3em}
 
@@
 
-\begin{equation}
+\vspace{-3em}
+\begin{align}
 n(v_{j},t+1) = \int_{L_{j}}^{U_{j}} k_{j}(v_{j},u_{j},\bar{w_{j}}(u_{j}))n(u_{j},t)
-\end{equation}
+\end{align}
+\vspace{-3em}
 
@@
 
-\begin{equation}
+\vspace{-3em}
+\begin{align}
 k_{j}(v_{j},u_{j},\bar{w_{j}}) = S_j(u_j, \bar{w_{j}}(u_{j}))G_j(v_{j},u_{j},\bar{w_{j}}(u_{j})) + R_j(v_{j},u_{j},\bar{w_{j}}),
-\end{equation}
+\end{align}
+\vspace{-3em}
 

### 2016-04-12 manuscript/manuscript.Rmd
updates quadrat information for population model

@@
 At the individual level we explicitly model three vital rates: growth (13 year-to-year transitions, 29 quadrats, 18,730 records), survival (13 years, 33 quadrats, 29,353 records), and recruitment (13 years, 33 quadrats, 304 records).
-At the quadrat level we model population growth as change in percent cover of quadrats with non-zero cover in year *t* and in year *t-1* (13 year-to-year transitions, 19 quadrats, 784 records).
-For modeling population growth at the quadrat level we ignore within-quadrat extirpation and colonization events because they are very rare in our time series ($N=9$ and $N=10$, respectively, across all species).
+At the quadrat level we model population growth as change in percent cover of quadrats with non-zero cover in year *t* and in year *t-1* (13 year-to-year transitions, 29 quadrats, 866 records).
+For modeling population growth at the quadrat level we ignore within-quadrat extirpation and colonization events because they are very rare in our time series ($N=16$ and $N=13$, respectively, across all species).
 Given the broad spatial distribution of the quadrats we are studying, it is safe to assume that these events are in fact rare enough to be ignored for our purposes.

### 2016-04-15 manuscript/manuscript.Rmd
adds new posterior climate effect densities figure

@@
 
+\begin{figure}[!ht]
+  \centering
+      \includegraphics[width=6in]{./components/lppds_suppfig.png}
+  \caption{Cross validation scores (summed log pointwise predictive density) plotted against the prior variance for $\boldsymbol{\beta}_C$.
+The optimal score for prediction for each species-vital rate combination is shown with a vertical red line.}
+\end{figure}
+
+\newpage{}
+
+\begin{figure}[!ht]
+  \centering
+      \includegraphics[width=5in]{./components/climate_posteriors_figure.png}
+  \caption{Posterior distributions of climate effects ($\boldsymbol{\beta}_C$) for each species and vital rate statistical model.
+Since our priors were constrained via ridge-regression, we highlight climate effects whose 80\% credible intervals do not overlap zero (red for negative coefficients, blue for positive coefficients).}
+\end{figure}
+
+\newpage{}
+
 ```{r figure_3, dependson="plot-options", fig.cap="Sensitivity of equilibrium cover simulated from the IPM to each climate scenario applied to individual and combined vital rates.

### 2016-04-15 FigureScripts/plot_climate_posteriors.R

@@ -1,0 +1,103 @@
+##  R Script to Plot Posterior Densities of Climate Effects
+##
+##  Author: Andrew Tredennick
+##  Last update: 4-15-2016
+
+
+# Clear the Workspace
+rm(list=ls())
+
+
+
+####
+####  LOAD LIBRARIES
+####
+library(ggplot2)
+library(ggthemes)
+library(plyr)
+library(reshape2)
+library(ggmcmc)
+library(gridExtra)
+
+
+
+####
+####  PRELIMINARIES
+####
+species <- c("BOGR", "HECO", "PASM", "POSE")
+vitals <- c("growth", "survival", "recruitment", "cover")
+path2ipms <- "../analysis/ipm/vitalRateRegs/"
+path2qbms <- "../analysis/quadBM/vitalRateRegressions/truncNormModel/"
+
+
+####
+#### LOOP OVER MODELS AND SPECIES; PLOT THEM
+####
+pdf(file = "clim_posteriors.pdf", onefile = TRUE, height=3.5, width=3)
+for(do_species in species){
+  for(do_vital in vitals){
+    if(do_vital=="cover") { path <- paste0(path2qbms, "popgrowth_stanmcmc_",do_species,".RDS") }
+    if(do_vital!="cover") { path <- paste0(path2ipms,do_vital,"/",do_vital,"_stanmcmc_",do_species,".RDS") }
+    if(do_vital=="survival"){ post_df <- ggs(readRDS(path)) }
+    if(do_vital=="recruitment"){ post_df <- ggs(readRDS(path)) }
+    if(do_vital=="cover"){ post_df <- readRDS(path) }
+    if(do_vital=="growth"){ post_df <- readRDS(path) }
+    
+    post_clims <- post_df[grep("b2",post_df$Parameter),]
+    post_clim_stats <- ddply(post_clims, .(Parameter), summarise,
+                             average = mean(value),
+                             upperci = quantile(value, 0.9),
+                             lowerci = quantile(value, 0.1))
+    post_clim_stats$col <- NA
+    for(i in 1:nrow(post_clim_stats)){
+      if(sign(post_clim_stats$lowerci[i]) == sign(post_clim_stats$upperci[i]) & post_clim_stats$average[i] > 0 ){
+        post_clim_stats$col[i] <- "Cblue"
+      }
+      if(sign(post_clim_stats$lowerci[i]) == sign(post_clim_stats$upperci[i]) & post_clim_stats$average[i] < 0 ){
[…]

### 2016-04-20 manuscript/components/error_table.tex
updates QBM forecast horizon results and figure; adds error table

\begin{table}[ht]
\centering
\caption{Accuracy (mean absolute error, MAE) and precision (90\% Distance) 
                  of out of sample predictions. Forecasts were made without random 
                  year effects; only climate covariates could explain year-to-year 
                  variation. 90\% Distance refers to the average distance between the 
                  upper and lower 90th percentiles of the 100 predicted values for 
                  each quadrat-year combination.} 
\begin{tabular}{llrrr}
  \hline
Species & Model & MAE & 90\% Distance & Mean Obs. Cover \\ 
  \hline
BOGR & IPM & 4.03 & 24.74 & 9.43 \\ 
  BOGR & QBM & 2.80 & 29.46 & 8.35 \\ 
  HECO & IPM & 0.34 & 1.87 & 1.15 \\ 
  HECO & QBM & 0.29 & 5.53 & 1.18 \\ 
  PASM & IPM & 0.14 & 0.50 & 0.42 \\ 
  PASM & QBM & 0.17 & 1.57 & 0.42 \\ 
  POSE & IPM & 0.45 & 1.63 & 1.25 \\ 
  POSE & QBM & 0.41 & 3.51 & 1.27 \\ 
   \hline
\end{tabular}
\end{table}

### 2016-04-20 FigureScripts/plot_forecast_horizons.R

@@ -54,12 +54,13 @@ out_qbm <- data.frame(yearstart=NA, mean_error=NA, rmse=NA, species=NA)
 for(ispp in 1:length(species_list)){
   tmp_file <- paste0(species_list[ispp],"_final_year_cover.RDS")
   tmp_sims <- readRDS(paste0(path2qbms,tmp_file))
+  tmp_sims <- tmp_sims[2:nrow(tmp_sims),]
   tmp_sims$error <- with(tmp_sims, finalcover-obs_finalcover)
   tmp_agg <- ddply(tmp_sims, .(yearstart), summarise,
                    mean_error = mean(error),
                    rmse = sqrt(mean(error^2)))
   tmp_agg$species <- species_list[ispp]
-  out_qpm <- rbind(out_ipm, tmp_agg)
+  out_qbm <- rbind(out_qbm, tmp_agg)
 }
 out_qbm <- out_qbm[2:nrow(out_qbm),] # remove first NA-filled row
 
@@ -67,6 +68,8 @@ out_qbm <- out_qbm[2:nrow(out_qbm),] # remove first NA-filled row
 out_ipm$model <- "IPM"
 out_ipm$yearsbefore <- with(out_ipm, max(startyear)-startyear+1)
 out_qbm$model <- "QBM"
+out_qbm$yearsbefore <- with(out_qbm, max(yearstart)-yearstart+1)
+names(out_qbm) <- names(out_ipm)
 out_all <- rbind(out_ipm, out_qbm)
 
 
@@ -74,11 +77,16 @@ out_all <- rbind(out_ipm, out_qbm)
 ####
 ####  MAKE THE PLOT
 ####
-ggplot(data=out_ipm, aes(x=yearsbefore, y=1-rmse))+
+ggplot(data=out_all, aes(x=yearsbefore, y=abs(mean_error)*100, linetype=model, shape=model))+
   geom_line(color="grey50")+
   geom_point(size=3)+
   facet_grid(species~., scales="free")+
   xlab("Years Before Forecast")+
-  ylab("Forecast Skill (1 - Root Mean Square Error)")+
+  # ylab(expression(atop("Forecast Skill", atop("(1 - Root Mean Square Error)", ""))))+
+  ylab("Mean Absolute Error (% Cover)")+
   scale_x_continuous(breaks=c(1:13))+
-  theme_few()
+  scale_shape_discrete(name="")+
+  scale_linetype_discrete(name="")+
+  theme_few()+
[…]

### 2016-04-20 analysis/quadBM/simulations/forecast_horizon_qbm.R

@@ -123,6 +123,8 @@ for(do_species in sppList){
   # Lognormal sigma (called tau here)
   tau <- fitthin[grep("tau", fitthin$Parameter),]
   
+  outall <- data.frame(quad=NA, finalcover=NA, sim=NA, 
+                       species=NA, yearstart=NA, obs_finalcover=NA)
   for(do_year in all_years){
     ####
     #### Run simulations -----------------------------------------------------
@@ -177,7 +179,8 @@ for(do_species in sppList){
     colnames(newNs)[1:2] <- c("quad", "finalcover")
     obs_by_quad <- subset(obs_data, year==45 & Species==do_species)[,c("quad","propCover")]
     colnames(obs_by_quad) <- c("quad", "obs_finalcover")
-    outall <- merge(newNs, obs_by_quad)
+    tmpoutall <- merge(newNs, obs_by_quad)
+    outall <- rbind(outall, tmpoutall)
   }#end year loop
   
   outname <- paste(do_species,"_final_year_cover.RDS", sep="")

### 2016-04-20 manuscript/manuscript.Rmd
updates forecast horizon plot

@@
 
-Data accessibility
-------------------
-Chart-quadrat data available from *Ecological Archives* at http://esapubs.org/archive/ecol/E092/143/.
-All code to reproduce our work is available on Dryad (link) and GitHub (https://github.com/atredennick/MicroMesoForecast).
-
 \pagebreak{}
-
-\setcounter{page}{29}
 
@@
 ------
-```{r table_1,results='asis',message=FALSE, comment=NA, echo=FALSE, eval=F}
-####
-##  Make a table for average residual from each model and year effects structure
-####
-
-##  IPM results
-# Read in data
-setwd("/Users/atredenn/Repos/MicroMesoForecast/manuscript/")
-ipm_onestep <- readRDS("../analysis/ipm/simulations/results/one_step_validation/ipm_loyo_forecasts_combined.rds")
-# Remove predictions >1 for fair comparison to [0,1] truncated quad model
-resets <- which(ipm_onestep[,"cover.t1"]>1)
-ipm_onestep[resets, "cover.t1"] <- 1
-# Average predictions over quad-year reps
-avg_ipm <- ddply(ipm_onestep, .(species, quad, t1), summarise,
-                 avg_prediction = median(cover.t1),
-                 observation = mean(cover.t0),
-                 quant_dist = quantile(cover.t1*100, probs = 0.95) - quantile(cover.t1, probs = 0.05))
-# Calculate error
-avg_ipm$error <- avg_ipm$observation*100 - avg_ipm$avg_prediction*100
-# Aggregate over species
-ipmstats <- ddply(avg_ipm, .(species), summarise,
-                  mae = mean(abs(error)),
-                  mean_cover = mean(observation*100),
-                  quant_dist = mean(quant_dist))
-ipmstats$model <- "IPM"
-
-##  QBM results
-# Read in data
-qbm_onestep <- readRDS("../analysis/quadBM/simulations/results/one_step_validation/qbm_one-step_forecasts_combined.rds")
-# Average predictions over quad-year reps
-avg_qbm <- ddply(qbm_onestep, .(species, quad, year), summarise,
-                 avg_prediction = median(cover.t1),
-                 observation = mean(cover.t0),
-                 quant_dist = quantile(cover.t1*100, probs = 0.95) - quantile(cover.t1, probs = 0.05))
-# Calculate error
-avg_qbm$error <- avg_qbm$observation*100 - avg_qbm$avg_prediction*100
-# Aggregate over species
-qbmstats <- ddply(avg_qbm, .(species), summarise,
-                  mae = mean(abs(error)),
-                  mean_cover = mean(observation*100),
-                  quant_dist = mean(quant_dist))
-qbmstats$model <- "QBM"
-
-mae_list <- list()
-mae_pvals <- list()
-species <- unique(qbmstats$species)
-for(spp in species){
-  tmp1 <- subset(avg_ipm, species==spp)
-  tmp2 <- subset(avg_qbm, species==spp)
-  err1 <- as.data.frame(abs(tmp1$observation - tmp1$avg_prediction))
-  err2 <- as.data.frame(abs(tmp2$observation - tmp2$avg_prediction))
-  colnames(err1) <- "resid"
-  err1$model <- "ipm"
-  colnames(err2) <- "resid"
-  err2$model <- "qbm"
-  errd <-  rbind(err1, err2)
-  mae_ttest <- with(errd, t.test(resid~model, alternative = "less"))
-  mae_list[[spp]] <- mae_ttest
-  mae_pvals[[spp]] <- mae_ttest[["p.value"]]
-}
-
-mae_ps <- melt(mae_pvals)
-
-stats_table <- rbind(ipmstats, qbmstats)
-stats_table <- stats_table[ order(stats_table[,"species"]), ]
-stats_table <- stats_table[,c("species", "model", "mae",
-                              "quant_dist","mean_cover")]
-
-colnames(stats_table) <- c("Species", "Model", "MAE", "90% Distance", "Mean Obs. Cover")
-print.xtable(xtable(stats_table, caption = "Accuracy (mean absolute error, MAE) and precision (90\\% Distance) of out of sample predictions.
-Forecasts were made without random year effects; only climate covariates could explain year-to-year variation.
-90\\% Distance refers to the average distance between the upper and lower 90th percentiles of the 100 predicted values for each quadrat-year combination."),type="latex", comment=FALSE,
-             include.rownames=FALSE, caption.placement="top")
-```
+
+\begin{table}[ht]
+\centering
+\caption{Accuracy (mean absolute error, MAE) and precision (90\% Distance) 
+                  of out of sample predictions. Forecasts were made without random 
+                  year effects; only climate covariates could explain year-to-year 
+                  variation. 90\% Distance refers to the average distance between the 
+                  upper and lower 90th percentiles of the 100 predicted values for 
+                  each quadrat-year combination.} 
+\begin{tabular}{llrrr}
+  \hline
+Species & Model & MAE & 90\% Distance & Mean Obs. Cover \\ 
+  \hline
+BOGR & IPM & 4.03 & 24.74 & 9.43 \\ 
+  BOGR & QBM & 2.80 & 29.46 & 8.35 \\ 
+  HECO & IPM & 0.34 & 1.87 & 1.15 \\ 
+  HECO & QBM & 0.29 & 5.53 & 1.18 \\ 
+  PASM & IPM & 0.14 & 0.50 & 0.42 \\ 
+  PASM & QBM & 0.17 & 1.57 & 0.42 \\ 
+  POSE & IPM & 0.45 & 1.63 & 1.25 \\ 
+  POSE & QBM & 0.41 & 3.51 & 1.27 \\ 
+   \hline
+\end{tabular}
+\end{table}
 

### 2016-04-26 manuscript/components/error_table.tex
adds new one step and climate change simulation results

@@
   \hline
-BOGR & IPM & 4.03 & 24.74 & 9.43 \\ 
-  BOGR & QBM & 2.80 & 29.46 & 8.35 \\ 
-  HECO & IPM & 0.34 & 1.87 & 1.15 \\ 
-  HECO & QBM & 0.29 & 5.53 & 1.18 \\ 
-  PASM & IPM & 0.14 & 0.50 & 0.42 \\ 
-  PASM & QBM & 0.17 & 1.57 & 0.42 \\ 
-  POSE & IPM & 0.45 & 1.63 & 1.25 \\ 
-  POSE & QBM & 0.41 & 3.51 & 1.27 \\ 
+BOGR & IPM & 4.61 & 22.41 & 8.26 \\ 
+  BOGR & QBM & 4.06 & 29.18 & 8.26 \\ 
+  HECO & IPM & 0.59 & 1.93 & 1.22 \\ 
+  HECO & QBM & 0.60 & 5.64 & 1.22 \\ 
+  PASM & IPM & 0.15 & 0.49 & 0.40 \\ 
+  PASM & QBM & 0.20 & 1.59 & 0.40 \\ 
+  POSE & IPM & 0.71 & 1.66 & 1.23 \\ 
+  POSE & QBM & 0.76 & 3.52 & 1.23 \\ 
    \hline

### 2016-04-26 manuscript/manuscript.Rmd
adds new one step and climate change simulation results

@@
   \centering
-      \includegraphics[width=5in]{./components/climate_posteriors_figure.png}
+      \includegraphics{./components/climate_posteriors_figure.png}
   \caption{Posterior distributions of climate effects ($\boldsymbol{\beta}_C$) for each species and vital rate statistical model.
-Since our priors were constrained via ridge-regression, we highlight climate effects whose 80\% credible intervals do not overlap zero (red for negative coefficients, blue for positive coefficients).}
+Since our priors were constrained via ridge-regression, we highlight climate effects whose 80\% credible intervals do not overlap zero (red for negative coefficients, blue for positive coefficients).
+Kernel bandwidths of posterior densities were adjusted by a factor of 4 for visual clarity.}
 \end{figure}

### 2016-04-26 FigureScripts/climate_change_figure.R

@@ -1,0 +1,26 @@
+##  R Script for Plotting Results from Climate Change Scenarios
+##
+##  Author: Andrew Tredennick
+##  Last update: 4-26-2016
+
+# Clear the workspace
+rm(list=ls())
+
+
+
+####
+####  PRELIMINARIES
+####
+path2qbms <- "../analysis/quadBM/simulations/results/climchange_varyparams/"
+path2ipms <- "../analysis/ipm/simulations/results/climchange_varyparams/"
+path2figs <- "../manuscript/components/"
+
+
+
+####
+####  LOAD LIBRARIES
+####
+library("plyr") # data summarising
+library("reshape2") # data wrangling
+library("ggplot2") # plotting
+library("ggthemes") # plotting themes

### 2016-04-26 FigureScripts/one_step_results_table.R

@@ -55,9 +55,9 @@ ipmstats$model <- "IPM"
 qbm_onestep <- readRDS(paste0(path2qbm, "qbm_one-step_forecasts_combined.RDS"))
 # Average predictions over quad-year reps
 avg_qbm <- ddply(qbm_onestep, .(species, quad, year), summarise,
-                 avg_prediction = median(cover.t1),
-                 observation = mean(cover.t0),
-                 quant_dist = quantile(cover.t1*100, probs = 0.95) - quantile(cover.t1, probs = 0.05))
+                 avg_prediction = median(pred_cover.t1),
+                 observation = mean(obs_cover.t1),
+                 quant_dist = quantile(pred_cover.t1*100, probs = 0.95) - quantile(pred_cover.t1, probs = 0.05))
 # Calculate error
 avg_qbm$error <- avg_qbm$observation*100 - avg_qbm$avg_prediction*100
 # Aggregate over species
@@ -66,6 +66,9 @@ qbmstats <- ddply(avg_qbm, .(species), summarise,
                   mean_cover = mean(observation*100),
                   quant_dist = mean(quant_dist))
 qbmstats$model <- "QBM"
+
+# REPLACE mean QBM cover with cover from IPM because it is the correct value
+qbmstats$mean_cover <- ipmstats$mean_cover
 
 
 testq <- subset(qbm_onestep, quad=="B1"&sim==1)
@@ -76,26 +79,85 @@ testi <- subset(ipm_onestep, quad=="B1"&rep==1)
 ####  CALCULATE P-VALUES FROM T-TESTS ON MODEL ERRORS
 ####  [SEE YE et al. 2015 (PNAS) FOR EXAMPLE USING MODEL ACCURACIES]
 ####
-mae_list <- list()
-mae_pvals <- list()
-species <- unique(qbmstats$species)
-for(spp in species){
-  tmp1 <- subset(avg_ipm, species==spp)
-  tmp2 <- subset(avg_qbm, species==spp)
-  err1 <- as.data.frame(abs(tmp1$observation - tmp1$avg_prediction))
-  err2 <- as.data.frame(abs(tmp2$observation - tmp2$avg_prediction))
-  colnames(err1) <- "resid"
-  err1$model <- "ipm"
-  colnames(err2) <- "resid"
-  err2$model <- "qbm"
-  errd <-  rbind(err1, err2)
[…]

### 2016-04-26 FigureScripts/plot_climate_posteriors.R

@@ -84,8 +84,8 @@ for(do_species in species){
       scale_y_continuous(breaks=NULL)+
       theme_few()+
       theme(strip.background=element_rect(fill="white"))+
-      theme(strip.text.y = element_text(size = 8, angle = 0))+
-      theme(axis.title=element_text(size=8), text=element_text(size=6))
+      theme(strip.text.y = element_text(size = 0, angle = 0))+
+      theme(axis.title=element_text(size=0), text=element_text(size=16))
     print(gout)
   }
 }

### 2016-04-26 analysis/ipm/simulations/03_climate_change_wrapper.R

@@ -4,7 +4,7 @@
 #### 2-11-2014
 #### Updated: 4-6-2016
 
-rm(list-ls())
+rm(list=ls())
 
 do_change <- 1 # Determines the level of climate change programmatically (not yet)
 
@@ -28,8 +28,8 @@ if(use_mean_params==FALSE){
 }
 
 ##  Global parameters for the simulations
-tlimit <- 250  # number of years to simulate
-burn.in <- 50    # years to cut before calculations
+tlimit <- 2500  # number of years to simulate
+burn.in <- 500    # years to cut before calculations
 spp_list <- c("BOGR","HECO","PASM","POSE")
 n_spp <- length(spp_list)
 
@@ -96,11 +96,11 @@ clim_data_obs <- clim_data_obs[,c("year", "pptLag","ppt1","ppt2","TmeanSpr1","Tm
 perturb_climate <- function(clim_alt, clim_var, clim_data_obs, do_spp){
   doSpp <- spp_list[do_spp]
   
-  if(is.na(clim_var)==TRUE){
+  if(is.na(clim_var[1])==TRUE){
     clim_data <- clim_data_obs
   }
   
-  if(is.na(clim_var)==FALSE){
+  if(is.na(clim_var[1])==FALSE){
     # Perturb climate time series
     clim_data <- clim_data_obs # re-assigns the temporary clim_data object
     for(ivar in 1:length(clim_var)){
@@ -112,7 +112,7 @@ perturb_climate <- function(clim_alt, clim_var, clim_data_obs, do_spp){
       clim_data[,vars] <- clim_data_obs[,vars]+tmp1[,vars] # applies ppt additions
     }
     # Make sure climate perturbations have been applied
-    if(mean(clim_data$pptLag)==mean(clim_data_obs$pptLag)) {
+    if(mean(clim_data[,vars[1]])==mean(clim_data_obs[,vars[1]])) {
       stop("climate perturbation not applied")
     }
   }
@@ -149,7 +149,7 @@ for(ispp in 1:length(spp_list)){
   
   # Throw error if something weird happens
   if(min(covSave)<0.00001) { stop("really low values; check IPM") }
-  if(max(covSave)>1) { stop("cover greater than 1; check IPM") }
+  # if(max(covSave)>1) { stop("cover greater than 1; check IPM") }
   
   # Set output filename
   filenameid <- paste(clim_var, collapse="")
@@ -204,8 +204,8 @@ for(ispp in 1:length(spp_list)){
[…]

### 2016-04-26 analysis/ipm/simulations/results/one_step_validation/check_1steps.R

@@ -4,13 +4,12 @@
 
 file <- "_sim_cover_1step_ahead.csv"
 spp_list <- c("BOGR", "HECO", "PASM", "POSE")
-all_d <- data.frame(quad=NA,t1=NA,t0=NA,rep=NA,cover.t0=NA,cover.t1=NA,obs.cover.t1=NA,species=NA)
+all_d <- list()
 for(i in 1:length(spp_list)){
   tmp_d <- read.csv(paste(spp_list[i],file,sep=""))
   tmp_d$species <- rep(spp_list[i],nrow(tmp_d))
   all_d <- rbind(all_d, tmp_d)
 }
-all_d <- all_d[2:nrow(all_d),]
 saveRDS(all_d,"ipm_loyo_forecasts_combined.RDS")
 
 # all_d$resid <- with(all_d, (cover.t1*100)-(obs.cover.t1*100))

### 2016-04-26 manuscript/manuscript.Rmd
updates ms figures and figure scripts; adds ipm mean forecast horizon results

@@
 -------
-```{r pvalues, echo=FALSE, eval=TRUE}
-##  IPM results
-# Read in data
-setwd("/Users/atredenn/Repos/MicroMesoForecast/manuscript/")
-ipm_onestep <- readRDS("../analysis/ipm/simulations/results/one_step_validation/ipm_loyo_forecasts_combined.rds")
-# Remove predictions >1 for fair comparison to [0,1] truncated quad model
-resets <- which(ipm_onestep[,"cover.t1"]>1)
-ipm_onestep[resets, "cover.t1"] <- 1
-# Average predictions over quad-year reps
-avg_ipm <- ddply(ipm_onestep, .(species, quad, t1), summarise,
-                 avg_prediction = median(cover.t1),
-                 observation = mean(cover.t0),
-                 quant_dist = quantile(cover.t1*100, probs = 0.95) - quantile(cover.t1, probs = 0.05))
-# Calculate error
-avg_ipm$error <- avg_ipm$observation*100 - avg_ipm$avg_prediction*100
-# Aggregate over species
-ipmstats <- ddply(avg_ipm, .(species), summarise,
-                  mae = mean(abs(error)),
-                  mean_cover = mean(observation*100),
-                  quant_dist = mean(quant_dist))
-ipmstats$model <- "IPM"
-
-##  QBM results
-# Read in data
-qbm_onestep <- readRDS("../analysis/quadBM/simulations/results/one_step_validation/qbm_one-step_forecasts_combined.rds")
-# Average predictions over quad-year reps
-avg_qbm <- ddply(qbm_onestep, .(species, quad, year), summarise,
-                 avg_prediction = median(cover.t1),
-                 observation = mean(cover.t0),
-                 quant_dist = quantile(cover.t1*100, probs = 0.95) - quantile(cover.t1, probs = 0.05))
-# Calculate error
-avg_qbm$error <- avg_qbm$observation*100 - avg_qbm$avg_prediction*100
-# Aggregate over species
-qbmstats <- ddply(avg_qbm, .(species), summarise,
-                  mae = mean(abs(error)),
-                  mean_cover = mean(observation*100),
-                  quant_dist = mean(quant_dist))
-qbmstats$model <- "QBM"
-
-mae_list <- list()
-mae_pvals <- list()
-species <- unique(qbmstats$species)
-for(spp in species){
-  tmp1 <- subset(avg_ipm, species==spp)
-  tmp2 <- subset(avg_qbm, species==spp)
-  err1 <- as.data.frame(abs(tmp1$observation - tmp1$avg_prediction))
-  err2 <- as.data.frame(abs(tmp2$observation - tmp2$avg_prediction))
-  colnames(err1) <- "resid"
-  err1$model <- "ipm"
-  colnames(err2) <- "resid"
-  err2$model <- "qbm"
-  errd <-  rbind(err1, err2)
-  mae_ttest <- with(errd, t.test(resid~model, alternative = "less"))
-  mae_list[[spp]] <- mae_ttest
-  mae_pvals[[spp]] <- mae_ttest[["p.value"]]
-}
-
-mae_ps <- melt(mae_pvals)
-```
-
 ### Comparison of forecast models
@@
 
-```{r get_results, echo=FALSE, eval=FALSE}
-setwd("~/Repos/MicroMesoForecast/analysis/")
-
-####
-##  Mean parameter runs
-####
-meanparam_ipm <- data.frame(cover=NA, species=NA, 
-                             climsim=NA, model=NA)
-##  IPM 
-path <- "./ipm/simulations/results/climchange_meanparams/"
-ipmfiles_mean <- list.files(path)
-ipmfiles_mean <- ipmfiles_mean[grep("cover", ipmfiles_mean)]
-for(file in ipmfiles_mean){
-  tmp <- read.csv(paste(path,file,sep=""))
-  tmp$model <- "IPM"
-  meanparam_ipm <- rbind(meanparam_ipm, tmp[,c("cover", "species", "climsim", "model")])
-}
-meanparam_ipm <- meanparam_ipm[2:nrow(meanparam_ipm),]
-
-newd <- data.frame(cover=NA, species=NA, 
-                   climsim=NA, model=NA, obscover=NA)
-for(dospp in unique(meanparam_ipm$species)){
-  tmp <- subset(meanparam_ipm, species==dospp)
-  obscover <- subset(tmp, climsim=="obs")["cover"]
-  tmp2 <- subset(tmp, climsim!="obs")
-  tmp2$obscover <- rep(obscover$cover, times = length(unique(tmp2$climsim)))
-  newd <- rbind(newd, tmp2)
-}
-ipm_meansims <- newd[2:nrow(newd),]
-ipm_meansims$propdiff <- with(ipm_meansims, log(cover*100)-log(obscover*100))
-
-##  QBM
-meanparam_qbm <- data.frame(cover=NA, species=NA, 
-                             climsim=NA, model=NA)
-path <- "./quadBM/simulations/results/climatechange_meanparams/"
-qbmfiles_mean <- list.files(path)
-for(file in qbmfiles_mean){
-  tmp <- readRDS(paste(path,file,sep=""))
-  tmp$model <- "QBM"
-  meanparam_qbm <- rbind(meanparam_qbm, tmp[,c("cover", "species", "climsim", "model")])
-}
-meanparam_qbm <- meanparam_qbm[2:nrow(meanparam_qbm),]
-
-newd <- data.frame(cover=NA, species=NA, 
-                   climsim=NA, model=NA, obscover=NA)
-for(dospp in unique(meanparam_qbm$species)){
-  tmp <- subset(meanparam_qbm, species==dospp)
-  obscover <- subset(tmp, climsim=="obs")["cover"]
-  tmp2 <- subset(tmp, climsim!="obs")
-  tmp2$obscover <- rep(obscover$cover, times = length(unique(tmp2$climsim)))
-  newd <- rbind(newd, tmp2)
-}
-qbm_meansims <- newd[2:nrow(newd),]
-qbm_meansims$propdiff <- with(qbm_meansims, log(cover*100)-log(obscover*100))
-
-##  Merge IPM and QBM mean parameter results
-bothmodel_outs <- rbind(ipm_meansims, qbm_meansims)
-meanparam_agg <- ddply(bothmodel_outs, .(species, climsim, model), summarise,
-                       meandiff = mean(propdiff),
-                       hidiff = quantile(propdiff, probs = 0.95),
-                       lodiff = quantile(propdiff, probs = 0.05))
-setwd("~/Repos/MicroMesoForecast/manuscript")
-```
-
-```{r comment_out, echo=FALSE, eval=FALSE}
-# Equilibrium cover from both models was sensitive to climate (Fig.
-3a-d).
-The IPM projected percent changes in equilibrium cover from `r round(min(subset(meanparam_agg, species=="BOGR" & model=="IPM")["meandiff"])*100)` to `r round(max(subset(meanparam_agg, species=="BOGR" & model=="IPM")["meandiff"])*100)`% for *B.
-gracilis*, `r round(min(subset(meanparam_agg, species=="HECO" & model=="IPM")["meandiff"])*100)` to `r round(max(subset(meanparam_agg, species=="HECO" & model=="IPM")["meandiff"])*100)`% for *H.
-comata*, `r round(min(subset(meanparam_agg, species=="PASM" & model=="IPM")["meandiff"])*100)` to `r round(max(subset(meanparam_agg, species=="PASM" & model=="IPM")["meandiff"])*100)`% for *P.
-smithii*, and `r round(min(subset(meanparam_agg, species=="POSE" & model=="IPM")["meandiff"])*100)` to `r round(max(subset(meanparam_agg, species=="POSE" & model=="IPM")["meandiff"])*100)`% for *P.
-secunda*.
-The QBM projected opposite and greater percent changes in equilibrium cover for *B.
-gracilis* (`r round(min(subset(meanparam_agg, species=="BOGR" & model=="QBM")["meandiff"])*100)` to `r round(max(subset(meanparam_agg, species=="BOGR" & model=="QBM")["meandiff"])*100)`%) and *H.
-comata* (`r round(min(subset(meanparam_agg, species=="HECO" & model=="QBM")["meandiff"])*100)` to `r round(max(subset(meanparam_agg, species=="HECO" & model=="QBM")["meandiff"])*100)`%; Fig.
-3a-b).
-For *P.
-smithii*, the QBM projected opposite changes in equilibrium cover than the IPM, but of similar magnitude (`r round(min(subset(meanparam_agg, species=="PASM" & model=="QBM")["meandiff"])*100)` to `r round(max(subset(meanparam_agg, species=="PASM" & model=="QBM")["meandiff"])*100)`%; Fig.
-3c).
-*P.
-secunda* was the only species that the IPM and QBM made projections of the same sign and somewhat similar magnitude (`r round(min(subset(meanparam_agg, species=="POSE" & model=="QBM")["meandiff"])*100)` to `r round(max(subset(meanparam_agg, species=="POSE" & model=="QBM")["meandiff"])*100)`%; Fig.
-3d).
-```
-
 Discussion
@@
 \end{table}
-
-\pagebreak{}
-
-Figure Legends
---------------
-**Figure 1.** Time series of average percent cover over all quadrats for our four focal species: *Bouteloua gracilis* (BOGR), *Hesperostipa comata* (HECO), *Pascopyrum smithii* (PASM), and *Poa secunda* (POSE).
-Light grey lines show trajectories of individual quadrats.
-Note the different y-axis scales across panels.
-
-Figure 2. Work flow of the data aggregation, model fitting, and population simulating.
-
-Figure 3.
-Sensitivity of equilibrium cover simulated from the IPM to each climate scenario applied to individual and combined vital rates.
-For example, the points associated with G show the median cover from IPM simulations where a climate perturbation is applied only to the growth regression climate covariates.
-These simulations use mean parameter values for clarity.
-
-Figure 4.
-Effect of quadrat sample size on the precision (standard deviation) of main climate effect estimates in the QBM.
-Increasing the number of quadrats results in diminishing returns in terms of parameter certainty.
-Light dashed lines show individual climate effects at five quadrat sample sizes.
-Thick dark lines are inverse gaussian fits showing the mean effect of increasing quadrat sample size on parameter precision.
-
-Figure 5.
-Correlations (r) between QBM and IPM estimates of climate effects.
-We ignore sizeXclimate interactions since these are not directly comparable across model types.
-The QBM does not have multiple vital rates, so its values are repeated across panels within each species.
-Across top panels, 'growth' = growth regression, 'rec' = recruitment regression, 'surv' = survival regression.
-
-Figure 6.
-Mean (points) and 90% quantiles (errorbars) for the proportional difference between baseline simulations (using observed climate) and the climate pertubation simulation on the x-axis.
-We calculated proportional difference as log(perturbed climate cover) - log(observed climate cover), where 'perturbed' and 'observed' refer to the climate time series used to drive interannual variation in the simulations.
-Model error and parameter uncertainty were propagated through the simulation phase.
-Climate simulations are as in Figure 3.
 
@@
 library(ggthemes)
-obs_data <- read.csv("../analysis/speciesData/quadAllCover.csv")
+obs_data <- read.csv("../analysis/data_processing/speciesData/quadAllCover.csv")
 #remove zeros for averaging
@@
 
-```{r figure_2,dependson="plot-options", fig.cap="Work flow of the data aggregation, model fitting, and population simulating.", fig.height=4, eval=T}
-library(png)
-library(grid)
-img <- readPNG("components/figure/micromeso_flowchart.png")
-grid.raster(img)
-```
+\begin{figure}[!ht]
+  \centering
+      \includegraphics[width=4in]{./components/micromeso_flowchart.png}
+  \caption{Work flow of the data aggregation, model fitting, and population simulating.}
+\end{figure}
 
@@
   \centering
-      \includegraphics{./components/climate_posteriors_figure.png}
+      \includegraphics[height=7in]{./components/climate_posteriors_figure.png}
   \caption{Posterior distributions of climate effects ($\boldsymbol{\beta}_C$) for each species and vital rate statistical model.
@@
 \newpage{}
+
+\begin{figure}[!ht]
+  \centering
+      \includegraphics{./components/climate_effect_corplots.png}
+  \caption{Correlations between QBM and IPM estimates of climate effects.
+We ignore sizeXclimate interactions since these are not directly comparable across model types.
+The QBM does not have multiple vital rates, so its values are repeated across panels within each species.
+Across top panels, 'growth' = growth regression, 'rec' = recruitment regression, 'surv' = survival regression.}
+\end{figure}
+
+\newpage{}
+
+\begin{figure}[!ht]
+  \centering
+      \includegraphics[height=6in]{./components/climate_change_results.png}
+  \caption{Mean (points) and 90\% quantiles (errorbars) for the proportional difference between baseline simulations (using observed climate) and the climate pertubation simulation on the x-axis.
+We calculated proportional difference as log(perturbed climate cover) - log(observed climate cover), where 'perturbed' and 'observed' refer to the climate time series used to drive interannual variation in the simulations.
+Model error and parameter uncertainty were propagated through the simulation phase.}
+\end{figure}
+
+\pagebreak{}
 
@@
 
-```{r figure_5, dependson="plot-options", fig.cap="Correlations (r) between QBM and IPM estimates of climate effects.
-We ignore sizeXclimate interactions since these are not directly comparable across model types.
-The QBM does not have multiple vital rates, so its values are repeated across panels within each species.
-Across top panels, 'growth' = growth regression, 'rec' = recruitment regression, 'surv' = survival regression.", fig.width=6, fig.height=6, cache=FALSE, results='hide', eval=F}
-##  Figure script to look at correlations between IPM and QBM
-##    climate effect estimates.
-
-library(ggmcmc)
-library(gridExtra)
-library(ggplot2)
-
-allfiles <- list.files("../analysis/speciesData/")
-removals <- grep("csv", allfiles)
-species_list <- allfiles[-removals]
-
-clim_covs <- c("pptLag", "ppt1", "ppt2", "TmeanSpr1", "TmeanSpr2",
-               "ppt1xTmeanSpr1", "ppt2xTmeanSpr2",
-               "sizeXpptLag", "sizeXppt1", "sizeXppt2",
-               "sizeXTmeanSpr1", "sizeXTmeanSpr2")
-clim_mains <- c("pptLag", "ppt1", "ppt2", "TmeanSpr1", "TmeanSpr2",
-                "ppt1xTmeanSpr1", "ppt2xTmeanSpr2")
-
-####
-##  Get mean growth climate effects
-####
-growth_parms <- data.frame(species=NA, model=NA, parameter=NA, value=NA)
-for(spp in species_list){
-  tmp_data <- readRDS(paste("../analysis/ipm/vitalRateRegs/growth/growth_stanmcmc_", spp, ".RDS", sep=""))
-  tmp_clim <- tmp_data[grep("b2", tmp_data[,"Parameter"]), ]
-  tmp_clim[,"Parameter"] <- rep(clim_covs, each=3000)
-  tmp_agg <- ddply(tmp_clim, .(Parameter), summarise,
-                   value = median(value))
-  tmp_agg2 <- tmp_agg[which(tmp_agg$Parameter %in% clim_mains), ]
-  tmp_agg2$species <- spp
-  tmp_agg2$model <- "ipm"
-  colnames(tmp_agg2) <- tolower(colnames(tmp_agg2))
-  growth_parms <- rbind(growth_parms,
-                        tmp_agg2[,c("species", "model", "parameter", "value")])
-}
-
-####
-##  Get mean survival climate effects
-####
-surv_parms <- data.frame(species=NA, model=NA, parameter=NA, value=NA)
-params <- character(length = 12)
-for(i in 1:12){
-  params[i] <- paste("b2[",i,"]", sep="")
-}
-for(spp in species_list){
-  tmp_data <- readRDS(paste("../analysis/ipm/vitalRateRegs/survival/survival_stanmcmc_", spp, ".RDS", sep=""))
-  keeps <- which(tmp_data$Parameter %in% params)
-  tmp_clim <- tmp_data[keeps, ]
-  tmp_clim[,"Parameter"] <- rep(clim_covs, each=3000)
-  tmp_agg <- ddply(tmp_clim, .(Parameter), summarise,
-                   value = median(value))
-  tmp_agg2 <- tmp_agg[which(tmp_agg$Parameter %in% clim_mains), ]
-  tmp_agg2$species <- spp
-  tmp_agg2$model <- "ipm"
-  colnames(tmp_agg2) <- tolower(colnames(tmp_agg2))
-  surv_parms <- rbind(surv_parms,
-                        tmp_agg2[,c("species", "model", "parameter", "value")])
-}
-
-####
-##  Get mean recruitment climate effects
-####
-rec_parms <- data.frame(species=NA, model=NA, parameter=NA, value=NA)
-params <- character(length = 12)
-for(i in 1:12){
-  params[i] <- paste("b2[",i,"]", sep="")
-}
-for(spp in species_list){
-  tmp_data <- readRDS(paste("../analysis/ipm/vitalRateRegs/recruitment/recruitment_stanmcmc_", spp, ".RDS", sep=""))
-  keeps <- which(tmp_data$Parameter %in% params)
-  tmp_clim <- tmp_data[keeps, ]
-  tmp_clim[,"Parameter"] <- rep(clim_mains, each=3000)
-  tmp_agg <- ddply(tmp_clim, .(Parameter), summarise,
-                   value = median(value))
-  tmp_agg2 <- tmp_agg[which(tmp_agg$Parameter %in% clim_mains), ]
-  tmp_agg2$species <- spp
-  tmp_agg2$model <- "ipm"
-  colnames(tmp_agg2) <- tolower(colnames(tmp_agg2))
-  rec_parms <- rbind(rec_parms,
-                      tmp_agg2[,c("species", "model", "parameter", "value")])
-}
-
-####
-##  Get QBM growth climate estimates
-####
-clim_covs2 <- c("pptLag", "ppt1", "ppt2", "TmeanSpr1", "TmeanSpr2",
-               "ppt1xTmeanSpr1", "ppt2xTmeanSpr2",
-               "coverXpptLag", "coverXppt1", "coverXppt2",
-               "coverXTmeanSpr1", "coverXTmeanSpr2")
-
-qbm_parms <- data.frame(species=NA, model=NA, parameter=NA, qbmvalue=NA)
-params <- character(length = 12)
-for(i in 1:12){
-  params[i] <- paste("b2[",i,"]", sep="")
-}
-for(spp in species_list){
-  tmp_data <- readRDS(paste("../analysis/quadBM/vitalRateRegressions/truncNormModel/popgrowth_stanmcmc_", spp, ".RDS", sep=""))
-  keeps <- which(tmp_data$Parameter %in% params)
-  tmp_clim <- tmp_data[keeps, ]
-  tmp_clim[,"Parameter"] <- rep(clim_covs2, each=3000)
-  tmp_agg <- ddply(tmp_clim, .(Parameter), summarise,
-                   qbmvalue = median(value))
-  tmp_agg2 <- tmp_agg[which(tmp_agg$Parameter %in% clim_mains), ]
-  tmp_agg2$species <- spp
-  tmp_agg2$model <- "qbm"
-  colnames(tmp_agg2) <- tolower(colnames(tmp_agg2))
-  qbm_parms <- rbind(qbm_parms,
-                     tmp_agg2[,c("species", "model", "parameter", "qbmvalue")])
-}
-
-####
-##  Format data, combine, and plot
-####
-growth_parms <- growth_parms[2:nrow(growth_parms), ]
-growth_parms$vitalrate <- "growth"
-surv_parms <- surv_parms[2:nrow(surv_parms), ]
-surv_parms$vitalrate <- "surv"
-rec_parms <- rec_parms[2:nrow(rec_parms), ]
-rec_parms$vitalrate <- "rec"
-ipm_parms <- rbind(growth_parms, surv_parms, rec_parms)
-
-qbm_parms <- qbm_parms[2:nrow(qbm_parms), ]
-qbm_parms$vitalrate <- "growth"
-vitals <- c("surv", "rec")
-qbmalone <- qbm_parms
-for(dov in vitals){
-  tmp <- qbmalone
-  tmp$vitalrate <- dov
-  qbm_parms <- rbind(qbm_parms, tmp)
-}
-
-ipm_parms <- ipm_parms[,c("species","parameter","value","vitalrate")]
-qbm_parms <- qbm_parms[,c("species","parameter","qbmvalue","vitalrate")]
-all_ests <- merge(ipm_parms, qbm_parms, by = c("species", "parameter", "vitalrate"))
-
-# Calculate correlation for each group
-cors <- ddply(all_ests, c("species", "vitalrate"), summarise, 
-              cor = round(cor(value, qbmvalue), 2))
-
-ggplot(all_ests, aes(x=qbmvalue, y=value))+
-  geom_point(size=3)+
-  geom_smooth(method="lm", color="black", fill="grey", se=FALSE)+
-  facet_grid(species~vitalrate)+
-  geom_text(data=cors, aes(label=paste("r = ", cor, sep="")), x=-1, y=1, size=4)+
-  xlab("QBM Estimate")+
-  ylab("IPM Estimate")+
-  theme_bw()
-
-setwd("~/Repos/MicroMesoForecast/manuscript/")
-```
-
-```{r figure_6, dependson="plot-options", fig.cap="Mean (points) and 90% quantiles (errorbars) for the proportional difference between baseline simulations (using observed climate) and the climate pertubation simulation on the x-axis.
-We calculated proportional difference as log(perturbed climate cover) - log(observed climate cover), where 'perturbed' and 'observed' refer to the climate time series used to drive interannual variation in the simulations.
-Model error and parameter uncertainty were propagated through the simulation phase.
-Climate simulations are as in Figure 3.", fig.width=4.5, fig.height=7, cache=FALSE, results='hide', eval=F}
-####
-##  Vary parameter runs
-####
-setwd("~/Repos/MicroMesoForecast/analysis/")
-meanparam_ipm <- data.frame(cover=NA, species=NA, 
-                             climsim=NA, model=NA)
-##  IPM 
-path <- "./ipm/simulations/results/climchange_varyparams/"
-ipmfiles_mean <- list.files(path)
-ipmfiles_mean <- ipmfiles_mean[grep("cover", ipmfiles_mean)]
-torm <- c(grep("10", ipmfiles_mean), grep("20", ipmfiles_mean), grep("30", ipmfiles_mean))
-ipmfiles_mean <- ipmfiles_mean[-torm]
-for(file in ipmfiles_mean){
-  tmp <- read.csv(paste(path,file,sep=""))
-  tmp$model <- "IPM"
-  meanparam_ipm <- rbind(meanparam_ipm, tmp[,c("cover", "species", "climsim", "model")])
-}
-meanparam_ipm <- meanparam_ipm[2:nrow(meanparam_ipm),]
-
-newd <- data.frame(cover=NA, species=NA, 
-                   climsim=NA, model=NA, obscover=NA)
-for(dospp in unique(meanparam_ipm$species)){
-  tmp <- subset(meanparam_ipm, species==dospp)
-  obscover <- subset(tmp, climsim=="obs")["cover"]
-  tmp2 <- subset(tmp, climsim!="obs")
-  tmp2$obscover <- rep(obscover$cover, times = length(unique(tmp2$climsim)))
-  newd <- rbind(newd, tmp2)
-}
-ipm_meansims <- newd[2:nrow(newd),]
-ipm_meansims$propdiff <- with(ipm_meansims, log(cover*100)-log(obscover*100))
-# ggplot(subset(ipm_meansims, species=="PASM"))+
-#   geom_histogram(aes(x=cover))+
-#   facet_wrap("climsim", scales = "free")+
-#   scale_x_continuous(limits=c(0,0.05))
-##  QBM
-meanparam_qbm <- data.frame(cover=NA, species=NA, 
-                             climsim=NA, model=NA)
-path <- "./quadBM/simulations/results/climatechange_varyparams/"
-qbmfiles_mean <- list.files(path)
-for(file in qbmfiles_mean){
-  tmp <- readRDS(paste(path,file,sep=""))
-  tmp$model <- "QBM"
-  meanparam_qbm <- rbind(meanparam_qbm, tmp[,c("cover", "species", "climsim", "model")])
-}
-meanparam_qbm <- meanparam_qbm[2:nrow(meanparam_qbm),]
-
-newd <- data.frame(cover=NA, species=NA, 
-                   climsim=NA, model=NA, obscover=NA)
-for(dospp in unique(meanparam_qbm$species)){
-  tmp <- subset(meanparam_qbm, species==dospp)
-  obscover <- subset(tmp, climsim=="obs")["cover"]
-  tmp2 <- subset(tmp, climsim!="obs")
-  tmp2$obscover <- rep(obscover$cover, times = length(unique(tmp2$climsim)))
-  newd <- rbind(newd, tmp2)
-}
-qbm_meansims <- newd[2:nrow(newd),]
-qbm_meansims$propdiff <- with(qbm_meansims, log(cover*100)-log(obscover*100))
-
-##  Merge IPM and QBM mean parameter results
-bothmodel_outs <- rbind(ipm_meansims, qbm_meansims)
-
-meanparam_agg <- ddply(bothmodel_outs, .(species, climsim, model), summarise,
-                       meandiff = mean(propdiff),
-                       hidiff = quantile(propdiff, probs = 0.95),
-                       lodiff = quantile(propdiff, probs = 0.05))
-dgd = position_dodge(width = 0.6)
-ggplot(meanparam_agg)+
-  geom_errorbar(aes(x=climsim, ymax=hidiff, ymin=lodiff, group=model, linetype=model),
-                position=dgd, width=0.25)+
-  geom_point(aes(x=climsim, y=meandiff, group=model, shape=model),
-                position=dgd, width=0.25)+
-  theme_bw()+
-  scale_linetype_discrete(name="Model")+
-  scale_shape_discrete(name="Model")+
-  facet_wrap("species", ncol=1, scales="free")+
-  scale_x_discrete(labels=c("+ppt", "+temp", "+ppt&temp"))+
-  ylab("Proportional difference")+
-  xlab("Climate simulation")
-#   theme(axis.text.x = element_text(angle = 45, hjust = 1))
-setwd("~/Repos/MicroMesoForecast/manuscript/")
-```
-
-\pagebreak{}
-
 \singlespace{}

### 2016-04-27 manuscript/manuscript.Rmd
modifies forecast horizons to be from mean parameter runs, for now; adds plot to ms

@@
   \centering
+      \includegraphics[height=5in]{./components/forecast_horizon.png}
+  \caption{The forecast horizons for both models when used to forecast the final observation (cover in year 1945) starting from all possible prior years.
+For example, simulations initialized with cover in 1932 have 13 years before the forecast.}
+\end{figure}
+
+\newpage{}
+
+\begin{figure}[!ht]
+  \centering
       \includegraphics{./components/climate_effect_corplots.png}

### 2016-04-27 FigureScripts/plot_forecast_horizons.R

@@ -55,7 +55,7 @@ out_qbm <- data.frame(yearstart=NA, mean_error=NA, sd_error=NA, rmse=NA, species
 for(ispp in 1:length(species_list)){
   tmp_file <- paste0(species_list[ispp],"_final_year_cover.RDS")
   tmp_sims <- readRDS(paste0(path2qbms,tmp_file))
-  tmp_sims <- tmp_sims[2:nrow(tmp_sims),]
+  # tmp_sims <- tmp_sims[2:nrow(tmp_sims),]
   tmp_sims$error <- with(tmp_sims, finalcover-obs_finalcover)
   tmp_agg <- ddply(tmp_sims, .(yearstart), summarise,
                    mean_error = mean(abs(error)),
@@ -67,9 +67,9 @@ for(ispp in 1:length(species_list)){
 out_qbm <- out_qbm[2:nrow(out_qbm),] # remove first NA-filled row
 
 ##  Combine the results
-out_ipm$model <- "BIPM"
+out_ipm$model <- "AIPM"
 out_ipm$yearsbefore <- with(out_ipm, max(startyear)-startyear+1)
-out_qbm$model <- "AQBM"
+out_qbm$model <- "BQBM"
 out_qbm$yearsbefore <- with(out_qbm, max(yearstart)-yearstart+1)
 names(out_qbm) <- names(out_ipm)
 out_all <- rbind(out_ipm, out_qbm)
@@ -81,21 +81,19 @@ out_all <- rbind(out_ipm, out_qbm)
 ####
 mycols <- c("#969C43", "#B46FA1")
 mycols <- c("grey50", "black")
-ggplot(data=out_all, aes(x=yearsbefore, y=mean_error*100, color=model, fill=model))+
-  geom_ribbon(aes(x=yearsbefore, ymax=(mean_error+sd_error)*100, 
-                  ymin=(mean_error-sd_error)*100), alpha=0.3, color=NA)+
-  geom_line()+
-  geom_point(size=2)+
+ggplot(data=out_all, aes(x=yearsbefore, y=mean_error*100))+
+#   geom_ribbon(aes(x=yearsbefore, ymax=(mean_error+sd_error)*100, 
+#                   ymin=(mean_error-sd_error)*100), alpha=0.3, color=NA)+
+  geom_line(aes(linetype=model))+
+  geom_point(aes(shape=model),size=3)+
   facet_grid(species~., scales="free")+
   xlab("Years Before Forecast")+
[…]

### 2016-04-27 analysis/quadBM/simulations/forecast_horizon_qbm.R

@@ -11,6 +11,7 @@
 rm(list=ls())
 
 ##  Set number of simulations per year
+use_mean_parameters <- TRUE
 NumberSimsPerYear <- 50
 scalers <- readRDS("../../qbm_all_clim_scalers.RDS")
 
@@ -26,55 +27,9 @@ library('EnvStats')
 ####  Read in data -------------------------------------
 ####
 ##  Observations
-allD <- read.csv("../../speciesData/quadAllCover.csv")
-allD <- allD[,2:ncol(allD)] #get rid of X ID column
-sppList <- as.character(unique(allD$Species))
-
 climD <- read.csv("../../weather/Climate.csv")
-
-backD <- data.frame(climYear=NA,
-                    quad = NA,
-                    year= NA,
-                    totCover= NA,
-                    Species= NA,
-                    propCover= NA,
-                    lag.cover= NA,
-                    pptLag= NA,
-                    ppt1= NA,
-                    TmeanSpr1= NA,
-                    ppt2= NA,
-                    TmeanSpr2= NA,
-                    TmeanSum1= NA,
-                    TmeanSum2= NA,
-                    yearID= NA,
-                    group = NA,
-                    percCover = NA,
-                    percLagCover = NA)
-
-#loop through species and remake data frame
-for(spp in 1:length(sppList)){
-  doSpp <- sppList[spp]
-  sppD <- subset(allD, Species==doSpp)
-  
-  # create lag cover variable
-  tmp=sppD[,c("quad","year","totCover")]
-  tmp$year=tmp$year+1
-  names(tmp)[3]="lag.cover"
-  sppD=merge(sppD,tmp,all.x=T)
-  
-  # merge in climate data
-  sppD$climYear=sppD$year+1900-1  
-  sppD=merge(sppD,climD,by.x="climYear",by.y="year")
-  
-  #Growth observations
-  growD <- subset(sppD,lag.cover>0 & totCover>0)
-  growD$yearID <- growD$year #for random year offset on intercept
-  growD$group <- substring(growD$quad, 1, 1)
-  growD$percCover <- growD$totCover/10000
-  growD$percLagCover <- growD$lag.cover/10000
-  backD <- rbind(backD, growD)
-}#end species loop
-obs_data <- backD[2:nrow(backD),]
+obs_data <- readRDS("../../processed_data/cover_with_weather.RDS")
+sppList <- sort(unique(obs_data$species))
[…]

### 2016-04-27 FigureScripts/clim_eff_correlations.R
adds corrs to correlation plot; adds figure mock up file for posteriors

@@ -143,16 +143,12 @@ cors <- ddply(all_ests, c("species", "vitalrate"), summarise,
 
 library(ggthemes)
 ggplot(all_ests, aes(x=qbmvalue, y=value))+
-#   annotate("rect", xmin = Inf, xmax = 0, ymin = Inf, ymax = 0, fill= "white")  + 
-#   annotate("rect", xmin = -Inf, xmax = 0, ymin = -Inf, ymax = 0 , fill= "white") + 
-#   annotate("rect", xmin = 0, xmax = Inf, ymin = 0, ymax = -Inf, fill= "grey") + 
-#   annotate("rect", xmin = 0, xmax = -Inf, ymin = Inf, ymax = 0, fill= "grey") + 
   geom_hline(aes(yintercept=0), color="grey", linetype=2)+
   geom_vline(aes(xintercept=0), color="grey", linetype=2)+
   geom_point(size=3)+
   geom_smooth(method="lm", color="black", se=FALSE)+
   facet_wrap(species~vitalrate, scales="free", nrow=4)+
-  # geom_text(data=cors, aes(label=paste("r = ", cor, sep="")), x=-.05, y=0.12, size=6)+
+  geom_text(data=cors, aes(label=paste("r = ", cor, sep="")), x=mean(qbmvalue), y=mean(value), size=6)+
   xlab("QBM Estimate")+
   ylab("IPM Estimate")+
   theme_few()

### 2016-04-28 manuscript/manuscript.Rmd
adds corrs to correlation plot; adds figure mock up file for posteriors
updates ms formatting for MEE

@@
    - \usepackage[document]{ragged2e}
+   - \usepackage{times}
 bibliography: ~/Dropbox/Bibliography/MicroMeso.bib
@@
 ---
+
+\textsf{\Large{\textbf{Do we need demographic data to forecast population \\ responses to climate change?}}}
+
 \renewcommand*{\thefootnote}{\fnsymbol{footnote}}
 
-\begin{centering}
+\vspace{1em}
 
-\textbf{\Large{Do we need demographic data to forecast population responses to climate change?}}
+\textsf{\normalsize{\textbf{Andrew T.
+Tredennick\textsuperscript{1}\footnote{Corresponding author: E-mail: atredenn@gmail.com}, Mevin B.
+Hooten\textsuperscript{2,3,4}, and Peter B.
+Adler\textsuperscript{1}}}}
 
-\textsc{\normalsize{Andrew T. Tredennick\footnote{Corresponding author: atredenn@gmail.com}\textsuperscript{1}, Peter B. Adler\textsuperscript{1}, and Mevin B. Hooten\textsuperscript{2,3,4}}}
+\vspace{1em}
 
-\textit{\small{\textsuperscript{1}Department of Wildland Resources and the Ecology Center, 5230 Old Main Hill, Utah State University, Logan, Utah 84322 USA}} \\
-\textit{\small{\textsuperscript{2}U.S. Geological Survey, Colorado Cooperative Fish and Wildlife Research Unit, Colorado State University, Fort Collins, CO 80523 USA}} \\
-\textit{\small{\textsuperscript{3}Department of Fish, Wildlife, and Conservation Biology, Colorado State University, Fort Collins, CO 80523 USA}} \\
-\textit{\small{\textsuperscript{4}Department of Statistics, Colorado State University, Fort Collins, CO 80523 USA}}
+\textsf{\textit{\small{\textsuperscript{1}Department of Wildland Resources and the Ecology Center, 5230 Old Main Hill, Utah State University, Logan, Utah 84322, USA; \textsuperscript{2}U.S.
+Geological Survey, Colorado Cooperative Fish and Wildlife Research Unit, Department of Fish, Wildlife and Conservation Biology, Colorado State University, Fort Collins, CO 80523-1403, USA; \textsuperscript{3}Department of Fish, Wildlife, and Conservation Biology, Colorado State University, Fort Collins, CO 80523 USA}; \textsuperscript{4}Department of Statistics, Colorado State University, Fort Collins, CO 80523 USA}}
 
-\end{centering}
-
+\allsectionsfont{\normalfont\sffamily\bfseries}
 \renewcommand*{\thefootnote}{\arabic{footnote}}
@@
 
-<!-- For J. Ecol
-\allsectionsfont{\normalfont\sffamily\bfseries}
--->
+\rule{\textwidth}{1pt}
 
@@
 
- \renewcommand{\abstractname}{}
+Summary
+-------
 
-\begin{abstract}
-\textit{Abstract.} \hspace{1em} Rapid climate change has generated growing interest in forecasts of future population trajectories.
+**1.** Rapid climate change has generated growing interest in forecasts of future population trajectories.
 Traditional population models, typically built using detailed demographic observations from one study site, can address climate change impacts at one location, but are difficult to scale up to the landscape and regional scales relevant to management decisions.
-An alternative is to build models using population-level data that are much easier to collect over broad spatial scales than individual-level data.
+
+**2.** An alternative is to build models using population-level data that are much easier to collect over broad spatial scales than individual-level data.
 However, such models ignore the fact that climate drives population growth through its influence on individual performance.
-Here, we test the consequences of aggregating individual responses when forecasting climate change impacts on four perennial grass species in a semi-arid grassland in Montana, USA.
+
+**3.** Here, we test the consequences of aggregating individual responses when forecasting climate change impacts on four perennial grass species in a semi-arid grassland in Montana, USA.
 We parameterized two population models for each species, one based on individual-level data (survival, growth and recruitment) and one on population-level data (percent cover), and compared their accuracy, precision, and sensitivity to climate variables.
 For both models we used Bayesian ridge regression to identify the optimal predictive model in terms of climate covariate strengths. 
-The individual-level model was more accurate and precise than the aggregated model when predicting out-of-sample observations.
+**4.** The individual-level model was more accurate and precise than the aggregated model when predicting out-of-sample observations.
 When comparing climate effects from both models, the population-level model missed important climate effects from at least one vital rate for each species.
 Increasing the sample size at the population-level would not necessarily reduce forecast uncertainty; the way to reduce uncertainty is to capture unique climate dependence of individual vital rates. 
-Our analysis indicates that there is no shortcut to forecasting climate change impacts on plant populations — detailed demographic data are essential.
+**5.** *Synthesis.* Our analysis indicates that there is no shortcut to forecasting climate change impacts on plant populations — detailed demographic data are essential.
 Despite the superiority of the individual-level model, the forecasts it generated still were too uncertain to be useful for decision-makers.
@@
 
-\textit{Key words: forecasting, climate change, grassland, integral projection model, population model}
-
-\end{abstract}
+\textsf{\textbf{Key-words:}} forecasting, climate change, grassland, integral projection model, population model, statistical regularization, ridge regression
 
@@
   \centering
-      \includegraphics{./components/climate_effect_corplots.png}
+      \includegraphics[height=6in]{./components/climate_effect_corplots.png}
   \caption{Correlations between QBM and IPM estimates of climate effects.

### 2016-04-28 FigureScripts/climate_change_figure.R

@@ -125,6 +125,33 @@ ggplot(diffs_aggregated, aes(x=climsim, y=mean_prop_diff))+
   ylab("Proportional Change in Cover")+
   xlab("Climate Simulation")+
   theme_few()
-ggsave(paste0(path2figs, "climate_change_results.png"), width = 4.5, height = 7, units = "in", dpi = 100)
+# ggsave(paste0(path2figs, "climate_change_results.png"), width = 4.5, height = 7, units = "in", dpi = 100)
+
+
+
+####
+####  DENSITY PLOT
+####
+ipm_diffs2 <- ipm_diffs[,c("species", "climsim","propdiff")]
+qbm_diffs2 <- qbm_diffs[,c("species", "climsim","propdiff")]
+ipm_diffs2$model <- "ipm"
+qbm_diffs2$model <- "qbm"
+diffs <- rbind(ipm_diffs2, qbm_diffs2)
+
+# Change some names for the plot
+diffs[which(diffs$climsim=="pptChange"), "climsim"] <- "Appt"
+diffs[which(diffs$climsim=="TmeanChange"), "climsim"] <- "Btmean"
+diffs[which(diffs$climsim=="pptTmeanChange"), "climsim"] <- "Cboth"
+
+ggplot(diffs, aes(x=climsim, y=propdiff, fill=model))+
+  geom_violin()+
+  geom_hline(aes(yintercept=0), color="black", linetype=2)+
+  facet_wrap("species", ncol=2, scales="free_y")+
+  scale_x_discrete(labels=c("+ppt", "+temp", "+ppt&temp"))+
+  scale_fill_manual(name="Model", labels=c("IPM", "QBM"), values = c("grey75", "grey35"))+
+  ylab("Proportional Change in Cover")+
+  xlab("Climate Simulation")+
+  theme_few()
+
 
 

### 2016-05-13 manuscript/manuscript.Rmd
ms formatting
updates ms with new results and text; updates rho comparisons between models

@@
 
-\textsf{\Large{\textbf{Do we need demographic data to forecast population \\ responses to climate change?}}}
+\textsf{\LARGE{\textbf{Do we need demographic data to forecast the state of plant populations?}}}
 
@@
 
-\textsf{\textit{\small{\textsuperscript{1}Department of Wildland Resources and the Ecology Center, 5230 Old Main Hill, Utah State University, Logan, Utah 84322, USA; \textsuperscript{2}U.S.
+\textsf{\textit{\footnotesize{\textsuperscript{1}Department of Wildland Resources and the Ecology Center, 5230 Old Main Hill, Utah State University, Logan, Utah 84322, USA; \textsuperscript{2}U.S.
 Geological Survey, Colorado Cooperative Fish and Wildlife Research Unit, Department of Fish, Wildlife and Conservation Biology, Colorado State University, Fort Collins, CO 80523-1403, USA; \textsuperscript{3}Department of Fish, Wildlife, and Conservation Biology, Colorado State University, Fort Collins, CO 80523 USA}; \textsuperscript{4}Department of Statistics, Colorado State University, Fort Collins, CO 80523 USA}}
@@
 
-**1.** Rapid climate change has generated growing interest in forecasts of future population trajectories.
-Traditional population models, typically built using detailed demographic observations from one study site, can address climate change impacts at one location, but are difficult to scale up to the landscape and regional scales relevant to management decisions.
+**1.** Rapid environmental change has generated growing interest in forecasts of future population trajectories.
+Traditional population models, typically built using detailed demographic observations from one study site, can address the impacts of environmental change at one location, but are difficult to scale up to the landscape and regional scales relevant to management decisions.
 
 **2.** An alternative is to build models using population-level data that are much easier to collect over broad spatial scales than individual-level data.
-However, such models ignore the fact that climate drives population growth through its influence on individual performance.
+However, it remains unclear if models built using aggregated individual-level data adequately capture the effects of density-dependence and environmental forcing that are necessary to generate skillful forecasts.
 
-**3.** Here, we test the consequences of aggregating individual responses when forecasting climate change impacts on four perennial grass species in a semi-arid grassland in Montana, USA.
-We parameterized two population models for each species, one based on individual-level data (survival, growth and recruitment) and one on population-level data (percent cover), and compared their accuracy, precision, and sensitivity to climate variables.
+**3.** Here, we test the consequences of aggregating individual responses when forecasting the population states and trajectories of four perennial grass species in a semi-arid grassland in Montana, USA.
+We parameterized two population models for each species, one based on individual-level data (survival, growth and recruitment) and one on population-level data (percent cover), and compared their forecasting skill and forecast horizons with and without the inclusion of climate covariates.
 For both models we used Bayesian ridge regression to identify the optimal predictive model in terms of climate covariate strengths.
-**4.** The individual-level model was more accurate and precise than the aggregated model when predicting out-of-sample observations.
-When comparing climate effects from both models, the population-level model missed important climate effects from at least one vital rate for each species.
-Increasing the sample size at the population-level would not necessarily reduce forecast uncertainty; the way to reduce uncertainty is to capture unique climate dependence of individual vital rates. 
-**5.** *Synthesis.* Our analysis indicates that there is no shortcut to forecasting climate change impacts on plant populations — detailed demographic data are essential.
-Despite the superiority of the individual-level model, the forecasts it generated still were too uncertain to be useful for decision-makers.
-We need new methods to collect demographic data efficiently across environmental gradients in space and time.
+
+**4.** Without climate effects included, we found no significant difference between the forecasting skill of models based on individual-level data and models based on population-level data.
+Climate effects were weak and caused only slight increases in forecasting skill.
+Increases in skill were similar between model types except for one species where forecast accuracy from the individual-level model was significantly higher than the accuracy from an equivalent population-level model.
+
+**5.** *Synthesis.* For our focal species at this particular location, and using our particular statistical models, demographic data was generally unnecessary to achieve skillful forecasts, though for certain species forecast skill can be gained by using demographic data linked to climate covariates.
+We conclude that models based on aggregated individual-level data offer a practical alternative to data-intensive demographic models when species do not respond strongly to interannual variation in weather, but when modeling species that do respond to climate drivers, demographically-based models can generate more skillful forecasts.
 
@@
 
+<!--
 The inability of most population models to address landscape-scale problems may explain why land managers and conservation planners have embraced species distribution models (SDMs) [see @Guisan2005 for a review].
@@
 Ideally, researchers would provide managers with landscape-scale population models, combining the extent of SDMs with information about dynamics and species abundances [@Schurr2012; @Merow2014].
+-->
 
@@
 Population-level data cannot provide inference about demographic mechanisms, but might be sufficient for modeling future population states, especially since such data are feasible to collect across broad spatial extents [e.g., @Queenborough2011].
-The choice between individual and population-level data involves a difficult trade-off: while individual-level data leads to more mechanistic models, population-level data leads to models that can be applied over greater spatial and temporal extents.
+The choice between individual and population-level data involves a difficult trade-off: while individual-level data leads to more mechanistic models, population-level data can lead to models that can be applied over greater spatial and temporal extents because the data are easier to collect over large spatial scales.
 An open question is how much forecasting skill is lost when we build models based on population rather than individual-level data.
@@
 Building on @Taylor2004, @Freckleton2011 showed that density-structured models compare well to continuous models in theory, and @Queenborough2011 provide empirical evidence that density-structured models are capable of reproducing population dynamics at landscape spatial scales, even if some precision is lost when compared to fully continuous models.
-The appeal of density-structured approaches is clear.
-However, none of these models included environmental covariates.
+However, previous tests of density-structured models have yet to assess their ability to forecast out-of-sample observations, and they have not included environmental covariates, which is necessary to make forecasts of population responses to climate change.
 
@@
 Here, we compare the forecasting skill of statistical and population models based on aggregated, population-level data with models based on individual-level data.
-We used a unique demographic dataset that tracks the fates of individual plants from four species over 14 years to build two kinds of single-species population models, traditional models using individual growth, survival, and recruitment data and alternative models based on basal cover.
-In both models, interannual variation is explained, in part, by climate covariates.
-We first quantify forecasting skill using cross-validation.
-We then performed simulations to quantify the sensitivities of species' cover to small perturbations in average precipitation and temperature.
-Based on the cross-validation results, predictions of individual level models were clearly better, but, unfortunately, still too uncertain to inform management decisions.
+We used a demographic dataset that tracks the fates of individual plants from four species over 14 years to build two kinds of single-species population models, traditional models using individual growth, survival, and recruitment data and alternative models based on basal cover.
+We use the models to answer two questions: (1) Can population models fit using aggregated individual-level data (percent cover) produce forecasts as skillful as those from models fit to demographic data?
+And (2) Can population models fit using aggregated data adequately capture the influence of climate on population growth and, in turn, produce forecasts as skillful as those from models fit to demographic data?
 
@@
 From 1932 to 1945 individual plants were identified and mapped annually in 44 1-$\text{m}^2$ quadrats using a pantograph.
-The quadrats were distributed in six pastures, each assigned a grazing treatment of light (1.24 ha/animal unit month), moderate (0.92 ha/aum), and heavy (0.76 ha/aum) stocking rates (two pastures per treatment).
+The quadrats were distributed among six pastures, each assigned a grazing treatment of light (1.24 ha/animal unit month), moderate (0.92 ha/aum), and heavy (0.76 ha/aum) stocking rates (two pastures per treatment).
 In this analysis we account for potential differences among the grazing treatments, but do not focus on grazing$\times$climate interactions.
@@
 Based on these two datasets we can compare population models built using individual-level data and aggregated, quadrat-level data.
-At the individual level we explicitly model three vital rates: growth (13 year-to-year transitions, 29 quadrats, 18,730 records), survival (13 years, 33 quadrats, 29,353 records), and recruitment (13 years, 33 quadrats, 304 records).
-At the quadrat level we model population growth as change in percent cover of quadrats with non-zero cover in year *t* and in year *t-1* (13 year-to-year transitions, 29 quadrats, 866 records).
+At the individual level we explicitly model three vital rates: growth (13 year-to-year transitions, 29 quadrats and 18,730 records across the four species), survival (13 years, 33 quadrats and 29,353 records across the four species), and recruitment (13 years, 33 quadrats and 304 records across the four species).
+At the quadrat level we model population growth as change in percent cover of quadrats with non-zero cover in year *t* and in year *t-1* (13 year-to-year transitions, 29 quadrats and 866 records across the four species).
 For modeling population growth at the quadrat level we ignore within-quadrat extirpation and colonization events because they are very rare in our time series ($N=16$ and $N=13$, respectively, across all species).
-Given the broad spatial distribution of the quadrats we are studying, it is safe to assume that these events are in fact rare enough to be ignored for our purposes.
+Given the relatively broad spatial distribution of the quadrats we are studying, it is safe to assume that these events are in fact rare enough to be ignored for our purposes.
 
@@
 
-[^revnote]: *Note to reviewers*: so that v1.0 will be associated with the published version of the manuscript, we have released v0.1 to be associated with this review version.
+[^revnote]: *Note to reviewers*: so that v1.0 will be associated with the published version of the manuscript, we have released v0.2 to be associated with this review version.
 
@@
 We describe the statistical models separately since fitting the models required different approaches.
-All models contain five climate covariates that we chose \emph{a priori}: "water year" precipitation at \emph{t}-2 (lagppt); April through June precipitation at \emph{t}-1 and \emph{t} (ppt1 and ppt2, respectively) and April through June temperature at \emph{t}-1 and \emph{t} (TmeanSpr1 and TmeanSpr2, respectively), where \emph{t}-1 to *t* is the transition of interest.
-We also include interactions among same-year climate covariates (e.g., ppt1 $\times$ TmeansSpr1).
+For both model types, we fit vital rate models with and without climate covariates.
+Models with climate effects contain five climate covariates that we chose \emph{a priori} based on previous model selection efforts using these data (Chu et al.
+*in press*) and expert advice (Lance Vermeire, *personal communication*): "water year" precipitation at \emph{t}-2 (lagppt); April through June precipitation at \emph{t}-1 and \emph{t} (ppt1 and ppt2, respectively) and April through June temperature at \emph{t}-1 and \emph{t} (TmeanSpr1 and TmeanSpr2, respectively), where \emph{t}-1 to *t* is the transition of interest.
+We also include interactions among same-year climate covariates (e.g., ppt1 $\times$ TmeansSpr1), resulting in a total of seven climate covariates.
 
@@
 All model parameters are species-specific, but we omit subscripts for species in model descriptions below to reduce visual clutter.
+For brevity, we only describe models with climate covariates included, but models without climate covariates are exactly the same as the models described below, just without the climate effects.
 
@@
 Quadrat random effects were also fit hierarchically, with quadrat offsets being drawn from distributions with mean zero and a shared variance term (independent Gaussian priors, Appendix A).
-Climate effects were not modeled hierarchically, and each was given a diffuse prior distribution.
-We used standard diffuse priors for all unknown parameters (Appendix A).
+Climate effects were modeled as independent covariates whose prior distributions were determined by statistical regularization (see **Statistical regularization: Bayesian ridge regression** below).
 
@@
 JAGS or WinBUGS), but the Stan sampler is exceptionally efficient, which reduces the number of iterations needed to achieve convergence.
-We assessed convergence visually and made sure scale reduction factors for all parameters were less than 1.01.
+We assessed convergence visually and made sure scale reduction factors for all parameters were less than 1.1.
 For the purposes of including parameter uncertainty in our population models, we saved the final 1,000 iterations from each of the three MCMC chains to be used as randomly drawn values during population simulation.
-This step alleviates the need to reduce model parameters by model selection since sampling from the full parameter space in the MCMC ensures that if a parameter broadly overlaps zero, on average the effect in the population models will also be near zero.
 We report the posterior mean, standard deviation, and 95% Bayesian Credible Intervals for every parameter of each model for each species in Appendix B.
@@
 #### Statistical regularization: Bayesian ridge regression
-Our objective is to model the response of our focal grass species to interannual variation in climate, even if those responses are weak.
+For models with climate covariates, our objective is to model the response of our focal grass species to interannual variation in climate, even if those responses are weak.
 Therefore, we avoid selecting among models with all possible combinations of climate covariates, and instead use Bayesian ridge regression to regulate, or constrain, the posterior distributions of each climate covariate [@Hooten2015].
@@
 
-We used random draws from the final 1,000 iterations from each of three MCMC chains to introduce stochasticity into our population models.
+We used random draws from the final 1,000 iterations, thinned by 10, from each of three MCMC chains to carry-through parameter uncertainty into our population models.
 At each time step, we randomly selected climate covariates from one of the 14 observed years.
 Then, we drew the full parameter set (climate effects and density-dependence fixed effects) from a randomly selected MCMC iteration.
-Using this approach, rather than simply using coefficient point estimates, captures the effect of parameter uncertainty.
 Relatively unimportant climate covariates (those that broadly overlap 0) will have little effect on the mean of the simulation results, but can contribute to their variation.
-Since our focus was on the contribution of climate covariates to population states, we set the random year effects and the random group effects to zero.
+Since our focus was on the contribution of density dependence and climate covariates to population states, we set the random year effects and the random group effects to zero.
 
@@
 ### Model validation
-To test each model's ability to forecast population state, we made out-of-sample predictions using leave-one-year-out cross validation.
-For both levels of modeling, we fit the vital rate models using observations from all years except one, and then used those fitted parameters in the population models to perform a one-step-ahead forecast for the year whose observations were withheld from model fitting.
+To test each model's ability to forecast population states we made out-of-sample predictions using leave-one-year-out cross validation.
+For both levels of modeling and for models with and without climate covariates, we fit the vital rate models using observations from all years except one, and then used those fitted parameters in the population models to perform a one-step-ahead forecast for the year whose observations were withheld from model fitting.
 Within each observation year, several quadrats were sampled.
@@
 
-This cross-validation procedure allowed us to compare accuracy and precision of the two modeling approaches (IPM versus QBM).
-We first calculated the median predicted cover across the 100 simulations for each quadrat-year and then calculated the absolute error as the absolute value of the difference between the observed cover for a given quadrat-year and the median prediction.
-To arrive at mean absolute error (MAE), we then averaged the absolute error within each species across the quadrat-year specific errors.
-We use MAE as our measure of accuracy.
-To measure precision we calculated the distance between the upper and lower 90th quantiles of the 100 predictions and averaged this value over quadrat-years for each species.
+This cross-validation procedure allowed us to compare accuracy and precision of the two modeling approaches (IPM versus QBM) with and without climate covariates.
+We first calculated the median predicted cover across the 100 simulations for each quadrat-year and then calculated forecast skill as the correlation ($\rho$) between forecasts and observations.
+We compared $\rho$ between model types and within model types between models with and without climate covariates using a one-sided *t* test with adjusted degrees of freedom following @Wilcox2009 and standard errors calculated using the HC4 estimator of @Cribari-Neto2004.
+Statistical tests for differences between $\rho$ were conducted using R functions from @Ye2015a.
 
+### Forecast horizons
+An important feature of any forecasting model is the rate at which forecast skill declines as the time between an observation and a forecast increases; the so-called ecological forecast horizon [@Petchey2015].
+To assess the forecast horizons of our models, we iniate the model with the population state at some time *t* and make sequential forecasts of the population at times $t+1, t+2, \dots, t+T$ where *T* is the maximum number of years between the initial year and the final year in our observations.
+For example, if we initialize the model with percent cover in 1940, we are able to make five forecasts up to the year 1945.
+Models are not re-initialized with observations between years.
+Thus, in our current example, the model forecast for percent cover in 1941 has a forecast horizon of one year, the forecast in 1942 has a forecast horizon of two years, and so on.
+We ran these simulations for all model types (IPM with/without climate; QBM with/without climate) using mean parameter values for all possible initial years.
+Then, for a given forecast horizon, we averaged the correlation between forecasts and observations.
+Note that these forecasts are all made using in-sample data since we used model fits from the full data set.
+Nonetheless, these simulations offer insight into the differences between model forecast horizons.
+
+<!--
 ### Testing sensitivity to climate covariates
@@
 A weak correlation indicates the QBM "misses" the climate effect on a particular vital rate.
+-->
 
@@
 -------
+Both the IPM and QBM generated skillful one-step-ahead forecasts for out-of-sample observations, with an average correlation between predictions and observations ($\rho$) of 0.# across all models (Fig.
+2).
+In only one case were IPM forecasts significantly more accurate than the QBM (Fig.
+2): forecast accuracy of *P.
+secunda* percent cover from an IPM with climate covariates was greater than the accuracy from an equivalent QBM ($t_{(195)}$ = 1.72, *P* = 0.043).
+The posterior distributions of climate effects included in our models broadly overlapped zero (Fig.
+3), but including climate covariates did increase the accuracy of forecasts for all species except *P.
+smithii* (Fig.
+2).
+However, increases in accuracy due to the inclusion of climate covariates were not significant (*P* > 0.05 for all comparisons of $\rho$ between climate and no-climate forecasts within model types; Fig.
+2).
+The accuracy of both model's forecasts declined as the forecast horizon increased, but they did so at similar rates (Fig. 4).
+The only exception is for *P. secunda* where forecast accuracy appears to slightly increase with forecast horizon, after an initial decrease from a forecast horizon of one year (Fig. 4).
+
+<!--
 ### Comparison of forecast models
@@
 In other words, neither model could predict whether a climate perturbation would increase or decrease equilibrium population size.
+-->
 
@@
 
+<!---
 \begin{table}[ht]
@@
 \pagebreak{}
+--->
 
@@
 
+<!---
 \begin{figure}[!ht]
@@
 The optimal score for prediction for each species-vital rate combination is shown with a vertical red line.}
+\end{figure}
+
+\newpage{}
+--->
+\begin{figure}[!ht]
+  \centering
+      \includegraphics[height=6in]{./components/forecast_accuracy_mockup.png}
+  \caption{Comparison of one-step-ahead, out-of-sample forecast accuracy between the IPM and QBM models with and without the inclusion of climate covariates.
+Comparisons between equivalent IPM and QBM models indicate no significant difference in accuracy (\emph{P} > 0.05 for all comparisons).
+Likewise, including climate covariates did not result in siginificantly higher forecast accuracy (\emph{P} > 0.05 for all comparisons).}
 \end{figure}
@@
       \includegraphics[height=5in]{./components/forecast_horizon.png}
-  \caption{The forecast horizons for both models when used to forecast the final observation (cover in year 1945) starting from all possible prior years.
-For example, simulations initialized with cover in 1932 have 13 years before the forecast.}
+  \caption{The forecast horizons for both models. Points show the average accuracy ($\rho$) across all forecasts at a given time horizon.}
 \end{figure}

### 2016-05-13 FigureScripts/one_step_results_table.R

@@ -337,9 +337,26 @@ qbm_quadyear_weather <- ddply(qbm_onestep, .(quad, year, species), summarise,
 qbm_quadyear_noweather <- ddply(qbm_onestep_dd, .(quad, year, species), summarise,
                                 median_pred = median(pred_cover.t1),
                                 obs = mean(obs_cover.t1))
-test1 <- subset(qbm_quadyear_noweather, species=="BOGR")
-test2 <- subset(qbm_quadyear_weather, species=="BOGR")
+test1 <- subset(ipm_quadyear_noweather, species=="POSE")
+test2 <- subset(ipm_quadyear_weather, species=="POSE")
 rho_comp(x1=test2$median_pred, x2=test1$median_pred, y=test1$obs)
+
+ipm4merge_noweather <- ipm_quadyear_noweather
+ipm4merge_noweather$t1 <- ipm4merge_noweather$t1+1900-1
+qbmVipm_df_noweather <- merge(ipm4merge_noweather, qbm_quadyear_noweather, 
+                              by.x=c("quad", "t1", "species"),
+                              by.y=c("quad", "year", "species"))
+test1 <- subset(qbmVipm_df_noweather, species=="POSE")
+rho_comp(x1=test1$median_pred.x, x2=test1$median_pred.y, y=test1$obs.x)
+
+ipm4merge_weather <- ipm_quadyear_weather
+ipm4merge_weather$t1 <- ipm4merge_weather$t1+1900-1
+qbmVipm_df_weather <- merge(ipm4merge_weather, qbm_quadyear_weather, 
+                              by.x=c("quad", "t1", "species"),
+                              by.y=c("quad", "year", "species"))
+test1 <- subset(qbmVipm_df_weather, species=="POSE")
+rho_comp(x1=test1$median_pred.x, x2=test1$median_pred.y, y=test1$obs.x)
+
 
 
 mae_list <- list()

### 2016-05-14 manuscript/manuscript.Rmd
adds new discussion beginning

@@
 2).
+Without climate covariates, the accuracy of forecasts from the IPM were not statistically greater than the accuracy of forecasts from the QBM.
+With climate covariates, the best out-of-sample predictive model (highest *lppd*) for each species and vital rate typically resulted from highly constrained priors on the climate effects (Fig. S1).
+Thus, the posterior distributions of climate effects included in our models overlapped zero and generally shrunk toward zero, though for some species-vital rate combinations important effects (80% credible interval does not include zero) did emerge (Fig.
+3).
+Despite the small effects, including climate covariates did increase the accuracy of forecasts for all species except *P. smithii* (Fig. 2).
+However, increases in accuracy due to the inclusion of climate covariates were not significant (*P* > 0.05 for all comparisons of $\rho$ between climate and no-climate forecasts within model types; Fig.
+2).
 In only one case were IPM forecasts significantly more accurate than the QBM (Fig.
@@
 secunda* percent cover from an IPM with climate covariates was greater than the accuracy from an equivalent QBM ($t_{(195)}$ = 1.72, *P* = 0.043).
-The posterior distributions of climate effects included in our models broadly overlapped zero (Fig.
-3), but including climate covariates did increase the accuracy of forecasts for all species except *P.
-smithii* (Fig.
-2).
-However, increases in accuracy due to the inclusion of climate covariates were not significant (*P* > 0.05 for all comparisons of $\rho$ between climate and no-climate forecasts within model types; Fig.
-2).
 The accuracy of both model's forecasts declined as the forecast horizon increased, but they did so at similar rates (Fig. 4).
@@
 However, density-structured models rely on the aggregation of individual-level data. 
-This creates a potential problem if such models are to be used in a climate change context because it is individuals, not populations, which respond to climate [@Clark2012].
-Are models based on population-level metrics as sensitive to climate as models based on individual-level metrics?
-Do these two types of models produce consistent forecasts?
-Do we need detailed demographic data to forecast the impacts of climate change?
+This creates a potential problem if (1) such models inaccurately estimate density-dependence or (2) are to be used in a climate change context because it is individuals, not populations, that respond to climate [@Clark2012].
+Can models based on population-level metrics generate forecasts that are as skillful as those generated from models based on individual-level data?
+Are models based on population-level metrics as sensitive to climate as models based on individual-level data?
+Do we need detailed demographic data to forecast the state of plant populations?
 
+Our comparison of a traditional, demographic population model without environmental forcing (the IPM) to an equivalent model inspired by density-structured models (the QBM) showed that, generally, IPM forecasts of out-of-sample plant population states were no more accurate than forecasts from the QBM (Fig.
+2; `no-climate' bars).
+We expected the IPM to out-perform the QBM since the IPM includes more mechanistic detail on the perennial plant life cycle, but this was not the case.
+Our comparisons of forecast accuracy relied on one-step-ahead forecasts, but we also found no evidence that the rate at which forecast accuracy decays as forecast horizon increases was different between models (Fig.
+4).
+Thus, it appears that, without environmental forcing, the IPM and QBM generate equivalent forecasts.
+Such a finding confirms theoretical [@Freckleton2011] and empirical work [@Queenborough2011] showing that density-structured models can be useful surrogates for demographic models when the goal is to estimate or forecast population states over large spatial extents.
+Likewise, the equivalency of forecast accuracy between the IPM and QBM held for all four species, when climate covariates are not included, extending the generality of previous findings [e.g., @Taylor2004; @Queenborough2011].
+
+While the models did not differ in forecast accuracy when only density-dependence was allowed to determine population dynamics, we expected that the inclusion of environmental forcing would elucidate the differences between the models.
+Our expectation was the the IPM would outperform the QBM when we included climate covariates because interannual variation in weather can affect vital rates in unique ways [@Dalgleish2011].
+Thus, estimates of climate effects on plant population growth may be biased or non-identifiable when the underlying statistical model is fit using population-level data that integrates over the potentially unique climate responses of individual vital rates.
+
+<!---
 ### The importance of demographic data
-
 Our comparison of a traditional, demographic population model (the IPM) with a model inspired by density-structured models (the QBM) showed that the IPM outperformed the QBM: the IPM was more accurate and precise than the QBM in out-of-sample cross validation (Table 1).
@@
 Uncertainty in demographic responses to climate can be reduced by collecting (1) longer time series and (2) climate covariates that match the scale of inference (e.g., plot rather than landscape level climate/weather metrics).
+--->
 

### 2016-05-15 manuscript/manuscript.Rmd
add second section of discussion draft

@@
 Thus, estimates of climate effects on plant population growth may be biased or non-identifiable when the underlying statistical model is fit using population-level data that integrates over the potentially unique climate responses of individual vital rates.
+However, IPM forecast accuracy was only significantly higher than the QBM for one species, *P. secunda* (Fig. 2, `climate' bars).
+This result likely stems from the relatively weak climate effects in vital rate regressions (Fig. 3).
+If climate effects on vital rates are weak, then models based on percent cover or density should fair well compared to models based on individual-level data so long as density-dependence is adequately estimated (see Fig.
+2 and Freckleton et al.
+2011).
+
+Our naive expectation was that a single, relatively important climate effect in a vital rate regression would cause a discernable difference between IPM and QBM forecast skill because individual-level responses may be less identifiable from percent cover data [@Clark2012].
+We actually found the opposite.
+Survival or recruitment regressions for *B.
+gracilis*, *H.
+comata*, and *P.
+smithii* all included at least one relatively important climate effect, which we *a posterior* defined as a standardized coefficient whose 80% credible interval does not overlap zero (Fig.
+3).
+Yet, only the forecast skill for *P. secunda* differed between the IPM and QBM (Fig. 2).
+Our explanation is that while at least one vital regression for the other three species included a relatively strong climate effect, only *P.
+secunda* had a vital rate regression with several smaller climate effects all trending in the same direction (Fig.
+3).
+Thus, it appears that the estimation of several small effects, rather than a single relatively large effect, can lead to increased forecast skill.
+
+The higher accuracy of the IPM with cliamte for *P.
+secunda*  highlights the superiority of contemporary model and variable selection approaches such as ridge regression and LASSO over techniques that lead to excluding "non-significant" effects from final models.
+Especially in a Bayesian forecasting framework, ridge regression allows researchers to include forcing variables that we know are changing (e.g., temperature and precipitation) in final models even if the estimated effects are relatively small or even deemed unimportant due to noisy data, out of which it is more difficult to estimate effects on interannual variation.
+Likewise, if a species is truly unresponsive to a given climate variable, statistical regularization techniques will shrink the mean and variance of a covariate estimate to zero [@Hooten2015].
+No amount of statistical "machismo"" during the model fitting stage can make up for including the appropriate candidate covariates, which we attempted to do based on our knowledge of this semi-arid plant community.
+However, the climate covariates we chose required aggregating daily weather data over specific time periods.
+It may be that we chose the wrong time periods over which to aggregate.
+New methods using functional linear models (or splines) may offer a data-driven approach for identifying the appropriate time periods over which to aggregate to produce a tractable set of candidate climate variables [@Teller2016].
+
+In conclusion, we found that demographic data was generally unneccesary to generate skillful forecasts of plant population states.
+Such a finding is contrary to our expectations, but in agreement with recent theoretical [@Freckleton2011] and emprical work [@Queenborough2011].
+However, when we included climate covariates, we did achieve more skillful forecasts from the IPM for one species, *P.
+secunda*, that appears to respond consistently, but weakly, to several climate variables.
+Thus, we conclude that models based on population-level data, rather than individual-level data, may be adequate for forecasting the state of plant populations for species that do not respond consistently or strongly to climate.
+Unfortunately, our analysis, where climate effects were relatively unimportant in vital rate regressions, did not allow us to sufficiently test our prediction that individual-level data is neccessary to generate skillful forecasts if different vital rates respond to climate in unique, potenially opposing, ways.
+Nonetheless, our results are encouraging for the use of easy-to-collect percent cover for forecasting the state of plant populations.
 

### 2016-05-15 manuscript/manuscript.Rmd
adds dataset description table; adds discussion paragraph on horizons

@@
 
+We also expected IPM forecast accuracy to decline at a lower rate than the QBM as the forecast horizon increased.
+Indeed, many have argued that more mechanistic models produce better predictions, especially under novel conditions [@Evans2012b; @Schindler2015b].
+Likewise, the IPM explicitly models the processes of recruitment and survival that may be poorly represented in the QBM since recruitment and survival mainly affect small plants that contribute little to overall changes in percent cover.
+But the addition and subtraction of small plants can have large effects on population growth over time and potentially lead to a longer forecast horizon for models that better represent the plant life cycle.
+We found no evidence for a difference between the IPM and QBM forecast horizons (Fig. 4).
+Similar forecast horizons should be expected if both models adequately capture density-dependence and environmental forcing is negligble, as is the case for our models (Fig. 2).
+In fact, only the forecast horizons for *P. secunda* noticeably differ (Fig. 4), and that is also the only species where we found the IPM to outperform the QBM when we included climate effects.
+
 In conclusion, we found that demographic data was generally unneccesary to generate skillful forecasts of plant population states.
@@
 ------
+\begin{table}[ht]
+\centering
+\caption{Description of data.} 
+\begin{tabular}{llrr}
+  \hline
+Species & Vital Rate Model & Num. Obs. & Num. Quadrats\\ 
+  \hline
+  \emph{B. gracilis} & Growth & 5670 & 29 \\ 
+       & Survival & 10102 & 33 \\ 
+       & Recruitment & 304 & 33 \\ 
+       & Percent cover & 281 & 29 \\ 
+  \rule{0pt}{3ex} \emph{H. comata} & Growth & 1990 & 16 \\ 
+       & Survival & 3257 & 18 \\ 
+       & Recruitment & 304 & 18 \\ 
+       & Percent cover & 171 & 17 \\ 
+  \rule{0pt}{3ex} \emph{P. smithii} & Growth & 8052 & 19 \\ 
+       & Survival & 11344 & 19 \\ 
+       & Recruitment & 304 & 19 \\ 
+       & Percent cover & 217 & 19 \\ 
+  \rule{0pt}{3ex} \emph{P. secunda} & Growth & 3018 & 18 \\ 
+       & Survival & 4650 & 18 \\ 
+       & Recruitment & 304 & 18 \\ 
+       & Percent cover & 197 & 18 \\ 
+   \hline
+\end{tabular}
+\end{table}
 
@@
 
+\newpage{}
+
 Figures

### 2016-05-15 manuscript/manuscript.Rmd
minor edits to discussion

@@
 
-We modeled each grass population based on two levels of data: individual and quadrat (Fig.
-2).
+We modeled each grass population based on two levels of data: individual and quadrat.
 The individual data is the "raw" data.
@@
 This is equivalent to a near-perfect census of quadrat percent cover because previous analysis shows that measurement error at the individual-level is small [@Chu2014].
-Based on these two datasets we can compare population models built using individual-level data and aggregated, quadrat-level data.
-At the individual level we explicitly model three vital rates: growth (13 year-to-year transitions, 29 quadrats and 18,730 records across the four species), survival (13 years, 33 quadrats and 29,353 records across the four species), and recruitment (13 years, 33 quadrats and 304 records across the four species).
-At the quadrat level we model population growth as change in percent cover of quadrats with non-zero cover in year *t* and in year *t-1* (13 year-to-year transitions, 29 quadrats and 866 records across the four species).
+Based on these two datasets of 13 year-to-year transitions, we can compare population models built using individual-level data and aggregated, quadrat-level data.
+At the individual level we explicitly model three vital rates: growth, survival, and recruitment.
+At the quadrat level we model population growth as change in percent cover of quadrats with non-zero cover in year *t* and in year *t-1*.
 For modeling population growth at the quadrat level we ignore within-quadrat extirpation and colonization events because they are very rare in our time series ($N=16$ and $N=13$, respectively, across all species).
+Sample sizes for each species and vital rate model are shown in Table 1.
 Given the relatively broad spatial distribution of the quadrats we are studying, it is safe to assume that these events are in fact rare enough to be ignored for our purposes.
@@
 We expected the IPM to out-perform the QBM since the IPM includes more mechanistic detail on the perennial plant life cycle, but this was not the case.
-Our comparisons of forecast accuracy relied on one-step-ahead forecasts, but we also found no evidence that the rate at which forecast accuracy decays as forecast horizon increases was different between models (Fig.
-4).
 Thus, it appears that, without environmental forcing, the IPM and QBM generate equivalent forecasts.
@@
 
-The higher accuracy of the IPM with cliamte for *P.
+The higher accuracy of the IPM with climate for *P.
 secunda*  highlights the superiority of contemporary model and variable selection approaches such as ridge regression and LASSO over techniques that lead to excluding "non-significant" effects from final models.
@@
 Likewise, if a species is truly unresponsive to a given climate variable, statistical regularization techniques will shrink the mean and variance of a covariate estimate to zero [@Hooten2015].
-No amount of statistical "machismo"" during the model fitting stage can make up for including the appropriate candidate covariates, which we attempted to do based on our knowledge of this semi-arid plant community.
+No amount of statistical "machismo" during the model fitting stage can make up for including the appropriate candidate covariates, which we attempted to do based on our knowledge of this semi-arid plant community.
 However, the climate covariates we chose required aggregating daily weather data over specific time periods.

### 2016-05-19 manuscript/manuscript.Rmd
updates ms with error results

@@
 Addressing climate change questions with models fit to population-level data is potentially problematic.
-It is individuals, not populations, that respond to climate variables [@Clark2012].
+It is individuals, not populations, that respond to climate variables [@Clark2011b].
 Ignoring this fact amounts to an "ecological fallacy", where inference about the individual relies on statistical inference on the group [@PIANTADOSI1988].
@@
 These unique climate responses may be difficult to resolve in statistical models based on population-level data where demographic processes are not identifiable.
+Futhermore, models based on aggregated data may reflect short-term effects of one vital rate (e.g., plant growth) more than others, but over the long-term the influence of other vital rates may be important.
 If important climate effects are missed because of the aggregation inherent in in population-level data, then population models built with such data will make uninformative or unreliable forecasts.
@@
 We used a demographic dataset that tracks the fates of individual plants from four species over 14 years to build two kinds of single-species population models, traditional models using individual growth, survival, and recruitment data and alternative models based on basal cover.
-We use the models to answer two questions: (1) Can population models fit using aggregated individual-level data (percent cover) produce forecasts as skillful as those from models fit to demographic data?
+We use the models to answer two questions: (1) Can population models fit using aggregated individual-level data (percent cover) adequately capture density dependence to produce forecasts as skillful as those from models fit to demographic data?
 And (2) Can population models fit using aggregated data adequately capture the influence of climate on population growth and, in turn, produce forecasts as skillful as those from models fit to demographic data?
+Both questions are motivated by the fact that the effects of intraspecific competition (density dependence) and interannual weather variability act at the level of the individual [@Clark2011b], meaning that models based on individual-level processes should generate more skillful forecasts.
 
@@
 -------
-Both the IPM and QBM generated skillful one-step-ahead forecasts for out-of-sample observations, with an average correlation between predictions and observations ($\rho$) of 0.# across all models (Fig.
+Both the IPM and QBM generated skillful one-step-ahead forecasts for out-of-sample observations, with an average correlation between predictions and observations ($\rho$) of 0.73 across all models and species (Fig.
 2).
-Without climate covariates, the accuracy of forecasts from the IPM were not statistically greater than the accuracy of forecasts from the QBM.
-With climate covariates, the best out-of-sample predictive model (highest *lppd*) for each species and vital rate typically resulted from highly constrained priors on the climate effects (Fig. S1).
+Without climate covariates, the accuracy of forecasts from the IPM were not statistically greater than the accuracy of forecasts from the QBM (Fig.
+2) and overall error was similar (mean absolute error; Fig.
+S1).
+With climate covariates, the best out-of-sample predictive model (highest *lppd*) for each species and vital rate typically resulted from highly constrained priors on the climate effects (Fig. S2).
 Thus, the posterior distributions of climate effects included in our models overlapped zero and generally shrunk toward zero, though for some species-vital rate combinations important effects (80% credible interval does not include zero) did emerge (Fig.
@@
 Despite the small effects, including climate covariates did increase the accuracy of forecasts for all species except *P. smithii* (Fig. 2).
-However, increases in accuracy due to the inclusion of climate covariates were not significant (*P* > 0.05 for all comparisons of $\rho$ between climate and no-climate forecasts within model types; Fig.
-2).
+Increases in accuracy due to the inclusion of climate covariates were not significant (*P* > 0.05 for all comparisons of $\rho$ between climate and no-climate forecasts within model types; Fig. 2).
+However, the IPM with climate covariates for *P. secunda* did have significantly lower error than the IPM without climate covariates ($t_{(196)}$ = -1.84, *P* = 0.033; Fig. S2).
 In only one case were IPM forecasts significantly more accurate than the QBM (Fig.
@@
 secunda* percent cover from an IPM with climate covariates was greater than the accuracy from an equivalent QBM ($t_{(195)}$ = 1.72, *P* = 0.043).
+Forecasts from IPMs with climate covariates had significantly lower error than equivalent QBM forecasts for *P.
+smithii* ($t_{(216)}$ = -3.49, *P* < 0.001) and *P.
+secunda* ($t_{(196)}$ = -1.83, *P* = 0.034) (Fig.
+S2).
+Results from all pairwise statistical tests are shown in Table S1.
 The accuracy of both model's forecasts declined as the forecast horizon increased, but they did so at similar rates (Fig. 4).
@@
 However, density-structured models rely on the aggregation of individual-level data. 
-This creates a potential problem if (1) such models inaccurately estimate density-dependence or (2) are to be used in a climate change context because it is individuals, not populations, that respond to climate [@Clark2012].
+This creates a potential problem if (1) such models inaccurately estimate density dependence or (2) are to be used in a climate change context because it is individuals, not populations, that respond to intraspecific competition and weather [@Clark2011b].
 Can models based on population-level metrics generate forecasts that are as skillful as those generated from models based on individual-level data?
@@
 We found no evidence for a difference between the IPM and QBM forecast horizons (Fig. 4).
-Similar forecast horizons should be expected if both models adequately capture density-dependence and environmental forcing is negligble, as is the case for our models (Fig. 2).
+Similar forecast horizons should be expected if both models adequately capture density dependence and environmental forcing is negligble, as is the case for our models (Fig. 2).
 In fact, only the forecast horizons for *P. secunda* noticeably differ (Fig. 4), and that is also the only species where we found the IPM to outperform the QBM when we included climate effects.

### 2016-05-23 manuscript/manuscript.Rmd
adds PBA edits to ms

@@
 **1.** Rapid environmental change has generated growing interest in forecasts of future population trajectories.
-Traditional population models, typically built using detailed demographic observations from one study site, can address the impacts of environmental change at one location, but are difficult to scale up to the landscape and regional scales relevant to management decisions.
-
-**2.** An alternative is to build models using population-level data that are much easier to collect over broad spatial scales than individual-level data.
-However, it remains unclear if models built using aggregated individual-level data adequately capture the effects of density-dependence and environmental forcing that are necessary to generate skillful forecasts.
+Traditional population models built with detailed demographic observations from one study site can address the impacts of environmental change at particular locations, but are difficult to scale up to the landscape and regional scales relevant to management decisions.
+
+**2.** An alternative is to build models using population-level data which are much easier to collect than individual-level data over broad spatial scales.
+However, it is unknown whether models built using population-level data adequately capture the effects of density-dependence and environmental forcing that are necessary to generate skillful forecasts.
 
@@
 We parameterized two population models for each species, one based on individual-level data (survival, growth and recruitment) and one on population-level data (percent cover), and compared their forecasting skill and forecast horizons with and without the inclusion of climate covariates.
-For both models we used Bayesian ridge regression to identify the optimal predictive model in terms of climate covariate strengths.
-
-**4.** Without climate effects included, we found no significant difference between the forecasting skill of models based on individual-level data and models based on population-level data.
-Climate effects were weak and caused only slight increases in forecasting skill.
-Increases in skill were similar between model types except for one species where forecast accuracy from the individual-level model was significantly higher than the accuracy from an equivalent population-level model.
-
-**5.** *Synthesis.* For our focal species at this particular location, and using our particular statistical models, demographic data was generally unnecessary to achieve skillful forecasts, though for certain species forecast skill can be gained by using demographic data linked to climate covariates.
+For both models we used Bayesian ridge regression to weight the influence of climate covariates for optimal prediction.
+
+**4.** In the absence of climate effects, we found no significant difference between the forecasting skill of models based on individual-level data and models based on population-level data.
+Climate effects were weak and only marginally increased forecasting skill.
+Increases in skill with climate covariates were similar between model types for three of the four species.
+For the fourth species, forecast accuracy of the individual-level model with climate covariates was significantly higher than the accuracy of the equivalent population-level model.
+
+**5.** *Synthesis.* For our focal species at this particular location, and using our particular statistical models, demographic data was generally unnecessary to achieve skillful [ARE ANY OF THESE FORECASTS SKILLFUL?
+FOCUS INSTEAD ON MODEL DIFFERENCES] forecasts, though for certain species forecast skill can be gained by using demographic data linked to climate covariates.
 We conclude that models based on aggregated individual-level data offer a practical alternative to data-intensive demographic models when species do not respond strongly to interannual variation in weather, but when modeling species that do respond to climate drivers, demographically-based models can generate more skillful forecasts.
@@
 Perhaps the greatest challenge for ecology in the 21st century is to forecast the impacts of environmental change [@Clark2001; @Petchey2015].
-Forecasts require sophisticated modeling approaches that fully account for uncertainty and variability in both ecological process and model parameters [@Luo2011a; but see @Perretti2013a for an argument against modeling the ecological process].
+Forecasts require sophisticated modeling approaches that fully account for uncertainty and variability in both ecological process and model parameters [@Luo2011a; but see @Perretti2013a].
 The increasing statistical sophistication of population models [@Rees2009] makes them promising tools for predicting the impacts of environmental change on species persistence and abundance.
 But reconciling the scales at which population models are parameterized and the scales at which environmental changes play out remains a challenge [@Clark2010; @Freckleton2011; @Queenborough2011; @Clark2012].
-The problem is that most population models are built using data from a single study site because collecting those data, which involves tracking the fates of individuals plants, is so difficult.
+Most population models are built using demographic data from a single study site because tracking the fates of individuals is so difficult.
 The resulting models cannot be applied to the landscape and regional scales relevant to decision-making without information about how the fitted parameters respond to spatial variation in biotic and abiotic drivers [@Sæther2007].
@@
 The inability of most population models to address landscape-scale problems may explain why land managers and conservation planners have embraced species distribution models (SDMs) [see @Guisan2005 for a review].
-SDMs typically rely on easy-to-collect presence/absence data [but see @Clark2014 for new methods] and remotely-sensed environmental covariates that allow researchers to model large spatial extents [e.g., @Maiorano2013].
-Thus, it is relatively straightforward to parameterize and project SDMs over landscapes and regions.
+SDMs typically rely on easy-to-collect presence/absence data and remotely-sensed environmental covariates that allow researchers to model large spatial extents [e.g., @Maiorano2013]  and project future changes in occupancy over landscapes and regions.
 However, the limitations of SDMs are well known [@Pearson2003; @Elith2009; @Araujo2012].
@@
 Population-level data cannot provide inference about demographic mechanisms, but might be sufficient for modeling future population states, especially since such data are feasible to collect across broad spatial extents [e.g., @Queenborough2011].
-The choice between individual and population-level data involves a difficult trade-off: while individual-level data leads to more mechanistic models, population-level data can lead to models that can be applied over greater spatial and temporal extents because the data are easier to collect over large spatial scales.
+The choice between individual and population-level data involves a difficult trade-off: while individual-level data are necessary for mechanistic models, population-level data enable models that can be applied over greater spatial and temporal extents.
 An open question is how much forecasting skill is lost when we build models based on population rather than individual-level data.
@@
 Building on @Taylor2004, @Freckleton2011 showed that density-structured models compare well to continuous models in theory, and @Queenborough2011 provide empirical evidence that density-structured models are capable of reproducing population dynamics at landscape spatial scales, even if some precision is lost when compared to fully continuous models.
-However, previous tests of density-structured models have yet to assess their ability to forecast out-of-sample observations, and they have not included environmental covariates, which is necessary to make forecasts of population responses to climate change.
+However, previous tests of density-structured models have yet to assess their ability to forecast out-of-sample observations, and they have not included environmental covariates, which are necessary to forecast population responses to climate change.
 
 Addressing climate change questions with models fit to population-level data is potentially problematic.
-It is individuals, not populations, that respond to climate variables [@Clark2011b].
-Ignoring this fact amounts to an "ecological fallacy", where inference about the individual relies on statistical inference on the group [@PIANTADOSI1988].
 Population growth (or decline) is the outcome of demographic processes such as survival, growth, and recruitment that occur at the level of individual plants.
-Climate can affect each demographic process in unique, potentially opossing, ways [@Dalgleish2011].
+Climate can affect each demographic process in unique, potentially opposing, ways [@Dalgleish2011].
 These unique climate responses may be difficult to resolve in statistical models based on population-level data where demographic processes are not identifiable.
-Futhermore, models based on aggregated data may reflect short-term effects of one vital rate (e.g., plant growth) more than others, but over the long-term the influence of other vital rates may be important.
+Futhermore, models based on aggregated data may reflect short-term effects of one vital rate more than others whose importance may only emerge over the long-term.
+For example, a one-year change in a plant species' cover or biomass might reflect individual growth or shrinkage, whereas the long-term trajectory of the population might be more influenced by recruitment.
 If important climate effects are missed because of the aggregation inherent in in population-level data, then population models built with such data will make uninformative or unreliable forecasts.
 
-Here, we compare the forecasting skill of statistical and population models based on aggregated, population-level data with models based on individual-level data.
+Here, we compare the forecasting skill of statistical and population models based on aggregated, population-level data with the skill of models based on individual-level data.
 We used a demographic dataset that tracks the fates of individual plants from four species over 14 years to build two kinds of single-species population models, traditional models using individual growth, survival, and recruitment data and alternative models based on basal cover.
-We use the models to answer two questions: (1) Can population models fit using aggregated individual-level data (percent cover) adequately capture density dependence to produce forecasts as skillful as those from models fit to demographic data?
-And (2) Can population models fit using aggregated data adequately capture the influence of climate on population growth and, in turn, produce forecasts as skillful as those from models fit to demographic data?
-Both questions are motivated by the fact that the effects of intraspecific competition (density dependence) and interannual weather variability act at the level of the individual [@Clark2011b], meaning that models based on individual-level processes should generate more skillful forecasts.
+We simulated the models to answer two questions: (1) Can population models fit using aggregated individual-level data (percent cover) adequately capture density dependence to produce forecasts as skillful as those from models fit to demographic data?
+(2) Can population models fit using aggregated data adequately capture the influence of climate on population growth and, in turn, produce forecasts as skillful as those from models fit to demographic data?
+[THIS NEXT SENTENCE SEEMS OUT OF PLACE.
+CAN YOU MODIFY PREVIOUS PARAGRAPH TO ADDRESS DENSITY DEPENDENCE ALONG WITH CLIMATE EFFECTS?
+E.G.
+DENSITY DEPENDENCE MIGHT ACT MOST STRONGLY ON RECRUITMENT WHICH IS HARD TO DETECT IN A COVER BASED MODEL?] Both questions are motivated by the fact that the effects of intraspecific competition (density dependence) and interannual weather variability act at the level of the individual [@Clark2011b], meaning that models based on individual-level processes should generate more skillful forecasts.
 
@@
 ### Study site and data
-Our demographic data come from the Fort Keogh Livestock and Range Research Laboratory in eastern Montana's northern mixed prairie near Miles City, Montana, USA ($46^{\circ}$ 19' N, $105^{\circ}$ 48' W).
-The dataset is freely available on Ecological Archives[^link] [@Anderson2011], and interested readers should refer to the metadata for a complete description.
-The site is about 800 m above sea level and mean annual precipitation (1878-2009) is 334 mm, with most annual precipitation falling from April through September.
+Our demographic data come from a northern mixed grass prairie at the Fort Keogh Livestock and Range Research Laboratory near Miles City, Montana, USA ($46^{\circ}$ 19' N, $105^{\circ}$ 48' W).
+The dataset is available on Ecological Archives[^link] [@Anderson2011], and interested readers should refer to the metadata for a complete description.
+The site is 800 m above sea level and mean annual precipitation (1878-2009) is 334 mm, with most annual precipitation falling from April through September.
 The community is grass-dominated, and we focused on the four most abundant grass species: \emph{Bouteloua gracilis} (BOGR), \emph{Hesperostipa comata} (HECO), \emph{Pascopyrum smithii} (PASM), and \emph{Poa secunda} (POSE) (Fig.
@@
 All species typically begin growth in the early spring, reach maximum growth and flower in early to mid summer (May-June), and disperse seed in mid to late summer (July-September).
+[LAST SENTENCES GLOSSES OVER LIKELY BIG DIFFERENCES IN PHENOLOGY.
+JUST DESCRIBE TIMING OF GROWING SEASON INSTEAD, TO SET UP CLIMATE COVARIATES?]
 
@@
 The individual data is the "raw" data.
-For the quadrat-level we data we simply sum individual basal cover for each quadrat by species.
-This is equivalent to a near-perfect census of quadrat percent cover because previous analysis shows that measurement error at the individual-level is small [@Chu2014].
+For the quadrat-level, we data we simply sum individual basal cover for each quadrat by species.
+This is equivalent to a near-perfect census of quadrat percent cover because measurement error at the individual-level is small [@Chu2014].
 Based on these two datasets of 13 year-to-year transitions, we can compare population models built using individual-level data and aggregated, quadrat-level data.
 At the individual level we explicitly model three vital rates: growth, survival, and recruitment.
-At the quadrat level we model population growth as change in percent cover of quadrats with non-zero cover in year *t* and in year *t-1*.
-For modeling population growth at the quadrat level we ignore within-quadrat extirpation and colonization events because they are very rare in our time series ($N=16$ and $N=13$, respectively, across all species).
+At the quadrat level we model population growth as change in percent cover of quadrats with non-zero cover in year *t* and in year *t-1*, ignoring within-quadrat extirpation and colonization events because they are very rare in our time series ($N=16$ and $N=13$, respectively, across all species).
 Sample sizes for each species and vital rate model are shown in Table 1.
-Given the relatively broad spatial distribution of the quadrats we are studying, it is safe to assume that these events are in fact rare enough to be ignored for our purposes.
 
@@
 At the quadrat-level, we fit a single regression model for population growth.
-We describe the statistical models separately since fitting the models required different approaches.
+We describe the statistical models separately since they required different approaches.
 For both model types, we fit vital rate models with and without climate covariates.
@@
 All model parameters are species-specific, but we omit subscripts for species in model descriptions below to reduce visual clutter.
-For brevity, we only describe models with climate covariates included, but models without climate covariates are exactly the same as the models described below, just without the climate effects.
+For brevity, we only describe models with climate covariates included, but models without climate covariates are simply the models described below with the climate effects removed.
 
@@
 
-where $x$ is the log of genet basal area, $\beta_{0,t}$ is a year specific intercept intercept, $\beta_Q$ is the random effect of quadrat group location, $\beta_{s,t}$ is the year-specific slope parameter for size, **z** is a vector of *i* climate covariates specific to year *t*, $\boldsymbol{\beta}_c$ is a vector of fixed climate effects of length *i*, $\beta_{d,1}$ is the effect of intraspecific crowding experienced by the focal genet (*w*), and $\beta_{d,2}$ is a size by crowding ($xw$) interaction effect.
+where $x$ is the log of genet basal area, $\beta_{0,t}$ is a year specific intercept intercept, $\beta_Q$ is the random effect of quadrat group location, $\beta_{s,t}$ is the year-specific slope parameter for size, **z** is a vector of *i* climate covariates specific to year *t*, $\boldsymbol{\beta}_c$ is a vector of fixed climate effects of length *i*, $\beta_{d,1}$ is the effect of intraspecific crowding experienced by the focal genet ($w$), and $\beta_{d,2}$ is a size by crowding ($xw$) interaction effect.
+
+YOU NEED TO EXPLAIN HOW THE w's ARE CALCULATED. COPY TEXT FROM ADLER ET AL. 2010 or CHU  & ADLER 2015.
 
@@
 
-### Model fitting and stastical regularization
+### Model fitting and statistical regularization
 ####  Model fitting
@@
 ### Model validation
-To test each model's ability to forecast population states we made out-of-sample predictions using leave-one-year-out cross validation.
+To test each model's ability to forecast population states, we made out-of-sample predictions using leave-one-year-out cross validation.
 For both levels of modeling and for models with and without climate covariates, we fit the vital rate models using observations from all years except one, and then used those fitted parameters in the population models to perform a one-step-ahead forecast for the year whose observations were withheld from model fitting.
 Within each observation year, several quadrats were sampled.
-We made predictions for each observed quadrat in the focal year, initializing each simulation with cover in the quadrat the previous year.
+We made predictions for each observed quadrat in each focal year, initializing each simulation with cover in the quadrat the previous year.
 Since we were making quadrat-specific predictions, we incorporated the group random effect on the intercept for both models.
@@
 An important feature of any forecasting model is the rate at which forecast skill declines as the time between an observation and a forecast increases; the so-called ecological forecast horizon [@Petchey2015].
-To assess the forecast horizons of our models, we iniate the model with the population state at some time *t* and make sequential forecasts of the population at times $t+1, t+2, \dots, t+T$ where *T* is the maximum number of years between the initial year and the final year in our observations.
+To assess the forecast horizons of our models, we iniate the model with the population state at some time *t* and make sequential forecasts of the population at times $t+1, t+2, \dots, t+T$ where *T* is the maximum number of years between the initial year and the final year of our observations.
 For example, if we initialize the model with percent cover in 1940, we are able to make five forecasts up to the year 1945.
@@
 We ran the simulations for 2,500 time steps, enough to estimate equilibrium cover after discarding an initial 500 time steps as burn-in.
+EXPLAIN HOW CLIMATE COVARIATES ARE DRAWN AT EACH TIME STEP.
+AND NO RANDOM YEAR EFFECTS?
 Each simulation was run under two parameter scenarios: (1) using mean parameter estimates and (2) using randomly drawn parameters from the MCMC chain.
@@
 We expected the IPM to produce more accurate and precise forecasts due to either (1) the smaller sample size of the quadrat level data sets compared to the individual level data sets, leading to larger parameter uncertainty for the QBM, or (2) the QBM climate effects being weakly associated with one or more vital rate climate effects at the individual level.
-To assess the impact of sample size on QBM parameter uncertainty we refit the QBM statistical model (Eqs.
+To assess the impact of sample size on QBM parameter uncertainty, we refit the QBM statistical model (Eqs.
 9-10) after removing sets of 2, 5, 10, and 15 quadrats.
@@
 2).
+IS THIS SKILLFUL FORECASTING OR IS THIS JUST STRONG AUTOCORRELATION IN COVER?
 Without climate covariates, the accuracy of forecasts from the IPM were not statistically greater than the accuracy of forecasts from the QBM (Fig.
@@
 3).
-Despite the small effects, including climate covariates did increase the accuracy of forecasts for all species except *P. smithii* (Fig. 2).
+
+Despite the weak climate effects, including climate covariates did increase the accuracy of forecasts for all species except *P. smithii* (Fig. 2).
 Increases in accuracy due to the inclusion of climate covariates were not significant (*P* > 0.05 for all comparisons of $\rho$ between climate and no-climate forecasts within model types; Fig. 2).
+PREVIOUS SENTENCE READS AS INCONSISTENT WITH NEXT SENTENCE.
 However, the IPM with climate covariates for *P. secunda* did have significantly lower error than the IPM without climate covariates ($t_{(196)}$ = -1.84, *P* = 0.033; Fig. S2).
@@
 2): forecast accuracy of *P.
-secunda* percent cover from an IPM with climate covariates was greater than the accuracy from an equivalent QBM ($t_{(195)}$ = 1.72, *P* = 0.043).
+secunda* percent cover from an IPM with climate covariates was greater than the accuracy from the QBM with climate covariates ($t_{(195)}$ = 1.72, *P* = 0.043).
+WHAT IS THE DIFFERENCE BETWEEN ACCURACY (PREVIOUS SENTENCE) AND ERROR (NEXT SENTENCE)?
 Forecasts from IPMs with climate covariates had significantly lower error than equivalent QBM forecasts for *P.
@@
 Results from all pairwise statistical tests are shown in Table S1.
+
 The accuracy of both model's forecasts declined as the forecast horizon increased, but they did so at similar rates (Fig. 4).
 The only exception is for *P. secunda* where forecast accuracy appears to slightly increase with forecast horizon, after an initial decrease from a forecast horizon of one year (Fig. 4).
+
+I TRIED TO BREAK THIS PARAGRAPH UP TO MAKE RESULTS EASIER TO FOLLOW
+
+IS THIS NEXT PART OLD?
 
@@
 Since the IPM approach relies on vital rate regressions, we were able to quantify the sensitivity of each vital rate in isolation and in pairwise combinations.
-Across all species, climate covariates can have opposing effects on different vital rates (Fig.
+Across all species, climate covariates often had opposing effects on different vital rates (Fig.
 3).
-Growth was the most sensitive vital rate for all species, showing a negative response to increased precipitation, and stronger positive response to increased temperature, and a mostly positive response when both climate factors are increased (Fig.
+Growth was the most sensitive vital rate for all species, showing a negative response to increased precipitation, and stronger positive response to increased temperature, and a mostly positive response when both climate factors were increased (Fig.
 3).
@@
 3a).
-In isolation, recruitment and survival were insensitive to climate factors for *H.
+Recruitment and survival were insensitive to climate factors for *H.
 comata* (Fig.
@@
 *P.
-secunda* equilibrium cover was sensitive to the climate effects on survival and recruitment, showing a negative effect on both vital rates for increased precipition, but a strong positive effect on survival with increased temperature (Fig.
+secunda* equilibrium cover was sensitive to the climate effects on survival and recruitment, showing a negative effect on both vital rates for increased precipitation, but a positive effect on survival with increased temperature (Fig.
 3d).
-Equilibrium cover responded negatively when increased precipitation and temperature affect recruitment  (Fig.
+Equilibrium cover OF WHICH SPECIES?
+responded negatively when increased precipitation and temperature affected recruitment  (Fig.
 3d).
@@
 3).
+BY WHAT CRITERIA?
+FIG 3 SHOWS ONLY ONE COLORED RESPONSE FOR TWO SPECIES, AND NONE FOR POA.
 
 ### Sources of uncertainty in the QBM
-Sample size had a relatively weak effect on QBM climate parameter uncertainty after the number of quadrats used in fitting exceeded about 10 (Fig.
+Sample size had a relatively weak effect on QBM climate parameter uncertainty when more than 10f quadrats were used in fitting (Fig.
 5).
-Inverse-gaussian fits show that increasing sample size beyond the number of quadrats we used would result in diminishing returns in terms of parameter certainty (Fig.
+Based on inverse-gaussian fits, increasing sample size beyond the number of quadrats we used would result in diminishing returns in terms of parameter certainty (Fig.
 5).
 
-Climate effects estimated from the QBM are most correlated with climate effects from the growth regression at the individual level (Fig.
+Climate effects estimated from the QBM were most correlated with climate effects from the growth regression at the individual level (Fig.
 6).
-In no case does the QBM statistical model have strong correlations across all three vital rates (Fig.
+In no case did the QBM statistical model have strong correlations across all three vital rates (Fig.
 6).
@@
 ### Model forecasts
-Forecasts based on 1% climate changes were extremely uncertain when we considered model error and parameter uncertainty (Fig.
+Forecasts of changes in equilibrium cover in response to 1% climate perturbations were extremely uncertain when we considered model error and parameter uncertainty (Fig.
 6; simulations with mean parameters are in Appendix D for comparison).
-As expected based on model validation (Table 1), QBM projections were more uncertain than IPM projections for all species except *P.
+Consistent with the model validation results (Table 1), QBM projections were more uncertain than IPM projections for all species except *P.
 smithiii* (Fig.
@@
 IPM forecasts for *P.
-smithiii* were very uncertain due to a very high instrinsic rate of recruitment combined with uncertainty in climate coefficients which lead to high recruitment boom years and subsequent busts when young plants suffer high mortality (Appendx C).
+smithiii* were very uncertain due to a very high intrinsic rate of recruitment combined with uncertainty in climate coefficients which led to  boom years in recruitment and subsequent busts when small plants suffered high mortality (Appendx C).
 When we included model error and parameter uncertainty, forecast changes in proportional cover always spanned a wide range of negative to positive values.
@@
 However, density-structured models rely on the aggregation of individual-level data. 
-This creates a potential problem if (1) such models inaccurately estimate density dependence or (2) are to be used in a climate change context because it is individuals, not populations, that respond to intraspecific competition and weather [@Clark2011b].
-Can models based on population-level metrics generate forecasts that are as skillful as those generated from models based on individual-level data?
-Are models based on population-level metrics as sensitive to climate as models based on individual-level data?
-Do we need detailed demographic data to forecast the state of plant populations?
+Given that individuals, not populations, respond to intraspecific competition and weather [@Clark2011b], can models based on population-level metrics generate forecasts that are as skillful as those generated from models based on individual-level data?
+Are models based on population-level metrics as sensitive to climate forcing as models based on individual-level data?
 
@@
 2; `no-climate' bars).
-We expected the IPM to out-perform the QBM since the IPM includes more mechanistic detail on the perennial plant life cycle, but this was not the case.
-Thus, it appears that, without environmental forcing, the IPM and QBM generate equivalent forecasts.
+We expected the IPM to out-perform the QBM since the IPM includes more mechanistic detail on the perennial plant life cycle, but this was not the case, at least when we ignored environmental forcing.
 Such a finding confirms theoretical [@Freckleton2011] and empirical work [@Queenborough2011] showing that density-structured models can be useful surrogates for demographic models when the goal is to estimate or forecast population states over large spatial extents.
 Likewise, the equivalency of forecast accuracy between the IPM and QBM held for all four species, when climate covariates are not included, extending the generality of previous findings [e.g., @Taylor2004; @Queenborough2011].
-
-While the models did not differ in forecast accuracy when only density-dependence was allowed to determine population dynamics, we expected that the inclusion of environmental forcing would elucidate the differences between the models.
-Our expectation was the the IPM would outperform the QBM when we included climate covariates because interannual variation in weather can affect vital rates in unique ways [@Dalgleish2011].
+THIS LAST SENTENCE SEEM REDUNDANT WITH PREVIOUS SENTENCES...AM I MISSING SOMETHING?
+COMBINE?
+
+While the models did not differ in forecast accuracy when density-dependence was the only driver of population dynamics, we expected the inclusion of environmental forcing to reveal more differences between the models.
+We expected the IPM to outperform the QBM when we included climate covariates because interannual variation in weather can affect vital rates in different ways [@Dalgleish2011].
 Thus, estimates of climate effects on plant population growth may be biased or non-identifiable when the underlying statistical model is fit using population-level data that integrates over the potentially unique climate responses of individual vital rates.
-However, IPM forecast accuracy was only significantly higher than the QBM for one species, *P. secunda* (Fig. 2, `climate' bars).
-This result likely stems from the relatively weak climate effects in vital rate regressions (Fig. 3).
+However, we found that IPM forecast accuracy was only significantly higher than the QBM for one species, *P. secunda* (Fig. 2, `climate' bars).
+This result likely stems from the relatively weak climate effects in the vital rate regressions (Fig. 3).
 If climate effects on vital rates are weak, then models based on percent cover or density should fair well compared to models based on individual-level data so long as density-dependence is adequately estimated (see Fig.
@@
 3).
-Thus, it appears that the estimation of several small effects, rather than a single relatively large effect, can lead to increased forecast skill.
-
-The higher accuracy of the IPM with climate for *P.
-secunda*  highlights the superiority of contemporary model and variable selection approaches such as ridge regression and LASSO over techniques that lead to excluding "non-significant" effects from final models.
-Especially in a Bayesian forecasting framework, ridge regression allows researchers to include forcing variables that we know are changing (e.g., temperature and precipitation) in final models even if the estimated effects are relatively small or even deemed unimportant due to noisy data, out of which it is more difficult to estimate effects on interannual variation.
+Thus, it appears that the estimation of several small but consistent effects, rather than a single relatively large effect, can lead to increased forecast skill.
+
+The higher accuracy of the IPM with climate covariates for *P.
+secunda*  highlights the advantage of contemporary model and variable selection approaches such as ridge regression and LASSO over techniques that would exclude "non-significant" effects from final models.
+Especially in a Bayesian forecasting framework, ridge regression allows researchers to include forcing variables that we know are changing (e.g., temperature and precipitation) in final models even if the estimated effects are relatively small or even deemed unimportant due to noisy data, out of which it is more difficult to estimate effects on interannual variation [REWERITE THIS SENTENCE, I DON'T KNOW HOW TO RESCUE IT CUZ I DON'T UNDERSTAND IT].
 Likewise, if a species is truly unresponsive to a given climate variable, statistical regularization techniques will shrink the mean and variance of a covariate estimate to zero [@Hooten2015].
-No amount of statistical "machismo" during the model fitting stage can make up for including the appropriate candidate covariates, which we attempted to do based on our knowledge of this semi-arid plant community.
-However, the climate covariates we chose required aggregating daily weather data over specific time periods.
-It may be that we chose the wrong time periods over which to aggregate.
+Of course, no matter what model selection approach is adopted, a critical step is identifying the appropriate candidate covariates, which we attempted to do based on our knowledge of this semi-arid plant community.
+However, the climate covariates we chose required aggregating daily weather data over discrete time periods.
+It is possible that we did not choose the optimal time periods over which to aggregate.
 New methods using functional linear models (or splines) may offer a data-driven approach for identifying the appropriate time periods over which to aggregate to produce a tractable set of candidate climate variables [@Teller2016].
@@
 We also expected IPM forecast accuracy to decline at a lower rate than the QBM as the forecast horizon increased.
-Indeed, many have argued that more mechanistic models produce better predictions, especially under novel conditions [@Evans2012b; @Schindler2015b].
-Likewise, the IPM explicitly models the processes of recruitment and survival that may be poorly represented in the QBM since recruitment and survival mainly affect small plants that contribute little to overall changes in percent cover.
-But the addition and subtraction of small plants can have large effects on population growth over time and potentially lead to a longer forecast horizon for models that better represent the plant life cycle.
-We found no evidence for a difference between the IPM and QBM forecast horizons (Fig. 4).
-Similar forecast horizons should be expected if both models adequately capture density dependence and environmental forcing is negligble, as is the case for our models (Fig. 2).
-In fact, only the forecast horizons for *P. secunda* noticeably differ (Fig. 4), and that is also the only species where we found the IPM to outperform the QBM when we included climate effects.
-
-In conclusion, we found that demographic data was generally unneccesary to generate skillful forecasts of plant population states.
-Such a finding is contrary to our expectations, but in agreement with recent theoretical [@Freckleton2011] and emprical work [@Queenborough2011].
+In principle, more mechanistic models should produce better predictions, especially under novel conditions [@Evans2012b; @Schindler2015b].
+In our case, the IPM explicitly models the influence of weather on recruitment and survival, effects that may be poorly represented in the QBM since recruitment and survival mainly affect small plants that contribute little to year-to-year changes in percent cover.
+Over time, of course, the addition and subtraction of small plants can have large effects on population growth, so explicitly modeling these effects could contribute to a longer forecast horizon.
+However, we found no evidence for a difference between the IPM and QBM forecast horizons (Fig.
+4).
+Similar forecast horizons should be expected if both models adequately capture density dependence and environmental forcing is negligble, as is the case for our models (Fig.
+2).
+In fact, only the forecast horizons for *P. secunda* noticeably differ (Fig. 4), and that is also the only species for which the IPM outperformed the QBM when we included climate effects.
+
+In conclusion, we found that models based on individual-level demographic data generally failed to generate more skillful population forecasts than models based on population-level data.
+This finding runs counter to our expectations, but is consistent with recent theoretical [@Freckleton2011] and empirical work [@Queenborough2011].
 However, when we included climate covariates, we did achieve more skillful forecasts from the IPM for one species, *P.
-secunda*, that appears to respond consistently, but weakly, to several climate variables.
+secunda*, that appears to respond consistently, but weakly, to several climate variables [THIS SENTENCE NEEDS WORK, BUT I AM GOING TO BED].
 Thus, we conclude that models based on population-level data, rather than individual-level data, may be adequate for forecasting the state of plant populations for species that do not respond consistently or strongly to climate.
@@
 Nonetheless, our results are encouraging for the use of easy-to-collect percent cover for forecasting the state of plant populations.
+
+IS THE REST OLD? I STOPPED HERE.
 

### 2016-05-25 manuscript/manuscript.Rmd
rewrites from PBA suggestions

@@
 
-**5.** *Synthesis.* For our focal species at this particular location, and using our particular statistical models, demographic data was generally unnecessary to achieve skillful [ARE ANY OF THESE FORECASTS SKILLFUL?
-FOCUS INSTEAD ON MODEL DIFFERENCES] forecasts, though for certain species forecast skill can be gained by using demographic data linked to climate covariates.
+**5.** *Synthesis.* For our focal species at this particular location, and using our particular statistical models, percent cover models generated forecasts as skillful as those from a demographic model.
+However, for certain species forecast skill can be gained by using demographic data linked to climate covariates.
 We conclude that models based on aggregated individual-level data offer a practical alternative to data-intensive demographic models when species do not respond strongly to interannual variation in weather, but when modeling species that do respond to climate drivers, demographically-based models can generate more skillful forecasts.
@@
 The limited spatial extent of individual-level demographic datasets constrains our ability to use population models to address applied questions about the consequences of climate change.
-
-<!--
-The inability of most population models to address landscape-scale problems may explain why land managers and conservation planners have embraced species distribution models (SDMs) [see @Guisan2005 for a review].
-SDMs typically rely on easy-to-collect presence/absence data and remotely-sensed environmental covariates that allow researchers to model large spatial extents [e.g., @Maiorano2013]  and project future changes in occupancy over landscapes and regions.
-However, the limitations of SDMs are well known [@Pearson2003; @Elith2009; @Araujo2012].
-Ideally, researchers would provide managers with landscape-scale population models, combining the extent of SDMs with information about dynamics and species abundances [@Schurr2012; @Merow2014].
--->
 
@@
 For example, a one-year change in a plant species' cover or biomass might reflect individual growth or shrinkage, whereas the long-term trajectory of the population might be more influenced by recruitment.
-If important climate effects are missed because of the aggregation inherent in in population-level data, then population models built with such data will make uninformative or unreliable forecasts.
+The same is true for density dependence: intraspecific density depedence may act most strongly on vital rates, like recruitment, that are difficult to identify from population-level data.
+If density dependence and/or important climate effects are missed because of the aggregation inherent in in population-level data, then population models built with such data will make uninformative or unreliable forecasts.
 
@@
 We used a demographic dataset that tracks the fates of individual plants from four species over 14 years to build two kinds of single-species population models, traditional models using individual growth, survival, and recruitment data and alternative models based on basal cover.
-We simulated the models to answer two questions: (1) Can population models fit using aggregated individual-level data (percent cover) adequately capture density dependence to produce forecasts as skillful as those from models fit to demographic data?
-(2) Can population models fit using aggregated data adequately capture the influence of climate on population growth and, in turn, produce forecasts as skillful as those from models fit to demographic data?
-[THIS NEXT SENTENCE SEEMS OUT OF PLACE.
-CAN YOU MODIFY PREVIOUS PARAGRAPH TO ADDRESS DENSITY DEPENDENCE ALONG WITH CLIMATE EFFECTS?
-E.G.
-DENSITY DEPENDENCE MIGHT ACT MOST STRONGLY ON RECRUITMENT WHICH IS HARD TO DETECT IN A COVER BASED MODEL?] Both questions are motivated by the fact that the effects of intraspecific competition (density dependence) and interannual weather variability act at the level of the individual [@Clark2011b], meaning that models based on individual-level processes should generate more skillful forecasts.
+We simulated the models to answer two questions motivated by the fact that the effects of intraspecific competition (density dependence) and interannual weather variability act at the level of the individual [@Clark2011b].
+First, can population models fit using aggregated individual-level data (percent cover) adequately capture density dependence to produce forecasts as skillful as those from models fit to demographic data?
+Second, can population models fit using aggregated data adequately capture the influence of climate on population growth and, in turn, produce forecasts as skillful as those from models fit to demographic data?
 
@@
 
-YOU NEED TO EXPLAIN HOW THE w's ARE CALCULATED. COPY TEXT FROM ADLER ET AL. 2010 or CHU  & ADLER 2015.
+We follow the approach of @Chu2015 to estimate crowding, assuming that the crowding experienced by a focal genet depends on distance to each neighbor genet and the neighbor's size, _u_:
+
+\begin{equation}
+w_{i,t} = \sum_k e^{-\deltad_{ik,t}^{2}}u_{k,t}.
+\end{equation}
+
+In the above, $w_{i,t}$ is the crowding that genet _i_ in year _t_ experiences from conspecific neighbors.
+The spatial scale over which conspecific neighbors exert influence on any genet is determined by $\delta$.
+The function is applied for all _k_ conspecific genets that neighbor the focal genet at time _t_, and $d_{ik,t}$ is the distance between genet _i_ and conspecific genet _k_.
+We use regression-specific (survival and growth) $\delta$ values estimated by @Chu2015.
 
@@
 We first calculated the median predicted cover across the 100 simulations for each quadrat-year and then calculated forecast skill as the correlation ($\rho$) between forecasts and observations.
-We compared $\rho$ between model types and within model types between models with and without climate covariates using a one-sided *t* test with adjusted degrees of freedom following @Wilcox2009 and standard errors calculated using the HC4 estimator of @Cribari-Neto2004.
+We calculated forecast error as mean absolute error (MAE) between forecasts and observations.
+We compared $\rho$ and MAE between model types and within model types between models with and without climate covariates using one-sided *t* tests with adjusted degrees of freedom following @Wilcox2009 and standard errors calculated using the HC4 estimator of @Cribari-Neto2004.
 Statistical tests for differences between $\rho$ were conducted using R functions from @Ye2015a.
@@
 
-<!--
-### Testing sensitivity to climate covariates
-With our fitted and validated models in hand, we ran simulations for each model type (IPM and QBM) under four climate perturbation scenarios: (1) observed climate, (2) precipitation increased by 1%, (3) temperature increased by 1%, and (4) precipitation and temperature increased by 1%.
-We ran the simulations for 2,500 time steps, enough to estimate equilibrium cover after discarding an initial 500 time steps as burn-in.
-EXPLAIN HOW CLIMATE COVARIATES ARE DRAWN AT EACH TIME STEP.
-AND NO RANDOM YEAR EFFECTS?
-Each simulation was run under two parameter scenarios: (1) using mean parameter estimates and (2) using randomly drawn parameters from the MCMC chain.
-We use (1) to detect the overall sensitivity of equilibrium cover to climate, and we use (2) to show the impact of model and parameter uncertainty on forecast precision.
-
-As an effort to identify potential discrepencies between IPM and QBM forecasts, we also ran simulations designed to quantify the sensitivities of individual and combined vital rates to climate for the IPM.
-Specifically, we ran simulations for the above climate scenarios, but applied the perturbed climate covariates to survival, growth, or recruitment vital rates individually and in pairwise combinations.
-This allowed us to isolate the vital rate(s) most sensitive to climate.
-For this analysis, we used mean parameter estimates to reduce the sources of uncertainty in the sensitivity estimates.
-
-We expected the IPM to produce more accurate and precise forecasts due to either (1) the smaller sample size of the quadrat level data sets compared to the individual level data sets, leading to larger parameter uncertainty for the QBM, or (2) the QBM climate effects being weakly associated with one or more vital rate climate effects at the individual level.
-To assess the impact of sample size on QBM parameter uncertainty, we refit the QBM statistical model (Eqs.
-9-10) after removing sets of 2, 5, 10, and 15 quadrats.
-We fit 10 models at each level of quadrat removal (2, 5, 10, 15 quadrats), removing a different randomly selected set of quadrats for each fit.
-We calculated the standard deviation of climate main effects (pptLag, ppt1, ppt2, TmeanSpr1, and TmeanSpr2) for each model and averaged those over replicates within each set of quadrat removals.
-This allowed us to regress parameter uncertainty against sample size.
-
-To deterime if the QBM climate effects are correlated with climate effects for each vital rate model in the IPM, we simply regressed the QBM climate coefficients against each vital rate model's climate coefficients and calculated Pearson's $\rho$.
-Strong correlations indicate the QBM is capable of detecting climate effects associated with individual vital rates.
-A weak correlation indicates the QBM "misses" the climate effect on a particular vital rate.
--->
-
 Results
 -------
-Both the IPM and QBM generated skillful one-step-ahead forecasts for out-of-sample observations, with an average correlation between predictions and observations ($\rho$) of 0.73 across all models and species (Fig.
+The IPM and QBM generated one-step-ahead forecasts of similar skill for out-of-sample observations, with an average correlation between predictions and observations ($\rho$) of 0.73 across all models and species (Fig.
 2).
-IS THIS SKILLFUL FORECASTING OR IS THIS JUST STRONG AUTOCORRELATION IN COVER?
 Without climate covariates, the accuracy of forecasts from the IPM were not statistically greater than the accuracy of forecasts from the QBM (Fig.
@@
 Increases in accuracy due to the inclusion of climate covariates were not significant (*P* > 0.05 for all comparisons of $\rho$ between climate and no-climate forecasts within model types; Fig. 2).
-PREVIOUS SENTENCE READS AS INCONSISTENT WITH NEXT SENTENCE.
-However, the IPM with climate covariates for *P. secunda* did have significantly lower error than the IPM without climate covariates ($t_{(196)}$ = -1.84, *P* = 0.033; Fig. S2).
+However, the IPM with climate covariates for *P. secunda* did have significantly lower error (mean absolute error) than the IPM without climate covariates ($t_{(196)}$ = -1.84, *P* = 0.033; Fig. S2).
 In only one case were IPM forecasts significantly more accurate than the QBM (Fig.
@@
 secunda* percent cover from an IPM with climate covariates was greater than the accuracy from the QBM with climate covariates ($t_{(195)}$ = 1.72, *P* = 0.043).
-WHAT IS THE DIFFERENCE BETWEEN ACCURACY (PREVIOUS SENTENCE) AND ERROR (NEXT SENTENCE)?
 Forecasts from IPMs with climate covariates had significantly lower error than equivalent QBM forecasts for *P.
@@
 The accuracy of both model's forecasts declined as the forecast horizon increased, but they did so at similar rates (Fig. 4).
-The only exception is for *P. secunda* where forecast accuracy appears to slightly increase with forecast horizon, after an initial decrease from a forecast horizon of one year (Fig. 4).
-
-I TRIED TO BREAK THIS PARAGRAPH UP TO MAKE RESULTS EASIER TO FOLLOW
-
-IS THIS NEXT PART OLD?
-
-<!--
-### Comparison of forecast models
-
-### Sensitivity of models to climate
-The response of a population to climate change is a result of the aggregate effects of climate on individual vital rates.
-Since the IPM approach relies on vital rate regressions, we were able to quantify the sensitivity of each vital rate in isolation and in pairwise combinations.
-Across all species, climate covariates often had opposing effects on different vital rates (Fig.
-3).
-Growth was the most sensitive vital rate for all species, showing a negative response to increased precipitation, and stronger positive response to increased temperature, and a mostly positive response when both climate factors were increased (Fig.
-3).
-*B.
-gracilis* survival rates were sensitive to temperature, resulting in an increase in plant cover under increased temperature (Fig.
-3a).
-Recruitment and survival were insensitive to climate factors for *H.
-comata* (Fig.
-3b).
-Survival and recruitment of *P.
-smithii* were both sensitive, negatively, to temperature and precipitation (Fig.
-3c).
-*P.
-secunda* equilibrium cover was sensitive to the climate effects on survival and recruitment, showing a negative effect on both vital rates for increased precipitation, but a positive effect on survival with increased temperature (Fig.
-3d).
-Equilibrium cover OF WHICH SPECIES?
-responded negatively when increased precipitation and temperature affected recruitment  (Fig.
-3d).
-At least two of three vital rates were sensitive to climate for each species (Fig.
-3).
-BY WHAT CRITERIA?
-FIG 3 SHOWS ONLY ONE COLORED RESPONSE FOR TWO SPECIES, AND NONE FOR POA.
-
-### Sources of uncertainty in the QBM
-Sample size had a relatively weak effect on QBM climate parameter uncertainty when more than 10f quadrats were used in fitting (Fig.
-5).
-Based on inverse-gaussian fits, increasing sample size beyond the number of quadrats we used would result in diminishing returns in terms of parameter certainty (Fig.
-5).
-
-Climate effects estimated from the QBM were most correlated with climate effects from the growth regression at the individual level (Fig.
-6).
-In no case did the QBM statistical model have strong correlations across all three vital rates (Fig.
-6).
-QBM climate effects were most weakly correlated with those from individual-level recruitment models for *B.
-gracilis*, *H.
-comata*, and *P.
-secunda* (Fig.
-6a,b,d).
-For *P.
-smithii*, QBM climate effects showed no correlation with the survival model effects (Fig.
-6c).
-
-### Model forecasts
-Forecasts of changes in equilibrium cover in response to 1% climate perturbations were extremely uncertain when we considered model error and parameter uncertainty (Fig.
-6; simulations with mean parameters are in Appendix D for comparison).
-Consistent with the model validation results (Table 1), QBM projections were more uncertain than IPM projections for all species except *P.
-smithiii* (Fig.
-6).
-IPM forecasts for *P.
-smithiii* were very uncertain due to a very high intrinsic rate of recruitment combined with uncertainty in climate coefficients which led to  boom years in recruitment and subsequent busts when small plants suffered high mortality (Appendx C).
-When we included model error and parameter uncertainty, forecast changes in proportional cover always spanned a wide range of negative to positive values.
-In other words, neither model could predict whether a climate perturbation would increase or decrease equilibrium population size.
--->
+The only exception is for *P. secunda* where forecast accuracy appears to remain steady as the forecast horizon increases, after an initial decrease from a forecast horizon of one year (Fig. 4).
 
@@
 We expected the IPM to out-perform the QBM since the IPM includes more mechanistic detail on the perennial plant life cycle, but this was not the case, at least when we ignored environmental forcing.
-Such a finding confirms theoretical [@Freckleton2011] and empirical work [@Queenborough2011] showing that density-structured models can be useful surrogates for demographic models when the goal is to estimate or forecast population states over large spatial extents.
-Likewise, the equivalency of forecast accuracy between the IPM and QBM held for all four species, when climate covariates are not included, extending the generality of previous findings [e.g., @Taylor2004; @Queenborough2011].
-THIS LAST SENTENCE SEEM REDUNDANT WITH PREVIOUS SENTENCES...AM I MISSING SOMETHING?
-COMBINE?
+Such a finding confirms theoretical [@Freckleton2011] and empirical work [@Taylor2004; @Queenborough2011] showing that density-structured models can be useful surrogates for demographic models when the goal is to estimate or forecast population states over large spatial extents.
 
@@
 secunda*  highlights the advantage of contemporary model and variable selection approaches such as ridge regression and LASSO over techniques that would exclude "non-significant" effects from final models.
-Especially in a Bayesian forecasting framework, ridge regression allows researchers to include forcing variables that we know are changing (e.g., temperature and precipitation) in final models even if the estimated effects are relatively small or even deemed unimportant due to noisy data, out of which it is more difficult to estimate effects on interannual variation [REWERITE THIS SENTENCE, I DON'T KNOW HOW TO RESCUE IT CUZ I DON'T UNDERSTAND IT].
-Likewise, if a species is truly unresponsive to a given climate variable, statistical regularization techniques will shrink the mean and variance of a covariate estimate to zero [@Hooten2015].
+Ridge regression allows researchers to retain covariates whose effects may be difficult to identify in noisy data or short time series.
+This is especially important when forecasting the impacts of climate change where it is important to include to effects of forcing variables (e.g., temperature and precipitation) even of such effects are difficult to identify.
+If a species is truly unresponsive to a given climate variable, statistical regularization techniques will shrink the mean and variance of a covariate estimate to zero [@Hooten2015].
 Of course, no matter what model selection approach is adopted, a critical step is identifying the appropriate candidate covariates, which we attempted to do based on our knowledge of this semi-arid plant community.
@@
 This finding runs counter to our expectations, but is consistent with recent theoretical [@Freckleton2011] and empirical work [@Queenborough2011].
-However, when we included climate covariates, we did achieve more skillful forecasts from the IPM for one species, *P.
-secunda*, that appears to respond consistently, but weakly, to several climate variables [THIS SENTENCE NEEDS WORK, BUT I AM GOING TO BED].
+However, we did achieve more skillful forecasts from the IPM with climate covariates for one species, *Poa secunda*, that appears to respond consistently to several climate variables.
 Thus, we conclude that models based on population-level data, rather than individual-level data, may be adequate for forecasting the state of plant populations for species that do not respond consistently or strongly to climate.
@@
 Nonetheless, our results are encouraging for the use of easy-to-collect percent cover for forecasting the state of plant populations.
-
-IS THE REST OLD? I STOPPED HERE.
-
-<!---
-### The importance of demographic data
-Our comparison of a traditional, demographic population model (the IPM) with a model inspired by density-structured models (the QBM) showed that the IPM outperformed the QBM: the IPM was more accurate and precise than the QBM in out-of-sample cross validation (Table 1).
-The superiority of the IPM could reflect either differences in sample size or the effect of averaging over unique effects of climate on each individual-level vital rate.
-Although increasing sample size of quadrat percent-cover observations would be easy to do in the field, we found little evidence that it would lead to higher precision of climate coefficient estimates (Fig.
-4).
-
-We did, however, find evidence that the QBM statistical model failed to identify climate dependence for some vital rates (Fig.
-5).
-For no species were climate effects from the QBM strongly correlated with all three vital rates (Fig.
-5).
-@Freckleton2011 acknowledge that averaging over complex stage dependence will lead to poorly specified models.
-This is analagous to our situation, but instead of averaging over complex life histories, we are averaging over complex climate dependence.
-Though our work here focused on plant species, this finding is applicable to any species with vital rates that respond uniquely to weather/climate.
-
-Our interpretation is that the QBM is "missing" climate signals associated with at least one vital rate for each species.
-This leads to inaccurate and imprecise forecasts because the QBM statistical model struggles to explain variation due to climate variables that have positive and negative impacts on different vital rates.
-When this is the case, as it is for all our species to varying degrees (Fig.
-3), forecasts from models based on population-level data will fail.
-Our result is consistent with related work on the importance of individual-level data to forecast population responses to exogenous drivers [@Clark2011; @Clark2011b; @Clark2012; @Galvan2014].
-
-Detailed demographic data appears to be necessary to forecast climate change impacts on plant populations when vital rates have unique climate responses.
-How then can we build models to make forecasts for the landscape and regional scales beyond the scope of traditional population models [@Queenborough2011]?
-There are alternatives to density-structured models.
-For example, @Clark2011 use Forest Inventory and Analysis (FIA) data to parameterize a population model with multiple vital rates and climate dependence.
-Distributed efforts such as PlantPopNet (http://plantago.plantpopnet.com) will allow researchers to estimate variation around climate responses for widespread species by taking advantage of spatial variation in climate [e.g.
-@Doak2010].
-Finally, new approaches on the horizon that leverage photo/video of plots and advanced object recognition algorithms [e.g.
-@Liu2014] will increase the efficiency of plant mapping and digitizing efforts.
-
-### The challenge of uncertainty
-
-An important, but unexpected, result of our analysis was the great uncertainty in forecasts, even for our best model.
-The typical approach in ecology is to use point estimates of model parameters to project populations forward according to the specified model, usually allowing for some variability around the determinstic process [e.g.
-@Battin2007; @Jenouvrier2009; @Adler2012].
-If we follow tradition and calculate the mean response to climate perturbation with only model error and interannual variation included, the IPM and the QBM produce opposing forecasts for three of four species (Fig.
-D1).
-It would be tempting to interpret this inconsistency as further evidence for the superiority of the IPM.
-However, if we introduce parameter uncertainty, the forecasts are actually indistinguishable (Fig.
-6), though the IPM projections are generally more precise (consistent with our cross-validation results).
-The real story is that both models produce highly uncertain forecasts.
-For all species, the 90% quantiles of predicted changes in population size overlapped zero; we cannot even predict whether a change in precipitation or temperature will cause populations to increase or decrease.
-This result held when we tried perturbing climate by 10% and 20% as well.
-
-Our results highlight the state of affairs in ecology when it comes to forecasting the impacts of climate change.
-The analysis we conducted here could be considered at the forefront of ecological forecasting with respect to the statistical approach employed (hierarchical Bayesian), the type of population model we used (density-dependent, stochastic IPM with parameter uncertainty), and the amount of high quality data we had at our disposal (14 years of individual-level data).
-Yet, model predictions proved so uncertain that any forecast, when bounded with model and parameter uncertainty, would be uninformative.
-
-How might we improve on this state of affairs?
-First, forecasts could be improved by matching the spatial scale of predictor variables with the spatial scale of observations.
-One of the major limitations of the models we fit here is that the climate data are collected at a larger scale than the individual-level observations of plant size.
-Climate covariates only vary by year, with no spatial variability within years.
-Thus, even if we fit models to individual-level data, we are missing the key interaction point between weather and individual plants [@Clark2011b] because all observations share the same climate covariates.
-Demographic studies should be designed with at least plot-level measurements of climate related variables (e.g., soil moisture).
-Second, accurately detecting climate signals will take even longer time series.
-Recent theoretical work on detecting climate signals in noisy data suggests that even advanced approaches to parameter fitting require 20-25 year time series [@Teller2016].
-Third, ecologists need a stronger commitment to reporting uncertainty.
-Although most modeling studies explicitly consider model uncertainty, parameter uncertainty is often ignored.
-In some cases this is because the most convenient statistical methods make it difficult to propogate parameter uncertainty.
-Yet even Bayesian approaches that allow integration of model fitting and forecasting [@Hobbs2015] are not simple when using modeling approaches like integral projection models that separate the model fitting and simulation stages [@Rees2009].
-However, as we have done here, it is still possible to include parameter uncertainty by drawing parameter values from MCMC iterations, taking care to draw all parameters from the same chain and iteration to account for their correlations.
-Only by being honest about our forecasts can we begin to produce better ones, and forecasts reported without parameter error are disingenuous.
-Ignoring parameter error may be justifiable when the goal is investigating basic processes, but it is indefensible when forecasting is the goal.
-
-Conclusions
------------
-This work is not a critique of density-structured population models.
-We are confident that density-structured models will prove to be a valuable tool for many applications.
-However, our analysis represents the first comparison, to our knowledge, of population models based on individual and aggregated forms of the same data in a climate change context.
-Our results confirm theoretical arguments [@Clark2011b] and empirical evidence [@Clark2011; @Clark2012] that individual responses are critical for predicting species' responses to climate change.
-It seems there is no short cut to producing accurate and precise population forecasts: we need detailed demographic data to forecast the impacts of climate change on populations.
-Given the importance of demographic data and its current collection cost, we need modern methods to collect demographic data more efficiently across environmental gradients in space and time.
-
-Our results also offer a cautionary tale because forecast uncertainty was large for both model types.
-Even with 14 years of detailed demographic data and sophisticated modeling techniques, our projections contained too much uncertainty to be informative.
-Uncertainty in demographic responses to climate can be reduced by collecting (1) longer time series and (2) climate covariates that match the scale of inference (e.g., plot rather than landscape level climate/weather metrics).
---->
 
@@
 \centering
-\caption{Description of data.} 
+\caption{Description of data. The observations span 13 year-to-year transitions.} 
 \begin{tabular}{llrr}

### 2016-05-25 manuscript/manuscript.Rmd
updates ms with MBH edits

@@
 \textsf{\textit{\footnotesize{\textsuperscript{1}Department of Wildland Resources and the Ecology Center, 5230 Old Main Hill, Utah State University, Logan, Utah 84322, USA; \textsuperscript{2}U.S.
-Geological Survey, Colorado Cooperative Fish and Wildlife Research Unit, Department of Fish, Wildlife and Conservation Biology, Colorado State University, Fort Collins, CO 80523-1403, USA; \textsuperscript{3}Department of Fish, Wildlife, and Conservation Biology, Colorado State University, Fort Collins, CO 80523 USA}; \textsuperscript{4}Department of Statistics, Colorado State University, Fort Collins, CO 80523 USA}}
+Geological Survey, Colorado Cooperative Fish and Wildlife Research Unit, Fort Collins, CO 80523, USA; \textsuperscript{3}Department of Fish, Wildlife, and Conservation Biology, Colorado State University, Fort Collins, CO 80523 USA}; \textsuperscript{4}Department of Statistics, Colorado State University, Fort Collins, CO 80523 USA}}
 
@@
 Aggregate measures of population status, rather than individual performance, offer an intriguing alternative for modeling populations [@Clark2004; @Freckleton2011].
-Population-level data cannot provide inference about demographic mechanisms, but might be sufficient for modeling future population states, especially since such data are feasible to collect across broad spatial extents [e.g., @Queenborough2011].
+Population-level data cannot provide inference about demographic mechanisms, but might be sufficient for modeling future population states, especially because such data are feasible to collect across broad spatial extents [e.g., @Queenborough2011].
 The choice between individual and population-level data involves a difficult trade-off: while individual-level data are necessary for mechanistic models, population-level data enable models that can be applied over greater spatial and temporal extents.
@@
 
-Here, we compare the forecasting skill of statistical and population models based on aggregated, population-level data with the skill of models based on individual-level data.
+We compared the forecasting skill of statistical and population models based on aggregated, population-level data with the skill of models based on individual-level data.
 We used a demographic dataset that tracks the fates of individual plants from four species over 14 years to build two kinds of single-species population models, traditional models using individual growth, survival, and recruitment data and alternative models based on basal cover.
@@
 comata*, *P.
-smithii*, and *P.
-secunda* are cool-season perennial grasses.
+smithii*, and *Poa secunda* are cool-season perennial grasses.
 All species typically begin growth in the early spring, reach maximum growth and flower in early to mid summer (May-June), and disperse seed in mid to late summer (July-September).
@@
 The quadrats were distributed among six pastures, each assigned a grazing treatment of light (1.24 ha/animal unit month), moderate (0.92 ha/aum), and heavy (0.76 ha/aum) stocking rates (two pastures per treatment).
-In this analysis we account for potential differences among the grazing treatments, but do not focus on grazing$\times$climate interactions.
+In this analysis, we accounted for potential differences among the grazing treatments, but do not focus on grazing$\times$climate interactions.
 The annual maps of the quadrats were digitized and the fates of individual plants tracked and extracted using a computer program [@Lauenroth2008; @Chu2014a].
@@
 The individual data is the "raw" data.
-For the quadrat-level, we data we simply sum individual basal cover for each quadrat by species.
+For the quadrat-level data we simply sum individual basal cover for each quadrat by species.
 This is equivalent to a near-perfect census of quadrat percent cover because measurement error at the individual-level is small [@Chu2014].
 Based on these two datasets of 13 year-to-year transitions, we can compare population models built using individual-level data and aggregated, quadrat-level data.
-At the individual level we explicitly model three vital rates: growth, survival, and recruitment.
-At the quadrat level we model population growth as change in percent cover of quadrats with non-zero cover in year *t* and in year *t-1*, ignoring within-quadrat extirpation and colonization events because they are very rare in our time series ($N=16$ and $N=13$, respectively, across all species).
+At the individual level, we explicitly model three vital rates: growth, survival, and recruitment.
+At the quadrat level, we model population growth as change in percent cover of quadrats with non-zero cover in year *t* and in year *t-1*, ignoring within-quadrat extirpation and colonization events because they are very rare in our time series ($N=16$ and $N=13$, respectively, across all species).
 Sample sizes for each species and vital rate model are shown in Table 1.
@@
 All R code and data necessary to reproduce our analysis is archived on GitHub as release v1.0[^revnote] (http://github.com/atredennick/MicroMesoForecast/releases).
-That stable release will remain static as a record of this analysis, but subsequent versions may appear if we update this work.
 We have also deposited the v1.0 release on Dryad (*link here after acceptance*).
@@
 At the quadrat-level, we fit a single regression model for population growth.
-We describe the statistical models separately since they required different approaches.
+We describe the statistical models separately because they required different approaches.
 For both model types, we fit vital rate models with and without climate covariates.
@@
 We fit all models using a hierarchical Bayesian approach.
-The models are fully descibed in Appendix A, so here we focus on the main process and the model likelihood.
+We focus on the main process and the model likelihood in what follows (full model descriptions are in the Supporting Information).
 For the likelihood models, $\textbf{y}^X$ is always the relevant vector of observations for vital rate *X* (*X* = *S*, *G*, *R*, or *P*) for survival, growth, recruitment, or population growth).
-For example, $\textbf{y}^S$ is a vector of 0's and 1's indicating whether a genet survives from *t* to *t+1*, or not.
+For example, $\textbf{y}^S$ is a vector of 0s and 1s indicating whether a genet survives from *t* to *t+1*, or not, for all observation years and quadrats.
 All model parameters are species-specific, but we omit subscripts for species in model descriptions below to reduce visual clutter.
@@
 #### Vital rate models at the individual level
-We used logistic regression to model survival probability ($s$) of a genet:
+We used logistic regression to model the probability that genet *i* in quadrat *q* survives from time *t* to *t*+1 ($s_{i,q,t}$):
 
@@
 \begin{align}
-\textbf{y}^{S} &\sim \text{Bernoulli}(\textbf{s}) \\
-\text{logit}[s(x,\textbf{z}_t,w)] &= \beta_{0,t} + \beta_{s,t}x + \beta_Q + \textbf{z}'_t \boldsymbol{\beta}_c + \beta_{d,1} w + \beta_{d,2} (xw)
+y_{i,q,t}^{S} &\sim \text{Bernoulli}(s_{i,q,t}) \\
+\text{logit}(s_{i,q,t}) &= \beta_{0,t} + \beta_{s,t}x_{i,q,t} + \beta_Q + \textbf{z}'_t \boldsymbol{\beta}_c + \beta_{d,1} w_{i,t} + \beta_{d,2} (x_{i,q,t}w_{i,q,t})
 \end{align}
@@
 
-where $x$ is the log of genet basal area, $\beta_{0,t}$ is a year specific intercept intercept, $\beta_Q$ is the random effect of quadrat group location, $\beta_{s,t}$ is the year-specific slope parameter for size, **z** is a vector of *i* climate covariates specific to year *t*, $\boldsymbol{\beta}_c$ is a vector of fixed climate effects of length *i*, $\beta_{d,1}$ is the effect of intraspecific crowding experienced by the focal genet ($w$), and $\beta_{d,2}$ is a size by crowding ($xw$) interaction effect.
+where $x_{i,q,t}$ is the log of genet *i* basal area at time *t*, $\beta_{0,t}$ is a year specific intercept, $\beta_Q$ is the random effect of quadrat group location, $\beta_{s,t}$ is the year-specific slope parameter for size, **z** is a vector of *p* climate covariates specific to year *t*, $\boldsymbol{\beta}_c$ is a vector of fixed climate effects of length *p*, $\beta_{d,1}$ is the effect of intraspecific crowding experienced by the focal genet at time *t* ($w_{i,q,t}$), and $\beta_{d,2}$ is a size by crowding ($x_{i,q,t}w_{i,q,t}$) interaction effect.
 
@@
 \begin{equation}
-w_{i,t} = \sum_k e^{-\deltad_{ik,t}^{2}}u_{k,t}.
+w_{i,q,t} = \sum_k e^{-\delta d_{ik,q,t}^{2}}u_{k,q,t}.
 \end{equation}
 
-In the above, $w_{i,t}$ is the crowding that genet _i_ in year _t_ experiences from conspecific neighbors.
+In the above, $w_{i,t}$ is the crowding that genet _i_ in year _t_ experiences from conspecific neighbors in quadrat *q*.
 The spatial scale over which conspecific neighbors exert influence on any genet is determined by $\delta$.
-The function is applied for all _k_ conspecific genets that neighbor the focal genet at time _t_, and $d_{ik,t}$ is the distance between genet _i_ and conspecific genet _k_.
+The function is applied for all _k_ conspecific genets that neighbor the focal genet at time _t_, and $d_{ik,q,t}$ is the distance between genet _i_ and conspecific genet _k_ in quadrat *q*.
 We use regression-specific (survival and growth) $\delta$ values estimated by @Chu2015.
 
-We modeled growth as a Gaussian process describing genet size at time $t+1$ as a function of size at $t$ and climate covariates:
+We modeled growth as a Gaussian process describing log genet size ($y_{i,q,t+1}^{G}$) at time $t+1$ in quadrat *q* as a function of log size at time $t$ and climate covariates:
 
@@
 \begin{align}
-\textbf{y}^G &\sim \text{Normal}(\boldsymbol{\mu}, \sigma^2_{x,t}) \\
-\mu(x,\textbf{z}_t,w) &= \beta_{0,t} + \beta_{s,t}x + \beta_Q + \textbf{z}'_t \boldsymbol{\beta}_c + \beta_{d,1} w + \beta_{d,2} (xw)
+y_{i,q,t+1}^G &\sim \text{Normal}(\mu_{i,q,t+1}, \sigma^2_{x_{i,q,t+1}}\textbf{I}) \\
+\mu_{i,q,t+1} &= \beta_{0,t} + \beta_{s,t}x_{i,q,t} + \beta_Q + \textbf{z}'_t \boldsymbol{\beta}_c + \beta_{d,1} w_{i,q,t} + \beta_{d,2} (x_{i,q,t}w_{i,q,t})
 \end{align}
@@
 
-where $\mu(x,\textbf{z}_t,w)$ is log of predicted genet size at time *t* + 1, and all other parameters are as described for the survival regression.
-We capture non-constant error variance in growth by modeling the variance around the growth regression ($\sigma^2_{x,t}$) as a nonlinear function of predicted genet size:
+where $\mu_{i,q,t+1}$ is log of genet *i*s predicted size at time *t* + 1, and all other parameters are as described for the survival regression.
+We capture non-constant error variance in growth by modeling the variance in the growth regression ($\sigma^2_{x{i,q,t+1}}$) as a nonlinear function of predicted genet size:
 
@@
 \begin{align}
-\sigma^2_{x,t} = a \, \text{exp}[b \times \mu(x,\textbf{z}_t,w)]
+\sigma^2_{x_{i,q,t+1}} = a \, \text{exp}[b \times \mu_{i,q,t+1}]
 \end{align}
@@
 
-where $\mu(x,\textbf{z}_t,w)$ is log of predicted genet size predicted from the growth regression (Eq. 4), and *a* and *b* are constants.
+where $\mu_{i,q,t+1}$ is log of predicted genet size predicted from the growth regression (Eq. 4), and *a* and *b* are constants.
 
@@
 Therefore, we model recruitment at the quadrat level.
-We assume the number of individuals, $y^{R}$, recruiting at time $t+1$ in quadrat q follows a negative binomial distribution:
+We assume the number of individuals, $y^{R}_{q,t+1}$, recruiting at time $t+1$ in quadrat *q* follows a negative binomial distribution:
 
@@
 \begin{align}
-\lambda_{q,t+1} = C'_{q,t} \, \text{exp}\left(\beta_{0,t} + \beta_Q + \textbf{z}'_t \boldsymbol{\beta}_c + \beta_{d}\sqrt{C'_{q,t}}\right)
+\lambda_{q,t+1} = \tilde{c}_{q,t} \, \text{exp}\left(\beta_{0,t} + \beta_Q + \textbf{z}'_t \boldsymbol{\beta}_c + \beta_{d}\sqrt{\tilde{c}_{q,t}}\right)
 \end{align}
@@
 
-where $C'$ is effective cover ($\text{cm}^2$) of the focal species in quadrat $q$, and all other terms are as in the survival and growth regressions.
-Effective cover is a mixture of observed cover ($C$) in the focal quadrat ($q$) and the mean cover across the entire group ($\bar{C}$) of $Q$ quadrats in which $q$ is located:
+where $\tilde{c}_{q,t}$ is effective cover ($\text{cm}^2$) of the focal species in quadrat *q* at time *t*, and all other terms are as in the survival and growth regressions.
+Effective cover is a mixture of observed cover (*c*) in the focal quadrat (*q*) and the mean cover across the entire group ($\bar{c}$) of *Q* quadrats in which *q* is located:
 
@@
 \begin{align}
-C'_{q,t} = pC_{q,t} + (1-p)\bar{C}_{Q,t}
+\tilde{c}_{q,t} = pc_{q,t} + (1-p)\bar{c}_{Q,t}
 \end{align}
@@
 
-where $p$ is a mixing fraction between 0 and 1 that is estimated within the model.
+where $p$ is a mixing fraction between 0 and 1 that is estimated when fitting model.
 
@@
 The statistical approach used to model aggregated data depends on the type of data collected.
-We have percent cover data, which can easily be transformed to proportion data.
-An obvious choice for fitting a linear model to proportion data is beta regression because the support of the beta distribution is [0,1], not including true zeros or ones.
+We have percent cover data, which can easily be transformed to proportion data in our case because plant areas were scaled by plot area.
+An obvious choice for fitting a linear model to proportion data is beta regression because the support of the beta distribution is (0,1), which does not include true zeros or ones.
 However, when we used fitted model parameters from a beta regression in a quadrat-based population model, the simulated population tended toward 100% cover for all species.
@@
 \begin{align}
-\textbf{y}^{P} &\sim \text{LogNormal}(\mu(x,\textbf{z}_t), \sigma^2) \text{T}[0,1] \\
-\mu(x,\textbf{z}_t) &= \beta_{0,t} + \beta_{s,t}x + \beta_Q + \textbf{z}'_t \boldsymbol{\beta}_c
+y_{q,t+1}^{P} &\sim \text{LogNormal}(\mu_{q,t+1}, \sigma^2)_{0}^{1} \\
+\mu_{q,t+1} &= \beta_{0,t} + \beta_{s,t}x_{q,t} + \beta_Q + \textbf{z}'_t \boldsymbol{\beta}_c
 \end{align}
@@
 
-where $\mu(x,\textbf{z}_t)$ is the log of proportional cover in quadrat $q$ at time $t+1$, and all other parameters are as in the individual-level growth model (Eq.
+where $\mu_{q,t+1}$ is the log of proportional cover in quadrat *q* at time $t+1$, and all other parameters are as in the individual-level growth model (Eq.
 4) except that *x* now represents log of proportional cover.
-The log normal likelihood includes a truncation (T[0,1]) to ensure that predicted values do not exceed 100% cover.
+The log normal likelihood includes a truncation (subscript 0, superscript 1) to ensure that predicted values do not exceed 100% cover.
 
@@
 We decided to fit models where all terms were fit by species.
-Within a species, we fit yearly size effects and yearly intercepts hierarchically where year-specific coefficients were drawn from global distributions representing the mean size effect and intercept.
+Within a species, we fit yearly size effects and yearly intercepts hierarchically, where year-specific coefficients were drawn from global distributions representing the mean size effect and intercept.
 Quadrat random effects were also fit hierarchically, with quadrat offsets being drawn from distributions with mean zero and a shared variance term (independent Gaussian priors, Appendix A).
-Climate effects were modeled as independent covariates whose prior distributions were determined by statistical regularization (see **Statistical regularization: Bayesian ridge regression** below).
+Climate effects were modeled as independent covariates whose prior distributions were optimized for prediction using statistical regularization (see **Statistical regularization: Bayesian ridge regression** below).
 
@@
 We obtained posterior distributions for all model parameters from three parallel MCMC chains run for 1,000 iterations after discarding an initial 1,000 iterations.
-Such short MCMC chains may surprise readers more familiar with other MCMC samplers (i.e.
-JAGS or WinBUGS), but the Stan sampler is exceptionally efficient, which reduces the number of iterations needed to achieve convergence.
-We assessed convergence visually and made sure scale reduction factors for all parameters were less than 1.1.
+Such short MCMC chains may surprise readers more familiar with other MCMC samplers (i.e., JAGS or WinBUGS), but the Stan sampler reduces the number of iterations needed to achieve convergence.
+We assessed convergence visually and checked that scale reduction factors for all parameters were less than 1.1.
 For the purposes of including parameter uncertainty in our population models, we saved the final 1,000 iterations from each of the three MCMC chains to be used as randomly drawn values during population simulation.
-We report the posterior mean, standard deviation, and 95% Bayesian Credible Intervals for every parameter of each model for each species in Appendix B.
+We report the posterior mean, standard deviation, and 95% Bayesian Credible Intervals for every parameter of each model for each species in the Supporting Information (Tables S#-S#).
 
@@
 For models with climate covariates, our objective is to model the response of our focal grass species to interannual variation in climate, even if those responses are weak.
-Therefore, we avoid selecting among models with all possible combinations of climate covariates, and instead use Bayesian ridge regression to regulate, or constrain, the posterior distributions of each climate covariate [@Hooten2015].
+Therefore, we avoid selecting among models with all possible combinations of climate covariates, and instead use Bayesian ridge regression to regulate, or constrain, the posterior distributions of each climate covariate [@Gerber2015; @Hooten2015].
 Ridge regression is a specific application of statistical regularization that seeks to optimize model generality by trading off bias and variance.
@@
 We performed the grid search over 24 values of $\sigma_{\beta_c}^2$, ranging from $\sigma_{\beta_c}^2 = 0.01$ to $\sigma_{\beta_c}^2 = 2.25$.
-For each statistical model and each species we fit $13\times24=312$ models (13 years to leave out for cross-validation and 24 values of $\sigma_{\beta_c}^2$) -- a total of 4,992 models.
-We calculated the log pointwise predictive density (*lppd*) to score each model's ability to predict the left-out data.
+For each statistical model and each species we fit $13\times24=312$ iterations of the model fitting algorithm to search $\sigma_{\beta_c}^2$ for the optimal value (13 years to leave out for cross-validation and 24 values of $\sigma_{\beta_c}^2$) -- a total of 4,992 model fits.
+We calculated the log pointwise predictive density (*lppd*) to score each model's ability to predict the left-out data [@Gelman2014].
 Thus, for training data $y_{train}$ and held-out data $y_{hold}$ at a given value of  $\sigma_\theta^2$ across all MCMC samples $s=1,2,...,S$ and all hold outs of data from year *t* to year *T*, and letting $\theta$ represent all unknowns, *lppd* is
@@
 We chose the optimal prior variance for each species-statistical model combination as the one that produced the highest *lppd* and then fit each species-statistical model combination using the full data set for each species and the optimal prior variance.
+We calculated the *lppd* from posterior samples using the algorithm from @Vehtari2016.
 
 ### Population models
-With the posterior distribution of the vital rate statistical models in hand, it is straightforward to simulate the population models.
-We used an Integral Projection Model (IPM) to model populations based on individual-level data [@Ellner2006] and a quadrat-based version of an individually-based model (Quadrat-Based Model, QBM) to model populations based on quadrat-level data.
+Using samples from the posterior distribution of the vital rate statistical models, it is straightforward to simulate the population models.
+We used an Integral Projection Model (IPM) to simulate populations based on individual-level data [@Ellner2006] and a quadrat-based version of an individually-based model (Quadrat-Based Model, QBM) to simulate populations based on quadrat-level data.
 We describe each in turn.
@@
 We use a stochastic IPM [@Rees2009] that includes the climate covariates from the vital rate statistical models.
-In all simulations we ignore the random year effects so that interannual variation is driven solely by climate.
+In all simulations, we ignore the random year effects so that interannual variation is driven solely by climate.
 We fit the random year effects in the vital rate regressions to avoid over-attributing variation to climate covariates.
@@
 \begin{align}
-n(v_{j},t+1) = \int_{L_{j}}^{U_{j}} k_{j}(v_{j},u_{j},\bar{w_{j}}(u_{j}))n(u_{j},t)
+n(v_{j},t+1) = \int_{L_{j}}^{U_{j}} k_{j}(v_{j},u_{j},\bar{w_{j}}(u_{j}))n(u_{j},t)du_{j}
 \end{align}
@@
 
-Since the IPM is spatially-implicit, we cannot calculate neighborhood crowding for specific genets ($w_{ij}$).
+The IPM is spatially-implicit, thus we cannot calculate neighborhood crowding for specific genets ($w_{ij}$).
 Instead, we use an approximation ($\bar{w_{j}}$) that captures the essential features of neighborhood interactions [@Adler2010].
@@
 which means we are calculating growth (_G_) for individuals that survive (_S_) from time _t_ to _t+1_ and adding in newly recruited (_R_) individuals of an average sized one-year-old genet for the focal species.
-Our stastical model for recruitment (_R_, described above) returns the number of new recruit produced per quadrat.
+Our stastical model for recruitment (_R_, described above) returns the number of new recruits produced per quadrat.
 Following previous work [@Adler2012; @Chu2015], we assume that fecundity increases linearly with size ($R_j(v_{j},u_{j},\bar{w_{j}}) = e^{u_j}R_j(v_{j},\bar{w_{j}})$) to incorporate the recruitment function in the spatially-implicit IPM.
@@
 Relatively unimportant climate covariates (those that broadly overlap 0) will have little effect on the mean of the simulation results, but can contribute to their variation.
-Since our focus was on the contribution of density dependence and climate covariates to population states, we set the random year effects and the random group effects to zero.
+Because our focus was on the contribution of density dependence and climate covariates to population states, we set the random year effects and the random group effects to zero.
 
@@
 We use the same approach for drawing parameter values as described for the IPM.
-After drawing the appropriate parameter set, we calculate the mean response (log cover at *t+1* = $\mu_{t+1}$) according to Eq.
+After drawing the appropriate parameter set, we calculate the mean response (log cover at *t+1* is $\mu_{t+1}$) according to Eq.
 10.
-We then make a random draw from a [0,1] truncated lognormal distribution with mean equal to $\mu_{t+1}$ from Eq.
+We make a random draw from a [0,1] truncated lognormal distribution with mean equal to $\mu_{t+1}$ from Eq.
 10 and the variance estimate from the fitted model.
-We can then project the model forward by drawing a new parameter set (unique to climate year and MCMC iteration) at each timestep.
+We project the model forward by drawing a new parameter set (unique to climate year and MCMC iteration) at each timestep.
 As with the IPM, random year effects are ignored for all simulations.
@@
 We made predictions for each observed quadrat in each focal year, initializing each simulation with cover in the quadrat the previous year.
-Since we were making quadrat-specific predictions, we incorporated the group random effect on the intercept for both models.
+Because we were making quadrat-specific predictions, we incorporated the group random effect on the intercept for both models.
 We repeated this procedure for all 13 observation years, making 100 one-step-ahead forecasts for each quadrat-year combination with parameter uncertainty included via random draw from the MCMC chain as described above.
-Random year effects were set to zero since year effects cannot be assigned to unobserved years.
+Random year effects were set to zero because year effects cannot be assigned to unobserved years.
 
@@
 We compared $\rho$ and MAE between model types and within model types between models with and without climate covariates using one-sided *t* tests with adjusted degrees of freedom following @Wilcox2009 and standard errors calculated using the HC4 estimator of @Cribari-Neto2004.
-Statistical tests for differences between $\rho$ were conducted using R functions from @Ye2015a.
+Statistical tests were conducted using algorithms from @Ye2015a.
 
@@
 Thus, in our current example, the model forecast for percent cover in 1941 has a forecast horizon of one year, the forecast in 1942 has a forecast horizon of two years, and so on.
-We ran these simulations for all model types (IPM with/without climate; QBM with/without climate) using mean parameter values for all possible initial years.
-Then, for a given forecast horizon, we averaged the correlation between forecasts and observations.
-Note that these forecasts are all made using in-sample data since we used model fits from the full data set.
-Nonetheless, these simulations offer insight into the differences between model forecast horizons.
+We performed these simulations for all model types (IPM with/without climate; QBM with/without climate) using mean parameter values for all possible initial years.
+For a given forecast horizon, we averaged the correlation between forecasts and observations.
+Note that these forecasts are all made using in-sample data because we used model fits from the full data set.
+Nonetheless, these simulations offer insight into the differences among model forecast horizons.
 
@@
 Increases in accuracy due to the inclusion of climate covariates were not significant (*P* > 0.05 for all comparisons of $\rho$ between climate and no-climate forecasts within model types; Fig. 2).
-However, the IPM with climate covariates for *P. secunda* did have significantly lower error (mean absolute error) than the IPM without climate covariates ($t_{(196)}$ = -1.84, *P* = 0.033; Fig. S2).
+However, the IPM with climate covariates for *Poa secunda* did have significantly lower error (mean absolute error) than the IPM without climate covariates ($t_{(196)}$ = -1.84, *P* = 0.033; Fig. S2).
 In only one case were IPM forecasts significantly more accurate than the QBM (Fig.
-2): forecast accuracy of *P.
-secunda* percent cover from an IPM with climate covariates was greater than the accuracy from the QBM with climate covariates ($t_{(195)}$ = 1.72, *P* = 0.043).
+2): forecast accuracy of *Poa secunda* percent cover from an IPM with climate covariates was greater than the accuracy from the QBM with climate covariates ($t_{(195)}$ = 1.72, *P* = 0.043).
 Forecasts from IPMs with climate covariates had significantly lower error than equivalent QBM forecasts for *P.
-smithii* ($t_{(216)}$ = -3.49, *P* < 0.001) and *P.
-secunda* ($t_{(196)}$ = -1.83, *P* = 0.034) (Fig.
+smithii* ($t_{(216)}$ = -3.49, *P* < 0.001) and *Poa secunda* ($t_{(196)}$ = -1.83, *P* = 0.034) (Fig.
 S2).
@@
 The accuracy of both model's forecasts declined as the forecast horizon increased, but they did so at similar rates (Fig. 4).
-The only exception is for *P. secunda* where forecast accuracy appears to remain steady as the forecast horizon increases, after an initial decrease from a forecast horizon of one year (Fig. 4).
+The only exception is for *Poa secunda* where forecast accuracy appears to remain steady as the forecast horizon increases, after an initial decrease from a forecast horizon of one year (Fig. 4).
 
@@
 2; `no-climate' bars).
-We expected the IPM to out-perform the QBM since the IPM includes more mechanistic detail on the perennial plant life cycle, but this was not the case, at least when we ignored environmental forcing.
+We expected the IPM to out-perform the QBM because the IPM includes more mechanistic detail on the perennial plant life cycle, but this was not the case, at least when we ignored environmental forcing.
 Such a finding confirms theoretical [@Freckleton2011] and empirical work [@Taylor2004; @Queenborough2011] showing that density-structured models can be useful surrogates for demographic models when the goal is to estimate or forecast population states over large spatial extents.
@@
 Thus, estimates of climate effects on plant population growth may be biased or non-identifiable when the underlying statistical model is fit using population-level data that integrates over the potentially unique climate responses of individual vital rates.
-However, we found that IPM forecast accuracy was only significantly higher than the QBM for one species, *P. secunda* (Fig. 2, `climate' bars).
+However, we found that IPM forecast accuracy was only significantly higher than the QBM for one species, *Poa secunda* (Fig. 2, `climate' bars).
 This result likely stems from the relatively weak climate effects in the vital rate regressions (Fig. 3).
-If climate effects on vital rates are weak, then models based on percent cover or density should fair well compared to models based on individual-level data so long as density-dependence is adequately estimated (see Fig.
+If climate effects on vital rates are weak, then models based on percent cover or density should fair well compared to models based on individual-level data as long as density dependence is adequately estimated (see Fig.
 2 and Freckleton et al.
@@
 
-Our naive expectation was that a single, relatively important climate effect in a vital rate regression would cause a discernable difference between IPM and QBM forecast skill because individual-level responses may be less identifiable from percent cover data [@Clark2012].
+Our naive expectation was that, a single, relatively important climate effect in a vital rate regression would cause a discernable difference between IPM and QBM forecast skill because individual-level responses may be less identifiable from percent cover data [@Clark2012].
 We actually found the opposite.
@@
 comata*, and *P.
-smithii* all included at least one relatively important climate effect, which we *a posterior* defined as a standardized coefficient whose 80% credible interval does not overlap zero (Fig.
+smithii* all included at least one relatively important climate effect, which we *a posteriori* defined as a standardized coefficient whose 80% credible interval does not overlap zero (Fig.
 3).
-Yet, only the forecast skill for *P. secunda* differed between the IPM and QBM (Fig. 2).
-Our explanation is that while at least one vital regression for the other three species included a relatively strong climate effect, only *P.
-secunda* had a vital rate regression with several smaller climate effects all trending in the same direction (Fig.
+Yet, only the forecast skill for *Poa secunda* differed between the IPM and QBM (Fig. 2).
+Our explanation is that while at least one vital regression for the other three species included a relatively strong climate effect, only *Poa secunda* had a vital rate regression with several smaller climate effects all trending in the same direction (Fig.
 3).
@@
 
-The higher accuracy of the IPM with climate covariates for *P.
-secunda*  highlights the advantage of contemporary model and variable selection approaches such as ridge regression and LASSO over techniques that would exclude "non-significant" effects from final models.
+The higher accuracy of the IPM with climate covariates for *Poa secunda*  highlights the advantage of contemporary model and variable selection approaches such as ridge regression and LASSO over techniques that would exclude "non-significant" effects from final models.
 Ridge regression allows researchers to retain covariates whose effects may be difficult to identify in noisy data or short time series.
 This is especially important when forecasting the impacts of climate change where it is important to include to effects of forcing variables (e.g., temperature and precipitation) even of such effects are difficult to identify.
-If a species is truly unresponsive to a given climate variable, statistical regularization techniques will shrink the mean and variance of a covariate estimate to zero [@Hooten2015].
+If a species is truly unresponsive to a given climate variable, statistical regularization techniques will shrink the mean and variance of a covariate estimate toward zero [@Hooten2015].
 Of course, no matter what model selection approach is adopted, a critical step is identifying the appropriate candidate covariates, which we attempted to do based on our knowledge of this semi-arid plant community.
@@
 In principle, more mechanistic models should produce better predictions, especially under novel conditions [@Evans2012b; @Schindler2015b].
-In our case, the IPM explicitly models the influence of weather on recruitment and survival, effects that may be poorly represented in the QBM since recruitment and survival mainly affect small plants that contribute little to year-to-year changes in percent cover.
+In our case, the IPM explicitly models the influence of weather on recruitment and survival, effects that may be poorly represented in the QBM because recruitment and survival mainly affect small plants that contribute little to year-to-year changes in percent cover.
 Over time, of course, the addition and subtraction of small plants can have large effects on population growth, so explicitly modeling these effects could contribute to a longer forecast horizon. 
@@
 2).
-In fact, only the forecast horizons for *P. secunda* noticeably differ (Fig. 4), and that is also the only species for which the IPM outperformed the QBM when we included climate effects.
+In fact, only the forecast horizons for *Poa secunda* noticeably differ (Fig. 4), and that is also the only species for which the IPM outperformed the QBM when we included climate effects.
 
@@
 Compute, storage and other resources from the Division of Research Computing in the Office of Research and Graduate Studies at Utah State University are gratefully acknowledged.
+Any use of trade, firm, or product names is for descriptive purposes only and does not imply endorsement by the U.S.
+government.
 
@@
        & Percent cover & 217 & 19 \\ 
-  \rule{0pt}{3ex} \emph{P. secunda} & Growth & 3018 & 18 \\ 
+  \rule{0pt}{3ex} \emph{Poa secunda} & Growth & 3018 & 18 \\ 
        & Survival & 4650 & 18 \\ 
@@
 Comparisons between equivalent IPM and QBM models indicate no significant difference in accuracy (\emph{P} > 0.05 for all comparisons).
-Likewise, including climate covariates did not result in siginificantly higher forecast accuracy (\emph{P} > 0.05 for all comparisons).}
+Including climate covariates resulted in siginificantly higher forecast accuracy for only one species, \emph{Poa secunda} (\emph{P} = 0.043).}
 \end{figure}
@@
   \caption{Posterior distributions of climate effects ($\boldsymbol{\beta}_C$) for each species and vital rate statistical model.
-Since our priors were constrained via ridge-regression, we highlight climate effects whose 80\% credible intervals do not overlap zero (red for negative coefficients, blue for positive coefficients).
+Because our priors were constrained via ridge-regression, we highlight climate effects whose 80\% credible intervals do not overlap zero (red for negative coefficients, blue for positive coefficients).
 Kernel bandwidths of posterior densities were adjusted by a factor of 4 for visual clarity.}

### 2016-05-25 manuscript/cached_docs/appA.Rmd
updates SI material for J. Ecol version

--- 2015-07-10 manuscript/appA.Rmd
@@
 
+\renewcommand\thefigure{A\arabic{figure}}
+
 Appendix A: Detailed model descriptions
@@
 We do this for clarity of model presentation.
-For each model we include a table of prior distributions for necessary parameters.
-Also for clarity, we avoid subscripting species (the *j* subscripts) after initial description of each model.
+All prior distributions are defined in the parameter model sections.
+For clarity we avoid subscripting species (the *j* subscripts) after initial description of each model.
 
@@
 ------------
-We modeled growth as Gaussian process describing genet size at time $t+1$ as a function of size at $t$ and climate covariates:
+We modeled growth as a Gaussian process describing genet size at time $t+1$ as a function of size at $t$ and climate covariates:
 
 \begin{align}
-x_{ijQ,t+1} &= \gamma^{G}_{j,t} + \phi^{G}_{jQ} + \beta^{G}_{j,t}x_{ij,t} + \omega^{G}_{j}w_{ij,t} + \nu^{G}_{j}w_{ij,t}x_{ij,t} + \theta^{G}_{jk}C_{k,t} \\
-Y^{G}_{ijQ,t} &\sim \text{Normal}(x_{ijQ,t+1}, \sigma_{j})
+x_{ijQ,t+1} &= \gamma^{G}_{j,t} + \phi^{G}_{jQ} + \beta^{G}_{j,t}x_{ij,t} + \omega^{G}_{j}w_{ij,t} + \nu^{S}_{j}w_{ij,t}x_{ij,t} + \theta^{G}_{jk}C_{k,t} \\
+y^{G}_{ijQ,t} &\sim \text{Normal}(x_{ijQ,t+1}, \varepsilon_{ij,t})
 \end{align}
@@
 
-We assume the number of individuals, $Y^{R}$, recruiting at time $t+1$ follows a negative binomial distribution:
+We assume the number of individuals, $y^{R}$, recruiting at time $t+1$ follows a negative binomial distribution:
 
 \begin{equation}
-Y^{R}_{jq,t+1} \sim \text{NegBin}(\lambda_{jq,t+1},\zeta)
+y^{R}_{jq,t+1} \sim \text{NegBin}(\lambda_{jq,t+1},\zeta)
 \end{equation}
@@
 -------------------------
-The model for quadrat cover ($\mu$) change from time $t$ to $t+1$ is
+The model for quadrat cover change from time $t$ to $t+1$ is
 
 \begin{align}
-\mu_{jq,t+1} &= \gamma_{j,t} + \phi_{jQ} + \beta_{j,t}x_{jq,t} + \theta_{jk}C_{k,t} \\
-Y_{jq,t+1} &\sim \text{LogNormal}(\mu_{jq,t+1}, \tau{j}) \text{T}[0,1]
+x_{jq,t+1} &= \gamma^{P}_{j,t} + \phi^{P}_{jQ} + \beta^{P}_{j,t}x_{jq,t} + \theta^{P}_{jk}C_{k,t} \\
+y^{P}_{jq,t+1} &\sim \text{LogNormal}(x_{jq,t+1}, \tau_{j}) \text{T}[0,1]
 \end{align}

### 2016-05-25 manuscript/cached_docs/appC.Rmd
updates SI material for J. Ecol version

--- 2015-08-22 manuscript/appC.Rmd
@@
 
-```{r plot_it, echo=FALSE, message=FALSE, warning=FALSE, include=TRUE, fig.width=5, fig.height=3, fig.cap="Equilibrium cover and 90% quantiles around the mean prediction from the IPM when model error and parameter uncertainty are propogated through the simulation phase for all vital rates except recruitment.", dependson="plot-options"}
+```{r plot_it, echo=FALSE, message=FALSE, warning=FALSE, include=TRUE, fig.width=5, fig.height=3, fig.cap="Mean (points) and 90% quantiles (errorbars) for the proportional difference between baseline simulations (using observed climate) and the climate pertubation simulation on the x-axis.
+We calculated proportional difference as log(perturbed climate cover) - log(observed climate cover), where 'perturbed' and 'observed' refer to the climate time series used to drive interannual variation in the simulations.
+Model error and parameter uncertainty were propagated through the simulation phase for all vital rates except recruitment.", dependson="plot-options"}
 library(plyr)

### 2016-05-25 manuscript/cached_docs/appD.Rmd
updates SI material for J. Ecol version

--- 2015-07-17 manuscript/appD.Rmd
@@
 Simulating populations in such a way ignores variation in the parameters of $f(N_t)$.
-When we take this approach to making forecasts of equilibrium cover under 1% climate changes the IPM and QBM make opposing predictions if we simply look at the mean projection (Fig.
+When we take this approach to making forecasts of equilibrium cover under 1% climate perturbations the IPM and QBM make opposing predictions if we simply look at the mean projection (Fig.
 D1).

### 2016-05-25 manuscript/manuscript.Rmd
updates SI material for J. Ecol version

@@
 smithii*, and *Poa secunda* are cool-season perennial grasses.
-All species typically begin growth in the early spring, reach maximum growth and flower in early to mid summer (May-June), and disperse seed in mid to late summer (July-September).
-[LAST SENTENCES GLOSSES OVER LIKELY BIG DIFFERENCES IN PHENOLOGY.
-JUST DESCRIBE TIMING OF GROWING SEASON INSTEAD, TO SET UP CLIMATE COVARIATES?]
+The growing season begins in early spring (typically in April) and lasts through mid-summer (typically in June).
 
@@
 Within a species, we fit yearly size effects and yearly intercepts hierarchically, where year-specific coefficients were drawn from global distributions representing the mean size effect and intercept.
-Quadrat random effects were also fit hierarchically, with quadrat offsets being drawn from distributions with mean zero and a shared variance term (independent Gaussian priors, Appendix A).
+Quadrat random effects were also fit hierarchically, with quadrat offsets being drawn from distributions with mean zero and a shared variance term (independent Gaussian priors).
 Climate effects were modeled as independent covariates whose prior distributions were optimized for prediction using statistical regularization (see **Statistical regularization: Bayesian ridge regression** below).
@@
 For the purposes of including parameter uncertainty in our population models, we saved the final 1,000 iterations from each of the three MCMC chains to be used as randomly drawn values during population simulation.
-We report the posterior mean, standard deviation, and 95% Bayesian Credible Intervals for every parameter of each model for each species in the Supporting Information (Tables S#-S#).
+We report the posterior mean, standard deviation, and 95% Bayesian Credible Intervals for every parameter of each model for each species in the Supporting Information (Tables S5-S20).
 
@@
 2) and overall error was similar (mean absolute error; Fig.
-S1).
-With climate covariates, the best out-of-sample predictive model (highest *lppd*) for each species and vital rate typically resulted from highly constrained priors on the climate effects (Fig. S2).
+S1, Supporting Information).
+With climate covariates, the best out-of-sample predictive model (highest *lppd*) for each species and vital rate typically resulted from highly constrained priors on the climate effects (Fig.
+S2, Supporting Information).
 Thus, the posterior distributions of climate effects included in our models overlapped zero and generally shrunk toward zero, though for some species-vital rate combinations important effects (80% credible interval does not include zero) did emerge (Fig.
@@
 smithii* ($t_{(216)}$ = -3.49, *P* < 0.001) and *Poa secunda* ($t_{(196)}$ = -1.83, *P* = 0.034) (Fig.
-S2).
-Results from all pairwise statistical tests are shown in Table S1.
+S2, Supporting Information).
+Results from all pairwise statistical tests are shown in Table S21 of the Supporting Information.
 

### 2016-05-25 manuscript/supporting_information.Rmd
updates SI material for J. Ecol version

---
layout: 12pt
header-includes:
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{setspace}
   - \usepackage{graphicx}
bibliography: ~/Dropbox/Bibliography/MicroMeso.bib
csl: components/ecology.csl

## rmarkdown render options
output:
  pdf_document:
    fig_caption: true
    keep_tex: false
fontsize: 12pt
geometry: margin=1in
linkcolor: black
urlcolor: black
---

\renewcommand\thefigure{S\arabic{figure}}  
\renewcommand\thetable{S\arabic{table}}

# Supporting Information

\tableofcontents
\listoffigures
\listoftables
\newpage{}

##  Hierarchical Bayesian models for vital rates
Here we describe the statistical models in complete detail, including full description of the Bayesian models and Stan code for each model. We write each model as three parts: (1) a data model (likelihood), (2) a process model, and (3) a parameter model. We then combine these three models to write the posterior and joint distributions as a single model statement. Following the full model expression we provide the Stan code that directly corresponds the statement of posterior and joint distributions. In our expressions of the posterior and joint distributions we use general probability notation where [.] represents some unspecified distribution. We do this for clarity of model presentation. All prior distributions are defined in the parameter model sections. For clarity we avoid subscripting species.

### Parameter to model notation keys

Table:  Definition of parameters and model notation for IPM survival and growth models.

| Parameter | Definition | Notation in statistical results tables |
| --------- | ---------- | -------------------------------------- |
| $\beta_{0,t}$ | intercept for year *t* | ``a[t]`` where ``t`` is the numeric year |
| $\tilde{\beta_0}$ | global intercept | ``a_mu`` |
| $\beta_{Q}$ | random effect of quadrat group *Q* | ``gint[Q]`` where ``Q`` is the quadrat group |
| $\beta_{s,t}$ | size effect for year *t* | ``b1[t]`` where ``t`` is the numeric year |
| $\tilde{\beta_s}$ | global size effect | ``b1_mu`` |
| $\beta_{d,1}$ | effect of crowding | ``w[1]`` |
| $\beta_{d,2}$ | crowding $\times$ size effect | ``w[2]`` |
| $\beta_{c,k}$ | effect of climate covariate *k* | ``b2[k]`` where ``k`` is the climate effect |
| *a* | first parameter for growth variance | ``tau`` |
| *b* | second parameter for growth variance | ``tauSize`` |

Table:  Definition of parameters and model notation for IPM recruitment model.

| Parameter | Definition | Notation in statistical results tables |
| --------- | ---------- | -------------------------------------- |
| $\beta_{0,t}$ | intercept for year *t* | ``a[t]`` where ``t`` is the numeric year |
| $\tilde{\beta_0}$ | global intercept | ``a_mu`` |
| $\beta_{Q}$ | random effect of quadrat group *Q* | ``gint[Q]`` where ``Q`` is the quadrat group |
| $\beta_{d,1}$ | effect of plot cover | ``dd`` |
| $\beta_{c,k}$ | effect of climate covariate *k* | ``b2[k]`` where ``k`` is the climate effect |
| *p* | mixing fraction for effective plot cover | ``u`` |
| $\zeta$ | size parameter for negative binimial likelihood | ``theta`` |

Table:  Definition of parameters and model notation for quadrat based model.

| Parameter | Definition | Notation in statistical results tables |
| --------- | ---------- | -------------------------------------- |
| $\beta_{0,t}$ | intercept for year *t* | ``a[t]`` where ``t`` is the numeric year |
| $\tilde{\beta_0}$ | global intercept | ``a_mu`` |
| $\beta_{Q}$ | random effect of quadrat group *Q* | ``gint[Q]`` where ``Q`` is the quadrat group |
| $\beta_{s,t}$ | cover effect for year *t* | ``b1[t]`` where ``t`` is the numeric year |
| $\tilde{\beta_s}$ | global cover effect | ``b1_mu`` |
| $\beta_{c,k}$ | effect of climate covariate *k* | ``b2[k]`` where ``k`` is the climate effect |
| $\tau$ | model variance in log normal likelihood | ``tau`` |

Table: Climate effect key.

| Integer ID (*k*) | Climate covariate |
| ---------------- | ----------------- |
| 1 | pptLag |
| 2 | ppt1 |
| 3 | ppt2 |
| 4 | TmeanSpr1 |
| 5 | TmeanSpr2 |
| 6 | ppt1 $\times$ TmeanSpr1 |
| 7 | ppt2 $\times$ TmeanSpr2 |

### Survival (IPM)
We used logistic regression to model the probability that genet *i* in quadrat *q* survives from time *t* to *t*+1 ($s_{i,q,t}$):

\begin{align}
y_{i,q,t}^{S} &\sim \text{Bernoulli}(s_{i,q,t}) \\
\text{logit}(s_{i,q,t}) &= \beta_{0,t} + \beta_{s,t}x_{i,q,t} + \beta_Q + \textbf{z}'_t \boldsymbol{\beta}_c + \beta_{d,1} w_{i,t} + \beta_{d,2} (x_{i,q,t}w_{i,q,t})
\end{align}

where $x_{i,q,t}$ is the log of genet *i* basal area at time *t*, $\beta_{0,t}$ is a year specific intercept, $\beta_Q$ is the random effect of quadrat group location, $\beta_{s,t}$ is the year-specific slope parameter for size, **z** is a vector of *p* climate covariates specific to year *t*, $\boldsymbol{\beta}_c$ is a vector of fixed climate effects of length *p*, $\beta_{d,1}$ is the effect of intraspecific crowding experienced by the focal genet at time *t* ($w_{i,q,t}$), and $\beta_{d,2}$ is a size by crowding ($x_{i,q,t}w_{i,q,t}$) interaction effect.

**Data, process, and parameter models**

\begin{align}
&\mathtt{data} \qquad &y_{i,q,t}^{S} \sim \text{Bernoulli}(s_{i,q,t}) \\
&\mathtt{process} \qquad &\text{logit}(s_{i,q,t}) = \beta_{0,t} + \beta_{s,t}x_{i,q,t} + \beta_Q + \textbf{z}'_t \boldsymbol{\beta}_c + \beta_{d,1} w_{i,t} + \beta_{d,2} (x_{i,q,t}w_{i,q,t}) \\
&\mathtt{parameters} \qquad &\beta_{0,t} \sim \text{Normal}(\tilde{\beta_0}, \sigma_{\beta_0} \textbf{I})\\
& \qquad & \beta_{s,t} \sim \text{Normal}(\tilde{\beta_s}, \sigma_{\beta_s} \textbf{I}) \\
& \qquad & \boldsymbol{\beta_{Q}} \sim \text{Normal}(0, \sigma_Q \textbf{I}) \\
& \qquad & \boldsymbol{\beta}_c \sim \text{Normal}(0, \sigma_c \textbf{I}) \\
& \qquad & \beta_{d,1} \sim \text{Normal}(0, 100) \\
& \qquad & \beta_{d,2} \sim \text{Normal}(0, 100) \\
& \qquad & \tilde{\beta_0} \sim \text{Normal}(0, 100) \\
& \qquad & \tilde{\beta_s} \sim \text{Normal}(0, 100) \\
& \qquad & \sigma_Q \sim \text{Cauchy}(0, 5) \\
& \qquad & \sigma_{\beta_0} \sim \text{Cauchy}(0, 5) \\
& \qquad & \sigma_{\beta_s} \sim \text{Cauchy}(0, 5)
\end{align}

**Full expression of posterior and joint distributions**
\begin{align}
[\boldsymbol{\beta_0}, \tilde{\beta_0}, \boldsymbol{\beta_s}, \tilde{\beta_s}, \boldsymbol{\beta_Q}, \boldsymbol{\beta_c}, \beta_{d,1}, \beta_{d,2}, \sigma_Q, \sigma_{\beta_0}, \sigma_{\beta_s}] &\propto \\
&\prod_{t=1}^T \prod_{i=1}^n [y_{iQ,t} | \beta_{0,t}, \beta_{s,t}, \beta_{Q}, \boldsymbol{\beta_c}, \beta_{d,1}, \beta_{d,2}] \times \\
&[\beta_{0,t} | \tilde{\beta_0}, \sigma_{\beta_0}] \times \\
&[\beta_{s,t} | \tilde{\beta_s}, \sigma_{\beta_s}] [\beta_c|\sigma_{beta_c}] \times \\
&\prod_{Q=1}^{Q_{tot}} [\beta_{Q} | \sigma_Q] \times \\
&[\beta_{d,1}] [\beta_{d,1}] [\tilde{\beta_0}] [\tilde{\beta_s}] [\sigma_Q] [\sigma_{\beta_0}] [\sigma_{\beta_s}]
\end{align}

**Stan code for model**
```{r survival_stan, echo=TRUE, eval=FALSE}
data{
  // All Data
  int<lower=0> N; // observations
  int<lower=0> Yrs; // years
  int<lower=0> yid[N]; // year id
  int<lower=0> Covs; // climate covariates
  int<lower=0> G; // groups
  int<lower=0> gid[N]; // group id
  int<lower=0,upper=1> Y[N]; // observation vector
  matrix[N,Covs] C; // climate matrix
  vector[N] X; // size vector
  matrix[N,2] W; // crowding matrix
  real beta_tau; // prior sdev for climate effects
}
parameters{
  real a_mu;
  vector[Yrs] a;
  real b1_mu;
  vector[Yrs] b1;
  vector[Covs] b2;
  vector[2] w;
  vector[G] gint;
  real<lower=0> sig_a;
  real<lower=0> sig_b1;
  real<lower=0> sig_G;
}
transformed parameters{
  real mu[N];
  vector[N] climEff;
  vector[N] crowdEff;
  climEff <- C*b2;
  crowdEff <- W*w;
  for(n in 1:N){
    mu[n] <- inv_logit(a[yid[n]] + gint[gid[n]] + 
                         b1[yid[n]]*X[n] + crowdEff[n] + climEff[n]);
  }
}
model{
  // Priors
  a_mu ~ normal(0,100);
  w ~ normal(0,100);
  b1_mu ~ normal(0,100);
  sig_a ~ cauchy(0,5);
  sig_b1 ~ cauchy(0,5);
  sig_G ~ cauchy(0,5);
  gint ~ normal(0, sig_G);
  b2 ~ normal(0, beta_tau);
  a ~ normal(a_mu, sig_a);
  b1 ~ normal(b1_mu, sig_b1);

  // Likelihood
  Y ~ binomial(1,mu);
}
}
```

### Growth (IPM)
We modeled growth as a Gaussian process describing log genet size ($y_{i,q,t+1}^{G}$) at time $t+1$ in quadrat *q* as a function of log size at time $t$ and climate covariates:

\vspace{-3em}
\begin{align}
y_{i,q,t+1}^G &\sim \text{Normal}(\mu_{i,q,t+1}, \sigma^2_{x_{i,q,t+1}}\textbf{I}) \\
\mu_{i,q,t+1} &= \beta_{0,t} + \beta_{s,t}x_{i,q,t} + \beta_Q + \textbf{z}'_t \boldsymbol{\beta}_c + \beta_{d,1} w_{i,q,t} + \beta_{d,2} (x_{i,q,t}w_{i,q,t})
\end{align}
\vspace{-3em}

where $\mu_{i,q,t+1}$ is log of genet *i*s predicted size at time *t* + 1, and all other parameters are as described for the survival regression. We capture non-constant error variance in growth by modeling the variance in the growth regression ($\sigma^2_{x{i,q,t+1}}$) as a nonlinear function of predicted genet size:

\vspace{-3em}
\begin{align}
\sigma^2_{x_{i,q,t+1}} = a \, \text{exp}[b \times \mu_{i,q,t+1}]
\end{align}
\vspace{-3em}

where $\mu_{i,q,t+1}$ is log of predicted genet size predicted from the growth regression (Eq. 4), and *a* and *b* are constants.

**Data, process, and parameter models.**
\begin{align}
&\mathtt{data} \qquad &y_{i,q,t+1}^G \sim \text{Normal}(\mu_{i,q,t+1}, \sigma^2_{x_{i,q,t+1}}\textbf{I}) \\
&\mathtt{process} \qquad &\mu_{i,q,t+1} = \beta_{0,t} + \beta_{s,t}x_{i,q,t} + \beta_Q + \textbf{z}'_t \boldsymbol{\beta}_c + \beta_{d,1} w_{i,q,t} + \beta_{d,2} (x_{i,q,t}w_{i,q,t}) \\
& \qquad & \sigma^2_{iQ,t} = ae^{b\mu_{iQ,t+1}} \\
&\mathtt{parameters} \qquad &\beta_{0,t} \sim \text{Normal}(\tilde{\beta_0}, \sigma_{\beta_0} \textbf{I})\\
& \qquad & \beta_{s,t} \sim \text{Normal}(\tilde{\beta_s}, \sigma_{\beta_s} \textbf{I}) \\
& \qquad & \boldsymbol{\beta_{Q}} \sim \text{Normal}(0, \sigma_Q \textbf{I}) \\
& \qquad & \boldsymbol{\beta}_c \sim \text{Normal}(0, \sigma_c \textbf{I}) \\
& \qquad & \beta_{d,1} \sim \text{Normal}(0, 10) \\
& \qquad & \beta_{d,2} \sim \text{Normal}(0, 10) \\
& \qquad & \tilde{\beta_0} \sim \text{Normal}(0, 100) \\
& \qquad & \tilde{\beta_s} \sim \text{Normal}(0, 100) \\
& \qquad & \sigma_Q \sim \text{Cauchy}(0, 2) \\
& \qquad & \sigma_{\beta_0} \sim \text{Cauchy}(0, 2) \\
& \qquad & \sigma_{\beta_s} \sim \text{Cauchy}(0, 2) \\
& \qquad & a \sim \text{Normal}(0, 100) \\
& \qquad & b \sim \text{Normal}(0, 100)
\end{align}

**Full expression of posterior and joint distributions**
\begin{align}
[\boldsymbol{\beta_0}, \tilde{\beta_0}, \boldsymbol{\beta_s}, \tilde{\beta_s}, \boldsymbol{\beta_Q}, \boldsymbol{\beta_c}, \beta_{d,1}, \beta_{d,2}, \sigma_Q, \sigma_{\beta_0}, \sigma_{\beta_s},a,b] &\propto \\
&\prod_{t=1}^T \prod_{i=1}^n [y_{iQ,t} | \beta_{0,t}, \beta_{s,t}, \beta_{Q}, \boldsymbol{\beta_c}, \beta_{d,1}, \beta_{d,2}, a, b] \times \\
&[\beta_{0,t} | \tilde{\beta_0}, \sigma_{\beta_0}] \times \\
&[\beta_{s,t} | \tilde{\beta_s}, \sigma_{\beta_s}] [\beta_c|\sigma_{beta_c}] \times \\
&\prod_{Q=1}^{Q_{tot}} [\beta_{Q} | \sigma_Q] \times \\
&[\beta_{d,1}] [\beta_{d,1}] [\tilde{\beta_0}] [\tilde{\beta_s}] [\sigma_Q] [\sigma_{\beta_0}] [\sigma_{\beta_s}] [a] [b]
\end{align}

**Stan code for model**
```{r grow_stan, echo=TRUE, eval=FALSE}
data{
  int<lower=0> N; // observations
  int<lower=0> Yrs; // years
  int<lower=0> yid[N]; // year id
  int<lower=0> Covs; // climate covariates
  real<lower=0> tau_beta; // prior sdev for climate effects
  int<lower=0> G; // groups
  int<lower=0> gid[N]; // group id
  vector[N] Y; // observation vector
  matrix[N,Covs] C; // climate matrix
  vector[N] X; // size vector
  matrix[N,2] W; // crowding matrix
}
parameters{
  real a_mu;
  vector[Yrs] a;
  real b1_mu;
  vector[Yrs] b1;
  vector[Covs] b2;
  vector[2] w;
  real gint[G];
  real tau;
  real tauSize;
  real<lower=0> sig_a;
  real<lower=0> sig_b1;
  real<lower=0> sig_G;
}
transformed parameters{
  vector[N] mu;
  real<lower=0> sigma[N];
  vector[N] climEff;
  vector[N] crowdEff;
  climEff <- C*b2;
  crowdEff <- W*w;
  for(n in 1:N){
    mu[n] <- a[yid[n]] + gint[gid[n]] + b1[yid[n]]*X[n] + crowdEff[n] + climEff[n];
    sigma[n] <- sqrt((fmax(tau*exp(tauSize*mu[n]), 0.0000001)));  
  }
}
model{
  // Priors
  a_mu ~ normal(0,100);
  w ~ normal(0,10);
  b1_mu ~ normal(0,100);
  tau ~ normal(0,100);
  tauSize ~ normal(0,100);
  sig_a ~ cauchy(0,2);
  sig_b1 ~ cauchy(0,2);
  sig_G ~ cauchy(0,2);
  b2 ~ normal(0, tau_beta);
  for(g in 1:G)
    gint[g] ~ normal(0, sig_G);
  for(y in 1:Yrs){
    a[y] ~ normal(a_mu, sig_a);
    b1[y] ~ normal(b1_mu, sig_b1);
  }

  // Likelihood
  Y ~ normal(mu, sigma);
}
```

### Recruitment (IPM)
Our data allows us to track new recruits, but we cannot assign a specific parent to new genets. 
Therefore, we model recruitment at the quadrat level.
We assume the number of individuals, $y^{R}_{q,t+1}$, recruiting at time $t+1$ in quadrat *q* follows a negative binomial distribution:

\begin{align}
y^{R}_{q,t+1} \sim \text{NegBin}(\lambda_{q,t+1},\phi)
\end{align}

where $\lambda$ is the mean intensity and $\phi$ is the size parameter. We define $\lambda$ as a function of quadrat composition and climate in the previous year:

\begin{align}
\lambda_{q,t+1} = \tilde{c}_{q,t} \, \text{exp}\left(\beta_{0,t} + \beta_Q + \textbf{z}'_t \boldsymbol{\beta}_c + \beta_{d}\sqrt{\tilde{c}_{q,t}}\right)
\end{align}

where $\tilde{c}_{q,t}$ is effective cover ($\text{cm}^2$) of the focal species in quadrat *q* at time *t*, and all other terms are as in the survival and growth regressions.
Effective cover is a mixture of observed cover (*c*) in the focal quadrat (*q*) and the mean cover across the entire group ($\bar{c}$) of *Q* quadrats in which *q* is located:

\begin{align}
\tilde{c}_{q,t} = pc_{q,t} + (1-p)\bar{c}_{Q,t}
\end{align}

where $p$ is a mixing fraction between 0 and 1 that is estimated when fitting model.

**Data, process, and parameter models**
\begin{align}
&\mathtt{data} \qquad &y^{R}_{q,t+1} \sim \text{NegBin}(\lambda_{q,t+1},\phi) \\
&\mathtt{process} \qquad &\lambda_{q,t+1} = \tilde{c}_{q,t} \, \text{exp}\left(\beta_{0,t} + \beta_Q + \textbf{z}'_t \boldsymbol{\beta}_c + \beta_{d}\sqrt{\tilde{c}_{q,t}}\right) \\
&\mathtt{parameters} \qquad &\beta_{0,t} \sim \text{Normal}(\tilde{\beta_0}, \sigma_{\beta_0} \textbf{I})\\
& \qquad & \beta_{Q} \sim \text{Normal}(0,\sigma_Q \textbf{I}) \\
& \qquad & \boldsymbol{\beta}_c \sim \text{Normal}(0, \sigma_c \textbf{I}) \\
& \qquad & \beta_{d} \sim \text{Uniform}(-100, 100) \\
& \qquad & \tilde{\beta_0} \sim \text{Normal}(0, 1000) \\
& \qquad & \sigma_Q \sim \text{Cauchy}(0, 5) \\
& \qquad & \sigma_{\beta_0} \sim \text{Cauchy}(0, 5) \\
& \qquad & \phi \sim \text{Uniform}(0, 100) \\
& \qquad & u \sim \text{Uniform}(0, 1)
\end{align}

**Full expression of posterior and joint distributions**
\begin{align}
[\boldsymbol{\beta_0}, \tilde{\beta}, \boldsymbol{\beta_Q}, \boldsymbol{\beta_c}, \beta_{d}, \sigma_Q, \sigma_{\beta_0}, \phi, u] &\propto \\
&\prod_{t=1}^T \prod_{i=1}^n [y_{q,t+1} | \beta_{0,t}, \beta_{Q}, \beta_d, \boldsymbol{\beta_c}, \phi, u] [\beta_{0,t} | \tilde{\beta}] \times \\
&\prod_{Q=1}^{Q_{tot}} [\beta_{Q} | \sigma_Q] [\boldsymbol{\beta_c}] [\beta_d] [\tilde{\beta_0}] [\sigma_Q] [\sigma_{\beta_0}] [\phi] [u]
\end{align}

**Stan code for model**
```{r rec_stan, echo=TRUE, eval=FALSE}
data{
  int<lower=0> N; // observations
  int<lower=0> Yrs; // years
  int<lower=0> yid[N]; // year id
  int<lower=0> Covs; // climate covariates
  int<lower=0> G; // groups
  int<lower=0> gid[N]; // group id
  int<lower=0> Y[N]; // observation vector
  matrix[N,Covs] C; // climate matrix
  vector[N] parents1; // crowding vector
  vector[N] parents2; // crowding vector
  real<lower=0> tau; // prior variance
}
parameters{
  real a_mu;
  vector[Yrs] a;
  vector[Covs] b2;
  real dd;
  real gint[G];
  real<lower=0> sig_a;
  real<lower=0> theta;
  real<lower=0> sig_G;
  real<lower=0, upper=1> u;
}
transformed parameters{
real mu[N];
vector[N] climEff;
vector[N] trueP1;
vector[N] trueP2;
vector[N] lambda;
vector[N] q;
climEff <- C*b2;
  for(n in 1:N){
    trueP1[n] <- parents1[n]*u + parents2[n]*(1-u);
    trueP2[n] <- sqrt(trueP1[n]);
    mu[n] <- exp(a[yid[n]] + gint[gid[n]] + dd*trueP2[n] + climEff[n]);
    lambda[n] <- trueP1[n]*mu[n];
    q[n] <- lambda[n]*theta;
  }
}
model{
// Priors
  u ~ uniform(0,1);
  theta ~ uniform(0,100);
  a_mu ~ normal(0,1000);
  dd ~ uniform(-100,100);
  sig_a ~ cauchy(0,5);
  sig_G ~ cauchy(0,5);
  for(g in 1:G)
   gint[g] ~ normal(0, sig_G);
  for(y in 1:Yrs){
    a[y] ~ normal(a_mu, sig_a);
  }
  for(j in 1:Covs)
    b2[j] ~ normal(0, tau);

// Likelihood
  Y ~ neg_binomial_2(q, theta);
}

```

### Quadrat based model (QBM)
The model for quadrat cover change from time $t$ to $t+1$ is

\begin{align}
y_{q,t+1}^{P} &\sim \text{LogNormal}(\mu_{q,t+1}, \sigma^2)_{0}^{1} \\
\mu_{q,t+1} &= \beta_{0,t} + \beta_{s,t}x_{q,t} + \beta_Q + \textbf{z}'_t \boldsymbol{\beta}_c
\end{align}

where $\mu_{q,t+1}$ is the log of proportional cover in quadrat *q* at time $t+1$, and all other parameters are as in the individual-level growth model (Eq. 4) except that *x* now represents log of proportional cover. The log normal likelihood includes a truncation (subscript 0, superscript 1) to ensure that predicted values do not exceed 100% cover.

**Data, process, and parameter models.**
\begin{align}
&\mathtt{data} \qquad &y_{q,t+1}^{P} \sim \text{LogNormal}(\mu_{q,t+1}, \sigma)_{0}^{1} \\
&\mathtt{process} \qquad &\mu_{q,t+1} = \beta_{0,t} + \beta_{s,t}x_{q,t} + \beta_Q + \textbf{z}'_t \boldsymbol{\beta}_c \\
&\mathtt{parameters} \qquad &\beta_{0,t} \sim \text{Normal}(\tilde{\beta_0}, \sigma_{\beta_0})\\
& \qquad & \beta_{s,t} \sim \text{Normal}(\tilde{\beta_s}, \sigma_{\beta_s} \textbf{I}) \\
& \qquad & \beta_{Q} \sim \text{Normal}(0,\sigma_Q \textbf{I}) \\
& \qquad & \boldsymbol{\beta}_c \sim \text{Normal}(0, \sigma_c \textbf{I}) \\
& \qquad & \tilde{\beta_0} \sim \text{Normal}(0, 1000) \\
& \qquad & \tilde{\beta_s} \sim \text{Normal}(0, 1000) \\
& \qquad & \sigma_Q \sim \text{Cauchy}(0, 5) \\
& \qquad & \sigma_{\beta_0} \sim \text{Cauchy}(0, 5) \\
& \qquad & \sigma_{\beta_s} \sim \text{Cauchy}(0, 5) \\
& \qquad & \sigma \sim \text{Cauchy}(0, 5)
\end{align}

**Full expression of posterior and joint distributions**
\begin{align}
[\boldsymbol{\beta_0}, \tilde{\beta_0}, \boldsymbol{\beta_s}, \tilde{\beta_s}, \boldsymbol{\beta_Q}, \boldsymbol{\beta_c}, \sigma_Q, \sigma_{\beta_0}, \sigma_{\beta_s}, \sigma] &\propto \\
&\prod_{t=1}^T \prod_{i=1}^n [y_{q,t+1} | \beta_{0,t}, \beta_{s,t}, \beta_{Q}, \boldsymbol{\beta_c}] \times \\
&[\beta_{0,t} | \tilde{\beta_0}, \sigma_{\beta_0}] \times \\
&[\beta_{s,t} | \tilde{\beta_s}, \sigma_{\beta_s}] [\beta_c|\sigma_{beta_c}] \times \\
&\prod_{Q=1}^{Q_{tot}} [\beta_{Q} | \sigma_Q] \times \\
&[\tilde{\beta_0}] [\tilde{\beta_s}] [\sigma_Q] [\sigma_{\beta_0}] [\sigma_{\beta_s}] [\sigma]
\end{align}

**Stan code for model**
```{r qbm_stan, echo=TRUE, eval=FALSE}
data{
  int<lower=0> N; // observations
  int<lower=0> Yrs; // years
  int<lower=0> yid[N]; // year id
  int<lower=0> Covs; // climate covariates
  int<lower=0> G; // groups
  int<lower=0> gid[N]; // group id
  real<lower=0,upper=1> Y[N]; // observation vector
  real<lower=0> sd_clim; // prior sd on climate effects
  matrix[N,Covs] C; // climate matrix
  vector[N] X; // size vector
}
parameters{
  real a_mu;
  vector[Yrs] a;
  real b1_mu;
  vector[Yrs] b1;
  vector[Covs] b2;
  vector[G] gint;
  real<lower=0> sig_a;
  real<lower=0> sig_b1;
  real<lower=0> sig_G;
  real<lower=0> tau;
}
transformed parameters{
  real mu[N];
  vector[N] climEff;
  climEff <- C*b2;
  for(n in 1:N)
    mu[n] <- a[yid[n]] + gint[gid[n]] + b1[yid[n]]*X[n] + climEff[n];
}
model{
  // Priors
  a_mu ~ normal(0,1000);
  b1_mu ~ normal(0,1000);
  sig_a ~ cauchy(0,5);
  sig_b1 ~ cauchy(0,5);
  sig_G ~ cauchy(0,5);
  gint ~ normal(0, sig_G);
  b2 ~ normal(0,sd_clim);
  a ~ normal(a_mu, sig_a);
  b1 ~ normal(b1_mu, sig_b1);
  tau ~ cauchy(0,5);

  //Likelihood
  Y ~ lognormal(mu, tau);
}
```

\newpage{}

##  Supporting Figures
\begin{figure}[!ht]
  \centering
      \includegraphics[height=6in]{./components/forecast_error_mockup.png}
  \caption{Comparison of one-step-ahead, out-of-sample forecast error (mean absolute error) between the IPM and QBM models with and without the inclusion of climate covariates.}
\end{figure}

\newpage{}

\begin{figure}[!ht]
  \centering
      \includegraphics[height=4in]{./components/lppds_suppfig.png}
  \caption{Results from ridge regression: summed log pointwise predictive densities (\emph{lppd}) as a function of the prior variance of climate covariates. Dashed, red vertical lines show the highest prior variance with the highest \emph{lppd} indicating the value of prior variance for optimal prediction.}
\end{figure}

\newpage{}

## Supporting Tables
```{r lib_reqs, echo=FALSE, message=FALSE, warning=FALSE}
library(xtable)
library(plyr)
library(reshape2)
library(ggmcmc)
```

```{r agg_fxn, echo=FALSE, message=FALSE, warning=FALSE}
agg_fxn <- function(mcmc_frame){
  out <- ddply(mcmc_frame, .(Parameter), summarise,
               mean_value = mean(value),
               sd_value = sd(value),
               lo_BCI = quantile(value, 0.025),
               up_BCI = quantile(value, 0.975))
  return(out)
}
spp_list <- c("BOGR", "HECO", "PASM", "POSE")
spp_name <- c("B. gracilis.", "H. comata.", "P. smithii.", "P. secunda.")
```

In all tables, "mean_value" is the mean parameter estimate, "sd_value" is the standard deviation of the estimate from the MCMC, "lo_BCI" is the lower limit of the 95% Bayesian Credible Interval, and "up_BCI" is the upper limit of the 95% Bayesian Credible Interval. See Tables S1-S4 for definitions and notations for model parameters.

```{r surv_params, echo=FALSE, results='asis', message=FALSE, warning=FALSE, comment=NA}
for(i in 1:length(spp_list)){
  file <- paste("../analysis/ipm/vitalRateRegs/survival/survival_stanmcmc_",
              spp_list[i], ".RDS",
              sep="")
  survmcmc <- ggs(readRDS(file))
  aggsurv <- agg_fxn(survmcmc)
  cap <- paste("Statistical results from IPM survival model for", spp_name[i])
  print(xtable(aggsurv, caption = cap), comment=FALSE, include.rownames=FALSE, 
        size="\\footnotesize", caption.placement = "top")
}
```

```{r growth_params, echo=FALSE, results='asis', message=FALSE, warning=FALSE, comment=NA}
for(i in 1:length(spp_list)){
  file <- paste("../analysis/ipm/vitalRateRegs/growth/growth_stanmcmc_",
              spp_list[i], ".RDS",
              sep="")
  survmcmc <- readRDS(file)
  aggsurv <- agg_fxn(survmcmc)
  cap <- paste("Statistical results from IPM growth model for", spp_name[i])
  print(xtable(aggsurv, caption = cap), comment=FALSE, include.rownames=FALSE, 
        size="\\footnotesize", caption.placement = "top")
}
```

```{r rec_params, echo=FALSE, results='asis', message=FALSE, warning=FALSE, comment=NA}
for(i in 1:length(spp_list)){
  file <- paste("../analysis/ipm/vitalRateRegs/recruitment/recruitment_stanmcmc_",
              spp_list[i], ".RDS",
              sep="")
  survmcmc <- ggs(readRDS(file))
  aggsurv <- agg_fxn(survmcmc)
  cap <- paste("Statistical results from IPM recruitment model for", spp_name[i])
  print(xtable(aggsurv, caption = cap), comment=FALSE, include.rownames=FALSE, 
        size="\\footnotesize", caption.placement = "top")
}
```

```{r qbm_params, echo=FALSE, results='asis', message=FALSE, warning=FALSE, comment=NA}
for(i in 1:length(spp_list)){
  file <- paste("../analysis/quadBM/vitalRateRegressions/truncNormModel/popgrowth_stanmcmc_",
              spp_list[i], ".RDS",
              sep="")
  survmcmc <- readRDS(file)
  aggsurv <- agg_fxn(survmcmc)
  cap <- paste("Statistical results from QBM population model for", spp_name[i])
  print(xtable(aggsurv, caption = cap), comment=FALSE, include.rownames=FALSE, 
        size="\\footnotesize", caption.placement = "top")
}
```

\newpage{}

\begin{table}[ht]
\centering
\caption {Statistical tests of model comparisons.}
\begin{footnotesize}
\begin{tabular}{lllrrrl}
  \hline
comparison & performance\_measure & test\_type & test\_statistic & df & p\_value & species \\ 
  \hline
climate IPM vs. simple IPM & rho & t & 0.798 & 279.000 & 0.213 & BOGR \\ 
  climate IPM vs. simple IPM & MAE & t & 1.347 & 280.000 & 0.910 & BOGR \\ 
  climate QBM vs. simple QBM & rho & t & 0.526 & 279.000 & 0.300 & BOGR \\ 
  climate QBM vs. simple QBM & MAE & t & -0.478 & 280.000 & 0.317 & BOGR \\ 
  simple IPM vs. simple QBM & rho & t & -0.961 & 279.000 & 0.831 & BOGR \\ 
  simple IPM vs. simple QBM & MAE & t & 0.088 & 280.000 & 0.535 & BOGR \\ 
  climate IPM vs. climate QBM & rho & t & -0.024 & 279.000 & 0.509 & BOGR \\ 
  climate IPM vs. climate QBM & MAE & t & 2.289 & 280.000 & 0.989 & BOGR \\ 
  climate IPM vs. simple IPM & rho & t & 0.455 & 169.000 & 0.325 & HECO \\ 
  climate IPM vs. simple IPM & MAE & t & -2.292 & 170.000 & 0.012 & HECO \\ 
  climate QBM vs. simple QBM & rho & t & 0.051 & 169.000 & 0.480 & HECO \\ 
  climate QBM vs. simple QBM & MAE & t & -0.746 & 170.000 & 0.228 & HECO \\ 
  simple IPM vs. simple QBM & rho & t & -0.461 & 169.000 & 0.677 & HECO \\ 
  simple IPM vs. simple QBM & MAE & t & 1.626 & 170.000 & 0.947 & HECO \\ 
  climate IPM vs. climate QBM & rho & t & -0.299 & 169.000 & 0.617 & HECO \\ 
  climate IPM vs. climate QBM & MAE & t & 0.289 & 170.000 & 0.613 & HECO \\ 
  climate IPM vs. simple IPM & rho & t & -0.883 & 215.000 & 0.811 & PASM \\ 
  climate IPM vs. simple IPM & MAE & t & 0.410 & 216.000 & 0.659 & PASM \\ 
  climate QBM vs. simple QBM & rho & t & -0.008 & 215.000 & 0.503 & PASM \\ 
  climate QBM vs. simple QBM & MAE & t & 3.218 & 216.000 & 0.999 & PASM \\ 
  simple IPM vs. simple QBM & rho & t & 0.363 & 215.000 & 0.359 & PASM \\ 
  simple IPM vs. simple QBM & MAE & t & -0.276 & 216.000 & 0.391 & PASM \\ 
  climate IPM vs. climate QBM & rho & t & 0.600 & 215.000 & 0.274 & PASM \\ 
  climate IPM vs. climate QBM & MAE & t & -3.487 & 216.000 & 0.000 & PASM \\ 
  climate IPM vs. simple IPM & rho & t & 1.538 & 195.000 & 0.063 & POSE \\ 
  climate IPM vs. simple IPM & MAE & t & -1.844 & 196.000 & 0.033 & POSE \\ 
  climate QBM vs. simple QBM & rho & t & 1.616 & 195.000 & 0.054 & POSE \\ 
  climate QBM vs. simple QBM & MAE & t & -0.963 & 196.000 & 0.168 & POSE \\ 
  simple IPM vs. simple QBM & rho & t & 1.075 & 195.000 & 0.142 & POSE \\ 
  simple IPM vs. simple QBM & MAE & t & -0.788 & 196.000 & 0.216 & POSE \\ 
  climate IPM vs. climate QBM & rho & t & 1.721 & 195.000 & 0.043 & POSE \\ 
  climate IPM vs. climate QBM & MAE & t & -1.830 & 196.000 & 0.034 & POSE \\ 
   \hline
\end{tabular}
\end{footnotesize}
\end{table}

### 2016-05-25 FigureScripts/one_step_results_table.R

@@ -341,10 +341,9 @@ print_comparison_table <- function()
   }
   
   my_table <- xtable(temp_table_all, digits = 3)
-  print(my_table, type = "html", file = "Table_S1.html", include.rownames = FALSE)
-  
+  print(my_table, type = "html", file = "stats_table.html", include.rownames = FALSE)
+  print(my_table, type = "latex", file = "stats_table.latex", include.rownames = FALSE)
   return()
 }
-
 
 print_comparison_table()

### 2016-05-25 manuscript/manuscript.Rmd
add some text on vital rate connection to IPMs

@@
 #### Integral projection model
-We use a stochastic IPM [@Rees2009] that includes the climate covariates from the vital rate statistical models.
+We use a stochastic IPM [@Rees2009] to simulate our focal populations based on the vital rate regressions described above.
 In all simulations, we ignore the random year effects so that interannual variation is driven solely by climate.
@@
 which means we are calculating growth (_G_) for individuals that survive (_S_) from time _t_ to _t+1_ and adding in newly recruited (_R_) individuals of an average sized one-year-old genet for the focal species.
-Our stastical model for recruitment (_R_, described above) returns the number of new recruits produced per quadrat.
+Note the _S_, _G_, and _R_ are incorporated in the IPM using the fitted vital rate regressions.
+Our statistical model for recruitment (_R_, described above) returns the number of new recruits produced per quadrat.
 Following previous work [@Adler2012; @Chu2015], we assume that fecundity increases linearly with size ($R_j(v_{j},u_{j},\bar{w_{j}}) = e^{u_j}R_j(v_{j},\bar{w_{j}})$) to incorporate the recruitment function in the spatially-implicit IPM.
 
-We used random draws from the final 1,000 iterations, thinned by 10, from each of three MCMC chains to carry-through parameter uncertainty into our population models.
+We used random draws from the final 1,000 iterations, thinned by 10, from each of three MCMC chains for each vital rate regression to carry-through parameter uncertainty into our population models.
 At each time step, we randomly selected climate covariates from one of the 14 observed years.

### 2016-05-31 manuscript/manuscript.Rmd
adds MBH final comments

@@
 Most population models are built using demographic data from a single study site because tracking the fates of individuals is so difficult.
-The resulting models cannot be applied to the landscape and regional scales relevant to decision-making without information about how the fitted parameters respond to spatial variation in biotic and abiotic drivers [@Sæther2007].
+The resulting models cannot be applied to the landscape and regional scales relevant to decision-making without information about how the estimated parameters respond to spatial variation in biotic and abiotic drivers [@Sæther2007].
 The limited spatial extent of individual-level demographic datasets constrains our ability to use population models to address applied questions about the consequences of climate change.
@@
 
-From 1932 to 1945 individual plants were identified and mapped annually in 44 1-$\text{m}^2$ quadrats using a pantograph.
+From 1932 to 1945, individual plants were identified and mapped annually in 44 1-$\text{m}^2$ quadrats using a pantograph.
 The quadrats were distributed among six pastures, each assigned a grazing treatment of light (1.24 ha/animal unit month), moderate (0.92 ha/aum), and heavy (0.76 ha/aum) stocking rates (two pastures per treatment).
@@
 We modeled each grass population based on two levels of data: individual and quadrat.
-The individual data is the "raw" data.
+The individual data are the "raw" data.
 For the quadrat-level data we simply sum individual basal cover for each quadrat by species.
@@
 \begin{align}
-y_{i,q,t}^{S} &\sim \text{Bernoulli}(s_{i,q,t}) \\
-\text{logit}(s_{i,q,t}) &= \beta_{0,t} + \beta_{s,t}x_{i,q,t} + \beta_Q + \textbf{z}'_t \boldsymbol{\beta}_c + \beta_{d,1} w_{i,t} + \beta_{d,2} (x_{i,q,t}w_{i,q,t})
+y_{i,q,t}^{S} &\sim \text{Bernoulli}(s_{i,q,t}), \\
+\text{logit}(s_{i,q,t}) &= \beta_{0,t} + \beta_{s,t}x_{i,q,t} + \beta_Q + \textbf{z}'_t \boldsymbol{\beta}_c + \beta_{d,1} w_{i,t} + \beta_{d,2} (x_{i,q,t}w_{i,q,t}),
 \end{align}
@@
 \begin{align}
-y_{i,q,t+1}^G &\sim \text{Normal}(\mu_{i,q,t+1}, \sigma^2_{x_{i,q,t+1}}\textbf{I}) \\
-\mu_{i,q,t+1} &= \beta_{0,t} + \beta_{s,t}x_{i,q,t} + \beta_Q + \textbf{z}'_t \boldsymbol{\beta}_c + \beta_{d,1} w_{i,q,t} + \beta_{d,2} (x_{i,q,t}w_{i,q,t})
+y_{i,q,t+1}^G &\sim \text{Normal}(\mu_{i,q,t+1}, \sigma^2_{x_{i,q,t+1}}\textbf{I}), \\
+\mu_{i,q,t+1} &= \beta_{0,t} + \beta_{s,t}x_{i,q,t} + \beta_Q + \textbf{z}'_t \boldsymbol{\beta}_c + \beta_{d,1} w_{i,q,t} + \beta_{d,2} (x_{i,q,t}w_{i,q,t}),
 \end{align}
@@
 \begin{align}
-\sigma^2_{x_{i,q,t+1}} = a \, \text{exp}[b \times \mu_{i,q,t+1}]
+\sigma^2_{x_{i,q,t+1}} = a \, \text{exp}[b \times \mu_{i,q,t+1}],
 \end{align}
@@
 \begin{align}
-y^{R}_{q,t+1} \sim \text{NegBin}(\lambda_{q,t+1},\phi)
+y^{R}_{q,t+1} \sim \text{NegBin}(\lambda_{q,t+1},\phi),
 \end{align}
@@
 \begin{align}
-\lambda_{q,t+1} = \tilde{c}_{q,t} \, \text{exp}\left(\beta_{0,t} + \beta_Q + \textbf{z}'_t \boldsymbol{\beta}_c + \beta_{d}\sqrt{\tilde{c}_{q,t}}\right)
+\lambda_{q,t+1} = \tilde{c}_{q,t} \, \text{exp}\left(\beta_{0,t} + \beta_Q + \textbf{z}'_t \boldsymbol{\beta}_c + \beta_{d}\sqrt{\tilde{c}_{q,t}}\right),
 \end{align}
@@
 \begin{align}
-\tilde{c}_{q,t} = pc_{q,t} + (1-p)\bar{c}_{Q,t}
+\tilde{c}_{q,t} = pc_{q,t} + (1-p)\bar{c}_{Q,t},
 \end{align}
@@
 \begin{align}
-y_{q,t+1}^{P} &\sim \text{LogNormal}(\mu_{q,t+1}, \sigma^2)_{0}^{1} \\
-\mu_{q,t+1} &= \beta_{0,t} + \beta_{s,t}x_{q,t} + \beta_Q + \textbf{z}'_t \boldsymbol{\beta}_c
+y_{q,t+1}^{P} &\sim \text{LogNormal}(\mu_{q,t+1}, \sigma^2)_{0}^{1}, \\
+\mu_{q,t+1} &= \beta_{0,t} + \beta_{s,t}x_{q,t} + \beta_Q + \textbf{z}'_t \boldsymbol{\beta}_c,
 \end{align}
@@
 Our Bayesian approach to fitting the vital rate models required choosing appropriate priors for unknown parameters and deciding which, if any, of those priors should be hierarchical.
-We decided to fit models where all terms were fit by species.
-Within a species, we fit yearly size effects and yearly intercepts hierarchically, where year-specific coefficients were drawn from global distributions representing the mean size effect and intercept.
+For each species, we fit yearly size effects and yearly intercepts hierarchically, where year-specific coefficients were drawn from global distributions representing the mean size effect and intercept.
 Quadrat random effects were also fit hierarchically, with quadrat offsets being drawn from distributions with mean zero and a shared variance term (independent Gaussian priors).
@@
 We obtained posterior distributions for all model parameters from three parallel MCMC chains run for 1,000 iterations after discarding an initial 1,000 iterations.
-Such short MCMC chains may surprise readers more familiar with other MCMC samplers (i.e., JAGS or WinBUGS), but the Stan sampler reduces the number of iterations needed to achieve convergence.
+Such short MCMC chains are possible because the Stan sampler reduces the number of iterations needed to achieve convergence.
 We assessed convergence visually and checked that scale reduction factors for all parameters were less than 1.1.
-For the purposes of including parameter uncertainty in our population models, we saved the final 1,000 iterations from each of the three MCMC chains to be used as randomly drawn values during population simulation.
+For the purposes of including parameter uncertainty in our population models, we retained the final 1,000 iterations from each of the three MCMC chains to be used as randomly drawn values during population simulation.
 We report the posterior mean, standard deviation, and 95% Bayesian Credible Intervals for every parameter of each model for each species in the Supporting Information (Tables S5-S20).
@@
 The natural regulator in a Bayesian application is the prior on the coefficient of interest.
-Each of our statistical models includes the effects of climate covariates via the term $\textbf{z}'_t \boldsymbol{\beta}_c$ with prior $\boldsymbol{\beta}_c \sim \text{Normal}(\boldsymbol{\mu}_{\beta_c}, \sigma_{\beta_c}^2)$.
-Since we standardized all climate covariates to have mean zero and variance one, we can set $\boldsymbol{\mu}_{\beta_c} = 0$ and let $\sigma_{\beta_c}^2$ serve as the regulator that can shrink covariates toward zero -- the smaller the prior variance, the more the posteriors of $\boldsymbol{\beta}_c$ are shrunk toward zero, and the stronger the penalty [@Hooten2015].
+Each of our statistical models includes the effects of climate covariates via the term $\textbf{z}'_t \boldsymbol{\beta}_c$ with prior $\boldsymbol{\beta}_c \sim \text{Normal}(\boldsymbol{\mu}_{\beta_c}, \sigma_{\beta_c}^2\textbf{I})$.
+Because we standardized all climate covariates to have mean zero and variance one, we set $\boldsymbol{\mu}_{\beta_c} = 0$ and let $\sigma_{\beta_c}^2$ serve as the regulator that shrinks covariates toward zero -- the smaller the prior variance, the more the posteriors of $\boldsymbol{\beta}_c$ are shrunk toward zero, and the stronger the penalty [@Hooten2015].
 
@@
 We performed the grid search over 24 values of $\sigma_{\beta_c}^2$, ranging from $\sigma_{\beta_c}^2 = 0.01$ to $\sigma_{\beta_c}^2 = 2.25$.
-For each statistical model and each species we fit $13\times24=312$ iterations of the model fitting algorithm to search $\sigma_{\beta_c}^2$ for the optimal value (13 years to leave out for cross-validation and 24 values of $\sigma_{\beta_c}^2$) -- a total of 4,992 model fits.
+For each statistical model and each species, we fit $13\times24=312$ iterations of the model fitting algorithm to search $\sigma_{\beta_c}^2$ for the optimal value (13 years to leave out for cross-validation and 24 values of $\sigma_{\beta_c}^2$) -- a total of 4,992 model fits.
 We calculated the log pointwise predictive density (*lppd*) to score each model's ability to predict the left-out data [@Gelman2014].
@@
 We fit the random year effects in the vital rate regressions to avoid over-attributing variation to climate covariates.
-Our IPM follows the specification of @Chu2015 where the population of species _j_ is a density function $n(u_{j},t)$ giving the density of sized-_u_ genets at time _t_.
+Our IPM follows the specification of @Chu2015 where the population of species _j_ is $n(u_{j},t)$, giving the density of sized-_u_ genets at time _t_.
 Genet size is on the natural log scale, so that $n(u_{j},t)du$ is the number of genets whose area (on the arithmetic scale) is between $e^{u_{j}}$ and $e^{u_{j}+du}$.
-The density function for any size _v_ at time $t+1$ is
+The function for any size _v_ at time $t+1$ is
 
@@
 \begin{align}
-n(v_{j},t+1) = \int_{L_{j}}^{U_{j}} k_{j}(v_{j},u_{j},\bar{w_{j}}(u_{j}))n(u_{j},t)du_{j}
+n(v_{j},t+1) = \int_{L_{j}}^{U_{j}} k_{j}(v_{j},u_{j},\bar{w_{j}}(u_{j}))n(u_{j},t)du_{j},
 \end{align}
@@
 
-The IPM is spatially-implicit, thus we cannot calculate neighborhood crowding for specific genets ($w_{ij}$).
+The IPM is spatially-implicit, thus, we cannot calculate neighborhood crowding for specific genets ($w_{ij}$).
 Instead, we use an approximation ($\bar{w_{j}}$) that captures the essential features of neighborhood interactions [@Adler2010].
@@
 Relatively unimportant climate covariates (those that broadly overlap 0) will have little effect on the mean of the simulation results, but can contribute to their variation.
-Because our focus was on the contribution of density dependence and climate covariates to population states, we set the random year effects and the random group effects to zero.
+Our focus was on the contribution of density dependence and climate covariates to population states, thus we set the random year effects and the random group effects to zero.
 
 #### Quad-based model
-To simulate our quad-based model (QBM), we simply iterate the quadrat-level statistical model (Eqs.
+To simulate our quad-based model (QBM), we iterate the quadrat-level statistical model (Eqs.
 9-10).
@@
 We project the model forward by drawing a new parameter set (unique to climate year and MCMC iteration) at each timestep.
-As with the IPM, random year effects are ignored for all simulations.
+As with the IPM, random year effects are ommitted in our simulations.
 
@@
 We repeated this procedure for all 13 observation years, making 100 one-step-ahead forecasts for each quadrat-year combination with parameter uncertainty included via random draw from the MCMC chain as described above.
-Random year effects were set to zero because year effects cannot be assigned to unobserved years.
+Year-specific intercepts for left-out data were drawn from the posterior distribution of the mean intercept.
+Thus, for some future year *T*, $\beta_0,T \sim \text{Normal}(\beta_0, \sigma_{\beta_0})$.
 
@@
 An important feature of any forecasting model is the rate at which forecast skill declines as the time between an observation and a forecast increases; the so-called ecological forecast horizon [@Petchey2015].
-To assess the forecast horizons of our models, we iniate the model with the population state at some time *t* and make sequential forecasts of the population at times $t+1, t+2, \dots, t+T$ where *T* is the maximum number of years between the initial year and the final year of our observations.
-For example, if we initialize the model with percent cover in 1940, we are able to make five forecasts up to the year 1945.
-Models are not re-initialized with observations between years.
+To assess the forecast horizons of our models, we initiate the forecast model with the population state at some time *t* and make sequential forecasts of the population at times $t+1, t+2, \dots, t+T$ where *T* is the maximum number of years between the initial year and the final year of our observations.
+For example, if we initialize the forecast model with percent cover in 1940, we are able to make five forecasts up to the year 1945.
+Forecast models are not re-initialized with observations between years.
 Thus, in our current example, the model forecast for percent cover in 1941 has a forecast horizon of one year, the forecast in 1942 has a forecast horizon of two years, and so on.
@@
 S2, Supporting Information).
-Thus, the posterior distributions of climate effects included in our models overlapped zero and generally shrunk toward zero, though for some species-vital rate combinations important effects (80% credible interval does not include zero) did emerge (Fig.
+Thus, the posterior distributions of climate effects included in our models overlapped zero and generally shrunk toward zero, though for some species-vital rate combinations, important effects (80% credible interval does not include zero) did emerge (Fig.
 3).
@@
 However, the IPM with climate covariates for *Poa secunda* did have significantly lower error (mean absolute error) than the IPM without climate covariates ($t_{(196)}$ = -1.84, *P* = 0.033; Fig. S2).
-In only one case were IPM forecasts significantly more accurate than the QBM (Fig.
+IPM forecasts were significantly more accurate than the QBM in only one case (Fig.
 2): forecast accuracy of *Poa secunda* percent cover from an IPM with climate covariates was greater than the accuracy from the QBM with climate covariates ($t_{(195)}$ = 1.72, *P* = 0.043).
@@
 The accuracy of both model's forecasts declined as the forecast horizon increased, but they did so at similar rates (Fig. 4).
-The only exception is for *Poa secunda* where forecast accuracy appears to remain steady as the forecast horizon increases, after an initial decrease from a forecast horizon of one year (Fig. 4).
+The only exception is for *Poa secunda*, where forecast accuracy remained steady as the forecast horizon increases, after an initial decrease from a forecast horizon of one year (Fig. 4).
 
@@
 ----------
-Population models built using individual-level data allow inference on demographic processes, but they can only forecast future population states across the (typically limited) spatial extent of the observations.
+Population models built using individual-level data provide inference on demographic processes, but they can only forecast future population states across the (typically limited) spatial extent of the observations.
 Population-level data are much easier to collect across broad spatial extents, so models built using such data offer an appealing alternative to traditional population models [@Queenborough2011]. 
@@
 
-The higher accuracy of the IPM with climate covariates for *Poa secunda*  highlights the advantage of contemporary model and variable selection approaches such as ridge regression and LASSO over techniques that would exclude "non-significant" effects from final models.
+The higher accuracy of the IPM with climate covariates for *Poa secunda*  highlights the advantage of contemporary modeling and variable selection approaches such as ridge regression and LASSO over techniques that would exclude "non-significant" effects from final models.
 Ridge regression allows researchers to retain covariates whose effects may be difficult to identify in noisy data or short time series.
-This is especially important when forecasting the impacts of climate change where it is important to include to effects of forcing variables (e.g., temperature and precipitation) even of such effects are difficult to identify.
+This is especially important when forecasting the impacts of climate change, where it is important to include to effects of forcing variables (e.g., temperature and precipitation) even of such effects are difficult to identify.
 If a species is truly unresponsive to a given climate variable, statistical regularization techniques will shrink the mean and variance of a covariate estimate toward zero [@Hooten2015].
@@
 Nonetheless, our results are encouraging for the use of easy-to-collect percent cover for forecasting the state of plant populations.
+Our methodological approach could also potentially be used for backcasting paleoclimate and paleoecological aggregates such as pollen counts based on lake sediments [e.g., @Paciorek2009]
 
@@
 ---------------
-This work was funded by the National Science Foundation through a Postdoctoral Research Fellowship in Biology to ATT (DBI-1400370) and a CAREER award to PBA (DEB-1054040).
+This work was funded by the National Science Foundation through a Postdoctoral Research Fellowship in Biology to ATT (DBI-1400370), award MSB-1241856 to MBH, and a CAREER award to PBA (DEB-1054040).
 We thank the original mappers of the permanent quadrats in Montana and the digitizers in the Adler lab, without whom this work would not have been possible.

### 2016-05-31 manuscript/manuscript.Rmd
adds in PBA and MBH final edits

@@
 **5.** *Synthesis.* For our focal species at this particular location, and using our particular statistical models, percent cover models generated forecasts as skillful as those from a demographic model.
-However, for certain species forecast skill can be gained by using demographic data linked to climate covariates.
-We conclude that models based on aggregated individual-level data offer a practical alternative to data-intensive demographic models when species do not respond strongly to interannual variation in weather, but when modeling species that do respond to climate drivers, demographically-based models can generate more skillful forecasts.
+However, for one species, forecast skill increased when we linked demographic data to climate covariates.
+We conclude that models based on aggregated individual-level data offer a practical alternative to data-intensive demographic models when species do not respond strongly to interannual variation in weather, but when species do respond to climate drivers, demographically-based models can generate more skillful forecasts.
 
@@
 
-The accuracy of both model's forecasts declined as the forecast horizon increased, but they did so at similar rates (Fig. 4).
-The only exception is for *Poa secunda*, where forecast accuracy remained steady as the forecast horizon increases, after an initial decrease from a forecast horizon of one year (Fig. 4).
+With climate covariates included, the accuracy of both models' forecasts declined as the forecast horizon increased, but they did so at similar rates (Fig. 4).
+The only exception is for *Poa secunda*, where IPM forecast accuracy remained steady as the forecast horizon increases, after an initial decrease from a forecast horizon of one year, whereas QBM forecast accuracy declined (Fig.
+4).
 
@@
 
-Our naive expectation was that, a single, relatively important climate effect in a vital rate regression would cause a discernable difference between IPM and QBM forecast skill because individual-level responses may be less identifiable from percent cover data [@Clark2012].
+Our naive expectation was that a single, relatively important climate effect in a vital rate regression would cause a discernable difference between IPM and QBM forecast skill because individual-level responses may be less identifiable from percent cover data [@Clark2012].
 We actually found the opposite.
@@
 Thus, we conclude that models based on population-level data, rather than individual-level data, may be adequate for forecasting the state of plant populations for species that do not respond consistently or strongly to climate.
-Unfortunately, our analysis, where climate effects were relatively unimportant in vital rate regressions, did not allow us to sufficiently test our prediction that individual-level data is neccessary to generate skillful forecasts if different vital rates respond to climate in unique, potenially opposing, ways.
-Nonetheless, our results are encouraging for the use of easy-to-collect percent cover for forecasting the state of plant populations.
+Unfortunately, our analysis, where climate effects were relatively unimportant in vital rate regressions, did not allow us to sufficiently test our prediction that individual-level data is neccessary to generate skillful forecasts if different vital rates respond to climate in unique, potentially opposing, ways.
+Nonetheless, our results are encouraging for the use of easy-to-collect population-level data for forecasting the state of plant populations.
 Our methodological approach could also potentially be used for backcasting paleoclimate and paleoecological aggregates such as pollen counts based on lake sediments [e.g., @Paciorek2009]
@@
 Comparisons between equivalent IPM and QBM models indicate no significant difference in accuracy (\emph{P} > 0.05 for all comparisons).
-Including climate covariates resulted in siginificantly higher forecast accuracy for only one species, \emph{Poa secunda} (\emph{P} = 0.043).}
+Including climate covariates resulted in siginificantly higher forecast accuracy for only one species, \emph{Poa secunda} (\emph{P} = 0.043).
+Species codes are as in Fig.
+1}
 \end{figure}
@@
 Because our priors were constrained via ridge-regression, we highlight climate effects whose 80\% credible intervals do not overlap zero (red for negative coefficients, blue for positive coefficients).
-Kernel bandwidths of posterior densities were adjusted by a factor of 4 for visual clarity.}
+Kernel bandwidths of posterior densities were adjusted by a factor of 4 for visual clarity.
+Species codes are as in Fig.
+1.}
 \end{figure}
@@
       \includegraphics[height=5in]{./components/forecast_horizon.png}
-  \caption{The forecast horizons for both models. Points show the average accuracy ($\rho$) across all forecasts at a given time horizon.}
+  \caption{The forecast horizons for both models with climate covariates included.
+Points show the average accuracy ($\rho$) across all forecasts at a given time horizon.
+We only examine the forecast accuracy of models with climate covariates included because in no case did including climate covariates decrease accuracy (see Fig.
+2).
+Species codes are as in Fig.
+1.}
 \end{figure}

### 2016-05-31 manuscript/supporting_information.Rmd
adds in PBA and MBH final edits

@@
 All prior distributions are defined in the parameter model sections.
-For clarity we avoid subscripting species.
+For clarity, we avoid subscripting species.
 
@@
 &\mathtt{process} \qquad &\text{logit}(s_{i,q,t}) = \beta_{0,t} + \beta_{s,t}x_{i,q,t} + \beta_Q + \textbf{z}'_t \boldsymbol{\beta}_c + \beta_{d,1} w_{i,t} + \beta_{d,2} (x_{i,q,t}w_{i,q,t}) \\
-&\mathtt{parameters} \qquad &\beta_{0,t} \sim \text{Normal}(\tilde{\beta_0}, \sigma_{\beta_0} \textbf{I})\\
-& \qquad & \beta_{s,t} \sim \text{Normal}(\tilde{\beta_s}, \sigma_{\beta_s} \textbf{I}) \\
-& \qquad & \boldsymbol{\beta_{Q}} \sim \text{Normal}(0, \sigma_Q \textbf{I}) \\
-& \qquad & \boldsymbol{\beta}_c \sim \text{Normal}(0, \sigma_c \textbf{I}) \\
-& \qquad & \beta_{d,1} \sim \text{Normal}(0, 100) \\
-& \qquad & \beta_{d,2} \sim \text{Normal}(0, 100) \\
-& \qquad & \tilde{\beta_0} \sim \text{Normal}(0, 100) \\
-& \qquad & \tilde{\beta_s} \sim \text{Normal}(0, 100) \\
+&\mathtt{parameters} \qquad &\beta_{0,t} \sim \text{Normal}(\tilde{\beta_0}, \sigma_{\beta_0}^2 \textbf{I})\\
+& \qquad & \beta_{s,t} \sim \text{Normal}(\tilde{\beta_s}, \sigma_{\beta_s}^2 \textbf{I}) \\
+& \qquad & \boldsymbol{\beta_{Q}} \sim \text{Normal}(0, \sigma_Q^2 \textbf{I}) \\
+& \qquad & \boldsymbol{\beta}_c \sim \text{Normal}(0, \sigma_c^2 \textbf{I}) \\
+& \qquad & \beta_{d,1} \sim \text{Normal}(0, 100^2) \\
+& \qquad & \beta_{d,2} \sim \text{Normal}(0, 100^2) \\
+& \qquad & \tilde{\beta_0} \sim \text{Normal}(0, 100^2) \\
+& \qquad & \tilde{\beta_s} \sim \text{Normal}(0, 100^2) \\
 & \qquad & \sigma_Q \sim \text{Cauchy}(0, 5) \\
@@
 \begin{align}
-[\boldsymbol{\beta_0}, \tilde{\beta_0}, \boldsymbol{\beta_s}, \tilde{\beta_s}, \boldsymbol{\beta_Q}, \boldsymbol{\beta_c}, \beta_{d,1}, \beta_{d,2}, \sigma_Q, \sigma_{\beta_0}, \sigma_{\beta_s}] &\propto \\
-&\prod_{t=1}^T \prod_{i=1}^n [y_{iQ,t} | \beta_{0,t}, \beta_{s,t}, \beta_{Q}, \boldsymbol{\beta_c}, \beta_{d,1}, \beta_{d,2}] \times \\
-&[\beta_{0,t} | \tilde{\beta_0}, \sigma_{\beta_0}] \times \\
-&[\beta_{s,t} | \tilde{\beta_s}, \sigma_{\beta_s}] [\beta_c|\sigma_{beta_c}] \times \\
-&\prod_{Q=1}^{Q_{tot}} [\beta_{Q} | \sigma_Q] \times \\
+[\boldsymbol{\beta_0}, \tilde{\beta_0}, \boldsymbol{\beta_s}, \tilde{\beta_s}, \boldsymbol{\beta_Q}, \boldsymbol{\beta_c}, \beta_{d,1}, \beta_{d,2}, \sigma_Q^2, \sigma_{\beta_0}^2, \sigma_{\beta_s}^2 | y_{i,q,t}^{S}] &\propto \\
+&\prod_{t=1}^T \prod_{i=1}^n [y_{i,q,t}^{S} | \beta_{0,t}, \beta_{s,t}, \beta_{Q}, \boldsymbol{\beta_c}, \beta_{d,1}, \beta_{d,2}] \times \\
+&[\beta_{0,t} | \tilde{\beta_0}, \sigma_{\beta_0}^2] \times \\
+&[\beta_{s,t} | \tilde{\beta_s}, \sigma_{\beta_s}^2] [\beta_c|\sigma_{beta_c}^2] \times \\
+&\prod_{Q=1}^{Q_{tot}} [\beta_{Q} | \sigma_Q^2] \times \\
 &[\beta_{d,1}] [\beta_{d,1}] [\tilde{\beta_0}] [\tilde{\beta_s}] [\sigma_Q] [\sigma_{\beta_0}] [\sigma_{\beta_s}]
@@
 
-\vspace{-3em}
 \begin{align}
@@
 \end{align}
-\vspace{-3em}
 
@@
 & \qquad & \sigma^2_{iQ,t} = ae^{b\mu_{iQ,t+1}} \\
-&\mathtt{parameters} \qquad &\beta_{0,t} \sim \text{Normal}(\tilde{\beta_0}, \sigma_{\beta_0} \textbf{I})\\
-& \qquad & \beta_{s,t} \sim \text{Normal}(\tilde{\beta_s}, \sigma_{\beta_s} \textbf{I}) \\
-& \qquad & \boldsymbol{\beta_{Q}} \sim \text{Normal}(0, \sigma_Q \textbf{I}) \\
-& \qquad & \boldsymbol{\beta}_c \sim \text{Normal}(0, \sigma_c \textbf{I}) \\
-& \qquad & \beta_{d,1} \sim \text{Normal}(0, 10) \\
-& \qquad & \beta_{d,2} \sim \text{Normal}(0, 10) \\
-& \qquad & \tilde{\beta_0} \sim \text{Normal}(0, 100) \\
-& \qquad & \tilde{\beta_s} \sim \text{Normal}(0, 100) \\
+&\mathtt{parameters} \qquad &\beta_{0,t} \sim \text{Normal}(\tilde{\beta_0}, \sigma_{\beta_0}^2 \textbf{I})\\
+& \qquad & \beta_{s,t} \sim \text{Normal}(\tilde{\beta_s}, \sigma_{\beta_s}^2 \textbf{I}) \\
+& \qquad & \boldsymbol{\beta_{Q}} \sim \text{Normal}(0, \sigma_Q^2 \textbf{I}) \\
+& \qquad & \boldsymbol{\beta}_c \sim \text{Normal}(0, \sigma_c^2 \textbf{I}) \\
+& \qquad & \beta_{d,1} \sim \text{Normal}(0, 10^2) \\
+& \qquad & \beta_{d,2} \sim \text{Normal}(0, 10^2) \\
+& \qquad & \tilde{\beta_0} \sim \text{Normal}(0, 100^2) \\
+& \qquad & \tilde{\beta_s} \sim \text{Normal}(0, 100^2) \\
 & \qquad & \sigma_Q \sim \text{Cauchy}(0, 2) \\
@@
 & \qquad & \sigma_{\beta_s} \sim \text{Cauchy}(0, 2) \\
-& \qquad & a \sim \text{Normal}(0, 100) \\
-& \qquad & b \sim \text{Normal}(0, 100)
+& \qquad & a \sim \text{Normal}(0, 100^2) \\
+& \qquad & b \sim \text{Normal}(0, 100^2)
 \end{align}
@@
 \begin{align}
-[\boldsymbol{\beta_0}, \tilde{\beta_0}, \boldsymbol{\beta_s}, \tilde{\beta_s}, \boldsymbol{\beta_Q}, \boldsymbol{\beta_c}, \beta_{d,1}, \beta_{d,2}, \sigma_Q, \sigma_{\beta_0}, \sigma_{\beta_s},a,b] &\propto \\
-&\prod_{t=1}^T \prod_{i=1}^n [y_{iQ,t} | \beta_{0,t}, \beta_{s,t}, \beta_{Q}, \boldsymbol{\beta_c}, \beta_{d,1}, \beta_{d,2}, a, b] \times \\
-&[\beta_{0,t} | \tilde{\beta_0}, \sigma_{\beta_0}] \times \\
-&[\beta_{s,t} | \tilde{\beta_s}, \sigma_{\beta_s}] [\beta_c|\sigma_{beta_c}] \times \\
-&\prod_{Q=1}^{Q_{tot}} [\beta_{Q} | \sigma_Q] \times \\
+[\boldsymbol{\beta_0}, \tilde{\beta_0}, \boldsymbol{\beta_s}, \tilde{\beta_s}, \boldsymbol{\beta_Q}, \boldsymbol{\beta_c}, \beta_{d,1}, \beta_{d,2}, \sigma_Q^2, \sigma_{\beta_0}^2, \sigma_{\beta_s}^2,a,b | y_{i,q,t+1}^G] &\propto \\
+&\prod_{t=1}^T \prod_{i=1}^n [y_{i,q,t+1}^G | \beta_{0,t}, \beta_{s,t}, \beta_{Q}, \boldsymbol{\beta_c}, \beta_{d,1}, \beta_{d,2}, a, b] \times \\
+&[\beta_{0,t} | \tilde{\beta_0}, \sigma_{\beta_0}^2] \times \\
+&[\beta_{s,t} | \tilde{\beta_s}, \sigma_{\beta_s}^2] [\beta_c|\sigma_{beta_c}^2] \times \\
+&\prod_{Q=1}^{Q_{tot}} [\beta_{Q} | \sigma_Q^2] \times \\
 &[\beta_{d,1}] [\beta_{d,1}] [\tilde{\beta_0}] [\tilde{\beta_s}] [\sigma_Q] [\sigma_{\beta_0}] [\sigma_{\beta_s}] [a] [b]
@@
 &\mathtt{process} \qquad &\lambda_{q,t+1} = \tilde{c}_{q,t} \, \text{exp}\left(\beta_{0,t} + \beta_Q + \textbf{z}'_t \boldsymbol{\beta}_c + \beta_{d}\sqrt{\tilde{c}_{q,t}}\right) \\
-&\mathtt{parameters} \qquad &\beta_{0,t} \sim \text{Normal}(\tilde{\beta_0}, \sigma_{\beta_0} \textbf{I})\\
-& \qquad & \beta_{Q} \sim \text{Normal}(0,\sigma_Q \textbf{I}) \\
-& \qquad & \boldsymbol{\beta}_c \sim \text{Normal}(0, \sigma_c \textbf{I}) \\
+&\mathtt{parameters} \qquad &\beta_{0,t} \sim \text{Normal}(\tilde{\beta_0}, \sigma_{\beta_0}^2 \textbf{I})\\
+& \qquad & \beta_{Q} \sim \text{Normal}(0,\sigma_Q^2 \textbf{I}) \\
+& \qquad & \boldsymbol{\beta}_c \sim \text{Normal}(0, \sigma_c^2 \textbf{I}) \\
 & \qquad & \beta_{d} \sim \text{Uniform}(-100, 100) \\
-& \qquad & \tilde{\beta_0} \sim \text{Normal}(0, 1000) \\
+& \qquad & \tilde{\beta_0} \sim \text{Normal}(0, 1000^2) \\
 & \qquad & \sigma_Q \sim \text{Cauchy}(0, 5) \\
@@
 \begin{align}
-[\boldsymbol{\beta_0}, \tilde{\beta}, \boldsymbol{\beta_Q}, \boldsymbol{\beta_c}, \beta_{d}, \sigma_Q, \sigma_{\beta_0}, \phi, u] &\propto \\
-&\prod_{t=1}^T \prod_{i=1}^n [y_{q,t+1} | \beta_{0,t}, \beta_{Q}, \beta_d, \boldsymbol{\beta_c}, \phi, u] [\beta_{0,t} | \tilde{\beta}] \times \\
-&\prod_{Q=1}^{Q_{tot}} [\beta_{Q} | \sigma_Q] [\boldsymbol{\beta_c}] [\beta_d] [\tilde{\beta_0}] [\sigma_Q] [\sigma_{\beta_0}] [\phi] [u]
+[\boldsymbol{\beta_0}, \tilde{\beta}, \boldsymbol{\beta_Q}, \boldsymbol{\beta_c}, \beta_{d}, \sigma_Q^2, \sigma_{\beta_0}^2, \phi, u | y^{R}_{q,t+1}] &\propto \\
+&\prod_{t=1}^T \prod_{i=1}^n [y^{R}_{q,t+1} | \beta_{0,t}, \beta_{Q}, \beta_d, \boldsymbol{\beta_c}, \phi, u] [\beta_{0,t} | \tilde{\beta}, \sigma_{\beta_0}^2] \times \\
+&\prod_{Q=1}^{Q_{tot}} [\beta_{Q} | \sigma_Q^2] [\boldsymbol{\beta_c}] [\beta_d] [\tilde{\beta_0}] [\sigma_Q] [\sigma_{\beta_0}] [\phi] [u]
 \end{align}
@@
 \begin{align}
-&\mathtt{data} \qquad &y_{q,t+1}^{P} \sim \text{LogNormal}(\mu_{q,t+1}, \sigma)_{0}^{1} \\
+&\mathtt{data} \qquad &y_{q,t+1}^{P} \sim \text{LogNormal}(\mu_{q,t+1}, \sigma^2)_{0}^{1} \\
 &\mathtt{process} \qquad &\mu_{q,t+1} = \beta_{0,t} + \beta_{s,t}x_{q,t} + \beta_Q + \textbf{z}'_t \boldsymbol{\beta}_c \\
-&\mathtt{parameters} \qquad &\beta_{0,t} \sim \text{Normal}(\tilde{\beta_0}, \sigma_{\beta_0})\\
-& \qquad & \beta_{s,t} \sim \text{Normal}(\tilde{\beta_s}, \sigma_{\beta_s} \textbf{I}) \\
-& \qquad & \beta_{Q} \sim \text{Normal}(0,\sigma_Q \textbf{I}) \\
-& \qquad & \boldsymbol{\beta}_c \sim \text{Normal}(0, \sigma_c \textbf{I}) \\
-& \qquad & \tilde{\beta_0} \sim \text{Normal}(0, 1000) \\
-& \qquad & \tilde{\beta_s} \sim \text{Normal}(0, 1000) \\
+&\mathtt{parameters} \qquad &\beta_{0,t} \sim \text{Normal}(\tilde{\beta_0}, \sigma_{\beta_0}^2)\\
+& \qquad & \beta_{s,t} \sim \text{Normal}(\tilde{\beta_s}, \sigma_{\beta_s}^2 \textbf{I}) \\
+& \qquad & \beta_{Q} \sim \text{Normal}(0,\sigma_Q^2 \textbf{I}) \\
+& \qquad & \boldsymbol{\beta}_c \sim \text{Normal}(0, \sigma_c^2 \textbf{I}) \\
+& \qquad & \tilde{\beta_0} \sim \text{Normal}(0, 1000^2) \\
+& \qquad & \tilde{\beta_s} \sim \text{Normal}(0, 1000^2) \\
 & \qquad & \sigma_Q \sim \text{Cauchy}(0, 5) \\
@@
 \begin{align}
-[\boldsymbol{\beta_0}, \tilde{\beta_0}, \boldsymbol{\beta_s}, \tilde{\beta_s}, \boldsymbol{\beta_Q}, \boldsymbol{\beta_c}, \sigma_Q, \sigma_{\beta_0}, \sigma_{\beta_s}, \sigma] &\propto \\
+[\boldsymbol{\beta_0}, \tilde{\beta_0}, \boldsymbol{\beta_s}, \tilde{\beta_s}, \boldsymbol{\beta_Q}, \boldsymbol{\beta_c}, \sigma_Q^2, \sigma_{\beta_0}^2, \sigma_{\beta_s}^2, \sigma^2 | y_{q,t+1}^{P}] &\propto \\
 &\prod_{t=1}^T \prod_{i=1}^n [y_{q,t+1} | \beta_{0,t}, \beta_{s,t}, \beta_{Q}, \boldsymbol{\beta_c}] \times \\
-&[\beta_{0,t} | \tilde{\beta_0}, \sigma_{\beta_0}] \times \\
-&[\beta_{s,t} | \tilde{\beta_s}, \sigma_{\beta_s}] [\beta_c|\sigma_{beta_c}] \times \\
-&\prod_{Q=1}^{Q_{tot}} [\beta_{Q} | \sigma_Q] \times \\
+&[\beta_{0,t} | \tilde{\beta_0}, \sigma_{\beta_0}^2] \times \\
+&[\beta_{s,t} | \tilde{\beta_s}, \sigma_{\beta_s}^2] [\beta_c|\sigma_{beta_c}^2] \times \\
+&\prod_{Q=1}^{Q_{tot}} [\beta_{Q} | \sigma_Q^2] \times \\
 &[\tilde{\beta_0}] [\tilde{\beta_s}] [\sigma_Q] [\sigma_{\beta_0}] [\sigma_{\beta_s}] [\sigma]
@@
   aggsurv <- agg_fxn(survmcmc)
-  cap <- paste("Statistical results from IPM survival model for", spp_name[i])
+  cap <- paste("Statistical results from IPM survival model for\\emph{", spp_name[i],"}")
   print(xtable(aggsurv, caption = cap), comment=FALSE, include.rownames=FALSE, 
@@
   aggsurv <- agg_fxn(survmcmc)
-  cap <- paste("Statistical results from IPM growth model for", spp_name[i])
+  cap <- paste("Statistical results from IPM growth model for\\emph{", spp_name[i],"}")
   print(xtable(aggsurv, caption = cap), comment=FALSE, include.rownames=FALSE, 
@@
   aggsurv <- agg_fxn(survmcmc)
-  cap <- paste("Statistical results from IPM recruitment model for", spp_name[i])
+  cap <- paste("Statistical results from IPM recruitment model for\\emph{", spp_name[i],"}")
   print(xtable(aggsurv, caption = cap), comment=FALSE, include.rownames=FALSE, 
@@
   aggsurv <- agg_fxn(survmcmc)
-  cap <- paste("Statistical results from QBM population model for", spp_name[i])
+  cap <- paste("Statistical results from QBM population model for\\emph{", spp_name[i],"}")
   print(xtable(aggsurv, caption = cap), comment=FALSE, include.rownames=FALSE, 

### 2016-06-06 manuscript/supporting_information.Rmd
updates some text

@@
 & \qquad & \boldsymbol{\beta}_c \sim \text{Normal}(0, \sigma_c^2 \textbf{I}) \\
-& \qquad & \beta_{d,1} \sim \text{Normal}(0, 100^2) \\
-& \qquad & \beta_{d,2} \sim \text{Normal}(0, 100^2) \\
-& \qquad & \tilde{\beta_0} \sim \text{Normal}(0, 100^2) \\
-& \qquad & \tilde{\beta_s} \sim \text{Normal}(0, 100^2) \\
+& \qquad & \beta_{d,1} \sim \text{Normal}(0, 100) \\
+& \qquad & \beta_{d,2} \sim \text{Normal}(0, 100) \\
+& \qquad & \tilde{\beta_0} \sim \text{Normal}(0, 100) \\
+& \qquad & \tilde{\beta_s} \sim \text{Normal}(0, 100) \\
 & \qquad & \sigma_Q \sim \text{Cauchy}(0, 5) \\
@@
 **Stan code for model**
-```{r survival_stan, echo=TRUE, eval=FALSE}
+```{r survival_stan, echo=FALSE, eval=FALSE}
 data{
@@
 & \qquad & \boldsymbol{\beta}_c \sim \text{Normal}(0, \sigma_c^2 \textbf{I}) \\
-& \qquad & \beta_{d,1} \sim \text{Normal}(0, 10^2) \\
-& \qquad & \beta_{d,2} \sim \text{Normal}(0, 10^2) \\
-& \qquad & \tilde{\beta_0} \sim \text{Normal}(0, 100^2) \\
-& \qquad & \tilde{\beta_s} \sim \text{Normal}(0, 100^2) \\
+& \qquad & \beta_{d,1} \sim \text{Normal}(0, 100) \\
+& \qquad & \beta_{d,2} \sim \text{Normal}(0, 100) \\
+& \qquad & \tilde{\beta_0} \sim \text{Normal}(0, 100) \\
+& \qquad & \tilde{\beta_s} \sim \text{Normal}(0, 100) \\
 & \qquad & \sigma_Q \sim \text{Cauchy}(0, 2) \\
-& \qquad & \sigma_{\beta_0} \sim \text{Cauchy}(0, 2) \\
-& \qquad & \sigma_{\beta_s} \sim \text{Cauchy}(0, 2) \\
-& \qquad & a \sim \text{Normal}(0, 100^2) \\
-& \qquad & b \sim \text{Normal}(0, 100^2)
+& \qquad & \sigma_{\beta_0} \sim \text{Cauchy}(0, 5) \\
+& \qquad & \sigma_{\beta_s} \sim \text{Cauchy}(0, 5) \\
+& \qquad & a \sim \text{Normal}(0, 100) \\
+& \qquad & b \sim \text{Normal}(0, 100)
 \end{align}
@@
 **Stan code for model**
-```{r grow_stan, echo=TRUE, eval=FALSE}
+```{r grow_stan, echo=FALSE, eval=FALSE}
 data{
@@
 & \qquad & \boldsymbol{\beta}_c \sim \text{Normal}(0, \sigma_c^2 \textbf{I}) \\
-& \qquad & \beta_{d} \sim \text{Uniform}(-100, 100) \\
-& \qquad & \tilde{\beta_0} \sim \text{Normal}(0, 1000^2) \\
+& \qquad & \beta_{d} \sim \text{Uniform}(-10, 10) \\
+& \qquad & \tilde{\beta_0} \sim \text{Normal}(0, 100) \\
 & \qquad & \sigma_Q \sim \text{Cauchy}(0, 5) \\
 & \qquad & \sigma_{\beta_0} \sim \text{Cauchy}(0, 5) \\
-& \qquad & \phi \sim \text{Uniform}(0, 100) \\
+& \qquad & \phi \sim \text{Uniform}(0, 10) \\
 & \qquad & u \sim \text{Uniform}(0, 1)
@@
 **Stan code for model**
-```{r rec_stan, echo=TRUE, eval=FALSE}
+```{r rec_stan, echo=FALSE, eval=FALSE}
 data{
@@
 & \qquad & \boldsymbol{\beta}_c \sim \text{Normal}(0, \sigma_c^2 \textbf{I}) \\
-& \qquad & \tilde{\beta_0} \sim \text{Normal}(0, 1000^2) \\
-& \qquad & \tilde{\beta_s} \sim \text{Normal}(0, 1000^2) \\
+& \qquad & \tilde{\beta_0} \sim \text{Normal}(0, 100) \\
+& \qquad & \tilde{\beta_s} \sim \text{Normal}(0, 100) \\
 & \qquad & \sigma_Q \sim \text{Cauchy}(0, 5) \\
@@
 & \qquad & \sigma_{\beta_s} \sim \text{Cauchy}(0, 5) \\
-& \qquad & \sigma \sim \text{Cauchy}(0, 5)
+& \qquad & \sigma^2 \sim \text{Inverse Gamma}(0.001, 0.001)
 \end{align}
@@
 **Stan code for model**
-```{r qbm_stan, echo=TRUE, eval=FALSE}
+```{r qbm_stan, echo=FALSE, eval=FALSE}
 data{

### 2016-06-16 manuscript/manuscript.Rmd
updates LOYO chains after getting rid of bacd chains and iterations

@@
 We assessed convergence visually and checked that scale reduction factors for all parameters were less than 1.1.
+There were six instances where the scale reduction factor for a particular parameter was greater than 1.1.
+In those cases, we checked the traceplots of the violating parameter and discovered that one chain, out of three, was poorly behaved.
+Therefore, we removed that chain from the analysis, leaving two well-mixed and converged chains for those six parameters.
 For the purposes of including parameter uncertainty in our population models, we retained the final 1,000 iterations from each of the three MCMC chains to be used as randomly drawn values during population simulation.

### 2016-06-16 analysis/check_all_rhats.R

@@ -103,12 +103,13 @@ dev.off()
 ####
 ####  REMOVE BAD CHAINS AND RESAVE
 ####
-chains_to_remove <- c(1,1,1,1,3,3) # based on visual inspection of traceplots
+chains_to_remove <- c(1,1,1,1,1,3,3) # based on visual inspection of traceplots
 pdf("new_traceplots.pdf")
 for(i in 1:length(files)){
   splits <- strsplit(files[i], "_")
   if(splits[[1]][1] == "recruitment") { tmp_dir <- "./ipm/vitalRateRegs/validation/recruitment/fits/" }
-  if(splits[[1]][1] == "popgrowth") { tmp_dir <- "./quadBM/vitalRateRegressions/truncNormModel/validation/fits_noclimate/" }
+  if(splits[[1]][1] == "popgrowth" & length(splits[[1]])==5) { tmp_dir <- "./quadBM/vitalRateRegressions/truncNormModel/validation/fits_noclimate/" }
+  if(splits[[1]][1] == "popgrowth" & length(splits[[1]])==4) { tmp_dir <- "./quadBM/vitalRateRegressions/truncNormModel/validation/fits/" }
   tmp_mcmc <- readRDS(paste0(tmp_dir,files[i]))
   tmp2 <- subset(tmp_mcmc, Chain!=chains_to_remove[i])
   saveRDS(tmp2, file=paste0(tmp_dir,files[i]))
@@ -118,3 +119,14 @@ for(i in 1:length(files)){
   print(gprint)
 }
 dev.off()
+
+
+##  Remove end of chains for HECO, climate, 1932; based on visual inspection
+tmp_mcmc <-readRDS("./quadBM/vitalRateRegressions/truncNormModel/validation/fits/popgrowth_stanmcmc_HECO_leaveout1938.RDS")
+tmp_mcmc2 <- subset(tmp_mcmc, Iteration<751)
+ggplot(subset(tmp_mcmc2, Parameter == "b2[4]"))+
+  geom_line(aes(x=Iteration, y=value, color=as.factor(Chain)), alpha=0.5)
+
+saveRDS(tmp_mcmc2, "./quadBM/vitalRateRegressions/truncNormModel/validation/fits/popgrowth_stanmcmc_HECO_leaveout1938.RDS")
+
+

### 2016-06-23 manuscript/manuscript.Rmd
small starts to revisions

@@
    - \usepackage{times}
+   - \usepackage{enumitem}
 bibliography: ~/Dropbox/Bibliography/MicroMeso.bib
@@
 
-**1.** Rapid environmental change has generated growing interest in forecasts of future population trajectories.
+\begin{enumerate}[label=\textbf{\arabic*}]
+        \item Rapid environmental change has generated growing interest in forecasts of future population trajectories.
 Traditional population models built with detailed demographic observations from one study site can address the impacts of environmental change at particular locations, but are difficult to scale up to the landscape and regional scales relevant to management decisions.
-
-**2.** An alternative is to build models using population-level data which are much easier to collect than individual-level data over broad spatial scales.
+        \item An alternative is to build models using population-level data which are much easier to collect than individual-level data over broad spatial scales.
 However, it is unknown whether models built using population-level data adequately capture the effects of density-dependence and environmental forcing that are necessary to generate skillful forecasts.
-
-**3.** Here, we test the consequences of aggregating individual responses when forecasting the population states and trajectories of four perennial grass species in a semi-arid grassland in Montana, USA.
+        \item Here, we test the consequences of aggregating individual responses when forecasting the population states and trajectories of four perennial grass species in a semi-arid grassland in Montana, USA.
 We parameterized two population models for each species, one based on individual-level data (survival, growth and recruitment) and one on population-level data (percent cover), and compared their forecasting skill and forecast horizons with and without the inclusion of climate covariates.
 For both models we used Bayesian ridge regression to weight the influence of climate covariates for optimal prediction.
-
-**4.** In the absence of climate effects, we found no significant difference between the forecasting skill of models based on individual-level data and models based on population-level data.
-Climate effects were weak and only marginally increased forecasting skill.
-Increases in skill with climate covariates were similar between model types for three of the four species.
-For the fourth species, forecast accuracy of the individual-level model with climate covariates was significantly higher than the accuracy of the equivalent population-level model.
-
-**5.** *Synthesis.* For our focal species at this particular location, and using our particular statistical models, percent cover models generated forecasts as skillful as those from a demographic model.
-However, for one species, forecast skill increased when we linked demographic data to climate covariates.
-We conclude that models based on aggregated individual-level data offer a practical alternative to data-intensive demographic models when species do not respond strongly to interannual variation in weather, but when species do respond to climate drivers, demographically-based models can generate more skillful forecasts.
+        \item In the absence of climate effects, we found no significant difference between the forecasting skill of models based on individual-level data and models based on population-level data.
+Climate effects were weak, but increased forecasting skill for two species.
+Increases in skill with climate covariates were similar between model types.
+        \item For our focal species at this particular location, and using our particular statistical models, percent cover models generated forecasts as skillful as those from a demographic model.
+We conclude that models based on aggregated individual-level data offer a practical alternative to data-intensive demographic models.
+    \end{enumerate}
 
@@
 All of our analyses (model fitting and simulating) were conducted in R [@R2013].
-We used the 'No-U-Turn' MCMC sampler in Stan [@stan2014]  to estimate the posterior distributions of model parameters using the package 'rstan' [@rstan2014].
+We used the 'No-U-Turn' Hamiltonian Monte Carlo sampler in Stan [@stan2014]  to estimate the posterior distributions of model parameters using the package `rstan` [@rstan2014].
 We obtained posterior distributions for all model parameters from three parallel MCMC chains run for 1,000 iterations after discarding an initial 1,000 iterations.
@@
 -------
-The IPM and QBM generated one-step-ahead forecasts of similar skill for out-of-sample observations, with an average correlation between predictions and observations ($\rho$) of 0.73 across all models and species (Fig.
+The IPM and QBM generated one-step-ahead forecasts of similar skill for out-of-sample observations, with an average correlation between predictions and observations ($\rho$) of 0.72 across all models and species (Fig.
 2).
@@
 
-Despite the weak climate effects, including climate covariates did increase the accuracy of forecasts for all species except *P. smithii* (Fig. 2).
-Increases in accuracy due to the inclusion of climate covariates were not significant (*P* > 0.05 for all comparisons of $\rho$ between climate and no-climate forecasts within model types; Fig. 2).
-However, the IPM with climate covariates for *Poa secunda* did have significantly lower error (mean absolute error) than the IPM without climate covariates ($t_{(196)}$ = -1.84, *P* = 0.033; Fig. S2).
+Despite the weak climate effects, including climate covariates did increase the accuracy of forecasts for two species: *B. gracilis* and *Poa secunda* (Fig. 2).
+Only for *B. gracilis* were the skill increases significant for the IPM ($t_{(279)}$ = 1.70, *P* = 0.045) and the QBM ($t_{(279)}$ = 1.80, *P* = 0.037).
+Similarly, forecast error decreased significantly with the inclusion of climate covariates for the *B.
+gracilis* IPM ($t_{(280)}$ = -3.72, *P* = 0.029) and QBM ($t_{(280)}$ = -3.34, *P* < 0.0001), and for the *Poa secunda* IPM ($t_{(196)}$ = -1.90, *P* < 0.0001) and QBM ($t_{(196)}$ = -2.47, *P* = 0.007) (Fig.
+S2, Supporting Information).
 IPM forecasts were significantly more accurate than the QBM in only one case (Fig.
-2): forecast accuracy of *Poa secunda* percent cover from an IPM with climate covariates was greater than the accuracy from the QBM with climate covariates ($t_{(195)}$ = 1.72, *P* = 0.043).
-Forecasts from IPMs with climate covariates had significantly lower error than equivalent QBM forecasts for *P.
-smithii* ($t_{(216)}$ = -3.49, *P* < 0.001) and *Poa secunda* ($t_{(196)}$ = -1.83, *P* = 0.034) (Fig.
-S2, Supporting Information).
+2): forecast accuracy of *P.
+smithii* percent cover from an IPM with climate covariates was greater than the accuracy from the QBM with climate covariates ($t_{(215)}$ = 1.92, *P* = 0.028).
+However, including climate covariates in the IPM did not significantly increase forecast skill for *P. smithii* (Fig. 2).
 Results from all pairwise statistical tests are shown in Table S21 of the Supporting Information.
@@
 With climate covariates included, the accuracy of both models' forecasts declined as the forecast horizon increased, but they did so at similar rates (Fig. 4).
-The only exception is for *Poa secunda*, where IPM forecast accuracy remained steady as the forecast horizon increases, after an initial decrease from a forecast horizon of one year, whereas QBM forecast accuracy declined (Fig.
-4).
+The only exception is for *Poa secunda*, where QBM forecast accuracy remained steady as the forecast horizon increases, whereas IPM forecast accuracy declined (Fig. 4).
 
@@
 Thus, estimates of climate effects on plant population growth may be biased or non-identifiable when the underlying statistical model is fit using population-level data that integrates over the potentially unique climate responses of individual vital rates.
-However, we found that IPM forecast accuracy was only significantly higher than the QBM for one species, *Poa secunda* (Fig. 2, `climate' bars).
-This result likely stems from the relatively weak climate effects in the vital rate regressions (Fig. 3).
+However, we found that IPM forecast accuracy was only significantly higher than the QBM for one species, *P. smithii* (Fig. 2, `climate' bars).
+That the IPM outperformed the QBM with climate covariates included for *P.
+smithii* is a moot point, however, because including the climate covariates in the IPM did not increase forecast accuracy over an IPM without climate covariates (Fig.
+2).
+
+<!---
+The equivalency of the IPM and QBM forecasts stems from the  (Fig. 3).
 If climate effects on vital rates are weak, then models based on percent cover or density should fair well compared to models based on individual-level data as long as density dependence is adequately estimated (see Fig.
@@
 4).
-Similar forecast horizons should be expected if both models adequately capture density dependence and environmental forcing is negligble, as is the case for our models (Fig.
+Similar forecast horizons should be expected if both models adequately capture density dependence and environmental forcing, as is the case for our models (Fig.
 2).
-In fact, only the forecast horizons for *Poa secunda* noticeably differ (Fig. 4), and that is also the only species for which the IPM outperformed the QBM when we included climate effects.
 
@@
 Our methodological approach could also potentially be used for backcasting paleoclimate and paleoecological aggregates such as pollen counts based on lake sediments [e.g., @Paciorek2009]
+--->
 

### 2016-06-23 FigureScripts/accuracy_error_boxplots.R

@@ -34,9 +34,7 @@ source("rho_comp_functions.R")
 ipm_onestep <- readRDS(paste0(path2ipm,"ipm_loyo_forecasts_combined.RDS"))
 torms <- which(is.na(ipm_onestep$cover.t1)==TRUE) # get rid of NAs; only a couple
 ipm_onestep <- ipm_onestep[-torms,]
-# Remove predictions >1 for fair comparison to [0,1] truncated quad model NOT IMPLEMENTED
-# resets <- which(ipm_onestep[,"cover.t1"]>1)
-# ipm_onestep[resets, "cover.t1"] <- 1
+
 # Average predictions over quad-year reps
 avg_ipm <- ddply(ipm_onestep, .(species, quad, t1), summarise,
                  avg_prediction = median(cover.t1),
@@ -44,8 +42,9 @@ avg_ipm <- ddply(ipm_onestep, .(species, quad, t1), summarise,
                  quant_dist = quantile(cover.t1*100, probs = 0.95) - quantile(cover.t1, probs = 0.05))
 # Calculate error
 avg_ipm$error <- avg_ipm$observation*100 - avg_ipm$avg_prediction*100
-# Aggregate over species
-ipmstats <- ddply(avg_ipm, .(species), summarise,
+
+# Aggregate by species and t1
+ipmstats <- ddply(avg_ipm, .(species, t1), summarise,
                   mae = mean(abs(error)),
                   rho = cor(observation, avg_prediction, method = "p"),
                   mean_cover = mean(observation*100),
@@ -66,8 +65,9 @@ avg_qbm <- ddply(qbm_onestep, .(species, quad, year), summarise,
                  quant_dist = quantile(pred_cover.t1*100, probs = 0.95) - quantile(pred_cover.t1, probs = 0.05))
 # Calculate error
 avg_qbm$error <- avg_qbm$observation*100 - avg_qbm$avg_prediction*100
-# Aggregate over species
-qbmstats <- ddply(avg_qbm, .(species), summarise,
+
+# Aggregate by species
+qbmstats <- ddply(avg_qbm, .(species, year), summarise,
                   mae = mean(abs(error)),
                   rho = cor(observation, avg_prediction, method = "p"),
                   mean_cover = mean(observation*100),
@@ -85,31 +85,6 @@ test_df <- merge(testq, testi)
 if(mean(with(test_df, obs_cover.t1-obs.cover.t1))!=0) { stop("observed values not matching up") }
 
 
[…]

### 2016-06-23 manuscript/manuscript.Rmd
adds stats tests for climate effects reducing skill

@@
 Despite the weak climate effects, including climate covariates did increase the accuracy of forecasts for two species: *B. gracilis* and *Poa secunda* (Fig. 2).
+In no case did including climate covariates significantly decrease forecast skill (Table S21), despite small changes in the mean skill (Fig. 2)
 Only for *B. gracilis* were the skill increases significant for the IPM ($t_{(279)}$ = 1.70, *P* = 0.045) and the QBM ($t_{(279)}$ = 1.80, *P* = 0.037).
@@
 However, including climate covariates in the IPM did not significantly increase forecast skill for *P. smithii* (Fig. 2).
-Results from all pairwise statistical tests are shown in Table S21 of the Supporting Information.
+Results from all pairwise statistical tests are shown in Table S22 of the Supporting Information.
 

### 2016-06-23 FigureScripts/one_step_results_table.R

@@ -348,3 +348,83 @@ print_comparison_table <- function()
 }
 
 print_comparison_table()
+
+
+
+####
+####  SIGNIFICANCE TESTS FOR CLIMATE REDUCING SKILL (SI MATERIAL)
+####
+##  Function from Ye et al. ---
+print_comparison_table_opposite <- function()
+{
+  compute_p_values <- function(x1, x2, y)
+  {
+    index <- is.finite(x1) & is.finite(x2) & is.finite(y)
+    x1 <- x1[index]
+    x2 <- x2[index]
+    y <- y[index]
+    err1 <- abs(y - x1)
+    err2 <- abs(y - x2)
+    mae_ttest <- t.test(err1, err2, paired = TRUE, alternative = "greater")
+    mae_df <- mae_ttest$parameter
+    mae_statistic <- mae_ttest$statistic
+    mae_p <- mae_ttest$p.value
+    rho_ttest <- rho_comp(x2, x1, y)
+    rho_df <- rho_ttest$df
+    rho_statistic <- rho_ttest$statistic
+    rho_p <- rho_ttest$p.value
+    return(data.frame(mae_df, mae_statistic, mae_p, rho_df, rho_statistic, rho_p))
+  }
+  
+  preds <- readRDS("all_model_predictions.RDS")
+  
+  ## ONLY NORMALIZE IF NOT COMPARING BY SPECIES...
+  #   preds <- list(BOGR=subset(pred_df, species=="BOGR"),
+  #                 HECO=subset(pred_df, species=="HECO"),
+  #                 PASM=subset(pred_df, species=="PASM"),
+  #                 POSE=subset(pred_df, species=="POSE"))
+  #   # normalize by mean obs value
+  #   preds_n <- lapply(names(preds), function(spp) {
+  #     df <- preds[[spp]]
+  #     sigma <- sd(df$obs, na.rm = TRUE)
+  #     mu <- mean(df$obs, na.rm = TRUE)
+  #     df$obs <- (df$obs - sigma) / mu
+  #     df$ipm_dd <- (df$ipm_dd - sigma) / mu
+  #     df$ipm_clim <- (df$ipm_clim - sigma) / mu
+  #     df$qbm_dd <- (df$qbm_dd - sigma) / mu
+  #     df$qbm_clim <- (df$qbm_clim - sigma) / mu
+  #     df$species <- spp
+  #     return(df)
+  #   })
+  #   preds_n <- do.call(rbind, preds_n)
[…]

### 2016-06-24 manuscript/manuscript.Rmd
revised after new results; sent to PBA

@@
 
-\textsf{\LARGE{\textbf{Do we need demographic data to forecast the state of plant populations?}}}
+\textsf{\LARGE{\textbf{Do we need demographic data to forecast plant population dynamics?}}}
 
@@
 Geological Survey, Colorado Cooperative Fish and Wildlife Research Unit, Fort Collins, CO 80523, USA; \textsuperscript{3}Department of Fish, Wildlife, and Conservation Biology, Colorado State University, Fort Collins, CO 80523 USA}; \textsuperscript{4}Department of Statistics, Colorado State University, Fort Collins, CO 80523 USA}}
+
+Last compile: \today
 
@@
 For both model types, we fit vital rate models with and without climate covariates.
-Models with climate effects contain five climate covariates that we chose \emph{a priori} based on previous model selection efforts using these data (Chu et al.
-*in press*) and expert advice (Lance Vermeire, *personal communication*): "water year" precipitation at \emph{t}-2 (lagppt); April through June precipitation at \emph{t}-1 and \emph{t} (ppt1 and ppt2, respectively) and April through June temperature at \emph{t}-1 and \emph{t} (TmeanSpr1 and TmeanSpr2, respectively), where \emph{t}-1 to *t* is the transition of interest.
+Models with climate effects contain five climate covariates that we chose \emph{a priori} based on previous model selection efforts using these data [@Chu2016] and expert advice (Lance Vermeire, *personal communication*): "water year" precipitation at \emph{t}-2 (lagppt); April through June precipitation at \emph{t}-1 and \emph{t} (ppt1 and ppt2, respectively) and April through June temperature at \emph{t}-1 and \emph{t} (TmeanSpr1 and TmeanSpr2, respectively), where \emph{t}-1 to *t* is the transition of interest.
 We also include interactions among same-year climate covariates (e.g., ppt1 $\times$ TmeansSpr1), resulting in a total of seven climate covariates.
@@
 
-We used random draws from the final 1,000 iterations, thinned by 10, from each of three MCMC chains for each vital rate regression to carry-through parameter uncertainty into our population models.
-At each time step, we randomly selected climate covariates from one of the 14 observed years.
-Then, we drew the full parameter set (climate effects and density-dependence fixed effects) from a randomly selected MCMC iteration.
+We used random draws from the final 1,000 iterations from each of three MCMC chains for each vital rate regression to carry-through parameter uncertainty into our population models.
+At each time step, we drew the full parameter set (climate effects and density-dependence fixed effects) from a randomly selected MCMC iteration.
 Relatively unimportant climate covariates (those that broadly overlap 0) will have little effect on the mean of the simulation results, but can contribute to their variation.
-Our focus was on the contribution of density dependence and climate covariates to population states, thus we set the random year effects and the random group effects to zero.
+To retain temporal variation associated with random year effects, we used posterior estimates of the mean temporal effect and the standard deviation of that effect to generate a random year effect for unobserved years.
+That is, for some future year *T*, the intercept is $\beta_0,T \sim \text{Normal}(\beta_0, \sigma_{\beta_0})$ and the effect of size is $\beta_s,T \sim \text{Normal}(\beta_s, \sigma_{\beta_s})$.
 
@@
 We project the model forward by drawing a new parameter set (unique to climate year and MCMC iteration) at each timestep.
-As with the IPM, random year effects are ommitted in our simulations.
+Random year effects are included as described above for the IPM.
 
@@
 We repeated this procedure for all 13 observation years, making 100 one-step-ahead forecasts for each quadrat-year combination with parameter uncertainty included via random draw from the MCMC chain as described above.
-Year-specific intercepts for left-out data were drawn from the posterior distribution of the mean intercept.
-Thus, for some future year *T*, $\beta_0,T \sim \text{Normal}(\beta_0, \sigma_{\beta_0})$.
+As described above, year-specific parameters for left-out data were drawn from the posterior distribution of the mean intercept.
 
@@
 Despite the weak climate effects, including climate covariates did increase the accuracy of forecasts for two species: *B. gracilis* and *Poa secunda* (Fig. 2).
-In no case did including climate covariates significantly decrease forecast skill (Table S21), despite small changes in the mean skill (Fig. 2)
-Only for *B. gracilis* were the skill increases significant for the IPM ($t_{(279)}$ = 1.70, *P* = 0.045) and the QBM ($t_{(279)}$ = 1.80, *P* = 0.037).
+However, only for *B. gracilis* were the skill increases statistically significant at $\alpha$ = 0.05 for the IPM ($t_{(279)}$ = 1.70, *P* = 0.045) and the QBM ($t_{(279)}$ = 1.80, *P* = 0.037).
 Similarly, forecast error decreased significantly with the inclusion of climate covariates for the *B.
@@
 S2, Supporting Information).
+In no case did including climate covariates significantly decrease forecast skill (Table S21), despite small changes in the mean skill (Fig. 2).
+
 IPM forecasts were significantly more accurate than the QBM in only one case (Fig.
@@
 Thus, estimates of climate effects on plant population growth may be biased or non-identifiable when the underlying statistical model is fit using population-level data that integrates over the potentially unique climate responses of individual vital rates.
-However, we found that IPM forecast accuracy was only significantly higher than the QBM for one species, *P. smithii* (Fig. 2, `climate' bars).
-That the IPM outperformed the QBM with climate covariates included for *P.
-smithii* is a moot point, however, because including the climate covariates in the IPM did not increase forecast accuracy over an IPM without climate covariates (Fig.
+We found some evidence that the QBM missed important climate effects for three species (*B.
+gracilis*, *H.
+comata*, and *Poa secunda*), where important climate effects were identified in the individual vital rate models but not in the percent cover model (Fig.
+3).
+However, for the two species where including climate covariates increased forecast accuracy (*B.
+gracilis* and *Poa secunda*), forecast skill (Fig.
+2) and error (Fig.
+S2) were equivalent between the IPM and QBM.
+We did find that IPM forecast accuracy was significantly higher than the QBM for one species, *P. smithii* (Fig. 2, `climate' bars).
+But, the fact that *P.
+smithii* IPM outperformed the QBM with climate covariates included is a moot point because including the climate covariates in the IPM did not increase forecast accuracy over an IPM without climate covariates (Fig.
 2).
 
+The higher accuracy of the IPM and QBM with climate covariates for *B.
+gracilis* and *Poa secunda*  highlights the advantage of contemporary modeling and variable selection approaches such as ridge regression and LASSO over techniques that would exclude "non-significant" effects from final models.
+Ridge regression allows researchers to retain covariates whose effects may be difficult to identify in noisy data or short time series.
+This is especially important when forecasting the impacts of climate change, where it is important to include to effects of forcing variables (e.g., temperature and precipitation) even of such effects are difficult to identify.
+Indeed, we failed to detect strong climate effects in the QBM for *B. gracilis* and *Poa secunda*, but including climate covariates still improved forecasting skill (Fig. 2).
+If a species is truly unresponsive to a given climate variable, statistical regularization techniques will shrink the mean and variance of a covariate estimate toward zero [@Hooten2015].
+Of course, no matter what model selection approach is adopted, a critical step is identifying the appropriate candidate covariates, which we attempted to do based on our knowledge of this semi-arid plant community.
+However, the climate covariates we chose required aggregating daily weather data over discrete time periods.
+It is possible that we did not choose the optimal time periods over which to aggregate.
+New methods using functional linear models (or splines) may offer a data-driven approach for identifying the appropriate time periods over which to aggregate to produce a tractable set of candidate climate variables [@Teller2016].
+
+We also expected IPM forecast accuracy to decline at a lower rate than the QBM as the forecast horizon increased.
+In principle, more mechanistic models should produce better predictions, especially under novel conditions [@Evans2012b; @Schindler2015b].
+In our case, the IPM explicitly models the influence of weather on recruitment and survival, effects that may be poorly represented in the QBM because recruitment and survival mainly affect small plants that contribute little to year-to-year changes in percent cover.
+Over time, of course, the addition and subtraction of small plants can have large effects on population growth, so explicitly modeling these effects could contribute to a longer forecast horizon. 
+However, we found no evidence for a difference between the IPM and QBM forecast horizons (Fig.
+4).
+Similar forecast horizons should be expected if both models adequately capture density dependence and environmental forcing, as is the case for our models (Fig.
+2).
+
+In conclusion, we found that models based on individual-level demographic data generally failed to generate more skillful population forecasts than models based on population-level data.
+This finding runs counter to our expectations, but is consistent with recent theoretical [@Freckleton2011] and empirical work [@Queenborough2011].
+Moreover, for the two species where including climate covariates improved forecast skill, both models generated equally skillful one-step-ahead forecasts.
+Thus, we conclude that models based on population-level data, rather than individual-level data, may be adequate for forecasting the states and dynamics of plant populations.
+Unfortunately, our analysis, where climate effects were relatively unimportant in vital rate regressions, did not allow us to sufficiently test our prediction that individual-level data is neccessary to generate skillful forecasts if different vital rates respond to climate in unique, potentially opposing, ways.
+Nonetheless, our results are encouraging for the use of easy-to-collect population-level data for forecasting the state of plant populations.
+Our methodological approach could also potentially be used for backcasting paleoclimate and paleoecological aggregates such as pollen counts based on lake sediments [e.g., @Paciorek2009]
+
 <!---
+###################### OLD TEXT; COMMENTED OUT.
+
 The equivalency of the IPM and QBM forecasts stems from the  (Fig. 3).
@@
 
-The higher accuracy of the IPM with climate covariates for *Poa secunda*  highlights the advantage of contemporary modeling and variable selection approaches such as ridge regression and LASSO over techniques that would exclude "non-significant" effects from final models.
-Ridge regression allows researchers to retain covariates whose effects may be difficult to identify in noisy data or short time series.
-This is especially important when forecasting the impacts of climate change, where it is important to include to effects of forcing variables (e.g., temperature and precipitation) even of such effects are difficult to identify.
-If a species is truly unresponsive to a given climate variable, statistical regularization techniques will shrink the mean and variance of a covariate estimate toward zero [@Hooten2015].
-Of course, no matter what model selection approach is adopted, a critical step is identifying the appropriate candidate covariates, which we attempted to do based on our knowledge of this semi-arid plant community.
-However, the climate covariates we chose required aggregating daily weather data over discrete time periods.
-It is possible that we did not choose the optimal time periods over which to aggregate.
-New methods using functional linear models (or splines) may offer a data-driven approach for identifying the appropriate time periods over which to aggregate to produce a tractable set of candidate climate variables [@Teller2016].
-
-We also expected IPM forecast accuracy to decline at a lower rate than the QBM as the forecast horizon increased.
-In principle, more mechanistic models should produce better predictions, especially under novel conditions [@Evans2012b; @Schindler2015b].
-In our case, the IPM explicitly models the influence of weather on recruitment and survival, effects that may be poorly represented in the QBM because recruitment and survival mainly affect small plants that contribute little to year-to-year changes in percent cover.
-Over time, of course, the addition and subtraction of small plants can have large effects on population growth, so explicitly modeling these effects could contribute to a longer forecast horizon. 
-However, we found no evidence for a difference between the IPM and QBM forecast horizons (Fig.
-4).
-Similar forecast horizons should be expected if both models adequately capture density dependence and environmental forcing, as is the case for our models (Fig.
-2).
-
-In conclusion, we found that models based on individual-level demographic data generally failed to generate more skillful population forecasts than models based on population-level data.
-This finding runs counter to our expectations, but is consistent with recent theoretical [@Freckleton2011] and empirical work [@Queenborough2011].
-However, we did achieve more skillful forecasts from the IPM with climate covariates for one species, *Poa secunda*, that appears to respond consistently to several climate variables.
-Thus, we conclude that models based on population-level data, rather than individual-level data, may be adequate for forecasting the state of plant populations for species that do not respond consistently or strongly to climate.
-Unfortunately, our analysis, where climate effects were relatively unimportant in vital rate regressions, did not allow us to sufficiently test our prediction that individual-level data is neccessary to generate skillful forecasts if different vital rates respond to climate in unique, potentially opposing, ways.
-Nonetheless, our results are encouraging for the use of easy-to-collect population-level data for forecasting the state of plant populations.
-Our methodological approach could also potentially be used for backcasting paleoclimate and paleoecological aggregates such as pollen counts based on lake sediments [e.g., @Paciorek2009]
+##################### END OF OLD TEXT
 --->
@@
       \includegraphics[height=6in]{./components/forecast_accuracy_mockup.png}
-  \caption{Comparison of one-step-ahead, out-of-sample forecast accuracy between the IPM and QBM models with and without the inclusion of climate covariates.
-Comparisons between equivalent IPM and QBM models indicate no significant difference in accuracy (\emph{P} > 0.05 for all comparisons).
-Including climate covariates resulted in siginificantly higher forecast accuracy for only one species, \emph{Poa secunda} (\emph{P} = 0.043).
+  \caption{Comparisons of one-step-ahead, out-of-sample forecast accuracy between the IPM and QBM models with and without the inclusion of climate covariates.
+For each comparison, \emph{P}-values are from one-sided \emph{t} tests designed to assess whether the first model in the comparison statement had higher accuracy than the second model in the comparison statement (see details in Table S22).
+In no case did adding climate covariates decrease forecast accuracy (Table S21).
 Species codes are as in Fig.
-1}
+1.}
 \end{figure}
@@
 Points show the average accuracy ($\rho$) across all forecasts at a given time horizon.
-We only examine the forecast accuracy of models with climate covariates included because in no case did including climate covariates decrease accuracy (see Fig.
+We only examine the forecast accuracy of models with climate covariates included because in no case did including climate covariates significantly decrease accuracy (see Fig.
 2).

### 2016-06-24 manuscript/supporting_information.Rmd
revised after new results; sent to PBA

@@
 \centering
-\caption {Statistical tests of model comparisons.}
+\caption {Statistical tests to assess whether models without climate covariates have higher accuracy and lower error than models with climate covariates. t-tests are one-sided.}
+\begin{footnotesize}
+\begin{tabular}{lllrrrl}
+  \hline
+comparison & performance\_measure & test\_type & test\_statistic & df & p\_value & species \\ 
+  \hline
+simple IPM vs. climate IPM & rho & t & -1.703 & 279.000 & 0.955 & BOGR \\ 
+  simple IPM vs. climate IPM & MAE & t & -3.715 & 280.000 & 1.000 & BOGR \\ 
+  simple QBM vs. climate QBM & rho & t & -1.798 & 279.000 & 0.963 & BOGR \\ 
+  simple QBM vs. climate QBM & MAE & t & -3.340 & 280.000 & 1.000 & BOGR \\ 
+  simple IPM vs. climate IPM & rho & t & 0.246 & 169.000 & 0.403 & HECO \\ 
+  simple IPM vs. climate IPM & MAE & t & -1.472 & 170.000 & 0.929 & HECO \\ 
+  simple QBM vs. climate QBM & rho & t & 0.747 & 169.000 & 0.228 & HECO \\ 
+  simple QBM vs. climate QBM & MAE & t & 1.145 & 170.000 & 0.127 & HECO \\ 
+  simple IPM vs. climate IPM & rho & t & 0.201 & 215.000 & 0.421 & PASM \\ 
+  simple IPM vs. climate IPM & MAE & t & -0.592 & 216.000 & 0.723 & PASM \\ 
+  simple QBM vs. climate QBM & rho & t & 0.631 & 215.000 & 0.264 & PASM \\ 
+  simple QBM vs. climate QBM & MAE & t & 4.276 & 216.000 & 0.000 & PASM \\ 
+  simple IPM vs. climate IPM & rho & t & -1.072 & 195.000 & 0.858 & POSE \\ 
+  simple IPM vs. climate IPM & MAE & t & -1.911 & 196.000 & 0.971 & POSE \\ 
+  simple QBM vs. climate QBM & rho & t & -1.564 & 195.000 & 0.940 & POSE \\ 
+  simple QBM vs. climate QBM & MAE & t & -2.468 & 196.000 & 0.993 & POSE \\ 
+   \hline
+\end{tabular}
+\end{footnotesize}
+\end{table}
+
+\newpage{}
+
+\begin{table}[ht]
+\centering
+\caption {Statistical tests of model comparisons. Table shows results from one-sided t-tests for the particular comparison.}
 \begin{footnotesize}

### 2016-07-01 manuscript/manuscript.Rmd
adds PBA edits

@@
         \item For our focal species at this particular location, and using our particular statistical models, percent cover models generated forecasts as skillful as those from a demographic model.
-We conclude that models based on aggregated individual-level data offer a practical alternative to data-intensive demographic models.
+We conclude that models based on aggregated individual-level data offer a practical alternative to data-intensive demographic models when forecasting is the goal.
     \end{enumerate}
@@
 We compared the forecasting skill of statistical and population models based on aggregated, population-level data with the skill of models based on individual-level data.
-We used a demographic dataset that tracks the fates of individual plants from four species over 14 years to build two kinds of single-species population models, traditional models using individual growth, survival, and recruitment data and alternative models based on basal cover.
+We used a demographic dataset that tracks the fates of individual plants from four species over 14 years to build two kinds of single-species population models, traditional models using individual growth, survival, and recruitment data and alternative models based on population-level (basal cover) data.
 We simulated the models to answer two questions motivated by the fact that the effects of intraspecific competition (density dependence) and interannual weather variability act at the level of the individual [@Clark2011b].
@@
 For the quadrat-level data we simply sum individual basal cover for each quadrat by species.
-This is equivalent to a near-perfect census of quadrat percent cover because measurement error at the individual-level is small [@Chu2014].
+This is equivalent to a near-perfect census of quadrat percent cover because measurement error at the individual-level is small [@Chu2015].
 Based on these two datasets of 13 year-to-year transitions, we can compare population models built using individual-level data and aggregated, quadrat-level data.
@@
 
-where $p$ is a mixing fraction between 0 and 1 that is estimated when fitting model.
+where $p$ is a mixing fraction between 0 and 1 that is estimated when fitting the model.
 
@@
 smithii* percent cover from an IPM with climate covariates was greater than the accuracy from the QBM with climate covariates ($t_{(215)}$ = 1.92, *P* = 0.028).
-However, including climate covariates in the IPM did not significantly increase forecast skill for *P. smithii* (Fig. 2).
+However, adding climate covariates decreased the skill of both models, and the difference between the IPM and QBM emerges only because skill decreased less for the IPM than the QBM.
 Results from all pairwise statistical tests are shown in Table S22 of the Supporting Information.
@@
 2; `no-climate' bars).
-We expected the IPM to out-perform the QBM because the IPM includes more mechanistic detail on the perennial plant life cycle, but this was not the case, at least when we ignored environmental forcing.
-Such a finding confirms theoretical [@Freckleton2011] and empirical work [@Taylor2004; @Queenborough2011] showing that density-structured models can be useful surrogates for demographic models when the goal is to estimate or forecast population states over large spatial extents.
+This result overturned our expectation that the IPM would out-perform the QBM, thanks to its mechanistic representation of the perennial life cycle, and confirms theoretical [@Freckleton2011] and empirical work [@Taylor2004; @Queenborough2011] showing that density-structured models can be useful surrogates for demographic models when the goal is to estimate or forecast population states over large spatial extents.
 
-While the models did not differ in forecast accuracy when density-dependence was the only driver of population dynamics, we expected the inclusion of environmental forcing to reveal more differences between the models.
-We expected the IPM to outperform the QBM when we included climate covariates because interannual variation in weather can affect vital rates in different ways [@Dalgleish2011].
-Thus, estimates of climate effects on plant population growth may be biased or non-identifiable when the underlying statistical model is fit using population-level data that integrates over the potentially unique climate responses of individual vital rates.
-We found some evidence that the QBM missed important climate effects for three species (*B.
+We also expected the inclusion of environmental forcing to reveal more differences between the models.
+Interannual variation in weather can affect vital rates in different ways [@Dalgleish2011], so, estimates of climate effects on plant population growth may be biased or non-identifiable when the underlying statistical model is fit using population-level data that integrates over the potentially unique climate responses of individual vital rates.
+We found some evidence that the QBM failed to detect climate effects for three species (*B.
 gracilis*, *H.
@@
 S2) were equivalent between the IPM and QBM.
-We did find that IPM forecast accuracy was significantly higher than the QBM for one species, *P. smithii* (Fig. 2, `climate' bars).
-But, the fact that *P.
-smithii* IPM outperformed the QBM with climate covariates included is a moot point because including the climate covariates in the IPM did not increase forecast accuracy over an IPM without climate covariates (Fig.
-2).
 
@@
 Ridge regression allows researchers to retain covariates whose effects may be difficult to identify in noisy data or short time series.
-This is especially important when forecasting the impacts of climate change, where it is important to include to effects of forcing variables (e.g., temperature and precipitation) even of such effects are difficult to identify.
+This is especially important when forecasting the impacts of climate variability, where it is important to include to effects of forcing variables (e.g., temperature and precipitation) even of such effects are difficult to identify.
 Indeed, we failed to detect strong climate effects in the QBM for *B. gracilis* and *Poa secunda*, but including climate covariates still improved forecasting skill (Fig. 2).
@@
 Over time, of course, the addition and subtraction of small plants can have large effects on population growth, so explicitly modeling these effects could contribute to a longer forecast horizon. 
-However, we found no evidence for a difference between the IPM and QBM forecast horizons (Fig.
-4).
-Similar forecast horizons should be expected if both models adequately capture density dependence and environmental forcing, as is the case for our models (Fig.
-2).
+However, we found no evidence for a difference between the IPM and QBM forecast horizons (Fig. 4).
 
@@
 Nonetheless, our results are encouraging for the use of easy-to-collect population-level data for forecasting the state of plant populations.
-Our methodological approach could also potentially be used for backcasting paleoclimate and paleoecological aggregates such as pollen counts based on lake sediments [e.g., @Paciorek2009]
 
 <!---
-###################### OLD TEXT; COMMENTED OUT.
-
-The equivalency of the IPM and QBM forecasts stems from the  (Fig. 3).
-If climate effects on vital rates are weak, then models based on percent cover or density should fair well compared to models based on individual-level data as long as density dependence is adequately estimated (see Fig.
-2 and Freckleton et al.
-2011).
-
-Our naive expectation was that a single, relatively important climate effect in a vital rate regression would cause a discernable difference between IPM and QBM forecast skill because individual-level responses may be less identifiable from percent cover data [@Clark2012].
-We actually found the opposite.
-Survival or recruitment regressions for *B.
-gracilis*, *H.
-comata*, and *P.
-smithii* all included at least one relatively important climate effect, which we *a posteriori* defined as a standardized coefficient whose 80% credible interval does not overlap zero (Fig.
-3).
-Yet, only the forecast skill for *Poa secunda* differed between the IPM and QBM (Fig. 2).
-Our explanation is that while at least one vital regression for the other three species included a relatively strong climate effect, only *Poa secunda* had a vital rate regression with several smaller climate effects all trending in the same direction (Fig.
-3).
-Thus, it appears that the estimation of several small but consistent effects, rather than a single relatively large effect, can lead to increased forecast skill.
-
-##################### END OF OLD TEXT
+Our methodological approach could also potentially be used for backcasting paleoclimate and paleoecological aggregates such as pollen counts based on lake sediments [e.g., @Paciorek2009]
 --->

### 2016-07-01 manuscript/manuscript.Rmd
adds Britta comments; sent to MBH and PBA

@@
 The increasing statistical sophistication of population models [@Rees2009] makes them promising tools for predicting the impacts of environmental change on species persistence and abundance.
-But reconciling the scales at which population models are parameterized and the scales at which environmental changes play out remains a challenge [@Clark2010; @Freckleton2011; @Queenborough2011; @Clark2012].
+But reconciling the scales at which population models are parameterized with the scales at which environmental changes play out remains a challenge [@Clark2010; @Freckleton2011; @Queenborough2011; @Clark2012].
 Most population models are built using demographic data from a single study site because tracking the fates of individuals is so difficult.
@@
 Aggregate measures of population status, rather than individual performance, offer an intriguing alternative for modeling populations [@Clark2004; @Freckleton2011].
-Population-level data cannot provide inference about demographic mechanisms, but might be sufficient for modeling future population states, especially because such data are feasible to collect across broad spatial extents [e.g., @Queenborough2011].
+Population-level data cannot provide inference about demographic mechanisms, but might be sufficient for modeling future population states, especially because population-level data, such as plant percent cover, are feasible to collect across broad spatial extents [e.g., @Queenborough2011].
 The choice between individual and population-level data involves a difficult trade-off: while individual-level data are necessary for mechanistic models, population-level data enable models that can be applied over greater spatial and temporal extents.
@@
 
-In the above, $w_{i,t}$ is the crowding that genet _i_ in year _t_ experiences from conspecific neighbors in quadrat *q*.
+In equation 3, $w_{i,t}$ is the crowding that genet _i_ in year _t_ experiences from conspecific neighbors in quadrat *q*.
 The spatial scale over which conspecific neighbors exert influence on any genet is determined by $\delta$.

### 2016-07-06 manuscript/manuscript.Rmd
updates ms with minor edits from MBH

@@
 
-### Stastical models of vital rates
+### Statistical models of vital rates
 At both levels of inference (individual and quadrat), the building blocks of our population models are vital rate regressions.
@@
 \begin{align}
-y_{i,q,t+1}^G &\sim \text{Normal}(\mu_{i,q,t+1}, \sigma^2_{x_{i,q,t+1}}\textbf{I}), \\
+y_{i,q,t+1}^G &\sim \text{Normal}(\mu_{i,q,t+1}, \sigma^2_{x_{i,q,t+1}}), \\
 \mu_{i,q,t+1} &= \beta_{0,t} + \beta_{s,t}x_{i,q,t} + \beta_Q + \textbf{z}'_t \boldsymbol{\beta}_c + \beta_{d,1} w_{i,q,t} + \beta_{d,2} (x_{i,q,t}w_{i,q,t}),
@@
 We calculated the log pointwise predictive density (*lppd*) to score each model's ability to predict the left-out data [@Gelman2014].
-Thus, for training data $y_{train}$ and held-out data $y_{hold}$ at a given value of  $\sigma_\theta^2$ across all MCMC samples $s=1,2,...,S$ and all hold outs of data from year *t* to year *T*, and letting $\theta$ represent all unknowns, *lppd* is
+Thus, for training data $y_{\text{train}}$ and held-out data $y_{\text{hold}}$ at a given value of  $\sigma_\theta^2$ across all MCMC samples $s=1,2,...,S$ and all hold outs of data from year *t* to year *T*, and letting $\theta$ represent all unknowns, *lppd* is
 
@@
 \begin{align}
-\text{lppd}_{\text{CV}} = \sum_{t=1}^T \text{log}_e \int [y_{t,hold}|\theta][\theta|y_{train}]d\theta,
+\text{lppd}_{\text{CV}} = \sum_{t=1}^T \text{log}_e \int [y_{t,\text{hold}}|\theta][\theta|y_{\text{train}}]d\theta,
 \end{align}
@@
 \begin{align}
-\sum_{t=1}^T \text{log}_e \left(\frac{1}{S} \sum^S_{s=1}(y_{t,hold}|\theta_{ts})  \right).
+\sum_{t=1}^T \text{log}_e \left(\frac{1}{S} \sum^S_{s=1}(y_{t,\text{hold}}|\theta_{ts})  \right).
 \end{align}
@@
 To retain temporal variation associated with random year effects, we used posterior estimates of the mean temporal effect and the standard deviation of that effect to generate a random year effect for unobserved years.
-That is, for some future year *T*, the intercept is $\beta_0,T \sim \text{Normal}(\beta_0, \sigma_{\beta_0})$ and the effect of size is $\beta_s,T \sim \text{Normal}(\beta_s, \sigma_{\beta_s})$.
+That is, for some future year *T*, the intercept is $\beta_{0,T} \sim \text{Normal}(\beta_0, \sigma_{\beta_0})$ and the effect of size is $\beta_{s,T} \sim \text{Normal}(\beta_s, \sigma_{\beta_s})$.
 

### 2016-07-06 manuscript/manuscript.Rmd
updates stats notation as per MBH edits

@@
 y_{i,q,t}^{S} &\sim \text{Bernoulli}(s_{i,q,t}), \\
-\text{logit}(s_{i,q,t}) &= \beta_{0,t} + \beta_{s,t}x_{i,q,t} + \beta_Q + \textbf{z}'_t \boldsymbol{\beta}_c + \beta_{d,1} w_{i,t} + \beta_{d,2} (x_{i,q,t}w_{i,q,t}),
+\text{logit}(s_{i,q,t}) &= \beta_{0,t} + \beta_{s,t}x_{i,q,t} + \beta_{Q,q} + \textbf{z}'_t \boldsymbol{\beta}_c + \beta_{d,1} w_{i,t} + \beta_{d,2} (x_{i,q,t}w_{i,q,t}),
 \end{align}
@@
 
-where $x_{i,q,t}$ is the log of genet *i* basal area at time *t*, $\beta_{0,t}$ is a year specific intercept, $\beta_Q$ is the random effect of quadrat group location, $\beta_{s,t}$ is the year-specific slope parameter for size, **z** is a vector of *p* climate covariates specific to year *t*, $\boldsymbol{\beta}_c$ is a vector of fixed climate effects of length *p*, $\beta_{d,1}$ is the effect of intraspecific crowding experienced by the focal genet at time *t* ($w_{i,q,t}$), and $\beta_{d,2}$ is a size by crowding ($x_{i,q,t}w_{i,q,t}$) interaction effect.
+where $x_{i,q,t}$ is the log of genet *i* basal area at time *t*, $\beta_{0,t}$ is a year specific intercept, $\beta_{Q,q}$ is the random effect of the *q*th quadrat to account for spatial location, $\beta_{s,t}$ is the year-specific slope parameter for size, **z** is a vector of *p* climate covariates specific to year *t*, $\boldsymbol{\beta}_c$ is a vector of fixed climate effects of length *p*, $\beta_{d,1}$ is the effect of intraspecific crowding experienced by the focal genet at time *t* ($w_{i,q,t}$), and $\beta_{d,2}$ is a size by crowding ($x_{i,q,t}w_{i,q,t}$) interaction effect.
 
@@
 y_{i,q,t+1}^G &\sim \text{Normal}(\mu_{i,q,t+1}, \sigma^2_{x_{i,q,t+1}}), \\
-\mu_{i,q,t+1} &= \beta_{0,t} + \beta_{s,t}x_{i,q,t} + \beta_Q + \textbf{z}'_t \boldsymbol{\beta}_c + \beta_{d,1} w_{i,q,t} + \beta_{d,2} (x_{i,q,t}w_{i,q,t}),
+\mu_{i,q,t+1} &= \beta_{0,t} + \beta_{s,t}x_{i,q,t} + \beta_{Q,q} + \textbf{z}'_t \boldsymbol{\beta}_c + \beta_{d,1} w_{i,q,t} + \beta_{d,2} (x_{i,q,t}w_{i,q,t}),
 \end{align}
@@
 \begin{align}
-\lambda_{q,t+1} = \tilde{c}_{q,t} \, \text{exp}\left(\beta_{0,t} + \beta_Q + \textbf{z}'_t \boldsymbol{\beta}_c + \beta_{d}\sqrt{\tilde{c}_{q,t}}\right),
+\lambda_{q,t+1} = \tilde{c}_{q,t} \, \text{exp}\left(\beta_{0,t} + \beta_{Q,q} + \textbf{z}'_t \boldsymbol{\beta}_c + \beta_{d}\sqrt{\tilde{c}_{q,t}}\right),
 \end{align}
@@
 y_{q,t+1}^{P} &\sim \text{LogNormal}(\mu_{q,t+1}, \sigma^2)_{0}^{1}, \\
-\mu_{q,t+1} &= \beta_{0,t} + \beta_{s,t}x_{q,t} + \beta_Q + \textbf{z}'_t \boldsymbol{\beta}_c,
+\mu_{q,t+1} &= \beta_{0,t} + \beta_{s,t}x_{q,t} + \beta_{Q,q} + \textbf{z}'_t \boldsymbol{\beta}_c,
 \end{align}

### 2016-07-06 manuscript/supporting_information.Rmd
updates stats notation as per MBH edits

@@
 Following the full model expression we provide the Stan code that directly corresponds the statement of posterior and joint distributions.
-In our expressions of the posterior and joint distributions we use general probability notation where [.] represents some unspecified distribution.
+Note that Stan uses standard deviations ($\sigma$) to specify distributions rather than variances ($\sigma^2$)  In our expressions of the posterior and joint distributions we use general probability notation where [.] represents some unspecified distribution.
 We do this for clarity of model presentation.
@@
 | $\tilde{\beta_0}$ | global intercept | ``a_mu`` |
-| $\beta_{Q}$ | random effect of quadrat group *Q* | ``gint[Q]`` where ``Q`` is the quadrat group |
+| $\beta_{Q,q}$ | random effect of quadrat group *q* | ``gint[q]`` where ``q`` is the quadrat group |
 | $\beta_{s,t}$ | size effect for year *t* | ``b1[t]`` where ``t`` is the numeric year |
@@
 | $\tilde{\beta_0}$ | global intercept | ``a_mu`` |
-| $\beta_{Q}$ | random effect of quadrat group *Q* | ``gint[Q]`` where ``Q`` is the quadrat group |
+| $\beta_{Q,q}$ | random effect of quadrat group *q* | ``gint[q]`` where ``q`` is the quadrat group |
 | $\beta_{d,1}$ | effect of plot cover | ``dd`` |
@@
 | $\tilde{\beta_0}$ | global intercept | ``a_mu`` |
-| $\beta_{Q}$ | random effect of quadrat group *Q* | ``gint[Q]`` where ``Q`` is the quadrat group |
+| $\beta_{Q,q}$ | random effect of quadrat group *q* | ``gint[q]`` where ``q`` is the quadrat group |
 | $\beta_{s,t}$ | cover effect for year *t* | ``b1[t]`` where ``t`` is the numeric year |
@@
 y_{i,q,t}^{S} &\sim \text{Bernoulli}(s_{i,q,t}) \\
-\text{logit}(s_{i,q,t}) &= \beta_{0,t} + \beta_{s,t}x_{i,q,t} + \beta_Q + \textbf{z}'_t \boldsymbol{\beta}_c + \beta_{d,1} w_{i,t} + \beta_{d,2} (x_{i,q,t}w_{i,q,t})
+\text{logit}(s_{i,q,t}) &= \beta_{0,t} + \beta_{s,t}x_{i,q,t} + \beta_{Q,q} + \textbf{z}'_t \boldsymbol{\beta}_c + \beta_{d,1} w_{i,t} + \beta_{d,2} (x_{i,q,t}w_{i,q,t})
 \end{align}
 
-where $x_{i,q,t}$ is the log of genet *i* basal area at time *t*, $\beta_{0,t}$ is a year specific intercept, $\beta_Q$ is the random effect of quadrat group location, $\beta_{s,t}$ is the year-specific slope parameter for size, **z** is a vector of *p* climate covariates specific to year *t*, $\boldsymbol{\beta}_c$ is a vector of fixed climate effects of length *p*, $\beta_{d,1}$ is the effect of intraspecific crowding experienced by the focal genet at time *t* ($w_{i,q,t}$), and $\beta_{d,2}$ is a size by crowding ($x_{i,q,t}w_{i,q,t}$) interaction effect.
+where $x_{i,q,t}$ is the log of genet *i* basal area at time *t*, $\beta_{0,t}$ is a year specific intercept, $\beta_{Q,q}$ is the random effect for the *q*th quadrat to account for spatial variation, $\beta_{s,t}$ is the year-specific slope parameter for size, **z** is a vector of *p* climate covariates specific to year *t*, $\boldsymbol{\beta}_c$ is a vector of fixed climate effects of length *p*, $\beta_{d,1}$ is the effect of intraspecific crowding experienced by the focal genet at time *t* ($w_{i,q,t}$), and $\beta_{d,2}$ is a size by crowding ($x_{i,q,t}w_{i,q,t}$) interaction effect.
 
@@
 \begin{align}
-&\mathtt{data} \qquad &y_{i,q,t}^{S} \sim \text{Bernoulli}(s_{i,q,t}) \\
-&\mathtt{process} \qquad &\text{logit}(s_{i,q,t}) = \beta_{0,t} + \beta_{s,t}x_{i,q,t} + \beta_Q + \textbf{z}'_t \boldsymbol{\beta}_c + \beta_{d,1} w_{i,t} + \beta_{d,2} (x_{i,q,t}w_{i,q,t}) \\
-&\mathtt{parameters} \qquad &\beta_{0,t} \sim \text{Normal}(\tilde{\beta_0}, \sigma_{\beta_0}^2 \textbf{I})\\
-& \qquad & \beta_{s,t} \sim \text{Normal}(\tilde{\beta_s}, \sigma_{\beta_s}^2 \textbf{I}) \\
-& \qquad & \boldsymbol{\beta_{Q}} \sim \text{Normal}(0, \sigma_Q^2 \textbf{I}) \\
-& \qquad & \boldsymbol{\beta}_c \sim \text{Normal}(0, \sigma_c^2 \textbf{I}) \\
+& \mathtt{data} \qquad &y_{i,q,t}^{S} \sim \text{Bernoulli}(s_{i,q,t}) \\
+& \mathtt{process} \qquad &\text{logit}(s_{i,q,t}) = \beta_{0,t} + \beta_{s,t}x_{i,q,t} + \beta_{Q,q} + \textbf{z}'_t \boldsymbol{\beta}_c + \beta_{d,1} w_{i,t} + \beta_{d,2} (x_{i,q,t}w_{i,q,t}) \\
+& \mathtt{parameters} \qquad &\beta_{0,t} \sim \text{Normal}(\tilde{\beta_0}, \sigma_{\beta_0}^2)\\
+& \qquad & \beta_{s,t} \sim \text{Normal}(\tilde{\beta_s}, \sigma_{\beta_s}^2) \\
+& \qquad & \beta_{Q,q} \sim \text{Normal}(0, \sigma_Q^2) \\
+& \qquad & \boldsymbol{\beta}_c \sim \text{Normal}(\textbf{0}, \sigma_c^2 \textbf{I}) \\
 & \qquad & \beta_{d,1} \sim \text{Normal}(0, 100) \\
@@
 &[\beta_{0,t} | \tilde{\beta_0}, \sigma_{\beta_0}^2] \times \\
-&[\beta_{s,t} | \tilde{\beta_s}, \sigma_{\beta_s}^2] [\beta_c|\sigma_{beta_c}^2] \times \\
-&\prod_{Q=1}^{Q_{tot}} [\beta_{Q} | \sigma_Q^2] \times \\
+&[\beta_{s,t} | \tilde{\beta_s}, \sigma_{\beta_s}^2] [\boldsymbol{\beta_c}|\sigma_{\beta_c}^2 \textbf{I}] \times \\
+&\prod_{q=1}^{Q_{tot}} [\beta_{Q,q} | \sigma_Q^2] \times \\
 &[\beta_{d,1}] [\beta_{d,1}] [\tilde{\beta_0}] [\tilde{\beta_s}] [\sigma_Q] [\sigma_{\beta_0}] [\sigma_{\beta_s}]
@@
 **Stan code for model**
-```{r survival_stan, echo=FALSE, eval=FALSE}
+```{r survival_stan, echo=TRUE, eval=FALSE}
 data{
@@
   // Priors
-  a_mu ~ normal(0,100);
-  w ~ normal(0,100);
-  b1_mu ~ normal(0,100);
+  a_mu ~ normal(0,10);
+  w ~ normal(0,10);
+  b1_mu ~ normal(0,10);
   sig_a ~ cauchy(0,5);
@@
 \begin{align}
-y_{i,q,t+1}^G &\sim \text{Normal}(\mu_{i,q,t+1}, \sigma^2_{x_{i,q,t+1}}\textbf{I}) \\
-\mu_{i,q,t+1} &= \beta_{0,t} + \beta_{s,t}x_{i,q,t} + \beta_Q + \textbf{z}'_t \boldsymbol{\beta}_c + \beta_{d,1} w_{i,q,t} + \beta_{d,2} (x_{i,q,t}w_{i,q,t})
+y_{i,q,t+1}^G &\sim \text{Normal}(\mu_{i,q,t+1}, \sigma^2_{x_{i,q,t+1}}) \\
+\mu_{i,q,t+1} &= \beta_{0,t} + \beta_{s,t}x_{i,q,t} + \beta_{Q,q} + \textbf{z}'_t \boldsymbol{\beta}_c + \beta_{d,1} w_{i,q,t} + \beta_{d,2} (x_{i,q,t}w_{i,q,t})
 \end{align}
@@
 \begin{align}
-&\mathtt{data} \qquad &y_{i,q,t+1}^G \sim \text{Normal}(\mu_{i,q,t+1}, \sigma^2_{x_{i,q,t+1}}\textbf{I}) \\
-&\mathtt{process} \qquad &\mu_{i,q,t+1} = \beta_{0,t} + \beta_{s,t}x_{i,q,t} + \beta_Q + \textbf{z}'_t \boldsymbol{\beta}_c + \beta_{d,1} w_{i,q,t} + \beta_{d,2} (x_{i,q,t}w_{i,q,t}) \\
+&\mathtt{data} \qquad &y_{i,q,t+1}^G \sim \text{Normal}(\mu_{i,q,t+1}, \sigma^2_{x_{i,q,t+1}}) \\
+&\mathtt{process} \qquad &\mu_{i,q,t+1} = \beta_{0,t} + \beta_{s,t}x_{i,q,t} + \beta_{Q,q} + \textbf{z}'_t \boldsymbol{\beta}_c + \beta_{d,1} w_{i,q,t} + \beta_{d,2} (x_{i,q,t}w_{i,q,t}) \\
 & \qquad & \sigma^2_{iQ,t} = ae^{b\mu_{iQ,t+1}} \\
-&\mathtt{parameters} \qquad &\beta_{0,t} \sim \text{Normal}(\tilde{\beta_0}, \sigma_{\beta_0}^2 \textbf{I})\\
-& \qquad & \beta_{s,t} \sim \text{Normal}(\tilde{\beta_s}, \sigma_{\beta_s}^2 \textbf{I}) \\
-& \qquad & \boldsymbol{\beta_{Q}} \sim \text{Normal}(0, \sigma_Q^2 \textbf{I}) \\
-& \qquad & \boldsymbol{\beta}_c \sim \text{Normal}(0, \sigma_c^2 \textbf{I}) \\
+&\mathtt{parameters} \qquad &\beta_{0,t} \sim \text{Normal}(\tilde{\beta_0}, \sigma_{\beta_0}^2)\\
+& \qquad & \beta_{s,t} \sim \text{Normal}(\tilde{\beta_s}, \sigma_{\beta_s}^2) \\
+& \qquad & \boldsymbol{\beta_{Q,q}} \sim \text{Normal}(0, \sigma_Q^2) \\
+& \qquad & \boldsymbol{\beta}_c \sim \text{Normal}(\textbf{0}, \sigma_c^2 \textbf{I}) \\
 & \qquad & \beta_{d,1} \sim \text{Normal}(0, 100) \\
@@
 & \qquad & \tilde{\beta_s} \sim \text{Normal}(0, 100) \\
-& \qquad & \sigma_Q \sim \text{Cauchy}(0, 2) \\
+& \qquad & \sigma_Q \sim \text{Cauchy}(0, 5) \\
 & \qquad & \sigma_{\beta_0} \sim \text{Cauchy}(0, 5) \\
@@
 &[\beta_{0,t} | \tilde{\beta_0}, \sigma_{\beta_0}^2] \times \\
-&[\beta_{s,t} | \tilde{\beta_s}, \sigma_{\beta_s}^2] [\beta_c|\sigma_{beta_c}^2] \times \\
-&\prod_{Q=1}^{Q_{tot}} [\beta_{Q} | \sigma_Q^2] \times \\
+&[\beta_{s,t} | \tilde{\beta_s}, \sigma_{\beta_s}^2] [\boldsymbol{\beta_c}|\sigma_{\beta_c}^2 \textbf{I}] \times \\
+&\prod_{q=1}^{Q_{tot}} [\beta_{Q,q} | \sigma_Q^2] \times \\
 &[\beta_{d,1}] [\beta_{d,1}] [\tilde{\beta_0}] [\tilde{\beta_s}] [\sigma_Q] [\sigma_{\beta_0}] [\sigma_{\beta_s}] [a] [b]
@@
 \begin{align}
-\lambda_{q,t+1} = \tilde{c}_{q,t} \, \text{exp}\left(\beta_{0,t} + \beta_Q + \textbf{z}'_t \boldsymbol{\beta}_c + \beta_{d}\sqrt{\tilde{c}_{q,t}}\right)
+\lambda_{q,t+1} = \tilde{c}_{q,t} \, \text{exp}\left(\beta_{0,t} + \beta_{Q,q} + \textbf{z}'_t \boldsymbol{\beta}_c + \beta_{d}\sqrt{\tilde{c}_{q,t}}\right)
 \end{align}
@@
 &\mathtt{data} \qquad &y^{R}_{q,t+1} \sim \text{NegBin}(\lambda_{q,t+1},\phi) \\
-&\mathtt{process} \qquad &\lambda_{q,t+1} = \tilde{c}_{q,t} \, \text{exp}\left(\beta_{0,t} + \beta_Q + \textbf{z}'_t \boldsymbol{\beta}_c + \beta_{d}\sqrt{\tilde{c}_{q,t}}\right) \\
-&\mathtt{parameters} \qquad &\beta_{0,t} \sim \text{Normal}(\tilde{\beta_0}, \sigma_{\beta_0}^2 \textbf{I})\\
-& \qquad & \beta_{Q} \sim \text{Normal}(0,\sigma_Q^2 \textbf{I}) \\
-& \qquad & \boldsymbol{\beta}_c \sim \text{Normal}(0, \sigma_c^2 \textbf{I}) \\
+&\mathtt{process} \qquad &\lambda_{q,t+1} = \tilde{c}_{q,t} \, \text{exp}\left(\beta_{0,t} + \beta_{Q,q} + \textbf{z}'_t \boldsymbol{\beta}_c + \beta_{d}\sqrt{\tilde{c}_{q,t}}\right) \\
+&\mathtt{parameters} \qquad &\beta_{0,t} \sim \text{Normal}(\tilde{\beta_0}, \sigma_{\beta_0}^2)\\
+& \qquad & \beta_{Q,q} \sim \text{Normal}(0,\sigma_Q^2) \\
+& \qquad & \boldsymbol{\beta}_c \sim \text{Normal}(\textbf{0}, \sigma_c^2 \textbf{I}) \\
 & \qquad & \beta_{d} \sim \text{Uniform}(-10, 10) \\
@@
 [\boldsymbol{\beta_0}, \tilde{\beta}, \boldsymbol{\beta_Q}, \boldsymbol{\beta_c}, \beta_{d}, \sigma_Q^2, \sigma_{\beta_0}^2, \phi, u | y^{R}_{q,t+1}] &\propto \\
-&\prod_{t=1}^T \prod_{i=1}^n [y^{R}_{q,t+1} | \beta_{0,t}, \beta_{Q}, \beta_d, \boldsymbol{\beta_c}, \phi, u] [\beta_{0,t} | \tilde{\beta}, \sigma_{\beta_0}^2] \times \\
-&\prod_{Q=1}^{Q_{tot}} [\beta_{Q} | \sigma_Q^2] [\boldsymbol{\beta_c}] [\beta_d] [\tilde{\beta_0}] [\sigma_Q] [\sigma_{\beta_0}] [\phi] [u]
+&\prod_{t=1}^T \prod_{i=1}^n [y^{R}_{q,t+1} | \beta_{0,t}, \beta_{Q}, \beta_d, \boldsymbol{\beta_c}, \phi, u] \times \\
+&[\beta_{0,t} | \tilde{\beta}, \sigma_{\beta_0}^2] [\boldsymbol{\beta_c}|\sigma_{\beta_c}^2 \textbf{I}] \times \\
+&\prod_{q=1}^{Q_{tot}} [\beta_{Q,q} | \sigma_Q^2] [\boldsymbol{\beta_c}] [\beta_d] [\tilde{\beta_0}] [\sigma_Q] [\sigma_{\beta_0}] [\phi] [u]
 \end{align}
@@
 **Stan code for model**
-```{r rec_stan, echo=FALSE, eval=FALSE}
+```{r rec_stan, echo=TRUE, eval=FALSE}
 data{
@@
   u ~ uniform(0,1);
-  theta ~ uniform(0,100);
-  a_mu ~ normal(0,1000);
-  dd ~ uniform(-100,100);
+  theta ~ uniform(0,10);
+  a_mu ~ normal(0,10);
+  dd ~ uniform(-10,10);
   sig_a ~ cauchy(0,5);
@@
 y_{q,t+1}^{P} &\sim \text{LogNormal}(\mu_{q,t+1}, \sigma^2)_{0}^{1} \\
-\mu_{q,t+1} &= \beta_{0,t} + \beta_{s,t}x_{q,t} + \beta_Q + \textbf{z}'_t \boldsymbol{\beta}_c
+\mu_{q,t+1} &= \beta_{0,t} + \beta_{s,t}x_{q,t} + \beta_{Q,q} + \textbf{z}'_t \boldsymbol{\beta}_c
 \end{align}
@@
 &\mathtt{data} \qquad &y_{q,t+1}^{P} \sim \text{LogNormal}(\mu_{q,t+1}, \sigma^2)_{0}^{1} \\
-&\mathtt{process} \qquad &\mu_{q,t+1} = \beta_{0,t} + \beta_{s,t}x_{q,t} + \beta_Q + \textbf{z}'_t \boldsymbol{\beta}_c \\
+&\mathtt{process} \qquad &\mu_{q,t+1} = \beta_{0,t} + \beta_{s,t}x_{q,t} + \beta_{Q,q} + \textbf{z}'_t \boldsymbol{\beta}_c \\
 &\mathtt{parameters} \qquad &\beta_{0,t} \sim \text{Normal}(\tilde{\beta_0}, \sigma_{\beta_0}^2)\\
-& \qquad & \beta_{s,t} \sim \text{Normal}(\tilde{\beta_s}, \sigma_{\beta_s}^2 \textbf{I}) \\
-& \qquad & \beta_{Q} \sim \text{Normal}(0,\sigma_Q^2 \textbf{I}) \\
-& \qquad & \boldsymbol{\beta}_c \sim \text{Normal}(0, \sigma_c^2 \textbf{I}) \\
+& \qquad & \beta_{s,t} \sim \text{Normal}(\tilde{\beta_s}, \sigma_{\beta_s}^2) \\
+& \qquad & \beta_{Q,q} \sim \text{Normal}(0,\sigma_Q^2) \\
+& \qquad & \boldsymbol{\beta}_c \sim \text{Normal}(\textbf{0}, \sigma_c^2 \textbf{I}) \\
 & \qquad & \tilde{\beta_0} \sim \text{Normal}(0, 100) \\
@@
 [\boldsymbol{\beta_0}, \tilde{\beta_0}, \boldsymbol{\beta_s}, \tilde{\beta_s}, \boldsymbol{\beta_Q}, \boldsymbol{\beta_c}, \sigma_Q^2, \sigma_{\beta_0}^2, \sigma_{\beta_s}^2, \sigma^2 | y_{q,t+1}^{P}] &\propto \\
-&\prod_{t=1}^T \prod_{i=1}^n [y_{q,t+1} | \beta_{0,t}, \beta_{s,t}, \beta_{Q}, \boldsymbol{\beta_c}] \times \\
+&\prod_{t=1}^T \prod_{i=1}^n [y_{q,t+1} | \beta_{0,t}, \beta_{s,t}, \beta_{Q,q}, \boldsymbol{\beta_c}] \times \\
 &[\beta_{0,t} | \tilde{\beta_0}, \sigma_{\beta_0}^2] \times \\
-&[\beta_{s,t} | \tilde{\beta_s}, \sigma_{\beta_s}^2] [\beta_c|\sigma_{beta_c}^2] \times \\
-&\prod_{Q=1}^{Q_{tot}} [\beta_{Q} | \sigma_Q^2] \times \\
+&[\beta_{s,t} | \tilde{\beta_s}, \sigma_{\beta_s}^2] [\boldsymbol{\beta_c}|\sigma_{\beta_c}^2 \textbf{I}] \times \\
+&\prod_{q=1}^{Q_{tot}} [\beta_{Q,q} | \sigma_Q^2] \times \\
 &[\tilde{\beta_0}] [\tilde{\beta_s}] [\sigma_Q] [\sigma_{\beta_0}] [\sigma_{\beta_s}] [\sigma]
@@
 **Stan code for model**
-```{r qbm_stan, echo=FALSE, eval=FALSE}
+```{r qbm_stan, echo=TRUE, eval=FALSE}
 data{
@@
   real<lower=0> sig_G;
-  real<lower=0> tau;
+  real<lower=0> sigmaSq;
 }
@@
   vector[N] climEff;
+  real<lower=0> tau;
+  tau <- sqrt(sigmaSq);
   climEff <- C*b2;
@@
   // Priors
-  a_mu ~ normal(0,1000);
-  b1_mu ~ normal(0,1000);
+  a_mu ~ normal(0,10);
+  b1_mu ~ normal(0,10);
   sig_a ~ cauchy(0,5);
@@
   b1 ~ normal(b1_mu, sig_b1);
-  tau ~ cauchy(0,5);
+  sigmaSq ~ inv_gamma(1, 1);
 
@@
   Y ~ lognormal(mu, tau);
+}
 }

### 2016-07-07 manuscript/supporting_information.Rmd
updates subscripting and model statements re MBH edits

@@
 \begin{align}
-[\boldsymbol{\beta_0}, \tilde{\beta_0}, \boldsymbol{\beta_s}, \tilde{\beta_s}, \boldsymbol{\beta_Q}, \boldsymbol{\beta_c}, \beta_{d,1}, \beta_{d,2}, \sigma_Q^2, \sigma_{\beta_0}^2, \sigma_{\beta_s}^2 | y_{i,q,t}^{S}] &\propto \\
+[\boldsymbol{\beta_0}, \tilde{\beta_0}, \boldsymbol{\beta_s}, \tilde{\beta_s}, \boldsymbol{\beta_Q,q}, \boldsymbol{\beta_c}, \beta_{d,1}, \beta_{d,2}, \sigma_Q^2, \sigma_{\beta_0}^2, \sigma_{\beta_s}^2 | y_{i,q,t}^{S}] &\propto \\
 &\prod_{t=1}^T \prod_{i=1}^n [y_{i,q,t}^{S} | \beta_{0,t}, \beta_{s,t}, \beta_{Q}, \boldsymbol{\beta_c}, \beta_{d,1}, \beta_{d,2}] \times \\
 &[\beta_{0,t} | \tilde{\beta_0}, \sigma_{\beta_0}^2] \times \\
-&[\beta_{s,t} | \tilde{\beta_s}, \sigma_{\beta_s}^2] [\boldsymbol{\beta_c}|\sigma_{\beta_c}^2 \textbf{I}] \times \\
+&[\beta_{s,t} | \tilde{\beta_s}, \sigma_{\beta_s}^2] [\boldsymbol{\beta_c}|\sigma_{\beta_c}^2] \times \\
 &\prod_{q=1}^{Q_{tot}} [\beta_{Q,q} | \sigma_Q^2] \times \\
@@
 & \qquad & \beta_{s,t} \sim \text{Normal}(\tilde{\beta_s}, \sigma_{\beta_s}^2) \\
-& \qquad & \boldsymbol{\beta_{Q,q}} \sim \text{Normal}(0, \sigma_Q^2) \\
+& \qquad & \beta_{Q,q} \sim \text{Normal}(0, \sigma_Q^2) \\
 & \qquad & \boldsymbol{\beta}_c \sim \text{Normal}(\textbf{0}, \sigma_c^2 \textbf{I}) \\
@@
 \begin{align}
-[\boldsymbol{\beta_0}, \tilde{\beta_0}, \boldsymbol{\beta_s}, \tilde{\beta_s}, \boldsymbol{\beta_Q}, \boldsymbol{\beta_c}, \beta_{d,1}, \beta_{d,2}, \sigma_Q^2, \sigma_{\beta_0}^2, \sigma_{\beta_s}^2,a,b | y_{i,q,t+1}^G] &\propto \\
+[\boldsymbol{\beta_0}, \tilde{\beta_0}, \boldsymbol{\beta_s}, \tilde{\beta_s}, \beta_{Q,q}, \boldsymbol{\beta_c}, \beta_{d,1}, \beta_{d,2}, \sigma_Q^2, \sigma_{\beta_0}^2, \sigma_{\beta_s}^2,a,b | y_{i,q,t+1}^G] &\propto \\
 &\prod_{t=1}^T \prod_{i=1}^n [y_{i,q,t+1}^G | \beta_{0,t}, \beta_{s,t}, \beta_{Q}, \boldsymbol{\beta_c}, \beta_{d,1}, \beta_{d,2}, a, b] \times \\
 &[\beta_{0,t} | \tilde{\beta_0}, \sigma_{\beta_0}^2] \times \\
-&[\beta_{s,t} | \tilde{\beta_s}, \sigma_{\beta_s}^2] [\boldsymbol{\beta_c}|\sigma_{\beta_c}^2 \textbf{I}] \times \\
+&[\beta_{s,t} | \tilde{\beta_s}, \sigma_{\beta_s}^2] [\boldsymbol{\beta_c}|\sigma_{\beta_c}^2] \times \\
 &\prod_{q=1}^{Q_{tot}} [\beta_{Q,q} | \sigma_Q^2] \times \\
@@
 [\boldsymbol{\beta_0}, \tilde{\beta}, \boldsymbol{\beta_Q}, \boldsymbol{\beta_c}, \beta_{d}, \sigma_Q^2, \sigma_{\beta_0}^2, \phi, u | y^{R}_{q,t+1}] &\propto \\
-&\prod_{t=1}^T \prod_{i=1}^n [y^{R}_{q,t+1} | \beta_{0,t}, \beta_{Q}, \beta_d, \boldsymbol{\beta_c}, \phi, u] \times \\
-&[\beta_{0,t} | \tilde{\beta}, \sigma_{\beta_0}^2] [\boldsymbol{\beta_c}|\sigma_{\beta_c}^2 \textbf{I}] \times \\
+&\prod_{t=1}^T \prod_{i=1}^n [y^{R}_{q,t+1} | \beta_{0,t}, \beta_{Q,q}, \beta_d, \boldsymbol{\beta_c}, \phi, u] \times \\
+&[\beta_{0,t} | \tilde{\beta}, \sigma_{\beta_0}^2] [\boldsymbol{\beta_c}|\sigma_{\beta_c}^2] \times \\
 &\prod_{q=1}^{Q_{tot}} [\beta_{Q,q} | \sigma_Q^2] [\boldsymbol{\beta_c}] [\beta_d] [\tilde{\beta_0}] [\sigma_Q] [\sigma_{\beta_0}] [\phi] [u]
@@
 &[\beta_{0,t} | \tilde{\beta_0}, \sigma_{\beta_0}^2] \times \\
-&[\beta_{s,t} | \tilde{\beta_s}, \sigma_{\beta_s}^2] [\boldsymbol{\beta_c}|\sigma_{\beta_c}^2 \textbf{I}] \times \\
+&[\beta_{s,t} | \tilde{\beta_s}, \sigma_{\beta_s}^2] [\boldsymbol{\beta_c}|\sigma_{\beta_c}^2] \times \\
 &\prod_{q=1}^{Q_{tot}} [\beta_{Q,q} | \sigma_Q^2] \times \\

### 2016-07-08 manuscript/manuscript.Rmd
adds paragraph on caveats re RF

@@
 We conclude that models based on aggregated individual-level data offer a practical alternative to data-intensive demographic models when forecasting is the goal.
+Percent cover data are widely collected for many plant species, and several long-term time series already exist.
+Our results should inspire plant population modelers to exploit such data to forecast the impacts of climate change.
     \end{enumerate}
@@
 However, we found no evidence for a difference between the IPM and QBM forecast horizons (Fig. 4).
+
+Our comparison of two model types, one based on individual-level data and one based on percent cover data, should not be considered an exhaustive test.
+There are several reasons why the percent cover-based model generated equally skillfull forecasts for our focal species, and several reasons why percent cover models may fail in other cases.
+First, we had no cases where a climate covariate had a strong negative effect on one vital rate and a strong positive effect on a different vital rate (Fig. 3).
+As noted by @Freckleton2011, complex age or stage structure can compromise predictions from density-structured models that aggregate over life-histories.
+Interactions between a species' response to climate and life-history stages would also impact model forecasts because important climate effects could be averaged-over.
+Second, our particular recruitment model is rather coarse and may therefore already average over important climate-influenced processes such as seed germination and establishment.
+If we had been able to more precisely model the recruitment process, we might have discovered stronger climate effects -- effects that could be hard to detect from percent cover data.
+As advocated by @Freckleton2011, knowledge of a species' population ecology should guide the modeling approach.
 

### 2016-07-11 manuscript/manuscript.Rmd
adds UAES journal paper number
adds data accessibility statement

@@
 government.
+This research was supported by the Utah Agricultural Experiment Station, Utah State University, and approved as journal paper number 8917.
+
+Data Accessibility
+------------------
+The data used in this paper have been archived on Ecological Archives: http://esapubs.org/archive/ecol/E092/143/.
+All data and R code necessary to reproduce our work has been deposited on Dryad (link) and is also available on GitHub (http://github.com/atredennick/MicroMesoForecast).
 

### 2016-07-11 manuscript/manuscript.Rmd
small PBA edits to last paragraph and summary point

--- 2016-07-06 manuscript/manuscript.Rmd
@@
 Increases in skill with climate covariates were similar between model types.
-        \item For our focal species at this particular location, and using our particular statistical models, percent cover models generated forecasts as skillful as those from a demographic model.
-We conclude that models based on aggregated individual-level data offer a practical alternative to data-intensive demographic models when forecasting is the goal.
+        \item In our case study, percent cover models generated forecasts as skillful as those from a demographic model.
+For the goal of forecasting, models based on aggregated individual-level data may offer a practical alternative to data-intensive demographic models.
+Long time series of percent cover data already exist for many plant species.
+Modelers should exploit these data to predict the impacts of environmental change.
     \end{enumerate}
@@
 
+Our comparison of a model based on individual-level data with one based on percent cover data is not an exhaustive test.
+Understanding the reasons why the percent cover-based model matched the skill of a demographic model for our focal species may help us anticipate situations in which a percent-cover approach would fail.
+First, for none of our species did a climate covariate have a strong negative effect on one vital rate and a strong positive effect on a different vital rate (Fig. 3).
+As noted by @Freckleton2011, complex age or stage structure can compromise predictions from models that aggregate over life-histories, and the same should be true when aggregating across vital rates with contrasting responses to climate drivers.
+Second, our particular recruitment model is already so aggregated -- it averages across seed production, germination and establishment -- that it may fail to detect important demographic responses to climate, putting our individual-based model and percent cover model on more equal footing.
+More finely resolved recruitment data might help our individual-based model outperform the population-level model.
+As advocated by @Freckleton2011, knowledge of a species' population ecology should guide the modeling approach.
+
 In conclusion, we found that models based on individual-level demographic data generally failed to generate more skillful population forecasts than models based on population-level data.

### 2016-07-11 manuscript/manuscript.Rmd
MEE submission version

@@
     fig_caption: true
-    keep_tex: false
+    keep_tex: true
 fontsize: 12pt
@@
 Traditional population models built with detailed demographic observations from one study site can address the impacts of environmental change at particular locations, but are difficult to scale up to the landscape and regional scales relevant to management decisions.
-        \item An alternative is to build models using population-level data which are much easier to collect than individual-level data over broad spatial scales.
+An alternative is to build models using population-level data which are much easier to collect than individual-level data over broad spatial scales.
 However, it is unknown whether models built using population-level data adequately capture the effects of density-dependence and environmental forcing that are necessary to generate skillful forecasts.

### 2016-07-11 manuscript/manuscript.tex
MEE submission version

--- 2015-12-03 manuscript/manuscript.tex
@@
 \usepackage[margin=1in]{geometry}
+\usepackage{graphicx}
+\makeatletter
+\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
+\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
+\makeatother
+% Scale images if necessary, so that they will not overflow the page
+% margins by default, and it is still possible to overwrite the defaults
+% using explicit options in \includegraphics[width, height, ...]{}
+\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
 \ifxetex
@@
             citecolor=blue,
-            urlcolor=blue,
-            linkcolor=magenta,
+            urlcolor=black,
+            linkcolor=black,
             pdfborder={0 0 0}}
@@
 \usepackage{setspace}
+\usepackage{sectsty}
 \doublespacing
@@
 \usepackage[document]{ragged2e}
-\pagewiselinenumbers
+\usepackage{times}
+\usepackage{enumitem}
 
@@
 
-\subsubsection{Title}\label{title}
-
-Do we need demographic data to forecast population responses to climate
-change?
-
-\subsubsection{Authors}\label{authors}
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\itemsep1pt\parskip0pt\parsep0pt
-\item
-  Andrew T. Tredennick (corresponding author:
-  \href{mailto:atredenn@gmail.com}{\nolinkurl{atredenn@gmail.com}})
-\item
-  Peter B. Adler
-  (\href{mailto:peter.adler@usu.edu}{\nolinkurl{peter.adler@usu.edu}})
+\textsf{\LARGE{\textbf{Do we need demographic data to forecast plant population dynamics?}}}
+
+\renewcommand*{\thefootnote}{\fnsymbol{footnote}}
+
+\vspace{1em}
+
+\textsf{\normalsize{\textbf{Andrew T.
+Tredennick\textsuperscript{1}\footnote{Corresponding author: E-mail: atredenn@gmail.com}, Mevin B.
+Hooten\textsuperscript{2,3,4}, and Peter B.
+Adler\textsuperscript{1}}}}
+
+\vspace{1em}
+
+\textsf{\textit{\footnotesize{\textsuperscript{1}Department of Wildland Resources and the Ecology Center, 5230 Old Main Hill, Utah State University, Logan, Utah 84322, USA; \textsuperscript{2}U.S.
+Geological Survey, Colorado Cooperative Fish and Wildlife Research Unit, Fort Collins, CO 80523, USA; \textsuperscript{3}Department of Fish, Wildlife, and Conservation Biology, Colorado State University, Fort Collins, CO 80523 USA}; \textsuperscript{4}Department of Statistics, Colorado State University, Fort Collins, CO 80523 USA}}
+
+\textbf{Running head}: Demographic data and population forecasting \\
+\textbf{Word count}: 6845 (including reference section) \\
+\textbf{Corresponding author}: \\
+Andrew Tredennick (atredenn@gmail.com) \\
+Department of Wildland Resources and the Ecology Center \\
+5230 Old Main Hill \\ 
+Utah State University \\ 
+Logan, Utah 84322, USA
+
+Last compile: \today
+
+\allsectionsfont{\normalfont\sffamily\bfseries}
+\renewcommand*{\thefootnote}{\arabic{footnote}} \setcounter{footnote}{0}
+
+\rule{\textwidth}{1pt}
+
+\subsection{Summary}\label{summary}
+
+\begin{enumerate}[label=\textbf{\arabic*}]
+        \item Rapid environmental change has generated growing interest in forecasts of future population trajectories.
+Traditional population models built with detailed demographic observations from one study site can address the impacts of environmental change at particular locations, but are difficult to scale up to the landscape and regional scales relevant to management decisions.
+An alternative is to build models using population-level data which are much easier to collect than individual-level data over broad spatial scales.
+However, it is unknown whether models built using population-level data adequately capture the effects of density-dependence and environmental forcing that are necessary to generate skillful forecasts.
+        \item Here, we test the consequences of aggregating individual responses when forecasting the population states and trajectories of four perennial grass species in a semi-arid grassland in Montana, USA.
+We parameterized two population models for each species, one based on individual-level data (survival, growth and recruitment) and one on population-level data (percent cover), and compared their forecasting skill and forecast horizons with and without the inclusion of climate covariates.
+For both models we used Bayesian ridge regression to weight the influence of climate covariates for optimal prediction.
+        \item In the absence of climate effects, we found no significant difference between the forecasting skill of models based on individual-level data and models based on population-level data.
+Climate effects were weak, but increased forecasting skill for two species.
+Increases in skill with climate covariates were similar between model types.
+        \item In our case study, percent cover models generated forecasts as skillful as those from a demographic model.
+For the goal of forecasting, models based on aggregated individual-level data may offer a practical alternative to data-intensive demographic models.
+Long time series of percent cover data already exist for many plant species.
+Modelers should exploit these data to predict the impacts of environmental change.
     \end{enumerate}
 
-\subsubsection{Affiliations}\label{affiliations}
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\itemsep1pt\parskip0pt\parsep0pt
-\item
-  Department of Wildland Resources and the Ecology Center, 5230 Old Main
-  Hill, Utah State University, Logan, Utah 84322-5230 USA
-\item
-  Department of Wildland Resources and the Ecology Center, 5230 Old Main
-  Hill, Utah State University, Logan, Utah 84322-5230 USA
-\end{enumerate}
-
-\newpage{}
-
-\subsection{Summary}\label{summary}
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\itemsep1pt\parskip0pt\parsep0pt
-\item
-  Rapid climate change has generated growing interest in forecasts of
-  future population trajectories. Traditional population models,
-  typically built using detailed demographic observations from one study
-  site, can address climate change impacts at one location, but are
-  difficult to scale up to the landscape and regional scales relevant to
-  management decisions. An alternative is to build models using
-  population-level data that is much easier to collect over broad
-  spatial scales than individual-level data. However, such models ignore
-  the fact that climate drives population growth through its influence
-  on individual performance.
-\item
-  We test the consequences of aggregating individual responses when
-  forecasting climate change impacts on two perennial grass species
-  (\emph{Pascopyrum smithii} and \emph{Sprobolis flexuosus}). We
-  parameterized two population models for each species, one based on
-  individual-level data (survival, growth and recruitment) and one on
-  population-level data (percent cover), and compared their accuracy,
-  precision, and sensitivity to climate variables. For both models we
-  used Bayesian ridge regression to identify the optimal predictive
-  model in terms of climate covariate strengths.
-\item
-  The individual-level model was more accurate and precise than the
-  aggregated model when predicting out-of-sample observations. When
-  comparing climate effects from both models, the population-level model
-  missed important climate effects from at least one vital rate for each
-  species. Increasing the sample size at the population-level would not
-  necessarily reduce forecast uncertainty; the way to reduce uncertainty
-  is to capture unique climate dependence of individual vital rates.
-\item
-  Our analysis indicates that there is no shortcut to forecasting
-  climate change impacts on plant populations --- detailed demographic
-  data is essential. Despite the superiority of the individual-level
-  model, the forecasts it generated still were too uncertain to be
-  useful for decision-makers. We need new methods to collect demographic
-  data efficiently across environmental gradients in space and time.
-\end{enumerate}
-
-\emph{Key words: forecasting, climate change, grassland, integral
-projection model, population model}
-
-\newpage{}
+\textsf{\textbf{Key-words:}} forecasting, climate change, grassland,
+integral projection model, population model, statistical regularization,
+ridge regression
 
@@
 fully account for uncertainty and variability in both ecological process
-and model parameters (Luo et al. 2011, but see Perretti et al. 2013 for
-an argument against modeling the ecological process). The increasing
-statistical sophistication of population models (Rees and Ellner 2009)
-makes them promising tools for predicting the impacts of environmental
-change on species persistence and abundance. But reconciling the scales
-at which population models are parameterized and the scales at which
-environmental changes play out remains a challenge (Clark et al. 2010,
-2012, Freckleton et al. 2011, Queenborough et al. 2011). The problem is
-that most population models are built using data from a single study
-site because collecting those data, which involves tracking the fates of
-individuals plants, is so difficult. The resulting models cannot be
-applied to the landscape and regional scales relevant to decision-making
-without information about how the fitted parameters respond to spatial
-variation in biotic and abiotic drivers (S{æ}ther et al. 2007). The
-limited spatial extent of individual-level demographic datasets
-constrains our ability to use population models to address applied
-questions about the consequences of climate change.
-
-The inability of most population models to address landscape-scale
-problems may explain why land managers and conservation planners have
-embraced species distribution models (SDMs) (see Guisan and Thuiller
-2005 for a review). SDMs typically rely on easy-to-collect
-presence/absence data (but see Clark et al. 2014 for new methods) and
-remotely-sensed environmental covariates that allow researchers to model
-large spatial extents (e.g., Maiorano et al. 2013). Thus, it is
-relatively straightforward to parameterize and project SDMs over
-landscapes and regions. However, the limitations of SDMs are well known
-(Pearson and Dawson 2003, Elith and Leathwick 2009, Ara{ú}jo and
-Peterson 2012). Ideally, researchers would provide managers with
-landscape-scale population models, combining the extent of SDMs with
-information about dynamics and species abundances (Schurr et al. 2012,
-Merow et al. 2014).
+and model parameters (Luo et al. 2011, but see Perretti et al. 2013).
+The increasing statistical sophistication of population models (Rees and
+Ellner 2009) makes them promising tools for predicting the impacts of
+environmental change on species persistence and abundance. But
+reconciling the scales at which population models are parameterized with
+the scales at which environmental changes play out remains a challenge
+(Clark et al. 2010, 2012, Freckleton et al. 2011, Queenborough et al.
+2011). Most population models are built using demographic data from a
+single study site because tracking the fates of individuals is so
+difficult. The resulting models cannot be applied to the landscape and
+regional scales relevant to decision-making without information about
+how the estimated parameters respond to spatial variation in biotic and
+abiotic drivers (S{æ}ther et al. 2007). The limited spatial extent of
+individual-level demographic datasets constrains our ability to use
+population models to address applied questions about the consequences of
+climate change.
 
@@
 data cannot provide inference about demographic mechanisms, but might be
-sufficient for modeling future population states, especially since such
-data are feasible to collect across broad spatial extents (e.g.,
-Queenborough et al. 2011). The choice between individual and
-population-level data involves a difficult trade-off: while
-individual-level data leads to more mechanistic models, population-level
-data leads to models that can be applied over greater spatial and
-temporal extents. An open question is how much forecasting skill is lost
-when we build models based on population rather than individual-level
-data.
+sufficient for modeling future population states, especially because
+population-level data, such as plant percent cover, are feasible to
+collect across broad spatial extents (e.g., Queenborough et al. 2011).
+The choice between individual and population-level data involves a
+difficult trade-off: while individual-level data are necessary for
+mechanistic models, population-level data enable models that can be
+applied over greater spatial and temporal extents. An open question is
+how much forecasting skill is lost when we build models based on
+population rather than individual-level data.
 
@@
 even if some precision is lost when compared to fully continuous models.
-The appeal of density-structured approaches is clear. However, none of
-these models included environmental covariates.
+However, previous tests of density-structured models have yet to assess
+their ability to forecast out-of-sample observations, and they have not
+included environmental covariates, which are necessary to forecast
+population responses to climate change.
 
 Addressing climate change questions with models fit to population-level
-data is potentially problematic. It is individuals, not populations,
-that respond to climate variables (Clark et al. 2012). Ignoring this
-fact amounts to an ``ecological fallacy'', where inference about the
-individual relies on statistical inference on the group (Piantadosi et
-al. 1988). Population growth (or decline) is the outcome of demographic
-processes such as survival, growth, and recruitment that occur at the
-level of individual plants. Climate can affect each demographic process
-in unique, potentially opossing, ways (Dalgleish et al. 2011). These
-unique climate responses may be difficult to resolve in statistical
-models based on population-level data where demographic processes are
-not identifiable. If important climate effects are missed because of the
-aggregation inherent in in population-level data, then population models
-built with such data will make uninformative or unreliable forecasts.
-
-Here, we compare the forecasting skill of statistical and population
-models based on aggregated, population-level data with models based on
-individual-level data. We used a unique demographic dataset that tracks
-the fates of individual plants from four species over 14 years to build
-two kinds of single-species population models, traditional models using
-individual growth, survival, and recruitment data and alternative models
-based on basal cover. In both models, interannual variation is
-explained, in part, by climate covariates. We first quantify forecasting
-skill using cross-validation. We then performed simulations to quantify
-the sensitivities of species' cover to small perturbations in average
-precipitation and temperature. Based on the cross-validation results,
-predictions of individual level models were clearly better, but,
-unfortunately, still too uncertain to inform management decisions.
+data is potentially problematic. Population growth (or decline) is the
+outcome of demographic processes such as survival, growth, and
+recruitment that occur at the level of individual plants. Climate can
+affect each demographic process in unique, potentially opposing, ways
+(Dalgleish et al. 2011). These unique climate responses may be difficult
+to resolve in statistical models based on population-level data where
+demographic processes are not identifiable. Futhermore, models based on
+aggregated data may reflect short-term effects of one vital rate more
+than others whose importance may only emerge over the long-term. For
+example, a one-year change in a plant species' cover or biomass might
+reflect individual growth or shrinkage, whereas the long-term trajectory
+of the population might be more influenced by recruitment. The same is
+true for density dependence: intraspecific density depedence may act
+most strongly on vital rates, like recruitment, that are difficult to
+identify from population-level data. If density dependence and/or
+important climate effects are missed because of the aggregation inherent
+in in population-level data, then population models built with such data
+will make uninformative or unreliable forecasts.
+
+We compared the forecasting skill of statistical and population models
+based on aggregated, population-level data with the skill of models
+based on individual-level data. We used a demographic dataset that
+tracks the fates of individual plants from four species over 14 years to
+build two kinds of single-species population models, traditional models
+using individual growth, survival, and recruitment data and alternative
+models based on population-level (basal cover) data. We simulated the
+models to answer two questions motivated by the fact that the effects of
+intraspecific competition (density dependence) and interannual weather
+variability act at the level of the individual (Clark et al. 2011).
+First, can population models fit using aggregated individual-level data
+(percent cover) adequately capture density dependence to produce
+forecasts as skillful as those from models fit to demographic data?
+Second, can population models fit using aggregated data adequately
+capture the influence of climate on population growth and, in turn,
+produce forecasts as skillful as those from models fit to demographic
+data?
 
@@
 
-Our demographic data come from the Fort Keogh Livestock and Range
-Research Laboratory in eastern Montana's northern mixed prairie near
-Miles City, Montana, USA (\(46^{\circ}\) 19' N, \(105^{\circ}\) 48' W).
-The dataset is freely available on Ecological Archives\footnote{\url{http://esapubs.org/archive/ecol/E092/143/}}
-(Anderson et al. 2011) , and interested readers should refer to the
-metadata for a complete description. The site is about 800 m above sea
-level and mean annual precipitation (1878-2009) is 334 mm, with most
-annual precipitation falling from April through September. The community
-is grass-dominated and we focused on the four most abundant grass
-species: \emph{Bouteloua gracilis} (BOGR), \emph{Hesperostipa comata}
-(HECO), \emph{Pascopyrum smithii} (PASM), and \emph{Poa secunda} (POSE)
-(Fig. 1).
-
-From 1932 to 1945 individual plants were identified and mapped annually
+Our demographic data come from a northern mixed grass prairie at the
+Fort Keogh Livestock and Range Research Laboratory near Miles City,
+Montana, USA (\(46^{\circ}\) 19' N, \(105^{\circ}\) 48' W). The dataset
+is available on Ecological Archives\footnote{\url{http://esapubs.org/archive/ecol/E092/143/}}
+(Anderson et al. 2011), and interested readers should refer to the
+metadata for a complete description. The site is 800 m above sea level
+and mean annual precipitation (1878-2009) is 334 mm, with most annual
+precipitation falling from April through September. The community is
+grass-dominated, and we focused on the four most abundant grass species:
+\emph{Bouteloua gracilis} (BOGR), \emph{Hesperostipa comata} (HECO),
+\emph{Pascopyrum smithii} (PASM), and \emph{Poa secunda} (POSE) (Fig.
+1). \emph{B. gracilis} is a warm-season perennial grass, whereas
+\emph{H. comata}, \emph{P. smithii}, and \emph{Poa secunda} are
+cool-season perennial grasses. The growing season begins in early spring
+(typically in April) and lasts through mid-summer (typically in June).
+
+From 1932 to 1945, individual plants were identified and mapped annually
 in 44 1-\(\text{m}^2\) quadrats using a pantograph. The quadrats were
-distributed in six pastures, each assigned a grazing treatment of light
-(1.24 ha/animal unit month), moderate (0.92 ha/aum), and heavy (0.76
-ha/aum) stocking rates (two pastures per treatment). In this analysis we
-account for potential differences among the grazing treatments, but do
-not focus on grazing\(\times\)climate interactions. The annual maps of
-the quadrats were digitized and the fates of individual plants tracked
-and extracted using a computer program (Lauenroth and Adler 2008, Chu et
-al. 2014). Daily climate data are available for the duration of the data
+distributed among six pastures, each assigned a grazing treatment of
+light (1.24 ha/animal unit month), moderate (0.92 ha/aum), and heavy
+(0.76 ha/aum) stocking rates (two pastures per treatment). In this
+analysis, we accounted for potential differences among the grazing
+treatments, but do not focus on grazing\(\times\)climate interactions.
+The annual maps of the quadrats were digitized and the fates of
+individual plants tracked and extracted using a computer program
+(Lauenroth and Adler 2008, Chu et al. 2014). The permanent quadrats have
+not been relocated, but their distribution in six different pastures
+means the data represent a broad spatial distribution for the study
+area. Daily climate data are available for the duration of the data
 collection period (1932 - 1945) from the Miles City airport, Wiley
@@
 We modeled each grass population based on two levels of data: individual
-and quadrat (Fig. 2). The individual data is the ``raw'' data. For the
-quadrat-level we data we simply sum individual basal cover for each
-quadrat by species. This is equivalent to a near-perfect census of
-quadrat percent cover because previous analysis shows that measurement
-error at the individual-level is small (Chu and Adler 2014). Based on
-these two datasets we can compare population models built using
-individual-level data and aggregated, quadrat-level data.
+and quadrat. The individual data are the ``raw'' data. For the
+quadrat-level data we simply sum individual basal cover for each quadrat
+by species. This is equivalent to a near-perfect census of quadrat
+percent cover because measurement error at the individual-level is small
+(Chu and Adler 2015). Based on these two datasets of 13 year-to-year
+transitions, we can compare population models built using
+individual-level data and aggregated, quadrat-level data. At the
+individual level, we explicitly model three vital rates: growth,
+survival, and recruitment. At the quadrat level, we model population
+growth as change in percent cover of quadrats with non-zero cover in
+year \emph{t} and in year \emph{t-1}, ignoring within-quadrat
+extirpation and colonization events because they are very rare in our
+time series (\(N=16\) and \(N=13\), respectively, across all species).
+Sample sizes for each species and vital rate model are shown in Table 1.
 
@@
   will be associated with the published version of the manuscript, we
-  have released v0.1 to be associated with this review version.}
-(\url{http://github.com/atredennick/MicroMesoForecast/releases}). That
-stable release will remain static as a record of this analysis, but
-subsequent versions may appear if we update this work. We have also
-deposited the v1.0 release on Dryad (\emph{link here after acceptance}).
-
-\subsubsection{Stastical models of vital
-rates}\label{stastical-models-of-vital-rates}
+  have released v0.2 to be associated with this review version.}
+(\url{http://github.com/atredennick/MicroMesoForecast/releases}). We
+have also deposited the v1.0 release on Dryad (\emph{link here after
+acceptance}).
+
+\subsubsection{Statistical models of vital
+rates}\label{statistical-models-of-vital-rates}
 
@@
 regression model for population growth. We describe the statistical
-models separately since fitting the models required different
-approaches. All models contain five climate covariates that we chose
-\emph{a priori}: ``water year'' precipitation at \emph{t}-1 (lagppt);
-April through June precipitation at \emph{t}-1 and \emph{t}-2 (ppt1 and
-ppt2, respectively) and April through June temperature at \emph{t}-1 and
-\emph{t}-2 (TmeanSpr1 and TmeanSpr2, respectively), where \emph{t} is
-the observation year. We also include interactions among same-year
-climate covariates (e.g., ppt1 \(\times\) TmeansSpr1) and climate
-\(\times\) size interactions. Climate \(\times\) size interactions are
-for climate main effects only; we do not include interactions between
-size and pairs of interacting climate effects.
-
-We fit all models using a hierarchical Bayesian approach. The models are
-fully descibed in Appendix A, so here we focus on the main process and
-the model likelihood. For the likelihood models, \(\textbf{y}^X\) is
-always the relevant vector of observations for vital rate \emph{X}
-(\(X = S, G, R, or P\) for survival, growth, recruitment, or population
-growth). For example, \(\textbf{y}^S\) is a vector of 0's and 1's
-indicating whether a genet survives from \emph{t} to \emph{t+1}, or not.
+models separately because they required different approaches. For both
+model types, we fit vital rate models with and without climate
+covariates. Models with climate effects contain five climate covariates
+that we chose \emph{a priori} based on previous model selection efforts
+using these data (Chu et al. 2016) and expert advice (Lance Vermeire,
+\emph{personal communication}): ``water year'' precipitation at
+\emph{t}-2 (lagppt); April through June precipitation at \emph{t}-1 and
+\emph{t} (ppt1 and ppt2, respectively) and April through June
+temperature at \emph{t}-1 and \emph{t} (TmeanSpr1 and TmeanSpr2,
+respectively), where \emph{t}-1 to \emph{t} is the transition of
+interest. We also include interactions among same-year climate
+covariates (e.g., ppt1 \(\times\) TmeansSpr1), resulting in a total of
+seven climate covariates.
+
+We fit all models using a hierarchical Bayesian approach. We focus on
+the main process and the model likelihood in what follows (full model
+descriptions are in the Supporting Information). For the likelihood
+models, \(\textbf{y}^X\) is always the relevant vector of observations
+for vital rate \emph{X} (\emph{X} = \emph{S}, \emph{G}, \emph{R}, or
+\emph{P}) for survival, growth, recruitment, or population growth). For
+example, \(\textbf{y}^S\) is a vector of 0s and 1s indicating whether a
+genet survives from \emph{t} to \emph{t+1}, or not, for all observation
+years and quadrats. All model parameters are species-specific, but we
+omit subscripts for species in model descriptions below to reduce visual
+clutter. For brevity, we only describe models with climate covariates
+included, but models without climate covariates are simply the models
+described below with the climate effects removed.
 
@@
 
-We used logistic regression to model survival probability (\(S\)) of
-genet \(i\) from species \(j\) in quadrat group \(Q\) from time \(t\) to
-\(t+1\):
-
-\begin{align}
-\text{logit}(S_{ijQ,t}) &= \gamma^{S}_{j,t} + \phi^{S}_{jQ} + \beta^{S}_{j,t}x_{ij,t} + \omega^{S}_{j}w_{ij,t} + \nu^{S}_{j}w_{ij,t}x_{ij,t} + \theta^{S}_{jk}C_{k,t} \\
-y^{S}_{ijQ,t} &\sim \text{Bernoulli}(S_{ijQ,t})
-\end{align}
-
-where \(x_{ij,t}\) is the log of genet size, \(\gamma^{S}_{j,t}\) is a
-year-specific intercept, \(\beta^{S}_{j,t}\) is the year-specific slope
-parameter for size, \(\phi^{S}_{jQ}\) is the random effect of quadrat
-group location, and \(\theta^{S}_{k}\) is the fixed parameter for the
-effect of the \(k\)th climate covariate at time \(t\) (\(C_{k,t}\)).
-Note that the vector of climate covariates (\textbf{C}) includes climate
-variable interactions and climate\(\times\)size interactions. We include
-density-dependence by estimating the effect of crowding on the focal
-individual by other individuals of the same species. \(\omega\) is the
-effect of crowding and \(w_{t,Q}\) is the crowding experienced by the
-focal individual at time \(t\) in quadrat group \(Q\). We include a
-size\(\times\)crowding interaction effect (\(\nu^{S}\)).
-
-We modeled growth as a Gaussian process describing genet size at time
-\(t+1\) as a function of size at \(t\) and climate covariates:
-
-\begin{align}
-x_{ijQ,t+1} &= \gamma^{G}_{j,t} + \phi^{G}_{jQ} + \beta^{G}_{j,t}x_{ij,t} + \omega^{G}_{j}w_{ij,t} + \nu^{S}_{j}w_{ij,t}x_{ij,t} + \theta^{G}_{jk}C_{k,t} \\
-y^{G}_{ijQ,t} &\sim \text{Normal}(x_{ijQ,t+1}, \varepsilon_{ij,t})
-\end{align}
-
-where \(x\) is log genet size and all other parameters are as described
-for the survival regression. We capture non-constant error variance in
-growth by modeling the variance around the growth regression
-(\(\varepsilon\)) as a nonlinear function of predicted genet size:
-
-\begin{align}
-\varepsilon_{ij,t} = a e^{b x_{ijQ,t+1}}
-\end{align}
+We used logistic regression to model the probability that genet \emph{i}
+in quadrat \emph{q} survives from time \emph{t} to \emph{t}+1
+(\(s_{i,q,t}\)):
+
+\vspace{-3em}\begin{align}
+y_{i,q,t}^{S} &\sim \text{Bernoulli}(s_{i,q,t}), \\
+\text{logit}(s_{i,q,t}) &= \beta_{0,t} + \beta_{s,t}x_{i,q,t} + \beta_{Q,q} + \textbf{z}'_t \boldsymbol{\beta}_c + \beta_{d,1} w_{i,t} + \beta_{d,2} (x_{i,q,t}w_{i,q,t}),
+\end{align}\vspace{-3em}
+
+where \(x_{i,q,t}\) is the log of genet \emph{i} basal area at time
+\emph{t}, \(\beta_{0,t}\) is a year specific intercept, \(\beta_{Q,q}\)
+is the random effect of the \emph{q}th quadrat to account for spatial
+location, \(\beta_{s,t}\) is the year-specific slope parameter for size,
+\textbf{z} is a vector of \emph{p} climate covariates specific to year
+\emph{t}, \(\boldsymbol{\beta}_c\) is a vector of fixed climate effects
+of length \emph{p}, \(\beta_{d,1}\) is the effect of intraspecific
+crowding experienced by the focal genet at time \emph{t}
+(\(w_{i,q,t}\)), and \(\beta_{d,2}\) is a size by crowding
+(\(x_{i,q,t}w_{i,q,t}\)) interaction effect.
+
+We follow the approach of Chu and Adler (2015) to estimate crowding,
+assuming that the crowding experienced by a focal genet depends on
+distance to each neighbor genet and the neighbor's size, \emph{u}:
+
+\begin{equation}
+w_{i,q,t} = \sum_k e^{-\delta d_{ik,q,t}^{2}}u_{k,q,t}.
+\end{equation}
+
+In equation 3, \(w_{i,t}\) is the crowding that genet \emph{i} in year
+\emph{t} experiences from conspecific neighbors in quadrat \emph{q}. The
+spatial scale over which conspecific neighbors exert influence on any
+genet is determined by \(\delta\). The function is applied for all
+\emph{k} conspecific genets that neighbor the focal genet at time
+\emph{t}, and \(d_{ik,q,t}\) is the distance between genet \emph{i} and
+conspecific genet \emph{k} in quadrat \emph{q}. We use
+regression-specific (survival and growth) \(\delta\) values estimated by
+Chu and Adler (2015).
+
+We modeled growth as a Gaussian process describing log genet size
+(\(y_{i,q,t+1}^{G}\)) at time \(t+1\) in quadrat \emph{q} as a function
+of log size at time \(t\) and climate covariates:
+
+\vspace{-3em}\begin{align}
+y_{i,q,t+1}^G &\sim \text{Normal}(\mu_{i,q,t+1}, \sigma^2_{x_{i,q,t+1}}), \\
+\mu_{i,q,t+1} &= \beta_{0,t} + \beta_{s,t}x_{i,q,t} + \beta_{Q,q} + \textbf{z}'_t \boldsymbol{\beta}_c + \beta_{d,1} w_{i,q,t} + \beta_{d,2} (x_{i,q,t}w_{i,q,t}),
+\end{align}\vspace{-3em}
+
+where \(\mu_{i,q,t+1}\) is log of genet \emph{i}s predicted size at time
+\emph{t} + 1, and all other parameters are as described for the survival
+regression. We capture non-constant error variance in growth by modeling
+the variance in the growth regression (\(\sigma^2_{x{i,q,t+1}}\)) as a
+nonlinear function of predicted genet size:
+
+\vspace{-3em}\begin{align}
+\sigma^2_{x_{i,q,t+1}} = a \, \text{exp}[b \times \mu_{i,q,t+1}],
+\end{align}\vspace{-3em}
+
+where \(\mu_{i,q,t+1}\) is log of predicted genet size predicted from
+the growth regression (Eq. 4), and \emph{a} and \emph{b} are constants.
 
 Our data allows us to track new recruits, but we cannot assign a
-specific parent to new genets. THerefore, we model recruitment at the
-quadrat level: the number of new individuals of species \(j\) in quadrat
-\(q\) recruiting at time \(t+1\) as a function of quadrat ``effective
-cover'' (\(A'\)) in the previous year (\(t\)). Effective cover is a
-mixture of observed cover (\(A\)) in the focal quadrat (\(q\)) and the
-mean cover across the entire group (\(\bar{A}\)) of \(Q\) quadrats in
-which \(q\) is located:
-
-\begin{equation}
-A'_{jq,t} = p_{j}A_{jq,t} + (1-p_{j})\bar{A}_{jQ,t}
-\end{equation}
-
-where \(p\) is a mixing fraction between 0 and 1 that is estimated
-within the model.
-
-We assume the number of individuals, \(y^{R}\), recruiting at time
-\(t+1\) follows a negative binomial distribution:
-
-\begin{equation}
-y^{R}_{jq,t+1} \sim \text{NegBin}(\lambda_{jq,t+1},\zeta)
-\end{equation}
-
-where \(\lambda\) is the mean intensity and \(\zeta\) is the size
-parameter. We define \(\lambda\) as:
-
-\begin{equation}
-\lambda_{jq,t+1} = A'_{jq,t}e^{(\gamma^{R}_{j,t} + \phi^{R}_{jQ} + \theta^{R}_{jk}C_{k,t} + \omega^{R}\sqrt{A'_{q,t}})}
-\end{equation}
-
-where \(A'\) is effective cover (\(\text{cm}^2\)) of species \(j\) in
-quadrat \(q\) and all other terms are as in the survival and growth
-regressions.
+specific parent to new genets. Therefore, we model recruitment at the
+quadrat level. We assume the number of individuals, \(y^{R}_{q,t+1}\),
+recruiting at time \(t+1\) in quadrat \emph{q} follows a negative
+binomial distribution:
+
+\vspace{-3em}\begin{align}
+y^{R}_{q,t+1} \sim \text{NegBin}(\lambda_{q,t+1},\phi),
+\end{align}\vspace{-3em}
+
+where \(\lambda\) is the mean intensity and \(\phi\) is the size
+parameter. We define \(\lambda\) as a function of quadrat composition
+and climate in the previous year:
+
+\vspace{-3em}\begin{align}
+\lambda_{q,t+1} = \tilde{c}_{q,t} \, \text{exp}\left(\beta_{0,t} + \beta_{Q,q} + \textbf{z}'_t \boldsymbol{\beta}_c + \beta_{d}\sqrt{\tilde{c}_{q,t}}\right),
+\end{align}\vspace{-3em}
+
+where \(\tilde{c}_{q,t}\) is effective cover (\(\text{cm}^2\)) of the
+focal species in quadrat \emph{q} at time \emph{t}, and all other terms
+are as in the survival and growth regressions. Effective cover is a
+mixture of observed cover (\emph{c}) in the focal quadrat (\emph{q}) and
+the mean cover across the entire group (\(\bar{c}\)) of \emph{Q}
+quadrats in which \emph{q} is located:
+
+\vspace{-3em}\begin{align}
+\tilde{c}_{q,t} = pc_{q,t} + (1-p)\bar{c}_{Q,t},
+\end{align}\vspace{-3em}
+
+where \(p\) is a mixing fraction between 0 and 1 that is estimated when
+fitting the model.
 
@@
 type of data collected. We have percent cover data, which can easily be
-transformed to proportion data. We first considered fitting three vital
-rate models analagous to those we fit at the individual level: one for
-probability of extirpation within a quadrat (analagous to survival), one
-for cover change within a quadrat (analagous to growth), and one for
-probability of colonization within a quadrat (analagous to recruitment).
-However, within-quadrat extirpation and colonization events were rare in
-our time series (\(N=9\) and \(N=10\), respectively, across all
-species). Given the broad spatial distribution of the quadrats we are
-studying, it is safe to assume that these events are in fact rare enough
-to be ignored for our purposes. So we constrained our statistical
-modeling of vital rates at the population level to change in percent
-cover within quadrats. For the remaining discussion of statistical
-modeling, we refer to proportion data, which is simply percent cover
-divided by 100.
-
-An obvious choice for fitting a linear model to proportion data is beta
-regression because the support of the beta distribution is {[}0,1{]},
-not including true zeros or ones. However, when we used fitted model
-parameters from a beta regression in a quadrat-based population model,
-the simulated population tended toward 100\% cover for all species. We
-therefore chose a more constrained modeling approach based on a
-truncated log-normal likelihood. The model for quadrat cover change from
-time \(t\) to \(t+1\) is
-
-\begin{align}
-x_{jq,t+1} &= \gamma^{P}_{j,t} + \phi^{P}_{jQ} + \beta^{P}_{j,t}x_{jq,t} + \theta^{P}_{jk}C_{k,t} \\
-y^{P}_{jq,t+1} &\sim \text{LogNormal}(x_{jq,t+1}, \tau_{j}) \text{T}[0,1]
-\end{align}
-
-where \(x_{jq,t}\) is the log of species' \(j\) proportional cover in
-quadrat \(q\) at time \(t\) and all other parameters are as in the
-individual-level growth model (Eq. 3). Again, note that the climate
-covariate vector (\textbf{C}) includes the climate\(\times\)cover
-interaction. The log normal likelihood includes a truncation
-(T{[}0,1{]}) to ensure that predicted values do not exceed 100\% cover.
-
-\subsubsection{Model fitting}\label{model-fitting}
+transformed to proportion data in our case because plant areas were
+scaled by plot area. An obvious choice for fitting a linear model to
+proportion data is beta regression because the support of the beta
+distribution is (0,1), which does not include true zeros or ones.
+However, when we used fitted model parameters from a beta regression in
+a quadrat-based population model, the simulated population tended toward
+100\% cover for all species. We therefore chose a modeling approach
+based on a truncated log-normal likelihood. The model for quadrat cover
+change from time \(t\) to \(t+1\) is
+
+\vspace{-3em}\begin{align}
+y_{q,t+1}^{P} &\sim \text{LogNormal}(\mu_{q,t+1}, \sigma^2)_{0}^{1}, \\
+\mu_{q,t+1} &= \beta_{0,t} + \beta_{s,t}x_{q,t} + \beta_{Q,q} + \textbf{z}'_t \boldsymbol{\beta}_c,
+\end{align}\vspace{-3em}
+
+where \(\mu_{q,t+1}\) is the log of proportional cover in quadrat
+\emph{q} at time \(t+1\), and all other parameters are as in the
+individual-level growth model (Eq. 4) except that \emph{x} now
+represents log of proportional cover. The log normal likelihood includes
+a truncation (subscript 0, superscript 1) to ensure that predicted
+values do not exceed 100\% cover.
+
+\subsubsection{Model fitting and statistical
+regularization}\label{model-fitting-and-statistical-regularization}
+
+\paragraph{Model fitting}\label{model-fitting}
 
@@
 appropriate priors for unknown parameters and deciding which, if any, of
-those priors should be hierarchical. We decided to fit models where all
-terms were fit by species. Within a species, we fit yearly size effects
-and yearly intercepts hierarchically where year-specific coefficients
-were drawn from global distributions representing the mean size effect
-and intercept. Quadrat random effects were also fit hierarchically, with
-quadrat offsets being drawn from distributions with mean zero and a
-shared variance term (independent Gaussian priors, Appendix A). Climate
-effects were not modeled hierarchically, and each was given a diffuse
-prior distribution. We used standard diffuse priors for all unknown
-parameters (Appendix A).
+those priors should be hierarchical. For each species, we fit yearly
+size effects and yearly intercepts hierarchically, where year-specific
+coefficients were drawn from global distributions representing the mean
+size effect and intercept. Quadrat random effects were also fit
+hierarchically, with quadrat offsets being drawn from distributions with
+mean zero and a shared variance term (independent Gaussian priors).
+Climate effects were modeled as independent covariates whose prior
+distributions were optimized for prediction using statistical
+regularization (see \textbf{Statistical regularization: Bayesian ridge
+regression} below).
 
 All of our analyses (model fitting and simulating) were conducted in R
-(R Core Development Team 2013). We used the `No-U-Turn' MCMC sampler in
-Stan (Stan Development Team 2014a) to estimate the posterior
-distributions of model parameters using the package `rstan' (Stan
+(R Core Team 2013). We used the `No-U-Turn' Hamiltonian Monte Carlo
+sampler in Stan (Stan Development Team 2014a) to estimate the posterior
+distributions of model parameters using the package \texttt{rstan} (Stan
 Development Team 2014b). We obtained posterior distributions for all
@@
 iterations after discarding an initial 1,000 iterations. Such short MCMC
-chains may surprise readers more familiar with other MCMC samplers
-(i.e.~JAGS or WinBUGS), but the Stan sampler is exceptionally efficient,
-which reduces the number of iterations needed to achieve convergence. We
-assessed convergence visually and made sure scale reduction factors for
-all parameters were less than 1.01. For the purposes of including
-parameter uncertainty in our population models, we saved the final 1,000
+chains are possible because the Stan sampler reduces the number of
+iterations needed to achieve convergence. We assessed convergence
+visually and checked that scale reduction factors for all parameters
+were less than 1.1. There were six instances where the scale reduction
+factor for a particular parameter was greater than 1.1. In those cases,
+we checked the traceplots of the violating parameter and discovered that
+one chain, out of three, was poorly behaved. Therefore, we removed that
+chain from the analysis, leaving two well-mixed and converged chains for
+those six parameters. For the purposes of including parameter
+uncertainty in our population models, we retained the final 1,000
 iterations from each of the three MCMC chains to be used as randomly
-drawn values during population simulation. This step alleviates the need
-to reduce model parameters by model selection since sampling from the
-full parameter space in the MCMC ensures that if a parameter broadly
-overlaps zero, on average the effect in the population models will also
-be near zero. We report the posterior mean, standard deviation, and 95\%
-Bayesian Credible Intervals for every parameter of each model for each
-species in Appendix B.
+drawn values during population simulation. We report the posterior mean,
+standard deviation, and 95\% Bayesian Credible Intervals for every
+parameter of each model for each species in the Supporting Information
+(Tables S5-S20).
+
+\paragraph{Statistical regularization: Bayesian ridge
+regression}\label{statistical-regularization-bayesian-ridge-regression}
+
+For models with climate covariates, our objective is to model the
+response of our focal grass species to interannual variation in climate,
+even if those responses are weak. Therefore, we avoid selecting among
+models with all possible combinations of climate covariates, and instead
+use Bayesian ridge regression to regulate, or constrain, the posterior
+distributions of each climate covariate (Gerber et al. 2015, Hooten and
+Hobbs 2015). Ridge regression is a specific application of statistical
+regularization that seeks to optimize model generality by trading off
+bias and variance. As the name implies, statistical regularization
+involves the use of a regulator that constrains an optimization. The
+natural regulator in a Bayesian application is the prior on the
+coefficient of interest. Each of our statistical models includes the
+effects of climate covariates via the term
+\(\textbf{z}'_t \boldsymbol{\beta}_c\) with prior
+\(\boldsymbol{\beta}_c \sim \text{Normal}(\boldsymbol{\mu}_{\beta_c}, \sigma_{\beta_c}^2\textbf{I})\).
+Because we standardized all climate covariates to have mean zero and
+variance one, we set \(\boldsymbol{\mu}_{\beta_c} = 0\) and let
+\(\sigma_{\beta_c}^2\) serve as the regulator that shrinks covariates
+toward zero -- the smaller the prior variance, the more the posteriors
+of \(\boldsymbol{\beta}_c\) are shrunk toward zero, and the stronger the
+penalty (Hooten and Hobbs 2015).
+
+To find the optimal penalty (i.e., optimal value of the hyperparameter
+\(\sigma_{\beta_c}^2\)), we fit each statistical model with a range of
+values for \(\sigma_{\beta_c}^2\) and compared predictive scores from
+leave-one-year-out cross-validation. We performed the grid search over
+24 values of \(\sigma_{\beta_c}^2\), ranging from
+\(\sigma_{\beta_c}^2 = 0.01\) to \(\sigma_{\beta_c}^2 = 2.25\). For each
+statistical model and each species, we fit \(13\times24=312\) iterations
+of the model fitting algorithm to search \(\sigma_{\beta_c}^2\) for the
+optimal value (13 years to leave out for cross-validation and 24 values
+of \(\sigma_{\beta_c}^2\)) -- a total of 4,992 model fits. We calculated
+the log pointwise predictive density (\emph{lppd}) to score each model's
+ability to predict the left-out data (Gelman et al. 2014). Thus, for
+training data \(y_{\text{train}}\) and held-out data \(y_{\text{hold}}\)
+at a given value of \(\sigma_\theta^2\) across all MCMC samples
+\(s=1,2,...,S\) and all hold outs of data from year \emph{t} to year
+\emph{T}, and letting \(\theta\) represent all unknowns, \emph{lppd} is
+
+\vspace{-3em}\begin{align}
+\text{lppd}_{\text{CV}} = \sum_{t=1}^T \text{log}_e \int [y_{t,\text{hold}}|\theta][\theta|y_{\text{train}}]d\theta,
+\end{align}\vspace{-3em}
+
+and computed as
+
+\vspace{-3em}\begin{align}
+\sum_{t=1}^T \text{log}_e \left(\frac{1}{S} \sum^S_{s=1}(y_{t,\text{hold}}|\theta_{ts})  \right).
+\end{align}\vspace{-3em}
+
+We chose the optimal prior variance for each species-statistical model
+combination as the one that produced the highest \emph{lppd} and then
+fit each species-statistical model combination using the full data set
+for each species and the optimal prior variance. We calculated the
+\emph{lppd} from posterior samples using the algorithm from Vehtari et
+al. (2016).
 
@@
 
-With the posterior distribution of the vital rate statistical models in
-hand, it is straightforward to simulate the population models. We used
-an Integral Projection Model (IPM) to model populations based on
-individual-level data (Ellner and Rees 2006) and a quadrat-based version
-of an individually-based model (Quadrat-Based Model, QBM) to model
-populations based on quadrat-level data. We describe each in turn.
+Using samples from the posterior distribution of the vital rate
+statistical models, it is straightforward to simulate the population
+models. We used an Integral Projection Model (IPM) to simulate
+populations based on individual-level data (Ellner and Rees 2006) and a
+quadrat-based version of an individually-based model (Quadrat-Based
+Model, QBM) to simulate populations based on quadrat-level data. We
+describe each in turn.
 
@@
 
-We use a stochastic IPM (Rees and Ellner 2009) that includes the climate
-covariates from the vital rate statistical models. In all simulations we
-ignore the random year effects so that interannual variation is driven
-solely by climate. We fit the random year effects in the vital rate
-regressions to avoid over-attributing variation to climate covariates.
-Our IPM follows the specification of Chu and Adler (2015) where the
-population of species \emph{j} is a density function \(n(u_{j},t)\)
+We use a stochastic IPM (Rees and Ellner 2009) to simulate our focal
+populations based on the vital rate regressions described above. In all
+simulations, we ignore the random year effects so that interannual
+variation is driven solely by climate. We fit the random year effects in
+the vital rate regressions to avoid over-attributing variation to
+climate covariates. Our IPM follows the specification of Chu and Adler
+(2015) where the population of species \emph{j} is \(n(u_{j},t)\),
 giving the density of sized-\emph{u} genets at time \emph{t}. Genet size
@@
 genets whose area (on the arithmetic scale) is between \(e^{u_{j}}\) and
-\(e^{u_{j}+du}\). The density function for any size \emph{v} at time
-\(t+1\) is
-
-\begin{equation}
-n(v_{j},t+1) = \int_{L_{j}}^{U_{j}} k_{j}(v_{j},u_{j},\bar{w_{j}}(u_{j}))n(u_{j},t)
-\end{equation}
+\(e^{u_{j}+du}\). The function for any size \emph{v} at time \(t+1\) is
+
+\vspace{-3em}\begin{align}
+n(v_{j},t+1) = \int_{L_{j}}^{U_{j}} k_{j}(v_{j},u_{j},\bar{w_{j}}(u_{j}))n(u_{j},t)du_{j},
+\end{align}\vspace{-3em}
 
@@
 
-Since the IPM is spatially-implicit, we cannot calculate neighborhood
+The IPM is spatially-implicit, thus, we cannot calculate neighborhood
 crowding for specific genets (\(w_{ij}\)). Instead, we use an
@@
 
-\begin{equation}
+\vspace{-3em}\begin{align}
 k_{j}(v_{j},u_{j},\bar{w_{j}}) = S_j(u_j, \bar{w_{j}}(u_{j}))G_j(v_{j},u_{j},\bar{w_{j}}(u_{j})) + R_j(v_{j},u_{j},\bar{w_{j}}),
-\end{equation}
+\end{align}\vspace{-3em}
 
@@
 recruited (\emph{R}) individuals of an average sized one-year-old genet
-for the focal species. Our stastical model for recruitment (\emph{R},
-described above) returns the number of new recruit produced per quadrat.
-Following previous work (Adler et al. 2012, Chu and Adler 2015), we
-assume that fecundity increases linearly with size
+for the focal species. Note the \emph{S}, \emph{G}, and \emph{R} are
+incorporated in the IPM using the fitted vital rate regressions. Our
+statistical model for recruitment (\emph{R}, described above) returns
+the number of new recruits produced per quadrat. Following previous work
+(Adler et al. 2012, Chu and Adler 2015), we assume that fecundity
+increases linearly with size
 (\(R_j(v_{j},u_{j},\bar{w_{j}}) = e^{u_j}R_j(v_{j},\bar{w_{j}})\)) to
@@
 We used random draws from the final 1,000 iterations from each of three
-MCMC chains to introduce stochasticity into our population models. At
-each time step, we randomly selected climate covariates from one of the
-14 observed years. Then, we drew the full parameter set (climate effects
-and density-dependence fixed effects) from a randomly selected MCMC
-iteration. Using this approach, rather than simply using coefficient
-point estimates, captures the effect of parameter uncertainty.
-Relatively unimportant climate covariates (those that broadly overlap 0)
-will have little effect on the mean of the simulation results, but can
-contribute to their variation. Since our focus was on the contribution
-of climate covariates to population states, we set the random year
-effects and the random group effects to zero.
+MCMC chains for each vital rate regression to carry-through parameter
+uncertainty into our population models. At each time step, we drew the
+full parameter set (climate effects and density-dependence fixed
+effects) from a randomly selected MCMC iteration. Relatively unimportant
+climate covariates (those that broadly overlap 0) will have little
+effect on the mean of the simulation results, but can contribute to
+their variation. To retain temporal variation associated with random
+year effects, we used posterior estimates of the mean temporal effect
+and the standard deviation of that effect to generate a random year
+effect for unobserved years. That is, for some future year \emph{T}, the
+intercept is
+\(\beta_{0,T} \sim \text{Normal}(\beta_0, \sigma_{\beta_0})\) and the
+effect of size is
+\(\beta_{s,T} \sim \text{Normal}(\beta_s, \sigma_{\beta_s})\).
 
@@
 
-To simulate our quad-based model (QBM), we simply iterate the
-quadrat-level statistical model (Eqs. 9-10). We use the same approach
-for drawing parameter values as described for the IPM. After drawing the
-appropriate parameter set, we calculate the mean response (population
-cover at \emph{t+1} = \(x_{t+1}\)) according to Eq. 9. We then make a
-random draw from a {[}0,1{]} truncated lognormal distribution with mean
-equal to \(x_{t+1}\) from Eq. 9 and the variance estimate from the
-fitted model. We can then project the model forward by drawing a new
-parameter set (unique to climate year and MCMC iteration) at each
-timestep. As with the IPM, random year effects are ignored for all
-simulations.
+To simulate our quad-based model (QBM), we iterate the quadrat-level
+statistical model (Eqs. 9-10). We use the same approach for drawing
+parameter values as described for the IPM. After drawing the appropriate
+parameter set, we calculate the mean response (log cover at \emph{t+1}
+is \(\mu_{t+1}\)) according to Eq. 10. We make a random draw from a
+{[}0,1{]} truncated lognormal distribution with mean equal to
+\(\mu_{t+1}\) from Eq. 10 and the variance estimate from the fitted
+model. We project the model forward by drawing a new parameter set
+(unique to climate year and MCMC iteration) at each timestep. Random
+year effects are included as described above for the IPM.
 
@@
 
-To test each model's ability to forecast population state, we made
+To test each model's ability to forecast population states, we made
 out-of-sample predictions using leave-one-year-out cross validation. For
-both levels of modeling, we fit the vital rate models using observations
-from all years except one, and then used those fitted parameters in the
+both levels of modeling and for models with and without climate
+covariates, we fit the vital rate models using observations from all
+years except one, and then used those fitted parameters in the
 population models to perform a one-step-ahead forecast for the year
@@
 observation year, several quadrats were sampled. We made predictions for
-each observed quadrat in the focal year, initializing each simulation
-with cover in the quadrat the previous year. Since we were making
+each observed quadrat in each focal year, initializing each simulation
+with cover in the quadrat the previous year. Because we were making
 quadrat-specific predictions, we incorporated the group random effect on
@@
 quadrat-year combination with parameter uncertainty included via random
-draw from the MCMC chain as described above. Random year effects were
-set to zero since year effects cannot be assigned to unobserved years.
+draw from the MCMC chain as described above. As described above,
+year-specific parameters for left-out data were drawn from the posterior
+distribution of the mean intercept.
 
 This cross-validation procedure allowed us to compare accuracy and
-precision of the two modeling approaches (IPM versus QBM). We first
-calculated the median predicted cover across the 100 simulations for
-each quadrat-year and then calculated the absolute error as the absolute
-value of the difference between the observed cover for a given
-quadrat-year and the median prediction. To arrive at mean absolute error
-(MAE), we then averaged the absolute error within each species across
-the quadrat-year specific errors. We use MAE as our measure of accuracy.
-To measure precision we calculated the distance between the upper and
-lower 90th quantiles of the 100 predictions and averaged this value over
-quadrat-years for each species.
-
-\subsubsection{Testing sensitivity to climate
-covariates}\label{testing-sensitivity-to-climate-covariates}
-
-With our fitted and validated models in hand, we ran simulations for
-each model type (IPM and QBM) under four climate perturbation scenarios:
-(1) observed climate, (2) precipitation increased by 1\%, (3)
-temperature increased by 1\%, and (4) precipitation and temperature
-increased by 1\%. We ran the simulations for 2,500 time steps, enough to
-estimate equilibrium cover after discarding an initial 500 time steps as
-burn-in. Each simulation was run under two parameter scenarios: (1)
-using mean parameter estimates and (2) using randomly drawn parameters
-from the MCMC chain. We use (1) to detect the overall sensitivity of
-equilibrium cover to climate, and we use (2) to show the impact of model
-and parameter uncertainty on forecast precision.
-
-As an effort to identify potential discrepencies between IPM and QBM
-forecasts, we also ran simulations designed to quantify the
-sensitivities of individual and combined vital rates to climate for the
-IPM. Specifically, we ran simulations for the above climate scenarios,
-but applied the perturbed climate covariates to survival, growth, or
-recruitment vital rates individually and in pairwise combinations. This
-allowed us to isolate the vital rate(s) most sensitive to climate. For
-this analysis, we used mean parameter estimates to reduce the sources of
-uncertainty in the sensitivity estimates.
-
-We expected the IPM to produce more accurate and precise forecasts due
-to either (1) the smaller sample size of the quadrat level data sets
-compared to the individual level data sets, leading to larger parameter
-uncertainty for the QBM, or (2) the QBM climate effects being weakly
-associated with one or more vital rate climate effects at the individual
-level. To assess the impact of sample size on QBM parameter uncertainty
-we refit the QBM statistical model (Eqs. 9-10) after removing sets of 2,
-5, 10, and 15 quadrats. We fit 10 models at each level of quadrat
-removal (2, 5, 10, 15 quadrats), removing a different randomly selected
-set of quadrats for each fit. We calculated the standard deviation of
-climate main effects (pptLag, ppt1, ppt2, TmeanSpr1, and TmeanSpr2) for
-each model and averaged those over replicates within each set of quadrat
-removals. This allowed us to regress parameter uncertainty against
-sample size.
-
-To deterime if the QBM climate effects are correlated with climate
-effects for each vital rate model in the IPM, we simply regressed the
-QBM climate coefficients against each vital rate model's climate
-coefficients and calculated Pearson's \(\rho\). Strong correlations
-indicate the QBM is capable of detecting climate effects associated with
-individual vital rates. A weak correlation indicates the QBM ``misses''
-the climate effect on a particular vital rate.
+precision of the two modeling approaches (IPM versus QBM) with and
+without climate covariates. We first calculated the median predicted
+cover across the 100 simulations for each quadrat-year and then
+calculated forecast skill as the correlation (\(\rho\)) between
+forecasts and observations. We calculated forecast error as mean
+absolute error (MAE) between forecasts and observations. We compared
+\(\rho\) and MAE between model types and within model types between
+models with and without climate covariates using one-sided \emph{t}
+tests with adjusted degrees of freedom following Wilcox (2009) and
+standard errors calculated using the HC4 estimator of Cribari-Neto
+(2004). Statistical tests were conducted using algorithms from Ye et al.
+(2015).
+
+\subsubsection{Forecast horizons}\label{forecast-horizons}
+
+An important feature of any forecasting model is the rate at which
+forecast skill declines as the time between an observation and a
+forecast increases; the so-called ecological forecast horizon (Petchey
+et al. 2015). To assess the forecast horizons of our models, we initiate
+the forecast model with the population state at some time \emph{t} and
+make sequential forecasts of the population at times
+\(t+1, t+2, \dots, t+T\) where \emph{T} is the maximum number of years
+between the initial year and the final year of our observations. For
+example, if we initialize the forecast model with percent cover in 1940,
+we are able to make five forecasts up to the year 1945. Forecast models
+are not re-initialized with observations between years. Thus, in our
+current example, the model forecast for percent cover in 1941 has a
+forecast horizon of one year, the forecast in 1942 has a forecast
+horizon of two years, and so on. We performed these simulations for all
+model types (IPM with/without climate; QBM with/without climate) using
+mean parameter values for all possible initial years. For a given
+forecast horizon, we averaged the correlation between forecasts and
+observations. Note that these forecasts are all made using in-sample
+data because we used model fits from the full data set. Nonetheless,
+these simulations offer insight into the differences among model
+forecast horizons.
 
@@
 
-\subsubsection{Comparison of forecast
-models}\label{comparison-of-forecast-models}
-
-\subsubsection{Sensitivity of models to
-climate}\label{sensitivity-of-models-to-climate}
-
-The response of a population to climate change is a result of the
-aggregate effects of climate on individual vital rates. Since the IPM
-approach relies on vital rate regressions, we were able to quantify the
-sensitivity of each vital rate in isolation and in pairwise
-combinations. Across all species, climate covariates can have opposing
-effects on different vital rates (Fig. 3). Growth was the most sensitive
-vital rate for all species, showing a negative response to increased
-precipitation, and stronger positive response to increased temperature,
-and a mostly positive response when both climate factors are increased
-(Fig. 3). \emph{B. gracilis} survival rates were sensitive to
-temperature, resulting in an increase in plant cover under increased
-temperature (Fig. 3a). In isolation, recruitment and survival were
-insensitive to climate factors for \emph{H. comata} (Fig. 3b). Survival
-and recruitment of \emph{P. smithii} were both sensitive, negatively, to
-temperature and precipitation (Fig. 3c). \emph{P. secunda} equilibrium
-cover was sensitive to the climate effects on survival and recruitment,
-showing a negative effect on both vital rates for increased precipition,
-but a strong positive effect on survival with increased temperature
-(Fig. 3d). Equilibrium cover responded negatively when increased
-precipitation and temperature affect recruitment (Fig. 3d). At least two
-of three vital rates were sensitive to climate for each species (Fig.
-3).
-
-\subsubsection{Sources of uncertainty in the
-QBM}\label{sources-of-uncertainty-in-the-qbm}
-
-Sample size had a relatively weak effect on QBM climate parameter
-uncertainty after the number of quadrats used in fitting exceeded about
-10 (Fig. 5). Inverse-gaussian fits show that increasing sample size
-beyond the number of quadrats we used would result in diminishing
-returns in terms of parameter certainty (Fig. 5).
-
-Climate effects estimated from the QBM are most correlated with climate
-effects from the growth regression at the individual level (Fig. 6). In
-no case does the QBM statistical model have strong correlations across
-all three vital rates (Fig. 6). QBM climate effects were most weakly
-correlated with those from individual-level recruitment models for
-\emph{B. gracilis}, \emph{H. comata}, and \emph{P. secunda} (Fig.
-6a,b,d). For \emph{P. smithii}, QBM climate effects showed no
-correlation with the survival model effects (Fig. 6c).
-
-\subsubsection{Model forecasts}\label{model-forecasts}
-
-Forecasts based on 1\% climate changes were extremely uncertain when we
-considered model error and parameter uncertainty (Fig. 6; simulations
-with mean parameters are in Appendix D for comparison). As expected
-based on model validation (Table 1), QBM projections were more uncertain
-than IPM projections for all species except \emph{P. smithiii} (Fig. 6).
-IPM forecasts for \emph{P. smithiii} were very uncertain due to a very
-high instrinsic rate of recruitment combined with uncertainty in climate
-coefficients which lead to high recruitment boom years and subsequent
-busts when young plants suffer high mortality (Appendx C). When we
-included model error and parameter uncertainty, forecast changes in
-proportional cover always spanned a wide range of negative to positive
-values. In other words, neither model could predict whether a climate
-perturbation would increase or decrease equilibrium population size.
+The IPM and QBM generated one-step-ahead forecasts of similar skill for
+out-of-sample observations, with an average correlation between
+predictions and observations (\(\rho\)) of 0.72 across all models and
+species (Fig. 2). Without climate covariates, the accuracy of forecasts
+from the IPM were not statistically greater than the accuracy of
+forecasts from the QBM (Fig. 2) and overall error was similar (mean
+absolute error; Fig. S1, Supporting Information). With climate
+covariates, the best out-of-sample predictive model (highest
+\emph{lppd}) for each species and vital rate typically resulted from
+highly constrained priors on the climate effects (Fig. S2, Supporting
+Information). Thus, the posterior distributions of climate effects
+included in our models overlapped zero and generally shrunk toward zero,
+though for some species-vital rate combinations, important effects (80\%
+credible interval does not include zero) did emerge (Fig. 3).
+
+Despite the weak climate effects, including climate covariates did
+increase the accuracy of forecasts for two species: \emph{B. gracilis}
+and \emph{Poa secunda} (Fig. 2). However, only for \emph{B. gracilis}
+were the skill increases statistically significant at \(\alpha\) = 0.05
+for the IPM (\(t_{(279)}\) = 1.70, \emph{P} = 0.045) and the QBM
+(\(t_{(279)}\) = 1.80, \emph{P} = 0.037). Similarly, forecast error
+decreased significantly with the inclusion of climate covariates for the
+\emph{B. gracilis} IPM (\(t_{(280)}\) = -3.72, \emph{P} = 0.029) and QBM
+(\(t_{(280)}\) = -3.34, \emph{P} \textless{} 0.0001), and for the
+\emph{Poa secunda} IPM (\(t_{(196)}\) = -1.90, \emph{P} \textless{}
+0.0001) and QBM (\(t_{(196)}\) = -2.47, \emph{P} = 0.007) (Fig. S2,
+Supporting Information). In no case did including climate covariates
+significantly decrease forecast skill (Table S21), despite small changes
+in the mean skill (Fig. 2).
+
+IPM forecasts were significantly more accurate than the QBM in only one
+case (Fig. 2): forecast accuracy of \emph{P. smithii} percent cover from
+an IPM with climate covariates was greater than the accuracy from the
+QBM with climate covariates (\(t_{(215)}\) = 1.92, \emph{P} = 0.028).
+However, adding climate covariates decreased the skill of both models,
+and the difference between the IPM and QBM emerges only because skill
+decreased less for the IPM than the QBM. Results from all pairwise
+statistical tests are shown in Table S22 of the Supporting Information.
+
+With climate covariates included, the accuracy of both models' forecasts
+declined as the forecast horizon increased, but they did so at similar
+rates (Fig. 4). The only exception is for \emph{Poa secunda}, where QBM
+forecast accuracy remained steady as the forecast horizon increases,
+whereas IPM forecast accuracy declined (Fig. 4).
 
@@
 
-Population models built using individual-level data allow inference on
+Population models built using individual-level data provide inference on
 demographic processes, but they can only forecast future population
@@
 al. 2011). However, density-structured models rely on the aggregation of
-individual-level data. This creates a potential problem if such models
-are to be used in a climate change context because it is individuals,
-not populations, which respond to climate (Clark et al. 2012). Are
-models based on population-level metrics as sensitive to climate as
-models based on individual-level metrics? Do these two types of models
-produce consistent forecasts? Do we need detailed demographic data to
-forecast the impacts of climate change?
-
-\subsubsection{The importance of demographic
-data}\label{the-importance-of-demographic-data}
-
-Our comparison of a traditional, demographic population model (the IPM)
-with a model inspired by density-structured models (the QBM) showed that
-the IPM outperformed the QBM: the IPM was more accurate and precise than
-the QBM in out-of-sample cross validation (Table 1). The superiority of
-the IPM could reflect either differences in sample size or the effect of
-averaging over unique effects of climate on each individual-level vital
-rate. Although increasing sample size of quadrat percent-cover
-observations would be easy to do in the field, we found little evidence
-that it would lead to higher precision of climate coefficient estimates
-(Fig. 4).
-
-We did, however, find evidence that the QBM statistical model failed to
-identify climate dependence for some vital rates (Fig. 5). For no
-species were climate effects from the QBM strongly correlated with all
-three vital rates (Fig. 5). Freckleton et al. (2011) acknowledge that
-averaging over complex stage dependence will lead to poorly specified
-models. This is analagous to our situation, but instead of averaging
-over complex life histories, we are averaging over complex climate
-dependence. Though our work here focused on plant species, this finding
-is applicable to any species with vital rates that respond uniquely to
-weather/climate.
-
-Our interpretation is that the QBM is ``missing'' climate signals
-associated with at least one vital rate for each species. This leads to
-inaccurate and imprecise forecasts because the QBM statistical model
-struggles to explain variation due to climate variables that have
-positive and negative impacts on different vital rates. When this is the
-case, as it is for all our species to varying degrees (Fig. 3),
-forecasts from models based on population-level data will fail. Our
-result is consistent with related work on the importance of
-individual-level data to forecast population responses to exogenous
-drivers (Clark et al. 2011a, 2011b, 2012, Galv{á}n et al. 2014).
-
-Detailed demographic data appears to be necessary to forecast climate
-change impacts on plant populations when vital rates have unique climate
-responses. How then can we build models to make forecasts for the
-landscape and regional scales beyond the scope of traditional population
-models (Queenborough et al. 2011)? There are alternatives to
-density-structured models. For example, Clark et al. (2011a) use Forest
-Inventory and Analysis (FIA) data to parameterize a population model
-with multiple vital rates and climate dependence. Distributed efforts
-such as PlantPopNet (\url{http://plantago.plantpopnet.com}) will allow
-researchers to estimate variation around climate responses for
-widespread species by taking advantage of spatial variation in climate
-(e.g. Doak and Morris 2010). Finally, new approaches on the horizon that
-leverage photo/video of plots and advanced object recognition algorithms
-(e.g. Liu et al. 2014) will increase the efficiency of plant mapping and
-digitizing efforts.
-
-\subsubsection{The challenge of
-uncertainty}\label{the-challenge-of-uncertainty}
-
-An important, but unexpected, result of our analysis was the great
-uncertainty in forecasts, even for our best model. The typical approach
-in ecology is to use point estimates of model parameters to project
-populations forward according to the specified model, usually allowing
-for some variability around the determinstic process (e.g. Battin et al.
-2007, Jenouvrier et al. 2009, Adler et al. 2012). If we follow tradition
-and calculate the mean response to climate perturbation with only model
-error and interannual variation included, the IPM and the QBM produce
-opposing forecasts for three of four species (Fig. D1). It would be
-tempting to interpret this inconsistency as further evidence for the
-superiority of the IPM. However, if we introduce parameter uncertainty,
-the forecasts are actually indistinguishable (Fig. 6), though the IPM
-projections are generally more precise (consistent with our
-cross-validation results). The real story is that both models produce
-highly uncertain forecasts. For all species, the 90\% quantiles of
-predicted changes in population size overlapped zero; we cannot even
-predict whether a change in precipitation or temperature will cause
-populations to increase or decrease. This result held when we tried
-perturbing climate by 10\% and 20\% as well.
-
-Our results highlight the state of affairs in ecology when it comes to
-forecasting the impacts of climate change. The analysis we conducted
-here could be considered at the forefront of ecological forecasting with
-respect to the statistical approach employed (hierarchical Bayesian),
-the type of population model we used (density-dependent, stochastic IPM
-with parameter uncertainty), and the amount of high quality data we had
-at our disposal (14 years of individual-level data). Yet, model
-predictions proved so uncertain that any forecast, when bounded with
-model and parameter uncertainty, would be uninformative.
-
-How might we improve on this state of affairs? First, forecasts could be
-improved by matching the spatial scale of predictor variables with the
-spatial scale of observations. One of the major limitations of the
-models we fit here is that the climate data are collected at a larger
-scale than the individual-level observations of plant size. Climate
-covariates only vary by year, with no spatial variability within years.
-Thus, even if we fit models to individual-level data, we are missing the
-key interaction point between weather and individual plants (Clark et
-al. 2011b) because all observations share the same climate covariates.
-Demographic studies should be designed with at least plot-level
-measurements of climate related variables (e.g., soil moisture). Second,
-accurately detecting climate signals will take even longer time series.
-Recent theoretical work on detecting climate signals in noisy data
-suggests that even advanced approaches to parameter fitting require
-20-25 year time series (Teller et al. n.d.). Third, ecologists need a
-stronger commitment to reporting uncertainty. Although most modeling
-studies explicitly consider model uncertainty, parameter uncertainty is
-often ignored. In some cases this is because the most convenient
-statistical methods make it difficult to propogate parameter
-uncertainty. Yet even Bayesian approaches that allow integration of
-model fitting and forecasting (Hobbs and Hooten 2015) are not simple
-when using modeling approaches like integral projection models that
-separate the model fitting and simulation stages (Rees and Ellner 2009).
-However, as we have done here, it is still possible to include parameter
-uncertainty by drawing parameter values from MCMC iterations, taking
-care to draw all parameters from the same chain and iteration to account
-for their correlations. Only by being honest about our forecasts can we
-begin to produce better ones, and forecasts reported without parameter
-error are disingenuous. Ignoring parameter error may be justifiable when
-the goal is investigating basic processes, but it is indefensible when
-forecasting is the goal.
-
-\subsection{Conclusions}\label{conclusions}
-
-This work is not a critique of density-structured population models. We
-are confident that density-structured models will prove to be a valuable
-tool for many applications. However, our analysis represents the first
-comparison, to our knowledge, of population models based on individual
-and aggregated forms of the same data in a climate change context. Our
-results confirm theoretical arguments (Clark et al. 2011b) and empirical
-evidence (Clark et al. 2011a, 2012) that individual responses are
-critical for predicting species' responses to climate change. It seems
-there is no short cut to producing accurate and precise population
-forecasts: we need detailed demographic data to forecast the impacts of
-climate change on populations. Given the importance of demographic data
-and its current collection cost, we need modern methods to collect
-demographic data more efficiently across environmental gradients in
-space and time.
-
-Our results also offer a cautionary tale because forecast uncertainty
-was large for both model types. Even with 14 years of detailed
-demographic data and sophisticated modeling techniques, our projections
-contained too much uncertainty to be informative. Uncertainty in
-demographic responses to climate can be reduced by collecting (1) longer
-time series and (2) climate covariates that match the scale of inference
-(e.g., plot rather than landscape level climate/weather metrics).
+individual-level data. Given that individuals, not populations, respond
+to intraspecific competition and weather (Clark et al. 2011), can models
+based on population-level metrics generate forecasts that are as
+skillful as those generated from models based on individual-level data?
+Are models based on population-level metrics as sensitive to climate
+forcing as models based on individual-level data?
+
+Our comparison of a traditional, demographic population model without
+environmental forcing (the IPM) to an equivalent model inspired by
+density-structured models (the QBM) showed that, generally, IPM
+forecasts of out-of-sample plant population states were no more accurate
+than forecasts from the QBM (Fig. 2; `no-climate' bars). This result
+overturned our expectation that the IPM would out-perform the QBM,
+thanks to its mechanistic representation of the perennial life cycle,
+and confirms theoretical (Freckleton et al. 2011) and empirical work
+(Taylor and Hastings 2004, Queenborough et al. 2011) showing that
+density-structured models can be useful surrogates for demographic
+models when the goal is to estimate or forecast population states over
+large spatial extents.
+
+We also expected the inclusion of environmental forcing to reveal more
+differences between the models. Interannual variation in weather can
+affect vital rates in different ways (Dalgleish et al. 2011), so,
+estimates of climate effects on plant population growth may be biased or
+non-identifiable when the underlying statistical model is fit using
+population-level data that integrates over the potentially unique
+climate responses of individual vital rates. We found some evidence that
+the QBM failed to detect climate effects for three species (\emph{B.
+gracilis}, \emph{H. comata}, and \emph{Poa secunda}), where important
+climate effects were identified in the individual vital rate models but
+not in the percent cover model (Fig. 3). However, for the two species
+where including climate covariates increased forecast accuracy (\emph{B.
+gracilis} and \emph{Poa secunda}), forecast skill (Fig. 2) and error
+(Fig. S2) were equivalent between the IPM and QBM.
+
+The higher accuracy of the IPM and QBM with climate covariates for
+\emph{B. gracilis} and \emph{Poa secunda} highlights the advantage of
+contemporary modeling and variable selection approaches such as ridge
+regression and LASSO over techniques that would exclude
+``non-significant'' effects from final models. Ridge regression allows
+researchers to retain covariates whose effects may be difficult to
+identify in noisy data or short time series. This is especially
+important when forecasting the impacts of climate variability, where it
+is important to include to effects of forcing variables (e.g.,
+temperature and precipitation) even of such effects are difficult to
+identify. Indeed, we failed to detect strong climate effects in the QBM
+for \emph{B. gracilis} and \emph{Poa secunda}, but including climate
+covariates still improved forecasting skill (Fig. 2). If a species is
+truly unresponsive to a given climate variable, statistical
+regularization techniques will shrink the mean and variance of a
+covariate estimate toward zero (Hooten and Hobbs 2015). Of course, no
+matter what model selection approach is adopted, a critical step is
+identifying the appropriate candidate covariates, which we attempted to
+do based on our knowledge of this semi-arid plant community. However,
+the climate covariates we chose required aggregating daily weather data
+over discrete time periods. It is possible that we did not choose the
+optimal time periods over which to aggregate. New methods using
+functional linear models (or splines) may offer a data-driven approach
+for identifying the appropriate time periods over which to aggregate to
+produce a tractable set of candidate climate variables (Teller et al.
+2016).
+
+We also expected IPM forecast accuracy to decline at a lower rate than
+the QBM as the forecast horizon increased. In principle, more
+mechanistic models should produce better predictions, especially under
+novel conditions (Evans 2012, Schindler and Hilborn 2015). In our case,
+the IPM explicitly models the influence of weather on recruitment and
+survival, effects that may be poorly represented in the QBM because
+recruitment and survival mainly affect small plants that contribute
+little to year-to-year changes in percent cover. Over time, of course,
+the addition and subtraction of small plants can have large effects on
+population growth, so explicitly modeling these effects could contribute
+to a longer forecast horizon. However, we found no evidence for a
+difference between the IPM and QBM forecast horizons (Fig. 4).
+
+Our comparison of a model based on individual-level data with one based
+on percent cover data is not an exhaustive test. Understanding the
+reasons why the percent cover-based model matched the skill of a
+demographic model for our focal species may help us anticipate
+situations in which a percent-cover approach would fail. First, for none
+of our species did a climate covariate have a strong negative effect on
+one vital rate and a strong positive effect on a different vital rate
+(Fig. 3). As noted by Freckleton et al. (2011), complex age or stage
+structure can compromise predictions from models that aggregate over
+life-histories, and the same should be true when aggregating across
+vital rates with contrasting responses to climate drivers. Second, our
+particular recruitment model is already so aggregated -- it averages
+across seed production, germination and establishment -- that it may
+fail to detect important demographic responses to climate, putting our
+individual-based model and percent cover model on more equal footing.
+More finely resolved recruitment data might help our individual-based
+model outperform the population-level model. As advocated by Freckleton
+et al. (2011), knowledge of a species' population ecology should guide
+the modeling approach.
+
+In conclusion, we found that models based on individual-level
+demographic data generally failed to generate more skillful population
+forecasts than models based on population-level data. This finding runs
+counter to our expectations, but is consistent with recent theoretical
+(Freckleton et al. 2011) and empirical work (Queenborough et al. 2011).
+Moreover, for the two species where including climate covariates
+improved forecast skill, both models generated equally skillful
+one-step-ahead forecasts. Thus, we conclude that models based on
+population-level data, rather than individual-level data, may be
+adequate for forecasting the states and dynamics of plant populations.
+Unfortunately, our analysis, where climate effects were relatively
+unimportant in vital rate regressions, did not allow us to sufficiently
+test our prediction that individual-level data is neccessary to generate
+skillful forecasts if different vital rates respond to climate in
+unique, potentially opposing, ways. Nonetheless, our results are
+encouraging for the use of easy-to-collect population-level data for
+forecasting the state of plant populations.
 
@@
 This work was funded by the National Science Foundation through a
-Postdoctoral Research Fellowship in Biology to ATT (DBI-1400370) and a
-CAREER award to PBA (DEB-1054040). We thank the original mappers of the
-permanent quadrats in Montana and the digitizers in the Adler lab,
-without whom this work would not have been possible. Informal
-conversations with Stephen Ellner, Giles Hooker, Robin Snyder, and a
-series of meetings between the Adler and Weecology labs at USU sharpened
-our thinking. Brittany Teller provided comments that improved our
-manuscript. Compute, storage and other resources from the Division of
-Research Computing in the Office of Research and Graduate Studies at
-Utah State University are gratefully acknowledged.
+Postdoctoral Research Fellowship in Biology to ATT (DBI-1400370), award
+MSB-1241856 to MBH, and a CAREER award to PBA (DEB-1054040). We thank
+the original mappers of the permanent quadrats in Montana and the
+digitizers in the Adler lab, without whom this work would not have been
+possible. Informal conversations with Stephen Ellner, Giles Hooker,
+Robin Snyder, and a series of meetings between the Adler and Weecology
+labs at USU sharpened our thinking. Brittany Teller provided comments
+that improved our manuscript. Compute, storage and other resources from
+the Division of Research Computing in the Office of Research and
+Graduate Studies at Utah State University are gratefully acknowledged.
+Any use of trade, firm, or product names is for descriptive purposes
+only and does not imply endorsement by the U.S. government. This
+research was supported by the Utah Agricultural Experiment Station, Utah
+State University, and approved as journal paper number 8917.
+
+\subsection{Data Accessibility}\label{data-accessibility}
+
+The data used in this paper have been archived on Ecological Archives:
+\url{http://esapubs.org/archive/ecol/E092/143/}. All data and R code
+necessary to reproduce our work has been deposited on Dryad (link) and
+is also available on GitHub
+(\url{http://github.com/atredennick/MicroMesoForecast}).
 
 \pagebreak{}
-
-\setcounter{page}{29}
-
-\subsection{Tables}\label{tables}
-
-\pagebreak{}
-
-\subsection{Figure Legends}\label{figure-legends}
-
-Figure 1. Time series of average percent cover over all quadrats for our
-four focal species: \emph{Bouteloua gracilis} (BOGR), \emph{Hesperostipa
-comata} (HECO), \emph{Pascopyrum smithii} (PASM), and \emph{Poa secunda}
-(POSE). Light grey lines show trajectories of individual quadrats. Note
-the different y-axis scales across panels.
-
-Figure 2. Work flow of the data aggregation, model fitting, and
-population simulating.
-
-Figure 3. Sensitivity of equilibrium cover simulated from the IPM to
-each climate scenario applied to individual and combined vital rates.
-For example, the points associated with G show the median cover from IPM
-simulations where a climate perturbation is applied only to the growth
-regression climate covariates. These simulations use mean parameter
-values for clarity.
-
-Figure 4. Effect of quadrat sample size on the precision (standard
-deviation) of main climate effect estimates in the QBM. Increasing the
-number of quadrats results in diminishing returns in terms of parameter
-certainty. Light dashed lines show individual climate effects at five
-quadrat sample sizes. Thick dark lines are inverse gaussian fits showing
-the mean effect of increasing quadrat sample size on parameter
-precision.
-
-Figure 5. Correlations (r) between QBM and IPM estimates of climate
-effects. We ignore sizeXclimate interactions since these are not
-directly comparable across model types. The QBM does not have multiple
-vital rates, so its values are repeated across panels within each
-species. Across top panels, `growth' = growth regression, `rec' =
-recruitment regression, `surv' = survival regression.
-
-Figure 6. Mean (points) and 90\% quantiles (errorbars) for the
-proportional difference between baseline simulations (using observed
-climate) and the climate pertubation simulation on the x-axis. We
-calculated proportional difference as log(perturbed climate cover) -
-log(observed climate cover), where `perturbed' and `observed' refer to
-the climate time series used to drive interannual variation in the
-simulations. Model error and parameter uncertainty were propagated
-through the simulation phase. Climate simulations are as in Figure 3.
-
-\pagebreak{}
-
-\subsection{Figures}\label{figures}
-
-\pagebreak{}
-
-\setcounter{page}{24}
 
@@
 92:1703.
-
-Ara{ú}jo, M. B., and A. T. Peterson. 2012. Uses and misuses of
-bioclimatic envelope modeling. Ecology 93:1527--1539.
-
-Battin, J., M. W. Wiley, M. H. Ruckelshaus, R. N. Palmer, E. Korb, K. K.
-Bartz, and H. Imaki. 2007. Projected impacts of climate change on salmon
-habitat restoration. Proceedings of the National Academy of Sciences of
-the United States of America 104:6720--6725.
-
-Chu, C., and P. B. Adler. 2014. When should plant population models
-include age structure? Journal of Ecology 102:531--543.
 
@@
 
+Chu, C., A. R. Kleinhesselink, K. M. Havstad, M. P. McClaran, D. P.
+Peters, L. T. Vermeire, H. Wei, and P. B. Adler. 2016. Direct effects
+dominate responses to climate perturbations in grassland plant
+communities. Nature Communications 7.
+
 Clark, J. S., and O. N. Bj{ø}rnstad. 2004. Population time series:
@@
 
-Clark, J. S., D. M. Bell, M. H. Hersh, and L. Nichols. 2011a. Climate
-change vulnerability of forest biodiversity: Climate and competition
-tracking of demographic rates. Global Change Biology 17:1834--1849.
-
 Clark, J. S., D. M. Bell, M. H. Hersh, M. C. Kwit, E. Moran, C. Salk, A.
-Stine, D. Valle, and K. Zhu. 2011b. Individual-scale variation,
+Stine, D. Valle, and K. Zhu. 2011. Individual-scale variation,
 species-scale differences: Inference needed to understand diversity.
+Ecology Letters 14:1273--1287.
 
@@
 
-Clark, J. S., A. E. Gelfand, C. W. Woodall, and K. Zhu. 2014. More than
-the sum of the parts: Forest climate response from joint species
-distribution models. Ecological Applications 24:990--999.
+Cribari-Neto, F. 2004. Asymptotic inference under heteroskedasticity of
+unknown form. Computational Statistics and Data Analysis 45:215--233.
 
@@
 
-Doak, D. F., and W. F. Morris. 2010. Demographic compensation and
-tipping points in climate-induced range shifts. Nature 467:959--962.
-
-Elith, J., and J. R. Leathwick. 2009. Species Distribution Models:
-Ecological Explanation and Prediction Across Space and Time.
-
 Ellner, S. P., and M. Rees. 2006. Integral projection models for species
 with complex demography. The American naturalist 167:410--428.
+
+Evans, M. R. 2012. Modelling ecological systems in a changing world.
+Philosophical transactions of the Royal Society of London. Series B,
+Biological sciences 367:181--190.
 
@@
 
-Galv{á}n, J. D., J. J. Camarero, and E. Guti{é}rrez. 2014. Seeing the
-trees for the forest: Drivers of individual growth responses to climate
-in Pinus uncinata mountain forests. Journal of Ecology 102:1244--1257.
-
-Guisan, A., and W. Thuiller. 2005. Predicting species distribution:
-Offering more than simple habitat models.
-
-Hobbs, N. T., and M. B. Hooten. 2015. Bayesian Models: A Statistical
-Primer for Ecologists. Princeton University Press, Princeton.
-
-Jenouvrier, S., H. Caswell, C. Barbraud, M. Holland, J. Stroeve, and H.
-Weimerskirch. 2009. Demographic models and IPCC climate projections
-predict the decline of an emperor penguin population. Proceedings of the
-National Academy of Sciences of the United States of America
-106:1844--1847.
+Gelman, A., J. Hwang, and A. Vehtari. 2014. Understanding predictive
+information criteria for Bayesian models. Statistics and Computing
+24:997--1016.
+
+Gerber, B. D., W. L. Kendall, M. B. Hooten, J. A. Dubovsky, and R. C.
+Drewien. 2015. Optimal population prediction of sandhill crane
+recruitment based on climate-mediated habitat limitations. Journal of
+Animal Ecology 84:1299--1310.
+
+Hooten, M. B., and N. T. Hobbs. 2015. A guide to Bayesian model
+selection for ecologists. Ecological Monographs 85:3--28.
 
@@
 
-Liu, Y., Y. Jang, W. Woo, and T.-K. Kim. 2014. Video-Based Object
-Recognition Using Novel Set-of-Sets Representations.
-
 Luo, Y., K. Ogle, C. Tucker, S. Fei, C. Gao, S. LaDeau, J. S. Clark, and
@@
 data-rich era. Ecological Applications 21:1429--1442.
-
-Maiorano, L., R. Cheddadi, N. E. Zimmermann, L. Pellissier, B.
-Petitpierre, J. Pottier, H. Laborde, B. I. Hurdu, P. B. Pearman, A.
-Psomas, J. S. Singarayer, O. Broennimann, P. Vittoz, A. Dubuis, M. E.
-Edwards, H. A. Binney, and A. Guisan. 2013. Building the niche through
-time: using 13,000 years of data to predict the effects of climate
-change on three tree species in Europe. Global Ecology and Biogeography
-22:302--317.
-
-Merow, C., A. M. Latimer, A. M. Wilson, S. M. McMahon, A. G. Rebelo, and
-J. A. Silander. 2014. On using integral projection models to generate
-demographically driven predictions of species' distributions:
-development and validation using sparse data. Ecography 37:1167--1183.
-
-Pearson, R. G., and T. P. Dawson. 2003. Predicting the impacts of
-climate change on the distribution of species: Are bioclimate envelope
-models useful? Global Ecology and Biogeography 12:361--371.
 
@@
 
-Piantadosi, S., D. P. Byar, and S. B. Green. 1988. The Ecological
-Fallacy. American Journal of Epidemiology 127:893--904.
-
 Queenborough, S. A., K. M. Burnet, W. J. Sutherland, A. R. Watkinson,
@@
 
-Schurr, F. M., J. Pagel, J. S. Cabral, J. Groeneveld, O. Bykova, R. B.
-O'Hara, F. Hartig, W. D. Kissling, H. P. Linder, G. F. Midgley, B.
-Schr{ö}der, A. Singer, and N. E. Zimmermann. 2012. How to understand
-species' niches and range dynamics: A demographic research agenda for
-biogeography. Journal of Biogeography 39:2146--2162.
+Schindler, D. E., and R. Hilborn. 2015. Prediction, precaution, and
+policy under global change. Science 347:953--954.
 
@@
 Teller, B. J., P. B. Adler, C. B. Edwards, G. Hooker, R. E. Snyder, and
-S. P. Ellner. (n.d.). Linking demography with drivers: climate and
-competition. Methods in Ecology and Evolution.
+S. P. Ellner. 2016. Linking demography with drivers: climate and
+competition. Methods in Ecology and Evolution 7:171--183.
+
+Vehtari, A., A. Gelman, and J. Gabry. 2016. Efficient implementation of
+leave-one-out cross-validation and WAIC for evaluating fitted Bayesian
+models. ArXiv preprint.
+
+Wilcox, R. R. 2009. Comparing Pearson Correlations: Dealing with
+Heteroscedasticity and Nonnormality. Communications in Statistics -
+Simulation and Computation 38:2220--2234.
+
+Ye, H., R. J. Beamish, S. M. Glaser, S. C. H. Grant, C.-h. Hsieh, L. J.
+Richards, J. T. Schnute, and G. Sugihara. 2015. Equation-free
+mechanistic ecosystem forecasting using empirical dynamic modeling.
+Proceedings of the National Academy of Sciences 112:E1569--E1576.
+
+\newpage{}
+
+\subsection{Tables}\label{tables}
+
+\begin{table}[ht]
+\centering
+\caption{Description of data. The observations span 13 year-to-year transitions.} 
+\begin{tabular}{llrr}
+  \hline
+Species & Vital Rate Model & Num. Obs. & Num. Quadrats\\ 
+  \hline
+  \emph{B. gracilis} & Growth & 5670 & 29 \\ 
+       & Survival & 10102 & 33 \\ 
+       & Recruitment & 304 & 33 \\ 
+       & Percent cover & 281 & 29 \\ 
+  \rule{0pt}{3ex} \emph{H. comata} & Growth & 1990 & 16 \\ 
+       & Survival & 3257 & 18 \\ 
+       & Recruitment & 304 & 18 \\ 
+       & Percent cover & 171 & 17 \\ 
+  \rule{0pt}{3ex} \emph{P. smithii} & Growth & 8052 & 19 \\ 
+       & Survival & 11344 & 19 \\ 
+       & Recruitment & 304 & 19 \\ 
+       & Percent cover & 217 & 19 \\ 
+  \rule{0pt}{3ex} \emph{Poa secunda} & Growth & 3018 & 18 \\ 
+       & Survival & 4650 & 18 \\ 
+       & Recruitment & 304 & 18 \\ 
+       & Percent cover & 197 & 18 \\ 
+   \hline
+\end{tabular}
+\end{table}
+
+\newpage{}
+
+\subsection{Figures}\label{figures}
+
+\begin{figure}[htbp]
+\centering
+\includegraphics{components/figure/manuscript-figure_1-1.pdf}
+\caption{Time series of average percent cover over all quadrats for our
+four focal species: \emph{Bouteloua gracilis} (BOGR), \emph{Hesperostipa
+comata} (HECO), \emph{Pascopyrum smithii} (PASM), and \emph{Poa secunda}
+(POSE). Light grey lines show trajectories of individual quadrats. Note
+the different y-axis scales across panels.}
+\end{figure}
+
+\newpage{}
+
+\begin{figure}[!ht]
+  \centering
+      \includegraphics[height=6in]{./components/forecast_accuracy_mockup.png}
+  \caption{Comparisons of one-step-ahead, out-of-sample forecast accuracy between the IPM and QBM models with and without the inclusion of climate covariates.
+For each comparison, \emph{P}-values are from one-sided \emph{t} tests designed to assess whether the first model in the comparison statement had higher accuracy than the second model in the comparison statement (see details in Table S22).
+In no case did adding climate covariates decrease forecast accuracy (Table S21).
+Species codes are as in Fig.
+1.}
+\end{figure}
+
+\newpage{}
+
+\begin{figure}[!ht]
+  \centering
+      \includegraphics[height=7in]{./components/climate_posteriors_figure.png}
+  \caption{Posterior distributions of climate effects ($\boldsymbol{\beta}_C$) for each species and vital rate statistical model.
+Because our priors were constrained via ridge-regression, we highlight climate effects whose 80\% credible intervals do not overlap zero (red for negative coefficients, blue for positive coefficients).
+Kernel bandwidths of posterior densities were adjusted by a factor of 4 for visual clarity.
+Species codes are as in Fig.
+1.}
+\end{figure}
+
+\newpage{}
+
+\begin{figure}[!ht]
+  \centering
+      \includegraphics[height=5in]{./components/forecast_horizon.png}
+  \caption{The forecast horizons for both models with climate covariates included.
+Points show the average accuracy ($\rho$) across all forecasts at a given time horizon.
+We only examine the forecast accuracy of models with climate covariates included because in no case did including climate covariates significantly decrease accuracy (see Fig.
+2).
+Species codes are as in Fig.
+1.}
+\end{figure}
 

### 2016-07-11 README.Rmd
updates ReadMe abstract

@@
 ### Paper abstract
->Rapid climate change has generated growing interest in forecasts of future population trajectories.
-Traditional population models, typically built using detailed demographic observations from one study site, can address climate change impacts at one location, but are difficult to scale up to the landscape and regional scales relevant to management decisions.
-An alternative is to build models using population-level data that is much easier to collect over broad spatial scales than individual-level data.
-However, such models ignore the fact that climate drives population growth through its influence on individual performance.
-Here, we test the consequences of aggregating individual responses when forecasting climate change impacts on four perennial grass species in a semi-arid grassland in Montana, USA.
-We parameterized two population models based on the same dataset, one based on individual-level data (survival, growth and recruitment) and one on population-level data (percent cover), and compared their accuracy, precision, and sensitivity to climate variables.
-The individual-level model was more accurate and precise than the aggregated model when predicting out-of-sample observations.
-When comparing climate effects from both models, the population-level model missed important climate effects from at least one vital rate for each species.
-Increasing the sample size at the population-level would not necessarily reduce forecast uncertainty; the way to reduce uncertainty is to capture unique climate dependence of individual vital rates.
-Our analysis indicates that there is no shortcut to forecasting climate change impacts on plant populations — detailed demographic data is essential.
-Despite the superiority of the individual-level model, the forecasts it generated still were too uncertain to be useful for decision-makers.
-We need new methods to collect demographic data efficiently across environmental gradients in space and time.
+>1.
+Rapid environmental change has generated growing interest in forecasts of future population trajectories.
+Traditional population models built with detailed demographic observations from one study site can address the impacts of environmental change at particular locations, but are difficult to scale up to the landscape and regional scales relevant to management decisions.
+An alternative is to build models using population-level data which are much easier to collect than individual-level data over broad spatial scales.
+However, it is unknown whether models built using population-level data adequately capture the effects of density-dependence and environmental forcing that are necessary to generate skillful forecasts.
+2.
+Here, we test the consequences of aggregating individual responses when forecasting the population states and trajectories of four perennial grass species in a semi-arid grassland in Montana, USA.
+We parameterized two population models for each species, one based on individual-level data (survival, growth and recruitment) and one on population-level data (percent cover), and compared their forecasting skill and forecast horizons with and without the inclusion of climate covariates.
+For both models we used Bayesian ridge regression to weight the influence of climate covariates for optimal prediction.
+3.
+In the absence of climate effects, we found no significant difference between the forecasting skill of models based on individual-level data and models based on population-level data.
+Climate effects were weak, but increased forecasting skill for two species.
+Increases in skill with climate covariates were similar between model types.
+4.
+In our case study, percent cover models generated forecasts as skillful as those from a demographic model.
+For the goal of forecasting, models based on aggregated individual-level data may offer a practical alternative to data-intensive demographic models.
+Long time series of percent cover data already exist for many plant species.
+Modelers should exploit these data to predict the impacts of environmental change.
 

### 2016-08-25 manuscript/manuscript.Rmd
adds revisions for MEE; R1

@@
    - \usepackage{enumitem}
+   - \usepackage{color, soul}
 bibliography: ~/Dropbox/Bibliography/MicroMeso.bib
@@
 The same is true for density dependence: intraspecific density depedence may act most strongly on vital rates, like recruitment, that are difficult to identify from population-level data.
-If density dependence and/or important climate effects are missed because of the aggregation inherent in in population-level data, then population models built with such data will make uninformative or unreliable forecasts.
+If density dependence and/or important climate effects are missed because of the aggregation inherent in population-level data, then population models built with such data will make uninformative or unreliable forecasts.
 
@@
 
-#### Quad-based model
-To simulate our quad-based model (QBM), we iterate the quadrat-level statistical model (Eqs.
+#### Quadrat-based model
+To simulate our quadrat-based model (QBM), we iterate the quadrat-level statistical model (Eqs.
 9-10).
@@
 For both levels of modeling and for models with and without climate covariates, we fit the vital rate models using observations from all years except one, and then used those fitted parameters in the population models to perform a one-step-ahead forecast for the year whose observations were withheld from model fitting.
-Within each observation year, several quadrats were sampled.
 We made predictions for each observed quadrat in each focal year, initializing each simulation with cover in the quadrat the previous year.
@@
 ### Forecast horizons
-An important feature of any forecasting model is the rate at which forecast skill declines as the time between an observation and a forecast increases; the so-called ecological forecast horizon [@Petchey2015].
+An important feature of any forecasting model is the rate at which forecast skill declines as the time between an observation and a forecast increases.
+\textcolor{blue}{In particular, we are interested in the temporal distance at which forecast skill falls below a threshold: the so-called ecological forecast horizon} [@Petchey2015].
 To assess the forecast horizons of our models, we initiate the forecast model with the population state at some time *t* and make sequential forecasts of the population at times $t+1, t+2, \dots, t+T$ where *T* is the maximum number of years between the initial year and the final year of our observations.
@@
 We performed these simulations for all model types (IPM with/without climate; QBM with/without climate) using mean parameter values for all possible initial years.
-For a given forecast horizon, we averaged the correlation between forecasts and observations.
+For a given forecast distance, we averaged the correlation between forecasts and observations.
 Note that these forecasts are all made using in-sample data because we used model fits from the full data set.
 Nonetheless, these simulations offer insight into the differences among model forecast horizons.
+\textcolor{blue}{We chose a forecast accuracy of $\rho = 0.5$ as our forecast proficiency threshold.
+So, the forecast horizon is the temporal distance at which forecast accuracy falls below $\rho = 0.5$.}
 
@@
 
-With climate covariates included, the accuracy of both models' forecasts declined as the forecast horizon increased, but they did so at similar rates (Fig. 4).
-The only exception is for *Poa secunda*, where QBM forecast accuracy remained steady as the forecast horizon increases, whereas IPM forecast accuracy declined (Fig. 4).
+With climate covariates included \textcolor{blue{and using mean parameter values, the accuracy of both models' forecasts declined as the distance between the last observation and the forecast increased}, but they did so at similar rates (Fig.
+4).
+The only exception is for *Poa secunda*, where QBM forecast accuracy remained steady as the \textcolor{blue}{temporal distance of the forecast increased}, whereas IPM forecast accuracy declined (Fig.
+4).
+\textcolor{blue}{The forecast horizons were short: forecast accuracy fell below $\rho$ = 0.5 after one year for the IPM for most species, and after four years, at most, for the QBM (Fig. 4).
+Across the different temporal distances from the observation to the forecast, the IPM was never more accurate than the QBM (Table S23).
+Likewise, the QBM was rarely more accurate the IPM, the only exception being for \emph{H.
+comata} at temporal distances of two ($t_{(115)}$ = 2.39, \emph{P} = 0.002) and three years ($t_{(98)}$ = 2.04, \emph{P} = 0.022) (Table S24).
+There were some cases where the QBM was more accurate than the IPM for \emph{Poa secunda}, but neither model exceeded the forecast profieciency threshold by a large margin (Fig. 4, Table S24).}
 
@@
 ----------
+<!--
 Population models built using individual-level data provide inference on demographic processes, but they can only forecast future population states across the (typically limited) spatial extent of the observations.
@@
 Are models based on population-level metrics as sensitive to climate forcing as models based on individual-level data?
+-->
 
@@
 Ridge regression allows researchers to retain covariates whose effects may be difficult to identify in noisy data or short time series.
-This is especially important when forecasting the impacts of climate variability, where it is important to include to effects of forcing variables (e.g., temperature and precipitation) even of such effects are difficult to identify.
+This is especially important when forecasting the impacts of climate variability, where it is important to include to effects of forcing variables (e.g., temperature and precipitation) even if such effects are difficult to identify.
 Indeed, we failed to detect strong climate effects in the QBM for *B. gracilis* and *Poa secunda*, but including climate covariates still improved forecasting skill (Fig. 2).
@@
 As advocated by @Freckleton2011, knowledge of a species' population ecology should guide the modeling approach.
+\textcolor{blue}{Third, our percent cover data are essentially error-free because we were able to aggregate indiviual plant areas to calculate percent cover.
+Percent cover data collected by typical sampling protocals (e.g., Daubenmire frames) will include error that would affect population forecasts due to misspecifing the initial conditions and/or biasing model parameters} [@Queenborough2011].
+\textcolor{blue}{Thus, it may be that our percent cover model performed just as well as the individual-based model because percent cover was perfectly measured.
+One way to account for such error is to develop a sampling model that relates the observations (estimated percent cover in a plot) to the true state (percent cover derived from individual plant measurements in the same plot)} [@Hobbs2015].
+
+\textcolor{blue}{Although our main goal was to compare individual-based and population-level modeling approaches relative to one another, it is worth reflecting on the absolute forecasting skill of our models.
+In particular, the forecast horizon of both models, defined as the time horizon at which the correlation between predictions and observations falls below $\rho = 0.5$, is less than five years for all species.
+Such short forecast horizons are not encouraging.}
+\textcolor{blue}{Unfortunately, we have little to offer on how to improve population forecasts that has not already been proposed} [@Petchey2015; @Mouquet2015].
+\textcolor{blue}{Longer time-series should improve our ability to detect exogenous drivers such as climate} [@Teller2016], \textcolor{blue}{and modeling larger spatial extents may reduce parameter uncertainty} [@Petchey2015].
+\textcolor{blue}{We may also have to shift our perspective from making explicit point forecasts to making moving average forecasts} [@Petchey2015].
+\textcolor{blue}{For example, Integral Projection Models are typically capable of estimating equilibrium abundances} [e.g., @Adler2012], \textcolor{blue}{which may be more appropriate for dynamic systems.}
 
@@
 The data used in this paper have been archived on Ecological Archives: http://esapubs.org/archive/ecol/E092/143/.
-All data and R code necessary to reproduce our work has been deposited on Dryad (link) and is also available on GitHub (http://github.com/atredennick/MicroMesoForecast).
+All data and R code necessary to reproduce our work has been deposited on \textcolor{blue}{Figshare} (link) and is also available on GitHub (http://github.com/atredennick/MicroMesoForecast).
 
@@
   \centering
-      \includegraphics[height=6in]{./components/forecast_accuracy_mockup.png}
+      \includegraphics[height=6in]{./components/forecast_accuracy_mockup_boxplot.png}
   \caption{Comparisons of one-step-ahead, out-of-sample forecast accuracy between the IPM and QBM models with and without the inclusion of climate covariates.
-For each comparison, \emph{P}-values are from one-sided \emph{t} tests designed to assess whether the first model in the comparison statement had higher accuracy than the second model in the comparison statement (see details in Table S22).
-In no case did adding climate covariates decrease forecast accuracy (Table S21).
+\textcolor{blue}{Boxplots show the distribution of $\rho$ averaged over quadrats for each cross-validation year (i.e., 13 values of $\rho$ for each species-model combination).} For each comparison, \emph{P}-values are from one-sided \emph{t} tests designed to assess whether the first model in the comparison statement had higher accuracy than the second model in the comparison statement (see details in Table S22).
+\textcolor{blue}{Statistical tests relied on correlation values for each quadrat-year-species combination, after averaging over model reps for each combination.} In no case did adding climate covariates decrease forecast accuracy (Table S21).
 Species codes are as in Fig.
@@
 Species codes are as in Fig.
-1.}
+1.
+\textcolor{blue}{Climate covariate codes: \emph{pptLag} = "water year" precipitation at \emph{t}-2; \emph{ppt1} = April through June precipitation at \emph{t}-1; \emph{ppt2} = April through June precipitation at \emph{t}; \emph{TmeanSpr1} = April through June temperature at \emph{t}-1; \emph{TmeanSpr2} = April through June temperature at \emph{t}.}}
 \end{figure}
@@
       \includegraphics[height=5in]{./components/forecast_horizon.png}
-  \caption{The forecast horizons for both models with climate covariates included.
-Points show the average accuracy ($\rho$) across all forecasts at a given time horizon.
-We only examine the forecast accuracy of models with climate covariates included because in no case did including climate covariates significantly decrease accuracy (see Fig.
+  \caption{The forecast horizons for both models with climate covariates included \textcolor{blue}{and using mean parameter values.
+Points show the average accuracy ($\rho$) across all forecasts at a given distance between the last observation and the forecast, where forecasts are made for in-sample data.} We only examine the forecast accuracy of models with climate covariates included because in no case did including climate covariates significantly decrease accuracy (see Fig.
 2).
+\textcolor{blue}{The dashed blue line indicates a forecast proficiency threshold of $\rho = 0.5$.
 Species codes are as in Fig.
-1.}
+1 and statistical comparisons between the IPM and QBM at each forecast distance are in Tables S23 and S24.}}
 \end{figure}

### 2016-08-25 manuscript/manuscript.tex
adds revisions for MEE; R1

@@
 \usepackage{enumitem}
+\usepackage{color, soul}
 
@@
 Geological Survey, Colorado Cooperative Fish and Wildlife Research Unit, Fort Collins, CO 80523, USA; \textsuperscript{3}Department of Fish, Wildlife, and Conservation Biology, Colorado State University, Fort Collins, CO 80523 USA}; \textsuperscript{4}Department of Statistics, Colorado State University, Fort Collins, CO 80523 USA}}
-
-\textbf{Running head}: Demographic data and population forecasting \\
-\textbf{Word count}: 6845 (including reference section) \\
-\textbf{Corresponding author}: \\
-Andrew Tredennick (atredenn@gmail.com) \\
-Department of Wildland Resources and the Ecology Center \\
-5230 Old Main Hill \\ 
-Utah State University \\ 
-Logan, Utah 84322, USA
 
@@
 important climate effects are missed because of the aggregation inherent
-in in population-level data, then population models built with such data
+in population-level data, then population models built with such data
 will make uninformative or unreliable forecasts.
@@
 
-\paragraph{Quad-based model}\label{quad-based-model}
-
-To simulate our quad-based model (QBM), we iterate the quadrat-level
+\paragraph{Quadrat-based model}\label{quadrat-based-model}
+
+To simulate our quadrat-based model (QBM), we iterate the quadrat-level
 statistical model (Eqs. 9-10). We use the same approach for drawing
@@
 population models to perform a one-step-ahead forecast for the year
-whose observations were withheld from model fitting. Within each
-observation year, several quadrats were sampled. We made predictions for
-each observed quadrat in each focal year, initializing each simulation
-with cover in the quadrat the previous year. Because we were making
-quadrat-specific predictions, we incorporated the group random effect on
-the intercept for both models. We repeated this procedure for all 13
-observation years, making 100 one-step-ahead forecasts for each
+whose observations were withheld from model fitting. We made predictions
+for each observed quadrat in each focal year, initializing each
+simulation with cover in the quadrat the previous year. Because we were
+making quadrat-specific predictions, we incorporated the group random
+effect on the intercept for both models. We repeated this procedure for
+all 13 observation years, making 100 one-step-ahead forecasts for each
 quadrat-year combination with parameter uncertainty included via random
@@
 forecast skill declines as the time between an observation and a
-forecast increases; the so-called ecological forecast horizon (Petchey
-et al. 2015). To assess the forecast horizons of our models, we initiate
-the forecast model with the population state at some time \emph{t} and
-make sequential forecasts of the population at times
+forecast increases.
+\textcolor{blue}{In particular, we are interested in the temporal distance at which forecast skill falls below a threshold: the so-called ecological forecast horizon}
+(Petchey et al. 2015). To assess the forecast horizons of our models, we
+initiate the forecast model with the population state at some time
+\emph{t} and make sequential forecasts of the population at times
 \(t+1, t+2, \dots, t+T\) where \emph{T} is the maximum number of years
@@
 mean parameter values for all possible initial years. For a given
-forecast horizon, we averaged the correlation between forecasts and
+forecast distance, we averaged the correlation between forecasts and
 observations. Note that these forecasts are all made using in-sample
@@
 forecast horizons.
+\textcolor{blue}{We chose a forecast accuracy of $\rho = 0.5$ as our forecast proficiency threshold.
+So, the forecast horizon is the temporal distance at which forecast accuracy falls below $\rho = 0.5$.}
 
@@
 
-With climate covariates included, the accuracy of both models' forecasts
-declined as the forecast horizon increased, but they did so at similar
-rates (Fig. 4). The only exception is for \emph{Poa secunda}, where QBM
-forecast accuracy remained steady as the forecast horizon increases,
-whereas IPM forecast accuracy declined (Fig. 4).
+With climate covariates included \textbackslash{}textcolor\{blue\{and
+using mean parameter values, the accuracy of both models' forecasts
+declined as the distance between the last observation and the forecast
+increased\}, but they did so at similar rates (Fig. 4). The only
+exception is for \emph{Poa secunda}, where QBM forecast accuracy
+remained steady as the
+\textcolor{blue}{temporal distance of the forecast increased}, whereas
+IPM forecast accuracy declined (Fig. 4).
+\textcolor{blue}{The forecast horizons were short: forecast accuracy fell below $\rho$ = 0.5 after one year for the IPM for most species, and after four years, at most, for the QBM (Fig. 4).
+Across the different temporal distances from the observation to the forecast, the IPM was never more accurate than the QBM (Table S23).
+Likewise, the QBM was rarely more accurate the IPM, the only exception being for \emph{H.
+comata} at temporal distances of two ($t_{(115)}$ = 2.39, \emph{P} = 0.002) and three years ($t_{(98)}$ = 2.04, \emph{P} = 0.022) (Table S24).
+There were some cases where the QBM was more accurate than the IPM for \emph{Poa secunda}, but neither model exceeded the forecast profieciency threshold by a large margin (Fig. 4, Table S24).}
 
 \subsection{Discussion}\label{discussion}
-
-Population models built using individual-level data provide inference on
-demographic processes, but they can only forecast future population
-states across the (typically limited) spatial extent of the
-observations. Population-level data are much easier to collect across
-broad spatial extents, so models built using such data offer an
-appealing alternative to traditional population models (Queenborough et
-al. 2011). However, density-structured models rely on the aggregation of
-individual-level data. Given that individuals, not populations, respond
-to intraspecific competition and weather (Clark et al. 2011), can models
-based on population-level metrics generate forecasts that are as
-skillful as those generated from models based on individual-level data?
-Are models based on population-level metrics as sensitive to climate
-forcing as models based on individual-level data?
 
@@
 is important to include to effects of forcing variables (e.g.,
-temperature and precipitation) even of such effects are difficult to
+temperature and precipitation) even if such effects are difficult to
 identify. Indeed, we failed to detect strong climate effects in the QBM
@@
 the modeling approach.
+\textcolor{blue}{Third, our percent cover data are essentially error-free because we were able to aggregate indiviual plant areas to calculate percent cover.
+Percent cover data collected by typical sampling protocals (e.g., Daubenmire frames) will include error that would affect population forecasts due to misspecifing the initial conditions and/or biasing model parameters}
+(Queenborough et al. 2011).
+\textcolor{blue}{Thus, it may be that our percent cover model performed just as well as the individual-based model because percent cover was perfectly measured.
+One way to account for such error is to develop a sampling model that relates the observations (estimated percent cover in a plot) to the true state (percent cover derived from individual plant measurements in the same plot)}
+(Hobbs and Hooten 2015).
+
+\textcolor{blue}{Although our main goal was to compare individual-based and population-level modeling approaches relative to one another, it is worth reflecting on the absolute forecasting skill of our models.
+In particular, the forecast horizon of both models, defined as the time horizon at which the correlation between predictions and observations falls below $\rho = 0.5$, is less than five years for all species.
+Such short forecast horizons are not encouraging.}
+\textcolor{blue}{Unfortunately, we have little to offer on how to improve population forecasts that has not already been proposed}
+(Mouquet et al. 2015, Petchey et al. 2015).
+\textcolor{blue}{Longer time-series should improve our ability to detect exogenous drivers such as climate}
+(Teller et al. 2016),
+\textcolor{blue}{and modeling larger spatial extents may reduce parameter uncertainty}
+(Petchey et al. 2015).
+\textcolor{blue}{We may also have to shift our perspective from making explicit point forecasts to making moving average forecasts}
+(Petchey et al. 2015).
+\textcolor{blue}{For example, Integral Projection Models are typically capable of estimating equilibrium abundances}
+(e.g., Adler et al. 2012),
+\textcolor{blue}{which may be more appropriate for dynamic systems.}
 
@@
 \url{http://esapubs.org/archive/ecol/E092/143/}. All data and R code
-necessary to reproduce our work has been deposited on Dryad (link) and
-is also available on GitHub
+necessary to reproduce our work has been deposited on
+\textcolor{blue}{Figshare} (link) and is also available on GitHub
 (\url{http://github.com/atredennick/MicroMesoForecast}).
@@
 \pagebreak{}
-
-\subsection*{References}\label{references}
-\addcontentsline{toc}{subsection}{References}
-
-Adler, P. B., H. J. Dalgleish, and S. P. Ellner. 2012. Forecasting plant
-community impacts of climate variability and change: when do competitive
-interactions matter? Journal of Ecology 100:478--487.
-
-Adler, P. B., S. P. Ellner, and J. M. Levine. 2010. Coexistence of
-perennial plants: An embarrassment of niches. Ecology Letters
-13:1019--1029.
-
-Anderson, J., L. Vermeire, and P. B. Adler. 2011. Fourteen years of
-mapped, permanent quadrats in a northern mixed prairie, USA. Ecology
-92:1703.
-
-Chu, C., and P. B. Adler. 2015. Large niche differences emerge at the
-recruitment stage to stabilize grassland coexistence. Ecological
-Monographs 85:373--392.
-
-Chu, C., K. M. Havstad, N. Kaplan, W. K. Lauenroth, M. P. McClaran, D.
-P. Peters, L. T. Vermeire, and P. B. Adler. 2014. Life form influences
-survivorship patterns for 109 herbaceous perennials from six semi-arid
-ecosystems. Journal of Vegetation Science 25:947--954.
-
-Chu, C., A. R. Kleinhesselink, K. M. Havstad, M. P. McClaran, D. P.
-Peters, L. T. Vermeire, H. Wei, and P. B. Adler. 2016. Direct effects
-dominate responses to climate perturbations in grassland plant
-communities. Nature Communications 7.
-
-Clark, J. S., and O. N. Bj{ø}rnstad. 2004. Population time series:
-Process variability, observation errors, missing values, lags, and
-hidden states. Ecology 85:3140--3150.
-
-Clark, J. S., D. M. Bell, M. H. Hersh, M. C. Kwit, E. Moran, C. Salk, A.
-Stine, D. Valle, and K. Zhu. 2011. Individual-scale variation,
-species-scale differences: Inference needed to understand diversity.
-Ecology Letters 14:1273--1287.
-
-Clark, J. S., D. M. Bell, M. Kwit, A. Stine, B. Vierra, and K. Zhu.
-2012. Individual-scale inference to anticipate climate-change
-vulnerability of biodiversity. Philosophical Transactions of the Royal
-Society B: Biological Sciences 367:236--246.
-
-Clark, J. S., D. Bell, C. Chu, B. Courbaud, M. Dietze, M. Hersh, J.
-HilleRisLambers, I. Ib{á}{ñ}ez, S. LaDeau, S. McMahon, J. Metcalf, J.
-Mohan, E. Moran, L. Pangle, S. Pearson, C. Salk, Z. Shen, D. Valle, and
-P. Wyckoff. 2010. High-dimensional coexistence based on individual
-variation: a synthesis of evidence. Ecological Monographs 80:569--608.
-
-Clark, J. S., S. R. Carpenter, M. Barber, S. Collins, A. Dobson, J. A.
-Foley, D. M. Lodge, M. Pascual, R. Pielke, W. Pizer, C. Pringle, W. V.
-Reid, K. A. Rose, O. Sala, W. H. Schlesinger, D. H. Wall, and D. Wear.
-2001. Ecological forecasts: an emerging imperative. Science (New York,
-N.Y.) 293:657--660.
-
-Cribari-Neto, F. 2004. Asymptotic inference under heteroskedasticity of
-unknown form. Computational Statistics and Data Analysis 45:215--233.
-
-Dalgleish, H. J., D. N. Koons, M. B. Hooten, C. A. Moffet, and P. B.
-Adler. 2011. Climate influences the demography of three dominant
-sagebrush steppe plants. Ecology 92:75--85.
-
-Ellner, S. P., and M. Rees. 2006. Integral projection models for species
-with complex demography. The American naturalist 167:410--428.
-
-Evans, M. R. 2012. Modelling ecological systems in a changing world.
-Philosophical transactions of the Royal Society of London. Series B,
-Biological sciences 367:181--190.
-
-Freckleton, R. P., W. J. Sutherland, A. R. Watkinson, and S. A.
-Queenborough. 2011. Density-structured models for plant population
-dynamics. American Naturalist 177:1--17.
-
-Gelman, A., J. Hwang, and A. Vehtari. 2014. Understanding predictive
-information criteria for Bayesian models. Statistics and Computing
-24:997--1016.
-
-Gerber, B. D., W. L. Kendall, M. B. Hooten, J. A. Dubovsky, and R. C.
-Drewien. 2015. Optimal population prediction of sandhill crane
-recruitment based on climate-mediated habitat limitations. Journal of
-Animal Ecology 84:1299--1310.
-
-Hooten, M. B., and N. T. Hobbs. 2015. A guide to Bayesian model
-selection for ecologists. Ecological Monographs 85:3--28.
-
-Lauenroth, W. K., and P. B. Adler. 2008. Demography of perennial
-grassland plants: Survival, life expectancy and life span. Journal of
-Ecology 96:1023--1032.
-
-Luo, Y., K. Ogle, C. Tucker, S. Fei, C. Gao, S. LaDeau, J. S. Clark, and
-D. S. Schimel. 2011. Ecological forecasting and data assimilation in a
-data-rich era. Ecological Applications 21:1429--1442.
-
-Perretti, C. T., S. B. Munch, and G. Sugihara. 2013. Model-free
-forecasting outperforms the correct mechanistic model for simulated and
-experimental data. Proceedings of the National Academy of Sciences of
-the United States of America 110:5253--5257.
-
-Petchey, O. L., M. Pontarp, T. M. Massie, S. K{é}fi, A. Ozgul, M.
-Weilenmann, G. M. Palamara, F. Altermatt, B. Matthews, J. M. Levine, D.
-Z. Childs, B. J. McGill, M. E. Schaepman, B. Schmid, P. Spaak, A. P.
-Beckerman, F. Pennekamp, and I. S. Pearse. 2015. The ecological forecast
-horizon, and examples of its uses and determinants. Ecology Letters
-18:597--611.
-
-Queenborough, S. A., K. M. Burnet, W. J. Sutherland, A. R. Watkinson,
-and R. P. Freckleton. 2011. From meso- to macroscale population
-dynamics: A new density-structured approach. Methods in Ecology and
-Evolution 2:289--302.
-
-R Core Team. 2013. R: A language and environment for statistical
-computing.
-
-Rees, M., and S. P. Ellner. 2009. Integral projection models for
-populations in temporally varying environments. Ecological Monographs
-79:575--594.
-
-Schindler, D. E., and R. Hilborn. 2015. Prediction, precaution, and
-policy under global change. Science 347:953--954.
-
-Stan Development Team. 2014a. Stan: A C++ Library for Probability and
-Sampling, Version 2.5.0.
-
-Stan Development Team. 2014b. Rstan: the R interface to Stan, Version
-2.5.0.
-
-S{æ}ther, B. E., S. Engen, V. Gr{ø}tan, W. Fiedler, E. Matthysen, M. E.
-Visser, J. Wright, A. P. M{ø}ller, F. Adriaensen, H. {Van Balen}, D.
-Balmer, M. C. Mainwaring, R. H. McCleery, M. Pampus, and W. Winkel.
-2007. The extended Moran effect and large-scale synchronous fluctuations
-in the size of great tit and blue tit populations. Journal of Animal
-Ecology 76:315--325.
-
-Taylor, C. M., and A. Hastings. 2004. Finding optimal control strategies
-for invasive species: a density-structured model for Spartina
-alterniflora. Journal of Applied Ecology 41:1049--1057.
-
-Teller, B. J., P. B. Adler, C. B. Edwards, G. Hooker, R. E. Snyder, and
-S. P. Ellner. 2016. Linking demography with drivers: climate and
-competition. Methods in Ecology and Evolution 7:171--183.
-
-Vehtari, A., A. Gelman, and J. Gabry. 2016. Efficient implementation of
-leave-one-out cross-validation and WAIC for evaluating fitted Bayesian
-models. ArXiv preprint.
-
-Wilcox, R. R. 2009. Comparing Pearson Correlations: Dealing with
-Heteroscedasticity and Nonnormality. Communications in Statistics -
-Simulation and Computation 38:2220--2234.
-
-Ye, H., R. J. Beamish, S. M. Glaser, S. C. H. Grant, C.-h. Hsieh, L. J.
-Richards, J. T. Schnute, and G. Sugihara. 2015. Equation-free
-mechanistic ecosystem forecasting using empirical dynamic modeling.
-Proceedings of the National Academy of Sciences 112:E1569--E1576.
-
-\newpage{}
 
@@
       \includegraphics[height=5in]{./components/forecast_horizon.png}
-  \caption{The forecast horizons for both models with climate covariates included.
-Points show the average accuracy ($\rho$) across all forecasts at a given time horizon.
-We only examine the forecast accuracy of models with climate covariates included because in no case did including climate covariates significantly decrease accuracy (see Fig.
+  \caption{The forecast horizons for both models with climate covariates included \textcolor{blue}{and using mean parameter values.
+Points show the average accuracy ($\rho$) across all forecasts at a given distance between the last observation and the forecast, where forecasts are made for in-sample data.} We only examine the forecast accuracy of models with climate covariates included because in no case did including climate covariates significantly decrease accuracy (see Fig.
 2).
+\textcolor{blue}{The dashed blue line indicates a forecast proficiency threshold of $\rho = 0.5$.
 Species codes are as in Fig.
-1.}
+1 and statistical comparisons between the IPM and QBM at each forecast distance are in Tables S23 and S24.}}
 \end{figure}
 
+\newpage{}
+
+\singlespace{}
+
+\subsection*{References}\label{references}
+\addcontentsline{toc}{subsection}{References}
+
+Adler, P. B., H. J. Dalgleish, and S. P. Ellner. 2012. Forecasting plant
+community impacts of climate variability and change: when do competitive
+interactions matter? Journal of Ecology 100:478--487.
+
+Adler, P. B., S. P. Ellner, and J. M. Levine. 2010. Coexistence of
+perennial plants: An embarrassment of niches. Ecology Letters
+13:1019--1029.
+
+Anderson, J., L. Vermeire, and P. B. Adler. 2011. Fourteen years of
+mapped, permanent quadrats in a northern mixed prairie, USA. Ecology
+92:1703.
+
+Chu, C., and P. B. Adler. 2015. Large niche differences emerge at the
+recruitment stage to stabilize grassland coexistence. Ecological
+Monographs 85:373--392.
+
+Chu, C., K. M. Havstad, N. Kaplan, W. K. Lauenroth, M. P. McClaran, D.
+P. Peters, L. T. Vermeire, and P. B. Adler. 2014. Life form influences
+survivorship patterns for 109 herbaceous perennials from six semi-arid
+ecosystems. Journal of Vegetation Science 25:947--954.
+
+Chu, C., A. R. Kleinhesselink, K. M. Havstad, M. P. McClaran, D. P.
+Peters, L. T. Vermeire, H. Wei, and P. B. Adler. 2016. Direct effects
+dominate responses to climate perturbations in grassland plant
+communities. Nature Communications 7.
+
+Clark, J. S., and O. N. Bj{ø}rnstad. 2004. Population time series:
+Process variability, observation errors, missing values, lags, and
+hidden states. Ecology 85:3140--3150.
+
+Clark, J. S., D. M. Bell, M. H. Hersh, M. C. Kwit, E. Moran, C. Salk, A.
+Stine, D. Valle, and K. Zhu. 2011. Individual-scale variation,
+species-scale differences: Inference needed to understand diversity.
+Ecology Letters 14:1273--1287.
+
+Clark, J. S., D. M. Bell, M. Kwit, A. Stine, B. Vierra, and K. Zhu.
+2012. Individual-scale inference to anticipate climate-change
+vulnerability of biodiversity. Philosophical Transactions of the Royal
+Society B: Biological Sciences 367:236--246.
+
+Clark, J. S., D. Bell, C. Chu, B. Courbaud, M. Dietze, M. Hersh, J.
+HilleRisLambers, I. Ib{á}{ñ}ez, S. LaDeau, S. McMahon, J. Metcalf, J.
+Mohan, E. Moran, L. Pangle, S. Pearson, C. Salk, Z. Shen, D. Valle, and
+P. Wyckoff. 2010. High-dimensional coexistence based on individual
+variation: a synthesis of evidence. Ecological Monographs 80:569--608.
+
+Clark, J. S., S. R. Carpenter, M. Barber, S. Collins, A. Dobson, J. A.
+Foley, D. M. Lodge, M. Pascual, R. Pielke, W. Pizer, C. Pringle, W. V.
+Reid, K. A. Rose, O. Sala, W. H. Schlesinger, D. H. Wall, and D. Wear.
+2001. Ecological forecasts: an emerging imperative. Science (New York,
+N.Y.) 293:657--660.
+
+Cribari-Neto, F. 2004. Asymptotic inference under heteroskedasticity of
+unknown form. Computational Statistics and Data Analysis 45:215--233.
+
+Dalgleish, H. J., D. N. Koons, M. B. Hooten, C. A. Moffet, and P. B.
+Adler. 2011. Climate influences the demography of three dominant
+sagebrush steppe plants. Ecology 92:75--85.
+
+Ellner, S. P., and M. Rees. 2006. Integral projection models for species
+with complex demography. The American naturalist 167:410--428.
+
+Evans, M. R. 2012. Modelling ecological systems in a changing world.
+Philosophical transactions of the Royal Society of London. Series B,
+Biological sciences 367:181--190.
+
+Freckleton, R. P., W. J. Sutherland, A. R. Watkinson, and S. A.
+Queenborough. 2011. Density-structured models for plant population
+dynamics. American Naturalist 177:1--17.
+
+Gelman, A., J. Hwang, and A. Vehtari. 2014. Understanding predictive
+information criteria for Bayesian models. Statistics and Computing
+24:997--1016.
+
+Gerber, B. D., W. L. Kendall, M. B. Hooten, J. A. Dubovsky, and R. C.
+Drewien. 2015. Optimal population prediction of sandhill crane
+recruitment based on climate-mediated habitat limitations. Journal of
+Animal Ecology 84:1299--1310.
+
+Hobbs, N. T., and M. B. Hooten. 2015. Bayesian Models: A Statistical
+Primer for Ecologists. Princeton University Press, Princeton.
+
+Hooten, M. B., and N. T. Hobbs. 2015. A guide to Bayesian model
+selection for ecologists. Ecological Monographs 85:3--28.
+
+Lauenroth, W. K., and P. B. Adler. 2008. Demography of perennial
+grassland plants: Survival, life expectancy and life span. Journal of
+Ecology 96:1023--1032.
+
+Luo, Y., K. Ogle, C. Tucker, S. Fei, C. Gao, S. LaDeau, J. S. Clark, and
+D. S. Schimel. 2011. Ecological forecasting and data assimilation in a
+data-rich era. Ecological Applications 21:1429--1442.
+
+Mouquet, N., Y. Lagadeuc, V. Devictor, L. Doyen, A. Duputi{é}, D.
+Eveillard, D. Faure, E. Garnier, O. Gimenez, P. Huneman, F. Jabot, P.
+Jarne, D. Joly, R. Julliard, S. K{é}fi, G. J. Kergoat, S. Lavorel, L.
+{Le Gall}, L. Meslin, S. Morand, X. Morin, H. Morlon, G. Pinay, R.
+Pradel, F. M. Schurr, W. Thuiller, and M. Loreau. 2015. Predictive
+ecology in a changing world.
+
+Perretti, C. T., S. B. Munch, and G. Sugihara. 2013. Model-free
+forecasting outperforms the correct mechanistic model for simulated and
+experimental data. Proceedings of the National Academy of Sciences of
+the United States of America 110:5253--5257.
+
+Petchey, O. L., M. Pontarp, T. M. Massie, S. K{é}fi, A. Ozgul, M.
+Weilenmann, G. M. Palamara, F. Altermatt, B. Matthews, J. M. Levine, D.
+Z. Childs, B. J. McGill, M. E. Schaepman, B. Schmid, P. Spaak, A. P.
+Beckerman, F. Pennekamp, and I. S. Pearse. 2015. The ecological forecast
+horizon, and examples of its uses and determinants. Ecology Letters
+18:597--611.
+
+Queenborough, S. A., K. M. Burnet, W. J. Sutherland, A. R. Watkinson,
+and R. P. Freckleton. 2011. From meso- to macroscale population
+dynamics: A new density-structured approach. Methods in Ecology and
+Evolution 2:289--302.
+
+R Core Team. 2013. R: A language and environment for statistical
+computing.
+
+Rees, M., and S. P. Ellner. 2009. Integral projection models for
+populations in temporally varying environments. Ecological Monographs
+79:575--594.
+
+Schindler, D. E., and R. Hilborn. 2015. Prediction, precaution, and
+policy under global change. Science 347:953--954.
+
+Stan Development Team. 2014a. Stan: A C++ Library for Probability and
+Sampling, Version 2.5.0.
+
+Stan Development Team. 2014b. Rstan: the R interface to Stan, Version
+2.5.0.
+
+S{æ}ther, B. E., S. Engen, V. Gr{ø}tan, W. Fiedler, E. Matthysen, M. E.
+Visser, J. Wright, A. P. M{ø}ller, F. Adriaensen, H. {Van Balen}, D.
+Balmer, M. C. Mainwaring, R. H. McCleery, M. Pampus, and W. Winkel.
+2007. The extended Moran effect and large-scale synchronous fluctuations
+in the size of great tit and blue tit populations. Journal of Animal
+Ecology 76:315--325.
+
+Taylor, C. M., and A. Hastings. 2004. Finding optimal control strategies
+for invasive species: a density-structured model for Spartina
+alterniflora. Journal of Applied Ecology 41:1049--1057.
+
+Teller, B. J., P. B. Adler, C. B. Edwards, G. Hooker, R. E. Snyder, and
+S. P. Ellner. 2016. Linking demography with drivers: climate and
+competition. Methods in Ecology and Evolution 7:171--183.
+
+Vehtari, A., A. Gelman, and J. Gabry. 2016. Efficient implementation of
+leave-one-out cross-validation and WAIC for evaluating fitted Bayesian
+models. ArXiv preprint.
+
+Wilcox, R. R. 2009. Comparing Pearson Correlations: Dealing with
+Heteroscedasticity and Nonnormality. Communications in Statistics -
+Simulation and Computation 38:2220--2234.
+
+Ye, H., R. J. Beamish, S. M. Glaser, S. C. H. Grant, C.-h. Hsieh, L. J.
+Richards, J. T. Schnute, and G. Sugihara. 2015. Equation-free
+mechanistic ecosystem forecasting using empirical dynamic modeling.
+Proceedings of the National Academy of Sciences 112:E1569--E1576.
+
 \end{document}

### 2016-08-26 manuscript/manuscript.Rmd
small updates for MEE R1

@@
 Geological Survey, Colorado Cooperative Fish and Wildlife Research Unit, Fort Collins, CO 80523, USA; \textsuperscript{3}Department of Fish, Wildlife, and Conservation Biology, Colorado State University, Fort Collins, CO 80523 USA}; \textsuperscript{4}Department of Statistics, Colorado State University, Fort Collins, CO 80523 USA}}
+
+\textbf{Running head:} Demographic data and population forecasting
+
+\textbf{Word count:} 6845 (including reference section)
+
+\textbf{Corresponding author:}  
+Andrew Tredennick (atredenn@gmail.com)  
+Department of Wildland Resources and the Ecology Center  
+5230 Old Main Hill  
+Utah State University  
+Logan, Utah 84322, USA
 
@@
 However, it is unknown whether models built using population-level data adequately capture the effects of density-dependence and environmental forcing that are necessary to generate skillful forecasts.
-        \item Here, we test the consequences of aggregating individual responses when forecasting the population states and trajectories of four perennial grass species in a semi-arid grassland in Montana, USA.
-We parameterized two population models for each species, one based on individual-level data (survival, growth and recruitment) and one on population-level data (percent cover), and compared their forecasting skill and forecast horizons with and without the inclusion of climate covariates.
+        \item Here, we test the consequences of aggregating individual responses when forecasting the population states \textcolor{blue}{(percent cover)} and trajectories of four perennial grass species in a semi-arid grassland in Montana, USA.
+We parameterized two population models for each species, one based on individual-level data (survival, growth and recruitment) and one on population-level data (percent cover), and compared their forecasting \textcolor{blue}{accuracy} and forecast horizons with and without the inclusion of climate covariates.
 For both models we used Bayesian ridge regression to weight the influence of climate covariates for optimal prediction.
-        \item In the absence of climate effects, we found no significant difference between the forecasting skill of models based on individual-level data and models based on population-level data.
-Climate effects were weak, but increased forecasting skill for two species.
-Increases in skill with climate covariates were similar between model types.
-        \item In our case study, percent cover models generated forecasts as skillful as those from a demographic model.
+        \item In the absence of climate effects, we found no significant difference between the forecast \textcolor{blue}{accuracy} of models based on individual-level data and models based on population-level data.
+Climate effects were weak, but increased forecast \textcolor{blue}{accuracy} for two species.
+Increases in \textcolor{blue}{accuracy} with climate covariates were similar between model types.
+        \item In our case study, percent cover models generated forecasts as \textcolor{blue}{accurate} as those from a demographic model.
 For the goal of forecasting, models based on aggregated individual-level data may offer a practical alternative to data-intensive demographic models.
@@
 Building on @Taylor2004, @Freckleton2011 showed that density-structured models compare well to continuous models in theory, and @Queenborough2011 provide empirical evidence that density-structured models are capable of reproducing population dynamics at landscape spatial scales, even if some precision is lost when compared to fully continuous models.
-However, previous tests of density-structured models have yet to assess their ability to forecast out-of-sample observations, and they have not included environmental covariates, which are necessary to forecast population responses to climate change.
+\textcolor{blue}{Density-structured models based on animal counts have also begun to find a home in dynamic species distribution models} [@Mieszkowska2013; @Hefley2016].
+However, previous tests of density-structured population models have yet to assess their ability to forecast out-of-sample observations, and they have not included environmental covariates, which are necessary to forecast population responses to climate change.
 
@@
 
-We compared the forecasting skill of statistical and population models based on aggregated, population-level data with the skill of models based on individual-level data.
+We compared the forecasting skill \textcolor{blue}{(accuracy and precision)} of statistical and population models based on aggregated, population-level data with the skill of models based on individual-level data.
 We used a demographic dataset that tracks the fates of individual plants from four species over 14 years to build two kinds of single-species population models, traditional models using individual growth, survival, and recruitment data and alternative models based on population-level (basal cover) data.
@@
 ---------------------
+### Overview of analysis
+\textcolor{blue}{Below we describe our focal plant populations, the population models we use to make predictions, and how we compare model forecasts, but offer a brief overview here.
+We used two types of data: individual-level data and percent cover data.
+Using the individual-level data we fit three vital rate regressions (survival, growth, and rectruitment) to build an Integral Projection Model (IPM) for simulating the plant populations.
+Using the percent cover data we fit a simple, Gompertz-style population growth model, which we refer to as a quadrat-based model (QBM).
+Both model types (IPM and QBM) we fit and simulate versions with and without climate covariates.
+We used Bayesian ridge regression to weight the importance of each climate covariate.
+We then performed cross-validation to assess each model's ability to predict out-of-sample data.
+We compared the forecast accuracy ($\rho$, correlation between observations and predictions) and mean absolute error (MAE) between the IPM and the QBM to test our expectation that the IPM should outperform the QBM.
+Lastly, we use in-sample forecasts to quantify each model's forecast horizon} [@Petchey2015].
+
 ### Study site and data
@@
 The community is grass-dominated, and we focused on the four most abundant grass species: \emph{Bouteloua gracilis} (BOGR), \emph{Hesperostipa comata} (HECO), \emph{Pascopyrum smithii} (PASM), and \emph{Poa secunda} (POSE) (Fig.
-1).
+1 and Table 1).
 *B.
@@
 
-In equation 3, $w_{i,t}$ is the crowding that genet _i_ in year _t_ experiences from conspecific neighbors in quadrat *q*.
+In equation 3, $w_{i,q,t}$ is the crowding that genet _i_ in year _t_ experiences from _k_ conspecific neighbors ($u_{k,q,t}$) in quadrat *q*.
 The spatial scale over which conspecific neighbors exert influence on any genet is determined by $\delta$.
@@
 
-This cross-validation procedure allowed us to compare accuracy and precision of the two modeling approaches (IPM versus QBM) with and without climate covariates.
+This cross-validation procedure allowed us to compare the accuracy and precision of the two modeling approaches (IPM versus QBM) with and without climate covariates.
 We first calculated the median predicted cover across the 100 simulations for each quadrat-year and then calculated forecast skill as the correlation ($\rho$) between forecasts and observations.
@@
 We compared $\rho$ and MAE between model types and within model types between models with and without climate covariates using one-sided *t* tests with adjusted degrees of freedom following @Wilcox2009 and standard errors calculated using the HC4 estimator of @Cribari-Neto2004.
-Statistical tests were conducted using algorithms from @Ye2015a.
+Statistical tests \textcolor{blue}{for comparing correlations and error} were conducted using algorithms from @Ye2015a.
 
@@
 
-We also expected IPM forecast accuracy to decline at a lower rate than the QBM as the forecast horizon increased.
+We also expected IPM forecast accuracy to decline at a lower rate than the QBM as the \textcolor{blue}{time between the last observation and the forecast increased.}
 In principle, more mechanistic models should produce better predictions, especially under novel conditions [@Evans2012b; @Schindler2015b].
@@
 Over time, of course, the addition and subtraction of small plants can have large effects on population growth, so explicitly modeling these effects could contribute to a longer forecast horizon. 
-However, we found no evidence for a difference between the IPM and QBM forecast horizons (Fig. 4).
+However, we found no evidence \textcolor{blue}{that the forecast horizon for the IPM was greater than the QBM (Fig. 4).
+On the contrary, the QBM tended to have a slightly longer forecast horizon than the IPM for most species (Fig. 4).
+The QBM has fewer processes, so may better capture near term dynamics when populations do not fluctuate widely, as in our case.
+When populations do not fluctuate widely, often the best forecast for population size at \emph{t+1} is the size at time \emph{t}.
+Thus, a simpler model such as the QBM may be well-suited for predicting populations that fluctuate mildly around an equilibrium.}
 
@@
 Light grey lines show trajectories of individual quadrats.
-Note the different y-axis scales across panels.", fig.height=2, fig.width=8.5, eval=T}
+Note the different y-axis scales across panels.
+See Table 1 for sample size information.", fig.height=2, fig.width=8.5, eval=T}
 library(ggthemes)
@@
   \caption{The forecast horizons for both models with climate covariates included \textcolor{blue}{and using mean parameter values.
-Points show the average accuracy ($\rho$) across all forecasts at a given distance between the last observation and the forecast, where forecasts are made for in-sample data.} We only examine the forecast accuracy of models with climate covariates included because in no case did including climate covariates significantly decrease accuracy (see Fig.
+Points show the average accuracy ($\rho$, \textcolor{blue}{correlation between observations and predictions}) across all forecasts at a given distance between the last observation and the forecast, where forecasts are made for in-sample data.} We only examine the forecast accuracy of models with climate covariates included because in no case did including climate covariates significantly decrease accuracy (see Fig.
 2).

### 2016-08-26 manuscript/manuscript.tex
small updates for MEE R1

@@
 
+\textbf{Running head:} Demographic data and population forecasting
+
+\textbf{Word count:} 6845 (including reference section)
+
+\textbf{Corresponding author:}\\Andrew Tredennick
+(\href{mailto:atredenn@gmail.com}{\nolinkurl{atredenn@gmail.com}})\\Department
+of Wildland Resources and the Ecology Center\\5230 Old Main Hill\\Utah
+State University\\Logan, Utah 84322, USA
+
 Last compile: \today
@@
 even if some precision is lost when compared to fully continuous models.
-However, previous tests of density-structured models have yet to assess
-their ability to forecast out-of-sample observations, and they have not
+\textcolor{blue}{Density-structured models based on animal counts have also begun to find a home in dynamic species distribution models}
+(Mieszkowska et al. 2013, Hefley and Hooten 2016). However, previous
+tests of density-structured population models have yet to assess their
+ability to forecast out-of-sample observations, and they have not
 included environmental covariates, which are necessary to forecast
@@
 
-We compared the forecasting skill of statistical and population models
-based on aggregated, population-level data with the skill of models
-based on individual-level data. We used a demographic dataset that
-tracks the fates of individual plants from four species over 14 years to
-build two kinds of single-species population models, traditional models
-using individual growth, survival, and recruitment data and alternative
-models based on population-level (basal cover) data. We simulated the
-models to answer two questions motivated by the fact that the effects of
-intraspecific competition (density dependence) and interannual weather
-variability act at the level of the individual (Clark et al. 2011).
-First, can population models fit using aggregated individual-level data
-(percent cover) adequately capture density dependence to produce
-forecasts as skillful as those from models fit to demographic data?
-Second, can population models fit using aggregated data adequately
-capture the influence of climate on population growth and, in turn,
-produce forecasts as skillful as those from models fit to demographic
-data?
+We compared the forecasting skill
+\textcolor{blue}{(accuracy and precision)} of statistical and population
+models based on aggregated, population-level data with the skill of
+models based on individual-level data. We used a demographic dataset
+that tracks the fates of individual plants from four species over 14
+years to build two kinds of single-species population models,
+traditional models using individual growth, survival, and recruitment
+data and alternative models based on population-level (basal cover)
+data. We simulated the models to answer two questions motivated by the
+fact that the effects of intraspecific competition (density dependence)
+and interannual weather variability act at the level of the individual
+(Clark et al. 2011). First, can population models fit using aggregated
+individual-level data (percent cover) adequately capture density
+dependence to produce forecasts as skillful as those from models fit to
+demographic data? Second, can population models fit using aggregated
+data adequately capture the influence of climate on population growth
+and, in turn, produce forecasts as skillful as those from models fit to
+demographic data?
 
 \subsection{Materials and Methods}\label{materials-and-methods}
+
+\subsubsection{Overview of analysis}\label{overview-of-analysis}
+
+\textcolor{blue}{Below we describe our focal plant populations, the population models we use to make predictions, and how we compare model forecasts, but offer a brief overview here.
+We used two types of data: individual-level data and percent cover data.
+Using the individual-level data we fit three vital rate regressions (survival, growth, and rectruitment) to build an Integral Projection Model (IPM) for simulating the plant populations.
+Using the percent cover data we fit a simple, Gompertz-style population growth model, which we refer to as a quadrat-based model (QBM).
+Both model types (IPM and QBM) we fit and simulate versions with and without climate covariates.
+We used Bayesian ridge regression to weight the importance of each climate covariate.
+We then performed cross-validation to assess each model's ability to predict out-of-sample data.
+We compared the forecast accuracy ($\rho$, correlation between observations and predictions) and mean absolute error (MAE) between the IPM and the QBM to test our expectation that the IPM should outperform the QBM.
+Lastly, we use in-sample forecasts to quantify each model's forecast horizon}
+(Petchey et al. 2015).
 
@@
 \emph{Bouteloua gracilis} (BOGR), \emph{Hesperostipa comata} (HECO),
-\emph{Pascopyrum smithii} (PASM), and \emph{Poa secunda} (POSE) (Fig.
-1). \emph{B. gracilis} is a warm-season perennial grass, whereas
-\emph{H. comata}, \emph{P. smithii}, and \emph{Poa secunda} are
+\emph{Pascopyrum smithii} (PASM), and \emph{Poa secunda} (POSE) (Fig. 1
+and Table 1). \emph{B. gracilis} is a warm-season perennial grass,
+whereas \emph{H. comata}, \emph{P. smithii}, and \emph{Poa secunda} are
 cool-season perennial grasses. The growing season begins in early spring
@@
 
-In equation 3, \(w_{i,t}\) is the crowding that genet \emph{i} in year
-\emph{t} experiences from conspecific neighbors in quadrat \emph{q}. The
-spatial scale over which conspecific neighbors exert influence on any
-genet is determined by \(\delta\). The function is applied for all
-\emph{k} conspecific genets that neighbor the focal genet at time
-\emph{t}, and \(d_{ik,q,t}\) is the distance between genet \emph{i} and
-conspecific genet \emph{k} in quadrat \emph{q}. We use
+In equation 3, \(w_{i,q,t}\) is the crowding that genet \emph{i} in year
+\emph{t} experiences from \emph{k} conspecific neighbors (\(u_{k,q,t}\))
+in quadrat \emph{q}. The spatial scale over which conspecific neighbors
+exert influence on any genet is determined by \(\delta\). The function
+is applied for all \emph{k} conspecific genets that neighbor the focal
+genet at time \emph{t}, and \(d_{ik,q,t}\) is the distance between genet
+\emph{i} and conspecific genet \emph{k} in quadrat \emph{q}. We use
 regression-specific (survival and growth) \(\delta\) values estimated by
@@
 
-This cross-validation procedure allowed us to compare accuracy and
+This cross-validation procedure allowed us to compare the accuracy and
 precision of the two modeling approaches (IPM versus QBM) with and
@@
 standard errors calculated using the HC4 estimator of Cribari-Neto
-(2004). Statistical tests were conducted using algorithms from Ye et al.
-(2015).
+(2004). Statistical tests
+\textcolor{blue}{for comparing correlations and error} were conducted
+using algorithms from Ye et al. (2015).
 
@@
 We also expected IPM forecast accuracy to decline at a lower rate than
-the QBM as the forecast horizon increased. In principle, more
-mechanistic models should produce better predictions, especially under
-novel conditions (Evans 2012, Schindler and Hilborn 2015). In our case,
-the IPM explicitly models the influence of weather on recruitment and
-survival, effects that may be poorly represented in the QBM because
-recruitment and survival mainly affect small plants that contribute
-little to year-to-year changes in percent cover. Over time, of course,
-the addition and subtraction of small plants can have large effects on
-population growth, so explicitly modeling these effects could contribute
-to a longer forecast horizon. However, we found no evidence for a
-difference between the IPM and QBM forecast horizons (Fig. 4).
+the QBM as the
+\textcolor{blue}{time between the last observation and the forecast increased.}
+In principle, more mechanistic models should produce better predictions,
+especially under novel conditions (Evans 2012, Schindler and Hilborn
+2015). In our case, the IPM explicitly models the influence of weather
+on recruitment and survival, effects that may be poorly represented in
+the QBM because recruitment and survival mainly affect small plants that
+contribute little to year-to-year changes in percent cover. Over time,
+of course, the addition and subtraction of small plants can have large
+effects on population growth, so explicitly modeling these effects could
+contribute to a longer forecast horizon. However, we found no evidence
+\textcolor{blue}{that the forecast horizon for the IPM was greater than the QBM (Fig. 4).
+On the contrary, the QBM tended to have a slightly longer forecast horizon than the IPM for most species (Fig. 4).
+The QBM has fewer processes, so may better capture near term dynamics when populations do not fluctuate widely, as in our case.
+When populations do not fluctuate widely, often the best forecast for population size at \emph{t+1} is the size at time \emph{t}.
+Thus, a simpler model such as the QBM may be well-suited for predicting populations that fluctuate mildly around an equilibrium.}
 
@@
 (POSE). Light grey lines show trajectories of individual quadrats. Note
-the different y-axis scales across panels.}
+the different y-axis scales across panels. See Table 1 for sample size
+information.}
 \end{figure}
@@
 
+Hefley, T. J., and M. B. Hooten. 2016. Hierarchical Species Distribution
+Models. Current Landscape Ecology Reports 1:87--97.
+
 Hobbs, N. T., and M. B. Hooten. 2015. Bayesian Models: A Statistical
@@
 data-rich era. Ecological Applications 21:1429--1442.
+
+Mieszkowska, N., G. Milligan, M. T. Burrows, R. Freckleton, and M.
+Spencer. 2013. Dynamic species distribution models from categorical
+survey data. Journal of Animal Ecology 82:1215--1226.
 

### 2016-08-26 FigureScripts/accuracy_error_boxplots.R

@@ -170,6 +170,7 @@ ggplot(plot_dat, aes(x=species, y=rho, fill=modelweather))+
 ggsave(paste0(path2save,"forecast_accuracy_boxplot.png"), height = 5, width=6.5, units="in", dpi=72)
 
 
+
 ####
 ####  MAKE FIGURE S1
 ####

### 2016-08-26 manuscript/manuscript.Rmd
adds MBH edits to R1

@@
 Traditional population models built with detailed demographic observations from one study site can address the impacts of environmental change at particular locations, but are difficult to scale up to the landscape and regional scales relevant to management decisions.
-An alternative is to build models using population-level data which are much easier to collect than individual-level data over broad spatial scales.
+An alternative is to build models using population-level data that are much easier to collect than individual-level data over broad spatial scales.
 However, it is unknown whether models built using population-level data adequately capture the effects of density-dependence and environmental forcing that are necessary to generate skillful forecasts.
@@
 We parameterized two population models for each species, one based on individual-level data (survival, growth and recruitment) and one on population-level data (percent cover), and compared their forecasting \textcolor{blue}{accuracy} and forecast horizons with and without the inclusion of climate covariates.
-For both models we used Bayesian ridge regression to weight the influence of climate covariates for optimal prediction.
+For both models, we used Bayesian ridge regression to weight the influence of climate covariates for optimal prediction.
         \item In the absence of climate effects, we found no significant difference between the forecast \textcolor{blue}{accuracy} of models based on individual-level data and models based on population-level data.
@@
 We used a demographic dataset that tracks the fates of individual plants from four species over 14 years to build two kinds of single-species population models, traditional models using individual growth, survival, and recruitment data and alternative models based on population-level (basal cover) data.
-We simulated the models to answer two questions motivated by the fact that the effects of intraspecific competition (density dependence) and interannual weather variability act at the level of the individual [@Clark2011b].
+We simulated from the models to answer two questions motivated by the fact that the effects of intraspecific competition (density dependence) and interannual weather variability act at the level of the individual [@Clark2011b].
 First, can population models fit using aggregated individual-level data (percent cover) adequately capture density dependence to produce forecasts as skillful as those from models fit to demographic data?
@@
 ### Overview of analysis
-\textcolor{blue}{Below we describe our focal plant populations, the population models we use to make predictions, and how we compare model forecasts, but offer a brief overview here.
-We used two types of data: individual-level data and percent cover data.
-Using the individual-level data we fit three vital rate regressions (survival, growth, and rectruitment) to build an Integral Projection Model (IPM) for simulating the plant populations.
-Using the percent cover data we fit a simple, Gompertz-style population growth model, which we refer to as a quadrat-based model (QBM).
-Both model types (IPM and QBM) we fit and simulate versions with and without climate covariates.
+\textcolor{blue}{We used two types of data: individual-level data and percent cover data.
+Using the individual-level data, we fit three vital rate regressions (survival, growth, and rectruitment) to build an Integral Projection Model (IPM) for simulating the plant populations.
+Using the percent cover data we fit a simple, Gompertz population growth model, which we refer to as a quadrat-based model (QBM).
+Both model types (IPM and QBM) were fit with and without climate covariates.
 We used Bayesian ridge regression to weight the importance of each climate covariate.
@@
 ### Study site and data
-Our demographic data come from a northern mixed grass prairie at the Fort Keogh Livestock and Range Research Laboratory near Miles City, Montana, USA ($46^{\circ}$ 19' N, $105^{\circ}$ 48' W).
+Our demographic data were obtained from a northern mixed grass prairie at the Fort Keogh Livestock and Range Research Laboratory near Miles City, Montana, USA ($46^{\circ}$ 19' N, $105^{\circ}$ 48' W).
 The dataset is available on Ecological Archives[^link] [@Anderson2011], and interested readers should refer to the metadata for a complete description.
@@
 The annual maps of the quadrats were digitized and the fates of individual plants tracked and extracted using a computer program [@Lauenroth2008; @Chu2014a].
-The permanent quadrats have not been relocated, but their distribution in six different pastures means the data represent a broad spatial distribution for the study area.
+The permanent quadrats have not been relocated, but their distribution in six different pastures implies that the data represent a broad spatial distribution for the study area.
 Daily climate data are available for the duration of the data collection period (1932 - 1945) from the Miles City airport, Wiley Field, 9 km from the study site.
@@
 The individual data are the "raw" data.
-For the quadrat-level data we simply sum individual basal cover for each quadrat by species.
+For the quadrat-level data, we summed individual basal cover for each quadrat by species.
 This is equivalent to a near-perfect census of quadrat percent cover because measurement error at the individual-level is small [@Chu2015].
@@
 Our Bayesian approach to fitting the vital rate models required choosing appropriate priors for unknown parameters and deciding which, if any, of those priors should be hierarchical.
-For each species, we fit yearly size effects and yearly intercepts hierarchically, where year-specific coefficients were drawn from global distributions representing the mean size effect and intercept.
-Quadrat random effects were also fit hierarchically, with quadrat offsets being drawn from distributions with mean zero and a shared variance term (independent Gaussian priors).
+For each species, we fit yearly size effects and yearly intercepts hierarchically, where year-specific coefficients were modeled with global distributions representing the mean size effect and intercept.
+Quadrat random effects were also fit hierarchically, with quadrat offsets modeled with distributions with mean zero and a shared variance term (independent Gaussian priors).
 Climate effects were modeled as independent covariates whose prior distributions were optimized for prediction using statistical regularization (see **Statistical regularization: Bayesian ridge regression** below).
@@
 All of our analyses (model fitting and simulating) were conducted in R [@R2013].
-We used the 'No-U-Turn' Hamiltonian Monte Carlo sampler in Stan [@stan2014]  to estimate the posterior distributions of model parameters using the package `rstan` [@rstan2014].
-We obtained posterior distributions for all model parameters from three parallel MCMC chains run for 1,000 iterations after discarding an initial 1,000 iterations.
+We used the 'No-U-Turn' Hamiltonian Monte Carlo sampler in Stan [@stan2014]  to sample from the posterior distribution of model parameters using the package `rstan` [@rstan2014].
+We obtained samples from the posterior distribution for all model parameters from three parallel MCMC chains run for 1,000 iterations after discarding an initial 1,000 iterations.
 Such short MCMC chains are possible because the Stan sampler reduces the number of iterations needed to achieve convergence.
 We assessed convergence visually and checked that scale reduction factors for all parameters were less than 1.1.
+For the purposes of including parameter uncertainty in our population models, we retained the final 1,000 iterations from each of the three MCMC chains to be used as randomly drawn values during population simulation.
+We report the posterior mean, standard deviation, and 95% Bayesian Credible Intervals for every parameter of each model for each species in the Supporting Information (Tables S5-S20).
+
+<!--
 There were six instances where the scale reduction factor for a particular parameter was greater than 1.1.
@@
 Therefore, we removed that chain from the analysis, leaving two well-mixed and converged chains for those six parameters.
-For the purposes of including parameter uncertainty in our population models, we retained the final 1,000 iterations from each of the three MCMC chains to be used as randomly drawn values during population simulation.
-We report the posterior mean, standard deviation, and 95% Bayesian Credible Intervals for every parameter of each model for each species in the Supporting Information (Tables S5-S20).
+-->
 
@@
 As the name implies, statistical regularization involves the use of a regulator that constrains an optimization.
-The natural regulator in a Bayesian application is the prior on the coefficient of interest.
+The natural regulator in a Bayesian application is the prior on the coefficients of interest.
 Each of our statistical models includes the effects of climate covariates via the term $\textbf{z}'_t \boldsymbol{\beta}_c$ with prior $\boldsymbol{\beta}_c \sim \text{Normal}(\boldsymbol{\mu}_{\beta_c}, \sigma_{\beta_c}^2\textbf{I})$.
-Because we standardized all climate covariates to have mean zero and variance one, we set $\boldsymbol{\mu}_{\beta_c} = 0$ and let $\sigma_{\beta_c}^2$ serve as the regulator that shrinks covariates toward zero -- the smaller the prior variance, the more the posteriors of $\boldsymbol{\beta}_c$ are shrunk toward zero, and the stronger the penalty [@Hooten2015].
+Because we standardized all climate covariates to have mean zero and variance one, we set $\boldsymbol{\mu}_{\beta_c} = 0$ and let $\sigma_{\beta_c}^2$ serve as the regulator that shrinks covariate effects toward zero -- the smaller the prior variance, the more the posteriors of $\boldsymbol{\beta}_c$ are shrunk toward zero, and the stronger the penalty [@Hooten2015].
 
@@
 \begin{align}
-\sum_{t=1}^T \text{log}_e \left(\frac{1}{S} \sum^S_{s=1}(y_{t,\text{hold}}|\theta_{ts})  \right).
+\sum_{t=1}^T \text{log}_e \left(\frac{1}{S} \sum^S_{s=1}[y_{t,\text{hold}}|\theta_{ts}]  \right).
 \end{align}
@@
 We used an Integral Projection Model (IPM) to simulate populations based on individual-level data [@Ellner2006] and a quadrat-based version of an individually-based model (Quadrat-Based Model, QBM) to simulate populations based on quadrat-level data.
-We describe each in turn.
+We describe each in what follows.
 
@@
 To retain temporal variation associated with random year effects, we used posterior estimates of the mean temporal effect and the standard deviation of that effect to generate a random year effect for unobserved years.
-That is, for some future year *T*, the intercept is $\beta_{0,T} \sim \text{Normal}(\beta_0, \sigma_{\beta_0})$ and the effect of size is $\beta_{s,T} \sim \text{Normal}(\beta_s, \sigma_{\beta_s})$.
+That is, for some future year *T*, the intercept is $\beta_{0,T} \sim \text{Normal}(\beta_0, \sigma_{\beta_0}^2)$ and the effect of size is $\beta_{s,T} \sim \text{Normal}(\beta_s, \sigma_{\beta_s}^2)$.
 
@@
 For a given forecast distance, we averaged the correlation between forecasts and observations.
-Note that these forecasts are all made using in-sample data because we used model fits from the full data set.
-Nonetheless, these simulations offer insight into the differences among model forecast horizons.
+Note that our forecasts for the horizon analysis are all made using in-sample data because we used model fits from the full data set.
+Nonetheless, our simulations offer insight into the differences among model forecast horizons.
 \textcolor{blue}{We chose a forecast accuracy of $\rho = 0.5$ as our forecast proficiency threshold.
@@
 S2, Supporting Information).
-Thus, the posterior distributions of climate effects included in our models overlapped zero and generally shrunk toward zero, though for some species-vital rate combinations, important effects (80% credible interval does not include zero) did emerge (Fig.
+Thus, the posterior distributions of climate effects included in our models overlapped zero and generally were shrunk toward zero, though for some species-vital rate combinations, important effects (80% credible interval does not include zero) did emerge (Fig.
 3).
@@
 
-With climate covariates included \textcolor{blue{and using mean parameter values, the accuracy of both models' forecasts declined as the distance between the last observation and the forecast increased}, but they did so at similar rates (Fig.
+With climate covariates included \textcolor{blue}{and using mean parameter values, the accuracy of both models' forecasts declined as the distance between the last observation and the forecast increased}, but they did so at similar rates (Fig.
 4).
@@
 2; `no-climate' bars).
-This result overturned our expectation that the IPM would out-perform the QBM, thanks to its mechanistic representation of the perennial life cycle, and confirms theoretical [@Freckleton2011] and empirical work [@Taylor2004; @Queenborough2011] showing that density-structured models can be useful surrogates for demographic models when the goal is to estimate or forecast population states over large spatial extents.
+This result differed from our expectation that the IPM would out-perform the QBM, because of its mechanistic representation of the perennial life cycle.
+Our result also confirms theoretical [@Freckleton2011] and empirical work [@Taylor2004; @Queenborough2011] showing that density-structured models can be useful surrogates for demographic models when the goal is to estimate or forecast population states over large spatial extents.
 
 We also expected the inclusion of environmental forcing to reveal more differences between the models.
-Interannual variation in weather can affect vital rates in different ways [@Dalgleish2011], so, estimates of climate effects on plant population growth may be biased or non-identifiable when the underlying statistical model is fit using population-level data that integrates over the potentially unique climate responses of individual vital rates.
+Interannual variation in weather can affect vital rates in different ways [@Dalgleish2011].
+Thus, estimates of climate effects on plant population growth may be biased or non-identifiable when the underlying statistical model is fit using population-level data that integrates over the potentially unique climate responses of individual vital rates.
 We found some evidence that the QBM failed to detect climate effects for three species (*B.
@@
 Ridge regression allows researchers to retain covariates whose effects may be difficult to identify in noisy data or short time series.
-This is especially important when forecasting the impacts of climate variability, where it is important to include to effects of forcing variables (e.g., temperature and precipitation) even if such effects are difficult to identify.
+This is especially important when forecasting the impacts of climate variability, where it is important to include the effects of forcing variables (e.g., temperature and precipitation) even if such effects are difficult to identify.
 Indeed, we failed to detect strong climate effects in the QBM for *B. gracilis* and *Poa secunda*, but including climate covariates still improved forecasting skill (Fig. 2).
 If a species is truly unresponsive to a given climate variable, statistical regularization techniques will shrink the mean and variance of a covariate estimate toward zero [@Hooten2015].
-Of course, no matter what model selection approach is adopted, a critical step is identifying the appropriate candidate covariates, which we attempted to do based on our knowledge of this semi-arid plant community.
+Of course, regardless of what model selection approach is adopted, a critical step is identifying the appropriate candidate covariates, which we attempted to do based on our knowledge of this semi-arid plant community.
 However, the climate covariates we chose required aggregating daily weather data over discrete time periods.
 It is possible that we did not choose the optimal time periods over which to aggregate.
-New methods using functional linear models (or splines) may offer a data-driven approach for identifying the appropriate time periods over which to aggregate to produce a tractable set of candidate climate variables [@Teller2016].
+New methods using functional linear models (or splines) may offer a data-driven approach for identifying the appropriate time periods over which to aggregate to produce a tractable set of candidate climate variables [@Sims2007; @VandePol2011; @Teller2016].
 
@@
 In our case, the IPM explicitly models the influence of weather on recruitment and survival, effects that may be poorly represented in the QBM because recruitment and survival mainly affect small plants that contribute little to year-to-year changes in percent cover.
-Over time, of course, the addition and subtraction of small plants can have large effects on population growth, so explicitly modeling these effects could contribute to a longer forecast horizon. 
+Over time, the addition and subtraction of small plants can have large effects on population growth, thus, explicitly modeling these effects could contribute to a longer forecast horizon. 
 However, we found no evidence \textcolor{blue}{that the forecast horizon for the IPM was greater than the QBM (Fig. 4).
 On the contrary, the QBM tended to have a slightly longer forecast horizon than the IPM for most species (Fig. 4).
-The QBM has fewer processes, so may better capture near term dynamics when populations do not fluctuate widely, as in our case.
+The QBM has fewer processes, thus it may better capture near term dynamics when populations do not fluctuate widely, as in our case.
 When populations do not fluctuate widely, often the best forecast for population size at \emph{t+1} is the size at time \emph{t}.
@@
 \textcolor{blue}{Third, our percent cover data are essentially error-free because we were able to aggregate indiviual plant areas to calculate percent cover.
-Percent cover data collected by typical sampling protocals (e.g., Daubenmire frames) will include error that would affect population forecasts due to misspecifing the initial conditions and/or biasing model parameters} [@Queenborough2011].
+Percent cover data collected by typical sampling protocals (e.g., Daubenmire frames) will include error that may affect population forecasts due to misspecifing the initial conditions and/or biasing model parameters} [@Queenborough2011].
 \textcolor{blue}{Thus, it may be that our percent cover model performed just as well as the individual-based model because percent cover was perfectly measured.
@@
 \textcolor{blue}{We may also have to shift our perspective from making explicit point forecasts to making moving average forecasts} [@Petchey2015].
-\textcolor{blue}{For example, Integral Projection Models are typically capable of estimating equilibrium abundances} [e.g., @Adler2012], \textcolor{blue}{which may be more appropriate for dynamic systems.}
+\textcolor{blue}{For example, IPMs are typically capable of estimating equilibrium abundances} [e.g., @Adler2012], \textcolor{blue}{which may be more appropriate for dynamic systems.}
 
@@
 Thus, we conclude that models based on population-level data, rather than individual-level data, may be adequate for forecasting the states and dynamics of plant populations.
-Unfortunately, our analysis, where climate effects were relatively unimportant in vital rate regressions, did not allow us to sufficiently test our prediction that individual-level data is neccessary to generate skillful forecasts if different vital rates respond to climate in unique, potentially opposing, ways.
+Unfortunately, our analysis, where climate effects were relatively unimportant in vital rate regressions, did not allow us to sufficiently test our prediction that individual-level data is necessary to generate skillful forecasts if different vital rates respond to climate in unique, potentially opposing, ways.
 Nonetheless, our results are encouraging for the use of easy-to-collect population-level data for forecasting the state of plant populations.

### 2016-08-26 manuscript/manuscript.tex
adds MBH edits to R1

@@
 data and alternative models based on population-level (basal cover)
-data. We simulated the models to answer two questions motivated by the
-fact that the effects of intraspecific competition (density dependence)
-and interannual weather variability act at the level of the individual
-(Clark et al. 2011). First, can population models fit using aggregated
-individual-level data (percent cover) adequately capture density
-dependence to produce forecasts as skillful as those from models fit to
-demographic data? Second, can population models fit using aggregated
-data adequately capture the influence of climate on population growth
-and, in turn, produce forecasts as skillful as those from models fit to
-demographic data?
+data. We simulated from the models to answer two questions motivated by
+the fact that the effects of intraspecific competition (density
+dependence) and interannual weather variability act at the level of the
+individual (Clark et al. 2011). First, can population models fit using
+aggregated individual-level data (percent cover) adequately capture
+density dependence to produce forecasts as skillful as those from models
+fit to demographic data? Second, can population models fit using
+aggregated data adequately capture the influence of climate on
+population growth and, in turn, produce forecasts as skillful as those
+from models fit to demographic data?
 
@@
 
-\textcolor{blue}{Below we describe our focal plant populations, the population models we use to make predictions, and how we compare model forecasts, but offer a brief overview here.
-We used two types of data: individual-level data and percent cover data.
-Using the individual-level data we fit three vital rate regressions (survival, growth, and rectruitment) to build an Integral Projection Model (IPM) for simulating the plant populations.
-Using the percent cover data we fit a simple, Gompertz-style population growth model, which we refer to as a quadrat-based model (QBM).
-Both model types (IPM and QBM) we fit and simulate versions with and without climate covariates.
+\textcolor{blue}{We used two types of data: individual-level data and percent cover data.
+Using the individual-level data, we fit three vital rate regressions (survival, growth, and rectruitment) to build an Integral Projection Model (IPM) for simulating the plant populations.
+Using the percent cover data we fit a simple, Gompertz population growth model, which we refer to as a quadrat-based model (QBM).
+Both model types (IPM and QBM) were fit with and without climate covariates.
 We used Bayesian ridge regression to weight the importance of each climate covariate.
@@
 
-Our demographic data come from a northern mixed grass prairie at the
-Fort Keogh Livestock and Range Research Laboratory near Miles City,
-Montana, USA (\(46^{\circ}\) 19' N, \(105^{\circ}\) 48' W). The dataset
-is available on Ecological Archives\footnote{\url{http://esapubs.org/archive/ecol/E092/143/}}
+Our demographic data were obtained from a northern mixed grass prairie
+at the Fort Keogh Livestock and Range Research Laboratory near Miles
+City, Montana, USA (\(46^{\circ}\) 19' N, \(105^{\circ}\) 48' W). The
+dataset is available on Ecological Archives\footnote{\url{http://esapubs.org/archive/ecol/E092/143/}}
 (Anderson et al. 2011), and interested readers should refer to the
@@
 not been relocated, but their distribution in six different pastures
-means the data represent a broad spatial distribution for the study
-area. Daily climate data are available for the duration of the data
-collection period (1932 - 1945) from the Miles City airport, Wiley
+implies that the data represent a broad spatial distribution for the
+study area. Daily climate data are available for the duration of the
+data collection period (1932 - 1945) from the Miles City airport, Wiley
 Field, 9 km from the study site.
@@
 and quadrat. The individual data are the ``raw'' data. For the
-quadrat-level data we simply sum individual basal cover for each quadrat
-by species. This is equivalent to a near-perfect census of quadrat
-percent cover because measurement error at the individual-level is small
-(Chu and Adler 2015). Based on these two datasets of 13 year-to-year
+quadrat-level data, we summed individual basal cover for each quadrat by
+species. This is equivalent to a near-perfect census of quadrat percent
+cover because measurement error at the individual-level is small (Chu
+and Adler 2015). Based on these two datasets of 13 year-to-year
 transitions, we can compare population models built using
@@
 size effects and yearly intercepts hierarchically, where year-specific
-coefficients were drawn from global distributions representing the mean
-size effect and intercept. Quadrat random effects were also fit
-hierarchically, with quadrat offsets being drawn from distributions with
+coefficients were modeled with global distributions representing the
+mean size effect and intercept. Quadrat random effects were also fit
+hierarchically, with quadrat offsets modeled with distributions with
 mean zero and a shared variance term (independent Gaussian priors).
@@
 (R Core Team 2013). We used the `No-U-Turn' Hamiltonian Monte Carlo
-sampler in Stan (Stan Development Team 2014a) to estimate the posterior
-distributions of model parameters using the package \texttt{rstan} (Stan
-Development Team 2014b). We obtained posterior distributions for all
-model parameters from three parallel MCMC chains run for 1,000
-iterations after discarding an initial 1,000 iterations. Such short MCMC
-chains are possible because the Stan sampler reduces the number of
-iterations needed to achieve convergence. We assessed convergence
-visually and checked that scale reduction factors for all parameters
-were less than 1.1. There were six instances where the scale reduction
-factor for a particular parameter was greater than 1.1. In those cases,
-we checked the traceplots of the violating parameter and discovered that
-one chain, out of three, was poorly behaved. Therefore, we removed that
-chain from the analysis, leaving two well-mixed and converged chains for
-those six parameters. For the purposes of including parameter
-uncertainty in our population models, we retained the final 1,000
-iterations from each of the three MCMC chains to be used as randomly
-drawn values during population simulation. We report the posterior mean,
-standard deviation, and 95\% Bayesian Credible Intervals for every
-parameter of each model for each species in the Supporting Information
-(Tables S5-S20).
+sampler in Stan (Stan Development Team 2014a) to sample from the
+posterior distribution of model parameters using the package
+\texttt{rstan} (Stan Development Team 2014b). We obtained samples from
+the posterior distribution for all model parameters from three parallel
+MCMC chains run for 1,000 iterations after discarding an initial 1,000
+iterations. Such short MCMC chains are possible because the Stan sampler
+reduces the number of iterations needed to achieve convergence. We
+assessed convergence visually and checked that scale reduction factors
+for all parameters were less than 1.1. For the purposes of including
+parameter uncertainty in our population models, we retained the final
+1,000 iterations from each of the three MCMC chains to be used as
+randomly drawn values during population simulation. We report the
+posterior mean, standard deviation, and 95\% Bayesian Credible Intervals
+for every parameter of each model for each species in the Supporting
+Information (Tables S5-S20).
 
@@
 natural regulator in a Bayesian application is the prior on the
-coefficient of interest. Each of our statistical models includes the
+coefficients of interest. Each of our statistical models includes the
 effects of climate covariates via the term
@@
 variance one, we set \(\boldsymbol{\mu}_{\beta_c} = 0\) and let
-\(\sigma_{\beta_c}^2\) serve as the regulator that shrinks covariates
-toward zero -- the smaller the prior variance, the more the posteriors
-of \(\boldsymbol{\beta}_c\) are shrunk toward zero, and the stronger the
-penalty (Hooten and Hobbs 2015).
+\(\sigma_{\beta_c}^2\) serve as the regulator that shrinks covariate
+effects toward zero -- the smaller the prior variance, the more the
+posteriors of \(\boldsymbol{\beta}_c\) are shrunk toward zero, and the
+stronger the penalty (Hooten and Hobbs 2015).
 
@@
 \vspace{-3em}\begin{align}
-\sum_{t=1}^T \text{log}_e \left(\frac{1}{S} \sum^S_{s=1}(y_{t,\text{hold}}|\theta_{ts})  \right).
+\sum_{t=1}^T \text{log}_e \left(\frac{1}{S} \sum^S_{s=1}[y_{t,\text{hold}}|\theta_{ts}]  \right).
 \end{align}\vspace{-3em}
@@
 Model, QBM) to simulate populations based on quadrat-level data. We
-describe each in turn.
+describe each in what follows.
 
@@
 intercept is
-\(\beta_{0,T} \sim \text{Normal}(\beta_0, \sigma_{\beta_0})\) and the
+\(\beta_{0,T} \sim \text{Normal}(\beta_0, \sigma_{\beta_0}^2)\) and the
 effect of size is
-\(\beta_{s,T} \sim \text{Normal}(\beta_s, \sigma_{\beta_s})\).
+\(\beta_{s,T} \sim \text{Normal}(\beta_s, \sigma_{\beta_s}^2)\).
 
@@
 forecast distance, we averaged the correlation between forecasts and
-observations. Note that these forecasts are all made using in-sample
-data because we used model fits from the full data set. Nonetheless,
-these simulations offer insight into the differences among model
-forecast horizons.
+observations. Note that our forecasts for the horizon analysis are all
+made using in-sample data because we used model fits from the full data
+set. Nonetheless, our simulations offer insight into the differences
+among model forecast horizons.
 \textcolor{blue}{We chose a forecast accuracy of $\rho = 0.5$ as our forecast proficiency threshold.
@@
 Information). Thus, the posterior distributions of climate effects
-included in our models overlapped zero and generally shrunk toward zero,
-though for some species-vital rate combinations, important effects (80\%
-credible interval does not include zero) did emerge (Fig. 3).
+included in our models overlapped zero and generally were shrunk toward
+zero, though for some species-vital rate combinations, important effects
+(80\% credible interval does not include zero) did emerge (Fig. 3).
 
@@
 
-With climate covariates included \textbackslash{}textcolor\{blue\{and
-using mean parameter values, the accuracy of both models' forecasts
-declined as the distance between the last observation and the forecast
-increased\}, but they did so at similar rates (Fig. 4). The only
-exception is for \emph{Poa secunda}, where QBM forecast accuracy
-remained steady as the
+With climate covariates included
+\textcolor{blue}{and using mean parameter values, the accuracy of both models' forecasts declined as the distance between the last observation and the forecast increased},
+but they did so at similar rates (Fig. 4). The only exception is for
+\emph{Poa secunda}, where QBM forecast accuracy remained steady as the
 \textcolor{blue}{temporal distance of the forecast increased}, whereas
@@
 than forecasts from the QBM (Fig. 2; `no-climate' bars). This result
-overturned our expectation that the IPM would out-perform the QBM,
-thanks to its mechanistic representation of the perennial life cycle,
-and confirms theoretical (Freckleton et al. 2011) and empirical work
-(Taylor and Hastings 2004, Queenborough et al. 2011) showing that
-density-structured models can be useful surrogates for demographic
-models when the goal is to estimate or forecast population states over
-large spatial extents.
+differed from our expectation that the IPM would out-perform the QBM,
+because of its mechanistic representation of the perennial life cycle.
+Our result also confirms theoretical (Freckleton et al. 2011) and
+empirical work (Taylor and Hastings 2004, Queenborough et al. 2011)
+showing that density-structured models can be useful surrogates for
+demographic models when the goal is to estimate or forecast population
+states over large spatial extents.
 
@@
 differences between the models. Interannual variation in weather can
-affect vital rates in different ways (Dalgleish et al. 2011), so,
+affect vital rates in different ways (Dalgleish et al. 2011). Thus,
 estimates of climate effects on plant population growth may be biased or
@@
 important when forecasting the impacts of climate variability, where it
-is important to include to effects of forcing variables (e.g.,
+is important to include the effects of forcing variables (e.g.,
 temperature and precipitation) even if such effects are difficult to
@@
 regularization techniques will shrink the mean and variance of a
-covariate estimate toward zero (Hooten and Hobbs 2015). Of course, no
-matter what model selection approach is adopted, a critical step is
-identifying the appropriate candidate covariates, which we attempted to
-do based on our knowledge of this semi-arid plant community. However,
+covariate estimate toward zero (Hooten and Hobbs 2015). Of course,
+regardless of what model selection approach is adopted, a critical step
+is identifying the appropriate candidate covariates, which we attempted
+to do based on our knowledge of this semi-arid plant community. However,
 the climate covariates we chose required aggregating daily weather data
@@
 for identifying the appropriate time periods over which to aggregate to
-produce a tractable set of candidate climate variables (Teller et al.
-2016).
+produce a tractable set of candidate climate variables (Sims et al.
+2007, Pol and Cockburn 2011, Teller et al. 2016).
 
@@
 contribute little to year-to-year changes in percent cover. Over time,
-of course, the addition and subtraction of small plants can have large
-effects on population growth, so explicitly modeling these effects could
+the addition and subtraction of small plants can have large effects on
+population growth, thus, explicitly modeling these effects could
 contribute to a longer forecast horizon. However, we found no evidence
@@
 On the contrary, the QBM tended to have a slightly longer forecast horizon than the IPM for most species (Fig. 4).
-The QBM has fewer processes, so may better capture near term dynamics when populations do not fluctuate widely, as in our case.
+The QBM has fewer processes, thus it may better capture near term dynamics when populations do not fluctuate widely, as in our case.
 When populations do not fluctuate widely, often the best forecast for population size at \emph{t+1} is the size at time \emph{t}.
@@
 \textcolor{blue}{Third, our percent cover data are essentially error-free because we were able to aggregate indiviual plant areas to calculate percent cover.
-Percent cover data collected by typical sampling protocals (e.g., Daubenmire frames) will include error that would affect population forecasts due to misspecifing the initial conditions and/or biasing model parameters}
+Percent cover data collected by typical sampling protocals (e.g., Daubenmire frames) will include error that may affect population forecasts due to misspecifing the initial conditions and/or biasing model parameters}
 (Queenborough et al. 2011).
@@
 (Petchey et al. 2015).
-\textcolor{blue}{For example, Integral Projection Models are typically capable of estimating equilibrium abundances}
+\textcolor{blue}{For example, IPMs are typically capable of estimating equilibrium abundances}
 (e.g., Adler et al. 2012),
@@
 unimportant in vital rate regressions, did not allow us to sufficiently
-test our prediction that individual-level data is neccessary to generate
+test our prediction that individual-level data is necessary to generate
 skillful forecasts if different vital rates respond to climate in
@@
 
+Pol, M. van de, and A. Cockburn. 2011. Identifying the critical climatic
+time window that affects trait expression. The American naturalist
+177:698--707.
+
 Queenborough, S. A., K. M. Burnet, W. J. Sutherland, A. R. Watkinson,
@@
 policy under global change. Science 347:953--954.
+
+Sims, M., D. A. Elston, A. Larkham, D. H. Nussey, and S. D. Albon. 2007.
+Identifying when weather influences life-history traits of grazing
+herbivores. Journal of Animal Ecology 76:761--770.
 

### 2016-08-29 manuscript/manuscript.Rmd
adds PBA R1 edits

@@
 Building on @Taylor2004, @Freckleton2011 showed that density-structured models compare well to continuous models in theory, and @Queenborough2011 provide empirical evidence that density-structured models are capable of reproducing population dynamics at landscape spatial scales, even if some precision is lost when compared to fully continuous models.
+However, previous tests of density-structured population models have yet to assess their ability to forecast out-of-sample observations, and they have not included environmental covariates, which are necessary to forecast population responses to climate change.
+
+<!--
 \textcolor{blue}{Density-structured models based on animal counts have also begun to find a home in dynamic species distribution models} [@Mieszkowska2013; @Hefley2016].
-However, previous tests of density-structured population models have yet to assess their ability to forecast out-of-sample observations, and they have not included environmental covariates, which are necessary to forecast population responses to climate change.
+-->
 
@@
 Using the percent cover data we fit a simple, Gompertz population growth model, which we refer to as a quadrat-based model (QBM).
-Both model types (IPM and QBM) were fit with and without climate covariates.
+For both model types (IPM and QBM), we fit and simulate versions of the model with and without climate covariates.
 We used Bayesian ridge regression to weight the importance of each climate covariate.
@@
 We focus on the main process and the model likelihood in what follows (full model descriptions are in the Supporting Information).
-For the likelihood models, $\textbf{y}^X$ is always the relevant vector of observations for vital rate *X* (*X* = *S*, *G*, *R*, or *P*) for survival, growth, recruitment, or population growth).
+For the likelihood models, $\textbf{y}^X$ is always the relevant vector of observations for vital rate *X* (*X* = *S*, *G*, *R*, or *P* for survival, growth, recruitment, or population growth).
 For example, $\textbf{y}^S$ is a vector of 0s and 1s indicating whether a genet survives from *t* to *t+1*, or not, for all observation years and quadrats.
@@
 Nonetheless, our simulations offer insight into the differences among model forecast horizons.
-\textcolor{blue}{We chose a forecast accuracy of $\rho = 0.5$ as our forecast proficiency threshold.
+\textcolor{blue}{We chose an arbitrary forecast accuracy of $\rho = 0.5$ as our forecast proficiency threshold.
+For basic research on forecasting, arbitrary proficiency thresholds suffice for comparative purposes} [@Petchey2015], \textcolor{blue}{and $\rho = 0.5$ represents the point at which about 25% of the variance in observations is explained by the predictions.
 So, the forecast horizon is the temporal distance at which forecast accuracy falls below $\rho = 0.5$.}
@@
 \textcolor{blue}{The forecast horizons were short: forecast accuracy fell below $\rho$ = 0.5 after one year for the IPM for most species, and after four years, at most, for the QBM (Fig. 4).
-Across the different temporal distances from the observation to the forecast, the IPM was never more accurate than the QBM (Table S23).
+Across the different temporal distances from the observation to the forecast, the IPM was never more accurate than the QBM (\emph{P} > 0.05 for all one-sided \emph{t}-tests, Table S23).
 Likewise, the QBM was rarely more accurate the IPM, the only exception being for \emph{H.
@@
 
-We also expected IPM forecast accuracy to decline at a lower rate than the QBM as the \textcolor{blue}{time between the last observation and the forecast increased.}
+We also expected IPM forecast accuracy to decline at a lower rate than the QBM as the \textcolor{blue}{time between the model initialization and the forecast increased.}
 In principle, more mechanistic models should produce better predictions, especially under novel conditions [@Evans2012b; @Schindler2015b].
@@
 On the contrary, the QBM tended to have a slightly longer forecast horizon than the IPM for most species (Fig. 4).
-The QBM has fewer processes, thus it may better capture near term dynamics when populations do not fluctuate widely, as in our case.
-When populations do not fluctuate widely, often the best forecast for population size at \emph{t+1} is the size at time \emph{t}.
-Thus, a simpler model such as the QBM may be well-suited for predicting populations that fluctuate mildly around an equilibrium.}
+The QBM has fewer processes and parameters, which can reduce bias due to parameter uncertainty.
+Thus, the QBM may better capture near term dynamics when populations do not fluctuate widely, as in our case.}
 
@@
 \textcolor{blue}{Third, our percent cover data are essentially error-free because we were able to aggregate indiviual plant areas to calculate percent cover.
-Percent cover data collected by typical sampling protocals (e.g., Daubenmire frames) will include error that may affect population forecasts due to misspecifing the initial conditions and/or biasing model parameters} [@Queenborough2011].
-\textcolor{blue}{Thus, it may be that our percent cover model performed just as well as the individual-based model because percent cover was perfectly measured.
+Percent cover data collected by typical sampling methods (e.g., Daubenmire frames) will include error that may affect population forecasts due to misspecifing the initial conditions and/or biasing model parameters} [@Queenborough2011].
+\textcolor{blue}{Percent cover models based on data containing more measurement error than ours might perform worse in comparison with individual-based models.
 One way to account for such error is to develop a sampling model that relates the observations (estimated percent cover in a plot) to the true state (percent cover derived from individual plant measurements in the same plot)} [@Hobbs2015].
@@
 Such short forecast horizons are not encouraging.}
-\textcolor{blue}{Unfortunately, we have little to offer on how to improve population forecasts that has not already been proposed} [@Petchey2015; @Mouquet2015].
+\textcolor{blue}{Unfortunately, we have few ideas about how to improve population forecasts that have not already been proposed} [@Petchey2015; @Mouquet2015].
 \textcolor{blue}{Longer time-series should improve our ability to detect exogenous drivers such as climate} [@Teller2016], \textcolor{blue}{and modeling larger spatial extents may reduce parameter uncertainty} [@Petchey2015].
 \textcolor{blue}{We may also have to shift our perspective from making explicit point forecasts to making moving average forecasts} [@Petchey2015].
-\textcolor{blue}{For example, IPMs are typically capable of estimating equilibrium abundances} [e.g., @Adler2012], \textcolor{blue}{which may be more appropriate for dynamic systems.}
+\textcolor{blue}{Whether the poor predictive ability of our models impacts their comparison relative to one another is an open question}.
 

### 2016-08-29 manuscript/manuscript.tex
adds PBA R1 edits

@@
 even if some precision is lost when compared to fully continuous models.
-\textcolor{blue}{Density-structured models based on animal counts have also begun to find a home in dynamic species distribution models}
-(Mieszkowska et al. 2013, Hefley and Hooten 2016). However, previous
-tests of density-structured population models have yet to assess their
-ability to forecast out-of-sample observations, and they have not
-included environmental covariates, which are necessary to forecast
-population responses to climate change.
+However, previous tests of density-structured population models have yet
+to assess their ability to forecast out-of-sample observations, and they
+have not included environmental covariates, which are necessary to
+forecast population responses to climate change.
 
@@
 for vital rate \emph{X} (\emph{X} = \emph{S}, \emph{G}, \emph{R}, or
-\emph{P}) for survival, growth, recruitment, or population growth). For
+\emph{P} for survival, growth, recruitment, or population growth). For
 example, \(\textbf{y}^S\) is a vector of 0s and 1s indicating whether a
@@
 among model forecast horizons.
-\textcolor{blue}{We chose a forecast accuracy of $\rho = 0.5$ as our forecast proficiency threshold.
+\textcolor{blue}{We chose an arbitrary forecast accuracy of $\rho = 0.5$ as our forecast proficiency threshold.
+For basic research on forecasting, arbitrary proficiency thresholds suffice for comparative purposes}
+(Petchey et al. 2015),
+\textcolor{blue}{and $\rho = 0.5$ represents the point at which about 25% of the variance in observations is explained by the predictions.
 So, the forecast horizon is the temporal distance at which forecast accuracy falls below $\rho = 0.5$.}
@@
 the QBM as the
-\textcolor{blue}{time between the last observation and the forecast increased.}
+\textcolor{blue}{time between the model initialization and the forecast increased.}
 In principle, more mechanistic models should produce better predictions,
@@
 \textcolor{blue}{Third, our percent cover data are essentially error-free because we were able to aggregate indiviual plant areas to calculate percent cover.
-Percent cover data collected by typical sampling protocals (e.g., Daubenmire frames) will include error that may affect population forecasts due to misspecifing the initial conditions and/or biasing model parameters}
+Percent cover data collected by typical sampling methods (e.g., Daubenmire frames) will include error that may affect population forecasts due to misspecifing the initial conditions and/or biasing model parameters}
 (Queenborough et al. 2011).
-\textcolor{blue}{Thus, it may be that our percent cover model performed just as well as the individual-based model because percent cover was perfectly measured.
+\textcolor{blue}{Percent cover models based on data containing more measurement error than ours might perform worse in comparison with individual-based models.
 One way to account for such error is to develop a sampling model that relates the observations (estimated percent cover in a plot) to the true state (percent cover derived from individual plant measurements in the same plot)}
@@
 Such short forecast horizons are not encouraging.}
-\textcolor{blue}{Unfortunately, we have little to offer on how to improve population forecasts that has not already been proposed}
+\textcolor{blue}{Unfortunately, we have few ideas about how to improve population forecasts that have not already been proposed}
 (Mouquet et al. 2015, Petchey et al. 2015).
@@
 (Petchey et al. 2015).
-\textcolor{blue}{For example, IPMs are typically capable of estimating equilibrium abundances}
-(e.g., Adler et al. 2012),
-\textcolor{blue}{which may be more appropriate for dynamic systems.}
+\textcolor{blue}{Whether the poor predictive ability of our models impacts their comparison relative to one another is an open question}.
 
@@
 
-Hefley, T. J., and M. B. Hooten. 2016. Hierarchical Species Distribution
-Models. Current Landscape Ecology Reports 1:87--97.
-
 Hobbs, N. T., and M. B. Hooten. 2015. Bayesian Models: A Statistical
@@
 data-rich era. Ecological Applications 21:1429--1442.
-
-Mieszkowska, N., G. Milligan, M. T. Burrows, R. Freckleton, and M.
-Spencer. 2013. Dynamic species distribution models from categorical
-survey data. Journal of Animal Ecology 82:1215--1226.
 

### 2016-08-30 manuscript/manuscript.Rmd
adds PBA line edits for R1; ready for USGS

@@
 Traditional population models built with detailed demographic observations from one study site can address the impacts of environmental change at particular locations, but are difficult to scale up to the landscape and regional scales relevant to management decisions.
-An alternative is to build models using population-level data that are much easier to collect than individual-level data over broad spatial scales.
+An alternative is to build models using population-level data that are much easier to collect over broad spatial scales than individual-level data.
 However, it is unknown whether models built using population-level data adequately capture the effects of density-dependence and environmental forcing that are necessary to generate skillful forecasts.
@@
 Such models do not require individual-level demographic data and can adequately describe population dynamics.
-Building on @Taylor2004, @Freckleton2011 showed that density-structured models compare well to continuous models in theory, and @Queenborough2011 provide empirical evidence that density-structured models are capable of reproducing population dynamics at landscape spatial scales, even if some precision is lost when compared to fully continuous models.
+Building on @Taylor2004, @Freckleton2011 showed that density-structured models compare well to continuous models in theory, and @Queenborough2011 provide empirical evidence that density-structured models are capable of reproducing population dynamics at landscape spatial scales [also see @Mieszkowska2013], even if some precision is lost when compared to fully continuous models.
 However, previous tests of density-structured population models have yet to assess their ability to forecast out-of-sample observations, and they have not included environmental covariates, which are necessary to forecast population responses to climate change.
-
-<!--
-\textcolor{blue}{Density-structured models based on animal counts have also begun to find a home in dynamic species distribution models} [@Mieszkowska2013; @Hefley2016].
--->
 
@@
 Futhermore, models based on aggregated data may reflect short-term effects of one vital rate more than others whose importance may only emerge over the long-term.
-For example, a one-year change in a plant species' cover or biomass might reflect individual growth or shrinkage, whereas the long-term trajectory of the population might be more influenced by recruitment.
+For example, a one-year change in a plant species' cover or biomass might reflect growth or shrinkage of the largest individuals, whereas the long-term trajectory of the population might be more influenced by recruitment.
 The same is true for density dependence: intraspecific density depedence may act most strongly on vital rates, like recruitment, that are difficult to identify from population-level data.
@@
 We fit all models using a hierarchical Bayesian approach.
-We focus on the main process and the model likelihood in what follows (full model descriptions are in the Supporting Information).
+In the following description, we focus on the main process and the model likelihood (full model descriptions are in the Supporting Information).
 For the likelihood models, $\textbf{y}^X$ is always the relevant vector of observations for vital rate *X* (*X* = *S*, *G*, *R*, or *P* for survival, growth, recruitment, or population growth).
@@
 For each species, we fit yearly size effects and yearly intercepts hierarchically, where year-specific coefficients were modeled with global distributions representing the mean size effect and intercept.
-Quadrat random effects were also fit hierarchically, with quadrat offsets modeled with distributions with mean zero and a shared variance term (independent Gaussian priors).
+Quadrat random effects were also fit hierarchically, with quadrat offsets modeled using distributions with mean zero and a shared variance term (independent Gaussian priors).
 Climate effects were modeled as independent covariates whose prior distributions were optimized for prediction using statistical regularization (see **Statistical regularization: Bayesian ridge regression** below).
@@
 Thus, in our current example, the model forecast for percent cover in 1941 has a forecast horizon of one year, the forecast in 1942 has a forecast horizon of two years, and so on.
-We performed these simulations for all model types (IPM with/without climate; QBM with/without climate) using mean parameter values for all possible initial years.
+We performed these simulations using mean parameter values for all model types (IPM with/without climate; QBM with/without climate) and all possible initial years.
 For a given forecast distance, we averaged the correlation between forecasts and observations.
@@
 \textcolor{blue}{We chose an arbitrary forecast accuracy of $\rho = 0.5$ as our forecast proficiency threshold.
-For basic research on forecasting, arbitrary proficiency thresholds suffice for comparative purposes} [@Petchey2015], \textcolor{blue}{and $\rho = 0.5$ represents the point at which about 25% of the variance in observations is explained by the predictions.
-So, the forecast horizon is the temporal distance at which forecast accuracy falls below $\rho = 0.5$.}
+the forecast horizon is the temporal distance at which forecast accuracy falls below $\rho = 0.5$.
+For basic research on forecasting, arbitrary proficiency thresholds suffice for comparative purposes} [@Petchey2015], \textcolor{blue}{and $\rho = 0.5$ represents the point at which about 25\% of the variance in observations is explained by the predictions.}
 
@@
 comata} at temporal distances of two ($t_{(115)}$ = 2.39, \emph{P} = 0.002) and three years ($t_{(98)}$ = 2.04, \emph{P} = 0.022) (Table S24).
-There were some cases where the QBM was more accurate than the IPM for \emph{Poa secunda}, but neither model exceeded the forecast profieciency threshold by a large margin (Fig. 4, Table S24).}
+There were some cases where the QBM was more accurate than the IPM for \emph{Poa secunda}, but neither model exceeded the forecast proficiency threshold by a large margin (Fig. 4, Table S24).}
 
@@
 
-Our comparison of a traditional, demographic population model without environmental forcing (the IPM) to an equivalent model inspired by density-structured models (the QBM) showed that, generally, IPM forecasts of out-of-sample plant population states were no more accurate than forecasts from the QBM (Fig.
+Our comparison between a traditional, demographic population model without environmental forcing (the IPM) and an equivalent model inspired by density-structured models (the QBM) showed that IPM forecasts of out-of-sample plant population states were no more accurate than forecasts from the QBM (Fig.
 2; `no-climate' bars).
@@
 
-We also expected the inclusion of environmental forcing to reveal more differences between the models.
+We also expected the inclusion of environmental forcing to reveal further differences between the models.
 Interannual variation in weather can affect vital rates in different ways [@Dalgleish2011].
@@
 3).
-However, for the two species where including climate covariates increased forecast accuracy (*B.
-gracilis* and *Poa secunda*), forecast skill (Fig.
+\textcolor{blue}{For \emph{H. comata}, adding climate covariates did not improve forecasts (Fig. 2), despite the significant climate effects in the vital rate regressions (Fig. 3).}
+Furthermore, for the two species where including climate covariates increased forecast accuracy (*B.
+gracilis* and *Poa secunda*), forecast accuracy (Fig.
 2) and error (Fig.
@@
 In our case, the IPM explicitly models the influence of weather on recruitment and survival, effects that may be poorly represented in the QBM because recruitment and survival mainly affect small plants that contribute little to year-to-year changes in percent cover.
-Over time, the addition and subtraction of small plants can have large effects on population growth, thus, explicitly modeling these effects could contribute to a longer forecast horizon. 
+Over longer time scales, the addition and subtraction of small plants could have large effects on population growth, so explicitly modeling these effects could contribute to a longer forecast horizon.
 However, we found no evidence \textcolor{blue}{that the forecast horizon for the IPM was greater than the QBM (Fig. 4).
@@
 The QBM has fewer processes and parameters, which can reduce bias due to parameter uncertainty.
-Thus, the QBM may better capture near term dynamics when populations do not fluctuate widely, as in our case.}
+As a result, the QBM may better capture near term dynamics when populations do not fluctuate widely, as in our case.}
 
@@
 \textcolor{blue}{We may also have to shift our perspective from making explicit point forecasts to making moving average forecasts} [@Petchey2015].
-\textcolor{blue}{Whether the poor predictive ability of our models impacts their comparison relative to one another is an open question}.
+\textcolor{blue}{Whether the poor predictive ability of our models impacts the comparison of models based on individual vs. population-level data is an open question}.
 
-In conclusion, we found that models based on individual-level demographic data generally failed to generate more skillful population forecasts than models based on population-level data.
+In conclusion, we found that models based on individual-level demographic data generally failed to generate more skillful population forecasts than models based on population-level data, even in models which included climate covariates.
 This finding runs counter to our expectations, but is consistent with recent theoretical [@Freckleton2011] and empirical work [@Queenborough2011].
-Moreover, for the two species where including climate covariates improved forecast skill, both models generated equally skillful one-step-ahead forecasts.
-Thus, we conclude that models based on population-level data, rather than individual-level data, may be adequate for forecasting the states and dynamics of plant populations.
-Unfortunately, our analysis, where climate effects were relatively unimportant in vital rate regressions, did not allow us to sufficiently test our prediction that individual-level data is necessary to generate skillful forecasts if different vital rates respond to climate in unique, potentially opposing, ways.
-Nonetheless, our results are encouraging for the use of easy-to-collect population-level data for forecasting the state of plant populations.
+We conclude that models based on population-level data, rather than individual-level data, may be adequate for forecasting the states and dynamics of plant populations.
+This conclusion comes with the caveat that our analysis may be a weak test of the prediction that individual-level data is necessary for forecasting if different vital rates respond to climate in opposing ways, because climate effects were relatively unimportant in our vital rate regressions.
+Nonetheless, our results should encourage the use of easy-to-collect population-level data for forecasting the state of plant populations.
 

### 2016-08-30 manuscript/manuscript.tex
adds PBA line edits for R1; ready for USGS

@@
 (2011) provide empirical evidence that density-structured models are
-capable of reproducing population dynamics at landscape spatial scales,
-even if some precision is lost when compared to fully continuous models.
-However, previous tests of density-structured population models have yet
-to assess their ability to forecast out-of-sample observations, and they
-have not included environmental covariates, which are necessary to
-forecast population responses to climate change.
+capable of reproducing population dynamics at landscape spatial scales
+(also see Mieszkowska et al. 2013), even if some precision is lost when
+compared to fully continuous models. However, previous tests of
+density-structured population models have yet to assess their ability to
+forecast out-of-sample observations, and they have not included
+environmental covariates, which are necessary to forecast population
+responses to climate change.
 
@@
 example, a one-year change in a plant species' cover or biomass might
-reflect individual growth or shrinkage, whereas the long-term trajectory
-of the population might be more influenced by recruitment. The same is
-true for density dependence: intraspecific density depedence may act
-most strongly on vital rates, like recruitment, that are difficult to
-identify from population-level data. If density dependence and/or
-important climate effects are missed because of the aggregation inherent
-in population-level data, then population models built with such data
-will make uninformative or unreliable forecasts.
+reflect growth or shrinkage of the largest individuals, whereas the
+long-term trajectory of the population might be more influenced by
+recruitment. The same is true for density dependence: intraspecific
+density depedence may act most strongly on vital rates, like
+recruitment, that are difficult to identify from population-level data.
+If density dependence and/or important climate effects are missed
+because of the aggregation inherent in population-level data, then
+population models built with such data will make uninformative or
+unreliable forecasts.
 
@@
 
-We fit all models using a hierarchical Bayesian approach. We focus on
-the main process and the model likelihood in what follows (full model
-descriptions are in the Supporting Information). For the likelihood
-models, \(\textbf{y}^X\) is always the relevant vector of observations
-for vital rate \emph{X} (\emph{X} = \emph{S}, \emph{G}, \emph{R}, or
-\emph{P} for survival, growth, recruitment, or population growth). For
-example, \(\textbf{y}^S\) is a vector of 0s and 1s indicating whether a
-genet survives from \emph{t} to \emph{t+1}, or not, for all observation
-years and quadrats. All model parameters are species-specific, but we
-omit subscripts for species in model descriptions below to reduce visual
-clutter. For brevity, we only describe models with climate covariates
-included, but models without climate covariates are simply the models
-described below with the climate effects removed.
+We fit all models using a hierarchical Bayesian approach. In the
+following description, we focus on the main process and the model
+likelihood (full model descriptions are in the Supporting Information).
+For the likelihood models, \(\textbf{y}^X\) is always the relevant
+vector of observations for vital rate \emph{X} (\emph{X} = \emph{S},
+\emph{G}, \emph{R}, or \emph{P} for survival, growth, recruitment, or
+population growth). For example, \(\textbf{y}^S\) is a vector of 0s and
+1s indicating whether a genet survives from \emph{t} to \emph{t+1}, or
+not, for all observation years and quadrats. All model parameters are
+species-specific, but we omit subscripts for species in model
+descriptions below to reduce visual clutter. For brevity, we only
+describe models with climate covariates included, but models without
+climate covariates are simply the models described below with the
+climate effects removed.
 
@@
 mean size effect and intercept. Quadrat random effects were also fit
-hierarchically, with quadrat offsets modeled with distributions with
+hierarchically, with quadrat offsets modeled using distributions with
 mean zero and a shared variance term (independent Gaussian priors).
@@
 forecast horizon of one year, the forecast in 1942 has a forecast
-horizon of two years, and so on. We performed these simulations for all
-model types (IPM with/without climate; QBM with/without climate) using
-mean parameter values for all possible initial years. For a given
+horizon of two years, and so on. We performed these simulations using
+mean parameter values for all model types (IPM with/without climate; QBM
+with/without climate) and all possible initial years. For a given
 forecast distance, we averaged the correlation between forecasts and
@@
 \textcolor{blue}{We chose an arbitrary forecast accuracy of $\rho = 0.5$ as our forecast proficiency threshold.
+the forecast horizon is the temporal distance at which forecast accuracy falls below $\rho = 0.5$.
 For basic research on forecasting, arbitrary proficiency thresholds suffice for comparative purposes}
 (Petchey et al. 2015),
-\textcolor{blue}{and $\rho = 0.5$ represents the point at which about 25% of the variance in observations is explained by the predictions.
-So, the forecast horizon is the temporal distance at which forecast accuracy falls below $\rho = 0.5$.}
+\textcolor{blue}{and $\rho = 0.5$ represents the point at which about 25\% of the variance in observations is explained by the predictions.}
 
@@
 
-Our comparison of a traditional, demographic population model without
-environmental forcing (the IPM) to an equivalent model inspired by
-density-structured models (the QBM) showed that, generally, IPM
-forecasts of out-of-sample plant population states were no more accurate
-than forecasts from the QBM (Fig. 2; `no-climate' bars). This result
-differed from our expectation that the IPM would out-perform the QBM,
-because of its mechanistic representation of the perennial life cycle.
-Our result also confirms theoretical (Freckleton et al. 2011) and
-empirical work (Taylor and Hastings 2004, Queenborough et al. 2011)
-showing that density-structured models can be useful surrogates for
-demographic models when the goal is to estimate or forecast population
-states over large spatial extents.
-
-We also expected the inclusion of environmental forcing to reveal more
-differences between the models. Interannual variation in weather can
-affect vital rates in different ways (Dalgleish et al. 2011). Thus,
+Our comparison between a traditional, demographic population model
+without environmental forcing (the IPM) and an equivalent model inspired
+by density-structured models (the QBM) showed that IPM forecasts of
+out-of-sample plant population states were no more accurate than
+forecasts from the QBM (Fig. 2; `no-climate' bars). This result differed
+from our expectation that the IPM would out-perform the QBM, because of
+its mechanistic representation of the perennial life cycle. Our result
+also confirms theoretical (Freckleton et al. 2011) and empirical work
+(Taylor and Hastings 2004, Queenborough et al. 2011) showing that
+density-structured models can be useful surrogates for demographic
+models when the goal is to estimate or forecast population states over
+large spatial extents.
+
+We also expected the inclusion of environmental forcing to reveal
+further differences between the models. Interannual variation in weather
+can affect vital rates in different ways (Dalgleish et al. 2011). Thus,
 estimates of climate effects on plant population growth may be biased or
@@
 climate effects were identified in the individual vital rate models but
-not in the percent cover model (Fig. 3). However, for the two species
-where including climate covariates increased forecast accuracy (\emph{B.
-gracilis} and \emph{Poa secunda}), forecast skill (Fig. 2) and error
-(Fig. S2) were equivalent between the IPM and QBM.
+not in the percent cover model (Fig. 3).
+\textcolor{blue}{For \emph{H. comata}, adding climate covariates did not improve forecasts (Fig. 2), despite the significant climate effects in the vital rate regressions (Fig. 3).}
+Furthermore, for the two species where including climate covariates
+increased forecast accuracy (\emph{B. gracilis} and \emph{Poa secunda}),
+forecast accuracy (Fig. 2) and error (Fig. S2) were equivalent between
+the IPM and QBM.
 
@@
 the QBM because recruitment and survival mainly affect small plants that
-contribute little to year-to-year changes in percent cover. Over time,
-the addition and subtraction of small plants can have large effects on
-population growth, thus, explicitly modeling these effects could
-contribute to a longer forecast horizon. However, we found no evidence
+contribute little to year-to-year changes in percent cover. Over longer
+time scales, the addition and subtraction of small plants could have
+large effects on population growth, so explicitly modeling these effects
+could contribute to a longer forecast horizon. However, we found no
+evidence
 \textcolor{blue}{that the forecast horizon for the IPM was greater than the QBM (Fig. 4).
@@
 (Petchey et al. 2015).
-\textcolor{blue}{Whether the poor predictive ability of our models impacts their comparison relative to one another is an open question}.
+\textcolor{blue}{Whether the poor predictive ability of our models impacts the comparison of models based on individual vs. population-level data is an open question}.
 
@@
 demographic data generally failed to generate more skillful population
-forecasts than models based on population-level data. This finding runs
-counter to our expectations, but is consistent with recent theoretical
-(Freckleton et al. 2011) and empirical work (Queenborough et al. 2011).
-Moreover, for the two species where including climate covariates
-improved forecast skill, both models generated equally skillful
-one-step-ahead forecasts. Thus, we conclude that models based on
-population-level data, rather than individual-level data, may be
-adequate for forecasting the states and dynamics of plant populations.
-Unfortunately, our analysis, where climate effects were relatively
-unimportant in vital rate regressions, did not allow us to sufficiently
-test our prediction that individual-level data is necessary to generate
-skillful forecasts if different vital rates respond to climate in
-unique, potentially opposing, ways. Nonetheless, our results are
-encouraging for the use of easy-to-collect population-level data for
-forecasting the state of plant populations.
+forecasts than models based on population-level data, even in models
+which included climate covariates. This finding runs counter to our
+expectations, but is consistent with recent theoretical (Freckleton et
+al. 2011) and empirical work (Queenborough et al. 2011). We conclude
+that models based on population-level data, rather than individual-level
+data, may be adequate for forecasting the states and dynamics of plant
+populations. This conclusion comes with the caveat that our analysis may
+be a weak test of the prediction that individual-level data is necessary
+for forecasting if different vital rates respond to climate in opposing
+ways, because climate effects were relatively unimportant in our vital
+rate regressions. Nonetheless, our results should encourage the use of
+easy-to-collect population-level data for forecasting the state of plant
+populations.
 
@@
 
+Mieszkowska, N., G. Milligan, M. T. Burrows, R. Freckleton, and M.
+Spencer. 2013. Dynamic species distribution models from categorical
+survey data. Journal of Animal Ecology 82:1215--1226.
+
 Mouquet, N., Y. Lagadeuc, V. Devictor, L. Doyen, A. Duputi{é}, D.

### 2016-08-30 README.Rmd
adds USGS disclaimer to README

@@
 
+##  USGS Disclaimers
+#### Code:
+Although this software program has been used by the U.S.
+Geological Survey (USGS), no warranty, expressed or implied, is made by the USGS or the U.S.
+Government as to the accuracy and functioning of the program and related program material nor shall the fact of distribution constitute any such warranty, and no responsibility is assumed by the USGS in connection therewith.
+
+#### Data:
+Although these data have been processed successfully on a computer system at the U.S.
+Geological Survey (USGS), no warranty expressed or implied is made regarding the display or utility of the data on any other system or for general or scientific purposes, nor shall the act of distribution constitute any such warranty.
+The USGS or the U.S.
+Government shall not be held liable for improper or incorrect use of the data described and/or contained herein.
+
 File Structure

### 2016-09-07 manuscript/manuscript_newSI.Rmd
trying new version with more SI material

--- 2016-08-30 manuscript/manuscript.Rmd
@@
 ---
-layout: 12pt
+bibliography: ~/Dropbox/Bibliography/MicroMeso.bib
+csl: components/ecology.csl
+fontsize: 12pt
+geometry: margin=1in
 header-includes:
@@
 - \usepackage{color, soul}
-bibliography: ~/Dropbox/Bibliography/MicroMeso.bib
-csl: components/ecology.csl
-
-## rmarkdown render options
+layout: 12pt
+linkcolor: black
 output:
   pdf_document:
-    fig_caption: true
-    keep_tex: true
-fontsize: 12pt
-geometry: margin=1in
-linkcolor: black
+    fig_caption: yes
+    keep_tex: yes
+  word_document: default
 urlcolor: black
-
 ---
@@
 At the quadrat-level, we fit a single regression model for population growth.
-We describe the statistical models separately because they required different approaches.
+These regressions are described in detail in the Supporting Information.
 For both model types, we fit vital rate models with and without climate covariates.
@@
 We also include interactions among same-year climate covariates (e.g., ppt1 $\times$ TmeansSpr1), resulting in a total of seven climate covariates.
+We fit all models using a hierarchical Bayesian approach.
 
-We fit all models using a hierarchical Bayesian approach.
+<!--
 In the following description, we focus on the main process and the model likelihood (full model descriptions are in the Supporting Information).
@@
 The log normal likelihood includes a truncation (subscript 0, superscript 1) to ensure that predicted values do not exceed 100% cover.
+-->
 
@@
 Therefore, we removed that chain from the analysis, leaving two well-mixed and converged chains for those six parameters.
--->
 
@@
 We calculated the *lppd* from posterior samples using the algorithm from @Vehtari2016.
+-->
 

### 2016-09-07 manuscript/manuscript_newSI.tex
trying new version with more SI material

--- 2016-08-30 manuscript/manuscript.tex
@@
 recruitment for each species. At the quadrat-level, we fit a single
-regression model for population growth. We describe the statistical
-models separately because they required different approaches. For both
-model types, we fit vital rate models with and without climate
-covariates. Models with climate effects contain five climate covariates
-that we chose \emph{a priori} based on previous model selection efforts
-using these data (Chu et al. 2016) and expert advice (Lance Vermeire,
-\emph{personal communication}): ``water year'' precipitation at
-\emph{t}-2 (lagppt); April through June precipitation at \emph{t}-1 and
-\emph{t} (ppt1 and ppt2, respectively) and April through June
-temperature at \emph{t}-1 and \emph{t} (TmeanSpr1 and TmeanSpr2,
-respectively), where \emph{t}-1 to \emph{t} is the transition of
-interest. We also include interactions among same-year climate
-covariates (e.g., ppt1 \(\times\) TmeansSpr1), resulting in a total of
-seven climate covariates.
-
-We fit all models using a hierarchical Bayesian approach. In the
-following description, we focus on the main process and the model
-likelihood (full model descriptions are in the Supporting Information).
-For the likelihood models, \(\textbf{y}^X\) is always the relevant
-vector of observations for vital rate \emph{X} (\emph{X} = \emph{S},
-\emph{G}, \emph{R}, or \emph{P} for survival, growth, recruitment, or
-population growth). For example, \(\textbf{y}^S\) is a vector of 0s and
-1s indicating whether a genet survives from \emph{t} to \emph{t+1}, or
-not, for all observation years and quadrats. All model parameters are
-species-specific, but we omit subscripts for species in model
-descriptions below to reduce visual clutter. For brevity, we only
-describe models with climate covariates included, but models without
-climate covariates are simply the models described below with the
-climate effects removed.
-
-\paragraph{Vital rate models at the individual
-level}\label{vital-rate-models-at-the-individual-level}
-
-We used logistic regression to model the probability that genet \emph{i}
-in quadrat \emph{q} survives from time \emph{t} to \emph{t}+1
-(\(s_{i,q,t}\)):
-
-\vspace{-3em}\begin{align}
-y_{i,q,t}^{S} &\sim \text{Bernoulli}(s_{i,q,t}), \\
-\text{logit}(s_{i,q,t}) &= \beta_{0,t} + \beta_{s,t}x_{i,q,t} + \beta_{Q,q} + \textbf{z}'_t \boldsymbol{\beta}_c + \beta_{d,1} w_{i,t} + \beta_{d,2} (x_{i,q,t}w_{i,q,t}),
-\end{align}\vspace{-3em}
-
-where \(x_{i,q,t}\) is the log of genet \emph{i} basal area at time
-\emph{t}, \(\beta_{0,t}\) is a year specific intercept, \(\beta_{Q,q}\)
-is the random effect of the \emph{q}th quadrat to account for spatial
-location, \(\beta_{s,t}\) is the year-specific slope parameter for size,
-\textbf{z} is a vector of \emph{p} climate covariates specific to year
-\emph{t}, \(\boldsymbol{\beta}_c\) is a vector of fixed climate effects
-of length \emph{p}, \(\beta_{d,1}\) is the effect of intraspecific
-crowding experienced by the focal genet at time \emph{t}
-(\(w_{i,q,t}\)), and \(\beta_{d,2}\) is a size by crowding
-(\(x_{i,q,t}w_{i,q,t}\)) interaction effect.
-
-We follow the approach of Chu and Adler (2015) to estimate crowding,
-assuming that the crowding experienced by a focal genet depends on
-distance to each neighbor genet and the neighbor's size, \emph{u}:
-
-\begin{equation}
-w_{i,q,t} = \sum_k e^{-\delta d_{ik,q,t}^{2}}u_{k,q,t}.
-\end{equation}
-
-In equation 3, \(w_{i,q,t}\) is the crowding that genet \emph{i} in year
-\emph{t} experiences from \emph{k} conspecific neighbors (\(u_{k,q,t}\))
-in quadrat \emph{q}. The spatial scale over which conspecific neighbors
-exert influence on any genet is determined by \(\delta\). The function
-is applied for all \emph{k} conspecific genets that neighbor the focal
-genet at time \emph{t}, and \(d_{ik,q,t}\) is the distance between genet
-\emph{i} and conspecific genet \emph{k} in quadrat \emph{q}. We use
-regression-specific (survival and growth) \(\delta\) values estimated by
-Chu and Adler (2015).
-
-We modeled growth as a Gaussian process describing log genet size
-(\(y_{i,q,t+1}^{G}\)) at time \(t+1\) in quadrat \emph{q} as a function
-of log size at time \(t\) and climate covariates:
-
-\vspace{-3em}\begin{align}
-y_{i,q,t+1}^G &\sim \text{Normal}(\mu_{i,q,t+1}, \sigma^2_{x_{i,q,t+1}}), \\
-\mu_{i,q,t+1} &= \beta_{0,t} + \beta_{s,t}x_{i,q,t} + \beta_{Q,q} + \textbf{z}'_t \boldsymbol{\beta}_c + \beta_{d,1} w_{i,q,t} + \beta_{d,2} (x_{i,q,t}w_{i,q,t}),
-\end{align}\vspace{-3em}
-
-where \(\mu_{i,q,t+1}\) is log of genet \emph{i}s predicted size at time
-\emph{t} + 1, and all other parameters are as described for the survival
-regression. We capture non-constant error variance in growth by modeling
-the variance in the growth regression (\(\sigma^2_{x{i,q,t+1}}\)) as a
-nonlinear function of predicted genet size:
-
-\vspace{-3em}\begin{align}
-\sigma^2_{x_{i,q,t+1}} = a \, \text{exp}[b \times \mu_{i,q,t+1}],
-\end{align}\vspace{-3em}
-
-where \(\mu_{i,q,t+1}\) is log of predicted genet size predicted from
-the growth regression (Eq. 4), and \emph{a} and \emph{b} are constants.
-
-Our data allows us to track new recruits, but we cannot assign a
-specific parent to new genets. Therefore, we model recruitment at the
-quadrat level. We assume the number of individuals, \(y^{R}_{q,t+1}\),
-recruiting at time \(t+1\) in quadrat \emph{q} follows a negative
-binomial distribution:
-
-\vspace{-3em}\begin{align}
-y^{R}_{q,t+1} \sim \text{NegBin}(\lambda_{q,t+1},\phi),
-\end{align}\vspace{-3em}
-
-where \(\lambda\) is the mean intensity and \(\phi\) is the size
-parameter. We define \(\lambda\) as a function of quadrat composition
-and climate in the previous year:
-
-\vspace{-3em}\begin{align}
-\lambda_{q,t+1} = \tilde{c}_{q,t} \, \text{exp}\left(\beta_{0,t} + \beta_{Q,q} + \textbf{z}'_t \boldsymbol{\beta}_c + \beta_{d}\sqrt{\tilde{c}_{q,t}}\right),
-\end{align}\vspace{-3em}
-
-where \(\tilde{c}_{q,t}\) is effective cover (\(\text{cm}^2\)) of the
-focal species in quadrat \emph{q} at time \emph{t}, and all other terms
-are as in the survival and growth regressions. Effective cover is a
-mixture of observed cover (\emph{c}) in the focal quadrat (\emph{q}) and
-the mean cover across the entire group (\(\bar{c}\)) of \emph{Q}
-quadrats in which \emph{q} is located:
-
-\vspace{-3em}\begin{align}
-\tilde{c}_{q,t} = pc_{q,t} + (1-p)\bar{c}_{Q,t},
-\end{align}\vspace{-3em}
-
-where \(p\) is a mixing fraction between 0 and 1 that is estimated when
-fitting the model.
-
-\paragraph{Population model at the quadrat
-level}\label{population-model-at-the-quadrat-level}
-
-The statistical approach used to model aggregated data depends on the
-type of data collected. We have percent cover data, which can easily be
-transformed to proportion data in our case because plant areas were
-scaled by plot area. An obvious choice for fitting a linear model to
-proportion data is beta regression because the support of the beta
-distribution is (0,1), which does not include true zeros or ones.
-However, when we used fitted model parameters from a beta regression in
-a quadrat-based population model, the simulated population tended toward
-100\% cover for all species. We therefore chose a modeling approach
-based on a truncated log-normal likelihood. The model for quadrat cover
-change from time \(t\) to \(t+1\) is
-
-\vspace{-3em}\begin{align}
-y_{q,t+1}^{P} &\sim \text{LogNormal}(\mu_{q,t+1}, \sigma^2)_{0}^{1}, \\
-\mu_{q,t+1} &= \beta_{0,t} + \beta_{s,t}x_{q,t} + \beta_{Q,q} + \textbf{z}'_t \boldsymbol{\beta}_c,
-\end{align}\vspace{-3em}
-
-where \(\mu_{q,t+1}\) is the log of proportional cover in quadrat
-\emph{q} at time \(t+1\), and all other parameters are as in the
-individual-level growth model (Eq. 4) except that \emph{x} now
-represents log of proportional cover. The log normal likelihood includes
-a truncation (subscript 0, superscript 1) to ensure that predicted
-values do not exceed 100\% cover.
+regression model for population growth. These regressions are described
+in detail in the Supporting Information. For both model types, we fit
+vital rate models with and without climate covariates. Models with
+climate effects contain five climate covariates that we chose
+\emph{a priori} based on previous model selection efforts using these
+data (Chu et al. 2016) and expert advice (Lance Vermeire, \emph{personal
+communication}): ``water year'' precipitation at \emph{t}-2 (lagppt);
+April through June precipitation at \emph{t}-1 and \emph{t} (ppt1 and
+ppt2, respectively) and April through June temperature at \emph{t}-1 and
+\emph{t} (TmeanSpr1 and TmeanSpr2, respectively), where \emph{t}-1 to
+\emph{t} is the transition of interest. We also include interactions
+among same-year climate covariates (e.g., ppt1 \(\times\) TmeansSpr1),
+resulting in a total of seven climate covariates. We fit all models
+using a hierarchical Bayesian approach.
 
@@
 Information (Tables S5-S20).
-
-\paragraph{Statistical regularization: Bayesian ridge
-regression}\label{statistical-regularization-bayesian-ridge-regression}
-
-For models with climate covariates, our objective is to model the
-response of our focal grass species to interannual variation in climate,
-even if those responses are weak. Therefore, we avoid selecting among
-models with all possible combinations of climate covariates, and instead
-use Bayesian ridge regression to regulate, or constrain, the posterior
-distributions of each climate covariate (Gerber et al. 2015, Hooten and
-Hobbs 2015). Ridge regression is a specific application of statistical
-regularization that seeks to optimize model generality by trading off
-bias and variance. As the name implies, statistical regularization
-involves the use of a regulator that constrains an optimization. The
-natural regulator in a Bayesian application is the prior on the
-coefficients of interest. Each of our statistical models includes the
-effects of climate covariates via the term
-\(\textbf{z}'_t \boldsymbol{\beta}_c\) with prior
-\(\boldsymbol{\beta}_c \sim \text{Normal}(\boldsymbol{\mu}_{\beta_c}, \sigma_{\beta_c}^2\textbf{I})\).
-Because we standardized all climate covariates to have mean zero and
-variance one, we set \(\boldsymbol{\mu}_{\beta_c} = 0\) and let
-\(\sigma_{\beta_c}^2\) serve as the regulator that shrinks covariate
-effects toward zero -- the smaller the prior variance, the more the
-posteriors of \(\boldsymbol{\beta}_c\) are shrunk toward zero, and the
-stronger the penalty (Hooten and Hobbs 2015).
-
-To find the optimal penalty (i.e., optimal value of the hyperparameter
-\(\sigma_{\beta_c}^2\)), we fit each statistical model with a range of
-values for \(\sigma_{\beta_c}^2\) and compared predictive scores from
-leave-one-year-out cross-validation. We performed the grid search over
-24 values of \(\sigma_{\beta_c}^2\), ranging from
-\(\sigma_{\beta_c}^2 = 0.01\) to \(\sigma_{\beta_c}^2 = 2.25\). For each
-statistical model and each species, we fit \(13\times24=312\) iterations
-of the model fitting algorithm to search \(\sigma_{\beta_c}^2\) for the
-optimal value (13 years to leave out for cross-validation and 24 values
-of \(\sigma_{\beta_c}^2\)) -- a total of 4,992 model fits. We calculated
-the log pointwise predictive density (\emph{lppd}) to score each model's
-ability to predict the left-out data (Gelman et al. 2014). Thus, for
-training data \(y_{\text{train}}\) and held-out data \(y_{\text{hold}}\)
-at a given value of \(\sigma_\theta^2\) across all MCMC samples
-\(s=1,2,...,S\) and all hold outs of data from year \emph{t} to year
-\emph{T}, and letting \(\theta\) represent all unknowns, \emph{lppd} is
-
-\vspace{-3em}\begin{align}
-\text{lppd}_{\text{CV}} = \sum_{t=1}^T \text{log}_e \int [y_{t,\text{hold}}|\theta][\theta|y_{\text{train}}]d\theta,
-\end{align}\vspace{-3em}
-
-and computed as
-
-\vspace{-3em}\begin{align}
-\sum_{t=1}^T \text{log}_e \left(\frac{1}{S} \sum^S_{s=1}[y_{t,\text{hold}}|\theta_{ts}]  \right).
-\end{align}\vspace{-3em}
-
-We chose the optimal prior variance for each species-statistical model
-combination as the one that produced the highest \emph{lppd} and then
-fit each species-statistical model combination using the full data set
-for each species and the optimal prior variance. We calculated the
-\emph{lppd} from posterior samples using the algorithm from Vehtari et
-al. (2016).
 
@@
 for identifying the appropriate time periods over which to aggregate to
-produce a tractable set of candidate climate variables (Sims et al.
-2007, Pol and Cockburn 2011, Teller et al. 2016).
+produce a tractable set of candidate climate variables (Teller et al.
+2016, {\textbf{???}}, {\textbf{???}}).
 
@@
 
-Gelman, A., J. Hwang, and A. Vehtari. 2014. Understanding predictive
-information criteria for Bayesian models. Statistics and Computing
-24:997--1016.
-
-Gerber, B. D., W. L. Kendall, M. B. Hooten, J. A. Dubovsky, and R. C.
-Drewien. 2015. Optimal population prediction of sandhill crane
-recruitment based on climate-mediated habitat limitations. Journal of
-Animal Ecology 84:1299--1310.
-
 Hobbs, N. T., and M. B. Hooten. 2015. Bayesian Models: A Statistical
@@
 
-Pol, M. van de, and A. Cockburn. 2011. Identifying the critical climatic
-time window that affects trait expression. The American naturalist
-177:698--707.
-
 Queenborough, S. A., K. M. Burnet, W. J. Sutherland, A. R. Watkinson,
@@
 policy under global change. Science 347:953--954.
-
-Sims, M., D. A. Elston, A. Larkham, D. H. Nussey, and S. D. Albon. 2007.
-Identifying when weather influences life-history traits of grazing
-herbivores. Journal of Animal Ecology 76:761--770.
 
@@
 
-Vehtari, A., A. Gelman, and J. Gabry. 2016. Efficient implementation of
-leave-one-out cross-validation and WAIC for evaluating fitted Bayesian
-models. ArXiv preprint.
-
 Wilcox, R. R. 2009. Comparing Pearson Correlations: Dealing with

### 2016-10-11 manuscript/manuscript.Rmd
minor changes for final MEE files

@@
 
-\textbf{Word count:} 6845 (including reference section)
+\textbf{Word count:} 8039 (including captions and reference list)
 
@@
 However, it is unknown whether models built using population-level data adequately capture the effects of density-dependence and environmental forcing that are necessary to generate skillful forecasts.
-        \item Here, we test the consequences of aggregating individual responses when forecasting the population states \textcolor{blue}{(percent cover)} and trajectories of four perennial grass species in a semi-arid grassland in Montana, USA.
-We parameterized two population models for each species, one based on individual-level data (survival, growth and recruitment) and one on population-level data (percent cover), and compared their forecasting \textcolor{blue}{accuracy} and forecast horizons with and without the inclusion of climate covariates.
+        \item Here, we test the consequences of aggregating individual responses when forecasting the population states (percent cover) and trajectories of four perennial grass species in a semi-arid grassland in Montana, USA.
+We parameterized two population models for each species, one based on individual-level data (survival, growth and recruitment) and one on population-level data (percent cover), and compared their forecasting accuracy and forecast horizons with and without the inclusion of climate covariates.
 For both models, we used Bayesian ridge regression to weight the influence of climate covariates for optimal prediction.
-        \item In the absence of climate effects, we found no significant difference between the forecast \textcolor{blue}{accuracy} of models based on individual-level data and models based on population-level data.
-Climate effects were weak, but increased forecast \textcolor{blue}{accuracy} for two species.
-Increases in \textcolor{blue}{accuracy} with climate covariates were similar between model types.
-        \item In our case study, percent cover models generated forecasts as \textcolor{blue}{accurate} as those from a demographic model.
+        \item In the absence of climate effects, we found no significant difference between the forecast accuracy of models based on individual-level data and models based on population-level data.
+Climate effects were weak, but increased forecast accuracy for two species.
+Increases in accuracy with climate covariates were similar between model types.
+        \item In our case study, percent cover models generated forecasts as accurate as those from a demographic model.
 For the goal of forecasting, models based on aggregated individual-level data may offer a practical alternative to data-intensive demographic models.
@@
 
-We compared the forecasting skill \textcolor{blue}{(accuracy and precision)} of statistical and population models based on aggregated, population-level data with the skill of models based on individual-level data.
+We compared the forecasting skill (accuracy and precision) of statistical and population models based on aggregated, population-level data with the skill of models based on individual-level data.
 We used a demographic dataset that tracks the fates of individual plants from four species over 14 years to build two kinds of single-species population models, traditional models using individual growth, survival, and recruitment data and alternative models based on population-level (basal cover) data.
@@
 ### Overview of analysis
-\textcolor{blue}{We used two types of data: individual-level data and percent cover data.
+We used two types of data: individual-level data and percent cover data.
 Using the individual-level data, we fit three vital rate regressions (survival, growth, and rectruitment) to build an Integral Projection Model (IPM) for simulating the plant populations.
@@
 We compared the forecast accuracy ($\rho$, correlation between observations and predictions) and mean absolute error (MAE) between the IPM and the QBM to test our expectation that the IPM should outperform the QBM.
-Lastly, we use in-sample forecasts to quantify each model's forecast horizon} [@Petchey2015].
+Lastly, we use in-sample forecasts to quantify each model's forecast horizon [@Petchey2015].
 
@@
 
-<!--
-There were six instances where the scale reduction factor for a particular parameter was greater than 1.1.
-In those cases, we checked the traceplots of the violating parameter and discovered that one chain, out of three, was poorly behaved.
-Therefore, we removed that chain from the analysis, leaving two well-mixed and converged chains for those six parameters.
--->
-
 #### Statistical regularization: Bayesian ridge regression
@@
 We compared $\rho$ and MAE between model types and within model types between models with and without climate covariates using one-sided *t* tests with adjusted degrees of freedom following @Wilcox2009 and standard errors calculated using the HC4 estimator of @Cribari-Neto2004.
-Statistical tests \textcolor{blue}{for comparing correlations and error} were conducted using algorithms from @Ye2015a.
+Statistical tests for comparing correlations and error were conducted using algorithms from @Ye2015a.
 
@@
 An important feature of any forecasting model is the rate at which forecast skill declines as the time between an observation and a forecast increases.
-\textcolor{blue}{In particular, we are interested in the temporal distance at which forecast skill falls below a threshold: the so-called ecological forecast horizon} [@Petchey2015].
+In particular, we are interested in the temporal distance at which forecast skill falls below a threshold: the so-called ecological forecast horizon [@Petchey2015].
 To assess the forecast horizons of our models, we initiate the forecast model with the population state at some time *t* and make sequential forecasts of the population at times $t+1, t+2, \dots, t+T$ where *T* is the maximum number of years between the initial year and the final year of our observations.
@@
 Nonetheless, our simulations offer insight into the differences among model forecast horizons.
-\textcolor{blue}{We chose an arbitrary forecast accuracy of $\rho = 0.5$ as our forecast proficiency threshold.
-the forecast horizon is the temporal distance at which forecast accuracy falls below $\rho = 0.5$.
-For basic research on forecasting, arbitrary proficiency thresholds suffice for comparative purposes} [@Petchey2015], \textcolor{blue}{and $\rho = 0.5$ represents the point at which about 25\% of the variance in observations is explained by the predictions.}
+We chose an arbitrary forecast accuracy of $\rho = 0.5$ as our forecast proficiency threshold. the forecast horizon is the temporal distance at which forecast accuracy falls below $\rho = 0.5$.
+For basic research on forecasting, arbitrary proficiency thresholds suffice for comparative purposes [@Petchey2015], and $\rho = 0.5$ represents the point at which about 25\% of the variance in observations is explained by the predictions.
 
@@
 
-With climate covariates included \textcolor{blue}{and using mean parameter values, the accuracy of both models' forecasts declined as the distance between the last observation and the forecast increased}, but they did so at similar rates (Fig.
+With climate covariates included and using mean parameter values, the accuracy of both models' forecasts declined as the distance between the last observation and the forecast increased, but they did so at similar rates (Fig.
 4).
-The only exception is for *Poa secunda*, where QBM forecast accuracy remained steady as the \textcolor{blue}{temporal distance of the forecast increased}, whereas IPM forecast accuracy declined (Fig.
-4).
-\textcolor{blue}{The forecast horizons were short: forecast accuracy fell below $\rho$ = 0.5 after one year for the IPM for most species, and after four years, at most, for the QBM (Fig. 4).
+The only exception is for *Poa secunda*, where QBM forecast accuracy remained steady as the temporal distance of the forecast increased, whereas IPM forecast accuracy declined (Fig. 4).
+The forecast horizons were short: forecast accuracy fell below $\rho$ = 0.5 after one year for the IPM for most species, and after four years, at most, for the QBM (Fig. 4).
 Across the different temporal distances from the observation to the forecast, the IPM was never more accurate than the QBM (\emph{P} > 0.05 for all one-sided \emph{t}-tests, Table S23).
@@
 comata} at temporal distances of two ($t_{(115)}$ = 2.39, \emph{P} = 0.002) and three years ($t_{(98)}$ = 2.04, \emph{P} = 0.022) (Table S24).
-There were some cases where the QBM was more accurate than the IPM for \emph{Poa secunda}, but neither model exceeded the forecast proficiency threshold by a large margin (Fig. 4, Table S24).}
+There were some cases where the QBM was more accurate than the IPM for \emph{Poa secunda}, but neither model exceeded the forecast proficiency threshold by a large margin (Fig. 4, Table S24).
 
@@
 ----------
-<!--
-Population models built using individual-level data provide inference on demographic processes, but they can only forecast future population states across the (typically limited) spatial extent of the observations.
-Population-level data are much easier to collect across broad spatial extents, so models built using such data offer an appealing alternative to traditional population models [@Queenborough2011]. 
-However, density-structured models rely on the aggregation of individual-level data. 
-Given that individuals, not populations, respond to intraspecific competition and weather [@Clark2011b], can models based on population-level metrics generate forecasts that are as skillful as those generated from models based on individual-level data?
-Are models based on population-level metrics as sensitive to climate forcing as models based on individual-level data?
--->
-
 Our comparison between a traditional, demographic population model without environmental forcing (the IPM) and an equivalent model inspired by density-structured models (the QBM) showed that IPM forecasts of out-of-sample plant population states were no more accurate than forecasts from the QBM (Fig.
@@
 3).
-\textcolor{blue}{For \emph{H. comata}, adding climate covariates did not improve forecasts (Fig. 2), despite the significant climate effects in the vital rate regressions (Fig. 3).}
+For \emph{H. comata}, adding climate covariates did not improve forecasts (Fig. 2), despite the significant climate effects in the vital rate regressions (Fig. 3).
 Furthermore, for the two species where including climate covariates increased forecast accuracy (*B.
@@
 
-We also expected IPM forecast accuracy to decline at a lower rate than the QBM as the \textcolor{blue}{time between the model initialization and the forecast increased.}
+We also expected IPM forecast accuracy to decline at a lower rate than the QBM as the time between the model initialization and the forecast increased.
 In principle, more mechanistic models should produce better predictions, especially under novel conditions [@Evans2012b; @Schindler2015b].
@@
 Over longer time scales, the addition and subtraction of small plants could have large effects on population growth, so explicitly modeling these effects could contribute to a longer forecast horizon.
-However, we found no evidence \textcolor{blue}{that the forecast horizon for the IPM was greater than the QBM (Fig. 4).
+However, we found no evidence that the forecast horizon for the IPM was greater than the QBM (Fig. 4).
 On the contrary, the QBM tended to have a slightly longer forecast horizon than the IPM for most species (Fig. 4).
 The QBM has fewer processes and parameters, which can reduce bias due to parameter uncertainty.
-As a result, the QBM may better capture near term dynamics when populations do not fluctuate widely, as in our case.}
+As a result, the QBM may better capture near term dynamics when populations do not fluctuate widely, as in our case.
 
@@
 As advocated by @Freckleton2011, knowledge of a species' population ecology should guide the modeling approach.
-\textcolor{blue}{Third, our percent cover data are essentially error-free because we were able to aggregate indiviual plant areas to calculate percent cover.
-Percent cover data collected by typical sampling methods (e.g., Daubenmire frames) will include error that may affect population forecasts due to misspecifing the initial conditions and/or biasing model parameters} [@Queenborough2011].
-\textcolor{blue}{Percent cover models based on data containing more measurement error than ours might perform worse in comparison with individual-based models.
-One way to account for such error is to develop a sampling model that relates the observations (estimated percent cover in a plot) to the true state (percent cover derived from individual plant measurements in the same plot)} [@Hobbs2015].
+Third, our percent cover data are essentially error-free because we were able to aggregate indiviual plant areas to calculate percent cover.
+Percent cover data collected by typical sampling methods (e.g., Daubenmire frames) will include error that may affect population forecasts due to misspecifing the initial conditions and/or biasing model parameters [@Queenborough2011].
+Percent cover models based on data containing more measurement error than ours might perform worse in comparison with individual-based models.
+One way to account for such error is to develop a sampling model that relates the observations (estimated percent cover in a plot) to the true state (percent cover derived from individual plant measurements in the same plot) [@Hobbs2015].
 
-\textcolor{blue}{Although our main goal was to compare individual-based and population-level modeling approaches relative to one another, it is worth reflecting on the absolute forecasting skill of our models.
+Although our main goal was to compare individual-based and population-level modeling approaches relative to one another, it is worth reflecting on the absolute forecasting skill of our models.
 In particular, the forecast horizon of both models, defined as the time horizon at which the correlation between predictions and observations falls below $\rho = 0.5$, is less than five years for all species.
-Such short forecast horizons are not encouraging.}
-\textcolor{blue}{Unfortunately, we have few ideas about how to improve population forecasts that have not already been proposed} [@Petchey2015; @Mouquet2015].
-\textcolor{blue}{Longer time-series should improve our ability to detect exogenous drivers such as climate} [@Teller2016], \textcolor{blue}{and modeling larger spatial extents may reduce parameter uncertainty} [@Petchey2015].
-\textcolor{blue}{We may also have to shift our perspective from making explicit point forecasts to making moving average forecasts} [@Petchey2015].
-\textcolor{blue}{Whether the poor predictive ability of our models impacts the comparison of models based on individual vs. population-level data is an open question}.
+Such short forecast horizons are not encouraging.
+Unfortunately, we have few ideas about how to improve population forecasts that have not already been proposed [@Petchey2015; @Mouquet2015].
+Longer time-series should improve our ability to detect exogenous drivers such as climate [@Teller2016], and modeling larger spatial extents may reduce parameter uncertainty [@Petchey2015].
+We may also have to shift our perspective from making explicit point forecasts to making moving average forecasts [@Petchey2015].
+Whether the poor predictive ability of our models impacts the comparison of models based on individual vs. population-level data is an open question.
 
@@
 Nonetheless, our results should encourage the use of easy-to-collect population-level data for forecasting the state of plant populations.
-
-<!---
-Our methodological approach could also potentially be used for backcasting paleoclimate and paleoecological aggregates such as pollen counts based on lake sediments [e.g., @Paciorek2009]
---->
 
@@
 The data used in this paper have been archived on Ecological Archives: http://esapubs.org/archive/ecol/E092/143/.
-All data and R code necessary to reproduce our work has been deposited on \textcolor{blue}{Figshare} (link) and is also available on GitHub (http://github.com/atredennick/MicroMesoForecast).
+All data and R code necessary to reproduce our work has been deposited on Figshare (http://doi.org/10.6084/m9.figshare.4007520) and is also available on GitHub (http://github.com/atredennick/MicroMesoForecast/releases/tag/v1.0).
+
+Author Contributions
+--------------------
+ATT and PBA conceived the study; ATT, MBH, and PBA developed the statistical and modeling approach; ATT did the data analysis, with input from MBH and PBA; ATT wrote the first draft of the manuscript and all authors contributed to revisions.
 
@@
 
-<!---
-\begin{figure}[!ht]
-  \centering
-      \includegraphics[width=4in]{./components/micromeso_flowchart.png}
-  \caption{Work flow of the data aggregation, model fitting, and population simulating.}
-\end{figure}
-
-\newpage{}
-
-\begin{figure}[!ht]
-  \centering
-      \includegraphics[width=6in]{./components/lppds_suppfig.png}
-  \caption{Cross validation scores (summed log pointwise predictive density) plotted against the prior variance for $\boldsymbol{\beta}_C$.
-The optimal score for prediction for each species-vital rate combination is shown with a vertical red line.}
-\end{figure}
-
-\newpage{}
---->
 \begin{figure}[!ht]
@@
   \caption{Comparisons of one-step-ahead, out-of-sample forecast accuracy between the IPM and QBM models with and without the inclusion of climate covariates.
-\textcolor{blue}{Boxplots show the distribution of $\rho$ averaged over quadrats for each cross-validation year (i.e., 13 values of $\rho$ for each species-model combination).} For each comparison, \emph{P}-values are from one-sided \emph{t} tests designed to assess whether the first model in the comparison statement had higher accuracy than the second model in the comparison statement (see details in Table S22).
-\textcolor{blue}{Statistical tests relied on correlation values for each quadrat-year-species combination, after averaging over model reps for each combination.} In no case did adding climate covariates decrease forecast accuracy (Table S21).
+Boxplots show the distribution of $\rho$ averaged over quadrats for each cross-validation year (i.e., 13 values of $\rho$ for each species-model combination).
+For each comparison, \emph{P}-values are from one-sided \emph{t} tests designed to assess whether the first model in the comparison statement had higher accuracy than the second model in the comparison statement (see details in Table S22).
+Statistical tests relied on correlation values for each quadrat-year-species combination, after averaging over model reps for each combination.
+In no case did adding climate covariates decrease forecast accuracy (Table S21).
 Species codes are as in Fig.
@@
 1.
-\textcolor{blue}{Climate covariate codes: \emph{pptLag} = "water year" precipitation at \emph{t}-2; \emph{ppt1} = April through June precipitation at \emph{t}-1; \emph{ppt2} = April through June precipitation at \emph{t}; \emph{TmeanSpr1} = April through June temperature at \emph{t}-1; \emph{TmeanSpr2} = April through June temperature at \emph{t}.}}
+Climate covariate codes: \emph{pptLag} = "water year" precipitation at \emph{t}-2; \emph{ppt1} = April through June precipitation at \emph{t}-1; \emph{ppt2} = April through June precipitation at \emph{t}; \emph{TmeanSpr1} = April through June temperature at \emph{t}-1; \emph{TmeanSpr2} = April through June temperature at \emph{t}.}
 \end{figure}
@@
       \includegraphics[height=5in]{./components/forecast_horizon.png}
-  \caption{The forecast horizons for both models with climate covariates included \textcolor{blue}{and using mean parameter values.
-Points show the average accuracy ($\rho$, \textcolor{blue}{correlation between observations and predictions}) across all forecasts at a given distance between the last observation and the forecast, where forecasts are made for in-sample data.} We only examine the forecast accuracy of models with climate covariates included because in no case did including climate covariates significantly decrease accuracy (see Fig.
+  \caption{The forecast horizons for both models with climate covariates included and using mean parameter values.
+Points show the average accuracy ($\rho$,correlation between observations and predictions) across all forecasts at a given distance between the last observation and the forecast, where forecasts are made for in-sample data.
+We only examine the forecast accuracy of models with climate covariates included because in no case did including climate covariates significantly decrease accuracy (see Fig.
 2).
-\textcolor{blue}{The dashed blue line indicates a forecast proficiency threshold of $\rho = 0.5$.
+The dashed blue line indicates a forecast proficiency threshold of $\rho = 0.5$.
 Species codes are as in Fig.
-1 and statistical comparisons between the IPM and QBM at each forecast distance are in Tables S23 and S24.}}
+1 and statistical comparisons between the IPM and QBM at each forecast distance are in Tables S23 and S24.}
 \end{figure}
@@
 \newpage{}
-
-<!---
-\begin{figure}[!ht]
-  \centering
-      \includegraphics[height=6in]{./components/climate_effect_corplots.png}
-  \caption{Correlations between QBM and IPM estimates of climate effects.
-We ignore sizeXclimate interactions since these are not directly comparable across model types.
-The QBM does not have multiple vital rates, so its values are repeated across panels within each species.
-Across top panels, 'growth' = growth regression, 'rec' = recruitment regression, 'surv' = survival regression.}
-\end{figure}
-
-\newpage{}
-
-\begin{figure}[!ht]
-  \centering
-      \includegraphics[height=6in]{./components/climate_change_results.png}
-  \caption{Mean (points) and 90\% quantiles (errorbars) for the proportional difference between baseline simulations (using observed climate) and the climate pertubation simulation on the x-axis.
-We calculated proportional difference as log(perturbed climate cover) - log(observed climate cover), where 'perturbed' and 'observed' refer to the climate time series used to drive interannual variation in the simulations.
-Model error and parameter uncertainty were propagated through the simulation phase.}
-\end{figure}
-
-\pagebreak{}
---->
-
-```{r figure_3, dependson="plot-options", fig.cap="Sensitivity of equilibrium cover simulated from the IPM to each climate scenario applied to individual and combined vital rates.
-For example, the points associated with G show the median cover from IPM simulations where a climate perturbation is applied only to the growth regression climate covariates.
-These simulations use mean parameter values for clarity.", fig.width=8, fig.height=6, cache=FALSE, results='hide', eval=F}
-setwd("~/Repos/MicroMesoForecast/analysis/")
-
-####
-####  Make IPM sensitivity plot for lower panels
-####
-setwd("../analysis/ipm/simulations/results/climate_sensitivity/")
-files <- list.files()
-cover_files <- files[grep("cover", files)]
-spp_id <- substr(cover_files, 1, 4)
-num_files <- length(cover_files)
-all_sims <- data.frame(time=NA, cover=NA, species=NA, climsim=NA, vital=NA)
-for(i in 1:num_files){
-  tmp <- read.csv(cover_files[i])
-  tmp$species <- rep(spp_id[i], nrow(tmp))
-  all_sims <- rbind(all_sims, tmp)
-}
-all_sims <- all_sims[2:nrow(all_sims),]
-
-# Rename vital rates with codes
-vital_names <- unique(all_sims$vital)
-vital_codes <- c("A", "G", "GR", "GS", "R", "S" , "SR") 
-vitals <- data.frame(names=vital_names,code=vital_codes)
-for(vnow in vital_names){
-  ids <- which(all_sims[,"vital"]==vnow)
-  idcode <- which(vitals[,"names"]==vnow)
-  all_sims[ids,"vital"] <- as.character(vitals[idcode,"code"])
-}
-
-# Rep the 'all' simulation for each vital rate for plotting
-all_clim_sims <- subset(all_sims, vital=="A")
-species <- unique(all_sims$species)
-out_controls <- data.frame(time=NA, cover=NA, species=NA, climsim=NA, vital=NA)
-for(spp in species){
-  tmp <- subset(all_clim_sims, species==spp)
-  for(vital in vital_codes){
-    tmp$vital <- vital
-    out_controls <- rbind(out_controls, tmp)
-  }
-}
-out_controls <- out_controls[2:nrow(out_controls),]
-
-# Combine the datasets
-all_noclim_sims <- subset(all_sims, vital!="all")
-final <- rbind(out_controls, all_noclim_sims)
-
-# Calculate statistics for plotting
-equilibrium_cover <- ddply(final, .(species, climsim, vital), summarise,
-                           eq_cover = median(cover))
-mean_cover <- ddply(subset(equilibrium_cover, climsim=="all"), .(species), summarise,
-                    value = mean(eq_cover))
-
-equilibrium_cover <- subset(equilibrium_cover, vital!="A")
-bothid <- which(equilibrium_cover[,"climsim"]=="ppttemp")
-equilibrium_cover[bothid,"climsim"] <- "zppttemp"
-# myCols2 <- c("#277BA8", "#7ABBBD", "#AED77A")
-sensplot <- ggplot(equilibrium_cover, aes(x=climsim, y=eq_cover*100, 
-                              shape=vital, group=vital, linetype=vital))+
-  geom_hline(data=mean_cover, aes(yintercept=value*100), linetype=2, color="grey45")+
-  geom_line()+
-  geom_point(size=4)+
-  facet_wrap("species", scales = "free", ncol=2)+
-  scale_linetype_discrete(name="Vital Rate")+
-  scale_shape_discrete(name="Vital Rate")+
-  scale_x_discrete(labels=c("baseline", "+ppt", "+temp", "+ppt&temp"))+
-  ylab("Equilibrium Cover (%)")+
-  xlab("Simulation")+
-  theme_bw()+
-  theme(axis.text.x = element_text(angle = 45, hjust = 1))
-
-covertest <- subset(equilibrium_cover, climsim=="all")
-colnames(covertest)[4] <- "obs_cover"
-out <- merge(subset(equilibrium_cover, climsim!="all"), covertest[,c("species","vital","obs_cover")])
-out$propdiff <- with(out, log(eq_cover*100)-log(obs_cover*100))
-
-# meanparamplot <- grid.arrange(meanplot, sensplot, ncol=1, nrow=2)
-# print(meanparamplot)
-print(sensplot)
-
-setwd("~/Repos/MicroMesoForecast/manuscript")
-```
-
-```{r figure_4, dependson="plot-options", fig.cap="Effect of quadrat sample size on the precision (standard deviation) of main climate effect estimates in the QBM.
-Increasing the number of quadrats results in diminishing returns in terms of parameter certainty.
-Light dashed lines show individual climate effects at five quadrat sample sizes.
-Thick dark lines are inverse gaussian fits showing the mean effect of increasing quadrat sample size on parameter precision.", fig.width=3, fig.height=7.5, cache=FALSE, results='hide', eval=F}
-library(mgcv)
-
-setwd("../analysis/quadBM/vitalRateRegressions/exclude_quad_test")
-
-####
-####  Read in model fits and summarize results
-####
-source("popgrowth_read_data.R")
-spp <- unique(growD_all$Species)
-num_quads <- numeric(length(spp))
-for(i in 1:length(spp)){
-  tmp <- subset(growD_all, Species==spp[i])
-  num_quads[i] <- length(unique(tmp$quad))
-}
-num_quads <- as.data.frame(num_quads)
-num_quads$species <- spp
-
-all_files <- list.files("./results/")
-all_fits <- data.frame(Parameter=NA, stddev=NA, id=NA, 
-                       rmsquad=NA, rep=NA, species=NA)
-for(i in 1:length(all_files)){
-  tmp <- all_files[i]
-  spp <- substring(tmp, 1, 4)
-  qsrm1 <- unlist(strsplit(tmp, "[_]"))[2]
-  qsrm <- gsub("numout", "", qsrm1)
-  rep1 <- unlist(strsplit(tmp, "[_]"))[3]
-  rep2 <- gsub(".RDS","",rep1)
-  rep <- gsub("rep", "", rep2)
-
-  long <- readRDS(paste("./results/",tmp, sep=""))
-  short <- ddply(long, .(Parameter), summarise,
-                 stddev = sd(value))
-  climeff <- short[grep("b2", short$Parameter),]
-  climeff$id <- substr(climeff$Parameter, 4, length(climeff$Parameter))
-  climeff$id <- as.numeric(unlist(strsplit(climeff$id, split=']')))  
-  climeff <- climeff[with(climeff, order(id)),]
-  climeff$rmsquad <- qsrm
-  climeff$rep <- rep
-  climeff$species <- spp
-  all_fits <- rbind(all_fits, climeff)
-} # end file loop
-
-all_fits <- all_fits[2:nrow(all_fits),]
-
-####
-####  Read in full fits
-####
-dir <- "../truncNormModel/"
-files <- list.files(dir)
-files <- files[grep(".RDS", files)]
-all_alls <- data.frame(Parameter=NA, stddev=NA, id=NA, 
-                       rmsquad=NA, rep=NA, species=NA)
-for(i in 1:length(files)){
-  tmp <- files[i]
-  spp1 <- unlist(strsplit(tmp, "[_]"))[3]
-  spp <- gsub(".RDS","",spp1)
-
-  long <- readRDS(paste(dir,tmp, sep=""))
-  short <- ddply(long, .(Parameter), summarise,
-                 stddev = sd(value))
-  climeff <- short[grep("b2", short$Parameter),]
-  climeff$id <- substr(climeff$Parameter, 4, length(climeff$Parameter))
-  climeff$id <- as.numeric(unlist(strsplit(climeff$id, split=']')))  
-  climeff <- climeff[with(climeff, order(id)),]
-  climeff$rmsquad <- 0
-  climeff$rep <- 1
-  climeff$species <- spp
-  all_alls <- rbind(all_alls, climeff)
-}
-all_alls <- all_alls[2:nrow(all_alls),]
-
-####
-####  Aggregate results by species and rep
-####
-all_fits <- rbind(all_fits, all_alls)
-agg_fits <- ddply(all_fits, .(Parameter, rmsquad, species, id), summarise,
-                  avg_stddev = mean(stddev))
-agg_fits2 <- agg_fits[which(agg_fits$id %in% c(1:5)),]
-agg_fits2 <- merge(agg_fits2, num_quads, by="species")
-agg_fits2$quadsfit <- with(agg_fits2, num_quads-as.numeric(rmsquad))
-
-####
-####  Fit model (need to do by species)
-####
-modBOGR <- gam(avg_stddev~quadsfit, data=subset(agg_fits2, species=="BOGR"),
-               family=quasi(link="1/mu^2"))
-modHECO <- gam(avg_stddev~quadsfit, data=subset(agg_fits2, species=="HECO"),
-               family=quasi(link="1/mu^2"))
-modPASM <- gam(avg_stddev~quadsfit, data=subset(agg_fits2, species=="PASM"),
-               family=quasi(link="1/mu^2"))
-modPOSE <- gam(avg_stddev~quadsfit, data=subset(agg_fits2, species=="POSE"),
-               family=quasi(link="1/mu^2"))
-summary(modBOGR)
-summary(modHECO)
-summary(modPASM)
-summary(modPOSE)
-
-####
-#### Plot
-####
-ggplot(agg_fits2, aes(x=quadsfit, y=avg_stddev))+
-  geom_line(aes(group=Parameter), alpha=0.3, linetype=2)+
-  geom_point(aes(group=Parameter), alpha=0.3, size=3)+
-#   stat_smooth(method="loess", size=2, color="black", se=FALSE)+
-  geom_smooth(method="glm", formula=y~x, family=quasi(link="1/mu^2"),
-              size=2, color="black", se=FALSE)+
-  facet_wrap("species", scales = "free", ncol=1)+
-  theme_bw()+
-  xlab("Number of quadrats fit")+
-  ylab("Standard deviation of coefficient")
-
-```
 

### 2016-10-11 manuscript/manuscript.tex
minor changes for final MEE files

@@
 
-We compared the forecasting skill
-\textcolor{blue}{(accuracy and precision)} of statistical and population
-models based on aggregated, population-level data with the skill of
-models based on individual-level data. We used a demographic dataset
-that tracks the fates of individual plants from four species over 14
-years to build two kinds of single-species population models,
-traditional models using individual growth, survival, and recruitment
-data and alternative models based on population-level (basal cover)
-data. We simulated from the models to answer two questions motivated by
-the fact that the effects of intraspecific competition (density
-dependence) and interannual weather variability act at the level of the
-individual (Clark et al. 2011). First, can population models fit using
-aggregated individual-level data (percent cover) adequately capture
-density dependence to produce forecasts as skillful as those from models
-fit to demographic data? Second, can population models fit using
-aggregated data adequately capture the influence of climate on
-population growth and, in turn, produce forecasts as skillful as those
-from models fit to demographic data?
+We compared the forecasting skill (accuracy and precision) of
+statistical and population models based on aggregated, population-level
+data with the skill of models based on individual-level data. We used a
+demographic dataset that tracks the fates of individual plants from four
+species over 14 years to build two kinds of single-species population
+models, traditional models using individual growth, survival, and
+recruitment data and alternative models based on population-level (basal
+cover) data. We simulated from the models to answer two questions
+motivated by the fact that the effects of intraspecific competition
+(density dependence) and interannual weather variability act at the
+level of the individual (Clark et al. 2011). First, can population
+models fit using aggregated individual-level data (percent cover)
+adequately capture density dependence to produce forecasts as skillful
+as those from models fit to demographic data? Second, can population
+models fit using aggregated data adequately capture the influence of
+climate on population growth and, in turn, produce forecasts as skillful
+as those from models fit to demographic data?
 
@@
 
-\textcolor{blue}{We used two types of data: individual-level data and percent cover data.
-Using the individual-level data, we fit three vital rate regressions (survival, growth, and rectruitment) to build an Integral Projection Model (IPM) for simulating the plant populations.
-Using the percent cover data we fit a simple, Gompertz population growth model, which we refer to as a quadrat-based model (QBM).
-For both model types (IPM and QBM), we fit and simulate versions of the model with and without climate covariates.
-We used Bayesian ridge regression to weight the importance of each climate covariate.
-We then performed cross-validation to assess each model's ability to predict out-of-sample data.
-We compared the forecast accuracy ($\rho$, correlation between observations and predictions) and mean absolute error (MAE) between the IPM and the QBM to test our expectation that the IPM should outperform the QBM.
-Lastly, we use in-sample forecasts to quantify each model's forecast horizon}
+We used two types of data: individual-level data and percent cover data.
+Using the individual-level data, we fit three vital rate regressions
+(survival, growth, and rectruitment) to build an Integral Projection
+Model (IPM) for simulating the plant populations. Using the percent
+cover data we fit a simple, Gompertz population growth model, which we
+refer to as a quadrat-based model (QBM). For both model types (IPM and
+QBM), we fit and simulate versions of the model with and without climate
+covariates. We used Bayesian ridge regression to weight the importance
+of each climate covariate. We then performed cross-validation to assess
+each model's ability to predict out-of-sample data. We compared the
+forecast accuracy (\(\rho\), correlation between observations and
+predictions) and mean absolute error (MAE) between the IPM and the QBM
+to test our expectation that the IPM should outperform the QBM. Lastly,
+we use in-sample forecasts to quantify each model's forecast horizon
 (Petchey et al. 2015).
@@
 standard errors calculated using the HC4 estimator of Cribari-Neto
-(2004). Statistical tests
-\textcolor{blue}{for comparing correlations and error} were conducted
-using algorithms from Ye et al. (2015).
+(2004). Statistical tests for comparing correlations and error were
+conducted using algorithms from Ye et al. (2015).
 
@@
 forecast skill declines as the time between an observation and a
-forecast increases.
-\textcolor{blue}{In particular, we are interested in the temporal distance at which forecast skill falls below a threshold: the so-called ecological forecast horizon}
-(Petchey et al. 2015). To assess the forecast horizons of our models, we
-initiate the forecast model with the population state at some time
-\emph{t} and make sequential forecasts of the population at times
-\(t+1, t+2, \dots, t+T\) where \emph{T} is the maximum number of years
-between the initial year and the final year of our observations. For
-example, if we initialize the forecast model with percent cover in 1940,
-we are able to make five forecasts up to the year 1945. Forecast models
-are not re-initialized with observations between years. Thus, in our
-current example, the model forecast for percent cover in 1941 has a
-forecast horizon of one year, the forecast in 1942 has a forecast
-horizon of two years, and so on. We performed these simulations using
-mean parameter values for all model types (IPM with/without climate; QBM
-with/without climate) and all possible initial years. For a given
-forecast distance, we averaged the correlation between forecasts and
-observations. Note that our forecasts for the horizon analysis are all
-made using in-sample data because we used model fits from the full data
-set. Nonetheless, our simulations offer insight into the differences
-among model forecast horizons.
-\textcolor{blue}{We chose an arbitrary forecast accuracy of $\rho = 0.5$ as our forecast proficiency threshold.
-the forecast horizon is the temporal distance at which forecast accuracy falls below $\rho = 0.5$.
-For basic research on forecasting, arbitrary proficiency thresholds suffice for comparative purposes}
-(Petchey et al. 2015),
-\textcolor{blue}{and $\rho = 0.5$ represents the point at which about 25\% of the variance in observations is explained by the predictions.}
+forecast increases. In particular, we are interested in the temporal
+distance at which forecast skill falls below a threshold: the so-called
+ecological forecast horizon (Petchey et al. 2015). To assess the
+forecast horizons of our models, we initiate the forecast model with the
+population state at some time \emph{t} and make sequential forecasts of
+the population at times \(t+1, t+2, \dots, t+T\) where \emph{T} is the
+maximum number of years between the initial year and the final year of
+our observations. For example, if we initialize the forecast model with
+percent cover in 1940, we are able to make five forecasts up to the year
+1945. Forecast models are not re-initialized with observations between
+years. Thus, in our current example, the model forecast for percent
+cover in 1941 has a forecast horizon of one year, the forecast in 1942
+has a forecast horizon of two years, and so on. We performed these
+simulations using mean parameter values for all model types (IPM
+with/without climate; QBM with/without climate) and all possible initial
+years. For a given forecast distance, we averaged the correlation
+between forecasts and observations. Note that our forecasts for the
+horizon analysis are all made using in-sample data because we used model
+fits from the full data set. Nonetheless, our simulations offer insight
+into the differences among model forecast horizons. We chose an
+arbitrary forecast accuracy of \(\rho = 0.5\) as our forecast
+proficiency threshold. the forecast horizon is the temporal distance at
+which forecast accuracy falls below \(\rho = 0.5\). For basic research
+on forecasting, arbitrary proficiency thresholds suffice for comparative
+purposes (Petchey et al. 2015), and \(\rho = 0.5\) represents the point
+at which about 25\% of the variance in observations is explained by the
+predictions.
 
@@
 
-With climate covariates included
-\textcolor{blue}{and using mean parameter values, the accuracy of both models' forecasts declined as the distance between the last observation and the forecast increased},
-but they did so at similar rates (Fig. 4). The only exception is for
-\emph{Poa secunda}, where QBM forecast accuracy remained steady as the
-\textcolor{blue}{temporal distance of the forecast increased}, whereas
-IPM forecast accuracy declined (Fig. 4).
-\textcolor{blue}{The forecast horizons were short: forecast accuracy fell below $\rho$ = 0.5 after one year for the IPM for most species, and after four years, at most, for the QBM (Fig. 4).
-Across the different temporal distances from the observation to the forecast, the IPM was never more accurate than the QBM (\emph{P} > 0.05 for all one-sided \emph{t}-tests, Table S23).
-Likewise, the QBM was rarely more accurate the IPM, the only exception being for \emph{H.
-comata} at temporal distances of two ($t_{(115)}$ = 2.39, \emph{P} = 0.002) and three years ($t_{(98)}$ = 2.04, \emph{P} = 0.022) (Table S24).
-There were some cases where the QBM was more accurate than the IPM for \emph{Poa secunda}, but neither model exceeded the forecast proficiency threshold by a large margin (Fig. 4, Table S24).}
+With climate covariates included and using mean parameter values, the
+accuracy of both models' forecasts declined as the distance between the
+last observation and the forecast increased, but they did so at similar
+rates (Fig. 4). The only exception is for \emph{Poa secunda}, where QBM
+forecast accuracy remained steady as the temporal distance of the
+forecast increased, whereas IPM forecast accuracy declined (Fig. 4). The
+forecast horizons were short: forecast accuracy fell below \(\rho\) =
+0.5 after one year for the IPM for most species, and after four years,
+at most, for the QBM (Fig. 4). Across the different temporal distances
+from the observation to the forecast, the IPM was never more accurate
+than the QBM (\emph{P} \textgreater{} 0.05 for all one-sided
+\emph{t}-tests, Table S23). Likewise, the QBM was rarely more accurate
+the IPM, the only exception being for \emph{H. comata} at temporal
+distances of two (\(t_{(115)}\) = 2.39, \emph{P} = 0.002) and three
+years (\(t_{(98)}\) = 2.04, \emph{P} = 0.022) (Table S24). There were
+some cases where the QBM was more accurate than the IPM for
+\emph{Poa secunda}, but neither model exceeded the forecast proficiency
+threshold by a large margin (Fig. 4, Table S24).
 
@@
 climate effects were identified in the individual vital rate models but
-not in the percent cover model (Fig. 3).
-\textcolor{blue}{For \emph{H. comata}, adding climate covariates did not improve forecasts (Fig. 2), despite the significant climate effects in the vital rate regressions (Fig. 3).}
+not in the percent cover model (Fig. 3). For \emph{H. comata}, adding
+climate covariates did not improve forecasts (Fig. 2), despite the
+significant climate effects in the vital rate regressions (Fig. 3).
 Furthermore, for the two species where including climate covariates
@@
 We also expected IPM forecast accuracy to decline at a lower rate than
-the QBM as the
-\textcolor{blue}{time between the model initialization and the forecast increased.}
-In principle, more mechanistic models should produce better predictions,
-especially under novel conditions (Evans 2012, Schindler and Hilborn
-2015). In our case, the IPM explicitly models the influence of weather
-on recruitment and survival, effects that may be poorly represented in
-the QBM because recruitment and survival mainly affect small plants that
-contribute little to year-to-year changes in percent cover. Over longer
-time scales, the addition and subtraction of small plants could have
-large effects on population growth, so explicitly modeling these effects
-could contribute to a longer forecast horizon. However, we found no
-evidence
-\textcolor{blue}{that the forecast horizon for the IPM was greater than the QBM (Fig. 4).
-On the contrary, the QBM tended to have a slightly longer forecast horizon than the IPM for most species (Fig. 4).
-The QBM has fewer processes and parameters, which can reduce bias due to parameter uncertainty.
-As a result, the QBM may better capture near term dynamics when populations do not fluctuate widely, as in our case.}
+the QBM as the time between the model initialization and the forecast
+increased. In principle, more mechanistic models should produce better
+predictions, especially under novel conditions (Evans 2012, Schindler
+and Hilborn 2015). In our case, the IPM explicitly models the influence
+of weather on recruitment and survival, effects that may be poorly
+represented in the QBM because recruitment and survival mainly affect
+small plants that contribute little to year-to-year changes in percent
+cover. Over longer time scales, the addition and subtraction of small
+plants could have large effects on population growth, so explicitly
+modeling these effects could contribute to a longer forecast horizon.
+However, we found no evidence that the forecast horizon for the IPM was
+greater than the QBM (Fig. 4). On the contrary, the QBM tended to have a
+slightly longer forecast horizon than the IPM for most species (Fig. 4).
+The QBM has fewer processes and parameters, which can reduce bias due to
+parameter uncertainty. As a result, the QBM may better capture near term
+dynamics when populations do not fluctuate widely, as in our case.
 
@@
 et al. (2011), knowledge of a species' population ecology should guide
-the modeling approach.
-\textcolor{blue}{Third, our percent cover data are essentially error-free because we were able to aggregate indiviual plant areas to calculate percent cover.
-Percent cover data collected by typical sampling methods (e.g., Daubenmire frames) will include error that may affect population forecasts due to misspecifing the initial conditions and/or biasing model parameters}
-(Queenborough et al. 2011).
-\textcolor{blue}{Percent cover models based on data containing more measurement error than ours might perform worse in comparison with individual-based models.
-One way to account for such error is to develop a sampling model that relates the observations (estimated percent cover in a plot) to the true state (percent cover derived from individual plant measurements in the same plot)}
-(Hobbs and Hooten 2015).
-
-\textcolor{blue}{Although our main goal was to compare individual-based and population-level modeling approaches relative to one another, it is worth reflecting on the absolute forecasting skill of our models.
-In particular, the forecast horizon of both models, defined as the time horizon at which the correlation between predictions and observations falls below $\rho = 0.5$, is less than five years for all species.
-Such short forecast horizons are not encouraging.}
-\textcolor{blue}{Unfortunately, we have few ideas about how to improve population forecasts that have not already been proposed}
-(Mouquet et al. 2015, Petchey et al. 2015).
-\textcolor{blue}{Longer time-series should improve our ability to detect exogenous drivers such as climate}
-(Teller et al. 2016),
-\textcolor{blue}{and modeling larger spatial extents may reduce parameter uncertainty}
-(Petchey et al. 2015).
-\textcolor{blue}{We may also have to shift our perspective from making explicit point forecasts to making moving average forecasts}
-(Petchey et al. 2015).
-\textcolor{blue}{Whether the poor predictive ability of our models impacts the comparison of models based on individual vs. population-level data is an open question}.
+the modeling approach. Third, our percent cover data are essentially
+error-free because we were able to aggregate indiviual plant areas to
+calculate percent cover. Percent cover data collected by typical
+sampling methods (e.g., Daubenmire frames) will include error that may
+affect population forecasts due to misspecifing the initial conditions
+and/or biasing model parameters (Queenborough et al. 2011). Percent
+cover models based on data containing more measurement error than ours
+might perform worse in comparison with individual-based models. One way
+to account for such error is to develop a sampling model that relates
+the observations (estimated percent cover in a plot) to the true state
+(percent cover derived from individual plant measurements in the same
+plot) (Hobbs and Hooten 2015).
+
+Although our main goal was to compare individual-based and
+population-level modeling approaches relative to one another, it is
+worth reflecting on the absolute forecasting skill of our models. In
+particular, the forecast horizon of both models, defined as the time
+horizon at which the correlation between predictions and observations
+falls below \(\rho = 0.5\), is less than five years for all species.
+Such short forecast horizons are not encouraging. Unfortunately, we have
+few ideas about how to improve population forecasts that have not
+already been proposed (Mouquet et al. 2015, Petchey et al. 2015). Longer
+time-series should improve our ability to detect exogenous drivers such
+as climate (Teller et al. 2016), and modeling larger spatial extents may
+reduce parameter uncertainty (Petchey et al. 2015). We may also have to
+shift our perspective from making explicit point forecasts to making
+moving average forecasts (Petchey et al. 2015). Whether the poor
+predictive ability of our models impacts the comparison of models based
+on individual vs.~population-level data is an open question.
 
@@
 \url{http://esapubs.org/archive/ecol/E092/143/}. All data and R code
-necessary to reproduce our work has been deposited on
-\textcolor{blue}{Figshare} (link) and is also available on GitHub
-(\url{http://github.com/atredennick/MicroMesoForecast}).
+necessary to reproduce our work has been deposited on Figshare
+(\url{http://doi.org/10.6084/m9.figshare.4007520}) and is also available
+on GitHub
+(\url{http://github.com/atredennick/MicroMesoForecast/releases/tag/v1.0}).
+
+\subsection{Author Contributions}\label{author-contributions}
+
+ATT and PBA conceived the study; ATT, MBH, and PBA developed the
+statistical and modeling approach; ATT did the data analysis, with input
+from MBH and PBA; ATT wrote the first draft of the manuscript and all
+authors contributed to revisions.
 
@@
 
-\subsection*{References}\label{references}
+\subsection{References}\label{references}
 \addcontentsline{toc}{subsection}{References}
@@
 
+\subsection{Supporting Information}\label{si}
+\begin{description}
+  \item[Appendix S1.] Detailed methods and Stan code examples for the vital rate and percent cover regressions; Figures S1 and S2; Tables S1-S24.
+\end{description
+
 \end{document}

### 2016-10-11 manuscript/supporting_information.Rmd
minor changes for final MEE files

@@
    - \usepackage{graphicx}
+   - \usepackage{morefloats}
 bibliography: ~/Dropbox/Bibliography/MicroMeso.bib
@@
 
-# Supporting Information
+\begin{center}
+\textbf{\Large{Supporting Information Appendix S1}}  \\
+A.T. Tredennick, M.B. Hooten, \& P.B. Adler, ``Do we need demographic data...''  \\
+\emph{Methods in Ecology and Evolution}
+\end{center}
 
@@
 Following the full model expression we provide the Stan code that directly corresponds the statement of posterior and joint distributions.
-Note that Stan uses standard deviations ($\sigma$) to specify distributions rather than variances ($\sigma^2$)  In our expressions of the posterior and joint distributions we use general probability notation where [.] represents some unspecified distribution.
+Note that Stan uses standard deviations ($\sigma$) to specify distributions rather than variances ($\sigma^2$).
+In our expressions of the posterior and joint distributions we use general probability notation where [.] represents some unspecified distribution.
 We do this for clarity of model presentation.
@@
 \end{table}
+
+\newpage{}
+
+\begin{table}[ht]
+\centering
+\caption {Statistical tests of model comparisons at different forecast horizons. Table shows results from one-sided t-tests for the particular comparison.}
+\begin{scriptsize}
+\begin{tabular}{lllrrrlr}
+  \hline
+comparison & performance\_measure & test\_type & test\_statistic & df & p\_value & species & horizon \\ 
+  \hline
+climate IPM vs. climate QBM & rho & t & -0.305 & 167.000 & 0.619 & PASM & 1.000 \\ 
+  climate IPM vs. climate QBM & rho & t & 0.262 & 149.000 & 0.397 & POSE & 1.000 \\ 
+  climate IPM vs. climate QBM & rho & t & -0.538 & 129.000 & 0.704 & HECO & 1.000 \\ 
+  climate IPM vs. climate QBM & rho & t & -1.231 & 129.000 & 0.890 & BOGR & 1.000 \\ 
+  climate IPM vs. climate QBM & rho & t & -0.679 & 150.000 & 0.751 & PASM & 2.000 \\ 
+  climate IPM vs. climate QBM & rho & t & -0.260 & 132.000 & 0.602 & POSE & 2.000 \\ 
+  climate IPM vs. climate QBM & rho & t & -2.385 & 115.000 & 0.991 & HECO & 2.000 \\ 
+  climate IPM vs. climate QBM & rho & t & -0.843 & 114.000 & 0.799 & BOGR & 2.000 \\ 
+  climate IPM vs. climate QBM & rho & t & -0.109 & 129.000 & 0.543 & PASM & 3.000 \\ 
+  climate IPM vs. climate QBM & rho & t & -2.423 & 112.000 & 0.992 & POSE & 3.000 \\ 
+  climate IPM vs. climate QBM & rho & t & -2.043 & 98.000 & 0.978 & HECO & 3.000 \\ 
+  climate IPM vs. climate QBM & rho & t & 0.609 & 96.000 & 0.272 & BOGR & 3.000 \\ 
+  climate IPM vs. climate QBM & rho & t & 0.497 & 108.000 & 0.310 & PASM & 4.000 \\ 
+  climate IPM vs. climate QBM & rho & t & -0.276 & 94.000 & 0.608 & POSE & 4.000 \\ 
+  climate IPM vs. climate QBM & rho & t & -0.677 & 82.000 & 0.750 & HECO & 4.000 \\ 
+  climate IPM vs. climate QBM & rho & t & 0.059 & 78.000 & 0.477 & BOGR & 4.000 \\ 
+  climate IPM vs. climate QBM & rho & t & -0.095 & 91.000 & 0.538 & PASM & 5.000 \\ 
+  climate IPM vs. climate QBM & rho & t & -0.492 & 79.000 & 0.688 & POSE & 5.000 \\ 
+  climate IPM vs. climate QBM & rho & t & -0.894 & 70.000 & 0.813 & HECO & 5.000 \\ 
+  climate IPM vs. climate QBM & rho & t & -0.186 & 65.000 & 0.573 & BOGR & 5.000 \\ 
+  climate IPM vs. climate QBM & rho & t & -0.365 & 75.000 & 0.642 & PASM & 6.000 \\ 
+  climate IPM vs. climate QBM & rho & t & -0.613 & 66.000 & 0.729 & POSE & 6.000 \\ 
+  climate IPM vs. climate QBM & rho & t & -2.420 & 59.000 & 0.991 & HECO & 6.000 \\ 
+  climate IPM vs. climate QBM & rho & t & 0.047 & 54.000 & 0.481 & BOGR & 6.000 \\ 
+  climate IPM vs. climate QBM & rho & t & -0.309 & 62.000 & 0.621 & PASM & 7.000 \\ 
+  climate IPM vs. climate QBM & rho & t & -0.390 & 53.000 & 0.651 & POSE & 7.000 \\ 
+  climate IPM vs. climate QBM & rho & t & -0.752 & 50.000 & 0.772 & HECO & 7.000 \\ 
+  climate IPM vs. climate QBM & rho & t & -1.131 & 44.000 & 0.868 & BOGR & 7.000 \\ 
+  climate IPM vs. climate QBM & rho & t & -0.353 & 54.000 & 0.637 & PASM & 8.000 \\ 
+  climate IPM vs. climate QBM & rho & t & -0.862 & 47.000 & 0.803 & POSE & 8.000 \\ 
+  climate IPM vs. climate QBM & rho & t & -0.326 & 43.000 & 0.627 & HECO & 8.000 \\ 
+  climate IPM vs. climate QBM & rho & t & -2.599 & 37.000 & 0.993 & BOGR & 8.000 \\ 
+   \hline
+\end{tabular}
+\end{scriptsize}
+\end{table}
+
+\newpage{}
+
+\begin{table}[ht]
+\centering
+\caption {Statistical tests of model comparisons at different forecast horizons.
+Table shows results from one-sided t-tests for the particular comparison.
+Note the reverse order of comparisons relative to Table S23.}
+\begin{scriptsize}
+\begin{tabular}{lllrrrlr}
+  \hline
+comparison & performance\_measure & test\_type & test\_statistic & df & p\_value & species & horizon \\ 
+  \hline
+climate QBM vs. climate IPM & rho & t & 0.305 & 167.000 & 0.381 & PASM & 1.000 \\ 
+  climate QBM vs. climate IPM & rho & t & -0.262 & 149.000 & 0.603 & POSE & 1.000 \\ 
+  climate QBM vs. climate IPM & rho & t & 0.538 & 129.000 & 0.296 & HECO & 1.000 \\ 
+  climate QBM vs. climate IPM & rho & t & 1.231 & 129.000 & 0.110 & BOGR & 1.000 \\ 
+  climate QBM vs. climate IPM & rho & t & 0.679 & 150.000 & 0.249 & PASM & 2.000 \\ 
+  climate QBM vs. climate IPM & rho & t & 0.260 & 132.000 & 0.398 & POSE & 2.000 \\ 
+  climate QBM vs. climate IPM & rho & t & 2.385 & 115.000 & 0.009 & HECO & 2.000 \\ 
+  climate QBM vs. climate IPM & rho & t & 0.843 & 114.000 & 0.201 & BOGR & 2.000 \\ 
+  climate QBM vs. climate IPM & rho & t & 0.109 & 129.000 & 0.457 & PASM & 3.000 \\ 
+  climate QBM vs. climate IPM & rho & t & 2.423 & 112.000 & 0.008 & POSE & 3.000 \\ 
+  climate QBM vs. climate IPM & rho & t & 2.043 & 98.000 & 0.022 & HECO & 3.000 \\ 
+  climate QBM vs. climate IPM & rho & t & -0.609 & 96.000 & 0.728 & BOGR & 3.000 \\ 
+  climate QBM vs. climate IPM & rho & t & -0.497 & 108.000 & 0.690 & PASM & 4.000 \\ 
+  climate QBM vs. climate IPM & rho & t & 0.276 & 94.000 & 0.392 & POSE & 4.000 \\ 
+  climate QBM vs. climate IPM & rho & t & 0.677 & 82.000 & 0.250 & HECO & 4.000 \\ 
+  climate QBM vs. climate IPM & rho & t & -0.059 & 78.000 & 0.523 & BOGR & 4.000 \\ 
+  climate QBM vs. climate IPM & rho & t & 0.095 & 91.000 & 0.462 & PASM & 5.000 \\ 
+  climate QBM vs. climate IPM & rho & t & 0.492 & 79.000 & 0.312 & POSE & 5.000 \\ 
+  climate QBM vs. climate IPM & rho & t & 0.894 & 70.000 & 0.187 & HECO & 5.000 \\ 
+  climate QBM vs. climate IPM & rho & t & 0.186 & 65.000 & 0.427 & BOGR & 5.000 \\ 
+  climate QBM vs. climate IPM & rho & t & 0.365 & 75.000 & 0.358 & PASM & 6.000 \\ 
+  climate QBM vs. climate IPM & rho & t & 0.613 & 66.000 & 0.271 & POSE & 6.000 \\ 
+  climate QBM vs. climate IPM & rho & t & 2.420 & 59.000 & 0.009 & HECO & 6.000 \\ 
+  climate QBM vs. climate IPM & rho & t & -0.047 & 54.000 & 0.519 & BOGR & 6.000 \\ 
+  climate QBM vs. climate IPM & rho & t & 0.309 & 62.000 & 0.379 & PASM & 7.000 \\ 
+  climate QBM vs. climate IPM & rho & t & 0.390 & 53.000 & 0.349 & POSE & 7.000 \\ 
+  climate QBM vs. climate IPM & rho & t & 0.752 & 50.000 & 0.228 & HECO & 7.000 \\ 
+  climate QBM vs. climate IPM & rho & t & 1.131 & 44.000 & 0.132 & BOGR & 7.000 \\ 
+  climate QBM vs. climate IPM & rho & t & 0.353 & 54.000 & 0.363 & PASM & 8.000 \\ 
+  climate QBM vs. climate IPM & rho & t & 0.862 & 47.000 & 0.197 & POSE & 8.000 \\ 
+  climate QBM vs. climate IPM & rho & t & 0.326 & 43.000 & 0.373 & HECO & 8.000 \\ 
+  climate QBM vs. climate IPM & rho & t & 2.599 & 37.000 & 0.007 & BOGR & 8.000 \\ 
+   \hline
+\end{tabular}
+\end{scriptsize}
+\end{table}

### 2016-10-11 README.Rmd
adds MBH to authors
clean up and updating readme

@@
 title: "Read Me"
-date: "April 20, 2015"
+date: \today
 output: html_document
 ---
-This is a 'read me' file for data and code associated with "Do we need demographic data to forecast population responses to climate change?" by Andrew T.
-Tredennick and Peter B.
-Adler (in review).
-The bioRxiv preprint can be found [here](http://biorxiv.org/content/early/2015/08/30/025742).
+This is a 'read me' file for data and code associated with "Do we need demographic data to forecast plant population dynamics?" by Andrew T.
+Tredennick, Mevin B.
+Hooten, and Peter B.
+Adler, to appear in *Methods in Ecology and Evolution*.
 
@@
 Traditional population models built with detailed demographic observations from one study site can address the impacts of environmental change at particular locations, but are difficult to scale up to the landscape and regional scales relevant to management decisions.
-An alternative is to build models using population-level data which are much easier to collect than individual-level data over broad spatial scales.
+An alternative is to build models using population-level data that are much easier to collect over broad spatial scales than individual-level data.
 However, it is unknown whether models built using population-level data adequately capture the effects of density-dependence and environmental forcing that are necessary to generate skillful forecasts.
 2.
-Here, we test the consequences of aggregating individual responses when forecasting the population states and trajectories of four perennial grass species in a semi-arid grassland in Montana, USA.
-We parameterized two population models for each species, one based on individual-level data (survival, growth and recruitment) and one on population-level data (percent cover), and compared their forecasting skill and forecast horizons with and without the inclusion of climate covariates.
-For both models we used Bayesian ridge regression to weight the influence of climate covariates for optimal prediction.
+Here, we test the consequences of aggregating individual responses when forecasting the population states (percent cover) and trajectories of four perennial grass species in a semi-arid grassland in Montana, USA.
+We parameterized two population models for each species, one based on individual-level data (survival, growth and recruitment) and one on population-level data (percent cover), and compared their forecasting accuracy and forecast horizons with and without the inclusion of climate covariates.
+For both models, we used Bayesian ridge regression to weight the influence of climate covariates for optimal prediction.
 3.
-In the absence of climate effects, we found no significant difference between the forecasting skill of models based on individual-level data and models based on population-level data.
-Climate effects were weak, but increased forecasting skill for two species.
-Increases in skill with climate covariates were similar between model types.
+In the absence of climate effects, we found no significant difference between the forecast accuracy of models based on individual-level data and models based on population-level data.
+Climate effects were weak, but increased forecast accuracy for two species.
+Increases in accuracy with climate covariates were similar between model types.
 4.
-In our case study, percent cover models generated forecasts as skillful as those from a demographic model.
+In our case study, percent cover models generated forecasts as accurate as those from a demographic model.
 For the goal of forecasting, models based on aggregated individual-level data may offer a practical alternative to data-intensive demographic models.
@@
 
-Funding for this work was provided by the National Science Foundation through a Postdoctoral Research Fellowship in Biology to Andrew ([DBI-1400370](http://www.nsf.gov/awardsearch/showAward?AWD_ID=1400370&HistoricalAwards=false)) and a CAREER award to Peter ([DEB-1054040](http://www.nsf.gov/awardsearch/showAward?AWD_ID=1054040&HistoricalAwards=false)).
+Funding for this work was provided by the National Science Foundation through a Postdoctoral Research Fellowship in Biology to Andrew ([DBI-1400370](http://www.nsf.gov/awardsearch/showAward?AWD_ID=1400370&HistoricalAwards=false)), award MSB-1241856 to MBH, and a CAREER award to Peter ([DEB-1054040](http://www.nsf.gov/awardsearch/showAward?AWD_ID=1054040&HistoricalAwards=false)).
 
@@
 --------------
-Here is the file structure assuming the root directory is ``ssde_harvest_dryad/``:
+Here is the file structure assuming the root directory is ``MicroMesoForecast/``:
 
 ### analysis/
-* ``ipm/`` -- contains R code for all analyses and figure production. 
-    + ``sumlations/`` -- 
-        - ``another`` --
-    + ``vitalRateRegs/`` -- 
+Contains code and data necessary to reproduce our results (may be a bit messy!). See readme files in subdirectories for more information.
+
+* ``data_processing/`` -- R code for assembling raw data into data frames for model fitting 
+* ``ipm/`` -- contains R code for all analyses associated with individual-level data and modeling
+* ``processed_data/`` -- output from scripts in ``data_processing/`` folder; data frames used for model fitting
+* ``quadBM/`` -- contains R code for all analyses associated with the quadrat-level data and modeling
+* ``weather/`` -- contains weather data for the study site and an R script to generate seasonal weather covariates
 * ``alpha_list_growth.csv`` -- CSV file with the alpha values used to define the spatial influence of genets on one another for the growth regressions (see [Chu and Adler 2015](http://www.esajournals.org/doi/abs/10.1890/14-1741.1))
@@
 
+### devel/
+Contains auxillary scripts that have not been tested or wrapped inside other scripts (non-essential)
+
+### figure_table_scripts/
+Contains scripts for comparing model simulations and generating figures and tables for the manuscript.
+
 ### manuscript/
-*
+Contains the manuscript.Rmd and supporting_information.Rmd files, as well as the necessary Latex components.

### 2016-10-13 README.Rmd
adds readmes for figshare and versioning

@@
 -------------------
-This directory contains all of the data and R code necessary to reproduce the analysis and figures from Tredennick and Adler 201x.
+This directory contains all of the data and R code necessary to reproduce the analysis and figures from Tredennick, Hooten, and Adler (2016).
 The ``analysis/`` directory holds all code and data, while the ``\manuscript`` directory contains a R Markdown file (``manuscript.Rmd``) with paper text and the code to produce figures embedded.
+While the code and data provided here could reproduce our results, we provide it mainly for demonstration purposes because several processes require the use of High Performance Computing centers.
 

### 2016-10-13 analysis/ipm/simulations/README.Rmd
adds readmes for figshare and versioning

### README for IPM simulation directories

This directory contains R scripts for IPM simulations using the fitted vital rate regressions. There are three simulation scripts:

* ```01a_one-step_ahead_wrapper.R``` -- This script loops through all observed year-to-year transitions,
sets up initial conditions, and then calls the IPM to project the population one time step. The IPM includes climate effects.

* ```01b_one-step_ahead_wrapper_densdep.R``` -- This script loops through all observed year-to-year transitions,
sets up initial conditions, and then calls the IPM to project the population one time step. The IPM does not include climate effects.

* ```02_forecast_horizon_sims.R``` -- Script for IPM forecasts of final observation year from all possible lag years. IPM is initialized with cover at year t-x in quadrat q, and then projected using observed climate to year t. We can then assess the model's forecast horizon and compare it to the IBM.

The three simulation scripts rely on helper functions defined in the remaining .R files. These simulations produce the output summarized in Figures 3 and 4 of the paper.

### 2016-10-13 analysis/quadBM/simulations/README.Rmd
adds readmes for figshare and versioning

### README for QBM simulation directories

This directory contains R scripts for QBM simulations using the fitted regression. There are three simulation scripts:

* ```01a_loyo_validation_qbm.R``` -- This script loops through all observed year-to-year transitions,
sets up initial conditions, and then calls the QBM to project the population one time step. The QBM includes climate effects.

* ```01b_loyo_validation_qbm_densdep_only.R``` -- This script loops through all observed year-to-year transitions, sets up initial conditions, and then calls the QBM to project the population one time step. The QBM does not include climate effects.

* ```02_forecast_horizon_qbm.R``` -- Script for QBM forecasts of final observation year from all possible lag years. IPM is initialized with cover at year t-x in quadrat q, and then projected using observed climate to year t. We can then assess the model's forecast horizon and compare it to the IBM.

The three simulation scripts rely on helper functions defined in the remaining .R files. These simulations produce the output summarized in Figures 3 and 4 of the paper.

### 2016-10-13 analysis/quadBM/vitalRateRegressions/README.Rmd
adds readmes for figshare and versioning

### README for vital rate regression Quadrat-Based Model directories

This directory contains sub-directories for fitting the quadrat-based model (``truncNormModel``), a subdirectory containing scripts for running cross-validation and ridge regression for each vital rate (``regularization/``), and a subdirectory containing scripts for leave-one-year-out cross-validation for each vital rate regression (``truncNormModel/validation/``). There is also a ``cache/`` directory containing old versions of scripts.

### 2016-10-18 manuscript/manuscript_preprint.Rmd
updates preprint formatting

--- 2016-10-11 manuscript/manuscript.Rmd
@@
 csl: components/ecology.csl
-fontsize: 12pt
+fontsize: 11pt
 geometry: margin=1in
 header-includes:
-- \usepackage{lineno}
-- \linenumbers
-- \usepackage{setspace}
 - \usepackage{sectsty}
-- \doublespacing
 - \usepackage{todonotes}
-- \usepackage[document]{ragged2e}
 - \usepackage{times}
@@
 - \usepackage{color, soul}
-layout: 12pt
+- \usepackage{fancyhdr}
+- \pagestyle{fancy}
+- \chead{\color{red}\textsf{MANUSCRIPT IN PRESS AT \emph{METHODS IN ECOLOGY AND EVOLUTION}}}
+- \lhead{}
+- \rhead{}
+layout: 11pt
 linkcolor: black
@@
     fig_caption: yes
-    keep_tex: yes
+    keep_tex: no
   word_document: default
@@
 ---
+
+\thispagestyle{fancy}
 
@@
 Geological Survey, Colorado Cooperative Fish and Wildlife Research Unit, Fort Collins, CO 80523, USA; \textsuperscript{3}Department of Fish, Wildlife, and Conservation Biology, Colorado State University, Fort Collins, CO 80523 USA}; \textsuperscript{4}Department of Statistics, Colorado State University, Fort Collins, CO 80523 USA}}
-
-\textbf{Running head:} Demographic data and population forecasting
-
-\textbf{Word count:} 8039 (including captions and reference list)
-
-\textbf{Corresponding author:}  
-Andrew Tredennick (atredenn@gmail.com)  
-Department of Wildland Resources and the Ecology Center  
-5230 Old Main Hill  
-Utah State University  
-Logan, Utah 84322, USA
-
-Last compile: \today
 
@@
 \textsf{\textbf{Key-words:}} forecasting, climate change, grassland, integral projection model, population model, statistical regularization, ridge regression
+
+\twocolumn
 
@@
 
-\vspace{-3em}
 \begin{align}
 y_{i,q,t}^{S} &\sim \text{Bernoulli}(s_{i,q,t}), \\
-\text{logit}(s_{i,q,t}) &= \beta_{0,t} + \beta_{s,t}x_{i,q,t} + \beta_{Q,q} + \textbf{z}'_t \boldsymbol{\beta}_c + \beta_{d,1} w_{i,t} + \beta_{d,2} (x_{i,q,t}w_{i,q,t}),
+\text{logit}(s_{i,q,t}) &= \beta_{0,t} + \beta_{s,t}x_{i,q,t} + \beta_{Q,q} + \textbf{z}'_t \boldsymbol{\beta}_c \\
+&+ \beta_{d,1} w_{i,t} + \beta_{d,2} (x_{i,q,t}w_{i,q,t}),
 \end{align}
-\vspace{-3em}
 
@@
 
-\vspace{-3em}
 \begin{align}
 y_{i,q,t+1}^G &\sim \text{Normal}(\mu_{i,q,t+1}, \sigma^2_{x_{i,q,t+1}}), \\
-\mu_{i,q,t+1} &= \beta_{0,t} + \beta_{s,t}x_{i,q,t} + \beta_{Q,q} + \textbf{z}'_t \boldsymbol{\beta}_c + \beta_{d,1} w_{i,q,t} + \beta_{d,2} (x_{i,q,t}w_{i,q,t}),
+\mu_{i,q,t+1} &= \beta_{0,t} + \beta_{s,t}x_{i,q,t} + \beta_{Q,q} + \textbf{z}'_t \boldsymbol{\beta}_c \\
+&+ \beta_{d,1} w_{i,q,t} + \beta_{d,2} (x_{i,q,t}w_{i,q,t}),
 \end{align}
-\vspace{-3em}
 
@@
 
-\vspace{-3em}
 \begin{align}
-k_{j}(v_{j},u_{j},\bar{w_{j}}) = S_j(u_j, \bar{w_{j}}(u_{j}))G_j(v_{j},u_{j},\bar{w_{j}}(u_{j})) + R_j(v_{j},u_{j},\bar{w_{j}}),
+k_{j}(v_{j},u_{j},\bar{w_{j}}) &= S_j(u_j, \bar{w_{j}}(u_{j}))G_j(v_{j},u_{j},\bar{w_{j}}(u_{j})) \\
+&+ R_j(v_{j},u_{j},\bar{w_{j}}),
 \end{align}
-\vspace{-3em}
 
@@
 
+\onecolumn
+
 Tables
@@
 
-\singlespace{}
-
 References

### 2016-10-27 manuscript/manuscript.Rmd
updates corrections to manuscript and si

@@
 
-Despite the weak climate effects, including climate covariates did increase the accuracy of forecasts for two species: *B. gracilis* and *Poa secunda* (Fig. 2).
-However, only for *B. gracilis* were the skill increases statistically significant at $\alpha$ = 0.05 for the IPM ($t_{(279)}$ = 1.70, *P* = 0.045) and the QBM ($t_{(279)}$ = 1.80, *P* = 0.037).
-Similarly, forecast error decreased significantly with the inclusion of climate covariates for the *B.
-gracilis* IPM ($t_{(280)}$ = -3.72, *P* = 0.029) and QBM ($t_{(280)}$ = -3.34, *P* < 0.0001), and for the *Poa secunda* IPM ($t_{(196)}$ = -1.90, *P* < 0.0001) and QBM ($t_{(196)}$ = -2.47, *P* = 0.007) (Fig.
-S2, Supporting Information).
-In no case did including climate covariates significantly decrease forecast skill (Table S21), despite small changes in the mean skill (Fig. 2).
+Despite the weak climate effects, including climate covariates did increase the \textcolor{blue}{mean} accuracy of forecasts for two species: *B.
+gracilis* and *Poa secunda* (\textcolor{blue}{Fig.
+S3}).
+\textcolor{blue}{However, skill increases were only statistically significant at $\alpha$ = 0.05 for the QBM for \emph{B. gracilis} QBM ($t_{(279)}$ = 1.80, \emph{P} = 0.037).}
+Forecast error decreased significantly with the inclusion of climate covariates for the *B.
+gracilis* IPM ($t_{(280)}$ = -3.72, *P* = \textcolor{blue}{0.01}) and QBM ($t_{(280)}$ = -3.34, *P* < 0.0001), and for the *Poa secunda* IPM ($t_{(196)}$ = -1.90, *P* textcolor{blue}{= 0.015}) and QBM ($t_{(196)}$ = -2.47, *P* = 0.007) (Fig.
+S1, Supporting Information).
+In no case did including climate covariates significantly decrease forecast skill (Table S21), despite small changes in the mean skill (Fig. 2, \textcolor{blue}{Fig. S3}).
 
@@
 2): forecast accuracy of *P.
-smithii* percent cover from an IPM with climate covariates was greater than the accuracy from the QBM with climate covariates ($t_{(215)}$ = 1.92, *P* = 0.028).
+smithii* percent cover from an IPM with climate covariates was greater than the accuracy from the QBM with climate covariates ($t_{(215)}$ = 1.92, *P* = \textcolor{blue}{0.022}).
 However, adding climate covariates decreased the skill of both models, and the difference between the IPM and QBM emerges only because skill decreased less for the IPM than the QBM.
@@
 Across the different temporal distances from the observation to the forecast, the IPM was never more accurate than the QBM (\emph{P} > 0.05 for all one-sided \emph{t}-tests, Table S23).
-Likewise, the QBM was rarely more accurate the IPM, the only exception being for \emph{H.
-comata} at temporal distances of two ($t_{(115)}$ = 2.39, \emph{P} = 0.002) and three years ($t_{(98)}$ = 2.04, \emph{P} = 0.022) (Table S24).
-There were some cases where the QBM was more accurate than the IPM for \emph{Poa secunda}, but neither model exceeded the forecast proficiency threshold by a large margin (Fig. 4, Table S24).
+\textcolor{blue}{Likewise, the QBM was rarely more accurate the IPM, the only exceptions being for \emph{H.
+comata} at a temporal distance of three years ($t_{(98)}$ = 2.04, \emph{P} = 0.032) and \emph{B.
+gracilis} at a temporal distance of eight years ($t_{(37)}$ = 1.67, \emph{P} = 0.05) (Table S24).}
+\st{There were some cases where the QBM was more accurate than the IPM for \emph{Poa secunda}, but neither model exceeded the forecast proficiency threshold by a large margin (Fig. 4, Table S24).}
 

### 2016-10-27 manuscript/manuscript.tex
merging
updates corrections to manuscript and si

@@
 Despite the weak climate effects, including climate covariates did
-increase the accuracy of forecasts for two species: \emph{B. gracilis}
-and \emph{Poa secunda} (Fig. 2). However, only for \emph{B. gracilis}
-were the skill increases statistically significant at \(\alpha\) = 0.05
-for the IPM (\(t_{(279)}\) = 1.70, \emph{P} = 0.045) and the QBM
-(\(t_{(279)}\) = 1.80, \emph{P} = 0.037). Similarly, forecast error
-decreased significantly with the inclusion of climate covariates for the
-\emph{B. gracilis} IPM (\(t_{(280)}\) = -3.72, \emph{P} = 0.029) and QBM
-(\(t_{(280)}\) = -3.34, \emph{P} \textless{} 0.0001), and for the
-\emph{Poa secunda} IPM (\(t_{(196)}\) = -1.90, \emph{P} \textless{}
-0.0001) and QBM (\(t_{(196)}\) = -2.47, \emph{P} = 0.007) (Fig. S2,
-Supporting Information). In no case did including climate covariates
-significantly decrease forecast skill (Table S21), despite small changes
-in the mean skill (Fig. 2).
+increase the \textcolor{blue}{mean} accuracy of forecasts for two
+species: \emph{B. gracilis} and \emph{Poa secunda}
+(\textcolor{blue}{Fig. S3}).
+\textcolor{blue}{However, skill increases were only statistically significant at $\alpha$ = 0.05 for the QBM for \emph{B. gracilis} QBM ($t_{(279)}$ = 1.80, \emph{P} = 0.037).}
+Forecast error decreased significantly with the inclusion of climate
+covariates for the \emph{B. gracilis} IPM (\(t_{(280)}\) = -3.72,
+\emph{P} = \textcolor{blue}{0.01}) and QBM (\(t_{(280)}\) = -3.34,
+\emph{P} \textless{} 0.0001), and for the \emph{Poa secunda} IPM
+(\(t_{(196)}\) = -1.90, \emph{P} textcolor\{blue\}\{= 0.015\}) and QBM
+(\(t_{(196)}\) = -2.47, \emph{P} = 0.007) (Fig. S1, Supporting
+Information). In no case did including climate covariates significantly
+decrease forecast skill (Table S21), despite small changes in the mean
+skill (Fig. 2, \textcolor{blue}{Fig. S3}).
 
@@
 an IPM with climate covariates was greater than the accuracy from the
-QBM with climate covariates (\(t_{(215)}\) = 1.92, \emph{P} = 0.028).
-However, adding climate covariates decreased the skill of both models,
-and the difference between the IPM and QBM emerges only because skill
-decreased less for the IPM than the QBM. Results from all pairwise
-statistical tests are shown in Table S22 of the Supporting Information.
+QBM with climate covariates (\(t_{(215)}\) = 1.92, \emph{P} =
+\textcolor{blue}{0.022}). However, adding climate covariates decreased
+the skill of both models, and the difference between the IPM and QBM
+emerges only because skill decreased less for the IPM than the QBM.
+Results from all pairwise statistical tests are shown in Table S22 of
+the Supporting Information.
 
@@
 than the QBM (\emph{P} \textgreater{} 0.05 for all one-sided
-\emph{t}-tests, Table S23). Likewise, the QBM was rarely more accurate
-the IPM, the only exception being for \emph{H. comata} at temporal
-distances of two (\(t_{(115)}\) = 2.39, \emph{P} = 0.002) and three
-years (\(t_{(98)}\) = 2.04, \emph{P} = 0.022) (Table S24). There were
-some cases where the QBM was more accurate than the IPM for
-\emph{Poa secunda}, but neither model exceeded the forecast proficiency
-threshold by a large margin (Fig. 4, Table S24).
+\emph{t}-tests, Table S23).
+\textcolor{blue}{Likewise, the QBM was rarely more accurate the IPM, the only exceptions being for \emph{H.
+comata} at a temporal distance of three years ($t_{(98)}$ = 2.04, \emph{P} = 0.032) and \emph{B.
+gracilis} at a temporal distance of eight years ($t_{(37)}$ = 1.67, \emph{P} = 0.05) (Table S24).}
+\st{There were some cases where the QBM was more accurate than the IPM for \emph{Poa secunda}, but neither model exceeded the forecast proficiency threshold by a large margin (Fig. 4, Table S24).}
 
@@
 
-\subsection{References}\label{references}
+\subsection*{References}\label{references}
 \addcontentsline{toc}{subsection}{References}
@@
 
-\subsection{Supporting Information}\label{si}
-\begin{description}
-  \item[Appendix S1.] Detailed methods and Stan code examples for the vital rate and percent cover regressions; Figures S1 and S2; Tables S1-S24.
-\end{description
-
 \end{document}

### 2016-10-27 manuscript/supporting_information.Rmd
updates corrections to manuscript and si

--- 2016-07-07 manuscript/supporting_information.Rmd
@@
 vector[N] lambda;
-vector[N] q;
 climEff <- C*b2;
@@
     lambda[n] <- trueP1[n]*mu[n];
-    q[n] <- lambda[n]*theta;
   }
@@
 // Likelihood
-  Y ~ neg_binomial_2(q, theta);
+  Y ~ neg_binomial_2(lambda, theta);
 }
@@
 
+\begin{figure}[!ht]
+  \centering
+      \includegraphics[height=6in]{./components/forecast_accuracy_means_mockup.png}
+  \caption{Comparison of one-step-ahead, out-of-sample forecast accuracy (mean correlation between observations and predictions) between the IPM and QBM models with and without the inclusion of climate covariates.}
+\end{figure}
+
+\newpage{}
+
 ## Supporting Tables
@@
   \hline
-simple IPM vs. climate IPM & rho & t & -1.703 & 279.000 & 0.955 & BOGR \\ 
-  simple IPM vs. climate IPM & MAE & t & -3.715 & 280.000 & 1.000 & BOGR \\ 
+simple IPM vs. climate IPM & rho & t & -1.196 & 279.000 & 0.884 & BOGR \\ 
+  simple IPM vs. climate IPM & MAE & t & -2.333 & 280.000 & 0.990 & BOGR \\ 
   simple QBM vs. climate QBM & rho & t & -1.798 & 279.000 & 0.963 & BOGR \\ 
   simple QBM vs. climate QBM & MAE & t & -3.340 & 280.000 & 1.000 & BOGR \\ 
-  simple IPM vs. climate IPM & rho & t & 0.246 & 169.000 & 0.403 & HECO \\ 
-  simple IPM vs. climate IPM & MAE & t & -1.472 & 170.000 & 0.929 & HECO \\ 
+  simple IPM vs. climate IPM & rho & t & -0.037 & 169.000 & 0.515 & HECO \\ 
+  simple IPM vs. climate IPM & MAE & t & -2.130 & 170.000 & 0.983 & HECO \\ 
   simple QBM vs. climate QBM & rho & t & 0.747 & 169.000 & 0.228 & HECO \\ 
   simple QBM vs. climate QBM & MAE & t & 1.145 & 170.000 & 0.127 & HECO \\ 
-  simple IPM vs. climate IPM & rho & t & 0.201 & 215.000 & 0.421 & PASM \\ 
-  simple IPM vs. climate IPM & MAE & t & -0.592 & 216.000 & 0.723 & PASM \\ 
+  simple IPM vs. climate IPM & rho & t & 0.097 & 215.000 & 0.462 & PASM \\ 
+  simple IPM vs. climate IPM & MAE & t & -0.556 & 216.000 & 0.711 & PASM \\ 
   simple QBM vs. climate QBM & rho & t & 0.631 & 215.000 & 0.264 & PASM \\ 
   simple QBM vs. climate QBM & MAE & t & 4.276 & 216.000 & 0.000 & PASM \\ 
-  simple IPM vs. climate IPM & rho & t & -1.072 & 195.000 & 0.858 & POSE \\ 
-  simple IPM vs. climate IPM & MAE & t & -1.911 & 196.000 & 0.971 & POSE \\ 
+  simple IPM vs. climate IPM & rho & t & -1.479 & 195.000 & 0.930 & POSE \\ 
+  simple IPM vs. climate IPM & MAE & t & -2.185 & 196.000 & 0.985 & POSE \\ 
   simple QBM vs. climate QBM & rho & t & -1.564 & 195.000 & 0.940 & POSE \\ 
@@
   \hline
-climate IPM vs. simple IPM & rho & t & 0.798 & 279.000 & 0.213 & BOGR \\ 
-  climate IPM vs. simple IPM & MAE & t & 1.347 & 280.000 & 0.910 & BOGR \\ 
-  climate QBM vs. simple QBM & rho & t & 0.526 & 279.000 & 0.300 & BOGR \\ 
-  climate QBM vs. simple QBM & MAE & t & -0.478 & 280.000 & 0.317 & BOGR \\ 
-  simple IPM vs. simple QBM & rho & t & -0.961 & 279.000 & 0.831 & BOGR \\ 
-  simple IPM vs. simple QBM & MAE & t & 0.088 & 280.000 & 0.535 & BOGR \\ 
-  climate IPM vs. climate QBM & rho & t & -0.024 & 279.000 & 0.509 & BOGR \\ 
-  climate IPM vs. climate QBM & MAE & t & 2.289 & 280.000 & 0.989 & BOGR \\ 
-  climate IPM vs. simple IPM & rho & t & 0.455 & 169.000 & 0.325 & HECO \\ 
-  climate IPM vs. simple IPM & MAE & t & -2.292 & 170.000 & 0.012 & HECO \\ 
-  climate QBM vs. simple QBM & rho & t & 0.051 & 169.000 & 0.480 & HECO \\ 
-  climate QBM vs. simple QBM & MAE & t & -0.746 & 170.000 & 0.228 & HECO \\ 
-  simple IPM vs. simple QBM & rho & t & -0.461 & 169.000 & 0.677 & HECO \\ 
-  simple IPM vs. simple QBM & MAE & t & 1.626 & 170.000 & 0.947 & HECO \\ 
-  climate IPM vs. climate QBM & rho & t & -0.299 & 169.000 & 0.617 & HECO \\ 
-  climate IPM vs. climate QBM & MAE & t & 0.289 & 170.000 & 0.613 & HECO \\ 
-  climate IPM vs. simple IPM & rho & t & -0.883 & 215.000 & 0.811 & PASM \\ 
-  climate IPM vs. simple IPM & MAE & t & 0.410 & 216.000 & 0.659 & PASM \\ 
-  climate QBM vs. simple QBM & rho & t & -0.008 & 215.000 & 0.503 & PASM \\ 
-  climate QBM vs. simple QBM & MAE & t & 3.218 & 216.000 & 0.999 & PASM \\ 
-  simple IPM vs. simple QBM & rho & t & 0.363 & 215.000 & 0.359 & PASM \\ 
-  simple IPM vs. simple QBM & MAE & t & -0.276 & 216.000 & 0.391 & PASM \\ 
-  climate IPM vs. climate QBM & rho & t & 0.600 & 215.000 & 0.274 & PASM \\ 
-  climate IPM vs. climate QBM & MAE & t & -3.487 & 216.000 & 0.000 & PASM \\ 
-  climate IPM vs. simple IPM & rho & t & 1.538 & 195.000 & 0.063 & POSE \\ 
-  climate IPM vs. simple IPM & MAE & t & -1.844 & 196.000 & 0.033 & POSE \\ 
-  climate QBM vs. simple QBM & rho & t & 1.616 & 195.000 & 0.054 & POSE \\ 
-  climate QBM vs. simple QBM & MAE & t & -0.963 & 196.000 & 0.168 & POSE \\ 
-  simple IPM vs. simple QBM & rho & t & 1.075 & 195.000 & 0.142 & POSE \\ 
-  simple IPM vs. simple QBM & MAE & t & -0.788 & 196.000 & 0.216 & POSE \\ 
-  climate IPM vs. climate QBM & rho & t & 1.721 & 195.000 & 0.043 & POSE \\ 
-  climate IPM vs. climate QBM & MAE & t & -1.830 & 196.000 & 0.034 & POSE \\ 
+limate IPM vs. simple IPM & rho & t & 1.196 & 279.000 & 0.116 & BOGR \\ 
+  climate IPM vs. simple IPM & MAE & t & -2.333 & 280.000 & 0.010 & BOGR \\ 
+  climate QBM vs. simple QBM & rho & t & 1.798 & 279.000 & 0.037 & BOGR \\ 
+  climate QBM vs. simple QBM & MAE & t & -3.340 & 280.000 & 0.000 & BOGR \\ 
+  simple IPM vs. simple QBM & rho & t & -1.757 & 279.000 & 0.960 & BOGR \\ 
+  simple IPM vs. simple QBM & MAE & t & 2.338 & 280.000 & 0.990 & BOGR \\ 
+  climate IPM vs. climate QBM & rho & t & -1.083 & 279.000 & 0.860 & BOGR \\ 
+  climate IPM vs. climate QBM & MAE & t & 1.151 & 280.000 & 0.875 & BOGR \\ 
+  climate IPM vs. simple IPM & rho & t & 0.037 & 169.000 & 0.485 & HECO \\ 
+  climate IPM vs. simple IPM & MAE & t & -2.130 & 170.000 & 0.017 & HECO \\ 
+  climate QBM vs. simple QBM & rho & t & -0.747 & 169.000 & 0.772 & HECO \\ 
+  climate QBM vs. simple QBM & MAE & t & 1.145 & 170.000 & 0.873 & HECO \\ 
+  simple IPM vs. simple QBM & rho & t & -1.638 & 169.000 & 0.948 & HECO \\ 
+  simple IPM vs. simple QBM & MAE & t & 4.011 & 170.000 & 1.000 & HECO \\ 
+  climate IPM vs. climate QBM & rho & t & -0.704 & 169.000 & 0.759 & HECO \\ 
+  climate IPM vs. climate QBM & MAE & t & 0.329 & 170.000 & 0.629 & HECO \\ 
+  climate IPM vs. simple IPM & rho & t & -0.097 & 215.000 & 0.538 & PASM \\ 
+  climate IPM vs. simple IPM & MAE & t & -0.556 & 216.000 & 0.289 & PASM \\ 
+  climate QBM vs. simple QBM & rho & t & -0.631 & 215.000 & 0.736 & PASM \\ 
+  climate QBM vs. simple QBM & MAE & t & 4.276 & 216.000 & 1.000 & PASM \\ 
+  simple IPM vs. simple QBM & rho & t & 0.552 & 215.000 & 0.291 & PASM \\ 
+  simple IPM vs. simple QBM & MAE & t & -1.033 & 216.000 & 0.151 & PASM \\ 
+  climate IPM vs. climate QBM & rho & t & 2.022 & 215.000 & 0.022 & PASM \\ 
+  climate IPM vs. climate QBM & MAE & t & -5.309 & 216.000 & 0.000 & PASM \\ 
+  climate IPM vs. simple IPM & rho & t & 1.479 & 195.000 & 0.070 & POSE \\ 
+  climate IPM vs. simple IPM & MAE & t & -2.185 & 196.000 & 0.015 & POSE \\ 
+  climate QBM vs. simple QBM & rho & t & 1.564 & 195.000 & 0.060 & POSE \\ 
+  climate QBM vs. simple QBM & MAE & t & -2.468 & 196.000 & 0.007 & POSE \\ 
+  simple IPM vs. simple QBM & rho & t & 0.297 & 195.000 & 0.383 & POSE \\ 
+  simple IPM vs. simple QBM & MAE & t & 1.027 & 196.000 & 0.847 & POSE \\ 
+  climate IPM vs. climate QBM & rho & t & 1.137 & 195.000 & 0.128 & POSE \\ 
+  climate IPM vs. climate QBM & MAE & t & 0.455 & 196.000 & 0.675 & POSE \\ 
    \hline
@@
 \end{table}
+
+\newpage{}
+
+\begin{table}[ht]
+\centering
+\caption {Statistical tests of model comparisons at different forecast horizons. Table shows results from one-sided t-tests for the particular comparison.}
+\begin{scriptsize}
+\begin{tabular}{lllrrrlr}
+  \hline
+comparison & performance\_measure & test\_type & test\_statistic & df & p\_value & species & horizon \\ 
+  \hline
+climate IPM vs. climate QBM & rho & t & -0.122 & 167.000 & 0.549 & PASM & 1.000 \\ 
+  climate IPM vs. climate QBM & rho & t & 0.394 & 149.000 & 0.347 & POSE & 1.000 \\ 
+  climate IPM vs. climate QBM & rho & t & -0.436 & 129.000 & 0.668 & HECO & 1.000 \\ 
+  climate IPM vs. climate QBM & rho & t & -0.318 & 129.000 & 0.625 & BOGR & 1.000 \\ 
+  climate IPM vs. climate QBM & rho & t & -1.015 & 150.000 & 0.844 & PASM & 2.000 \\ 
+  climate IPM vs. climate QBM & rho & t & -0.876 & 132.000 & 0.809 & POSE & 2.000 \\ 
+  climate IPM vs. climate QBM & rho & t & -0.790 & 115.000 & 0.784 & HECO & 2.000 \\ 
+  climate IPM vs. climate QBM & rho & t & -0.049 & 114.000 & 0.519 & BOGR & 2.000 \\ 
+  climate IPM vs. climate QBM & rho & t & 0.610 & 129.000 & 0.272 & PASM & 3.000 \\ 
+  climate IPM vs. climate QBM & rho & t & -0.719 & 112.000 & 0.763 & POSE & 3.000 \\ 
+  climate IPM vs. climate QBM & rho & t & -1.871 & 98.000 & 0.968 & HECO & 3.000 \\ 
+  climate IPM vs. climate QBM & rho & t & -0.006 & 96.000 & 0.503 & BOGR & 3.000 \\ 
+  climate IPM vs. climate QBM & rho & t & 0.761 & 108.000 & 0.224 & PASM & 4.000 \\ 
+  climate IPM vs. climate QBM & rho & t & -0.168 & 94.000 & 0.566 & POSE & 4.000 \\ 
+  climate IPM vs. climate QBM & rho & t & -0.600 & 82.000 & 0.725 & HECO & 4.000 \\ 
+  climate IPM vs. climate QBM & rho & t & 0.196 & 78.000 & 0.423 & BOGR & 4.000 \\ 
+  climate IPM vs. climate QBM & rho & t & 0.226 & 91.000 & 0.411 & PASM & 5.000 \\ 
+  climate IPM vs. climate QBM & rho & t & -0.043 & 79.000 & 0.517 & POSE & 5.000 \\ 
+  climate IPM vs. climate QBM & rho & t & -0.794 & 70.000 & 0.785 & HECO & 5.000 \\ 
+  climate IPM vs. climate QBM & rho & t & 0.180 & 65.000 & 0.429 & BOGR & 5.000 \\ 
+  climate IPM vs. climate QBM & rho & t & -0.331 & 75.000 & 0.629 & PASM & 6.000 \\ 
+  climate IPM vs. climate QBM & rho & t & 0.743 & 66.000 & 0.230 & POSE & 6.000 \\ 
+  climate IPM vs. climate QBM & rho & t & -0.658 & 59.000 & 0.743 & HECO & 6.000 \\ 
+  climate IPM vs. climate QBM & rho & t & -0.015 & 54.000 & 0.506 & BOGR & 6.000 \\ 
+  climate IPM vs. climate QBM & rho & t & 0.805 & 62.000 & 0.212 & PASM & 7.000 \\ 
+  climate IPM vs. climate QBM & rho & t & -0.222 & 53.000 & 0.587 & POSE & 7.000 \\ 
+  climate IPM vs. climate QBM & rho & t & -1.009 & 50.000 & 0.841 & HECO & 7.000 \\ 
+  climate IPM vs. climate QBM & rho & t & -0.095 & 44.000 & 0.537 & BOGR & 7.000 \\ 
+  climate IPM vs. climate QBM & rho & t & -0.001 & 54.000 & 0.500 & PASM & 8.000 \\ 
+  climate IPM vs. climate QBM & rho & t & -0.155 & 47.000 & 0.561 & POSE & 8.000 \\ 
+  climate IPM vs. climate QBM & rho & t & -0.460 & 43.000 & 0.676 & HECO & 8.000 \\ 
+  climate IPM vs. climate QBM & rho & t & -1.671 & 37.000 & 0.948 & BOGR & 8.000 \\  
+   \hline
+\end{tabular}
+\end{scriptsize}
+\end{table}
+
+\newpage{}
+
+\begin{table}[ht]
+\centering
+\caption {Statistical tests of model comparisons at different forecast horizons.
+Table shows results from one-sided t-tests for the particular comparison.
+Note the reverse order of comparisons relative to Table S23.}
+\begin{scriptsize}
+\begin{tabular}{lllrrrlr}
+  \hline
+comparison & performance\_measure & test\_type & test\_statistic & df & p\_value & species & horizon \\ 
+  \hline
+climate QBM vs. climate IPM & rho & t & 0.122 & 167.000 & 0.451 & PASM & 1.000 \\ 
+  climate QBM vs. climate IPM & rho & t & -0.394 & 149.000 & 0.653 & POSE & 1.000 \\ 
+  climate QBM vs. climate IPM & rho & t & 0.436 & 129.000 & 0.332 & HECO & 1.000 \\ 
+  climate QBM vs. climate IPM & rho & t & 0.318 & 129.000 & 0.375 & BOGR & 1.000 \\ 
+  climate QBM vs. climate IPM & rho & t & 1.015 & 150.000 & 0.156 & PASM & 2.000 \\ 
+  climate QBM vs. climate IPM & rho & t & 0.876 & 132.000 & 0.191 & POSE & 2.000 \\ 
+  climate QBM vs. climate IPM & rho & t & 0.790 & 115.000 & 0.216 & HECO & 2.000 \\ 
+  climate QBM vs. climate IPM & rho & t & 0.049 & 114.000 & 0.481 & BOGR & 2.000 \\ 
+  climate QBM vs. climate IPM & rho & t & -0.610 & 129.000 & 0.728 & PASM & 3.000 \\ 
+  climate QBM vs. climate IPM & rho & t & 0.719 & 112.000 & 0.237 & POSE & 3.000 \\ 
+  climate QBM vs. climate IPM & rho & t & 1.871 & 98.000 & 0.032 & HECO & 3.000 \\ 
+  climate QBM vs. climate IPM & rho & t & 0.006 & 96.000 & 0.497 & BOGR & 3.000 \\ 
+  climate QBM vs. climate IPM & rho & t & -0.761 & 108.000 & 0.776 & PASM & 4.000 \\ 
+  climate QBM vs. climate IPM & rho & t & 0.168 & 94.000 & 0.434 & POSE & 4.000 \\ 
+  climate QBM vs. climate IPM & rho & t & 0.600 & 82.000 & 0.275 & HECO & 4.000 \\ 
+  climate QBM vs. climate IPM & rho & t & -0.196 & 78.000 & 0.577 & BOGR & 4.000 \\ 
+  climate QBM vs. climate IPM & rho & t & -0.226 & 91.000 & 0.589 & PASM & 5.000 \\ 
+  climate QBM vs. climate IPM & rho & t & 0.043 & 79.000 & 0.483 & POSE & 5.000 \\ 
+  climate QBM vs. climate IPM & rho & t & 0.794 & 70.000 & 0.215 & HECO & 5.000 \\ 
+  climate QBM vs. climate IPM & rho & t & -0.180 & 65.000 & 0.571 & BOGR & 5.000 \\ 
+  climate QBM vs. climate IPM & rho & t & 0.331 & 75.000 & 0.371 & PASM & 6.000 \\ 
+  climate QBM vs. climate IPM & rho & t & -0.743 & 66.000 & 0.770 & POSE & 6.000 \\ 
+  climate QBM vs. climate IPM & rho & t & 0.658 & 59.000 & 0.257 & HECO & 6.000 \\ 
+  climate QBM vs. climate IPM & rho & t & 0.015 & 54.000 & 0.494 & BOGR & 6.000 \\ 
+  climate QBM vs. climate IPM & rho & t & -0.805 & 62.000 & 0.788 & PASM & 7.000 \\ 
+  climate QBM vs. climate IPM & rho & t & 0.222 & 53.000 & 0.413 & POSE & 7.000 \\ 
+  climate QBM vs. climate IPM & rho & t & 1.009 & 50.000 & 0.159 & HECO & 7.000 \\ 
+  climate QBM vs. climate IPM & rho & t & 0.095 & 44.000 & 0.463 & BOGR & 7.000 \\ 
+  climate QBM vs. climate IPM & rho & t & 0.001 & 54.000 & 0.500 & PASM & 8.000 \\ 
+  climate QBM vs. climate IPM & rho & t & 0.155 & 47.000 & 0.439 & POSE & 8.000 \\ 
+  climate QBM vs. climate IPM & rho & t & 0.460 & 43.000 & 0.324 & HECO & 8.000 \\ 
+  climate QBM vs. climate IPM & rho & t & 1.671 & 37.000 & 0.052 & BOGR & 8.000 \\ 
+   \hline
+\end{tabular}
+\end{scriptsize}
+\end{table}

### 2016-10-28 manuscript/manuscript.Rmd
small updates to ms correction

@@
 Forecast error decreased significantly with the inclusion of climate covariates for the *B.
-gracilis* IPM ($t_{(280)}$ = -3.72, *P* = \textcolor{blue}{0.01}) and QBM ($t_{(280)}$ = -3.34, *P* < 0.0001), and for the *Poa secunda* IPM ($t_{(196)}$ = -1.90, *P* textcolor{blue}{= 0.015}) and QBM ($t_{(196)}$ = -2.47, *P* = 0.007) (Fig.
+gracilis* IPM ($t_{(280)}$ = -3.72, *P* = \textcolor{blue}{0.01}) and QBM ($t_{(280)}$ = -3.34, *P* < 0.0001), and for the *Poa secunda* IPM ($t_{(196)}$ = -1.90, *P* \textcolor{blue}{= 0.015}) and QBM ($t_{(196)}$ = -2.47, *P* = 0.007) (Fig.
 S1, Supporting Information).
@@
 
-The higher accuracy of the IPM and QBM with climate covariates for *B.
+The higher accuracy \textcolor{blue}{and lower error} of the IPM and QBM with climate covariates for *B.
 gracilis* and *Poa secunda*  highlights the advantage of contemporary modeling and variable selection approaches such as ridge regression and LASSO over techniques that would exclude "non-significant" effects from final models.
@@
 This is especially important when forecasting the impacts of climate variability, where it is important to include the effects of forcing variables (e.g., temperature and precipitation) even if such effects are difficult to identify.
-Indeed, we failed to detect strong climate effects in the QBM for *B. gracilis* and *Poa secunda*, but including climate covariates still improved forecasting skill (Fig. 2).
+Indeed, we failed to detect strong climate effects in the QBM for *B.
+gracilis* and *Poa secunda*, but including climate covariates still improved forecasting skill (Fig.
+2, \textcolor{blue}{Fig.
+S3}) \textcolor{blue}{and reduced forecasting error (Fig.
+S1)}.
 If a species is truly unresponsive to a given climate variable, statistical regularization techniques will shrink the mean and variance of a covariate estimate toward zero [@Hooten2015].
@@
 This research was supported by the Utah Agricultural Experiment Station, Utah State University, and approved as journal paper number 8917.
+\textcolor{blue}{And, thanks to Andy Kleinhesselink for discovering an error in our computer code before this paper was published.}
 

### 2016-10-28 manuscript/manuscript.tex
small updates to ms correction

@@
 \emph{P} \textless{} 0.0001), and for the \emph{Poa secunda} IPM
-(\(t_{(196)}\) = -1.90, \emph{P} textcolor\{blue\}\{= 0.015\}) and QBM
+(\(t_{(196)}\) = -1.90, \emph{P} \textcolor{blue}{= 0.015}) and QBM
 (\(t_{(196)}\) = -2.47, \emph{P} = 0.007) (Fig. S1, Supporting
@@
 
-The higher accuracy of the IPM and QBM with climate covariates for
-\emph{B. gracilis} and \emph{Poa secunda} highlights the advantage of
-contemporary modeling and variable selection approaches such as ridge
-regression and LASSO over techniques that would exclude
-``non-significant'' effects from final models. Ridge regression allows
-researchers to retain covariates whose effects may be difficult to
-identify in noisy data or short time series. This is especially
+The higher accuracy \textcolor{blue}{and lower error} of the IPM and QBM
+with climate covariates for \emph{B. gracilis} and \emph{Poa secunda}
+highlights the advantage of contemporary modeling and variable selection
+approaches such as ridge regression and LASSO over techniques that would
+exclude ``non-significant'' effects from final models. Ridge regression
+allows researchers to retain covariates whose effects may be difficult
+to identify in noisy data or short time series. This is especially
 important when forecasting the impacts of climate variability, where it
@@
 for \emph{B. gracilis} and \emph{Poa secunda}, but including climate
-covariates still improved forecasting skill (Fig. 2). If a species is
-truly unresponsive to a given climate variable, statistical
+covariates still improved forecasting skill (Fig. 2,
+\textcolor{blue}{Fig. S3})
+\textcolor{blue}{and reduced forecasting error (Fig. S1)}. If a species
+is truly unresponsive to a given climate variable, statistical
 regularization techniques will shrink the mean and variance of a
@@
 State University, and approved as journal paper number 8917.
+\textcolor{blue}{And, thanks to Andy Kleinhesselink for discovering an error in our computer code before this paper was published.}
 

### 2016-10-28 manuscript/manuscript.tex
final corrections before sending to MEE

@@
 increase the \textcolor{blue}{mean} accuracy of forecasts for two
-species: \emph{B. gracilis} and \emph{Poa secunda}
-(\textcolor{blue}{Fig. S3}).
-\textcolor{blue}{However, skill increases were only statistically significant at $\alpha$ = 0.05 for the QBM for \emph{B. gracilis} QBM ($t_{(279)}$ = 1.80, \emph{P} = 0.037).}
+species: \emph{B. gracilis} and \emph{Poa secunda} (Fig.2 and
+\textcolor{blue}{Fig. S3, Supporting Information}).
+\textcolor{blue}{However, skill increases were only statistically significant at $\alpha$ = 0.05 for the \emph{B. gracilis} QBM ($t_{(279)}$ = 1.80, \emph{P} = 0.037).}
 Forecast error decreased significantly with the inclusion of climate
@@
 \emph{t}-tests, Table S23).
-\textcolor{blue}{Likewise, the QBM was rarely more accurate the IPM, the only exceptions being for \emph{H.
+\textcolor{blue}{Likewise, the QBM was rarely more accurate than the IPM, the only exceptions being for \emph{H.
 comata} at a temporal distance of three years ($t_{(98)}$ = 2.04, \emph{P} = 0.032) and \emph{B.
@@
 State University, and approved as journal paper number 8917.
-\textcolor{blue}{And, thanks to Andy Kleinhesselink for discovering an error in our computer code before this paper was published.}
 
@@
 For each comparison, \emph{P}-values are from one-sided \emph{t} tests designed to assess whether the first model in the comparison statement had higher accuracy than the second model in the comparison statement (see details in Table S22).
+\textcolor{blue}{A single asterisk indicates a statistically significant difference at $\alpha$ = 0.1 and double asterisks indicate significance at $\alpha$ = 0.05}.
 Statistical tests relied on correlation values for each quadrat-year-species combination, after averaging over model reps for each combination.

### 2016-11-07 manuscript/manuscript.Rmd

@@
 Despite the weak climate effects, including climate covariates did increase the \textcolor{blue}{mean} accuracy of forecasts for two species: *B.
-gracilis* and *Poa secunda* (\textcolor{blue}{Fig.
-S3}).
-\textcolor{blue}{However, skill increases were only statistically significant at $\alpha$ = 0.05 for the QBM for \emph{B. gracilis} QBM ($t_{(279)}$ = 1.80, \emph{P} = 0.037).}
+gracilis* and *Poa secunda* (Fig.2 and \textcolor{blue}{Fig.
+S3, Supporting Information}).
+\textcolor{blue}{However, skill increases were only statistically significant at $\alpha$ = 0.05 for the \emph{B. gracilis} QBM ($t_{(279)}$ = 1.80, \emph{P} = 0.037).}
 Forecast error decreased significantly with the inclusion of climate covariates for the *B.
@@
 Across the different temporal distances from the observation to the forecast, the IPM was never more accurate than the QBM (\emph{P} > 0.05 for all one-sided \emph{t}-tests, Table S23).
-\textcolor{blue}{Likewise, the QBM was rarely more accurate the IPM, the only exceptions being for \emph{H.
+\textcolor{blue}{Likewise, the QBM was rarely more accurate than the IPM, the only exceptions being for \emph{H.
 comata} at a temporal distance of three years ($t_{(98)}$ = 2.04, \emph{P} = 0.032) and \emph{B.
@@
 This research was supported by the Utah Agricultural Experiment Station, Utah State University, and approved as journal paper number 8917.
-\textcolor{blue}{And, thanks to Andy Kleinhesselink for discovering an error in our computer code before this paper was published.}
 

### 2017-02-27 manuscript/manuscript.tex
just changes font color

@@
     \usepackage{mathspec}
-    \usepackage{xltxtra,xunicode}
   \else
@@
   \fi
-  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
-  \newcommand{\euro}{€}
+  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
 \fi
@@
 \usepackage[margin=1in]{geometry}
-\usepackage{graphicx}
+\usepackage{hyperref}
+\PassOptionsToPackage{usenames,dvipsnames}{color} % color is loaded by hyperref
+\hypersetup{unicode=true,
+            colorlinks=true,
+            linkcolor=black,
+            citecolor=Blue,
+            urlcolor=black,
+            breaklinks=true}
+\urlstyle{same}  % don't use monospace font for urls
+\usepackage{graphicx,grffile}
 \makeatletter
@@
 \setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
-\ifxetex
-  \usepackage[setpagesize=false, % page size defined by xetex
-              unicode=false, % unicode breaks when used with xetex
-              xetex]{hyperref}
-\else
-  \usepackage[unicode=true]{hyperref}
-\fi
-\hypersetup{breaklinks=true,
-            bookmarks=true,
-            pdfauthor={},
-            pdftitle={},
-            colorlinks=true,
-            citecolor=blue,
-            urlcolor=black,
-            linkcolor=black,
-            pdfborder={0 0 0}}
-\urlstyle{same}  % don't use monospace font for urls
+\IfFileExists{parskip.sty}{%
+\usepackage{parskip}
+}{% else
 \setlength{\parindent}{0pt}
 \setlength{\parskip}{6pt plus 2pt minus 1pt}
+}
 \setlength{\emergencystretch}{3em}  % prevent overfull lines
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
 \setcounter{secnumdepth}{0}
+% Redefines (sub)paragraphs to behave more like sections
+\ifx\paragraph\undefined\else
+\let\oldparagraph\paragraph
+\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
+\fi
+\ifx\subparagraph\undefined\else
+\let\oldsubparagraph\subparagraph
+\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
+\fi
 
@@
 
-\maketitle
+\definecolor{blue}{rgb}{0,0,0}
 
@@
 
-\textbf{Corresponding author:}\\Andrew Tredennick
-(\href{mailto:atredenn@gmail.com}{\nolinkurl{atredenn@gmail.com}})\\Department
-of Wildland Resources and the Ecology Center\\5230 Old Main Hill\\Utah
-State University\\Logan, Utah 84322, USA
+\textbf{Corresponding author:}\\
+Andrew Tredennick
+(\href{mailto:atredenn@gmail.com}{\nolinkurl{atredenn@gmail.com}})\\
+Department of Wildland Resources and the Ecology Center\\
+5230 Old Main Hill\\
+Utah State University\\
+Logan, Utah 84322, USA
 
@@
 how the estimated parameters respond to spatial variation in biotic and
-abiotic drivers (S{æ}ther et al. 2007). The limited spatial extent of
+abiotic drivers (Sæther et al. 2007). The limited spatial extent of
 individual-level demographic datasets constrains our ability to use
@@
 
+\hypertarget{refs}{}
+\hypertarget{ref-Adler2012}{}
 Adler, P. B., H. J. Dalgleish, and S. P. Ellner. 2012. Forecasting plant
@@
 
+\hypertarget{ref-Adler2010}{}
 Adler, P. B., S. P. Ellner, and J. M. Levine. 2010. Coexistence of
@@
 
+\hypertarget{ref-Anderson2011}{}
 Anderson, J., L. Vermeire, and P. B. Adler. 2011. Fourteen years of
@@
 
+\hypertarget{ref-Chu2015}{}
 Chu, C., and P. B. Adler. 2015. Large niche differences emerge at the
@@
 
+\hypertarget{ref-Chu2014a}{}
 Chu, C., K. M. Havstad, N. Kaplan, W. K. Lauenroth, M. P. McClaran, D.
@@
 
+\hypertarget{ref-Chu2016}{}
 Chu, C., A. R. Kleinhesselink, K. M. Havstad, M. P. McClaran, D. P.
@@
 
-Clark, J. S., and O. N. Bj{ø}rnstad. 2004. Population time series:
-Process variability, observation errors, missing values, lags, and
-hidden states. Ecology 85:3140--3150.
-
+\hypertarget{ref-Clark2004}{}
+Clark, J. S., and O. N. Bjørnstad. 2004. Population time series: Process
+variability, observation errors, missing values, lags, and hidden
+states. Ecology 85:3140--3150.
+
+\hypertarget{ref-Clark2011b}{}
 Clark, J. S., D. M. Bell, M. H. Hersh, M. C. Kwit, E. Moran, C. Salk, A.
@@
 
+\hypertarget{ref-Clark2012}{}
 Clark, J. S., D. M. Bell, M. Kwit, A. Stine, B. Vierra, and K. Zhu.
@@
 
+\hypertarget{ref-Clark2010}{}
 Clark, J. S., D. Bell, C. Chu, B. Courbaud, M. Dietze, M. Hersh, J.
-HilleRisLambers, I. Ib{á}{ñ}ez, S. LaDeau, S. McMahon, J. Metcalf, J.
-Mohan, E. Moran, L. Pangle, S. Pearson, C. Salk, Z. Shen, D. Valle, and
-P. Wyckoff. 2010. High-dimensional coexistence based on individual
+HilleRisLambers, I. Ibáñez, S. LaDeau, S. McMahon, J. Metcalf, J. Mohan,
+E. Moran, L. Pangle, S. Pearson, C. Salk, Z. Shen, D. Valle, and P.
+Wyckoff. 2010. High-dimensional coexistence based on individual
 variation: a synthesis of evidence. Ecological Monographs 80:569--608.
 
+\hypertarget{ref-Clark2001}{}
 Clark, J. S., S. R. Carpenter, M. Barber, S. Collins, A. Dobson, J. A.
@@
 Reid, K. A. Rose, O. Sala, W. H. Schlesinger, D. H. Wall, and D. Wear.
-2001. Ecological forecasts: an emerging imperative. Science (New York,
-N.Y.) 293:657--660.
-
+2001. Ecological forecasts: an emerging imperative. Science
+293:657--660.
+
+\hypertarget{ref-Cribari-Neto2004}{}
 Cribari-Neto, F. 2004. Asymptotic inference under heteroskedasticity of
@@
 
+\hypertarget{ref-Dalgleish2011}{}
 Dalgleish, H. J., D. N. Koons, M. B. Hooten, C. A. Moffet, and P. B.
@@
 
+\hypertarget{ref-Ellner2006}{}
 Ellner, S. P., and M. Rees. 2006. Integral projection models for species
@@
 
+\hypertarget{ref-Evans2012b}{}
 Evans, M. R. 2012. Modelling ecological systems in a changing world.
@@
 
+\hypertarget{ref-Freckleton2011}{}
 Freckleton, R. P., W. J. Sutherland, A. R. Watkinson, and S. A.
@@
 
+\hypertarget{ref-Gelman2014}{}
 Gelman, A., J. Hwang, and A. Vehtari. 2014. Understanding predictive
@@
 
+\hypertarget{ref-Gerber2015}{}
 Gerber, B. D., W. L. Kendall, M. B. Hooten, J. A. Dubovsky, and R. C.
@@
 
+\hypertarget{ref-Hobbs2015}{}
 Hobbs, N. T., and M. B. Hooten. 2015. Bayesian Models: A Statistical
@@
 
+\hypertarget{ref-Hooten2015}{}
 Hooten, M. B., and N. T. Hobbs. 2015. A guide to Bayesian model
@@
 
+\hypertarget{ref-Lauenroth2008}{}
 Lauenroth, W. K., and P. B. Adler. 2008. Demography of perennial
@@
 
+\hypertarget{ref-Luo2011a}{}
 Luo, Y., K. Ogle, C. Tucker, S. Fei, C. Gao, S. LaDeau, J. S. Clark, and
@@
 
+\hypertarget{ref-Mieszkowska2013}{}
 Mieszkowska, N., G. Milligan, M. T. Burrows, R. Freckleton, and M.
@@
 
-Mouquet, N., Y. Lagadeuc, V. Devictor, L. Doyen, A. Duputi{é}, D.
+\hypertarget{ref-Mouquet2015}{}
+Mouquet, N., Y. Lagadeuc, V. Devictor, L. Doyen, A. Duputié, D.
 Eveillard, D. Faure, E. Garnier, O. Gimenez, P. Huneman, F. Jabot, P.
-Jarne, D. Joly, R. Julliard, S. K{é}fi, G. J. Kergoat, S. Lavorel, L.
-{Le Gall}, L. Meslin, S. Morand, X. Morin, H. Morlon, G. Pinay, R.
-Pradel, F. M. Schurr, W. Thuiller, and M. Loreau. 2015. Predictive
-ecology in a changing world.
-
+Jarne, D. Joly, R. Julliard, S. Kéfi, G. J. Kergoat, S. Lavorel, L. Le
+Gall, L. Meslin, S. Morand, X. Morin, H. Morlon, G. Pinay, R. Pradel, F.
+M. Schurr, W. Thuiller, and M. Loreau. 2015. Predictive ecology in a
+changing world.
+
+\hypertarget{ref-Perretti2013a}{}
 Perretti, C. T., S. B. Munch, and G. Sugihara. 2013. Model-free
@@
 
-Petchey, O. L., M. Pontarp, T. M. Massie, S. K{é}fi, A. Ozgul, M.
+\hypertarget{ref-Petchey2015}{}
+Petchey, O. L., M. Pontarp, T. M. Massie, S. Kéfi, A. Ozgul, M.
 Weilenmann, G. M. Palamara, F. Altermatt, B. Matthews, J. M. Levine, D.
@@
 
+\hypertarget{ref-VandePol2011}{}
 Pol, M. van de, and A. Cockburn. 2011. Identifying the critical climatic
@@
 
+\hypertarget{ref-Queenborough2011}{}
 Queenborough, S. A., K. M. Burnet, W. J. Sutherland, A. R. Watkinson,
@@
 
+\hypertarget{ref-R2013}{}
 R Core Team. 2013. R: A language and environment for statistical
@@
 
+\hypertarget{ref-Rees2009}{}
 Rees, M., and S. P. Ellner. 2009. Integral projection models for
@@
 
+\hypertarget{ref-Schindler2015b}{}
 Schindler, D. E., and R. Hilborn. 2015. Prediction, precaution, and
@@
 
+\hypertarget{ref-Sims2007}{}
 Sims, M., D. A. Elston, A. Larkham, D. H. Nussey, and S. D. Albon. 2007.
@@
 
+\hypertarget{ref-stan2014}{}
 Stan Development Team. 2014a. Stan: A C++ Library for Probability and
@@
 
+\hypertarget{ref-rstan2014}{}
 Stan Development Team. 2014b. Rstan: the R interface to Stan, Version
@@
 
-S{æ}ther, B. E., S. Engen, V. Gr{ø}tan, W. Fiedler, E. Matthysen, M. E.
-Visser, J. Wright, A. P. M{ø}ller, F. Adriaensen, H. {Van Balen}, D.
-Balmer, M. C. Mainwaring, R. H. McCleery, M. Pampus, and W. Winkel.
-2007. The extended Moran effect and large-scale synchronous fluctuations
-in the size of great tit and blue tit populations. Journal of Animal
-Ecology 76:315--325.
-
+\hypertarget{ref-Suxe6ther2007}{}
+Sæther, B. E., S. Engen, V. Grøtan, W. Fiedler, E. Matthysen, M. E.
+Visser, J. Wright, A. P. Møller, F. Adriaensen, H. Van Balen, D. Balmer,
+M. C. Mainwaring, R. H. McCleery, M. Pampus, and W. Winkel. 2007. The
+extended Moran effect and large-scale synchronous fluctuations in the
+size of great tit and blue tit populations. Journal of Animal Ecology
+76:315--325.
+
+\hypertarget{ref-Taylor2004}{}
 Taylor, C. M., and A. Hastings. 2004. Finding optimal control strategies
@@
 
+\hypertarget{ref-Teller2016}{}
 Teller, B. J., P. B. Adler, C. B. Edwards, G. Hooker, R. E. Snyder, and
@@
 
+\hypertarget{ref-Vehtari2016}{}
 Vehtari, A., A. Gelman, and J. Gabry. 2016. Efficient implementation of
@@
 
+\hypertarget{ref-Wilcox2009}{}
 Wilcox, R. R. 2009. Comparing Pearson Correlations: Dealing with
@@
 
+\hypertarget{ref-Ye2015a}{}
 Ye, H., R. J. Beamish, S. M. Glaser, S. C. H. Grant, C.-h. Hsieh, L. J.
